### 2011-01-01 doc/hotpar2011/softxmt-hotpar2011.tex
first checkin of hotpar outline

\documentclass{acm_proc_article-sp}
\begin{document}

\title{Latency-tolerant runtimes for large graph computations}

\numberofauthors{7}
\author{
(authors)
}

\maketitle
\begin{abstract}
Multithreading is good
\end{abstract}

\section{Introduction}

\section{Graph problems}

\section{Multithreading on commodity processors}

\subsection{Memory concurrency}
\subsection{Synchronization}

\section{Results}

\section{Future work}

\section{Conclusion}

\end{document}

### 2011-01-03 doc/hotpar2011/softxmt-hotpar2011.tex
outline

@@
 
-\section{Graph problems}
+sparse graph problems important
 
-\section{Multithreading on commodity processors}
+perform badly on commodity processors, since most accesses are misses
+
+context-swtiching for latency tolerance has been successful with MTA, but single thread performance sucks
+
+we want latency tolerance on commodity processors for best of both worlds
+
+could just wait for wider multicores, but if context switching works, we can get enough concurrency on fewer cores and shut off the rest.
+
+\section{Programming model for graph problems}
+
+many many threads
+
+prefetch and switch on
+
+full/empty bits
+
+\subsection{Opportunities}
+
+little FP use ==> don't worry about FP registers
+
+would like enough concurrency to keep pipeline filled, but commodity CPUs don't have enough external bandwidth. So, get enough to saturate bandwidth.
+
+\section{Approach}
+
+Describe general approach.
+
+\subsection{Low-latency context switch}
+
+switch only stack pointer and program counter; depend on compiler to save/restore anything else. in general case, hopefully this will be fast.
 
 \subsection{Memory concurrency}
+
+Commodity procs don't provide enough. use software? FPGA?
+
 \subsection{Synchronization}
 
-\section{Results}
+use existing atomic instructions? Use FPGA listening snoops on QPI?
 
-\section{Future work}
+\section{Current Implementation and Results}
+
+current: just context switch on single node
+
+linked lists:
+
+there is some room for memory concurrency
+
+thread contexts don't take up too much space in cache
+
+sparse matmul:
+
+speedup?
+
+\section{Related work}
+
+hardware:
+
+XMT/MTA
+
+cyclops
+
+niagra
+
+software:
+
+qthreads
+
+cilk? tbb?
 

### 2011-01-04 doc/affiliates/greenery.tex

\documentclass{beamer}
\usetheme{Dresden}
\usepackage{enumerate,listings}

\setbeamertemplate{navigation symbols}{}
\begin{document}
\begin{frame}
\frametitle{libgreenery}
Goal: expose more memory references through independent contexts.
\begin{enumerate}
  \item Overhead between threads' references can't outweigh gain from prefetch
  \item Context switch must be as fast as possible ($<10$ns)
  \item Many threads must fit in same cache (tiny stacks)
  \item Dramatic limits on what you can run inside these
  \end{enumerate}
\begin{block}{API}
  Minimal as possible:
  \begin{enumerate}
  \item \texttt{thread\_spawn}, \texttt{thread\_yield}
  \item \texttt{thread\_block} (mostly avoid explicit synchronization)
  \item \texttt{prefetch\_and\_switch} (the key)
  \end{enumerate}
\end{block}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Example usage}
Unlike hardware threads, we \emph{want} these to cooperate tightly.
\begin{lstlisting}
// shared
int examine_size, i = 0;
vertex examine[], next[];
// many threads run this:
while (i < examine_size) {
  vertex v = examine[i];
  vertex edges[] = prefetch_and_switch(edge_lists[v]);
  for (j = 0; j < edges.size; ++j) {
     vertex target = prefetch_and_switch(edges[j]);
     next.push_back(target);
  }
}
thread_block(barrier)
// swap queues, repeat
\end{lstlisting}
Rely mostly on lack of preemption for synchronization, work on shared
data, and prefetch/switch on any reference with poor locality.
\end{frame}
\end{document}

### 2011-01-11 doc/hotpar2011/softxmt-hotpar2011.tex
abstract

@@
 \begin{abstract}
-Multithreading is good
+
+  Large irregular applications such as sparse graph problems often
+  perform badly on commodity clusters. These application often make
+  many nonsequential accesses to global data structures with low
+  degrees of locality, which interact badly with commodity processors'
+  memory systems, which are optimized for local, sequential memory
+  accesses.
+
+  Custom hardware optimized for these irregular applications has been
+  built: it exhibits excellent performance and energy efficiency for
+  these applications, but performs badly on standard sequential
+  workloads.
+
+  We believe we can have the best of both worlds. We argue that modern
+  commodity processors have many of the features necessary to give
+  good performance on irregular problems, and that a latency-tolerant
+  runtime designed to exploit these features can enable speedup for
+  irregular applications. For the areas where commodity hardware is
+  lacking, we suggest an approach for adding a small amount of
+  additional hardware to make up the difference.
+
+  In this paper, we describe the key features of such a model and
+  runtime system. We show results from simple benchmarks run on a
+  preliminary implementation of a latency-tolerant runtime that show
+  this approach has promise.
+
 \end{abstract}
@@
 sparse graph problems important
+
+Important 
+Social networks
+Facebook: 500M active users, 130 friends each on average. for
+computing friend suggestions, 40-node cluster, 72GB/node, run every 2
+days.
+
+Most commodity processors sold today have multiple cores, and their
 
@@
 \subsection{Opportunities}
+
+characteristics:
 
@@
 use existing atomic instructions? Use FPGA listening snoops on QPI?
+
+parition remote memory on each node; have one thread access each
+partition, making sync easier
 

### 2011-01-11 doc/hotpar2011/softxmt-hotpar2011.tex
abstract

--- 2011-01-03 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \begin{abstract}
-Multithreading is good
+
+  Large irregular applications such as sparse graph problems often
+  perform badly on commodity clusters. These application make many
+  non-sequential accesses to global data structures with low degrees of
+  locality, interacting badly with commodity processors' memory
+  systems, which are optimized for local, sequential memory
+  accesses. Special-purpose hardware optimized for these irregular
+  applications has been built; the Cray XMT is one such machine. It
+  uses multithreading to tolerate the latency of low-locality memory
+  accesses, and while it exhibits excellent performance and energy
+  efficiency for irregular applications, it performs badly on standard
+  sequential applications. We believe we can have the best of both
+  worlds: we argue that modern commodity processors have many of the
+  features necessary to give good performance on irregular problems,
+  and that a latency-tolerant runtime designed to exploit these
+  features can enable speedup for irregular applications on commodity
+  processors. In this paper, we describe the key features of such a
+  runtime system, and what a cluster built using our runtime might
+  look like. Using a preliminary single-node implementation of our
+  runtime, we obtained speedups on simple list chasing and sparse
+  matrix multiply benchmarks. Our results show promise for a
+  multi-node implementation.
+
 \end{abstract}
@@
 sparse graph problems important
+
+Important 
+Social networks
+Facebook: 500M active users, 130 friends each on average. for
+computing friend suggestions, 40-node cluster, 72GB/node, run every 2
+days.
+
+- difficult to partition
+
+Most commodity processors sold today have multiple cores, and their
 

### 2011-01-12 doc/hotpar2011/softxmt-hotpar2011.tex
comments (\todo command) on abstract

@@
+%% move stuff to defs file
+\newcommand{\todo}[1]{{\bfseries [[#1]]}}
+%% To disable, just uncomment this line
+%% \renewcommand{\todo}[1]{\relax}
+
 \documentclass{acm_proc_article-sp}
@@
   worlds: we argue that modern commodity processors have many of the
-  features necessary to give good performance on irregular problems,
+  features necessary \todo{Brandon: we will have to be able to list these as features in the paper or use a different phrase; my feeling is that these are aspects of the system which are not optimized for this but are sufficient to suggest potential for enough memory concurrency} to give good performance on irregular problems,
   and that a latency-tolerant runtime designed to exploit these
@@
   look like. Using a preliminary single-node implementation of our
-  runtime, we obtained speedups on simple list chasing and sparse
-  matrix multiply benchmarks. Our results show promise for a
-  multi-node implementation.
+  runtime, we obtained speedups on simple list chasing benchmarks. Our results show promise for a multi-node implementation.
 
@@
 
-sparse graph problems important
+sparse graph problems important (our specific area is sparse graphs with low diameter, like social networks)
 

### 2011-01-13 doc/hotpar2011/softxmt-hotpar2011.tex
conclusion

--- 2011-01-11 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \documentclass{acm_proc_article-sp}
+\usepackage{url}
 \begin{document}
@@
 
+Many problems in bioinformatics and social networks
+
 sparse graph problems important
+
+Many
+
+Social networks are an
 
@@
 Social networks
-Facebook: 500M active users, 130 friends each on average. for
-computing friend suggestions, 40-node cluster, 72GB/node, run every 2
-days.
+
+Social networks are
+
+And social networkIn January 2011, Facebook had 500 million active users with an
+average of 130 friends each \cite{Facebook:2011p91}. Facebook's
+friend suggestion algorithm in 2010 ran on a 40-node cluster with 72GB
+memory per node. Even with that capacity, new suggestions could only
+be computed every two days \cite{Backstrom:2010p90}.
+
+We
 
@@
 
-could just wait for wider multicores, but if context switching works, we can get enough concurrency on fewer cores and shut off the rest.
+could just wait for wider multicores, but if context switching works,
+we can get enough concurrency on fewer cores and shut off the rest.
+
+replicate or partition
 
@@
 
+Large irregular problems found in areas such as social networking and
+bioinformatics tend to have poor performance on commodity clusters due
+to freqent low-locality memory accesses. We present a programming
+model and runtime system that uses lightweight context switching to
+tolerate long-latency accesses and increase performance on these
+applications. Our preliminary evaluation shows that the overhead of
+context switching is not prohibitive for thread counts needed for a
+large-scale implementation, but memory concurrency is limited by
+hardware structures in commodity processors. We describe how we plan
+to address the challenges of large-scale memory concurrency and
+lightweight synchronization.
+
+\bibliographystyle{acm}
+\bibliography{softxmt-hotpar2011}
+
 \end{document}

### 2011-01-13 doc/hotpar2011/softxmt-hotpar2011.tex

--- 2011-01-12 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \documentclass{acm_proc_article-sp}
+\usepackage{url}
 \begin{document}
 
-\title{Latency-tolerant runtimes for large graph computations}
+\title{Latency-tolerant runtime for large graph computations}
 
@@
 
-  Large irregular applications such as sparse graph problems often
+  Large irregular applications such as sparse, low diameter graph problems often
   perform badly on commodity clusters. These application make many
@@
 
-sparse graph problems important (our specific area is sparse graphs with low diameter, like social networks)
+Graphs have proven to be a useful abstraction for solving real-world
+problems in social and biological network analysis. But many graphs in
+social and biological network analysis are sparse, with low diameter
+and a degree distribution that follows a power law. These properties
+make it difficult to take advantage of the locality assumptions built
+into the memory hierarchy of modern processors, so that even simple
+traversals are dominated by memory latency.
 
-Important 
-Social networks
-Facebook: 500M active users, 130 friends each on average. for
-computing friend suggestions, 40-node cluster, 72GB/node, run every 2
-days.
+And these graphs can be large---too large for a single node. In
+January 2011, Facebook had 500 million active users with an average of
+130 friends each \cite{Facebook:2011p91}. A natural approach would be
+to try to partition the graph across the nodes in a cluster. But with
+low diameter graphs, a traversal starting at any node is likely to
+quickly leave its partition and travel over the network, incurring
+high latency costs. And power law graphs have a small number of nodes
+that generate significantly more work that other nodes, leading to
+computational imbalance.
 
-- difficult to partition
+% Facebook's friend suggestion
+% algorithm in 2010 ran on a 40-node cluster with 72GB memory per
+% node. Even with that capacity, new suggestions could only be computed
+% every two days \cite{Backstrom:2010p90}.
 
-Most commodity processors sold today have multiple cores, and their
+Multithreading is a technique that has been used successfully to
+implement efficient computations for these graphs. The Cray XMT is an
+example of such a solution: it tries to solve the memory latency
+problem through concurrency rather than caching. Each XMT processor
+supports 128 hardware contexts and $~$1000 outstanding memory
+operations and is able to switch contexts every cycle. But this
+approach has a limitation: each context is significantly slower than a
+modern commodity processor, yielding bad single-threaded
+performance. While the XMT is fast at sparse, low-degree graph
+problems, it is slow at other scientific codes that can take advantage
+of commodity processors' memory hierarchies. Thus, few XMTs have been
+built.
 
-perform badly on commodity processors, since most accesses are misses
+We want the best of both worlds---we want to build a system that has
+good performance on both these low-locality graph codes as well as on
+common scientific codes. Our approach is inspired by the XMT, but
+implemented in software on commodity microprocessors. The core of our
+approach is a lightweight coroutine library designed to overlap
+long-latency memory prefetches with other contexts' computation. This
+document describes our approach.
 
-context-swtiching for latency tolerance has been successful with MTA, but single thread performance sucks
+\todo{do we want to mention hardware at this point? If so, we should
+  perhaps call it a hybrid approach.}
 
-we want latency tolerance on commodity processors for best of both worlds
-
-could just wait for wider multicores, but if context switching works, we can get enough concurrency on fewer cores and shut off the rest.
-
-\section{Programming model for graph problems}
+\section{Programming model}
 
@@
 
-Describe general approach.
+Describe general approach, explaining
 
@@
 partition, making sync easier
+
+\subsection{Extending to a multi-node system}
 
@@
 
+why not wait for more cores?
+
+hyperthreading?
+
 \section{Conclusion}
 
+Large irregular problems found in areas such as social networking and
+bioinformatics tend to have poor performance on commodity clusters due
+to frequent low-locality memory accesses. We present a programming
+model and runtime system that uses lightweight context switching to
+tolerate long-latency accesses and increase performance on these
+applications. Our preliminary evaluation shows that the overhead of
+context switching is not prohibitive for thread counts needed for a
+large-scale implementation, but memory concurrency is limited by
+hardware structures in commodity processors. We describe how we plan
+to address the challenges of large-scale memory concurrency and
+lightweight synchronization.
+
+\bibliographystyle{acm}
+\bibliography{softxmt-hotpar2011}
+
 \end{document}

### 2011-01-13 doc/hotpar2011/softxmt-hotpar2011.tex
intro and conclusion
minor comment  in intro

@@
 
-\title{Latency-tolerant runtime for large graph computations}
+\title{A latency-tolerant runtime for large graph computations}
 
@@
 
-And these graphs can be large---too large for a single node. In
-January 2011, Facebook had 500 million active users with an average of
+And these graphs can be large---too large for a single node \todo{may want to avoid saying 'node' since graph 'node' is also used in this paragraph}.
+In January 2011, Facebook had 500 million active users with an average of
 130 friends each \cite{Facebook:2011p91}. A natural approach would be
@@
 high latency costs. And power law graphs have a small number of nodes
-that generate significantly more work that other nodes, leading to
+that generate significantly more work than other nodes, leading to
 computational imbalance.

### 2011-01-13 doc/hotpar2011/softxmt-hotpar2011.tex
comment on abstract

@@
   efficiency for irregular applications, it performs badly on standard
-  sequential applications. We believe we can have the best of both
+  sequential applications \todo{Brandon: should this be more specific or worded differently?
+Its true sequential perform badly since switching between many threads, but also does XMT also perform  poorly (relative to alternatives) on highly parallel but regular apps with lots of temporal and spatial locality?}.
+We believe we can have the best of both
   worlds: we argue that modern commodity processors have many of the

### 2011-01-13 doc/hotpar2011/evaluation.tex
add eval rough draft

To study the memory concurrency in the system, we did experiments that approximate large irregular applications. The experiments consist of multiple threads each walking a number of separate linked lists. Each memory reference of a node is highly likely to result in an access to main memory, and we control the number of concurrent requests offered by a thread by varying the number of lists each thread walks. \todo{Brandon: somewhere we might explicitly say that this experiment involves no synchronization, and relate to plan to build this as mentioned in the components of runtime section} \todo{Brandon: is it important to mention that for isolation we use huge pages to reduce TLB misses?} Our experiments were run on a Xeon X5660, which uses the Nehalem microarchitecture with 2 sockets, 6 cores per socket, and 12GB of DRAM per socket. For the results listed memory allocations were pinned to socket0's DRAM.
\subsection{Results} 
	We found that a single core is bandwidth-limited by the size of the L1 cache line-fill buffer (keeps track of outstanding misses of the core); bandwidth plateaus at about 10 concurrent misses, the size of the buffer. Running multiple cores, the system reaches a bandwidth plateau with fewer outstanding misses per core. In fact, the maximum number of total memory references in flight on the socket is consistently 36 (see figure). When using all six cores on the socket, bandwidth peaks at 275Mref/s. Since this is less than even a conservative bound on available pin bandwidth, there is some bottleneck to performance in the memory system. A likely place is the global read-tracker queue containing references to L3 cache; it has a capacity of 32 references, which concurs with the observed limit of 36 in-flight references. In addition, we performed measurements using performance counters, which consistently showed an average queue occupancy of 27, also close to the capacity.
 \todo{Brandon: it is great to identify why the bandwidth is limited, but in paper need to connect to why it is important. in general, we get an idea of what might need to be changed, especially as latency goes up and we need more concurrency}  \todo{Brandon: may cite nehalem bandwidth numbers here as well to validate our bw numbers with a second source}
	Bandwidth of memory requests from cores on a single socket seems to peak at about 275Mreq/s; however adding requests to socket0's memory from socket1 cores increases the bandwidth to about	350Mreq/s, showing that there is additional room in the memory system that is simply reserved for remote requests.

\subsection{Using green threads}
	If a core is to generate some number of concurrent memory requests, it may need multiple contexts \todo{Brandon: this idea likely goes elsewhere in describing overall approach}.  We ran our experiments with varying numbers of green threads per core to use multiple contexts to achieve concurrency. Our results showed (see fig) that using green threads gives little to no overhead due to context switching. \todo{Brandon: somehow also tie in prefetch-and-switch; also might elaborate on the results and yield-only if space}

\subsection{Considering the 'multi-node case'}
	\todo{needs introduction, which will depend on what the system description section looks like}
	We assume that the QPI link to the network interface will be the minimum bandwidth and not the network. On the Nehalem, the QPI link to the network interface provides a bandwidth 12.8GB/s each way. This as much as 200Mref/s for 64 byte references. Given that half of the bandwidth will go towards satisfying remote requests, leaving half to making requests \todo{n-1/n req vs. sum-over-n(1/(n-1))}.  So, the cores must put out a bandwidth of 100Mref/s for full utilization of the link.  Assuming a reasonable network latency of $1-2\mu s$, by Little's Law, the number of outstanding references is $100Mref/s*1-2\mu s = 100-200$. This is the amount of memory concurrency that needs to be generated for the latency of memory requests to be hidden.
We consider two issues: first, can we support 100-200 misses?, and second, can the caches support enough contexts? \todo{Brandon: might want to be more explicit about the interplay between cache pressure, number of contexts, amount of references that can be generated by X number of contexts, etc}
	Can we support 100-200 misses? If the remote requests are reads that go to the global queue, then the number of outstanding references is limited to about 36. If the remote requests do not need to be stored in the global queue, then the line-fill buffers of the cores can support a total of $6 cores * 10 ref/core = 60 references$, still not quite enough. If, however, prefetches are messages--memory stores--requesting a piece of remote memory then many more outstanding references could be supported, either with a software data structure or a memory management device implemented in an FPGA.
	Can the caches support enough contexts? To start exploring this question, we did cache pressure experiments measuring the effect of working set size and number of software contexts on performance. Our experiments suggest that with up to about 8 contexts/core, we can achieve almost the same max bandwidth, and for one update per remote reference (expected for basic graph comps?), working set size/thread of up to 64KiB scales well. \todo{Brandon: need to be more specific.  Also, are there important highlights regarding how the plot scales as local updates increases}.  Threads do not take up much space in the cache, shown by XXX.

\todo{should we mention the lack of synchronization in the experiments and how intend to explore it in next step, given that it is part of our proposed runtime?}

\todo{order seems okay, but rework organization of subsections}

### 2011-01-16 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   softxmt-hotpar2011.tex

@@
 
+While the ideal programming model is one that facilitates easy
+expression in the problem domain as well as efficient execution on the
+available hardware, our programming model today is motivated by only
+efficient execution: it exists as a rudimentary library for concept
+demonstration, not as a language suitable for development.  The
+programming model we use today is therefore only partially formed. It
+exists as an augmentation of existing threaded programming models such
+as pthreads or openmp that are readily available on most computer
+systems.
+
+Though not strictly necessary, for ease of exposition we imagine a
+one-to-one binding between threads and cores.  Our model manifests
+within any thread independently of the others as a ``fray'' [[for lack
+of better term ]].  Within the thread, a fray instance is the
+instantiation of a set of coroutines and a scheduler.  Instantiation
+is at user-level and the operating system sees the entire fray as as a
+single user-level thread.
+
+[[ describe calls that perform thread & scheduler instantiation; and schedule activation ]]
+
+Each coroutine executes non-preemptively, utilizing the core to which
+its ``parent'' thread has been assigned by the operating system until it
+reaches a yield point.  Typically, the coroutine yields after
+initiating a prefetch operation on data that the programmer or
+compiler anticipates would probably not complete before the data is
+referenced by a subsequent load operation.  In this way, the core
+continues to execute instructions for other coroutines when otherwise
+it would likely stall.  The programming model thus tolerates latency.
+
+[[ describe calls to prefetch-and-yield ]]
+
+When a thread frays, for the sake of expediency, a fixed portion of
+its stack is divided amongst the coroutines.  Given a finite stack,
+each coroutine is constrained to execute a call tree of depth known to
+be finite at compilation time: recursion is not yet permitted.
+
+[[ discussion of coroutines & synchronization ]] Were a coroutine to
+block on a synchronization variable shared with other threads, the
+entire fray would suspend execution.  This can lead to deadlock when,
+for example, one coroutine waits to consume from another thread that
+is waiting to consume what only another coroutine in this first fray
+can produce.  Instead, coroutines must yield on failed synchronization
+events, spinning rather than blocking, where were they bonafide
+thereads, blocking might be more efficient.
+
+[[ say something about ``synchronization'' within a fray ]]
+
+Coroutines completing their work yield without adding themselves to
+the scheduling queue.  The last coroutine to exit in this way returns
+as the main thread, as in the common  fork-join model of parallelism.
+
 many many threads
@@
 \section{Current Implementation and Results}
+
+\todo{Begin eval insert}
+\input{evaluation}
+\todo{End eval insert}
 

### 2011-01-17 doc/hotpar2011/softxmt-hotpar2011.tex
minor formatting edits

@@
 problem through concurrency rather than caching. Each XMT processor
-supports 128 hardware contexts and $~$1000 outstanding memory
+supports 128 hardware contexts and $\sim$1000 outstanding memory
 operations and is able to switch contexts every cycle. But this
@@
 one-to-one binding between threads and cores.  Our model manifests
-within any thread independently of the others as a ``fray'' [[for lack
-of better term ]].  Within the thread, a fray instance is the
+within any thread independently of the others as a ``fray'' \todo{for lack
+of better term }.  Within the thread, a fray instance is the
 instantiation of a set of coroutines and a scheduler.  Instantiation
@@
 
-[[ describe calls that perform thread & scheduler instantiation; and schedule activation ]]
+\todo{ describe calls that perform thread \& scheduler instantiation; and schedule activation }
 
@@
 
-[[ describe calls to prefetch-and-yield ]]
+\todo{ describe calls to prefetch-and-yield }
 
@@
 
-[[ discussion of coroutines & synchronization ]] Were a coroutine to
+\todo{ discussion of coroutines \& synchronization } Were a coroutine to
 block on a synchronization variable shared with other threads, the
@@
 
-[[ say something about ``synchronization'' within a fray ]]
+\todo{ say something about ``synchronization'' within a fray }
 

### 2011-01-21 doc/hotpar2011/softxmt-hotpar2011.tex

@@
 
-\title{A latency-tolerant runtime for large graph computations}
+\title{Can We Crunch the Social Graph on the Cheap?}
 
@@
 approach is a lightweight coroutine library designed to overlap
-long-latency memory prefetches with other contexts' computation. This
-document describes our approach.
+long-latency memory prefetches with other contexts' computation.
+
+\todo{something like: project still in early stages. this document lays out
+  problems we think a system like this needs to solve, and describes
+  some implementation strategies we want to explore. we have a basic
+  initial implementation with some experiments that suggest there is
+  promise here.}
 
@@
 can produce.  Instead, coroutines must yield on failed synchronization
-events, spinning rather than blocking, where were they bonafide
-thereads, blocking might be more efficient.
+events, spinning rather than blocking, where were they bona fide
+threads, blocking might be more efficient.
 
@@
 
-many many threads
-
-prefetch and switch on
-
-full/empty bits
-
-\subsection{Opportunities}
-
-characteristics:
-
-little FP use ==> don't worry about FP registers
-
-would like enough concurrency to keep pipeline filled, but commodity CPUs don't have enough external bandwidth. So, get enough to saturate bandwidth.
-
 \section{Approach}
 
-Describe general approach, explaining
+Our goal is to enable high throughput on graphs with two challenging
+properties. First, the graphs' combination of sparseness and low
+degree means there is little locality to be exploited. Second, the
+graphs are larger than a single node's memory and do not partition
+well \todo{explain?}. We must spread the graph across multiple nodes
+and send memory operations over a network. If memory requests are
+uniformly distributed over the working set, a given node will make as
+many accesses to a distant node's memory as it does to the local node.
+
+We would like to run as fast as current hardware will let us. Pin
+bandwidth is the tightest resource: Intel's QuickPath Interconnect
+(QPI) allows approximately 100 million requests per second
+\todo{cite}, and recent interconnect products allow similar request
+rates, with a round-trip latency of approximately 3$\mu$s
+\todo{cite}. Little's law tells us then that we must support on the
+order of 300 concurrent memory operations, and thus coroutines, to
+cover this latency.
+
+These requirements drive the design of our system. There are three key
+components.
 
@@
 
-switch only stack pointer and program counter; depend on compiler to save/restore anything else. in general case, hopefully this will be fast.
+There are three main concerns in designing the coroutine
+mechanism. First, contexts must be small, so many coroutines can be
+active without monopolizing the cache. Second, context switches must
+be fast, so that overhead does not swamp actual work. Third, when a
+coroutine yields, we must choose a good coroutine to run next.
 
-\subsection{Memory concurrency}
+We minimize context size by treating context switches as function
+calls, as in \cite{charm}. This allows us to save and restore the
+minimum number of registers allowed by the ABI, and depend on the
+compiler to save and restore any other registers in use. We minimize
+context switch time both by keeping contexts small and by doing all
+switching and scheduling in user space, as in \todo{cite other green
+  threads packages}.
 
-Commodity procs don't provide enough. use software? FPGA?
+The choice of which coroutine to run next presents us with a
+tradeoff. The simplest scheduler chooses coroutines to execute in
+round-robin fashion. But this may lead to performance problems: a
+coroutine might be scheduled before the event which caused it to yield
+has completed, stalling the pipeline and keeping other coroutines from
+progressing. A coroutine might also be scheduled long after the event
+which caused it to yield has completed; while this wouldn't block
+other coroutines from proceeding, it would add unnecessary latency to
+the yielding coroutine. A more intelligent scheduler might allow
+coroutines to execute only when they were ready, but this could
+increase the overhead of context switching. Our initial implementation
+takes the simple approach, but more investigation is warranted.
+
+\subsection{Accessing Global Memory}
+
+% Our goal is to support many concurrent accesses to a global address
+% space spread across nodes in a cluster. We can draw on hardware
+% support developed for the Partitioned Global Address Space family of
+% languages to accomplish this.
+
+% We draw on libraries and network support 
+% goal. Languages in the Partitioned Global Address Space family such as
+% X10 and Chapel have similar goals,
+
+% It is also useful to access this low-locality data without disturbing
+% the thread contexts and local storage in the cache. Current x86
+% processors support non-temporal loads and stores. These memory
+% operations are quite relevant for our problem: they are designed to
+% move data without polluting the cache. Using these instructions works
+% better than mapping the low-locality data as uncached, since that
+% restricts the processors' ability to reorder operations.
+
+Following the lead of the PGAS family of languages, we will use
+the RDMA functionality provided by modern Infiniband networks to
+access our global memory. \todo{rethink?}
+
+These networks allow us to read and write memory directly from remote
+nodes, but they do so through a low-level interface designed to be
+hidden behind a library. Loads and stores to global memory must be
+replaced with calls to a library. This library causes the network
+interface to fetch the remote data and place it in a temporary
+buffer. When the requesting coroutine is scheduled again, data is
+returned from the buffer and execution proceeds.
+
+In order to minimize memory access overhead, the memory operation
+library calls do not interact with the network interface
+directly. Instead, they queue their operations to another thread, the
+{\em memory manager thread}, running on another core in the CPU. This thread
+translates the global address into a network destination and issues
+the network operation, checks for completion, and manages the returned
+data buffers.
+
+But we have to be careful---directing memory requests from multiple
+compute threads to a single memory manager thread may reduce memory
+concurrency. And we've increased the overhead of a single memory
+operation---what in the single node case was a simple read, is now
+multiple queuing operations between threads, and multiple reads and
+writes over QPI to the network interface. Adding more memory manager
+threads may enable more memory concurrency, but the overhead is still high.
+
+Instead, we propose hardware acceleration with a {\em memory manager
+  FPGA}, a coprocessor FPGA inserted into a Xeon socket so it can
+participate in the node's coherence domain. This coprocessor does no
+computation---it would only manage the global memory requests of the
+main CPU, performing the function of the memory manager thread
+described earlier.
+
+By participating in the coherence domain, the memory manager FPGA can
+act as a proxy for the rest of global memory. It advertises the
+address space of the rest of the system, and translates local loads
+and stores to addresses on remote nodes into the network operations
+required.
+
+Unfortunately, current processors do not provide sufficient hardware
+resources for us to reach our concurrency target using this
+approach. Each core in Intel's X5650 Xeon supports only 10 outstanding
+misses from its private L2; the entire chip is limited to 32 L2 misses
+to local memory and 12 L2 misses for memory reached over the QPI
+bus. A simple translation of a local load to a network access would
+tie up one of these slots for the entire multi-microsecond latency of
+the request. We must bypass these structures to allow the amount of
+concurrency we need; the CPU needs to direct the memory manager FPGA
+to do work on its behalf without tying up its own resources.
+
+To accomplish this, we encode commands in the high-order bits of
+remote memory addresses. The memory manager FPGA advertises each
+remote address twice. The first address is used for blocking reads and
+writes---any access incurs the full latency of the remote
+access. \todo{verify writes block?} The second address is used for
+prefetches and asynchronous writes. These addresses differ by one
+high-order bit, easily set with a mask in a macro. A read from the
+prefetch address starts the network transaction, but immediately
+returns a dummy value. When the value arrives, it is stored in a
+temporary buffer; when the requesting coroutine is scheduled again, it
+executes the blocking read, and the memory manager immediately returns
+the value from the buffer. \todo{this isn't fully baked---do we want
+  to include this sort of detail?}
 
@@
 
-use existing atomic instructions? Use FPGA listening snoops on QPI?
+In systems with hundreds of threads per node, lightweight
+synchronization is important. While we have not implemented any
+synchronization primitives yet, full-empty bits are a natural choice
+given our programming model. Two questions then arise: at what
+granularity should full-empty bits work, and what component should do
+the synchronization?
 
-parition remote memory on each node; have one thread access each
-partition, making sync easier
+We expect most data accesses to be to 64-bit words. It is possible to
+support full-empty bit synchronization on arbitrary words by
+allocating additional storage for the full-empty bits, and translating
+each atomic full-empty operation into a sequence of memory operations
+that modify the data and full-empty bit atomically.
 
-\subsection{Extending to a multi-node system}
+One potential optimization is to limit full-empty synchronization to
+pointers to aligned 64-bit words. This leaves the bottom three bits
+free to store the full-empty bits. These bits would be masked out when
+the pointer was returned to the user.
+
+These techniques can be implemented in software; \cite{qthreads} is
+one example. Alternatively, atomic operations could be delegated to a
+memory manager FPGA as described in the previous section. The address
+command encoding would be extended to support the various full-empty
+operations.
+
+It might seem natural to use the FPGA as a memory controller as
+well, storing all full-empty-bit-capable memory in DRAMs attached to
+the FPGA. While this is conceptually simple, it does make the
+programming model and the design more complicated. Another possibility
+would be to store only the full-empty bits in memories attached to the
+FPGA, and leave data in memory attached to the CPU. We believe that
+starting simple is best: both data and full-empty bits should be
+stored in the CPU's memory. \todo{discuss QPI bandwidth differences}
 
@@
 \todo{End eval insert}
-
-current: just context switch on single node
-
-linked lists:
-
-there is some room for memory concurrency
-
-thread contexts don't take up too much space in cache
-
-sparse matmul:
-
-speedup?
 

### 2011-01-22 doc/hotpar2011/evaluation.tex
added threed eval figs

@@
 Each memory reference of a node is highly likely to result in an access to main memory, and we control the number of concurrent requests offered by a thread by varying the number of lists each thread walks.
-\todo{Brandon: somewhere we might explicitly say that this experiment involves no synchronization, and relate to plan to build this as mentioned in the components of runtime section} \todo{Brandon: is it important to mention that for isolation we use huge pages to reduce TLB misses?} Our experiments were run on a Xeon X5660, which uses the Nehalem microarchitecture with 2 sockets, 6 cores per socket, and 12GB of DRAM per socket.
-For the results listed memory allocations were pinned to socket0's DRAM.
+\todo{Brandon: somewhere we might explicitly say that this experiment involves no synchronization, and relate to plan to build this as mentioned in the components of runtime section} \todo{Brandon: is it important to mention that for isolation we use huge pages to reduce TLB misses?} Our experiments were run on a Xeon X5660, which uses the Nehalem microarchitecture, with 2 sockets, 6 cores per socket, and 12GB of DRAM per socket.
+For the results listed, memory allocations were pinned to socket0's DRAM.
 \subsection{Results}
-	We found that a single core is bandwidth-limited by the size of the L1 cache line-fill buffer (keeps track of outstanding misses of the core); bandwidth plateaus at about 10 concurrent misses, the size of the buffer.
+
+\todo{crop the pdfs}
+\todo{enlarge font of tableau plot since pdf is getting scaled down}
+\begin{figure*}
+	\begin{center}
+ 		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-maxrates.pdf}
+	\end{center}
+	\caption{The max rates for processing list nodes with varying numbers of cores.
+Each core walks one list, and each node access results in a reference to main memory.
+All memory is local to socket0.
+Cores 1-6 are on socket0, and cores 7-12 are on socket1.}
+	\label{fig:listwalk-maxrates}
+\end{figure*}
+
+\begin{figure*}
+	\begin{center}
+ 		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-outflats.pdf}
+	\end{center}
+	\caption{}
+	\label{fig:listwalk-outflats}
+\end{figure*}
+
+\todo{choose best phrase to use for outstanding/concurrent miss/refs/reads and use consistently unless there is a place one makes most sense}
+\todo{if mentioning the cores 7-12 then need to talk about QPI link probably}
+	 Figure~\ref{fig:listwalk-outflats} shows the total number of outstanding misses the system can handle before performance plateaus.
+We found that a single core is bandwidth-limited by the size of the L1 cache line-fill buffer (keeps track of outstanding misses of the core).
+Bandwidth \todo{word:performance?} plateaus at about 10 concurrent misses, the size of the buffer.
 Running multiple cores, the system reaches a bandwidth plateau with fewer outstanding misses per core.
-In fact, the maximum number of total memory references in flight on the socket is consistently 36 (see figure).
-When using all six cores on the socket, bandwidth peaks at 275Mref/s.
+In fact, the maximum number of total memory references in flight on a single socket is consistently 36 (see figure).
+When using all six cores on the socket, bandwidth peaks at about 275Mref/s.
 Since this is less than even a conservative bound on available pin bandwidth, there is some bottleneck to performance in the memory system.
-A likely place is the global read-tracker queue containing references to L3 cache; it has a capacity of 32 references, which concurs with the observed limit of 36 in-flight references.
-In addition, we performed measurements using performance counters, which consistently showed an average queue occupancy of 27, also close to the capacity.
+A likely place is the global read-tracker queue containing references to the L3 cache; it has a capacity of 32 references, which concurs with the observed limit of 36 in-flight references.
+In addition, we performed measurements using performance counters, which consistently showed an average queue occupancy of 27, also close to the capacity 32.
  \todo{Brandon: it is great to identify why the bandwidth is limited, but in paper need to connect to why it is important.
 in general, we get an idea of what might need to be changed, especially as latency goes up and we need more concurrency}  \todo{Brandon: may cite nehalem bandwidth numbers here as well to validate our bw numbers with a second source}
-	Bandwidth of memory requests from cores on a single socket seems to peak at about 275Mreq/s; however adding requests to socket0's memory from socket1 cores increases the bandwidth to about	350Mreq/s, showing that there is additional room in the memory system that is simply reserved for remote requests.
+	Bandwidth of memory requests from cores on a single socket seems to peak at about 275Mreq/s; however adding requests to socket0's memory from socket1 cores increases the bandwidth to about	350Mreq/s, showing that there is additional room in the memory system that is simply reserved for remote requests.\todo{Currently a bit of a
+hanging statement, but it is intended to show more evidence of limit being queue.}
+
+\begin{figure*}
+	\begin{center}
+ 		\includegraphics[width=0.5\textwidth]{figures/multi-core-green-thread-1gb.pdf}
+	\end{center}
+	\caption{}
+	\label{fig:listwalk-green-threads}
+\end{figure*}
 
 \subsection{Using green threads}
-	If a core is to generate some number of concurrent memory requests, it may need multiple contexts \todo{Brandon: this idea likely goes elsewhere in describing overall approach}.
+	If a core is to generate some number of concurrent memory requests to tolerate latency and keep busy, it may need multiple contexts \todo{Brandon: this idea likely goes elsewhere in describing overall approach}.
 We ran our experiments with varying numbers of green threads per core to use multiple contexts to achieve concurrency.
-Our results showed (see fig) that using green threads gives little to no overhead due to context switching.
+Our results, displayed in Figure~\ref{fig:listwalk-green-threads} \todo{decide whether max/avg rate as well as how much info/whether to make consistent with the other plots format}, show that using green threads gives little to no overhead due to context switching.
 \todo{Brandon: somehow also tie in prefetch-and-switch; also might elaborate on the results and yield-only if space}
@@
 If, however, prefetches are messages--memory stores--requesting a piece of remote memory then many more outstanding references could be supported, either with a software data structure or a memory management device implemented in an FPGA.
+\todo{talk about the simulated delay experiments?
+mention that there are still the same memory system limits like queues so results don't show what perf would look like beyond a certain point, but do show that by adding contexts we can achieve the same performance when latency goes up.}
 	Can the caches support enough contexts?

### 2011-01-22 doc/hotpar2011/evaluation.tex
edits, rewriting, images

@@
-To study the memory concurrency in the system, we did experiments that approximate large irregular applications.
+To study memory concurrency in commodity systems and begin building a runtime, we did experiments that approximate large irregular applications.
 The experiments consist of multiple threads each walking a number of separate linked lists.
 Each memory reference of a node is highly likely to result in an access to main memory, and we control the number of concurrent requests offered by a thread by varying the number of lists each thread walks.
-\todo{Brandon: somewhere we might explicitly say that this experiment involves no synchronization, and relate to plan to build this as mentioned in the components of runtime section} \todo{Brandon: is it important to mention that for isolation we use huge pages to reduce TLB misses?} Our experiments were run on a Xeon X5660, which uses the Nehalem microarchitecture, with 2 sockets, 6 cores per socket, and 12GB of DRAM per socket.
+\todo{Brandon: somewhere we might explicitly say that this experiment involves no synchronization on list nodes, and relate to plan to build this as mentioned in the components of runtime section} Our experiments were run on a Xeon X5660, which uses the Nehalem microarchitecture, and has 2 sockets, 6 cores per socket, and 12GB of DRAM per socket.
 For the results listed, memory allocations were pinned to socket0's DRAM.
+Memory references from socket1's cores must be satisfied by sending requests over the QPI link between the sockets.
+\todo{Brandon: is it important to mention that for isolation we use huge pages to reduce TLB misses?}
 \subsection{Results}
 
-\todo{crop the pdfs}
-\todo{enlarge font of tableau plot since pdf is getting scaled down}
-\begin{figure*}
+\todo{Might still consider the reqrate v.
+conc miss graph, because of showing the relation of performance going up as there is more concurrency, or else make sure to describe this clearly in the text}.
+\begin{figure}[h]
 	\begin{center}
- 		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-maxrates.pdf}
+		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-maxrates-edited.pdf}
 	\end{center}
@@
 	\label{fig:listwalk-maxrates}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}
+\begin{figure}[h]
 	\begin{center}
- 		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-outflats.pdf}
+		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-outflats-edited.pdf}
 	\end{center}
-	\caption{}
+	\caption{This is my caption about multi-listwak-outflats-edited, it shows the number of misses when rate plateaus...}
 	\label{fig:listwalk-outflats}
-\end{figure*}
+\end{figure}
 
 \todo{choose best phrase to use for outstanding/concurrent miss/refs/reads and use consistently unless there is a place one makes most sense}
-\todo{if mentioning the cores 7-12 then need to talk about QPI link probably}
 	 Figure~\ref{fig:listwalk-outflats} shows the total number of outstanding misses the system can handle before performance plateaus.
-We found that a single core is bandwidth-limited by the size of the L1 cache line-fill buffer (keeps track of outstanding misses of the core).
-Bandwidth \todo{word:performance?} plateaus at about 10 concurrent misses, the size of the buffer.
+We found that for a single core, bandwidth plateaus once there are 10 concurrent misses; this is because the L1 cache line-buffer, which keeps track of outstanding misses of the core, is size 10.
+If concurrency is limited, there is less latency tolerance and performance is limited.
+
 	Running multiple cores, the system reaches a bandwidth plateau with fewer outstanding misses per core.
-In fact, the maximum number of total memory references in flight on a single socket is consistently 36 (see figure).
-When using all six cores on the socket, bandwidth peaks at about 275Mref/s.
-Since this is less than even a conservative bound on available pin bandwidth, there is some bottleneck to performance in the memory system.
+In fact, the maximum number of total memory references in flight on a single socket is consistently 36.
+When using all six cores on the socket, bandwidth peaks at about 275Mref/s (see Figure~\ref{fig:listwalk-maxrates}).
+Since this is less than even a conservative bound on available pin bandwidth to DRAM, there is some bottleneck to performance in the memory system.
 A likely place is the global read-tracker queue containing references to the L3 cache; it has a capacity of 32 references, which concurs with the observed limit of 36 in-flight references.
@@
  \todo{Brandon: it is great to identify why the bandwidth is limited, but in paper need to connect to why it is important.
-in general, we get an idea of what might need to be changed, especially as latency goes up and we need more concurrency}  \todo{Brandon: may cite nehalem bandwidth numbers here as well to validate our bw numbers with a second source}
+in general, we get an idea of what might need to be changed, especially as latency goes up and we need more concurrency}  \todo{cite ``multi-socket memory concurrency" to validate bandwidth numbers}
 	Bandwidth of memory requests from cores on a single socket seems to peak at about 275Mreq/s; however adding requests to socket0's memory from socket1 cores increases the bandwidth to about 350Mreq/s, showing that there is additional room in the memory system that is simply reserved for remote requests.\todo{Currently a bit of a
@@
 \subsection{Using green threads}
-	If a core is to generate some number of concurrent memory requests to tolerate latency and keep busy, it may need multiple contexts \todo{Brandon: this idea likely goes elsewhere in describing overall approach}.
+	If a core is to have the number of concurrent memory requests required to tolerate latency and keep busy, it may need multiple contexts \todo{Brandon: this idea likely goes elsewhere in describing overall approach}.
 We ran our experiments with varying numbers of green threads per core to use multiple contexts to achieve concurrency.

### 2011-01-23 doc/hotpar2011/evaluation.tex
multi-node case / delay sims

@@
 
-\subsection{Considering the 'multi-node case'}
-	\todo{needs introduction, which will depend on what the system description section looks like}
-	We assume that the QPI link to the network interface will be the minimum bandwidth and not the network.
+\subsection{Expanding to Multiple Compute Nodes}
+	\todo{better title?}\\
+	\todo{needs introduction, which will depend on what the system description section looks like}\\
+	We use the single processor experiments to give insight into concurrency requirements in a multi-node system that uses a network with relatively long latency.
+
 	On the Nehalem, the QPI link to the network interface provides a bandwidth 12.8GB/s each way.
-This as much as 200Mref/s for 64 byte references.
-Given that half of the bandwidth will go towards satisfying remote requests, leaving half to making requests \todo{n-1/n req vs.
-sum-over-n(1/(n-1))}.
-So, the cores must put out a bandwidth of 100Mref/s for full utilization of the link.
-Assuming a reasonable network latency of $1-2\mu s$, by Little's Law, the number of outstanding references is $100Mref/s*1-2\mu s = 100-200$.
-This is the amount of memory concurrency that needs to be generated for the latency of memory requests to be hidden.
+This as much as 200Mref/s for 64-byte messages.
+For uniformly random distributed references across the entire shared memory space, we can expect half of this outgoing bandwidth will go towards satisfying remote requests, leaving half for making requests.
+So, the cores can put out a bandwidth of 100Mref/s at full utilization of the link.
+Recent networks have suggested a max issue rate of 30M-90M messages per second is attainable \todo{cite mellanox press release and/or XT3/Eldorado} and this seems to be gradually increasing, so we use 100Mref/s as a good approximation to work with.
+Assuming a round-trip network latency of $1-2\mu s$, then by Little's Law, the number of outstanding references is $100Mref/s*1-2\mu s = 100-200$.
+\todo{Redo this calculation with proper numbers latency} This is the amount of memory concurrency that needs to exist for the latency of memory requests to be hidden when there is a rate of 100Mref/s.
 We consider two issues: first, can we support 100-200 misses?, and second, can the caches support enough contexts? 
-\todo{Brandon: might want to be more explicit about the interplay between cache pressure, number of contexts, amount of references that can be generated by X number of contexts, etc}
+In order to keep the cores busy, there must be sufficient memory concurrency, and the number of contexts required to achieve this will vary.
+As there are more contexts in a core, there are more workings sets, including both data and coroutine stacks, that should fit in the cache to maintain performance.
+
 	Can we support 100-200 misses?
@@
 If the remote requests do not need to be stored in the global queue, then the line-fill buffers of the cores can support a total of $6 cores * 10 ref/core = 60 references$, still not quite enough.
-If, however, prefetches are messages--memory stores--requesting a piece of remote memory then many more outstanding references could be supported, either with a software data structure or a memory management device implemented in an FPGA.
-\todo{talk about the simulated delay experiments?
-mention that there are still the same memory system limits like queues so results don't show what perf would look like beyond a certain point, but do show that by adding contexts we can achieve the same performance when latency goes up.}
+If, however, prefetches are messages--communicated with memory stores--requesting a piece of remote memory then many more outstanding references could be supported, either with a software data structure or a memory management device implemented in an FPGA.
+	We performed some experiments with green threads on listwalk, where we simulate the network latency by not allowing a coroutine to continue after it does a prefetch-and-switch until some specified amount of time has passed.
+As seen in Figure \todo{ref delay plot} it is still possible to achieve the maximum throughput: it just requires more contexts to supply enough concurrent references, as expected.
+This simulation is of course limited both by the fact that prefetched data is actually sitting in cache longer than it should (increasing the effective working set size) and that the memory system still caps performance, but it is a useful validation of adding software contexts to achieve the level of concurrency that is necessary.
+
 	Can the caches support enough contexts?
+Having too many separate working sets that do not fit into cache will cause performance degradation and thrashing.
 To start exploring this question, we did cache pressure experiments measuring the effect of working set size and number of software contexts on performance.
-Our experiments suggest that with up to about 8 contexts/core, we can achieve almost the same max bandwidth, and for one update per remote reference (expected for basic graph comps?), working set size/thread of up to 64KiB scales well.
+\todo{see carlson fig} Our experiments suggest that with up to about 8 contexts/core, we can achieve almost the same max bandwidth \todo{ as}, and for one update per remote reference \todo{expected for basic graph comps?}, working set size/thread of up to 64KiB scales well.
 \todo{Brandon: need to be more specific.
@@
 
-\todo{should we mention the lack of synchronization in the experiments and how intend to explore it in next step, given that it is part of our proposed runtime?}
-
-\todo{order seems okay, but rework organization of subsections}
+\todo{should we mention the lack of synchronization in the experiments and how intend to explore it in next step, given that it is part of our proposed runtime?
+Can we evaluate design choices qualitatively right now?}

### 2011-01-24 doc/hotpar2011/evaluation.tex

@@
 \todo{Brandon: somewhere we might explicitly say that this experiment involves no synchronization on list nodes, and relate to plan to build this as mentioned in the components of runtime section} Our experiments were run on a Xeon X5660, which uses the Nehalem microarchitecture, and has 2 sockets, 6 cores per socket, and 12GB of DRAM per socket.
-For the results listed, memory allocations were pinned to socket0's DRAM.
-Memory references from socket1's cores must be satisfied by sending requests over the QPI link between the sockets.
+For the results listed, memory allocations were pinned to the DRAM of the first socket (``socket0").
+Memory references from cores of socket1 must be satisfied by sending requests over the QPI link between the sockets.
 \todo{Brandon: is it important to mention that for isolation we use huge pages to reduce TLB misses?}
@@
 
-\todo{Might still consider the reqrate v.
-conc miss graph, because of showing the relation of performance going up as there is more concurrency, or else make sure to describe this clearly in the text}.
-\begin{figure}[h]
-	\begin{center}
-		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-maxrates-edited.pdf}
-	\end{center}
-	\caption{The max rates for processing list nodes with varying numbers of cores.
-Each core walks one list, and each node access results in a reference to main memory.
-All memory is local to socket0.
-Cores 1-6 are on socket0, and cores 7-12 are on socket1.}
-	\label{fig:listwalk-maxrates}
-\end{figure}
+\todo{might also include simple bar graph with max 1 socket, max 2 sockets, to membind=0}
 
@@
 	\begin{center}
-		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-outflats-edited.pdf}
+		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-totalconc-edited.pdf}
 	\end{center}
-	\caption{This is my caption about multi-listwak-outflats-edited, it shows the number of misses when rate plateaus...}
-	\label{fig:listwalk-outflats}
+	\caption{Throughput of listwalk versus total number of concurrent references offered by the cores.
+Data points for running 1 core are shown as circles, and data points for running 6 cores are shown as diamonds.
+Notice that the throughput for a single core levels out at 10 concurrent references.
+For 4 to 6 cores, the throughput levels off at around 36 references, which seems to be the most memory concurrency a socket can handle.}
+	\label{fig:listwalk-totalconc}
 \end{figure}
@@
 \todo{choose best phrase to use for outstanding/concurrent miss/refs/reads and use consistently unless there is a place one makes most sense}
-	 Figure~\ref{fig:listwalk-outflats} shows the total number of outstanding misses the system can handle before performance plateaus.
+	 Figure~\ref{fig:listwalk-totalconc} shows the total number of outstanding misses the system can handle before performance plateaus.
 We found that for a single core, bandwidth plateaus once there are 10 concurrent misses; this is because the L1 cache line-buffer, which keeps track of outstanding misses of the core, is size 10.
@@
 
-\begin{figure*}
-	\begin{center}
- 		\includegraphics[width=0.5\textwidth]{figures/multi-core-green-thread-1gb.pdf}
-	\end{center}
-	\caption{}
-	\label{fig:listwalk-green-threads}
-\end{figure*}
+%\begin{figure*}
+%	\begin{center}
+%		\includegraphics[width=0.5\textwidth]{figures/multi-core-green-thread-1gb.pdf}
+%	\end{center}
+%	\caption{}
+%	\label{fig:listwalk-green-threads}
+%\end{figure*}
 
@@
 Recent networks have suggested a max issue rate of 30M-90M messages per second is attainable \todo{cite mellanox press release and/or XT3/Eldorado} and this seems to be gradually increasing, so we use 100Mref/s as a good approximation to work with.
-Assuming a round-trip network latency of $1-2\mu s$, then by Little's Law, the number of outstanding references is $100Mref/s*1-2\mu s = 100-200$.
-\todo{Redo this calculation with proper numbers latency} This is the amount of memory concurrency that needs to exist for the latency of memory requests to be hidden when there is a rate of 100Mref/s.
-We consider two issues: first, can we support 100-200 misses?, and second, can the caches support enough contexts? 
-In order to keep the cores busy, there must be sufficient memory concurrency, and the number of contexts required to achieve this will vary.
+Assuming a round-trip network latency of 1 to 2$\mu s$, then by Little's Law, the number of outstanding references is $(100Mref/s) * $(2 to 3 $\mu s$) = 200 to 300.
+This is the amount of memory concurrency that needs to exist for the latency of memory requests to be hidden when there is a rate of 100Mref/s.
+We consider two issues: first, can we support 200-300 misses?, and second, can the caches support enough contexts? 
+\todo{how many contexts required for that much memory concurrency.
+Is it app dependent?} In order to keep the cores busy, there must be sufficient memory concurrency, and the number of contexts required to achieve this will vary.
 As there are more contexts in a core, there are more workings sets, including both data and coroutine stacks, that should fit in the cache to maintain performance.
 
-	Can we support 100-200 misses?
+	Can we support 200-300 misses?
 If the remote requests are reads that go to the global queue, then the number of outstanding references is limited to about 36.
@@
 If, however, prefetches are messages--communicated with memory stores--requesting a piece of remote memory then many more outstanding references could be supported, either with a software data structure or a memory management device implemented in an FPGA.
+	\begin{figure}[h]
+	\begin{center}
+ 		\includegraphics[width=0.5\textwidth]{figures/multi-green-delay7800-edited.pdf}
+	\end{center}
+	\caption{}
+	\label{fig:listwalk-green-delay}
+\end{figure}
 	We performed some experiments with green threads on listwalk, where we simulate the network latency by not allowing a coroutine to continue after it does a prefetch-and-switch until some specified amount of time has passed.

### 2011-01-24 doc/hotpar2011/evaluation.tex
reworked evaluation

@@
 \todo{might also include simple bar graph with max 1 socket, max 2 sockets, to membind=0}
-
-\begin{figure}[h]
-	\begin{center}
-		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-totalconc-edited.pdf}
-	\end{center}
-	\caption{Throughput of listwalk versus total number of concurrent references offered by the cores.
-Data points for running 1 core are shown as circles, and data points for running 6 cores are shown as diamonds.
-Notice that the throughput for a single core levels out at 10 concurrent references.
-For 4 to 6 cores, the throughput levels off at around 36 references, which seems to be the most memory concurrency a socket can handle.}
-	\label{fig:listwalk-totalconc}
-\end{figure}
 

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
reworked evaluation

@@
 \usepackage{url}
+\usepackage{fancyvrb}
+\usepackage{alltt}
 \begin{document}
@@
 \author{
-(authors)
+Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Luis Ceze\\
+University of Washington\\
+\email{\{nelson, bdmyers, ahh, djg, luisceze\}@cs.washington.edu}\\ \\
+Simon Kahan \\ 
+Pacific Northwest National Lab\\
+\email{\{skahan\}@cs.washington.edu}
 }
@@
 \begin{abstract}
+
+Graph analytics are very important.
+
+Irregular applications present a challenge for parallel machines:
+obtaining performance is not straightforward.
+
+Multithreading is a technique that has been successfully used to
+extract performance from these problems.
+
+Our goal is to build a system with XMT-like properties, implemented in
+software with commodity procesors
+
+In this paper, we describe a runtime for latency tolerance. Our
+analysis shows the feasibility of our approach. We end with our plans
+for adding synchronization support and scaling up to multiple nodes.
 
@@
 
-\section{Current Implementation and Results}
+\section{Evaluation}
 
-\todo{Begin eval insert}
-\input{evaluation}
-\todo{End eval insert}
+To evaluate the feasibility of our proposal, we ran pointer chasing
+benchmarks on a single-node implementation of our runtime. These
+pointer chasing experiments are intended to model a particular
+``wost-case'' behavior of irregular applications, where each memory
+reference causes a cache miss.
+
+We ran these experiments on a Dell PowerEdge R410 with two Xeon X5650
+processors and 24GB of RAM, with hyperthreading disabled. These
+processors use a NUMA memory architecture, where each processor has
+its own integrated memory controller and DIMMs; references to other
+processors are carried over Intel's cache-coherent QuickPath
+Interconnect (QPI).
+
+Our evaulation consists of three parts. First, we describe the
+performance of the test machine's memory system to use as a
+baseline. Then, we demonstrate that the same performance is achievable
+using our runtime system: we measure pointer chasing performance and
+the effects of cache pressure due to the number of contexts
+required. Finally, we simulate the effects of fetching memory over a
+network by adding delays to our runtime.
+
+\subsection{Memory system performance}
+
+We first characterize the performance of our test machine's memory
+system. We measured three parameters: the maximum random reference
+rate that can be issued by the cores in one processor, the maximum random
+reference rate that can be serviced by one processor's memory controller,
+and the maximum random reference rate that can be carried over the QPI
+link and serviced by another processor's memory controller.
+
+To find the maximum random reference rate that one processor's cores
+can issue, we ran pointer chasing code following the model of
+Figure~\ref{fig:pointernocoro}. Each core issues $n$ list traversals
+in a loop; we call $n * \textit{number of cores}$ the number of
+{\em concurrent offered references}, since the memory system may not
+be able to satisfy them all in parallel. Since our goal is to find a
+baseline for evaluating our coroutine library, we depend on the
+processor's ILP for memory concurrency, rather than our coroutine
+library.
+
+The lists were sized to exceed the last level cache and were laid out
+randomly with pointers spaced at a cache line granularity, maximizing
+the probability of each reference being a miss. We allocated the lists
+on 1GB huge pages to minimize TLB refill overhead. The lists were
+allocated in the memory attached to the processor containing the cores
+doing the traversal. We ran 10 traversals and recorded the maximum
+rate.
+
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-totalconc-edited.pdf}
+	\end{center}
+	\caption{Throughput of listwalk versus total number of
+          concurrent references offered by the cores. Data points for
+          running 1 core are shown as circles, and data points for
+          running 6 cores are shown as diamonds. 
+%Notice that the throughput for a single core levels out at 10
+%concurrent references. For 4 to 6 cores, the throughput levels off at
+%around 36 references, which seems to be the most memory concurrency a
+%processor can handle.
+        }
+	\label{fig:listwalk-totalconc}
+\end{figure}
+
+Figure~\ref{fig:listwalk-totalconc} shows the result. Each point represents
+the maximum rate pointers are traversed for a given number of
+concurrent offered references. We see a maximum rate of 277 million
+references per second; this is achieved when the number of offered
+references is 42. Note that a single core cannot support this level of
+memory concurrency; the maximum reference rate for a single core is
+107M. We believe this limit is due to the core's {\em line fill
+  buffer} \todo{cite nehalem perf docs}, which limits the core to 10
+concurrent L2 misses.
+
+The memory controller in the processor has more bandwidth than its
+cores can saturate. To measure the memory controller's maximum random
+reference rate, we extended the previous experiment so that cores in
+both processors were traversing lists allocated in the first processor's
+memory. With this configuration, we observed a maximum rate of 360
+million references per second. We believe this difference is due to
+the configuration of the processor's {\em Global Queue} \todo{cite nehalem
+  docs}, which limits the number of concurrently-executing read misses
+from a processor's cores to its local last-level cache and memory to 32.
+
+\begin{figure}[h]
+  \begin{center}
+    \includegraphics[width=0.5\textwidth]{figures/qpi_bw-edited.pdf}
+  \end{center}
+  \caption{Pointer chasing in a remote processor's memory.}
+  \label{fig:listwalk-qpi}
+\end{figure}
+
+Finally, we measured the rate at which the cores can make requests
+over the QPI link. For this experiment, we allocated the lists in the
+second processor, and ran the traversals on the first
+processor. Figure~\ref{listwalk-qpi} shows the result. We found that use
+of the QPI link limits us to 175 million references per second.
+
+\begin{figure}
+\begin{alltt}
+  while (count-- > 0) \{
+    list1 = list1->next;
+    list2 = list2->next;
+    \ldots
+    list\(n\) = list\(n\)->next;
+  \}
+\end{alltt}
+\caption{Pseudocode for pointer chasing without coroutines.}
+\label{fig:pointernocoro}
+\end{figure}
+
+\subsection{Runtime performance}
+
+To characterize the performance of our runtime, we investigated three
+effects: the maximum reference rate using coroutines
+to obtain memory concurrency, the effects of cache pressure from the
+coroutines' context storage, and the maximum reference rate obtainable
+using a simulated network delay.
+
+\begin{figure}
+\begin{alltt}
+  while (count-- > 0) \{
+     prefetch(&(list1->next));
+     prefetch(&(list2->next));
+     \ldots
+     prefetch(&(list\(n\)->next));
+     switch();
+     list1 = read(&(list1->next));
+     list2 = read(&(list2->next));
+     \ldots
+     list\(n\) = read(&(list\(n\)->next));
+ \}
+\end{alltt}
+\caption{Pseudocode for pointer chasing using coroutines.}
+\label{fig:pointercoro}
+\end{figure}
+
+To find the maximum reference rate obtainable using our coroutine
+library, we modified our pointer chasing benchmark as shown in
+figure~\ref{fig:pointercoro}. Now there are two sources of offered
+concurrent references: the ILP exploited by the processor, and the
+memory concurrency enabled by prefetching and switching to a new
+coroutine. As with our first experiment, we allocated the lists in the
+same processor as the cores doing the traversal.
+
+\begin{figure}[h]
+  \begin{center}
+    \includegraphics[width=0.5\textwidth]{figures/multi-green-edited.pdf}
+  \end{center}
+  \caption{Pointer chasing with coroutines, with one reference per
+    coroutine.}
+  \label{fig:multi-green}
+\end{figure}
+
+Figure~\ref{fig:multi-green} shows the result. For this
+experiment, we limited memory concurrency to one concurrent miss per
+coroutine, so the only source of memory concurrency is the use of
+coroutines. We are able to obtain a rate of 275 million references per
+second with 48 concurrent misses, or 8 coroutines per core. These
+results are very close to the ILP-only experiment.
+
+We observe a gradual decrease in reference rate once the number of
+concurrent references per core exceeds 10; we believe this is due to
+later prefetches squashing earlier ones in the line fill buffer. We
+also observe many samples below the maximum rate even with enough
+memory concurrency; this is harder to explain, but we suspect it is
+due to resource contention in the cores' pipelines once the memory
+system is full of requests.
+
+\begin{figure}
+\begin{alltt}
+  while (count-- > 0) \{
+    prefetch(&(list1->next));
+    switch();
+    list1 = read(&(list1->next));
+    for( i in 1 to num_local_updates ) \{
+      local->data++;
+      local = local->next;
+   \}
+  \}
+\end{alltt}
+\caption{Pseudocode for pointer chasing with local updates.}
+\label{pointerupdate}
+\end{figure}
+
+The stacks for the coroutines are stored in the data cache, where they
+will compete for space with an application's data. To characterize the
+effects of this cache pressure, we modified our list chasing benchmark
+to include random updates to a per-coroutine local data structure that
+is small enough to fit in cache. Figure~\ref{pointerupdate} shows the
+general idea.
+
+\begin{figure}[h]
+  \begin{center}
+    \includegraphics[width=0.5\textwidth]{figures/cache-pressure-edited.pdf}
+  \end{center}
+  \caption{Cache pressure with coroutines.}
+  \label{fig:cache-pressure}
+\end{figure}
+
+We ran this code on six cores with all data allocated in the same
+processor's memory. We varied the size of the local working set from 2KB
+to 4MB, and ran both 0 and 32 local updates for each pointer
+traversal. We measured the overall reference rate, including both the
+
+Figure~\ref{fig:cache-pressure} shows the result. On the left of the
+figure, we see the list-traversal-only case, where we allocate
+per-coroutine space, but make no references to it. As we would expect,
+the reference rates grow with the number of coroutines just as in
+Figure~\ref{fig:multi-green}. On the right of the figure, we see the
+case with 32 updates for each list traversal. While performance
+decreases for large working set sizes and large number of coroutines,
+we see speedup for up to 128KB working set size with 8 coroutines per
+core.
+
+Finally, to simulate the performance of pointer chasing in a
+multi-node system, we modified the scheduler of our coroutine library
+to include a delay before a coroutine can be rescheduled, imitating
+the network transit delay even though we are referencing local
+memory. Recently published measurements of Infiniband latencies
+\todo{cite} suggest we might expect 1.1 $\mu$s source-to-destination
+latencies between two computers connected back-to-back; we assume a
+switch transit delay of 400 ns in each direction and thus estimate a 3
+$\mu$s round-trip delay. In this experiment, once a coroutine has
+yielded, we use the core's timestamp counter to ensure it does not get
+rescheduled for 3 $\mu$s.
+
+\begin{figure}[h]
+  \begin{center}
+    \includegraphics[width=0.5\textwidth]{figures/multi-green-delay7800-edited.pdf}
+  \end{center}
+  \caption{Pointer chasing with simulated network delay.}
+  \label{fig:network-delay}
+\end{figure}
+
+We ran these experiments with two sources of memory concurrency: each
+core ran up to 250 coroutines, and each coroutine made between 1 and 4
+concurrent references. The results are shown in
+Figure~\ref{fig:network-delay}. We are still able to reach a
+near-maximum rate of 273 million references per second, but at least 3
+concurrent references per coroutine are required. Since a single
+reference per coroutine achieves 236 million references per
+second. Any network interface would be connected through a QPI link,
+which we measured to have a bandwidth limit of 175 million references
+per second, our runtime appears to have the capacity to chase pointers
+in a remote memory.
+
+\todo{may want to rerun with memory on remote core to incorporate QPI limit?}
+
+%\todo{Begin eval insert}
+%\input{evaluation}
+%\todo{End eval insert}
 

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
approach changes and comments

--- 2011-01-21 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-And these graphs can be large---too large for a single node \todo{may want to avoid saying 'node' since graph 'node' is also used in this paragraph}.
-In January 2011, Facebook had 500 million active users with an average of
+And these graphs can be large---too large for a single node. In January 2011, Facebook had 500 million active users with an average of
 130 friends each \cite{Facebook:2011p91}. A natural approach would be
 to try to partition the graph across the nodes in a cluster. But with
-low diameter graphs, a traversal starting at any node is likely to
+low diameter graphs, a traversal starting at any vertex is likely to
 quickly leave its partition and travel over the network, incurring
-high latency costs. And power law graphs have a small number of nodes
-that generate significantly more work than other nodes, leading to
+high latency costs. And power law graphs have a small number of vertices
+that generate significantly more work than other vertices, leading to
 computational imbalance.
@@
 
-Each coroutine executes non-preemptively, utilizing the core to which
+Each coroutine executes non-preemptively \todo{seems like non-premptively isn't ``how" it executes, rather how it is scheduled, maybe say ``cooperatively"?}, utilizing the core to which
 its ``parent'' thread has been assigned by the operating system until it
@@
 
-Our goal is to enable high throughput on graphs with two challenging
+Our goal is to enable high throughput on graphs having two challenging
 properties. First, the graphs' combination of sparseness and low
@@
 graphs are larger than a single node's memory and do not partition
-well \todo{explain?}. We must spread the graph across multiple nodes
+well because of low diameter \todo{enough explanation now?}. We must spread the graph across multiple nodes
 and send memory operations over a network. If memory requests are
 uniformly distributed over the working set, a given node will make as
-many accesses to a distant node's memory as it does to the local node.
+many accesses to remote nodes' memory as there are remote requests to the local node's memory coming in.
 
@@
 \todo{cite}. Little's law tells us then that we must support on the
-order of 300 concurrent memory operations, and thus coroutines, to
+order of 300 concurrent memory operations--and thus, up to this many coroutines--to
 cover this latency.
@@
 components.
+First, the system needs lightweight context switches (coroutines) to be able to support many contexts--thereby increasing memory concurrency available--with minimal impact on performance.
+Second, a memory management subsystem must support many concurrent accesses to a global address space.
+Third, the system must provide synchronization at a fine-granularity \todo{this last one sounds incomplete because no reason given}.
 
-\subsection{Low-latency context switch}
+\subsection{Lightweight context switch}
+\todo{can we include in the title something that means lightweight and good scheduling?}
 
@@
 be fast, so that overhead does not swamp actual work. Third, when a
-coroutine yields, we must choose a good coroutine to run next.
+coroutine yields, we must choose a suitable coroutine to run next.
 
@@
 round-robin fashion. But this may lead to performance problems: a
-coroutine might be scheduled before the event which caused it to yield
+coroutine might be scheduled before the event which caused it to yield (e.g., a long latency memory access)
 has completed, stalling the pipeline and keeping other coroutines from
@@
 
+\subsubsection{Software management}
 In order to minimize memory access overhead, the memory operation
@@
 compute threads to a single memory manager thread may reduce memory
-concurrency. And we've increased the overhead of a single memory
-operation---what in the single node case was a simple read, is now
-multiple queuing operations between threads, and multiple reads and
+concurrency. And we have increased the overhead of a single memory
+operation: what in the single node case was a simple read, is now
+multiple queuing operations between threads and multiple reads and
 writes over QPI to the network interface. Adding more memory manager
-threads may enable more memory concurrency, but the overhead is still high.
+threads may enable more memory concurrency, but the overhead is still high \todo{why?
+because lots of code to run?
+so accel by impl in hardware.
+Maybe this is answered by the "what in...is now" already}
 
-Instead, we propose hardware acceleration with a {\em memory manager
+\subsubsection{Hardware management}
+\todo{might want to call it accelerator throughout after saying FPGA once}
+Considering this overhead, we propose hardware acceleration with a {\em memory manager
   FPGA}, a coprocessor FPGA inserted into a Xeon socket so it can
@@
 tie up one of these slots for the entire multi-microsecond latency of
-the request. We must bypass these structures to allow the amount of
-concurrency we need; the CPU needs to direct the memory manager FPGA
+the request. To allow the amount of concurrency we need, we must bypass these structures; the CPU needs to direct the memory manager FPGA
 to do work on its behalf without tying up its own resources.
@@
 remote memory addresses. The memory manager FPGA advertises each
-remote address twice. The first address is used for blocking reads and
+remote address twice \todo{this sounded like an active thing the FPGA is doing each time you make remote request, first time I read it.
+Consider just sticking to cmd as upper order address bits than saying advertise each address twice}.
+The first address is used for blocking reads and
 writes---any access incurs the full latency of the remote
@@
 executes the blocking read, and the memory manager immediately returns
-the value from the buffer. \todo{this isn't fully baked---do we want
+the value from the buffer \todo{is this necessarily restricted to the memory manager returning value from buffer rather than buffer in local memory?
+maybe it is, since if buffer was in local mem, would have to also write a full bit or make two reads: valid bit and data}.
+\todo{this isn't fully baked---do we want
   to include this sort of detail?}
@@
 synchronization primitives yet, full-empty bits are a natural choice
-given our programming model. Two questions then arise: at what
+given our programming model \todo{say why natural choice unless programming model section makes this point already}. Two questions then arise: at what
 granularity should full-empty bits work, and what component should do
@@
 
+% 2TB global memory => 68GB of distributed FE bits.  This seems reasonable, given that a 2TB system might have 64 nodes, so ~1GB per memory manager if one manager per node.
+
 \section{Current Implementation and Results}
-
-\todo{Begin eval insert}
 \input{evaluation}
-\todo{End eval insert}
 
@@
 why not wait for more cores?
+- perhaps memory subsystems on a chip are not keeping up with the bandwidth of so many cores.
 

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex

--- 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 stored in the CPU's memory. \todo{discuss QPI bandwidth differences}
+
+% 2TB global memory => 68GB of distributed FE bits.  This seems reasonable, given that a 2TB system might have 64 nodes, so ~1GB per memory manager if one manager per node.
+
+\section{Current Implementation and Results}
+\input{evaluation}
 

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
reviewed evaluation draft in paper

@@
 %% To disable, just uncomment this line
-%% \renewcommand{\todo}[1]{\relax}
+ %%\renewcommand{\todo}[1]{\relax}
 
@@
 
-\section{Current Implementation and Results}
-\input{evaluation}
-
 \section{Evaluation}
@@
 
+\begin{figure}
+\begin{alltt}
+  while (count-- > 0) \{
+    list1 = list1->next;
+    list2 = list2->next;
+    \ldots
+    list\(n\) = list\(n\)->next;
+  \}
+\end{alltt}
+\caption{Pseudocode for pointer chasing without coroutines.}
+\label{fig:pointernocoro}
+\end{figure}
+
 The lists were sized to exceed the last level cache and were laid out
@@
 allocated in the memory attached to the processor containing the cores
-doing the traversal. We ran 10 traversals and recorded the maximum
-rate.
+doing the traversal. We ran 10 traversals, measuring average rate in each, and recorded the maximum rate.
 
@@
 references per second; this is achieved when the number of offered
-references is 42. Note that a single core cannot support this level of
+references is 42 \todo{yes, max is achieved here but essentially levels out earlier like 36; do we want a percentage metric for defining a plateau}.
+Note that a single core cannot support this level of
 memory concurrency; the maximum reference rate for a single core is
@@
 
+% take this figure out if we need the space
+% the main point is room for remote references to local memory
 \begin{figure}[h]
@@
 second processor, and ran the traversals on the first
-processor. Figure~\ref{listwalk-qpi} shows the result. We found that use
+processor. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use
 of the QPI link limits us to 175 million references per second.
-
-\begin{figure}
-\begin{alltt}
-  while (count-- > 0) \{
-    list1 = list1->next;
-    list2 = list2->next;
-    \ldots
-    list\(n\) = list\(n\)->next;
-  \}
-\end{alltt}
-\caption{Pseudocode for pointer chasing without coroutines.}
-\label{fig:pointernocoro}
-\end{figure}
 
@@
 later prefetches squashing earlier ones in the line fill buffer. We
-also observe many samples below the maximum rate even with enough
+also observe some repeatable outliers below the maximum rate even with enough
 memory concurrency; this is harder to explain, but we suspect it is
@@
 to 4MB, and ran both 0 and 32 local updates for each pointer
-traversal. We measured the overall reference rate, including both the
+traversal. We measured the overall reference rate, including both the local updates and remote references.
 
@@
 figure, we see the list-traversal-only case, where we allocate
-per-coroutine space, but make no references to it. As we would expect,
+per-coroutine space but make no references to it. As we would expect,
 the reference rates grow with the number of coroutines just as in
@@
 decreases for large working set sizes and large number of coroutines,
-we see speedup for up to 128KB working set size with 8 coroutines per
+we see speedup as number coroutines increases for up to 128KB working set size with 8 coroutines per
 core.
@@
 We ran these experiments with two sources of memory concurrency: each
-core ran up to 250 coroutines, and each coroutine made between 1 and 4
+core ran up to 167 coroutines, and each coroutine made between 1 and 4
 concurrent references. The results are shown in
@@
 near-maximum rate of 273 million references per second, but at least 3
-concurrent references per coroutine are required. Since a single
+concurrent references per coroutine are required. Running with a single
 reference per coroutine achieves 236 million references per
@@
 which we measured to have a bandwidth limit of 175 million references
-per second, our runtime appears to have the capacity to chase pointers
+per second, so our runtime appears to have the capacity to chase pointers
 in a remote memory.
 
+\todo{do we want to include that the memory system imposes max limit so really more throughput is still achievable if there are enough ILP memory concurrency?}
 \todo{may want to rerun with memory on remote core to incorporate QPI limit?}

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
abstract, intro

--- 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-Graph analytics are very important.
+ Techniques from graph analytics have found important applications in
+  areas such as social networks and bioinformatics. These irregular
+  problems present a challenge for parallel machines: obtaining
+  performance is not straightforward. Multithreading works well for
+  such problems, but large-scale implementations of the technique like
+  the Cray XMT are expensive, non-commodity machines.
 
-Irregular applications present a challenge for parallel machines:
-obtaining performance is not straightforward.
-
-Multithreading is a technique that has been successfully used to
-extract performance from these problems.
-
-Our goal is to build a system with XMT-like properties, implemented in
-software with commodity procesors
-
-In this paper, we describe a runtime for latency tolerance. Our
-analysis shows the feasibility of our approach. We end with our plans
-for adding synchronization support and scaling up to multiple nodes.
-
-  Large irregular applications such as sparse, low diameter graph problems often
-  perform badly on commodity clusters. These application make many
-  non-sequential accesses to global data structures with low degrees of
-  locality, interacting badly with commodity processors' memory
-  systems, which are optimized for local, sequential memory
-  accesses. Special-purpose hardware optimized for these irregular
-  applications has been built; the Cray XMT is one such machine. It
-  uses multithreading to tolerate the latency of low-locality memory
-  accesses, and while it exhibits excellent performance and energy
-  efficiency for irregular applications, it performs badly on standard
-  sequential applications \todo{Brandon: should this be more specific or worded differently?
-Its true sequential perform badly since switching between many threads, but also does XMT also perform  poorly (relative to alternatives) on highly parallel but regular apps with lots of temporal and spatial locality?}.
-We believe we can have the best of both
-  worlds: we argue that modern commodity processors have many of the
-  features necessary \todo{Brandon: we will have to be able to list these as features in the paper or use a different phrase; my feeling is that these are aspects of the system which are not optimized for this but are sufficient to suggest potential for enough memory concurrency} to give good performance on irregular problems,
-  and that a latency-tolerant runtime designed to exploit these
-  features can enable speedup for irregular applications on commodity
-  processors. In this paper, we describe the key features of such a
-  runtime system, and what a cluster built using our runtime might
-  look like. Using a preliminary single-node implementation of our
-  runtime, we obtained speedups on simple list chasing benchmarks. Our results show promise for a multi-node implementation.
-
+  Our goal is to build a system with XMT-like properties, implemented
+  with commodity processors. This paper presents a runtime for latency
+  tolerance, along with our plans for scaling up to multiple
+  nodes. Our analysis shows the feasibility of our approach.
 \end{abstract}
@@
 
-Graphs have proven to be a useful abstraction for solving real-world
-problems in social and biological network analysis. But many graphs in
-social and biological network analysis are sparse, with low diameter
-and a degree distribution that follows a power law. These properties
-make it difficult to take advantage of the locality assumptions built
-into the memory hierarchy of modern processors, so that even simple
-traversals are dominated by memory latency.
+Techniques from graph analytics have found important applications in
+areas such as social networks and bioinformatics. \todo{more!}
 
-And these graphs can be large---too large for a single node. In January 2011, Facebook had 500 million active users with an average of
-130 friends each \cite{Facebook:2011p91}. A natural approach would be
-to try to partition the graph across the nodes in a cluster. But with
-low diameter graphs, a traversal starting at any vertex is likely to
-quickly leave its partition and travel over the network, incurring
-high latency costs. And power law graphs have a small number of vertices
-that generate significantly more work than other vertices, leading to
-computational imbalance.
-
+% In January 2011, Facebook had 500 million active users with an average
+% of 130 friends each \cite{Facebook:2011p91}.
 % Facebook's friend suggestion
@@
 
+The most interesting computational challenge comes from large,
+low-diameter, power law graphs: this combination makes extracting
+performance difficult. Their size keeps them from fitting in a single
+commodity machine's memory. Their low diameter means that they are
+difficult to lay out with locality. Their power law distribution means
+that a few vertices generate much more work than the rest, leading to
+computational imbalance and even less locality. Together, these
+properties lead to a difficult conclusion: each edge traversal is
+likely to be a very expensive cache miss.
+
 Multithreading is a technique that has been used successfully to
 implement efficient computations for these graphs. The Cray XMT is an
-example of such a solution: it tries to solve the memory latency
-problem through concurrency rather than caching. Each XMT processor
-supports 128 hardware contexts and $\sim$1000 outstanding memory
-operations and is able to switch contexts every cycle. But this
-approach has a limitation: each context is significantly slower than a
-modern commodity processor, yielding bad single-threaded
-performance. While the XMT is fast at sparse, low-degree graph
-problems, it is slow at other scientific codes that can take advantage
-of commodity processors' memory hierarchies. Thus, few XMTs have been
-built.
+example of such an approach: it solves the memory latency problem
+through concurrency rather than caching. Each XMT processor supports
+128 hardware contexts and 1024 outstanding memory operations, and is
+able to switch contexts every cycle. This ability comes at a cost: the
+XMT is an expensive, non-commodity machine.
 
-We want the best of both worlds---we want to build a system that has
-good performance on both these low-locality graph codes as well as on
-common scientific codes. Our approach is inspired by the XMT, but
-implemented in software on commodity microprocessors. The core of our
-approach is a lightweight coroutine library designed to overlap
-long-latency memory prefetches with other contexts' computation.
+We want the best of both worlds. Our goal is to build a system that
+has good performance on these low-locality graph codes but is
+implemented with commodity processors. The core of our approach is a
+lightweight coroutine library designed to overlap long-latency memory
+prefetches with other contexts' computation.
 
-\todo{something like: project still in early stages. this document lays out
-  problems we think a system like this needs to solve, and describes
-  some implementation strategies we want to explore. we have a basic
-  initial implementation with some experiments that suggest there is
-  promise here.}
+In this paper, we describe our single-node runtime for latency
+tolerance, along with our plans for scaling to multiple nodes.  We
+evaluate the performance of our runtime using a set of list traversal
+benchmarks designed to model certain worst-case behaviors of these
+large graph problems.
 
-\todo{do we want to mention hardware at this point? If so, we should
-  perhaps call it a hybrid approach.}
+The rest of the paper is organized as follows. We present related work
+in section~\ref{sec:related}. We describe the programming model we are
+implementing in section~\ref{sec:model}, and describe our approach for
+implementing this model in section~\ref{sec:approach}. We evaluate our
+latency-tolerant runtime in section~\ref{sec:evaluation} and conclude
+in section~\ref{sec:conclusion}.
+
+\section{Related work}
+\label{sec:related}
+
+\todo{actually write this}
+
+Multithreading for latency tolerance has been explored in a number of directions.
+
+Hardware approaches: The Tera MTA and Cray XMT, the MIT Alewife
+project, ??? use hardware context switches to tolerate memory latency
+for large systems. More recently, Simultaneous Multi-Threading,
+Niagara, Cyclops, and even GPUs switch between multiple contexts to
+tolerate memory latency, for memory that is hundreds of cycles of
+away, rather than the thousands we must tolerate.
+
+software:
+
+Mowry's SCM
+
+Cullers' TAM
+
+qthreads
+
+cilk? tbb?
+
+why not wait for more cores?
 
 \section{Programming model}
+\label{sec:model}
+
+\todo{Change to discuss API:
+
+context switches:
+  yield()
+  discuss coroutines
+
+mem:
+  alloc()
+  read(p) 
+  write(p, d)
+  prefetch(p)
+
+sync:
+ {read, write} {EE, EF, FE, FF}
+}
 
@@
 
-Each coroutine executes non-preemptively \todo{seems like non-premptively isn't ``how" it executes, rather how it is scheduled, maybe say ``cooperatively"?}, utilizing the core to which
+Each coroutine executes non-preemptively, utilizing the core to which
 its ``parent'' thread has been assigned by the operating system until it
@@
 \section{Approach}
+\label{sec:approach}
 
-Our goal is to enable high throughput on graphs having two challenging
+Our goal is to enable high throughput on graphs with two challenging
 properties. First, the graphs' combination of sparseness and low
@@
 graphs are larger than a single node's memory and do not partition
-well because of low diameter \todo{enough explanation now?}. We must spread the graph across multiple nodes
+well \todo{explain?}. We must spread the graph across multiple nodes
 and send memory operations over a network. If memory requests are
 uniformly distributed over the working set, a given node will make as
-many accesses to remote nodes' memory as there are remote requests to the local node's memory coming in.
+many accesses to a distant node's memory as it does to the local node.
 
@@
 \todo{cite}. Little's law tells us then that we must support on the
-order of 300 concurrent memory operations--and thus, up to this many coroutines--to
+order of 300 concurrent memory operations, and thus coroutines, to
 cover this latency.
@@
 components.
-First, the system needs lightweight context switches (coroutines) to be able to support many contexts--thereby increasing memory concurrency available--with minimal impact on performance.
-Second, a memory management subsystem must support many concurrent accesses to a global address space.
-Third, the system must provide synchronization at a fine-granularity \todo{this last one sounds incomplete because no reason given}.
 
-\subsection{Lightweight context switch}
-\todo{can we include in the title something that means lightweight and good scheduling?}
+\subsection{Low-latency context switch}
 
@@
 be fast, so that overhead does not swamp actual work. Third, when a
-coroutine yields, we must choose a suitable coroutine to run next.
+coroutine yields, we must choose a good coroutine to run next.
 
@@
 round-robin fashion. But this may lead to performance problems: a
-coroutine might be scheduled before the event which caused it to yield (e.g., a long latency memory access)
+coroutine might be scheduled before the event which caused it to yield
 has completed, stalling the pipeline and keeping other coroutines from
@@
 
-\subsubsection{Software management}
 In order to minimize memory access overhead, the memory operation
@@
 compute threads to a single memory manager thread may reduce memory
-concurrency. And we have increased the overhead of a single memory
-operation: what in the single node case was a simple read, is now
-multiple queuing operations between threads and multiple reads and
+concurrency. And we've increased the overhead of a single memory
+operation---what in the single node case was a simple read, is now
+multiple queuing operations between threads, and multiple reads and
 writes over QPI to the network interface. Adding more memory manager
-threads may enable more memory concurrency, but the overhead is still high \todo{why?
-because lots of code to run?
-so accel by impl in hardware.
-Maybe this is answered by the "what in...is now" already}
+threads may enable more memory concurrency, but the overhead is still high.
 
-\subsubsection{Hardware management}
-\todo{might want to call it accelerator throughout after saying FPGA once}
-Considering this overhead, we propose hardware acceleration with a {\em memory manager
+Instead, we propose hardware acceleration with a {\em memory manager
   FPGA}, a coprocessor FPGA inserted into a Xeon socket so it can
@@
 tie up one of these slots for the entire multi-microsecond latency of
-the request. To allow the amount of concurrency we need, we must bypass these structures; the CPU needs to direct the memory manager FPGA
+the request. We must bypass these structures to allow the amount of
+concurrency we need; the CPU needs to direct the memory manager FPGA
 to do work on its behalf without tying up its own resources.
@@
 remote memory addresses. The memory manager FPGA advertises each
-remote address twice \todo{this sounded like an active thing the FPGA is doing each time you make remote request, first time I read it.
-Consider just sticking to cmd as upper order address bits than saying advertise each address twice}.
-The first address is used for blocking reads and
+remote address twice. The first address is used for blocking reads and
 writes---any access incurs the full latency of the remote
@@
 executes the blocking read, and the memory manager immediately returns
-the value from the buffer \todo{is this necessarily restricted to the memory manager returning value from buffer rather than buffer in local memory?
-maybe it is, since if buffer was in local mem, would have to also write a full bit or make two reads: valid bit and data}.
-\todo{this isn't fully baked---do we want
+the value from the buffer. \todo{this isn't fully baked---do we want
   to include this sort of detail?}
@@
 synchronization primitives yet, full-empty bits are a natural choice
-given our programming model \todo{say why natural choice unless programming model section makes this point already}. Two questions then arise: at what
+given our programming model. Two questions then arise: at what
 granularity should full-empty bits work, and what component should do
@@
 
-% 2TB global memory => 68GB of distributed FE bits.  This seems reasonable, given that a 2TB system might have 64 nodes, so ~1GB per memory manager if one manager per node.
-
-\section{Current Implementation and Results}
-\input{evaluation}
-
 \section{Evaluation}
+\label{sec:evaluation}
 
@@
 
-\section{Related work}
-
-hardware:
-
-XMT/MTA
-
-cyclops
-
-niagra
-
-software:
-
-qthreads
-
-cilk? tbb?
-
-why not wait for more cores?
-- perhaps memory subsystems on a chip are not keeping up with the bandwidth of so many cores.
-
-hyperthreading?
-
 \section{Conclusion}
-
-Large irregular problems found in areas such as social networking and
-bioinformatics tend to have poor performance on commodity clusters due
-to frequent low-locality memory accesses. We present a programming
-model and runtime system that uses lightweight context switching to
-tolerate long-latency accesses and increase performance on these
-applications. Our preliminary evaluation shows that the overhead of
-context switching is not prohibitive for thread counts needed for a
-large-scale implementation, but memory concurrency is limited by
-hardware structures in commodity processors. We describe how we plan
-to address the challenges of large-scale memory concurrency and
-lightweight synchronization.
+\label{sec:conclusion}
 

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex

--- 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \section{Approach}
+\label{sec:approach}
 
@@
 \section{Evaluation}
+\label{sec:evaluation}
 

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
citations for mandal, qpi

--- 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 (QPI) allows approximately 100 million requests per second
-\todo{cite}, and recent interconnect products allow similar request
+\cite{quickpath:website} \todo{specifically 25.6GB/s two way, which we expect 6.4GB/s for one way for outgoing request}, and recent interconnect products allow similar request
 rates, with a round-trip latency of approximately 3$\mu$s
@@
 concurrent offered references. We see a maximum rate of 277 million
-references per second; this is achieved when the number of offered
+references per second, which agrees with the measurements found by \cite{Mandal:2010} with a similar machine. This rate is achieved when the number of offered
 references is 42 \todo{yes, max is achieved here but essentially levels out earlier like 36; do we want a percentage metric for defining a plateau}.

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
hardware related work

@@
 \label{sec:related}
+The concept of using independent parallel contexts to tolerate latency
+is not novel.  The inspiration for our project is of course the Cray
+XMT \cite{feo-xmt}, which implemented a hardware version of our
+programming model (in what would today be considered exceptionally
+slow silicon.)  The XMT can be seen as simultaneous multithreading
+\cite{tullsen-smt} writ large; the technique and motivation is
+essentially the same, but the XMT allows hundreds of contexts (modern
+Intel SMT implementations are generally limited to two.) The Alewife
+system \cite{agarwal-alewife} also switches contexts (in hardware) to
+hide latency on remote-node accesses.
+
+Modern programmable GPUs can also be interpreted in using an XMT-like
+model under the hood;
+% Do I need to include a citation here?
+while they generate large numbers of available execution contexts via
+a (semi-)explicit SIMD model, those SIMD instructions keep functional
+units satisfied by overlapping execution with the loads required for
+other contexts.
+
+Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
+  multithreading}.  As opposed to our system of explicitly marked
+remote references, SCM assumes a high hit rate and traps to an
+interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
+short enough latencies that using more than two contexts generally
+produced slowdown.
+ b. TAM
+ c. qthreads
+Of course, our system is proceeded by the hardware
+\emph{software controlled multithreading},
 
 \todo{actually write this}
-
-Multithreading for latency tolerance has been explored in a number of directions.
-
-Hardware approaches: The Tera MTA and Cray XMT, the MIT Alewife
-project, ??? use hardware context switches to tolerate memory latency
-for large systems. More recently, Simultaneous Multi-Threading,
-Niagara, Cyclops, and even GPUs switch between multiple contexts to
-tolerate memory latency, for memory that is hundreds of cycles of
-away, rather than the thousands we must tolerate.
-
-software:
-
-Mowry's SCM
-
-Cullers' TAM
-
-qthreads
-
-cilk? tbb?
-
-why not wait for more cores?
 

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
more related work

@@
 system \cite{agarwal-alewife} also switches contexts (in hardware) to
-hide latency on remote-node accesses.
+hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
+allows hundreds of independently executing contexts to share FPUs,
+caches, and paths to memory, but does not context switch or share
+basic execution hardware.
 
-Modern programmable GPUs can also be interpreted in using an XMT-like
+Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
 model under the hood;
@@
 produced slowdown.
- b. TAM
- c. qthreads
-Of course, our system is proceeded by the hardware
-\emph{software controlled multithreading},
 
-\todo{actually write this}
+Partitioned global address space models, such as that used in Unified Parallel C,
+% definitely need a citation here
+ share our approach of exposing all data (even remote data)
+ transparently to any threads and have largely solved the problems of
+ how to efficiently translate those remote reads into message
+ passing.  However, their programming models emphasize locality more
+ than our work does; while a program can examine any vertex in a UPC
+ program at any time, it probably shouldn't (whereas our kernels rely on our
+ latency hiding technique to make common remote references fast.
 
+The value of the information in extremely large and sparse graphs is
+rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
+sophisticated analysis on data from Twitter, identifying important
+actors and their networks of influence, and it can scale to the full
+size of graphs such as Facebook's friend network.
 \section{Programming model}

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   softxmt-hotpar2011.tex

--- 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \usepackage{alltt}
+\usepackage{comment}
 \begin{document}
@@
 
+Our programming model partitions the address space into a shared
+global shared address space and per node private address spaces.
+Locality may be exploited directly by the programmer in the node
+private address spaces, just as in a conventional cache-coherent
+multiprocessor node.  In contrast, the global space is explicitly
+uncached.  There, every processor can access any word of memory
+directly via its unique global address.
+
+Whereas in the PGAS model coherency of global copies is implictly
+maintained across the system, we forego caching of globally shared
+data.  A node can exploit locality within a chunk of data stored in
+the global space only by first copying it into the node's local space.
+In this way, a node can access data at random across the system while
+incurring no hardware coherence overhead: consistency is enforced by
+our mechanisms for explicit global synchronization.
+
+The model promises efficiency subject to the presence of sufficient
+concurrency.  Exactly how that concurrency is expressed by the
+programmer is language dependent.  Our programming model serves to
+augment threaded programming languages so that they become latency
+tolerant on large address spaces.  Software multithreading is used to
+provide latency tolerance: a computation that is about to execute a
+long latency operation such as a remote memory reference initiates the
+operation and then yields to the scheduler, which quickly resumes
+execution of another computation.  Concurrency is also used -- via
+aggregation of requests to a common node destination -- to make more
+efficient use of network bandwidth and in some cases reduce bandwidth
+demands.
+
+Synchronization on locations in the node private address space is the
+same as on conventional systems.  Synchronization on global addresses
+works differently: we provide atomic operations such as {\tt
+  int\_fetch\_add} as well as operations using full-bits, as on the
+Cray XMT.  As is true for other long latency operations,
+synchronization latency is tolerated by yielding to the scheduler.
+Even non-blocking global atomics yield: atomic operations executed as
+x86 instructions sabbotage concurrency by forcing all outstanding
+memory references to complete before new ones can be issued.  By
+yielding, the core can switch to another computation while the
+synchronization is accomplished outside of the pipeline.  In addition,
+the implementation can exploit aggregation when concurrency is
+abundant to increase the throughput of global synchronization.
+
+Thus, the programming model provides the programmer with options: express
+locality within the private space, and it will be exploited to reduce
+latency and bandwidth; express concurrency, and it will be used toward
+tolerating latency and reducing the cost of bandwidth.
+
+\begin {comment}
 \todo{Change to discuss API:
@@
 as the main thread, as in the common  fork-join model of parallelism.
+
+\end{comment}
 

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   softxmt-hotpar2011.tex

@@
 
-Our programming model partitions the address space into a shared
-global shared address space and per node private address spaces.
-Locality may be exploited directly by the programmer in the node
-private address spaces, just as in a conventional cache-coherent
+Our shared-memory programming model partitions the overall address
+space into a global shared address space and per node private address
+spaces.  Locality may be exploited directly by the programmer in the
+node private address spaces, just as in a conventional cache-coherent
 multiprocessor node.  In contrast, the global space is explicitly
@@
 maintained across the system, we forego caching of globally shared
-data.  A node can exploit locality within a chunk of data stored in
+data.  A node can exploit temporal locality of data stored in
 the global space only by first copying it into the node's local space.

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   softxmt-hotpar2011.tex
system overview and intro; additional authors

--- 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \usepackage{alltt}
-\usepackage{comment}
 \begin{document}
@@
 \author{
-Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Luis Ceze\\
+Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
+Carl Ebeling, Luis Ceze\\
 University of Washington\\
-\email{\{nelson, bdmyers, ahh, djg, luisceze\}@cs.washington.edu}\\ \\
+\email{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}\\ \\
 Simon Kahan \\ 
@@
 \email{\{skahan\}@cs.washington.edu}
+Preston Briggs\\ 
+Cray Inc.\\
+\email{\{preston.briggs\}@gmail.com}
 }
@@
 
+\begin{figure*}[htbp]
+  \begin{center}
+    \includegraphics[width=0.9\textwidth]{figures/system-overview.pdf}
+	\end{center}
+	\caption{Overview of our proposed system design. The base
+          system is a cluster of commodity nodes and interconnect. We
+          add a latency-tolerant runtime and a global memory manager
+          to obtain efficient access to the global address space. Only
+          the shaded components may require hardware design.}
+	\label{fig:system-overview}
+\end{figure*}
+
 We want the best of both worlds. Our goal is to build a system that
 has good performance on these low-locality graph codes but is
-implemented with commodity processors. The core of our approach is a
-lightweight coroutine library designed to overlap long-latency memory
-prefetches with other contexts' computation.
+implemented with cheap commodity processors.
 
-In this paper, we describe our single-node runtime for latency
-tolerance, along with our plans for scaling to multiple nodes.  We
-evaluate the performance of our runtime using a set of list traversal
-benchmarks designed to model certain worst-case behaviors of these
-large graph problems.
+Figure~\ref{fig:system-overview} is a schematic of such a
+system. It is composed of multiple nodes built with commodity
+processors, communicating over an Infiniband network. We add two
+components: a runtime system, responsible for executing the user's threads,
+and a global memory manager, responsible for facilitating memory
+operations to the non-coherent global memory shared across the nodes.
+This memory manager is the only component that might require hardware,
+but we are exploring a mix of hardware and software implementations.
 
-The rest of the paper is organized as follows. We present related work
-in section~\ref{sec:related}. We describe the programming model we are
-implementing in section~\ref{sec:model}, and describe our approach for
-implementing this model in section~\ref{sec:approach}. We evaluate our
-latency-tolerant runtime in section~\ref{sec:evaluation} and conclude
-in section~\ref{sec:conclusion}.
+This paper explores the feasibility of our multi-node goal using a
+single-node implementation of our lightweight threading runtime. We
+use prefetch instructions together with lightweight threads to
+tolerate the latency of DRAM access on the local node node. We use
+pointer chasing benchmarks to verify our runtime's overhead is
+acceptable. We also model the effects of pointer chasing on a remote
+node.
+
+The rest of the paper is organized as follows. We describe our
+proposal from a programmer's viewpoint in section~\ref{sec:model}. We
+describe the implementation of our latency-tolerant runtime and plans
+for extending to multiple nodes in section~\ref{sec:approach}. We
+evaluate our runtime in section~\ref{sec:evaluation}.  We present
+related work in section~\ref{sec:related} and conclude in
+section~\ref{sec:conclusion}.
 
@@
 
-Our programming model partitions the address space into a shared
-global shared address space and per node private address spaces.
-Locality may be exploited directly by the programmer in the node
-private address spaces, just as in a conventional cache-coherent
-multiprocessor node.  In contrast, the global space is explicitly
-uncached.  There, every processor can access any word of memory
-directly via its unique global address.
-
-Whereas in the PGAS model coherency of global copies is implictly
-maintained across the system, we forego caching of globally shared
-data.  A node can exploit locality within a chunk of data stored in
-the global space only by first copying it into the node's local space.
-In this way, a node can access data at random across the system while
-incurring no hardware coherence overhead: consistency is enforced by
-our mechanisms for explicit global synchronization.
-
-The model promises efficiency subject to the presence of sufficient
-concurrency.  Exactly how that concurrency is expressed by the
-programmer is language dependent.  Our programming model serves to
-augment threaded programming languages so that they become latency
-tolerant on large address spaces.  Software multithreading is used to
-provide latency tolerance: a computation that is about to execute a
-long latency operation such as a remote memory reference initiates the
-operation and then yields to the scheduler, which quickly resumes
-execution of another computation.  Concurrency is also used -- via
-aggregation of requests to a common node destination -- to make more
-efficient use of network bandwidth and in some cases reduce bandwidth
-demands.
-
-Synchronization on locations in the node private address space is the
-same as on conventional systems.  Synchronization on global addresses
-works differently: we provide atomic operations such as {\tt
-  int\_fetch\_add} as well as operations using full-bits, as on the
-Cray XMT.  As is true for other long latency operations,
-synchronization latency is tolerated by yielding to the scheduler.
-Even non-blocking global atomics yield: atomic operations executed as
-x86 instructions sabbotage concurrency by forcing all outstanding
-memory references to complete before new ones can be issued.  By
-yielding, the core can switch to another computation while the
-synchronization is accomplished outside of the pipeline.  In addition,
-the implementation can exploit aggregation when concurrency is
-abundant to increase the throughput of global synchronization.
-
-Thus, the programming model provides the programmer with options: express
-locality within the private space, and it will be exploited to reduce
-latency and bandwidth; express concurrency, and it will be used toward
-tolerating latency and reducing the cost of bandwidth.
-
-\begin {comment}
 \todo{Change to discuss API:
@@
 as the main thread, as in the common  fork-join model of parallelism.
-
-\end{comment}
 

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
system overview and intro; additional authors

@@
 Pacific Northwest National Lab\\
-\email{\{skahan\}@cs.washington.edu}
+\email{skahan@cs.washington.edu}\\ \\
 Preston Briggs\\ 
 Cray Inc.\\
-\email{\{preston.briggs\}@gmail.com}
+\email{preston.briggs@gmail.com}\\
 }
@@
 
+Our shared-memory programming model partitions the overall address
+space into a global shared address space and per node private address
+spaces.  Locality may be exploited directly by the programmer in the
+node private address spaces, just as in a conventional cache-coherent
+multiprocessor node.  In contrast, locality cannot be exploited in 
+the global space.
+
+Whereas in the PGAS model coherency of global copies is implictly
+maintained across the system, we forego caching of globally shared
+data.  A node can exploit temporal locality of data stored in
+the global space only by first copying it into the node's local space.
+In this way, a node can access data at random across the system while
+incurring no hardware coherence overhead: consistency is enforced by
+our mechanisms for explicit global synchronization.
+
+The model promises efficiency subject to the presence of sufficient
+concurrency.  Exactly how that concurrency is expressed by the
+programmer is language dependent.  Our programming model serves to
+augment threaded programming languages so that they become latency
+tolerant on large address spaces.  Software multithreading is used to
+provide latency tolerance: a computation that is about to execute a
+long latency operation such as a remote memory reference initiates the
+operation and then yields to the scheduler, which quickly resumes
+execution of another computation.  Concurrency is also used -- via
+aggregation of requests to a common node destination -- to make more
+efficient use of network bandwidth and in some cases reduce bandwidth
+demands.
+
+Synchronization on locations in the node private address space is the
+same as on conventional systems.  Synchronization on global addresses
+works differently: we provide atomic operations such as {\tt
+  int\_fetch\_add} as well as operations using full-bits, as on the
+Cray XMT.  As is true for other long latency operations,
+synchronization latency is tolerated by yielding to the scheduler.
+Even non-blocking global atomics yield: atomic operations executed as
+x86 instructions sabbotage concurrency by forcing all outstanding
+memory references to complete before new ones can be issued.  By
+yielding, the core can switch to another computation while the
+synchronization is accomplished outside of the pipeline.  In addition,
+the implementation can exploit aggregation when concurrency is
+abundant to increase the throughput of global synchronization.
+
+Thus, the programming model provides the programmer with options: express
+locality within the private space, and it will be exploited to reduce
+latency and bandwidth; express concurrency, and it will be used toward
+tolerating latency and reducing the cost of bandwidth.
+
+\begin {comment}
 \todo{Change to discuss API:

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
evaluation, post-meeting pass

--- 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \label{sec:related}
-The concept of using independent parallel contexts to tolerate latency
-is not novel.  The inspiration for our project is of course the Cray
-XMT \cite{feo-xmt}, which implemented a hardware version of our
-programming model (in what would today be considered exceptionally
-slow silicon.)  The XMT can be seen as simultaneous multithreading
-\cite{tullsen-smt} writ large; the technique and motivation is
-essentially the same, but the XMT allows hundreds of contexts (modern
-Intel SMT implementations are generally limited to two.) The Alewife
-system \cite{agarwal-alewife} also switches contexts (in hardware) to
-hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
-allows hundreds of independently executing contexts to share FPUs,
-caches, and paths to memory, but does not context switch or share
-basic execution hardware.
 
-Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
-model under the hood;
-% Do I need to include a citation here?
-while they generate large numbers of available execution contexts via
-a (semi-)explicit SIMD model, those SIMD instructions keep functional
-units satisfied by overlapping execution with the loads required for
-other contexts.
+\todo{actually write this}
 
-Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
-  multithreading}.  As opposed to our system of explicitly marked
-remote references, SCM assumes a high hit rate and traps to an
-interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
-short enough latencies that using more than two contexts generally
-produced slowdown.
+Multithreading for latency tolerance has been explored in a number of directions.
 
-Partitioned global address space models, such as that used in Unified Parallel C,
-% definitely need a citation here
- share our approach of exposing all data (even remote data)
- transparently to any threads and have largely solved the problems of
- how to efficiently translate those remote reads into message
- passing.  However, their programming models emphasize locality more
- than our work does; while a program can examine any vertex in a UPC
- program at any time, it probably shouldn't (whereas our kernels rely on our
- latency hiding technique to make common remote references fast.
+Hardware approaches: The Tera MTA and Cray XMT, the MIT Alewife
+project, ??? use hardware context switches to tolerate memory latency
+for large systems. More recently, Simultaneous Multi-Threading,
+Niagara, Cyclops, and even GPUs switch between multiple contexts to
+tolerate memory latency, for memory that is hundreds of cycles of
+away, rather than the thousands we must tolerate.
 
-The value of the information in extremely large and sparse graphs is
-rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
-sophisticated analysis on data from Twitter, identifying important
-actors and their networks of influence, and it can scale to the full
-size of graphs such as Facebook's friend network.
+software:
+
+Mowry's SCM
+
+Cullers' TAM
+
+qthreads
+
+cilk? tbb?
+
+why not wait for more cores?
+
 \section{Programming model}
@@
 
-Our shared-memory programming model partitions the overall address
-space into a global shared address space and per node private address
-spaces.  Locality may be exploited directly by the programmer in the
-node private address spaces, just as in a conventional cache-coherent
-multiprocessor node.  In contrast, the global space is explicitly
-uncached.  There, every processor can access any word of memory
-directly via its unique global address.
-
-Whereas in the PGAS model coherency of global copies is implictly
-maintained across the system, we forego caching of globally shared
-data.  A node can exploit temporal locality of data stored in
-the global space only by first copying it into the node's local space.
-In this way, a node can access data at random across the system while
-incurring no hardware coherence overhead: consistency is enforced by
-our mechanisms for explicit global synchronization.
-
-The model promises efficiency subject to the presence of sufficient
-concurrency.  Exactly how that concurrency is expressed by the
-programmer is language dependent.  Our programming model serves to
-augment threaded programming languages so that they become latency
-tolerant on large address spaces.  Software multithreading is used to
-provide latency tolerance: a computation that is about to execute a
-long latency operation such as a remote memory reference initiates the
-operation and then yields to the scheduler, which quickly resumes
-execution of another computation.  Concurrency is also used -- via
-aggregation of requests to a common node destination -- to make more
-efficient use of network bandwidth and in some cases reduce bandwidth
-demands.
-
-Synchronization on locations in the node private address space is the
-same as on conventional systems.  Synchronization on global addresses
-works differently: we provide atomic operations such as {\tt
-  int\_fetch\_add} as well as operations using full-bits, as on the
-Cray XMT.  As is true for other long latency operations,
-synchronization latency is tolerated by yielding to the scheduler.
-Even non-blocking global atomics yield: atomic operations executed as
-x86 instructions sabbotage concurrency by forcing all outstanding
-memory references to complete before new ones can be issued.  By
-yielding, the core can switch to another computation while the
-synchronization is accomplished outside of the pipeline.  In addition,
-the implementation can exploit aggregation when concurrency is
-abundant to increase the throughput of global synchronization.
-
-Thus, the programming model provides the programmer with options: express
-locality within the private space, and it will be exploited to reduce
-latency and bandwidth; express concurrency, and it will be used toward
-tolerating latency and reducing the cost of bandwidth.
-
-\begin {comment}
 \todo{Change to discuss API:
@@
 
-\end{comment}
-
 \section{Approach}
@@
 rates, with a round-trip latency of approximately 3$\mu$s
-\todo{cite}. Little's law tells us then that we must support on the
+\todo{cite} \todo{we calculate the 3us in the evaluation based on some numbers for sw-to-sw plus switches, should we pull that up here?}. Little's law tells us then that we must support on the
 order of 300 concurrent memory operations--and thus, up to this many coroutines--to
@@
 writes over QPI to the network interface. Adding more memory manager
-threads may enable more memory concurrency, but the overhead is still high \todo{why?
-because lots of code to run?
-so accel by impl in hardware.
-Maybe this is answered by the "what in...is now" already}
-
+threads may enable more memory concurrency, but the overhead is still high \todo{reword so overhead still high sounds supported as above}
 \subsubsection{Hardware management}
@@
 
-To evaluate the feasibility of our proposal, we ran pointer chasing
-benchmarks on a single-node implementation of our runtime. These
-pointer chasing experiments are intended to model a particular
-``wost-case'' behavior of irregular applications, where each memory
-reference causes a cache miss.
+Our goal is to evaluate the feasibility of our proposed runtime.
+We want to determine two things: whether coroutines can generate memory concurrency without incurring minimal performance overhead, and whether the system can support the level of concurrency required to tolerate the latency that will be present in a multi-node system.
 
-We ran these experiments on a Dell PowerEdge R410 with two Xeon X5650
-processors and 24GB of RAM, with hyperthreading disabled. These
-processors use a NUMA memory architecture, where each processor has
+We focus the evaluation on one component of the runtime: lightweight context switching.
+We ran pointer-chasing benchmarks on a single-node implementation of our runtime.
+These pointer chasing experiments are intended to model a particular ``worst-case'' behavior of irregular applications, where each memory reference causes a cache miss.
+
+We ran these experiments on a Dell PowerEdge R410 with two Xeon X5650 (Nehalem microarchitecture)
+chips and 24GB of RAM, with hyperthreading disabled. These
+chips use a NUMA memory architecture, where each chip has
 its own integrated memory controller and DIMMs; references to other
-processors are carried over Intel's cache-coherent QuickPath
+chips' memory are carried over Intel's cache-coherent QuickPath
 Interconnect (QPI).
 
-Our evaulation consists of three parts. First, we describe the
-performance of the test machine's memory system to use as a
-baseline. Then, we demonstrate that the same performance is achievable
-using our runtime system: we measure pointer chasing performance and
-the effects of cache pressure due to the number of contexts
-required. Finally, we simulate the effects of fetching memory over a
-network by adding delays to our runtime.
+Our evaulation consists of two parts.
+First, we demonstrate that the runtime system can achieve the same performance as when there is explicit memory concurrency available: we measure pointer chasing performance and
+the effects of cache pressure due to the number of contexts required.
+Second, we look at the runtime within the multi-node picture and simulate the effects of fetching memory over a network to show that the runtime can reach the level of concurrency required to hide the latency.
+In each part we begin by studying relevant aspects of our test machine's memory system so we can interpret our runtime results.
 
-\subsection{Memory system performance}
+\subsection{Single-node}
 
 We first characterize the performance of our test machine's memory
-system. We measured three parameters: the maximum random reference
-rate that can be issued by the cores in one processor, the maximum random
-reference rate that can be serviced by one processor's memory controller,
-and the maximum random reference rate that can be carried over the QPI
-link and serviced by another processor's memory controller.
+system. Then we use these results to evaluate the performance of our runtime within a single node.
 
-To find the maximum random reference rate that one processor's cores
+\subsubsection{Base memory system performance}
+\todo{title might be improved}
+
+We measured two parameters: the maximum random reference
+rate that can be issued by the cores in one chip and the maximum random
+reference rate that can be serviced by one chip's memory controller.\todo{The 360M  "serviced" number could perhaps be an "aside" kind of paragraph since not supporting a major point}
+
+To find the maximum random reference rate that one chip's cores
 can issue, we ran pointer chasing code following the model of
-Figure~\ref{fig:pointernocoro}. Each core issues $n$ list traversals
+Figure~\ref{fig:code}a. Each core issues $n$ list traversals
 in a loop; we call $n * \textit{number of cores}$ the number of
@@
 baseline for evaluating our coroutine library, we depend on the
-processor's ILP for memory concurrency, rather than our coroutine
+cores' exploitation of ILP for memory concurrency, rather than our coroutine
 library.
 
-\begin{figure}
+\begin{figure*}[ht]
+%list walk ILP
+\begin{minipage}[b]{0.3\linewidth}
+\centering
 \begin{alltt}
@@
   \}
+  \center{{\bf (a)}}
 \end{alltt}
-\caption{Pseudocode for pointer chasing without coroutines.}
+%\caption{Pseudocode for pointer chasing without coroutines.}
 \label{fig:pointernocoro}
-\end{figure}
+\end{minipage}
+%listwalk green threads
+\begin{minipage}[b]{0.35\linewidth}
+\centering
+\begin{alltt}
+  while (count-- > 0) \{
+     prefetch(&(list1->next));
+     switch();
+     list1 = read(&(list1->next));
+ \}
+  \center{{\bf (b)}}
+\end{alltt}
+  %%n lists
+  % while (count-- > 0) \{
+  % prefetch(&(list1->next));
+  %   prefetch(&(list2->next));
+   %  \ldots
+   %  prefetch(&(list\(n\)->next));
+   %  switch();
+   %  list1 = read(&(list1->next));
+   %  list2 = read(&(list2->next));
+   %  \ldots
+   %  list\(n\) = read(&(list\(n\)->next));
+%\caption{Pseudocode for pointer chasing using coroutines.}
+\label{fig:pointercoro}
+\end{minipage}
+%cache pressure
+\begin{minipage}[b]{0.32\linewidth}
+\centering
+\begin{alltt}
+  while (count-- > 0) \{
+    prefetch(&(list1->next));
+    switch();
+    list1 = read(&(list1->next));
+    for( i in 1 to num_local_updates ) \{
+      local->data++;
+      local = local->next;
+   \}
+  \}
+   \center{{\bf (c)}}
+\end{alltt}
+%\caption{Pseudocode for pointer chasing with local updates.}
+\label{fig:pointerupdate}
+\end{minipage}
+\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.}
+\label{fig:code}
+\end{figure*}
 
@@
 concurrent offered references. We see a maximum rate of 277 million
-references per second, which agrees with the measurements found by \cite{Mandal:2010} with a similar machine. This rate is achieved when the number of offered
-references is 42 \todo{yes, max is achieved here but essentially levels out earlier like 36; do we want a percentage metric for defining a plateau}.
+references per second, which agrees with the measurements found by \cite{Mandal:2010} with a similar machine.
+This rate is achieved when the number of offered references is 36.
 Note that a single core cannot support this level of
 memory concurrency; the maximum reference rate for a single core is
-107M. We believe this limit is due to the core's {\em line fill
-  buffer} \todo{cite nehalem perf docs}, which limits the core to 10
-concurrent L2 misses.
+107M. We believe this limit is due to the core's limited number of {\em line fill
+  buffers} \cite{nehalem:arch}, which limits the core to 10
+concurrent private cache misses.
 
-The memory controller in the processor has more bandwidth than its
+The memory controller in the chip has more bandwidth than its
 cores can saturate. To measure the memory controller's maximum random
@@
 million references per second. We believe this difference is due to
-the configuration of the processor's {\em Global Queue} \todo{cite nehalem
-  docs}, which limits the number of concurrently-executing read misses
-from a processor's cores to its local last-level cache and memory to 32.
+the configuration of the processor's {\em Global Queue} \cite{nehalem:perf}, which limits the number of concurrently-executing read misses from a chips cores to its local last-level cache and memory to 32.
 
-% take this figure out if we need the space
-% the main point is room for remote references to local memory
-\begin{figure}[h]
-  \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/qpi_bw-edited.pdf}
-  \end{center}
-  \caption{Pointer chasing in a remote processor's memory.}
-  \label{fig:listwalk-qpi}
-\end{figure}
+\subsubsection{Runtime performance}
 
-Finally, we measured the rate at which the cores can make requests
-over the QPI link. For this experiment, we allocated the lists in the
-second processor, and ran the traversals on the first
-processor. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use
-of the QPI link limits us to 175 million references per second.
-
-\subsection{Runtime performance}
-
-To characterize the performance of our runtime, we investigated three
+To evaluate the performance of our runtime, we investigated two
 effects: the maximum reference rate using coroutines
-to obtain memory concurrency, the effects of cache pressure from the
-coroutines' context storage, and the maximum reference rate obtainable
-using a simulated network delay.
-
-\begin{figure}
-\begin{alltt}
-  while (count-- > 0) \{
-     prefetch(&(list1->next));
-     prefetch(&(list2->next));
-     \ldots
-     prefetch(&(list\(n\)->next));
-     switch();
-     list1 = read(&(list1->next));
-     list2 = read(&(list2->next));
-     \ldots
-     list\(n\) = read(&(list\(n\)->next));
- \}
-\end{alltt}
-\caption{Pseudocode for pointer chasing using coroutines.}
-\label{fig:pointercoro}
-\end{figure}
+to obtain memory concurrency and the effects of cache pressure from the
+coroutines' context storage.
 
@@
 library, we modified our pointer chasing benchmark as shown in
-figure~\ref{fig:pointercoro}. Now there are two sources of offered
+figure~\ref{fig:code}b. Now there are two sources of offered
 concurrent references: the ILP exploited by the processor, and the
@@
   \caption{Pointer chasing with coroutines, with one reference per
-    coroutine.}
+    coroutine. Number of coroutines per core is $\textit{total concurrent references} / \textit{number of cores}$.}
   \label{fig:multi-green}
@@
 second with 48 concurrent misses, or 8 coroutines per core. These
-results are very close to the ILP-only experiment.
+results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36.
+When the number of references per coroutine is larger than 1, fewer coroutines are required to reach full performance.
 
@@
 concurrent references per core exceeds 10; we believe this is due to
-later prefetches squashing earlier ones in the line fill buffer. We
-also observe some repeatable outliers below the maximum rate even with enough
-memory concurrency; this is harder to explain, but we suspect it is
+later prefetches squashing earlier ones in the line fill buffers. We
+also observe some points that do not follow the trend and look like noise, but these results are actually repeatable. This effect is harder to explain, but we suspect it is
 due to resource contention in the cores' pipelines once the memory
@@
 
-\begin{figure}
-\begin{alltt}
-  while (count-- > 0) \{
-    prefetch(&(list1->next));
-    switch();
-    list1 = read(&(list1->next));
-    for( i in 1 to num_local_updates ) \{
-      local->data++;
-      local = local->next;
-   \}
-  \}
-\end{alltt}
-\caption{Pseudocode for pointer chasing with local updates.}
-\label{pointerupdate}
-\end{figure}
+%Such points are repeatably lower, and looks like the std dev of such points (the diamond ones than are low) is also higher than other "normal" points: around 35M versus 13M or less, which seems to suggest a weird effect at those numbers}
 
@@
 to include random updates to a per-coroutine local data structure that
-is small enough to fit in cache. Figure~\ref{pointerupdate} shows the
+is small enough to fit in cache. Figure~\ref{fig:code}c shows the
 general idea.
@@
 
-Finally, to simulate the performance of pointer chasing in a
+\subsection{Multi-node}
+
+To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.
+First we measure the QPI bandwidth between the two chips of the test machine, since the chip would talk to the network interface over a QPI link.
+
+% take this figure out if we need the space
+% the main point is room for remote references to local memory
+\begin{figure}[h]
+  \begin{center}
+    \includegraphics[width=0.5\textwidth]{figures/qpi_bw-edited.pdf}
+  \end{center}
+  \caption{Pointer chasing in a remote processor's memory.}
+  \label{fig:listwalk-qpi}
+\end{figure}
+%%%%%%%%%%
+
+We measured the rate at which the cores can make requests
+over the QPI link. For this experiment, we allocated the lists in the
+second chip's memory and ran the traversals on the first chip's cores. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use of the QPI link limits us to 175 million references per second.
+
+To simulate the performance of pointer chasing in a
 multi-node system, we modified the scheduler of our coroutine library
@@
 latencies between two computers connected back-to-back; we assume a
-switch transit delay of 400 ns in each direction and thus estimate a 3
+switch transit delay of 400 ns \todo{how many hops} in each direction and thus estimate a 3
 $\mu$s round-trip delay. In this experiment, once a coroutine has
@@
 Figure~\ref{fig:network-delay}. We are still able to reach a
-near-maximum rate of 273 million references per second, but at least 3
+near-maximum rate of 273 million references per second, but at least 2
 concurrent references per coroutine are required. Running with a single
@@
 
-\todo{do we want to include that the memory system imposes max limit so really more throughput is still achievable if there are enough ILP memory concurrency?}
-\todo{may want to rerun with memory on remote core to incorporate QPI limit?}
+\todo{do we want to include that the memory system imposes max limit so really more throughput is still achievable if there are enough ILP memory concurrency?}\\
+\todo{also have results that remote cores over QPI plus delay 3us can make it up to the 175M, with sufficient number of coroutines}
 

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex

@@
 \label{sec:related}
+The concept of using independent parallel contexts to tolerate latency
+is not novel.  The inspiration for our project is of course the Cray
+XMT \cite{feo-xmt}, which implemented a hardware version of our
+programming model (in what would today be considered exceptionally
+slow silicon.)  The XMT can be seen as simultaneous multithreading
+\cite{tullsen-smt} writ large; the technique and motivation is
+essentially the same, but the XMT allows hundreds of contexts (modern
+Intel SMT implementations are generally limited to two.) The Alewife
+system \cite{agarwal-alewife} also switches contexts (in hardware) to
+hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
+allows hundreds of independently executing contexts to share FPUs,
+caches, and paths to memory, but does not context switch or share
+basic execution hardware.
 
-\todo{actually write this}
+Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
+model under the hood;
+% Do I need to include a citation here?
+while they generate large numbers of available execution contexts via
+a (semi-)explicit SIMD model, those SIMD instructions keep functional
+units satisfied by overlapping execution with the loads required for
+other contexts.
 
-Multithreading for latency tolerance has been explored in a number of directions.
+Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
+  multithreading}.  As opposed to our system of explicitly marked
+remote references, SCM assumes a high hit rate and traps to an
+interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
+short enough latencies that using more than two contexts generally
+produced slowdown.
 
-Hardware approaches: The Tera MTA and Cray XMT, the MIT Alewife
-project, ??? use hardware context switches to tolerate memory latency
-for large systems. More recently, Simultaneous Multi-Threading,
-Niagara, Cyclops, and even GPUs switch between multiple contexts to
-tolerate memory latency, for memory that is hundreds of cycles of
-away, rather than the thousands we must tolerate.
+Partitioned global address space models, such as that used in Unified Parallel C,
+% definitely need a citation here
+ share our approach of exposing all data (even remote data)
+ transparently to any threads and have largely solved the problems of
+ how to efficiently translate those remote reads into message
+ passing.  However, their programming models emphasize locality more
+ than our work does; while a program can examine any vertex in a UPC
+ program at any time, it probably shouldn't (whereas our kernels rely on our
+ latency hiding technique to make common remote references fast.
 
-software:
-
-Mowry's SCM
-
-Cullers' TAM
-
-qthreads
-
-cilk? tbb?
-
-why not wait for more cores?
-
+The value of the information in extremely large and sparse graphs is
+rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
+sophisticated analysis on data from Twitter, identifying important
+actors and their networks of influence, and it can scale to the full
+size of graphs such as Facebook's friend network.
 \section{Programming model}

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
finish up carls eval comments

@@
 \end{minipage}
-\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.}
+\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.
+Coroutine code, as in (b), might also include ILP like what (a) depends on for memory concurrency, e.g., by doing multiple prefetches to independent lists, switch, then multiple reads.}
 \label{fig:code}
@@
 library, we modified our pointer chasing benchmark as shown in
-figure~\ref{fig:code}b. Now there are two sources of offered
+Figure~\ref{fig:code}b.
+Recall that the baseline code in Figure~\ref{fig:code}a relied solely on ILP to achieve memory concurrency, which may not be abundant in real applications.
+Using coroutines, there are two sources of offered
 concurrent references: the ILP exploited by the processor, and the
@@
 latencies between two computers connected back-to-back; we assume a
-switch transit delay of 400 ns \todo{how many hops} in each direction and thus estimate a 3
+switch transit delay of 400 ns \todo{how many hops? 1or3?} in each direction and thus estimate a 3
 $\mu$s round-trip delay. In this experiment, once a coroutine has
@@
 reference per coroutine achieves 236 million references per
-second. Any network interface would be connected through a QPI link,
+second.
+Note that the throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support.
+This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.
+
+We also observe that to achieve a throughput equal to the expected max outgoing network bandwidth (100 million references per second), \todo{reference earlier section calculation?} when there is only one reference per context available, we require 320 coroutines.
+This is close to the minimum of 300 concurrent references needed to tolerate $3 \mu s$ latency.
+
+Any network interface would be connected through a QPI link,
 which we measured to have a bandwidth limit of 175 million references

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
implementation

--- 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \author{
-Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Luis Ceze\\
+Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
+Carl Ebeling, Luis Ceze\\
 University of Washington\\
-\email{\{nelson, bdmyers, ahh, djg, luisceze\}@cs.washington.edu}\\ \\
+\email{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}\\ \\
 Simon Kahan \\ 
 Pacific Northwest National Lab\\
-\email{\{skahan\}@cs.washington.edu}
+\email{skahan@cs.washington.edu}\\ \\
+Preston Briggs\\ 
+Cray Inc.\\
+\email{preston.briggs@gmail.com}\\
 }
@@
 commodity machine's memory. Their low diameter means that they are
-difficult to lay out with locality. Their power law distribution means
-that a few vertices generate much more work than the rest, leading to
-computational imbalance and even less locality. Together, these
-properties lead to a difficult conclusion: each edge traversal is
-likely to be a very expensive cache miss.
+difficult to lay out with locality: every node needs to be near every
+other node. Their power law distribution means that a few vertices
+generate much more work than the rest, leading to computational
+imbalance and even less locality. Together, these properties lead to a
+difficult conclusion: each edge traversal is likely to be a very
+expensive cache miss.
 
@@
 
+\begin{figure*}[htbp]
+  \begin{center}
+    \includegraphics[width=0.9\textwidth]{figures/system-overview.pdf}
+	\end{center}
+	\caption{Overview of our proposed system design. The base
+          system is a cluster of commodity nodes and interconnect. We
+          add a latency-tolerant runtime and a global memory manager
+          to obtain efficient access to the global address space. Only
+          the shaded components may require hardware design.}
+	\label{fig:system-overview}
+\end{figure*}
+
 We want the best of both worlds. Our goal is to build a system that
 has good performance on these low-locality graph codes but is
-implemented with commodity processors. The core of our approach is a
-lightweight coroutine library designed to overlap long-latency memory
-prefetches with other contexts' computation.
+implemented with cheap commodity processors.
 
-In this paper, we describe our single-node runtime for latency
-tolerance, along with our plans for scaling to multiple nodes.  We
-evaluate the performance of our runtime using a set of list traversal
-benchmarks designed to model certain worst-case behaviors of these
-large graph problems.
+Figure~\ref{fig:system-overview} shows the structure of our
+proposal. It is composed of multiple nodes built with commodity
+processors, communicating over an Infiniband network. We add two
+components: a runtime system, responsible for executing the user's
+threads, and a global memory manager, responsible for facilitating
+memory requests to the non-coherent global memory shared across the
+nodes.  This memory manager is the only component that might require
+hardware, but we are exploring a mix of hardware and software
+implementations.
 
-The rest of the paper is organized as follows. We present related work
-in section~\ref{sec:related}. We describe the programming model we are
-implementing in section~\ref{sec:model}, and describe our approach for
-implementing this model in section~\ref{sec:approach}. We evaluate our
-latency-tolerant runtime in section~\ref{sec:evaluation} and conclude
-in section~\ref{sec:conclusion}.
+This paper explores the feasibility of our multi-node goal using a
+single-node implementation of our lightweight threading runtime. We
+use prefetch instructions together with lightweight threads to
+tolerate the latency of DRAM access on the local node node. We use
+pointer chasing benchmarks to verify our runtime's overhead is
+acceptable. We also model the effects of pointer chasing on a remote
+node.
+
+The rest of the paper is organized as follows. We describe our
+proposal from a programmer's viewpoint in section~\ref{sec:model}. We
+describe the implementation of our latency-tolerant runtime and plans
+for extending to multiple nodes in section~\ref{sec:approach}. We
+evaluate our runtime in section~\ref{sec:evaluation}.  We present
+related work in section~\ref{sec:related} and conclude in
+section~\ref{sec:conclusion}.
 
@@
 the global space.
+
+\todo{explain why global address space is good}
 
@@
 
-\section{Approach}
+\section{Implementation}
 \label{sec:approach}
 
-Our goal is to enable high throughput on graphs having two challenging
-properties. First, the graphs' combination of sparseness and low
-degree means there is little locality to be exploited. Second, the
-graphs are larger than a single node's memory and do not partition
-well because of low diameter \todo{enough explanation now?}. We must spread the graph across multiple nodes
-and send memory operations over a network. If memory requests are
-uniformly distributed over the working set, a given node will make as
-many accesses to remote nodes' memory as there are remote requests to the local node's memory coming in.
+The goal of our implementation is to use multithreading to tolerate
+the latency of random accesses to memory in a global address space
+spread across multiple nodes. This goal has three key challenges:
 
-We would like to run as fast as current hardware will let us. Pin
-bandwidth is the tightest resource: Intel's QuickPath Interconnect
-(QPI) allows approximately 100 million requests per second
-\cite{quickpath:website} \todo{specifically 25.6GB/s two way, which we expect 6.4GB/s for one way for outgoing request}, and recent interconnect products allow similar request
-rates, with a round-trip latency of approximately 3$\mu$s
-\todo{cite} \todo{we calculate the 3us in the evaluation based on some numbers for sw-to-sw plus switches, should we pull that up here?}. Little's law tells us then that we must support on the
-order of 300 concurrent memory operations--and thus, up to this many coroutines--to
-cover this latency.
+\begin{itemize}
+\item {\bf Lightweight multithreading}
 
-These requirements drive the design of our system. There are three key
-components.
-First, the system needs lightweight context switches (coroutines) to be able to support many contexts--thereby increasing memory concurrency available--with minimal impact on performance.
-Second, a memory management subsystem must support many concurrent accesses to a global address space.
-Third, the system must provide synchronization at a fine-granularity \todo{this last one sounds incomplete because no reason given}.
+  We expect to need hundreds of
+  threads to tolerated a cluster's network latency, so our threading
+  implementation must have low overhead.
 
-\subsection{Lightweight context switch}
-\todo{can we include in the title something that means lightweight and good scheduling?}
+\item {\bf Concurrency in global memory references}
+
+  A thread making a long-latency global memory reference must be able
+  to yield and allow other threads to make progress.
+
+\item {\bf Low overhead synchronization}
+
+  In a system with high degrees of concurrency, lightweight
+  synchronization primitives are important.
+\end{itemize}
+
+We discuss our approach to solving each of these challenges in
+turn. For each one, we discuss both our current single-node
+implementation and ideas for a multi-node implementation.
+
+\subsection{Lightweight multithreading}
+
+Our approach to supporting many lightweight threads uses a cooperative,
+user-level multitasking system built using coroutines.
 
 There are three main concerns in designing the coroutine
-mechanism. First, contexts must be small, so many coroutines can be
-active without monopolizing the cache. Second, context switches must
-be fast, so that overhead does not swamp actual work. Third, when a
-coroutine yields, we must choose a suitable coroutine to run next.
+mechanism. First, the working set of each coroutine must be small, so
+many coroutines can be active without monopolizing the cache. Second,
+context switches must be fast, so that overhead does not swamp actual
+work. Third, when a coroutine yields, we must choose a suitable
+coroutine to run next.
 
 We minimize context size by treating context switches as function
-calls, as in \cite{charm}. This allows us to save and restore the
+calls, as in~\cite{charm}. This allows us to save and restore the
 minimum number of registers allowed by the ABI, and depend on the
@@
 
-The choice of which coroutine to run next presents us with a
-tradeoff. The simplest scheduler chooses coroutines to execute in
-round-robin fashion. But this may lead to performance problems: a
-coroutine might be scheduled before the event which caused it to yield (e.g., a long latency memory access)
-has completed, stalling the pipeline and keeping other coroutines from
-progressing. A coroutine might also be scheduled long after the event
-which caused it to yield has completed; while this wouldn't block
-other coroutines from proceeding, it would add unnecessary latency to
-the yielding coroutine. A more intelligent scheduler might allow
-coroutines to execute only when they were ready, but this could
-increase the overhead of context switching. Our initial implementation
-takes the simple approach, but more investigation is warranted.
+There are a number of possibilites for coroutine scheduling.
+Ideally, a thread would be rescheduled when its request had
+completed, but this may complicate the scheduling decision and slow
+down context switches. A simple round-robin scheduler may be sufficient.
 
-\subsection{Accessing Global Memory}
+\subsubsection{Single-node implementation}
+We implemented coroutines as described, with a round-robin
+scheduler. If a prefetched data value is no longer in the cache when
+the coroutine is rescheduled and the value is used, the coroutine pays
+the full cost of the fetch.
 
-% Our goal is to support many concurrent accesses to a global address
-% space spread across nodes in a cluster. We can draw on hardware
-% support developed for the Partitioned Global Address Space family of
-% languages to accomplish this.
+\subsubsection{Multi-node ideas}
+We will explore how the memory manager and coroutine scheduler may
+interact, so that threads waiting on long-latency memory operations or
+synchronization operations will be scheduled soon after their
+operations have completed.
 
-% We draw on libraries and network support 
-% goal. Languages in the Partitioned Global Address Space family such as
-% X10 and Chapel have similar goals,
+\subsection{Concurrency in global memory references}
 
-% It is also useful to access this low-locality data without disturbing
-% the thread contexts and local storage in the cache. Current x86
-% processors support non-temporal loads and stores. These memory
-% operations are quite relevant for our problem: they are designed to
-% move data without polluting the cache. Using these instructions works
-% better than mapping the low-locality data as uncached, since that
-% restricts the processors' ability to reorder operations.
+To enable concurrency in memory references, we break each read or
+write into two operations: the {\em issue} operation and the {\em
+  complete} operation. The read issue operation starts data moving
+towards the processor and returns immediately, like a prefetch; the
+read complete operation blocks until data is available. The write
+issue operation takes data along with an address, and starts an
+asynchronous write; the write complete operation blocks until the
+write is committed.
 
-Following the lead of the PGAS family of languages, we will use
-the RDMA functionality provided by modern Infiniband networks to
-access our global memory. \todo{rethink?}
+A read operation in user code consists of three steps: a {\em read
+  issue}, which causes the data to start moving; a {\em yield}, which
+causes another thread to start executing, overlapping the read
+latency; and, once the reading thread has been rescheduled, a {\em
+  read complete} blocks until the data is available. A write operation
+follow the same pattern.
 
-These networks allow us to read and write memory directly from remote
-nodes, but they do so through a low-level interface designed to be
-hidden behind a library. Loads and stores to global memory must be
-replaced with calls to a library. This library causes the network
-interface to fetch the remote data and place it in a temporary
-buffer. When the requesting coroutine is scheduled again, data is
-returned from the buffer and execution proceeds.
+\subsubsection{Single-node implementation}
+We use prefetch instructions for the issue operation, and regular
+blocking reads and writes for the complete operation. Since we expect
+these global references to have low locality, we use the
+``non-temporal'' form of prefetch, which tries to bring the data near
+the core without disturbing the state of the cache
+hierarchy. \todo{cite intel docs}
 
-\subsubsection{Software management}
-In order to minimize memory access overhead, the memory operation
-library calls do not interact with the network interface
-directly. Instead, they queue their operations to another thread, the
-{\em memory manager thread}, running on another core in the CPU. This thread
-translates the global address into a network destination and issues
-the network operation, checks for completion, and manages the returned
-data buffers.
+\subsubsection{Multi-node ideas}
 
-But we have to be careful---directing memory requests from multiple
-compute threads to a single memory manager thread may reduce memory
-concurrency. And we have increased the overhead of a single memory
-operation: what in the single node case was a simple read, is now
-multiple queuing operations between threads and multiple reads and
-writes over QPI to the network interface. Adding more memory manager
-threads may enable more memory concurrency, but the overhead is still high \todo{reword so overhead still high sounds supported as above}
-\subsubsection{Hardware management}
-\todo{might want to call it accelerator throughout after saying FPGA once}
-Considering this overhead, we propose hardware acceleration with a {\em memory manager
-  FPGA}, a coprocessor FPGA inserted into a Xeon socket so it can
-participate in the node's coherence domain. This coprocessor does no
-computation---it would only manage the global memory requests of the
-main CPU, performing the function of the memory manager thread
-described earlier.
+We believe we will need hundreds of outstanding memory references per processor to
+cover the latency of remote references in a multi-node
+system. Unfortunately, commodity processors have much smaller limits
+on memory concurrency: section \todo{cite evaluation} finds a limit of
+36 concurrent loads. We will have to manage memory concurrency on our
+own to avoid this limit. We describe approaches in both software and
+hardware.
 
-By participating in the coherence domain, the memory manager FPGA can
-act as a proxy for the rest of global memory. It advertises the
-address space of the rest of the system, and translates local loads
-and stores to addresses on remote nodes into the network operations
-required.
+One approach is for user code to delegate global references to a
+special {\em global memory manager thread}, running on another core in
+the chip. This thread translates the global address into a network
+destination, makes the request through the network card to the remote
+node, checks for completion, and delivers the returned data to the
+requesting thread. On the remote node, the remote memory manager
+thread would perform the memory operation and return the data to the
+requesting node.
 
-Unfortunately, current processors do not provide sufficient hardware
-resources for us to reach our concurrency target using this
-approach. Each core in Intel's X5650 Xeon supports only 10 outstanding
-misses from its private L2; the entire chip is limited to 32 L2 misses
-to local memory and 12 L2 misses for memory reached over the QPI
-bus. A simple translation of a local load to a network access would
-tie up one of these slots for the entire multi-microsecond latency of
-the request. To allow the amount of concurrency we need, we must bypass these structures; the CPU needs to direct the memory manager FPGA
-to do work on its behalf without tying up its own resources.
+This approach has the benefit of enabling our entire system to be
+build using commodity parts, but its potential for performance is not
+clear. We have increased the overhead of a single memory operation:
+what in the single node case was a simple read, is now multiple
+queuing operations between threads and multiple reads and writes over
+the processor's link to the network interface.
 
-To accomplish this, we encode commands in the high-order bits of
-remote memory addresses. The memory manager FPGA advertises each
-remote address twice \todo{this sounded like an active thing the FPGA is doing each time you make remote request, first time I read it.
-Consider just sticking to cmd as upper order address bits than saying advertise each address twice}.
-The first address is used for blocking reads and
-writes---any access incurs the full latency of the remote
-access. \todo{verify writes block?} The second address is used for
-prefetches and asynchronous writes. These addresses differ by one
-high-order bit, easily set with a mask in a macro. A read from the
-prefetch address starts the network transaction, but immediately
-returns a dummy value. When the value arrives, it is stored in a
-temporary buffer; when the requesting coroutine is scheduled again, it
-executes the blocking read, and the memory manager immediately returns
-the value from the buffer \todo{is this necessarily restricted to the memory manager returning value from buffer rather than buffer in local memory?
-maybe it is, since if buffer was in local mem, would have to also write a full bit or make two reads: valid bit and data}.
-\todo{this isn't fully baked---do we want
-  to include this sort of detail?}
+Another approach moves the global memory management to a hardware
+device. We can add a coprocessor FPGA in a processor socket, listening
+to coherence messages on the bus. Note that this accelerator is only a
+memory proxy, not a compute coprocessor. When it detects a reference
+to memory stored on a remote node, the accelerator would issue the
+request through the network controller and track its progress,
+delivering the data to the requesting thread when it was ready.
 
-\subsection{Synchronization}
+All global shared memory would be mapped into a segment of the local
+address space on each node. The accelerator would do the
+address-to-node translation. In addition, to support the issue and
+complete operations for concurrency support, commands would be encoded
+in the upper bits of the address: then each remote word would have multiple
+addresses on the local node. Reading from the {\em issue} address would
+immediately return a dummy value, but the accelerator would issue the
+request over the network. Reading from the {\em complete} address
+would block until the data was available.\todo{adjust with carl's suggestions}
 
-In systems with hundreds of threads per node, lightweight
-synchronization is important. While we have not implemented any
-synchronization primitives yet, full-empty bits are a natural choice
-given our programming model \todo{say why natural choice unless programming model section makes this point already}. Two questions then arise: at what
-granularity should full-empty bits work, and what component should do
-the synchronization?
+With either approach, the question of request aggregation will be
+important. Each memory request we make is small: 8 bytes is a likely
+size. But most networks are optimized for bulk transfers of a few
+thousand bytes. To improve network utilization, we will explore the
+aggregation of multiple memory operations to different addresses on
+the same remote node.
 
-We expect most data accesses to be to 64-bit words. It is possible to
-support full-empty bit synchronization on arbitrary words by
-allocating additional storage for the full-empty bits, and translating
-each atomic full-empty operation into a sequence of memory operations
-that modify the data and full-empty bit atomically.
+\subsection{Low-overhead synchronization}
+
+Just as with reads and writes, we enable concurrency around
+synchronization operations by splitting them into {\em issue} and {\em
+  complete} phases.
+
+Implementing full-empty bit support efficiently on a platform not
+designed for them is a challenge. It is possible to support full-empty
+bit synchronization on arbitrary words by allocating additional
+storage for the full-empty bits, and translating each atomic
+full-empty operation into a sequence of memory operations that modify
+the data and full-empty bit atomically, but this may have high
+overhead. \todo{cite qthreads somehow}
 
 One potential optimization is to limit full-empty synchronization to
-pointers to aligned 64-bit words. This leaves the bottom three bits
-free to store the full-empty bits. These bits would be masked out when
-the pointer was returned to the user.
+pointers to aligned data types, and reuse wasted space in the
+low-order bits of the pointer for full-empty bit storage. These bits
+would be masked out when the pointer was returned to the user.  This
+allows us to synchronize any data type through one level of
+indirection.
 
-These techniques can be implemented in software; \cite{qthreads} is
-one example. Alternatively, atomic operations could be delegated to a
-memory manager FPGA as described in the previous section. The address
-command encoding would be extended to support the various full-empty
-operations.
+\subsubsection{Single-node system}
+Just as with reads and writes, we prefetch and yield before performing
+a synchronization operation. We support only the synchronization
+operations supported by our development platform; we have not
+implemented full-empty bits.
 
-It might seem natural to use the FPGA as a memory controller as
-well, storing all full-empty-bit-capable memory in DRAMs attached to
-the FPGA. While this is conceptually simple, it does make the
-programming model and the design more complicated. Another possibility
-would be to store only the full-empty bits in memories attached to the
-FPGA, and leave data in memory attached to the CPU. We believe that
-starting simple is best: both data and full-empty bits should be
-stored in the CPU's memory. \todo{discuss QPI bandwidth differences}
+\subsubsection{Multi-node ideas}
+
+Synchronization operations can also be delegated to a memory manager
+thread or accelerator. This may simplify the implementation of
+synchronization operations: if only a single core (or single
+accelerator) is accessing the data being synchronized, the use of
+memory fences may be reduced or eliminated.
+
+Note that we are not advocating using an FPGA as a memory controller
+with attached DRAM and additional storage for full-empty bits. We
+believe this unnecessarily complicates the design for such a
+component. Our goal is to minimize the amount of hardware design our
+proposal requires.
 
@@
 
-Our goal is to evaluate the feasibility of our proposed runtime.
-We want to determine two things: whether coroutines can generate memory concurrency without incurring minimal performance overhead, and whether the system can support the level of concurrency required to tolerate the latency that will be present in a multi-node system.
+To evaluate the feasibility of our proposal, we ran pointer chasing
+benchmarks on a single-node implementation of our runtime. These
+pointer chasing experiments are intended to model a particular
+``wost-case'' behavior of irregular applications, where each memory
+reference causes a cache miss.
 
-We focus the evaluation on one component of the runtime: lightweight context switching.
-We ran pointer-chasing benchmarks on a single-node implementation of our runtime.
-These pointer chasing experiments are intended to model a particular ``worst-case'' behavior of irregular applications, where each memory reference causes a cache miss.
-
-We ran these experiments on a Dell PowerEdge R410 with two Xeon X5650 (Nehalem microarchitecture)
-chips and 24GB of RAM, with hyperthreading disabled. These
-chips use a NUMA memory architecture, where each chip has
+We ran these experiments on a Dell PowerEdge R410 with two Xeon X5650
+processors and 24GB of RAM, with hyperthreading disabled. These
+processors use a NUMA memory architecture, where each processor has
 its own integrated memory controller and DIMMs; references to other
-chips' memory are carried over Intel's cache-coherent QuickPath
+processors are carried over Intel's cache-coherent QuickPath
 Interconnect (QPI).
 
-Our evaulation consists of two parts.
-First, we demonstrate that the runtime system can achieve the same performance as when there is explicit memory concurrency available: we measure pointer chasing performance and
-the effects of cache pressure due to the number of contexts required.
-Second, we look at the runtime within the multi-node picture and simulate the effects of fetching memory over a network to show that the runtime can reach the level of concurrency required to hide the latency.
-In each part we begin by studying relevant aspects of our test machine's memory system so we can interpret our runtime results.
+Our evaulation consists of three parts. First, we describe the
+performance of the test machine's memory system to use as a
+baseline. Then, we demonstrate that the same performance is achievable
+using our runtime system: we measure pointer chasing performance and
+the effects of cache pressure due to the number of contexts
+required. Finally, we simulate the effects of fetching memory over a
+network by adding delays to our runtime.
 
-\subsection{Single-node}
+\subsection{Memory system performance}
 
 We first characterize the performance of our test machine's memory
-system. Then we use these results to evaluate the performance of our runtime within a single node.
+system. We measured three parameters: the maximum random reference
+rate that can be issued by the cores in one processor, the maximum random
+reference rate that can be serviced by one processor's memory controller,
+and the maximum random reference rate that can be carried over the QPI
+link and serviced by another processor's memory controller.
 
-\subsubsection{Base memory system performance}
-\todo{title might be improved}
-
-We measured two parameters: the maximum random reference
-rate that can be issued by the cores in one chip and the maximum random
-reference rate that can be serviced by one chip's memory controller.\todo{The 360M  "serviced" number could perhaps be an "aside" kind of paragraph since not supporting a major point}
-
-To find the maximum random reference rate that one chip's cores
+To find the maximum random reference rate that one processor's cores
 can issue, we ran pointer chasing code following the model of
-Figure~\ref{fig:code}a. Each core issues $n$ list traversals
+Figure~\ref{fig:pointernocoro}. Each core issues $n$ list traversals
 in a loop; we call $n * \textit{number of cores}$ the number of
@@
 baseline for evaluating our coroutine library, we depend on the
-cores' exploitation of ILP for memory concurrency, rather than our coroutine
+processor's ILP for memory concurrency, rather than our coroutine
 library.
 
-\begin{figure*}[ht]
-%list walk ILP
-\begin{minipage}[b]{0.3\linewidth}
-\centering
+\begin{figure}
 \begin{alltt}
@@
   \}
-  \center{{\bf (a)}}
 \end{alltt}
-%\caption{Pseudocode for pointer chasing without coroutines.}
+\caption{Pseudocode for pointer chasing without coroutines.}
 \label{fig:pointernocoro}
-\end{minipage}
-%listwalk green threads
-\begin{minipage}[b]{0.35\linewidth}
-\centering
-\begin{alltt}
-  while (count-- > 0) \{
-     prefetch(&(list1->next));
-     switch();
-     list1 = read(&(list1->next));
- \}
-  \center{{\bf (b)}}
-\end{alltt}
-  %%n lists
-  % while (count-- > 0) \{
-  % prefetch(&(list1->next));
-  %   prefetch(&(list2->next));
-   %  \ldots
-   %  prefetch(&(list\(n\)->next));
-   %  switch();
-   %  list1 = read(&(list1->next));
-   %  list2 = read(&(list2->next));
-   %  \ldots
-   %  list\(n\) = read(&(list\(n\)->next));
-%\caption{Pseudocode for pointer chasing using coroutines.}
-\label{fig:pointercoro}
-\end{minipage}
-%cache pressure
-\begin{minipage}[b]{0.32\linewidth}
-\centering
-\begin{alltt}
-  while (count-- > 0) \{
-    prefetch(&(list1->next));
-    switch();
-    list1 = read(&(list1->next));
-    for( i in 1 to num_local_updates ) \{
-      local->data++;
-      local = local->next;
-   \}
-  \}
-   \center{{\bf (c)}}
-\end{alltt}
-%\caption{Pseudocode for pointer chasing with local updates.}
-\label{fig:pointerupdate}
-\end{minipage}
-\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.}
-\label{fig:code}
-\end{figure*}
+\end{figure}
 
@@
 concurrent offered references. We see a maximum rate of 277 million
-references per second, which agrees with the measurements found by \cite{Mandal:2010} with a similar machine.
-This rate is achieved when the number of offered references is 36.
+references per second, which agrees with the measurements found by \cite{Mandal:2010} with a similar machine. This rate is achieved when the number of offered
+references is 42 \todo{yes, max is achieved here but essentially levels out earlier like 36; do we want a percentage metric for defining a plateau}.
 Note that a single core cannot support this level of
 memory concurrency; the maximum reference rate for a single core is
-107M. We believe this limit is due to the core's limited number of {\em line fill
-  buffers} \cite{nehalem:arch}, which limits the core to 10
-concurrent private cache misses.
+107M. We believe this limit is due to the core's {\em line fill
+  buffer} \todo{cite nehalem perf docs}, which limits the core to 10
+concurrent L2 misses.
 
-The memory controller in the chip has more bandwidth than its
+The memory controller in the processor has more bandwidth than its
 cores can saturate. To measure the memory controller's maximum random
@@
 million references per second. We believe this difference is due to
-the configuration of the processor's {\em Global Queue} \cite{nehalem:perf}, which limits the number of concurrently-executing read misses from a chips cores to its local last-level cache and memory to 32.
+the configuration of the processor's {\em Global Queue} \todo{cite nehalem
+  docs}, which limits the number of concurrently-executing read misses
+from a processor's cores to its local last-level cache and memory to 32.
 
-\subsubsection{Runtime performance}
+% take this figure out if we need the space
+% the main point is room for remote references to local memory
+\begin{figure}[h]
+  \begin{center}
+    \includegraphics[width=0.5\textwidth]{figures/qpi_bw-edited.pdf}
+  \end{center}
+  \caption{Pointer chasing in a remote processor's memory.}
+  \label{fig:listwalk-qpi}
+\end{figure}
 
-To evaluate the performance of our runtime, we investigated two
+Finally, we measured the rate at which the cores can make requests
+over the QPI link. For this experiment, we allocated the lists in the
+second processor, and ran the traversals on the first
+processor. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use
+of the QPI link limits us to 175 million references per second.
+
+\subsection{Runtime performance}
+
+To characterize the performance of our runtime, we investigated three
 effects: the maximum reference rate using coroutines
-to obtain memory concurrency and the effects of cache pressure from the
-coroutines' context storage.
+to obtain memory concurrency, the effects of cache pressure from the
+coroutines' context storage, and the maximum reference rate obtainable
+using a simulated network delay.
+
+\begin{figure}
+\begin{alltt}
+  while (count-- > 0) \{
+     prefetch(&(list1->next));
+     prefetch(&(list2->next));
+     \ldots
+     prefetch(&(list\(n\)->next));
+     switch();
+     list1 = read(&(list1->next));
+     list2 = read(&(list2->next));
+     \ldots
+     list\(n\) = read(&(list\(n\)->next));
+ \}
+\end{alltt}
+\caption{Pseudocode for pointer chasing using coroutines.}
+\label{fig:pointercoro}
+\end{figure}
 
@@
 library, we modified our pointer chasing benchmark as shown in
-figure~\ref{fig:code}b. Now there are two sources of offered
+figure~\ref{fig:pointercoro}. Now there are two sources of offered
 concurrent references: the ILP exploited by the processor, and the
@@
   \caption{Pointer chasing with coroutines, with one reference per
-    coroutine. Number of coroutines per core is $\textit{total concurrent references} / \textit{number of cores}$.}
+    coroutine.}
   \label{fig:multi-green}
@@
 second with 48 concurrent misses, or 8 coroutines per core. These
-results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36.
-When the number of references per coroutine is larger than 1, fewer coroutines are required to reach full performance.
+results are very close to the ILP-only experiment.
 
@@
 concurrent references per core exceeds 10; we believe this is due to
-later prefetches squashing earlier ones in the line fill buffers. We
-also observe some points that do not follow the trend and look like noise, but these results are actually repeatable. This effect is harder to explain, but we suspect it is
+later prefetches squashing earlier ones in the line fill buffer. We
+also observe some repeatable outliers below the maximum rate even with enough
+memory concurrency; this is harder to explain, but we suspect it is
 due to resource contention in the cores' pipelines once the memory
@@
 
-%Such points are repeatably lower, and looks like the std dev of such points (the diamond ones than are low) is also higher than other "normal" points: around 35M versus 13M or less, which seems to suggest a weird effect at those numbers}
+\begin{figure}
+\begin{alltt}
+  while (count-- > 0) \{
+    prefetch(&(list1->next));
+    switch();
+    list1 = read(&(list1->next));
+    for( i in 1 to num_local_updates ) \{
+      local->data++;
+      local = local->next;
+   \}
+  \}
+\end{alltt}
+\caption{Pseudocode for pointer chasing with local updates.}
+\label{pointerupdate}
+\end{figure}
 
@@
 to include random updates to a per-coroutine local data structure that
-is small enough to fit in cache. Figure~\ref{fig:code}c shows the
+is small enough to fit in cache. Figure~\ref{pointerupdate} shows the
 general idea.
@@
 
-\subsection{Multi-node}
-
-To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.
-First we measure the QPI bandwidth between the two chips of the test machine, since the chip would talk to the network interface over a QPI link.
-
-% take this figure out if we need the space
-% the main point is room for remote references to local memory
-\begin{figure}[h]
-  \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/qpi_bw-edited.pdf}
-  \end{center}
-  \caption{Pointer chasing in a remote processor's memory.}
-  \label{fig:listwalk-qpi}
-\end{figure}
-%%%%%%%%%%
-
-We measured the rate at which the cores can make requests
-over the QPI link. For this experiment, we allocated the lists in the
-second chip's memory and ran the traversals on the first chip's cores. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use of the QPI link limits us to 175 million references per second.
-
-To simulate the performance of pointer chasing in a
+Finally, to simulate the performance of pointer chasing in a
 multi-node system, we modified the scheduler of our coroutine library
@@
 latencies between two computers connected back-to-back; we assume a
-switch transit delay of 400 ns \todo{how many hops} in each direction and thus estimate a 3
+switch transit delay of 400 ns in each direction and thus estimate a 3
 $\mu$s round-trip delay. In this experiment, once a coroutine has
@@
 Figure~\ref{fig:network-delay}. We are still able to reach a
-near-maximum rate of 273 million references per second, but at least 2
+near-maximum rate of 273 million references per second, but at least 3
 concurrent references per coroutine are required. Running with a single
@@
 
-\todo{do we want to include that the memory system imposes max limit so really more throughput is still achievable if there are enough ILP memory concurrency?}\\
-\todo{also have results that remote cores over QPI plus delay 3us can make it up to the 175M, with sufficient number of coroutines}
+\todo{do we want to include that the memory system imposes max limit so really more throughput is still achievable if there are enough ILP memory concurrency?}
+\todo{may want to rerun with memory on remote core to incorporate QPI limit?}
 

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex

@@
   \}
+  \center{{\bf (a)}}
 \end{alltt}
-\caption{Pseudocode for pointer chasing without coroutines.}
+%\caption{Pseudocode for pointer chasing without coroutines.}
 \label{fig:pointernocoro}
-\end{figure}
+\end{minipage}
+%listwalk green threads
+\begin{minipage}[b]{0.35\linewidth}
+\centering
+\begin{alltt}
+  while (count-- > 0) \{
+     prefetch(&(list1->next));
+     switch();
+     list1 = read(&(list1->next));
+ \}
+  \center{{\bf (b)}}
+\end{alltt}
+  %%n lists
+  % while (count-- > 0) \{
+  % prefetch(&(list1->next));
+  %   prefetch(&(list2->next));
+   %  \ldots
+   %  prefetch(&(list\(n\)->next));
+   %  switch();
+   %  list1 = read(&(list1->next));
+   %  list2 = read(&(list2->next));
+   %  \ldots
+   %  list\(n\) = read(&(list\(n\)->next));
+%\caption{Pseudocode for pointer chasing using coroutines.}
+\label{fig:pointercoro}
+\end{minipage}
+%cache pressure
+\begin{minipage}[b]{0.32\linewidth}
+\centering
+\begin{alltt}
+  while (count-- > 0) \{
+    prefetch(&(list1->next));
+    switch();
+    list1 = read(&(list1->next));
+    for( i in 1 to num_local_updates ) \{
+      local->data++;
+      local = local->next;
+   \}
+  \}
+   \center{{\bf (c)}}
+\end{alltt}
+%\caption{Pseudocode for pointer chasing with local updates.}
+\label{fig:pointerupdate}
+\end{minipage}
+\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.
+Coroutine code, as in (b), might also include ILP like what (a) depends on for memory concurrency, e.g., by doing multiple prefetches to independent lists, switch, then multiple reads.}
+\label{fig:code}
+\end{figure*}
 
@@
 library, we modified our pointer chasing benchmark as shown in
-figure~\ref{fig:pointercoro}. Now there are two sources of offered
+Figure~\ref{fig:code}b.
+Recall that the baseline code in Figure~\ref{fig:code}a relied solely on ILP to achieve memory concurrency, which may not be abundant in real applications.
+Using coroutines, there are two sources of offered
 concurrent references: the ILP exploited by the processor, and the
@@
 latencies between two computers connected back-to-back; we assume a
-switch transit delay of 400 ns in each direction and thus estimate a 3
+switch transit delay of 400 ns \todo{how many hops? 1or3?} in each direction and thus estimate a 3
 $\mu$s round-trip delay. In this experiment, once a coroutine has
@@
 Figure~\ref{fig:network-delay}. We are still able to reach a
-near-maximum rate of 273 million references per second, but at least 3
+near-maximum rate of 273 million references per second, but at least 2
 concurrent references per coroutine are required. Running with a single
 reference per coroutine achieves 236 million references per
-second. Any network interface would be connected through a QPI link,
+second.
+Note that the throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support.
+This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.
+
+We also observe that to achieve a throughput equal to the expected max outgoing network bandwidth (100 million references per second), \todo{reference earlier section calculation?} when there is only one reference per context available, we require 320 coroutines.
+This is close to the minimum of 300 concurrent references needed to tolerate $3 \mu s$ latency.
+
+Any network interface would be connected through a QPI link,
 which we measured to have a bandwidth limit of 175 million references

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   softxmt-hotpar2011.tex

@@
   performance is not straightforward. Multithreading works well for
-  such problems, but large-scale implementations of the technique like
-  the Cray XMT are expensive, non-commodity machines.
+  such problems, but the only large-scale implementation of the technique
+  is the Cray XMT, an expensive, non-commodity machine.
 
@@
 memory requests to the non-coherent global memory shared across the
-nodes.  This memory manager is the only component that might require
-hardware, but we are exploring a mix of hardware and software
-implementations.
+nodes.  We are exploring a mix of hardware and software
+implementations of the memory manager.
 
@@
 
-\section{Related work}
-\label{sec:related}
-The concept of using independent parallel contexts to tolerate latency
-is not novel.  The inspiration for our project is of course the Cray
-XMT \cite{feo-xmt}, which implemented a hardware version of our
-programming model (in what would today be considered exceptionally
-slow silicon.)  The XMT can be seen as simultaneous multithreading
-\cite{tullsen-smt} writ large; the technique and motivation is
-essentially the same, but the XMT allows hundreds of contexts (modern
-Intel SMT implementations are generally limited to two.) The Alewife
-system \cite{agarwal-alewife} also switches contexts (in hardware) to
-hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
-allows hundreds of independently executing contexts to share FPUs,
-caches, and paths to memory, but does not context switch or share
-basic execution hardware.
-
-Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
-model under the hood;
-% Do I need to include a citation here?
-while they generate large numbers of available execution contexts via
-a (semi-)explicit SIMD model, those SIMD instructions keep functional
-units satisfied by overlapping execution with the loads required for
-other contexts.
-
-Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
-  multithreading}.  As opposed to our system of explicitly marked
-remote references, SCM assumes a high hit rate and traps to an
-interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
-short enough latencies that using more than two contexts generally
-produced slowdown.
-
-Partitioned global address space models, such as that used in Unified Parallel C,
-% definitely need a citation here
- share our approach of exposing all data (even remote data)
- transparently to any threads and have largely solved the problems of
- how to efficiently translate those remote reads into message
- passing.  However, their programming models emphasize locality more
- than our work does; while a program can examine any vertex in a UPC
- program at any time, it probably shouldn't (whereas our kernels rely on our
- latency hiding technique to make common remote references fast.
-
-The value of the information in extremely large and sparse graphs is
-rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
-sophisticated analysis on data from Twitter, identifying important
-actors and their networks of influence, and it can scale to the full
-size of graphs such as Facebook's friend network.
 \section{Programming model}
@@
 node private address spaces, just as in a conventional cache-coherent
-multiprocessor node.  In contrast, locality cannot be exploited in 
-the global space.
-
-\todo{explain why global address space is good}
+multiprocessor node.  In contrast, locality cannot be exploited in the
+global space: its value is in providing high bandwidth random access
+to any word in a shared data structure by any processor.
 
@@
 mechanism. First, the working set of each coroutine must be small, so
-many coroutines can be active without monopolizing the cache. Second,
+many coroutines can be active without trashing the cache. Second,
 context switches must be fast, so that overhead does not swamp actual
@@
 
+\section{Related work}
+\label{sec:related}
+The concept of using independent parallel contexts to tolerate latency
+is not novel.  The inspiration for our project is of course the Cray
+XMT \cite{feo-xmt}, which implemented a hardware version of our
+programming model (in what would today be considered exceptionally
+slow silicon.)  The XMT can be seen as simultaneous multithreading
+\cite{tullsen-smt} writ large; the technique and motivation is
+essentially the same, but the XMT allows hundreds of contexts (modern
+Intel SMT implementations are generally limited to two.) The Alewife
+system \cite{agarwal-alewife} also switches contexts (in hardware) to
+hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
+allows hundreds of independently executing contexts to share FPUs,
+caches, and paths to memory, but does not context switch or share
+basic execution hardware.
+
+Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
+model under the hood;
+% Do I need to include a citation here?
+while they generate large numbers of available execution contexts via
+a (semi-)explicit SIMD model, those SIMD instructions keep functional
+units satisfied by overlapping execution with the loads required for
+other contexts.
+
+Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
+  multithreading}.  As opposed to our system of explicitly marked
+remote references, SCM assumes a high hit rate and traps to an
+interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
+short enough latencies that using more than two contexts generally
+produced slowdown.
+
+Partitioned global address space models, such as that used in Unified Parallel C,
+% definitely need a citation here
+ share our approach of exposing all data (even remote data)
+ transparently to any threads and have largely solved the problems of
+ how to efficiently translate those remote reads into message
+ passing.  However, their programming models emphasize locality more
+ than our work does; while a program can examine any vertex in a UPC
+ program at any time, it probably shouldn't (whereas our kernels rely on our
+ latency hiding technique to make common remote references fast.
+
+The value of the information in extremely large and sparse graphs is
+rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
+sophisticated analysis on data from Twitter, identifying important
+actors and their networks of influence, and it can scale to the full
+size of graphs such as Facebook's friend network.
+
 \section{Conclusion}

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
added small fixes and references up to 3.2.1

@@
 commodity machine's memory. Their low diameter means that they are
-difficult to lay out with locality: every node needs to be near every
-other node. Their power law distribution means that a few vertices
+difficult to lay out with locality: every vertex needs to be near every
+other vertex. Their power law distribution means that a few vertices
 generate much more work than the rest, leading to computational
@@
 128 hardware contexts and 1024 outstanding memory operations, and is
-able to switch contexts every cycle. This ability comes at a cost: the
-XMT is an expensive, non-commodity machine.
+able to switch contexts every cycle \todo{cite XMT description}. This ability comes at a cost: the
+XMT is an expensive, non-commodity machine. \todo{don't we want to also say XMT has poor performance on x kind of apps?}
 
@@
 Whereas in the PGAS model coherency of global copies is implictly
-maintained across the system, we forego caching of globally shared
+maintained across the system \todo{cite pgas?}, we forego caching of globally shared
 data.  A node can exploit temporal locality of data stored in
@@
 locality within the private space, and it will be exploited to reduce
-latency and bandwidth; express concurrency, and it will be used toward
+latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
 tolerating latency and reducing the cost of bandwidth.
@@
 switching and scheduling in user space, as in \todo{cite other green
-  threads packages}.
+  threads packages, like qthreads}.
 
@@
 the core without disturbing the state of the cache
-hierarchy. \todo{cite intel docs}
+hierarchy \cite{intel:swdev}.
 
@@
 cover the latency of remote references in a multi-node
-system. Unfortunately, commodity processors have much smaller limits
-on memory concurrency: section \todo{cite evaluation} finds a limit of
+system\todo{ can we make a stronger statement? we know it will be 300$+$ according to our calculations}. Unfortunately, commodity processors have much smaller limits
+on memory concurrency: section~\ref{subsubsec:evalsinglebase} finds a limit of
 36 concurrent loads. We will have to manage memory concurrency on our
@@
 \subsubsection{Base memory system performance}
+\label{subsubsec:evalsinglebase}
 \todo{title might be improved}

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
editing and refs up to eval

@@
 proposal requires.
+\todo{but are we eliminating the possibility of FE-only storage on the FPGA?}
 
@@
 chips' memory are carried over Intel's cache-coherent QuickPath
-Interconnect (QPI).
+Interconnect (QPI) \cite{quickpath:website}.
 

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
branch for sigplan template

@@
 %% To disable, just uncomment this line
- %%\renewcommand{\todo}[1]{\relax}
+ \renewcommand{\todo}[1]{\relax}
 
-\documentclass{acm_proc_article-sp}
+%\documentclass{acm_proc_article-sp}
+\documentclass[10pt,nocopyrightspace,preprint]{sigplanconf}
 \usepackage{url}
@@
 \usepackage{alltt}
+\usepackage{graphicx}
 \usepackage{comment}
@@
 
-\numberofauthors{7}
-\author{
-Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
-Carl Ebeling, Luis Ceze\\
-University of Washington\\
-\email{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}\\ \\
-Simon Kahan \\ 
-Pacific Northwest National Lab\\
-\email{skahan@cs.washington.edu}\\ \\
-Preston Briggs\\ 
-Cray Inc.\\
-\email{preston.briggs@gmail.com}\\
-}
+%\numberofauthors{7}
+%\author{
+%Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
+%Carl Ebeling, Luis Ceze\\
+%University of Washington\\
+%\email{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}\\ \\
+%Simon Kahan \\ 
+%Pacific Northwest National Lab\\
+%\email{skahan@cs.washington.edu}\\ \\
+%Preston Briggs\\ 
+%Cray Inc.\\
+%\email{preston.briggs@gmail.com}\\
+%}
+\authorinfo{Mr Nobody}{University of Washington}{nobody.cs.washington.edu}
 
@@
 
-\todo{do we want to include that the memory system imposes max limit so really more throughput is still achievable if there are enough ILP memory concurrency?}\\
+\todo{do we want to include that the memory system imposes max limit so really more throughput is still achievable if there are enough ILP memory concurrency?}
 \todo{also have results that remote cores over QPI plus delay 3us can make it up to the 175M, with sufficient number of coroutines}

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
some more comments and fixes

--- 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 spread across multiple nodes. This goal has three key challenges:
-
 \begin{itemize}
@@
 \end{itemize}
-
 We discuss our approach to solving each of these challenges in
@@
 
+% cache pressure figure
 \begin{figure}[h]
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/cache-pressure-edited.pdf}
+    \includegraphics[width=0.42\textwidth]{figures/cache-pressure-edited.pdf}
   \end{center}
@@
 core.
+\todo{preempt questions about drop at 16 which is much less than the about 50 needed per core by saying that the drop is likely the same drop effect seen earlier, which is likely to be prefetch squashing since it happens at number of line fill buffers}
 
@@
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/qpi_bw-edited.pdf}
+    \includegraphics[width=0.42\textwidth]{figures/qpi_bw-edited.pdf}
   \end{center}
@@
 
+%simulated delay plot
 \begin{figure}[h]
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/multi-green-delay7800-edited.pdf}
+    \includegraphics[width=0.47\textwidth]{figures/multi-green-delay7800-edited.pdf}
   \end{center}

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
fixed authors

--- 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 %}
-\authorinfo{Mr Nobody}{University of Washington}{nobody.cs.washington.edu}
+
+%todo format the UW names horizontally
+\authorinfo{Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
+Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
+\authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
+\authorinfo{Preston Briggs}{Cray Inc.}{preston.briggs@gmail.com}
 

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
comments

@@
 %% To disable, just uncomment this line
- \renewcommand{\todo}[1]{\relax}
+%% \renewcommand{\todo}[1]{\relax}
 
@@
 	\begin{center}
-		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-totalconc-edited.pdf}
+		\includegraphics[width=0.47\textwidth]{figures/multi-listwalk-totalconc-edited.pdf}
 	\end{center}
@@
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/multi-green-edited.pdf}
+    \includegraphics[width=0.47\textwidth]{figures/multi-green-edited.pdf}
   \end{center}
@@
 
-\todo{do we want to include that the memory system imposes max limit so really more throughput is still achievable if there are enough ILP memory concurrency?}
 \todo{also have results that remote cores over QPI plus delay 3us can make it up to the 175M, with sufficient number of coroutines}

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   ../greenery/Makefile

@@
 \authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
-\authorinfo{Preston Briggs}{Cray Inc.}{preston.briggs@gmail.com}
+\authorinfo{Preston Briggs}{}{preston.briggs@gmail.com}
 
@@
   Our goal is to build a system with XMT-like properties, implemented
-  with commodity processors. This paper presents a runtime for latency
+  with commodity components. This paper presents a runtime for latency
   tolerance, along with our plans for scaling up to multiple
@@
 low-diameter, power law graphs: this combination makes extracting
-performance difficult. Their size keeps them from fitting in a single
+performance difficult. They do not fit in a single
 commodity machine's memory. Their low diameter means that they are
@@
 other vertex. Their power law distribution means that a few vertices
-generate much more work than the rest, leading to computational
-imbalance and even less locality. Together, these properties lead to a
-difficult conclusion: each edge traversal is likely to be a very
-expensive cache miss.
+generate much more work than the rest, leading to load
+imbalances. Together, these properties lead to a
+difficult conclusion: each edge traversal is likely to induce a very
+expensive cache miss or remote access.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
Mref/s

@@
 %% \renewcommand{\todo}[1]{\relax}
+
+\newcommand{\mrps}[0]{Mref/s}
 
@@
 the maximum rate pointers are traversed for a given number of
-concurrent offered references. We see a maximum rate of 277 million
-references per second, which agrees with the measurements found by \cite{Mandal:2010} with a similar machine.
+concurrent offered references.
+We see a maximum rate of 277 million references per second (\mrps), which agrees with the measurements found by \cite{Mandal:2010} with a similar machine.
 This rate is achieved when the number of offered references is 36.
@@
 memory. With this configuration, we observed a maximum rate of 360
-million references per second. We believe this difference is due to
+\mrps. We believe this difference is due to
 the configuration of the processor's {\em Global Queue} \cite{nehalem:perf}, which limits the number of concurrently-executing read misses from a chips cores to its local last-level cache and memory to 32.
@@
 coroutine, so the only source of memory concurrency is the use of
-coroutines. We are able to obtain a rate of 275 million references per
-second with 48 concurrent misses, or 8 coroutines per core. These
+coroutines. We are able to obtain a rate of 275 \mrps with 48 concurrent misses, or 8 coroutines per core. These
 results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36.
@@
 over the QPI link. For this experiment, we allocated the lists in the
-second chip's memory and ran the traversals on the first chip's cores. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use of the QPI link limits us to 175 million references per second.
+second chip's memory and ran the traversals on the first chip's cores. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use of the QPI link limits us to 175 \mrps.
 
@@
 Figure~\ref{fig:network-delay}. We are still able to reach a
-near-maximum rate of 273 million references per second, but at least 2
+near-maximum rate of 273 \mrps, but at least 2
 concurrent references per coroutine are required. Running with a single
-reference per coroutine achieves 236 million references per
-second.
+reference per coroutine achieves 236 \mrps.
 Note that the throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support.
@@
 
-We also observe that to achieve a throughput equal to the expected max outgoing network bandwidth (100 million references per second), \todo{reference earlier section calculation?} when there is only one reference per context available, we require 320 coroutines.
+We also observe that to achieve a throughput equal to the expected max outgoing network bandwidth (100 \mrps), \todo{reference earlier section calculation?} when there is only one reference per context available, we require 320 coroutines.
 This is close to the minimum of 300 concurrent references needed to tolerate $3 \mu s$ latency.
@@
 Any network interface would be connected through a QPI link,
-which we measured to have a bandwidth limit of 175 million references
-per second, so our runtime appears to have the capacity to chase pointers
+which we measured to have a bandwidth limit of 175 \mrps, so our runtime appears to have the capacity to chase pointers
 in a remote memory.

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   softxmt-hotpar2011.tex

--- 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \label{sec:approach}
-
 The goal of our implementation is to use multithreading to tolerate
@@
   We expect to need hundreds of
-  threads to tolerated a cluster's network latency, so our threading
-  implementation must have low overhead.
+  threads to tolerate network latency, so our threading
+  implementation must have low memory and context switching overheads.
 
@@
 
-  A thread making a long-latency global memory reference must be able
-  to yield and allow other threads to make progress.
+  We will need to keep track of the hundreds of outstanding memory fetch operations required to utilize bandwidth well.
 
@@
 
-  In a system with high degrees of concurrency, lightweight
-  synchronization primitives are important.
+  We must support global synchronization that can proceed without stalling the processing cores.
+
 \end{itemize}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
rewritten abstract

--- 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 %todo format the UW names horizontally
-\authorinfo{Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
+\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Dan Grossman, Mark Oskin,
 Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
@@
 
- Techniques from graph analytics have found important applications in
-  areas such as social networks and bioinformatics. These irregular
-  problems present a challenge for parallel machines: obtaining
-  performance is not straightforward. Multithreading works well for
-  such problems, but the only large-scale implementation of the technique
-  is the Cray XMT, an expensive, non-commodity machine.
+  Large, sparse, irregular graphs map poorly to the memory systems of
+  commodity hardware. The scale of emerging real world datasets
+  demands new techniques; especially in the parallel case,
+  straightforward algorithms will run inefficiently due to many
+  high-latency accesses.  Pervasive parallelism can tolerate that
+  latency, but the only large-scale implementation of the technique is
+  the Cray XMT, an expensive, non-commodity machine.
 
-  Our goal is to build a system with XMT-like properties, implemented
-  with commodity components. This paper presents a runtime for latency
-  tolerance, along with our plans for scaling up to multiple
-  nodes. Our analysis shows the feasibility of our approach.
+  If we can adapt the XMT's programming model to commodity hardware,
+  we can dramatically improve performance without resorting to
+  expensive custom systems. This paper presents a runtime for latency
+  tolerance through the use of parallel lightweight contexts, along
+  with our plans for scaling up to multiple nodes. Our analysis shows
+  the feasibility of our approach.
 \end{abstract}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
rewritten abstract (mark 2)

@@
 \begin{abstract}
-
   Large, sparse, irregular graphs map poorly to the memory systems of
   commodity hardware. The scale of emerging real world datasets
-  demands new techniques; especially in the parallel case,
-  straightforward algorithms will run inefficiently due to many
-  high-latency accesses.  Pervasive parallelism can tolerate that
-  latency, but the only large-scale implementation of the technique is
-  the Cray XMT, an expensive, non-commodity machine.
+  presents a challenge for straightforward algorithms which will run
+  inefficiently due to many high-latency accesses.  Pervasive
+  parallelism can tolerate that latency, but the only large-scale
+  implementation of the technique is the Cray XMT, an expensive,
+  non-commodity machine.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
reworked related work

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 %todo format the UW names horizontally
-\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Dan Grossman, Mark Oskin,
+\authorinfo{Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
 Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
 \authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
-\authorinfo{Preston Briggs}{}{preston.briggs@gmail.com}
+\authorinfo{Preston Briggs}{Cray Inc.}{preston.briggs@gmail.com}
 
@@
 
-  Large, sparse, irregular graphs map poorly to the memory systems of
-  commodity hardware. The scale of emerging real world datasets
-  demands new techniques; especially in the parallel case,
-  straightforward algorithms will run inefficiently due to many
-  high-latency accesses.  Pervasive parallelism can tolerate that
-  latency, but the only large-scale implementation of the technique is
-  the Cray XMT, an expensive, non-commodity machine.
+ Techniques from graph analytics have found important applications in
+  areas such as social networks and bioinformatics. These irregular
+  problems present a challenge for parallel machines: obtaining
+  performance is not straightforward. Multithreading works well for
+  such problems, but the only large-scale implementation of the technique
+  is the Cray XMT, an expensive, non-commodity machine.
 
-  If we can adapt the XMT's programming model to commodity hardware,
-  we can dramatically improve performance without resorting to
-  expensive custom systems. This paper presents a runtime for latency
-  tolerance through the use of parallel lightweight contexts, along
-  with our plans for scaling up to multiple nodes. Our analysis shows
-  the feasibility of our approach.
+  Our goal is to build a system with XMT-like properties, implemented
+  with commodity processors. This paper presents a runtime for latency
+  tolerance, along with our plans for scaling up to multiple
+  nodes. Our analysis shows the feasibility of our approach.
 \end{abstract}
@@
 low-diameter, power law graphs: this combination makes extracting
-performance difficult. They do not fit in a single
+performance difficult. Their size keeps them from fitting in a single
 commodity machine's memory. Their low diameter means that they are
@@
 other vertex. Their power law distribution means that a few vertices
-generate much more work than the rest, leading to load
-imbalances. Together, these properties lead to a
-difficult conclusion: each edge traversal is likely to induce a very
-expensive cache miss or remote access.
+generate much more work than the rest, leading to computational
+imbalance and even less locality. Together, these properties lead to a
+difficult conclusion: each edge traversal is likely to be a very
+expensive cache miss.
 
@@
 
-%\todo{Begin eval insert}
-%\input{evaluation}
-%\todo{End eval insert}
-
 \section{Related work}
 \label{sec:related}
-The concept of using independent parallel contexts to tolerate latency
-is not novel.  The inspiration for our project is of course the Cray
-XMT \cite{feo-xmt}, which implemented a hardware version of our
-programming model (in what would today be considered exceptionally
-slow silicon.)  The XMT can be seen as simultaneous multithreading
-\cite{tullsen-smt} writ large; the technique and motivation is
-essentially the same, but the XMT allows hundreds of contexts (modern
-Intel SMT implementations are generally limited to two.) The Alewife
-system \cite{agarwal-alewife} also switches contexts (in hardware) to
-hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
-allows hundreds of independently executing contexts to share FPUs,
-caches, and paths to memory, but does not context switch or share
-basic execution hardware.
 
-Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
-model under the hood;
-% Do I need to include a citation here?
-while they generate large numbers of available execution contexts via
-a (semi-)explicit SIMD model, those SIMD instructions keep functional
-units satisfied by overlapping execution with the loads required for
-other contexts.
+Much work exists on using multithreading to tolerate latency. Hardware
+implementations include the Tera MTA~\cite{tera} and Cray
+XMT~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
+MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
+even GPUs \cite{gpus}. Software-only approaches exist as well; the
+Software Controlled Multithreading project \cite{mowry-scm}, QThreads
+\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
+use variants of this idea.
+%In contrast to all of these, our goal is to tolerate
+%latencies across a multi-node system using mostly software.
 
-Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
-  multithreading}.  As opposed to our system of explicitly marked
-remote references, SCM assumes a high hit rate and traps to an
-interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
-short enough latencies that using more than two contexts generally
-produced slowdown.
+Our goal of presenting a global view of distributed memory to the
+programmer is shared by the PGAS community, and is used in languages
+like Chapel \cite{chapel}, X10 \cite{X10}, and UPC \cite{upc}. They
+get performance by minimizing references to remote nodes, whereas we
+optimize for this case.
 
-Partitioned global address space models, such as that used in Unified Parallel C,
-% definitely need a citation here
- share our approach of exposing all data (even remote data)
- transparently to any threads and have largely solved the problems of
- how to efficiently translate those remote reads into message
- passing.  However, their programming models emphasize locality more
- than our work does; while a program can examine any vertex in a UPC
- program at any time, it probably shouldn't (whereas our kernels rely on our
- latency hiding technique to make common remote references fast.
-
-The value of the information in extremely large and sparse graphs is
-rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
-sophisticated analysis on data from Twitter, identifying important
-actors and their networks of influence, and it can scale to the full
-size of graphs such as Facebook's friend network.
+The idea of processing large graphs on distributed machines has been
+exploredby projects such as Pregel \cite{pregel}, the Parallel Boost
+Graph Library \cite{parallelbgl}, and CGMLib \cite{cgmlib}. Our goal
+is to develop infrastructure to aid implementation of similar
+libraries.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
condensed eval up thru base system performance, condesed captions whitespace, fixed 3 codes figure

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \usepackage{comment}
+\usepackage[labelfont=bf,font=small,belowskip=-7pt,aboveskip=-10pt]{caption}
 \begin{document}
@@
 %todo format the UW names horizontally
-\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Dan Grossman, Mark Oskin,
+\authorinfo{Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
 Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
@@
 \begin{abstract}
-  Large, sparse, irregular graphs map poorly to the memory systems of
-  commodity hardware. The scale of emerging real world datasets
-  presents a challenge for straightforward algorithms which will run
-  inefficiently due to many high-latency accesses.  Pervasive
-  parallelism can tolerate that latency, but the only large-scale
-  implementation of the technique is the Cray XMT, an expensive,
-  non-commodity machine.
 
-  If we can adapt the XMT's programming model to commodity hardware,
-  we can dramatically improve performance without resorting to
-  expensive custom systems. This paper presents a runtime for latency
-  tolerance through the use of parallel lightweight contexts, along
-  with our plans for scaling up to multiple nodes. Our analysis shows
-  the feasibility of our approach.
+ Techniques from graph analytics have found important applications in
+  areas such as social networks and bioinformatics. These irregular
+  problems present a challenge for parallel machines: obtaining
+  performance is not straightforward. Multithreading works well for
+  such problems, but the only large-scale implementation of the technique
+  is the Cray XMT, an expensive, non-commodity machine.
+
+  Our goal is to build a system with XMT-like properties, implemented
+  with commodity components. This paper presents a runtime for latency
+  tolerance, along with our plans for scaling up to multiple
+  nodes. Our analysis shows the feasibility of our approach.
 \end{abstract}
@@
 \label{sec:approach}
+
 The goal of our implementation is to use multithreading to tolerate
@@
   We expect to need hundreds of
-  threads to tolerate network latency, so our threading
-  implementation must have low memory and context switching overheads.
+  threads to tolerated a cluster's network latency, so our threading
+  implementation must have low overhead.
 
@@
 
-  We will need to keep track of the hundreds of outstanding memory fetch operations required to utilize bandwidth well.
+  A thread making a long-latency global memory reference must be able
+  to yield and allow other threads to make progress.
 
@@
 
-  We must support global synchronization that can proceed without stalling the processing cores.
-
+  In a system with high degrees of concurrency, lightweight
+  synchronization primitives are important.
 \end{itemize}
@@
 Our evaulation consists of two parts.
-First, we demonstrate that the runtime system can achieve the same performance as when there is explicit memory concurrency available: we measure pointer chasing performance and
-the effects of cache pressure due to the number of contexts required.
-Second, we look at the runtime within the multi-node picture and simulate the effects of fetching memory over a network to show that the runtime can reach the level of concurrency required to hide the latency.
+First, we demonstrate that the runtime system can achieve the same performance as when there is explicit memory concurrency available.
+Second, we look at the runtime within the multi-node picture and show that it can tolerate large latencies.
 In each part we begin by studying relevant aspects of our test machine's memory system so we can interpret our runtime results.
@@
 \label{subsubsec:evalsinglebase}
-\todo{title might be improved}
 
@@
 rate that can be issued by the cores in one chip and the maximum random
-reference rate that can be serviced by one chip's memory controller.\todo{The 360M  "serviced" number could perhaps be an "aside" kind of paragraph since not supporting a major point}
+reference rate that can be serviced by one chip's memory controller.
 
@@
 \centering
-\begin{alltt}
+\begin{alltt}{\small
   while (count-- > 0) \{
@@
   \}
-  \center{{\bf (a)}}
+  }\centering{\bf (a)}
 \end{alltt}
@@
  \}
-  \center{{\bf (b)}}
+ }\centering{\bf (b)}
 \end{alltt}
@@
       local = local->next;
-   \}
-  \}
-   \center{{\bf (c)}}
+   \} \}
+   }\centering{\bf (c)}
 \end{alltt}
@@
 \end{minipage}
+\vspace{10pt}
 \caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.
@@
 
-The lists were sized to exceed the last level cache and were laid out
-randomly with pointers spaced at a cache line granularity, maximizing
+The lists were laid out randomly with pointers spaced at a cache line granularity, maximizing
 the probability of each reference being a miss. We allocated the lists
-on 1GB huge pages to minimize TLB refill overhead. The lists were
-allocated in the memory attached to the processor containing the cores
-doing the traversal. We ran 10 traversals, measuring average rate in each, and recorded the maximum rate.
+on 1GB huge pages to minimize TLB refill overhead.
 
@@
 the maximum rate pointers are traversed for a given number of
-concurrent offered references. We see a maximum rate of 277 million
-references per second, which agrees with the measurements found by \cite{Mandal:2010} with a similar machine.
+concurrent offered references.
+We see a maximum rate of 277 million references per second (\mrps), which agrees with the measurements found by \cite{Mandal:2010}.
 This rate is achieved when the number of offered references is 36.
@@
 memory concurrency; the maximum reference rate for a single core is
-107M. We believe this limit is due to the core's limited number of {\em line fill
-  buffers} \cite{nehalem:arch}, which limits the core to 10
-concurrent private cache misses.
+107M. We believe this limit is due to the core having only enough {\em line fill
+  buffers} \cite{nehalem:arch} to store 10 concurrent private cache misses.
 
@@
 memory. With this configuration, we observed a maximum rate of 360
-million references per second. We believe this difference is due to
-the configuration of the processor's {\em Global Queue} \cite{nehalem:perf}, which limits the number of concurrently-executing read misses from a chips cores to its local last-level cache and memory to 32.
+\mrps.
+We believe this difference is due to the chip having only 32 buffers in its {\em Global Queue} \cite{nehalem:perf} for tracking concurrently-executing read misses from all the cores' private caches.
 
-\subsubsection{Runtime performance}
+\subsubsection{Coroutine performance}
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
one word change

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \begin{abstract}
+  Large, sparse, irregular graphs map poorly to the memory systems of
+  commodity hardware. The scale of emerging real world datasets
+  presents a challenge for straightforward algorithms which will run
+  inefficiently due to many high-latency accesses.  Pervasive
+  parallelism can tolerate that latency, but the only large-scale
+  implementation of the technique is the Cray XMT, an expensive,
+  non-commodity machine.
 
- Techniques from graph analytics have found important applications in
-  areas such as social networks and bioinformatics. These irregular
-  problems present a challenge for parallel machines: obtaining
-  performance is not straightforward. Multithreading works well for
-  such problems, but the only large-scale implementation of the technique
-  is the Cray XMT, an expensive, non-commodity machine.
-
-  Our goal is to build a system with XMT-like properties, implemented
-  with commodity components. This paper presents a runtime for latency
-  tolerance, along with our plans for scaling up to multiple
-  nodes. Our analysis shows the feasibility of our approach.
+  If we can adapt the XMT's programming model to commodity hardware,
+  we can dramatically improve performance without resorting to
+  expensive custom systems. This paper presents a runtime for latency
+  tolerance through the use of parallel lightweight contexts, along
+  with our plans for scaling up to multiple nodes. Our analysis shows
+  the feasibility of our approach.
 \end{abstract}
@@
 like Chapel \cite{chapel}, X10 \cite{X10}, and UPC \cite{upc}. They
-get performance by minimizing references to remote nodes, whereas we
+obtain performance by minimizing references to remote nodes, whereas we
 optimize for this case.

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
edits to intro and sec 2

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-Techniques from graph analytics have found important applications in
-areas such as social networks and bioinformatics. \todo{more!}
+Many important emerging applications such as social network analysis,
+bioinformatics and analysing vast sensor data from actual networks
+(e.g, water, electrical grid) are based on crunching very large
+graphs. Unfortunately, the computational cost of these applications
+gets quickly out of hand. For example, in January 2011, Facebook had
+500 million active users with an average of 130 friends each
+\cite{Facebook:2011p91. Its friend suggestion algorithm in 2010 ran on
+  a 40-node cluster with 72GB memory per node, and still, new
+  suggestions could only be computed every two days
+  \cite{Backstrom:2010p90}. Speeding up these applications at a low
+  cost would have a large impact in how we deal and react to the fast
+  growth of the data-set and would make the results even more
+  valuable.
+
+% Techniques from graph analytics have found important applications in
+% areas such as social networks and bioinformatics. \todo{more!}
 
@@
 
-The most interesting computational challenge comes from large,
-low-diameter, power law graphs: this combination makes extracting
-performance difficult. They do not fit in a single
+In this context, the most interesting computational challenge comes
+from large, low-diameter, power law graphs: this combination makes
+extracting performance difficult. They do not fit in a single
 commodity machine's memory. Their low diameter means that they are
-difficult to lay out with locality: every vertex needs to be near every
-other vertex. Their power law distribution means that a few vertices
-generate much more work than the rest, leading to load
-imbalances. Together, these properties lead to a
-difficult conclusion: each edge traversal is likely to induce a very
-expensive cache miss or remote access.
+difficult to lay out with locality: every vertex needs to be near
+every other vertex. Their power law distribution means that a few
+vertices generate much more work than the rest, leading to load
+imbalances. Together, these properties lead to a difficult
+conclusion: each edge traversal is likely to lead to expensive
+inter-node communication. %cache miss or remote access.
 
@@
 implement efficient computations for these graphs. The Cray XMT is an
-example of such an approach: it solves the memory latency problem
-through concurrency rather than caching. Each XMT processor supports
-128 hardware contexts and 1024 outstanding memory operations, and is
-able to switch contexts every cycle \todo{cite XMT description}. This ability comes at a cost: the
-XMT is an expensive, non-commodity machine. \todo{don't we want to also say XMT has poor performance on x kind of apps?}
+example of such an approach: it solves the memory latency problem (or
+communication problem) through concurrency rather than caching. Each
+XMT processor supports 128 hardware contexts and 1024 outstanding
+memory operations, and is able to switch contexts every cycle.
+% \todo{cite XMT description}.  %% LUIS: there is enough information here.
+This ability comes at a cost: the XMT is an expensive, non-commodity
+machine. Moreover, arguably a lot of XMT's advantages for these type
+of applications come at the cost of performance in general-purpose
+applications.
+
+% \todo{don't we want to also say XMT has poor performance on x
+%   kind of apps?}
 
@@
 
-We want the best of both worlds. Our goal is to build a system that
-has good performance on these low-locality graph codes but is
-implemented with cheap commodity processors.
+%We want the best of both worlds. 
+We believe we can build a system mostly based on commodity parts that
+can attain XMT-like performance with a familiar, also XMT-like,
+programming model, but at a fraction of the cost. Therefore, our goal
+is to build a system that has good performance on low-locality graph
+codes but is implemented with cheap commodity processors and simple
+additional support in an FPGA. This approach also has the added
+benefit of not sacrificing general-purpose performance.
 
@@
 memory requests to the non-coherent global memory shared across the
-nodes.  We are exploring a mix of hardware and software
+nodes.  We are exploring a mix of hardware (FPGA) and software
 implementations of the memory manager.
 
-This paper explores the feasibility of our multi-node goal using a
-single-node implementation of our lightweight threading runtime. We
-use prefetch instructions together with lightweight threads to
-tolerate the latency of DRAM access on the local node node. We use
-pointer chasing benchmarks to verify our runtime's overhead is
-acceptable. We also model the effects of pointer chasing on a remote
-node.
+This paper describes the vision of our system and explores the
+feasibility of our ideas using a single-node implementation of a
+lightweight threading runtime. We use prefetch instructions together
+with lightweight threads to tolerate the latency of DRAM access on the
+local node node. We use adversarial pointer chasing benchmarks to
+verify our runtime's overhead is acceptable. We also model the effects
+of pointer chasing on a remote node by artificially increasing
+latencys to the order of several micro-seconds.
 
-The rest of the paper is organized as follows. We describe our
-proposal from a programmer's viewpoint in section~\ref{sec:model}. We
-describe the implementation of our latency-tolerant runtime and plans
-for extending to multiple nodes in section~\ref{sec:approach}. We
-evaluate our runtime in section~\ref{sec:evaluation}.  We present
-related work in section~\ref{sec:related} and conclude in
+The rest of the paper is organized as follows. We quickly describe our
+programming model in section~\ref{sec:model}. We describe the
+implementation of our latency-tolerant runtime and plans for extending
+to multiple nodes in section~\ref{sec:approach}. We evaluate our
+runtime in section~\ref{sec:evaluation}.  We present related work in
+section~\ref{sec:related} and conclude in
 section~\ref{sec:conclusion}.
@@
 
-Our shared-memory programming model partitions the overall address
+Our goal is to preserve the XMT programming model: large shared global
+address space, explicit concurrency with a large number of threads,
+and full-empty-bits synchronization.  We partition the overall address
 space into a global shared address space and per node private address
@@
 
-Whereas in the PGAS model coherency of global copies is implictly
-maintained across the system \todo{cite pgas?}, we forego caching of globally shared
-data.  A node can exploit temporal locality of data stored in
-the global space only by first copying it into the node's local space.
-In this way, a node can access data at random across the system while
+% Whereas in the PGAS model coherency of global copies is implictly
+% maintained across the system, % \todo{cite pgas?}
+% we completely forego
+% caching of globally shared data.
+
+A node can exploit temporal locality of data stored in the global
+space only by first copying it into the node's local space.  In this
+way, a node can access data at random across the system while
 incurring no hardware coherence overhead: consistency is enforced by
@@
 concurrency.  Exactly how that concurrency is expressed by the
-programmer is language dependent.  Our programming model serves to
-augment threaded programming languages so that they become latency
-tolerant on large address spaces.  Software multithreading is used to
-provide latency tolerance: a computation that is about to execute a
-long latency operation such as a remote memory reference initiates the
-operation and then yields to the scheduler, which quickly resumes
-execution of another computation.  Concurrency is also used -- via
-aggregation of requests to a common node destination -- to make more
-efficient use of network bandwidth and in some cases reduce bandwidth
-demands.
+programmer is language dependent. The goal of our system is to support
+that concurrency and consequently provide latency tolerant on a large
+address space. We do so via software multithreading: computation that
+is about to execute a long latency operation (e.g. a remote memory
+reference) initiates the operation and then yields to the scheduler,
+which quickly resumes execution of another computation.  Concurrency
+is also used -- via aggregation of requests to a common node
+destination -- to make more efficient use of network bandwidth and in
+some cases reduce bandwidth demands.
 
@@
 works differently: we provide atomic operations such as {\tt
-  int\_fetch\_add} as well as operations using full-bits, as on the
-Cray XMT.  As is true for other long latency operations,
+  int\_fetch\_add} as well as operations using full-empty-bits, as on
+the Cray XMT.  As is true for other long latency operations,
 synchronization latency is tolerated by yielding to the scheduler.
@@
 
-Thus, the programming model provides the programmer with options: express
-locality within the private space, and it will be exploited to reduce
-latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
-tolerating latency and reducing the cost of bandwidth.
+%% I dont think we need the paragraph below.
+
+% Thus, the programming model provides the programmer with options: express
+% locality within the private space, and it will be exploited to reduce
+% latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
+% tolerating latency and reducing the cost of bandwidth.
 
@@
 
+%\todo{Begin eval insert}
+%\input{evaluation}
+%\todo{End eval insert}
+
 \section{Related work}
 \label{sec:related}
+The concept of using independent parallel contexts to tolerate latency
+is not novel.  The inspiration for our project is of course the Cray
+XMT \cite{feo-xmt}, which implemented a hardware version of our
+programming model (in what would today be considered exceptionally
+slow silicon.)  The XMT can be seen as simultaneous multithreading
+\cite{tullsen-smt} writ large; the technique and motivation is
+essentially the same, but the XMT allows hundreds of contexts (modern
+Intel SMT implementations are generally limited to two.) The Alewife
+system \cite{agarwal-alewife} also switches contexts (in hardware) to
+hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
+allows hundreds of independently executing contexts to share FPUs,
+caches, and paths to memory, but does not context switch or share
+basic execution hardware.
 
-Much work exists on using multithreading to tolerate latency. Hardware
-implementations include the Tera MTA~\cite{tera} and Cray
-XMT~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
-MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
-even GPUs \cite{gpus}. Software-only approaches exist as well; the
-Software Controlled Multithreading project \cite{mowry-scm}, QThreads
-\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
-use variants of this idea.
-%In contrast to all of these, our goal is to tolerate
-%latencies across a multi-node system using mostly software.
+Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
+model under the hood;
+% Do I need to include a citation here?
+while they generate large numbers of available execution contexts via
+a (semi-)explicit SIMD model, those SIMD instructions keep functional
+units satisfied by overlapping execution with the loads required for
+other contexts.
 
-Our goal of presenting a global view of distributed memory to the
-programmer is shared by the PGAS community, and is used in languages
-like Chapel \cite{chapel}, X10 \cite{X10}, and UPC \cite{upc}. They
-get performance by minimizing references to remote nodes, whereas we
-optimize for this case.
+Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
+  multithreading}.  As opposed to our system of explicitly marked
+remote references, SCM assumes a high hit rate and traps to an
+interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
+short enough latencies that using more than two contexts generally
+produced slowdown.
 
-The idea of processing large graphs on distributed machines has been
-exploredby projects such as Pregel \cite{pregel}, the Parallel Boost
-Graph Library \cite{parallelbgl}, and CGMLib \cite{cgmlib}. Our goal
-is to develop infrastructure to aid implementation of similar
-libraries.
+Partitioned global address space models, such as that used in Unified Parallel C,
+% definitely need a citation here
+ share our approach of exposing all data (even remote data)
+ transparently to any threads and have largely solved the problems of
+ how to efficiently translate those remote reads into message
+ passing.  However, their programming models emphasize locality more
+ than our work does; while a program can examine any vertex in a UPC
+ program at any time, it probably shouldn't (whereas our kernels rely on our
+ latency hiding technique to make common remote references fast.
+
+The value of the information in extremely large and sparse graphs is
+rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
+sophisticated analysis on data from Twitter, identifying important
+actors and their networks of influence, and it can scale to the full
+size of graphs such as Facebook's friend network.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex

@@
 \begin{abstract}
+% Techniques from graph analytics have found important applications in
+% areas such as social networks and bioinformatics. These irregular
+% problems present a challenge for parallel machines: obtaining
+% performance is not straightforward. Multithreading works well for
+% such problems, but the only large-scale implementation of the technique
+% is the Cray XMT, an expensive, non-commodity machine.
 
-Techniques from graph analytics have found important applications in
-areas such as social networks and bioinformatics. These irregular
-problems present a challenge for parallel machines: obtaining
-performance is not straightforward. Multithreading works well for
-such problems, but the only large-scale implementation of the technique
-is the Cray XMT, an expensive, non-commodity machine.
+% Our goal is to build a system with XMT-like properties, implemented
+% with commodity components. This paper presents a runtime for latency
+% tolerance, along with our plans for scaling up to multiple
+% nodes. Our analysis shows the feasibility of our approach.
 
-Our goal is to build a system with XMT-like properties, implemented
-with commodity components. This paper presents a runtime for latency
-tolerance, along with our plans for scaling up to multiple
-nodes. Our analysis shows the feasibility of our approach.
+  Large, sparse, irregular graphs map poorly to the memory systems of
+  commodity hardware. The scale of emerging real world datasets
+  presents a challenge for straightforward algorithms which will run
+  inefficiently due to many high-latency accesses.  Pervasive
+  parallelism can tolerate that latency, but the only large-scale
+  implementation of the technique is the Cray XMT, an expensive,
+  non-commodity machine.
+
+  If we can adapt the XMT's programming model to commodity hardware,
+  we can dramatically improve performance without resorting to
+  expensive custom systems. This paper presents a runtime for latency
+  tolerance through the use of parallel lightweight contexts, along
+  with our plans for scaling up to multiple nodes. Our analysis shows
+  the feasibility of our approach.
 \end{abstract}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
authors

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \begin{document}
+
+%\textheight 9in
+%\textwidth 6.5in
+%\columnsep 1pc
+\renewcommand{\baselinestretch}{0.9}
 
@@
 %todo format the UW names horizontally
-\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Dan Grossman, Mark Oskin,
-Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
-\authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
-\authorinfo{Preston Briggs}{}{preston.briggs@gmail.com}
+\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
+  Dan Grossman, Mark Oskin, Carl Ebeling, Simon Kahan, Luis
+  Ceze}{University of Washington}{\{nelson, bdmyers, ahh, preston,
+  djg, oskin, ebeling, skahan, luisceze\}@cs.washington.edu}
 
 \maketitle
+
 \begin{abstract}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-%\todo{Begin eval insert}
-%\input{evaluation}
-%\todo{End eval insert}
-
 \section{Related work}
 \label{sec:related}
-The concept of using independent parallel contexts to tolerate latency
-is not novel.  The inspiration for our project is of course the Cray
-XMT \cite{feo-xmt}, which implemented a hardware version of our
-programming model (in what would today be considered exceptionally
-slow silicon.)  The XMT can be seen as simultaneous multithreading
-\cite{tullsen-smt} writ large; the technique and motivation is
-essentially the same, but the XMT allows hundreds of contexts (modern
-Intel SMT implementations are generally limited to two.) The Alewife
-system \cite{agarwal-alewife} also switches contexts (in hardware) to
-hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
-allows hundreds of independently executing contexts to share FPUs,
-caches, and paths to memory, but does not context switch or share
-basic execution hardware.
 
-Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
-model under the hood;
-% Do I need to include a citation here?
-while they generate large numbers of available execution contexts via
-a (semi-)explicit SIMD model, those SIMD instructions keep functional
-units satisfied by overlapping execution with the loads required for
-other contexts.
+Much work exists on using multithreading to tolerate latency. Hardware
+implementations include the Tera MTA~\cite{tera} and Cray
+XMT~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
+MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
+even GPUs \cite{gpus}. Software-only approaches exist as well; the
+Software Controlled Multithreading project \cite{mowry-scm}, QThreads
+\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
+use variants of this idea.
+%In contrast to all of these, our goal is to tolerate
+%latencies across a multi-node system using mostly software.
 
-Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
-  multithreading}.  As opposed to our system of explicitly marked
-remote references, SCM assumes a high hit rate and traps to an
-interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
-short enough latencies that using more than two contexts generally
-produced slowdown.
+Our goal of presenting a global view of distributed memory to the
+programmer is shared by the PGAS community, and is used in languages
+like Chapel \cite{chapel}, X10 \cite{X10}, and UPC \cite{upc}. They
+obtain performance by minimizing references to remote nodes, whereas we
+optimize for this case.
 
-Partitioned global address space models, such as that used in Unified Parallel C,
-% definitely need a citation here
- share our approach of exposing all data (even remote data)
- transparently to any threads and have largely solved the problems of
- how to efficiently translate those remote reads into message
- passing.  However, their programming models emphasize locality more
- than our work does; while a program can examine any vertex in a UPC
- program at any time, it probably shouldn't (whereas our kernels rely on our
- latency hiding technique to make common remote references fast.
-
-The value of the information in extremely large and sparse graphs is
-rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
-sophisticated analysis on data from Twitter, identifying important
-actors and their networks of influence, and it can scale to the full
-size of graphs such as Facebook's friend network.
+The idea of processing large graphs on distributed machines has been
+exploredby projects such as Pregel \cite{pregel}, the Parallel Boost
+Graph Library \cite{parallelbgl}, and CGMLib \cite{cgmlib}. Our goal
+is to develop infrastructure to aid implementation of similar
+libraries.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
new abstract

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-  Large, sparse, irregular graphs map poorly to the memory systems of
-  commodity hardware. The scale of emerging real world datasets
-  presents a challenge for straightforward algorithms which will run
-  inefficiently due to many high-latency accesses.  Pervasive
-  parallelism can tolerate that latency, but the only large-scale
-  implementation of the technique is the Cray XMT, an expensive,
-  non-commodity machine.
+  Crunching large graphs is the basis of many emerging applications,
+  such as social network analysis and bioinformatics. Graph analytics
+  algorithms exhibit little locality and therefore present significant
+  performance challenges. Hardware multithreading systems (e.g, Cray
+  XMT) showed that with enough concurrency, we can tolerate long
+  latencies. Unfortunately, this solution is not available with
+  commodity parts.
 
-  If we can adapt the XMT's programming model to commodity hardware,
-  we can dramatically improve performance without resorting to
-  expensive custom systems. This paper presents a runtime for latency
-  tolerance through the use of parallel lightweight contexts, along
-  with our plans for scaling up to multiple nodes. Our analysis shows
-  the feasibility of our approach.
+  Our goal is to develop a latency-tolerant system built out of
+  commodity parts and mostly in software. The proposed system includes
+  a runtime that supports a large number of lightweight contexts,
+  full-empty synchronization and a memory manager that provides a
+  high-latency but high-bandwidth global shared memory. This paper
+  lays out the vision for our system, and justifies its feasibility
+  with a performance analysis of the runtime for latency tolerance.
+
+% Our analysis shows the feasibility of our
+%   approach.
+
+% To that end we developed a new runtime system that supports
+%   a large number of lightweight contexts, together with full-empty
+%   synchronization primitives and a large global address space.
+
+% To that end, we propose a
+%   new runtime system that is able to sustain a large number of active
+%   threads
+
+%  XMT-like properties, implemented
+%   with commodity components. This paper presents a runtime for latency
+%   tolerance, along with our plans for scaling up to multiple
+%   nodes. Our analysis shows the feasibility of our approach.
+
+%   Large, sparse, irregular graphs map poorly to the memory systems of
+%   commodity hardware. The scale of emerging real world datasets
+%   presents a challenge for straightforward algorithms which will run
+%   inefficiently due to many high-latency accesses. Pervasive
+%   parallelism can tolerate that latency, but the only large-scale
+%   implementation of the technique is the Cray XMT, an expensive,
+%   non-commodity machine.
+
+%   If we can adapt the XMT's programming model to commodity hardware,
+%   we can dramatically improve performance without resorting to
+%   expensive custom systems. This paper presents a runtime for latency
+%   tolerance through the use of parallel lightweight contexts, along
+%   with our plans for scaling up to multiple nodes. Our analysis shows
+%   the feasibility of our approach.
 \end{abstract}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
spacing

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-\title{Can We Crunch the Social Graph on the Cheap?}
+%\title{Can We Crunch the Social Graph on the Cheap?}
+\title{Crunching Large Graphs with Commodity Processors}
 
@@
 500 million active users with an average of 130 friends each
-\cite{Facebook:2011p91. Its friend suggestion algorithm in 2010 ran on
+\cite{Facebook:2011p91}. Its friend suggestion algorithm in 2010 ran on
   a 40-node cluster with 72GB memory per node, and still, new
@@
   \begin{center}
-    \includegraphics[width=0.9\textwidth]{figures/system-overview.pdf}
+    \vspace{-0.25in}
+    \includegraphics[width=0.7\textwidth]{figures/system-overview.pdf}
+    \vspace{-0.2in}
 	\end{center}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
almost all of eval condensed

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 %% To disable, just uncomment this line
-%% \renewcommand{\todo}[1]{\relax}
+ \renewcommand{\todo}[1]{\relax}
+
+\newcommand{\mrps}[0]{Mref/s}
 
@@
 \begin{abstract}
-% Techniques from graph analytics have found important applications in
-% areas such as social networks and bioinformatics. These irregular
-% problems present a challenge for parallel machines: obtaining
-% performance is not straightforward. Multithreading works well for
-% such problems, but the only large-scale implementation of the technique
-% is the Cray XMT, an expensive, non-commodity machine.
 
-% Our goal is to build a system with XMT-like properties, implemented
-% with commodity components. This paper presents a runtime for latency
-% tolerance, along with our plans for scaling up to multiple
-% nodes. Our analysis shows the feasibility of our approach.
+ Techniques from graph analytics have found important applications in
+  areas such as social networks and bioinformatics. These irregular
+  problems present a challenge for parallel machines: obtaining
+  performance is not straightforward. Multithreading works well for
+  such problems, but the only large-scale implementation of the technique
+  is the Cray XMT, an expensive, non-commodity machine.
 
-  Crunching large graphs is the basis of many emerging applications,
-  such as social network analysis and bioinformatics. Graph analytics
-  algorithms exhibit little locality and therefore present significant
-  performance challenges. Hardware multithreading systems (e.g, Cray
-  XMT) showed that with enough concurrency, we can tolerate long
-  latencies. Unfortunately, this solution is not available with
-  commodity parts.
-
-  Our goal is to develop a latency-tolerant system built out of
-  commodity parts and mostly in software. The proposed system includes
-  a runtime that supports a large number of lightweight contexts,
-  full-empty synchronization and a memory manager that provides a
-  high-latency but high-bandwidth global shared memory. This paper
-  lays out the vision for our system, and justifies its feasibility
-  with a performance analysis of the runtime for latency tolerance.
-
-% Our analysis shows the feasibility of our
-%   approach.
-
-% To that end we developed a new runtime system that supports
-%   a large number of lightweight contexts, together with full-empty
-%   synchronization primitives and a large global address space.
-
-% To that end, we propose a
-%   new runtime system that is able to sustain a large number of active
-%   threads
-
-%  XMT-like properties, implemented
-%   with commodity components. This paper presents a runtime for latency
-%   tolerance, along with our plans for scaling up to multiple
-%   nodes. Our analysis shows the feasibility of our approach.
-
-%   Large, sparse, irregular graphs map poorly to the memory systems of
-%   commodity hardware. The scale of emerging real world datasets
-%   presents a challenge for straightforward algorithms which will run
-%   inefficiently due to many high-latency accesses. Pervasive
-%   parallelism can tolerate that latency, but the only large-scale
-%   implementation of the technique is the Cray XMT, an expensive,
-%   non-commodity machine.
-
-%   If we can adapt the XMT's programming model to commodity hardware,
-%   we can dramatically improve performance without resorting to
-%   expensive custom systems. This paper presents a runtime for latency
-%   tolerance through the use of parallel lightweight contexts, along
-%   with our plans for scaling up to multiple nodes. Our analysis shows
-%   the feasibility of our approach.
+  Our goal is to build a system with XMT-like properties, implemented
+  with commodity components. This paper presents a runtime for latency
+  tolerance, along with our plans for scaling up to multiple
+  nodes. Our analysis shows the feasibility of our approach.
 \end{abstract}
@@
 
-Many important emerging applications such as social network analysis,
-bioinformatics and analysing vast sensor data from actual networks
-(e.g, water, electrical grid) are based on crunching very large
-graphs. Unfortunately, the computational cost of these applications
-gets quickly out of hand. For example, in January 2011, Facebook had
-500 million active users with an average of 130 friends each
-\cite{Facebook:2011p91. Its friend suggestion algorithm in 2010 ran on
-  a 40-node cluster with 72GB memory per node, and still, new
-  suggestions could only be computed every two days
-  \cite{Backstrom:2010p90}. Speeding up these applications at a low
-  cost would have a large impact in how we deal and react to the fast
-  growth of the data-set and would make the results even more
-  valuable.
-
-% Techniques from graph analytics have found important applications in
-% areas such as social networks and bioinformatics. \todo{more!}
+Techniques from graph analytics have found important applications in
+areas such as social networks and bioinformatics. \todo{more!}
 
@@
 
-In this context, the most interesting computational challenge comes
-from large, low-diameter, power law graphs: this combination makes
-extracting performance difficult. They do not fit in a single
+The most interesting computational challenge comes from large,
+low-diameter, power law graphs: this combination makes extracting
+performance difficult. They do not fit in a single
 commodity machine's memory. Their low diameter means that they are
-difficult to lay out with locality: every vertex needs to be near
-every other vertex. Their power law distribution means that a few
-vertices generate much more work than the rest, leading to load
-imbalances. Together, these properties lead to a difficult
-conclusion: each edge traversal is likely to lead to expensive
-inter-node communication. %cache miss or remote access.
+difficult to lay out with locality: every vertex needs to be near every
+other vertex. Their power law distribution means that a few vertices
+generate much more work than the rest, leading to load
+imbalances. Together, these properties lead to a
+difficult conclusion: each edge traversal is likely to induce a very
+expensive cache miss or remote access.
 
@@
 implement efficient computations for these graphs. The Cray XMT is an
-example of such an approach: it solves the memory latency problem (or
-communication problem) through concurrency rather than caching. Each
-XMT processor supports 128 hardware contexts and 1024 outstanding
-memory operations, and is able to switch contexts every cycle.
-% \todo{cite XMT description}.  %% LUIS: there is enough information here.
-This ability comes at a cost: the XMT is an expensive, non-commodity
-machine. Moreover, arguably a lot of XMT's advantages for these type
-of applications come at the cost of performance in general-purpose
-applications.
-
-% \todo{don't we want to also say XMT has poor performance on x
-%   kind of apps?}
+example of such an approach: it solves the memory latency problem
+through concurrency rather than caching. Each XMT processor supports
+128 hardware contexts and 1024 outstanding memory operations, and is
+able to switch contexts every cycle \todo{cite XMT description}. This ability comes at a cost: the
+XMT is an expensive, non-commodity machine. \todo{don't we want to also say XMT has poor performance on x kind of apps?}
 
@@
 
-%We want the best of both worlds. 
-We believe we can build a system mostly based on commodity parts that
-can attain XMT-like performance with a familiar, also XMT-like,
-programming model, but at a fraction of the cost. Therefore, our goal
-is to build a system that has good performance on low-locality graph
-codes but is implemented with cheap commodity processors and simple
-additional support in an FPGA. This approach also has the added
-benefit of not sacrificing general-purpose performance.
+We want the best of both worlds. Our goal is to build a system that
+has good performance on these low-locality graph codes but is
+implemented with cheap commodity processors.
 
@@
 memory requests to the non-coherent global memory shared across the
-nodes.  We are exploring a mix of hardware (FPGA) and software
+nodes.  We are exploring a mix of hardware and software
 implementations of the memory manager.
 
-This paper describes the vision of our system and explores the
-feasibility of our ideas using a single-node implementation of a
-lightweight threading runtime. We use prefetch instructions together
-with lightweight threads to tolerate the latency of DRAM access on the
-local node node. We use adversarial pointer chasing benchmarks to
-verify our runtime's overhead is acceptable. We also model the effects
-of pointer chasing on a remote node by artificially increasing
-latencys to the order of several micro-seconds.
+This paper explores the feasibility of our multi-node goal using a
+single-node implementation of our lightweight threading runtime. We
+use prefetch instructions together with lightweight threads to
+tolerate the latency of DRAM access on the local node node. We use
+pointer chasing benchmarks to verify our runtime's overhead is
+acceptable. We also model the effects of pointer chasing on a remote
+node.
 
-The rest of the paper is organized as follows. We quickly describe our
-programming model in section~\ref{sec:model}. We describe the
-implementation of our latency-tolerant runtime and plans for extending
-to multiple nodes in section~\ref{sec:approach}. We evaluate our
-runtime in section~\ref{sec:evaluation}.  We present related work in
-section~\ref{sec:related} and conclude in
+The rest of the paper is organized as follows. We describe our
+proposal from a programmer's viewpoint in section~\ref{sec:model}. We
+describe the implementation of our latency-tolerant runtime and plans
+for extending to multiple nodes in section~\ref{sec:approach}. We
+evaluate our runtime in section~\ref{sec:evaluation}.  We present
+related work in section~\ref{sec:related} and conclude in
 section~\ref{sec:conclusion}.
@@
 
-Our goal is to preserve the XMT programming model: large shared global
-address space, explicit concurrency with a large number of threads,
-and full-empty-bits synchronization.  We partition the overall address
+Our shared-memory programming model partitions the overall address
 space into a global shared address space and per node private address
@@
 
-% Whereas in the PGAS model coherency of global copies is implictly
-% maintained across the system, % \todo{cite pgas?}
-% we completely forego
-% caching of globally shared data.
-
-A node can exploit temporal locality of data stored in the global
-space only by first copying it into the node's local space.  In this
-way, a node can access data at random across the system while
+Whereas in the PGAS model coherency of global copies is implictly
+maintained across the system \todo{cite pgas?}, we forego caching of globally shared
+data.  A node can exploit temporal locality of data stored in
+the global space only by first copying it into the node's local space.
+In this way, a node can access data at random across the system while
 incurring no hardware coherence overhead: consistency is enforced by
@@
 concurrency.  Exactly how that concurrency is expressed by the
-programmer is language dependent. The goal of our system is to support
-that concurrency and consequently provide latency tolerant on a large
-address space. We do so via software multithreading: computation that
-is about to execute a long latency operation (e.g. a remote memory
-reference) initiates the operation and then yields to the scheduler,
-which quickly resumes execution of another computation.  Concurrency
-is also used -- via aggregation of requests to a common node
-destination -- to make more efficient use of network bandwidth and in
-some cases reduce bandwidth demands.
+programmer is language dependent.  Our programming model serves to
+augment threaded programming languages so that they become latency
+tolerant on large address spaces.  Software multithreading is used to
+provide latency tolerance: a computation that is about to execute a
+long latency operation such as a remote memory reference initiates the
+operation and then yields to the scheduler, which quickly resumes
+execution of another computation.  Concurrency is also used -- via
+aggregation of requests to a common node destination -- to make more
+efficient use of network bandwidth and in some cases reduce bandwidth
+demands.
 
@@
 works differently: we provide atomic operations such as {\tt
-  int\_fetch\_add} as well as operations using full-empty-bits, as on
-the Cray XMT.  As is true for other long latency operations,
+  int\_fetch\_add} as well as operations using full-bits, as on the
+Cray XMT.  As is true for other long latency operations,
 synchronization latency is tolerated by yielding to the scheduler.
@@
 
-%% I dont think we need the paragraph below.
-
-% Thus, the programming model provides the programmer with options: express
-% locality within the private space, and it will be exploited to reduce
-% latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
-% tolerating latency and reducing the cost of bandwidth.
+Thus, the programming model provides the programmer with options: express
+locality within the private space, and it will be exploited to reduce
+latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
+tolerating latency and reducing the cost of bandwidth.
 
@@
 concurrent references per core exceeds 10; we believe this is due to
-later prefetches squashing earlier ones in the line fill buffers. We
-also observe some points that do not follow the trend and look like noise, but these results are actually repeatable. This effect is harder to explain, but we suspect it is
+later prefetches squashing earlier ones in the line fill buffers.
+In the real runtime we expect this effect will not exist, since prefetch will not be implemented with built-in non-temporal prefetch instructions.
+We
+also observe many points that do not follow the trend are below the max, but these results are repeatable. This effect is harder to explain, but we suspect it is
 due to resource contention in the cores' pipelines once the memory
@@
   \end{center}
-  \caption{Cache pressure with coroutines.}
+  \caption{Cache pressure with coroutines, on six cores.}
   \label{fig:cache-pressure}
@@
 
-We ran this code on six cores with all data allocated in the same
-processor's memory. We varied the size of the local working set from 2KB
+We varied the size of the local working set from 2KB
 to 4MB, and ran both 0 and 32 local updates for each pointer
@@
 
-Figure~\ref{fig:cache-pressure} shows the result. On the left of the
-figure, we see the list-traversal-only case, where we allocate
-per-coroutine space but make no references to it. As we would expect,
-the reference rates grow with the number of coroutines just as in
-Figure~\ref{fig:multi-green}. On the right of the figure, we see the
-case with 32 updates for each list traversal. While performance
-decreases for large working set sizes and large number of coroutines,
+Figure~\ref{fig:cache-pressure} shows the result. As we would expect,
+when there are no references to the local structures, the reference rates grow with the number of coroutines just as in
+Figure~\ref{fig:multi-green}. For 32 local updates per traversal, while performance
+decreases for large working set sizes and large numbers of coroutines,
 we see speedup as number coroutines increases for up to 128KB working set size with 8 coroutines per
@@
 
-\subsection{Multi-node}
+\subsection{Performance with multiple nodes}
 
 To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.
-First we measure the QPI bandwidth between the two chips of the test machine, since the chip would talk to the network interface over a QPI link.
+Since the network interface is accessed over QPI, estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine.
 
 % take this figure out if we need the space
+% maybe even take out whole QPI thing and just relate the 100M?
 % the main point is room for remote references to local memory
-\begin{figure}[h]
-  \begin{center}
-    \includegraphics[width=0.42\textwidth]{figures/qpi_bw-edited.pdf}
-  \end{center}
-  \caption{Pointer chasing in a remote processor's memory.}
-  \label{fig:listwalk-qpi}
-\end{figure}
+%\begin{figure}[h]
+ % \begin{center}
+  %  \includegraphics[width=0.42\textwidth]{figures/qpi_bw-edited.pdf}
+ % \end{center}
+ % \caption{Pointer chasing in a remote processor's memory.}
+ % \label{fig:listwalk-qpi}
+%\end{figure}
 %%%%%%%%%%
 
-We measured the rate at which the cores can make requests
-over the QPI link. For this experiment, we allocated the lists in the
-second chip's memory and ran the traversals on the first chip's cores. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use of the QPI link limits us to 175 million references per second.
+For this experiment, we allocated the lists in the
+second chip's memory and ran the traversals on the first chip's cores. We found that use of the QPI link limits us to 175 \mrps.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex

@@
 
-\title{Can We Crunch the Social Graph on the Cheap?}
+%\textheight 9in
+%\textwidth 6.5in
+%\columnsep 1pc
+\renewcommand{\baselinestretch}{0.9}
+
+%\title{Can We Crunch the Social Graph on the Cheap?}
+\title{Crunching Large Graphs with Commodity Processors}
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
conclusion

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 %% To disable, just uncomment this line
- \renewcommand{\todo}[1]{\relax}
-
-\newcommand{\mrps}[0]{Mref/s}
+%% \renewcommand{\todo}[1]{\relax}
 
@@
 \usepackage{comment}
-\usepackage[labelfont=bf,font=small,belowskip=-7pt,aboveskip=-10pt]{caption}
 \begin{document}
@@
 Our evaulation consists of two parts.
-First, we demonstrate that the runtime system can achieve the same performance as when there is explicit memory concurrency available.
-Second, we look at the runtime within the multi-node picture and show that it can tolerate large latencies.
+First, we demonstrate that the runtime system can achieve the same performance as when there is explicit memory concurrency available: we measure pointer chasing performance and
+the effects of cache pressure due to the number of contexts required.
+Second, we look at the runtime within the multi-node picture and simulate the effects of fetching memory over a network to show that the runtime can reach the level of concurrency required to hide the latency.
 In each part we begin by studying relevant aspects of our test machine's memory system so we can interpret our runtime results.
@@
 \label{subsubsec:evalsinglebase}
+\todo{title might be improved}
 
@@
 rate that can be issued by the cores in one chip and the maximum random
-reference rate that can be serviced by one chip's memory controller.
+reference rate that can be serviced by one chip's memory controller.\todo{The 360M  "serviced" number could perhaps be an "aside" kind of paragraph since not supporting a major point}
 
@@
 \centering
-\begin{alltt}{\small
+\begin{alltt}
   while (count-- > 0) \{
@@
   \}
-  }\centering{\bf (a)}
+  \center{{\bf (a)}}
 \end{alltt}
@@
  \}
- }\centering{\bf (b)}
+  \center{{\bf (b)}}
 \end{alltt}
@@
       local = local->next;
-   \} \}
-   }\centering{\bf (c)}
+   \}
+  \}
+   \center{{\bf (c)}}
 \end{alltt}
@@
 \end{minipage}
-\vspace{10pt}
 \caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.
@@
 
-The lists were laid out randomly with pointers spaced at a cache line granularity, maximizing
+The lists were sized to exceed the last level cache and were laid out
+randomly with pointers spaced at a cache line granularity, maximizing
 the probability of each reference being a miss. We allocated the lists
-on 1GB huge pages to minimize TLB refill overhead.
+on 1GB huge pages to minimize TLB refill overhead. The lists were
+allocated in the memory attached to the processor containing the cores
+doing the traversal. We ran 10 traversals, measuring average rate in each, and recorded the maximum rate.
 
@@
 the maximum rate pointers are traversed for a given number of
-concurrent offered references.
-We see a maximum rate of 277 million references per second (\mrps), which agrees with the measurements found by \cite{Mandal:2010}.
+concurrent offered references. We see a maximum rate of 277 million
+references per second, which agrees with the measurements found by \cite{Mandal:2010} with a similar machine.
 This rate is achieved when the number of offered references is 36.
@@
 memory concurrency; the maximum reference rate for a single core is
-107M. We believe this limit is due to the core having only enough {\em line fill
-  buffers} \cite{nehalem:arch} to store 10 concurrent private cache misses.
+107M. We believe this limit is due to the core's limited number of {\em line fill
+  buffers} \cite{nehalem:arch}, which limits the core to 10
+concurrent private cache misses.
 
@@
 memory. With this configuration, we observed a maximum rate of 360
-\mrps.
-We believe this difference is due to the chip having only 32 buffers in its {\em Global Queue} \cite{nehalem:perf} for tracking concurrently-executing read misses from all the cores' private caches.
+million references per second. We believe this difference is due to
+the configuration of the processor's {\em Global Queue} \cite{nehalem:perf}, which limits the number of concurrently-executing read misses from a chips cores to its local last-level cache and memory to 32.
 
-\subsubsection{Coroutine performance}
+\subsubsection{Runtime performance}
 
@@
 coroutine, so the only source of memory concurrency is the use of
-coroutines. We are able to obtain a rate of 275 \mrps with 48 concurrent misses, or 8 coroutines per core. These
+coroutines. We are able to obtain a rate of 275 million references per
+second with 48 concurrent misses, or 8 coroutines per core. These
 results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36.
@@
 concurrent references per core exceeds 10; we believe this is due to
-later prefetches squashing earlier ones in the line fill buffers.
-In the real runtime we expect this effect will not exist, since prefetch will not be implemented with built-in non-temporal prefetch instructions.
-We
-also observe many points that do not follow the trend are below the max, but these results are repeatable. This effect is harder to explain, but we suspect it is
+later prefetches squashing earlier ones in the line fill buffers. We
+also observe some points that do not follow the trend and look like noise, but these results are actually repeatable. This effect is harder to explain, but we suspect it is
 due to resource contention in the cores' pipelines once the memory
@@
   \end{center}
-  \caption{Cache pressure with coroutines, on six cores.}
+  \caption{Cache pressure with coroutines.}
   \label{fig:cache-pressure}
@@
 
-We varied the size of the local working set from 2KB
+We ran this code on six cores with all data allocated in the same
+processor's memory. We varied the size of the local working set from 2KB
 to 4MB, and ran both 0 and 32 local updates for each pointer
@@
 
-Figure~\ref{fig:cache-pressure} shows the result. As we would expect,
-when there are no references to the local structures, the reference rates grow with the number of coroutines just as in
-Figure~\ref{fig:multi-green}. For 32 local updates per traversal, while performance
-decreases for large working set sizes and large numbers of coroutines,
+Figure~\ref{fig:cache-pressure} shows the result. On the left of the
+figure, we see the list-traversal-only case, where we allocate
+per-coroutine space but make no references to it. As we would expect,
+the reference rates grow with the number of coroutines just as in
+Figure~\ref{fig:multi-green}. On the right of the figure, we see the
+case with 32 updates for each list traversal. While performance
+decreases for large working set sizes and large number of coroutines,
 we see speedup as number coroutines increases for up to 128KB working set size with 8 coroutines per
@@
 
-\subsection{Performance with multiple nodes}
+\subsection{Multi-node}
 
 To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.
-Since the network interface is accessed over QPI, estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine.
+First we measure the QPI bandwidth between the two chips of the test machine, since the chip would talk to the network interface over a QPI link.
 
 % take this figure out if we need the space
-% maybe even take out whole QPI thing and just relate the 100M?
 % the main point is room for remote references to local memory
-%\begin{figure}[h]
- % \begin{center}
-  %  \includegraphics[width=0.42\textwidth]{figures/qpi_bw-edited.pdf}
- % \end{center}
- % \caption{Pointer chasing in a remote processor's memory.}
- % \label{fig:listwalk-qpi}
-%\end{figure}
+\begin{figure}[h]
+  \begin{center}
+    \includegraphics[width=0.42\textwidth]{figures/qpi_bw-edited.pdf}
+  \end{center}
+  \caption{Pointer chasing in a remote processor's memory.}
+  \label{fig:listwalk-qpi}
+\end{figure}
 %%%%%%%%%%
 
-For this experiment, we allocated the lists in the
-second chip's memory and ran the traversals on the first chip's cores. We found that use of the QPI link limits us to 175 \mrps.
+We measured the rate at which the cores can make requests
+over the QPI link. For this experiment, we allocated the lists in the
+second chip's memory and ran the traversals on the first chip's cores. Figure~\ref{fig:listwalk-qpi} shows the result. We found that use of the QPI link limits us to 175 million references per second.
 
@@
 Figure~\ref{fig:network-delay}. We are still able to reach a
-near-maximum rate of 273 \mrps, but at least 2
+near-maximum rate of 273 million references per second, but at least 2
 concurrent references per coroutine are required. Running with a single
-reference per coroutine achieves 236 \mrps.
+reference per coroutine achieves 236 million references per
+second.
 Note that the throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support.
@@
 
-We also observe that to achieve a throughput equal to the expected max outgoing network bandwidth (100 \mrps), \todo{reference earlier section calculation?} when there is only one reference per context available, we require 320 coroutines.
+We also observe that to achieve a throughput equal to the expected max outgoing network bandwidth (100 million references per second), \todo{reference earlier section calculation?} when there is only one reference per context available, we require 320 coroutines.
 This is close to the minimum of 300 concurrent references needed to tolerate $3 \mu s$ latency.
@@
 Any network interface would be connected through a QPI link,
-which we measured to have a bandwidth limit of 175 \mrps, so our runtime appears to have the capacity to chase pointers
+which we measured to have a bandwidth limit of 175 million references
+per second, so our runtime appears to have the capacity to chase pointers
 in a remote memory.
@@
 
+We presented our plans on building a system for large graph
+computation using commodity parts. We leverage decades-old ideas on
+using a large ammount of concurrency to tolerate latencies; but do so
+mostly in software. We developed a runtime system for latency
+tolerance and our results showed that it is able to manage enough
+concurrency to tolerate latencies long enough to amortize the cost of
+expensive accesses to our large global memory system (high-latency,
+but high-bandwidth).
+
 \bibliographystyle{acm}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
finish condensing eval

@@
 
-Many important emerging applications such as social network analysis,
-bioinformatics and analysing vast sensor data from actual networks
-(e.g, water, electrical grid) are based on crunching very large
-graphs. Unfortunately, the computational cost of these applications
-gets quickly out of hand. For example, in January 2011, Facebook had
-500 million active users with an average of 130 friends each
-\cite{Facebook:2011p91. Its friend suggestion algorithm in 2010 ran on
-  a 40-node cluster with 72GB memory per node, and still, new
-  suggestions could only be computed every two days
-  \cite{Backstrom:2010p90}. Speeding up these applications at a low
-  cost would have a large impact in how we deal and react to the fast
-  growth of the data-set and would make the results even more
-  valuable.
-
-% Techniques from graph analytics have found important applications in
-% areas such as social networks and bioinformatics. \todo{more!}
+Techniques from graph analytics have found important applications in
+areas such as social networks and bioinformatics. \todo{more}
 
@@
 To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.
-Since the network interface is accessed over QPI, estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine.
+
+Since the network interface is accessed over QPI, we estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine.
+To do this, we allocated the lists in the second chip's memory and ran the traversals on the first chip's cores.
+We found that use of the QPI link limits us to 175 \mrps.
 
@@
 
-For this experiment, we allocated the lists in the
-second chip's memory and ran the traversals on the first chip's cores. We found that use of the QPI link limits us to 175 \mrps.
-
 To simulate the performance of pointer chasing in a
@@
 switch transit delay of 400 ns \todo{how many hops? 1or3?} in each direction and thus estimate a 3
-$\mu$s round-trip delay. In this experiment, once a coroutine has
-yielded, we use the core's timestamp counter to ensure it does not get
-rescheduled for 3 $\mu$s.
+$\mu$s round-trip delay.
 
@@
   \end{center}
-  \caption{Pointer chasing with simulated network delay.}
+  \vspace{3pt}
+  \caption{Pointer chasing with simulated network delay. Each core ran up to 167 coroutines, and each coroutine made 1 to 4 concurrent references.}
   \label{fig:network-delay}
@@
 
-We ran these experiments with two sources of memory concurrency: each
-core ran up to 167 coroutines, and each coroutine made between 1 and 4
-concurrent references. The results are shown in
+The results of the experiment are shown in
 Figure~\ref{fig:network-delay}. We are still able to reach a
-near-maximum rate of 273 \mrps, but at least 2
+near-maximum rate of 273 \mrps\footnote{The throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support.
+This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, but at least 2
 concurrent references per coroutine are required. Running with a single
 reference per coroutine achieves 236 \mrps.
-Note that the throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support.
-This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.
 
@@
 
-We presented our plans on building a system for large graph
-computation using commodity parts. We leverage decades-old ideas on
-using a large ammount of concurrency to tolerate latencies; but do so
-mostly in software. We developed a runtime system for latency
-tolerance and our results showed that it is able to manage enough
-concurrency to tolerate latencies long enough to amortize the cost of
-expensive accesses to our large global memory system (high-latency,
-but high-bandwidth).
-
 \bibliographystyle{acm}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
resolved conflicts

@@
 \label{sec:approach}
-
-The goal of our implementation is to use multithreading to tolerate
+Our goal is to use multithreading to tolerate
 the latency of random accesses to memory in a global address space
-spread across multiple nodes. This goal has three key challenges:
+spread across multiple nodes.  Our implementation must provide three features:
 \begin{itemize}
@@
 
-  In a system with high degrees of concurrency, lightweight
-  synchronization primitives are important.
+  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
 \end{itemize}
@@
 user-level multitasking system built using coroutines.
-
-There are three main concerns in designing the coroutine
-mechanism. First, the working set of each coroutine must be small, so
-many coroutines can be active without trashing the cache. Second,
-context switches must be fast, so that overhead does not swamp actual
-work. Third, when a coroutine yields, we must choose a suitable
-coroutine to run next.
+Much work has been done on similar user-level systems \cite{ult,capriccio}, but we have stringent requirements on size and performance, since we have large numbers of contexts rapidly issuing prefetches and passing control.
+Hence coroutines must be small, so
+many coroutines can be active without trashing the cache, and
+context switches must be fast (a small fraction of memory latency) so that we can overlap memory requests and achieve concurrency.
 
@@
 compiler to save and restore any other registers in use. We minimize
-context switch time both by keeping contexts small and by doing all
-switching and scheduling in user space, as in \todo{cite other green
-  threads packages, like qthreads}.
-
-There are a number of possibilites for coroutine scheduling.
-Ideally, a thread would be rescheduled when its request had
-completed, but this may complicate the scheduling decision and slow
-down context switches. A simple round-robin scheduler may be sufficient.
+context switch time by doing all switching and scheduling in user space (kernel traps are at least an order of magnitude slower.)
 
@@
 We implemented coroutines as described, with a round-robin
-scheduler. If a prefetched data value is no longer in the cache when
-the coroutine is rescheduled and the value is used, the coroutine pays
-the full cost of the fetch.
+scheduler.
+Coroutines do local work, issue a remote prefetch, and switch to the next in turn.
+When the coroutine is reactivated, it assumes the requested data is available and issues a read.
+This is a disadvantage of the simple implementation; if the data is not yet received and in cache, the core (and all its coroutines) can stall.
 
@@
 synchronization operations will be scheduled soon after their
-operations have completed.
+operations have completed (and not before).
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
rebooting

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 %% To disable, just uncomment this line
- \renewcommand{\todo}[1]{\relax}
+% \renewcommand{\todo}[1]{\relax}
 
@@
 %todo format the UW names horizontally
-\authorinfo{Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
-Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
-\authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
-\authorinfo{Preston Briggs}{}{preston.briggs@gmail.com}
+\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
+  Dan Grossman, Mark Oskin, Carl Ebeling, Simon Kahan, Luis
+  Ceze}{University of Washington}{\{nelson, bdmyers, ahh, preston,
+  djg, oskin, ebeling, skahan, luisceze\}@cs.washington.edu}
 
 \maketitle
+
 \begin{abstract}
+% Techniques from graph analytics have found important applications in
+% areas such as social networks and bioinformatics. These irregular
+% problems present a challenge for parallel machines: obtaining
+% performance is not straightforward. Multithreading works well for
+% such problems, but the only large-scale implementation of the technique
+% is the Cray XMT, an expensive, non-commodity machine.
 
- Techniques from graph analytics have found important applications in
-  areas such as social networks and bioinformatics. These irregular
-  problems present a challenge for parallel machines: obtaining
-  performance is not straightforward. Multithreading works well for
-  such problems, but the only large-scale implementation of the technique
-  is the Cray XMT, an expensive, non-commodity machine.
+% Our goal is to build a system with XMT-like properties, implemented
+% with commodity components. This paper presents a runtime for latency
+% tolerance, along with our plans for scaling up to multiple
+% nodes. Our analysis shows the feasibility of our approach.
 
-  Our goal is to build a system with XMT-like properties, implemented
-  with commodity components. This paper presents a runtime for latency
-  tolerance, along with our plans for scaling up to multiple
-  nodes. Our analysis shows the feasibility of our approach.
+  Crunching large graphs is the basis of many emerging applications,
+  such as social network analysis and bioinformatics. Graph analytics
+  algorithms exhibit little locality and therefore present significant
+  performance challenges. Hardware multithreading systems (e.g, Cray
+  XMT) showed that with enough concurrency, we can tolerate long
+  latencies. Unfortunately, this solution is not available with
+  commodity parts.
+
+  Our goal is to develop a latency-tolerant system built out of
+  commodity parts and mostly in software. The proposed system includes
+  a runtime that supports a large number of lightweight contexts,
+  full-empty synchronization and a memory manager that provides a
+  high-latency but high-bandwidth global shared memory. This paper
+  lays out the vision for our system, and justifies its feasibility
+  with a performance analysis of the runtime for latency tolerance.
+
+% Our analysis shows the feasibility of our
+%   approach.
+
+% To that end we developed a new runtime system that supports
+%   a large number of lightweight contexts, together with full-empty
+%   synchronization primitives and a large global address space.
+
+% To that end, we propose a
+%   new runtime system that is able to sustain a large number of active
+%   threads
+
+%  XMT-like properties, implemented
+%   with commodity components. This paper presents a runtime for latency
+%   tolerance, along with our plans for scaling up to multiple
+%   nodes. Our analysis shows the feasibility of our approach.
+
+%   Large, sparse, irregular graphs map poorly to the memory systems of
+%   commodity hardware. The scale of emerging real world datasets
+%   presents a challenge for straightforward algorithms which will run
+%   inefficiently due to many high-latency accesses. Pervasive
+%   parallelism can tolerate that latency, but the only large-scale
+%   implementation of the technique is the Cray XMT, an expensive,
+%   non-commodity machine.
+
+%   If we can adapt the XMT's programming model to commodity hardware,
+%   we can dramatically improve performance without resorting to
+%   expensive custom systems. This paper presents a runtime for latency
+%   tolerance through the use of parallel lightweight contexts, along
+%   with our plans for scaling up to multiple nodes. Our analysis shows
+%   the feasibility of our approach.
 \end{abstract}
@@
 
-Techniques from graph analytics have found important applications in
-areas such as social networks and bioinformatics. \todo{more}
+Many important emerging applications such as social network analysis,
+bioinformatics and analysing vast sensor data from actual networks
+(e.g, water, electrical grid) are based on crunching very large
+graphs. Unfortunately, the computational cost of these applications
+gets quickly out of hand. For example, in January 2011, Facebook had
+500 million active users with an average of 130 friends each
+\cite{Facebook:2011p91}. Its friend suggestion algorithm in 2010 ran on
+  a 40-node cluster with 72GB memory per node, and still, new
+  suggestions could only be computed every two days
+  \cite{Backstrom:2010p90}. Speeding up these applications at a low
+  cost would have a large impact in how we deal and react to the fast
+  growth of the data-set and would make the results even more
+  valuable.
+
+% Techniques from graph analytics have found important applications in
+% areas such as social networks and bioinformatics. \todo{more!}
 
@@
 To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.
-
-Since the network interface is accessed over QPI, we estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine.
-To do this, we allocated the lists in the second chip's memory and ran the traversals on the first chip's cores.
-We found that use of the QPI link limits us to 175 \mrps.
+Since the network interface is accessed over QPI, estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine.
 
@@
 
+For this experiment, we allocated the lists in the
+second chip's memory and ran the traversals on the first chip's cores. We found that use of the QPI link limits us to 175 \mrps.
+
 To simulate the performance of pointer chasing in a
@@
 switch transit delay of 400 ns \todo{how many hops? 1or3?} in each direction and thus estimate a 3
-$\mu$s round-trip delay.
+$\mu$s round-trip delay. In this experiment, once a coroutine has
+yielded, we use the core's timestamp counter to ensure it does not get
+rescheduled for 3 $\mu$s.
 
@@
   \end{center}
-  \vspace{3pt}
-  \caption{Pointer chasing with simulated network delay. Each core ran up to 167 coroutines, and each coroutine made 1 to 4 concurrent references.}
+  \caption{Pointer chasing with simulated network delay.}
   \label{fig:network-delay}
@@
 
-The results of the experiment are shown in
+We ran these experiments with two sources of memory concurrency: each
+core ran up to 167 coroutines, and each coroutine made between 1 and 4
+concurrent references. The results are shown in
 Figure~\ref{fig:network-delay}. We are still able to reach a
-near-maximum rate of 273 \mrps\footnote{The throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support.
-This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, but at least 2
+near-maximum rate of 273 \mrps, but at least 2
 concurrent references per coroutine are required. Running with a single
 reference per coroutine achieves 236 \mrps.
+Note that the throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support.
+This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.
 
@@
 
-%\todo{Begin eval insert}
-%\input{evaluation}
-%\todo{End eval insert}
-
 \section{Related work}
 \label{sec:related}
-The concept of using independent parallel contexts to tolerate latency
-is not novel.  The inspiration for our project is of course the Cray
-XMT \cite{feo-xmt}, which implemented a hardware version of our
-programming model (in what would today be considered exceptionally
-slow silicon.)  The XMT can be seen as simultaneous multithreading
-\cite{tullsen-smt} writ large; the technique and motivation is
-essentially the same, but the XMT allows hundreds of contexts (modern
-Intel SMT implementations are generally limited to two.) The Alewife
-system \cite{agarwal-alewife} also switches contexts (in hardware) to
-hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
-allows hundreds of independently executing contexts to share FPUs,
-caches, and paths to memory, but does not context switch or share
-basic execution hardware.
 
-Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
-model under the hood;
-% Do I need to include a citation here?
-while they generate large numbers of available execution contexts via
-a (semi-)explicit SIMD model, those SIMD instructions keep functional
-units satisfied by overlapping execution with the loads required for
-other contexts.
+Much work exists on using multithreading to tolerate latency. Hardware
+implementations include the Tera MTA~\cite{tera} and Cray
+XMT~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
+MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
+even GPUs \cite{gpus}. Software-only approaches exist as well; the
+Software Controlled Multithreading project \cite{mowry-scm}, QThreads
+\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
+use variants of this idea.
+%In contrast to all of these, our goal is to tolerate
+%latencies across a multi-node system using mostly software.
 
-Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
-  multithreading}.  As opposed to our system of explicitly marked
-remote references, SCM assumes a high hit rate and traps to an
-interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
-short enough latencies that using more than two contexts generally
-produced slowdown.
+There has been much work on user level threading; the First-Class User
+Level Threading project \cite{ult} is one example. Our goal is
+slightly different; we want many contexts to tolerate memory latency.
 
-Partitioned global address space models, such as that used in Unified Parallel C,
-% definitely need a citation here
- share our approach of exposing all data (even remote data)
- transparently to any threads and have largely solved the problems of
- how to efficiently translate those remote reads into message
- passing.  However, their programming models emphasize locality more
- than our work does; while a program can examine any vertex in a UPC
- program at any time, it probably shouldn't (whereas our kernels rely on our
- latency hiding technique to make common remote references fast.
+Our goal of presenting a global view of distributed memory to the
+programmer is shared by the PGAS community, and is used in languages
+like Chapel \cite{chapel}, X10 \cite{X10}, and UPC \cite{upc}. They
+obtain performance by minimizing references to remote nodes, whereas we
+optimize for this case.
 
-The value of the information in extremely large and sparse graphs is
-rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
-sophisticated analysis on data from Twitter, identifying important
-actors and their networks of influence, and it can scale to the full
-size of graphs such as Facebook's friend network.
+The idea of processing large graphs on distributed machines has been
+exploredby projects such as Pregel \cite{pregel}, the Parallel Boost
+Graph Library \cite{parallelbgl}, and CGMLib \cite{cgmlib}. Our goal
+is to develop infrastructure to aid implementation of similar
+libraries.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
crash recovery

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 %todo format the UW names horizontally
-\authorinfo{Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
-Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
-\authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
-\authorinfo{Preston Briggs}{}{preston.briggs@gmail.com}
+\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
+  Dan Grossman, Mark Oskin, Carl Ebeling, Simon Kahan, Luis
+  Ceze}{University of Washington}{\{nelson, bdmyers, ahh, preston,
+  djg, oskin, ebeling, skahan, luisceze\}@cs.washington.edu}
 
 \maketitle
+
 \begin{abstract}
+   Crunching large graphs is the basis of many emerging applications,
+   such as social network analysis and bioinformatics. Graph analytics
+   algorithms exhibit little locality and therefore present significant
+   performance challenges. Hardware multithreading systems (e.g, Cray
+   XMT) showed that with enough concurrency, we can tolerate long
+   latencies. Unfortunately, this solution is not available with
+   commodity parts.
 
- Techniques from graph analytics have found important applications in
-  areas such as social networks and bioinformatics. These irregular
-  problems present a challenge for parallel machines: obtaining
-  performance is not straightforward. Multithreading works well for
-  such problems, but the only large-scale implementation of the technique
-  is the Cray XMT, an expensive, non-commodity machine.
+   Our goal is to develop a latency-tolerant system built out of
+   commodity parts and mostly in software. The proposed system includes
+   a runtime that supports a large number of lightweight contexts,
+   full-empty synchronization and a memory manager that provides a
+   high-latency but high-bandwidth global shared memory. This paper
+   lays out the vision for our system, and justifies its feasibility
+   with a performance analysis of the runtime for latency tolerance.
 
-  Our goal is to build a system with XMT-like properties, implemented
-  with commodity components. This paper presents a runtime for latency
-  tolerance, along with our plans for scaling up to multiple
-  nodes. Our analysis shows the feasibility of our approach.
 \end{abstract}
@@
 
-Techniques from graph analytics have found important applications in
-areas such as social networks and bioinformatics. \todo{more}
+Many important emerging applications such as social network analysis,
+bioinformatics and analysing vast sensor data from actual networks
+(e.g, water, electrical grid) are based on crunching very large
+graphs. Unfortunately, the computational cost of these applications
+gets quickly out of hand. For example, in January 2011, Facebook had
+500 million active users with an average of 130 friends each
+\cite{Facebook:2011p91}. Its friend suggestion algorithm in 2010 ran on
+  a 40-node cluster with 72GB memory per node, and still, new
+  suggestions could only be computed every two days
+  \cite{Backstrom:2010p90}. Speeding up these applications at a low
+  cost would have a large impact in how we deal and react to the fast
+  growth of the data-set and would make the results even more
+  valuable.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
space saving
small change

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 %University of Washington\\
-%\email{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}\\ \\
+%\email{\{nelson, bdmyers, ahh, djg, oskin, ebeling, luisceze\}@cs.washington.edu}\\ \\
 %Simon Kahan \\ 
@@
 \authorinfo{Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
-Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
+Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebeling, luisceze\}@cs.washington.edu}
 \authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
@@
 \label{sec:approach}
-Our goal is to use multithreading to tolerate
+
+The goal of our implementation is to use multithreading to tolerate
 the latency of random accesses to memory in a global address space
-spread across multiple nodes.  Our implementation must provide three features:
+spread across multiple nodes. This goal has three key challenges:
 \begin{itemize}
@@
 
-  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
+  In a system with high degrees of concurrency, lightweight
+  synchronization primitives are important.
 \end{itemize}
@@
 user-level multitasking system built using coroutines.
-Much work has been done on similar user-level systems \cite{ult,capriccio}, but we have stringent requirements on size and performance, since we have large numbers of contexts rapidly issuing prefetches and passing control.
-Hence coroutines must be small, so
-many coroutines can be active without trashing the cache, and
-context switches must be fast (a small fraction of memory latency) so that we can overlap memory requests and achieve concurrency.
+
+There are three main concerns in designing the coroutine
+mechanism. First, the working set of each coroutine must be small, so
+many coroutines can be active without trashing the cache. Second,
+context switches must be fast, so that overhead does not swamp actual
+work. Third, when a coroutine yields, we must choose a suitable
+coroutine to run next.
 
@@
 compiler to save and restore any other registers in use. We minimize
-context switch time by doing all switching and scheduling in user space (kernel traps are at least an order of magnitude slower.)
+context switch time both by keeping contexts small and by doing all
+switching and scheduling in user space, as in \todo{cite other green
+  threads packages, like qthreads}.
+
+There are a number of possibilites for coroutine scheduling.
+Ideally, a thread would be rescheduled when its request had
+completed, but this may complicate the scheduling decision and slow
+down context switches. A simple round-robin scheduler may be sufficient.
 
@@
 We implemented coroutines as described, with a round-robin
-scheduler.
-Coroutines do local work, issue a remote prefetch, and switch to the next in turn.
-When the coroutine is reactivated, it assumes the requested data is available and issues a read.
-This is a disadvantage of the simple implementation; if the data is not yet received and in cache, the core (and all its coroutines) can stall.
+scheduler. If a prefetched data value is no longer in the cache when
+the coroutine is rescheduled and the value is used, the coroutine pays
+the full cost of the fetch.
 
@@
 synchronization operations will be scheduled soon after their
-operations have completed (and not before).
+operations have completed.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
a few tweaks

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 \usepackage{comment}
-\usepackage[labelfont=bf,font=small,belowskip=-7pt,aboveskip=-10pt]{caption}
+\usepackage[labelfont=bf,font=small,belowskip=-3pt,aboveskip=-5pt]{caption}
+\usepackage[compact]{titlesec}
+\usepackage{mdwlist}
+
 \begin{document}
+
+%\titlespacing{\section}{0pt}{*0}{*0}
+%\titlespacing{\subsection}{0pt}{*0}{*0}
+%\titlespacing{\subsubsection}{0pt}{*0}{*0}
 
@@
 %\columnsep 1pc
-\renewcommand{\baselinestretch}{0.9}
+\renewcommand{\baselinestretch}{0.85}
 
@@
 
-%todo format the UW names horizontally
 \authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
@@
 128 hardware contexts and 1024 outstanding memory operations, and is
-able to switch contexts every cycle \todo{cite XMT description}. This ability comes at a cost: the
-XMT is an expensive, non-commodity machine. \todo{don't we want to also say XMT has poor performance on x kind of apps?}
+able to switch contexts every cycle \cite{tera, feo-xmt}. This ability
+comes at a cost: the XMT is an expensive, non-commodity machine.
 
@@
     \includegraphics[width=0.7\textwidth]{figures/system-overview.pdf}
-    \vspace{-0.2in}
+    \vspace{-0.1in}
 	\end{center}
@@
 
-\begin {comment}
-\todo{Change to discuss API:
-
-context switches:
-  yield()
-  discuss coroutines
-
-mem:
-  alloc()
-  read(p) 
-  write(p, d)
-  prefetch(p)
-
-sync:
- {read, write} {EE, EF, FE, FF}
-}
-
-While the ideal programming model is one that facilitates easy
-expression in the problem domain as well as efficient execution on the
-available hardware, our programming model today is motivated by only
-efficient execution: it exists as a rudimentary library for concept
-demonstration, not as a language suitable for development.  The
-programming model we use today is therefore only partially formed. It
-exists as an augmentation of existing threaded programming models such
-as pthreads or openmp that are readily available on most computer
-systems.
-
-Though not strictly necessary, for ease of exposition we imagine a
-one-to-one binding between threads and cores.  Our model manifests
-within any thread independently of the others as a ``fray'' \todo{for lack
-of better term }.  Within the thread, a fray instance is the
-instantiation of a set of coroutines and a scheduler.  Instantiation
-is at user-level and the operating system sees the entire fray as as a
-single user-level thread.
-
-\todo{ describe calls that perform thread \& scheduler instantiation; and schedule activation }
-
-Each coroutine executes non-preemptively \todo{seems like non-premptively isn't ``how" it executes, rather how it is scheduled, maybe say ``cooperatively"?}, utilizing the core to which
-its ``parent'' thread has been assigned by the operating system until it
-reaches a yield point.  Typically, the coroutine yields after
-initiating a prefetch operation on data that the programmer or
-compiler anticipates would probably not complete before the data is
-referenced by a subsequent load operation.  In this way, the core
-continues to execute instructions for other coroutines when otherwise
-it would likely stall.  The programming model thus tolerates latency.
-
-\todo{ describe calls to prefetch-and-yield }
-
-When a thread frays, for the sake of expediency, a fixed portion of
-its stack is divided amongst the coroutines.  Given a finite stack,
-each coroutine is constrained to execute a call tree of depth known to
-be finite at compilation time: recursion is not yet permitted.
-
-\todo{ discussion of coroutines \& synchronization } Were a coroutine to
-block on a synchronization variable shared with other threads, the
-entire fray would suspend execution.  This can lead to deadlock when,
-for example, one coroutine waits to consume from another thread that
-is waiting to consume what only another coroutine in this first fray
-can produce.  Instead, coroutines must yield on failed synchronization
-events, spinning rather than blocking, where were they bona fide
-threads, blocking might be more efficient.
-
-\todo{ say something about ``synchronization'' within a fray }
-
-Coroutines completing their work yield without adding themselves to
-the scheduling queue.  The last coroutine to exit in this way returns
-as the main thread, as in the common  fork-join model of parallelism.
-
-\end{comment}
-
 \section{Implementation}
@@
 spread across multiple nodes.  Our implementation must provide three features:
-\begin{itemize}
-\item {\bf Lightweight multithreading}
-
-  We expect to need hundreds of
+\begin{description}
+\item[Lightweight multithreading:] We expect to need hundreds of
   threads to tolerated a cluster's network latency, so our threading
@@
 
-\item {\bf Concurrency in global memory references}
+\item[Concurrency in global memory references:] A thread making a
+  long-latency global memory reference must be able to yield and allow
+  other threads to make progress.
 
-  A thread making a long-latency global memory reference must be able
-  to yield and allow other threads to make progress.
-
-\item {\bf Low overhead synchronization}
-
+\item[Low overhead synchronization:] 
   We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
-\end{itemize}
+\end{description}
 We discuss our approach to solving each of these challenges in
@@
 these global references to have low locality, we use the
-``non-temporal'' form of prefetch, which tries to bring the data near
-the core without disturbing the state of the cache
-hierarchy \cite{intel:swdev}.
+``non-temporal'' form of prefetch to avoid disturbing the state of the
+cache hierarchy \cite{intel:swdev}.
 
@@
 
-We believe we will need hundreds of outstanding memory references per processor to
-cover the latency of remote references in a multi-node
-system\todo{ can we make a stronger statement? we know it will be 300$+$ according to our calculations}. Unfortunately, commodity processors have much smaller limits
-on memory concurrency: section~\ref{subsubsec:evalsinglebase} finds a limit of
-36 concurrent loads. We will have to manage memory concurrency on our
-own to avoid this limit. We describe approaches in both software and
-hardware.
+We believe we will need hundreds of outstanding memory references per
+processor to cover the latency of remote references in a multi-node
+system. Unfortunately, commodity processors have much smaller limits
+on memory concurrency: section~\ref{subsubsec:evalsinglebase} finds a
+limit of 36 concurrent loads. We will have to manage memory
+concurrency on our own to avoid this limit. We describe approaches in
+both software and hardware.
 
@@
 
-This approach has the benefit of enabling our entire system to be
-build using commodity parts, but its potential for performance is not
-clear. We have increased the overhead of a single memory operation:
-what in the single node case was a simple read, is now multiple
-queuing operations between threads and multiple reads and writes over
-the processor's link to the network interface.
+%This approach has the benefit of enabling our entire system to be
+%build using commodity parts, but its potential for performance is not
+%clear. We have increased the overhead of a single memory operation:
+%what in the single node case was a simple read, is now multiple
+%queuing operations between threads and multiple reads and writes over
+%the processor's link to the network interface.
 
@@
 device. We can add a coprocessor FPGA in a processor socket, listening
-to coherence messages on the bus. Note that this accelerator is only a
-memory proxy, not a compute coprocessor. When it detects a reference
-to memory stored on a remote node, the accelerator would issue the
-request through the network controller and track its progress,
-delivering the data to the requesting thread when it was ready.
+to coherence messages on the bus, similar to \cite{mogill}. Note that
+this accelerator is only a memory proxy, not a compute
+coprocessor. The accelerator wold map all global shared memory into a
+segment of each node's local address space; when it detects a
+reference to memory on a remote node, it would do the address-to-node
+translation and issue the request through the network controller,
+delivering the data to the requesting thread when ready. We would
+encode commands in the upper bits of the address; a read from a
+location's {\em issue} address start the remote request and return
+immediately, and a read from a location's {\em complete} address would
+block until the data was available.
 
-All global shared memory would be mapped into a segment of the local
-address space on each node. The accelerator would do the
-address-to-node translation. In addition, to support the issue and
-complete operations for concurrency support, commands would be encoded
-in the upper bits of the address: then each remote word would have multiple
-addresses on the local node. Reading from the {\em issue} address would
-immediately return a dummy value, but the accelerator would issue the
-request over the network. Reading from the {\em complete} address
-would block until the data was available.\todo{adjust with carl's suggestions}
+%To support the issue and complete operations for concurrency support,
+%commands would be encoded in the upper bits of the address: then each
+%remote word would have multiple addresses on the local node. Reading
+%from the {\em issue} address would immediately return a value
+%indicating that the accelerator is issuing the request over the
+%network. Reading from the {\em complete} address would block until the
+%data was available.
 
@@
 Implementing full-empty bit support efficiently on a platform not
-designed for them is a challenge. It is possible to support full-empty
-bit synchronization on arbitrary words by allocating additional
-storage for the full-empty bits, and translating each atomic
-full-empty operation into a sequence of memory operations that modify
-the data and full-empty bit atomically, but this may have high
-overhead. \todo{cite qthreads somehow}
+designed for them is a challenge. Previous work \cite{qthreads} has
+enabled support full-empty bit synchronization on arbitrary words by
+allocating additional storage for the full-empty bits and implementing
+atomic full-empty operations as library routines.
 
@@
 Synchronization operations can also be delegated to a memory manager
-thread or accelerator. This may simplify the implementation of
-synchronization operations: if only a single core (or single
-accelerator) is accessing the data being synchronized, the use of
-memory fences may be reduced or eliminated.
-
-Note that we are not advocating using an FPGA as a memory controller
-with attached DRAM and additional storage for full-empty bits. We
-believe this unnecessarily complicates the design for such a
-component. Our goal is to minimize the amount of hardware design our
-proposal requires.
-\todo{but are we eliminating the possibility of FE-only storage on the FPGA?}
+thread or accelerator. Sychronization on remote data would be
+performed by the remote memory manager. This may simplify the
+implementation: if only a single core (or single accelerator) is
+accessing the data being synchronized, the use of memory fences may be
+reduced or eliminated.
 
@@
 \begin{figure*}[ht]
+\vspace{-0.2in}
 %list walk ILP
-\begin{minipage}[b]{0.3\linewidth}
+\begin{minipage}[b]{0.3\linewidth}{\small
 \centering
-\begin{alltt}{\small
+\begin{alltt}{\scriptsize
   while (count-- > 0) \{
@@
 \label{fig:pointernocoro}
-\end{minipage}
+}\end{minipage}
 %listwalk green threads
-\begin{minipage}[b]{0.35\linewidth}
+\begin{minipage}[b]{0.35\linewidth}{\small
 \centering
-\begin{alltt}{\small
+\begin{alltt}{\scriptsize
   while (count-- > 0) \{
@@
 \label{fig:pointercoro}
-\end{minipage}
+}\end{minipage}
 %cache pressure
-\begin{minipage}[b]{0.32\linewidth}
+\begin{minipage}[b]{0.32\linewidth}{\small
 \centering
-\begin{alltt}{\small
+\begin{alltt}{\scriptsize
   while (count-- > 0) \{
@@
 \label{fig:pointerupdate}
-\end{minipage}
-\vspace{10pt}
+}\end{minipage}
+%\vspace{5pt}
 \caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.
@@
 \begin{figure}[h]
+  \vspace{-.1in}
 	\begin{center}
@@
 
-To simulate the performance of pointer chasing in a
-multi-node system, we modified the scheduler of our coroutine library
-to include a delay before a coroutine can be rescheduled, imitating
-the network transit delay even though we are referencing local
-memory. Recently published measurements of Infiniband latencies
-\todo{cite} suggest we might expect 1.1 $\mu$s source-to-destination
-latencies between two computers connected back-to-back; we assume a
-switch transit delay of 400 ns \todo{how many hops? 1or3?} in each direction and thus estimate a 3
-$\mu$s round-trip delay.
+To simulate the performance of pointer chasing in a multi-node system,
+we modified the scheduler of our coroutine library to include a delay
+before a coroutine can be rescheduled, imitating the network transit
+delay even though we are referencing local memory. We assume 1.1
+$\mu$s interface and 400 ns switch latencies in each direction
+\cite{infiniband, mellanox:site}, and thus estimate a 3 $\mu$s round-trip
+delay.
 
@@
 
-There has been much work on user level threading; the First-Class User
-Level Threading project \cite{ult} is one example. Our goal is
+There has also been much work on user level threading; the First-Class
+User Level Threading project \cite{ult} is one example. Our goal is
 slightly different; we want many contexts to tolerate memory latency.

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
more references

@@
 %% To disable, just uncomment this line
-% \renewcommand{\todo}[1]{\relax}
+ \renewcommand{\todo}[1]{\relax}
 
@@
 
-Whereas in the PGAS model coherency of global copies is implictly
-maintained across the system \todo{cite pgas?}, we forego caching of globally shared
-data.  A node can exploit temporal locality of data stored in
-the global space only by first copying it into the node's local space.
-In this way, a node can access data at random across the system while
-incurring no hardware coherence overhead: consistency is enforced by
-our mechanisms for explicit global synchronization.
+% Whereas in the PGAS model coherency of global copies is implictly
+% maintained across the system \todo{cite pgas?}, we forego caching of globally shared
+% data.  A node can exploit temporal locality of data stored in
+% the global space only by first copying it into the node's local space.
+% In this way, a node can access data at random across the system while
+% incurring no hardware coherence overhead: consistency is enforced by
+% our mechanisms for explicit global synchronization.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
programming model

@@
 % Whereas in the PGAS model coherency of global copies is implictly
-% maintained across the system \todo{cite pgas?}, we forego caching of globally shared
-% data.  A node can exploit temporal locality of data stored in
-% the global space only by first copying it into the node's local space.
-% In this way, a node can access data at random across the system while
+% maintained across the system, % \todo{cite pgas?}
+% we completely forego
+% caching of globally shared data.
+
+% A node can exploit temporal locality of data stored in the global
+% space only by first copying it into the node's local space.  In this
+% way, a node can access data at random across the system while
 % incurring no hardware coherence overhead: consistency is enforced by
@@
 concurrency.  Exactly how that concurrency is expressed by the
-programmer is language dependent.  Our programming model serves to
-augment threaded programming languages so that they become latency
-tolerant on large address spaces.  Software multithreading is used to
-provide latency tolerance: a computation that is about to execute a
-long latency operation such as a remote memory reference initiates the
-operation and then yields to the scheduler, which quickly resumes
-execution of another computation.  Concurrency is also used -- via
-aggregation of requests to a common node destination -- to make more
-efficient use of network bandwidth and in some cases reduce bandwidth
-demands.
+programmer is language dependent. The goal of our system is to support
+that concurrency and consequently provide latency tolerant on a large
+address space. We do so via software multithreading: computation that
+is about to execute a long latency operation (e.g. a remote memory
+reference) initiates the operation and then yields to the scheduler,
+which quickly resumes execution of another computation.  Concurrency
+is also used---via aggregation of requests to a common node
+destination---to make more efficient use of network bandwidth and in
+some cases reduce bandwidth demands.
 
@@
 works differently: we provide atomic operations such as {\tt
-  int\_fetch\_add} as well as operations using full-bits, as on the
-Cray XMT.  As is true for other long latency operations,
+  int\_fetch\_add} as well as operations using full-empty-bits, as on
+the Cray XMT.  As is true for other long latency operations,
 synchronization latency is tolerated by yielding to the scheduler.
-Even non-blocking global atomics yield: atomic operations executed as
-x86 instructions sabbotage concurrency by forcing all outstanding
-memory references to complete before new ones can be issued.  By
-yielding, the core can switch to another computation while the
-synchronization is accomplished outside of the pipeline.  In addition,
-the implementation can exploit aggregation when concurrency is
-abundant to increase the throughput of global synchronization.
-
-Thus, the programming model provides the programmer with options: express
-locality within the private space, and it will be exploited to reduce
-latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
-tolerating latency and reducing the cost of bandwidth.
+Even non-blocking global atomics yield so the core can switch to
+another computation while the synchronization is accomplished outside
+of the pipeline.  In addition, the implementation can exploit
+aggregation when concurrency is abundant to increase the throughput of
+global synchronization.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
removed a little space from a figure
removed a little more leading
compression

@@
 %\columnsep 1pc
-\renewcommand{\baselinestretch}{0.85}
+\renewcommand{\baselinestretch}{0.8}
 
@@
 
-\subsubsection{Single-node implementation}
-We implemented coroutines as described, with a round-robin
-scheduler.
-Coroutines do local work, issue a remote prefetch, and switch to the next in turn.
-When the coroutine is reactivated, it assumes the requested data is available and issues a read.
-This is a disadvantage of the simple implementation; if the data is not yet received and in cache, the core (and all its coroutines) can stall.
+\paragraph{Single-node implementation.}  We implemented coroutines
+as described, with a round-robin scheduler. Coroutines do local work,
+issue a remote prefetch, and switch to the next in turn.  When the
+coroutine is reactivated, it assumes the requested data is available
+and issues a read.  This is a disadvantage of the simple
+implementation; if the data is not yet received and in cache, the core
+(and all its coroutines) can stall.
 
-\subsubsection{Multi-node ideas}
+\paragraph{Multi-node ideas.}
 We will explore how the memory manager and coroutine scheduler may
@@
 
-\subsubsection{Single-node implementation}
+\paragraph{Single-node implementation.}
 We use prefetch instructions for the issue operation, and regular
@@
 
-\subsubsection{Multi-node ideas}
-
+\paragraph{Multi-node ideas.}
 We believe we will need hundreds of outstanding memory references per
@@
 
-\subsubsection{Single-node system}
+\paragraph{Single-node system.}
 Just as with reads and writes, we prefetch and yield before performing
@@
 operations supported by our development platform; we have not
-implemented full-empty bits.
+implemented full-empty bits yet.
 
-\subsubsection{Multi-node ideas}
-
+\paragraph{Multi-node ideas.}
 Synchronization operations can also be delegated to a memory manager
@@
   \end{center}
-  \vspace{3pt}
+  \vspace{-0.3in}
   \caption{Pointer chasing with simulated network delay. Each core ran up to 167 coroutines, and each coroutine made 1 to 4 concurrent references.}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
intro fixes?

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 imbalances. Together, these properties lead to a
-difficult conclusion: each edge traversal is likely to induce a very
-expensive cache miss or remote access.
+difficult conclusion: each edge traversal is likely to lead to a very
+expensive inter-node communication.
 
@@
 able to switch contexts every cycle \cite{tera, feo-xmt}. This ability
-comes at a cost: the XMT is an expensive, non-commodity machine.
+comes at a cost: the XMT is an expensive, non-commodity
+machine.
 
@@
 
-We want the best of both worlds. Our goal is to build a system that
-has good performance on these low-locality graph codes but is
-implemented with cheap commodity processors.
+We believe we can build a system based mostly on commodity parts that
+can attain XMT-like performance with a familiar, also XMT-like,
+programming model, but at a fraction of the cost. Therefore, our goal
+is to build a system that has good performance on low-locality graph
+codes but is implemented with cheap commodity processors and simple
+additional support in an FPGA. This approach also has the added
+benefit of not sacrificing general-purpose performance.
 
@@
 memory requests to the non-coherent global memory shared across the
-nodes.  We are exploring a mix of hardware and software
+nodes.  We are exploring a mix of hardware (FPGA) and software
 implementations of the memory manager.
 
-This paper explores the feasibility of our multi-node goal using a
-single-node implementation of our lightweight threading runtime. We
-use prefetch instructions together with lightweight threads to
-tolerate the latency of DRAM access on the local node node. We use
-pointer chasing benchmarks to verify our runtime's overhead is
-acceptable. We also model the effects of pointer chasing on a remote
-node.
+This paper describes the vision of our system and explores the
+feasibility of our ideas using a single-node implementation of a
+lightweight threading runtime. We use prefetch instructions together
+with lightweight threads to tolerate the latency of DRAM access on the
+local node node. We use adversarial pointer chasing benchmarks to
+verify our runtime's overhead is acceptable. We also model the effects
+of pointer chasing on a remote node by artificially increasing
+latenciess to the order of several micro-seconds.
 
-The rest of the paper is organized as follows. We describe our
-proposal from a programmer's viewpoint in section~\ref{sec:model}. We
-describe the implementation of our latency-tolerant runtime and plans
-for extending to multiple nodes in section~\ref{sec:approach}. We
-evaluate our runtime in section~\ref{sec:evaluation}.  We present
-related work in section~\ref{sec:related} and conclude in
+The rest of the paper is organized as follows. We quickly describe our
+programming model in section~\ref{sec:model}. We describe the
+implementation of our latency-tolerant runtime and plans for extending
+to multiple nodes in section~\ref{sec:approach}. We evaluate our
+runtime in section~\ref{sec:evaluation}.  We present related work in
+section~\ref{sec:related} and conclude in
 section~\ref{sec:conclusion}.

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
inifiniband press release and conclude multi-ode

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 imbalances. Together, these properties lead to a
-difficult conclusion: each edge traversal is likely to lead to a very
-expensive inter-node communication.
+difficult conclusion: each edge traversal is likely to induce a very
+expensive cache miss or remote access.
 
@@
 able to switch contexts every cycle \cite{tera, feo-xmt}. This ability
-comes at a cost: the XMT is an expensive, non-commodity
-machine.
+comes at a cost: the XMT is an expensive, non-commodity machine.
 
@@
 
-We believe we can build a system based mostly on commodity parts that
-can attain XMT-like performance with a familiar, also XMT-like,
-programming model, but at a fraction of the cost. Therefore, our goal
-is to build a system that has good performance on low-locality graph
-codes but is implemented with cheap commodity processors and simple
-additional support in an FPGA. This approach also has the added
-benefit of not sacrificing general-purpose performance.
+We want the best of both worlds. Our goal is to build a system that
+has good performance on these low-locality graph codes but is
+implemented with cheap commodity processors.
 
@@
 memory requests to the non-coherent global memory shared across the
-nodes.  We are exploring a mix of hardware (FPGA) and software
+nodes.  We are exploring a mix of hardware and software
 implementations of the memory manager.
 
-This paper describes the vision of our system and explores the
-feasibility of our ideas using a single-node implementation of a
-lightweight threading runtime. We use prefetch instructions together
-with lightweight threads to tolerate the latency of DRAM access on the
-local node node. We use adversarial pointer chasing benchmarks to
-verify our runtime's overhead is acceptable. We also model the effects
-of pointer chasing on a remote node by artificially increasing
-latenciess to the order of several micro-seconds.
+This paper explores the feasibility of our multi-node goal using a
+single-node implementation of our lightweight threading runtime. We
+use prefetch instructions together with lightweight threads to
+tolerate the latency of DRAM access on the local node node. We use
+pointer chasing benchmarks to verify our runtime's overhead is
+acceptable. We also model the effects of pointer chasing on a remote
+node.
 
-The rest of the paper is organized as follows. We quickly describe our
-programming model in section~\ref{sec:model}. We describe the
-implementation of our latency-tolerant runtime and plans for extending
-to multiple nodes in section~\ref{sec:approach}. We evaluate our
-runtime in section~\ref{sec:evaluation}.  We present related work in
-section~\ref{sec:related} and conclude in
+The rest of the paper is organized as follows. We describe our
+proposal from a programmer's viewpoint in section~\ref{sec:model}. We
+describe the implementation of our latency-tolerant runtime and plans
+for extending to multiple nodes in section~\ref{sec:approach}. We
+evaluate our runtime in section~\ref{sec:evaluation}.  We present
+related work in section~\ref{sec:related} and conclude in
 section~\ref{sec:conclusion}.
@@
 
-We also observe that to achieve a throughput equal to the expected max outgoing network bandwidth (100 \mrps), \todo{reference earlier section calculation?} when there is only one reference per context available, we require 320 coroutines.
-This is close to the minimum of 300 concurrent references needed to tolerate $3 \mu s$ latency.
+These results suggest that the runtime should be able to tolerate large latency well enough to exceed the max issue rates of both modern network interfaces (about 100 \mrps) \cite{mellanox:press} and the QPI link that connects the chip to the network interface.
 
-Any network interface would be connected through a QPI link,
-which we measured to have a bandwidth limit of 175 \mrps, so our runtime appears to have the capacity to chase pointers
-in a remote memory.
+%Any network interface would be connected through a QPI link,
+%which we measured to have a bandwidth limit of 175 \mrps, so our runtime appears to have the capacity to chase pointers
+%in a remote memory.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
more space edits

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-Our evaulation consists of two parts.
-First, we demonstrate that the runtime system can achieve the same performance as when there is explicit memory concurrency available.
-Second, we look at the runtime within the multi-node picture and show that it can tolerate large latencies.
-In each part we begin by studying relevant aspects of our test machine's memory system so we can interpret our runtime results.
+Our evaulation consists of two parts. First, we demonstrate that the
+runtime system can achieve the same performance as when there is
+explicit memory concurrency available. Second, we look at the runtime
+within the multi-node picture and show that it can tolerate large
+latencies. In each part we begin by studying relevant aspects of our
+test machine's memory system so we can interpret our runtime results.
 
@@
 
-We first characterize the performance of our test machine's memory
-system. Then we use these results to evaluate the performance of our runtime within a single node.
+% We first characterize the performance of our test machine's memory
+% system. Then we use these results to evaluate the performance of our runtime within a single node.
 
@@
 There has also been much work on user level threading; the First-Class
-User Level Threading project \cite{ult} is one example. Our goal is
-slightly different; we want many contexts to tolerate memory latency.
+User Level Threading project \cite{ult} is one example. We have a
+different goal; we want many lightweight contexts to tolerate memory
+latency.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
jjj

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
    performance challenges. Hardware multithreading systems (e.g, Cray
-   XMT) showed that with enough concurrency, we can tolerate long
+   XMT) have proven that with enough concurrency, we can tolerate long
    latencies. Unfortunately, this solution is not available with
@@
 Many important emerging applications such as social network analysis,
-bioinformatics and analysing vast sensor data from actual networks
+bioinformatics and analyzing vast sensor data from actual networks
 (e.g, water, electrical grid) are based on crunching very large
@@
 \cite{Facebook:2011p91}. Its friend suggestion algorithm in 2010 ran on
-  a 40-node cluster with 72GB memory per node, and still, new
+  a 40-node cluster with 72GB memory per node, and, still, new
   suggestions could only be computed every two days
@@
 memory requests to the non-coherent global memory shared across the
-nodes.  We are exploring a mix of hardware (FPGA) and software
+nodes.  We are exploring a mix of hardware and software
 implementations of the memory manager.
 
-This paper describes the vision of our system and explores the
-feasibility of our ideas using a single-node implementation of a
-lightweight threading runtime. We use prefetch instructions together
-with lightweight threads to tolerate the latency of DRAM access on the
-local node node. We use adversarial pointer chasing benchmarks to
-verify our runtime's overhead is acceptable. We also model the effects
-of pointer chasing on a remote node by artificially increasing
-latenciess to the order of several micro-seconds.
+This paper explores the feasibility of our multi-node goal using a
+single-node implementation of our lightweight threading runtime. We
+use prefetch instructions together with lightweight threads to
+tolerate the latency of DRAM access on the local node. We use
+pointer chasing benchmarks to verify our runtime's overhead is
+acceptable. We also model the effects of pointer chasing on a remote
+node.
 
-The rest of the paper is organized as follows. We quickly describe our
-programming model in section~\ref{sec:model}. We describe the
-implementation of our latency-tolerant runtime and plans for extending
-to multiple nodes in section~\ref{sec:approach}. We evaluate our
-runtime in section~\ref{sec:evaluation}.  We present related work in
-section~\ref{sec:related} and conclude in
+The rest of the paper is organized as follows. We describe our
+proposal from a programmer's viewpoint in section~\ref{sec:model}. We
+describe the implementation of our latency-tolerant runtime and plans
+for extending to multiple nodes in section~\ref{sec:approach}. We
+evaluate our runtime in section~\ref{sec:evaluation}.  We present
+related work in section~\ref{sec:related} and conclude in
 section~\ref{sec:conclusion}.
@@
 address space, explicit concurrency with a large number of threads,
-and full-empty-bits synchronization.  We partition the overall address
+and full-bit synchronization.  We partition the overall address
 space into a global shared address space and per node private address
@@
 programmer is language dependent. The goal of our system is to support
-that concurrency and consequently provide latency tolerant on a large
+that concurrency and consequently provide latency tolerance on a large
 address space. We do so via software multithreading: computation that
@@
 works differently: we provide atomic operations such as {\tt
-  int\_fetch\_add} as well as operations using full-empty-bits, as on
+  int\_fetch\_add} as well as operations using full-bits, as on
 the Cray XMT.  As is true for other long latency operations,
@@
 \item[Lightweight multithreading:] We expect to need hundreds of
-  threads to tolerated a cluster's network latency, so our threading
+  threads to tolerate a cluster's network latency, so our threading
   implementation must have low overhead.
@@
 
-Our approach to supporting many lightweight threads uses a cooperative,
-user-level multitasking system built using coroutines.
-Much work has been done on similar user-level systems \cite{ult,capriccio}, but we have stringent requirements on size and performance, since we have large numbers of contexts rapidly issuing prefetches and passing control.
-Hence coroutines must be small, so
-many coroutines can be active without trashing the cache, and
-context switches must be fast (a small fraction of memory latency) so that we can overlap memory requests and achieve concurrency.
+Our approach to supporting many lightweight threads uses a
+cooperative, user-level multitasking system built using
+coroutines. Much work has been done on similar user-level systems
+\cite{ult,capriccio}, but we have more stringent requirements:
+coroutines must use little space so that many can be active without
+trashing the cache; and context switches must be fast (a small
+fraction of memory latency) so that we can overlap memory requests and
+achieve concurrency.
 
@@
 
-\paragraph{Single-node implementation.}  We implemented coroutines
-as described, with a round-robin scheduler. Coroutines do local work,
-issue a remote prefetch, and switch to the next in turn.  When the
-coroutine is reactivated, it assumes the requested data is available
-and issues a read.  This is a disadvantage of the simple
-implementation; if the data is not yet received and in cache, the core
-(and all its coroutines) can stall.
+\subsubsection{Single-node implementation}
+We implemented coroutines as described, with a round-robin
+scheduler.
+Coroutines do local work, issue a remote prefetch, and switch to the next in turn.
+When the coroutine is reactivated, it assumes the requested data is available and issues a read.
+This is a disadvantage of the simple implementation; if the data is not yet received and in cache, the core (and all its coroutines) can stall.
 
-\paragraph{Multi-node ideas.}
+\subsubsection{Multi-node ideas}
 We will explore how the memory manager and coroutine scheduler may
@@
 
-\paragraph{Single-node implementation.}
+\subsubsection{Single-node implementation}
 We use prefetch instructions for the issue operation, and regular
@@
 
-\paragraph{Multi-node ideas.}
+\subsubsection{Multi-node ideas}
+
 We believe we will need hundreds of outstanding memory references per
@@
 
-\paragraph{Single-node system.}
+\subsubsection{Single-node system}
 Just as with reads and writes, we prefetch and yield before performing
@@
 operations supported by our development platform; we have not
-implemented full-empty bits yet.
+implemented full-empty bits.
 
-\paragraph{Multi-node ideas.}
+\subsubsection{Multi-node ideas}
+
 Synchronization operations can also be delegated to a memory manager
@@
 
-Our evaulation consists of two parts. First, we demonstrate that the
-runtime system can achieve the same performance as when there is
-explicit memory concurrency available. Second, we look at the runtime
-within the multi-node picture and show that it can tolerate large
-latencies. In each part we begin by studying relevant aspects of our
-test machine's memory system so we can interpret our runtime results.
+Our evaulation consists of two parts.
+First, we demonstrate that the runtime system can achieve the same performance as when there is explicit memory concurrency available.
+Second, we look at the runtime within the multi-node picture and show that it can tolerate large latencies.
+In each part we begin by studying relevant aspects of our test machine's memory system so we can interpret our runtime results.
 
@@
 
-% We first characterize the performance of our test machine's memory
-% system. Then we use these results to evaluate the performance of our runtime within a single node.
+We first characterize the performance of our test machine's memory
+system. Then we use these results to evaluate the performance of our runtime within a single node.
 
@@
 There has also been much work on user level threading; the First-Class
-User Level Threading project \cite{ult} is one example. We have a
-different goal; we want many lightweight contexts to tolerate memory
-latency.
+User Level Threading project \cite{ult} is one example. Our goal is
+slightly different; we want many contexts to tolerate memory latency.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 with lightweight threads to tolerate the latency of DRAM access on the
-local node node. We use adversarial pointer chasing benchmarks to
+local node. We use worst-case pointer chasing benchmarks to
 verify our runtime's overhead is acceptable. We also model the effects
 of pointer chasing on a remote node by artificially increasing
-latenciess to the order of several micro-seconds.
+latencies to several micro-seconds.
 
-The rest of the paper is organized as follows. We quickly describe our
+The rest of the paper is organized as follows. We briefly describe our
 programming model in section~\ref{sec:model}. We describe the

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
tense

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
    performance challenges. Hardware multithreading systems (e.g, Cray
-   XMT) showed that with enough concurrency, we can tolerate long
+   XMT) show that with enough concurrency, we can tolerate long
    latencies. Unfortunately, this solution is not available with

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
a few bits of space saving

@@
    performance challenges. Hardware multithreading systems (e.g, Cray
-   XMT) show that with enough concurrency, we can tolerate long
+   XMT) showed that with enough concurrency, we can tolerate long
    latencies. Unfortunately, this solution is not available with
@@
 
-\paragraph{Single-node implementation.}  We implemented coroutines
+%\paragraph{Single-node implementation.}  
+In our single-node implementation, we implemented coroutines
 as described, with a round-robin scheduler. Coroutines do local work,
@@
 
-\paragraph{Multi-node ideas.}
-We will explore how the memory manager and coroutine scheduler may
+%\paragraph{Multi-node ideas.}
+As we develop our multi-node implementation, we will explore how the memory manager and coroutine scheduler may
 interact, so that threads waiting on long-latency memory operations or
 synchronization operations will be scheduled soon after their
-operations have completed (and not before).
+operations have completed.
 
@@
 
-\paragraph{Single-node implementation.}
-We use prefetch instructions for the issue operation, and regular
+%\paragraph{Single-node implementation.}
+In our single-node implementation, we use prefetch instructions for the issue operation, and regular
 blocking reads and writes for the complete operation.  Since we expect
@@
 
-\paragraph{Multi-node ideas.}
-We believe we will need hundreds of outstanding memory references per
+%\paragraph{Multi-node ideas.}
+As we develop our multi-node implementation, we believe we will need hundreds of outstanding memory references per
 processor to cover the latency of remote references in a multi-node
@@
 
-\paragraph{Single-node system.}
-Just as with reads and writes, we prefetch and yield before performing
+%\paragraph{Single-node system.}
+In our single-node implementation, we prefetch and yield before performing
 a synchronization operation. We support only the synchronization
@@
 
-\paragraph{Multi-node ideas.}
-Synchronization operations can also be delegated to a memory manager
+%\paragraph{Multi-node ideas.}
+In a multi-node implementation, synchronization operations can be delegated to a memory manager
 thread or accelerator. Sychronization on remote data would be

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
    performance challenges. Hardware multithreading systems (e.g, Cray
-   XMT) have proven that with enough concurrency, we can tolerate long
+   XMT) show that with enough concurrency, we can tolerate long
    latencies. Unfortunately, this solution is not available with

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
modified: softxmt-hotpar2011.tex

@@
    performance challenges. Hardware multithreading systems (e.g, Cray
-   XMT) show that with enough concurrency, we can tolerate long
+   XMT) have proven that with enough concurrency, we can tolerate long
    latencies. Unfortunately, this solution is not available with
@@
 
-We minimize context size by treating context switches as function
-calls, as in~\cite{charm}. This allows us to save and restore the
-minimum number of registers allowed by the ABI, and depend on the
-compiler to save and restore any other registers in use. We minimize
+%We minimize context size by treating context switches as function
+%calls, as in~\cite{charm}. This allows us 
+%to save and restore the
+%minimum number of registers allowed by the ABI, and 
+%Simon:  not really true -- we could save fewer if we knew which of the
+%callee save registers weren't going to be clobbered.
+We treat context switches as function calls, as in~\cite{charm}. This allows us 
+to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers:  the size of a coroutine context is hardly more than the stack space needed to save these registers.
+We minimize
 context switch time by doing all switching and scheduling in user space (kernel traps are at least an order of magnitude slower.)
 
-%\paragraph{Single-node implementation.}  
-In our single-node implementation, we implemented coroutines
+\paragraph{Single-node implementation.}  We implemented coroutines
 as described, with a round-robin scheduler. Coroutines do local work,
@@
 
-%\paragraph{Multi-node ideas.}
-As we develop our multi-node implementation, we will explore how the memory manager and coroutine scheduler may
+\paragraph{Multi-node ideas.}
+We will explore how the memory manager and coroutine scheduler may
 interact, so that threads waiting on long-latency memory operations or
 synchronization operations will be scheduled soon after their
-operations have completed.
+operations have completed (and not before).
 
@@
 
-%\paragraph{Single-node implementation.}
-In our single-node implementation, we use prefetch instructions for the issue operation, and regular
+\paragraph{Single-node implementation.}
+We use prefetch instructions for the issue operation, and regular
 blocking reads and writes for the complete operation.  Since we expect
@@
 
-%\paragraph{Multi-node ideas.}
-As we develop our multi-node implementation, we believe we will need hundreds of outstanding memory references per
+\paragraph{Multi-node ideas.}
+We believe we will need hundreds of outstanding memory references per
 processor to cover the latency of remote references in a multi-node
@@
 
-%\paragraph{Single-node system.}
-In our single-node implementation, we prefetch and yield before performing
+\paragraph{Single-node system.}
+Just as with reads and writes, we prefetch and yield before performing
 a synchronization operation. We support only the synchronization
@@
 
-%\paragraph{Multi-node ideas.}
-In a multi-node implementation, synchronization operations can be delegated to a memory manager
+\paragraph{Multi-node ideas.}
+Synchronization operations can also be delegated to a memory manager
 thread or accelerator. Sychronization on remote data would be

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
Reword introduction

@@
 Many important emerging applications such as social network analysis,
-bioinformatics and analyzing vast sensor data from actual networks
-(e.g, water, electrical grid) are based on crunching very large
+bioinformatics, and sensor networks rely on crunching very large
 graphs. Unfortunately, the computational cost of these applications
@@
   \cite{Backstrom:2010p90}. Speeding up these applications at a low
-  cost would have a large impact in how we deal and react to the fast
-  growth of the data-set and would make the results even more
-  valuable.
+  cost would have a large impact in how we deal with large data-sets and make them even more valuable.
 
@@
 We believe we can build a system based mostly on commodity parts that
-can attain XMT-like performance with a familiar, also XMT-like,
+can attain XMT-like performance with a familiar, XMT-like,
 programming model, but at a fraction of the cost. Therefore, our goal
@@
 
-%We minimize context size by treating context switches as function
-%calls, as in~\cite{charm}. This allows us 
-%to save and restore the
-%minimum number of registers allowed by the ABI, and 
-%Simon:  not really true -- we could save fewer if we knew which of the
-%callee save registers weren't going to be clobbered.
-We treat context switches as function calls, as in~\cite{charm}. This allows us 
-to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers:  the size of a coroutine context is hardly more than the stack space needed to save these registers.
-We minimize
+We minimize context size by treating context switches as function
+calls, as in~\cite{charm}. This allows us to save and restore the
+minimum number of registers allowed by the ABI, and depend on the
+compiler to save and restore any other registers in use. We minimize
 context switch time by doing all switching and scheduling in user space (kernel traps are at least an order of magnitude slower.)

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex

@@
 
-We minimize context size by treating context switches as function
-calls, as in~\cite{charm}. This allows us to save and restore the
-minimum number of registers allowed by the ABI, and depend on the
-compiler to save and restore any other registers in use. We minimize
+%We minimize context size by treating context switches as function
+%calls, as in~\cite{charm}. This allows us 
+%to save and restore the
+%minimum number of registers allowed by the ABI, and 
+%Simon:  not really true -- we could save fewer if we knew which of the
+%callee save registers weren't going to be clobbered.
+We treat context switches as function calls, as in~\cite{charm}. This allows us 
+to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers:  the size of a coroutine context is hardly more than the stack space needed to save these registers.
+We minimize
 context switch time by doing all switching and scheduling in user space (kernel traps are at least an order of magnitude slower.)

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
figures move up

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 Many important emerging applications such as social network analysis,
-bioinformatics, and sensor networks rely on crunching very large
+bioinformatics and analyzing vast sensor data from actual networks
+(e.g, water, electrical grid) are based on crunching very large
 graphs. Unfortunately, the computational cost of these applications
@@
   \cite{Backstrom:2010p90}. Speeding up these applications at a low
-  cost would have a large impact in how we deal with large data-sets and make them even more valuable.
+  cost would have a large impact in how we deal and react to the fast
+  growth of the data-set and would make the results even more
+  valuable.
 
@@
 We believe we can build a system based mostly on commodity parts that
-can attain XMT-like performance with a familiar, XMT-like,
+can attain XMT-like performance with a familiar, also XMT-like,
 programming model, but at a fraction of the cost. Therefore, our goal
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \vspace{-.1in}
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \begin{center}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
some sec4/ref/conclusion changes

@@
   \end{center}
+  \vspace{-0.3in}
   \caption{Cache pressure with coroutines, on six cores.}
@@
 decreases for large working set sizes and large numbers of coroutines,
-we see speedup as number coroutines increases for up to 128KB working set size with 8 coroutines per
+we see speedup as the number of coroutines increases up to 128KB working set size with 8 coroutines per
 core.
@@
 This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, but at least 2
-concurrent references per coroutine are required. Running with a single
+concurrent references per coroutine are required. Running with one
 reference per coroutine achieves 236 \mrps.
 
-These results suggest that the runtime should be able to tolerate large latency well enough to exceed the max issue rates of both modern network interfaces (about 100 \mrps) \cite{mellanox:press} and the QPI link that connects the chip to the network interface.
+These results suggest that the runtime should be able to tolerate
+large latency well enough to exceed the max issue rates of both modern
+network interfaces (about 100 \mrps) \cite{mellanox:press} and the QPI
+link used to reach the network interface.
 
@@
 
-There has also been much work on user level threading; the First-Class
-User Level Threading project \cite{ult} is one example. We have a
+There has also been much work on user level threading, including the First-Class
+User Level Threading project \cite{ult} and Capriccio \cite{capriccio}. We have a
 different goal; we want many lightweight contexts to tolerate memory
@@
 The idea of processing large graphs on distributed machines has been
-exploredby projects such as Pregel \cite{pregel}, the Parallel Boost
+explored by projects such as Pregel \cite{pregel}, the Parallel Boost
 Graph Library \cite{parallelbgl}, and CGMLib \cite{cgmlib}. Our goal
@@
 
-We presented our plans on building a system for large graph
+We presented our plans for building a system for large graph
 computation using commodity parts. We leverage decades-old ideas on
-using a large ammount of concurrency to tolerate latencies; but do so
+using a large amount of concurrency to tolerate latencies; but we do so
 mostly in software. We developed a runtime system for latency

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   softxmt-hotpar2011.tex

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
    a runtime that supports a large number of lightweight contexts,
-   full-empty synchronization and a memory manager that provides a
+   full-bit synchronization and a memory manager that provides a
    high-latency but high-bandwidth global shared memory. This paper
@@
 
-%\paragraph{Single-node implementation.}  
-In our single-node implementation, we implemented coroutines
-as described, with a round-robin scheduler. Coroutines do local work,
-issue a remote prefetch, and switch to the next in turn.  When the
+\paragraph{Single-node implementation.}  We implemented coroutines
+with a round-robin scheduler. Coroutines do local work,
+issue a remote prefetch, and yield. When a
 coroutine is reactivated, it assumes the requested data is available
-and issues a read.  This is a disadvantage of the simple
-implementation; if the data is not yet received and in cache, the core
+and issues a read.  If the data is not yet received and in cache, the core
 (and all its coroutines) can stall.
 
-%\paragraph{Multi-node ideas.}
-As we develop our multi-node implementation, we will explore how the memory manager and coroutine scheduler may
+\paragraph{Multi-node ideas.}
+We will explore how the memory manager and coroutine scheduler may
 interact, so that threads waiting on long-latency memory operations or
 synchronization operations will be scheduled soon after their
-operations have completed.
+operations have completed (and not before).
 
@@
 this accelerator is only a memory proxy, not a compute
-coprocessor. The accelerator wold map all global shared memory into a
+coprocessor. The accelerator would map all global shared memory into a
 segment of each node's local address space; when it detects a
@@
 
-Implementing full-empty bit support efficiently on a platform not
+Implementing full-bit support efficiently on a platform not
 designed for them is a challenge. Previous work \cite{qthreads} has
-enabled support full-empty bit synchronization on arbitrary words by
-allocating additional storage for the full-empty bits and implementing
-atomic full-empty operations as library routines.
+enabled support for full-bit synchronization on arbitrary words by
+allocating additional storage for the full-bits and implementing
+atomic full-bit operations as library routines.
 
-One potential optimization is to limit full-empty synchronization to
+One potential optimization is to limit full-bit synchronization to
 pointers to aligned data types, and reuse wasted space in the
-low-order bits of the pointer for full-empty bit storage. These bits
-would be masked out when the pointer was returned to the user.  This
+low-order bits of the pointer for full-bit storage. These bits
+would be masked out when the pointer is returned to the user.  This
 allows us to synchronize any data type through one level of
@@
 
-%\paragraph{Single-node system.}
-In our single-node implementation, we prefetch and yield before performing
+\paragraph{Single-node system.}
+Just as with reads and writes, we prefetch and yield before performing
 a synchronization operation. We support only the synchronization
 operations supported by our development platform; we have not
-implemented full-empty bits yet.
+implemented full-bits yet.
 
-%\paragraph{Multi-node ideas.}
-In a multi-node implementation, synchronization operations can be delegated to a memory manager
+\paragraph{Multi-node ideas.}
+Synchronization operations can also be delegated to a memory manager
 thread or accelerator. Sychronization on remote data would be
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
   \vspace{-.1in}
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
   \begin{center}

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
some sec4/ref/conclusion changes

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-Our goal is to evaluate the feasibility of our proposed runtime.
-We want to determine two things: whether coroutines can generate memory concurrency without incurring minimal performance overhead, and whether the system can support the level of concurrency required to tolerate the latency that will be present in a multi-node system.
+Our goal is to evaluate the feasibility of our proposed runtime. We
+want to determine two things: whether coroutines can generate memory
+concurrency while incurring minimal performance overhead, and whether the system can support the level of concurrency required to tolerate the latency that will be present in a multi-node system.
 
@@
 Our evaulation consists of two parts. First, we demonstrate that the
-runtime system can achieve the same performance as when there is
-explicit memory concurrency available. Second, we look at the runtime
+runtime system can achieve the same performance as when 
+explicit memory concurrency is available. Second, we look at the runtime
 within the multi-node picture and show that it can tolerate large
@@
 
-\subsection{Single-node}
+\subsection{Single-node performance}
 
@@
 
-\subsubsection{Base memory system performance}
+\subsection{Base memory system performance}
 \label{subsubsec:evalsinglebase}
@@
 coroutine, so the only source of memory concurrency is the use of
-coroutines. We are able to obtain a rate of 275 \mrps with 48 concurrent misses, or 8 coroutines per core. These
+coroutines. We are able to obtain a rate of 275 \mrps\ with 48 concurrent misses, or 8 coroutines per core. These
 results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36.

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex

@@
 coroutine is reactivated, it assumes the requested data is available
-and issues a read.  This is a disadvantage of the simple
-implementation; if the data is not yet received and in cache, the core
+and issues a read.  If the data is not yet received and in cache, the core
 (and all its coroutines) can stall.
@@
 operations supported by our development platform; we have not
-implemented full-empty bits yet.
+implemented full-bits yet.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   softxmt-hotpar2011.tex
spelling

@@
 
-% Whereas in the PGAS model coherency of global copies is implictly
+% Whereas in the PGAS model coherency of global copies is implicitly
 % maintained across the system, % \todo{cite pgas?}
@@
 \item[Low overhead synchronization:] 
-  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
+%  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
+  A thread that is synchronizing must yield its core to allow others to make progress.
 \end{description}
@@
 In a multi-node implementation, synchronization operations can be delegated to a memory manager
-thread or accelerator. Sychronization on remote data would be
+thread or accelerator. Synchronization on remote data would be
 performed by the remote memory manager. This may simplify the
@@
 
-Our evaulation consists of two parts. First, we demonstrate that the
+Our evaluation consists of two parts. First, we demonstrate that the
 runtime system can achieve the same performance as when 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
subsubroutine error
micro chages related work

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-%\paragraph{Single-node implementation.}  
-In our single-node implementation, we implemented coroutines
-as described, with a round-robin scheduler. Coroutines do local work,
-issue a remote prefetch, and switch to the next in turn.  When the
+\paragraph{Single-node implementation.}  We implemented coroutines
+with a round-robin scheduler. Coroutines do local work,
+issue a remote prefetch, and yield. When a
 coroutine is reactivated, it assumes the requested data is available
@@
 
-Our goal is to evaluate the feasibility of our proposed runtime. We
-want to determine two things: whether coroutines can generate memory
-concurrency while incurring minimal performance overhead, and whether the system can support the level of concurrency required to tolerate the latency that will be present in a multi-node system.
+Our goal is to evaluate the feasibility of our proposed runtime.
+We want to determine two things: whether coroutines can generate memory concurrency without incurring minimal performance overhead, and whether the system can support the level of concurrency required to tolerate the latency that will be present in a multi-node system.
 
@@
 Our evaulation consists of two parts. First, we demonstrate that the
-runtime system can achieve the same performance as when 
-explicit memory concurrency is available. Second, we look at the runtime
+runtime system can achieve the same performance as when there is
+explicit memory concurrency available. Second, we look at the runtime
 within the multi-node picture and show that it can tolerate large
@@
 
-\subsection{Single-node performance}
+\subsection{Single-node}
 
@@
 
-\subsection{Base memory system performance}
+\subsubsection{Base memory system performance}
 \label{subsubsec:evalsinglebase}
@@
 coroutine, so the only source of memory concurrency is the use of
-coroutines. We are able to obtain a rate of 275 \mrps\ with 48 concurrent misses, or 8 coroutines per core. These
+coroutines. We are able to obtain a rate of 275 \mrps with 48 concurrent misses, or 8 coroutines per core. These
 results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36.
@@
   \end{center}
-  \vspace{-0.3in}
   \caption{Cache pressure with coroutines, on six cores.}
@@
 decreases for large working set sizes and large numbers of coroutines,
-we see speedup as the number of coroutines increases up to 128KB working set size with 8 coroutines per
+we see speedup as number coroutines increases for up to 128KB working set size with 8 coroutines per
 core.
@@
 This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, but at least 2
-concurrent references per coroutine are required. Running with one
+concurrent references per coroutine are required. Running with a single
 reference per coroutine achieves 236 \mrps.
 
-These results suggest that the runtime should be able to tolerate
-large latency well enough to exceed the max issue rates of both modern
-network interfaces (about 100 \mrps) \cite{mellanox:press} and the QPI
-link used to reach the network interface.
+These results suggest that the runtime should be able to tolerate large latency well enough to exceed the max issue rates of both modern network interfaces (about 100 \mrps) \cite{mellanox:press} and the QPI link that connects the chip to the network interface.
 
@@
 
-There has also been much work on user level threading, including the First-Class
-User Level Threading project \cite{ult} and Capriccio \cite{capriccio}. We have a
+There has also been much work on user level threading; the First-Class
+User Level Threading project \cite{ult} is one example. We have a
 different goal; we want many lightweight contexts to tolerate memory
@@
 obtain performance by minimizing references to remote nodes, whereas we
-optimize for this case.
+design for remote references.
 
@@
 
-We presented our plans for building a system for large graph
+We presented our plans on building a system for large graph
 computation using commodity parts. We leverage decades-old ideas on
-using a large amount of concurrency to tolerate latencies; but we do so
+using a large ammount of concurrency to tolerate latencies; but do so
 mostly in software. We developed a runtime system for latency

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
no space

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 We treat context switches as function calls, as in~\cite{charm}. This allows us 
-to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers:  the size of a coroutine context is hardly more than the stack space needed to save these registers.
-We minimize
-context switch time by doing all switching and scheduling in user space (kernel traps are at least an order of magnitude slower.)
+to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers.  We minimize
+context switch time by doing all switching and scheduling in user space.
 

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
almost

@@
 tolerance and our results showed that it is able to manage enough
-concurrency to tolerate latencies long enough to amortize the cost of
+concurrency to tolerate long latencies to amortize the cost of
 expensive accesses to our large global memory system (high-latency,

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
modified:   softxmt-hotpar2011.tex

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
-  Dan Grossman, Mark Oskin, Carl Ebeling, Simon Kahan, Luis
-  Ceze}{University of Washington}{\{nelson, bdmyers, ahh, preston,
+\authorinfo{Jacob Nelson$^{\dagger}$, Brandon Myers$^{\dagger}$, A.H. Hunter$^{\dagger}$, Preston Briggs$^{\dagger}$,
+  Dan Grossman$^{\dagger}$, Mark Oskin$^{\dagger}$, Carl Ebeling$^{\dagger}$, Simon Kahan$^{{\dagger \ddagger}}$, Luis
+  Ceze$^{\dagger}$}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{\{nelson, bdmyers, ahh, preston,
   djg, oskin, ebeling, skahan, luisceze\}@cs.washington.edu}
@@
 
-% Whereas in the PGAS model coherency of global copies is implicitly
+% Whereas in the PGAS model coherency of global copies is implictly
 % maintained across the system, % \todo{cite pgas?}
@@
 We treat context switches as function calls, as in~\cite{charm}. This allows us 
-to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers.  We minimize
-context switch time by doing all switching and scheduling in user space.
+to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers:  the size of a coroutine context is hardly more than the stack space needed to save these registers.
+We minimize
+context switch time by doing all switching and scheduling in user space (kernel traps are at least an order of magnitude slower.)
 
@@
 In a multi-node implementation, synchronization operations can be delegated to a memory manager
-thread or accelerator. Synchronization on remote data would be
+thread or accelerator. Sychronization on remote data would be
 performed by the remote memory manager. This may simplify the
@@
 
-Our evaluation consists of two parts. First, we demonstrate that the
+Our evaulation consists of two parts. First, we demonstrate that the
 runtime system can achieve the same performance as when 
@@
 obtain performance by minimizing references to remote nodes, whereas we
-design for remote references.
+optimize for this case.
 

### 2011-03-09 doc/hotpar2011/softxmt-hotpar2011.tex
new author order

--- 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
-\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
-  Dan Grossman, Mark Oskin, Carl Ebeling, Simon Kahan, Luis
-  Ceze}{University of Washington}{\{nelson, bdmyers, ahh, preston,
-  djg, oskin, ebeling, skahan, luisceze\}@cs.washington.edu}
+\authorinfo{Jacob Nelson$^{\dagger}$, Brandon Myers$^{\dagger}$, A.H. Hunter$^{\dagger}$, Luis
+  Ceze$^{\dagger}$,
+  Dan Grossman$^{\dagger}$, Mark Oskin$^{\dagger}$, Carl Ebeling$^{\dagger}$, Simon Kahan$^{{\dagger \ddagger}}$, Preston Briggs$^{\dagger}$}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{\{nelson, bdmyers, ahh, luisceze, djg, oskin, ebeling, skahan, preston\}@cs.washington.edu}
 

### 2011-04-01 jumpthreads/illustration.tex
An illustration of the technique (documentation)

\documentclass{article}
\usepackage{tikz}
\usetikzlibrary{positioning}
\thispagestyle{empty}
\begin{document}
\begin{tikzpicture}[node distance=0.1cm]
  \node [,circle] (n1) at (0,0) { };
  \node [text width=4.5cm,above] at (n1.north) {
    \tt while (pointer1 != NULL) \{\\
    \ \ \ \ prefetch(pointer1);
  };
  \node [,circle] (s1) [below=of n1] { };
  \node [text width=4.5cm,below] at (s1.south) {
    \tt\ \ \ \ pointer1 = *pointer1;\\
    \}
  };

  \node [,circle] (n2) at (6,0) { };
  \node [text width=4.5cm,above] at (n2.north) {
    \tt while (pointer2 != NULL) \{\\
    \ \ \ \ prefetch(pointer2);
  };
  \node [,circle] (s2) [below=of n2] { };
  \node [text width=4.5cm,below] at (s2.south) {
    \tt\ \ \ \ pointer2 = *pointer2;\\
    \}
  };

  \node [,circle] (n3) at (6,-4.5) { };
  \node [text width=4.5cm,above] at (n3.north) {
    \tt while (pointer3 != NULL) \{\\
    \ \ \ \ prefetch(pointer3);
  };
  \node [,circle] (s3) [below=of n3] { };
  \node [text width=4.5cm,below] at (s3.south) {
    \tt\ \ \ \ pointer3 = *pointer3;\\
    \}
  };

  \node [,circle] (n4) at (0,-4.5) { };
  \node [text width=4.5cm,above] at (n4.north) {
    \tt while (pointer4 != NULL) \{\\
    \ \ \ \ prefetch(pointer4);
  };
  \node [,circle] (s4) [below=of n4] { };
  \node [text width=4.5cm,below] at (s4.south) {
    \tt\ \ \ \ pointer4 = *pointer4;\\
    \}
  };
 \draw[->, thick] (n1.center) -- (s2.center);
 \draw[->, thick] (n3.center) -- (s4.center);
 \draw[->, thick] (n2.center) .. controls +(right:4) and +(right:4)
 .. (s3.center);
 \draw[->, thick] (n4.center) .. controls +(left:4.5) and +(left:4.5) .. (s1.center);
\end{tikzpicture}
\end{document}

### 2011-04-11 doc/hotpar2011/softxmt-hotpar2011.tex
post draft submission comments and typo fixes

@@
 %% move stuff to defs file
+
+%command inserts todo/comment
 \newcommand{\todo}[1]{{\bfseries [[#1]]}}
 %% To disable, just uncomment this line
- \renewcommand{\todo}[1]{\relax}
+% \renewcommand{\todo}[1]{\relax}
 
@@
 generate much more work than the rest, leading to load
-imbalances. Together, these properties lead to a
+imbalances. Together, these properties \todo{including power law distr?} lead to a
 difficult conclusion: each edge traversal is likely to lead to a very
@@
 which quickly resumes execution of another computation.  Concurrency
-is also used---via aggregation of requests to a common node
+is also used\todo{vague}---via aggregation of requests to a common node
 destination---to make more efficient use of network bandwidth and in
@@
 %  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
-  A thread that is synchronizing must yield its core to allow others to make progress.
+  A thread that is synchronizing must yield its core to allow others to make progress. \todo{repetitive with the previous item}
 \end{description}
@@
 %callee save registers weren't going to be clobbered.
+\todo{decide whether to change these 2 paragraphs to be inclusive of other lightweight implementations, like jump table}
 We treat context switches as function calls, as in~\cite{charm}. This allows us 
@@
   read complete} blocks until the data is available. A write operation
-follow the same pattern.
+follows the same pattern.
 
@@
 We
-also observe many points that do not follow the trend are below the max, but these results are repeatable. This effect is harder to explain, but we suspect it is
+also observe many points that do not follow the trend are below the max, and these results are repeatable. This effect is harder to explain, but we suspect it is
 due to resource contention in the cores' pipelines once the memory
@@
 tolerance and our results showed that it is able to manage enough
-concurrency to tolerate long latencies to amortize the cost of
+concurrency to tolerate long latencies to hide the cost of
 expensive accesses to our large global memory system (high-latency,

### 2011-04-19 doc/hotpar2011/softxmt-hotpar2011.tex
remove page numbers
some edits all over. will do another pass after eval is done

@@
 
-Figure~\ref{fig:system-overview} shows the structure of our
-proposal. It is composed of multiple nodes built with commodity
-processors, communicating over an Infiniband network. We add two
-components: a runtime system, responsible for executing the user's
-threads, and a global memory manager, responsible for facilitating
-memory requests to the non-coherent global memory shared across the
-nodes.  We are exploring a mix of hardware (FPGA) and software
-implementations of the memory manager.
+Figure~\ref{fig:system-overview} shows an overview of our proposal. It
+is composed of multiple nodes built with commodity processors,
+communicating over an Infiniband network. We add two components: a
+runtime system, responsible for executing and managing user threads,
+and a global memory manager, responsible for facilitating memory
+requests to the global memory space shared across the nodes.
+We are exploring a mix of hardware (FPGA) and software implementations
+of the memory manager.
 
@@
 of pointer chasing on a remote node by artificially increasing
-latencies to several micro-seconds.
+latencies to several microseconds.
 
@@
 reference) initiates the operation and then yields to the scheduler,
-which quickly resumes execution of another computation.  Concurrency
-is also used\todo{vague}---via aggregation of requests to a common node
-destination---to make more efficient use of network bandwidth and in
-some cases reduce bandwidth demands.
+which quickly resumes execution of another computation. Concurrency is
+also used to make more efficient use of network bandwidth: it
+enables aggregation of short memory requests into coarser requests,
+which leads to large network messages and consequently better
+bandwidth utilization.
 
-Synchronization on locations in the node private address space is the
-same as on conventional systems.  Synchronization on global addresses
-works differently: we provide atomic operations such as {\tt
-  int\_fetch\_add} as well as operations using full-bits, as on
+% \todo{vague}---via aggregation of requests to a common node
+% destination--- and in
+% some cases reduce bandwidth demands.
+
+Synchronization on locations in the private address space of a node
+works the the same way as on conventional systems.  Synchronization on
+global addresses works differently: we provide atomic operations such
+as {\tt int\_fetch\_add} as well as operations using full-bits, as on
 the Cray XMT.  As is true for other long latency operations,
@@
 another computation while the synchronization is accomplished outside
-of the pipeline.  In addition, the implementation can exploit
-aggregation when concurrency is abundant to increase the throughput of
-global synchronization.
+of the pipeline.  In addition, the system can also exploit aggregation
+when concurrency is abundant to increase the throughput of global
+synchronization.
 
@@
 %  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
-  A thread that is synchronizing must yield its core to allow others to make progress. \todo{repetitive with the previous item}
+  Similarly to a regular memory access, long-latency synchronization
+  should also cause a thread to yield its core  and allow others to make
+  progress. %\todo{repetitive with the previous item}
 \end{description}
@@
 %callee save registers weren't going to be clobbered.
-\todo{decide whether to change these 2 paragraphs to be inclusive of other lightweight implementations, like jump table}
-We treat context switches as function calls, as in~\cite{charm}. This allows us 
-to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers.  We minimize
-context switch time by doing all switching and scheduling in user space.
+\todo{decide whether to change these 2 paragraphs to be inclusive of
+  other lightweight implementations, like jump table. I support the
+  idea of making it more inclusive but I can't think of what to
+  actually change.}  We treat context switches as function calls, as
+in~\cite{charm}. This allows us to depend on the compiler to save and
+restore caller-save registers while we explicitly save and restore all
+the callee-save registers.  We minimize context switch time by doing
+all switching and scheduling in user space.
 

### 2011-04-19 doc/hotpar2011/softxmt-hotpar2011.tex
remove ALL page numbers
code segments polished and clarified

@@
 \maketitle
+
+% Use the following at camera-ready time to suppress page numbers.
+% Comment it out when you first submit the paper for review.
+\thispagestyle{empty}
+\pagestyle{empty}
 
@@
   while (count-- > 0) \{
-     prefetch(&(list1->next));
-     switch();
-     list1 = read(&(list1->next));
+     readIssue(&(list->next));
+     yield();
+     list = readCompete(&(list->next));
  \}
@@
   while (count-- > 0) \{
-    prefetch(&(list1->next));
-    switch();
-    list1 = read(&(list1->next));
+    readIssue(&(remoteList->next));
+    yield();
+    remoteList = readComplete(&(remoteList->next));
     for( i in 1 to num_local_updates ) \{
-      local->data++;
-      local = local->next;
+      localList->data++;
+      localList = localList->next;
    \} \}
@@
 
-To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.
+To evaluate the potential for the lightweight contexts to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.
 
@@
 
-To simulate the performance of pointer chasing in a multi-node system,
-we modified the scheduler of our coroutine library to include a delay
+To simulate the performance of pointer chasing in a multi-node system
+we access the other chip's memory and also have the scheduler of our coroutine library include a delay
 before a coroutine can be rescheduled, imitating the network transit
-delay even though we are referencing local memory. We assume 1.1
-$\mu$s interface and 400 ns switch latencies in each direction
+delay even though we are referencing local memory. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
 \cite{infiniband, mellanox:site}, and thus estimate a 3 $\mu$s round-trip

### 2011-04-19 doc/hotpar2011/softxmt-hotpar2011.tex
insert new cache,delaynet figs (not final versions); draft their captions; rewrite cache pressure and latency experiment parts of eval, but still needs polish.  Section 4.1.2 still needs changes.

@@
 concurrent offered references.
-We see a maximum rate of 277 million references per second (\mrps), which agrees with the measurements found by \cite{Mandal:2010}.
+We see a maximum rate of 277 million references per second (\mrps), which agrees with the measurements in \cite{Mandal:2010}.
 This rate is achieved when the number of offered references is 36.
@@
 memory concurrency; the maximum reference rate for a single core is
-107M. We believe this limit is due to the core having only enough {\em line fill
+107 \mrps. We believe this limit is due to the core having only enough {\em line fill
   buffers} \cite{nehalem:arch} to store 10 concurrent private cache misses.
@@
   \begin{center}
-    \includegraphics[width=0.42\textwidth]{figures/cache-pressure-edited.pdf}
+    \includegraphics[width=0.45\textwidth]{figures/cachepressurenewview-edited.pdf}
   \end{center}
-  \vspace{-0.3in}
-  \caption{Cache pressure with coroutines, on six cores.}
+  \vspace{0.0in}
+  \caption{Cache pressure with coroutines, on six cores.
+The vertical axis is remote reference rate.
+(a) with 8 coroutines per core.
+(b) with 48 coroutines per core, about the number required for the network rate in simulated delay experiments.
+}
   \label{fig:cache-pressure}
@@
 
-We varied the size of the local working set from 2KB
-to 4MB, and ran both 0 and 32 local updates for each pointer
-traversal. We measured the overall reference rate, including both the local updates and remote references.
+We varied the size of the per-thread local working set from 2KB
+to 4MB, and ran 0 to 32 local updates for each pointer
+traversal. We measured the remote reference rate (i.e. accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).
 
-Figure~\ref{fig:cache-pressure} shows the result. As we would expect,
-when there are no references to the local structures, the reference rates grow with the number of coroutines just as in
-Figure~\ref{fig:multi-green}. For 32 local updates per traversal, while performance
-decreases for large working set sizes and large numbers of coroutines,
-we see speedup as the number of coroutines increases up to 128KB working set size with 8 coroutines per
-core.
-\todo{preempt questions about drop at 16 which is much less than the about 50 needed per core by saying that the drop is likely the same drop effect seen earlier, which is likely to be prefetch squashing since it happens at number of line fill buffers}
+Figure~\ref{fig:cache-pressure}a shows the result for 8 coroutines per core. 
+With 1 to 4 updates per remote reference for smaller working sets, the runtime achieves near maximum throughput.
+As working set size increases, cache misses in the local space become a bottleneck.
+When there are several updates to large local spaces, the data of other coroutines is likely to be kicked out and performance suffers.
+
+%As we would expect, when there are no references to the local structures, the reference rates grow with the number of coroutines, just as in Figure~\ref{fig:multi-green}.
+For 32 local updates per traversal, while performance decreases for large working set sizes and large numbers of coroutines, we see speedup as the number of coroutines increases up to 128KB working set size with 8 coroutines per core.
 
@@
 To simulate the performance of pointer chasing in a multi-node system
-we access the other chip's memory and also have the scheduler of our coroutine library include a delay
+we referenced the second chip's memory and also had coroutine scheduler include a delay
 before a coroutine can be rescheduled, imitating the network transit
-delay even though we are referencing local memory. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
+delay even though we are accessing local memory. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
 \cite{infiniband, mellanox:site}, and thus estimate a 3 $\mu$s round-trip
@@
 %simulated delay plot
+%TODO: replace with the thru QPI version
 \begin{figure}[h]
   \begin{center}
-    \includegraphics[width=0.47\textwidth]{figures/multi-green-delay7800-edited.pdf}
+    \includegraphics[width=0.47\textwidth]{figures/delay7800-remote-edited.pdf}
   \end{center}
   \vspace{-0.3in}
-  \caption{Pointer chasing with simulated network delay. Each core ran up to 167 coroutines, and each coroutine made 1 to 4 concurrent references.}
+  \caption{Pointer chasing across QPI link and with simulated network delay. Used six cores and varied the number of coroutines per core.}
   \label{fig:network-delay}
@@
 The results of the experiment are shown in
-Figure~\ref{fig:network-delay}. We are still able to reach a
-near-maximum rate of 273 \mrps\footnote{The throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support.
-This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, but at least 2
-concurrent references per coroutine are required. Running with one
-reference per coroutine achieves 236 \mrps.
+Figure~\ref{fig:network-delay}. The runtime can still reach a
+maximum rate of 173 \mrps\footnote{The throughput keeps increasing up to around 564 total concurrent references offered (at which point QPI is saturated), far more than the 36 we know the chip to support.
+This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, saturating the QPI link.
+It reaches the estimated network rate of 100 \mrps with about 53 coroutines per core.
 
@@
 link used to reach the network interface.
+\todo{integrate this paragraph with previous ones instead}
+
+Figure~\ref{fig:cache-pressure}b shows the results of the cache pressure experiment for 48 coroutines per core, about the number of contexts needed to reach the network rate in the delay experiment.
+We observe that when performing up to 8 local updates per remote reference, smaller working sets still allow the desired throughput.
+When performing 1 to 2 updates, we can have working sets of up to 32KB per coroutine while maintaining throughput.
 
@@
 %in a remote memory.
-
-\todo{also have results that remote cores over QPI plus delay 3us can make it up to the 175M, with sufficient number of coroutines}
 
@@
 User Level Threading project \cite{ult} and Capriccio \cite{capriccio}. We have a
-different goal; we want many lightweight contexts to tolerate memory
+different goal \todo{we didnt say what their goal is}; we want many lightweight contexts to tolerate memory
 latency.

### 2011-04-19 doc/hotpar2011/softxmt-hotpar2011.tex
terrible rework of intro

--- 2011-04-19 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 
+% Use the following at camera-ready time to suppress page numbers.
+% Comment it out when you first submit the paper for review.
+\thispagestyle{empty}
+\pagestyle{empty}
+
 \begin{abstract}
@@
 graphs. Unfortunately, the computational cost of these applications
-gets quickly out of hand. For example, in January 2011, Facebook had
-500 million active users with an average of 130 friends each
-\cite{Facebook:2011p91}. Its friend suggestion algorithm in 2010 ran on
-  a 40-node cluster with 72GB memory per node, and, still, new
-  suggestions could only be computed every two days
-  \cite{Backstrom:2010p90}. Speeding up these applications at a low
-  cost would have a large impact in how we deal with large data-sets and make them even more valuable.
+gets quickly out of hand. A Yahoo snapshot of the World Wide Web graph
+from 2002 has more than 1.4 billion vertices and 6.6 billion edges
+\cite{pegasus, hadi}; the social network Facebook has over 500 million
+active users with 130 friends each on average
+\cite{Facebook:2011p91}. Applications that process these large
+datasets are slow and power-hungry. Speeding up such applications at a
+low cost would have a large impact in how we deal with large data-sets
+and make them even more valuable.
+
+%For example, in January 2011, Facebook had
+%500 million active users with an average of 130 friends each
+%\cite{Facebook:2011p91}. Its friend suggestion algorithm in 2010 ran on
+%  a 40-node cluster with 72GB memory per node, and, still, new
+%  suggestions could only be computed every two days
+%  \cite{Backstrom:2010p90}. Speeding up these applications at a low
+%  cost would have a large impact in how we deal with large data-sets and make them even more valuable.
 
@@
 low-diameter, power law graphs: this combination makes extracting
-performance difficult. They do not fit in a single
-commodity machine's memory. Their low diameter means that they are
-difficult to lay out with locality: every vertex needs to be near every
-other vertex. Their power law distribution means that a few vertices
-generate much more work than the rest, leading to load
-imbalances. Together, these properties \todo{including power law distr?} lead to a
-difficult conclusion: each edge traversal is likely to lead to a very
-expensive inter-node communication.
+performance difficult. The Web graph mentioned eariler requires 120GB
+of storage \cite{pegasus}, difficult to fit in a single commodity machine's
+memory. It has an average diameter of 15.6 \cite{hadi}: this low diameter makes it
+difficult to lay out with locality, or to partition across a
+cluster. Graphs with power law distributions are also difficult to
+partition in a balanced way \cite{lang, leskovec}; their few
+high-degree vertices also generate much more work than the rest,
+leading to load imbalances.
+
+%Together, these properties \todo{including power law distr?} lead to a
+%difficult conclusion: each edge traversal is likely to lead to a very
+%expensive inter-node communication.
 
@@
   Similarly to a regular memory access, long-latency synchronization
-  should also cause a thread to yield its core  and allow others to make
+  should also cause a thread to yield and allow others to make
   progress. %\todo{repetitive with the previous item}

### 2011-04-19 doc/hotpar2011/softxmt-hotpar2011.tex
reworked the network latency and cache pressure eval sections.  Change Mref/s command to properly typeset units.

@@
 % \renewcommand{\todo}[1]{\relax}
-
-\newcommand{\mrps}[0]{Mref/s}
 
@@
 \usepackage{mdwlist}
+\usepackage{units}
+
+% Mref/s
+\newcommand{\mrps}[1]{\unit[#1]{Mref/s}}
 
@@
 concurrent offered references.
-We see a maximum rate of 277 million references per second (\mrps), which agrees with the measurements in \cite{Mandal:2010}.
+We see a maximum rate of 277 million references per second (\mrps{}), which agrees with the measurements in \cite{Mandal:2010}.
 This rate is achieved when the number of offered references is 36.
@@
 memory concurrency; the maximum reference rate for a single core is
-107 \mrps. We believe this limit is due to the core having only enough {\em line fill
+\mrps{107}. We believe this limit is due to the core having only enough {\em line fill
   buffers} \cite{nehalem:arch} to store 10 concurrent private cache misses.
@@
 both processors were traversing lists allocated in the first processor's
-memory. With this configuration, we observed a maximum rate of 360
-\mrps.
+memory.
+With this configuration, we observed a maximum rate of \mrps{360}.
 We believe this difference is due to the chip having only 32 buffers in its {\em Global Queue} \cite{nehalem:perf} for tracking concurrently-executing read misses from all the cores' private caches.
@@
 coroutine, so the only source of memory concurrency is the use of
-coroutines. We are able to obtain a rate of 275 \mrps\ with 48 concurrent misses, or 8 coroutines per core. These
+coroutines. We are able to obtain a rate of \mrps{275} with 48 concurrent misses, or 8 coroutines per core. These
 results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36.
@@
 later prefetches squashing earlier ones in the line fill buffers.
-In the real runtime we expect this effect will not exist, since prefetch will not be implemented with built-in non-temporal prefetch instructions.
+In the real runtime we expect this effect will not exist, since prefetch will not be implemented with built-in non-temporal prefetch instructions.\todo{pull that sentence and replace with something about no-prefetch results}.
 We
@@
 
-We varied the size of the per-thread local working set from 2KB
+We varied the size of the per-coroutine local working set from 2KB
 to 4MB, and ran 0 to 32 local updates for each pointer
@@
 As working set size increases, cache misses in the local space become a bottleneck.
-When there are several updates to large local spaces, the data of other coroutines is likely to be kicked out and performance suffers.
+When there are several updates and large working sets, the data of other coroutines is likely to be evicted from cache and performance suffers.
 
@@
 To do this, we allocated the lists in the second chip's memory and ran the traversals on the first chip's cores.
-We found that use of the QPI link limits us to 175 \mrps.
+We found that use of the QPI link limits us to \mrps{175}.
 
@@
 To simulate the performance of pointer chasing in a multi-node system
-we referenced the second chip's memory and also had coroutine scheduler include a delay
+we referenced the second chip's memory and also had our coroutine scheduler include a delay
 before a coroutine can be rescheduled, imitating the network transit
-delay even though we are accessing local memory. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
+delay. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
 \cite{infiniband, mellanox:site}, and thus estimate a 3 $\mu$s round-trip
-delay.
+delay. In analyzing the results, we assume a network issue rate of \mrps{100}, based on modern network interfaces \cite{mellanox:press}.
 
@@
   \vspace{-0.3in}
-  \caption{Pointer chasing across QPI link and with simulated network delay. Used six cores and varied the number of coroutines per core.}
+  \caption{Pointer chasing across QPI link and with simulated network delay of $3\mu s$, six cores.
+Using six cores, 53 coroutines per core are required to reach the assumed network rate of \mrps{100}.}
   \label{fig:network-delay}
@@
 
-The results of the experiment are shown in
+The results of the experiment with six cores are shown in
 Figure~\ref{fig:network-delay}. The runtime can still reach a
-maximum rate of 173 \mrps\footnote{The throughput keeps increasing up to around 564 total concurrent references offered (at which point QPI is saturated), far more than the 36 we know the chip to support.
+maximum rate of \mrps{173}\footnote{The throughput keeps increasing up to around 564 total concurrent references offered (at which point QPI is saturated), far more than the 36 we know the chip to support.
 This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, saturating the QPI link.
-It reaches the estimated network rate of 100 \mrps with about 53 coroutines per core.
+Little's law predicts that 300 concurrent references are required to achieve the estimated network rate of \mrps{100} when there is $3\mu s$ latency.
+We observe this: \mrps{100} is reached with about 53 coroutines per core.
+Although there may be other overheads in a full runtime, this experiment suggests there is potential to tolerate multi-node system latencies.
 
-These results suggest that the runtime should be able to tolerate
-large latency well enough to exceed the max issue rates of both modern
-network interfaces (about 100 \mrps) \cite{mellanox:press} and the QPI
-link used to reach the network interface.
-\todo{integrate this paragraph with previous ones instead}
-
-Figure~\ref{fig:cache-pressure}b shows the results of the cache pressure experiment for 48 coroutines per core, about the number of contexts needed to reach the network rate in the delay experiment.
+Figure~\ref{fig:cache-pressure}b shows the results of the cache pressure experiment for supporting 48 coroutines per core, about the number of contexts needed to tolerate latency in the network experiment.
 We observe that when performing up to 8 local updates per remote reference, smaller working sets still allow the desired throughput.
-When performing 1 to 2 updates, we can have working sets of up to 32KB per coroutine while maintaining throughput.
+When performing 1 to 2 updates, the runtime can maintain throughput for working sets of up to 32KB.
 

### 2011-04-19 doc/hotpar2011/softxmt-hotpar2011.tex
minor eval change

@@
 Little's law predicts that 300 concurrent references are required to achieve the estimated network rate of \mrps{100} when there is $3\mu s$ latency.
-We observe this: \mrps{100} is reached with about 53 coroutines per core.
-Although there may be other overheads in a full runtime, this experiment suggests there is potential to tolerate multi-node system latencies.
+The results agree: \mrps{100} is reached with about 53 coroutines per core.
+Although there will be other overheads in a full runtime, this experiment suggests there is potential to tolerate multi-node system latencies.
 

### 2011-04-19 doc/hotpar2011/softxmt-hotpar2011.tex
eval: prefetches paragraph revised
rewrite some fig captions

@@
 \usepackage{fancyvrb}
+\usepackage{fancyhdr}
 \usepackage{alltt}
@@
 	\end{center}
-	\caption{Throughput of listwalk versus total number of
-          concurrent references offered by the cores. Data points for
-          running 1 core are shown as circles, and data points for
-          running 6 cores are shown as diamonds. 
+	\caption{Pointer chasing throughput versus total number of
+          concurrent references offered by the cores. Total concurrent references offered is \emph{number of lists per core * number of cores}.
 %Notice that the throughput for a single core levels out at 10
@@
 later prefetches squashing earlier ones in the line fill buffers.
-In the real runtime we expect this effect will not exist, since prefetch will not be implemented with built-in non-temporal prefetch instructions.\todo{pull that sentence and replace with something about no-prefetch results}.
-We
-also observe many points that do not follow the trend are below the max, and these results are repeatable. This effect is harder to explain, but we suspect it is
+In the multi-node runtime, \texttt{readIssue} will not be implemented with a prefetch instruction.
+The single-node coroutine library does depend on prefetches: other experiments show that without them, the overhead of context switches limits concurrency when there are more than two coroutines per core.
+Lastly, we observe repeatable data points that are off the trend and fall below the max.
+This effect is harder to explain, but we suspect it may be
 due to resource contention in the cores' pipelines once the memory

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
new intro, more references

@@
 graphs. Unfortunately, the computational cost of these applications
-gets quickly out of hand. A Yahoo snapshot of the World Wide Web graph
-from 2002 has more than 1.4 billion vertices and 6.6 billion edges
-\cite{pegasus, hadi}; the social network Facebook has over 500 million
-active users with 130 friends each on average
-\cite{Facebook:2011p91}. Applications that process these large
-datasets are slow and power-hungry. Speeding up such applications at a
-low cost would have a large impact in how we deal with large data-sets
-and make them even more valuable.
+gets quickly out of hand. Tools to analyze social networks and query
+semantic graphs with billions of entities exist
+today~\cite{joslyn:hpsf, ediger:msda, brocheler:cosi, pegasus,
+  pregel}, but graphs of interest to defense applications are expected
+have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
+up these applications at a low cost would have a large impact in how
+we deal with large data-sets and make them even more valuable.
 
+%2002 snapshot of the web has 1.4 billion vertices and 6.6 billion
+%edges \cite{pegasus}. Facebook has over 500M active users who interact
+%with 900M groups, events, and communities; the average user has 30 friends \cite{Facebook:2011p91}. 
 %For example, in January 2011, Facebook had
@@
 
+% A Yahoo snapshot of the World Wide Web graph
+% from 2002 has more than 1.4 billion vertices and 6.6 billion edges
+% \cite{pegasus, hadi}; the social network Facebook has over 500 million
+% active users with 130 friends each on average
+% \cite{Facebook:2011p91}. Applications that process these large
+% datasets are slow and power-hungry. Speeding up such applications at a
+% low cost would have a large impact in how we deal with large data-sets
+% and make them even more valuable.
+
 % In January 2011, Facebook had 500 million active users with an average
@@
 
+%  The most interesting computational challenge comes from large,
+%  low-diameter, power law graphs: this combination makes extracting
+% performance difficult. They do not fit in a single
+% commodity machine's memory. Their low diameter means that they are
+% difficult to lay out with locality: every vertex needs to be near every
+% other vertex. Their power law distribution means that a few vertices
+% generate much more work than the rest, leading to load
+% imbalances. Together, these properties \todo{including power law distr?} lead to a
+% difficult conclusion: each edge traversal is likely to lead to a very
+% expensive inter-node communication.
+
 The most interesting computational challenge comes from large,
 low-diameter, power law graphs: this combination makes extracting
-performance difficult. The Web graph mentioned eariler requires 120GB
-of storage \cite{pegasus}, difficult to fit in a single commodity machine's
-memory. It has an average diameter of 15.6 \cite{hadi}: this low diameter makes it
-difficult to lay out with locality, or to partition across a
-cluster. Graphs with power law distributions are also difficult to
-partition in a balanced way \cite{lang, leskovec}; their few
-high-degree vertices also generate much more work than the rest,
-leading to load imbalances.
+performance difficult. They do not fit in a single commodity machine's
+memory. They are difficult lay out with locality, or to partition in a
+balanced way~\cite{lang, leskovec}. Expensive inter-node communication
+is a limiting factor in scalability~\cite{yoo:parallelbfs}.
+
+% Their low diameter means that they are
+% difficult to lay out with locality: every vertex needs to be near every
+% other vertex. Their power law distribution means that a few vertices
+% generate much more work than the rest, leading to load
+% imbalances. 
+% Together, these properties \todo{including power law distr?} lead to a
+% difficult conclusion: each edge traversal is likely to lead to a very
+% expensive inter-node communication.
+
+% They are difficult to partition effi
+
+% Common techniques to 
+% Often
+
+% The most interesting computational challenge comes from large,
+% low-diameter, power law graphs: this combination makes extracting
+% performance difficult. The Web graph mentioned eariler requires 120GB
+% of storage \cite{pegasus}, difficult to fit in a single commodity machine's
+% memory. It has an average diameter of 15.6 \cite{hadi}: this low diameter makes it
+% difficult to lay out with locality, or to partition across a
+% cluster. Graphs with power law distributions are also difficult to
+% partition in a balanced way \cite{lang, leskovec}; their few
+% high-degree vertices also generate much more work than the rest,
+% leading to load imbalances.
 
@@
 Multithreading is a technique that has been used successfully to
-implement efficient computations for these graphs. The Cray XMT is an
-example of such an approach: it solves the memory latency problem
-through concurrency rather than caching. Each XMT processor supports
-128 hardware contexts and 1024 outstanding memory operations, and is
-able to switch contexts every cycle \cite{tera, feo-xmt}. This ability
-comes at a cost: the XMT is an expensive, non-commodity
-machine.
+implement efficient computations for these
+graphs~\cite{bader:bfsmta}. The Cray XMT is an example of such an
+approach: it solves the memory latency problem through concurrency
+rather than caching. Each XMT processor supports 128 hardware contexts
+and 1024 outstanding memory operations, and is able to switch contexts
+every cycle~\cite{tera, feo-xmt}. This ability comes at a cost: the
+XMT is an expensive, non-commodity machine.
 
@@
 codes but is implemented with cheap commodity processors and simple
-additional support in an FPGA. This approach also has the added
+additional support in an FPGA. This approach has the added
 benefit of not sacrificing general-purpose performance.

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
fix spacings
grammer intro

@@
   pregel}, but graphs of interest to defense applications are expected
-have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
+to have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
 up these applications at a low cost would have a large impact in how
@@
 verify our runtime's overhead is acceptable. We also model the effects
-of pointer chasing on a remote node by artificially increasing
+of pointer chasing with a remote node by artificially increasing
 latencies to several microseconds.
@@
 }\end{minipage}
-%\vspace{5pt}
+\vspace{5pt}
 \caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.
@@
   \begin{center}
-    \includegraphics[width=0.45\textwidth]{figures/cachepressurenewview-edited.pdf}
+    \includegraphics[width=0.49\textwidth]{figures/cachepressurenewview-edited.pdf}
   \end{center}

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
intro one more time

--- 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
@@
 %\documentclass{acm_proc_article-sp}
-\documentclass[10pt,nocopyrightspace,preprint]{sigplanconf}
+\documentclass[10pt,nocopyrightspace]{sigplanconf}
 \usepackage{url}
@@
 \renewcommand{\baselinestretch}{0.8}
+%\renewcommand{\baselinestretch}{0}
 
@@
   Dan Grossman$^{\dagger}$, Mark Oskin$^{\dagger}$, Carl Ebeling$^{\dagger}$, Simon Kahan$^{{\dagger \ddagger}}$, Preston Briggs$^{\dagger}$}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{\{nelson, bdmyers, ahh, luisceze, djg, oskin, ebeling, skahan, preston\}@cs.washington.edu}
+
+%\lfoot{USENIX HotPar 2011, Berkeley}
 
@@
 performance difficult. They do not fit in a single commodity machine's
-memory. They are difficult lay out with locality, or to partition in a
-balanced way~\cite{lang, leskovec}. Expensive inter-node communication
-is a limiting factor in scalability~\cite{yoo:parallelbfs}.
+memory. They are difficult lay out with locality, since every vertex
+needs to be near every other vertex. They are difficult to partition
+in a balanced way~\cite{lang, leskovec}, leading to hotspots and load
+imbalance. High-latency inter-node communication is a limiting factor
+in scalability~\cite{yoo:parallelbfs}.
 
@@
 %callee save registers weren't going to be clobbered.
-\todo{decide whether to change these 2 paragraphs to be inclusive of
-  other lightweight implementations, like jump table. I support the
-  idea of making it more inclusive but I can't think of what to
-  actually change.}  We treat context switches as function calls, as
+%\todo{decide whether to change these 2 paragraphs to be inclusive of
+%  other lightweight implementations, like jump table. I support the
+%  idea of making it more inclusive but I can't think of what to
+%  actually change.}  
+We treat context switches as function calls, as
 in~\cite{charm}. This allows us to depend on the compiler to save and
@@
 \bibliographystyle{acm}
+{\small
 \bibliography{softxmt-hotpar2011}
-
+}
 \end{document}

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
wording change in QPI section
minor edits to intro

@@
 graphs. Unfortunately, the computational cost of these applications
-gets quickly out of hand. Tools to analyze social networks and query
-semantic graphs with billions of entities exist
+gets quickly out of hand. While tools to analyze social networks and
+query semantic graphs with billions of entities exist
 today~\cite{joslyn:hpsf, ediger:msda, brocheler:cosi, pegasus,
-  pregel}, but graphs of interest to defense applications are expected
-to have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
+  pregel}, graphs of interest to defense applications are expected to
+have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
 up these applications at a low cost would have a large impact in how
@@
 in a balanced way~\cite{lang, leskovec}, leading to hotspots and load
-imbalance. High-latency inter-node communication is a limiting factor
-in scalability~\cite{yoo:parallelbfs}.
+imbalance. Consequently, high-latency inter-node communication becomes
+an important limiting factor in scalability~\cite{yoo:parallelbfs}.
 

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
fix spacings and cache/net figs move to top
some modifications

@@
 %\paragraph{Multi-node ideas.}
-As we develop our multi-node implementation, we will explore how the memory manager and coroutine scheduler may
-interact, so that threads waiting on long-latency memory operations or
-synchronization operations will be scheduled soon after their
-operations have completed.
+%As we develop our multi-node implementation, we will explore how the memory manager and coroutine scheduler may
+%interact, so that threads waiting on long-latency memory operations or
+%synchronization operations will be scheduled soon after their
+%operations have completed.
+As we develop our multi-node implementation, we will improve our
+scheduler to activate threads after their long-latency operations
+have completed.
+
+% will explore how the memory manager and coroutine scheduler may
+%interact, so that threads waiting on long-latency memory operations or
+%synchronization operations will be scheduled soon after their
+%operations have completed.
 
@@
 In our single-node implementation, we use prefetch instructions for the issue operation, and regular
-blocking reads and writes for the complete operation.  Since we expect
-these global references to have low locality, we use the
-``non-temporal'' form of prefetch to avoid disturbing the state of the
-cache hierarchy \cite{intel:swdev}.
+blocking reads and writes for the complete operation.
+%  Since we expect
+%these global references to have low locality, we use the
+%``non-temporal'' form of prefetch to avoid disturbing the state of the
+%cache hierarchy \cite{intel:swdev}.
 
@@
 With either approach, the question of request aggregation will be
-important. Each memory request we make is small: 8 bytes is a likely
-size. But most networks are optimized for bulk transfers of a few
+important. Each memory request we make is small---perhaps 8
+bytes---but most networks are optimized for bulk transfers of a few
 thousand bytes. To improve network utilization, we will explore the
@@
      yield();
-     list = readCompete(&(list->next));
+     list = readComplete(&(list->next));
  \}
@@
 
-Figure~\ref{fig:multi-green} shows the result. For this
-experiment, we limited memory concurrency to one concurrent miss per
-coroutine, so the only source of memory concurrency is the use of
-coroutines. We are able to obtain a rate of \mrps{275} with 48 concurrent misses, or 8 coroutines per core. These
-results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36.
-When the number of references per coroutine is larger than 1, fewer coroutines are required to reach full performance.
+Figure~\ref{fig:multi-green} shows the result. For this experiment,
+the only source of memory concurrency is the use of coroutines. We are
+able to obtain a rate of \mrps{275} with 48 concurrent misses, or 8
+coroutines per core. More concurrent requests are required to saturate
+memory bandwidth than in the ILP-only experiment.
 
@@
 concurrent references per core exceeds 10; we believe this is due to
-later prefetches squashing earlier ones in the line fill buffers.
-In the multi-node runtime, \texttt{readIssue} will not be implemented with a prefetch instruction.
-The single-node coroutine library does depend on prefetches: other experiments show that without them, the overhead of context switches limits concurrency when there are more than two coroutines per core.
+later prefetches squashing earlier ones in the line fill buffers. In
+the multi-node runtime, \texttt{readIssue} will not be implemented
+with a prefetch instruction. The single-node coroutine library does
+depend on prefetches: further experiments show that without them, the overhead of context switches limits concurrency when there are more than two coroutines per core.
 Lastly, we observe repeatable data points that are off the trend and fall below the max.
@@
 
-Figure~\ref{fig:cache-pressure}a shows the result for 8 coroutines per core. 
-With 1 to 4 updates per remote reference for smaller working sets, the runtime achieves near maximum throughput.
-As working set size increases, cache misses in the local space become a bottleneck.
-When there are several updates and large working sets, the data of other coroutines is likely to be evicted from cache and performance suffers.
+Figure~\ref{fig:cache-pressure}a shows the result for 8 coroutines per
+core.  With 1 to 4 updates per remote reference for smaller working
+sets, the runtime achieves near maximum throughput. As working set
+size increases, cache misses in the local space become a
+bottleneck. When there are several updates to large working sets, the
+data of other coroutines is likely to be evicted from cache and
+performance suffers.
 
@@
 
-To evaluate the potential for the lightweight contexts to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.
+To evaluate the potential for the lightweight contexts to perform in
+the multi-node case, we simulated a network delay to see whether the
+runtime could tolerate the larger latency.
 
-Since the network interface is accessed over QPI, we estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine.
-To do this, we allocated the lists in the second chip's memory and ran the traversals on the first chip's cores.
-We found that use of the QPI link limits us to \mrps{175}.
+Any network communication must travel over the QPI link in our test
+system. We estimated the bandwidth of this link by allocating lists in
+the first chip's memory and traversing those lists on the second
+chip's cores. We found that use of the QPI link limits us to
+\mrps{175}.
 
@@
 User Level Threading project \cite{ult} and Capriccio \cite{capriccio}. We have a
-different goal \todo{we didnt say what their goal is}; we want many lightweight contexts to tolerate memory
+different goal; we want many lightweight contexts to tolerate memory
 latency.

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
six pages again!

@@
 \vspace{5pt}
-\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.
-Coroutine code, as in (b), might also include ILP like what (a) depends on for memory concurrency, e.g., by doing multiple prefetches to independent lists, switch, then multiple reads.}
+\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates.}
 \label{fig:code}
@@
 	\end{center}
+	\vspace{-9pt}
 	\caption{Pointer chasing throughput versus total number of
@@
   \end{center}
+  \vspace{-9pt}
   \caption{Pointer chasing with coroutines, with one reference per
@@
 % cache pressure figure
-\begin{figure}[h]
+\begin{figure}[t]
   \begin{center}
-    \includegraphics[width=0.49\textwidth]{figures/cachepressurenewview-edited.pdf}
+    \includegraphics[width=0.48\textwidth]{figures/cachepressurenewview-edited.pdf}
   \end{center}
-  \vspace{0.0in}
-  \caption{Cache pressure with coroutines, on six cores.
-The vertical axis is remote reference rate.
-(a) with 8 coroutines per core.
-(b) with 48 coroutines per core, about the number required for the network rate in simulated delay experiments.
-}
+  \vspace{-10pt} %space with fig
+  \caption{Cache pressure with coroutines, on six cores. The vertical
+    axis is remote reference rate, with (a) 8 coroutines per core, and
+    (b) 48 coroutines per core, about the number required for the
+    network rate in the simulated delay experiments. }
+  \vspace{-1.5pt} % space with text
   \label{fig:cache-pressure}
@@
 
-We varied the size of the per-coroutine local working set from 2KB
+We varied the size of the per-coroutine local working set from 2KiB
 to 4MB, and ran 0 to 32 local updates for each pointer
@@
 core.  With 1 to 4 updates per remote reference for smaller working
-sets, the runtime achieves near maximum throughput. As working set
-size increases, cache misses in the local space become a
-bottleneck. When there are several updates to large working sets, the
-data of other coroutines is likely to be evicted from cache and
-performance suffers.
+sets, the runtime achieves near maximum throughput. 
+%As working set
+%size increases, cache misses in the local space become a
+%bottleneck. 
+%When there are several updates to large working sets, 
+As the number of updates and the working set size increase, the data
+of other coroutines is likely to be evicted from cache and performance
+suffers.
 
@@
 %TODO: replace with the thru QPI version
-\begin{figure}[h]
+\begin{figure}[t]
   \begin{center}
@@
   \vspace{-0.3in}
-  \caption{Pointer chasing across QPI link and with simulated network delay of $3\mu s$, six cores.
-Using six cores, 53 coroutines per core are required to reach the assumed network rate of \mrps{100}.}
+  \caption{Pointer chasing across QPI link with simulated network delay of $3\mu s$, six cores. Using six cores, 53 coroutines per core are required to reach the assumed network rate of \mrps{100}.}
   \label{fig:network-delay}
@@
 We observe that when performing up to 8 local updates per remote reference, smaller working sets still allow the desired throughput.
-When performing 1 to 2 updates, the runtime can maintain throughput for working sets of up to 32KB.
+When performing 1 to 2 updates, the runtime can maintain throughput for working sets of up to 32KiB.
 
@@
 
-\bibliographystyle{acm}
-{\small
+%\bibliographystyle{acm}
+\bibliographystyle{plain}
+\renewcommand{\bibfont}{\footnotesize}  
 \bibliography{softxmt-hotpar2011}
-}
 \end{document}

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
final

@@
 performance difficult. They do not fit in a single commodity machine's
-memory. They are difficult lay out with locality, since every vertex
+memory. They are difficult to lay out with locality, since every vertex
 needs to be near every other vertex. They are difficult to partition
@@
 reference rate, we extended the previous experiment so that cores in
-both processors were traversing lists allocated in the first processor's
+both processors traversed lists allocated in the first processor's
 memory.
@@
 We varied the size of the per-coroutine local working set from 2KiB
-to 4MB, and ran 0 to 32 local updates for each pointer
+to 4MiB, and ran 0 to 32 local updates for each pointer
 traversal. We measured the remote reference rate (i.e. accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
final
author list

@@
 
-\authorinfo{Jacob Nelson$^{\dagger}$, Brandon Myers$^{\dagger}$, A.H. Hunter$^{\dagger}$, Luis
-  Ceze$^{\dagger}$,
-  Dan Grossman$^{\dagger}$, Mark Oskin$^{\dagger}$, Carl Ebeling$^{\dagger}$, Simon Kahan$^{{\dagger \ddagger}}$, Preston Briggs$^{\dagger}$}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{\{nelson, bdmyers, ahh, luisceze, djg, oskin, ebeling, skahan, preston\}@cs.washington.edu}
+\authorinfo{Jacob Nelson$^{\dagger}$,
+  Brandon Myers$^{\dagger}$,
+  A.H. Hunter$^{\dagger}$,
+  Preston Briggs$^{\dagger}$
+  Luis Ceze$^{\dagger}$,
+  Carl Ebeling$^{\dagger}$,\\
+  Dan Grossman$^{\dagger}$,
+  Simon Kahan$^{{\dagger \ddagger}}$,
+  Mark Oskin$^{\dagger}$,
+}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
+  \{nelson, bdmyers, ahh, preston, luisceze, ebeling, djg, skahan, oskin\}@cs.washington.edu}
 
@@
 the first chip's memory and traversing those lists on the second
-chip's cores. We found that use of the QPI link limits us to
+chip's cores. We found that use of the QPI link limits throughput to
 \mrps{175}.

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
most of preston's changes

@@
 %\columnsep 1pc
-\renewcommand{\baselinestretch}{0.8}
 %\renewcommand{\baselinestretch}{0}
@@
   Brandon Myers$^{\dagger}$,
-  A.H. Hunter$^{\dagger}$,
+  A. H. Hunter$^{\dagger}$,
   Preston Briggs$^{\dagger}$
@@
    algorithms exhibit little locality and therefore present significant
-   performance challenges. Hardware multithreading systems (e.g, Cray
-   XMT) show that with enough concurrency, we can tolerate long
+   performance challenges. Hardware multithreading systems (e.g., Cray
+   {\small XMT}) show that with enough concurrency, we can tolerate long
    latencies. Unfortunately, this solution is not available with
@@
    high-latency but high-bandwidth global shared memory. This paper
-   lays out the vision for our system, and justifies its feasibility
+   lays out the vision for our system and justifies its feasibility
    with a performance analysis of the runtime for latency tolerance.
@@
 \end{abstract}
+
+\renewcommand{\baselinestretch}{0.8}
 
@@
 gets quickly out of hand. While tools to analyze social networks and
-query semantic graphs with billions of entities exist
+query semantic graphs with billions of vertices and edges exist
 today~\cite{joslyn:hpsf, ediger:msda, brocheler:cosi, pegasus,
@@
 up these applications at a low cost would have a large impact in how
-we deal with large data-sets and make them even more valuable.
+we analyze large data-sets and make them even more valuable.
 
@@
 implement efficient computations for these
-graphs~\cite{bader:bfsmta}. The Cray XMT is an example of such an
+graphs~\cite{bader:bfsmta}. The Cray {\small XMT} is an example of such an
 approach: it solves the memory latency problem through concurrency
-rather than caching. Each XMT processor supports 128 hardware contexts
+rather than caching. Each {\small XMT} processor supports 128 hardware contexts
 and 1024 outstanding memory operations, and is able to switch contexts
 every cycle~\cite{tera, feo-xmt}. This ability comes at a cost: the
-XMT is an expensive, non-commodity machine.
+{\small XMT} is an expensive, non-commodity machine.
 
@@
 We believe we can build a system based mostly on commodity parts that
-can attain XMT-like performance with a familiar, XMT-like,
+can attain {\small XMT}-like performance with a familiar, {\small XMT}-like,
 programming model, but at a fraction of the cost. Therefore, our goal
@@
 
-Our goal is to preserve the XMT programming model: large shared global
+Our goal is to preserve the {\small XMT} programming model: large shared global
 address space, explicit concurrency with a large number of threads,
@@
 address space. We do so via software multithreading: computation that
-is about to execute a long latency operation (e.g. a remote memory
+is about to execute a long latency operation (e.g., a remote memory
 reference) initiates the operation and then yields to the scheduler,
@@
 as {\tt int\_fetch\_add} as well as operations using full-bits, as on
-the Cray XMT.  As is true for other long latency operations,
+the Cray {\small XMT}.  As is true for other long latency operations,
 synchronization latency is tolerated by yielding to the scheduler.
@@
 \begin{description}
-\item[Lightweight multithreading:] We expect to need hundreds of
+\item[Lightweight multithreading] We expect to need hundreds of
   threads to tolerate a cluster's network latency, so our threading
@@
 
-\item[Concurrency in global memory references:] A thread making a
+\item[Concurrency in global memory references] A thread making a
   long-latency global memory reference must be able to yield and allow
@@
 
-\item[Low overhead synchronization:] 
+\item[Low overhead synchronization] 
 %  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
@@
 a synchronization operation. We support only the synchronization
-operations supported by our development platform; we have not
-implemented full-bits yet.
+operations supported by our development platform; we have not yet
+implemented full-bits.
 
@@
 within the multi-node picture and show that it can tolerate large
-latencies. In each part we begin by studying relevant aspects of our
+latencies. In each part, we begin by studying relevant aspects of our
 test machine's memory system so we can interpret our runtime results.
@@
 in a loop; we call $n * \textit{number of cores}$ the number of
-{\em concurrent offered references}, since the memory system may not
+{\em concurrent offered references} since the memory system may not
 be able to satisfy them all in parallel. Since our goal is to find a
@@
 the probability of each reference being a miss. We allocated the lists
-on 1GB huge pages to minimize TLB refill overhead.
+on 1~GB huge pages to minimize TLB refill overhead.
 
@@
 
-We varied the size of the per-coroutine local working set from 2KiB
-to 4MiB, and ran 0 to 32 local updates for each pointer
-traversal. We measured the remote reference rate (i.e. accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).
+We varied the size of the per-coroutine local working set from 2~KB
+to 4~MB, and ran 0 to 32 local updates for each pointer
+traversal. We measured the remote reference rate (i.e., accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).
 
-Figure~\ref{fig:cache-pressure}a shows the result for 8 coroutines per
-core.  With 1 to 4 updates per remote reference for smaller working
+Figure~\ref{fig:cache-pressure}a shows the result for 8~coroutines per
+core.  With 1~to 4~updates per remote reference for smaller working
 sets, the runtime achieves near maximum throughput. 
@@
 Figure~\ref{fig:network-delay}. The runtime can still reach a
-maximum rate of \mrps{173}\footnote{The throughput keeps increasing up to around 564 total concurrent references offered (at which point QPI is saturated), far more than the 36 we know the chip to support.
-This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, saturating the QPI link.
-Little's law predicts that 300 concurrent references are required to achieve the estimated network rate of \mrps{100} when there is $3\mu s$ latency.
+maximum rate of \mrps{173},\footnote{The throughput keeps increasing
+  up to around 564 total concurrent references offered (at which point
+  QPI is saturated), far more than the 36 we know the chip to
+  support. This is because we have modeled the network latency by
+  forcing coroutines to wait the extra time before using the data, so
+  the physical miss buffers on our machine are not tied up any longer
+  than usual.} saturating the QPI link.
+Little's Law predicts that 300 concurrent references are required to achieve the estimated network rate of \mrps{100} when there is $3\mu s$ latency.
 The results agree: \mrps{100} is reached with about 53 coroutines per core.
@@
 We observe that when performing up to 8 local updates per remote reference, smaller working sets still allow the desired throughput.
-When performing 1 to 2 updates, the runtime can maintain throughput for working sets of up to 32KiB.
+When performing 1 to 2 updates, the runtime can maintain throughput for working sets of up to 32~KB.
 
@@
 implementations include the Tera MTA~\cite{tera} and Cray
-XMT~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
+{\small XMT}~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
 MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
@@
 but high-bandwidth).
-
+\newpage
 %\bibliographystyle{acm}

### 2011-04-21 doc/hotpar2011/softxmt-hotpar2011.tex
some more modifications

@@
 every cycle~\cite{tera, feo-xmt}. This ability comes at a cost: the
-{\small XMT} is an expensive, non-commodity machine.
+{\small XMT} is an expensive, non-commodity machine with low
+single-threaded performance.
 
@@
 We believe we can build a system based mostly on commodity parts that
-can attain {\small XMT}-like performance with a familiar, {\small XMT}-like,
-programming model, but at a fraction of the cost. Therefore, our goal
-is to build a system that has good performance on low-locality graph
-codes but is implemented with cheap commodity processors and simple
-additional support in an FPGA. This approach has the added
-benefit of not sacrificing general-purpose performance.
+can attain {\small XMT}-like performance with a familiar, {\small
+  XMT}-like, programming model, but at a fraction of the
+cost. Therefore, our goal is to build a system that has good
+performance on low-locality graph codes but is implemented using cheap
+commodity processors with the possible additional support of an FPGA. This
+approach has the added benefit of not sacrificing general-purpose
+performance.
 
@@
 reference) initiates the operation and then yields to the scheduler,
-which quickly resumes execution of another computation. Concurrency is
-also used to make more efficient use of network bandwidth: it
+which quickly resumes execution of another computation. Concurrency 
+beyond that required for latency tolerance
+can also be used to make more efficient use of network bandwidth: it
 enables aggregation of short memory requests into coarser requests,
-which leads to large network messages and consequently better
+leading to large network messages and consequently better
 bandwidth utilization.
@@
 Synchronization on locations in the private address space of a node
-works the the same way as on conventional systems.  Synchronization on
+works the same way as on conventional systems.  Synchronization on
 global addresses works differently: we provide atomic operations such
@@
 \begin{description}
-\item[Lightweight multithreading] We expect to need hundreds of
+\item[Concurrency in global memory references\quad] We must support many
+  outstanding references to our long-latency global memory to fully
+  utilize its bandwidth.
+%A thread making a
+%  long-latency global memory reference must be able to yield and allow
+%  other threads to make progress.
+
+\item[Lightweight multithreading\quad] We expect to need hundreds of
   threads to tolerate a cluster's network latency, so our threading
@@
 
-\item[Concurrency in global memory references] A thread making a
-  long-latency global memory reference must be able to yield and allow
-  other threads to make progress.
+\item[Low-overhead synchronization\quad] 
+%  We must support global synchronization between threads that allows
+%  unrelated threads to proceed unblocked.
+Long-latency synchronization operations should not block the processor's
+pipeline, so that other threads can proceed.
 
-\item[Low overhead synchronization] 
-%  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
-  Similarly to a regular memory access, long-latency synchronization
-  should also cause a thread to yield and allow others to make
-  progress. %\todo{repetitive with the previous item}
+%Similarly to a regular memory access, long-latency synchronization
+%  should also cause a thread to yield and allow others to make
+%  progress. %\todo{repetitive with the previous item}
 \end{description}
@@
 implementation and ideas for a multi-node implementation.
+
+\subsection{Concurrency in global memory references}
+
+To enable concurrency in memory references, we break each read or
+write into two operations: the {\em issue} operation and the {\em
+  complete} operation. The read issue operation starts data moving
+towards the processor and returns immediately, like a prefetch; the
+read complete operation blocks until data is available. The write
+issue operation takes data along with an address, and starts an
+asynchronous write; the write complete operation blocks until the
+write is committed.
+
+A latency-tolerant read operation in threaded user code then consists of three steps: a {\em read
+  issue}, which causes the data to start moving; a {\em yield}, which
+causes another thread to start executing, overlapping the read
+latency; and, once the reading thread has been rescheduled, a {\em
+  read complete} blocks until the data is available. A write operation
+follows the same pattern, blocking until the data has been committed.
+
+%\paragraph{Single-node implementation.}
+In our single-node implementation, we use prefetch instructions for
+the issue operation, and regular blocking reads and writes for the
+complete operation. 
+Writes use the prefetch instruction to gain
+cache line ownership before modification.
+%  Since we expect
+%these global references to have low locality, we use the
+%``non-temporal'' form of prefetch to avoid disturbing the state of the
+%cache hierarchy \cite{intel:swdev}.
+
+%\paragraph{Multi-node ideas.}
+As we develop our multi-node implementation, we believe we will need hundreds of outstanding memory references per
+processor to cover the latency of remote references in a multi-node
+system. Unfortunately, commodity processors have much smaller limits
+on memory concurrency: section~\ref{subsubsec:evalsinglebase} finds a
+limit of 36 concurrent loads. We will have to manage memory
+concurrency on our own to bypass this limit. We describe approaches in
+both software and hardware.
+
+One approach is for user code to delegate global references to a
+special {\em global memory manager thread}, running on another core in
+the chip. This thread translates the global address into a network
+destination, makes the request through the network card to the remote
+node, checks for completion, and delivers the returned data to the
+requesting thread. On the remote node, the remote memory manager
+thread performs the memory operation and return the data to the
+requesting node.
+
+%This approach has the benefit of enabling our entire system to be
+%build using commodity parts, but its potential for performance is not
+%clear. We have increased the overhead of a single memory operation:
+%what in the single node case was a simple read, is now multiple
+%queuing operations between threads and multiple reads and writes over
+%the processor's link to the network interface.
+
+Another approach moves the global memory management to a hardware
+device. We can add a coprocessor FPGA in a processor socket, listening
+to coherence messages on the bus, similar to \cite{mogill}. Note that
+this accelerator is only a memory proxy, not a compute
+coprocessor. The accelerator maps all global shared memory into a
+segment of each node's local address space; when it detects a
+reference to memory on a remote node, it does the address-to-node
+translation and issues the request through the network controller,
+delivering the data to the requesting thread when ready. We 
+encode commands in the upper bits of the address; a read from a
+location's {\em issue} address starts the remote request and returns
+immediately, and a read from a location's {\em complete} address 
+blocks until the data is available.
+
+%To support the issue and complete operations for concurrency support,
+%commands would be encoded in the upper bits of the address: then each
+%remote word would have multiple addresses on the local node. Reading
+%from the {\em issue} address would immediately return a value
+%indicating that the accelerator is issuing the request over the
+%network. Reading from the {\em complete} address would block until the
+%data was available.
+
+With either approach, the question of request aggregation will be
+important. Each memory request we make is small---perhaps 8
+bytes---but most networks are optimized for bulk transfers of a few
+thousand bytes. To improve network utilization, we will explore the
+aggregation of multiple memory operations to different addresses on
+the same remote node.
 
@@
 In our single-node implementation, we implemented coroutines
-as described, with a round-robin scheduler. Coroutines do local work,
-issue a remote prefetch, and switch to the next in turn.  When the
-coroutine is reactivated, it assumes the requested data is available
-and issues a read.  If the data is not yet received and in cache, the core
-(and all its coroutines) can stall.
+as described, with a round-robin scheduler. 
+%Coroutines do local work,
+%issue a remote prefetch, and switch to the next in turn.  When the
+%coroutine is reactivated, it assumes the requested data is available
+%and issues a read.  If the data is not yet received and in cache, the core
+%(and all its coroutines) can stall.
 
@@
 %operations have completed.
-As we develop our multi-node implementation, we will improve our
-scheduler to activate threads after their long-latency operations
+As we develop our multi-node implementation, we will modify our
+scheduler to reactivate threads only after their long-latency operations
 have completed.
@@
 
-\subsection{Concurrency in global memory references}
-
-To enable concurrency in memory references, we break each read or
-write into two operations: the {\em issue} operation and the {\em
-  complete} operation. The read issue operation starts data moving
-towards the processor and returns immediately, like a prefetch; the
-read complete operation blocks until data is available. The write
-issue operation takes data along with an address, and starts an
-asynchronous write; the write complete operation blocks until the
-write is committed.
-
-A read operation in user code consists of three steps: a {\em read
-  issue}, which causes the data to start moving; a {\em yield}, which
-causes another thread to start executing, overlapping the read
-latency; and, once the reading thread has been rescheduled, a {\em
-  read complete} blocks until the data is available. A write operation
-follows the same pattern.
-
-%\paragraph{Single-node implementation.}
-In our single-node implementation, we use prefetch instructions for the issue operation, and regular
-blocking reads and writes for the complete operation.
-%  Since we expect
-%these global references to have low locality, we use the
-%``non-temporal'' form of prefetch to avoid disturbing the state of the
-%cache hierarchy \cite{intel:swdev}.
-
-%\paragraph{Multi-node ideas.}
-As we develop our multi-node implementation, we believe we will need hundreds of outstanding memory references per
-processor to cover the latency of remote references in a multi-node
-system. Unfortunately, commodity processors have much smaller limits
-on memory concurrency: section~\ref{subsubsec:evalsinglebase} finds a
-limit of 36 concurrent loads. We will have to manage memory
-concurrency on our own to avoid this limit. We describe approaches in
-both software and hardware.
-
-One approach is for user code to delegate global references to a
-special {\em global memory manager thread}, running on another core in
-the chip. This thread translates the global address into a network
-destination, makes the request through the network card to the remote
-node, checks for completion, and delivers the returned data to the
-requesting thread. On the remote node, the remote memory manager
-thread would perform the memory operation and return the data to the
-requesting node.
-
-%This approach has the benefit of enabling our entire system to be
-%build using commodity parts, but its potential for performance is not
-%clear. We have increased the overhead of a single memory operation:
-%what in the single node case was a simple read, is now multiple
-%queuing operations between threads and multiple reads and writes over
-%the processor's link to the network interface.
-
-Another approach moves the global memory management to a hardware
-device. We can add a coprocessor FPGA in a processor socket, listening
-to coherence messages on the bus, similar to \cite{mogill}. Note that
-this accelerator is only a memory proxy, not a compute
-coprocessor. The accelerator would map all global shared memory into a
-segment of each node's local address space; when it detects a
-reference to memory on a remote node, it would do the address-to-node
-translation and issue the request through the network controller,
-delivering the data to the requesting thread when ready. We would
-encode commands in the upper bits of the address; a read from a
-location's {\em issue} address start the remote request and return
-immediately, and a read from a location's {\em complete} address would
-block until the data was available.
-
-%To support the issue and complete operations for concurrency support,
-%commands would be encoded in the upper bits of the address: then each
-%remote word would have multiple addresses on the local node. Reading
-%from the {\em issue} address would immediately return a value
-%indicating that the accelerator is issuing the request over the
-%network. Reading from the {\em complete} address would block until the
-%data was available.
-
-With either approach, the question of request aggregation will be
-important. Each memory request we make is small---perhaps 8
-bytes---but most networks are optimized for bulk transfers of a few
-thousand bytes. To improve network utilization, we will explore the
-aggregation of multiple memory operations to different addresses on
-the same remote node.
-
 \subsection{Low-overhead synchronization}
@@
 synchronization operations by splitting them into {\em issue} and {\em
-  complete} phases.
+  complete} operations.
 
@@
 
-We focus the evaluation on one component of the runtime: lightweight context switching.
+We focused the evaluation on one component of the runtime: lightweight context switching.
 We ran pointer-chasing benchmarks on a single-node implementation of our runtime.
@@
 the probability of each reference being a miss. We allocated the lists
-on 1~GB huge pages to minimize TLB refill overhead.
+on 1~GB huge pages to minimize overhead due to TLB refills.
 
@@
 the multi-node runtime, \texttt{readIssue} will not be implemented
-with a prefetch instruction. The single-node coroutine library does
-depend on prefetches: further experiments show that without them, the overhead of context switches limits concurrency when there are more than two coroutines per core.
-Lastly, we observe repeatable data points that are off the trend and fall below the max.
-This effect is harder to explain, but we suspect it may be
+with a prefetch instruction. 
+%The single-node coroutine library does
+%depend on prefetches: further experiments show that without them, the
+%overhead of context switches limits concurrency when there are more
+%than two coroutines per core. 
+Lastly, we observe repeatable data points that are off the trend and fall below the max. This effect is harder to explain, but we suspect it may be
 due to resource contention in the cores' pipelines once the memory
@@
 
-We varied the size of the per-coroutine local working set from 2~KB
+We varied the size of the per-coroutine local working set from 64~B
 to 4~MB, and ran 0 to 32 local updates for each pointer
@@
 To simulate the performance of pointer chasing in a multi-node system
-we referenced the second chip's memory and also had our coroutine scheduler include a delay
+we referenced the first chip's memory from the second chip's cores and also had our coroutine scheduler include a delay
 before a coroutine can be rescheduled, imitating the network transit
@@
 computation using commodity parts. We leverage decades-old ideas on
-using a large amount of concurrency to tolerate latencies; but we do so
-mostly in software. We developed a runtime system for latency
-tolerance and our results showed that it is able to manage enough
-concurrency to tolerate long latencies to hide the cost of
-expensive accesses to our large global memory system (high-latency,
-but high-bandwidth).
-\newpage
+using a large amount of concurrency to tolerate latencies, but we do
+so mostly in software. We developed a runtime system for latency
+tolerance; our results showed that it supports enough concurrency
+to tolerate the long latencies of our large high-bandwidth global
+memory system.
+
 %\bibliographystyle{acm}

### 2011-04-21 doc/hotpar2011/softxmt-hotpar2011.tex
bib updates

@@
   A. H. Hunter$^{\dagger}$,
-  Preston Briggs$^{\dagger}$
+  Preston Briggs$^{\dagger}$,
   Luis Ceze$^{\dagger}$,
@@
   Simon Kahan$^{{\dagger \ddagger}}$,
-  Mark Oskin$^{\dagger}$,
+  Mark Oskin$^{\dagger}$
 }{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
@@
 have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
-up these applications at a low cost would have a large impact in how
+up these applications at a low cost would have a significant impact in how
 we analyze large data-sets and make them even more valuable.
@@
 {\small XMT} is an expensive, non-commodity machine with low
-single-threaded performance.
+single-thread performance.
 
@@
 %than two coroutines per core. 
-Lastly, we observe repeatable data points that are off the trend and fall below the max. This effect is harder to explain, but we suspect it may be
+We also observe repeatable data points that are off the trend and fall below the max. This effect is harder to explain, but we suspect it may be
 due to resource contention in the cores' pipelines once the memory
@@
   \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/cachepressurenewview-edited.pdf}
+%    \includegraphics[width=0.48\textwidth]{figures/cachepressurenewview-edited.pdf}
+    \includegraphics[width=0.48\textwidth]{figures/cachepressure-64to4M-edited2.pdf}
   \end{center}
-  \vspace{-10pt} %space with fig
+%  \vspace{-10pt} %space with fig
   \caption{Cache pressure with coroutines, on six cores. The vertical
@@
 We varied the size of the per-coroutine local working set from 64~B
-to 4~MB, and ran 0 to 32 local updates for each pointer
+to 4~MB, and ran 0 to 32 local updates for each remote pointer
 traversal. We measured the remote reference rate (i.e., accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).
@@
 As the number of updates and the working set size increase, the data
-of other coroutines is likely to be evicted from cache and performance
-suffers.
+of other coroutines is likely to be evicted from cache, leading to
+decreased performance.
 
@@
 we referenced the first chip's memory from the second chip's cores and also had our coroutine scheduler include a delay
-before a coroutine can be rescheduled, imitating the network transit
+before a coroutine can be reactivated, imitating the network transit
 delay. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
@@
 
-Figure~\ref{fig:cache-pressure}b shows the results of the cache pressure experiment for supporting 48 coroutines per core, about the number of contexts needed to tolerate latency in the network experiment.
-We observe that when performing up to 8 local updates per remote reference, smaller working sets still allow the desired throughput.
-When performing 1 to 2 updates, the runtime can maintain throughput for working sets of up to 32~KB.
+Figure~\ref{fig:cache-pressure}b shows the results of the cache
+pressure experiment for supporting 48 coroutines per core, about the
+number of contexts needed to tolerate latency in the network
+experiment. We observe that when performing up to 16~local updates per
+remote reference, smaller working sets still allow the desired
+throughput. When performing 1~to 2~updates, the runtime can maintain
+throughput for working sets of up to 32~KB.
 
@@
 Much work exists on using multithreading to tolerate latency. Hardware
-implementations include the Tera MTA~\cite{tera} and Cray
+implementations include the Tera {\small MTA}~\cite{tera} and Cray
 {\small XMT}~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
@@
 so mostly in software. We developed a runtime system for latency
-tolerance; our results showed that it supports enough concurrency
+tolerance, and our results showed that it supports enough concurrency
 to tolerate the long latencies of our large high-bandwidth global

### 2011-04-21 doc/hotpar2011/softxmt-hotpar2011.tex
enlarge "network rate" in cache plot; resize, formant
a few bits of cleanup

@@
   pregel}, graphs of interest to defense applications are expected to
-have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
+have trillions of vertices and edges~\cite{hartley:mssg, kolda:imkd}. Speeding
 up these applications at a low cost would have a significant impact in how
@@
 reference rate, we extended the previous experiment so that cores in
-both processors traversed lists allocated in the first processor's
+both chips traversed lists allocated in the first chip's
 memory.
@@
 coroutine. As with our first experiment, we allocated the lists in the
-same processor as the cores doing the traversal.
+same chip as the cores doing the traversal.
 
@@
 Figure~\ref{fig:cache-pressure}a shows the result for 8~coroutines per
-core.  With 1~to 4~updates per remote reference for smaller working
-sets, the runtime achieves near maximum throughput. 
+core.  With 1, 2, and 4~updates per remote reference for smaller working
+sets, the runtime is able to achieve near maximum throughput. 
 %As working set
@@
 To simulate the performance of pointer chasing in a multi-node system
-we referenced the first chip's memory from the second chip's cores and also had our coroutine scheduler include a delay
+we referenced the first chip's memory from the second chip's cores and
+modified our coroutine scheduler to include a delay
 before a coroutine can be reactivated, imitating the network transit

### 2011-04-21 doc/hotpar2011/softxmt-hotpar2011.tex
memory concurrency small edit

--- 2011-04-21 doc/hotpar2011/softxmt-hotpar2011.tex
@@
   complete} operation. The read issue operation starts data moving
-towards the processor and returns immediately, like a prefetch; the
+towards the processor and returns immediately, like a prefetch. The
 read complete operation blocks until data is available. The write
 issue operation takes data along with an address, and starts an
-asynchronous write; the write complete operation blocks until the
+asynchronous write. The write complete operation blocks until the
 write is committed.
@@
   \begin{center}
+  \vspace{-10pt}
 %    \includegraphics[width=0.48\textwidth]{figures/cachepressurenewview-edited.pdf}
-    \includegraphics[width=0.48\textwidth]{figures/cachepressure-64to4M-edited2.pdf}
+    \includegraphics[width=0.50\textwidth]{figures/cachepressurenewview-edited.pdf}
   \end{center}
+  \vspace{2pt}
 %  \vspace{-10pt} %space with fig

### 2012-07-10 doc/asplos13/grappa.tex
Initial ASPLOS 2013 draft

\section{Grappa} \label{sec:grappa}

In this section we describe the Grappa runtime system (Figure~\ref{fig:grappa}).  Grappa is a software runtime system that one adds to a commodity HPC cluster.  It provides the following software abstractions to applications: \textbf{memory}, a global distributed shared memory and \textbf{tasking}, co-operative multithreading with automatic load balancing.  Under the hood a third component, \textbf{communication}, a network packet scheduler/aggregator is key to achieving high performance, but largely invisible to applications.  We will now describe each of these components.

\subsection{Memory}

\subsection{Tasking}

\subsection{Communication}

### 2012-07-10 doc/asplos13/intro.tex
Initial ASPLOS 2013 draft

\section{Introduction} \label{sec:intro}

Computing on large graph datasets is both important and challenging.  Underlying many applications critical for national security, market analysis, and social computing are algorithms that process large graphs.  Three basic factors make these applications challenging for traditional computing systems: (1) the datasets are enormous -- far more than can fit within a single computing system's memory; (2) the computation to communication ratio is low -- applications spend most of their time fetching nodes and edges of the graph, and the computation performed on that data often takes a trivial amount of time by comparison; and (3) graph datasets have little to no locality, defying traditional mechanisms architects have designed, such as caches and prefetching, for improving memory system performance.

Traditionally, when organizations had to process large graphs they would turn to the Cray XMT~\cite{xmt} line of machines.  The XMT is well suited to the processing of large graphs: it has a large distributed shared memory, and instead of relying on caches to exploit locality, it relies on fine-grained multithreading to tolerate access latency.  Locality in the data access stream is not required, if sufficient threads exist to overlap computation with communication.  The XMT machines have two basic detractions, however: (1) they are not inexpensive because customers cannot exploit the economics of commodity hardware and software; and (2) applications designed to execute on commodity clusters cannot be automatically ported to the XMT with good performance.  Users of the XMT need to commit to its style and language of software development.

In this paper we describe Grappa, a software runtime system built for your typical high-performance compute cluster, that makes graph algorithms fast and scalable.  Grappa is inspired by the XMT hardware, but implements the key mechanisms of that design in software, and adds additional software components to tolerate the differences between a typical compute cluster and the more specialized XMT hardware.  A useful analogy is that Grappa is to graph processing, as Hadoop~\cite{hadoop} is to map-reduce.

By design, Grappa is built on as much freely available and commodity infrastructure as possible.  We use unmodified Linux for the operating system.  An off the shelf user-mode infiniband device driver stack~\cite{Melonox?} is used for network interfacing.  MPI~\cite{mpi} is used for process setup and tere down.  Finally, GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes and active message invocation.

To this commodity hardware and software mix Grappa adds three main software components: (1) a lightweight tasking layer that supports a context switch in as few as \checkme{40ns} and distributed load balancing; (2) a distributed shared memory layer that supports customary operations such as read and write as well as remote operations such as fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that mitigates the problem where commodity networks are designed to achieve peek bandwidth only on large packet sizes, yet graph algorithms tend to want to fetch only a handful of bytes at a time.  Each of these components will be described in more detail in Section~\ref{sec:grappa}.

Grappa achieves its design goal, which is to make graph algorithms execute quickly and scalably on commodity clusters.  Our yardstick for comparison is the XMT hardware itself.  Using the same number of processors (in our case cores on a single multicore processor), Grappa is in the same ballpark as the XMT: on some graphs, such as the \checkme{WTF} Grappa is \checkme{2X} faster, while on others, such as the \checkme{WTF2} Grappa is \checkme{2X} slower.  In Section~\ref{sec:results} we explore the factors that underpin this performance.  But most importantly, application performance \emph{scales} with Grappa.  For significantly less real world cost, users can \emph{add} more processors to a commodity cluster and easily outperform the XMT hardware on graph processing, while still saving significant dollars.

\TODO{Add roadmap here}

### 2012-07-10 doc/asplos13/paper.tex
Initial ASPLOS 2013 draft

\input{header}

\begin{document}

\title{All Your Graphs Belong to Us}
\author{TBD}

\maketitle
\begin{abstract}
\input{abstract}
\end{abstract}

\input{intro}

\input{whynot}

\input{grappa}

\subsection{References}
\bibliographystyle{abbrv}
\bibliography{paper}

\end{document}

### 2012-07-10 doc/asplos13/whynot.tex
Initial ASPLOS 2013 draft

\section{Motivation} \label{sec:motivation}

Grappa is built on many existing widely available hardware and software packages and designed for the typical high performance computing (HPC) cluster.  Hardware nodes consist of rack mount servers that each have a multicore processor and multiple gigabytes of memory.  These servers communicate with each other through both switched 10G ethernet and switched Infiniband.  Individual nodes boot Linux, and except for the high performance network that ties them together are otherwise indistinguishable from an individual server operating in isolation.  Freely available MPI~\cite{mpi} and GASnet~\cite{gasnet} software packages are installed on the system, as well as usermode device driver support for the Infiniband network.

\noindent\textbf{Why do graph applications perform poorly on clusters?}

### 2012-07-11 doc/asplos13/grappa.tex
updated memory

@@
 
+Applications written for Grappa utilize two forms of memory: local and global.
+
+Local memory is exactly what its name implies.
+Accesses occur through conventional pointers.
+The compiler emits an access and the memory is manipulated directly.
+Obviously, local pointers cannot access memory on other system nodes.
+Applications use local accesses for a number of things in Grappa: the stack associated with a task, accesses to localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
+
+Global memory is a distributed block cyclic heap.
+Applications allocate memory in this heap and a global pointer is returned.
+Global pointers are utilized in one of two ways: (1) via a \emph{cache} or (2) via a \emph{delegate}.
+
+\paragraph{Cache:} With a cache, applications can instruct Grappa to fetch a global pointer (of any length) and return a local pointer to a cached copy of the global memory.
+Under the hood, Grappa performs the mechanics of gathering chunks of data from multiple system nodes and presenting a conventional appearing linear block of memory as a pointer into a cache.
+Caches are either read-only or read-write and alternatively incoherent or coherent.
+Read-only and and read-write caches are what their names imply.
+Under the hood, the distinction between these two forms of caches lays in how they interact with the directories that maintain the coherence state for coherent access to the memory the global pointer points to.
+For coherent caches, these directories maintain a simple coherence protocol that permits multiple outstanding read-only cached copies of data to exist, or alternatively, only a single outstanding read-write cached copy to exist.
+Unlike hardware caches, these software caches require explicit release operations issued in software; i.e., applications must explicitly inform Grappa when the local pointer that points into a local cached copy of global memory is no longer needed.
+Grappa then uses this release operation to interact with the directories that control the global memory in question.
+For incoherent accesses, the directories are bypassed.
+Applications acquire a local pointer to a cached copy of the global data and can expressly push the data from this local copy back to global memory in order to post any writes.
+Obviously, applications should only use incoherent accesses when they can \emph{prove} accesses across multiple threads and system nodes are disjoint, or solely read-only, or (in rarer cases) maintaining a consistent view of memory is irrelevant to application correctness, or for debugging purposes.
+
+\paragraph{Delegates:} Alternatively applications can dispatch computation to be performed on individual machine-word sized chunks of global memory to the memory system itself (e.g., \emph{fetch-and-add}).
+Applications provide a pointer to a function and a global pointer to Grappa, which takes care to execute that function on the system node that actually stores the memory pointed to by the global pointer.
+In such cases, localization of the pointer is more efficient.
+Delegate operations are most often used for synchronization primitives, but really any computation is permissible.
+
+\TODO{Check: Do delegate operations interact with the directories?  Are they coherent?}
+
+Figure~\ref{fig:globalmemory} depicts the global memory architecture and access interface.
+
 \subsection{Tasking}

### 2012-07-11 doc/asplos13/grappa.tex
communication v0

@@
 \subsection{Communication}
+
+With Grappa, from the perspective of the application, all communication between tasks occurs through global shared memory.
+These requests are ultimately translated into explicit inter system node communication requests that are dispatched to GASnet.
+Under the hood there exists a key component within Grappa, the message aggregator, that makes fine-grained communication efficient on commodity Infiniband networks.
+Infiniband achieves its peek bisection bandwidth \emph{only} when the packet sizes are relatively large (a kilobyte or more).
+In our experience, graph algorithms request objects through the caching system (previously described) with a median object size of \checkme{16} bytes.
+Our measurements, confirm manufactures published data~\cite{infinibandbandwidth}, that with this packet size the bisection bandwidth is only a small fraction, less than ~\checkme{5\%} of the peak bisection bandwidth.
+The reason for this discrepancy is the combination of overheads associated with handling each packet (in terms of bytes that form the actual Infiniband packet, processing time at the card and processing on the CPU within the driver stack).
+
+To work around this problem we perform packet aggregation.
+Instead of dispatching a packet immediately to the Infiniband hardware, Grappa holds onto an individual communication request.
+Pending requests that are to be sent to the same remote node are queued.
+Once a sufficient number of them (in terms of bytes) or a timeout occurs, they are dispatched as a single packet to GASnet and onward to the Infiniband network.
+In Section~\ref{sec:results} we explore the sensitivity of performance to these two key parameters, but the net result is that performance is maximized by aiming to aggregate for \checkme{1024} bytes and \checkme{1ms}.

### 2012-07-13 doc/asplos13/grappa.tex
Trival change to test repository move

@@
 \section{Grappa} \label{sec:grappa}
-
+[trivial change]
 In this section we describe the Grappa runtime system (Figure~\ref{fig:grappa}).
@@
 
+Grappa provides its own task library to applications.
+Applications begin execution with a single task, bound to an existing thread stack.
+Tasks can spawn new tasks, and those new tasks are initially not bound to a thread stack, but instead are queued for future execution.
+
+Grappa context switches between tasks non-preemptively.
+Most calls into the Grappa library, however, do initiate a context switch, and tasks can explicitly yield control.
+Grappa switches between tasks entirely within usermode, and in our experiments we have measured the context switch time to be only \checkme{40ns}.
+This time is kept small because it is done entirely in usermode and only registers specified as caller saved by the compiler need be saved/restored.
+Tasks become bound to a thread stack when no other task is runnable, or Grappa determines it is profitable to do so.
+\TODO{How does Grappa determine this?}.
+
+Tasks that are not bound to a stack are steelable by the work steeling system within Grappa.
+This load balances work across the collection of nodes executing an application built on Grappa.
+\TODO{Talk about the work steeling algorithm that we use}.
+
 \subsection{Communication}

### 2012-07-13 doc/asplos13/method.tex
initial methodology section

\section{Metholody} \label{sec:method}

To explore the performance of the Grappa runtime we have implemented three algorithms: breadth first search, betweenness centrality, and unbalanced tree search.  These algorithms were implemented for the Cray XMT (our baseline) and for Grappa.  Performance results for Grappa were obtained on a 144 node cluster.  Individual nodes of this cluster contain the hardware depicted in Table~\ref{table:grappanode}.  Nodes are interconnected with both 10G Ethernet and Infiniband.  For all but startup and configuration, they are configured to communicate over the Infiniband network for these experiments.

The metric we use is algorithmic time, which means startup and loading of the graph (from disk) is not included in the measurement.  Data is collected on real systems, which means minor variations in runtime exist from run to run.  The average of multiple runs are used, and where appropriate, confidence in the result is reported (error bars, standard deviation, etc).

One question that must be answered when making a comparison between the Cray XMT and a typical HPC cluster is what is a fair comparison -- these systems are quite different.  Three options immediately come to mind: equal number of processing cores, equal number of network interfaces, and equal dollars.  We discounted the last option fairly quickly, because it isn't a lasting data-point and, as will be evident later, would be hugely unfair to the XMT.  The first option, cores, has some merit, but Grappa is designed for applications that have no locality in their computation.  This means almost all of their memory accesses are remote.  Hence, the factor that limits their performance is not processing, but communicating.  Hence, we have chosen the middle option, network interfaces as the way to normalize across the XMT and our cluster.  Each processor in the XMT system has its own network interface to access shared memory.  In the HPC cluster we use, each processor (which contains 12 cores), has a single Infiniband interface.  Hence, for our results we scale up XMT processors one for one with full system nodes.

### 2012-07-14 doc/asplos13/implementation.tex
Added initial implementation discussion

\section{Implementation} \label{sec:implementation}

We now discuss implementation details for the major system components discussed in Section~\ref{sec:grappa}.

\subsection{Tasking}

\subsubsection{Tasks and workers}

Grappa tasks are 32-byte entities: a 64-bit function pointer plus
three 64-bit arguments. We use three arguments because tasks are
usually generated as part of a parallel loop decomposition, and thus
each task needs three kinds of data:
\begin{description}
\item[Function pointer] This 64-bit value indicates what routine
  should run. We depend on the fact that all processes in the system
  load the same binary so that function pointers are consistent across
  processes.
\item[Private argument] This 64-bit value is used for task-specific
  data, and is often a loop index.
\item[Shared argument] This 64-bit value is used for data shared
  between all tasks that are part of a loop, and is usually a global
  pointer to a struct. Each task uses its private argument to 
\item[Synchronization argument] This 64-bit value is used to determine
  when all tasks that are part of a loop have finished. It is usually
  a global pointer to a synchronization object allocated at the core
  that spawned the task.
\end{description}
While these are the most common uses of the three task arguments, they
are treated as arbitray 64-bit values in the runtime, and can be used
to store whatever the user wants.

Tasks are not allocated any execution resources until the scheduler
decides to run them; when this occurs, tasks are matched with {\em
  workers}. Each worker simply a collection of status bits and a
stack, allocated at a particular core. Workers

\subsubsection{Context switching}

Context switching in Grappa is non-preemptive. Any Grappa routine that
does a potentially-long-latency operation calls a context-switch
routine that gives control of the core back to the scheduler. Users
may also explicity yield control to the scheduler.

As with other cooperative multithreading systems, we treat context
switches as function calls, saving and restoring only the callee-saved
state as specified in the x86-64 ABI \TODO{cite}. This involves saving
six general-purpose 64-bit registers and the stack pointer, as well as
the 16-bit x87 floating point control word and the SSE context/status
register. Thus, the minimum amount of state a cooperative context
switch routine must save according to the ABI is 62 bytes.

Since the compiler sees all calls to the context switch routine, we
can save even less state. Our context switch routine appears to the
compiler as inline assembly; we can declare all the registers we need
to save as ``clobbered'' by the inline assembly routine, and the
compiler will issue its own save and restore code. This allows the
compiler to avoid saving any registers that are not used, or are used
for temporary values that are not needed after the context switch. In
particular, only one of our applications does floating point
operations; all the other applications can safely ignore the x87 and
SSE2 registers as part of context switching.

\subsubsection{Scheduling}

Grappa's scheduler is simple. When

\subsubsection{Work stealing}

\subsection{Communication}

\subsection{Aggregation}

\subsection{Synchronization}

\subsection{Consistency}

### 2012-07-15 doc/asplos13/implementation.tex
added scheduling description

@@
 
-Grappa's scheduler is simple. When
+Each core in the Grappa system has its own independent scheduler. Each
+scheduler has three main tasks to perform.  First, it must ensure that
+its communication resources are serviced reasonably often. Second, it
+must ensure that if a running task was waiting on a long-latency
+operation and that operation has completed, the task will be
+rescheduled. Third, if there are tasks waiting to be run and spare
+execution resources, the tasks must be matched with idle workers.
+
+Each scheduler has three queues:
+\begin{description}
+\item{Ready worker queue} This is a FIFO queue of tasks that are
+  matched with workers and are ready to execute.
+\item{Private task queue} This is a FIFO queue of tasks that must run on this core.
+\item{Public task queue} This is a LIFO queue of tasks that are
+  waiting to be matched with workers. Tasks may be stolen from this
+  queue by other cores in the system; this is described in the next section.
+\end{description}
+
+Whenever a task yields or suspends, the scheduler makes a decision
+about what to do next. First, it determines if it is time to service
+the communication resources should be serviced. This is done on a
+periodic basis. Second, it determines if any workers with running
+tasks are ready to execute; if so, one is scheduled. Finally, if there
+are no workers ready to run, but there are tasks waiting to be matched
+with workers, an idle worker is woken (or a new worker is spawned),
+matched with a task, and scheduled.
+
+The scheduler uses its core's timestamp counter to track time. The
+timestamp counter is read once per scheduling decision. Timestamp
+values are not shared between cores. We use timestamp values only to
+ensure period execution of communication maintenance tasks; being
+wrong will only change the rate of these events with a potential
+performance cost, and no correctness violation. Nevertheless, we
+assume Grappa will run on machines with modern timestamp counter
+implementations that are monotonic and frequency-compensated.
 
 \subsubsection{Work stealing}
+
+\TODO{BrandonM or Jacob}
 

### 2012-07-15 doc/asplos13/evaluation.tex
add evaluation section, start to UTS subsection/plot

\section{Evaluation} \label{sec:evaluation}
%todo put method in here, if it stays small

\subsection{Memory-based Unbalanced tree search}
%Traversal of an unbalanced, unpartitioned tree captures the foundations for implementing most forms of irregular parallelism. 
...
%Recursion is found in every irregular divide and conquer algorithm. Loops, even regular ones, must be dynamically scheduled to get good load balance when run on a large, non-uniform system. Nested parallelism involves fine-grain creation of dynamic amounts of work. Dataflow..

Unbalanced tree search (UTS), is a benchmark for evaluating the programmability and performance of systems for parallel applications that require dynamic load balancing \cite{Olivier:uts2006}. It involves traversing an unbalanced implicit tree: at each vertex, its number of children is sampled from some probability distribution, and this number of new nodes are added to a work queue to be visited. While this benchmark captures irregular, dynamic parallelism, we want to evaluate performance on algorithms that touch existing data. We augment UTS by using the traversal to create a large tree in memory, and then we traverse the in-memory tree. We call this benchmark UTS-mem, and the timed portion is this traversal of the in-memory tree. The assumption is that in general the size of each subtree is unknown until that subtree has actually been traversed.

%todo: say tree explore is same as a search except we are additionally needing to synchronize on all vertices visited

The Grappa version of tree search UTS-mem

\lstset{language=C++,
       basicstyle=\footnotesize,
       tabsize=2}
\begin{lstlisting}
// base of shared vertex_t[]
GlobalAddress<vertex_t> Vertices;

// base of shared int64_t[]
GlobalAddress<int64_t> ChildrenPointers;

void search_vertex( id ) {
  GlobalAddress<vertex_t> v_addr = Vertices + id;
  Incoherent<vertex_t>::RO v( v_add, 1 );

  // start index of my children pointers
  childIndex = v->childIndex;

  // how many children I have
  numChildren = v->numChildren;

  // parallel loop over the child list for this vertex
  parallel_for(fn=&search_children, start=childIndex, iters=numChildren);
}

void search_children( start, iters ) {
  // take advantage of spatial locality in the array of children
  GlobalAddress<int64_t> child_base_addr = ChildrenPointers + start;
  Incoherent<int64_t>::RO childIds( child_base_addr, iters );

  // spawn a task to visit each child
  for (i = 0..iters) {
    SoftXMT_publicTask( fn=&search_vertex, id=childIds[i] );
  }
}
\end{lstlisting}

We ran UTS-mem on Grappa and the XMT with two different 100M-vertex trees with a geometric distribution numChildren distribution. Figure~\ref{fig:grappa-xmt-uts} shows the performance in terms of number of vertices visited per second versus number of compute nodes. The Grappa results are for the best parameter values from a limited search over \flushtimeout~and \asyncforthr~\TODO{want to say this in general at beginning of eval?}. Grappa achieves 3x better performance for 12 nodes. Scaling up, Grappa adds 2Mvert/s/node versus XMT's 850Kvert/s/node.
\TODO{why is XMT getting different results for T1L,T2L; would expect it to be the machine with the most uniform results.}

\TODO{sensitivity study for \asyncforthr parameter, ie caching amount. Although we can take advantage of locality in children array and the XMT cannot, we still win without it}

\begin{figure}[h]
    \begin{center}
        \includegraphics[width=0.5\textwidth]{figs/grappa-xmt-uts.pdf}
    \end{center}
    \caption{describe...}
    \label{fig:grappa-xmt-uts}
\end{figure}

### 2012-07-16 doc/asplos13/evaluation.tex
moving UTS discussion about

@@
 
-\subsection{Memory-based Unbalanced tree search}
-%Traversal of an unbalanced, unpartitioned tree captures the foundations for implementing most forms of irregular parallelism. 
-...
-%Recursion is found in every irregular divide and conquer algorithm.
-Loops, even regular ones, must be dynamically scheduled to get good load balance when run on a large, non-uniform system.
-Nested parallelism involves fine-grain creation of dynamic amounts of work.
-Dataflow..
-
-Unbalanced tree search (UTS), is a benchmark for evaluating the programmability and performance of systems for parallel applications that require dynamic load balancing \cite{Olivier:uts2006}.
-It involves traversing an unbalanced implicit tree: at each vertex, its number of children is sampled from some probability distribution, and this number of new nodes are added to a work queue to be visited.
-While this benchmark captures irregular, dynamic parallelism, we want to evaluate performance on algorithms that touch existing data.
-We augment UTS by using the traversal to create a large tree in memory, and then we traverse the in-memory tree.
-We call this benchmark UTS-mem, and the timed portion is this traversal of the in-memory tree.
-The assumption is that in general the size of each subtree is unknown until that subtree has actually been traversed.
-
-%todo: say tree explore is same as a search except we are additionally needing to synchronize on all vertices visited
-
-The Grappa version of tree search UTS-mem
-
-\lstset{language=C++,
-       basicstyle=\footnotesize,
-       tabsize=2}
-\begin{lstlisting}
-// base of shared vertex_t[]
-GlobalAddress<vertex_t> Vertices;
-
-// base of shared int64_t[]
-GlobalAddress<int64_t> ChildrenPointers;
-
-void search_vertex( id ) {
-  GlobalAddress<vertex_t> v_addr = Vertices + id;
-  Incoherent<vertex_t>::RO v( v_add, 1 );
-
-  // start index of my children pointers
-  childIndex = v->childIndex;
-
-  // how many children I have
-  numChildren = v->numChildren;
-
-  // parallel loop over the child list for this vertex
-  parallel_for(fn=&search_children, start=childIndex, iters=numChildren);
-}
-
-void search_children( start, iters ) {
-  // take advantage of spatial locality in the array of children
-  GlobalAddress<int64_t> child_base_addr = ChildrenPointers + start;
-  Incoherent<int64_t>::RO childIds( child_base_addr, iters );
-
-  // spawn a task to visit each child
-  for (i = 0..iters) {
-    SoftXMT_publicTask( fn=&search_vertex, id=childIds[i] );
-  }
-}
-\end{lstlisting}
+\subsection{Unbalanced tree search}
 

### 2012-07-16 doc/asplos13/method.tex
moving UTS discussion about

@@
 Hence, for our results we scale up XMT processors one for one with full system nodes.
+
+\subsection{Applications}
+%Traversal of an unbalanced, unpartitioned tree captures the foundations for implementing most forms of irregular parallelism. 
+%Recursion is found in every irregular divide and conquer algorithm.
+Loops, even regular ones, must be dynamically scheduled to get good load balance when run on a large, non-uniform system.
+Nested parallelism involves fine-grain creation of dynamic amounts of work.
+Dataflow..
+
+\paragraph{Unbalanced tree search (UTS)}, is a benchmark for evaluating the programmability and performance of systems for parallel applications that require dynamic load balancing \cite{Olivier:uts2006}.
+It involves traversing an unbalanced implicit tree: at each vertex, its number of children is sampled from some probability distribution, and this number of new nodes are added to a work queue to be visited.
+While this benchmark captures irregular, dynamic parallelism, we want to evaluate performance on algorithms that touch existing data.
+We augment UTS by using the traversal to create a large tree in memory, and then we traverse the in-memory tree.
+We call this benchmark UTS-mem, and the timed portion is this traversal of the in-memory tree.
+The assumption is that in general the size of each subtree is unknown until that subtree has actually been traversed.
+
+%todo: say tree explore is same as a search except we are additionally needing to synchronize on all vertices visited
+
+\comment{The Grappa version of tree search UTS-mem
+
+\lstset{language=C++,
+       basicstyle=\footnotesize,
+       tabsize=2}
+\begin{lstlisting}
+// base of shared vertex_t[]
+GlobalAddress<vertex_t> Vertices;
+
+// base of shared int64_t[]
+GlobalAddress<int64_t> ChildrenPointers;
+
+void search_vertex( id ) {
+  GlobalAddress<vertex_t> v_addr = Vertices + id;
+  Incoherent<vertex_t>::RO v( v_add, 1 );
+
+  // start index of my children pointers
+  childIndex = v->childIndex;
+
+  // how many children I have
+  numChildren = v->numChildren;
+
+  // parallel loop over the child list for this vertex
+  parallel_for(fn=&search_children, start=childIndex, iters=numChildren);
+}
+
+void search_children( start, iters ) {
+  // take advantage of spatial locality in the array of children
+  GlobalAddress<int64_t> child_base_addr = ChildrenPointers + start;
+  Incoherent<int64_t>::RO childIds( child_base_addr, iters );
+
+  // spawn a task to visit each child
+  for (i = 0..iters) {
+    SoftXMT_publicTask( fn=&search_vertex, id=childIds[i] );
+  }
+}
+\end{lstlisting}
+}

### 2012-07-17 doc/asplos13/method.tex
comment out UTS code snippet in eval

@@
 
-The metric we use is algorithmic time, which means startup and loading of the graph (from disk) is not included in the measurement.
+The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.
 Data is collected on real systems, which means minor variations in runtime exist from run to run.
@@
 %todo: say tree explore is same as a search except we are additionally needing to synchronize on all vertices visited
+
+The Grappa version of the in-memory tree search uses the asynchronous parallel for loop over a visited vertex's children list.
+A larger threshold can be set to take advantage of this locality by processing more child pointers sequentially in a single task.
 
@@
 }
+
+\paragraph{Breadth-first-search (BFS)} This is the primary kernel for the Graph500 benchmark and is what currently determines the ranking of machines on the Graph500 list~\cite{graph500list}.
+As a whole, the Graph500 benchmark suite is designed to bring the focus of system design on data-intensive workloads, particularly large-scale graph analysis problems, that are important among cybersecurity, informatics, and network understanding workloads.
+Graph500's definition of BFS of a graph consists of computing a search tree containing parent nodes for each traversed vertex during the search.
+While this is a relatively simple problem to solve, it exercises the random-access and fine-grained synchronization capabilities of a system as well as being a primitive in many other graph algorithms.
+Performance is measured in \emph{traversed edges per second} (TEPS), where the number of edges is the edges making up the generated BFS tree.
+One of the reference implementations of Graph500 BFS is for the XMT.
+We compare this against a straightforward Grappa implementation that algorithmically matches the XMT reference code to allow us to fairly compare the execution systems.
+We do not employ algorithmic improvements beyond the reference code, though there are many \cite{Beamer:Graph500,Yoo:FixedPointGraph500}.

### 2012-07-17 doc/asplos13/grappa.tex
Removing trivial change
grappa system diagram
initial workstealing details + impl

@@
 \section{Grappa} \label{sec:grappa}
-[trivial change]
+
+\begin{figure}[t]
+\begin{center}
+  \includegraphics[width=0.5\columnwidth]{figs/grappa-crappy}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
 In this section we describe the Grappa runtime system (Figure~\ref{fig:grappa}).
@@
 
-Grappa context switches between tasks non-preemptively.
-Most calls into the Grappa library, however, do initiate a context switch, and tasks can explicitly yield control.
-Grappa switches between tasks entirely within usermode, and in our experiments we have measured the context switch time to be only \checkme{40ns}.
-This time is kept small because it is done entirely in usermode and only registers specified as caller saved by the compiler need be saved/restored.
-Tasks become bound to a thread stack when no other task is runnable, or Grappa determines it is profitable to do so.
-\TODO{How does Grappa determine this?}.
+Grappa context switches between tasks non-preemptively. \TODO{too
+    vague: Most calls
+    into the Grappa library, however, do initiate a context switch}, and
+tasks can explicitly yield control.  Grappa switches between tasks
+entirely within usermode, and in our experiments we have measured the
+context switch time to be only \checkme{40ns}.  This time is kept
+small because it is done entirely in usermode and only registers
+specified as caller saved by the compiler need be saved/restored.
+Tasks become bound to a thread stack when no other task is runnable,
+or Grappa determines it is profitable to do so.  \TODO{How does Grappa
+    determine this? BrandonM: right now we don't actually spawn beyond
+the starting number of stacks, so Grappa only bounds when there exist
+idle worker and unbound task}.
 
-Tasks that are not bound to a stack are steelable by the work steeling system within Grappa.
-This load balances work across the collection of nodes executing an application built on Grappa.
-\TODO{Talk about the work steeling algorithm that we use}.
+\TODO{some of this first sentence might be in the part where we discuss our high
+    level solutions to irregular dynamic parallelism}
+The combination of running tasks on a large non-uniform system and
+dynamic parallelism creates the potential for an imbalance of tasks
+across the system. Some form of dynamic load balancing is necessary.
+Grappa employs work stealing: if a \checkme{grappa-process} determines it
+does not have enough work, it takes some from a \checkme{grappa-process}
+that has public, unstarted tasks.
 

### 2012-07-17 doc/asplos13/implementation.tex
initial workstealing details + impl

@@
 decides to run them; when this occurs, tasks are matched with {\em
-  workers}. Each worker simply a collection of status bits and a
+  workers}. Each worker is simply a collection of status bits and a
 stack, allocated at a particular core. Workers
@@
 Whenever a task yields or suspends, the scheduler makes a decision
-about what to do next. First, it determines if it is time to service
+about what to do next. First, it determines if
 the communication resources should be serviced. This is done on a
@@
 
-\subsubsection{Work stealing}
+\subsubsection{Work stealing} \label{subsec:implementation-worksteal}
+When a \TODO{grappa process} runs
+out of work, it becomes a ``thief'' and asks a random other
+\TODO{grappa process} (``victim'') for work. If the victim has tasks
+to spare in its public queue, they will be transferred to the thief's public queue.
 
-\TODO{BrandonM or Jacob}
+Most study of work stealing assumes that a processor has one worker
+thread and so it only steals when utilization would go to zero. 
+%that is,when there are no more local tasks available. 
+Grappa, on the other hand, relies on having many worker threads per core. The
+core would be fully-utilized if there is always something on the ready
+queue. Even if there are many active tasks, if they are all suspended
+for long-latency network requests, then the core is underutilized.
+
+choices we've made
+-random victims
+-steal from FIFO side of queue
+-steal when at the moment readyQ empty and task queues empty
+-steal min(chunk\_size, half)
+
+Recent work has explored scalable work stealing in terms of efficient
+termination detection. 
+\TODO{we should reference or put our global task joiner here}
 
 \subsection{Communication}
+
+\TODO{Jacob}
 

### 2012-07-17 doc/asplos13/implementation.tex
work steal implementation: discussion of when to steal + termination

@@
 \item{Public task queue} This is a LIFO queue of tasks that are
-  waiting to be matched with workers. Tasks may be stolen from this
-  queue by other cores in the system; this is described in the next section.
+  waiting to be matched with workers. It is a local partition of a shared
+  task pool; this is described in the next section.
 \end{description}
@@
 \subsubsection{Work stealing} \label{subsec:implementation-worksteal}
-When a \TODO{grappa process} runs
-out of work, it becomes a ``thief'' and asks a random other
-\TODO{grappa process} (``victim'') for work. If the victim has tasks
+When a \checkme{grappa-process} runs
+out of work, it becomes a ``thief'' and asks another other
+\checkme{grappa-process} (``victim'') for work. If the victim has tasks
 to spare in its public queue, they will be transferred to the thief's public queue.
@@
 for long-latency network requests, then the core is underutilized.
+So, there is a choice: should the core use some of this idle time to
+do perform work stealing or should it just wait that local tasks will
+create more work. We choose to have a core inititiate a steal when all
+of the following conditions hold: no workers are ready to run, 
+%\checkme{there are idle worker threads --not required to mention since
+%technically would happen if have right number}
+the unstarted task queues are empty, and there are no outstanding steal requests. 
+Having one outstanding steal request allowed throttles steals to
+prevent flooding the network, but other scalable quieting mechanisms
+would be possible, such as voting tree\cite{scalable work stealing or
+cilk98} or lifelines \cite{lifelines}. Termination detection is not
+built into the Grappa task scheduler, rather, it is considered a programming error for
+the program to exit without syncing all tasks.
+
+\todo{where to put the following related}
+Fork join computations: implement with futures or continuation passing
+style
+feed-forward computations: implement with scalable task counting (we
+focus on this style in our benchmarks)
+(breifly describe global task joiner)
 
@@
 termination detection.
-\TODO{we should reference or put our global task joiner here}
 

### 2012-07-17 doc/asplos13/implementation.tex
Fixed up broken imeplemention.tex and Makefile from last commit

@@
 prevent flooding the network, but other scalable quieting mechanisms
-would be possible, such as voting tree\cite{scalable work stealing or
-cilk98} or lifelines \cite{lifelines}. Termination detection is not
+would be possible, such as voting tree\cite{scalableWorkStealingOrCilk98}
+or lifelines \cite{lifelines}. Termination detection is not
 built into the Grappa task scheduler, rather, it is considered a programming error for
@@
 
-\todo{where to put the following related}
+\TODO{where to put the following related}
 Fork join computations: implement with futures or continuation passing

### 2012-07-17 doc/asplos13/method.tex
First stab at BFS methodology section.
Minor rewording of BFS

@@
 As a whole, the Graph500 benchmark suite is designed to bring the focus of system design on data-intensive workloads, particularly large-scale graph analysis problems, that are important among cybersecurity, informatics, and network understanding workloads.
-Graph500's definition of BFS of a graph consists of computing a search tree containing parent nodes for each traversed vertex during the search.
+The BFS benchmark builds a search tree containing parent nodes for each traversed vertex during the search.
 While this is a relatively simple problem to solve, it exercises the random-access and fine-grained synchronization capabilities of a system as well as being a primitive in many other graph algorithms.
@@
 One of the reference implementations of Graph500 BFS is for the XMT.
-We compare this against a straightforward Grappa implementation that algorithmically matches the XMT reference code to allow us to fairly compare the execution systems.
+We compare this against a straightforward Grappa implementation that algorithmically matches the XMT reference code to allow us to fairly compare the execution of the two systems.
 We do not employ algorithmic improvements beyond the reference code, though there are many \cite{Beamer:Graph500,Yoo:FixedPointGraph500}.

### 2012-07-17 doc/asplos13/implementation.tex
initial communication implementation section

@@
 \begin{description}
-\item{Ready worker queue} This is a FIFO queue of tasks that are
+\item{\bf Ready worker queue} This is a FIFO queue of tasks that are
   matched with workers and are ready to execute.
-\item{Private task queue} This is a FIFO queue of tasks that must run on this core.
-\item{Public task queue} This is a LIFO queue of tasks that are
+\item{\bf Private task queue} This is a FIFO queue of tasks that must run on this core.
+\item{\bf Public task queue} This is a LIFO queue of tasks that are
   waiting to be matched with workers. It is a local partition of a shared
@@
 
-\TODO{Jacob}
+In order to mitigate the low message injection rate limits of commodity
+networks, Grappa's communication has two layers.
 
-\subsection{Aggregation}
+At the upper layer, Grappa implements asynchronous active messages
+\cite{vonEicken92}. Each message consists of a function pointer, an
+optional argument payload, and an optional data payload. When a task
+sends a message, the message is copied to a send queue associated with
+the message's destination and the task continues execution.
 
-\subsection{Synchronization}
+Grappa's lower networking layer aggregates the upper layer's messages
+to improve performance. In our applications we see that the bulk of
+the messages sent are small. \TODO{include stats} Commodity networks,
+even ones designed for high-performance computing, provide low
+bandwidth utilization at small message sizes. \TODO{include stats?} To
+make the best use of the network, we must convert our small messages
+into large ones.
 
-\subsection{Consistency}
+Grappa leverages its lightweight context switching to do this
+conversion. When a task sends a message, the message is not
+immediately sent. Instead, it is stored in a queue associated with its
+destination, to be sent later. If the task requires a response, it
+suspends itself and the scheduler finds other work to run while the
+task is waiting. When the reply arrives, the suspended task is woken.
+
+There are three situations in which a queue of aggregated messages is
+sent. First, each queue has a message size threshold, chosen to give
+reasonable network performance. If the size in bytes of a queue is
+above the threshold, the contents of the queue are sent
+immediately. Second, each queue has a wait time threshhold. If the
+oldest message in a queue has been waiting longer than this threshold,
+the contents of the queue are sent immediately, even if the queue size
+is lower than the message size threshold.  Third, queues may be
+explicitly flushed in situations where the programmer wants to
+minimize the latency of a message at the cost of bandwidth
+utilization.
+
+The network layer is serviced by polling. Periodically when a context
+switch occurs, the Grappa scheduler switches to the network polling
+thread. This thread has three responsibilities. First, it polls the
+lower-level network layer to ensure it makes progress. Second, it
+deaggregates received messages and executes active message
+handlers. Third, it checks to see if any aggregation queues have
+messages that have been waiting longer than the threshold; if so, it
+sends them.
+
+Underneath the aggregation layer, Grappa uses the GASNet communication
+library \cite{GASNet} to actually move data. All interprocess
+communication, whether on or off a cluster node, is handled by the
+GASNet library. GASNet is able to take advantage of many communication
+mechanism, including Ethernet and Infiniband between nodes, as well as
+shared memory within a node.
+
+Some networks provide access to a remote machine's memory
+directly. This would seem to be a good fit for a programming model
+focused on global shared memory, but in fact we do not use it. RDMA
+operations are subject to the same message rate limitations as all
+other messages on these cards, and thus using raw RDMA operations for
+our small messages would make inefficient use of bandwidth. Instead,
+we implement remote memory operations with active messages as
+described in the next section. This means that Grappa is not limited
+to RDMA-capable networks.
+
+\subsection{Memory}
+
+remote access
+
+heap
+
+allocation
+
+synchronization
+
+consistency
+>>>>>>> Stashed changes

### 2012-07-18 doc/asplos13/relatedwork.tex
related work starts

\section{Related work}

Ways to organize this:
Approaches to graphs: special purpose XMT -> cluster -> distributed frameworks {pregel,graphlab} -> limited programming model -> grappa gen purpose

dsm, pgas
lightweight threading
other graph solutions: 
-graph parallel, distributed
--pregel - but is synchronous and thus inefficient
--graphlab - asynchrounous, esp useful for problems that have been shown empirically to converge faster with asynchrony

\subsection{Multithreading to tolerate latency}
hardware: tera, cray
smt,
alewife
gpu,
software threads

\subsection{Programming models for distributed graph processing}

Distributed graph processing frameworks like Pregel \cite{pregel:2010} and Distributed GraphLab \cite{distgraphlab:vldb12} provide graph-parallel, vertex-centric programming abstractions that free the application writer from solving distributed system issues like scheduling parallelism, handling fault tolerance, and scaling communication. Pregel adopts a bulk-synchronous parallel (BSP) execution model, which makes it inefficient on workloads that could prioritize vertices. GraphLab, on the other hand, schedules vertex computations individually, allowing prioritization, which gives faster convergence in a variety of iterative algorithms. Grappa also supports dynamic parallelism with asynchronous execution, but has a more general purpose programming model, where parallelism is expressed as tasks or loop iterations. \todo{and express locality? mention pgas e.g. Chapel language implementation? How much to talk about how graphlab achieves scalable performance?}

\subsection{Partitioned Global Address space}
The goal of presenting a global view of distributed memory to the programmer is shared by the PGAS community, and is used in languages like Chapel \cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC}. In these programming models, access to shared structures look like normal memory references, but the programmer tries to minimize references to remote nodes. Unlike the usual programming goal in PGAS, since we target problems with poor locality, we design Grappa for remote references as a common case. Thread private data is still local, and unlike the XMT, we exploit locality where it does exist in an application (such as spatial locality in a graph edgelist), but we design the system to perform well in the midst of mostly remote accesses to large shared data structures. We would like to implement PGAS languages that support dynamic parallelism, like Chapel, in Grappa.

### 2012-07-18 doc/asplos13/intro.tex
draft rewrite draft rewrite

@@
 
-Computing on large graph datasets is both important and challenging.
-Underlying many applications critical for national security, market analysis, and social computing are algorithms that process large graphs.
-Three basic factors make these applications challenging for traditional computing systems: (1) the datasets are enormous -- far more than can fit within a single computing system's memory; (2) the computation to communication ratio is low -- applications spend most of their time fetching nodes and edges of the graph, and the computation performed on that data often takes a trivial amount of time by comparison; and (3) graph datasets have little to no locality, defying traditional mechanisms architects have designed, such as caches and prefetching, for improving memory system performance.
+Irregular applications are those whose parallel execution generates tasks whose work, interdependences, or memory accesses are highly sensitive to input.
+Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow.
+Contemporary examples include real-time mining or querying large unstructured data sets in the business, national security, machine learning, data-driven science, and social network computing domains.
+For these emerging applications, fast response -- given the sheer amount of data -- requires multinode systems.
+The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband.
+Our vision is to facilitate scaling of irregular applications efficiently on these mass market systems.
 
-Traditionally, when organizations had to process large graphs they would turn to the Cray XMT~\cite{xmt} line of machines.
-The XMT is well suited to the processing of large graphs: it has a large distributed shared memory, and instead of relying on caches to exploit locality, it relies on fine-grained multithreading to tolerate access latency.
-Locality in the data access stream is not required, if sufficient threads exist to overlap computation with communication.
-The XMT machines have two basic detractions, however: (1) they are not inexpensive because customers cannot exploit the economics of commodity hardware and software; and (2) applications designed to execute on commodity clusters cannot be automatically ported to the XMT with good performance.
-Users of the XMT need to commit to its style and language of software development.
+What makes our vision challenging is that irregular applications do not readily exploit the features of commodity systems.
+This is true even for single node systems.
+It is not atypical for any given task's data references to be spread randomly across the entire memory of the system.
+Yet, for commodity processors' caches to be effective, computations must exploit spatial locality.
+Furthermore, the addresses of references are often determined in the task only just before their values are used.
+Unfortunately, mass market prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed.
+Because commodity processors' pipeline utilization relies on spatial and temporal locality -- or, at least the ability to predict in advance what is needed -- they stall often when executing irregular applications.
 
-In this paper we describe Grappa, a software runtime system built for your typical high-performance compute cluster, that makes graph algorithms fast and scalable.
-Grappa is inspired by the XMT hardware, but implements the key mechanisms of that design in software, and adds additional software components to tolerate the differences between a typical compute cluster and the more specialized XMT hardware.
-A useful analogy is that Grappa is to graph processing, as Hadoop~\cite{hadoop} is to map-reduce.
+On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.
+Irregular applications also present a challenge to mass market network technology, which is designed to transfer blocks of data, not the word-sized references emitted by irregular application tasks.
+Injection rate into the networks is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.
+While some irregular applications can be restructured to better exploit locality, aggregate requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.
 
-By design, Grappa is built on as much freely available and commodity infrastructure as possible.
+An idealized alternative to demanding this expertise of programmers was provided by the fully custom Tera MTA-2~\cite{mta-2} system.
+The MTA was well-suited to extremely irregular applications: it had a large distributed shared memory with no caches, relying on multithreaded processors to tolerate memory latency.
+On every clock cycle, each processor would execute a ready instruction chosen from one of its 128 hardware thread contexts.
+Any application with sufficient thread parallelism would fully utilize the processors independent of the memory access pattern (modulo hot-spots).
+The network was designed with an single word injection rate that matched the processor clock frequency and sufficient bandwidth to sustain a reference from every processor on every clock cycle.
+Unfortunately, while an excellent match to extremely irregular applications, the MTA was not cost-effective on applications that could exploit locality and had very poor single threaded performance, making it a commercial failure.
+The Cray XMT approximates the Tera MTA-2, substantially reducing its cost but not overcoming its narrow range of applicability.
+
+In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.
+This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating with inexpensive mass market hardware components the approach taken by Tera.
+We exploit the increased aggregate instruction rate per socket relative to chip bandwidth.
+That is, where locality friendly computations make use of instructions to get work done in cache, we use what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.
+Ultimately, the opportunity is to cover the spectrum of irregular to regular computation:  where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.
+Thus we make best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.
+
+Thus far, we have developed Grappa:  a latency tolerant runtime for scaling irregular applications on x86 clusters.
+Grappa leverages as much freely available and commodity infrastructure as possible.
 We use unmodified Linux for the operating system.
 An off the shelf user-mode infiniband device driver stack~\cite{Melonox?} is used for network interfacing.
-MPI~\cite{mpi} is used for process setup and tere down.
-Finally, GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes and active message invocation.
+MPI~\cite{mpi} is used for process setup and tear down.
+Finally, GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
 
-To this commodity hardware and software mix Grappa adds three main software components: (1) a lightweight tasking layer that supports a context switch in as few as \checkme{40ns} and distributed load balancing; (2) a distributed shared memory layer that supports customary operations such as read and write as well as remote operations such as fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that mitigates the problem where commodity networks are designed to achieve peek bandwidth only on large packet sizes, yet graph algorithms tend to want to fetch only a handful of bytes at a time.
+To this commodity hardware and software mix Grappa adds three main software components: (1) a lightweight tasking layer that supports a context switch in as few as \checkme{40ns} and distributed load balancing; (2) a distributed shared memory layer that supports customary operations such as read and write as well as remote operations such as fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that mitigates the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
 Each of these components will be described in more detail in Section~\ref{sec:grappa}.
 
-Grappa achieves its design goal, which is to make graph algorithms execute quickly and scalably on commodity clusters.
+{\em Provide evidence we have succeeded -- will rewrite as necessary based on final results:}
+Grappa achieves its design goal, which is to make irregular applications execute quickly and scalably on commodity clusters.
 Our yardstick for comparison is the XMT hardware itself.
@@
 \TODO{Add roadmap here}
+
+---

### 2012-07-18 doc/asplos13/implementation.tex
oops, fix build break
initial memory section

@@
 At the upper layer, Grappa implements asynchronous active messages
-\cite{vonEicken92}. Each message consists of a function pointer, an
+\TODO{cite vonEicken92}. Each message consists of a function pointer, an
 optional argument payload, and an optional data payload. When a task
@@
 Underneath the aggregation layer, Grappa uses the GASNet communication
-library \cite{GASNet} to actually move data. All interprocess
+library \TODO{cite GASNet} to actually move data. All interprocess
 communication, whether on or off a cluster node, is handled by the
@@
 
-remote access
+Grappa implements a software distributed shared memory. This memory is
+exposed through two address spaces; one optimized for global access to
+node-local data, and another optimized for high aggregate random
+access bandwidth to low locality data.
 
-heap
+All global addresses in Grappa are 64-bit values. Each byte of global
+memory is associated with a {\em home core}; this core is responsible
+for allocating and modifying that memory. Since today's commodity
+processors support a 48-bit virtual address space \TODO{cite?}, Grappa
+uses the spare bits to distinguish which address space a global
+address belongs to, and to represent the home core of a global
+address.
 
-allocation
+The first global address space is the {\em two-dimensional} address
+space. This uses a traditional PGAS addressing model, where each
+address is a tuple of a rank in the job (or global process ID) and an
+address in that process. The lower 48 bits of the address hold a
+virtual address in the process. The top bit is set to indicate that
+this a 2D address. This leaves 15 bits for network endpoint ID, which
+limits our scalability to $2^15$ endpoints.
 
-synchronization
+Any local data can be made accessible by other nodes in the system by
+wrapping the address and node ID into a 2D global address. This
+address can then be accessed with delegate or cache operations. At the
+destination the address is converted into a canonical x86 address by
+replacing the upper bits with the sign-exteded upper bit of the
+virtual address.
 
-consistency
->>>>>>> Stashed changes
+The second global address space is the {\em linear} address space, so
+named because nearby addresses can be treated as adjacent even if they
+are stored on different nodes. Addresses in the linear address space
+are partitioned across the cluster in a block-cyclic fashion. Choosing
+the block size involves trading off sequential bandwidth agains
+aggregate random access bandwidth. Smaller block sizes help spread
+data across all the memory controllers in the cluster, but larger
+block sizes allow the locality-optimized memory controllers to provide
+increased sequential bandwidth. The block size is typically set to 64
+bytes, or the size of a single hardware cache line, in order to
+exploit spatial locality when available.
+
+All linear addresses refer to data allocated from a global heap. The
+heap metadata is stored on a single node. Currently all heap
+operations serialize through this node; while this has been sufficient
+for our benchmarks, parallel performance can be achieved through
+combining \TODO{ cite MAMA and flat combining }. 2D addresses may
+refer to memory allocated from a single processes' heap or from a
+task's stack.
+
+All global memory operations are implemented with active
+messages. Requestors construct a descriptor holding the current
+task/worker ID and space for a response, issue their request active
+message to the node given by the request address, and suspend
+themselves. When the request arrives at the address' home node, the
+operation is performed and a response active message is sent. When the
+response arrives at the requesting node, the descriptor is filled in
+with the results of the operation and the requesting task is woken.
+
+Delegate operations are used for syncronization. Their modifications
+are always executed at the home core of their address. Each task may
+have only one outstanding delegate operation. We limit delegate
+operations to operate objects in the 2D address space or objects that
+fit in a single block of the linear address space. No context switches
+are allowed while the data it being modified. Given these
+restrictions, we can ensure that delegate operations from multiple
+requestors are always serialized through a single core in the system,
+providing atomic semantics without using atomic operations.
+
+Cache operations are used to exploit spatial or temporal
+locality. They allow a single task to buffer a range of data, read or
+modify it, and write the modified version back to its home node. All
+acquires and releases are expressed explicitly by the
+programmer. Cache operations may operate on data that spans multiple
+blocks of the linear address space; the programmer provides the
+starting address and length, and the cache issues as many active
+messages as necessary to retrieve all necessary data.
+
+All current cache operations are incoherent; two tasks may
+concurrently modify the same block with their writes interfereing
+unless protected by synchronization through delegate operations. While
+this has been sufficient for our benchmarks, it is possible to embed
+synchronization in the cache active messages and reduce the number of
+network round trips necessary.

### 2012-07-18 doc/asplos13/evaluation.tex
comment out UTS code snippet in eval
added paragraph making note of Cray XMT UTS results

@@
 Scaling up, Grappa adds 2Mvert/s/node versus XMT's 850Kvert/s/node.
+
 \TODO{why is XMT getting different results for T1L,T2L; would expect it to be the machine with the most uniform results.}
+
+{\em you can rewrite this or blow it off -- just trying to get some thoughts in before leaving}
+Despite our efforts to tune the UTS implementation specific to Cray XMT, performance does not scale well with increasing processor count, flattening out around sixty processors.
+When we increase the size of the tree, we find that performance does not improve, suggesting that performance is not limited by task parallelism.
+Cray's performance tools show that an increasing number of memory retry operations are generated by code within the runtime system.
+Retries are performed by the memory controller when remote synchronization operations fail to find the full-bit associated with each memory location in the unavailable state.
+Retries are issued at low priority relative to new memory operations issued by the processor by other contexts, so they consume what would otherwise be unused injection bandwidth.
+On a full-bandwidth system such as the MTA-2, retries have no impact on the progress of tasks other than their own.
+On a Cray XMT, network bandwidth is limited, so retries create congestion.
+In comparison, Grappa performs synchronization without retries, delaying response at the receiving end until ready to notify the sender to proceed.
+This saves bandwidth and permits scaling of tasks performing synchronization even on low injection rate networks.
 

### 2012-07-18 doc/asplos13/implementation.tex
Rearranged implementation and made a pass on it.

@@
 We now discuss implementation details for the major system components discussed in Section~\ref{sec:grappa}.
+
+\subsection{Memory}
+
+Grappa implements a software distributed shared memory. This memory is
+exposed through two address spaces; one optimized for global access to
+node-local data, and another optimized for high aggregate random
+access bandwidth to low locality data.
+
+\paragraph{2D global addresses:} All global addresses in Grappa are 64-bit values. Each byte of global
+memory is associated with a {\em home core}; this core is responsible
+for allocating and modifying that memory. Since today's commodity
+processors support a 48-bit virtual address space \TODO{cite?}, Grappa
+uses the spare bits to distinguish which address space a global
+address belongs to, and to represent the home core of a global
+address.
+
+The first global address space is the {\em two-dimensional} address
+space. This uses a traditional PGAS addressing model, where each
+address is a tuple of a rank in the job (or global process ID) and an
+address in that process. The lower 48 bits of the address hold a
+virtual address in the process. The top bit is set to indicate that
+this a 2D address. This leaves 15 bits for network endpoint ID, which
+limits our scalability to $2^15$ endpoints.
+
+Any local data can be made accessible by other nodes in the system by
+wrapping the address and node ID into a 2D global address. This
+address can then be accessed with delegate or cache operations. At the
+destination the address is converted into a canonical x86 address by
+replacing the upper bits with the sign-exteded upper bit of the
+virtual address.
+
+\paragraph{Linear global addresses:} The second global address space is the {\em linear} address space, so
+named because nearby addresses can be treated as adjacent even if they
+are stored on different nodes. Addresses in the linear address space
+are partitioned across the cluster in a block-cyclic fashion. Choosing
+the block size involves trading off sequential bandwidth against
+aggregate random access bandwidth. Smaller block sizes help spread
+data across all the memory controllers in the cluster, but larger
+block sizes allow the locality-optimized memory controllers to provide
+increased sequential bandwidth. The block size, which is configurable, is typically set to 64
+bytes, or the size of a single hardware cache line, in order to
+exploit spatial locality when available.
+
+All linear addresses refer to data allocated from a global heap. The
+heap metadata is stored on a single node. Currently all heap
+operations serialize through this node; while this has been sufficient
+for our benchmarks, parallel performance can be achieved through
+combining \TODO{ cite MAMA and flat combining }. 2D addresses may
+refer to memory allocated from a single processes' heap or from a
+task's stack.
+
+\TODO{Add a figure that depicts both forms of global addresses.}
+
+All global memory operations are implemented with active
+messages. Requestors construct a descriptor holding the current
+task/worker ID and space for a response, issue their request active
+message to the node given by the request address, and suspend
+themselves. When the request arrives at the address' home node, the
+operation is performed and a response active message is sent. When the
+response arrives at the requesting node, the descriptor is filled in
+with the results of the operation and the requesting task is woken.
+
+\paragraph{Delegates:} Delegate operations are used for syncronization. Their modifications
+are always executed at the home core of their address. Each task may
+have only one outstanding delegate operation. We limit delegate
+operations to operate on objects in the 2D address space or objects that
+fit in a single block of the linear address space. No context switches
+are allowed while the data it being modified. Given these
+restrictions, we can ensure that delegate operations for the same address from multiple
+requestors are always serialized through a single core in the system,
+providing atomic semantics without using atomic operations.
+
+Cache operations are used to exploit spatial or temporal
+locality. They allow a single task to buffer a range of data, read or
+modify it, and write the modified version back to its home node. All
+acquires and releases are expressed explicitly by the
+programmer. Cache operations may operate on data that spans multiple
+blocks of the linear address space; the programmer provides the
+starting address and length, and the cache issues as many active
+messages as necessary to retrieve all necessary data.
+
+\comment{If anything, this is just a footnote. -M
+
+All current cache operations are incoherent; two tasks may
+concurrently modify the same block with their writes interfereing
+unless protected by synchronization through delegate operations. While
+this has been sufficient for our benchmarks, it is possible to embed
+synchronization in the cache active messages and reduce the number of
+network round trips necessary.}
 
@@
 \item[Function pointer] This 64-bit value indicates what routine
-  should run. We depend on the fact that all processes in the system
-  load the same binary so that function pointers are consistent across
-  processes.
-\item[Private argument] This 64-bit value is used for task-specific
-  data, and is often a loop index.
-\item[Shared argument] This 64-bit value is used for data shared
-  between all tasks that are part of a loop, and is usually a global
-  pointer to a struct. Each task uses its private argument to 
-\item[Synchronization argument] This 64-bit value is used to determine
+  should run. We configure the system nodes to disable address randomization, so that we can depend on the fact that function pointers are valid across process images.
+\item[Private argument] This 64-bit value is typically used for task-specific
+  data (e.g., a loop index).
+\item[Shared argument] This 64-bit value is typically used for data shared
+  between all tasks that are part of a loop.
+\item[Synchronization argument] This 64-bit value is typically used to determine
   when all tasks that are part of a loop have finished. It is usually
@@
 are treated as arbitray 64-bit values in the runtime, and can be used
-to store whatever the user wants.
+for any purpose.
 
@@
   workers}. Each worker is simply a collection of status bits and a
-stack, allocated at a particular core. Workers
+stack, allocated at a particular core.
 
@@
 can save even less state. Our context switch routine appears to the
-compiler as inline assembly; we can declare all the registers we need
+compiler as inline assembly; we declare all the registers we need
 to save as ``clobbered'' by the inline assembly routine, and the
-compiler will issue its own save and restore code. This allows the
+compiler will issue its own save and restore code as needed. This allows the
 compiler to avoid saving any registers that are not used, or are used
-for temporary values that are not needed after the context switch. In
+for temporary values that are not needed after the context switch.
+\comment{I highly doubt this is important in real apps beyond our kernels.  -M
+
+In
 particular, only one of our applications does floating point
 operations; all the other applications can safely ignore the x87 and
-SSE2 registers as part of context switching.
+SSE2 registers as part of context switching.}
 
@@
 
+\TODO{What is a worker?  We don't realyl describe this.  -M}
+
 Whenever a task yields or suspends, the scheduler makes a decision
@@
 values are not shared between cores. We use timestamp values only to
-ensure period execution of communication maintenance tasks; being
+ensure periodic execution of communication maintenance tasks; being
 wrong will only change the rate of these events with a potential
-performance cost, and no correctness violation. Nevertheless, we
+performance cost, but does not cause a correctness violation. Nevertheless, we
 assume Grappa will run on machines with modern timestamp counter
@@
 thread and so it only steals when utilization would go to zero. 
-%that is,when there are no more local tasks available. 
 Grappa, on the other hand, relies on having many worker threads per core. The
@@
 So, there is a choice: should the core use some of this idle time to
-do perform work stealing or should it just wait that local tasks will
+perform work stealing or should it just wait as it is possible that local tasks will
 create more work. We choose to have a core inititiate a steal when all
 of the following conditions hold: no workers are ready to run, 
-%\checkme{there are idle worker threads --not required to mention since
-%technically would happen if have right number}
 the unstarted task queues are empty, and there are no outstanding steal requests. 
-Having one outstanding steal request allowed throttles steals to
+Having only one outstanding steal request throttles steals to
 prevent flooding the network, but other scalable quieting mechanisms
-would be possible, such as voting tree\cite{scalableWorkStealingOrCilk98}
+would be possible, such as a voting tree\cite{scalableWorkStealingOrCilk98}
 or lifelines \cite{lifelines}. Termination detection is not
@@
 In order to mitigate the low message injection rate limits of commodity
-networks, Grappa's communication has two layers.
+networks, Grappa's communication stack has two layers.
 
@@
 GASNet library. GASNet is able to take advantage of many communication
-mechanism, including Ethernet and Infiniband between nodes, as well as
+mechanisms, including Ethernet and Infiniband between nodes, as well as
 shared memory within a node.
@@
 our small messages would make inefficient use of bandwidth. Instead,
-we implement remote memory operations with active messages as
-described in the next section. This means that Grappa is not limited
+we implement remote memory operations with active messages. A byproduct of this design decision is that Grappa is not limited
 to RDMA-capable networks.
-
-\subsection{Memory}
-
-Grappa implements a software distributed shared memory. This memory is
-exposed through two address spaces; one optimized for global access to
-node-local data, and another optimized for high aggregate random
-access bandwidth to low locality data.
-
-All global addresses in Grappa are 64-bit values. Each byte of global
-memory is associated with a {\em home core}; this core is responsible
-for allocating and modifying that memory. Since today's commodity
-processors support a 48-bit virtual address space \TODO{cite?}, Grappa
-uses the spare bits to distinguish which address space a global
-address belongs to, and to represent the home core of a global
-address.
-
-The first global address space is the {\em two-dimensional} address
-space. This uses a traditional PGAS addressing model, where each
-address is a tuple of a rank in the job (or global process ID) and an
-address in that process. The lower 48 bits of the address hold a
-virtual address in the process. The top bit is set to indicate that
-this a 2D address. This leaves 15 bits for network endpoint ID, which
-limits our scalability to $2^15$ endpoints.
-
-Any local data can be made accessible by other nodes in the system by
-wrapping the address and node ID into a 2D global address. This
-address can then be accessed with delegate or cache operations. At the
-destination the address is converted into a canonical x86 address by
-replacing the upper bits with the sign-exteded upper bit of the
-virtual address.
-
-The second global address space is the {\em linear} address space, so
-named because nearby addresses can be treated as adjacent even if they
-are stored on different nodes. Addresses in the linear address space
-are partitioned across the cluster in a block-cyclic fashion. Choosing
-the block size involves trading off sequential bandwidth agains
-aggregate random access bandwidth. Smaller block sizes help spread
-data across all the memory controllers in the cluster, but larger
-block sizes allow the locality-optimized memory controllers to provide
-increased sequential bandwidth. The block size is typically set to 64
-bytes, or the size of a single hardware cache line, in order to
-exploit spatial locality when available.
-
-All linear addresses refer to data allocated from a global heap. The
-heap metadata is stored on a single node. Currently all heap
-operations serialize through this node; while this has been sufficient
-for our benchmarks, parallel performance can be achieved through
-combining \TODO{ cite MAMA and flat combining }. 2D addresses may
-refer to memory allocated from a single processes' heap or from a
-task's stack.
-
-All global memory operations are implemented with active
-messages. Requestors construct a descriptor holding the current
-task/worker ID and space for a response, issue their request active
-message to the node given by the request address, and suspend
-themselves. When the request arrives at the address' home node, the
-operation is performed and a response active message is sent. When the
-response arrives at the requesting node, the descriptor is filled in
-with the results of the operation and the requesting task is woken.
-
-Delegate operations are used for syncronization. Their modifications
-are always executed at the home core of their address. Each task may
-have only one outstanding delegate operation. We limit delegate
-operations to operate objects in the 2D address space or objects that
-fit in a single block of the linear address space. No context switches
-are allowed while the data it being modified. Given these
-restrictions, we can ensure that delegate operations from multiple
-requestors are always serialized through a single core in the system,
-providing atomic semantics without using atomic operations.
-
-Cache operations are used to exploit spatial or temporal
-locality. They allow a single task to buffer a range of data, read or
-modify it, and write the modified version back to its home node. All
-acquires and releases are expressed explicitly by the
-programmer. Cache operations may operate on data that spans multiple
-blocks of the linear address space; the programmer provides the
-starting address and length, and the cache issues as many active
-messages as necessary to retrieve all necessary data.
-
-All current cache operations are incoherent; two tasks may
-concurrently modify the same block with their writes interfereing
-unless protected by synchronization through delegate operations. While
-this has been sufficient for our benchmarks, it is possible to embed
-synchronization in the cache active messages and reduce the number of
-network round trips necessary.

### 2012-07-18 doc/asplos13/evaluation.tex
description of UTS trees in eval

@@
 
-We ran UTS-mem on Grappa and the XMT with two different 100M-vertex trees with a geometric distribution numChildren distribution.
-Figure~\ref{fig:grappa-xmt-uts} shows the performance in terms of number of vertices visited per second versus number of compute nodes.
-The Grappa results are for the best parameter values from a limited search over \flushtimeout~and \asyncforthr~\TODO{want to say this in general at beginning of eval?}.
-Grappa achieves 3x better performance for 12 nodes.
-Scaling up, Grappa adds 2Mvert/s/node versus XMT's 850Kvert/s/node.
+We ran UTS-mem on Grappa and the XMT with a binomial and two geometric
+100M-vertex trees and a geometric 1.6B-vertex tree
+\footnote{T3L: -t 0 -b 2000 -q 0.200014 -m 5 -r 7\\
+          T2L: -t 1 -a 2 -d 23 -b 7 -r 220\\
+          T1L: -t 1 -a 3 -d 13 -b 4 -r 29\\
+          T1XL: -t 1 -a 3 -d 15 -b 4 -r 29\\}.
+Both systems do relatively poorly on the binomial tree because it is very deep and a
+traversal produces little parallelism. Figure~\ref{fig:grappa-xmt-uts}
+shows the performance in terms of number of vertices visited per
+second versus number of compute nodes. The Grappa results are for the
+best parameter values from a limited search over \flushtimeout~and
+\asyncforthr~\TODO{want to say this in general at beginning of eval?}.
+Grappa with 16 machines is faster than the entire XMT. Grappa achieves \checkme{81Mvert/s} with \checkme{60 machines} and the
+XMT achieves only 50Mvert/s. Scaling up, Grappa adds 2Mvert/s/nodes
+and the XMT adds 850Kvert/s/node.
+
+Grappa
+\TODO{Preston/Simon: why the xmt stops scaling}
 

### 2012-07-19 doc/asplos13/abstract.tex
pass on abs,intro

Grappa is a latency tolerant runtime system for commodity multinode multicore
platforms that addresses the challenges posed in scaling irregular parallel
applications to large datasets. These applications are of growing importance
as interactive analysis of large unstructured data sets emerges as a
commercial and government priority. Grappa serves both as a C++ user library
and as a foundation upon which higher level languages can be developed or
adapted to support development of irregular parallel applications. Grappa
tolerates the delays to distant memory by multiplexing hundreds to thousands
of lightweight tasks to each processor core; balances load via fine-grained
workstealing; and provides efficient synchronization and consistency across
its single shared address space by delegating all accesses to any given global
location to a corresponding processor core. We present a detailed description
of the Grappa system, programming examples using the library interface, and
compare performance for several irregular benchmarks to the Cray XMT, a custom
system used to target the real time graph analytics market.

### 2012-07-19 doc/asplos13/intro.tex
pass on abs,intro

@@
 
-Irregular applications are those whose parallel execution generates tasks whose work, interdependences, or memory accesses are highly sensitive to input.
+Irregular applications are those whose parallel execution generates tasks with work, interdependences, or memory accesses that are highly sensitive to input.
 Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow.
@@
 The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband.
-Our vision is to facilitate scaling of irregular applications efficiently on these mass market systems.
+Our goal is to facilitate scaling of irregular applications efficiently on these mass market systems.
 
-What makes our vision challenging is that irregular applications do not readily exploit the features of commodity systems.
+What makes our goal challenging is that irregular applications do not readily exploit the features of commodity systems.
 This is true even for single node systems.
@@
 Furthermore, the addresses of references are often determined in the task only just before their values are used.
-Unfortunately, mass market prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed.
+Unfortunately, commodity prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed.
 Because commodity processors' pipeline utilization relies on spatial and temporal locality -- or, at least the ability to predict in advance what is needed -- they stall often when executing irregular applications.
@@
 On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.
-Irregular applications also present a challenge to mass market network technology, which is designed to transfer blocks of data, not the word-sized references emitted by irregular application tasks.
+Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.
 Injection rate into the networks is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.
@@
 
-Thus far, we have developed Grappa:  a latency tolerant runtime for scaling irregular applications on x86 clusters.
+We have developed Grappa:  a latency tolerant runtime for scaling irregular applications on x86 clusters.
 Grappa leverages as much freely available and commodity infrastructure as possible.
@@
 MPI~\cite{mpi} is used for process setup and tear down.
-Finally, GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
-
-To this commodity hardware and software mix Grappa adds three main software components: (1) a lightweight tasking layer that supports a context switch in as few as \checkme{40ns} and distributed load balancing; (2) a distributed shared memory layer that supports customary operations such as read and write as well as remote operations such as fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that mitigates the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
+GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
+To this commodity hardware and software mix Grappa adds three main software components: (1) a lightweight tasking layer that supports a context switch in as few as \checkme{40ns} and distributed load balancing; (2) a distributed shared memory layer that supports customary operations such as read and write as well as remote operations such as fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
 Each of these components will be described in more detail in Section~\ref{sec:grappa}.
@@
 
-\TODO{Add roadmap here}
-
----
+\TODO{Add roadmap here. Luis: do we really want a roadmap here? We have enough about what we have done I think, so we might not need to sell things we plan to do.}

### 2012-07-19 doc/asplos13/grappa.tex
Rework of "Grappa" section

--- 2012-07-11 doc/asplos13/grappa.tex
@@
 
-In this section we describe the Grappa runtime system (Figure~\ref{fig:grappa}).
-Grappa is a software runtime system that one adds to a commodity HPC cluster.
-It provides the following software abstractions to applications: \textbf{memory}, a global distributed shared memory and \textbf{tasking}, co-operative multithreading with automatic load balancing.
-Under the hood a third component, \textbf{communication}, a network packet scheduler/aggregator is key to achieving high performance, but largely invisible to applications.
-We will now describe each of these components.
+\begin{figure}[t]
+\begin{center}
+  \includegraphics[width=0.5\columnwidth]{figs/grappa-crappy}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
 
-\subsection{Memory}
+In this section we describe the Grappa runtime system
+(Figure~\ref{fig:grappa}). Grappa is a software runtime system that
+allows a commodity distributed-memory HPC cluster to be programmed as
+if it were a single large shared-memory machine. Grappa is designed to
+smooth over some of the performance discontinuities in commodity
+hardware, giving good performance when there is little locality to be
+exploited while allowing the programmer to exploit it when it is
+available.
 
-Applications written for Grappa utilize two forms of memory: local and global.
+Grappa provides three key capabilities. The first is a distributed
+shared \textbf{global memory} with high aggregate random access
+bandwidth. The second is a \textbf{tasking} system with lightweight
+multithreading and automatic load balancing through work
+stealing. Finally, under the hood, a \textbf{communication} layer uses
+network packet scheduling and aggregation to achieve high performance
+while remaining largely invisible to applications.
 
-Local memory is exactly what its name implies.
-Accesses occur through conventional pointers.
-The compiler emits an access and the memory is manipulated directly.
-Obviously, local pointers cannot access memory on other system nodes.
-Applications use local accesses for a number of things in Grappa: the stack associated with a task, accesses to localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
+In the following sections, we discuss the programmer's view of
+Grappa's main capabilities, including a subset of the Grappa API. In
+section~\ref{implementation}, we discuss the details of their
+implementation.
 
-Global memory is a distributed block cyclic heap.
-Applications allocate memory in this heap and a global pointer is returned.
-Global pointers are utilized in one of two ways: (1) via a \emph{cache} or (2) via a \emph{delegate}.
+\subsection{Tasks}
 
-\paragraph{Cache:} With a cache, applications can instruct Grappa to fetch a global pointer (of any length) and return a local pointer to a cached copy of the global memory.
-Under the hood, Grappa performs the mechanics of gathering chunks of data from multiple system nodes and presenting a conventional appearing linear block of memory as a pointer into a cache.
-Caches are either read-only or read-write and alternatively incoherent or coherent.
-Read-only and and read-write caches are what their names imply.
-Under the hood, the distinction between these two forms of caches lays in how they interact with the directories that maintain the coherence state for coherent access to the memory the global pointer points to.
-For coherent caches, these directories maintain a simple coherence protocol that permits multiple outstanding read-only cached copies of data to exist, or alternatively, only a single outstanding read-write cached copy to exist.
-Unlike hardware caches, these software caches require explicit release operations issued in software; i.e., applications must explicitly inform Grappa when the local pointer that points into a local cached copy of global memory is no longer needed.
-Grappa then uses this release operation to interact with the directories that control the global memory in question.
-For incoherent accesses, the directories are bypassed.
-Applications acquire a local pointer to a cached copy of the global data and can expressly push the data from this local copy back to global memory in order to post any writes.
-Obviously, applications should only use incoherent accesses when they can \emph{prove} accesses across multiple threads and system nodes are disjoint, or solely read-only, or (in rarer cases) maintaining a consistent view of memory is irrelevant to application correctness, or for debugging purposes.
+%The basic unit of locality in Grappa is the {\em core}. Each core is
+%responsible for a section of the global memory
 
-\paragraph{Delegates:} Alternatively applications can dispatch computation to be performed on individual machine-word sized chunks of global memory to the memory system itself (e.g., \emph{fetch-and-add}).
-Applications provide a pointer to a function and a global pointer to Grappa, which takes care to execute that function on the system node that actually stores the memory pointed to by the global pointer.
-In such cases, localization of the pointer is more efficient.
-Delegate operations are most often used for synchronization primitives, but really any computation is permissible.
+The basic unit of execution in Grappa is the {\em task}. Tasks are
+small; just a function pointer and some arguments. Tasks are only
+queued when they are spawned; later, when resources are free, they are
+allocated a stack, bound to a core, and executed.
 
-\TODO{Check: Do delegate operations interact with the directories?  Are they coherent?}
+What makes Grappa's tasks special is that when they perform
+long-latency operations they can give up control of their core,
+allowing the core to remain busy while waiting for the operation to
+complete. This most often done implicitly inside a call into Grappa's
+API, but it can also be done explicitly by the programmer using the
+calls shown in Figure~\ref{fig:scheduling}.
 
-Figure~\ref{fig:globalmemory} depicts the global memory architecture and access interface.
+All Grappa scheduling is done in userspace, and minimal state is
+saved, in order to minimize the context switch time. In our
+experiments, we saw context switch times as low as \TODO{40ns}.
 
-\subsection{Tasking}
+This lightweight context switching is Grappa's key enabling
+feature. It gives us the ability to tolerate latency. Given that
+ability, we are able to trade latency for throughput: by {\em
+  increasing} latency in key components in the system we are able to
+increase our aggregate random access bandwidth, our synchronization
+bandwidth, and our ability to tolerate load.
 
-\subsection{Communication}
+\begin{figure}[htbp]
+  \begin{center}
+    \begin{description}\small
+    \item[ \texttt{ yield() } ] \hfill \\
+      Gives up control of core to scheduler, queuing task to be scheduled again soon
+    \item[ \texttt{ suspend() } ] \hfill \\
+      Gives up control of core to scheduler
+    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
+      Enqueues $t$ to be scheduled again soon
+    \end{description}
+    \begin{minipage}{0.95\columnwidth}
+      \caption{\label{fig:scheduling} Grappa API: scheduling} %{-4ex}}
+    \end{minipage}
+    %\vspace{-3ex}
+  \end{center}
+\end{figure}
 
-With Grappa, from the perspective of the application, all communication between tasks occurs through global shared memory.
-These requests are ultimately translated into explicit inter system node communication requests that are dispatched to GASnet.
-Under the hood there exists a key component within Grappa, the message aggregator, that makes fine-grained communication efficient on commodity Infiniband networks.
-Infiniband achieves its peek bisection bandwidth \emph{only} when the packet sizes are relatively large (a kilobyte or more).
-In our experience, graph algorithms request objects through the caching system (previously described) with a median object size of \checkme{16} bytes.
-Our measurements, confirm manufactures published data~\cite{infinibandbandwidth}, that with this packet size the bisection bandwidth is only a small fraction, less than ~\checkme{5\%} of the peak bisection bandwidth.
-The reason for this discrepancy is the combination of overheads associated with handling each packet (in terms of bytes that form the actual Infiniband packet, processing time at the card and processing on the CPU within the driver stack).
+\subsection{Expressing parallelism}
 
-To work around this problem we perform packet aggregation.
-Instead of dispatching a packet immediately to the Infiniband hardware, Grappa holds onto an individual communication request.
-Pending requests that are to be sent to the same remote node are queued.
-Once a sufficient number of them (in terms of bytes) or a timeout occurs, they are dispatched as a single packet to GASnet and onward to the Infiniband network.
-In Section~\ref{sec:results} we explore the sensitivity of performance to these two key parameters, but the net result is that performance is maximized by aiming to aggregate for \checkme{1024} bytes and \checkme{1ms}.
+A programmer's goal in coding with Grappa should be to express as much
+parallelism as possible without worrying about where it will execute.
+Grappa then chooses where and when to execute this parallelism,
+scheduling as much work as necessary to tolerate network latencies and
+keep the systems' cores busy.
+
+Grappa provides multiple methods for expressing parallelism, shown in
+Figure~\ref{fig:expressing-parallelism}. The first is explicit task
+spawns. When the programmer identifies work that can be done in
+parallel, the work may be wrapped up in a function and queued with its
+arguments for later execution using a \texttt{spawn}. Sometimes the
+programmer may want to spawn a task on a specific core in the system,
+or at the home core of a particular memory location. Grappa provides a
+\texttt{spawn\_on} call for this purpose; the task is bound the
+specified core immediately and will never be stolen.
+
+The other method for expressing parallelism is a Cilk-like {\em
+  recursive loop decomposition} \TODO{cite}. The programmer specifies
+a function pointer along with start and end indices and an optional
+threshold; Grappa decomposes the iterations into a
+logarithmically-sized tree of tasks, stopping to execute when the
+number of iterations is below the threshold.
+
+Finally, a programmer may want to run a small piece of code on a
+particular core in the system without waiting for execution resources
+to be available. Grappa provides the \texttt{call\_on} call for this
+purpose.
+
+\begin{figure}[htbp]
+  \begin{center}
+    \begin{description}\small
+    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
+      Creates a new stealable task
+    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
+      Creates a new private task that will run on a specific core 
+    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
+      Executes iterations of a loop as stealable tasks 
+    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
+      Runs a limited function on a specific core without consuming
+      execution resources
+    \end{description}
+    \begin{minipage}{0.95\columnwidth}
+      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism} % \vspace{-4ex}}
+    \end{minipage}
+    %\vspace{-3ex}
+  \end{center}
+\end{figure}
+
+\subsection{Accessing memory}
+
+Applications written for Grappa utilize two forms of memory: local and
+global.
+
+Local memory is local to a single core in the system.  Accesses occur
+through conventional pointers.  The compiler emits an access and the
+memory is manipulated directly.  Applications use local accesses for a
+number of things in Grappa: the stack associated with a task, accesses
+to localized global memory in caches (see below), and accesses to
+debugging infrastructure that is local to each system node.  Local
+pointers cannot access memory on other cores, and are valid only on
+their home core.
+
+Large, low-locality data is stored in Grappa's global memory. All
+global data must be accessed through calls into Grappa's API, shown in
+Figure~\ref{fig:accessing-memory}.
+
+Grappa provides two methods for storing data in the global memory. The
+first is a distributed heap striped across all the machines in the
+system. The \texttt{global\_malloc} and \texttt{global\_free} calls
+are used to allocate and deallocate memory in the global heap; on
+allocation, a global pointer is returned. Grappa also allows any local
+data on the core's stacks or heap to be exported to the global address
+space and accessed by other cores across the system.
+
+There are two approaches to accessing global memory. When the
+programmer expects global data to have spatial locality to exploit,
+{\em cache} operations may be used. When there is no locality to
+exploit, {\em delegate} operations are used. Since these operations
+are expected communicate with other nodes and have high latency,
+
+With a cache, applications can instruct Grappa to fetch a global
+pointer of any length and return a local pointer to a cached copy of
+the global memory. Under the hood, Grappa performs the mechanics of
+gathering chunks of data from multiple system nodes and presenting a
+conventional appearing linear block of memory as a pointer into a
+cache. Grappa cache operations have the usual read-only and read-write
+variants, along with a write-only variant used to initialize data
+structures.
+
+When accessing and modifying low-locality data, it is more efficient
+to modify the data on its home core rather than bringing a copy to the
+requesting core and returning it after modification. Delegate
+operations provide this capability. Applications can dispatch
+computation to be performed on individual machine-word sized chunks of
+global memory to the memory system itself (e.g.,
+\emph{fetch-and-add}).
+
+Delegate operations are also the primary synchronization method in
+Grappa. Each piece of global memory is paired with a single core in
+the system; all accesses to that memory are done by that core. This
+allows delegate operations to be performed atomically on the core
+simply by making sure no other code interleaves; since context
+switches in Grappa are cooperative, this is trivial. Thus, delegate
+operations are able to provide atomic semantics without using atomic
+operations.
+
+\begin{figure}[htbp]
+  \begin{center}
+    \begin{description}\small
+      \item[ \texttt{ global\_address global\_malloc( size )} ] \hfill \\
+      \item[ \texttt{ global\_free( global\_address )} ] \hfill \\
+        Allocates and frees memory in the global heap
+      \item[ \texttt{ delegate\_read( global\_address, local\_var )} ] 
+      \item[ \texttt{ delegate\_write( global\_address, local\_var )} ] \vspace{-2ex}
+      \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] \vspace{-2ex}
+      \item[ \texttt{ delegate\_inc( global\_address, local\_var )} ] \vspace{-2ex} \hfill \\
+        Performs a memory operation at the home core of a global address
+      \item[ \texttt{ cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} ]
+      \item[ \texttt{ cache\_release( global\_address, local\_buf )} ] \vspace{-2ex} \hfill \\
+        Perform cache operations \TODO{expand}
+    \end{description}
+    \begin{minipage}{0.95\columnwidth}
+      \caption{\label{fig:accessing-memory} Grappa API: accessing memory} %{-4ex}}
+    \end{minipage}
+    %\vspace{-3ex}
+  \end{center}
+\end{figure}
+
+\subsection{Under the hood}
+
+Grappa contains two key components that are not directly visible to
+the programmer but are key to providing good performance. Both depend
+on Grappa's ability to tolerate latency in order to obtain this
+performance.
+
+The first is a load balancing mechanism that uses work stealing to
+move tasks from busy nodes to idle nodes. When a core detects that it
+has idle execution resources and no new tasks to run, it issues steal
+requests over the network. The core continues to execute remaining
+active work while the steal requests are in flight. When a steal
+request succeeds, it returns with more tasks that can be matched with
+execution resources and scheduled.
+
+The second is a network packet scheduling and aggregation system. From
+the perspective of the application, all communication between tasks in
+Grappa takes place either through global shared memory or remote task
+spawns. Both of these requests are ultimately translated into explicit
+inter-system-node/core communication requests that are dispatched to
+the Garnet library. Most of our requests are small, and commodity
+networks do not transport small packets efficiently. In order to make
+more efficient use of bandwidth, Grappa aggregates packets going to the
+same destination until it can form a larger packet that will make more
+efficient use of the network. In this way, Grappa trades per-request
+latency for application-level throughput.

### 2012-07-19 doc/asplos13/implementation.tex
Rework of "Grappa" section

@@
 
-\paragraph{2D global addresses:} All global addresses in Grappa are 64-bit values. Each byte of global
+\paragraph{2D global addresses} All global addresses in Grappa are 64-bit values. Each byte of global
 memory is associated with a {\em home core}; this core is responsible
@@
 destination the address is converted into a canonical x86 address by
-replacing the upper bits with the sign-exteded upper bit of the
+replacing the upper bits with the sign-extended upper bit of the
 virtual address.
 
-\paragraph{Linear global addresses:} The second global address space is the {\em linear} address space, so
+\paragraph{Linear global addresses} The second global address space is the {\em linear} address space, so
 named because nearby addresses can be treated as adjacent even if they
@@
 All global memory operations are implemented with active
-messages. Requestors construct a descriptor holding the current
+messages. Requesters construct a descriptor holding the current
 task/worker ID and space for a response, issue their request active
@@
 
-\paragraph{Delegates:} Delegate operations are used for syncronization. Their modifications
+\paragraph{Delegates} Delegate operations are used for low-locality accesses and synchronization. Their modifications
 are always executed at the home core of their address. Each task may
@@
 restrictions, we can ensure that delegate operations for the same address from multiple
-requestors are always serialized through a single core in the system,
+requesters are always serialized through a single core in the system,
 providing atomic semantics without using atomic operations.
 
-Cache operations are used to exploit spatial or temporal
+\paragraph{Caches} Cache operations are used to exploit spatial or temporal
 locality. They allow a single task to buffer a range of data, read or
@@
 All current cache operations are incoherent; two tasks may
-concurrently modify the same block with their writes interfereing
+concurrently modify the same block with their writes interfering
 unless protected by synchronization through delegate operations. While
@@
 While these are the most common uses of the three task arguments, they
-are treated as arbitray 64-bit values in the runtime, and can be used
+are treated as arbitrary 64-bit values in the runtime, and can be used
 for any purpose.
@@
 routine that gives control of the core back to the scheduler. Users
-may also explicity yield control to the scheduler.
+may also explicit yield control to the scheduler.
 
@@
 
-\TODO{What is a worker?  We don't realyl describe this.  -M}
+\TODO{What is a worker?  We don't really describe this.  -M}
 
@@
 perform work stealing or should it just wait as it is possible that local tasks will
-create more work. We choose to have a core inititiate a steal when all
+create more work. We choose to have a core initiate a steal when all
 of the following conditions hold: no workers are ready to run, 
@@
 focus on this style in our benchmarks)
-(breifly describe global task joiner)
+(briefly describe global task joiner)
 
@@
 Grappa's lower networking layer aggregates the upper layer's messages
-to improve performance. In our applications we see that the bulk of
-the messages sent are small. \TODO{include stats} Commodity networks,
-even ones designed for high-performance computing, provide low
-bandwidth utilization at small message sizes. \TODO{include stats?} To
-make the best use of the network, we must convert our small messages
-into large ones.
+to improve performance. Commodity networks including Infiniband
+achieves their peak injection bandwidth \emph{only} when the packet
+sizes are relatively large---on the order of multiple kilobytes. The
+reason for this discrepancy is the combination of overheads associated
+with handling each packet (in terms of bytes that form the actual
+Infiniband packet, processing time at the card and processing on the
+CPU within the driver stack). Our measurements confirm manufactures
+published data~\cite{infinibandbandwidth}, that with this packet size
+the bisection bandwidth is only a small fraction, less than
+~\checkme{5\%} of the peak bisection bandwidth.
+
+In our experiments the vast majority of requests were smaller than 255
+bytes \TODO{more precise? median?}, far too small to make efficient
+use of the network. To make the best use of the network, we must
+convert our small messages into large ones.
 
@@
 above the threshold, the contents of the queue are sent
-immediately. Second, each queue has a wait time threshhold. If the
+immediately. Second, each queue has a wait time threshold. If the
 oldest message in a queue has been waiting longer than this threshold,

### 2012-07-19 doc/asplos13/evaluation.tex
more description of trees and performance to uts eval

@@
 Both systems do relatively poorly on the binomial tree because it is very deep and a
-traversal produces little parallelism. Figure~\ref{fig:grappa-xmt-uts}
+traversal produces little parallelism, which both systems depend on.
+\todo{data from our trace would be nice here}. Figure~\ref{fig:grappa-xmt-uts}
 shows the performance in terms of number of vertices visited per
@@
 \asyncforthr~\TODO{want to say this in general at beginning of eval?}.
-Grappa with 16 machines is faster than the entire XMT. Grappa achieves \checkme{81Mvert/s} with \checkme{60 machines} and the
+Grappa with 16 machines is faster than the entire XMT. Grappa achieves
+\checkme{112Mvert/s} with \checkme{60 machines} and the
 XMT achieves only 50Mvert/s. Scaling up, Grappa adds 2Mvert/s/nodes
-and the XMT adds 850Kvert/s/node.
+and the XMT adds 850Kvert/s/node. Grappa's scaling slows down at
+\checkme{90} machines because of the unscalability of our aggregation
+mechanism as there are more network endpoints.
 
 Grappa
-\TODO{Preston/Simon: why the xmt stops scaling}
+\TODO{Preston/Simon: why the xmt stops scaling, readfe retries?}
+
+The XMT plataeus around \checkme{60} processors for 100M, 1.6B, and
+4.2B-vertex trees, so performance is not limited by availability of
+parallelism for this implementation of UTS-mem.
 

### 2012-07-19 doc/asplos13/method.tex
description of systems PAL and XMT

@@
 
+\subsection{Systems}
+For measurements, we run Grappa on a 144-node cluster of AMD
+Interlagos processors, located at Pacific Northwest National Lab
+(PNNL). Nodes have 32 core (every pair share a floating-point
+unit) 2.1-GHz processors, and 64GB of memory, and \TODO{xxx 40Gb network card}. \todo{cluster
+    uses xxx Infiniband switch}. We configure the nodes to have 32 1-GB
+hugepages to minimize TLB misses for the random access patterns we
+expect from irregular applications. The results we present are for
+this machine, but we have also run Grappa on our own 12-node Intel
+Xeon Westmere cluster.
+
+For the MTA comparison, we use a 128-node Cray XMT (3rd generation
+of the MTA), also located at PNNL. Each node consists of a 500-MHz MTA Threadstorm multithreaded
+processor that supports 128 streams. The machine uses Cray's proprietary
+SeaStar2 interconnection network.
+
 \subsection{Applications}
@@
 
-\paragraph{Unbalanced tree search (UTS)}, is a benchmark for evaluating the programmability and performance of systems for parallel applications that require dynamic load balancing \cite{Olivier:uts2006}.
-It involves traversing an unbalanced implicit tree: at each vertex, its number of children is sampled from some probability distribution, and this number of new nodes are added to a work queue to be visited.
-While this benchmark captures irregular, dynamic parallelism, we want to evaluate performance on algorithms that touch existing data.
-We augment UTS by using the traversal to create a large tree in memory, and then we traverse the in-memory tree.
-We call this benchmark UTS-mem, and the timed portion is this traversal of the in-memory tree.
+\paragraph{Unbalanced tree search in-memory (UTS-Mem)}, is a benchmark
+for evaluating the programmability and performance of systems for
+parallel applications that require dynamic load balancing
+\cite{Olivier:uts2006}. It involves traversing an unbalanced implicit
+tree: at each vertex, its number of children is sampled from some
+probability distribution, and this number of new nodes are added to a
+work queue to be visited. While this benchmark captures irregular,
+dynamic parallelism, we want to evaluate performance on algorithms
+that touch existing data. We augment UTS by using the traversal to
+create a large tree in memory, and then we traverse the in-memory
+tree.
+We call this benchmark UTS-Mem, and the timed portion is this traversal of the in-memory tree.
 The assumption is that in general the size of each subtree is unknown until that subtree has actually been traversed.

### 2012-07-19 doc/asplos13/method.tex
change uses of "todo" --> "TODO"

@@
-\section{Metholody} \label{sec:method}
+\section{Methodolody} \label{sec:method}
 
@@
 (PNNL). Nodes have 32 core (every pair share a floating-point
-unit) 2.1-GHz processors, and 64GB of memory, and \TODO{xxx 40Gb network card}. \todo{cluster
+unit) 2.1-GHz processors, and 64GB of memory, and \TODO{xxx 40Gb network card}. \TODO{cluster
     uses xxx Infiniband switch}. We configure the nodes to have 32 1-GB
@@
 
-\paragraph{Unbalanced tree search in-memory (UTS-Mem)}, is a benchmark
+\paragraph{Unbalanced tree search in-memory (UTS-Mem)}. Unbalanced
+Search Tree (UTS) is a benchmark
 for evaluating the programmability and performance of systems for
@@
 dynamic parallelism, we want to evaluate performance on algorithms
-that touch existing data. We augment UTS by using the traversal to
+that touch existing data.
+
+We augment UTS by using the traversal to
 create a large tree in memory, and then we traverse the in-memory

### 2012-07-20 doc/asplos13/implementation.tex
comments on implementation 'workers'

@@
 messages. Requesters construct a descriptor holding the current
-task/worker ID and space for a response, issue their request active
+worker \TODO{reorder this paragraph after defined} ID and space for a response, issue their request active
 message to the node given by the request address, and suspend
@@
 are always executed at the home core of their address. Each task may
-have only one outstanding delegate operation. We limit delegate
+have only one outstanding delegate operation \TODO{this is not an
+    inherent limitation, it would be better to discuss in terms of
+    fence -BM}. We limit delegate
 operations to operate on objects in the 2D address space or objects that
@@
 \comment{If anything, this is just a footnote. -M
-
 All current cache operations are incoherent; two tasks may
@@
 usually generated as part of a parallel loop decomposition, and thus
-each task needs three kinds of data:
+each task needs three kinds of data.
 \begin{description}
@@
 \item[Shared argument] This 64-bit value is typically used for data shared
-  between all tasks that are part of a loop.
+  between all tasks that are part of a loop or to specify the number
+  of loop iterations.
 \item[Synchronization argument] This 64-bit value is typically used to determine
@@
 decides to run them; when this occurs, tasks are matched with {\em
-  workers}. Each worker is simply a collection of status bits and a
+  worker} threads. Each worker is simply a collection of status bits and a
 stack, allocated at a particular core.
@@
 \comment{I highly doubt this is important in real apps beyond our kernels.  -M
+Why not? liveness is a common compiler analysis -BM
 
@@
 \end{description}
-
-\TODO{What is a worker?  We don't really describe this.  -M}
 
@@
 In order to mitigate the low message injection rate limits of commodity
-networks, Grappa's communication stack has two layers.
+networks, Grappa's communication stack has two layers: one for
+user-level messages and one for network-level messages.
 
@@
 Infiniband packet, processing time at the card and processing on the
-CPU within the driver stack). Our measurements confirm manufactures
+CPU within the driver stack). Our measurements confirm manufacturers
 published data~\cite{infinibandbandwidth}, that with this packet size
@@
 
-In our experiments the vast majority of requests were smaller than 255
+In our experiments the vast majority of requests were smaller than
+\checkme{255}
 bytes \TODO{more precise? median?}, far too small to make efficient
@@
 convert our small messages into large ones.
+\TODO{reference to aggregation evaluation later in the paper}
 

### 2012-07-20 doc/asplos13/method.tex
refinements to UTS methodology as in-memory. Still need to discuss representation of irregular parallelism, as well as tree search

@@
 
-One question that must be answered when making a comparison between the Cray XMT and a typical HPC cluster is what is a fair comparison -- these systems are quite different.
-Three options immediately come to mind: equal number of processing cores, equal number of network interfaces, and equal dollars.
-We discounted the last option fairly quickly, because it isn't a lasting data-point and, as will be evident later, would be hugely unfair to the XMT.
-The first option, cores, has some merit, but Grappa is designed for applications that have no locality in their computation.
-This means almost all of their memory accesses are remote.
-Hence, the factor that limits their performance is not processing, but communicating.
-Hence, we have chosen the middle option, network interfaces as the way to normalize across the XMT and our cluster.
-Each processor in the XMT system has its own network interface to access shared memory.
-In the HPC cluster we use, each processor (which contains 12 cores), has a single Infiniband interface.
-Hence, for our results we scale up XMT processors one for one with full system nodes.
+One question that must be answered when making a comparison between
+the Cray XMT and a typical HPC cluster is what is a fair comparison --
+these systems are quite different.  Three options immediately come to
+mind: equal number of processing cores, equal number of network
+interfaces, and equal dollars.  \checkme{We discounted the last option fairly
+quickly, because it isn't a lasting data-point and, as will be evident
+later, would be hugely unfair to the XMT.}  The first option, cores,
+has some merit, but Grappa is designed for applications that have no
+locality in their computation.  This means almost all of their memory
+accesses are remote.  Hence, the factor that limits their performance
+is not processing, but communicating.  Hence, we have chosen the
+middle option, network interfaces as the way to normalize across the
+XMT and our cluster.  Each processor in the XMT system has its own
+network interface to access shared memory.  In the HPC cluster we use,
+each processor (which contains up to 32 cores), has a single Infiniband interface.  Hence, for our results we scale up XMT processors one for one with full system nodes.
 
@@
 \subsection{Applications}
-%Traversal of an unbalanced, unpartitioned tree captures the foundations for implementing most forms of irregular parallelism. 
-%Recursion is found in every irregular divide and conquer algorithm.
-Loops, even regular ones, must be dynamically scheduled to get good load balance when run on a large, non-uniform system.
-Nested parallelism involves fine-grain creation of dynamic amounts of work.
+\TODO{
+    Traversal of an unbalanced, unpartitioned tree captures the foundations for implementing most forms of irregular parallelism. 
+Recursion is found in every irregular divide and conquer algorithm.
+Loops, even regular ones, must be dynamically scheduled to get good
+load balance when run on a large, non-uniform system. Nested
+parallelism involves fine-grain creation of dynamic amounts of work.
 Dataflow..
+}
 
-\paragraph{Unbalanced tree search in-memory (UTS-Mem)}. Unbalanced
-Search Tree (UTS) is a benchmark
+\paragraph{Unbalanced tree search in-memory (UTS-Mem)}
+Unbalanced Tree Search (UTS) is a benchmark
 for evaluating the programmability and performance of systems for
@@
 dynamic parallelism, we want to evaluate performance on algorithms
-that touch existing data.
+with irregular memory access patterns.
 
-We augment UTS by using the traversal to
+We augment UTS by using the existing traversal code to
 create a large tree in memory, and then we traverse the in-memory
-tree.
-We call this benchmark UTS-Mem, and the timed portion is this traversal of the in-memory tree.
-The assumption is that in general the size of each subtree is unknown until that subtree has actually been traversed.
+tree. We call this benchmark UTS-Mem, and the timed portion is this
+traversal of the in-memory tree. This in-memory traversal is
+unknowledgeable of the tree structure beforehand.
 
@@
 
-The Grappa version of the in-memory tree search uses the asynchronous parallel for loop over a visited vertex's children list.
-A larger threshold can be set to take advantage of this locality by processing more child pointers sequentially in a single task.
+The Grappa version of the in-memory tree search uses the
+asynchronous parallel for loop over a visited vertex's children 
+list. A larger threshold can be set to take advantage of this 
+locality by processing more child pointers sequentially in a single task.
 

### 2012-07-20 doc/asplos13/implementation.tex
Change uses of gasnet to \gasnet; remove redundant ack from method.tex:Systems

@@
 
-Underneath the aggregation layer, Grappa uses the GASNet communication
+Underneath the aggregation layer, Grappa uses the \gasnet~communication
 library \TODO{cite GASNet} to actually move data. All interprocess
 communication, whether on or off a cluster node, is handled by the
-GASNet library. GASNet is able to take advantage of many communication
+\gasnet~library. \gasnet~is able to take advantage of many communication
 mechanisms, including Ethernet and Infiniband between nodes, as well as

### 2012-07-20 doc/asplos13/method.tex
Change uses of gasnet to \gasnet; remove redundant ack from method.tex:Systems

@@
 For measurements, we run Grappa on a 144-node cluster of AMD
-Interlagos processors, located at Pacific Northwest National Lab
-(PNNL). Nodes have 32 core (every pair share a floating-point
-unit) 2.1-GHz processors, and 64GB of memory, and \TODO{xxx 40Gb network card}. \TODO{cluster
+Interlagos processors. Nodes have 32-core (every pair share a floating-point
+unit) 2.1-GHz processors, 64GB of memory, and \TODO{xxx 40Gb network card}. \TODO{cluster
     uses xxx Infiniband switch}. We configure the nodes to have 32 1-GB
@@
 
-For the MTA comparison, we use a 128-node Cray XMT (3rd generation
-of the MTA), also located at PNNL. Each node consists of a 500-MHz MTA Threadstorm multithreaded
+We compare to the MTA using a 128-node Cray XMT (3rd generation MTA). 
+Each node consists of a 500-MHz MTA Threadstorm multithreaded
 processor that supports 128 streams. The machine uses Cray's proprietary

### 2012-07-20 doc/asplos13/grappa.tex

@@
 API, but it can also be done explicitly by the programmer using the
-calls shown in Figure~\ref{fig:scheduling}.
+calls shown in Figure~\ref{fig:scheduling}. \TODO{cite UPC split phase
+    reads and writes, but include that we provide a way to overlap
+    computation}
 
@@
 
-Grappa provides multiple methods for expressing parallelism, shown in
+Grappa provides four methods for expressing parallelism, shown in
 Figure~\ref{fig:expressing-parallelism}. The first is explicit task
@@
 or at the home core of a particular memory location. Grappa provides a
-\texttt{spawn\_on} call for this purpose; the task is bound the
-specified core immediately and will never be stolen.
+\texttt{spawn\_on} call for this purpose; the task is immediately
+enqueued on the specified core and is not available to other cores.
 
-The other method for expressing parallelism is a Cilk-like {\em
-  recursive loop decomposition} \TODO{cite}. The programmer specifies
+The next method for expressing parallelism is a parallel for loop,
+where the number of iterations must be known at loop entry. The programmer specifies
 a function pointer along with start and end indices and an optional
-threshold; Grappa decomposes the iterations into a
-logarithmically-sized tree of tasks, stopping to execute when the
-number of iterations is below the threshold.
+threshold to control parallel overhead. Grappa does {\em recursive decomposition} of iterations,
+ as in the implementation of Cilk's cilk\_for construct \TODO{cite cilk\_for implementation:
+     could only find slides page6 of http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-Cilk++.pdf}.
+It generates a logarithmically-deep tree of tasks, stopping to execute
+the loop body when the number of iterations is below the threshold (``grainsize'' in Cilk).
+\TODO{we may want to make our classification purer: the for loop is
+    not Grappa's core, and is built atop tasks as a useful construct; in some sense the only way
+to express parallelism in Grappa is tasks and perhaps call\_on. -BM}
 
@@
       Runs a limited function on a specific core without consuming
-      execution resources
+      Grappa execution resources 
     \end{description}
@@
 
-Large, low-locality data is stored in Grappa's global memory. All
+Large data that is expected to be accessed with low locality is stored in Grappa's global memory. All
 global data must be accessed through calls into Grappa's API, shown in
@@
 
-Grappa provides two methods for storing data in the global memory. The
+Grappa provides two methods for {\emph storing} data in the global memory. The
 first is a distributed heap striped across all the machines in the
@@
 allocation, a global pointer is returned. Grappa also allows any local
-data on the core's stacks or heap to be exported to the global address
-space and accessed by other cores across the system.
+data on a core's stacks or heap to be exported to the global address
+space to be made accessible to other cores across the system.
 
-There are two approaches to accessing global memory. When the
-programmer expects global data to have spatial locality to exploit,
-{\em cache} operations may be used. When there is no locality to
+There are also two approaches to {\emph accessing} global memory. When the
+programmer expects a computation on shared data to have spatial
+locality to exploit, {\em cache} operations may be used. When there is no locality to
 exploit, {\em delegate} operations are used. Since these operations
 are expected communicate with other nodes and have high latency,
+\TODO{...}
 
@@
 variants, along with a write-only variant used to initialize data
-structures.
+structures. Cache operations exploit spatial locality by reducing the number of small network
+messages accessing contguous data. Languages for distributed
+shared memory systems have done optimizations to achieve a similar goal. The UPC compiler
+coalesces struct and array accesses into 
+remote get/put \cite{Chen:2005}, or Fortran D compiler's message
+vectorization that hoists small messages out of a loop
+\cite{FortranD:1992}. Caching in Grappa additionally provides a
+mechanism for exploiting temporal locality by operating on the data locally.
 
-When accessing and modifying low-locality data, it is more efficient
+When the access pattern has low-locality, it is more efficient
 to modify the data on its home core rather than bringing a copy to the
@@
 
-Delegate operations are also the primary synchronization method in
+Delegate operations, proposed in \cite{Nelson:hotpar11}, are also the primary synchronization method in
 Grappa. Each piece of global memory is paired with a single core in
@@
 switches in Grappa are cooperative, this is trivial. Thus, delegate
-operations are able to provide atomic semantics without using atomic
-operations.
+operations are able to provide atomic semantics to memory owned by one core without using atomic
+operations. Using delegation to implement isolation for any subset of
+memory has been explored in \cite{delegated:oopsla11}.
 
@@
 
-The first is a load balancing mechanism that uses work stealing to
+The first is a dynamic load balancing mechanism that uses work stealing to
 move tasks from busy nodes to idle nodes. When a core detects that it
@@
 inter-system-node/core communication requests that are dispatched to
-the Garnet library. Most of our requests are small, and commodity
+the \gasnet~library. Most of \checkme{our requests} are small, and commodity
 networks do not transport small packets efficiently. In order to make

### 2012-07-20 doc/asplos13/method.tex

@@
-\section{Methodolody} \label{sec:method}
+\section{Methodology} \label{sec:method}
 
@@
 
-\paragraph{Breadth-first-search (BFS)} This is the primary kernel for the Graph500 benchmark and is what currently determines the ranking of machines on the Graph500 list~\cite{graph500list}.
+\paragraph{Breadth-first-search (BFS)}
+This is the primary kernel for the Graph500 benchmark and is what currently determines the ranking of machines on the Graph500 list~\cite{graph500list}.
 As a whole, the Graph500 benchmark suite is designed to bring the focus of system design on data-intensive workloads, particularly large-scale graph analysis problems, that are important among cybersecurity, informatics, and network understanding workloads.
@@
 We do not employ algorithmic improvements beyond the reference code, though there are many \cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
+
+\paragraph{Betweenness Centrality}
+An important measure of the importance of particular vertices in a network is betweenness centrality (BC) \cite{freeman1979centrality}.
+By this measure, the "importance" of each vertex is computed by finding the fraction of shortest paths that pass through it, which can optionally be approximated by only computing shortest paths using a subset of the vertices as starting points.
+BC can be useful for understanding which vertices may have the greatest impact, so in social networks this could be the primary person linking two communities.
+Because it requires multiple breadth-first traversals across the entire graph and in reverse, on power-law degree graphs, this algorithm exercises random accesses rate, and load balancing.
+It also requires fine-grained synchronization on updates to vertex centrality values because multiple paths will update the same vertex.
+BC is a kernel in the DARPA High Performance Computing Systems (HPCS) Scalable Synthetic Compact Applications graph analysis (SSCA\#2) benchmark\cite{ssca2}.
+Performance for BC is also measured in TEPS, where the number of traversed edges is the total number of edges in the graph multiplied by the number of random starting vertices used in the approximate computation.
+We use the XMT implementation of BC implemented as part of the GraphCT~\cite{GraphCT} library and a comparable Grappa implementation, both of which use the parallel BC algorithm developed by Bader et al.~\cite{bader:bc}.

### 2012-07-20 doc/asplos13/grappa.tex
pass on section 3

@@
 \begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}}
+  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}\TODO{should we use the canonical Grappa picture here agumented withe the software stack?}}
 \end{minipage}
@@
 In this section we describe the Grappa runtime system
-(Figure~\ref{fig:grappa}). Grappa is a software runtime system that
-allows a commodity distributed-memory HPC cluster to be programmed as
-if it were a single large shared-memory machine. Grappa is designed to
-smooth over some of the performance discontinuities in commodity
-hardware, giving good performance when there is little locality to be
-exploited while allowing the programmer to exploit it when it is
-available.
-
+(Figure~\ref{fig:grappa}). 
+%%% Moved to intro
+% Grappa is a software runtime system that allows a commodity distributed-memory
+% HPC cluster to be programmed as if it were a single large shared-memory
+% machine.
+% Grappa is designed to
+% smooth over some of the performance discontinuities in commodity
+% hardware, giving good performance when there is little locality to be
+% exploited while allowing the programmer to exploit it when it is
+% available.
 Grappa provides three key capabilities. The first is a distributed
@@
 network packet scheduling and aggregation to achieve high performance
-while remaining largely invisible to applications.
+while remaining largely invisible to applications. We now discuss the programmer's view of
+Grappa's main capabilities, including a subset of the Grappa API. In
+Section~\ref{implementation}, we discuss the details of their
+implementation.
 
-In the following sections, we discuss the programmer's view of
-Grappa's main capabilities, including a subset of the Grappa API. In
-section~\ref{implementation}, we discuss the details of their
-implementation.
+\TODO{Should we have an example of code here, and walk thru as we explain the
+components and parts of the API?}
 
@@
 The basic unit of execution in Grappa is the {\em task}. Tasks are
-small; just a function pointer and some arguments. Tasks are only
+small; just a function pointer and arguments. Tasks are only
 queued when they are spawned; later, when resources are free, they are
@@
 experiments, we saw context switch times as low as \TODO{40ns}.
-
 This lightweight context switching is Grappa's key enabling
@@
 parallelism as possible without worrying about where it will execute.
-Grappa then chooses where and when to execute this parallelism,
+Grappa then chooses where and when to exploit this parallelism,
 scheduling as much work as necessary to tolerate network latencies and
-keep the systems' cores busy.
+keep cores busy.
 
@@
 
-The next method for expressing parallelism is a parallel for loop,
-where the number of iterations must be known at loop entry. The programmer specifies
-a function pointer along with start and end indices and an optional
-threshold to control parallel overhead. Grappa does {\em recursive decomposition} of iterations,
- as in the implementation of Cilk's cilk\_for construct \TODO{cite cilk\_for implementation:
-     could only find slides page6 of http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-Cilk++.pdf}.
-It generates a logarithmically-deep tree of tasks, stopping to execute
-the loop body when the number of iterations is below the threshold (``grainsize'' in Cilk).
-\TODO{we may want to make our classification purer: the for loop is
-    not Grappa's core, and is built atop tasks as a useful construct; in some sense the only way
-to express parallelism in Grappa is tasks and perhaps call\_on. -BM}
+The next method for expressing parallelism is a parallel for loop, where the
+number of iterations must be known at loop entry. The programmer specifies a
+function pointer along with start and end indices and an optional threshold to
+control parallel overhead. Grappa does {\em recursive decomposition} of
+iterations, as in the implementation of Cilk's cilk\_for construct \TODO{cite
+cilk\_for implementation: could only find slides page6 of
+http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-Cilk++.pdf}.
+It generates a logarithmically-deep tree of tasks, stopping to execute the
+loop body when the number of iterations is below the threshold (``grainsize''
+in Cilk). \TODO{we may want to make our classification purer: the for loop is
+not Grappa's core, and is built atop tasks as a useful construct; in some
+sense the only way to express parallelism in Grappa is tasks and perhaps
+call\_on. -BM}
 
@@
 
-\subsection{Accessing memory}
+\subsection{Memory}
 
@@
 
-Large data that is expected to be accessed with low locality is stored in Grappa's global memory. All
+Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory. All
 global data must be accessed through calls into Grappa's API, shown in
@@
 
+\paragraph{Global memory allocation}
 Grappa provides two methods for {\emph storing} data in the global memory. The
@@
 
-There are also two approaches to {\emph accessing} global memory. When the
+\paragraph{Global memory access}
+There are two approaches to {\emph accessing} global memory. When the
 programmer expects a computation on shared data to have spatial
@@
 are expected communicate with other nodes and have high latency,
-\TODO{...}
+\TODO{the runtime systems might context switch? what do we want to say here?}
 
-With a cache, applications can instruct Grappa to fetch a global
+\paragraph{Explicit caching}
+With Grappa's explicit caching support, applications can instruct Grappa to fetch a global
 pointer of any length and return a local pointer to a cached copy of
@@
 structures. Cache operations exploit spatial locality by reducing the number of small network
-messages accessing contguous data. Languages for distributed
+messages accessing contiguous data. Languages for distributed
 shared memory systems have done optimizations to achieve a similar goal. The UPC compiler
@@
 
+\paragraph{Delegate operations}
 When the access pattern has low-locality, it is more efficient
@@
 global memory to the memory system itself (e.g.,
-\emph{fetch-and-add}).
+\emph{fetch-and-add}). \TODO{should we bring up delegation in the intro?}
 
@@
 
-\subsection{Under the hood}
+\subsection{Load Balancing and Network Aggregation}
 
@@
 
-The first is a dynamic load balancing mechanism that uses work stealing to
+The first is a distributed dynamic load balancing mechanism that uses work stealing to
 move tasks from busy nodes to idle nodes. When a core detects that it
@@
 efficient use of the network. In this way, Grappa trades per-request
-latency for application-level throughput.
+latency for application-level throughput. \TODO{Jacob, can you add more details here on how this works?}

### 2012-07-20 doc/asplos13/intro.tex
pass on section 3

@@
 
-We have developed Grappa:  a latency tolerant runtime for scaling irregular applications on x86 clusters.
-Grappa leverages as much freely available and commodity infrastructure as possible.
-We use unmodified Linux for the operating system.
-An off the shelf user-mode infiniband device driver stack~\cite{Melonox?} is used for network interfacing.
-MPI~\cite{mpi} is used for process setup and tear down.
-GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
-To this commodity hardware and software mix Grappa adds three main software components: (1) a lightweight tasking layer that supports a context switch in as few as \checkme{40ns} and distributed load balancing; (2) a distributed shared memory layer that supports customary operations such as read and write as well as remote operations such as fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
+In this paper we introduce Grappa, a software runtime system that allows a
+commodity x86 distributed-memory HPC cluster to be programmed as if it were a
+single large shared-memory machine and provides scalable performance for
+irregular applications. Grappa is designed to smooth over some of the
+performance discontinuities in commodity hardware, giving good performance
+when there is little locality to be exploited while allowing the programmer to
+exploit it when it is available. Grappa leverages as much freely available and
+commodity infrastructure as possible. We use unmodified Linux for the
+operating system. An off the shelf user-mode infiniband device driver
+stack~\cite{Melonox?} is used for network interfacing. MPI~\cite{mpi} is used
+for process setup and tear down. GASnet~\cite{gasnet} is used as the
+underlying mechanism for remote memory reads and writes using active message
+invocations. To this commodity hardware and software mix Grappa adds three
+main software components: (1) a lightweight tasking layer that supports a
+context switch in as few as \checkme{40ns} and distributed load balancing; (2)
+a distributed shared memory layer that supports customary operations such as
+read and write as well as remote operations such as
+fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that
+combines short messages to mitigate the aforementioned problem that commodity
+networks are designed to achieve peak bandwidth only on large packet sizes,
+yet irregular applications tend to fetch only a handful of bytes at a time.
+
 Each of these components will be described in more detail in Section~\ref{sec:grappa}.

### 2012-07-20 doc/asplos13/grappa.tex
quick pass on sec 4. will do another pass on 3-4 once the TODOS are done.

@@
 \begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}\TODO{should we use the canonical Grappa picture here agumented withe the software stack?}}
+  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}\TODO{should we use the canonical Grappa picture here augmented withe the software stack?}}
 \end{minipage}
@@
 (Figure~\ref{fig:grappa}). 
-%%% Moved to intro
-% Grappa is a software runtime system that allows a commodity distributed-memory
-% HPC cluster to be programmed as if it were a single large shared-memory
-% machine.
-% Grappa is designed to
-% smooth over some of the performance discontinuities in commodity
-% hardware, giving good performance when there is little locality to be
-% exploited while allowing the programmer to exploit it when it is
-% available.
 Grappa provides three key capabilities. The first is a distributed
@@
 
-\TODO{Should we have an example of code here, and walk thru as we explain the
+\TODO{Should we have a code example here, and walk thru it as we explain the
 components and parts of the API?}
@@
 
-The basic unit of execution in Grappa is the {\em task}. Tasks are
+The basic unit of execution in Grappa is a {\em task}. Tasks are
 small; just a function pointer and arguments. Tasks are only
@@
 
-Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory. All
-global data must be accessed through calls into Grappa's API, shown in
-Figure~\ref{fig:accessing-memory}.
+Large data that is expected to be shared and accessed with low locality is
+stored in Grappa's global memory. All global data must be accessed through
+calls into Grappa's API, shown in Figure~\ref{fig:accessing-memory}.
 

### 2012-07-20 doc/asplos13/implementation.tex
quick pass on sec 4. will do another pass on 3-4 once the TODOS are done.

@@
 
-Grappa implements a software distributed shared memory. This memory is
-exposed through two address spaces; one optimized for global access to
-node-local data, and another optimized for high aggregate random
+Grappa implements a software distributed shared memory. This memory is exposed
+through two distinct, non-overlapping address spaces; one optimized for global
+access to node-local data, and another optimized for high aggregate random
 access bandwidth to low locality data.
 
-\paragraph{2D global addresses} All global addresses in Grappa are 64-bit values. Each byte of global
-memory is associated with a {\em home core}; this core is responsible
-for allocating and modifying that memory. Since today's commodity
-processors support a 48-bit virtual address space \TODO{cite?}, Grappa
-uses the spare bits to distinguish which address space a global
-address belongs to, and to represent the home core of a global
-address.
+All global addresses in Grappa are 64-bit
+values. Each byte of global memory is associated with a {\em home core}; this
+core is responsible for allocating and modifying that memory. Since today's
+commodity processors support a 48-bit virtual address space \TODO{cite?},
+Grappa uses the spare bits to distinguish which address space a global address
+belongs to, and to represent the home core of a global address.
 
-The first global address space is the {\em two-dimensional} address
-space. This uses a traditional PGAS addressing model, where each
-address is a tuple of a rank in the job (or global process ID) and an
-address in that process. The lower 48 bits of the address hold a
-virtual address in the process. The top bit is set to indicate that
-this a 2D address. This leaves 15 bits for network endpoint ID, which
-limits our scalability to $2^15$ endpoints.
+\paragraph{2D global addresses} The first global address space is the {\em
+two-dimensional} address space. This uses a traditional PGAS addressing model,
+where each address is a tuple of a rank in the job (or global process ID) and
+an address in that process. The lower 48 bits of the address hold a virtual
+address in the process. The top bit is set to indicates that the reference is
+a 2D address. This leaves 15 bits for network endpoint ID, which limits our
+scalability to $2^15$ endpoints.
 
-Any local data can be made accessible by other nodes in the system by
-wrapping the address and node ID into a 2D global address. This
-address can then be accessed with delegate or cache operations. At the
-destination the address is converted into a canonical x86 address by
-replacing the upper bits with the sign-extended upper bit of the
-virtual address.
+Any node-local data can be made accessible by other nodes in the system by
+wrapping the address and node ID into a 2D global address. This address can
+then be accessed with delegate and can also be cached by other nodes. At the
+destination the address is converted into a canonical x86 address by replacing
+the upper bits with the sign-extended upper bit of the virtual address.
 
@@
 
-\paragraph{Caches} Cache operations are used to exploit spatial or temporal
-locality. They allow a single task to buffer a range of data, read or
-modify it, and write the modified version back to its home node. All
-acquires and releases are expressed explicitly by the
-programmer. Cache operations may operate on data that spans multiple
-blocks of the linear address space; the programmer provides the
-starting address and length, and the cache issues as many active
-messages as necessary to retrieve all necessary data.
+\paragraph{Explicit caches} Grappa cache operations are used to exploit
+spatial or temporal locality. They allow a single task to buffer a range of
+data, read or modify it, and write the modified version back to its home node.
+All fetch and update operations of cache data are expressed explicitly by the
+programmer. Cache operations may operate on data that spans multiple blocks of
+the linear address space; the programmer provides the starting address and
+length, and the cache issues as many active messages as necessary to retrieve
+all necessary data.
 

### 2012-07-21 doc/asplos13/evaluation.tex
change uses of "todo" --> "TODO"
pass on section 3
a few notes on evaluation; adding plots soon

@@
 \section{Evaluation} \label{sec:evaluation}
-%todo put method in here, if it stays small
 
@@
 traversal produces little parallelism, which both systems depend on.
-\todo{data from our trace would be nice here}. Figure~\ref{fig:grappa-xmt-uts}
+\TODO{data from our trace would be nice here}. Figure~\ref{fig:grappa-xmt-uts}
 shows the performance in terms of number of vertices visited per
@@
 
-Grappa
-\TODO{Preston/Simon: why the xmt stops scaling, readfe retries?}
-
 The XMT plataeus around \checkme{60} processors for 100M, 1.6B, and
@@
 parallelism for this implementation of UTS-mem.
-
-\TODO{why is XMT getting different results for T1L,T2L; would expect it to be the machine with the most uniform results.}
 
@@
 
+\TODO{why is XMT getting different results for T1L,T2L; would expect it to be the machine with the most uniform results.}
+
 \TODO{sensitivity study for \asyncforthr parameter, ie caching amount. Although we can take advantage of locality in children array and the XMT cannot, we still win without it}
@@
 \end{figure}
+
+\subsection{Sensitivity}
+
+\subsection{Latency tolerance parameters}
+
+\paragraph{Benefit of aggregator}
+what happens with aggregator on and off
+
+\paragraph{Latency}
+aggregation latency and delegate latency
+
+\paragraph{Number of workers}
+We need enough workers to cover the network+aggregation latency, but more just slow things down.
+
+UTS
+BFS
+
+\subsubsection{Work stealing parameters}
+
+\paragraph{Chunk size}
+
+it's important to steal multiple tasks to 
+amortize cost of stealing
+spread work out quickly in large system
+
+there are times when it's useful to steal lots of stuff, even though
+we steal only a few things most of the time.
+
+UTS
+
+\paragraph{Parallel loop threshold}
+
+allows us to amoritize task creation overhead and exploit locality
+Gets us a 2x bump in performance
+not too sensitive
+
+UTS
+BFS

### 2012-07-21 doc/asplos13/evaluation.tex
some notes about plots

@@
 XMT achieves only 50Mvert/s. Scaling up, Grappa adds 2Mvert/s/nodes
-and the XMT adds 850Kvert/s/node. Grappa's scaling slows down at
-\checkme{90} machines because of the unscalability of our aggregation
-mechanism as there are more network endpoints.
+and the XMT adds 850Kvert/s/node. Grappa keeps scaling up through 128
+machines, although scaling declines because of the unscalability of our aggregation
+mechanism as number of network endpoints increases.
 
@@
 BFS
+
+%\begin{figure*}[ht]
+%    \begin{minipage}{0.3\linewidth}
+%        \centering
+%        \includegraphics[width=\textwidth]{figs/chunksize-uts.pdf}
+%        \caption{chunksize caption}
+%        \label{fig:chunksize-uts}
+%    \end{minipage}
+%    \begin{minipage}{0.3\linewidth}
+%        \centering
+%        \includegraphics[width=\textwidth]{figs/workers-uts.pdf}
+%        \caption{workers caption}
+%        \label{fig:workers-uts}
+%    \end{minipage}
+%    \begin{minipage}{0.3\linewidth}
+%        \centering
+%        \includegraphics[width=\textwidth]{figs/thresh-uts.pdf}
+%        \caption{threshold caption}
+%        \label{fig:thresh-uts}
+%    \end{minipage}
+%\end{figure*}
+
+Chunksize:
+amortizing cost of stealing
+spreading work out more
+most of the time we still little  (about 10 avg) but sometimes a lot
+(difference with BFS)

### 2012-07-21 doc/asplos13/relatedwork.tex
related work notes

@@
 \subsection{Multithreading to tolerate latency}
-hardware: tera, cray
-smt,
-alewife
-gpu,
+Using multithreading to tolerate memory latency is well-covered in the literature.
+Hardware implementations include the Tera MTA \cite{Tera}, Cray XMT \cite{}, Simultaneous multithreading \cite{}, MIT Alewife \cite{}, Cyclops \cite{}, and even GPUs \cite{fatahalian}.
+
+\subsection{systems?}
+RAMCloud - low latency; RPC call communication abstraction
+
+sDSM - most are page based and deal with maintaining consistency of data; built on shared memory programming models rather than something more PGAS.
+They can use prefetching, based on prediction to overlap communication and computation, but does this work for irregular apps?
+Difference from txDSM is that we don't try to provide txns over multiple nodes, only within one core's memory.
+
 software threads
@@
 Grappa also supports dynamic parallelism with asynchronous execution, but has a more general purpose programming model, where parallelism is expressed as tasks or loop iterations.
-\todo{and express locality?
-mention pgas e.g.
-Chapel language implementation?
+\TODO{and express locality?
 How much to talk about how graphlab achieves scalable performance?}

### 2012-07-21 doc/asplos13/grappa.tex
a few changes to grappa and implementation sections; add figures for system overview and memory structure

@@
 \begin{center}
-  \includegraphics[width=0.5\columnwidth]{figs/grappa-crappy}
+  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
 \begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}\TODO{should we use the canonical Grappa picture here augmented withe the software stack?}}
+  \caption{\label{fig:grappa} Grappa system overview}
 \end{minipage}
@@
 programmer expects a computation on shared data to have spatial
-locality to exploit, {\em cache} operations may be used. When there is no locality to
-exploit, {\em delegate} operations are used. Since these operations
-are expected communicate with other nodes and have high latency,
-\TODO{the runtime systems might context switch? what do we want to say here?}
+locality to exploit, {\em cache} operations may be used. When there is
+no locality to exploit, {\em delegate} operations are used. Since
+these operations are expected to communicate with other nodes and have
+high latency, cache and delegate operations interact with the
+scheduler; when issuing a long-latency request, they suspend the
+requesting task and allow the core to execute other work until their
+response arrives.
 
@@
 
-Grappa contains two key components that are not directly visible to
+Grappa contains two components that are not directly visible to
 the programmer but are key to providing good performance. Both depend
@@
 The first is a distributed dynamic load balancing mechanism that uses work stealing to
-move tasks from busy nodes to idle nodes. When a core detects that it
+move tasks from busy cores to idle cores. When a core detects that it
 has idle execution resources and no new tasks to run, it issues steal
@@
 inter-system-node/core communication requests that are dispatched to
-the \gasnet~library. Most of \checkme{our requests} are small, and commodity
-networks do not transport small packets efficiently. In order to make
-more efficient use of bandwidth, Grappa aggregates packets going to the
-same destination until it can form a larger packet that will make more
-efficient use of the network. In this way, Grappa trades per-request
-latency for application-level throughput. \TODO{Jacob, can you add more details here on how this works?}
+the \gasnet~library. In our benchmarks, we found that most of our
+packets were small (44 bytes on average for BFS). Commodity network
+interfaces do not transport small packets efficiently (see
+Section~\ref{sec:bandwidth-versus-packet-size}): the rate at which the
+network interface will accept new packets of any size is limited, so
+the only way to use a respectable fraction of the bandwidth is to send
+larger messages.
+
+Grappa's latency tolerance allows us to leverage the parallelism in
+our problem domain to build larger packets. Tasks can generate
+requests faster than the network will accept new packets, so instead
+of immediately sending every message, Grappa queues messages heading
+for the same destination. Once enough messages are queued, Grappa
+bundles all the messages into a single packet and sends it to the
+destination core. Tasks that require a response can suspend, allowing
+their core to do other work while waiting for a reply.
+
+While queuing adds latency to each individual request, it is a much
+lighter-weight operation than sending a packet and allows us to
+amoritize the cost of sending the packet across multiple tasks, while
+also making better use of available network bandwidth. In this way,
+Grappa trades per-request latency for application-level
+throughput.

### 2012-07-21 doc/asplos13/implementation.tex
a few changes to grappa and implementation sections; add figures for system overview and memory structure

@@
 
+\begin{figure}[t]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:memory-structure} Grappa memory structure}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
 \paragraph{2D global addresses} The first global address space is the {\em
@@
 a 2D address. This leaves 15 bits for network endpoint ID, which limits our
-scalability to $2^15$ endpoints.
+scalability to $2^{15}$ endpoints.
 
@@
 for our benchmarks, parallel performance can be achieved through
-combining \TODO{ cite MAMA and flat combining }. 2D addresses may
-refer to memory allocated from a single processes' heap or from a
-task's stack.
+combining \TODO{ cite MAMA and flat combining}. 2D addresses may refer
+to memory allocated from a single processes' heap or from a task's
+stack. Figure~\ref{fig:memory-structure} shows how 2D and linear
+addresses can refer to other cores' memory.
 
@@
 All global memory operations are implemented with active
-messages. Requesters construct a descriptor holding the current
-worker \TODO{reorder this paragraph after defined} ID and space for a response, issue their request active
-message to the node given by the request address, and suspend
-themselves. When the request arrives at the address' home node, the
-operation is performed and a response active message is sent. When the
-response arrives at the requesting node, the descriptor is filled in
-with the results of the operation and the requesting task is woken.
+messages. Requesters construct a descriptor holding the current task
+ID and space for a response, issue their request active message to the
+node given by the request address, and suspend themselves. When the
+request arrives at the address' home node, the operation is performed
+and a response active message is sent. When the response arrives at
+the requesting node, the descriptor is filled in with the results of
+the operation and the requesting task is woken.
 
 \paragraph{Delegates} Delegate operations are used for low-locality accesses and synchronization. Their modifications
-are always executed at the home core of their address. Each task may
-have only one outstanding delegate operation \TODO{this is not an
-    inherent limitation, it would be better to discuss in terms of
-    fence -BM}. We limit delegate
-operations to operate on objects in the 2D address space or objects that
-fit in a single block of the linear address space. No context switches
-are allowed while the data it being modified. Given these
-restrictions, we can ensure that delegate operations for the same address from multiple
-requesters are always serialized through a single core in the system,
-providing atomic semantics without using atomic operations.
+are always executed at the home core of their address, and while
+arbitrary memory operations can be delegated, we restrict the use of
+delegate operations in three ways to make them more useful for
+synchronization. First, we limit each task to one outstanding delegate
+operation to avoid the possibility of reordering in the
+network. Second, we limit delegate operations to operate on objects in
+the 2D address space or objects that fit in a single block of the
+linear address space so they can be satisfied with a single network
+request. Finally, no context switches are allowed while the data is
+being modified. Given these restrictions, we can ensure that delegate
+operations for the same address from multiple requesters are always
+serialized through a single core in the system, providing atomic
+semantics without using atomic operations.
 

### 2012-07-22 doc/asplos13/relatedwork.tex
more related work, on SDSMs, gasnet

@@
 
-dsm, pgas
-lightweight threading
-other graph solutions: 
--graph parallel, distributed
---pregel - but is synchronous and thus inefficient
---graphlab - asynchrounous, esp useful for problems that have been shown empirically to converge faster with asynchrony
-
 \subsection{Multithreading to tolerate latency}
@@
 
-\subsection{systems?}
-RAMCloud - low latency; RPC call communication abstraction
+software threads
+lightweight threading
 
-sDSM - most are page based and deal with maintaining consistency of data; built on shared memory programming models rather than something more PGAS.
-They can use prefetching, based on prediction to overlap communication and computation, but does this work for irregular apps?
-Difference from txDSM is that we don't try to provide txns over multiple nodes, only within one core's memory.
+RAMCloud takes a different approach and argues for optimizing networked systems for very low latency with the motivation of getting good performance regardless of access pattern, as well as reducing latency of recovery to reduce backup storage costs.
 
-software threads
+\subsection{Distribute Shared Memory}
+
+Grappa includes a software distributed shared memory (SDSM).
+Many traditional SDSM systems are page based [Treadmarks,...], and a lot of work deals with maintaining consistency of data efficiently and optimizing for sharing.
+Grappa's shared memory is implemented by Active Messages and there is no caching under the hood.
+Every piece of memory in the shared address space belongs to a single core, and data can be accessed incoherently at any granularity.
+Grappa provides an API for explicit incoherent caching to take advantage of locality where it exists, as well as sharing.
+
+Shared memory systems often try to reduce the cost of memory access by predicting good prefetches to overlap memory access and computation; however, this is less effective for irregular applications with fine-grained, data-dependent memory accesses.
+Grappa depends on high amounts of concurrency to tolerate latency with multithreading.
+
+SDSM systems are usually built to be programmed with flat shared memory programming models like OpenMP, rather than models that make locality explicit like PGAS.
+
+GASNet is a networking library for supporting portable implmentations of global address space applications or languages, like UPC and recently Chapel...
+Grappa currently uses GASNet for networking, but does not use gasnet's mapped shared memory segments for supporting RDMA.
+Rather, the global address space is implemented at a higher level using Active Messages.
+
+RAMCloud - RPC call communication abstraction like us because it supports more flexibility,...
+
+Difference from txDSM \cite{sdsm with txn coherence} is that we don't try to provide txns over multiple nodes, only within one core's memory.
+
+\subsection{Partitioned Global Address space}
+The goal of presenting a global view of distributed memory to the programmer is shared by the PGAS community, and is used in languages like Chapel \cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC}.
+In these programming models, access to shared structures look like normal memory references, but the programmer tries to minimize references to remote nodes.
+Unlike the usual programming goal in PGAS, since we target problems with poor locality, we design Grappa for remote references as a common case.
+Thread private data is still local, and unlike the XMT, we exploit locality where it does exist in an application (such as spatial locality in a graph edgelist), but we design the system to perform well in the midst of mostly remote accesses to large shared data structures.
+We would like to implement PGAS languages that support dynamic parallelism, like Chapel, in Grappa.
 
@@
 
-\subsection{Partitioned Global Address space}
-The goal of presenting a global view of distributed memory to the programmer is shared by the PGAS community, and is used in languages like Chapel \cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC}.
-In these programming models, access to shared structures look like normal memory references, but the programmer tries to minimize references to remote nodes.
-Unlike the usual programming goal in PGAS, since we target problems with poor locality, we design Grappa for remote references as a common case.
-Thread private data is still local, and unlike the XMT, we exploit locality where it does exist in an application (such as spatial locality in a graph edgelist), but we design the system to perform well in the midst of mostly remote accesses to large shared data structures.
-We would like to implement PGAS languages that support dynamic parallelism, like Chapel, in Grappa.
+\TODO{should we mention GreenMarl and how its first distributed try is compile down to Pregel}

### 2012-07-22 doc/asplos13/abstract.tex
Pass on paper, less related work and results

@@
 workstealing; and provides efficient synchronization and consistency across
-its single shared address space by delegating all accesses to any given global
-location to a corresponding processor core. We present a detailed description
+its single shared address space. We present a detailed description
 of the Grappa system, programming examples using the library interface, and

### 2012-07-22 doc/asplos13/implementation.tex
Pass on paper, less related work and results

@@
 
-\subsection{Memory}
-
-Grappa implements a software distributed shared memory. This memory is exposed
-through two distinct, non-overlapping address spaces; one optimized for global
-access to node-local data, and another optimized for high aggregate random
-access bandwidth to low locality data.
-
-All global addresses in Grappa are 64-bit
-values. Each byte of global memory is associated with a {\em home core}; this
-core is responsible for allocating and modifying that memory. Since today's
-commodity processors support a 48-bit virtual address space \TODO{cite?},
-Grappa uses the spare bits to distinguish which address space a global address
-belongs to, and to represent the home core of a global address.
-
-\begin{figure}[t]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:memory-structure} Grappa memory structure}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-\paragraph{2D global addresses} The first global address space is the {\em
-two-dimensional} address space. This uses a traditional PGAS addressing model,
-where each address is a tuple of a rank in the job (or global process ID) and
-an address in that process. The lower 48 bits of the address hold a virtual
-address in the process. The top bit is set to indicates that the reference is
-a 2D address. This leaves 15 bits for network endpoint ID, which limits our
-scalability to $2^{15}$ endpoints.
-
-Any node-local data can be made accessible by other nodes in the system by
-wrapping the address and node ID into a 2D global address. This address can
-then be accessed with delegate and can also be cached by other nodes. At the
-destination the address is converted into a canonical x86 address by replacing
-the upper bits with the sign-extended upper bit of the virtual address.
-
-\paragraph{Linear global addresses} The second global address space is the {\em linear} address space, so
-named because nearby addresses can be treated as adjacent even if they
-are stored on different nodes. Addresses in the linear address space
-are partitioned across the cluster in a block-cyclic fashion. Choosing
-the block size involves trading off sequential bandwidth against
-aggregate random access bandwidth. Smaller block sizes help spread
-data across all the memory controllers in the cluster, but larger
-block sizes allow the locality-optimized memory controllers to provide
-increased sequential bandwidth. The block size, which is configurable, is typically set to 64
-bytes, or the size of a single hardware cache line, in order to
-exploit spatial locality when available.
-
-All linear addresses refer to data allocated from a global heap. The
-heap metadata is stored on a single node. Currently all heap
-operations serialize through this node; while this has been sufficient
-for our benchmarks, parallel performance can be achieved through
-combining \TODO{ cite MAMA and flat combining}. 2D addresses may refer
-to memory allocated from a single processes' heap or from a task's
-stack. Figure~\ref{fig:memory-structure} shows how 2D and linear
-addresses can refer to other cores' memory.
-
-\TODO{Add a figure that depicts both forms of global addresses.}
-
-All global memory operations are implemented with active
-messages. Requesters construct a descriptor holding the current task
-ID and space for a response, issue their request active message to the
-node given by the request address, and suspend themselves. When the
-request arrives at the address' home node, the operation is performed
-and a response active message is sent. When the response arrives at
-the requesting node, the descriptor is filled in with the results of
-the operation and the requesting task is woken.
-
-\paragraph{Delegates} Delegate operations are used for low-locality accesses and synchronization. Their modifications
-are always executed at the home core of their address, and while
-arbitrary memory operations can be delegated, we restrict the use of
-delegate operations in three ways to make them more useful for
-synchronization. First, we limit each task to one outstanding delegate
-operation to avoid the possibility of reordering in the
-network. Second, we limit delegate operations to operate on objects in
-the 2D address space or objects that fit in a single block of the
-linear address space so they can be satisfied with a single network
-request. Finally, no context switches are allowed while the data is
-being modified. Given these restrictions, we can ensure that delegate
-operations for the same address from multiple requesters are always
-serialized through a single core in the system, providing atomic
-semantics without using atomic operations.
-
-\paragraph{Explicit caches} Grappa cache operations are used to exploit
-spatial or temporal locality. They allow a single task to buffer a range of
-data, read or modify it, and write the modified version back to its home node.
-All fetch and update operations of cache data are expressed explicitly by the
-programmer. Cache operations may operate on data that spans multiple blocks of
-the linear address space; the programmer provides the starting address and
-length, and the cache issues as many active messages as necessary to retrieve
-all necessary data.
-
-\comment{If anything, this is just a footnote. -M
-All current cache operations are incoherent; two tasks may
-concurrently modify the same block with their writes interfering
-unless protected by synchronization through delegate operations. While
-this has been sufficient for our benchmarks, it is possible to embed
-synchronization in the cache active messages and reduce the number of
-network round trips necessary.}
-
-\subsection{Tasking}
+\subsection{Tasks}
 
@@
 
-Context switching in Grappa is non-preemptive. Any Grappa routine that
-does a potentially-long-latency operation calls a context-switch
-routine that gives control of the core back to the scheduler. Users
-may also explicit yield control to the scheduler.
-
-As with other cooperative multithreading systems, we treat context
-switches as function calls, saving and restoring only the callee-saved
-state as specified in the x86-64 ABI \TODO{cite}. This involves saving
-six general-purpose 64-bit registers and the stack pointer, as well as
-the 16-bit x87 floating point control word and the SSE context/status
-register. Thus, the minimum amount of state a cooperative context
-switch routine must save according to the ABI is 62 bytes.
+Context switching in Grappa is non-preemptive. As with other cooperative
+multithreading systems, we treat context switches as function calls,
+saving and restoring only the callee-saved state as specified in the
+x86-64 ABI \cite{someone}. This involves saving six general-purpose
+64-bit registers and the stack pointer, as well as the 16-bit x87
+floating point control word and the SSE context/status register. Thus,
+the minimum amount of state a cooperative context switch routine must
+save according to the ABI is 62 bytes.
 
@@
 Why not? liveness is a common compiler analysis -BM
+Because it has to be live across the entire application.  All real applications do some form of floating point, however minor... -M
 
@@
   waiting to be matched with workers. It is a local partition of a shared
-  task pool; this is described in the next section.
+  task pool.
 \end{description}
 
-Whenever a task yields or suspends, the scheduler makes a decision
-about what to do next. First, it determines if
-the communication resources should be serviced. This is done on a
-periodic basis. Second, it determines if any workers with running
-tasks are ready to execute; if so, one is scheduled. Finally, if there
-are no workers ready to run, but there are tasks waiting to be matched
-with workers, an idle worker is woken (or a new worker is spawned),
-matched with a task, and scheduled.
+Whenever a task yields or suspends, the scheduler makes a decision about
+what to do next. First, it determines if the communication resources
+should be serviced. This is done on a periodic basis. Second, it
+determines if any workers with running tasks are ready to execute; if
+so, one is scheduled. Finally, if there are no workers ready to run, but
+there are tasks waiting to be matched with workers, an idle worker is
+woken (or a new worker is spawned), matched with a task, and scheduled.
+The scheduler ensures that communication maintenance tasks are executed periodically.
 
-The scheduler uses its core's timestamp counter to track time. The
-timestamp counter is read once per scheduling decision. Timestamp
-values are not shared between cores. We use timestamp values only to
-ensure periodic execution of communication maintenance tasks; being
-wrong will only change the rate of these events with a potential
-performance cost, but does not cause a correctness violation. Nevertheless, we
-assume Grappa will run on machines with modern timestamp counter
-implementations that are monotonic and frequency-compensated.
+\subsection{Memory}
+
+\TODO{Is there an off by 1 error in here?
+Need to read the AMD64 docs carefully to see if we can sign extend the top bit of the 48 bits of real virtual address space, or if we need to maintain bit 49.
+-M}
+
+Grappa implements a software distributed shared memory. All global
+addresses in Grappa are 64-bit values. Each byte of global memory is
+associated with a {\em home core}; this core is responsible for
+allocating and modifying that memory. Since today's commodity processors
+support a 48-bit virtual address space \cite{AMD64}, Grappa uses the
+spare bits to distinguish which address space a global address belongs
+to, and to represent the home core of a global address.
+
+Global memory in Grappa is exposed through two distinct, non-overlapping
+address spaces; \textbf{2D global addresses}, optimized for global
+access to node-local data, and \textbf{linear global addresses},
+optimized for high aggregate random access bandwidth to low locality
+data.
+
+\begin{figure}[t]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:memory-structure} Grappa memory structure}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
+\paragraph{2D global addresses} The first global address space is the {\em
+two-dimensional} address space. This uses a traditional PGAS addressing model,
+where each address is a tuple of a rank in the job (or global process ID) and
+an address in that process. The lower 48 bits of the address hold a virtual
+address in the process. The top bit is set to indicates that the reference is
+a 2D address. This leaves 15 bits for network endpoint ID, which limits our
+scalability to $2^{15}$ endpoints.
+
+Any node-local data can be made accessible by other nodes in the system
+by wrapping the address and node ID into a 2D global address. This
+address can then be accessed with a delegate and can also be cached by
+other nodes. At the destination the address is converted into a
+canonical x86 address by replacing the upper bits with the sign-extended
+upper bit of the virtual address. 2D addresses may refer to memory
+allocated from a single processes' heap or from a task's stack.
+Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can
+refer to other cores' memory.
+
+\paragraph{Linear global addresses} The second global address space is
+the {\em linear} address space, so named because nearby addresses can be
+treated as adjacent even if they are stored on different nodes.
+Addresses in the linear address space are partitioned across the cluster
+in a block-cyclic fashion. Choosing the block size involves trading off
+sequential bandwidth against aggregate random access bandwidth. Smaller
+block sizes help spread data across all the memory controllers in the
+cluster, but larger block sizes allow the locality-optimized memory
+controllers to provide increased sequential bandwidth. The block size,
+which is configurable, is typically set to 64 bytes, or the size of a
+single hardware cache line, in order to exploit spatial locality when
+available.
+
+All linear addresses refer to data allocated from a global heap. The
+heap metadata is stored on a single node. Currently all heap operations
+serialize through this node; while this has been sufficient for our
+benchmarks, in the future Grappa will provide parallel performance
+through combining~\cite{MAMA,flatcombining}.
+
+\TODO{Add a figure that depicts both forms of global addresses.}
+
+All global memory operations are implemented with active
+messages. Requesters construct a descriptor holding the current task
+ID and space for a response, issue their request active message to the
+node given by the request address, and suspend themselves. When the
+request arrives at the address' home node, the operation is performed
+and a response active message is sent. When the response arrives at
+the requesting node, the descriptor is filled in with the results of
+the operation and the requesting task is placed back on the ready queue.
+
+\paragraph{Delegates} Delegate operations are used for low-locality
+accesses and synchronization. Their modifications are always executed at
+the home core of their address, and while arbitrary memory operations
+can be delegated, we restrict the use of delegate operations in three
+ways to make them more useful for synchronization. First, we limit each
+task to one outstanding delegate operation to avoid the possibility of
+reordering in the network. Second, we limit delegate operations to
+operate on objects in the 2D address space or objects that fit in a
+single block of the linear address space so they can be satisfied with a
+single network request. Finally, no context switches are allowed while
+the data is being modified. Given these restrictions, we can ensure that
+delegate operations for the same address from multiple requesters are
+always serialized through a single core in the system, providing atomic
+semantics without using atomic operations.
+
+\paragraph{Explicit caches} Grappa cache operations are used to exploit
+spatial or temporal locality. They allow a single task to buffer a range
+of data, read or modify it, and write the modified version back to its
+home node(s). All fetch and update operations of cache data are
+expressed explicitly by the programmer. Cache operations may operate on
+data that spans multiple blocks of the linear address space; the
+programmer provides the starting address and length, and the cache
+issues as many active messages as necessary to retrieve all necessary
+data.
+
+\comment{If anything, this is just a footnote. -M
+All current cache operations are incoherent; two tasks may
+concurrently modify the same block with their writes interfering
+unless protected by synchronization through delegate operations. While
+this has been sufficient for our benchmarks, it is possible to embed
+synchronization in the cache active messages and reduce the number of
+network round trips necessary.}
+
+\subsection{Load Balancing and Network Aggregation}
 
 \subsubsection{Work stealing} \label{subsec:implementation-worksteal}
-When a \checkme{grappa-process} runs
-out of work, it becomes a ``thief'' and asks another other
-\checkme{grappa-process} (``victim'') for work. If the victim has tasks
-to spare in its public queue, they will be transferred to the thief's public queue.
 
-Most study of work stealing assumes that a processor has one worker
-thread and so it only steals when utilization would go to zero. 
-Grappa, on the other hand, relies on having many worker threads per core. The
+When a \checkme{grappa-process} runs out of work, it becomes a ``thief''
+and asks another other \checkme{grappa-process} (``victim'') for work.
+If the victim has tasks to spare in its public queue, they will be
+transferred to the thief's public queue.
+
+Most prior work on work stealing assumes that a processor has one worker
+thread and so it only steals when utilization would go to zero. Grappa,
+on the other hand, relies on having many worker threads per core. The
 core would be fully-utilized if there is always something on the ready
 queue. Even if there are many active tasks, if they are all suspended
-for long-latency network requests, then the core is underutilized.
-So, there is a choice: should the core use some of this idle time to
-perform work stealing or should it just wait as it is possible that local tasks will
-create more work. We choose to have a core initiate a steal when all
-of the following conditions hold: no workers are ready to run, 
-the unstarted task queues are empty, and there are no outstanding steal requests. 
-Having only one outstanding steal request throttles steals to
+for long-latency network requests, then the core is underutilized. So,
+there is a choice: should the core use some of this idle time to perform
+work stealing or should it just wait as it is possible that local tasks
+will create more work. We choose to have a core initiate a steal when
+all of the following conditions hold: no workers are ready to run, the
+unstarted task queues are empty, and there are no outstanding steal
+requests. Having only one outstanding steal request throttles steals to
 prevent flooding the network, but other scalable quieting mechanisms
-would be possible, such as a voting tree\cite{scalableWorkStealingOrCilk98}
-or lifelines \cite{lifelines}. Termination detection is not
-built into the Grappa task scheduler, rather, it is considered a programming error for
-the program to exit without syncing all tasks.
+would be possible, such as a voting
+tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
+Termination detection is not built into the Grappa task scheduler,
+rather, it is considered a programming error for the program to exit
+without syncing all tasks.
 
-\TODO{where to put the following related}
+\TODO{Someone check this paragraph to make sure it is true. :-)  -Mark}
+
+When a grappa node decides to issue a steal request it chooses the
+victim node at random.  It also requests a block of tasks at once.  If
+this block exceeds the amount of stealable tasks on the victim node,
+half of the stealable tasks are stolen instead.  While we did study the
+number of tasks to steal at once, we found performance was maximized
+with \checkme{128} tasks with little sensitivity around this number,
+hence we omit results for this from Section~\ref{sec:results} and simply
+use \checkme{128} as the parameter for all experiments.
+
+\comment{
+where to put the following related? -BM
+
 Fork join computations: implement with futures or continuation passing
@@
 choices we've made
--random victims
--steal from FIFO side of queue
--steal when at the moment readyQ empty and task queues empty
--steal min(chunk\_size, half)
+done -random victims
+hmm..  -steal from FIFO side of queue
+hmmm... -steal when at the moment readyQ empty and task queues empty
+done -steal min(chunk\_size, half)
 
-Recent work has explored scalable work stealing in terms of efficient
+omit - Recent work has explored scalable work stealing in terms of efficient
 termination detection. 
+}
 
-\subsection{Communication}
+\subsubsection{Network Aggregation}
 
@@
 At the upper layer, Grappa implements asynchronous active messages
-\TODO{cite vonEicken92}. Each message consists of a function pointer, an
+\cite{vonEicken92}. Each message consists of a function pointer, an
 optional argument payload, and an optional data payload. When a task
@@
 Grappa's lower networking layer aggregates the upper layer's messages
-to improve performance. Commodity networks including Infiniband
-achieves their peak injection bandwidth \emph{only} when the packet
+to improve performance. Commodity networks including infiniband
+achieves their peak bisection bandwidth \emph{only} when the packet
 sizes are relatively large---on the order of multiple kilobytes. The
@@
 with handling each packet (in terms of bytes that form the actual
-Infiniband packet, processing time at the card and processing on the
+packet, processing time at the card and processing on the
 CPU within the driver stack). Our measurements confirm manufacturers
@@
 In our experiments the vast majority of requests were smaller than
-\checkme{255}
-bytes \TODO{more precise? median?}, far too small to make efficient
-use of the network. To make the best use of the network, we must
-convert our small messages into large ones.
-\TODO{reference to aggregation evaluation later in the paper}
-
-Grappa leverages its lightweight context switching to do this
-conversion. When a task sends a message, the message is not
-immediately sent. Instead, it is stored in a queue associated with its
-destination, to be sent later. If the task requires a response, it
-suspends itself and the scheduler finds other work to run while the
-task is waiting. When the reply arrives, the suspended task is woken.
+\checkme{255} bytes \TODO{more precise? median?}, far too small to make
+efficient use of the network. To make the best use of the network, we
+must convert our small messages into large ones
+(Section~\ref{sec:results-aggregation}).  When a task sends a message,
+it is not immediately sent, but rather placed in a queue specific to the
+destination.
 
@@
 sent. First, each queue has a message size threshold, chosen to give
-reasonable network performance. If the size in bytes of a queue is
-above the threshold, the contents of the queue are sent
-immediately. Second, each queue has a wait time threshold. If the
-oldest message in a queue has been waiting longer than this threshold,
-the contents of the queue are sent immediately, even if the queue size
-is lower than the message size threshold.  Third, queues may be
-explicitly flushed in situations where the programmer wants to
-minimize the latency of a message at the cost of bandwidth
-utilization.
+reasonable network performance (\checkme{2048 bytes}). If the size in
+bytes of a queue is above the threshold, the contents of the queue are
+sent immediately. Second, each queue has a wait time threshold
+(\checkme{200ms}). If the oldest message in a queue has been waiting
+longer than this threshold, the contents of the queue are sent
+immediately, even if the queue size is lower than the message size
+threshold.  Third, queues may be explicitly flushed in situations where
+the programmer wants to minimize the latency of a message at the cost of
+bandwidth utilization.
 
@@
 Underneath the aggregation layer, Grappa uses the \gasnet~communication
-library \TODO{cite GASNet} to actually move data. All interprocess
+library~\cite{gasnet} to actually move data. All interprocess
 communication, whether on or off a cluster node, is handled by the
 \gasnet~library. \gasnet~is able to take advantage of many communication
-mechanisms, including Ethernet and Infiniband between nodes, as well as
+mechanisms, including ethernet and infiniband between nodes, as well as
 shared memory within a node.
 
-Some networks provide access to a remote machine's memory
-directly. This would seem to be a good fit for a programming model
-focused on global shared memory, but in fact we do not use it. RDMA
-operations are subject to the same message rate limitations as all
-other messages on these cards, and thus using raw RDMA operations for
-our small messages would make inefficient use of bandwidth. Instead,
-we implement remote memory operations with active messages. A byproduct of this design decision is that Grappa is not limited
-to RDMA-capable networks.
+Some networks provide access to a remote machine's memory directly. This
+would seem to be a good fit for a programming model focused on global
+shared memory, but in fact we do not use it. In our experiments, we
+found that RDMA operations are subject to the same message rate
+limitations as all other messages on these cards, and thus using raw
+RDMA operations for our small messages would make inefficient use of
+bandwidth. Instead, we implement remote memory operations with active
+messages. A byproduct of this design decision is that Grappa is not
+limited to RDMA-capable networks.

### 2012-07-22 doc/asplos13/grappa.tex

@@
 In this section we describe the Grappa runtime system
-(Figure~\ref{fig:grappa}). 
-Grappa provides three key capabilities. The first is a distributed
-shared \textbf{global memory} with high aggregate random access
-bandwidth. The second is a \textbf{tasking} system with lightweight
-multithreading and automatic load balancing through work
+(Figure~\ref{fig:grappa}). Grappa provides three key capabilities. The
+first is a distributed shared \textbf{global memory} with high aggregate
+random access bandwidth. The second is a \textbf{tasking} system with
+lightweight multithreading and automatic load balancing through work
 stealing. Finally, under the hood, a \textbf{communication} layer uses
 network packet scheduling and aggregation to achieve high performance
-while remaining largely invisible to applications. We now discuss the programmer's view of
-Grappa's main capabilities, including a subset of the Grappa API. In
-Section~\ref{implementation}, we discuss the details of their
-implementation.
+while remaining largely invisible to applications. We now discuss the
+programmer's view of Grappa's main capabilities, including a subset of
+the Grappa API. In Section~\ref{implementation}, we discuss the details
+of their implementation.
 
@@
 
-What makes Grappa's tasks special is that when they perform
-long-latency operations they can give up control of their core,
-allowing the core to remain busy while waiting for the operation to
-complete. This most often done implicitly inside a call into Grappa's
-API, but it can also be done explicitly by the programmer using the
-calls shown in Figure~\ref{fig:scheduling}. \TODO{cite UPC split phase
-    reads and writes, but include that we provide a way to overlap
-    computation}
+Grappa's tasks give up control of their core when when they perform
+long-latency operations, allowing the processor to remain busy while
+waiting for the operation to complete. This is most often done
+implicitly inside a call into Grappa's API, but it can also be done
+explicitly by the programmer using the calls shown in
+Figure~\ref{fig:scheduling}. \TODO{cite UPC split phase reads and
+writes, but include that we provide a way to overlap computation}
 
-All Grappa scheduling is done in userspace, and minimal state is
-saved, in order to minimize the context switch time. In our
-experiments, we saw context switch times as low as \TODO{40ns}.
-This lightweight context switching is Grappa's key enabling
-feature. It gives us the ability to tolerate latency. Given that
-ability, we are able to trade latency for throughput: by {\em
-  increasing} latency in key components in the system we are able to
-increase our aggregate random access bandwidth, our synchronization
-bandwidth, and our ability to tolerate load.
+All Grappa scheduling is done in userspace, and minimal state is saved,
+in order to minimize the context switch time. In our experiments, we saw
+context switch times as low as \checkme{40ns}. This lightweight context
+switching is Grappa's key enabling feature. It gives us the ability to
+tolerate latency. Given that ability, we are able to \textbf{trade
+latency for throughput}: by {\em increasing} latency in key components
+in the system we are able to increase our aggregate random access
+bandwidth, our synchronization bandwidth, and our ability to tolerate
+load.
 
@@
 parallel, the work may be wrapped up in a function and queued with its
-arguments for later execution using a \texttt{spawn}. Sometimes the
-programmer may want to spawn a task on a specific core in the system,
-or at the home core of a particular memory location. Grappa provides a
-\texttt{spawn\_on} call for this purpose; the task is immediately
-enqueued on the specified core and is not available to other cores.
+arguments for later execution using a \texttt{spawn}. If the
+programmer wishes to spawn a task on a specific core in the system,
+or at the home core of a particular memory location, Grappa provides a
+\texttt{spawn\_on} call for this purpose.
 
-The next method for expressing parallelism is a parallel for loop, where the
-number of iterations must be known at loop entry. The programmer specifies a
-function pointer along with start and end indices and an optional threshold to
-control parallel overhead. Grappa does {\em recursive decomposition} of
-iterations, as in the implementation of Cilk's cilk\_for construct \TODO{cite
-cilk\_for implementation: could only find slides page6 of
-http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-Cilk++.pdf}.
-It generates a logarithmically-deep tree of tasks, stopping to execute the
-loop body when the number of iterations is below the threshold (``grainsize''
-in Cilk). \TODO{we may want to make our classification purer: the for loop is
-not Grappa's core, and is built atop tasks as a useful construct; in some
-sense the only way to express parallelism in Grappa is tasks and perhaps
-call\_on. -BM}
-
-Finally, a programmer may want to run a small piece of code on a
-particular core in the system without waiting for execution resources
-to be available. Grappa provides the \texttt{call\_on} call for this
-purpose.
+The next method for expressing parallelism is a parallel for loop, where
+the number of iterations must be known at loop entry. The programmer
+specifies a function pointer along with start and end indices and an
+optional threshold to control parallel overhead. Grappa does {\em
+recursive decomposition} of iterations, similar to Cilk's cilk\_for
+construct~\cite {cilkforimplementation} \comment{could only find slides
+page6 of
+http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-
+Cilk++.pdf    ---- Good enough.  Cite it.  -Mark}.  It generates a
+logarithmically-deep tree of tasks, stopping to execute the loop body
+when the number of iterations is below the required threshold. Finally,
+a programmer may want to run a small piece of code on a particular core
+in the system without waiting for execution resources to be available.
+Grappa provides the \texttt{call\_on} call for this purpose.
 
@@
 
-\paragraph{Global memory access}
-There are two approaches to {\emph accessing} global memory. When the
-programmer expects a computation on shared data to have spatial
-locality to exploit, {\em cache} operations may be used. When there is
-no locality to exploit, {\em delegate} operations are used. Since
-these operations are expected to communicate with other nodes and have
-high latency, cache and delegate operations interact with the
-scheduler; when issuing a long-latency request, they suspend the
-requesting task and allow the core to execute other work until their
-response arrives.
+\paragraph{Global memory access} There are two approaches to {\emph
+accessing} global memory. When the programmer expects a computation on
+shared data to have spatial locality to exploit, {\em cache} operations
+may be used. When there is no locality to exploit, {\em delegate}
+operations are used. Since these operations are expected to communicate
+with other nodes and have high latency, cache and delegate operations
+interact with the scheduler; when issuing a long-latency request, they
+suspend the requesting task and allow the core to execute other work
+until their response arrives.
 
-\paragraph{Explicit caching}
-With Grappa's explicit caching support, applications can instruct Grappa to fetch a global
-pointer of any length and return a local pointer to a cached copy of
-the global memory. Under the hood, Grappa performs the mechanics of
-gathering chunks of data from multiple system nodes and presenting a
-conventional appearing linear block of memory as a pointer into a
-cache. Grappa cache operations have the usual read-only and read-write
-variants, along with a write-only variant used to initialize data
-structures. Cache operations exploit spatial locality by reducing the number of small network
-messages accessing contiguous data. Languages for distributed
-shared memory systems have done optimizations to achieve a similar goal. The UPC compiler
-coalesces struct and array accesses into 
-remote get/put \cite{Chen:2005}, or Fortran D compiler's message
-vectorization that hoists small messages out of a loop
-\cite{FortranD:1992}. Caching in Grappa additionally provides a
-mechanism for exploiting temporal locality by operating on the data locally.
+\paragraph{Explicit caching} With Grappa's explicit caching support,
+applications can instruct Grappa to fetch a global pointer of any length
+and return a local pointer to a cached copy of the global memory. Under
+the hood, Grappa performs the mechanics of gathering chunks of data from
+multiple system nodes and presenting a conventional appearing linear
+block of memory as a pointer into a cache. Grappa cache operations have
+the usual read-only and read-write variants, along with a write-only
+variant used to initialize data structures. Cache operations exploit
+spatial locality by reducing the number of small network messages
+accessing contiguous data. Languages for distributed shared memory
+systems have done optimizations to achieve a similar goal. For example,
+the UPC compiler coalesces struct and array accesses into remote get/put
+\cite{Chen:2005}, and Fortran D compiler's message vectorization hoists
+small messages out of a loop \cite{FortranD:1992}. Caching in Grappa
+additionally provides a mechanism for exploiting temporal locality by
+operating on the data locally.
 
@@
 global memory to the memory system itself (e.g.,
-\emph{fetch-and-add}). \TODO{should we bring up delegation in the intro?}
+\emph{fetch-and-add}). \comment{should we bring up delegation in the intro?}
 
-Delegate operations, proposed in \cite{Nelson:hotpar11}, are also the primary synchronization method in
-Grappa. Each piece of global memory is paired with a single core in
-the system; all accesses to that memory are done by that core. This
-allows delegate operations to be performed atomically on the core
-simply by making sure no other code interleaves; since context
-switches in Grappa are cooperative, this is trivial. Thus, delegate
-operations are able to provide atomic semantics to memory owned by one core without using atomic
-operations. Using delegation to implement isolation for any subset of
-memory has been explored in \cite{delegated:oopsla11}.
+Delegate operations, proposed in \cite{Nelson:hotpar11}, are also the
+primary synchronization method in Grappa. Each piece of global memory is
+paired with a single core in the system; all accesses to that actual
+memory location are done by that core. This allows delegate operations
+to be performed atomically on the core simply by making sure no other
+code interleaves their operation; and, since context switches in Grappa
+are cooperative, this is trivial. Thus, delegate operations are able to
+provide atomic semantics to memory owned by one core without using
+atomic operations. Using delegation to implement isolation for any
+subset of memory has been explored in \cite{delegated:oopsla11}.
 
@@
 
-The first is a distributed dynamic load balancing mechanism that uses work stealing to
-move tasks from busy cores to idle cores. When a core detects that it
-has idle execution resources and no new tasks to run, it issues steal
-requests over the network. The core continues to execute remaining
-active work while the steal requests are in flight. When a steal
-request succeeds, it returns with more tasks that can be matched with
-execution resources and scheduled.
+The first is a distributed dynamic load balancing mechanism that uses
+work stealing to move tasks from busy cores to idle cores. When a core
+detects that it has idle execution resources and no new tasks to run, it
+issues steal requests over the network. The core continues to execute
+remaining active work while the steal requests are in flight. When a
+steal request succeeds, it returns with more tasks that can be matched
+with execution resources and scheduled.
 

### 2012-07-22 doc/asplos13/intro.tex

@@
 Yet, for commodity processors' caches to be effective, computations must exploit spatial locality.
-Furthermore, the addresses of references are often determined in the task only just before their values are used.
-Unfortunately, commodity prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed.
-Because commodity processors' pipeline utilization relies on spatial and temporal locality -- or, at least the ability to predict in advance what is needed -- they stall often when executing irregular applications.
+Unfortunately, commodity prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed or the accesses follow a predictable pattern, neither of which occurs in irregular applications.
+As a consequence, commodity microprocessors stall often when executing irregular applications.
 
@@
 Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.
-Injection rate into the networks is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.
-While some irregular applications can be restructured to better exploit locality, aggregate requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.
+Injection rate into the network is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.
+While some irregular applications can be restructured to better exploit locality, aggregating requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.
 
-An idealized alternative to demanding this expertise of programmers was provided by the fully custom Tera MTA-2~\cite{mta-2} system.
+An alternative to demanding programmers manage locality in their applications, was provided by the fully custom Tera MTA-2~\cite{mta-2} system.
 The MTA was well-suited to extremely irregular applications: it had a large distributed shared memory with no caches, relying on multithreaded processors to tolerate memory latency.
@@
 This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating with inexpensive mass market hardware components the approach taken by Tera.
-We exploit the increased aggregate instruction rate per socket relative to chip bandwidth.
-That is, where locality friendly computations make use of instructions to get work done in cache, we use what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.
+We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.
 Ultimately, the opportunity is to cover the spectrum of irregular to regular computation:  where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.

### 2012-07-22 doc/asplos13/method.tex

@@
 Individual nodes of this cluster contain the hardware depicted in Table~\ref{table:grappanode}.
-Nodes are interconnected with both 10G Ethernet and Infiniband.
-For all but startup and configuration, they are configured to communicate over the Infiniband network for these experiments.
+Nodes are interconnected with both 10G ethernet and infiniband.
+For all but startup and configuration, they are configured to communicate over the infiniband network for these experiments.
 
@@
 Data is collected on real systems, which means minor variations in runtime exist from run to run.
-The average of multiple runs are used, and where appropriate, confidence in the result is reported (error bars, standard deviation, etc).
+The average of multiple runs are used, and where appropriate, confidence in the result is reported.
 
-One question that must be answered when making a comparison between
-the Cray XMT and a typical HPC cluster is what is a fair comparison --
-these systems are quite different.  Three options immediately come to
-mind: equal number of processing cores, equal number of network
-interfaces, and equal dollars.  \checkme{We discounted the last option fairly
-quickly, because it isn't a lasting data-point and, as will be evident
-later, would be hugely unfair to the XMT.}  The first option, cores,
-has some merit, but Grappa is designed for applications that have no
-locality in their computation.  This means almost all of their memory
-accesses are remote.  Hence, the factor that limits their performance
-is not processing, but communicating.  Hence, we have chosen the
-middle option, network interfaces as the way to normalize across the
-XMT and our cluster.  Each processor in the XMT system has its own
-network interface to access shared memory.  In the HPC cluster we use,
-each processor (which contains up to 32 cores), has a single Infiniband interface.  Hence, for our results we scale up XMT processors one for one with full system nodes.
+One question that must be answered when making a comparison between the
+Cray XMT and a typical HPC cluster is what is a fair comparison -- these
+systems are quite different.  Three options immediately come to mind:
+equal number of processing cores, equal number of network interfaces,
+and equal dollars.  We discounted the last option fairly quickly,
+because it isn't a lasting data-point -- the cost of hardware shifts
+over time, skewing the interpretation of results.  The first option,
+cores, has some merit, but Grappa is designed for applications that have
+no locality in their computation.  This means almost all of their memory
+accesses are remote.  The factor that limits their performance is not
+processing, but communicating.  Hence, we have chosen the middle option,
+network interfaces as the way to normalize across the XMT and our
+cluster.  Each processor in the XMT system has its own network interface
+to access shared memory.  In the HPC cluster we use, each processor
+(which contains up to 32 cores, although we only use \checkme{6} in our
+experiments), has a single infiniband interface.  Hence, for our results
+we scale up XMT processors one for one with full system nodes.
 
 \subsection{Systems}
-For measurements, we run Grappa on a 144-node cluster of AMD
-Interlagos processors. Nodes have 32-core (every pair share a floating-point
-unit) 2.1-GHz processors, 64GB of memory, and \TODO{xxx 40Gb network card}. \TODO{cluster
-    uses xxx Infiniband switch}. We configure the nodes to have 32 1-GB
-hugepages to minimize TLB misses for the random access patterns we
-expect from irregular applications. The results we present are for
-this machine, but we have also run Grappa on our own 12-node Intel
-Xeon Westmere cluster.
+
+For measurements, we run Grappa on a 144-node cluster of AMD Interlagos
+processors. Nodes have 32-core (every pair share a floating-point unit)
+2.1-GHz processors, 64GB of memory, and \checkme{a 40Gb Super-terrific
+Happy-Lucky} infiniband network card.   The cluster uses a
+\checkme{McNugget} infiniband switch. We configure the nodes to have 32
+1-GB hugepages to minimize TLB misses for the random access patterns we
+expect from irregular applications.\comment{The results we present are for this
+machine, but we have also run Grappa on our own 12-node Intel Xeon
+Westmere cluster, which performs similarly.}
 
@@
 \subsection{Applications}
-\TODO{
+
+We have used three benchmarks to explore the performance of Grappa:
+
+\comment{
     Traversal of an unbalanced, unpartitioned tree captures the foundations for implementing most forms of irregular parallelism. 
@@
 
-\paragraph{Unbalanced tree search in-memory (UTS-Mem)}
-Unbalanced Tree Search (UTS) is a benchmark
-for evaluating the programmability and performance of systems for
-parallel applications that require dynamic load balancing
-\cite{Olivier:uts2006}. It involves traversing an unbalanced implicit
-tree: at each vertex, its number of children is sampled from some
-probability distribution, and this number of new nodes are added to a
-work queue to be visited. While this benchmark captures irregular,
-dynamic parallelism, we want to evaluate performance on algorithms
-with irregular memory access patterns.
-
-We augment UTS by using the existing traversal code to
-create a large tree in memory, and then we traverse the in-memory
-tree. We call this benchmark UTS-Mem, and the timed portion is this
-traversal of the in-memory tree. This in-memory traversal is
-unknowledgeable of the tree structure beforehand.
+\paragraph{Unbalanced tree search in-memory (UTS-Mem)} Unbalanced Tree
+Search (UTS) is a benchmark for evaluating the programmability and
+performance of systems for parallel applications that require dynamic
+load balancing \cite{Olivier:uts2006}. It involves traversing an
+unbalanced implicit tree: at each vertex, its number of children is
+sampled from some probability distribution, and this number of new nodes
+are added to a work queue to be visited. While this benchmark captures
+irregular, dynamic \emph{computation}, we actually want to evaluate
+performance of algorithms with irregular \emph{memory} access patterns. 
+Thus we augment UTS by using the existing traversal code to create a
+large tree in memory, and then we traverse the in-memory tree. We call
+this benchmark UTS-Mem, and the timed portion is this traversal of the
+in-memory tree. This in-memory traversal has no knowledgeable of the
+tree structure beforehand.
 
@@
 asynchronous parallel for loop over a visited vertex's children 
-list. A larger threshold can be set to take advantage of this 
-locality by processing more child pointers sequentially in a single task.
+list.  \comment{WTF? -M
+A larger threshold can be set to take advantage of this 
+locality by processing more child pointers sequentially in a single task.}
 
@@
 An important measure of the importance of particular vertices in a network is betweenness centrality (BC) \cite{freeman1979centrality}.
-By this measure, the "importance" of each vertex is computed by finding the fraction of shortest paths that pass through it, which can optionally be approximated by only computing shortest paths using a subset of the vertices as starting points.
+By this measure, the ``importance'' of each vertex is computed by finding the fraction of shortest paths that pass through it, which can optionally be approximated by only computing shortest paths using a subset of the vertices as starting points.
 BC can be useful for understanding which vertices may have the greatest impact, so in social networks this could be the primary person linking two communities.

### 2012-07-22 doc/asplos13/evaluation.tex
the beginnings of an evaluation

@@
 
-\subsection{Unbalanced tree search}
+\subsection{Performance}
+
+\subsubsection{Unbalanced tree search}
 
@@
 
+\subsubsection{BFS}
+
+\begin{figure}[tH]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/bfs_compare}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:bfs-performance} BFS performance}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:bfs-performance} shows Grappa's performance on BFS
+compared to that of the XMT\footnote{We do not know the reason for the
+  XMT's performance dip at 112 nodes, but it is repeatable}. There are
+two things to observe. First, At 128 XMT processors/cluster nodes,
+the XMT is  4.17 times faster than Grappa. This is likely due to the
+XMT's faster network: the XMT is able to issue 100 million network
+requests per second per node; at the 128 nodes datapoint, Grappa is
+aggregating 25 million network requests per second, or 4 times slower.
+Second, Grappa is scaling as we add processors.
+
+\subsubsection{Centrality}
+
+\begin{figure}[tH]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/centrality-performance}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:centrality-performance} Centrality performance}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:centrality-performance} shows Grappa's performance on
+Betweenness Centrality compared to that of the XMT. At 128 XMT
+processors/cluster nodes, the XMT is 1.75 times faster than Grappa.
+
+\subsection{Benefit of aggregator}
+
+\begin{figure}[htbH]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/bfs_aggregator_value}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:bfs-aggregator} BFS performance with and without aggregator enabled \TODO{This graph sucks; rerun on actual data point and include in performance comparison}}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/aggregator_ping}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:aggregator-ping} Bandwidth versus message size
+    unidirectional ping test for Grappa with aggregation, Grappa with
+    raw GASNet messages, and MPI. Aggregation provides an 11x
+    bandwidth benefit at our common operating point.}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
+To evaluate the benefit of the aggregator, we ran two
+experiments. First, we ran a simple unidirectionaly ping test to see
+the maximum benefit the aggregator can provide in terms of improved
+network efficiency. Second, we ran BFS with the aggregator disabled in
+order to measure its benefit on an application.
+
+To implement the ping test, we wrote a simple Grappa application where
+the cores of one node send messages as fast as possible to the cores
+of another node. We vary the size of the payload up the maximum
+payload size supported by the aggregator (nearly 4KB). Each core has a
+single task sending to a single destination, so this is a best case
+scenario for the aggregator. To see the benefit of the aggregator, we
+added a bypass that lets us send messages directly through GASNet. We
+also compare against the OSU \texttt{osu\_mbw\_mr} benchmark
+\TODO{cite?}  compiled against OpenMPI 1.5.3 \TODO{cite?}; this
+benchmark has the same pattern of communication but doesn't have the
+overhead of Grappa's context switching.
+
+The results are shown in Figure~\ref{fig:aggregator-ping}. There are
+two key observations.
+
+First, small message performance against the existing libraries is
+terrible. The MPI application test shows us that peak per node
+bandwidth supported by our Infiniband card is 3.4GB/s. This is
+achievable only with large messages; we must send 16KB packets to get
+within 5 percent of peak bandwidth. But in our benchmarks, we saw
+average message between 32 and 64 bytes. At 32 bytes, the MPI test is
+using less than 7 percent of its peak bandwidth. Grappa sending
+messages directly through GASNet uses less than 3 percent of the peak
+bandwidth.
+
+Second, aggregation has the potential to improve this situation by an
+order of magnitude. With aggregation, Grappa is able to send 32-byte
+messags over 12 times faster than using GASNet directly. This is a
+more respectable 32 percent of peak bandwidth. Due to expedient design
+decisions, Grappa's aggregator limits its aggregation to 4KB; this
+limits its peak achievable bandwidth to to 75 percent of the actual
+peak.
+
+But this comparison is the best possible case for the aggregator. In
+order to verify that the aggregator still has value on actual
+applications at scale, we ran a small (scale 20 \TODO{run larger}) BFS
+with the aggregator disabled, with both 32 and 96
+nodes. Figure~\ref{fig:bfs-aggregator} shows the results. With 96
+nodes, the aggregator improves our application performance by 4 times.
+
 \subsection{Sensitivity}
 
-\subsection{Latency tolerance parameters}
+\paragraph{Aggreator timeout}
 
-\paragraph{Benefit of aggregator}
-what happens with aggregator on and off
+\begin{figure}[htb]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/bfs_sweep_flushticks}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:bfs-sweep-flushticks} Sensitivity to aggregation delay}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
 
-\paragraph{Latency}
-aggregation latency and delegate latency
+One of the key parameters of the aggregator is the message
+timeout. All messages that are queued must eventually be sent in order
+to ensure progress. In the best case, we are able to aggregate enough
+messages to fill an aggregation buffer and cause it to be sent, but as
+we scale up, the average rate of messages heading to a common
+destination decreases, and this gets harder. To bound the problem, the
+aggregator includes a timeout. Any packet waiting this long must be sent.
 
-\paragraph{Number of workers}
-We need enough workers to cover the network+aggregation latency, but more just slow things down.
+Figure~\ref{fig:bfs-sweep-flushticks} shows a sweep of this parameter
+for 32-node and 96-node BFS runs. The optimal timeout value varies
+with the size of the job. In the 32-node job, performance peaks with a
+2 ms timeout; with the 96-node job, performance peaks with a 14 ms
+timeout.
 
-UTS
-BFS
+\paragraph{Number of active tasks}
+
+\begin{figure}[htb]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/bfs_sweep_workers}
+\begin{minipage}{0.95\columnwidth} 
+  \caption{\label{fig:bfs-sweep-workers} Sensitivity to maximum active tasks}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
+When a task issues a request that requires a response, it blocks to
+allow other tasks to utilize its core. These tasks may also block. To
+support the many milleseconds of latency aggregation adds, we need to
+support many thousands of blocked tasks. One of the key parameters of
+the runtime is the number of blocked tasks allowed; we need enough to
+cover the network and aggregation latency, but too many running tasks
+can add extra latency as they all must be multiplexed onto the same core.
+
+Figure~\ref{fig:bfs-sweep-workers} shows a sweep of the maximum number
+of active tasks per core, holding all other parameters
+constant. Again, the optimal value varies with the size of the
+run. For 32 nodes, a maximum of 3072 active tasks per core gives the
+best result; for 96 nodes, 16000 tasks is best.
 
@@
 
+\TODO{incorporate UTS chunk size after rerun}
 it's important to steal multiple tasks to 
@@
 
-UTS
+amortizing cost of stealing
+spreading work out more
+most of the time we still little  (about 10 avg) but sometimes a lot
+(difference with BFS)
 
@@
 
-allows us to amoritize task creation overhead and exploit locality
-Gets us a 2x bump in performance
-not too sensitive
+\begin{figure}[htb]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/bfs_sweep_threshold}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:bfs-sweep-threshold} Sensitivity to parallel loop threshold. Note the log scale.}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
 
-UTS
-BFS
+We see benefit to limiting the amount of parallelism created by a
+recursive loop decoposition. The decomposition threshold parameter the
+runtime when to stop creating new tasks and just execute iterations
+serially. This allows us to amoritize the overhead of task creation
+and potentially exploit locality when data for adjacent iterations is
+also adjacent in memory.
+
+Figure~\ref{fig:bfs-sweep-threshold} shows a sweep of this parameter
+for a 32-node BFS run. We saw the best results serializing 32 or fewer
+iterations.
 
@@
 %\end{figure*}
-
-Chunksize:
-amortizing cost of stealing
-spreading work out more
-most of the time we still little  (about 10 avg) but sometimes a lot
-(difference with BFS)

### 2012-07-22 doc/asplos13/method.tex
the beginnings of an evaluation

@@
 Performance is measured in \emph{traversed edges per second} (TEPS), where the number of edges is the edges making up the generated BFS tree.
-One of the reference implementations of Graph500 BFS is for the XMT.
-We compare this against a straightforward Grappa implementation that algorithmically matches the XMT reference code to allow us to fairly compare the execution of the two systems.
-We do not employ algorithmic improvements beyond the reference code, though there are many \cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
+One of the reference implementations of Graph500 BFS is for the XMT; this code fails to scale past 16 XMT processors because it does not expose enough parallelism, so we modified the code to use a recursive loop decomposition similar to Grappa's.
+We compare this modified version against a straightforward Grappa implementation.
+We do not employ algorithmic improvements, though there are many \cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
 

### 2012-07-22 doc/asplos13/evaluation.tex
include uts plots, xmt uts temp plot: todo overlay

@@
 
-\subsubsection{Unbalanced tree search}
+\subsubsection{Unbalanced tree search in memory}
+We ran UTS-mem on Grappa and the XMT with a binomial and two geometric
+100M-vertex trees (T3L,T1L,T2L), a geometric 1.6B-vertex tree (T1XL), and a geometric
+4.2B-vertex tree (T1XXL).
+Both systems do relatively poorly on the binomial tree because it is very deep and a
+traversal produces too little parallelism to get good bandwidth.
+Figure~\ref{fig:grappa-xmt-uts}
+shows the performance in terms of number of vertices visited per
+second versus number of compute nodes. The Grappa results shown are for 5
+cores per node and the best parameter values from a search over
+runtime and application parameters.
+Grappa with \checkme{16} machines is faster than the entire XMT. Grappa achieves
+\checkme{188Mvert/s} with \checkme{128 machines} and the
+XMT achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes,
+Grappa adds 1.4 Mvert/s/node. The XMT scales at 850 Kvert/s/node,
+until it plateaus. Grappa keeps scaling up through 128---the maximum we
+could use---although scaling declines because of the unscalability of our aggregation
+mechanism as number of network endpoints increases. Using 6 cores per
+node gave Grappa up to a 20\% benefit at each data point, but Grappa could
+not run well beyond 90 machines with this core configuration.
 
-We ran UTS-mem on Grappa and the XMT with a binomial and two geometric
-100M-vertex trees and a geometric 1.6B-vertex tree
-\footnote{T3L: -t 0 -b 2000 -q 0.200014 -m 5 -r 7\\
-          T2L: -t 1 -a 2 -d 23 -b 7 -r 220\\
-          T1L: -t 1 -a 3 -d 13 -b 4 -r 29\\
-          T1XL: -t 1 -a 3 -d 15 -b 4 -r 29\\}.
-Both systems do relatively poorly on the binomial tree because it is very deep and a
-traversal produces little parallelism, which both systems depend on.
-\TODO{data from our trace would be nice here}. Figure~\ref{fig:grappa-xmt-uts}
-shows the performance in terms of number of vertices visited per
-second versus number of compute nodes. The Grappa results are for the
-best parameter values from a limited search over \flushtimeout~and
-\asyncforthr~\TODO{want to say this in general at beginning of eval?}.
-Grappa with 16 machines is faster than the entire XMT. Grappa achieves
-\checkme{112Mvert/s} with \checkme{60 machines} and the
-XMT achieves only 50Mvert/s. Scaling up, Grappa adds 2Mvert/s/nodes
-and the XMT adds 850Kvert/s/node. Grappa keeps scaling up through 128
-machines, although scaling declines because of the unscalability of our aggregation
-mechanism as number of network endpoints increases.
+%% UTS: performance comparison
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/uts_compare.pdf}
+    \end{center}
+    \caption{Performance of in-memory unbalanced tree search. Grappa
+        keeps scaling through 128 nodes and beats the XMT with 16
+        nodes.}
+    \label{fig:uts_compare}
+\end{figure}
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/uts_xmt_temp.pdf}
+    \end{center}
+    \caption{This is XMT uts result, \TODO{overlay the plots}}
+    \label{fig:uts_compare_temp}
+\end{figure}
 
-The XMT plataeus around \checkme{60} processors for 100M, 1.6B, and
-4.2B-vertex trees, so performance is not limited by availability of
-parallelism for this implementation of UTS-mem.
-
-{\em you can rewrite this or blow it off -- just trying to get some thoughts in before leaving}
-Despite our efforts to tune the UTS implementation specific to Cray XMT, performance does not scale well with increasing processor count, flattening out around sixty processors.
-When we increase the size of the tree, we find that performance does not improve, suggesting that performance is not limited by task parallelism.
+Despite our efforts to tune the UTS implementation specific to Cray
+XMT, performance does not scale well with increasing processor count,
+flattening out around sixty processors.  When we increase the size of
+the tree from 100M to 4.2B, we find that performance does not improve, suggesting that performance is not limited by task parallelism.
 Cray's performance tools show that an increasing number of memory retry operations are generated by code within the runtime system.
@@
 
-\TODO{why is XMT getting different results for T1L,T2L; would expect it to be the machine with the most uniform results.}
-
-\TODO{sensitivity study for \asyncforthr parameter, ie caching amount. Although we can take advantage of locality in children array and the XMT cannot, we still win without it}
-
-\begin{figure}[h]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/grappa-xmt-uts.pdf}
-    \end{center}
-    \caption{describe...}
-    \label{fig:grappa-xmt-uts}
-\end{figure}
+\TODO{ we still win with th=1 }
 
@@
 
+%% UTS: chunk size
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/uts_chunksize.pdf}
+    \end{center}
+    \caption{Performance of UTS-Mem with varying maximum chunk size of
+    steals, run with 30 nodes, 6 cores per node, 4000 workers,
+    \checkme{6M flush ticks}}
+    \label{fig:uts_chunksize}
+\end{figure}
+
 amortizing cost of stealing
@@
 
+% uts threshold
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/uts_threshold.pdf}
+    \end{center}
+    \caption{Performance of UTS-Mem with varying parallel loop
+        threshol, run with 30 nodes, 6 cores per node, 4000 workers,
+    \checkme{6M flush ticks}}
+    \label{fig:uts_threshold}
+\end{figure}
+
 %\begin{figure*}[ht]

### 2012-07-22 doc/asplos13/grappa.tex
fixed small issues

@@
       \item[ \texttt{ delegate\_read( global\_address, local\_var )} ] 
-      \item[ \texttt{ delegate\_write( global\_address, local\_var )} ] \vspace{-2ex}
-      \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] \vspace{-2ex}
-      \item[ \texttt{ delegate\_inc( global\_address, local\_var )} ] \vspace{-2ex} \hfill \\
+      \item[ \texttt{ delegate\_write( global\_address, local\_var )} ] %\vspace{-2ex}
+      \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] %\vspace{-2ex}
+      \item[ \texttt{ delegate\_inc( global\_address, local\_var )} ] %\vspace{-2ex} 
+\hfill \\
         Performs a memory operation at the home core of a global address
       \item[ \texttt{ cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} ]
-      \item[ \texttt{ cache\_release( global\_address, local\_buf )} ] \vspace{-2ex} \hfill \\
+      \item[ \texttt{ cache\_release( global\_address, local\_buf )} ] %\vspace{-2ex} 
+\hfill \\
         Perform cache operations \TODO{expand}

### 2012-07-22 doc/asplos13/evaluation.tex
Minor edits to eval

@@
 \subsubsection{Unbalanced tree search in memory}
+
 We ran UTS-mem on Grappa and the XMT with a binomial and two geometric
-100M-vertex trees (T3L,T1L,T2L), a geometric 1.6B-vertex tree (T1XL), and a geometric
-4.2B-vertex tree (T1XXL).
-Both systems do relatively poorly on the binomial tree because it is very deep and a
-traversal produces too little parallelism to get good bandwidth.
-Figure~\ref{fig:grappa-xmt-uts}
-shows the performance in terms of number of vertices visited per
-second versus number of compute nodes. The Grappa results shown are for 5
-cores per node and the best parameter values from a search over
-runtime and application parameters.
-Grappa with \checkme{16} machines is faster than the entire XMT. Grappa achieves
-\checkme{188Mvert/s} with \checkme{128 machines} and the
-XMT achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes,
-Grappa adds 1.4 Mvert/s/node. The XMT scales at 850 Kvert/s/node,
-until it plateaus. Grappa keeps scaling up through 128---the maximum we
-could use---although scaling declines because of the unscalability of our aggregation
-mechanism as number of network endpoints increases. Using 6 cores per
-node gave Grappa up to a 20\% benefit at each data point, but Grappa could
-not run well beyond 90 machines with this core configuration.
+100M-vertex trees (T3L,T1L,T2L), a geometric 1.6B-vertex tree (T1XL),
+and a geometric 4.2B-vertex tree (T1XXL). Both systems do relatively
+poorly on the binomial tree because it is very deep and a traversal
+produces too little parallelism. Figure~\ref{fig:grappa-xmt-uts} shows
+the performance in terms of number of vertices visited per second versus
+number of compute nodes. The Grappa results shown are for 5 cores per
+node and the best parameter values from a search over runtime and
+application parameters. Grappa with \checkme{16} machines is faster than
+the entire XMT of 128 processors. Grappa achieves \checkme{188Mvert/s}
+with \checkme{128 machines} and the XMT achieves only 50Mvert/s,
+plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
+The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
+scaling up through 128---the maximum we could use---although scaling
+declines because of the unscalability of our aggregation mechanism as
+number of network endpoints increases. Using 6 cores per node gave
+Grappa up to a 20\% benefit at each data point, but Grappa could not run
+well beyond 90 machines with this core configuration because of the
+current network aggregator design.
 
@@
 
-Despite our efforts to tune the UTS implementation specific to Cray
+Despite our efforts to tune the UTS implementation specific to the 
 XMT, performance does not scale well with increasing processor count,
@@
 On a Cray XMT, network bandwidth is limited, so retries create congestion.
-In comparison, Grappa performs synchronization without retries, delaying response at the receiving end until ready to notify the sender to proceed.
+In comparison, Grappa performs synchronization without retries, delaying responses at the receiving end until ready to notify the sender to proceed.
 This saves bandwidth and permits scaling of tasks performing synchronization even on low injection rate networks.
@@
 Figure~\ref{fig:bfs-performance} shows Grappa's performance on BFS
-compared to that of the XMT\footnote{We do not know the reason for the
-  XMT's performance dip at 112 nodes, but it is repeatable}. There are
+compared to that of the XMT\footnote{The performance dip at 112 nodes is not an anomaly.
+It is highly repeatable.
+Despite extensive investigation on our part, we can find no reasonable explanation for it.}.
+There are
 two things to observe. First, At 128 XMT processors/cluster nodes,
-the XMT is  4.17 times faster than Grappa. This is likely due to the
+the XMT is  4.17 times faster than Grappa. This is due to the
 XMT's faster network: the XMT is able to issue 100 million network
@@
 aggregating 25 million network requests per second, or 4 times slower.
-Second, Grappa is scaling as we add processors.
+\checkme{Second, Grappa is scaling...}
 
@@
 processors/cluster nodes, the XMT is 1.75 times faster than Grappa.
+\TODO{Add more here...}
 
-\subsection{Benefit of aggregator}
+\subsection{Network aggregation performance}
 
@@
 
-To evaluate the benefit of the aggregator, we ran two
-experiments. First, we ran a simple unidirectionaly ping test to see
-the maximum benefit the aggregator can provide in terms of improved
-network efficiency. Second, we ran BFS with the aggregator disabled in
-order to measure its benefit on an application.
+To evaluate the benefits of network aggregation, we ran two experiments.
+First, we ran a simple unidirectional ping test to see the maximum
+benefit the aggregator can provide in terms of improved network
+efficiency. Second, we ran BFS with the aggregator disabled in order to
+measure its benefit on an application.
 
@@
 
-First, small message performance against the existing libraries is
-terrible. The MPI application test shows us that peak per node
-bandwidth supported by our Infiniband card is 3.4GB/s. This is
+First, small message performance against the existing libraries is, as expected, poor. The MPI application test shows us that peak per node
+bandwidth supported by our infiniband card is 3.4GB/s. This is
 achievable only with large messages; we must send 16KB packets to get
@@
 order of magnitude. With aggregation, Grappa is able to send 32-byte
-messags over 12 times faster than using GASNet directly. This is a
+messages over 12 times faster than using GASNet directly. This is a
 more respectable 32 percent of peak bandwidth. Due to expedient design
 decisions, Grappa's aggregator limits its aggregation to 4KB; this
-limits its peak achievable bandwidth to to 75 percent of the actual
+limits its peak achievable bandwidth to 75 percent of the actual
 peak.
@@
 
-\paragraph{Aggreator timeout}
+\paragraph{Aggregator timeout}
 

### 2012-07-22 doc/asplos13/intro.tex
pass on intro

@@
 
-Irregular applications are those whose parallel execution generates tasks with work, interdependences, or memory accesses that are highly sensitive to input.
-Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow.
-Contemporary examples include real-time mining or querying large unstructured data sets in the business, national security, machine learning, data-driven science, and social network computing domains.
-For these emerging applications, fast response -- given the sheer amount of data -- requires multinode systems.
-The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband.
-Our goal is to facilitate scaling of irregular applications efficiently on these mass market systems.
+Irregular applications generates tasks with work, interdependences, or memory
+accesses that are highly sensitive to input. Classic examples of irregular
+applications include branch and bound optimization, SPICE circuit simulation,
+contact algorithms in car crash analysis, and network flow. Important
+contemporary examples include processing large graphs in the business,
+national security, machine learning, data-driven science, and social network
+computing domains. For these emerging applications, fast response -- given the
+sheer amount of data -- requires multinode systems. The most broadly available
+multinode systems are those built from x86 commodity computing nodes
+interconnected via ethernet or infiniband. Our goal is to enable scalable
+performance of irregular applications on these mass market systems.
 
-What makes our goal challenging is that irregular applications do not readily exploit the features of commodity systems.
-This is true even for single node systems.
-It is not atypical for any given task's data references to be spread randomly across the entire memory of the system.
-Yet, for commodity processors' caches to be effective, computations must exploit spatial locality.
-Unfortunately, commodity prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed or the accesses follow a predictable pattern, neither of which occurs in irregular applications.
-As a consequence, commodity microprocessors stall often when executing irregular applications.
+Our goals is challenging for two key reasons:
 
-On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.
-Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.
-Injection rate into the network is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.
-While some irregular applications can be restructured to better exploit locality, aggregating requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.
+\noindent{\bf Irregular applications observe little spatial locality.} It is
+not atypical for any given task's data references to be spread randomly across
+the entire memory of the system. This makes current memory hierarchy features
+ineffective. Unfortunately, commodity prefetching hardware is effective only
+when addresses are known many cycles before the data is consumed or the
+accesses follow a predictable pattern, neither of which occurs in irregular
+applications. As a consequence, commodity microprocessors stall often when
+executing irregular applications.
 
-An alternative to demanding programmers manage locality in their applications, was provided by the fully custom Tera MTA-2~\cite{mta-2} system.
-The MTA was well-suited to extremely irregular applications: it had a large distributed shared memory with no caches, relying on multithreaded processors to tolerate memory latency.
-On every clock cycle, each processor would execute a ready instruction chosen from one of its 128 hardware thread contexts.
-Any application with sufficient thread parallelism would fully utilize the processors independent of the memory access pattern (modulo hot-spots).
-The network was designed with an single word injection rate that matched the processor clock frequency and sufficient bandwidth to sustain a reference from every processor on every clock cycle.
-Unfortunately, while an excellent match to extremely irregular applications, the MTA was not cost-effective on applications that could exploit locality and had very poor single threaded performance, making it a commercial failure.
-The Cray XMT approximates the Tera MTA-2, substantially reducing its cost but not overcoming its narrow range of applicability.
+% What makes our goal challenging is that irregular applications do not
+% readily exploit the features of commodity systems.	This is true even for
+% single node systems.	It is not atypical for any given task's data
+% references to be spread randomly across the entire memory of the system.
+% Yet, for commodity processors' caches to be effective, computations must
+% exploit spatial locality.	Unfortunately, commodity prefetching hardware is
+% effective only when addresses are known many cycles before the data is
+% consumed or the accesses follow a predictable pattern, neither of which
+% occurs in irregular applications.	As a consequence, commodity
+% microprocessors stall often when executing irregular applications.
+
+\noindent{\bf Irregular applications frequently request off-node data with
+small messages.} On multinode systems, the challenges presented by low
+locality are analogous, and exacerbated by the increased latency of going
+off-node. Irregular applications also present a challenge to mass market
+network technology, which is designed to transfer large blocks of data, not
+the word-sized references emitted by irregular application tasks. Injection
+rate into the network is insufficient to utilize wire bandwidth when blocks
+are below about two-kilobytes, so any straightforward communication strategy
+severely under-utilizes the network.
+
+While some irregular applications can be restructured to better exploit
+locality, aggregating requests to increase message size, and manage the
+additional challenges of load balance and synchronization across multinode
+systems, the work to do so is formidable and requires knowledge and skills
+pertaining to distributed systems far beyond those of most application
+programmers.
+
+Luckily, many of the important irregular applications (e.g., graph processing,
+our focus on this paper) naturally offer large amounts of concurrency. This
+immediately suggests taking advantage of concurrency tolerate latencies of
+dealing with data movement. The fully custom Tera MTA-2~\cite{mta-2} system is
+classic example of supporting irregular applications by using concurrency to
+hide latencies. The MTA was well-suited to extremely irregular applications:
+it had a large distributed shared memory with no caches, relying on
+multithreaded processors to tolerate memory latency. On every clock cycle,
+each processor would execute a ready instruction chosen from one of its 128
+hardware thread contexts. Any application with sufficient thread parallelism
+would fully utilize the processors independent of the memory access pattern
+(modulo hot-spots). The network was designed with an single word injection
+rate that matched the processor clock frequency and sufficient bandwidth to
+sustain a reference from every processor on every clock cycle. Unfortunately,
+while an excellent match to extremely irregular applications, the MTA was not
+cost-effective on applications that could exploit locality and had very poor
+single threaded performance, making it a commercial failure. The Cray XMT
+approximates the Tera MTA-2, substantially reducing its cost but not
+overcoming its narrow range of applicability.
+
+% Two opportunities help overcome these challenges without putting more burden on the programmer: (1) Many interesting irregular applications (e.g., graph processing, our focus on this paper) naturally observe large amounts of concurrency; (2)
+
+% In addition to large amounts of concurrency in irregular applications, 
+% 
+% Commodity systems offer a large number of cores with good single-thread performance.
+Concurrency can be used to hide network latencies involved in fetching data from remote notes, mitigating the lack of locality.
+% 
+% 
+% ; (3) Commodity networks improved significantly and offer high bandwidth at a relatively low cost
 
@@
 invocations. To this commodity hardware and software mix Grappa adds three
-main software components: (1) a lightweight tasking layer that supports a
+main software components: (1) a \emph{lightweight tasking} layer that supports a
 context switch in as few as \checkme{40ns} and distributed load balancing; (2)
-a distributed shared memory layer that supports customary operations such as
-read and write as well as remote operations such as
-fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that
+a \emph{distributed shared memory} layer that supports customary operations such as
+\emph{read and write as well as remote operations} such as
+fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message aggregation} layer that
 combines short messages to mitigate the aforementioned problem that commodity
@@
 
-Each of these components will be described in more detail in Section~\ref{sec:grappa}.
+% 
+% Each of these components will be described in more detail in Section~\ref{sec:grappa}.
+% 
+% {\em Provide evidence we have succeeded -- will rewrite as necessary based on final results:}
 
-{\em Provide evidence we have succeeded -- will rewrite as necessary based on final results:}
-Grappa achieves its design goal, which is to make irregular applications execute quickly and scalably on commodity clusters.
-Our yardstick for comparison is the XMT hardware itself.
+Our evaluation of Grappa shows that it runs several graph-crunching applications (classic examples of irregular behavior) very efficiently on a commodity cluster  Our yardstick for comparison is the XMT hardware itself.
 Using the same number of processors (in our case cores on a single multicore processor), Grappa is in the same ballpark as the XMT: on some graphs, such as the \checkme{WTF} Grappa is \checkme{2X} faster, while on others, such as the \checkme{WTF2} Grappa is \checkme{2X} slower.
@@
 For significantly less real world cost, users can \emph{add} more processors to a commodity cluster and easily outperform the XMT hardware on graph processing, while still saving significant dollars.
-
-\TODO{Add roadmap here. Luis: do we really want a roadmap here? We have enough about what we have done I think, so we might not need to sell things we plan to do.}
+% 
+% \TODO{Add roadmap here. Luis: do we really want a roadmap here? We have enough about what we have done I think, so we might not need to sell things we plan to do.}

### 2012-07-22 doc/asplos13/grappa.tex
Small changes.

--- 2012-07-22 doc/asplos13/grappa.tex
@@
 
-In this section we describe the Grappa runtime system
-(Figure~\ref{fig:grappa}). Grappa provides three key capabilities. The
-first is a distributed shared \textbf{global memory} with high aggregate
-random access bandwidth. The second is a \textbf{tasking} system with
-lightweight multithreading and automatic load balancing through work
-stealing. Finally, under the hood, a \textbf{communication} layer uses
-network packet scheduling and aggregation to achieve high performance
-while remaining largely invisible to applications. We now discuss the
+The Grappa runtime system (Figure~\ref{fig:grappa}) supports scaling of irregular applications via three key components:
+a \textbf{distributed shared global memory} providing high aggregate random access bandwidth for both normal and synchronized operations;  
+a \textbf{tasking system} with lightweight multithreading to tolerate latency and workstealing to automate load balance;
+and a \textbf{communication layer} using network packet scheduling and aggregation to achieve high performance while remaining largely invisible to applications.
+Unlike many runtime systems where the task scheduler, memory access manager, and the communication  are separable subsystems, in Grappa these three components are tightly integrated and the performance of each depends heavily on that of the other two.
+For example, eficiency of the communication layer requires that the tasking system multiplex tasks -- and the memory manager fetch data -- sufficiently rapidly to aggregate hundreds of individual requests into each packet; otherwise, the network will be poorly utilized.
+Nonetheless, for simplicity of exposition, we describe each component in turn.
+
+We now discuss the
 programmer's view of Grappa's main capabilities, including a subset of

### 2012-07-22 doc/asplos13/relatedwork.tex
work around bad citation

--- 2012-07-18 doc/asplos13/relatedwork.tex
@@
-\section{Related work}
+\section{Background}
 
-Ways to organize this:
-Approaches to graphs: special purpose XMT -> cluster -> distributed frameworks {pregel,graphlab} -> limited programming model -> grappa gen purpose
+The goal of the Grappa framework is to simplify the task of implementing irregular applications.
+Grappa is built on many existing ideas in programming languages, systems and architecture.
+In this section we discuss related frameworks and key enabling technologies that Grappa builds upon.
 
-dsm, pgas
-lightweight threading
-other graph solutions: 
--graph parallel, distributed
---pregel - but is synchronous and thus inefficient
---graphlab - asynchrounous, esp useful for problems that have been shown empirically to converge faster with asynchrony
+\subsection{Comparative Frameworks}
 
-\subsection{Multithreading to tolerate latency}
-hardware: tera, cray
-smt,
-alewife
-gpu,
-software threads
-
-\subsection{Programming models for distributed graph processing}
-
-Distributed graph processing frameworks like Pregel \cite{pregel:2010} and Distributed GraphLab \cite{distgraphlab:vldb12} provide graph-parallel, vertex-centric programming abstractions that free the application writer from solving distributed system issues like scheduling parallelism, handling fault tolerance, and scaling communication.
+Distributed graph processing frameworks like Pregel \cite{pregel:2010} and Distributed GraphLab \cite{distgraphlab:vldb12} share similar goals.
 Pregel adopts a bulk-synchronous parallel (BSP) execution model, which makes it inefficient on workloads that could prioritize vertices.
 GraphLab, on the other hand, schedules vertex computations individually, allowing prioritization, which gives faster convergence in a variety of iterative algorithms.
-Grappa also supports dynamic parallelism with asynchronous execution, but has a more general purpose programming model, where parallelism is expressed as tasks or loop iterations.
-\todo{and express locality?
-mention pgas e.g.
-Chapel language implementation?
-How much to talk about how graphlab achieves scalable performance?}
+Grappa also supports dynamic parallelism with asynchronous execution, but parallelism is expressed as tasks or loop iterations.
 
-\subsection{Partitioned Global Address space}
-The goal of presenting a global view of distributed memory to the programmer is shared by the PGAS community, and is used in languages like Chapel \cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC}.
-In these programming models, access to shared structures look like normal memory references, but the programmer tries to minimize references to remote nodes.
-Unlike the usual programming goal in PGAS, since we target problems with poor locality, we design Grappa for remote references as a common case.
-Thread private data is still local, and unlike the XMT, we exploit locality where it does exist in an application (such as spatial locality in a graph edgelist), but we design the system to perform well in the midst of mostly remote accesses to large shared data structures.
-We would like to implement PGAS languages that support dynamic parallelism, like Chapel, in Grappa.
+\subsection{Global memory}
+
+Grappa implements a software distributed shared memory DSM system.
+Many traditional software DSM systems are page based~\cite{Treadmarks} and aim to hide the fact that they are built in software from applications by exploiting the processor's paging mechanisms.
+Instead Grappa, like other partitioned global address space (PGAS) models, implements its DSM at the language, rather than system level.
+Languages such as Chapel \cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC} make accesses to shared structures look like normal memory references.
+As we describe later, Grappa chooses a middle ground, where global addresses are explicit in the API and local accesses are emitted conventionally by the compiler.
+Similar to DSM systems Grappa provides a caching and coherence model to ensure high performance and consistent results.
+Cache operations are exposed to in the API, however, allowing software to optimize their usage.
+
+\subsection{Multithreading}
+
+Grappa uses multithreading to tolerate memory latency.
+This is a well known technique.
+Hardware implementations include the Tera MTA \cite{Tera}, Cray XMT \cite{XMT}, Simultaneous multithreading \cite{SMT}, MIT Alewife \cite{Alewife}, Cyclops \cite{Cyclops}, and even GPUs \cite{fatahalian}.
+As we describe in this paper, Grappa use a lightweight user-mode task scheduler to multiplex thousands of tasks on a single processing core.
+The large number of tasks is required because of the extremely large internode latency.

### 2012-07-22 doc/asplos13/relatedwork.tex
some minor edits

@@
 
-The goal of the Grappa framework is to simplify the task of implementing irregular applications.
-Grappa is built on many existing ideas in programming languages, systems and architecture.
-In this section we discuss related frameworks and key enabling technologies that Grappa builds upon.
+% The goal of the Grappa framework is to simplify the task of implementing
+% irregular applications.
+Grappa is built on many existing ideas in programming languages, systems and
+architecture. In this section we discuss related frameworks and key enabling
+technologies that Grappa builds upon.
 
@@
 
-Distributed graph processing frameworks like Pregel \cite{pregel:2010} and Distributed GraphLab \cite{distgraphlab:vldb12} share similar goals.
-Pregel adopts a bulk-synchronous parallel (BSP) execution model, which makes it inefficient on workloads that could prioritize vertices.
-GraphLab, on the other hand, schedules vertex computations individually, allowing prioritization, which gives faster convergence in a variety of iterative algorithms.
-Grappa also supports dynamic parallelism with asynchronous execution, but parallelism is expressed as tasks or loop iterations.
+Distributed graph processing frameworks like Pregel \cite{pregel:2010} and
+Distributed GraphLab \cite{distgraphlab:vldb12} share similar goals. Pregel
+adopts a bulk-synchronous parallel (BSP) execution model, which makes it
+inefficient on workloads that could prioritize vertices. GraphLab, on the
+other hand, schedules vertex computations individually, allowing
+prioritization, which gives faster convergence in a variety of iterative
+algorithms.\TODO{GraphLab is not general. Myers will expand on that here. Make
+a strong point about generality.} Grappa also supports dynamic parallelism
+with asynchronous execution, but parallelism is expressed as tasks or loop
+iterations.
 
@@
 
-Grappa implements a software distributed shared memory DSM system.
-Many traditional software DSM systems are page based~\cite{Treadmarks} and aim to hide the fact that they are built in software from applications by exploiting the processor's paging mechanisms.
-Instead Grappa, like other partitioned global address space (PGAS) models, implements its DSM at the language, rather than system level.
-Languages such as Chapel \cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC} make accesses to shared structures look like normal memory references.
-As we describe later, Grappa chooses a middle ground, where global addresses are explicit in the API and local accesses are emitted conventionally by the compiler.
-Similar to DSM systems Grappa provides a caching and coherence model to ensure high performance and consistent results.
-Cache operations are exposed to in the API, however, allowing software to optimize their usage.
+Grappa implements a software distributed shared memory DSM system. Many
+traditional software DSM systems are page based~\cite{Treadmarks,munin} and
+aim to hide the fact that they are built in software from applications by
+exploiting the processor's paging mechanisms. Instead Grappa, like other
+partitioned global address space (PGAS) models, implements its DSM at the
+language, rather than system level. Languages such as Chapel \cite{Chapel},
+X10 \cite{X10}, and UPC \cite{UPC} make accesses to shared structures look
+like normal memory references. As we describe later, Grappa chooses a middle
+ground, where global addresses are explicit in the API and local accesses are
+emitted conventionally by the compiler. Similar to DSM systems Grappa provides
+a caching and coherence model to ensure high performance and consistent
+results. Cache operations are exposed to in the API, however, allowing
+software to optimize their usage.
 
@@
 
-Grappa uses multithreading to tolerate memory latency.
-This is a well known technique.
-Hardware implementations include the Tera MTA \cite{Tera}, Cray XMT \cite{XMT}, Simultaneous multithreading \cite{SMT}, MIT Alewife \cite{Alewife}, Cyclops \cite{Cyclops}, and even GPUs \cite{fatahalian}.
-As we describe in this paper, Grappa use a lightweight user-mode task scheduler to multiplex thousands of tasks on a single processing core.
-The large number of tasks is required because of the extremely large internode latency.
+Grappa uses multithreading to tolerate memory latency. This is a well known
+technique. Hardware implementations include the Tera MTA \cite{Tera}, Cray XMT
+\cite{XMT}, Simultaneous multithreading \cite{SMT}, MIT Alewife
+\cite{Alewife}, Cyclops \cite{Cyclops}, and even GPUs \cite{fatahalian}. As we
+describe in this paper, Grappa use a lightweight user-mode task scheduler to
+multiplex thousands of tasks on a single processing core. The large number of
+tasks is required because of the extremely large internode latency.

### 2012-07-22 doc/asplos13/relatedwork.tex
some minor edits

@@
 
-\subsection{Comparative Frameworks}
-
+\paragraph{Comparable frameworks}
 Distributed graph processing frameworks like Pregel \cite{pregel:2010} and
@@
 
-\subsection{Global memory}
-
+\paragraph{Global memory}
 Grappa implements a software distributed shared memory DSM system. Many
@@
 
-\subsection{Multithreading}
-
+\paragraph{Multithreading}
 Grappa uses multithreading to tolerate memory latency. This is a well known

### 2012-07-22 doc/asplos13/relatedwork.tex
more todos

@@
 with asynchronous execution, but parallelism is expressed as tasks or loop
-iterations.
+iterations.\TODO{Should RAMcloud go here?}.
 

### 2012-07-22 doc/asplos13/intro.tex
fixed up intro

@@
 
-Irregular applications generates tasks with work, interdependences, or memory
+Irregular applications generate tasks with work, interdependences, or memory
 accesses that are highly sensitive to input. Classic examples of irregular
@@
 
-\noindent{\bf Irregular applications observe little spatial locality.} It is
-not atypical for any given task's data references to be spread randomly across
-the entire memory of the system. This makes current memory hierarchy features
-ineffective. Unfortunately, commodity prefetching hardware is effective only
-when addresses are known many cycles before the data is consumed or the
-accesses follow a predictable pattern, neither of which occurs in irregular
-applications. As a consequence, commodity microprocessors stall often when
-executing irregular applications.
+\noindent{\bf Irregular applications exhibit little spatial locality.}
+It is not atypical for any given task's data references to be spread
+randomly across the entire memory of the system. This makes current
+memory hierarchy features ineffective. Caches are of little assistance
+with such low data re-use and spatial locality. Commodity prefetching
+hardware is effective only when addresses are known many cycles before
+the data is consumed or the accesses follow a predictable pattern,
+neither of which occurs in irregular applications. As a consequence,
+commodity microprocessors stall often when executing irregular
+applications.
 
-% What makes our goal challenging is that irregular applications do not
-% readily exploit the features of commodity systems.	This is true even for
-% single node systems.	It is not atypical for any given task's data
-% references to be spread randomly across the entire memory of the system.
-% Yet, for commodity processors' caches to be effective, computations must
-% exploit spatial locality.	Unfortunately, commodity prefetching hardware is
-% effective only when addresses are known many cycles before the data is
-% consumed or the accesses follow a predictable pattern, neither of which
-% occurs in irregular applications.	As a consequence, commodity
-% microprocessors stall often when executing irregular applications.
-
-\noindent{\bf Irregular applications frequently request off-node data with
-small messages.} On multinode systems, the challenges presented by low
+\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low
 locality are analogous, and exacerbated by the increased latency of going
@@
 
-Luckily, many of the important irregular applications (e.g., graph processing,
-our focus on this paper) naturally offer large amounts of concurrency. This
-immediately suggests taking advantage of concurrency tolerate latencies of
-dealing with data movement. The fully custom Tera MTA-2~\cite{mta-2} system is
-classic example of supporting irregular applications by using concurrency to
-hide latencies. The MTA was well-suited to extremely irregular applications:
-it had a large distributed shared memory with no caches, relying on
-multithreaded processors to tolerate memory latency. On every clock cycle,
-each processor would execute a ready instruction chosen from one of its 128
-hardware thread contexts. Any application with sufficient thread parallelism
-would fully utilize the processors independent of the memory access pattern
-(modulo hot-spots). The network was designed with an single word injection
-rate that matched the processor clock frequency and sufficient bandwidth to
-sustain a reference from every processor on every clock cycle. Unfortunately,
-while an excellent match to extremely irregular applications, the MTA was not
-cost-effective on applications that could exploit locality and had very poor
-single threaded performance, making it a commercial failure. The Cray XMT
-approximates the Tera MTA-2, substantially reducing its cost but not
-overcoming its narrow range of applicability.
+Luckily, many of the important irregular applications (e.g., graph
+processing, our focus in this paper) naturally offer large amounts of
+concurrency. This immediately suggests taking advantage of concurrency
+to tolerate the latency of data movement. The fully custom Tera
+MTA-2~\cite{mta-2} system is a classic example of supporting irregular
+applications by using concurrency to hide latencies. It had a large
+distributed shared memory with no caches.  On every clock cycle, each
+processor would execute a ready instruction chosen from one of its 128
+hardware thread contexts, a sufficient number to fully hide memory
+access latency.  The network was designed with a single word injection
+rate that matched the processor clock frequency and sufficient bandwidth
+to sustain a reference from every processor on every clock cycle.
+Unfortunately, while an excellent match to extremely irregular
+applications, the MTA was not cost-effective on applications that could
+exploit locality and had very poor single threaded performance, making
+it a commercial failure. The Cray XMT approximates the Tera MTA-2,
+substantially reducing its cost but not overcoming its narrow range of
+applicability.
 
@@
 In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.
-This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating with inexpensive mass market hardware components the approach taken by Tera.
+This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating in software and inexpensive mass market hardware, the approach taken by Tera.
 We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.
 Ultimately, the opportunity is to cover the spectrum of irregular to regular computation: where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.
-Thus we make best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.
+Thus we make the best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.
 
-In this paper we introduce Grappa, a software runtime system that allows a
-commodity x86 distributed-memory HPC cluster to be programmed as if it were a
-single large shared-memory machine and provides scalable performance for
-irregular applications. Grappa is designed to smooth over some of the
-performance discontinuities in commodity hardware, giving good performance
-when there is little locality to be exploited while allowing the programmer to
-exploit it when it is available. Grappa leverages as much freely available and
-commodity infrastructure as possible. We use unmodified Linux for the
-operating system. An off the shelf user-mode infiniband device driver
-stack~\cite{Melonox?} is used for network interfacing. MPI~\cite{mpi} is used
-for process setup and tear down. GASnet~\cite{gasnet} is used as the
-underlying mechanism for remote memory reads and writes using active message
-invocations. To this commodity hardware and software mix Grappa adds three
-main software components: (1) a \emph{lightweight tasking} layer that supports a
-context switch in as few as \checkme{40ns} and distributed load balancing; (2)
-a \emph{distributed shared memory} layer that supports customary operations such as
-\emph{read and write as well as remote operations} such as
-fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message aggregation} layer that
-combines short messages to mitigate the aforementioned problem that commodity
-networks are designed to achieve peak bandwidth only on large packet sizes,
-yet irregular applications tend to fetch only a handful of bytes at a time.
+In this paper we introduce Grappa, a software runtime system that allows
+a commodity x86 distributed-memory HPC cluster to be programmed as if it
+were a single large shared-memory machine and provides scalable
+performance for irregular applications. Grappa is designed to smooth
+over some of the performance discontinuities in commodity hardware,
+giving good performance when there is little locality to be exploited
+while allowing the programmer to exploit it when it is available. Grappa
+leverages as much freely available and commodity infrastructure as
+possible. We use unmodified Linux for the operating system. An off the
+shelf user-mode infiniband device driver stack~\cite{Melonox?} is used
+for network interfacing. MPI~\cite{mpi} is used for process setup and
+tear down. GASnet~\cite{gasnet} is used as the underlying mechanism for
+remote memory reads and writes using active message invocations. To this
+commodity hardware and software mix Grappa adds three main software
+components: (1) a \emph{lightweight tasking} layer that supports a
+context switch in as few as 38ns and distributed load balancing; (2) a
+\emph{distributed shared memory} layer that supports customary
+operations such as \emph{read and write as well as remote operations}
+such as fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message
+aggregation} layer that combines short messages to mitigate the
+aforementioned problem that commodity networks are designed to achieve
+peak bandwidth only on large packet sizes, yet irregular applications
+tend to fetch only a handful of bytes at a time.
 

### 2012-07-22 doc/asplos13/intro.tex
commit to merge
single threaded -> single-thread

@@
 applications, the MTA was not cost-effective on applications that could
-exploit locality and had very poor single threaded performance, making
+exploit locality and had very poor single-thread performance, making
 it a commercial failure. The Cray XMT approximates the Tera MTA-2,

### 2012-07-22 doc/asplos13/grappa.tex
Hacked out and simplified grappa and restructured implementation

@@
 \end{figure}
-
-\subsection{Load Balancing and Network Aggregation}
-
-Grappa contains two components that are not directly visible to
-the programmer but are key to providing good performance. Both depend
-on Grappa's ability to tolerate latency in order to obtain this
-performance.
-
-The first is a distributed dynamic load balancing mechanism that uses
-work stealing to move tasks from busy cores to idle cores. When a core
-detects that it has idle execution resources and no new tasks to run, it
-issues steal requests over the network. The core continues to execute
-remaining active work while the steal requests are in flight. When a
-steal request succeeds, it returns with more tasks that can be matched
-with execution resources and scheduled.
-
-The second is a network packet scheduling and aggregation system. From
-the perspective of the application, all communication between tasks in
-Grappa takes place either through global shared memory or remote task
-spawns. Both of these requests are ultimately translated into explicit
-inter-system-node/core communication requests that are dispatched to
-the \gasnet~library. In our benchmarks, we found that most of our
-packets were small (44 bytes on average for BFS). Commodity network
-interfaces do not transport small packets efficiently (see
-Section~\ref{sec:bandwidth-versus-packet-size}): the rate at which the
-network interface will accept new packets of any size is limited, so
-the only way to use a respectable fraction of the bandwidth is to send
-larger messages.
-
-Grappa's latency tolerance allows us to leverage the parallelism in
-our problem domain to build larger packets. Tasks can generate
-requests faster than the network will accept new packets, so instead
-of immediately sending every message, Grappa queues messages heading
-for the same destination. Once enough messages are queued, Grappa
-bundles all the messages into a single packet and sends it to the
-destination core. Tasks that require a response can suspend, allowing
-their core to do other work while waiting for a reply.
-
-While queuing adds latency to each individual request, it is a much
-lighter-weight operation than sending a packet and allows us to
-amoritize the cost of sending the packet across multiple tasks, while
-also making better use of available network bandwidth. In this way,
-Grappa trades per-request latency for application-level
-throughput.

### 2012-07-22 doc/asplos13/implementation.tex
Hacked out and simplified grappa and restructured implementation

@@
 
-\subsubsection{Tasks and workers}
-
-Grappa tasks are 32-byte entities: a 64-bit function pointer plus
-three 64-bit arguments. We use three arguments because tasks are
-usually generated as part of a parallel loop decomposition, and thus
-each task needs three kinds of data.
+\paragraph{Tasks and workers} Grappa tasks are 32-byte entities: a
+64-bit function pointer plus three 64-bit arguments. We use three
+arguments because tasks are usually generated as part of a parallel loop
+decomposition, and thus each task needs three kinds of data.
 \begin{description}
@@
 
-\subsubsection{Context switching}
-
-Context switching in Grappa is non-preemptive. As with other cooperative
-multithreading systems, we treat context switches as function calls,
-saving and restoring only the callee-saved state as specified in the
-x86-64 ABI \cite{someone}. This involves saving six general-purpose
-64-bit registers and the stack pointer, as well as the 16-bit x87
-floating point control word and the SSE context/status register. Thus,
-the minimum amount of state a cooperative context switch routine must
-save according to the ABI is 62 bytes.
+\paragraph{Context switching} Grappa context switches between tasks
+non-preemptively. As with other cooperative multithreading systems, we
+treat context switches as function calls, saving and restoring only the
+callee-saved state as specified in the x86-64 ABI \cite{someone}. This
+involves saving six general-purpose 64-bit registers and the stack
+pointer, as well as the 16-bit x87 floating point control word and the
+SSE context/status register. Thus, the minimum amount of state a
+cooperative context switch routine must save according to the ABI is 62
+bytes.
 
@@
 for temporary values that are not needed after the context switch.
-\comment{I highly doubt this is important in real apps beyond our kernels.  -M
-Why not? liveness is a common compiler analysis -BM
-Because it has to be live across the entire application.  All real applications do some form of floating point, however minor... -M
 
-In
-particular, only one of our applications does floating point
-operations; all the other applications can safely ignore the x87 and
-SSE2 registers as part of context switching.}
-
-\subsubsection{Scheduling}
-
-Each core in the Grappa system has its own independent scheduler. Each
-scheduler has three main tasks to perform.  First, it must ensure that
-its communication resources are serviced reasonably often. Second, it
-must ensure that if a running task was waiting on a long-latency
-operation and that operation has completed, the task will be
-rescheduled. Third, if there are tasks waiting to be run and spare
-execution resources, the tasks must be matched with idle workers.
+\paragraph{Scheduling} Each core in the Grappa system has its own
+independent scheduler. Each scheduler has three main tasks to perform. 
+First, it must ensure that its communication resources are serviced
+reasonably often. Second, it must ensure that if a running task was
+waiting on a long-latency operation and that operation has completed,
+the task will be rescheduled. Third, if there are tasks waiting to be
+run and spare execution resources, the tasks must be matched with idle
+workers.
 
@@
 The scheduler ensures that communication maintenance tasks are executed periodically.
+
+\paragraph{Work stealing} When a Grappa node runs out of work, it
+becomes a ``thief'' and asks another other Grappa nodes (``victim'') for
+work. If the victim has tasks to spare in its public queue, they will be
+transferred to the thief's public queue.
+
+Most prior work on work stealing assumes that a processor has one worker
+thread and so it only steals when utilization would go to zero. Grappa,
+on the other hand, relies on having many worker threads per core. The
+core would be fully-utilized if there is always something on the ready
+queue. Even if there are many active tasks, if they are all suspended
+for long-latency network requests, then the core is underutilized. So,
+there is a choice: should the core use some of this idle time to perform
+work stealing or should it just wait as it is possible that local tasks
+will create more work. We choose to have a core initiate a steal when
+all of the following conditions hold: no workers are ready to run, the
+unstarted task queues are empty, and there are no outstanding steal
+requests. Having only one outstanding steal request throttles steals to
+prevent flooding the network, but other scalable quieting mechanisms
+would be possible, such as a voting
+tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
+Termination detection is not built into the Grappa task scheduler,
+rather, it is considered a programming error for the program to exit
+without syncing all tasks.
+
+\TODO{Someone check this paragraph to make sure it is true. :-)  -Mark}
+
+When a grappa node decides to issue a steal request it chooses the
+victim node at random.  It also requests a block of tasks at once.  If
+this block exceeds the amount of stealable tasks on the victim node,
+half of the stealable tasks are stolen instead.  While we did study the
+number of tasks to steal at once, we found performance was maximized
+with \checkme{128} tasks with little sensitivity around this number,
+hence we omit results for this from Section~\ref{sec:results} and simply
+use \checkme{128} as the parameter for all experiments.
 
@@
 
-\subsection{Load Balancing and Network Aggregation}
-
-\subsubsection{Work stealing} \label{subsec:implementation-worksteal}
-
-When a \checkme{grappa-process} runs out of work, it becomes a ``thief''
-and asks another other \checkme{grappa-process} (``victim'') for work.
-If the victim has tasks to spare in its public queue, they will be
-transferred to the thief's public queue.
-
-Most prior work on work stealing assumes that a processor has one worker
-thread and so it only steals when utilization would go to zero. Grappa,
-on the other hand, relies on having many worker threads per core. The
-core would be fully-utilized if there is always something on the ready
-queue. Even if there are many active tasks, if they are all suspended
-for long-latency network requests, then the core is underutilized. So,
-there is a choice: should the core use some of this idle time to perform
-work stealing or should it just wait as it is possible that local tasks
-will create more work. We choose to have a core initiate a steal when
-all of the following conditions hold: no workers are ready to run, the
-unstarted task queues are empty, and there are no outstanding steal
-requests. Having only one outstanding steal request throttles steals to
-prevent flooding the network, but other scalable quieting mechanisms
-would be possible, such as a voting
-tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
-Termination detection is not built into the Grappa task scheduler,
-rather, it is considered a programming error for the program to exit
-without syncing all tasks.
-
-\TODO{Someone check this paragraph to make sure it is true. :-)  -Mark}
-
-When a grappa node decides to issue a steal request it chooses the
-victim node at random.  It also requests a block of tasks at once.  If
-this block exceeds the amount of stealable tasks on the victim node,
-half of the stealable tasks are stolen instead.  While we did study the
-number of tasks to steal at once, we found performance was maximized
-with \checkme{128} tasks with little sensitivity around this number,
-hence we omit results for this from Section~\ref{sec:results} and simply
-use \checkme{128} as the parameter for all experiments.
-
-\comment{
-where to put the following related? -BM
-
-Fork join computations: implement with futures or continuation passing
-style
-feed-forward computations: implement with scalable task counting (we
-focus on this style in our benchmarks)
-(briefly describe global task joiner)
-
-choices we've made
-done -random victims
-hmm..  -steal from FIFO side of queue
-hmmm... -steal when at the moment readyQ empty and task queues empty
-done -steal min(chunk\_size, half)
-
-omit - Recent work has explored scalable work stealing in terms of efficient
-termination detection. 
-}
-
-\subsubsection{Network Aggregation}
+\subsection{Network Aggregation}
 
@@
 limited to RDMA-capable networks.
+
+\comment{
+
+Text that was in section 3 - Grappa.  Seems entirely redundant.  Not adding this anywhere.  -Mark
+
+\subsection{Load Balancing and Network Aggregation}
+
+Grappa contains two components that are not directly visible to
+the programmer but are key to providing good performance. Both depend
+on Grappa's ability to tolerate latency in order to obtain this
+performance.
+
+The first is a distributed dynamic load balancing mechanism that uses
+work stealing to move tasks from busy cores to idle cores. When a core
+detects that it has idle execution resources and no new tasks to run, it
+issues steal requests over the network. The core continues to execute
+remaining active work while the steal requests are in flight. When a
+steal request succeeds, it returns with more tasks that can be matched
+with execution resources and scheduled.
+
+The second is a network packet scheduling and aggregation system. From
+the perspective of the application, all communication between tasks in
+Grappa takes place either through global shared memory or remote task
+spawns. Both of these requests are ultimately translated into explicit
+inter-system-node/core communication requests that are dispatched to
+the \gasnet~library. In our benchmarks, we found that most of our
+packets were small (44 bytes on average for BFS). Commodity network
+interfaces do not transport small packets efficiently (see
+Section~\ref{sec:bandwidth-versus-packet-size}): the rate at which the
+network interface will accept new packets of any size is limited, so
+the only way to use a respectable fraction of the bandwidth is to send
+larger messages.
+
+Grappa's latency tolerance allows us to leverage the parallelism in
+our problem domain to build larger packets. Tasks can generate
+requests faster than the network will accept new packets, so instead
+of immediately sending every message, Grappa queues messages heading
+for the same destination. Once enough messages are queued, Grappa
+bundles all the messages into a single packet and sends it to the
+destination core. Tasks that require a response can suspend, allowing
+their core to do other work while waiting for a reply.
+
+While queuing adds latency to each individual request, it is a much
+lighter-weight operation than sending a packet and allows us to
+amoritize the cost of sending the packet across multiple tasks, while
+also making better use of available network bandwidth. In this way,
+Grappa trades per-request latency for application-level
+throughput.
+}

### 2012-07-22 doc/asplos13/grappa.tex
Further removing implementation from PL view

@@
 may be used. When there is no locality to exploit, {\em delegate}
-operations are used. Since these operations are expected to communicate
-with other nodes and have high latency, cache and delegate operations
-interact with the scheduler; when issuing a long-latency request, they
-suspend the requesting task and allow the core to execute other work
-until their response arrives.
+operations are used.
 
-\paragraph{Explicit caching} With Grappa's explicit caching support,
-applications can instruct Grappa to fetch a global pointer of any length
-and return a local pointer to a cached copy of the global memory. Under
-the hood, Grappa performs the mechanics of gathering chunks of data from
-multiple system nodes and presenting a conventional appearing linear
-block of memory as a pointer into a cache. Grappa cache operations have
-the usual read-only and read-write variants, along with a write-only
-variant used to initialize data structures. Cache operations exploit
-spatial locality by reducing the number of small network messages
-accessing contiguous data. Languages for distributed shared memory
-systems have done optimizations to achieve a similar goal. For example,
-the UPC compiler coalesces struct and array accesses into remote get/put
+\textbf{Explicit caching.} Grappa provides an API to fetch a global
+pointer of any length and return a local pointer to a cached copy of the
+global memory.  Grappa cache operations have the usual read-only and
+read-write variants, along with a write-only variant used to initialize
+data structures. Languages for distributed shared memory systems have
+done optimizations to achieve a similar goal. For example, the UPC
+compiler coalesces struct and array accesses into remote get/put
 \cite{Chen:2005}, and Fortran D compiler's message vectorization hoists
@@
 
-\paragraph{Delegate operations}
-When the access pattern has low-locality, it is more efficient
-to modify the data on its home core rather than bringing a copy to the
-requesting core and returning it after modification. Delegate
-operations provide this capability. Applications can dispatch
-computation to be performed on individual machine-word sized chunks of
-global memory to the memory system itself (e.g.,
-\emph{fetch-and-add}). \comment{should we bring up delegation in the intro?}
-
-Delegate operations, proposed in \cite{Nelson:hotpar11}, are also the
-primary synchronization method in Grappa. Each piece of global memory is
-paired with a single core in the system; all accesses to that actual
-memory location are done by that core. This allows delegate operations
-to be performed atomically on the core simply by making sure no other
-code interleaves their operation; and, since context switches in Grappa
-are cooperative, this is trivial. Thus, delegate operations are able to
-provide atomic semantics to memory owned by one core without using
-atomic operations. Using delegation to implement isolation for any
-subset of memory has been explored in \cite{delegated:oopsla11}.
+\textbf{Delegate operations.}
+When the access pattern has low-locality, it is more efficient to modify
+the data on its home core rather than bringing a copy to the requesting
+core and returning it after modification. Delegate operations provide
+this capability. Applications can dispatch computation to be performed
+on individual machine-word sized chunks of global memory to the memory
+system itself (e.g., \emph{fetch-and-add}).  Delegate operations,
+proposed in \cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the primary synchronization method in Grappa.
 

### 2012-07-22 doc/asplos13/implementation.tex
Further removing implementation from PL view

--- 2012-07-22 doc/asplos13/implementation.tex
@@
 
-All global memory operations are implemented with active
-messages. Requesters construct a descriptor holding the current task
-ID and space for a response, issue their request active message to the
-node given by the request address, and suspend themselves. When the
-request arrives at the address' home node, the operation is performed
-and a response active message is sent. When the response arrives at
-the requesting node, the descriptor is filled in with the results of
-the operation and the requesting task is placed back on the ready queue.
+All global memory operations are implemented with active messages.
+Requesters construct a descriptor holding the current task ID and space
+for a response, issue their request active message to the node given by
+the request address, and suspend themselves. When the request arrives at
+the address' home node, the operation is performed and a response active
+message is sent. When the response arrives at the requesting node, the
+descriptor is filled in with the results of the operation and the
+requesting task is placed back on the ready queue.
+
+\paragraph{Explicit caching.} Under the hood, Grappa performs the
+mechanics of gathering chunks of data from multiple system nodes and
+presenting a conventional appearing linear block of memory as a local
+pointer into a cache. The strategy employed is to issue all the
+constituent requests of a cache access request and then yield until all
+responses have occurred.  Currently, Grappa caches are \emph{not}
+coherent, requiring the programmer to maintain consistent access to
+data.  Future work will develop a software directory based coherence
+scheme to simplify consistent access to global data.
 
@@
 
-\paragraph{Explicit caches} Grappa cache operations are used to exploit
-spatial or temporal locality. They allow a single task to buffer a range
-of data, read or modify it, and write the modified version back to its
-home node(s). All fetch and update operations of cache data are
-expressed explicitly by the programmer. Cache operations may operate on
-data that spans multiple blocks of the linear address space; the
-programmer provides the starting address and length, and the cache
-issues as many active messages as necessary to retrieve all necessary
-data.
-
-\comment{If anything, this is just a footnote. -M
-All current cache operations are incoherent; two tasks may
-concurrently modify the same block with their writes interfering
-unless protected by synchronization through delegate operations. While
-this has been sufficient for our benchmarks, it is possible to embed
-synchronization in the cache active messages and reduce the number of
-network round trips necessary.}
-
-\subsection{Load Balancing and Network Aggregation}
-
-\subsubsection{Work stealing} \label{subsec:implementation-worksteal}
-
-When a \checkme{grappa-process} runs out of work, it becomes a ``thief''
-and asks another other \checkme{grappa-process} (``victim'') for work.
-If the victim has tasks to spare in its public queue, they will be
-transferred to the thief's public queue.
-
-Most prior work on work stealing assumes that a processor has one worker
-thread and so it only steals when utilization would go to zero. Grappa,
-on the other hand, relies on having many worker threads per core. The
-core would be fully-utilized if there is always something on the ready
-queue. Even if there are many active tasks, if they are all suspended
-for long-latency network requests, then the core is underutilized. So,
-there is a choice: should the core use some of this idle time to perform
-work stealing or should it just wait as it is possible that local tasks
-will create more work. We choose to have a core initiate a steal when
-all of the following conditions hold: no workers are ready to run, the
-unstarted task queues are empty, and there are no outstanding steal
-requests. Having only one outstanding steal request throttles steals to
-prevent flooding the network, but other scalable quieting mechanisms
-would be possible, such as a voting
-tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
-Termination detection is not built into the Grappa task scheduler,
-rather, it is considered a programming error for the program to exit
-without syncing all tasks.
-
-\TODO{Someone check this paragraph to make sure it is true. :-)  -Mark}
-
-When a grappa node decides to issue a steal request it chooses the
-victim node at random.  It also requests a block of tasks at once.  If
-this block exceeds the amount of stealable tasks on the victim node,
-half of the stealable tasks are stolen instead.  While we did study the
-number of tasks to steal at once, we found performance was maximized
-with \checkme{128} tasks with little sensitivity around this number,
-hence we omit results for this from Section~\ref{sec:results} and simply
-use \checkme{128} as the parameter for all experiments.
-
-\comment{
-where to put the following related? -BM
-
-Fork join computations: implement with futures or continuation passing
-style
-feed-forward computations: implement with scalable task counting (we
-focus on this style in our benchmarks)
-(briefly describe global task joiner)
-
-choices we've made
-done -random victims
-hmm..  -steal from FIFO side of queue
-hmmm... -steal when at the moment readyQ empty and task queues empty
-done -steal min(chunk\_size, half)
-
-omit - Recent work has explored scalable work stealing in terms of efficient
-termination detection. 
-}
-
-\subsubsection{Network Aggregation}
+\subsection{Network Aggregation}
 

### 2012-07-22 doc/asplos13/grappa.tex
simplifying intro

@@
 
-The Grappa runtime system (Figure~\ref{fig:grappa}) supports scaling of irregular applications via three key components:
-a \textbf{distributed shared global memory} providing high aggregate random access bandwidth for both normal and synchronized operations;  
-a \textbf{tasking system} with lightweight multithreading to tolerate latency and workstealing to automate load balance;
-and a \textbf{communication layer} using network packet scheduling and aggregation to achieve high performance while remaining largely invisible to applications.
-Unlike many runtime systems where the task scheduler, memory access manager, and the communication  are separable subsystems, in Grappa these three components are tightly integrated and the performance of each depends heavily on that of the other two.
-For example, eficiency of the communication layer requires that the tasking system multiplex tasks -- and the memory manager fetch data -- sufficiently rapidly to aggregate hundreds of individual requests into each packet; otherwise, the network will be poorly utilized.
-Nonetheless, for simplicity of exposition, we describe each component in turn.
-
-We now discuss the
-programmer's view of Grappa's main capabilities, including a subset of
-the Grappa API. In Section~\ref{implementation}, we discuss the details
-of their implementation.
+From the perspective of a software developer, Grappa
+(Figure~\ref{fig:grappa}) provides two main software components: a
+\textbf{tasking system} with lightweight multithreading to tolerate
+latency and workstealing to automate load balance; and a
+\textbf{distributed shared global memory} providing high aggregate
+random access bandwidth for both normal and synchronized operations. We
+now discuss the programmer's view of Grappa's main capabilities,
+including a subset of the Grappa API. In Section~\ref{implementation},
+we discuss the details of their implementation as well as a third
+component, the network communication layer that is key to performance
+but largely invisible to software developers.
 

### 2012-07-23 doc/asplos13/grappa.tex
add delegate/cache figure

@@
       \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] %\vspace{-2ex}
-      \item[ \texttt{ delegate\_inc( global\_address, local\_var )} ] %\vspace{-2ex} 
+      \item[ \texttt{ delegate\_fetch\_inc( global\_address, local\_var )} ] %\vspace{-2ex} 
 \hfill \\
@@
 \end{figure}
+
+\paragraph{Example}
+
+\begin{figure*}[htb]
+\begin{center}
+  \includegraphics[width=1.5\columnwidth]{figs/delegate-cache}
+\begin{minipage}{1.5\columnwidth}
+  \caption{\label{fig:delegate-cache} Delegation and cache example.}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure*}
+
+Figure~\ref{fig:delegate-cache} shows an example of delegate and cache
+operations. In step 1, a core allocates an array in the global
+heap. It then spawns two tasks on remote cores to increment elements
+of the array.
+
+The first task increments two elements of the array using cache
+operations. In step 2, the task is invoked. A cache request is issued
+for two adjacent integers starting at the second element of the
+array. Since these element are stored in the memories of two different
+cores, this requires the sending of two messages in step 3. The task
+is suspended until both responses arrive in step 4. The data carried
+in these responses is stored in the local buffer. The elements are
+then increment in the buffer. Then in step 5, the modified data is
+sent back to the home node. Acknoledgements are returned in step 6 so
+the task knows when the writes are complete.
+
+The second task increments an element of the array with a delegate
+operation. In step 7, the task is invoked. A delegate request is sent
+to the home core of the array element with the increment value. The
+task suspends until the response is received. In step 8, the increment
+is executed on the remote core. A response is returned in step 9 with
+the previous value of the array element.

### 2012-07-23 doc/asplos13/communication.tex
Super redux

--- 2012-07-22 doc/asplos13/implementation.tex
@@
-\section{Implementation} \label{sec:implementation}
-
-We now discuss implementation details for the major system components discussed in Section~\ref{sec:grappa}.
-
-\subsection{Tasks}
-
-\paragraph{Tasks and workers} Grappa tasks are 32-byte entities: a
-64-bit function pointer plus three 64-bit arguments. We use three
-arguments because tasks are usually generated as part of a parallel loop
-decomposition, and thus each task needs three kinds of data.
-\begin{description}
-\item[Function pointer] This 64-bit value indicates what routine
-  should run. We configure the system nodes to disable address randomization, so that we can depend on the fact that function pointers are valid across process images.
-\item[Private argument] This 64-bit value is typically used for task-specific
-  data (e.g., a loop index).
-\item[Shared argument] This 64-bit value is typically used for data shared
-  between all tasks that are part of a loop or to specify the number
-  of loop iterations.
-\item[Synchronization argument] This 64-bit value is typically used to determine
-  when all tasks that are part of a loop have finished. It is usually
-  a global pointer to a synchronization object allocated at the core
-  that spawned the task.
-\end{description}
-While these are the most common uses of the three task arguments, they
-are treated as arbitrary 64-bit values in the runtime, and can be used
-for any purpose.
-
-Tasks are not allocated any execution resources until the scheduler
-decides to run them; when this occurs, tasks are matched with {\em
-  worker} threads. Each worker is simply a collection of status bits and a
-stack, allocated at a particular core.
-
-\paragraph{Context switching} Grappa context switches between tasks
-non-preemptively. As with other cooperative multithreading systems, we
-treat context switches as function calls, saving and restoring only the
-callee-saved state as specified in the x86-64 ABI \cite{someone}. This
-involves saving six general-purpose 64-bit registers and the stack
-pointer, as well as the 16-bit x87 floating point control word and the
-SSE context/status register. Thus, the minimum amount of state a
-cooperative context switch routine must save according to the ABI is 62
-bytes.
-
-Since the compiler sees all calls to the context switch routine, we
-can save even less state. Our context switch routine appears to the
-compiler as inline assembly; we declare all the registers we need
-to save as ``clobbered'' by the inline assembly routine, and the
-compiler will issue its own save and restore code as needed. This allows the
-compiler to avoid saving any registers that are not used, or are used
-for temporary values that are not needed after the context switch.
-
-\paragraph{Scheduling} Each core in the Grappa system has its own
-independent scheduler. Each scheduler has three main tasks to perform. 
-First, it must ensure that its communication resources are serviced
-reasonably often. Second, it must ensure that if a running task was
-waiting on a long-latency operation and that operation has completed,
-the task will be rescheduled. Third, if there are tasks waiting to be
-run and spare execution resources, the tasks must be matched with idle
-workers.
-
-Each scheduler has three queues:
-\begin{description}
-\item{\bf Ready worker queue} This is a FIFO queue of tasks that are
-  matched with workers and are ready to execute.
-\item{\bf Private task queue} This is a FIFO queue of tasks that must run on this core.
-\item{\bf Public task queue} This is a LIFO queue of tasks that are
-  waiting to be matched with workers. It is a local partition of a shared
-  task pool.
-\end{description}
-
-Whenever a task yields or suspends, the scheduler makes a decision about
-what to do next. First, it determines if the communication resources
-should be serviced. This is done on a periodic basis. Second, it
-determines if any workers with running tasks are ready to execute; if
-so, one is scheduled. Finally, if there are no workers ready to run, but
-there are tasks waiting to be matched with workers, an idle worker is
-woken (or a new worker is spawned), matched with a task, and scheduled.
-The scheduler ensures that communication maintenance tasks are executed periodically.
-
-\paragraph{Work stealing} When a Grappa node runs out of work, it
-becomes a ``thief'' and asks another other Grappa nodes (``victim'') for
-work. If the victim has tasks to spare in its public queue, they will be
-transferred to the thief's public queue.
-
-Most prior work on work stealing assumes that a processor has one worker
-thread and so it only steals when utilization would go to zero. Grappa,
-on the other hand, relies on having many worker threads per core. The
-core would be fully-utilized if there is always something on the ready
-queue. Even if there are many active tasks, if they are all suspended
-for long-latency network requests, then the core is underutilized. So,
-there is a choice: should the core use some of this idle time to perform
-work stealing or should it just wait as it is possible that local tasks
-will create more work. We choose to have a core initiate a steal when
-all of the following conditions hold: no workers are ready to run, the
-unstarted task queues are empty, and there are no outstanding steal
-requests. Having only one outstanding steal request throttles steals to
-prevent flooding the network, but other scalable quieting mechanisms
-would be possible, such as a voting
-tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
-Termination detection is not built into the Grappa task scheduler,
-rather, it is considered a programming error for the program to exit
-without syncing all tasks.
-
-\TODO{Someone check this paragraph to make sure it is true. :-)  -Mark}
-
-When a grappa node decides to issue a steal request it chooses the
-victim node at random.  It also requests a block of tasks at once.  If
-this block exceeds the amount of stealable tasks on the victim node,
-half of the stealable tasks are stolen instead.  While we did study the
-number of tasks to steal at once, we found performance was maximized
-with \checkme{128} tasks with little sensitivity around this number,
-hence we omit results for this from Section~\ref{sec:results} and simply
-use \checkme{128} as the parameter for all experiments.
-
-\subsection{Memory}
-
-\TODO{Is there an off by 1 error in here?
-Need to read the AMD64 docs carefully to see if we can sign extend the top bit of the 48 bits of real virtual address space, or if we need to maintain bit 49.
--M}
-
-Grappa implements a software distributed shared memory. All global
-addresses in Grappa are 64-bit values. Each byte of global memory is
-associated with a {\em home core}; this core is responsible for
-allocating and modifying that memory. Since today's commodity processors
-support a 48-bit virtual address space \cite{AMD64}, Grappa uses the
-spare bits to distinguish which address space a global address belongs
-to, and to represent the home core of a global address.
-
-Global memory in Grappa is exposed through two distinct, non-overlapping
-address spaces; \textbf{2D global addresses}, optimized for global
-access to node-local data, and \textbf{linear global addresses},
-optimized for high aggregate random access bandwidth to low locality
-data.
-
-\begin{figure}[t]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:memory-structure} Grappa memory structure}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-\paragraph{2D global addresses} The first global address space is the {\em
-two-dimensional} address space. This uses a traditional PGAS addressing model,
-where each address is a tuple of a rank in the job (or global process ID) and
-an address in that process. The lower 48 bits of the address hold a virtual
-address in the process. The top bit is set to indicates that the reference is
-a 2D address. This leaves 15 bits for network endpoint ID, which limits our
-scalability to $2^{15}$ endpoints.
-
-Any node-local data can be made accessible by other nodes in the system
-by wrapping the address and node ID into a 2D global address. This
-address can then be accessed with a delegate and can also be cached by
-other nodes. At the destination the address is converted into a
-canonical x86 address by replacing the upper bits with the sign-extended
-upper bit of the virtual address. 2D addresses may refer to memory
-allocated from a single processes' heap or from a task's stack.
-Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can
-refer to other cores' memory.
-
-\paragraph{Linear global addresses} The second global address space is
-the {\em linear} address space, so named because nearby addresses can be
-treated as adjacent even if they are stored on different nodes.
-Addresses in the linear address space are partitioned across the cluster
-in a block-cyclic fashion. Choosing the block size involves trading off
-sequential bandwidth against aggregate random access bandwidth. Smaller
-block sizes help spread data across all the memory controllers in the
-cluster, but larger block sizes allow the locality-optimized memory
-controllers to provide increased sequential bandwidth. The block size,
-which is configurable, is typically set to 64 bytes, or the size of a
-single hardware cache line, in order to exploit spatial locality when
-available.
-
-All linear addresses refer to data allocated from a global heap. The
-heap metadata is stored on a single node. Currently all heap operations
-serialize through this node; while this has been sufficient for our
-benchmarks, in the future Grappa will provide parallel performance
-through combining~\cite{MAMA,flatcombining}.
-
-\TODO{Add a figure that depicts both forms of global addresses.}
-
-All global memory operations are implemented with active messages.
-Requesters construct a descriptor holding the current task ID and space
-for a response, issue their request active message to the node given by
-the request address, and suspend themselves. When the request arrives at
-the address' home node, the operation is performed and a response active
-message is sent. When the response arrives at the requesting node, the
-descriptor is filled in with the results of the operation and the
-requesting task is placed back on the ready queue.
-
-\paragraph{Explicit caching.} Under the hood, Grappa performs the
-mechanics of gathering chunks of data from multiple system nodes and
-presenting a conventional appearing linear block of memory as a local
-pointer into a cache. The strategy employed is to issue all the
-constituent requests of a cache access request and then yield until all
-responses have occurred.  Currently, Grappa caches are \emph{not}
-coherent, requiring the programmer to maintain consistent access to
-data.  Future work will develop a software directory based coherence
-scheme to simplify consistent access to global data.
-
-\paragraph{Delegates} Delegate operations are used for low-locality
-accesses and synchronization. Their modifications are always executed at
-the home core of their address, and while arbitrary memory operations
-can be delegated, we restrict the use of delegate operations in three
-ways to make them more useful for synchronization. First, we limit each
-task to one outstanding delegate operation to avoid the possibility of
-reordering in the network. Second, we limit delegate operations to
-operate on objects in the 2D address space or objects that fit in a
-single block of the linear address space so they can be satisfied with a
-single network request. Finally, no context switches are allowed while
-the data is being modified. Given these restrictions, we can ensure that
-delegate operations for the same address from multiple requesters are
-always serialized through a single core in the system, providing atomic
-semantics without using atomic operations.
-
-\subsection{Network Aggregation}
+\section{Communication}
 

### 2012-07-23 doc/asplos13/memory.tex
Super redux

--- 2012-07-23 doc/asplos13/grappa.tex
@@
-\section{Grappa} \label{sec:grappa}
-
-\begin{figure}[t]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:grappa} Grappa system overview}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-From the perspective of a software developer, Grappa
-(Figure~\ref{fig:grappa}) provides two main software components: a
-\textbf{tasking system} with lightweight multithreading to tolerate
-latency and workstealing to automate load balance; and a
-\textbf{distributed shared global memory} providing high aggregate
-random access bandwidth for both normal and synchronized operations. We
-now discuss the programmer's view of Grappa's main capabilities,
-including a subset of the Grappa API. In Section~\ref{implementation},
-we discuss the details of their implementation as well as a third
-component, the network communication layer that is key to performance
-but largely invisible to software developers.
-
-\TODO{Should we have a code example here, and walk thru it as we explain the
-components and parts of the API?}
-
-\subsection{Tasks}
-
-%The basic unit of locality in Grappa is the {\em core}. Each core is
-%responsible for a section of the global memory
-
-The basic unit of execution in Grappa is a {\em task}. Tasks are
-small; just a function pointer and arguments. Tasks are only
-queued when they are spawned; later, when resources are free, they are
-allocated a stack, bound to a core, and executed.
-
-Grappa's tasks give up control of their core when when they perform
-long-latency operations, allowing the processor to remain busy while
-waiting for the operation to complete. This is most often done
-implicitly inside a call into Grappa's API, but it can also be done
-explicitly by the programmer using the calls shown in
-Figure~\ref{fig:scheduling}. \TODO{cite UPC split phase reads and
-writes, but include that we provide a way to overlap computation}
-
-All Grappa scheduling is done in userspace, and minimal state is saved,
-in order to minimize the context switch time. In our experiments, we saw
-context switch times as low as \checkme{40ns}. This lightweight context
-switching is Grappa's key enabling feature. It gives us the ability to
-tolerate latency. Given that ability, we are able to \textbf{trade
-latency for throughput}: by {\em increasing} latency in key components
-in the system we are able to increase our aggregate random access
-bandwidth, our synchronization bandwidth, and our ability to tolerate
-load.
-
-\begin{figure}[htbp]
-  \begin{center}
-    \begin{description}\small
-    \item[ \texttt{ yield() } ] \hfill \\
-      Gives up control of core to scheduler, queuing task to be scheduled again soon
-    \item[ \texttt{ suspend() } ] \hfill \\
-      Gives up control of core to scheduler
-    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
-      Enqueues $t$ to be scheduled again soon
-    \end{description}
-    \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:scheduling} Grappa API: scheduling} %{-4ex}}
-    \end{minipage}
-    %\vspace{-3ex}
-  \end{center}
-\end{figure}
-
-\subsection{Expressing parallelism}
-
-A programmer's goal in coding with Grappa should be to express as much
-parallelism as possible without worrying about where it will execute.
-Grappa then chooses where and when to exploit this parallelism,
-scheduling as much work as necessary to tolerate network latencies and
-keep cores busy.
-
-Grappa provides four methods for expressing parallelism, shown in
-Figure~\ref{fig:expressing-parallelism}. The first is explicit task
-spawns. When the programmer identifies work that can be done in
-parallel, the work may be wrapped up in a function and queued with its
-arguments for later execution using a \texttt{spawn}. If the
-programmer wishes to spawn a task on a specific core in the system,
-or at the home core of a particular memory location, Grappa provides a
-\texttt{spawn\_on} call for this purpose.
-
-The next method for expressing parallelism is a parallel for loop, where
-the number of iterations must be known at loop entry. The programmer
-specifies a function pointer along with start and end indices and an
-optional threshold to control parallel overhead. Grappa does {\em
-recursive decomposition} of iterations, similar to Cilk's cilk\_for
-construct~\cite {cilkforimplementation} \comment{could only find slides
-page6 of
-http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-
-Cilk++.pdf    ---- Good enough.  Cite it.  -Mark}.  It generates a
-logarithmically-deep tree of tasks, stopping to execute the loop body
-when the number of iterations is below the required threshold. Finally,
-a programmer may want to run a small piece of code on a particular core
-in the system without waiting for execution resources to be available.
-Grappa provides the \texttt{call\_on} call for this purpose.
-
-\begin{figure}[htbp]
-  \begin{center}
-    \begin{description}\small
-    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
-      Creates a new stealable task
-    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
-      Creates a new private task that will run on a specific core 
-    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
-      Executes iterations of a loop as stealable tasks 
-    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
-      Runs a limited function on a specific core without consuming
-      Grappa execution resources 
-    \end{description}
-    \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism} % \vspace{-4ex}}
-    \end{minipage}
-    %\vspace{-3ex}
-  \end{center}
-\end{figure}
-
-\subsection{Memory}
+\section{Memory}
 
@@
 
-\paragraph{Global memory access} There are two approaches to {\emph
-accessing} global memory. When the programmer expects a computation on
-shared data to have spatial locality to exploit, {\em cache} operations
-may be used. When there is no locality to exploit, {\em delegate}
-operations are used.
-
-\textbf{Explicit caching.} Grappa provides an API to fetch a global
-pointer of any length and return a local pointer to a cached copy of the
-global memory.  Grappa cache operations have the usual read-only and
-read-write variants, along with a write-only variant used to initialize
-data structures. Languages for distributed shared memory systems have
-done optimizations to achieve a similar goal. For example, the UPC
-compiler coalesces struct and array accesses into remote get/put
-\cite{Chen:2005}, and Fortran D compiler's message vectorization hoists
-small messages out of a loop \cite{FortranD:1992}. Caching in Grappa
-additionally provides a mechanism for exploiting temporal locality by
-operating on the data locally.
-
-\textbf{Delegate operations.}
-When the access pattern has low-locality, it is more efficient to modify
-the data on its home core rather than bringing a copy to the requesting
-core and returning it after modification. Delegate operations provide
-this capability. Applications can dispatch computation to be performed
-on individual machine-word sized chunks of global memory to the memory
-system itself (e.g., \emph{fetch-and-add}).  Delegate operations,
-proposed in \cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the primary synchronization method in Grappa.
-
 \begin{figure}[htbp]
@@
 \end{figure}
+
+\paragraph{Global memory access} There are two approaches to {\emph
+accessing} global memory. When the programmer expects a computation on
+shared data to have spatial locality to exploit, {\em cache} operations
+may be used. When there is no locality to exploit, {\em delegate}
+operations are used.
+
+All global addresses in Grappa are 64-bit values. Each byte of global
+memory is associated with a {\em home core}; this core is responsible
+for allocating and modifying that memory. Since today's commodity
+processors support a 48-bit virtual address space \cite{AMD64}, Grappa
+uses the spare bits to distinguish which address space a global address
+belongs to, and to represent the home core of a global address.
+
+Global memory in Grappa is exposed through two distinct, non-overlapping
+address spaces; \textbf{2D global addresses}, optimized for global
+access to node-local data, and \textbf{linear global addresses},
+optimized for high aggregate random access bandwidth to low locality
+data.
+
+\begin{figure}[t]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:memory-structure} Grappa memory structure}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
+\paragraph{2D global addresses} The first global address space is the {\em
+two-dimensional} address space. This uses a traditional PGAS addressing model,
+where each address is a tuple of a rank in the job (or global process ID) and
+an address in that process. The lower 48 bits of the address hold a virtual
+address in the process. The top bit is set to indicates that the reference is
+a 2D address. This leaves 15 bits for network endpoint ID, which limits our
+scalability to $2^{15}$ endpoints.
+
+Any node-local data can be made accessible by other nodes in the system
+by wrapping the address and node ID into a 2D global address. This
+address can then be accessed with a delegate and can also be cached by
+other nodes. At the destination the address is converted into a
+canonical x86 address by replacing the upper bits with the sign-extended
+upper bit of the virtual address. 2D addresses may refer to memory
+allocated from a single processes' heap or from a task's stack.
+Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can
+refer to other cores' memory.
+
+\paragraph{Linear global addresses} The second global address space is
+the {\em linear} address space, so named because nearby addresses can be
+treated as adjacent even if they are stored on different nodes.
+Addresses in the linear address space are partitioned across the cluster
+in a block-cyclic fashion. Choosing the block size involves trading off
+sequential bandwidth against aggregate random access bandwidth. Smaller
+block sizes help spread data across all the memory controllers in the
+cluster, but larger block sizes allow the locality-optimized memory
+controllers to provide increased sequential bandwidth. The block size,
+which is configurable, is typically set to 64 bytes, or the size of a
+single hardware cache line, in order to exploit spatial locality when
+available.
+
+All linear addresses refer to data allocated from a global heap. The
+heap metadata is stored on a single node. Currently all heap operations
+serialize through this node; while this has been sufficient for our
+benchmarks, in the future Grappa will provide parallel performance
+through combining~\cite{MAMA,flatcombining}.
+
+\TODO{Add a figure that depicts both forms of global addresses.}
+
+All global memory operations are implemented with active messages.
+Requesters construct a descriptor holding the current task ID and space
+for a response, issue their request active message to the node given by
+the request address, and suspend themselves. When the request arrives at
+the address' home node, the operation is performed and a response active
+message is sent. When the response arrives at the requesting node, the
+descriptor is filled in with the results of the operation and the
+requesting task is placed back on the ready queue.
+
+\textbf{Explicit caching.} Grappa provides an API to fetch a global
+pointer of any length and return a local pointer to a cached copy of the
+global memory.  Grappa cache operations have the usual read-only and
+read-write variants, along with a write-only variant used to initialize
+data structures. Languages for distributed shared memory systems have
+done optimizations to achieve a similar goal. For example, the UPC
+compiler coalesces struct and array accesses into remote get/put
+\cite{Chen:2005}, and Fortran D compiler's message vectorization hoists
+small messages out of a loop \cite{FortranD:1992}. Caching in Grappa
+additionally provides a mechanism for exploiting temporal locality by
+operating on the data locally.
+
+Under the hood, Grappa performs the mechanics of gathering chunks of
+data from multiple system nodes and presenting a conventional appearing
+linear block of memory as a local pointer into a cache. The strategy
+employed is to issue all the constituent requests of a cache access
+request and then yield until all responses have occurred.  Currently,
+Grappa caches are \emph{not} coherent, requiring the programmer to
+maintain consistent access to data.  Future work will develop a software
+directory based coherence scheme to simplify consistent access to global
+data.
+
+\textbf{Delegate operations.}
+When the access pattern has low-locality, it is more efficient to modify
+the data on its home core rather than bringing a copy to the requesting
+core and returning it after modification. Delegate operations provide
+this capability. Applications can dispatch computation to be performed
+on individual machine-word sized chunks of global memory to the memory
+system itself (e.g., \emph{fetch-and-add}).  Delegate operations,
+proposed in \cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the primary synchronization method in Grappa.
+
+Delegate operations are always executed at the home core of their
+address, and while arbitrary memory operations can be delegated, we
+restrict the use of delegate operations in three ways to make them more
+useful for synchronization. First, we limit each task to one outstanding
+delegate operation to avoid the possibility of reordering in the
+network. Second, we limit delegate operations to operate on objects in
+the 2D address space or objects that fit in a single block of the linear
+address space so they can be satisfied with a single network request.
+Finally, no context switches are allowed while the data is being
+modified. Given these restrictions, we can ensure that delegate
+operations for the same address from multiple requesters are always
+serialized through a single core in the system, providing atomic
+semantics without using atomic operations.
 

### 2012-07-23 doc/asplos13/overview.tex
Super redux

\section{Grappa Overview}

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa system overview}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

Grappa (Figure~\ref{fig:grappa}) provides three main software components:
a \textbf{tasking system} with lightweight multithreading to tolerate
latency and workstealing to automate load balance;
a \textbf{distributed shared global memory} providing high aggregate
random access bandwidth for both normal and synchronized operations;
and a \textbf{communication layer}, which is largely invisible to the software developer but helps to improve network performance when applications only read and write small pieces of data.
In the next three sections we describe both the programmer's view of Grappa's main capabilities and how they are implemented.

### 2012-07-23 doc/asplos13/tasks.tex
Super redux

--- 2012-07-22 doc/asplos13/implementation.tex
@@
-\section{Implementation} \label{sec:implementation}
+\section{Tasks}
 
-We now discuss implementation details for the major system components discussed in Section~\ref{sec:grappa}.
+The basic unit of execution in Grappa is a {\em task}. Tasks are
+small; just a function pointer and arguments. Tasks are only
+queued when they are spawned; later, when resources are free, they are
+allocated a stack, bound to a core, and executed.
 
-\subsection{Tasks}
+Grappa's tasks give up control of their core when when they perform
+long-latency operations, allowing the processor to remain busy while
+waiting for the operation to complete. This is most often done
+implicitly inside a call into Grappa's API, but it can also be done
+explicitly by the programmer using the calls shown in
+Figure~\ref{fig:scheduling}. \TODO{cite UPC split phase reads and
+writes, but include that we provide a way to overlap computation}
+
+All Grappa scheduling is done in userspace, and minimal state is saved,
+in order to minimize the context switch time. In our experiments, we saw
+context switch times as low as \checkme{40ns}. This lightweight context
+switching is Grappa's key enabling feature. It gives us the ability to
+tolerate latency. Given that ability, we are able to \textbf{trade
+latency for throughput}: by {\em increasing} latency in key components
+in the system we are able to increase our aggregate random access
+bandwidth, our synchronization bandwidth, and our ability to tolerate
+load.
+
+\begin{figure}[htbp]
+  \begin{center}
+    \begin{description}\small
+    \item[ \texttt{ yield() } ] \hfill \\
+      Gives up control of core to scheduler, queuing task to be scheduled again soon
+    \item[ \texttt{ suspend() } ] \hfill \\
+      Gives up control of core to scheduler
+    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
+      Enqueues $t$ to be scheduled again soon
+    \end{description}
+    \begin{minipage}{0.95\columnwidth}
+      \caption{\label{fig:scheduling} Grappa API: scheduling} %{-4ex}}
+    \end{minipage}
+    %\vspace{-3ex}
+  \end{center}
+\end{figure}
+
+\subsection{Expressing parallelism}
+
+A programmer's goal in coding with Grappa should be to express as much
+parallelism as possible without worrying about where it will execute.
+Grappa then chooses where and when to exploit this parallelism,
+scheduling as much work as necessary to tolerate network latencies and
+keep cores busy.
+
+Grappa provides four methods for expressing parallelism, shown in
+Figure~\ref{fig:expressing-parallelism}. The first is explicit task
+spawns. When the programmer identifies work that can be done in
+parallel, the work may be wrapped up in a function and queued with its
+arguments for later execution using a \texttt{spawn}. If the
+programmer wishes to spawn a task on a specific core in the system,
+or at the home core of a particular memory location, Grappa provides a
+\texttt{spawn\_on} call for this purpose.
+
+The next method for expressing parallelism is a parallel for loop, where
+the number of iterations must be known at loop entry. The programmer
+specifies a function pointer along with start and end indices and an
+optional threshold to control parallel overhead. Grappa does {\em
+recursive decomposition} of iterations, similar to Cilk's cilk\_for
+construct~\cite {cilkforimplementation} \comment{could only find slides
+page6 of
+http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-
+Cilk++.pdf    ---- Good enough.  Cite it.  -Mark}.  It generates a
+logarithmically-deep tree of tasks, stopping to execute the loop body
+when the number of iterations is below the required threshold. Finally,
+a programmer may want to run a small piece of code on a particular core
+in the system without waiting for execution resources to be available.
+Grappa provides the \texttt{call\_on} call for this purpose.
+
+\begin{figure}[htbp]
+  \begin{center}
+    \begin{description}\small
+    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
+      Creates a new stealable task
+    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
+      Creates a new private task that will run on a specific core 
+    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
+      Executes iterations of a loop as stealable tasks 
+    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
+      Runs a limited function on a specific core without consuming
+      Grappa execution resources 
+    \end{description}
+    \begin{minipage}{0.95\columnwidth}
+      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism} % \vspace{-4ex}}
+    \end{minipage}
+    %\vspace{-3ex}
+  \end{center}
+\end{figure}
+
+\subsection{Implementation tasks}
 
@@
 use \checkme{128} as the parameter for all experiments.
-
-\subsection{Memory}
-
-\TODO{Is there an off by 1 error in here?
-Need to read the AMD64 docs carefully to see if we can sign extend the top bit of the 48 bits of real virtual address space, or if we need to maintain bit 49.
--M}
-
-Grappa implements a software distributed shared memory. All global
-addresses in Grappa are 64-bit values. Each byte of global memory is
-associated with a {\em home core}; this core is responsible for
-allocating and modifying that memory. Since today's commodity processors
-support a 48-bit virtual address space \cite{AMD64}, Grappa uses the
-spare bits to distinguish which address space a global address belongs
-to, and to represent the home core of a global address.
-
-Global memory in Grappa is exposed through two distinct, non-overlapping
-address spaces; \textbf{2D global addresses}, optimized for global
-access to node-local data, and \textbf{linear global addresses},
-optimized for high aggregate random access bandwidth to low locality
-data.
-
-\begin{figure}[t]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:memory-structure} Grappa memory structure}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-\paragraph{2D global addresses} The first global address space is the {\em
-two-dimensional} address space. This uses a traditional PGAS addressing model,
-where each address is a tuple of a rank in the job (or global process ID) and
-an address in that process. The lower 48 bits of the address hold a virtual
-address in the process. The top bit is set to indicates that the reference is
-a 2D address. This leaves 15 bits for network endpoint ID, which limits our
-scalability to $2^{15}$ endpoints.
-
-Any node-local data can be made accessible by other nodes in the system
-by wrapping the address and node ID into a 2D global address. This
-address can then be accessed with a delegate and can also be cached by
-other nodes. At the destination the address is converted into a
-canonical x86 address by replacing the upper bits with the sign-extended
-upper bit of the virtual address. 2D addresses may refer to memory
-allocated from a single processes' heap or from a task's stack.
-Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can
-refer to other cores' memory.
-
-\paragraph{Linear global addresses} The second global address space is
-the {\em linear} address space, so named because nearby addresses can be
-treated as adjacent even if they are stored on different nodes.
-Addresses in the linear address space are partitioned across the cluster
-in a block-cyclic fashion. Choosing the block size involves trading off
-sequential bandwidth against aggregate random access bandwidth. Smaller
-block sizes help spread data across all the memory controllers in the
-cluster, but larger block sizes allow the locality-optimized memory
-controllers to provide increased sequential bandwidth. The block size,
-which is configurable, is typically set to 64 bytes, or the size of a
-single hardware cache line, in order to exploit spatial locality when
-available.
-
-All linear addresses refer to data allocated from a global heap. The
-heap metadata is stored on a single node. Currently all heap operations
-serialize through this node; while this has been sufficient for our
-benchmarks, in the future Grappa will provide parallel performance
-through combining~\cite{MAMA,flatcombining}.
-
-\TODO{Add a figure that depicts both forms of global addresses.}
-
-All global memory operations are implemented with active messages.
-Requesters construct a descriptor holding the current task ID and space
-for a response, issue their request active message to the node given by
-the request address, and suspend themselves. When the request arrives at
-the address' home node, the operation is performed and a response active
-message is sent. When the response arrives at the requesting node, the
-descriptor is filled in with the results of the operation and the
-requesting task is placed back on the ready queue.
-
-\paragraph{Explicit caching.} Under the hood, Grappa performs the
-mechanics of gathering chunks of data from multiple system nodes and
-presenting a conventional appearing linear block of memory as a local
-pointer into a cache. The strategy employed is to issue all the
-constituent requests of a cache access request and then yield until all
-responses have occurred.  Currently, Grappa caches are \emph{not}
-coherent, requiring the programmer to maintain consistent access to
-data.  Future work will develop a software directory based coherence
-scheme to simplify consistent access to global data.
-
-\paragraph{Delegates} Delegate operations are used for low-locality
-accesses and synchronization. Their modifications are always executed at
-the home core of their address, and while arbitrary memory operations
-can be delegated, we restrict the use of delegate operations in three
-ways to make them more useful for synchronization. First, we limit each
-task to one outstanding delegate operation to avoid the possibility of
-reordering in the network. Second, we limit delegate operations to
-operate on objects in the 2D address space or objects that fit in a
-single block of the linear address space so they can be satisfied with a
-single network request. Finally, no context switches are allowed while
-the data is being modified. Given these restrictions, we can ensure that
-delegate operations for the same address from multiple requesters are
-always serialized through a single core in the system, providing atomic
-semantics without using atomic operations.
-
-\subsection{Network Aggregation}
-
-In order to mitigate the low message injection rate limits of commodity
-networks, Grappa's communication stack has two layers: one for
-user-level messages and one for network-level messages.
-
-At the upper layer, Grappa implements asynchronous active messages
-\cite{vonEicken92}. Each message consists of a function pointer, an
-optional argument payload, and an optional data payload. When a task
-sends a message, the message is copied to a send queue associated with
-the message's destination and the task continues execution.
-
-Grappa's lower networking layer aggregates the upper layer's messages
-to improve performance. Commodity networks including infiniband
-achieves their peak bisection bandwidth \emph{only} when the packet
-sizes are relatively large---on the order of multiple kilobytes. The
-reason for this discrepancy is the combination of overheads associated
-with handling each packet (in terms of bytes that form the actual
-packet, processing time at the card and processing on the
-CPU within the driver stack). Our measurements confirm manufacturers
-published data~\cite{infinibandbandwidth}, that with this packet size
-the bisection bandwidth is only a small fraction, less than
-~\checkme{5\%} of the peak bisection bandwidth.
-
-In our experiments the vast majority of requests were smaller than
-\checkme{255} bytes \TODO{more precise? median?}, far too small to make
-efficient use of the network. To make the best use of the network, we
-must convert our small messages into large ones
-(Section~\ref{sec:results-aggregation}).  When a task sends a message,
-it is not immediately sent, but rather placed in a queue specific to the
-destination.
-
-There are three situations in which a queue of aggregated messages is
-sent. First, each queue has a message size threshold, chosen to give
-reasonable network performance (\checkme{2048 bytes}). If the size in
-bytes of a queue is above the threshold, the contents of the queue are
-sent immediately. Second, each queue has a wait time threshold
-(\checkme{200ms}). If the oldest message in a queue has been waiting
-longer than this threshold, the contents of the queue are sent
-immediately, even if the queue size is lower than the message size
-threshold.  Third, queues may be explicitly flushed in situations where
-the programmer wants to minimize the latency of a message at the cost of
-bandwidth utilization.
-
-The network layer is serviced by polling. Periodically when a context
-switch occurs, the Grappa scheduler switches to the network polling
-thread. This thread has three responsibilities. First, it polls the
-lower-level network layer to ensure it makes progress. Second, it
-deaggregates received messages and executes active message
-handlers. Third, it checks to see if any aggregation queues have
-messages that have been waiting longer than the threshold; if so, it
-sends them.
-
-Underneath the aggregation layer, Grappa uses the \gasnet~communication
-library~\cite{gasnet} to actually move data. All interprocess
-communication, whether on or off a cluster node, is handled by the
-\gasnet~library. \gasnet~is able to take advantage of many communication
-mechanisms, including ethernet and infiniband between nodes, as well as
-shared memory within a node.
-
-Some networks provide access to a remote machine's memory directly. This
-would seem to be a good fit for a programming model focused on global
-shared memory, but in fact we do not use it. In our experiments, we
-found that RDMA operations are subject to the same message rate
-limitations as all other messages on these cards, and thus using raw
-RDMA operations for our small messages would make inefficient use of
-bandwidth. Instead, we implement remote memory operations with active
-messages. A byproduct of this design decision is that Grappa is not
-limited to RDMA-capable networks.

### 2012-07-23 doc/asplos13/memory.tex
First crack at normalizing memory after the redux

@@
 Applications written for Grappa utilize two forms of memory: local and
-global.
-
-Local memory is local to a single core in the system.  Accesses occur
+global.  Local memory is local to a single core in the system.  Accesses occur
 through conventional pointers.  The compiler emits an access and the
@@
 
-\paragraph{Global memory allocation}
-Grappa provides two methods for {\emph storing} data in the global memory. The
-first is a distributed heap striped across all the machines in the
-system. The \texttt{global\_malloc} and \texttt{global\_free} calls
-are used to allocate and deallocate memory in the global heap; on
-allocation, a global pointer is returned. Grappa also allows any local
-data on a core's stacks or heap to be exported to the global address
-space to be made accessible to other cores across the system.
+\paragraph{Global memory addressing} Grappa provides two methods for
+storing data in the global memory. The first is a distributed heap
+striped across all the machines in the system in a block cyclic fashion.
+The \texttt{global\_malloc} and \texttt{global\_free} calls are used to
+allocate and deallocate memory in the global heap.  Addresses to memory
+in the global heap use \textbf{linear addresses}.  Choosing the block
+size involves trading off sequential bandwidth against aggregate random
+access bandwidth. Smaller block sizes help spread data across all the
+memory controllers in the cluster, but larger block sizes allow the
+locality-optimized memory controllers to provide increased sequential
+bandwidth. The block size, which is configurable, is typically set to 64
+bytes, or the size of a single hardware cache line, in order to exploit
+spatial locality when available. The heap metadata is stored on a single
+node. Currently all heap operations serialize through this node; while
+this has been sufficient for our benchmarks, in the future Grappa will
+provide parallel performance through combining~\cite{MAMA,flatcombining}.
+
+Grappa also allows any local data on a core's stacks or heap to be
+exported to the global address space to be made accessible to other
+cores across the system. Addresses to global memory allocated in this
+way use \textbf{2D global addresses}.  This uses a traditional PGAS
+addressing model, where each address is a tuple of a rank in the job (or
+global process ID) and an address in that process. The lower 48 bits of
+the address hold a virtual address in the process. The top bit is set to
+indicates that the reference is a 2D address (as opposed to linear
+address). This leaves 15 bits for network endpoint ID, which limits our
+scalability to $2^{15}$ endpoints. Any node-local data can be made
+accessible by other nodes in the system by wrapping the address and node
+ID into a 2D global address. This address can then be accessed with a
+delegate and can also be cached by other nodes. At the destination the
+address is converted into a canonical x86 address by replacing the upper
+bits with the sign-extended upper bit of the virtual address. 2D
+addresses may refer to memory allocated from a single processes' heap or
+from a task's stack. Figure~\ref{fig:memory-structure} shows how 2D and
+linear addresses can refer to other cores' memory.
 
@@
 
-\paragraph{Global memory access} There are two approaches to {\emph
-accessing} global memory. When the programmer expects a computation on
-shared data to have spatial locality to exploit, {\em cache} operations
-may be used. When there is no locality to exploit, {\em delegate}
-operations are used.
-
-All global addresses in Grappa are 64-bit values. Each byte of global
-memory is associated with a {\em home core}; this core is responsible
-for allocating and modifying that memory. Since today's commodity
-processors support a 48-bit virtual address space \cite{AMD64}, Grappa
-uses the spare bits to distinguish which address space a global address
-belongs to, and to represent the home core of a global address.
-
-Global memory in Grappa is exposed through two distinct, non-overlapping
-address spaces; \textbf{2D global addresses}, optimized for global
-access to node-local data, and \textbf{linear global addresses},
-optimized for high aggregate random access bandwidth to low locality
-data.
-
 \begin{figure}[t]
@@
 
-\paragraph{2D global addresses} The first global address space is the {\em
-two-dimensional} address space. This uses a traditional PGAS addressing model,
-where each address is a tuple of a rank in the job (or global process ID) and
-an address in that process. The lower 48 bits of the address hold a virtual
-address in the process. The top bit is set to indicates that the reference is
-a 2D address. This leaves 15 bits for network endpoint ID, which limits our
-scalability to $2^{15}$ endpoints.
-
-Any node-local data can be made accessible by other nodes in the system
-by wrapping the address and node ID into a 2D global address. This
-address can then be accessed with a delegate and can also be cached by
-other nodes. At the destination the address is converted into a
-canonical x86 address by replacing the upper bits with the sign-extended
-upper bit of the virtual address. 2D addresses may refer to memory
-allocated from a single processes' heap or from a task's stack.
-Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can
-refer to other cores' memory.
-
-\paragraph{Linear global addresses} The second global address space is
-the {\em linear} address space, so named because nearby addresses can be
-treated as adjacent even if they are stored on different nodes.
-Addresses in the linear address space are partitioned across the cluster
-in a block-cyclic fashion. Choosing the block size involves trading off
-sequential bandwidth against aggregate random access bandwidth. Smaller
-block sizes help spread data across all the memory controllers in the
-cluster, but larger block sizes allow the locality-optimized memory
-controllers to provide increased sequential bandwidth. The block size,
-which is configurable, is typically set to 64 bytes, or the size of a
-single hardware cache line, in order to exploit spatial locality when
-available.
-
-All linear addresses refer to data allocated from a global heap. The
-heap metadata is stored on a single node. Currently all heap operations
-serialize through this node; while this has been sufficient for our
-benchmarks, in the future Grappa will provide parallel performance
-through combining~\cite{MAMA,flatcombining}.
-
-\TODO{Add a figure that depicts both forms of global addresses.}
-
-All global memory operations are implemented with active messages.
-Requesters construct a descriptor holding the current task ID and space
-for a response, issue their request active message to the node given by
-the request address, and suspend themselves. When the request arrives at
-the address' home node, the operation is performed and a response active
-message is sent. When the response arrives at the requesting node, the
-descriptor is filled in with the results of the operation and the
-requesting task is placed back on the ready queue.
+\paragraph{Global memory access} There are two general approaches
+Grappa applications use to {\emph access} global memory. When the
+programmer expects a computation on shared data to have spatial locality
+to exploit, {\em cache} operations may be used. When there is no
+locality to exploit, {\em delegate} operations are used.
 
@@
 employed is to issue all the constituent requests of a cache access
-request and then yield until all responses have occurred.  Currently,
-Grappa caches are \emph{not} coherent, requiring the programmer to
-maintain consistent access to data.  Future work will develop a software
-directory based coherence scheme to simplify consistent access to global
-data.
+request  (as Active Messages) and then yield until all responses have
+occurred.  Currently, Grappa caches are \emph{not} coherent, requiring
+the programmer to maintain consistent access to data.  Future work will
+develop a software directory based coherence scheme to simplify
+consistent access to global data.
 
-\textbf{Delegate operations.}
-When the access pattern has low-locality, it is more efficient to modify
-the data on its home core rather than bringing a copy to the requesting
-core and returning it after modification. Delegate operations provide
-this capability. Applications can dispatch computation to be performed
-on individual machine-word sized chunks of global memory to the memory
-system itself (e.g., \emph{fetch-and-add}).  Delegate operations,
-proposed in \cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the primary synchronization method in Grappa.
+\textbf{Delegate operations.} When the access pattern has low-locality,
+it is more efficient to modify the data on its home core rather than
+bringing a copy to the requesting core and returning it after
+modification. Delegate operations provide this capability. Applications
+can dispatch computation to be performed on individual machine-word
+sized chunks of global memory to the memory system itself (e.g.,
+\emph{fetch-and-add}).  Delegate operations, proposed in
+\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the
+primary synchronization method in Grappa.
 
@@
 
-\paragraph{Example}
-
 \begin{figure*}[htb]
@@
 
-Figure~\ref{fig:delegate-cache} shows an example of delegate and cache
-operations. In step 1, a core allocates an array in the global
-heap. It then spawns two tasks on remote cores to increment elements
-of the array.
+\paragraph{Example} Figure~\ref{fig:delegate-cache} shows an example of
+delegate and cache operations. In step 1, a core allocates an array in
+the global heap. It then spawns two tasks on remote cores to increment
+elements of the array.
 

### 2012-07-23 doc/asplos13/overview.tex

@@
 latency and workstealing to automate load balance;
-a \textbf{distributed shared global memory} providing high aggregate
+a \textbf{distributed shared memory} (DSM) providing high aggregate
 random access bandwidth for both normal and synchronized operations;
-and a \textbf{communication layer}, which is largely invisible to the software developer but helps to improve network performance when applications only read and write small pieces of data.
+and a \textbf{communication layer}, which is largely invisible to the software developer but helps to improve network performance when applications read and write only small pieces of data.
 In the next three sections we describe both the programmer's view of Grappa's main capabilities and how they are implemented.

### 2012-07-23 doc/asplos13/tasks.tex
Revised section Tasks

@@
 
-The basic unit of execution in Grappa is a {\em task}. Tasks are
-small; just a function pointer and arguments. Tasks are only
-queued when they are spawned; later, when resources are free, they are
-allocated a stack, bound to a core, and executed.
+The basic unit of execution in Grappa is a {\em task}. Each task is
+represented by a function pointer and its arguments and is enqueued,
+but not instantiated, when spawned.  Later, when resources are free,
+the task is allocated a stack, bound to a core, and executed.
 
-Grappa's tasks give up control of their core when when they perform
-long-latency operations, allowing the processor to remain busy while
-waiting for the operation to complete. This is most often done
-implicitly inside a call into Grappa's API, but it can also be done
-explicitly by the programmer using the calls shown in
-Figure~\ref{fig:scheduling}. \TODO{cite UPC split phase reads and
+During execution, a task yields control of its core whenever it performs
+a long-latency operation, allowing the processor to remain busy while
+waiting for the operation to complete.  In addition,  a programmer can
+direct scheduling explicitly via the Grappa API calls shown
+in Figure~\ref{fig:scheduling}.
+\TODO{cite UPC split phase reads and
 writes, but include that we provide a way to overlap computation}
 
-All Grappa scheduling is done in userspace, and minimal state is saved,
-in order to minimize the context switch time. In our experiments, we saw
-context switch times as low as \checkme{40ns}. This lightweight context
-switching is Grappa's key enabling feature. It gives us the ability to
-tolerate latency. Given that ability, we are able to \textbf{trade
-latency for throughput}: by {\em increasing} latency in key components
-in the system we are able to increase our aggregate random access
-bandwidth, our synchronization bandwidth, and our ability to tolerate
-load.
+To minimize yield overhead, all Grappa scheduling is done in
+user-space and saves minimal state, leading to context switch times as
+low as \checkme{40ns} even when switching amongst thousands of tasks.
+This fast context switching amongst many tasks is crucial to Grappa's
+effectiveness for two reasons.  First, latency can be tolerated only
+when the net cost of context switching is dwarfed by the loss incurred
+by the alternative: stalling.  Second, because we can switch amongst
+so many tasks efficiently, we can \textbf{trade latency for throughput}:
+by {\em increasing} latency in key components of the
+system we are able to increase our aggregate random access bandwidth,
+our synchronization bandwidth, and our ability to tolerate load
+imbalance.
+\TODO{check to see we actually explain these three somewhere.
+Ie, that aggregating increases throughput and latency, that delegating synchronization increases the rate and the latency at which we can eg atomically increment, and that stealing work increases the latency of some tasks (eg when their data is largely at the originating node) but provides greater throughput overall.}
 
@@
     \item[ \texttt{ yield() } ] \hfill \\
-      Gives up control of core to scheduler, queuing task to be scheduled again soon
+      Yields core to scheduler, enqueuing caller to be scheduled again soon
     \item[ \texttt{ suspend() } ] \hfill \\
-      Gives up control of core to scheduler
+      Yields core to scheduler, enqueuing caller only once another task calls wake
     \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
-      Enqueues $t$ to be scheduled again soon
+      Enqueues some other task $t$ to be scheduled again soon
     \end{description}

### 2012-07-23 doc/asplos13/tasks.tex
modified Expressing Parallelism but will need another pass.

@@
 
-A programmer's goal in coding with Grappa should be to express as much
-parallelism as possible without worrying about where it will execute.
+Grappa programmers focus on expressing as much
+parallelism as possible without concern for where it will execute.
 Grappa then chooses where and when to exploit this parallelism,
-scheduling as much work as necessary to tolerate network latencies and
-keep cores busy.
+scheduling as much work as is necessary on each core to keep it busy
+in the presence of system latencies and task dependences.
 
 Grappa provides four methods for expressing parallelism, shown in
-Figure~\ref{fig:expressing-parallelism}. The first is explicit task
-spawns. When the programmer identifies work that can be done in
-parallel, the work may be wrapped up in a function and queued with its
-arguments for later execution using a \texttt{spawn}. If the
-programmer wishes to spawn a task on a specific core in the system,
-or at the home core of a particular memory location, Grappa provides a
-\texttt{spawn\_on} call for this purpose.
+Figure~\ref{fig:expressing-parallelism}:
 
-The next method for expressing parallelism is a parallel for loop, where
-the number of iterations must be known at loop entry. The programmer
+First, when the programmer identifies work
+that can be done in parallel, the work may be wrapped up in a function
+and queued with its arguments for later execution using a
+\texttt{spawn}.
+
+Second, a programmer can use \texttt{spawn\_on} to spawn a task on a
+specific core in the system or at the home core of a particular memory
+location.
+
+Third, the programmer can invoke a parallel for loop, provided that
+the trip count is known at loop entry. The programmer
 specifies a function pointer along with start and end indices and an
@@
 logarithmically-deep tree of tasks, stopping to execute the loop body
-when the number of iterations is below the required threshold. Finally,
-a programmer may want to run a small piece of code on a particular core
+when the number of iterations is below the required threshold.
+
+Fourth, a programmer may want to run a small piece of code on a particular core
 in the system without waiting for execution resources to be available.

### 2012-07-23 doc/asplos13/relatedwork.tex
added todos to related
another pass on related work. still needs generality claims

@@
 
-\paragraph{Comparable frameworks}
-Distributed graph processing frameworks like Pregel \cite{pregel:2010} and
-Distributed GraphLab \cite{distgraphlab:vldb12} share similar goals. Pregel
-adopts a bulk-synchronous parallel (BSP) execution model, which makes it
-inefficient on workloads that could prioritize vertices. GraphLab, on the
-other hand, schedules vertex computations individually, allowing
-prioritization, which gives faster convergence in a variety of iterative
-algorithms.\TODO{GraphLab is not general. Myers will expand on that here. Make
-a strong point about generality.} Grappa also supports dynamic parallelism
-with asynchronous execution, but parallelism is expressed as tasks or loop
-iterations.\TODO{Should RAMcloud go here?}.
+\paragraph{Comparable frameworks} Distributed graph processing frameworks like
+Pregel \cite{pregel:2010} and Distributed GraphLab \cite{distgraphlab:vldb12}
+share similar goals as Grappa. Pregel adopts a bulk-synchronous parallel (BSP)
+execution model, which makes it inefficient on workloads that could prioritize
+vertices. GraphLab, on the other hand, schedules vertex computations
+individually, allowing prioritization, which gives faster convergence in a
+variety of iterative algorithms.\TODO{GraphLab is not general. Myers will
+expand on that here. Make a strong point about generality.} Grappa also
+supports dynamic parallelism with asynchronous execution, but parallelism is
+expressed as tasks or loop iterations.\TODO{Should RAMcloud go here?}.
 
-\paragraph{Global memory}
-Grappa implements a software distributed shared memory DSM system. Many
-traditional software DSM systems are page based~\cite{Treadmarks,munin} and
-aim to hide the fact that they are built in software from applications by
-exploiting the processor's paging mechanisms. Instead Grappa, like other
-partitioned global address space (PGAS) models, implements its DSM at the
-language, rather than system level. Languages such as Chapel \cite{Chapel},
-X10 \cite{X10}, and UPC \cite{UPC} make accesses to shared structures look
-like normal memory references. As we describe later, Grappa chooses a middle
-ground, where global addresses are explicit in the API and local accesses are
-emitted conventionally by the compiler. Similar to DSM systems Grappa provides
-a caching and coherence model to ensure high performance and consistent
-results. Cache operations are exposed to in the API, however, allowing
-software to optimize their usage.
+\paragraph{Global memory} Grappa includes a custom implementation of a
+software distributed shared memory (DSM) system. Many traditional software DSM
+systems are page based~\cite{Treadmarks,munin} and aim to hide the fact that
+they are built in software from applications by exploiting the processor's
+paging mechanisms, therefore relying heavily on locality. Instead, Grappa,
+like other partitioned global address space (PGAS) models, implements its DSM
+at the language, rather than system level. Languages such as Chapel
+\cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC} make accesses to shared
+structures look like normal memory references. As we describe later, Grappa
+chooses a middle ground, where global addresses are explicit in the API and
+local accesses are emitted conventionally by the compiler. Similar to DSM
+systems, Grappa provides a consistency model that ensure high performance and
+with good programmability properties. % Caching is Cache operations are explicit in the  exposed to in the
+% API, however, allowing software to optimize their usage.
+
+We decided to build a custom distributed shared memory implementation, as
+opposed to using an existing implementation, for two key reasons: (1) in order
+to support the large number of tasks required for latency tolerance, the
+implementation of the global memory system is tightly coupled to the task
+scheduler; and (2) since commodity networks only do efficient RDMA or ordinary
+transmission of small messages, the implementation of the global memory system
+must make use of the network message aggregator.
+
+%
 

### 2012-07-23 doc/asplos13/tasks.tex
cite amd64 abi and qthreads mtgl; should that be in background section?

@@
 treat context switches as function calls, saving and restoring only the
-callee-saved state as specified in the x86-64 ABI \cite{someone}. This
+callee-saved state as specified in the x86-64 ABI \cite{amd64:abi:2012}. This
 involves saving six general-purpose 64-bit registers and the stack

### 2012-07-23 doc/asplos13/evaluation.tex

@@
     \label{fig:uts_compare}
-\end{figure}
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/uts_xmt_temp.pdf}
-    \end{center}
-    \caption{This is XMT uts result, \TODO{overlay the plots}}
-    \label{fig:uts_compare_temp}
 \end{figure}
@@
 
-\TODO{incorporate UTS chunk size after rerun}
-it's important to steal multiple tasks to 
-amortize cost of stealing
-spread work out quickly in large system
+It is important to steal multiple tasks at a time to both amortize the
+cost of stealing over the network and to spread out work quickly in a
+large system. Figure~\ref{fig:ut_chunksize} shows performance and
+stealing statistics for UTS on \checkme{30} nodes as we increase the stealing chunk size. Recall
+that a thief will take a number of tasks equal to the minimum of half
+the available work or the chunk size; steals fail only when the victim
+has fewer than 2 available tasks. As the scheduler is allowed to
+steal more work beyond 1 task, we see that performance increases up to 6x. This
+shows that the heuristic of stealing the oldest task from victims is
+insufficient alone when a tree-structured computation is imbalanced,
+as observed in \cite{UTS}. By observing sampled state in the execution
+trace, we find that a chunk size as low as 1 allows stealing to
+spread the load evenly across the cluster but cores spend much time
+underutilized as multiple workers wait for steal replies that
+utlimately return little new work.
 
-there are times when it's useful to steal lots of stuff, even though
-we steal only a few things most of the time.
+Performance plateaus before maximum steal amount is limited by the
+size of the victim's task queue. This indicates that artificially limiting steals
+to \checkme{128} tasks does not limit performance. Although a lower
+chunk size limits how quickly work spreads, for sufficient chunk size,
+the heuristic of stealing the oldest tasks from victims in tree-based computations allows for
+stolen work to expand quickly.
 
@@
 
-amortizing cost of stealing
-spreading work out more
-most of the time we still little  (about 10 avg) but sometimes a lot
-(difference with BFS)
+\TODO{(difference with BFS)}
 
@@
 
-We see benefit to limiting the amount of parallelism created by a
-recursive loop decoposition. The decomposition threshold parameter the
-runtime when to stop creating new tasks and just execute iterations
-serially. This allows us to amoritize the overhead of task creation
-and potentially exploit locality when data for adjacent iterations is
+Parallel overhead---in the form of context switches, task spawns, and
+synchronization---can reduce the performance benefit of parallelism.
+Grappa sees a benefit to limiting the amount of parallelism created by
+a recursive loop decomposition. The parallel loop threshold (``parallel
+granularity'') parameter tells the runtime when to stop creating new tasks and just execute iterations
+sequentially. This allows us to amoritize the overhead of task
+creation. In addition, assigning sequential iterations to a single
+task provides the potential to exploit locality when data for adjacent iterations is
 also adjacent in memory.
@@
     \caption{Performance of UTS-Mem with varying parallel loop
-        threshol, run with 30 nodes, 6 cores per node, 4000 workers,
+        threshold, run with 30 nodes, 6 cores per node, 4000 workers,
     \checkme{6M flush ticks}}

### 2012-07-23 doc/asplos13/intro.tex

@@
 
-Our goals is challenging for two key reasons:
+Our goal is challenging for two key reasons:
 
@@
 to tolerate the latency of data movement. The fully custom Tera
-MTA-2~\cite{mta-2} system is a classic example of supporting irregular
+MTA-2~\cite{tera} \TODO{this is mta-1 citation} system is a classic example of supporting irregular
 applications by using concurrency to hide latencies. It had a large
@@
 shelf user-mode infiniband device driver stack~\cite{Melonox?} is used
-for network interfacing. MPI~\cite{mpi} is used for process setup and
-tear down. GASnet~\cite{gasnet} is used as the underlying mechanism for
+for network interfacing. MPI is used for process setup and
+tear down. GASnet~\cite{gasnet:spec} is used as the underlying mechanism for
 remote memory reads and writes using active message invocations. To this

### 2012-07-23 doc/asplos13/relatedwork.tex

@@
 at the language, rather than system level. Languages such as Chapel
-\cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC} make accesses to shared
+\cite{Chamberlain:2007}, X10 \cite{X10:2005}, and UPC \cite{upc:2005} make accesses to shared
 structures look like normal memory references. As we describe later, Grappa
@@
 Grappa uses multithreading to tolerate memory latency. This is a well known
-technique. Hardware implementations include the Tera MTA \cite{Tera}, Cray XMT
-\cite{XMT}, Simultaneous multithreading \cite{SMT}, MIT Alewife
-\cite{Alewife}, Cyclops \cite{Cyclops}, and even GPUs \cite{fatahalian}. As we
+technique. Hardware implementations include the Tera MTA \cite{tera:mta1}, Cray XMT
+\cite{feo:xmt}, Simultaneous multithreading \cite{tullsen:smt}, MIT Alewife
+\cite{agarwal:alewife}, Cyclops \cite{almasi:cyclops}, and even GPUs \cite{gpus}. As we
 describe in this paper, Grappa use a lightweight user-mode task scheduler to

### 2012-07-23 doc/asplos13/memory.tex
Moved example text to figure

@@
 semantics without using atomic operations. 
+Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
+cache operations interact.
 
@@
   \includegraphics[width=1.5\columnwidth]{figs/delegate-cache}
-\begin{minipage}{1.5\columnwidth}
-  \caption{\label{fig:delegate-cache} Delegation and cache example.}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure*}
-
-\paragraph{Example} Figure~\ref{fig:delegate-cache} shows an example of
-delegate and cache operations. In step 1, a core allocates an array in
+\begin{minipage}{1.9\columnwidth}
+  \caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in
 the global heap. It then spawns two tasks on remote cores to increment
-elements of the array.
-
-The first task increments two elements of the array using cache
+elements of the array.  The first task increments two elements of the array using cache
 operations. In step 2, the task is invoked. A cache request is issued
@@
 sent back to the home node. Acknoledgements are returned in step 6 so
-the task knows when the writes are complete.
-
-The second task increments an element of the array with a delegate
+the task knows when the writes are complete. The second task increments an element of the array with a delegate
 operation. In step 7, the task is invoked. A delegate request is sent
@@
 the previous value of the array element.
+}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure*}

### 2012-07-23 doc/asplos13/tasks.tex

@@
 The basic unit of execution in Grappa is a {\em task}. Each task is
-represented by a function pointer and its arguments and is enqueued,
-but not instantiated, when spawned.  Later, when resources are free,
+represented by a function pointer and its arguments and is enqueued when
+spawned, but not instantiated.  Later, when resources are free,
 the task is allocated a stack, bound to a core, and executed.
@@
 
-To minimize yield overhead, all Grappa scheduling is done in
-user-space and saves minimal state, leading to context switch times as
+To minimize yield overhead, the Grappa scheduler operates entirely in
+user-space and does little more than store register state of one task and load that of another. Context switch times are as
 low as \checkme{40ns} even when switching amongst thousands of tasks.
-This fast context switching amongst many tasks is crucial to Grappa's
-effectiveness for two reasons.  First, latency can be tolerated only
-when the net cost of context switching is dwarfed by the loss incurred
-by the alternative: stalling.  Second, because we can switch amongst
-so many tasks efficiently, we can \textbf{trade latency for throughput}:
-by {\em increasing} latency in key components of the
-system we are able to increase our aggregate random access bandwidth,
+Fast context switching amongst many tasks is crucial to Grappa for two reasons.
+
+First, only when the net cost of context switching is dominated by the loss incurred by stalling does a latency tolerant approach work to increase throughput on computations with intense random access.
+Random access latencies to node local memory are on the order of 100ns.
+Were the scheduler to add more overhead than this, it would actually slow down rather than speed up computations that randomly access node local data.
+
+Second, because we can switch amongst so many tasks without dramatically increasing scheduling overhead, Grappa can tolerate system latencies orders of magnitude greater than that of local memory.
+Network ping latency alone is approximately 2us, but we can tolerate much greater latency than this.
+Indeed, we can \textbf{trade latency for throughput}:  by {\em increasing} latency in key components of the system we are able to increase our aggregate random access bandwidth,
 our synchronization bandwidth, and our ability to tolerate load
@@
 Ie, that aggregating increases throughput and latency, that delegating synchronization increases the rate and the latency at which we can eg atomically increment, and that stealing work increases the latency of some tasks (eg when their data is largely at the originating node) but provides greater throughput overall.}
+
+Furthermore, by overprovisioning task parallelism, Grappa promises to provide robust throughput that is relatively insensitive to variations in latencies within and across our target systems.
 

### 2012-07-23 doc/asplos13/intro.tex
patched up citations

@@
 to tolerate the latency of data movement. The fully custom Tera
-MTA-2~\cite{tera} \TODO{this is mta-1 citation} system is a classic example of supporting irregular
+MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular
 applications by using concurrency to hide latencies. It had a large
@@
 possible. We use unmodified Linux for the operating system. An off the
-shelf user-mode infiniband device driver stack~\cite{Melonox?} is used
+shelf user-mode infiniband device driver stack~\cite{OFED} is used
 for network interfacing. MPI is used for process setup and

### 2012-07-23 doc/asplos13/relatedwork.tex
fixed up related work

--- 2012-07-23 doc/asplos13/relatedwork.tex
@@
 
-% The goal of the Grappa framework is to simplify the task of implementing
-% irregular applications.
 Grappa is built on many existing ideas in programming languages, systems and
@@
 
-\paragraph{Comparable frameworks} Distributed graph processing frameworks like
-Pregel \cite{pregel:2010} and Distributed GraphLab \cite{distgraphlab:vldb12}
-share similar goals as Grappa. Pregel adopts a bulk-synchronous parallel (BSP)
-execution model, which makes it inefficient on workloads that could prioritize
-vertices. GraphLab, on the other hand, schedules vertex computations
-individually, allowing prioritization, which gives faster convergence in a
-variety of iterative algorithms.\TODO{GraphLab is not general. Myers will
-expand on that here. Make a strong point about generality.} Grappa also
-supports dynamic parallelism with asynchronous execution, but parallelism is
-expressed as tasks or loop iterations.\TODO{Should RAMcloud go here?}.
+\paragraph{Comparable frameworks} Distributed graph processing
+frameworks like Pregel \cite{pregel:2010} and Distributed GraphLab
+\cite{distgraphlab:vldb12} share similar goals as Grappa. Pregel adopts
+a bulk-synchronous parallel (BSP) execution model, which makes it
+inefficient on workloads that could prioritize vertices. GraphLab, on
+the other hand, schedules vertex computations individually, allowing
+prioritization, which gives faster convergence in a variety of iterative
+algorithms.  GraphLab, however, imposes a rigid computation model where
+programmers must express computation as transformations on a vertex and
+its edge list only, with only information from adjacent vertexes. Pregel
+is only slightly less restrictive, as the input data can be any vertex
+in the graph.  Grappa also supports dynamic parallelism with
+asynchronous execution, but parallelism is expressed as tasks or loop
+iterations. By comparison, this is a far more general programming model
+for irregular computation tasks.
 
 \paragraph{Global memory} Grappa includes a custom implementation of a
-software distributed shared memory (DSM) system. Many traditional software DSM
-systems are page based~\cite{Treadmarks,munin} and aim to hide the fact that
-they are built in software from applications by exploiting the processor's
-paging mechanisms, therefore relying heavily on locality. Instead, Grappa,
-like other partitioned global address space (PGAS) models, implements its DSM
-at the language, rather than system level. Languages such as Chapel
-\cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC} make accesses to shared
-structures look like normal memory references. As we describe later, Grappa
-chooses a middle ground, where global addresses are explicit in the API and
-local accesses are emitted conventionally by the compiler. Similar to DSM
-systems, Grappa provides a consistency model that ensure high performance and
-with good programmability properties. % Caching is Cache operations are explicit in the  exposed to in the
-% API, however, allowing software to optimize their usage.
+software distributed shared memory (DSM) system. Many traditional
+software DSM systems are page based~\cite{Treadmarks,munin} and aim to
+hide the fact that they are built in software from applications by
+exploiting the processor's paging mechanisms, therefore relying heavily
+on locality. Instead, Grappa, like other partitioned global address
+space (PGAS) models, implements its DSM at the language, rather than
+system level. Languages such as Chapel \cite{Chamberlain:2007}, X10
+\cite{X10:2005}, and UPC \cite{upc:2005} make accesses to shared
+structures look like normal memory references. As we describe later,
+Grappa chooses a middle ground, where global addresses are explicit in
+the API and local accesses are emitted conventionally by the compiler. 
+While Grappa's DSM system is conceptually similar to prior work, its
+implementation is tuned for irregular computations.  Past DSM work,
+being page-based, could exploit the RDMA capabilities of network
+hardware to move large page-sized blocks of data from node to node.  In
+our experience, when these networks move small blocks of data (a few
+bytes), only a fraction of the available bandwidth is achieved.  In
+addition, the DSM system in Grappa ends up being tightly coupled to the
+task scheduler in order to overlap long latency memory operations with
+useful computation.  For these two reasons it became necessary to build
+a new DSM system specifically for Grappa.
 
-We decided to build a custom distributed shared memory implementation, as
-opposed to using an existing implementation, for two key reasons: (1) in order
-to support the large number of tasks required for latency tolerance, the
-implementation of the global memory system is tightly coupled to the task
-scheduler; and (2) since commodity networks only do efficient RDMA or ordinary
-transmission of small messages, the implementation of the global memory system
-must make use of the network message aggregator.
+%% We could be stronger here on why Grappa's task library is different,
+%% but we are running out of time.  -Mark
 
-%
-
-\paragraph{Multithreading}
-Grappa uses multithreading to tolerate memory latency. This is a well known
-technique. Hardware implementations include the Tera MTA \cite{Tera}, Cray XMT
-\cite{XMT}, Simultaneous multithreading \cite{SMT}, MIT Alewife
-\cite{Alewife}, Cyclops \cite{Cyclops}, and even GPUs \cite{fatahalian}. As we
-describe in this paper, Grappa use a lightweight user-mode task scheduler to
-multiplex thousands of tasks on a single processing core. The large number of
-tasks is required because of the extremely large internode latency.
+\paragraph{Multithreading} Grappa uses multithreading to tolerate memory
+latency. This is a well known technique. Hardware implementations
+include the Tera MTA \cite{tera:mta1}, Cray XMT \cite{feo:xmt},
+Simultaneous multithreading \cite{tullsen:smt}, MIT Alewife
+\cite{agarwal:alewife}, Cyclops \cite{almasi:cyclops}, and even GPUs
+\cite{gpus}. As we describe in this paper, Grappa use a lightweight
+user-mode task scheduler to multiplex \emph{thousands} of tasks on a
+single processing core. The large number of tasks is required because of
+the extremely high internode latency Grappa is mitigating.  Grappa's
+task library employs several optimizations: an extremely fast task
+switch, a small task size, and judicious use of hardware prefetching to
+bring task state into the cache long before that task is actually
+scheduled.

### 2012-07-23 doc/asplos13/abstract.tex
pass on abstract

@@
-Grappa is a latency tolerant runtime system for commodity multinode multicore
-platforms that addresses the challenges posed in scaling irregular parallel
-applications to large datasets. These applications are of growing importance
-as interactive analysis of large unstructured data sets emerges as a
-commercial and government priority. Grappa serves both as a C++ user library
-and as a foundation upon which higher level languages can be developed or
-adapted to support development of irregular parallel applications. Grappa
-tolerates the delays to distant memory by multiplexing hundreds to thousands
-of lightweight tasks to each processor core; balances load via fine-grained
-workstealing; and provides efficient synchronization and consistency across
-its single shared address space. We present a detailed description
-of the Grappa system, programming examples using the library interface, and
-compare performance for several irregular benchmarks to the Cray XMT, a custom
-system used to target the real time graph analytics market.
+Grappa is a latency tolerant runtime system for commodity clusters of
+multicore computers, its goal is to provide scalable performance for irregular
+parallel applications. These applications are of growing importance as
+interactive analysis of large data sets emerges as a commercial and government
+priority, typically in the form of graph processing. Grappa serves both as a
+C++ user library and as a foundation upon which higher level languages can be
+developed or adapted to support development of irregular parallel
+applications. Grappa tolerates the delays to distant memory by multiplexing
+thousands of lightweight tasks to each processor core; balances load via
+fine-grained distributed work-stealing; takes full advantage of network
+characteristics by aggregating smaller requests into large ones; and provides
+efficient synchronization and remote operations across its single shared
+address space. We present a detailed description of the Grappa system,
+programming examples using the library interface, and compare performance for
+several irregular benchmarks to the Cray XMT, a custom system used to target
+the real time graph analytics market.

### 2012-07-23 doc/asplos13/communication.tex
Fixed checkmes in comm

@@
 the bisection bandwidth is only a small fraction, less than
-~\checkme{5\%} of the peak bisection bandwidth.
+3\% of the peak bisection bandwidth.
 
 In our experiments the vast majority of requests were smaller than
-\checkme{255} bytes \TODO{more precise? median?}, far too small to make
+44 bytes, far too small to make
 efficient use of the network. To make the best use of the network, we
@@
 sent. First, each queue has a message size threshold, chosen to give
-reasonable network performance (\checkme{2048 bytes}). If the size in
+reasonable network performance 4096. If the size in
 bytes of a queue is above the threshold, the contents of the queue are
 sent immediately. Second, each queue has a wait time threshold
-(\checkme{200ms}). If the oldest message in a queue has been waiting
+($\approx${1ms}). If the oldest message in a queue has been waiting
 longer than this threshold, the contents of the queue are sent

### 2012-07-23 doc/asplos13/abstract.tex
pass on abstract and intro.

@@
 Grappa is a latency tolerant runtime system for commodity clusters of
-multicore computers, its goal is to provide scalable performance for irregular
-parallel applications. These applications are of growing importance as
-interactive analysis of large data sets emerges as a commercial and government
-priority, typically in the form of graph processing. Grappa serves both as a
-C++ user library and as a foundation upon which higher level languages can be
-developed or adapted to support development of irregular parallel
-applications. Grappa tolerates the delays to distant memory by multiplexing
-thousands of lightweight tasks to each processor core; balances load via
-fine-grained distributed work-stealing; takes full advantage of network
-characteristics by aggregating smaller requests into large ones; and provides
-efficient synchronization and remote operations across its single shared
-address space. We present a detailed description of the Grappa system,
+multicore computers. Grappa's goal is to provide scalable performance for
+irregular parallel applications. These applications are of growing importance
+as interactive analysis of large data sets emerges as a commercial and
+government priority, typically in the form of graph processing. Grappa serves
+both as a C++ user library and as a foundation upon which higher level
+languages can be developed or adapted to support development of irregular
+parallel applications. Grappa tolerates the delays to distant memory by
+multiplexing thousands of lightweight tasks to each processor core; balances
+load via fine-grained distributed work-stealing; takes full advantage of
+network characteristics by aggregating smaller requests into large ones; and
+provides efficient synchronization and remote operations across its single
+shared address space. We present a detailed description of the Grappa system,
 programming examples using the library interface, and compare performance for

### 2012-07-23 doc/asplos13/intro.tex
pass on abstract and intro.

@@
 
-Our evaluation of Grappa shows that it runs several graph-crunching applications (classic examples of irregular behavior) very efficiently on a commodity cluster  Our yardstick for comparison is the XMT hardware itself.
-Using the same number of processors (in our case cores on a single multicore processor), Grappa is in the same ballpark as the XMT: on some graphs, such as the \checkme{WTF} Grappa is \checkme{2X} faster, while on others, such as the \checkme{WTF2} Grappa is \checkme{2X} slower.
-In Section~\ref{sec:results} we explore the factors that underpin this performance.
-But most importantly, application performance \emph{scales} with Grappa.
-For significantly less real world cost, users can \emph{add} more processors to a commodity cluster and easily outperform the XMT hardware on graph processing, while still saving significant dollars.
+Our evaluation of Grappa shows that it runs several graph-crunching
+applications (classic examples of irregular behavior) very efficiently on a
+commodity cluster. Our yardstick for comparison is the XMT hardware itself.
+Using the same number of processors (in our case cores on a single multicore
+processor), Grappa is in the same ballpark as the XMT: on some graphs, such as
+the \checkme{WTF} Grappa is \checkme{2X} faster, while on others, such as the
+\checkme{WTF2} Grappa is \checkme{2X} slower. In Section~\ref{sec:results} we
+explore the factors that underpin this performance. But most importantly,
+application performance \emph{scales} with Grappa. For significantly less real
+world cost, users can \emph{add} more processors to a commodity cluster and
+easily outperform the XMT hardware on graph processing, while still saving
+significant dollars.
+
 % 

### 2012-07-23 doc/asplos13/relatedwork.tex
minor edits

@@
 asynchronous execution, but parallelism is expressed as tasks or loop
-iterations. By comparison, this is a far more general programming model
+iterations, which is a far more general programming model
 for irregular computation tasks.

### 2012-07-23 doc/asplos13/method.tex

@@
 These algorithms were implemented for the Cray XMT (our baseline) and for Grappa.
-Performance results for Grappa were obtained on a 144 node cluster.
-Individual nodes of this cluster contain the hardware depicted in Table~\ref{table:grappanode}.
-Nodes are interconnected with both 10G ethernet and infiniband.
-For all but startup and configuration, they are configured to communicate over the infiniband network for these experiments.
-
 The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.
@@
 to access shared memory.  In the HPC cluster we use, each processor
-(which contains up to 32 cores, although we only use \checkme{6} in our
+(which contains up to 32 cores, although we only use 6 in our
 experiments), has a single infiniband interface.  Hence, for our results
@@
 processors. Nodes have 32-core (every pair share a floating-point unit)
-2.1-GHz processors, 64GB of memory, and \checkme{a 40Gb Super-terrific
-Happy-Lucky} infiniband network card.   The cluster uses a
-\checkme{McNugget} infiniband switch. We configure the nodes to have 32
-1-GB hugepages to minimize TLB misses for the random access patterns we
-expect from irregular applications.\comment{The results we present are for this
-machine, but we have also run Grappa on our own 12-node Intel Xeon
-Westmere cluster, which performs similarly.}
+2.1-GHz processors, 64GB of memory, and 40Gb Mellanox ConnectX-2
+infiniband network card.   The cluster uses a QLogic infiniband switch.
+We configure the nodes to have 32 1-GB hugepages to minimize TLB misses
+for the random access patterns we expect from irregular applications.
 
 We compare to the MTA using a 128-node Cray XMT (3rd generation MTA).
-Each node consists of a 500-MHz MTA Threadstorm multithreaded
-processor that supports 128 streams. The machine uses Cray's proprietary
-SeaStar2 interconnection network.
+Each node consists of a 500-MHz MTA Threadstorm multithreaded processor
+that supports 128 streams. The machine uses Cray's proprietary SeaStar2
+interconnection network.
 
@@
 We have used three benchmarks to explore the performance of Grappa:
-
-\comment{
-    Traversal of an unbalanced, unpartitioned tree captures the foundations for implementing most forms of irregular parallelism. 
-Recursion is found in every irregular divide and conquer algorithm.
-Loops, even regular ones, must be dynamically scheduled to get good
-load balance when run on a large, non-uniform system. Nested
-parallelism involves fine-grain creation of dynamic amounts of work.
-Dataflow..
-}
 
@@
 in-memory tree. This in-memory traversal has no knowledgeable of the
-tree structure beforehand.
-
-%todo: say tree explore is same as a search except we are additionally needing to synchronize on all vertices visited
-
-The Grappa version of the in-memory tree search uses the
-asynchronous parallel for loop over a visited vertex's children 
-list.  \comment{WTF? -M
-A larger threshold can be set to take advantage of this 
-locality by processing more child pointers sequentially in a single task.}
+tree structure beforehand.  The Grappa version of the in-memory tree
+search uses the asynchronous parallel for loop over a visited vertex's
+children list.
 

### 2012-07-23 doc/asplos13/overview.tex
pass on overview

@@
 
-Grappa (Figure~\ref{fig:grappa}) provides three main software components:
-a \textbf{tasking system} with lightweight multithreading to tolerate
-latency and workstealing to automate load balance;
-a \textbf{distributed shared memory} (DSM) providing high aggregate
-random access bandwidth for both normal and synchronized operations;
-and a \textbf{communication layer}, which is largely invisible to the software developer but helps to improve network performance when applications read and write only small pieces of data.
+Grappa (Figure~\ref{fig:grappa}) has three main software components:
+\begin{description}
+
+\item [Tasking system.] Our tasking system supports lightweight
+multithreading to tolerate communication latency and global distributed
+workstealing (i.e., tasks can be stolen from any node in the system), which
+provides automated load balancing.
+
+\item[Distributed shared memory.] Our DSM system provides support for
+fine-grain access to data anywhere in the system. It supports synchronization
+operations on global data, explicit local caching of any local in the system,
+and support for operation on remote data (delegating operations to home node).
+By tight integration with the tasking system described about and the
+communication layer (below), our DSM system offers high aggregate random
+access bandwidth for accessing remote data.
+
+\item[Communication layer.] As discussed earlier, modern commodity networks
+support high bandwidth only for large messages. Since irregular applications
+tend to need frequent communication of small requests, the main goal of our
+communication layer is to aggregate small messages into large ones better
+exploit what the network can offer. It is largely invisible to the application
+programmer.
+
+% software developer but helps to improve network performance when applications read and write only small pieces of data.
+
+\end{description}
+
 In the next three sections we describe both the programmer's view of Grappa's main capabilities and how they are implemented.

### 2012-07-23 doc/asplos13/conclude.tex

\section{Conclusion}

Irregular computations are both important and challenging to execute quickly. Scaling these applications easily to commodity hardware has been a historical challenge. Grappa is a runtime framework that simplifies this task for software developers and compiler writers.  This paper describes the Grappa framework and its three main components: a task library, a distributed shared memory system, and a network aggregator to make commodity networks efficient with small message sizes.  We explored the performance of Grappa on three algorithms: unbalanced tree search, breadth first search, and betweenness centrality.  Performance comparisons to the Cray XMT for a small number of nodes (16 nodes / 96 cores) show great promise: Grappa performs on-par with the far more expensive and custom XMT system, sometimes achieving better performance and sometimes worse.  Our initial scaling experiments to a higher number of nodes suggests Grappa does scale, but there is room for further improvement.  Due to limitations in the initial aggregator design Grappa cannot utilize more than 6 cores/node, and this limits not only peak performance at a small number of nodes, but scaling performance at a larger number of nodes.  Current work is ongoing on a new aggregator design.

### 2012-07-23 doc/asplos13/overview.tex
fixed up overview

@@
 fine-grain access to data anywhere in the system. It supports synchronization
-operations on global data, explicit local caching of any local in the system,
+operations on global data, explicit local caching of any memory in the system,
 and support for operation on remote data (delegating operations to home node).
-By tight integration with the tasking system described about and the
-communication layer (below), our DSM system offers high aggregate random
+By tight integration with the tasking system and the
+communication layer, our DSM system offers high aggregate random
 access bandwidth for accessing remote data.
@@
 tend to need frequent communication of small requests, the main goal of our
-communication layer is to aggregate small messages into large ones better
+communication layer is to aggregate small messages into large ones to better
 exploit what the network can offer. It is largely invisible to the application
@@
 
-In the next three sections we describe both the programmer's view of Grappa's main capabilities and how they are implemented.
+In the next three sections we describe both Grappa's main capabilities and how they are implemented.

### 2012-07-23 doc/asplos13/tasks.tex
fussing with overview and tasks

@@
 The basic unit of execution in Grappa is a {\em task}. Each task is
-represented by a function pointer and its arguments and is enqueued when
-spawned, but not instantiated.  Later, when resources are free,
-the task is allocated a stack, bound to a core, and executed.
+represented by a function pointer and its arguments. A new task is enqueued at
+spawn time; when resources are free, it is allocated a stack, bound to a core,
+and executed.
 
@@
 in Figure~\ref{fig:scheduling}.
-\TODO{cite UPC split phase reads and
-writes, but include that we provide a way to overlap computation}
+% \TODO{cite UPC split phase reads and
+% writes, but include that we provide a way to overlap computation}
 
 To minimize yield overhead, the Grappa scheduler operates entirely in
-user-space and does little more than store register state of one task and load that of another. Context switch times are as
-low as \checkme{40ns} even when switching amongst thousands of tasks.
-Fast context switching amongst many tasks is crucial to Grappa for two reasons.
-
-First, only when the net cost of context switching is dominated by the loss incurred by stalling does a latency tolerant approach work to increase throughput on computations with intense random access.
-Random access latencies to node local memory are on the order of 100ns.
-Were the scheduler to add more overhead than this, it would actually slow down rather than speed up computations that randomly access node local data.
-
-Second, because we can switch amongst so many tasks without dramatically increasing scheduling overhead, Grappa can tolerate system latencies orders of magnitude greater than that of local memory.
-Network ping latency alone is approximately 2us, but we can tolerate much greater latency than this.
-Indeed, we can \textbf{trade latency for throughput}:  by {\em increasing} latency in key components of the system we are able to increase our aggregate random access bandwidth,
-our synchronization bandwidth, and our ability to tolerate load
-imbalance.
-\TODO{check to see we actually explain these three somewhere.
-Ie, that aggregating increases throughput and latency, that delegating synchronization increases the rate and the latency at which we can eg atomically increment, and that stealing work increases the latency of some tasks (eg when their data is largely at the originating node) but provides greater throughput overall.}
-
-Furthermore, by overprovisioning task parallelism, Grappa promises to provide robust throughput that is relatively insensitive to variations in latencies within and across our target systems.
+user-space and does little more than store register state of one task and load
+that of another. Context switch times are as low as 38ns even when switching
+amongst thousands of tasks. Fast context switching amongst many tasks is
+crucial to Grappa for two reasons. First, only when the net cost of context
+switching is dominated by the loss incurred by stalling does a latency
+tolerant approach work to increase throughput on computations with intense
+random access. Random access latencies to node local memory are on the order
+of 100ns. Were the scheduler to add more overhead than this, it would actually
+slow down rather than speed up computations that randomly access node local
+data. Second, because we can switch amongst so many tasks without dramatically
+increasing scheduling overhead, Grappa can tolerate system latencies many
+orders of magnitude greater than that of local memory. Network ping latency
+alone is approximately 2us, but we can tolerate latencies in the millisecond
+scale. Furthermore, by over-provisioning task parallelism, Grappa promises to
+provide robust throughput that is relatively insensitive to variations in
+latencies within and across our target systems.
 
@@
 recursive decomposition} of iterations, similar to Cilk's cilk\_for
-construct~\cite {cilkforimplementation} \comment{could only find slides
+construct~\cite {cilkforimplementation}.\comment{could only find slides
 page6 of
 http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-
-Cilk++.pdf    ---- Good enough.  Cite it.  -Mark}.  It generates a
+Cilk++.pdf    ---- Good enough.  Cite it.  -Mark.}  It generates a
 logarithmically-deep tree of tasks, stopping to execute the loop body

### 2012-07-23 doc/asplos13/tasks.tex
edits to Implementation tasks -- still needs work -- anyone?

--- 2012-07-23 doc/asplos13/tasks.tex
@@
 http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-
-Cilk++.pdf    ---- Good enough.  Cite it.  -Mark}.  It generates a
+Cilk++.pdf    ---- Good enough.  Cite it.  -Mark} and TBB's {\tt parallel\_for}~\cite{intel_tbb}.  It generates a
 logarithmically-deep tree of tasks, stopping to execute the loop body
@@
 
-Fourth, a programmer may want to run a small piece of code on a particular core
-in the system without waiting for execution resources to be available.
-Grappa provides the \texttt{call\_on} call for this purpose.
+Fourth, a programmer may want to execute an active message; that is,
+to run a small piece of code on a particular core in the system
+without waiting for execution resources to be available.  For example,
+custom synchronization primitives execute this way, as a function
+executed on the core where the data lives.  Grappa provides the
+\texttt{call\_on} call for this purpose.
 
@@
 64-bit function pointer plus three 64-bit arguments. We use three
-arguments because tasks are usually generated as part of a parallel loop
-decomposition, and thus each task needs three kinds of data.
+arguments because tasks are often generated via a parallel loop
+decomposition, in which each task needs three kinds of data.
 \begin{description}
-\item[Function pointer] This 64-bit value indicates what routine
-  should run. We configure the system nodes to disable address randomization, so that we can depend on the fact that function pointers are valid across process images.
-\item[Private argument] This 64-bit value is typically used for task-specific
-  data (e.g., a loop index).
-\item[Shared argument] This 64-bit value is typically used for data shared
-  between all tasks that are part of a loop or to specify the number
-  of loop iterations.
-\item[Synchronization argument] This 64-bit value is typically used to determine
-  when all tasks that are part of a loop have finished. It is usually
-  a global pointer to a synchronization object allocated at the core
-  that spawned the task.
+\item[Function pointer] addresses the routine to run. We configure the system nodes to disable address randomization, so that function pointers are valid across process images.
+\item[Private argument] often a loop index.
+\item[Shared argument] typically data shared amongst a group of tasks, or the number of loop iterations to be performed by this task
+\item[Synchronization argument] often used to determine
+  when all tasks that are part of a loop have finished; a global pointer to a synchronization object allocated at the core that spawned a group of tasks.
 \end{description}
@@
 \paragraph{Scheduling} Each core in the Grappa system has its own
-independent scheduler. Each scheduler has three main tasks to perform. 
-First, it must ensure that its communication resources are serviced
-reasonably often. Second, it must ensure that if a running task was
-waiting on a long-latency operation and that operation has completed,
-the task will be rescheduled. Third, if there are tasks waiting to be
-run and spare execution resources, the tasks must be matched with idle
-workers.
+independent scheduler. Each scheduler has three main tasks to perform:
+servicing communication requests as described in
+Section~\ref{sec:communication}; rescheduling tasks that have been
+waiting on long-latency operations; and assigning ready tasks to
+worker resources that have become idle.
 
@@
 
-Whenever a task yields or suspends, the scheduler makes a decision about
-what to do next. First, it determines if the communication resources
-should be serviced. This is done on a periodic basis. Second, it
-determines if any workers with running tasks are ready to execute; if
-so, one is scheduled. Finally, if there are no workers ready to run, but
-there are tasks waiting to be matched with workers, an idle worker is
-woken (or a new worker is spawned), matched with a task, and scheduled.
-The scheduler ensures that communication maintenance tasks are executed periodically.
+Whenever a task yields or suspends, the scheduler makes a decision
+about what to do next. Servicing communication requests is given priority
+to ensure responsiveness, but to minimize overhead should context switches be frequent, servicing is performed only if sufficient time has elapsed.
+Second, the scheduler determines if any workers with running tasks are
+ready to execute; if so, one is scheduled. Finally, if there are no
+workers ready to run, but there are tasks waiting to be matched with
+workers, an idle worker is woken (or a new worker is spawned), matched
+with a task, and scheduled.
 
-\paragraph{Work stealing} When a Grappa node runs out of work, it
-becomes a ``thief'' and asks another other Grappa nodes (``victim'') for
-work. If the victim has tasks to spare in its public queue, they will be
-transferred to the thief's public queue.
+\paragraph{Work stealing}
 
-Most prior work on work stealing assumes that a processor has one worker
-thread and so it only steals when utilization would go to zero. Grappa,
-on the other hand, relies on having many worker threads per core. The
+When the scheduler finds no work to assign to its workers, it
+commences to steal work from other cores using an asynchronous
+\tt{call_on} active message.  It chooses a victim at random until it
+finds one with a non-zero amount of work in its public task queue.  Up
+to a tunable limit, the scheduler steals half of the tasks it finds.
+We did study the impact of this limit on performance and found
+\checkme{128} tasks to be optimal with little sensitivity around this
+number.  Hence we omit results for this study from
+Section~\ref{sec:results} and simply use \checkme{128} as the
+parameter for all experiments.
+
+\TODO{Someone check this paragraph to make sure it is true. :-)  -Mark}
+\TODO{Anything after this I'm not sure adds to the paper -- if it does, please edit -Simon}
+The
 core would be fully-utilized if there is always something on the ready
@@
 without syncing all tasks.
-
-\TODO{Someone check this paragraph to make sure it is true. :-)  -Mark}
-
-When a grappa node decides to issue a steal request it chooses the
-victim node at random.  It also requests a block of tasks at once.  If
-this block exceeds the amount of stealable tasks on the victim node,
-half of the stealable tasks are stolen instead.  While we did study the
-number of tasks to steal at once, we found performance was maximized
-with \checkme{128} tasks with little sensitivity around this number,
-hence we omit results for this from Section~\ref{sec:results} and simply
-use \checkme{128} as the parameter for all experiments.

### 2012-07-23 doc/asplos13/tasks.tex

@@
 recursive decomposition} of iterations, similar to Cilk's cilk\_for
-construct~\cite {cilkforimplementation} \comment{could only find slides
+construct~\cite {cilkforimplementation}.\comment{could only find slides
 page6 of

### 2012-07-23 doc/asplos13/method.tex
Preston's changes to methodology.

@@
 
-To explore the performance of the Grappa runtime we have implemented three algorithms: breadth first search, betweenness centrality, and unbalanced tree search.
-These algorithms were implemented for the Cray XMT (our baseline) and for Grappa.
-The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.
-Data is collected on real systems, which means minor variations in runtime exist from run to run.
-The average of multiple runs are used, and where appropriate, confidence in the result is reported.
+To explore the performance of the Grappa runtime we have implemented
+three algorithms: breadth first search, betweenness centrality, and
+unbalanced tree search.  These algorithms were implemented for the
+Cray XMT (our baseline) and for Grappa.  The metric we use is
+algorithmic time, which means startup and loading of the data
+structure (from disk) is not included in the measurement.  Data is
+collected on real systems, which means minor variations in runtime
+exist from run to run.  The average of multiple runs are used, and
+where appropriate, confidence in the result is reported.
 
@@
 to access shared memory.  In the HPC cluster we use, each processor
-(which contains up to 32 cores, although we only use 6 in our
+(which contains up to 32~cores, although we only use 6 in our
 experiments), has a single infiniband interface.  Hence, for our results
@@
 Each node consists of a 500-MHz MTA Threadstorm multithreaded processor
-that supports 128 streams. The machine uses Cray's proprietary SeaStar2
+that supports 128~streams. The machine uses Cray's proprietary SeaStar2
 interconnection network.
@@
 performance of systems for parallel applications that require dynamic
-load balancing \cite{Olivier:uts2006}. It involves traversing an
+load balancing~\cite{UTS}. It involves traversing an
 unbalanced implicit tree: at each vertex, its number of children is
@@
 \paragraph{Breadth-first-search (BFS)}
-This is the primary kernel for the Graph500 benchmark and is what currently determines the ranking of machines on the Graph500 list~\cite{graph500list}.
-As a whole, the Graph500 benchmark suite is designed to bring the focus of system design on data-intensive workloads, particularly large-scale graph analysis problems, that are important among cybersecurity, informatics, and network understanding workloads.
-The BFS benchmark builds a search tree containing parent nodes for each traversed vertex during the search.
-While this is a relatively simple problem to solve, it exercises the random-access and fine-grained synchronization capabilities of a system as well as being a primitive in many other graph algorithms.
-Performance is measured in \emph{traversed edges per second} (TEPS), where the number of edges is the edges making up the generated BFS tree.
-One of the reference implementations of Graph500 BFS is for the XMT; this code fails to scale past 16 XMT processors because it does not expose enough parallelism, so we modified the code to use a recursive loop decomposition similar to Grappa's.
-We compare this modified version against a straightforward Grappa implementation.
-We do not employ algorithmic improvements, though there are many \cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
+This is the primary kernel for the Graph500 benchmark and is what
+currently determines the ranking of machines on the Graph500
+list~\cite{graph500list}. As a whole, the Graph500 benchmark suite is
+designed to bring the focus of system design on data-intensive
+workloads, particularly large-scale graph analysis problems, that are
+important among cybersecurity, informatics, and network understanding
+workloads. The BFS benchmark builds a search tree containing parent
+nodes for each traversed vertex during the search.  While this is a
+relatively simple problem to solve, it exercises the random-access and
+fine-grained synchronization capabilities of a system as well as being
+a primitive in many other graph algorithms. Performance is measured in
+\emph{traversed edges per second} (TEPS), where the number of edges is
+the edges making up the generated BFS tree. One of the reference
+implementations of Graph500 BFS is for the XMT; this code fails to
+scale past 16 XMT processors because it does not expose enough
+parallelism, so we modified the code to use a recursive loop
+decomposition similar to Grappa's. We compare this modified version
+against a straightforward Grappa implementation. We do not employ
+algorithmic improvements, though there are
+many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
 
 \paragraph{Betweenness Centrality}
-An important measure of the importance of particular vertices in a network is betweenness centrality (BC) \cite{freeman1979centrality}.
-By this measure, the ``importance'' of each vertex is computed by finding the fraction of shortest paths that pass through it, which can optionally be approximated by only computing shortest paths using a subset of the vertices as starting points.
-BC can be useful for understanding which vertices may have the greatest impact, so in social networks this could be the primary person linking two communities.
-Because it requires multiple breadth-first traversals across the entire graph and in reverse, on power-law degree graphs, this algorithm exercises random accesses rate, and load balancing.
-It also requires fine-grained synchronization on updates to vertex centrality values because multiple paths will update the same vertex.
-BC is a kernel in the DARPA High Performance Computing Systems (HPCS) Scalable Synthetic Compact Applications graph analysis (SSCA\#2) benchmark\cite{ssca2}.
-Performance for BC is also measured in TEPS, where the number of traversed edges is the total number of edges in the graph multiplied by the number of random starting vertices used in the approximate computation.
-We use the XMT implementation of BC implemented as part of the GraphCT~\cite{GraphCT} library and a comparable Grappa implementation, both of which use the parallel BC algorithm developed by Bader et al.~\cite{bader:bc}.
+An important measure of the importance of particular vertices in a
+network is betweenness centrality
+(BC)~\cite{freeman1979centrality}. By this measure, the ``importance''
+of each vertex is computed by finding the fraction of shortest paths
+that pass through it, which can optionally be approximated by only
+computing shortest paths using a subset of the vertices as starting
+points. BC can be useful for understanding which vertices may have the
+greatest impact, so in social networks this could be the primary
+person linking two communities. Because it requires multiple
+breadth-first traversals across the entire graph and in reverse, on
+power-law degree graphs, this algorithm exercises random accesses
+rate, and load balancing. It also requires fine-grained
+synchronization on updates to vertex centrality values because
+multiple paths will update the same vertex. BC is a kernel in the
+DARPA High Performance Computing Systems (HPCS) Scalable Synthetic
+Compact Applications graph analysis (SSCA\#2)
+benchmark~\cite{ssca2}. Performance for BC is also measured in TEPS,
+where the number of traversed edges is the total number of edges in
+the graph multiplied by the number of random starting vertices used in
+the approximate computation. We use the XMT implementation of BC
+implemented as part of the GraphCT~\cite{GraphCT} library and a
+comparable Grappa implementation, both of which use the parallel BC
+algorithm developed by Bader et al.~\cite{bader:bc}.

### 2012-07-23 doc/asplos13/overview.tex

@@
 
+\paragraph{Exploiting Latency Tolerance} As we will show later, Grappa can tolerate latencies way beyond that of the network.
+Therefore, Grappa can afford to \emph{trade latency for throughput}: by {\em increasing} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
+
+% \TODO{check to see we actually explain these three somewhere.
+Ie, that aggregating increases throughput and latency, that delegating synchronization increases the rate and the latency at which we can eg atomically increment, and that stealing work increases the latency of some tasks (eg when their data is largely at the originating node) but provides greater throughput overall.}
+
 In the next three sections we describe both Grappa's main capabilities and how they are implemented.

### 2012-07-23 doc/asplos13/evaluation.tex
split UTS scaling and 1-16 sections. Trim the XMT retries discussion down

@@
 \subsubsection{Unbalanced tree search in memory}
+
+%Using 6 cores per node gave
+%Grappa up to a 20\% benefit at each data point, but Grappa could not run
+%well beyond 90 machines with this core configuration because of the
+%current network aggregator designa.
 
@@
 the performance in terms of number of vertices visited per second versus
-number of compute nodes. The Grappa results shown are for 5 cores per
+number of compute nodes. The Grappa results shown are for 6 cores per
 node and the best parameter values from a search over runtime and
-application parameters. Grappa with \checkme{16} machines is faster than
-the entire XMT of 128 processors. Grappa achieves \checkme{188Mvert/s}
-with \checkme{128 machines} and the XMT achieves only 50Mvert/s,
-plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
-The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
-scaling up through 128---the maximum we could use---although scaling
-declines because of the unscalability of our aggregation mechanism as
-number of network endpoints increases. Using 6 cores per node gave
-Grappa up to a 20\% benefit at each data point, but Grappa could not run
-well beyond 90 machines with this core configuration because of the
-current network aggregator design.
+application parameters.
 
@@
 \end{figure}
-
-Despite our efforts to tune the UTS implementation specific to the 
-XMT, performance does not scale well with increasing processor count,
-flattening out around sixty processors.  When we increase the size of
-the tree from 100M to 4.2B, we find that performance does not improve, suggesting that performance is not limited by task parallelism.
-Cray's performance tools show that an increasing number of memory retry operations are generated by code within the runtime system.
-Retries are performed by the memory controller when remote synchronization operations fail to find the full-bit associated with each memory location in the unavailable state.
-Retries are issued at low priority relative to new memory operations issued by the processor by other contexts, so they consume what would otherwise be unused injection bandwidth.
-On a full-bandwidth system such as the MTA-2, retries have no impact on the progress of tasks other than their own.
-On a Cray XMT, network bandwidth is limited, so retries create congestion.
-In comparison, Grappa performs synchronization without retries, delaying responses at the receiving end until ready to notify the sender to proceed.
-This saves bandwidth and permits scaling of tasks performing synchronization even on low injection rate networks.
 
@@
 
+\subsection{Scaling}
+\paragraph{UTS-Mem}
+We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
+(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
+nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
+Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
+achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
+The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
+scaling up through 128 nodes, although scaling
+declines because of the unscalability of our aggregation mechanism as
+number of network endpoints increases.
+
+Despite our efforts to tune the UTS implementation specific to the 
+XMT, performance does not scale well with increasing processor count,
+flattening out around 60 processors.  When we increase the size of
+the tree from 100M to 4.2B, we find that performance does not improve,
+suggesting that performance is not limited by task parallelism.
+Cray's performance tools show an increasing number of memory
+retry operations for failed synchronization operations generated by
+the runtime, which create network contention.
+%Retries are performed by the memory controller when remote synchronization operations fail to find the full-bit associated with each memory location in the unavailable state.
+Retries are issued at low priority relative to new memory operations issued by the processor by other contexts, so they consume what would otherwise be unused injection bandwidth.
+On a full-bandwidth system such as the MTA-2, retries have no impact on the progress of tasks other than their own.
+On a Cray XMT, network bandwidth is limited, so retries create congestion.
+In comparison, Grappa performs synchronization without retries, delaying responses at the receiving end until ready to notify the sender to proceed.
+This saves bandwidth and permits scaling of tasks performing synchronization even on low injection rate networks.
+
 %\begin{figure*}[ht]

### 2012-07-23 doc/asplos13/intro.tex
moved latency paragraph

@@
 
-In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.
-This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating in software and inexpensive mass market hardware, the approach taken by Tera.
-We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.
-Ultimately, the opportunity is to cover the spectrum of irregular to regular computation: where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.
-Thus we make the best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.
+In the twenty years that have elapsed since the Tera MTA, commodity
+microprocessors have become much faster and multicore has driven down the
+absolute price of computation; commodity network price-performance has
+improved as well. This shift has afforded us the opportunity to attack the
+challenges posed by irregular applications by emulating in software and
+inexpensive mass market hardware, the approach taken by Tera. We exploit the
+increased aggregate instruction rate per socket relative to chip bandwidth,
+using what would otherwise be wasted instructions to manage the multiplexing
+of as many as several thousand tasks per core, thus tolerating memory latency,
+reducing stalls, and making better use of available bandwidth. Ultimately, the
+opportunity is to cover the spectrum of irregular to regular computation:
+where tasks exhibit locality, multiplex fewer tasks and expend fewer
+instructions on context switching; where locality is lacking, multiplex more
+tasks at a higher rate to tolerate latency. Thus we make the best use of task
+parallelism -- either to scale to more cores or to tolerate latency -- and of
+caches -- either to exploit application locality or to house more task
+contexts.
 
@@
 giving good performance when there is little locality to be exploited
-while allowing the programmer to exploit it when it is available. Grappa
-leverages as much freely available and commodity infrastructure as
-possible. We use unmodified Linux for the operating system. An off the
-shelf user-mode infiniband device driver stack~\cite{OFED} is used
-for network interfacing. MPI is used for process setup and
-tear down. GASnet~\cite{gasnet:spec} is used as the underlying mechanism for
-remote memory reads and writes using active message invocations. To this
-commodity hardware and software mix Grappa adds three main software
-components: (1) a \emph{lightweight tasking} layer that supports a
-context switch in as few as 38ns and distributed load balancing; (2) a
-\emph{distributed shared memory} layer that supports customary
-operations such as \emph{read and write as well as remote operations}
-such as fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message
-aggregation} layer that combines short messages to mitigate the
-aforementioned problem that commodity networks are designed to achieve
-peak bandwidth only on large packet sizes, yet irregular applications
-tend to fetch only a handful of bytes at a time.
+while allowing the programmer to exploit it when it is available.
+
+Grappa leverages as much freely available and commodity infrastructure as
+possible. We use unmodified Linux for the operating system and an
+off-the-shelf user-mode infiniband device driver stack~\cite{OFED}. MPI is
+used for process setup and tear down. GASnet~\cite{gasnet:spec} is used as the
+underlying mechanism for remote memory reads and writes using active message
+invocations. To this commodity hardware and software mix Grappa adds three
+main software components: (1) a \emph{lightweight tasking} layer that supports
+a context switch in as few as 38ns and distributed global load balancing; (2)
+a \emph{distributed shared memory} layer that supports customary operations
+such as \emph{read and write as well as remote operations} such as
+fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message aggregation} layer
+that combines short messages to mitigate the aforementioned problem that
+commodity networks are designed to achieve peak bandwidth only on large packet
+sizes, yet irregular applications tend to fetch only a handful of bytes at a
+time. As we will show later, Grappa can tolerate latencies way beyond that of
+the network. Therefore, Grappa can afford to \emph{trade latency for
+throughput}: by {\em increasing} latency in key components of the system we
+are able to increase our aggregate random access bandwidth (by delaying and
+aggregating messages), our synchronization bandwidth (by delegating operations
+to remote nodes), and our ability to improve load imbalance (work stealing
+increases latency).
 

### 2012-07-23 doc/asplos13/tasks.tex
remove stuff from tasks and todos

@@
 in Figure~\ref{fig:scheduling}.
-% \TODO{cite UPC split phase reads and
-% writes, but include that we provide a way to overlap computation}
-
 To minimize yield overhead, the Grappa scheduler operates entirely in
@@
 that of another. Context switch times are as low as 38ns even when switching
-amongst thousands of tasks. Fast context switching amongst many tasks is
+amongst thousands of tasks.
+
+\comment{
+Delete the spin and move the 2us and ms thing to results if we have data to show it. -Mark
+
+Fast context switching amongst many tasks is
 crucial to Grappa for two reasons. First, only when the net cost of context
@@
 orders of magnitude greater than that of local memory. Network ping latency
-alone is approximately 2us, but we can tolerate latencies in the millisecond
+alone is approximately 2us, but our data suggests we can tolerate latencies in the millisecond
 scale. Furthermore, by over-provisioning task parallelism, Grappa promises to
@@
 latencies within and across our target systems.
+}
 
@@
 \paragraph{Work stealing} 
-
 When the scheduler finds no work to assign to its workers, it
 commences to steal work from other cores using an asynchronous
-\tt{call_on} active message.  It chooses a victim at random until it
-finds one with a non-zero amount of work in its public task queue.  Up
-to a tunable limit, the scheduler steals half of the tasks it finds.
-We did study the impact of this limit on performance and found
-\checkme{128} tasks to be optimal with little sensitivity around this
-number.  Hence we omit results for this study from
-Section~\ref{sec:results} and simply use \checkme{128} as the
-parameter for all experiments.
+\texttt{call\_on} active message.  It chooses a victim at random until it
+finds one with a non-zero amount of work in its public task queue.  The scheduler steals half of the tasks it finds at the victim.
 
-\TODO{Someone check this paragraph to make sure it is true. :-)  -Mark}
+\comment{Commenting out.  Simon things it doesn't add much I worried about its usefulness too.  -M
+
 \TODO{Anything after this I'm not sure adds to the paper -- if it does, please edit -Simon}
@@
 without syncing all tasks.
+}

### 2012-07-23 doc/asplos13/intro.tex
f
minor edits

@@
 off-the-shelf user-mode infiniband device driver stack~\cite{OFED}. MPI is
-used for process setup and tear down. GASnet~\cite{gasnet:spec} is used as the
+used for process setup and tear down. GASnet~\cite{gasnet} is used as the
 underlying mechanism for remote memory reads and writes using active message
@@
 throughput}: by {\em increasing} latency in key components of the system we
-are able to increase our aggregate random access bandwidth (by delaying and
-aggregating messages), our synchronization bandwidth (by delegating operations
+are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations
 to remote nodes), and our ability to improve load imbalance (work stealing

### 2012-07-23 doc/asplos13/conclude.tex
edits to conclusion

@@
 Irregular computations are both important and challenging to execute quickly.
-Scaling these applications easily to commodity hardware has been a historical challenge.
-Grappa is a runtime framework that simplifies this task for software developers and compiler writers.
-This paper describes the Grappa framework and its three main components: a task library, a distributed shared memory system, and a network aggregator to make commodity networks efficient with small message sizes.
-We explored the performance of Grappa on three algorithms: unbalanced tree search, breadth first search, and betweenness centrality.
-Performance comparisons to the Cray XMT for a small number of nodes (16 nodes / 96 cores) show great promise: Grappa performs on-par with the far more expensive and custom XMT system, sometimes achieving better performance and sometimes worse.
-Our initial scaling experiments to a higher number of nodes suggests Grappa does scale, but there is room for further improvement.
-Due to limitations in the initial aggregator design Grappa cannot utilize more than 6 cores/node, and this limits not only peak performance at a small number of nodes, but scaling performance at a larger number of nodes.
-Current work is ongoing on a new aggregator design.
+Scaling these applications easily to commodity hardware has been a historical
+challenge. Grappa is a runtime framework that simplifies this task for
+software developers and compiler writers. This paper describes the Grappa
+framework and its three main components: a task library, a distributed shared
+memory system, and a network aggregator to make commodity networks efficient
+with small message sizes. Grappa key aspect is extreme latency tolerance,
+which not only hides network latency but also enables the system to spend time
+on sophisticated work stealing and network optimizations, trading latency for
+even more throughput.
+
+We explored the performance of Grappa on three algorithms: unbalanced tree
+search, breadth first search, and betweenness centrality. Performance
+comparisons to the Cray XMT for a small number of nodes (16 nodes / 96 cores)
+show great promise: Grappa performs on-par with the far more expensive and
+custom XMT system, sometimes achieving better performance and sometimes worse.
+Our initial scaling experiments to a higher number of nodes suggests Grappa
+does scale, but there is some room for further improvement. Due to limitations
+in the initial aggregator design Grappa cannot utilize more than 6 cores/node,
+and this limits not only peak performance at a small number of nodes, but
+scaling performance at a larger number of nodes. Current work is ongoing on a
+new aggregator design.

### 2012-07-23 doc/asplos13/conclude.tex
c

@@
 custom XMT system, sometimes achieving better performance and sometimes worse.
-Our initial scaling experiments to a higher number of nodes suggests Grappa
-does scale, but there is some room for further improvement. Due to limitations
-in the initial aggregator design Grappa cannot utilize more than 6 cores/node,
-and this limits not only peak performance at a small number of nodes, but
-scaling performance at a larger number of nodes. Current work is ongoing on a
-new aggregator design.
+Our initial large scaling experiments to a higher number of nodes shows both promise and room for improvement.
+Grappa trounces XMT on UTS, but has a ways to go with achieving the same scaling performance on betweenness centrality.
+Grappa and XMT scale at similar rates on BFS, a key benchmark in this space.

### 2012-07-23 doc/asplos13/evaluation.tex
add the no aggreagiton uts plot`
added some plots

@@
     \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/uts_compare.pdf}
+      \includegraphics[width=0.5\textwidth]{figs/uts_performance.pdf}
     \end{center}
-    \caption{Performance of in-memory unbalanced tree search. Grappa
-        keeps scaling through 128 nodes and beats the XMT with 16
-        nodes.}
+    \caption{Performance of in-memory unbalanced tree search.}
     \label{fig:uts_compare}
@@
 \begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/bfs_compare}
+  \includegraphics[width=0.95\columnwidth]{figs/bfs_performance}
 \begin{minipage}{0.95\columnwidth}
@@
 \begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/centrality-performance}
+  \includegraphics[width=0.95\columnwidth]{figs/centrality_performance}
 \begin{minipage}{0.95\columnwidth}
@@
 
+\begin{figure}[htb]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/no_aggregation_uts.pdf}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:no-aggregation-uts} Performance of UTS on 16
+      nodes with and without Grappa's aggregation.}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
 \paragraph{Number of active tasks}
@@
 \begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/bfs_sweep_workers}
+  \includegraphics[width=0.95\columnwidth]{figs/worker_sweep}
 \begin{minipage}{0.95\columnwidth} 

### 2012-07-23 doc/asplos13/conclude.tex

@@
 Our initial large scaling experiments to a higher number of nodes shows both promise and room for improvement.
-Grappa trounces XMT on UTS, but has a ways to go with achieving the same scaling performance on betweenness centrality.
-Grappa and XMT scale at similar rates on BFS, a key benchmark in this space.
+Grappa trounces XMT on UTS, but while Grappa performance scales on BFS and betweenness centrality, limitations in the current aggregator design give the performance edge to XMT.
+Work is ongoing on further refining this component.

### 2012-07-23 doc/asplos13/evaluation.tex
prestons edits

@@
 poorly on the binomial tree because it is very deep and a traversal
-produces too little parallelism. Figure~\ref{fig:grappa-xmt-uts} shows
+produces too little parallelism. Figure~\ref{fig:uts_compare} shows
 the performance in terms of number of vertices visited per second versus
@@
 Figure~\ref{fig:bfs-performance} shows Grappa's performance on BFS
-compared to that of the XMT\footnote{The performance dip at 112 nodes is not an anomaly.
-It is highly repeatable.
-Despite extensive investigation on our part, we can find no reasonable explanation for it.}.
-There are
+compared to that of the XMT.\footnote{The performance dip at 112 nodes is not an anomaly; it is highly repeatable.
+Despite extensive investigation on our part, we can find no reasonable explanation for it.} There are
 two things to observe. First, At 128 XMT processors/cluster nodes,

### 2012-07-23 doc/asplos13/evaluation.tex
threshold discussion

@@
 \begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/bfs_sweep_flushticks}
+  \includegraphics[width=0.95\columnwidth]{figs/flushticks_sweep}
 \begin{minipage}{0.95\columnwidth}
@@
 
-\subsubsection{Work stealing parameters}
-
-\paragraph{Chunk size}
-
-It is important to steal multiple tasks at a time to both amortize the
-cost of stealing over the network and to spread out work quickly in a
-large system. Figure~\ref{fig:ut_chunksize} shows performance and
-stealing statistics for UTS on \checkme{30} nodes as we increase the stealing chunk size. Recall
-that a thief will take a number of tasks equal to the minimum of half
-the available work or the chunk size; steals fail only when the victim
-has fewer than 2 available tasks. As the scheduler is allowed to
-steal more work beyond 1 task, we see that performance increases up to 6x. This
-shows that the heuristic of stealing the oldest task from victims is
-insufficient alone when a tree-structured computation is imbalanced,
-as observed in \cite{UTS}. By observing sampled state in the execution
-trace, we find that a chunk size as low as 1 allows stealing to
-spread the load evenly across the cluster but cores spend much time
-underutilized as multiple workers wait for steal replies that
-utlimately return little new work.
-
-Performance plateaus before maximum steal amount is limited by the
-size of the victim's task queue. This indicates that artificially limiting steals
-to \checkme{128} tasks does not limit performance. Although a lower
-chunk size limits how quickly work spreads, for sufficient chunk size,
-the heuristic of stealing the oldest tasks from victims in tree-based computations allows for
-stolen work to expand quickly.
+%\subsubsection{Work stealing parameters}
+%
+%\paragraph{Chunk size}
+%
+%It is important to steal multiple tasks at a time to both amortize the
+%cost of stealing over the network and to spread out work quickly in a
+%large system. Figure~\ref{fig:ut_chunksize} shows performance and
+%stealing statistics for UTS on \checkme{30} nodes as we increase the stealing chunk size. Recall
+%that a thief will take a number of tasks equal to the minimum of half
+%the available work or the chunk size; steals fail only when the victim
+%has fewer than 2 available tasks. As the scheduler is allowed to
+%steal more work beyond 1 task, we see that performance increases up to 6x. This
+%shows that the heuristic of stealing the oldest task from victims is
+%insufficient alone when a tree-structured computation is imbalanced,
+%as observed in \cite{UTS}. By observing sampled state in the execution
+%trace, we find that a chunk size as low as 1 allows stealing to
+%spread the load evenly across the cluster but cores spend much time
+%underutilized as multiple workers wait for steal replies that
+%utlimately return little new work.
+%
+%Performance plateaus before maximum steal amount is limited by the
+%size of the victim's task queue. This indicates that artificially limiting steals
+%to \checkme{128} tasks does not limit performance. Although a lower
+%chunk size limits how quickly work spreads, for sufficient chunk size,
+%the heuristic of stealing the oldest tasks from victims in tree-based computations allows for
+%stolen work to expand quickly.
 
 %% UTS: chunk size
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/uts_chunksize.pdf}
-    \end{center}
-    \caption{Performance of UTS-Mem with varying maximum chunk size of
-    steals, run with 30 nodes, 6 cores per node, 4000 workers,
-    \checkme{6M flush ticks}}
-    \label{fig:uts_chunksize}
-\end{figure}
+%\begin{figure}[ht]
+%    \begin{center}
+%      \includegraphics[width=0.5\textwidth]{figs/uts_chunksize.pdf}
+%    \end{center}
+%    \caption{Performance of UTS-Mem with varying maximum chunk size of
+%    steals, run with 30 nodes, 6 cores per node, 4000 workers,
+%    \checkme{6M flush ticks}}
+%    \label{fig:uts_chunksize}
+%\end{figure}
 
@@
 task provides the potential to exploit locality when data for adjacent iterations is
-also adjacent in memory.
+also adjacent in memory. The ability to exploit this locality that
+exists in the application is an important advantage. We found that in UTS and BFS, increasing the
+threshold from 1 up to 8 or 16, respectively, increases performance by
+more than 60\%.
 
@@
 % uts threshold
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/uts_threshold.pdf}
-    \end{center}
-    \caption{Performance of UTS-Mem with varying parallel loop
-        threshold, run with 30 nodes, 6 cores per node, 4000 workers,
-    \checkme{6M flush ticks}}
-    \label{fig:uts_threshold}
-\end{figure}
+%\begin{figure}[ht]
+%    \begin{center}
+%      \includegraphics[width=0.5\textwidth]{figs/uts_threshold.pdf}
+%    \end{center}
+%    \caption{Performance of UTS-Mem with varying parallel loop
+%        threshold, run with 30 nodes, 6 cores per node, 4000 workers,
+%    \checkme{6M flush ticks}}
+%    \label{fig:uts_threshold}
+%\end{figure}
 

### 2012-07-23 doc/asplos13/intro.tex
revised intro

@@
 Our evaluation of Grappa shows that it runs several graph-crunching
-applications (classic examples of irregular behavior) very efficiently on a
-commodity cluster. Our yardstick for comparison is the XMT hardware itself.
-Using the same number of processors (in our case cores on a single multicore
-processor), Grappa is in the same ballpark as the XMT: on some graphs, such as
-the \checkme{WTF} Grappa is \checkme{2X} faster, while on others, such as the
-\checkme{WTF2} Grappa is \checkme{2X} slower. In Section~\ref{sec:results} we
-explore the factors that underpin this performance. But most importantly,
-application performance \emph{scales} with Grappa. For significantly less real
-world cost, users can \emph{add} more processors to a commodity cluster and
-easily outperform the XMT hardware on graph processing, while still saving
-significant dollars.
+applications (classic examples of irregular behavior) very efficiently
+on a commodity cluster. Our yardstick for comparison is the XMT hardware
+itself. Using the same number of network interfaces, Grappa is in the
+same ballpark as the XMT: For unbalanced tree search, Grappa is over 3X
+faster and shows greatly improved scalability; conversely, for breadth
+first search and betweenness centrality Grappa is 2.5X slower.  In
+Section~\ref{sec:results} we explore the factors that underpin this
+performance. Most importantly, however, for significantly less real
+world cost, users can \emph{add} significantly more processors to a
+commodity cluster than an XMT machine and use Grappa to achieve scalable
+performance.
 

### 2012-07-23 doc/asplos13/evaluation.tex
e
Hacked "scaling" in evaluation.

--- 2012-07-23 doc/asplos13/evaluation.tex
@@
 \subsection{Scaling}
-\paragraph{UTS-Mem}
-We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
-(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
-nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
-Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
-achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
-The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
-scaling up through 128 nodes, although scaling
-declines because of the unscalability of our aggregation mechanism as
-number of network endpoints increases.
+% Got rid of this discussion of UTS by itself, tried to work the highlights in below
+%\paragraph{UTS-Mem}
+%We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
+%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
+%nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
+%Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
+%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
+%The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
+%scaling up through 128 nodes, although scaling
+%declines because of the unscalability of our aggregation mechanism as
+%number of network endpoints increases. 
+%
+%Despite our efforts to tune the UTS implementation specific to the 
+%XMT, performance does not scale well with increasing processor count,
+%flattening out around 60 processors.  When we increase the size of
+%the tree from 100M to 4.2B, we find that performance does not improve,
+%suggesting that performance is not limited by task parallelism.
+%Cray's performance tools show an increasing number of memory
+%retry operations for failed synchronization operations generated by
+%the runtime, which create network contention.
+%
+To determine how Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
+For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
+For Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
+All of the benchmarks continue to improve out to 128 nodes for Grappa.
+UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while Grappa handles this by suspending tasks until messages return.
+For BFS and Centrality, the XMT scales approximately a constant factor better than Grappa.
+Fundamentally, aggregation as it is done in the current implementation of Grappa will not scale indefinitely, as increasing the number of destinations (individual cores across the cluster) does not scale because it requires more workers or time to form large enough messages.
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/scaling_cropped.pdf}
+    \end{center}
+    \caption{Scaling number of nodes: Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
+    \label{fig:uts_threshold}
+\end{figure}
 
-Despite our efforts to tune the UTS implementation specific to the 
-XMT, performance does not scale well with increasing processor count,
-flattening out around 60 processors.  When we increase the size of
-the tree from 100M to 4.2B, we find that performance does not improve,
-suggesting that performance is not limited by task parallelism.
-Cray's performance tools show an increasing number of memory
-retry operations for failed synchronization operations generated by
-the runtime, which create network contention.
 %Retries are performed by the memory controller when remote synchronization operations fail to find the full-bit associated with each memory location in the unavailable state.

### 2012-07-23 doc/asplos13/evaluation.tex

--- 2012-07-23 doc/asplos13/evaluation.tex
@@
 \subsection{Network aggregation performance}
-
-\begin{figure}[htbH]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/bfs_aggregator_value}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:bfs-aggregator} BFS performance with and without aggregator enabled \TODO{This graph sucks; rerun on actual data point and include in performance comparison}}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
 

### 2012-07-23 doc/asplos13/evaluation.tex
basic performance stuff

@@
 \section{Evaluation} \label{sec:evaluation}
+
+We evaluate Grappa in four ways. First, we evaluate its performance
+with respect to the XMT. Second, we explore the benefit that
+aggregation gives us. Third, we investigate sensitivity of two key
+parameters in the runtime. Fourth, we investigate scaling.
 
@@
 
-\subsubsection{Unbalanced tree search in memory}
-
-%Using 6 cores per node gave
-%Grappa up to a 20\% benefit at each data point, but Grappa could not run
-%well beyond 90 machines with this core configuration because of the
-%current network aggregator designa.
-
-We ran UTS-mem on Grappa and the XMT with a binomial and two geometric
-100M-vertex trees (T3L,T1L,T2L), a geometric 1.6B-vertex tree (T1XL),
-and a geometric 4.2B-vertex tree (T1XXL). Both systems do relatively
-poorly on the binomial tree because it is very deep and a traversal
-produces too little parallelism. Figure~\ref{fig:uts_compare} shows
-the performance in terms of number of vertices visited per second versus
-number of compute nodes. The Grappa results shown are for 6 cores per
-node and the best parameter values from a search over runtime and
-application parameters.
+To evaluate Grappa's preformance with respect to the XMT, we ran each
+of our three benchmarks on up to 16 nodes of each machine. Grappa used
+6 cores per node, with the best parameters chosen for each point. The
+XMT could not run on 2 nodes, so the point is omitted.
 
@@
 
-\TODO{ we still win with th=1 }
-
-\subsubsection{BFS}
+\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex tree
+(T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
+number of vertices visited per second versus number of compute
+nodes. Grappa is 3.2 times faster than the XMT at 16 nodes.
 
@@
 
-Figure~\ref{fig:bfs-performance} shows Grappa's performance on BFS
-compared to that of the XMT.\footnote{The performance dip at 112 nodes is not an anomaly; it is highly repeatable.
-Despite extensive investigation on our part, we can find no reasonable explanation for it.} There are
-two things to observe. First, At 128 XMT processors/cluster nodes,
-the XMT is  4.17 times faster than Grappa. This is due to the
-XMT's faster network: the XMT is able to issue 100 million network
-requests per second per node; at the 128 nodes datapoint, Grappa is
-aggregating 25 million network requests per second, or 4 times slower.
-\checkme{Second, Grappa is scaling...}
+\paragraph{BFS}
 
-\subsubsection{Centrality}
+We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and
+$2^{29}$ edges (25 GB of data). Figure~\ref{fig:bfs-performance} shows
+our performance in terms of graph edges traversed per second . The XMT
+is 2.5 times faster than Grappa at 16 nodes.
 
@@
 
-Figure~\ref{fig:centrality-performance} shows Grappa's performance on
-Betweenness Centrality compared to that of the XMT. At 128 XMT
+\subsubsection{Centrality}
+
+We ran Betweenness Centrality on the same scale 25 Kronecker graph as
+we did for BFS. Figure~\ref{fig:centrality-performance} shows our
+performance in terms of graph edges traversed per second. At 16 XMT
 processors/cluster nodes, the XMT is 1.75 times faster than Grappa.
-\TODO{Add more here...}
 

### 2012-07-23 doc/asplos13/relatedwork.tex
Minor spacing fixes (Preston)

@@
 \paragraph{Comparable frameworks} Distributed graph processing
-frameworks like Pregel \cite{pregel:2010} and Distributed GraphLab
-\cite{distgraphlab:vldb12} share similar goals as Grappa. Pregel adopts
-a bulk-synchronous parallel (BSP) execution model, which makes it
-inefficient on workloads that could prioritize vertices. GraphLab, on
-the other hand, schedules vertex computations individually, allowing
-prioritization, which gives faster convergence in a variety of iterative
-algorithms.  GraphLab, however, imposes a rigid computation model where
-programmers must express computation as transformations on a vertex and
-its edge list only, with only information from adjacent vertexes. Pregel
-is only slightly less restrictive, as the input data can be any vertex
-in the graph.  Grappa also supports dynamic parallelism with
-asynchronous execution, but parallelism is expressed as tasks or loop
-iterations, which is a far more general programming model
-for irregular computation tasks.
+frameworks like Pregel~\cite{pregel:2010} and Distributed
+GraphLab~\cite{distgraphlab:vldb12} share similar goals as
+Grappa. Pregel adopts a bulk-synchronous parallel (BSP) execution
+model, which makes it inefficient on workloads that could prioritize
+vertices. GraphLab, on the other hand, schedules vertex computations
+individually, allowing prioritization, which gives faster convergence
+in a variety of iterative algorithms.  GraphLab, however, imposes a
+rigid computation model where programmers must express computation as
+transformations on a vertex and its edge list only, with only
+information from adjacent vertexes. Pregel is only slightly less
+restrictive, as the input data can be any vertex in the graph.  Grappa
+also supports dynamic parallelism with asynchronous execution, but
+parallelism is expressed as tasks or loop iterations, which is a far
+more general programming model for irregular computation tasks.
 
@@
 hide the fact that they are built in software from applications by
-exploiting the processor's paging mechanisms, therefore relying heavily
-on locality. Instead, Grappa, like other partitioned global address
-space (PGAS) models, implements its DSM at the language, rather than
-system level. Languages such as Chapel \cite{Chamberlain:2007}, X10
-\cite{X10:2005}, and UPC \cite{upc:2005} make accesses to shared
-structures look like normal memory references. As we describe later,
-Grappa chooses a middle ground, where global addresses are explicit in
-the API and local accesses are emitted conventionally by the compiler. 
-While Grappa's DSM system is conceptually similar to prior work, its
+exploiting the processor's paging mechanisms, therefore relying
+heavily on locality. Instead, Grappa, like other partitioned global
+address space (PGAS) models, implements its DSM at the language,
+rather than system level. Languages such as
+Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and
+UPC~\cite{upc:2005} make accesses to shared structures look like
+normal memory references. As we describe later, Grappa chooses a
+middle ground, where global addresses are explicit in the API and
+local accesses are emitted conventionally by the compiler.  While
+Grappa's DSM system is conceptually similar to prior work, its
 implementation is tuned for irregular computations.  Past DSM work,
 being page-based, could exploit the RDMA capabilities of network
-hardware to move large page-sized blocks of data from node to node.  In
-our experience, when these networks move small blocks of data (a few
-bytes), only a fraction of the available bandwidth is achieved.  In
-addition, the DSM system in Grappa ends up being tightly coupled to the
-task scheduler in order to overlap long latency memory operations with
-useful computation.  For these two reasons it became necessary to build
-a new DSM system specifically for Grappa.
+hardware to move large page-sized blocks of data from node to node.
+In our experience, when these networks move small blocks of data (a
+few bytes), only a fraction of the available bandwidth is achieved.
+In addition, the DSM system in Grappa ends up being tightly coupled to
+the task scheduler in order to overlap long latency memory operations
+with useful computation.  For these two reasons it became necessary to
+build a new DSM system specifically for Grappa.
 
@@
 
-\paragraph{Multithreading} Grappa uses multithreading to tolerate memory
-latency. This is a well known technique. Hardware implementations
-include the Tera MTA \cite{tera:mta1}, Cray XMT \cite{feo:xmt},
-Simultaneous multithreading \cite{tullsen:smt}, MIT Alewife
-\cite{agarwal:alewife}, Cyclops \cite{almasi:cyclops}, and even GPUs
-\cite{gpus}. As we describe in this paper, Grappa use a lightweight
-user-mode task scheduler to multiplex \emph{thousands} of tasks on a
-single processing core. The large number of tasks is required because of
-the extremely high internode latency Grappa is mitigating.  Grappa's
-task library employs several optimizations: an extremely fast task
-switch, a small task size, and judicious use of hardware prefetching to
-bring task state into the cache long before that task is actually
-scheduled.
+\paragraph{Multithreading} Grappa uses multithreading to tolerate
+memory latency. This is a well known technique. Hardware
+implementations include the Tera MTA~\cite{tera:mta1}, Cray
+XMT~\cite{feo:xmt}, Simultaneous multithreading~\cite{tullsen:smt},
+MIT Alewife~\cite{agarwal:alewife}, Cyclops~\cite{almasi:cyclops}, and
+even GPUs~\cite{gpus}. As we describe in this paper, Grappa use a
+lightweight user-mode task scheduler to multiplex \emph{thousands\/} of
+tasks on a single processing core. The large number of tasks is
+required because of the extremely high internode latency Grappa is
+mitigating.  Grappa's task library employs several optimizations: an
+extremely fast task switch, a small task size, and judicious use of
+hardware prefetching to bring task state into the cache long before
+that task is actually scheduled.

### 2012-07-23 doc/asplos13/evaluation.tex
tiny change
scaling v2

@@
 For BFS and Centrality, the XMT scales approximately a constant factor better than Grappa.
-Fundamentally, aggregation as it is done in the current implementation of Grappa will not scale indefinitely, as increasing the number of destinations (individual cores across the cluster) does not scale because it requires more workers or time to form large enough messages.
+We attribute this to a limitation in the current aggregator design and network stack that Grappa uses.
+This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases} performance).
+Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
+Work is ongoing to rework the infiniband driver stack and aggregation interface to it, to remove this limitation.
+
 \begin{figure}[ht]

### 2012-07-23 doc/asplos13/evaluation.tex
Tiny changes.

--- 2012-07-23 doc/asplos13/evaluation.tex
@@
 
-To evaluate Grappa's preformance with respect to the XMT, we ran each
+To evaluate Grappa's performance with respect to the XMT, we ran each
 of our three benchmarks on up to 16 nodes of each machine. Grappa used
-6 cores per node, with the best parameters chosen for each point. The
-XMT could not run on 2 nodes, so the point is omitted.
+6 cores per node, with the best parameters chosen for each point. In
+some cases, the XMT could not run the benchmark with 2 nodes, so the
+point is omitted.
 
@@
 $2^{29}$ edges (25 GB of data). Figure~\ref{fig:bfs-performance} shows
-our performance in terms of graph edges traversed per second . The XMT
+our performance in terms of graph edges traversed per second. The XMT
 is 2.5 times faster than Grappa at 16 nodes.

### 2012-07-23 doc/asplos13/evaluation.tex
ev3
minor changes to the end of the scaling discussion

@@
 aggregation gives us. Third, we investigate sensitivity of two key
-parameters in the runtime. Fourth, we investigate scaling.
+parameters in the runtime. Fourth, we investigate scaling to more system nodes.
 
@@
 
-\subsubsection{Centrality}
+\paragraph{Centrality}
 
@@
 Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
-Work is ongoing to rework the infiniband driver stack and aggregation interface to it, to remove this limitation.
+Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
 

### 2012-07-23 doc/asplos13/evaluation.tex
spelling

@@
 allow other tasks to utilize its core. These tasks may also block. To
-support the many milleseconds of latency aggregation adds, we need to
+support the many milliseconds of latency aggregation adds, we need to
 support many thousands of blocked tasks. One of the key parameters of
@@
 granularity'') parameter tells the runtime when to stop creating new tasks and just execute iterations
-sequentially. This allows us to amoritize the overhead of task
+sequentially. This allows us to amortize the overhead of task
 creation. In addition, assigning sequential iterations to a single

### 2012-07-23 doc/asplos13/evaluation.tex

@@
 also compare against the OSU \texttt{osu\_mbw\_mr} benchmark
-\TODO{cite?}  compiled against OpenMPI 1.5.3 \TODO{cite?}; this
+\cite{osu:mpi}  compiled against OpenMPI 1.5.3; this
 benchmark has the same pattern of communication but doesn't have the
@@
 
-But this comparison is the best possible case for the aggregator. In
+This comparison is the best possible case for the aggregator. In
 order to verify that the aggregator still has value on actual
-applications at scale, we ran a small (scale 20 \TODO{run larger}) BFS
-with the aggregator disabled, with both 32 and 96
-nodes. Figure~\ref{fig:bfs-aggregator} shows the results. With 96
-nodes, the aggregator improves our application performance by 4 times.
+applications at scale, we ran a small (100M node tree) UTS-Mem
+with the aggregator disabled, on 16 nodes.
+Figure~\ref{fig:no-aggregation-uts}. At this configuration, the aggregator
+improves our application performance by 10x.
+
+\begin{figure}[htb]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{figs/no_aggregation_uts.pdf}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:no-aggregation-uts} Performance of UTS on 16
+      nodes with and without Grappa's aggregation.}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
 
@@
 timeout.
-
-\begin{figure}[htb]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/no_aggregation_uts.pdf}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:no-aggregation-uts} Performance of UTS on 16
-      nodes with and without Grappa's aggregation.}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
 

### 2012-07-23 doc/asplos13/evaluation.tex
e3

@@
 number of vertices visited per second versus number of compute
-nodes. Grappa is 3.2 times faster than the XMT at 16 nodes.
+nodes.
+Grappa is 3.2 times faster than the XMT at 16 nodes.
+As we will show later, the performance advantage Grappa has over XMT increases as more nodes are added.
+The main reason Grappa performs better is the software-based delegate synchronization obviates the need for the retry-based synchronization that XMT uses.
 
@@
 our performance in terms of graph edges traversed per second. The XMT
-is 2.5 times faster than Grappa at 16 nodes.
+is 2.5 times faster than Grappa at 16 nodes.  Performance does scale at a constant rate for Grappa, suggesting that adding more nodes will increase performance.
 

### 2012-07-23 doc/asplos13/evaluation.tex
eval scaling

@@
 destination decreases, and this gets harder. To bound the problem, the
-aggregator includes a timeout. Any packet waiting this long must be sent.
+aggregator includes a timeout. Any packet waiting this long is sent
+the next time the communications layer is serviced.
 
 Figure~\ref{fig:bfs-sweep-flushticks} shows a sweep of this parameter
-for 32-node and 96-node BFS runs. The optimal timeout value varies
-with the size of the job. In the 32-node job, performance peaks with a
-2 ms timeout; with the 96-node job, performance peaks with a 14 ms
-timeout.
-
-\paragraph{Number of active tasks}
+for UTS, BFS, and Betweenness Centrality on 16 nodes, using the
+datasets described previously. The maximum number of workers is fixed
+at 2048. All the benchmarks show a performance peak with a 2
+millisecond timeout; at this point we are delaying long enough to
+aggregate the largest packets we can; setting the parameter higher
+causes tasks to wait longer for responses, but few new requests are
+being generated.
 
@@
 
+\paragraph{Number of active tasks}
+
 When a task issues a request that requires a response, it blocks to
@@
 Figure~\ref{fig:bfs-sweep-workers} shows a sweep of the maximum number
-of active tasks per core, holding all other parameters
-constant. Again, the optimal value varies with the size of the
-run. For 32 nodes, a maximum of 3072 active tasks per core gives the
-best result; for 96 nodes, 16000 tasks is best.
+of active tasks (workers) per core for each of our three benchmarks on
+16 nodes. The aggregator timeout is set at 1 ms for UTS and 2 ms for
+BFS and Betweenness Centrality. The performance peak shifts in this
+case, with UTS peaking at 1536 workers, BFS peaking at 2048 workers,
+and Betweenness Centrality peaking at 3072 workers. This is the point
+where we have enough workers to cover the latency of aggreation. The
+different values reflect the different amounts of work done by a task
+in each benchmark; UTS does the least, while Betweenness Centrality does the most.
 
@@
 
-\TODO{(difference with BFS)}
+%\TODO{(difference with BFS)}
 

### 2012-07-24 doc/asplos13/intro.tex
i
fixed what i think were typos in 2nd to last paragraph.

@@
 a context switch in as few as 38ns and distributed global load balancing; (2)
-a \emph{distributed shared memory} layer that supports customary operations
-such as \emph{read and write as well as remote operations} such as
-fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message aggregation} layer
+a \emph{distributed shared memory} layer that supports normal access operations
+such as \emph{read} and \emph{write} as well as synchronizing operations such as \emph{fetch-and-add}\cite{fetchandadd}; and (3) a \emph{message aggregation} layer
 that combines short messages to mitigate the aforementioned problem that
@@
 first search and betweenness centrality Grappa is 2.5X slower.  In
-Section~\ref{sec:results} we explore the factors that underpin this
+Section~\ref{sec:evaluation} we explore the factors that underpin this
 performance. Most importantly, however, for significantly less real

### 2012-07-24 doc/asplos13/intro.tex
attempt to humour Preston.

@@
 invocations. To this commodity hardware and software mix Grappa adds three
-main software components: (1) a \emph{lightweight tasking} layer that supports
+main software components: (1) a \emph{lightweight tasking\/} layer that supports
 a context switch in as few as 38ns and distributed global load balancing; (2)
-a \emph{distributed shared memory} layer that supports normal access operations
-such as \emph{read} and \emph{write} as well as synchronizing operations such as \emph{fetch-and-add}\cite{fetchandadd}; and (3) a \emph{message aggregation} layer
+a \emph{distributed shared memory\/} layer that supports normal access operations
+such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer
 that combines short messages to mitigate the aforementioned problem that
@@
 the network. Therefore, Grappa can afford to \emph{trade latency for
-throughput}: by {\em increasing} latency in key components of the system we
+throughput\/}: by {\em increasing\/} latency in key components of the system we
 are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations
@@
 performance. Most importantly, however, for significantly less real
-world cost, users can \emph{add} significantly more processors to a
+world cost, users can \emph{add\/} significantly more processors to a
 commodity cluster than an XMT machine and use Grappa to achieve scalable

### 2012-10-16 doc/asplos13/rebuttal.txt
First pass at ASPLOS rebuttal

We thank our reviewers for their thoughtful comments.

Some reviewers (1, 4, 5) did not see a novel contribution in our
work. We believe that our system is the first runtime designed
specifically to tolerate latency by switching between thousands of
concurrent tasks per core. We are able to keep cores busy while
performing remote memory, synchronization, and work-stealing
operations. While the individual ideas used to build Grappa are
generally well-known, we know of no other runtime that combines these
ideas to trade latency for throughput.

Reviewer 3 takes issue with our characterization of our performance as
"on par" with that of the XMT. We are able to provide a simple,
general programming model similar to that of the XMT on
already-installed commodity clusters with at most a 2.5x slowdown,
even though the XMT has a 10x faster network injection rate and 20x
faster context switches than we do. We believe this slowdown is within
the realm of optimization---improvements to our network layer should
allow us to double the number of cores per node we use, and
improvements to our scheduler should allow us to increase the rate at
which each core injects messages into the network by ~30 percent.

Reviewer 2 asks about algorithmic modifications such as Beamer's BFS
transformation. Both the XMT and Grappa can benefit from such
transformations. Beamer's transformation increases BFS performance in
two ways: first, by reducing the number of edges traversed, and
second, by reducing the number of synchronizing operations
executed. Both the XMT and Grappa would benefit from the reduction in
edges traversed, but since Grappa's remote references are relatively
more costly than the XMT's, we would expect Grappa to benefit
more. Atomic operations on both platforms are no more expensive than
normal memory operations, so we would expect no significant benefit
from the reduction in synchronizing operations. Many transformations
also benefit by exploiting locality: Grappa provides mechanisms to do
so easily, while the hashed addressing of the XMT makes exploiting
locality difficult.

Reviewer 1 suggests comparing Grappa against other benchmark
implementations on commodity clusters in order to evaluate the benefit
of the programming model. In this work, we focused on providing a
similar programming model to the XMT, which is known to be simple
relative to distributed memory machines. We will explore programmer
productivity and compare against other commodity-cluster-based
implementations in future work.

Our evaluation was written in haste; in a final version we would
discuss key metrics including scheduling overhead and aggregator
performance limitations <Reviwer 2>.

We thank Reviewer 4 for expanding our understanding of the related
work. We will incorporate the suggestions.

[scalability?]

[citation for current memory hierarchy != irregular apps?]

### 2012-10-16 doc/asplos13/rebuttal.txt
myers first quick rebuttal pass

@@
 
-Some reviewers (1, 4, 5) did not see a novel contribution in our
-work. We believe that our system is the first runtime designed
-specifically to tolerate latency by switching between thousands of
-concurrent tasks per core. We are able to keep cores busy while
-performing remote memory, synchronization, and work-stealing
-operations. While the individual ideas used to build Grappa are
-generally well-known, we know of no other runtime that combines these
-ideas to trade latency for throughput.
+Some reviewers (1, 4, 5) did not see a novel contribution in our work. We
+believe that our system is the first runtime designed specifically to tolerate
+latency by switching between thousands of concurrent tasks per core. We are
+able to keep cores busy while performing remote memory, synchronization, and
+work-stealing operations. We aggregate small shared memory references into
+larger network messages to better utilize the network. While the individual
+ideas used to build Grappa are generally well-known, we know of no other
+runtime that combines these ideas to trade latency for throughput.
 
-Reviewer 3 takes issue with our characterization of our performance as
-"on par" with that of the XMT. We are able to provide a simple,
-general programming model similar to that of the XMT on
-already-installed commodity clusters with at most a 2.5x slowdown,
-even though the XMT has a 10x faster network injection rate and 20x
-faster context switches than we do. We believe this slowdown is within
-the realm of optimization---improvements to our network layer should
-allow us to double the number of cores per node we use, and
-improvements to our scheduler should allow us to increase the rate at
-which each core injects messages into the network by ~30 percent.
+Reviewer 3 takes issue with our characterization of our performance as "on
+par" with that of the XMT. We are able to provide a simple, general
+programming model similar to that of the XMT on already-installed commodity
+clusters with at most a 2.5x slowdown, even though the XMT has a 10x faster
+network injection rate and 20x faster context switches than we do. We believe
+this slowdown is within the realm of optimization of the runtime features
+already covered in the paper---improvements to our
+network layer should allow us to double the number of cores per node we use,
+and improvements to our scheduler should allow us to increase the rate at
+which each core injects messages into the network by ~30 percent.  [We should
+be precise about "similar programming model" to avoid more overstated claims,
+because in Grappa's current form it is not quite the same. The main difference
+is not a difficult one but it exists: in our benchmarks we take some locality into account: e.g. caching edgelist chunks]
 
@@
 
-Our evaluation was written in haste; in a final version we would
-discuss key metrics including scheduling overhead and aggregator
-performance limitations <Reviwer 2>.
+Our evaluation was written in haste; in a final version we would discuss key
+metrics including scheduling overhead and aggregator performance limitations
+<Reviwer 2>. The evaluation of the scheduler would break down scheduling
+overhead into checking the polling thread, checking other
+queues, doing load balancing, and context switch time, and compare the average overhead to the time a task executes between yields.
 

### 2012-10-16 doc/asplos13/rebuttal.txt
rebuttal, TODOs regarding evaluation

@@
 queues, doing load balancing, and context switch time, and compare the average overhead to the time a task executes between yields.
+[TODO: what we can say in rebuttal without more data:
+1. Why BC and BFS are worse while UTS is better than XMT. I think the story is
+that our injection rate is not as good and UTS wins because of XMT not scaling
+2. Why scaling of apps is BC < BFS < UTS.]
+[TODO: Mention that we already have the means to give a more explanatory
+evaluation: Grappa is instrumented and we have a lightweight tracing
+infrastructure.]
 

### 2012-10-16 doc/asplos13/rebuttal.txt
one small rebuttal change
some edits, reformatting

--- 2012-10-16 doc/asplos13/rebuttal.txt
@@
-We thank our reviewers for their thoughtful comments.
+We thank the reviewers for their thoughtful comments.
 
-Some reviewers (1, 4, 5) did not see a novel contribution in our work. We
-believe that our system is the first runtime designed specifically to tolerate
-latency by switching between thousands of concurrent tasks per core. We are
-able to keep cores busy while performing remote memory, synchronization, and
-work-stealing operations. We aggregate small shared memory references into
-larger network messages to better utilize the network. While the individual
-ideas used to build Grappa are generally well-known, we know of no other
-runtime that combines these ideas to trade latency for throughput.
+>>> What is the novelty of the work? (Reviewers 1, 4, 5)
 
-Reviewer 3 takes issue with our characterization of our performance as "on
-par" with that of the XMT. We are able to provide a simple, general
-programming model similar to that of the XMT on already-installed commodity
-clusters with at most a 2.5x slowdown, even though the XMT has a 10x faster
-network injection rate and 20x faster context switches than we do. We believe
-this slowdown is within the realm of optimization of the runtime features
-already covered in the paper---improvements to our
-network layer should allow us to double the number of cores per node we use,
-and improvements to our scheduler should allow us to increase the rate at
-which each core injects messages into the network by ~30 percent.  [We should
-be precise about "similar programming model" to avoid more overstated claims,
+We believe that our system is the first runtime designed specifically to
+tolerate latency by switching between thousands of concurrent tasks per core.
+We are able to keep cores busy while performing remote memory,
+synchronization, and work-stealing operations. We aggregate messages from
+unrelated threads into larger network messages to better utilize the network.
+While the individual ideas used to build Grappa are generally well-known, we
+know of no other runtime that combines these ideas to trade latency for
+throughput to build a general-purposed parallel runtime for large-scale applications.
+
+>>> Why a performance comparison with XMT? (Reviewer 3)
+
+We are able to provide a simple, general programming model similar to that of
+the XMT on ubiquitous commodity clusters with at most a 2.5x slowdown (and at
+best TODO faster), even though the XMT has a 10x faster network injection rate
+and 20x faster context switches than we do. We believe this slowdown is within
+the realm of optimization of the runtime features already covered in the
+paper---improvements to our network layer should allow us to double the number
+of cores per node we use, and improvements to our scheduler should allow us to
+increase the rate at which each core injects messages into the network by ~30
+percent.
+
+[This would just complicate things. I would leave it out... We should be
+precise about "similar programming model" to avoid more overstated claims,
 because in Grappa's current form it is not quite the same. The main difference
-is not a difficult one but it exists: in our benchmarks we take some locality into account: e.g. caching edgelist chunks]
+is not a difficult one but it exists: in our benchmarks we take some locality
+into account: e.g. caching edgelist chunks]
 
-Reviewer 2 asks about algorithmic modifications such as Beamer's BFS
-transformation. Both the XMT and Grappa can benefit from such
-transformations. Beamer's transformation increases BFS performance in
-two ways: first, by reducing the number of edges traversed, and
-second, by reducing the number of synchronizing operations
-executed. Both the XMT and Grappa would benefit from the reduction in
-edges traversed, but since Grappa's remote references are relatively
-more costly than the XMT's, we would expect Grappa to benefit
-more. Atomic operations on both platforms are no more expensive than
-normal memory operations, so we would expect no significant benefit
-from the reduction in synchronizing operations. Many transformations
-also benefit by exploiting locality: Grappa provides mechanisms to do
-so easily, while the hashed addressing of the XMT makes exploiting
-locality difficult.
+>>> Why not compare with other benchmarks that running commodity clusters?
+(Reviewer 1)
 
-Reviewer 1 suggests comparing Grappa against other benchmark
-implementations on commodity clusters in order to evaluate the benefit
-of the programming model. In this work, we focused on providing a
-similar programming model to the XMT, which is known to be simple
-relative to distributed memory machines. We will explore programmer
-productivity and compare against other commodity-cluster-based
-implementations in future work.
+In this work, we focused on providing a similar programming model to the XMT,
+which is known to be simple relative to distributed memory machines. We will
+explore programmer productivity and compare against other
+commodity-cluster-based implementations in future work.
 
-Our evaluation was written in haste; in a final version we would discuss key
-metrics including scheduling overhead and aggregator performance limitations
-<Reviwer 2>. The evaluation of the scheduler would break down scheduling
-overhead into checking the polling thread, checking other
-queues, doing load balancing, and context switch time, and compare the average overhead to the time a task executes between yields.
+[Can we say something more about how we would compare based on published results?]
 
-We thank Reviewer 4 for expanding our understanding of the related
-work. We will incorporate the suggestions.
+>>> Evaluation section is unpolished and incomplete. (Reviewer 2)
+
+We will cleanup and expand our evaluation. In the final version we will
+discuss key metrics including scheduling overhead and aggregator performance
+limitations. The evaluation of the scheduler would break down scheduling
+overhead into checking the polling thread, checking other queues, doing load
+balancing, and context switch time, and compare the average overhead to the
+time a task executes between yields.
+We will expand the evaluation with a detailed study of where time goes, since Grappa has a lightweight tracing infrastructure --- we did not add this to the evaluation due to lack of time.
+
+[You can explain all these points in the rebuttal. You just can't add more benchamrks or major results. TODO: what we can say in rebuttal without more data: 1. Why BC and BFS are
+worse while UTS is better than XMT. I think the story is that our injection
+rate is not as good and UTS wins because of XMT not scaling 2. Why scaling of
+apps is BC < BFS < UTS.]
+
+>>> What would be the impact of better algorithms (e.g., Beamer's BFS) in the
+performance comparisons? (Reviewer 2)
+
+Both the XMT and Grappa can benefit from such transformations. Beamer's
+transformation increases BFS performance in two ways: first, by reducing the
+number of edges traversed, and second, by reducing the number of synchronizing
+operations executed. Both the XMT and Grappa would benefit from the reduction
+in edges traversed, but since Grappa's remote references are relatively more
+costly than the XMT's, we would expect Grappa to benefit more. Atomic
+operations on both platforms are no more expensive than normal memory
+operations, so we would expect no significant benefit from the reduction in
+synchronizing operations. Many transformations also benefit by exploiting
+locality: Grappa provides mechanisms to do so easily, while the hashed
+addressing of the XMT makes exploiting locality difficult.
+
+>>> Missing related work. (Reviewer 4)
+
+Thanks for pointing to extra related work, we will add these to the final version.
 

### 2012-10-18 doc/asplos13/rebuttal.txt
submitted rebuttal

--- 2012-10-16 doc/asplos13/rebuttal.txt
@@
-We thank our reviewers for their thoughtful comments.
+Submitted:
 
-Some reviewers (1, 4, 5) did not see a novel contribution in our work. We
-believe that our system is the first runtime designed specifically to tolerate
-latency by switching between thousands of concurrent tasks per core. We are
-able to keep cores busy while performing remote memory, synchronization, and
-work-stealing operations. We aggregate small shared memory references into
-larger network messages to better utilize the network. While the individual
-ideas used to build Grappa are generally well-known, we know of no other
-runtime that combines these ideas to trade latency for throughput.
+We thank the reviewers for their thoughtful comments.
 
-Reviewer 3 takes issue with our characterization of our performance as "on
-par" with that of the XMT. We are able to provide a simple, general
-programming model similar to that of the XMT on already-installed commodity
-clusters with at most a 2.5x slowdown, even though the XMT has a 10x faster
-network injection rate and 20x faster context switches than we do. We believe
-this slowdown is within the realm of optimization of the runtime features
-already covered in the paper---improvements to our
-network layer should allow us to double the number of cores per node we use,
-and improvements to our scheduler should allow us to increase the rate at
-which each core injects messages into the network by ~30 percent.  [We should
-be precise about "similar programming model" to avoid more overstated claims,
-because in Grappa's current form it is not quite the same. The main difference
-is not a difficult one but it exists: in our benchmarks we take some locality into account: e.g. caching edgelist chunks]
+>>> What is the novelty of the work? (Reviewers 1, 3, 4, 5)
 
-Reviewer 2 asks about algorithmic modifications such as Beamer's BFS
-transformation. Both the XMT and Grappa can benefit from such
-transformations. Beamer's transformation increases BFS performance in
-two ways: first, by reducing the number of edges traversed, and
-second, by reducing the number of synchronizing operations
-executed. Both the XMT and Grappa would benefit from the reduction in
-edges traversed, but since Grappa's remote references are relatively
-more costly than the XMT's, we would expect Grappa to benefit
-more. Atomic operations on both platforms are no more expensive than
-normal memory operations, so we would expect no significant benefit
-from the reduction in synchronizing operations. Many transformations
-also benefit by exploiting locality: Grappa provides mechanisms to do
-so easily, while the hashed addressing of the XMT makes exploiting
-locality difficult.
+Conventional wisdom holds that to obtain performance on irregular applications one should minimize latency of individual memory and scheduling operations.
+In contrast, our approach delays and bundles these operations, exploiting the wealth of concurrency in the problem domain to overlap this additional latency with other computation and consequently achieve high overall throughput.
+We use this idea to obtain high throughput not only on remote memory operations, but also on remote synchronization and distributed work-stealing operations.
+We believe that our system is the first runtime designed specifically to tolerate multi-millisecond latencies by switching between millions of threads on hundreds to thousands of cores.
 
-Reviewer 1 suggests comparing Grappa against other benchmark
-implementations on commodity clusters in order to evaluate the benefit
-of the programming model. In this work, we focused on providing a
-similar programming model to the XMT, which is known to be simple
-relative to distributed memory machines. We will explore programmer
-productivity and compare against other commodity-cluster-based
-implementations in future work.
+While the individual ideas used to build Grappa are generally well-known, we know of no other runtime that combines these ideas with this goal.
+We combine ideas from: (1) distributed shared memory, but we use it for applications with little locality, i.e., caches are not necessarily valuable; (2) network message aggregation, but we use it for unrelated remote memory accesses instead of explicit messages, as in prior work; and (3) distributed work stealing, but with asynchronous steals that are overlapped with other computation via multithreading.
 
-Our evaluation was written in haste; in a final version we would discuss key
-metrics including scheduling overhead and aggregator performance limitations
-<Reviwer 2>. The evaluation of the scheduler would break down scheduling
-overhead into checking the polling thread, checking other
-queues, doing load balancing, and context switch time, and compare the average overhead to the time a task executes between yields.
+The original goal of the project was to mimic the Cray XMT with commodity processors using a combination of hardware and software support.
+We thought hardware support would be necessary due to the vast difference in hardware capability.
+Each Cray XMT processor supports 1024 outstanding memory operations, while the processors in our cluster support a few dozen.
+The XMT processors have dedicated storage for thread contexts, whereas ours must share the data cache.
+The XMT includes extensive support for fine-grained synchronization, including full-bits and the ability to atomically increment a word anywhere in the system with a single instruction, while our cluster’s atomic operations are limited to a single node’s shared memory domain and may stall the pipeline to enforce the memory consistency model.
+We find it encouraging that Grappa performs as well as it does compared to the XMT without depending on specialized hardware.
 
-We thank Reviewer 4 for expanding our understanding of the related
-work. We will incorporate the suggestions.
+>>> Why a performance comparison with XMT? (Reviewers 1, 3, 4)
 
-[scalability?]
+In this work, we focused on providing a programming model similar to that of the XMT, which is known to be simple relative to that of distributed memory machines.
+We are able to provide this programming model on ubiquitous commodity clusters with at worst a 2.5x slowdown (and at best 3.2x speedup) on 16 nodes, even though the XMT has a 10x faster network injection rate and zero-cost context switches.
+By comparing against the XMT, we can hold the programming model constant and evaluate the performance difference between custom and commodity hardware under that model.
+Future work will compare Grappa against other commodity solutions in terms of performance as well as ease of programming.
 
-[citation for current memory hierarchy != irregular apps?]
+>>> Is a 2.5x performance gap good or bad? What optimization opportunities exist? (Reviewers 1, 3)
+
+We believe our slowdown with respect to the XMT is within the realm of optimization of the runtime features discussed in the paper.
+We were limited to less than half of the available cores in our system by contention in our network layer; improving that should allow us to double our network injection rate per node.
+Experiments with our context switch code suggest we can obtain a further ~30 percent increase in network injection rate per core.
+
+Automatically tuning the runtime parameters is an interesting area of future work.
+We believe we can use core idleness and aggregated message sizes as signals to tune the worker count and aggregator delay, but it is not straightforward---low-work phases of execution as well as synchronization events may require temporary changes to these parameters to ensure rapid progress.
+A similar flavor of parameter exploration has been studied in previous work in task parallelism: e.g., parallel for-loops use granularity to trade off overhead of spawns, and adaptive work stealing schedulers switch between “work-first” and “help-first” policies.
+
+>>> Evaluation section is unpolished and incomplete. What about efficiency, scheduling overhead, and aggregator performance limitations? (Reviewers 1, 2, 3, 4)
+
+We will clean up and expand our evaluation.
+Grappa includes support for lightweight tracing and statistics collection, which was invaluable for debugging the performance of the runtime and benchmarks.
+We will use this to provide a detailed study of where time goes.
+An expanded evaluation will include analysis of scheduler overheads and execution efficiency, global memory and synchronization performance, and scaling challenges.
+We did not include this previously due to lack of time.
+
+>>> What would be the impact of better algorithms (e.g., Beamer's BFS) in the performance comparisons? (Reviewer 2)
+
+Since both the XMT and Grappa support a general programming model, both can benefit from such algorithmic modifications.
+Beamer's transformation increases BFS performance in two ways: first, by reducing the number of edges traversed, and second, by reducing the number of synchronizing operations executed.
+Both the XMT and Grappa would benefit from the reduction in edges traversed, but since Grappa's remote references are relatively more costly than the XMT's, we would expect Grappa to benefit more.
+Atomic operations on both platforms are no more expensive than normal memory operations, so we would expect no significant benefit from the reduction in synchronizing operations.
+Many transformations also benefit by exploiting locality: Grappa provides mechanisms to do so easily, while the hashed addressing of the XMT makes exploiting locality difficult.
+
+We will consider implementing Beamer’s approach for the final version.
+
+>>> Synchronization support (Reviewer 1)
+
+Grappa does have a synchronization construct like Cilk_sync which blocks a task until all its subtasks have completed.
+This is used to implement Grappa's parallel loop support.
+Grappa also supports fine-grained synchronization primitives, such as condition variables, via delegate operations, without memory fences or expensive retries as in the XMT.
+
+>>> Missing related work; reference for memory hierarchy challenges (Reviewers 4, 5)
+
+Thanks for pointing to extra related work; we will add these to the final version.
+
+We will also cite “Improving Memory Hierarchy Performance of Irregular Applications Using Data and Computation Reorderings” by Mellor-Crummey, Whalley, and Kennedy from IJPP 2001, which discusses the challenges of mapping irregular computation to commodity memory hierarchies.
+
+Unsubmitted additional material for future use:
+
+Finally, whether or not a purely software-based solution for commodity systems can offer customers of the Cray XMT a cost-effective alternative has been a long-standing question for those customers.
+The answer has been anything but obvious:  that is why they have been eager to support this work.
+The lack of immediacy in verdict is attributed to four technical challenges.
+
+First was the question of whether or not a commodity processor could tolerate the latency of large systems without additional hardware support.
+The obvious answer is “No:  each core is capable of issuing at most 10 memory operations before it stalls;  furthermore, each Nehalem memory controller can execute at most 32 memory operations (including software prefetches) before it stalls;  a Cray XMT processor with a network latency of only 2us and running at only 500MHz can issue up to 1024 memory references and still sometimes stalls;  clearly a 3GHz processor will need more than 32 outstanding memory references to tolerate the much greater latency of commodity networks.”  For quite some time, this response stood as the rationale for requesting that Intel processors be redesigned to get more references in flight.
+Indeed, there have been proposals to use FPGAs as surrogate memory controllers as a means of increasing memory concurrency to remote memory locations.
+Apparently -- surprisingly to some -- Grappa demonstrates that it is possible to effect the needed concurrency (several thousand remote memory references in flight) with sufficiently low overhead that a commodity system might in fact serve those customers’ purposes.
+
+Second was the question of whether or not those thousands of threads’ contexts would “blow out cache”, defeating the purpose of latency tolerance.
+Considering that a thousand threads might each execute many instructions each before any given thread that has yielded regains control of a core, one can imagine that both local and remote state that it had depended upon finding in cache to sustain performance might by then have been evicted.
+Again, it is surprising that this is not the case:  that while there is some performance degradation due to such evictions, degradation does not sufficiently degrade overall performance to negate the gains of latency tolerance.
+
+Third was the question of synchronization.
+The Cray XMT architecture includes several instructions each of which perform some form of full-bit synchronization and an int_fetch_add operation.
+With just one instruction, the XMT can increment a word of memory anywhere in the system.
+Hardware supports retry operations when full-bit operations find a word in the “wrong” state.
+Again, it is not obvious how to compete against this with a software-only system relying on processors whose atomic operations not only are limited in functionality but also induce fences, stalling the cpu pipelines until all outstanding stores are flushed.
+And, again, some folks have proposed FPGA solutions.
+Indeed, synchronization operations are dense in high-concurrency programs.
+We realized we could not use the x86 built-ins even for node-local data with any frequency without interfering with the very pipeline throughput we were aiming to improve through the use of concurrency.
+But, somewhat surprisingly considering all the additional operations involved, our software-only solution in which every synchronizing and every global reference passes through a Grappa delegate appears sufficiently efficient to compete.
+
+Fourth was the question of global memory model.
+The Cray XMT guarantees sequential consistency for race free programs.
+There are no caches so there are no coherency concerns.
+The compiler injects a “lookahead” number into every memory operation that expresses how many more operations from that thread the pipeline may execute before it must wait for the memory operation to complete.
+x86 processors have caches.
+There is no hardware support for lookahead numbers.
+We interweave threads outside the view of the compiler.
+What the x86 does provide is total store order.
+Putting all this together, it is not immediately clear what we get in the way of a consistency model.
+However, what we believe we have is the same sequential consistency for race free programs as is offered by the Cray XMT, using off the shelf compilers and no hardware support specific to multiprocessor systems.
+We believe without proof, yet, that we get this as a result of the total store order of individual processors combined with our delegation scheme that serializes all synchronizing and global requests.
+
+In summary, while there may well be no novelty in any one technique, we believe there is novelty in the fact that Grappa works as well as it does, likely offering a satisfactory commodity-based alternative to Cray XMT customers, even though we still have a great deal of work ahead.

### 2012-10-18 doc/asplos13/rebuttal.txt
add programming UTS-mem

@@
 
+--Comparing to XMT
+
 Finally, whether or not a purely software-based solution for commodity systems can offer customers of the Cray XMT a cost-effective alternative has been a long-standing question for those customers.
@@
 In summary, while there may well be no novelty in any one technique, we believe there is novelty in the fact that Grappa works as well as it does, likely offering a satisfactory commodity-based alternative to Cray XMT customers, even though we still have a great deal of work ahead.
+
+--Programming model 
+As an example, we discuss what a UPC implementation of our
+simplest benchmark, UTS-mem, (which only performs remote reads) would look
+like. UTS-mem is a parallel search of an unbalanced tree in memory, assuming
+no prior knowledge of the tree structure. UTS implementations use a work
+stack, containing addresses of nodes to visit. The UPC Thread would deqeue the
+global address and perform a high-latency remote read to get the node. With
+the node information, it then performs another remote read to get the node's
+edge list. The elements in the edge list can be added to the work stack. This
+simple solution will spend most of the cpu's time waiting for remote reads. An
+aggressive and fair implementation would need to overlap multiple reads for
+vertices and edge lists.** This would involve a scheduling loop that manages
+three structures: the work stack of unstarted vertices, the pool of
+outstanding reads, and the pool of received vertices and edge lists.  The
+Thread initiates nonblocking gets, polls for received messages (try_sync), and
+executes received vertices and edge lists.  The need of the extra data
+structures indicates that this level of communication overlap is too complex
+to be extracted from the simple version by UPC's compiler optimizations. The
+solution is further complicated by very large edge lists, which need to broken
+down into constant size chunks to prevent serialization.  These optimizations
+diverge from the UPC shared memory programming model.
+
+There is also a
+performance concern: the UPC solution is that just using nonblocking reads is
+not sufficient. Commodity networks do not get good bandwidth for small
+messages, and there is a limit to the number of outstanding RDMA operations.
+The UPC solution could address this by aggregating reads to the same node
+using active messages, but then you are building a general aggregation
+mechanism that Grappa provides in the network layer.
+
+**It might actually be possible to try manually unrolling the main loop of dequeue,read vertex,read edgelist,enqueue edges.
+By putting all the reads together in the unroll, the UPC compiler might be able to convert this gather into nonblocking ops.
+If so then it could be efficient in terms of the individual dependence from vertex to its edges.
+But all the work in one iteration may have to finish before the next iteration, in which case performance degradation is determined by the variability of non-uniform access times in the gather ops (dataflow within iteration, but synchronized iterations).
+Aggregation could be added by increasing the amount of unrolling in the loop.

### 2012-11-06 doc/asplos13/communication.tex

@@
-\section{Communication}
+\section{Communication}\label{sec:communication}
 
@@
 efficient use of the network. To make the best use of the network, we
-must convert our small messages into large ones
-(Section~\ref{sec:results-aggregation}).  When a task sends a message,
+must convert our small messages into large ones.
+When a task sends a message,
 it is not immediately sent, but rather placed in a queue specific to the
@@
 There are three situations in which a queue of aggregated messages is
-sent. First, each queue has a message size threshold, chosen to give
-reasonable network performance 4096. If the size in
+sent. First, each queue has a message size threshold of 4096 bytes, chosen to give reasonable network performance. If the size in
 bytes of a queue is above the threshold, the contents of the queue are

### 2012-12-11 doc/nvm2013/paper.tex
first commit of extended abstract for NVMW 2013

\input{header}
\usepackage{microtype}
\begin{document}

\title{Pomace: a Solid-State Grappa}
\authorinfo{Jacob Nelson$^{\dagger}$,
  Brandon Holt$^{\dagger}$,
  Brandon Myers$^{\dagger}$,
  Preston Briggs$^{\dagger}$,
  Luis Ceze$^{\dagger}$,
  Simon Kahan$^{{\dagger \ddagger}}$,
  Mark Oskin$^{\dagger}$
}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu}

\maketitle
%\begin{abstract}
%\input{abstract}
%\end{abstract}

We would like to present our software system Grappa, designed to scale
up emerging irregular applications such as graph analytics on
commodity clusters, and ask this community: What would be required to
adapt Grappa to exploit SSD, and how would doing so benefit this
application space?

\section{What is Grappa?}
Grappa is a latency tolerant runtime system for off-the-shelf clusters
that targets irregular applications, such as graph analytics.  These
applications are notoriously difficult to scale-up due to
data-dependent execution and memory access behaviors. The resulting
irregularity in computational demands hinders efforts to balance load
and to maintain spatial and temporal locality.  Resulting
inefficiencies are exacerbated at scale, as wasted energy rises with
the number of idling nodes, undersized network packets, and cache
misses.  Consequently, few irregular applications today make good use
of high-performance computers.  Those that do use complex, application
specific techniques to overlap computation with communication and were
developed by expert programmers who are in short supply.

Yet there is increasing demand for irregular applications.  Evidence
is the growing number of publications, conferences, and companies
focused on graph benchmarks, semantic databases, machine learning, and
“big data” discovery.

Grappa aims to mitigate the inefficiencies plaguing development and
execution of these important applications.  Grappa facilitates
expression of fine-grained task parallelism at the application
programming level.  Programmers attend only to expressing the
parallelism in their program, not to scheduling it.  At execution
time, Grappa’s task scheduler dynamically maps these tasks to x86
processor cores using work stealing heuristics similar to those
described in the literature.  What distinguishes Grappa from other
such runtime systems is that tasks implicitly yield at every long
latency operation, resulting in a context switch to another task.  In
this way, processors remain well-utilized even in the presence of
frequent references to remote memory.  Grappa thus tolerates many
microseconds of latency by multiplexing thousands of tasks per core.
In addition, Grappa aggregates references into packets destined for
common remote node locations.  In this way, Grappa’s design provides
for up to approximately 100 million remote references per second per
node using today’s commodity network hardware.  While similar latency
tolerance and aggregation techniques can be manually coded into some
applications, Grappa relieves the programmer of this burden.  In
addition, Grappa provides an API for reader-writer semantics for the
purpose of manipulating a local copy of any contiguous portion of the
shared address space, enabling programmers to exploit spatial and
temporal locality while maintaining a consistent global view.  With
these features, we have demonstrated performance on commodity systems
for highly irregular applications approaching that of the Cray XMT, a
fully custom system.

\subsection{What is the potential we imagine?}

While Grappa’s current design scales to the limit of available network
bandwidth, it does so only for applications whose working set resides
in RAM.  Unfortunately, provisioning systems with petabytes of RAM is
costly, energy inefficient, and likely to increase the time to prepare
for and recover from faults.

Emerging non-volatile memory technologies offer hope for an
alternative.  We imagine placing the entire data set on SSD or PCM and
either using Grappa directly to tolerate the latency much in the same
way as it is used now to tolerate network latency or to stage blocks
of data into RAM, using Grappa to tolerate the latency incurred by the
transfers.  We imagine the energy per available byte being much lower.
We imagine recovery from faults with most of the data residing in
non-volatile memory being faster.

Prior work only partially explores this potentiality. GraphChi
represents a class of approaches that use windowing methods in the
irregular application space, structuring the data on secondary storage
to support filling RAM with multiple interrelated subblocks sufficient
for each of a series of phases of computation.  Such approaches are
algorithmically efficient only when each successive phase requires
access to the entire data set.  MAGT addresses this limitation
introducing support for asynchronous computation on graphs under the
assumption that RAM is sized to accommodate data structures
proportional to vertex-count, still leaving edges on non-volatile
storage.  In addition, both methods apply only to single node systems:
Grappa provides a global view of memory on multinode systems.

\section{Applying non-volatile memory to Grappa}
The natural point of integration for non-volatile memories in Grappa
is to store the global heap in nonvolatile memory, since the heap
usually contains data that is infrequently modified. All accesses to
this shared heap in Grappa programs are done through an API. If the
access is to local memory, Grappa issues a prefetch for the address
and context-switches to other work while the prefetch is executing. If
the access is to memory stored on another node, Grappa issues a
message to that node requesting the data and context-switches to other
work until the reply arrives. Non-volatile memory would be accessed in
the same way: if a request is to memory stored in non-volatile memory,
Grappa would initiate the request and context-switch to other work
until the request had completed.

\subsection{Challenges in using commodity SSDs}
Grappa’s goal is to provide good irregular application performance on
commodity hardware. Unfortunately, commodity non-volatile memories are
difficult to use in this way for three reasons:

\paragraph{SSDs assume locality.} In our benchmarks, we find the average request size to be
less than 64 bytes. Today’s SSDs are optimized for reads in the
kilobyte range.

\paragraph{SSDs have a low request rate.} The fastest (cite) can
perform on the order of 1 million random reads per second, and many
are limited to less than 100K/s. In contrast, cores may be idle in
Grappa if each node cannot make 100 million random reads per second of
the global heap. Even the individual flash chips are slow: 50K/s (cite
micron).

\paragraph{SSDs have high overhead.} Commodity SSDs are usually treated
as normal disks---accesses must occur through the kernel, incurring a
high context switch overhead. Grappa depends on user-mode access to
the network layer. This capability has been explored in NVM research
(cite Moneta).

\subsection{Exploring the design space}

Previously we discussed the mechanism by which memory references in
Grappa could be issued to non-volatile memory devices. In this
section, we explore a number of different designs for interfacing the
non-volatile memory hardware with Grappa nodes. Each design has
tradeoffs, which we briefly discuss.

For simplicity, we omit discussion of write and erase operations. We
assume a non-volatile Grappa would store frequently-written temporary
data in DRAM and would use its standard latency-tolerance techniques
to cover the cost of infrequent NVM writes.

\subsubsection{Tolerating Flash-based SSD latency} 
A natural first step would be to add one or more SSDs to each Grappa
cluster node, and use Grappa's global memory API to expose concurrent
accesses and context-switch to tolerate the latency of SSD
access. Unfortunately this is unlikely to work well, for two
reasons. First, the SSD must accept requests at a rate sufficient to
keep the cores busy. Grappa's target is for each cluster node to
generate 100 million random 8-byte requests per second, but most SSDs
can accept requests at a rate two to four orders of magnitude slower.
Second, the SSD must be able to support many requests operating in
parallel. Little's Law suggests that if requests take 20 $\mu$s for
the SSD to service, and if the node generates requests at a rate of
100 million per second, the SSDs must support 2000 concurrent
requests. Again, existing SSDs support orders of magnitude less request concurrency.

We see three ways to address this disparity.
\paragraph{Scaling up the Flash array}
Existing Flash-based SSDs are designed for large block requests. An
SSD for Grappa could be designed to support high concurrency, small
requests. Where a traditional SSD might put multiple Flash chips on
the same bus, or use Flash chips in parallel to build a wider,
high-bandwidth bus, a Grappa SSD would try to maximize parallelism by
putting each chip on its own bus. Serial Flash might be used to
minimize pin count requirements, at the cost of some additional
latency. This design could scale to request rates, but the challenge
would be chip count: to support 100M reads/second, an array of
20$\mu$s flash would require 2000 chips. This is possible, but expensive.

\paragraph{Scaling down the processors}
Alternatively, rather than scaling up the flash array, processors
could be scaled down to match the flash request rate. A low-power
processor with few cores might generate only 1 million references per
second. This could be satisfied with only 20 20$\mu$s Flash chips.

There are two problems with this design. First, while each node
requires fewer chips, the overall cluster performance goal, and thus
chip requirement, is the same. Second,

\paragraph{Buffering requests to increase locality.}
Grappa trades latency for better throughput by aggregating network
communications for better utilization and by migrating synchronization
operations for reduced contention. The same idea could be applied to
increase locality and amortize the cost of Flash accesses. Requests
to common flash blocks would be bucketed by flash block address for a
limited time. When a sufficient number of requests arrived for a block
or when requests timed out, the requests would be issued to the Flash
layer.

While similar techniques have provided benefit in prior work in
restricted situations (cite), they depend on a sufficient request
density to each block to get a benefit. As datasets scale up, the
request density may decrease and the benefit of this technique may
decrease.

\subsubsection{Using future memory technologies}

While Grappa's goal so far has been to leverage commodity hardware, it
is worth exploring the potential benefit of emerging non-volatile
memory technologies like Phase-Change Memory. The advantages of PCM
are twofold: first, read latency is much closer to that of DRAM, and
second, PCM supports finer-grained accesses.

\paragraph{PCM as Main Memory}
Much recent research has explored the use of PCM as a main memory
replacement. Grappa would easily be able to take advantage of such a
development. When a Grappa program references a low-locality program,
a prefetch request is issued for that data a few cycles before the
thread is context-switched in\cite{Nelson:hotpar2011}. A PCM-based
Grappa would do the same, but would issue the prefetches earlier to
cover the longer latency of PCM.

\paragraph{PCM as attached storage}

Grappa could also support a PCM array in a separate address space (for
instance, on the PCIe bus). PCM chips available today (cite) support a
64-bit read in 200 ns; this means that 20 chips acting independently
would be sufficient to reach our target request rate.

Just as in network communications, issuing individual PCIe bus
operations for each memory request has too high overhead to get a
reasonable request rate.  Instead, Grappa would delay threads
performing operations in the PCM array until it could send a buffer
with enough requests to amortize the overhead. The PCM array would
perform the operations and buffer the results. When a sufficient
amount have completed, the PCM array would write the results to DRAM,
and one core would deliver the results to the requesting threads,
waking them.

An ideal array would support limited atomic operations on 64-bit
words, including increment and compare-and-swap.

This system could be built today, with the only downside being
density---currently available PCM chips are significantly less dense
than Flash or even DRAM. PCM's technology scaling curve suggests that
this will be an ideal solution in the future.

\section{Conclusion}
Efficient execution of irregular applications on large datasets is of
growing importance.  Grappa promises to address this need on mass
market clusters.  Addition of non-volatile storage is a natural
extension of such systems.  We ask, what would developing a Pomace for
Grappa -- a Grappa system exploiting non-volatile memory -- entail and
what capability and efficiency improvements might it bring to Grappa’s
application space.  Answering this question is an opportunity for
collaboration between our group and the NVMW community.

%\subsection{References}
\bibliographystyle{abbrv}
\bibliography{paper}

\end{document}

### 2012-12-11 doc/nvm2013/paper.tex
edits mostly to first section

@@
 
-\title{Pomace: a Solid-State Grappa}
+\title{Pomace: a Grappa for Non-Volatile Memory}
 \authorinfo{Jacob Nelson$^{\dagger}$,
@@
 \maketitle
-%\begin{abstract}
-%\input{abstract}
-%\end{abstract}
-
-We would like to present our software system Grappa, designed to scale
-up emerging irregular applications such as graph analytics on
-commodity clusters, and ask this community: What would be required to
-adapt Grappa to exploit SSD, and how would doing so benefit this
-application space?
+\begin{abstract}
+We revisit Grappa, our software platform that scales in-memory irregular
+applications on commodity clusters.
+We contemplate options for adapting Grappa to use non-volatile memory technologies.
+\end{abstract}
 
@@
 Grappa is a latency tolerant runtime system for off-the-shelf clusters
-that targets irregular applications, such as graph analytics.  These
+that targets in-memory irregular applications, such as interactive graph analytics.  These
 applications are notoriously difficult to scale-up due to
@@
 misses.  Consequently, few irregular applications today make good use
-of high-performance computers.  Those that do use complex, application
-specific techniques to overlap computation with communication and were
-developed by expert programmers who are in short supply.
+of high-performance computers.  Those that do, use complex, application-specific techniques to overlap computation with communication and are
+developed by expert programmers, who are in short supply.
 
-Yet there is increasing demand for irregular applications.  Evidence
+Yet, there is increasing demand for irregular applications. Evidence
 is the growing number of publications, conferences, and companies
@@
 “big data” discovery.
+Grappa aims to mitigate the inefficiencies plaguing development and execution of these important applications.
+The system (Figure~\ref{fig:grappa}) has three main software components:
 
-Grappa aims to mitigate the inefficiencies plaguing development and
-execution of these important applications.  Grappa facilitates
+\begin{figure}[t]
+\begin{center}
+  \includegraphics[width=0.95\columnwidth]{../asplos13/figs/system-overview}
+\begin{minipage}{0.95\columnwidth}
+  \caption{\label{fig:grappa} Grappa system overview}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
+
+\begin{description}
+
+\item [Tasking system.] Our tasking system supports lightweight
+multithreading to tolerate communication latency and global distributed
+workstealing (i.e., tasks can be stolen from any node in the system), which
+provides automated load balancing.  Grappa thus tolerates many
+microseconds of latency by multiplexing thousands of tasks per x86 core.
+
+\item[Distributed shared memory.] Our DSM system provides support for
+fine-grain access to data anywhere in the system. It supports synchronization
+operations on global data, explicit local caching of any memory in the system,
+and support for operations on remote data (delegating operations to home node).
+By tight integration with the tasking system and the
+communication layer, our DSM system offers high aggregate random
+access bandwidth to remote data.
+
+\item[Communication layer.] Modern commodity networks
+support high bandwidth only for large messages and irregular applications
+frequently issue small requests.
+Our communication layer aggregates these small requests into larger messages to better exploit network bandwidth, targeting 100 million remote references per second per node.
+Aggregation is largely invisible to the application programmer.
+
+% software developer but helps to improve network performance when applications read and write only small pieces of data.
+
+\end{description}
+
+Grappa thus facilitates
 expression of fine-grained task parallelism at the application
-programming level.  Programmers attend only to expressing the
-parallelism in their program, not to scheduling it.  At execution
-time, Grappa’s task scheduler dynamically maps these tasks to x86
-processor cores using work stealing heuristics similar to those
-described in the literature.  What distinguishes Grappa from other
-such runtime systems is that tasks implicitly yield at every long
-latency operation, resulting in a context switch to another task.  In
-this way, processors remain well-utilized even in the presence of
-frequent references to remote memory.  Grappa thus tolerates many
-microseconds of latency by multiplexing thousands of tasks per core.
-In addition, Grappa aggregates references into packets destined for
-common remote node locations.  In this way, Grappa’s design provides
-for up to approximately 100 million remote references per second per
-node using today’s commodity network hardware.  While similar latency
+programming level, employing that concurrency as needed to sustain throughput in the presence of long latency operations and to better utilize commodity networks.
+Programmers attend only to expressing the parallelism in their program, not to scheduling it.
+While similar latency
 tolerance and aggregation techniques can be manually coded into some
@@
 
-\subsection{What is the potential we imagine?}
+\subsection{What's the problem?}
 
@@
 alternative.  We imagine placing the entire data set on SSD or PCM and
-either using Grappa directly to tolerate the latency much in the same
-way as it is used now to tolerate network latency or to stage blocks
+either using Grappa directly to tolerate the latency, much in the same
+way as it is used now to tolerate network latency, or to stage blocks
 of data into RAM, using Grappa to tolerate the latency incurred by the
@@
 algorithmically efficient only when each successive phase requires
-access to the entire data set.  MAGT addresses this limitation
+access to the entire data set.  MAGT addresses this limitation by
 introducing support for asynchronous computation on graphs under the
@@
 requires fewer chips, the overall cluster performance goal, and thus
-chip requirement, is the same. Second,
+chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
 

### 2012-12-11 doc/nvm2013/paper.tex
changes (largely) to section 1

@@
 
-While Grappa’s current design scales to the limit of available network
+While Grappa's current design scales to the limit of available network
 bandwidth, it does so only for applications whose working set resides
@@
 \subsection{Challenges in using commodity SSDs}
-Grappa’s goal is to provide good irregular application performance on
+Grappa's goal is to provide good irregular application performance on
 commodity hardware. Unfortunately, commodity non-volatile memories are
@@
 
-\subsubsection{Using future memory technologies}
+v\subsubsection{Using future memory technologies}
 
@@
 Grappa -- a Grappa system exploiting non-volatile memory -- entail and
-what capability and efficiency improvements might it bring to Grappa’s
+what capability and efficiency improvements might it bring to Grappa's
 application space.  Answering this question is an opportunity for

### 2012-12-11 doc/nvm2013/paper.tex
get rid of last bad apostrophe
add graphchi and magt citations

@@
 
-Prior work only partially explores this potentiality. GraphChi
+Prior work only partially explores this potentiality. GraphChi~\cite{graphchi:osdi12}
 represents a class of approaches that use windowing methods in the
@@
 algorithmically efficient only when each successive phase requires
-access to the entire data set.  MAGT addresses this limitation by
+access to the entire data set.  MAGT~\cite{magt:2010} addresses this limitation by
 introducing support for asynchronous computation on graphs under the
@@
 \paragraph{SSDs assume locality.} In our benchmarks, we find the average request size to be
-less than 64 bytes. Today’s SSDs are optimized for reads in the
+less than 64 bytes. Today's SSDs are optimized for reads in the
 kilobyte range.
@@
 
-v\subsubsection{Using future memory technologies}
+\subsubsection{Using future memory technologies}
 

### 2012-12-11 doc/nvm2013/paper.tex
holt's comments on sec 1 & a bit of whitespace removal

@@
 
-\title{Pomace: a Grappa for Non-Volatile Memory}
+\title{Pomace: a Grappa for Non-Volatile Memory \vspace{-.2in}}
 \authorinfo{Jacob Nelson$^{\dagger}$,
@@
 }{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
-  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu}
+  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu \vspace{-1in}
+}
 
@@
 \end{abstract}
+\vspace{-.1in}
 
@@
 
-\item[Distributed shared memory.] Our DSM system provides support for
-fine-grain access to data anywhere in the system. It supports synchronization
-operations on global data, explicit local caching of any memory in the system,
-and support for operations on remote data (delegating operations to home node).
-By tight integration with the tasking system and the
-communication layer, our DSM system offers high aggregate random
-access bandwidth to remote data.
-
 \item[Communication layer.] Modern commodity networks
@@
 Aggregation is largely invisible to the application programmer.
+
+\item[Distributed shared memory.] Our DSM system provides support for
+fine-grain access to data anywhere in the system. It supports synchronization
+operations on global data, explicit local caching of any memory in the system,
+and support for operations on remote data (delegating operations to owner node).
+By tight integration with the tasking system and the
+communication layer, our DSM system offers high aggregate random
+access bandwidth to remote data.
 
@@
 applications, Grappa relieves the programmer of this burden.  In
-addition, Grappa provides an API for reader-writer semantics for the
+addition, Grappa provides an API for reader-writer lock semantics for the
 purpose of manipulating a local copy of any contiguous portion of the
@@
 access is to local memory, Grappa issues a prefetch for the address
-and context-switches to other work while the prefetch is executing. If
+and context-switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If
 the access is to memory stored on another node, Grappa issues a
@@
 
-\paragraph{SSDs have a low request rate.} The fastest (cite) can
+\paragraph{SSDs have a low request rate.} The fastest~\cite{fusionio} can
 perform on the order of 1 million random reads per second, and many
@@
 Grappa if each node cannot make 100 million random reads per second of
-the global heap. Even the individual flash chips are slow: 50K/s (cite
-micron).
+the global heap. Even the individual flash chips are slow, on the order of 40K/s~\cite{micronFlash}.
 
@@
 high context switch overhead. Grappa depends on user-mode access to
-the network layer. This capability has been explored in NVM research
-(cite Moneta).
+the network layer. This capability has been explored in recent research~\cite{caulfield:2012}.
 
@@
 data in DRAM and would use its standard latency-tolerance techniques
-to cover the cost of infrequent NVM writes.
+to cover the cost of infrequent writes.
 
@@
 
-There are two problems with this design. First, while each node
+There are two problems with this design for our purpose. First, while each node
 requires fewer chips, the overall cluster performance goal, and thus
@@
 While similar techniques have provided benefit in prior work in
-restricted situations (cite), they depend on a sufficient request
+restricted situations~\cite{magt:2010,graphchi:osdi12}, they depend on a sufficient request
 density to each block to get a benefit. As datasets scale up, the
@@
 a prefetch request is issued for that data a few cycles before the
-thread is context-switched in\cite{Nelson:hotpar2011}. A PCM-based
+thread is context-switched in. A PCM-based
 Grappa would do the same, but would issue the prefetches earlier to
@@
 Grappa could also support a PCM array in a separate address space (for
-instance, on the PCIe bus). PCM chips available today (cite) support a
+instance, on the PCIe bus). PCM chips available today~\cite{micronPCM} support a
 64-bit read in 200 ns; this means that 20 chips acting independently
@@
 
-An ideal array would support limited atomic operations on 64-bit
-words, including increment and compare-and-swap.
+%An ideal array would support limited atomic operations on 64-bit
+%words, including increment and compare-and-swap.
 

### 2012-12-11 doc/nvm2013/paper.tex
final bib stuff

--- 2012-12-11 doc/nvm2013/paper.tex
@@
 high context switch overhead. Grappa depends on user-mode access to
-the network layer. This capability has been explored in NVM research
-(cite Moneta).
+the network layer. This capability has been explored in NVM research~\cite{caulfield:2012}.
 
@@
 While similar techniques have provided benefit in prior work in
-restricted situations (cite), they depend on a sufficient request
+restricted situations~\cite{magt:2010,graphchi:osdi2012}, they depend on a sufficient request
 density to each block to get a benefit. As datasets scale up, the
@@
 a prefetch request is issued for that data a few cycles before the
-thread is context-switched in\cite{Nelson:hotpar2011}. A PCM-based
+thread is context-switched in~\cite{Nelson:hotpar2011}. A PCM-based
 Grappa would do the same, but would issue the prefetches earlier to

### 2012-12-11 doc/nvm2013/paper.tex
narrow margins to fit on two pages
a few fixes

--- 2012-12-11 doc/nvm2013/paper.tex
@@
 
-\title{Pomace: a Grappa for Non-Volatile Memory \vspace{-.2in}}
+\title{Pomace: a Grappa for Non-Volatile Memory}
 \authorinfo{Jacob Nelson$^{\dagger}$,
@@
 }{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
-  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu \vspace{-1in}
-}
+  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu}
 
@@
 \end{abstract}
-\vspace{-.1in}
 
@@
 
+\item[Distributed shared memory.] Our DSM system provides support for
+fine-grain access to data anywhere in the system. It supports synchronization
+operations on global data, explicit local caching of any memory in the system,
+and support for operations on remote data (delegating operations to home node).
+By tight integration with the tasking system and the
+communication layer, our DSM system offers high aggregate random
+access bandwidth to remote data.
+
 \item[Communication layer.] Modern commodity networks
@@
 Aggregation is largely invisible to the application programmer.
-
-\item[Distributed shared memory.] Our DSM system provides support for
-fine-grain access to data anywhere in the system. It supports synchronization
-operations on global data, explicit local caching of any memory in the system,
-and support for operations on remote data (delegating operations to owner node).
-By tight integration with the tasking system and the
-communication layer, our DSM system offers high aggregate random
-access bandwidth to remote data.
 
@@
 applications, Grappa relieves the programmer of this burden.  In
-addition, Grappa provides an API for reader-writer lock semantics for the
+addition, Grappa provides an API for reader-writer semantics for the
 purpose of manipulating a local copy of any contiguous portion of the

### 2012-12-11 doc/nvm2013/paper.tex
squish some more

@@
 \begin{document}
+\addtolength{\topmargin}{-.7in}
+\addtolength{\textheight}{.7in}
+\addtolength{\oddsidemargin}{-.25in}
+\addtolength{\evensidemargin}{-.25in}
+\addtolength{\textwidth}{.5in}
+\linespread{.94}
 
-\title{Pomace: a Grappa for Non-Volatile Memory}
+\title{Pomace: a Grappa for Non-Volatile Memory \vspace{-.2in}}
 \authorinfo{Jacob Nelson$^{\dagger}$,

### 2012-12-12 doc/nvm2013/paper.tex
incorporate Preston's comments

@@
 
-\title{Pomace: a Grappa for Non-Volatile Memory \vspace{-.2in}}
+\title{Pomace: A Grappa for Non-Volatile Memory \vspace{-.2in}}
 \authorinfo{Jacob Nelson$^{\dagger}$,
@@
 \section{What is Grappa?}
-Grappa is a latency tolerant runtime system for off-the-shelf clusters
+Grappa is a latency-tolerant runtime system for off-the-shelf clusters
 that targets in-memory irregular applications, such as interactive graph analytics.  These
-applications are notoriously difficult to scale-up due to
+applications are notoriously difficult to scale up due to
 data-dependent execution and memory access behaviors. The resulting
@@
 focused on graph benchmarks, semantic databases, machine learning, and
-“big data” discovery.
+``big data'' discovery.
 Grappa aims to mitigate the inefficiencies plaguing development and execution of these important applications.
@@
 
-\item[Distributed shared memory.] Our DSM system provides support for
+\item[Distributed shared memory.] Our \textls*{\small DSM} system provides support for
 fine-grain access to data anywhere in the system. It supports synchronization
@@
 By tight integration with the tasking system and the
-communication layer, our DSM system offers high aggregate random
-access bandwidth to remote data.
+communication layer, our \textls*{\small DSM} system offers high aggregate random-access bandwidth to remote data.
 
@@
 \end{description}
-
 Grappa thus facilitates
@@
 Programmers attend only to expressing the parallelism in their program, not to scheduling it.
-While similar latency
-tolerance and aggregation techniques can be manually coded into some
+While similar latency-tolerance and aggregation techniques can be manually coded into some
 applications, Grappa relieves the programmer of this burden.  In
-addition, Grappa provides an API for reader-writer lock semantics for the
+addition, Grappa provides an \textls*{\small API} for reader-writer lock semantics for the
 purpose of manipulating a local copy of any contiguous portion of the
@@
 these features, we have demonstrated performance on commodity systems
-for highly irregular applications approaching that of the Cray XMT, a
+for highly irregular applications approaching that of the Cray \textls*{\small XMT}~\cite{feo:xmt}, a
 fully custom system.
@@
 bandwidth, it does so only for applications whose working set resides
-in RAM.  Unfortunately, provisioning systems with petabytes of RAM is
+in \textls*{\small RAM}\@.  Unfortunately, provisioning systems with petabytes of \textls*{\small RAM} is
 costly, energy inefficient, and likely to increase the time to prepare
@@
 Emerging non-volatile memory technologies offer hope for an
-alternative.  We imagine placing the entire data set on SSD or PCM and
+alternative.  We imagine placing the entire data set on \textls*{\small SSD} or \textls*{\small PCM} and
 either using Grappa directly to tolerate the latency, much in the same
 way as it is used now to tolerate network latency, or to stage blocks
-of data into RAM, using Grappa to tolerate the latency incurred by the
+of data into \textls*{\small RAM}, using Grappa to tolerate the latency incurred by the
 transfers.  We imagine the energy per available byte being much lower.
@@
 irregular application space, structuring the data on secondary storage
-to support filling RAM with multiple interrelated subblocks sufficient
+to support filling \textls*{\small RAM} with multiple interrelated subblocks sufficient
 for each of a series of phases of computation.  Such approaches are
 algorithmically efficient only when each successive phase requires
-access to the entire data set.  MAGT~\cite{magt:2010} addresses this limitation by
+access to the entire data set.  Pearce et al.~\cite{magt:2010} address this limitation by
 introducing support for asynchronous computation on graphs under the
-assumption that RAM is sized to accommodate data structures
+assumption that \textls*{\small RAM} is sized to accommodate data structures
 proportional to vertex-count, still leaving edges on non-volatile
@@
 usually contains data that is infrequently modified. All accesses to
-this shared heap in Grappa programs are done through an API. If the
+this shared heap in Grappa programs are done through an \textls*{\small API}\@. If the
 access is to local memory, Grappa issues a prefetch for the address
-and context-switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If
+and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If
 the access is to memory stored on another node, Grappa issues a
-message to that node requesting the data and context-switches to other
+message to that node requesting the data and context switches to other
 work until the reply arrives. Non-volatile memory would be accessed in
-the same way: if a request is to memory stored in non-volatile memory,
-Grappa would initiate the request and context-switch to other work
+the same way: If a request is to memory stored in non-volatile memory,
+Grappa would initiate the request and context switch to other work
 until the request had completed.
 
-\subsection{Challenges in using commodity SSDs}
+\subsection{Challenges in using commodity \textls*{\small SSD}s}
 Grappa's goal is to provide good irregular application performance on
@@
 
-\paragraph{SSDs assume locality.} In our benchmarks, we find the average request size to be
-less than 64 bytes. Today's SSDs are optimized for reads in the
+\paragraph{\textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
+less than 64 bytes. Today's \textls*{\small SSD}s are optimized for reads in the
 kilobyte range.
 
-\paragraph{SSDs have a low request rate.} The fastest~\cite{fusionio} can
+\paragraph{\textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
 perform on the order of 1 million random reads per second, and many
@@
 
-\paragraph{SSDs have high overhead.} Commodity SSDs are usually treated
-as normal disks---accesses must occur through the kernel, incurring a
-high context switch overhead. Grappa depends on user-mode access to
+\paragraph{\textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
+as normal disks: accesses must occur through the kernel, incurring a
+high context-switch overhead. Grappa depends on user-mode access to
 the network layer. This capability has been explored in recent research~\cite{caulfield:2012}.
@@
 For simplicity, we omit discussion of write and erase operations. We
-assume a non-volatile Grappa would store frequently-written temporary
-data in DRAM and would use its standard latency-tolerance techniques
+assume a non-volatile Grappa would store frequently written temporary
+data in \textls*{\small DRAM} and would use its standard latency-tolerance techniques
 to cover the cost of infrequent writes.
 
-\subsubsection{Tolerating Flash-based SSD latency} 
-A natural first step would be to add one or more SSDs to each Grappa
-cluster node, and use Grappa's global memory API to expose concurrent
-accesses and context-switch to tolerate the latency of SSD
+\subsubsection{Tolerating Flash-based \textls*{\small SSD} latency} 
+A natural first step would be to add one or more \textls*{\small SSD}s to each Grappa
+cluster node, and use Grappa's global memory \textls*{\small API} to expose concurrent
+accesses and context switch to tolerate the latency of \textls*{\small SSD}
 access. Unfortunately this is unlikely to work well, for two
-reasons. First, the SSD must accept requests at a rate sufficient to
+reasons. First, the \textls*{\small SSD} must accept requests at a rate sufficient to
 keep the cores busy. Grappa's target is for each cluster node to
-generate 100 million random 8-byte requests per second, but most SSDs
+generate 100 million random 8-byte requests per second, but most \textls*{\small SSD}s
 can accept requests at a rate two to four orders of magnitude slower.
-Second, the SSD must be able to support many requests operating in
+Second, the \textls*{\small SSD} must be able to support many requests operating in
 parallel. Little's Law suggests that if requests take 20 $\mu$s for
-the SSD to service, and if the node generates requests at a rate of
-100 million per second, the SSDs must support 2000 concurrent
-requests. Again, existing SSDs support orders of magnitude less request concurrency.
+the \textls*{\small SSD} to service, and if the node generates requests at a rate of
+100 million per second, the \textls*{\small SSD}s must support 2000 concurrent
+requests. Again, existing \textls*{\small SSD}s support orders of magnitude less request concurrency.
 
@@
 \paragraph{Scaling up the Flash array}
-Existing Flash-based SSDs are designed for large block requests. An
-SSD for Grappa could be designed to support high concurrency, small
-requests. Where a traditional SSD might put multiple Flash chips on
+Existing Flash-based \textls*{\small SSD}s are designed for large block requests. An
+\textls*{\small SSD} for Grappa could be designed to support high concurrency, small
+requests. Where a traditional \textls*{\small SSD} might put multiple Flash chips on
 the same bus, or use Flash chips in parallel to build a wider,
-high-bandwidth bus, a Grappa SSD would try to maximize parallelism by
+high-bandwidth bus, a Grappa \textls*{\small SSD} would try to maximize parallelism by
 putting each chip on its own bus. Serial Flash might be used to
@@
 latency. This design could scale to request rates, but the challenge
-would be chip count: to support 100M reads/second, an array of
+would be chip count: To support 100M reads/second, an array of
 20$\mu$s flash would require 2000 chips. This is possible, but expensive.
@@
 processor with few cores might generate only 1 million references per
-second. This could be satisfied with only 20 20$\mu$s Flash chips.
+second. This could be satisfied with only 20\,x\,20$\mu$s Flash chips.
 
@@
 requires fewer chips, the overall cluster performance goal, and thus
-chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
+chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency~\cite{feo:xmt}.
 
@@
 is worth exploring the potential benefit of emerging non-volatile
-memory technologies like Phase-Change Memory. The advantages of PCM
-are twofold: first, read latency is much closer to that of DRAM, and
-second, PCM supports finer-grained accesses.
+memory technologies like Phase-Change Memory. The advantages of \textls*{\small PCM}
+are twofold: first, read latency is much closer to that of \textls*{\small DRAM}, and
+second, \textls*{\small PCM} supports finer-grained accesses.
 
-\paragraph{PCM as Main Memory}
-Much recent research has explored the use of PCM as a main memory
+\paragraph{\textls*{\small PCM} as Main Memory}
+Much recent research has explored the use of \textls*{\small PCM} as a main memory
 replacement. Grappa would easily be able to take advantage of such a
@@
 a prefetch request is issued for that data a few cycles before the
-thread is context-switched in. A PCM-based
+thread is context switched in. A \textls*{\small PCM}-based
 Grappa would do the same, but would issue the prefetches earlier to
-cover the longer latency of PCM.
+cover the longer latency of \textls*{\small PCM}.
 
-\paragraph{PCM as attached storage}
+\paragraph{\textls*{\small PCM} as attached storage}
 
-Grappa could also support a PCM array in a separate address space (for
-instance, on the PCIe bus). PCM chips available today~\cite{micronPCM} support a
+Grappa could also support a \textls*{\small PCM} array in a separate address space (for
+instance, on the \textls*{\small PCIe} bus). \textls*{\small PCM} chips available today~\cite{micronPCM} support a
 64-bit read in 200 ns; this means that 20 chips acting independently
@@
 
-Just as in network communications, issuing individual PCIe bus
+Just as in network communications, issuing individual \textls*{\small PCIe} bus
 operations for each memory request has too high overhead to get a
 reasonable request rate.  Instead, Grappa would delay threads
-performing operations in the PCM array until it could send a buffer
-with enough requests to amortize the overhead. The PCM array would
+performing operations in the \textls*{\small PCM} array until it could send a buffer
+with enough requests to amortize the overhead. The \textls*{\small PCM} array would
 perform the operations and buffer the results. When a sufficient
-amount have completed, the PCM array would write the results to DRAM,
+amount have completed, the \textls*{\small PCM} array would write the results to \textls*{\small DRAM}\@,
 and one core would deliver the results to the requesting threads,
@@
 This system could be built today, with the only downside being
-density---currently available PCM chips are significantly less dense
-than Flash or even DRAM. PCM's technology scaling curve suggests that
+density---currently available \textls*{\small PCM} chips are significantly less dense
+than Flash or even \textls*{\small DRAM}\@. \textls*{\small PCM}'s technology scaling curve suggests that
 this will be an ideal solution in the future.
@@
 application space.  Answering this question is an opportunity for
-collaboration between our group and the NVMW community.
+collaboration between our group and the \textls*{\small NVMW} community.
 

### 2012-12-12 doc/nvm2013/paper.tex
incorporate comments to section 2

@@
 \usepackage{microtype}
+\usepackage{placeins}
 \begin{document}
@@
 
+\FloatBarrier
 \begin{figure}[t]
@@
 usually contains data that is infrequently modified. All accesses to
-this shared heap in Grappa programs are done through an \textls*{\small API}\@. If the
+this shared heap in Grappa programs are done through an \textls*{\small API}, so this change could be made transparent to the programmer. If the
 access is to local memory, Grappa issues a prefetch for the address
@@
 latency. This design could scale to request rates, but the challenge
-would be chip count: To support 100M reads/second, an array of
+would be chip count: to support 100M reads per second, an array of
 20$\mu$s flash would require 2000 chips. This is possible, but expensive.
 
-\paragraph{Scaling down the processors}
+\paragraph{Scaling down the processors.}
 Alternatively, rather than scaling up the flash array, processors
@@
 increase locality and amortize the cost of Flash accesses. Requests
-to common flash blocks would be bucketed by flash block address for a
-limited time. When a sufficient number of requests arrived for a block
-or when requests timed out, the requests would be issued to the Flash
-layer.
+to common flash pages would be bucketed by flash page address for a
+limited time. When a sufficient number of requests have arrived for a page
+or when the requests time out, a single page request would be issued to the Flash
+layer; its result would be used to satisfy the individual requests.
 
-While similar techniques have provided benefit in prior work in
-restricted situations~\cite{magt:2010,graphchi:osdi12}, they depend on a sufficient request
-density to each block to get a benefit. As datasets scale up, the
-request density may decrease and the benefit of this technique may
-decrease.
+While similar techniques have provided benefit for traversal-oriented applications in
+restricted situations~\cite{magt:2010}, they depend on a sufficient request
+density to each flash page to get a benefit. As datasets scale up, this
+density may decrease, limiting the benefit of this technique.
+
+For iterative applications where the entire graph is regularily
+visited, an windowed approach similar to~\cite{graphchi:osdi12} may
+provide more benefit. In this approach, graph data is constantly
+streamed between SSD and DRAM. Requests wait to execute until the
+relevant fragment of the graph is in memory. Request density is not a
+concern for these applications; instead, bandwidth between the SSD and
+DRAM is the limiting factor.
 
@@
 replacement. Grappa would easily be able to take advantage of such a
-development. When a Grappa program references a low-locality program,
+development. When a Grappa program references low-locality data,
 a prefetch request is issued for that data a few cycles before the
@@
 This system could be built today, with the only downside being
-density---currently available \textls*{\small PCM} chips are significantly less dense
+density: currently available \textls*{\small PCM} chips are significantly less dense
 than Flash or even \textls*{\small DRAM}\@. \textls*{\small PCM}'s technology scaling curve suggests that

### 2012-12-12 doc/nvm2013/paper.tex
fix figure position
squeeze some more space

@@
 \end{minipage}
-\vspace{-3ex}
+\vspace{-6ex}
 \end{center}
@@
 
-Emerging non-volatile memory technologies offer hope for an
-alternative.  We imagine placing the entire data set on \textls*{\small SSD} or \textls*{\small PCM} and
-either using Grappa directly to tolerate the latency, much in the same
-way as it is used now to tolerate network latency, or to stage blocks
-of data into \textls*{\small RAM}, using Grappa to tolerate the latency incurred by the
-transfers.  We imagine the energy per available byte being much lower.
-We imagine recovery from faults with most of the data residing in
-non-volatile memory being faster.
+Emerging non-volatile memory technologies offer an alternative strategy:
+place the entire data set on \textls*{\small SSD} or \textls*{\small PCM};
+use Grappa to tolerate the latency directly, much in the same
+way as it is used now to tolerate network latency, or indirectly when transferring blocks of working-set
+data into \textls*{\small RAM}; save energy per available byte;
+recover from faults faster with most of the data preserved in non-volatile memory.
 
@@
 
-\paragraph{\textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
+%\paragraph{\textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
+\noindent{\bf \textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
 less than 64 bytes. Today's \textls*{\small SSD}s are optimized for reads in the
@@
 
-\paragraph{\textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
+%\paragraph{\textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
+\noindent{\bf \textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
 perform on the order of 1 million random reads per second, and many
@@
 
-\paragraph{\textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
+%\paragraph{\textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
+\noindent{\bf \textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
 as normal disks: accesses must occur through the kernel, incurring a
@@
 We see three ways to address this disparity.
-\paragraph{Scaling up the Flash array}
+
+%\paragraph{Scaling up the Flash array}
+\noindent{\bf Scaling up the Flash array}
 Existing Flash-based \textls*{\small SSD}s are designed for large block requests. An
@@
 
-\paragraph{Scaling down the processors.}
+%\paragraph{Scaling down the processors.}
+\noindent{\bf Scaling down the processors.}
 Alternatively, rather than scaling up the flash array, processors
@@
 
-\paragraph{Buffering requests to increase locality.}
+%\paragraph{Buffering requests to increase locality.}
+\noindent{\bf Buffering requests to increase locality.}
 Grappa trades latency for better throughput by aggregating network
@@
 
-\paragraph{\textls*{\small PCM} as Main Memory}
+%\paragraph{\textls*{\small PCM} as Main Memory}
+\noindent{\bf \textls*{\small PCM} as Main Memory}
 Much recent research has explored the use of \textls*{\small PCM} as a main memory
@@
 
-\paragraph{\textls*{\small PCM} as attached storage}
-
+%\paragraph{\textls*{\small PCM} as attached storage}
+\noindent{\bf \textls*{\small PCM} as attached storage}
 Grappa could also support a \textls*{\small PCM} array in a separate address space (for
@@
 
-%\subsection{References}
+\begin{spacing}{0.8}
 \bibliographystyle{abbrv}
 \bibliography{paper}
+\end{spacing}
 

### 2012-12-12 doc/nvm2013/paper.tex
mods to conclusion

@@
 \begin{document}
-\addtolength{\topmargin}{-.7in}
+\addtolength{\topmargin}{-.5in}
 \addtolength{\textheight}{.7in}
@@
 \addtolength{\textwidth}{.5in}
+\addtolength{\topsep}{-3pt}
 \linespread{.94}
@@
 \end{minipage}
-\vspace{-3ex}
+%\vspace{-6ex}
 \end{center}
@@
 \paragraph{\textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
+%\noindent{\bf \textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
 less than 64 bytes. Today's \textls*{\small SSD}s are optimized for reads in the
@@
 \paragraph{\textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
+%\noindent{\bf \textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
 perform on the order of 1 million random reads per second, and many
@@
 \paragraph{\textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
+%\noindent{\bf \textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
 as normal disks: accesses must occur through the kernel, incurring a
@@
 We see three ways to address this disparity.
-\paragraph{Scaling up the Flash array}
+
+\paragraph{Scaling up the Flash array.}
+%\noindent{\bf Scaling up the Flash array.}
 Existing Flash-based \textls*{\small SSD}s are designed for large block requests. An
@@
 \paragraph{Scaling down the processors.}
+%\noindent{\bf Scaling down the processors.}
 Alternatively, rather than scaling up the flash array, processors
@@
 \paragraph{Buffering requests to increase locality.}
+%\noindent{\bf Buffering requests to increase locality.}
 Grappa trades latency for better throughput by aggregating network
@@
 
-\paragraph{\textls*{\small PCM} as Main Memory}
+\paragraph{\textls*{\small PCM} as Main Memory.}
+%\noindent{\bf \textls*{\small PCM} as Main Memory.}
 Much recent research has explored the use of \textls*{\small PCM} as a main memory
@@
 
-\paragraph{\textls*{\small PCM} as attached storage}
-
+\paragraph{\textls*{\small PCM} as attached storage.}
+%\noindent{\bf \textls*{\small PCM} as attached storage.}
 Grappa could also support a \textls*{\small PCM} array in a separate address space (for
@@
 what capability and efficiency improvements might it bring to Grappa's
-application space.  Answering this question is an opportunity for
-collaboration between our group and the \textls*{\small NVMW} community.
+application space?  Answering this question is an opportunity for
+collaboration between our group and the non-volatile memory research community.
 
-%\subsection{References}
+\begin{spacing}{0.8}
 \bibliographystyle{abbrv}
 \bibliography{paper}
+\end{spacing}
 

### 2012-12-12 doc/nvm2013/paper.tex
reduce top sep

@@
 what capability and efficiency improvements might it bring to Grappa's
-application space?  Answering this question is an opportunity for
-collaboration between our group and the non-volatile memory research community.
+application space.  Answering this question is an opportunity for
+collaboration between our group and the \textls*{\small NVMW} community.
 

### 2012-12-12 doc/nvm2013/paper.tex
added space between SSD & 's'

--- 2012-12-12 doc/nvm2013/paper.tex
@@
 
-\subsection{Challenges in using commodity \textls*{\small SSD}s}
+\subsection{Challenges in using commodity \textls*{\small SSD}\hspace{0.5pt}s}
 Grappa's goal is to provide good irregular application performance on
@@
 
-\paragraph{\textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
-%\noindent{\bf \textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
-less than 64 bytes. Today's \textls*{\small SSD}s are optimized for reads in the
+%\paragraph{\textls*{\small SSD}\hspace{0.5pt}s assume locality.} In our benchmarks, we find the average request size to be
+\noindent{\bf \textls*{\small SSD}\hspace{0.5pt}s assume locality.} In our benchmarks, we find the average request size to be
+less than 64 bytes. Today's \textls*{\small SSD}\hspace{0.5pt}s are optimized for reads in the
 kilobyte range.
 
-\paragraph{\textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
-%\noindent{\bf \textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
+%\paragraph{\textls*{\small SSD}\hspace{0.5pt}s have a low request rate.} The fastest~\cite{fusionio} can
+\noindent{\bf \textls*{\small SSD}\hspace{0.5pt}s have a low request rate.} The fastest~\cite{fusionio} can
 perform on the order of 1 million random reads per second, and many
@@
 
-\paragraph{\textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
-%\noindent{\bf \textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
+%\paragraph{\textls*{\small SSD}\hspace{0.5pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.5pt}s are usually treated
+\noindent{\bf \textls*{\small SSD}\hspace{0.5pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.5pt}s are usually treated
 as normal disks: accesses must occur through the kernel, incurring a
@@
 \subsubsection{Tolerating Flash-based \textls*{\small SSD} latency} 
-A natural first step would be to add one or more \textls*{\small SSD}s to each Grappa
+A natural first step would be to add one or more \textls*{\small SSD}\hspace{0.5pt}s to each Grappa
 cluster node, and use Grappa's global memory \textls*{\small API} to expose concurrent
@@
 keep the cores busy. Grappa's target is for each cluster node to
-generate 100 million random 8-byte requests per second, but most \textls*{\small SSD}s
+generate 100 million random 8-byte requests per second, but most \textls*{\small SSD}\hspace{0.5pt}s
 can accept requests at a rate two to four orders of magnitude slower.
@@
 the \textls*{\small SSD} to service, and if the node generates requests at a rate of
-100 million per second, the \textls*{\small SSD}s must support 2000 concurrent
-requests. Again, existing \textls*{\small SSD}s support orders of magnitude less request concurrency.
+100 million per second, the \textls*{\small SSD}\hspace{0.5pt}s must support 2000 concurrent
+requests. Again, existing \textls*{\small SSD}\hspace{0.5pt}s support orders of magnitude less request concurrency.
 
@@
 
-\paragraph{Scaling up the Flash array.}
-%\noindent{\bf Scaling up the Flash array.}
-Existing Flash-based \textls*{\small SSD}s are designed for large block requests. An
+%\paragraph{Scaling up the Flash array}
+\noindent{\bf Scaling up the Flash array}
+Existing Flash-based \textls*{\small SSD}\hspace{0.5pt}s are designed for large block requests. An
 \textls*{\small SSD} for Grappa could be designed to support high concurrency, small

### 2012-12-12 doc/nvm2013/paper.tex
upped SSD s space to .75pt

--- 2012-12-12 doc/nvm2013/paper.tex
@@
 
-\subsection{Challenges in using commodity \textls*{\small SSD}s}
+\subsection{Challenges in using commodity \textls*{\small SSD}\hspace{0.75pt}s}
 Grappa's goal is to provide good irregular application performance on
@@
 
-\paragraph{\textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
-%\noindent{\bf \textls*{\small SSD}s assume locality.} In our benchmarks, we find the average request size to be
-less than 64 bytes. Today's \textls*{\small SSD}s are optimized for reads in the
+%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s assume locality.} In our benchmarks, we find the average request size to be
+\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s assume locality.} In our benchmarks, we find the average request size to be
+less than 64 bytes. Today's \textls*{\small SSD}\hspace{0.75pt}s are optimized for reads in the
 kilobyte range.
 
-\paragraph{\textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
-%\noindent{\bf \textls*{\small SSD}s have a low request rate.} The fastest~\cite{fusionio} can
+%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
+\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
 perform on the order of 1 million random reads per second, and many
@@
 
-\paragraph{\textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
-%\noindent{\bf \textls*{\small SSD}s have high overhead.} Commodity \textls*{\small SSD}s are usually treated
+%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
+\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
 as normal disks: accesses must occur through the kernel, incurring a
@@
 \subsubsection{Tolerating Flash-based \textls*{\small SSD} latency} 
-A natural first step would be to add one or more \textls*{\small SSD}s to each Grappa
+A natural first step would be to add one or more \textls*{\small SSD}\hspace{0.75pt}s to each Grappa
 cluster node, and use Grappa's global memory \textls*{\small API} to expose concurrent
@@
 keep the cores busy. Grappa's target is for each cluster node to
-generate 100 million random 8-byte requests per second, but most \textls*{\small SSD}s
+generate 100 million random 8-byte requests per second, but most \textls*{\small SSD}\hspace{0.75pt}s
 can accept requests at a rate two to four orders of magnitude slower.
@@
 the \textls*{\small SSD} to service, and if the node generates requests at a rate of
-100 million per second, the \textls*{\small SSD}s must support 2000 concurrent
-requests. Again, existing \textls*{\small SSD}s support orders of magnitude less request concurrency.
+100 million per second, the \textls*{\small SSD}\hspace{0.75pt}s must support 2000 concurrent
+requests. Again, existing \textls*{\small SSD}\hspace{0.75pt}s support orders of magnitude less request concurrency.
 
@@
 
-\paragraph{Scaling up the Flash array.}
-%\noindent{\bf Scaling up the Flash array.}
-Existing Flash-based \textls*{\small SSD}s are designed for large block requests. An
+%\paragraph{Scaling up the Flash array}
+\noindent{\bf Scaling up the Flash array}
+Existing Flash-based \textls*{\small SSD}\hspace{0.75pt}s are designed for large block requests. An
 \textls*{\small SSD} for Grappa could be designed to support high concurrency, small

### 2012-12-13 doc/nvm2013/paper.tex
edits and more squeezing

@@
 \begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:grappa} Grappa system overview}
+  \nocaptionrule\caption{\label{fig:grappa} Grappa system overview}
 \end{minipage}
-%\vspace{-6ex}
+\vspace{-5ex}
 \end{center}
@@
 \item [Tasking system.] Our tasking system supports lightweight
-multithreading to tolerate communication latency and global distributed
+multithreading to tolerate communication latency; it also supports global distributed
 workstealing (i.e., tasks can be stolen from any node in the system), which
@@
 \item[Communication layer.] Modern commodity networks
-support high bandwidth only for large messages and irregular applications
+support high bandwidth only for large messages, but irregular applications
 frequently issue small requests.
-Our communication layer aggregates these small requests into larger messages to better exploit network bandwidth, targeting 100 million remote references per second per node.
+Our communication layer aggregates these small requests into larger messages to better exploit network bandwidth, targeting 100 million 8-byte remote references per second per node.
 Aggregation is largely invisible to the application programmer.
@@
 operations on global data, explicit local caching of any memory in the system,
-and support for operations on remote data (delegating operations to owner node).
+and support for operations on remote data (delegating operations to the data's owner node).
 By tight integration with the tasking system and the
@@
 Emerging non-volatile memory technologies offer an alternative strategy:
-place the entire data set on \textls*{\small SSD} or \textls*{\small PCM};
-use Grappa to tolerate the latency directly, much in the same
-way as it is used now to tolerate network latency, or indirectly when transferring blocks of working-set
-data into \textls*{\small RAM}; save energy per available byte;
-recover from faults faster with most of the data preserved in non-volatile memory.
+place the entire data set on Flash or \textls*{\small PCM}, and
+%% use Grappa to tolerate the latency either directly, much in the same
+%% way as it is used now to tolerate network latency, or indirectly when transferring blocks of working-set
+%% data into \textls*{\small RAM}. This could save energy per available byte and allow
+%% faster fault recovery with most of the data preserved in non-volatile memory.
+use Grappa to tolerate the access latency. This could save energy per available byte and allow
+faster fault recovery with most of the data preserved in non-volatile memory.
 
@@
 The natural point of integration for non-volatile memories in Grappa
-is to store the global heap in nonvolatile memory, since the heap
-usually contains data that is infrequently modified. All accesses to
-this shared heap in Grappa programs are done through an \textls*{\small API}, so this change could be made transparent to the programmer. If the
-access is to local memory, Grappa issues a prefetch for the address
+is to store the global distributed shared heap in nonvolatile memory, since for irregular applications the heap
+often holds data that is infrequently modified. All accesses to
+this shared heap in Grappa programs are done through an \textls*{\small API}, so this change could be made transparent to the programmer. 
+If an 
+access is to a chunk of the heap stored in a node's local memory, Grappa issues a prefetch for the address
 and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If
@@
 Grappa's goal is to provide good irregular application performance on
-commodity hardware. Unfortunately, commodity non-volatile memories are
+mass market hardware. Unfortunately, commodity \textls*{\small SSD}\hspace{0.75pt}s are
 difficult to use in this way for three reasons:
@@
 perform on the order of 1 million random reads per second, and many
-are limited to less than 100K/s. In contrast, cores may be idle in
+are limited to less than 100K/s. Even the individual flash chips are slow, on the order of 40K/s~\cite{micronFlash}. In contrast, cores may be idle in
 Grappa if each node cannot make 100 million random reads per second of
-the global heap. Even the individual flash chips are slow, on the order of 40K/s~\cite{micronFlash}.
+the global heap.
 
@@
 high context-switch overhead. Grappa depends on user-mode access to
-the network layer. This capability has been explored in recent research~\cite{caulfield:2012}.
+the network layer. This approach has been applied to \textls*{\small SSD}\hspace{0.75pt}s in recent research~\cite{caulfield:2012}.
 
@@
 data in \textls*{\small DRAM} and would use its standard latency-tolerance techniques
-to cover the cost of infrequent writes.
+to cover the cost of infrequent writes to non-volatile memory.
 
 \subsubsection{Tolerating Flash-based \textls*{\small SSD} latency} 
-A natural first step would be to add one or more \textls*{\small SSD}\hspace{0.75pt}s to each Grappa
+A natural first step would be to add one or more mass market \textls*{\small SSD}\hspace{0.75pt}s to each Grappa
 cluster node, and use Grappa's global memory \textls*{\small API} to expose concurrent
@@
 access. Unfortunately this is unlikely to work well, for two
-reasons. First, the \textls*{\small SSD} must accept requests at a rate sufficient to
-keep the cores busy. Grappa's target is for each cluster node to
-generate 100 million random 8-byte requests per second, but most \textls*{\small SSD}\hspace{0.75pt}s
-can accept requests at a rate two to four orders of magnitude slower.
+reasons. First, as discussed earlier, the \textls*{\small SSD} must accept requests at a rate sufficient to
+keep the cores busy; most \textls*{\small SSD}\hspace{0.75pt}s
+can accept requests at a rate two to four orders of magnitude slower than required.
 Second, the \textls*{\small SSD} must be able to support many requests operating in
-parallel. Little's Law suggests that if requests take 20 $\mu$s for
+parallel. Little's Law suggests that if requests take 25 $\mu$s for
 the \textls*{\small SSD} to service, and if the node generates requests at a rate of
-100 million per second, the \textls*{\small SSD}\hspace{0.75pt}s must support 2000 concurrent
+100 million per second, the \textls*{\small SSD}\hspace{0.75pt}s must support 2500 concurrent
 requests. Again, existing \textls*{\small SSD}\hspace{0.75pt}s support orders of magnitude less request concurrency.
@@
 
-%\paragraph{Scaling up the Flash array}
-\noindent{\bf Scaling up the Flash array}
+\paragraph{Scaling up the Flash array.}
+%\noindent{\bf Scaling up the Flash array}
 Existing Flash-based \textls*{\small SSD}\hspace{0.75pt}s are designed for large block requests. An
@@
 latency. This design could scale to request rates, but the challenge
-would be chip count: to support 100M reads per second, an array of
-20$\mu$s flash would require 2000 chips. This is possible, but expensive.
+would be chip count: to support 100 million reads per second, an array of
+25$\mu$s flash would require 2500 chips. This is possible, but expensive.
 
@@
 processor with few cores might generate only 1 million references per
-second. This could be satisfied with only 20\,x\,20$\mu$s Flash chips.
-
+second. This could be satisfied with only 25\,x\,25$\mu$s Flash chips.
 There are two problems with this design for our purpose. First, while each node
 requires fewer chips, the overall cluster performance goal, and thus
-chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency~\cite{feo:xmt}.
+chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
 
@@
 
-While similar techniques have provided benefit for traversal-oriented applications in
+While similar techniques have provided benefit for graph traversal applications in
 restricted situations~\cite{magt:2010}, they depend on a sufficient request
-density to each flash page to get a benefit. As datasets scale up, this
+density to each flash page to get a benefit. As datasets scale up and machines scale out, this
 density may decrease, limiting the benefit of this technique.
@@
 provide more benefit. In this approach, graph data is constantly
-streamed between SSD and DRAM. Requests wait to execute until the
+streamed between \textls*{\small SSD} and \textls*{\small DRAM}. Requests wait to execute until the
 relevant fragment of the graph is in memory. Request density is not a
-concern for these applications; instead, bandwidth between the SSD and
-DRAM is the limiting factor.
+concern for these applications; instead, bandwidth between the \textls*{\small SSD} and \textls*{\small DRAM}
+is the limiting factor.
 
@@
 %\noindent{\bf \textls*{\small PCM} as Main Memory.}
-Much recent research has explored the use of \textls*{\small PCM} as a main memory
-replacement. Grappa would easily be able to take advantage of such a
+Recent research has explored the use of \textls*{\small PCM} as a main memory
+replacement~\cite{pcm:isca09}. Grappa would easily be able to take advantage of such a
 development. When a Grappa program references low-locality data,
@@
 %\noindent{\bf \textls*{\small PCM} as attached storage.}
-Grappa could also support a \textls*{\small PCM} array in a separate address space (for
+Grappa could also transparently support a \textls*{\small PCM} array in a separate address space (for
 instance, on the \textls*{\small PCIe} bus). \textls*{\small PCM} chips available today~\cite{micronPCM} support a
@@
 perform the operations and buffer the results. When a sufficient
-amount have completed, the \textls*{\small PCM} array would write the results to \textls*{\small DRAM}\@,
+number have completed, the \textls*{\small PCM} array would write the results to \textls*{\small DRAM}\@,
 and one core would deliver the results to the requesting threads,

### 2012-12-13 doc/nvm2013/paper.tex
spelling

@@
 multithreading to tolerate communication latency; it also supports global distributed
-workstealing (i.e., tasks can be stolen from any node in the system), which
+work stealing (i.e., tasks can be stolen from any node in the system), which
 provides automated load balancing.  Grappa thus tolerates many
@@
 The natural point of integration for non-volatile memories in Grappa
-is to store the global distributed shared heap in nonvolatile memory, since for irregular applications the heap
+is to store the global distributed shared heap in non-volatile memory, since for irregular applications the heap
 often holds data that is infrequently modified. All accesses to
@@
 
-For iterative applications where the entire graph is regularily
+For iterative applications where the entire graph is regularly
 visited, an windowed approach similar to~\cite{graphchi:osdi12} may

### 2013-01-29 doc/xps2013/header.tex
XPS 2013 rough in

\documentclass[11pt]{article}

\usepackage{times}
\usepackage{fullpage}
\usepackage[override]{cmtt}
\usepackage{sectsty}
\usepackage{graphicx}
\usepackage{color}
\usepackage{xspace}
\usepackage[lmargin=1in, rmargin=1in,tmargin=1in,bmargin=1in]{geometry}
\usepackage{url}
\usepackage{cite}
\usepackage{microtype}
\usepackage{listings}
\usepackage{subfig}

\newcommand{\TODO}[1]{\textcolor{red}{{\bf TODO:} #1}}
\newcommand{\ie}{{\em i.e.}}
\newcommand{\eg}{{\em e.g.}}
\newcommand{\benchmark}[1]{{\small\texttt{#1}}}
\newcommand{\ilcode}[1]{\texttt{#1}}
\newcommand{\comment}[1]{}

\newcommand{\gasnet}{GASNet}

\setlength{\parskip}{6pt}

\pagestyle{plain}

\begin{document}

### 2013-01-29 doc/xps2013/proposal.tex
XPS 2013 rough in

%% NSF XPS Proposal 2013

\input{header}

\begin{center}
  \textbf{\Large XPS: Towards an Architecture for Large-Scale, Irregular and Low-Locality Applications}\\ \vspace{5mm}
  \textbf{Luis Ceze ~~~~~ Simon Kahan ~~~~~ Mark Oskin (PI)}\\
  \textbf{University of Washington}
\end{center}

\section{Introduction}

Irregular applications generate tasks with work, interdependences, or memory
accesses that are highly sensitive to input. Classic examples of irregular
applications include branch and bound optimization, SPICE circuit simulation,
contact algorithms in car crash analysis, and network flow. Important
contemporary examples include processing large graphs in the business,
national security, machine learning, data-driven science, and social network
computing domains. For these emerging applications, fast response -- given the
sheer amount of data -- requires multinode systems. The most broadly available
multinode systems are those built from x86 commodity computing nodes
interconnected via ethernet or infiniband. Our goal is to enable scalable
performance of irregular applications on these mass market systems.

Our goal is challenging for two key reasons:

\noindent{\bf Irregular applications exhibit little spatial locality.}
It is not atypical for any given task's data references to be spread
randomly across the entire memory of the system. This makes current
memory hierarchy features ineffective. Caches are of little assistance
with such low data re-use and spatial locality. Commodity prefetching
hardware is effective only when addresses are known many cycles before
the data is consumed or the accesses follow a predictable pattern,
neither of which occurs in irregular applications. As a consequence,
commodity microprocessors stall often when executing irregular
applications.

\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low
locality are analogous, and exacerbated by the increased latency of going
off-node. Irregular applications also present a challenge to mass market
network technology, which is designed to transfer large blocks of data, not
the word-sized references emitted by irregular application tasks. Injection
rate into the network is insufficient to utilize wire bandwidth when blocks
are below about two-kilobytes, so any straightforward communication strategy
severely under-utilizes the network.

While some irregular applications can be restructured to better exploit
locality, aggregating requests to increase message size, and manage the
additional challenges of load balance and synchronization across multinode
systems, the work to do so is formidable and requires knowledge and skills
pertaining to distributed systems far beyond those of most application
programmers.

Luckily, many of the important irregular applications (e.g., graph
processing, machine learning, ...) naturally offer large amounts of
concurrency. This immediately suggests taking advantage of concurrency
to tolerate the latency of data movement. The fully custom Tera
MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular
applications by using concurrency to hide latencies. It had a large
distributed shared memory with no caches.  On every clock cycle, each
processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully hide memory
access latency.  The network was designed with a single word injection
rate that matched the processor clock frequency and sufficient bandwidth
to sustain a reference from every processor on every clock cycle.
Unfortunately, while an excellent match to extremely irregular
applications, the MTA was not cost-effective on applications that could
exploit locality and had very poor single-thread performance, making
it a commercial failure. The Cray XMT approximates the Tera MTA-2,
substantially reducing its cost but not overcoming its narrow range of
applicability.

Our preliminary work on Grappa shows that it runs several graph-crunching
applications (classic examples of irregular behavior) very efficiently
on a commodity cluster. Our yardstick for comparison is the XMT hardware
itself. Using the same number of network interfaces, Grappa is in the
same ballpark as the XMT: For unbalanced tree search, Grappa is over 3X
faster and shows greatly improved scalability; conversely, for breadth
first search and betweenness centrality Grappa is 2.5X slower.  In
Section~\ref{sec:evaluation} we explore the factors that underpin this
performance. Most importantly, however, for significantly less real
world cost, users can \emph{add\/} significantly more processors to a
commodity cluster than an XMT machine and use Grappa to achieve scalable
performance.

\section{Related work}

\section{Description of the Current Grappa System}

In the twenty years that have elapsed since the Tera MTA, commodity
microprocessors have become much faster and multicore has driven down the
absolute price of computation; commodity network price-performance has
improved as well. This shift has afforded us the opportunity to attack the
challenges posed by irregular applications by emulating in software and
inexpensive mass market hardware, the approach taken by Tera. We exploit the
increased aggregate instruction rate per socket relative to chip bandwidth,
using what would otherwise be wasted instructions to manage the multiplexing
of as many as several thousand tasks per core, thus tolerating memory latency,
reducing stalls, and making better use of available bandwidth. Ultimately, the
opportunity is to cover the spectrum of irregular to regular computation:
where tasks exhibit locality, multiplex fewer tasks and expend fewer
instructions on context switching; where locality is lacking, multiplex more
tasks at a higher rate to tolerate latency. Thus we make the best use of task
parallelism -- either to scale to more cores or to tolerate latency -- and of
caches -- either to exploit application locality or to house more task
contexts.

Grappa is a software runtime system that allows
a commodity x86 distributed-memory HPC cluster to be programmed as if it
were a single large shared-memory machine and provides scalable
performance for irregular applications. Grappa is designed to smooth
over some of the performance discontinuities in commodity hardware,
giving good performance when there is little locality to be exploited
while allowing the programmer to exploit it when it is available.

Grappa leverages as much freely available and commodity infrastructure as
possible. We use unmodified Linux for the operating system and an
off-the-shelf user-mode infiniband device driver stack~\cite{OFED}. MPI is
used for process setup and tear down. GASnet~\cite{gasnet} is used as the
underlying mechanism for remote memory reads and writes using active message
invocations. To this commodity hardware and software mix Grappa adds three
main software components: (1) a \emph{lightweight tasking\/} layer that supports
a context switch in as few as 38ns and distributed global load balancing; (2)
a \emph{distributed shared memory\/} layer that supports normal access operations
such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer
that combines short messages to mitigate the aforementioned problem that
commodity networks are designed to achieve peak bandwidth only on large packet
sizes, yet irregular applications tend to fetch only a handful of bytes at a
time. As we will show later, Grappa can tolerate latencies way beyond that of
the network. Therefore, Grappa can afford to \emph{trade latency for
throughput\/}: by {\em increasing\/} latency in key components of the system we
are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations
to remote nodes), and our ability to improve load imbalance (work stealing
increases latency).

Grappa (Figure~\ref{fig:grappa}) has three main software components:
\begin{description}

\item [Tasking system.] Our tasking system supports lightweight
multithreading to tolerate communication latency and global distributed
workstealing (i.e., tasks can be stolen from any node in the system), which
provides automated load balancing.

\item[Distributed shared memory.] Our DSM system provides support for
fine-grain access to data anywhere in the system. It supports synchronization
operations on global data, explicit local caching of any memory in the system,
and support for operation on remote data (delegating operations to home node).
By tight integration with the tasking system and the
communication layer, our DSM system offers high aggregate random
access bandwidth for accessing remote data.

\item[Communication layer.] As discussed earlier, modern commodity networks
support high bandwidth only for large messages. Since irregular applications
tend to need frequent communication of small requests, the main goal of our
communication layer is to aggregate small messages into large ones to better
exploit what the network can offer. It is largely invisible to the application
programmer.

\end{description}

\subsection{Tasks}

The basic unit of execution in Grappa is a {\em task}. Each task is
represented by a function pointer and its arguments. A new task is enqueued at
spawn time; when resources are free, it is allocated a stack, bound to a core,
and executed.

During execution, a task yields control of its core whenever it performs
a long-latency operation, allowing the processor to remain busy while
waiting for the operation to complete.  In addition,  a programmer can
direct scheduling explicitly via the Grappa API calls shown
in Figure~\ref{fig:scheduling}.
To minimize yield overhead, the Grappa scheduler operates entirely in
user-space and does little more than store register state of one task and load
that of another. Context switch times are as low as 38ns even when switching
amongst thousands of tasks.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{ yield() } ] \hfill \\
      Yields core to scheduler, enqueuing caller to be scheduled again soon
    \item[ \texttt{ suspend() } ] \hfill \\
      Yields core to scheduler, enqueuing caller only once another task calls wake
    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
      Enqueues some other task $t$ to be scheduled again soon
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:scheduling} Grappa API: scheduling} %{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

\subsection{Expressing parallelism}

Grappa programmers focus on expressing as much parallelism as possible without concern for where it will execute. Grappa then chooses where and when to exploit this parallelism, scheduling as much work as is necessary on each core to keep it busy in the presence of system latencies and task dependences.

Grappa provides four methods for expressing parallelism, shown in Figure~\ref{fig:expressing-parallelism}:

First, when the programmer identifies work that can be done in parallel, the work may be wrapped up in a function and queued with its arguments for later execution using a \texttt{spawn}.

Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific core in the system or at the home core of a particular memory location.

Third, the programmer can invoke a parallel for loop, provided that the trip count is known at loop entry. The programmer specifies a function pointer along with start and end indices and an optional threshold to control parallel overhead. Grappa does {\em recursive decomposition} of iterations, similar to Cilk's cilk\_for construct~\cite {cilkforimplementation} and TBB's {\tt parallel\_for}~\cite{intel_tbb}.  It generates a logarithmically-deep tree of tasks, stopping to execute the loop body when the number of iterations is below the required threshold.

Fourth, a programmer may want to execute an active message; that is, to run a small piece of code on a particular core in the system without waiting for execution resources to be available.  For example, custom synchronization primitives execute this way, as a function executed on the core where the data lives.  Grappa provides the \texttt{call\_on} call for this purpose.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
      Creates a new stealable task
    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
      Creates a new private task that will run on a specific core 
    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
      Executes iterations of a loop as stealable tasks 
    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
      Runs a limited function on a specific core without consuming
      Grappa execution resources 
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism} % \vspace{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

\subsection{Memory}

Applications written for Grappa utilize two forms of memory: local and
global.  Local memory is local to a single core in the system.  Accesses occur
through conventional pointers.  The compiler emits an access and the
memory is manipulated directly.  Applications use local accesses for a
number of things in Grappa: the stack associated with a task, accesses
to localized global memory in caches (see below), and accesses to
debugging infrastructure that is local to each system node.  Local
pointers cannot access memory on other cores, and are valid only on
their home core.

Large data that is expected to be shared and accessed with low locality is
stored in Grappa's global memory. All global data must be accessed through
calls into Grappa's API, shown in Figure~\ref{fig:accessing-memory}.

\paragraph{Global memory addressing} Grappa provides two methods for
storing data in the global memory. The first is a distributed heap
striped across all the machines in the system in a block cyclic fashion.
The \texttt{global\_malloc} and \texttt{global\_free} calls are used to
allocate and deallocate memory in the global heap.  Addresses to memory
in the global heap use \textbf{linear addresses}.  Choosing the block
size involves trading off sequential bandwidth against aggregate random
access bandwidth. Smaller block sizes help spread data across all the
memory controllers in the cluster, but larger block sizes allow the
locality-optimized memory controllers to provide increased sequential
bandwidth. The block size, which is configurable, is typically set to 64
bytes, or the size of a single hardware cache line, in order to exploit
spatial locality when available. The heap metadata is stored on a single
node. Currently all heap operations serialize through this node; while
this has been sufficient for our benchmarks, in the future Grappa will
provide parallel performance through combining~\cite{MAMA,flatcombining}.

Grappa also allows any local data on a core's stacks or heap to be
exported to the global address space to be made accessible to other
cores across the system. Addresses to global memory allocated in this
way use \textbf{2D global addresses}.  This uses a traditional PGAS
addressing model, where each address is a tuple of a rank in the job (or
global process ID) and an address in that process. The lower 48 bits of
the address hold a virtual address in the process. The top bit is set to
indicates that the reference is a 2D address (as opposed to linear
address). This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^{15}$ endpoints. Any node-local data can be made
accessible by other nodes in the system by wrapping the address and node
ID into a 2D global address. This address can then be accessed with a
delegate and can also be cached by other nodes. At the destination the
address is converted into a canonical x86 address by replacing the upper
bits with the sign-extended upper bit of the virtual address. 2D
addresses may refer to memory allocated from a single processes' heap or
from a task's stack. Figure~\ref{fig:memory-structure} shows how 2D and
linear addresses can refer to other cores' memory.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
      \item[ \texttt{ global\_address global\_malloc( size )} ] \hfill \\
      \item[ \texttt{ global\_free( global\_address )} ] \hfill \\
        Allocates and frees memory in the global heap
      \item[ \texttt{ delegate\_read( global\_address, local\_var )} ] 
      \item[ \texttt{ delegate\_write( global\_address, local\_var )} ] %\vspace{-2ex}
      \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] %\vspace{-2ex}
      \item[ \texttt{ delegate\_fetch\_inc( global\_address, local\_var )} ] %\vspace{-2ex} 
\hfill \\
        Performs a memory operation at the home core of a global address
      \item[ \texttt{ cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} ]
      \item[ \texttt{ cache\_release( global\_address, local\_buf )} ] %\vspace{-2ex} 
\hfill \\
Perform cache operations to acquire/release global data.  Acquire, returns a local pointer after all data has been copied to the local node.  Release, optionally writes data back to global memory and frees up management resources.
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:accessing-memory} Grappa API: accessing memory} %{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:memory-structure} Grappa memory structure}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\paragraph{Global memory access} There are two general approaches
Grappa applications use to {\emph access} global memory. When the
programmer expects a computation on shared data to have spatial locality
to exploit, {\em cache} operations may be used. When there is no
locality to exploit, {\em delegate} operations are used.

\textbf{Explicit caching.} Grappa provides an API to fetch a global
pointer of any length and return a local pointer to a cached copy of the
global memory.  Grappa cache operations have the usual read-only and
read-write variants, along with a write-only variant used to initialize
data structures. Languages for distributed shared memory systems have
done optimizations to achieve a similar goal. For example, the UPC
compiler coalesces struct and array accesses into remote get/put
\cite{Chen:2005}, and Fortran D compiler's message vectorization hoists
small messages out of a loop \cite{FortranD:1992}. Caching in Grappa
additionally provides a mechanism for exploiting temporal locality by
operating on the data locally.

Under the hood, Grappa performs the mechanics of gathering chunks of
data from multiple system nodes and presenting a conventional appearing
linear block of memory as a local pointer into a cache. The strategy
employed is to issue all the constituent requests of a cache access
request  (as Active Messages) and then yield until all responses have
occurred.  Currently, Grappa caches are \emph{not} coherent, requiring
the programmer to maintain consistent access to data.  Future work will
develop a software directory based coherence scheme to simplify
consistent access to global data.

\textbf{Delegate operations.} When the access pattern has low-locality,
it is more efficient to modify the data on its home core rather than
bringing a copy to the requesting core and returning it after
modification. Delegate operations provide this capability. Applications
can dispatch computation to be performed on individual machine-word
sized chunks of global memory to the memory system itself (e.g.,
\emph{fetch-and-add}).  Delegate operations, proposed in
\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the
primary synchronization method in Grappa.

Delegate operations are always executed at the home core of their
address, and while arbitrary memory operations can be delegated, we
restrict the use of delegate operations in three ways to make them more
useful for synchronization. First, we limit each task to one outstanding
delegate operation to avoid the possibility of reordering in the
network. Second, we limit delegate operations to operate on objects in
the 2D address space or objects that fit in a single block of the linear
address space so they can be satisfied with a single network request.
Finally, no context switches are allowed while the data is being
modified. Given these restrictions, we can ensure that delegate
operations for the same address from multiple requesters are always
serialized through a single core in the system, providing atomic
semantics without using atomic operations. 
Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
cache operations interact.

\begin{figure*}[htb]
\begin{center}
  \includegraphics[width=1.5\columnwidth]{figs/delegate-cache}
\begin{minipage}{1.9\columnwidth}
  \caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in
the global heap. It then spawns two tasks on remote cores to increment
elements of the array.  The first task increments two elements of the array using cache
operations. In step 2, the task is invoked. A cache request is issued
for two adjacent integers starting at the second element of the
array. Since these element are stored in the memories of two different
cores, this requires the sending of two messages in step 3. The task
is suspended until both responses arrive in step 4. The data carried
in these responses is stored in the local buffer. The elements are
then increment in the buffer. Then in step 5, the modified data is
sent back to the home node. Acknoledgements are returned in step 6 so
the task knows when the writes are complete. The second task increments an element of the array with a delegate
operation. In step 7, the task is invoked. A delegate request is sent
to the home core of the array element with the increment value. The
task suspends until the response is received. In step 8, the increment
is executed on the remote core. A response is returned in step 9 with
the previous value of the array element.
}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure*}

\subsection{Communication}\label{sec:communication}

In order to mitigate the low message injection rate limits of commodity
networks, Grappa's communication stack has two layers: one for
user-level messages and one for network-level messages.

At the upper layer, Grappa implements asynchronous active messages
\cite{vonEicken92}. Each message consists of a function pointer, an
optional argument payload, and an optional data payload. When a task
sends a message, the message is copied to a send queue associated with
the message's destination and the task continues execution.

Grappa's lower networking layer aggregates the upper layer's messages
to improve performance. Commodity networks including infiniband
achieves their peak bisection bandwidth \emph{only} when the packet
sizes are relatively large---on the order of multiple kilobytes. The
reason for this discrepancy is the combination of overheads associated
with handling each packet (in terms of bytes that form the actual
packet, processing time at the card and processing on the
CPU within the driver stack). Our measurements confirm manufacturers
published data~\cite{infinibandbandwidth}, that with this packet size
the bisection bandwidth is only a small fraction, less than
3\% of the peak bisection bandwidth.

In our experiments the vast majority of requests were smaller than
44 bytes, far too small to make
efficient use of the network. To make the best use of the network, we
must convert our small messages into large ones.
When a task sends a message,
it is not immediately sent, but rather placed in a queue specific to the
destination.

There are three situations in which a queue of aggregated messages is
sent. First, each queue has a message size threshold of 4096 bytes, chosen to give reasonable network performance. If the size in
bytes of a queue is above the threshold, the contents of the queue are
sent immediately. Second, each queue has a wait time threshold
($\approx${1ms}). If the oldest message in a queue has been waiting
longer than this threshold, the contents of the queue are sent
immediately, even if the queue size is lower than the message size
threshold.  Third, queues may be explicitly flushed in situations where
the programmer wants to minimize the latency of a message at the cost of
bandwidth utilization.

The network layer is serviced by polling. Periodically when a context
switch occurs, the Grappa scheduler switches to the network polling
thread. This thread has three responsibilities. First, it polls the
lower-level network layer to ensure it makes progress. Second, it
deaggregates received messages and executes active message
handlers. Third, it checks to see if any aggregation queues have
messages that have been waiting longer than the threshold; if so, it
sends them.

Underneath the aggregation layer, Grappa uses the \gasnet~communication
library~\cite{gasnet} to actually move data. All interprocess
communication, whether on or off a cluster node, is handled by the
\gasnet~library. \gasnet~is able to take advantage of many communication
mechanisms, including ethernet and infiniband between nodes, as well as
shared memory within a node.

Some networks provide access to a remote machine's memory directly. This
would seem to be a good fit for a programming model focused on global
shared memory, but in fact we do not use it. In our experiments, we
found that RDMA operations are subject to the same message rate
limitations as all other messages on these cards, and thus using raw
RDMA operations for our small messages would make inefficient use of
bandwidth. Instead, we implement remote memory operations with active
messages. A byproduct of this design decision is that Grappa is not
limited to RDMA-capable networks.

\subsection{Performance}

To evaluate Grappa's performance with respect to the XMT, we ran each
of our three benchmarks on up to 16 nodes of each machine. Grappa used
6 cores per node, with the best parameters chosen for each point. In
some cases, the XMT could not run the benchmark with 2 nodes, so the
point is omitted.

%% UTS: performance comparison
\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.5\textwidth]{figs/uts_performance.pdf}
    \end{center}
    \caption{Performance of in-memory unbalanced tree search.}
    \label{fig:uts_compare}
\end{figure}

\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex tree
(T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
number of vertices visited per second versus number of compute
nodes. Grappa is 3.2 times faster than the XMT at 16 nodes.  As we will show later, the performance advantage Grappa has over XMT increases as more nodes are added.  The main reason Grappa performs better is the software-based delegate synchronization obviates the need for the retry-based synchronization that XMT uses.

\begin{figure}[tH]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/bfs_performance}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:bfs-performance} BFS performance}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\paragraph{BFS}

We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and
$2^{29}$ edges (25 GB of data). Figure~\ref{fig:bfs-performance} shows
our performance in terms of graph edges traversed per second. The XMT
is 2.5 times faster than Grappa at 16 nodes.  Performance does scale at a constant rate for Grappa, suggesting that adding more nodes will increase performance.

\begin{figure}[tH]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/centrality_performance}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:centrality-performance} Centrality performance}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\paragraph{Centrality}

We ran Betweenness Centrality on the same scale 25 Kronecker graph as
we did for BFS. Figure~\ref{fig:centrality-performance} shows our
performance in terms of graph edges traversed per second. At 16 XMT
processors/cluster nodes, the XMT is 1.75 times faster than Grappa.

\input{footer}

### 2013-02-01 doc/xps2013/proposal.tex
partial commit for Luis the sick boy

@@
 
-Irregular applications generate tasks with work, interdependences, or memory
-accesses that are highly sensitive to input. Classic examples of irregular
-applications include branch and bound optimization, SPICE circuit simulation,
-contact algorithms in car crash analysis, and network flow. Important
-contemporary examples include processing large graphs in the business,
-national security, machine learning, data-driven science, and social network
-computing domains. For these emerging applications, fast response -- given the
-sheer amount of data -- requires multinode systems. The most broadly available
-multinode systems are those built from x86 commodity computing nodes
-interconnected via ethernet or infiniband. Our goal is to enable scalable
-performance of irregular applications on these mass market systems.
-
+Irregular applications generate tasks with work, interdependences, or memory accesses that are highly sensitive to input.
+Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow.
+Important contemporary examples include processing large graphs in the business, national security, machine learning, data-driven science, and social network computing domains.
+For these emerging applications, fast response -- given the sheer amount of data -- requires multinode systems.
+The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband.
+Our goal is to enable scalable performance of irregular applications on these mass market systems.
 Our goal is challenging for two key reasons:
 
-\noindent{\bf Irregular applications exhibit little spatial locality.}
-It is not atypical for any given task's data references to be spread
-randomly across the entire memory of the system. This makes current
-memory hierarchy features ineffective. Caches are of little assistance
-with such low data re-use and spatial locality. Commodity prefetching
-hardware is effective only when addresses are known many cycles before
-the data is consumed or the accesses follow a predictable pattern,
-neither of which occurs in irregular applications. As a consequence,
-commodity microprocessors stall often when executing irregular
-applications.
+\noindent{\bf Irregular applications exhibit little spatial locality.} It is not atypical for any given task's data references to be spread randomly across the entire memory of the system.
+This makes current memory hierarchy features ineffective.
+Caches are of little assistance with such low data re-use and spatial locality.
+Commodity prefetching hardware is effective only when addresses are known many cycles before the data is consumed or the accesses follow a predictable pattern, neither of which occurs in irregular applications.
+As a consequence, commodity microprocessors stall often when executing irregular applications.
 
-\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low
-locality are analogous, and exacerbated by the increased latency of going
-off-node. Irregular applications also present a challenge to mass market
-network technology, which is designed to transfer large blocks of data, not
-the word-sized references emitted by irregular application tasks. Injection
-rate into the network is insufficient to utilize wire bandwidth when blocks
-are below about two-kilobytes, so any straightforward communication strategy
-severely under-utilizes the network.
+\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.
+Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.
+Injection rate into the network is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.
 
-While some irregular applications can be restructured to better exploit
-locality, aggregating requests to increase message size, and manage the
-additional challenges of load balance and synchronization across multinode
-systems, the work to do so is formidable and requires knowledge and skills
-pertaining to distributed systems far beyond those of most application
-programmers.
+While some irregular applications can be restructured to better exploit locality, aggregating requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.
 
-Luckily, many of the important irregular applications (e.g., graph
-processing, machine learning, ...) naturally offer large amounts of
-concurrency. This immediately suggests taking advantage of concurrency
-to tolerate the latency of data movement. The fully custom Tera
-MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular
-applications by using concurrency to hide latencies. It had a large
-distributed shared memory with no caches.  On every clock cycle, each
-processor would execute a ready instruction chosen from one of its 128
-hardware thread contexts, a sufficient number to fully hide memory
-access latency.  The network was designed with a single word injection
-rate that matched the processor clock frequency and sufficient bandwidth
-to sustain a reference from every processor on every clock cycle.
-Unfortunately, while an excellent match to extremely irregular
-applications, the MTA was not cost-effective on applications that could
-exploit locality and had very poor single-thread performance, making
-it a commercial failure. The Cray XMT approximates the Tera MTA-2,
-substantially reducing its cost but not overcoming its narrow range of
-applicability.
+Luckily, many of the important irregular applications (e.g., graph processing, machine learning, ...) naturally offer large amounts of concurrency.
+This immediately suggests taking advantage of concurrency to tolerate the latency of data movement.
+The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular applications by using concurrency to hide latencies.
+It had a large distributed shared memory with no caches.
+On every clock cycle, each processor would execute a ready instruction chosen from one of its 128 hardware thread contexts, a sufficient number to fully hide memory access latency.
+The network was designed with a single word injection rate that matched the processor clock frequency and sufficient bandwidth to sustain a reference from every processor on every clock cycle.
+Unfortunately, while an excellent match to extremely irregular applications, the MTA was not cost-effective on applications that could exploit locality and had very poor single-thread performance, making it a commercial failure.
+The Cray XMT approximates the Tera MTA-2, substantially reducing its cost but not overcoming its narrow range of applicability.
 
-Our preliminary work on Grappa shows that it runs several graph-crunching
-applications (classic examples of irregular behavior) very efficiently
-on a commodity cluster. Our yardstick for comparison is the XMT hardware
-itself. Using the same number of network interfaces, Grappa is in the
-same ballpark as the XMT: For unbalanced tree search, Grappa is over 3X
-faster and shows greatly improved scalability; conversely, for breadth
-first search and betweenness centrality Grappa is 2.5X slower.  In
-Section~\ref{sec:evaluation} we explore the factors that underpin this
-performance. Most importantly, however, for significantly less real
-world cost, users can \emph{add\/} significantly more processors to a
-commodity cluster than an XMT machine and use Grappa to achieve scalable
-performance.
-
-\section{Related work}
+Our preliminary work on Grappa shows that it runs several graph-crunching applications (classic examples of irregular behavior) very efficiently on a commodity cluster.
+Our yardstick for comparison is the XMT hardware itself.
+Using the same number of network interfaces, Grappa is in the same ballpark as the XMT: For unbalanced tree search, Grappa is over 3X faster and shows greatly improved scalability; conversely, for breadth first search and betweenness centrality Grappa is 2.5X slower.
+In Section~\ref{sec:evaluation} we explore the factors that underpin this performance.
+Most importantly, however, for significantly less real world cost, users can \emph{add\/} significantly more processors to a commodity cluster than an XMT machine and use Grappa to achieve scalable performance.
 
@@
 
-In the twenty years that have elapsed since the Tera MTA, commodity
-microprocessors have become much faster and multicore has driven down the
-absolute price of computation; commodity network price-performance has
-improved as well. This shift has afforded us the opportunity to attack the
-challenges posed by irregular applications by emulating in software and
-inexpensive mass market hardware, the approach taken by Tera. We exploit the
-increased aggregate instruction rate per socket relative to chip bandwidth,
-using what would otherwise be wasted instructions to manage the multiplexing
-of as many as several thousand tasks per core, thus tolerating memory latency,
-reducing stalls, and making better use of available bandwidth. Ultimately, the
-opportunity is to cover the spectrum of irregular to regular computation:
-where tasks exhibit locality, multiplex fewer tasks and expend fewer
-instructions on context switching; where locality is lacking, multiplex more
-tasks at a higher rate to tolerate latency. Thus we make the best use of task
-parallelism -- either to scale to more cores or to tolerate latency -- and of
-caches -- either to exploit application locality or to house more task
-contexts.
+In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.
+This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating in software and inexpensive mass market hardware, the approach taken by Tera.
+We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.
+Ultimately, the opportunity is to cover the spectrum of irregular to regular computation: where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.
+Thus we make the best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.
 
-Grappa is a software runtime system that allows
-a commodity x86 distributed-memory HPC cluster to be programmed as if it
-were a single large shared-memory machine and provides scalable
-performance for irregular applications. Grappa is designed to smooth
-over some of the performance discontinuities in commodity hardware,
-giving good performance when there is little locality to be exploited
-while allowing the programmer to exploit it when it is available.
+Grappa is a software runtime system that allows a commodity x86 distributed-memory HPC cluster to be programmed as if it were a single large shared-memory machine and provides scalable performance for irregular applications.
+Grappa is designed to smooth over some of the performance discontinuities in commodity hardware, giving good performance when there is little locality to be exploited while allowing the programmer to exploit it when it is available.
 
-Grappa leverages as much freely available and commodity infrastructure as
-possible. We use unmodified Linux for the operating system and an
-off-the-shelf user-mode infiniband device driver stack~\cite{OFED}. MPI is
-used for process setup and tear down. GASnet~\cite{gasnet} is used as the
-underlying mechanism for remote memory reads and writes using active message
-invocations. To this commodity hardware and software mix Grappa adds three
-main software components: (1) a \emph{lightweight tasking\/} layer that supports
-a context switch in as few as 38ns and distributed global load balancing; (2)
-a \emph{distributed shared memory\/} layer that supports normal access operations
-such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer
-that combines short messages to mitigate the aforementioned problem that
-commodity networks are designed to achieve peak bandwidth only on large packet
-sizes, yet irregular applications tend to fetch only a handful of bytes at a
-time. As we will show later, Grappa can tolerate latencies way beyond that of
-the network. Therefore, Grappa can afford to \emph{trade latency for
-throughput\/}: by {\em increasing\/} latency in key components of the system we
-are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations
-to remote nodes), and our ability to improve load imbalance (work stealing
-increases latency).
+Grappa leverages as much freely available and commodity infrastructure as possible.
+We use unmodified Linux for the operating system and an off-the-shelf user-mode infiniband device driver stack~\cite{OFED}.
+MPI is used for process setup and tear down.
+GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
+To this commodity hardware and software mix Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as few as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
+As we will show later, Grappa can tolerate latencies way beyond that of the network.
+Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
 
-Grappa (Figure~\ref{fig:grappa}) has three main software components:
-\begin{description}
+Grappa (Figure~\ref{fig:grappa}) has three main software components: \begin{description}
 
-\item [Tasking system.] Our tasking system supports lightweight
-multithreading to tolerate communication latency and global distributed
-workstealing (i.e., tasks can be stolen from any node in the system), which
-provides automated load balancing.
+\item [Tasking system.] Our tasking system supports lightweight multithreading to tolerate communication latency and global distributed workstealing (i.e., tasks can be stolen from any node in the system), which provides automated load balancing.
 
-\item[Distributed shared memory.] Our DSM system provides support for
-fine-grain access to data anywhere in the system. It supports synchronization
-operations on global data, explicit local caching of any memory in the system,
-and support for operation on remote data (delegating operations to home node).
-By tight integration with the tasking system and the
-communication layer, our DSM system offers high aggregate random
-access bandwidth for accessing remote data.
+\item[Distributed shared memory.] Our DSM system provides support for fine-grain access to data anywhere in the system.
+It supports synchronization operations on global data, explicit local caching of any memory in the system, and support for operation on remote data (delegating operations to home node).
+By tight integration with the tasking system and the communication layer, our DSM system offers high aggregate random access bandwidth for accessing remote data.
 
-\item[Communication layer.] As discussed earlier, modern commodity networks
-support high bandwidth only for large messages. Since irregular applications
-tend to need frequent communication of small requests, the main goal of our
-communication layer is to aggregate small messages into large ones to better
-exploit what the network can offer. It is largely invisible to the application
-programmer.
+\item[Communication layer.] As discussed earlier, modern commodity networks support high bandwidth only for large messages.
+Since irregular applications tend to need frequent communication of small requests, the main goal of our communication layer is to aggregate small messages into large ones to better exploit what the network can offer.
+It is largely invisible to the application programmer.
 
@@
 
-The basic unit of execution in Grappa is a {\em task}. Each task is
-represented by a function pointer and its arguments. A new task is enqueued at
-spawn time; when resources are free, it is allocated a stack, bound to a core,
-and executed.
+The basic unit of execution in Grappa is a {\em task}.
+Each task is represented by a function pointer and its arguments.
+A new task is enqueued at spawn time; when resources are free, it is allocated a stack, bound to a core, and executed.
 
-During execution, a task yields control of its core whenever it performs
-a long-latency operation, allowing the processor to remain busy while
-waiting for the operation to complete.  In addition,  a programmer can
-direct scheduling explicitly via the Grappa API calls shown
-in Figure~\ref{fig:scheduling}.
-To minimize yield overhead, the Grappa scheduler operates entirely in
-user-space and does little more than store register state of one task and load
-that of another. Context switch times are as low as 38ns even when switching
-amongst thousands of tasks.
-
-\begin{figure}[htbp]
-  \begin{center}
-    \begin{description}\small
-    \item[ \texttt{ yield() } ] \hfill \\
-      Yields core to scheduler, enqueuing caller to be scheduled again soon
-    \item[ \texttt{ suspend() } ] \hfill \\
-      Yields core to scheduler, enqueuing caller only once another task calls wake
-    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
-      Enqueues some other task $t$ to be scheduled again soon
-    \end{description}
-    \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:scheduling} Grappa API: scheduling} %{-4ex}}
-    \end{minipage}
-    %\vspace{-3ex}
-  \end{center}
-\end{figure}
+During execution, a task yields control of its core whenever it performs a long-latency operation, allowing the processor to remain busy while waiting for the operation to complete.
+In addition,  a programmer can direct scheduling explicitly via the Grappa API calls.
+To minimize yield overhead, the Grappa scheduler operates entirely in user-space and does little more than store register state of one task and load that of another.
+Context switch times are as low as 38ns even when switching amongst thousands of tasks.
 
@@
 
-Grappa provides four methods for expressing parallelism, shown in Figure~\ref{fig:expressing-parallelism}:
-
+Grappa provides four methods for expressing parallelism.
 First, when the programmer identifies work that can be done in parallel, the work may be wrapped up in a function and queued with its arguments for later execution using a \texttt{spawn}.
-
 Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific core in the system or at the home core of a particular memory location.
-
 Third, the programmer can invoke a parallel for loop, provided that the trip count is known at loop entry.
@@
 It generates a logarithmically-deep tree of tasks, stopping to execute the loop body when the number of iterations is below the required threshold.
-
 Fourth, a programmer may want to execute an active message; that is, to run a small piece of code on a particular core in the system without waiting for execution resources to be available.
 For example, custom synchronization primitives execute this way, as a function executed on the core where the data lives.
-Grappa provides the \texttt{call\_on} call for this purpose.
-
-\begin{figure}[htbp]
-  \begin{center}
-    \begin{description}\small
-    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
-      Creates a new stealable task
-    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
-      Creates a new private task that will run on a specific core 
-    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
-      Executes iterations of a loop as stealable tasks 
-    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
-      Runs a limited function on a specific core without consuming
-      Grappa execution resources 
-    \end{description}
-    \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism} % \vspace{-4ex}}
-    \end{minipage}
-    %\vspace{-3ex}
-  \end{center}
-\end{figure}
 
@@
 
-Applications written for Grappa utilize two forms of memory: local and
-global.  Local memory is local to a single core in the system.  Accesses occur
-through conventional pointers.  The compiler emits an access and the
-memory is manipulated directly.  Applications use local accesses for a
-number of things in Grappa: the stack associated with a task, accesses
-to localized global memory in caches (see below), and accesses to
-debugging infrastructure that is local to each system node.  Local
-pointers cannot access memory on other cores, and are valid only on
-their home core.
+Applications written for Grappa utilize two forms of memory: local and global.
+Local memory is local to a single core in the system.
+Accesses occur through conventional pointers.
+The compiler emits an access and the memory is manipulated directly.
+Applications use local accesses for a number of things in Grappa: the stack associated with a task, accesses to localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
+Local pointers cannot access memory on other cores, and are valid only on their home core.
 
-Large data that is expected to be shared and accessed with low locality is
-stored in Grappa's global memory. All global data must be accessed through
-calls into Grappa's API, shown in Figure~\ref{fig:accessing-memory}.
+Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory.
+All global data must be accessed through calls into Grappa's API, shown in Figure~\ref{fig:accessing-memory}.
 
-\paragraph{Global memory addressing} Grappa provides two methods for
-storing data in the global memory. The first is a distributed heap
-striped across all the machines in the system in a block cyclic fashion.
-The \texttt{global\_malloc} and \texttt{global\_free} calls are used to
-allocate and deallocate memory in the global heap.  Addresses to memory
-in the global heap use \textbf{linear addresses}.  Choosing the block
-size involves trading off sequential bandwidth against aggregate random
-access bandwidth. Smaller block sizes help spread data across all the
-memory controllers in the cluster, but larger block sizes allow the
-locality-optimized memory controllers to provide increased sequential
-bandwidth. The block size, which is configurable, is typically set to 64
-bytes, or the size of a single hardware cache line, in order to exploit
-spatial locality when available. The heap metadata is stored on a single
-node. Currently all heap operations serialize through this node; while
-this has been sufficient for our benchmarks, in the future Grappa will
-provide parallel performance through combining~\cite{MAMA,flatcombining}.
+\paragraph{Global memory addressing} Grappa provides two methods for storing data in the global memory.
+The first is a distributed heap striped across all the machines in the system in a block cyclic fashion.
+The \texttt{global\_malloc} and \texttt{global\_free} calls are used to allocate and deallocate memory in the global heap.
+Addresses to memory in the global heap use \textbf{linear addresses}.
+Choosing the block size involves trading off sequential bandwidth against aggregate random access bandwidth.
+Smaller block sizes help spread data across all the memory controllers in the cluster, but larger block sizes allow the locality-optimized memory controllers to provide increased sequential bandwidth.
+The block size, which is configurable, is typically set to 64 bytes, or the size of a single hardware cache line, in order to exploit spatial locality when available.
+The heap metadata is stored on a single node.
+Currently all heap operations serialize through this node; while this has been sufficient for our benchmarks, in the future Grappa will provide parallel performance through combining~\cite{MAMA,flatcombining}.
 
-Grappa also allows any local data on a core's stacks or heap to be
-exported to the global address space to be made accessible to other
-cores across the system. Addresses to global memory allocated in this
-way use \textbf{2D global addresses}.  This uses a traditional PGAS
-addressing model, where each address is a tuple of a rank in the job (or
-global process ID) and an address in that process. The lower 48 bits of
-the address hold a virtual address in the process. The top bit is set to
-indicates that the reference is a 2D address (as opposed to linear
-address). This leaves 15 bits for network endpoint ID, which limits our
-scalability to $2^{15}$ endpoints. Any node-local data can be made
-accessible by other nodes in the system by wrapping the address and node
-ID into a 2D global address. This address can then be accessed with a
-delegate and can also be cached by other nodes. At the destination the
-address is converted into a canonical x86 address by replacing the upper
-bits with the sign-extended upper bit of the virtual address. 2D
-addresses may refer to memory allocated from a single processes' heap or
-from a task's stack. Figure~\ref{fig:memory-structure} shows how 2D and
-linear addresses can refer to other cores' memory.
+Grappa also allows any local data on a core's stacks or heap to be exported to the global address space to be made accessible to other cores across the system.
+Addresses to global memory allocated in this way use \textbf{2D global addresses}.
+This uses a traditional PGAS addressing model, where each address is a tuple of a rank in the job (or global process ID) and an address in that process.
+The lower 48 bits of the address hold a virtual address in the process.
+The top bit is set to indicates that the reference is a 2D address (as opposed to linear address).
+This leaves 15 bits for network endpoint ID, which limits our scalability to $2^{15}$ endpoints.
+Any node-local data can be made accessible by other nodes in the system by wrapping the address and node ID into a 2D global address.
+This address can then be accessed with a delegate and can also be cached by other nodes.
+At the destination the address is converted into a canonical x86 address by replacing the upper bits with the sign-extended upper bit of the virtual address.
+2D addresses may refer to memory allocated from a single processes' heap or from a task's stack.
 
-\begin{figure}[htbp]
-  \begin{center}
-    \begin{description}\small
-      \item[ \texttt{ global\_address global\_malloc( size )} ] \hfill \\
-      \item[ \texttt{ global\_free( global\_address )} ] \hfill \\
-        Allocates and frees memory in the global heap
-      \item[ \texttt{ delegate\_read( global\_address, local\_var )} ] 
-      \item[ \texttt{ delegate\_write( global\_address, local\_var )} ] %\vspace{-2ex}
-      \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] %\vspace{-2ex}
-      \item[ \texttt{ delegate\_fetch\_inc( global\_address, local\_var )} ] %\vspace{-2ex} 
-\hfill \\
-        Performs a memory operation at the home core of a global address
-      \item[ \texttt{ cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} ]
-      \item[ \texttt{ cache\_release( global\_address, local\_buf )} ] %\vspace{-2ex} 
-\hfill \\
-Perform cache operations to acquire/release global data.
-Acquire, returns a local pointer after all data has been copied to the local node.
-Release, optionally writes data back to global memory and frees up management resources.
-    \end{description}
-    \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:accessing-memory} Grappa API: accessing memory} %{-4ex}}
-    \end{minipage}
-    %\vspace{-3ex}
-  \end{center}
-\end{figure}
+\paragraph{Global memory access} There are two general approaches Grappa applications use to {\emph access} global memory.
+When the programmer expects a computation on shared data to have spatial locality to exploit, {\em cache} operations may be used.
+When there is no locality to exploit, {\em delegate} operations are used.
 
-\begin{figure}[t]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:memory-structure} Grappa memory structure}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
+\textbf{Explicit caching.} Grappa provides an API to fetch a global pointer of any length and return a local pointer to a cached copy of the global memory.
+Grappa cache operations have the usual read-only and read-write variants, along with a write-only variant used to initialize data structures.
+Languages for distributed shared memory systems have done optimizations to achieve a similar goal.
+For example, the UPC compiler coalesces struct and array accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's message vectorization hoists small messages out of a loop \cite{FortranD:1992}.
+Caching in Grappa additionally provides a mechanism for exploiting temporal locality by operating on the data locally.
 
-\paragraph{Global memory access} There are two general approaches
-Grappa applications use to {\emph access} global memory. When the
-programmer expects a computation on shared data to have spatial locality
-to exploit, {\em cache} operations may be used. When there is no
-locality to exploit, {\em delegate} operations are used.
+Under the hood, Grappa performs the mechanics of gathering chunks of data from multiple system nodes and presenting a conventional appearing linear block of memory as a local pointer into a cache.
+The strategy employed is to issue all the constituent requests of a cache access request  (as Active Messages) and then yield until all responses have occurred.
+Currently, Grappa caches are \emph{not} coherent, requiring the programmer to maintain consistent access to data.
+Future work will develop a software directory based coherence scheme to simplify consistent access to global data.
 
-\textbf{Explicit caching.} Grappa provides an API to fetch a global
-pointer of any length and return a local pointer to a cached copy of the
-global memory.  Grappa cache operations have the usual read-only and
-read-write variants, along with a write-only variant used to initialize
-data structures. Languages for distributed shared memory systems have
-done optimizations to achieve a similar goal. For example, the UPC
-compiler coalesces struct and array accesses into remote get/put
-\cite{Chen:2005}, and Fortran D compiler's message vectorization hoists
-small messages out of a loop \cite{FortranD:1992}. Caching in Grappa
-additionally provides a mechanism for exploiting temporal locality by
-operating on the data locally.
+\textbf{Delegate operations.} When the access pattern has low-locality, it is more efficient to modify the data on its home core rather than bringing a copy to the requesting core and returning it after modification.
+Delegate operations provide this capability.
+Applications can dispatch computation to be performed on individual machine-word sized chunks of global memory to the memory system itself (e.g., \emph{fetch-and-add}).
+Delegate operations, proposed in \cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the primary synchronization method in Grappa.
 
-Under the hood, Grappa performs the mechanics of gathering chunks of
-data from multiple system nodes and presenting a conventional appearing
-linear block of memory as a local pointer into a cache. The strategy
-employed is to issue all the constituent requests of a cache access
-request  (as Active Messages) and then yield until all responses have
-occurred.  Currently, Grappa caches are \emph{not} coherent, requiring
-the programmer to maintain consistent access to data.  Future work will
-develop a software directory based coherence scheme to simplify
-consistent access to global data.
-
-\textbf{Delegate operations.} When the access pattern has low-locality,
-it is more efficient to modify the data on its home core rather than
-bringing a copy to the requesting core and returning it after
-modification. Delegate operations provide this capability. Applications
-can dispatch computation to be performed on individual machine-word
-sized chunks of global memory to the memory system itself (e.g.,
-\emph{fetch-and-add}).  Delegate operations, proposed in
-\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the
-primary synchronization method in Grappa.
-
-Delegate operations are always executed at the home core of their
-address, and while arbitrary memory operations can be delegated, we
-restrict the use of delegate operations in three ways to make them more
-useful for synchronization. First, we limit each task to one outstanding
-delegate operation to avoid the possibility of reordering in the
-network. Second, we limit delegate operations to operate on objects in
-the 2D address space or objects that fit in a single block of the linear
-address space so they can be satisfied with a single network request.
-Finally, no context switches are allowed while the data is being
-modified. Given these restrictions, we can ensure that delegate
-operations for the same address from multiple requesters are always
-serialized through a single core in the system, providing atomic
-semantics without using atomic operations. 
-Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
-cache operations interact.
+Delegate operations are always executed at the home core of their address, and while arbitrary memory operations can be delegated, we restrict the use of delegate operations in three ways to make them more useful for synchronization.
+First, we limit each task to one outstanding delegate operation to avoid the possibility of reordering in the network.
+Second, we limit delegate operations to operate on objects in the 2D address space or objects that fit in a single block of the linear address space so they can be satisfied with a single network request.
+Finally, no context switches are allowed while the data is being modified.
+Given these restrictions, we can ensure that delegate operations for the same address from multiple requesters are always serialized through a single core in the system, providing atomic semantics without using atomic operations.
+Figure~\ref{fig:delegate-cache} depicts an example of how delegate and cache operations interact.
 
@@
 \begin{center}
-  \includegraphics[width=1.5\columnwidth]{figs/delegate-cache}
-\begin{minipage}{1.9\columnwidth}
-  \caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in
-the global heap. It then spawns two tasks on remote cores to increment
-elements of the array.  The first task increments two elements of the array using cache
-operations. In step 2, the task is invoked. A cache request is issued
-for two adjacent integers starting at the second element of the
-array. Since these element are stored in the memories of two different
-cores, this requires the sending of two messages in step 3. The task
-is suspended until both responses arrive in step 4. The data carried
-in these responses is stored in the local buffer. The elements are
-then increment in the buffer. Then in step 5, the modified data is
-sent back to the home node. Acknoledgements are returned in step 6 so
-the task knows when the writes are complete. The second task increments an element of the array with a delegate
-operation. In step 7, the task is invoked. A delegate request is sent
-to the home core of the array element with the increment value. The
-task suspends until the response is received. In step 8, the increment
-is executed on the remote core. A response is returned in step 9 with
-the previous value of the array element.
+  \includegraphics[width=6in]{figs/delegate-cache}
+\begin{minipage}{\columnwidth}
+
+\caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in the global heap.
+It then spawns two tasks on remote cores to increment elements of the array.
+The first task increments two elements of the array using cache operations.
+In step 2, the task is invoked.
+A cache request is issued for two adjacent integers starting at the second element of the array.
+Since these element are stored in the memories of two different cores, this requires the sending of two messages in step 3.
+The task is suspended until both responses arrive in step 4.
+The data carried in these responses is stored in the local buffer.
+The elements are then increment in the buffer.
+Then in step 5, the modified data is sent back to the home node.
+Acknoledgements are returned in step 6 so the task knows when the writes are complete.
+The second task increments an element of the array with a delegate operation.
+In step 7, the task is invoked.
+A delegate request is sent to the home core of the array element with the increment value.
+The task suspends until the response is received.
+In step 8, the increment is executed on the remote core.
+A response is returned in step 9 with the previous value of the array element.
 }
+
 \end{minipage}
@@
 
-In order to mitigate the low message injection rate limits of commodity
-networks, Grappa's communication stack has two layers: one for
-user-level messages and one for network-level messages.
+In order to mitigate the low message injection rate limits of commodity networks, Grappa's communication stack has two layers: one for user-level messages and one for network-level messages.
 
-At the upper layer, Grappa implements asynchronous active messages
-\cite{vonEicken92}. Each message consists of a function pointer, an
-optional argument payload, and an optional data payload. When a task
-sends a message, the message is copied to a send queue associated with
-the message's destination and the task continues execution.
+At the upper layer, Grappa implements asynchronous active messages \cite{vonEicken92}.
+Each message consists of a function pointer, an optional argument payload, and an optional data payload.
+When a task sends a message, the message is copied to a send queue associated with the message's destination and the task continues execution.
 
-Grappa's lower networking layer aggregates the upper layer's messages
-to improve performance. Commodity networks including infiniband
-achieves their peak bisection bandwidth \emph{only} when the packet
-sizes are relatively large---on the order of multiple kilobytes. The
-reason for this discrepancy is the combination of overheads associated
-with handling each packet (in terms of bytes that form the actual
-packet, processing time at the card and processing on the
-CPU within the driver stack). Our measurements confirm manufacturers
-published data~\cite{infinibandbandwidth}, that with this packet size
-the bisection bandwidth is only a small fraction, less than
-3\% of the peak bisection bandwidth.
+Grappa's lower networking layer aggregates the upper layer's messages to improve performance.
+Commodity networks including infiniband achieves their peak bisection bandwidth \emph{only} when the packet sizes are relatively large---on the order of multiple kilobytes.
+The reason for this discrepancy is the combination of overheads associated with handling each packet (in terms of bytes that form the actual packet, processing time at the card and processing on the CPU within the driver stack).
+Our measurements confirm manufacturers published data~\cite{infinibandbandwidth}, that with this packet size the bisection bandwidth is only a small fraction, less than 3\% of the peak bisection bandwidth.
 
-In our experiments the vast majority of requests were smaller than
-44 bytes, far too small to make
-efficient use of the network. To make the best use of the network, we
-must convert our small messages into large ones.
-When a task sends a message,
-it is not immediately sent, but rather placed in a queue specific to the
-destination.
+In our experiments the vast majority of requests were smaller than 44 bytes, far too small to make efficient use of the network.
+To make the best use of the network, we must convert our small messages into large ones.
+When a task sends a message, it is not immediately sent, but rather placed in a queue specific to the destination.
 
-There are three situations in which a queue of aggregated messages is
-sent. First, each queue has a message size threshold of 4096 bytes, chosen to give reasonable network performance. If the size in
-bytes of a queue is above the threshold, the contents of the queue are
-sent immediately. Second, each queue has a wait time threshold
-($\approx${1ms}). If the oldest message in a queue has been waiting
-longer than this threshold, the contents of the queue are sent
-immediately, even if the queue size is lower than the message size
-threshold.  Third, queues may be explicitly flushed in situations where
-the programmer wants to minimize the latency of a message at the cost of
-bandwidth utilization.
+There are three situations in which a queue of aggregated messages is sent.
+First, each queue has a message size threshold of 4096 bytes, chosen to give reasonable network performance.
+If the size in bytes of a queue is above the threshold, the contents of the queue are sent immediately.
+Second, each queue has a wait time threshold ($\approx${1ms}).
+If the oldest message in a queue has been waiting longer than this threshold, the contents of the queue are sent immediately, even if the queue size is lower than the message size threshold.
+Third, queues may be explicitly flushed in situations where the programmer wants to minimize the latency of a message at the cost of bandwidth utilization.
 
-The network layer is serviced by polling. Periodically when a context
-switch occurs, the Grappa scheduler switches to the network polling
-thread. This thread has three responsibilities. First, it polls the
-lower-level network layer to ensure it makes progress. Second, it
-deaggregates received messages and executes active message
-handlers. Third, it checks to see if any aggregation queues have
-messages that have been waiting longer than the threshold; if so, it
-sends them.
+The network layer is serviced by polling.
+Periodically when a context switch occurs, the Grappa scheduler switches to the network polling thread.
+This thread has three responsibilities.
+First, it polls the lower-level network layer to ensure it makes progress.
+Second, it deaggregates received messages and executes active message handlers.
+Third, it checks to see if any aggregation queues have messages that have been waiting longer than the threshold; if so, it sends them.
 
-Underneath the aggregation layer, Grappa uses the \gasnet~communication
-library~\cite{gasnet} to actually move data. All interprocess
-communication, whether on or off a cluster node, is handled by the
-\gasnet~library. \gasnet~is able to take advantage of many communication
-mechanisms, including ethernet and infiniband between nodes, as well as
-shared memory within a node.
+Underneath the aggregation layer, Grappa uses the \gasnet~communication library~\cite{gasnet} to actually move data.
+All interprocess communication, whether on or off a cluster node, is handled by the \gasnet~library.
+\gasnet~is able to take advantage of many communication mechanisms, including ethernet and infiniband between nodes, as well as shared memory within a node.
 
-Some networks provide access to a remote machine's memory directly. This
-would seem to be a good fit for a programming model focused on global
-shared memory, but in fact we do not use it. In our experiments, we
-found that RDMA operations are subject to the same message rate
-limitations as all other messages on these cards, and thus using raw
-RDMA operations for our small messages would make inefficient use of
-bandwidth. Instead, we implement remote memory operations with active
-messages. A byproduct of this design decision is that Grappa is not
-limited to RDMA-capable networks.
+Some networks provide access to a remote machine's memory directly.
+This would seem to be a good fit for a programming model focused on global shared memory, but in fact we do not use it.
+In our experiments, we found that RDMA operations are subject to the same message rate limitations as all other messages on these cards, and thus using raw RDMA operations for our small messages would make inefficient use of bandwidth.
+Instead, we implement remote memory operations with active messages.
+A byproduct of this design decision is that Grappa is not limited to RDMA-capable networks.
 
@@
 
-To evaluate Grappa's performance with respect to the XMT, we ran each
-of our three benchmarks on up to 16 nodes of each machine. Grappa used
-6 cores per node, with the best parameters chosen for each point. In
-some cases, the XMT could not run the benchmark with 2 nodes, so the
-point is omitted.
+To evaluate Grappa's performance with respect to the XMT, we ran each of our three benchmarks on up to 16 nodes of each machine.
+Grappa used 6 cores per node, with the best parameters chosen for each point.
+In some cases, the XMT could not run the benchmark with 2 nodes, so the point is omitted.
 
-%% UTS: performance comparison
-\begin{figure}[ht]
+\begin{figure}
 \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/uts_performance.pdf}
+
+\hspace{-1in}\begin{minipage}{0.3\textwidth}
+\begin{center}
+\includegraphics[width=3in]{figs/uts_performance.pdf}
+\caption{\label{fig:uts_compare} Performance of in-memory unbalanced tree search.}
 \end{center}
-    \caption{Performance of in-memory unbalanced tree search.}
-    \label{fig:uts_compare}
+\end{minipage}
+\hspace{1.25in}\
+\begin{minipage}{0.3\textwidth}
+\begin{center}
+\includegraphics[width=3in]{figs/bfs_performance}
+\caption{\label{fig:bfs-performance} Performance of breadth-first searcch.}
+\end{center}
+\end{minipage}
+
+\end{center}
 \end{figure}
 
-\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex tree
-(T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
-number of vertices visited per second versus number of compute
-nodes.
+\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex tree (T1L).
+Figure~\ref{fig:uts_compare} shows the performance in terms of number of vertices visited per second versus number of compute nodes.
 Grappa is 3.2 times faster than the XMT at 16 nodes.
@@
 
-\begin{figure}[tH]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/bfs_performance}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:bfs-performance} BFS performance}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
+\paragraph{BFS} We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and $2^{29}$ edges (25 GB of data).
+Figure~\ref{fig:bfs-performance} shows our performance in terms of graph edges traversed per second.
+The XMT is 2.5 times faster than Grappa at 16 nodes.
+Performance does scale at a constant rate for Grappa, suggesting that adding more nodes will increase performance.
 
-\paragraph{BFS}
+\section{Related work}
 
-We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and
-$2^{29}$ edges (25 GB of data). Figure~\ref{fig:bfs-performance} shows
-our performance in terms of graph edges traversed per second. The XMT
-is 2.5 times faster than Grappa at 16 nodes.  Performance does scale at a constant rate for Grappa, suggesting that adding more nodes will increase performance.
+Our efforts are built on many existing ideas in programming languages, systems and architecture. In this section we discuss related frameworks and key enabling technologies that Grappa builds upon.
 
-\begin{figure}[tH]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/centrality_performance}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:centrality-performance} Centrality performance}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
+\paragraph{Comparable frameworks} Distributed graph processing frameworks like Pregel~\cite{pregel:2010} and Distributed GraphLab~\cite{distgraphlab:vldb12} share similar goals as Grappa.
+Pregel adopts a bulk-synchronous parallel (BSP) execution model, which makes it inefficient on workloads that could prioritize vertices.
+GraphLab, on the other hand, schedules vertex computations individually, allowing prioritization, which gives faster convergence in a variety of iterative algorithms.
+GraphLab, however, imposes a rigid computation model where programmers must express computation as transformations on a vertex and its edge list only, with information only from adjacent vertexes.
+Pregel is only slightly less restrictive, as the input data can be any vertex in the graph.
+Grappa also supports dynamic parallelism with asynchronous execution, but parallelism is expressed as tasks or loop iterations, which is a far more general programming model for irregular computation tasks.
 
-\paragraph{Centrality}
+\paragraph{Global memory} Grappa includes a custom implementation of a software distributed shared memory (DSM) system.
+Many traditional software DSM systems are page based~\cite{Treadmarks,munin} and aim to hide the fact that they are built in software from applications by exploiting the processor's paging mechanisms, therefore relying heavily on locality.
+Instead, Grappa, like other partitioned global address space (PGAS) models, implements its DSM at the language, rather than system level.
+Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and UPC~\cite{upc:2005} make accesses to shared structures look like normal memory references.
+As we describe later, Grappa chooses a middle ground, where global addresses are explicit in the API and local accesses are emitted conventionally by the compiler.
+While Grappa's DSM system is conceptually similar to prior work, its implementation is tuned for irregular computations.
+Past DSM work, being page-based, could exploit the RDMA capabilities of network hardware to move large page-sized blocks of data from node to node.
+In our experience, when these networks move small blocks of data (a few bytes), only a fraction of the available bandwidth is achieved.
+In addition, the DSM system in Grappa ends up being tightly coupled to the task scheduler in order to overlap long latency memory operations with useful computation.
+For these two reasons it became necessary to build a new DSM system specifically for Grappa.
 
-We ran Betweenness Centrality on the same scale 25 Kronecker graph as
-we did for BFS. Figure~\ref{fig:centrality-performance} shows our
-performance in terms of graph edges traversed per second. At 16 XMT
-processors/cluster nodes, the XMT is 1.75 times faster than Grappa.
+\paragraph{Multithreading} Grappa uses multithreading to tolerate memory latency.
+This is a well known technique.
+Hardware implementations include the Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife}, Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}.
+As we describe in this paper, Grappa use a lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks on a single processing core.
+The large number of tasks is required because of the extremely high internode latency Grappa is mitigating.
+Grappa's task library employs several optimizations: an extremely fast task switch, a small task size, and judicious use of hardware prefetching to bring task state into the cache long before that task is actually scheduled.
+
+\section{Proposed Research}
+
+\subsection{Improved Networking layer}
+
+\paragraph{Virtual topology:} Currently Grappa uses a flat network topology where each processing core in the system is individually addressed.
+What this means, for example, is in a 128 node system, where each node is comprised of 16 processing cores, the network has 2048 end-points.
+The aggregator treats each of these end-points individually.
+This design choice is simple, and avoids lock contention within the underlying \gasnet layer, but does not scale.
+As more system nodes are added, a quadratic increase in the number of threads is required in order to sustain aggregator performance.
+
+We need to rewrite the aggregator to build in a hierarchical virtual network layer.
+The first layer of this hierarchy is to treat all processors at a given system node as a single destination.
+We intend to continue to use the lock-free single-threaded \gasnet communication layer, as we have found the lock-based one to incur too high of an overhead for our purposes.
+Thus the aggregator will need to very carefully utilize lock-free data-structures.
+
+The second, and upper layers of the virtual topology are built from clustering system nodes into groups.
+Instead of aggregating for all end-points, aggregation will occur for system nodes within a group.
+A designated (and dedicated) node will act as a router, processing incoming and outgoing traffic to system nodes within the group and other groups.
+
+Hierarchy will reduce scaling requirements from \BigO{n^2} threads to \BigO{n log(n)}.
+Of course, it takes what used to be a single network hop for communication and turns it into \BigO{log(n)} hops.
+More threads will be required to overcome this increased latency, but we believe only \BigO{log(n)} more, hence a real scaling benefit is achievable with the a virtual hierarchical network topology.
+
+\paragraph{User-level RDMA:} Certain commodity networking hardware today supports remote direct memory access (RDMA).
+Unfortunately, \emph{how} it is implemented is so crude that it is difficult or impossible to efficiently use in practice.
+We start with a short description of the current semantics of RDMA implementations.
+Next we describe the research we wish to undertake to make RDMA more useful.
+
+Networking gear supports remote read and write operations.
+Read operations take a source address and key (on a remote node), a destination address and key (on the local node), and a length.
+Cards support a address translation for regions of memory that are registered with them.
+These translations, indexed by the supplied key, are used to perform virtual to physical address translation.
+The physical address of the destination address is similarly translated using the translation table at the local node.
+Cards also support remote write operations with a similar set of semantics, where addresses are translated on cards using the cards local translation table.
+Both read and write operations are asynchronous.
+The requesting processor initiates the operation and then can check for completion by querying a completion queue at the local card.
+\emph{Notably}, the processor(s) on the remote node are provided a polling interface whereby they can learn if a remote write operation occurred.
+No such interface exists to learn about remote read or atomic operations.
+
+This interface suffers from multiple shortcomings.
+Ultimately, moving data to and from a remote node requires the interaction of two different operating system processes existing on two different system nodes.
+Those processes each maintain a collection of runtime data structures and moving data among them means interacting with those data structures constructively.
+Often such
+
+\subsection{SSD-only data structures}
+
+\subsection{Graph benchmark suite}
 

### 2013-02-01 doc/xps2013/proposal.tex
partial commit of research section

@@
 This interface suffers from multiple shortcomings.
-Ultimately, moving data to and from a remote node requires the interaction of two different operating system processes existing on two different system nodes.
-Those processes each maintain a collection of runtime data structures and moving data among them means interacting with those data structures constructively.
-Often such
+The biggest of which is that ultimately, moving data to and from a remote node constructively requires higher level semantic operations than simple reads and writes.
+These higher level semantics can, of course, be synthesized from the underlying read and write primitives, but it wastes round trips and synchronization time.
+One possible alternative solution is to borrow ideas from Active Messages~\cite{active-messages} and enable remote nodes to dispatch, essentially, user-level function calls.
+The challenge here is to make those calls interoperate securely with a modern operating system and multicore system and be efficient.
+\comment{This seems weak.
+Need some more ideas here.
+-Mark}
+
+The second challenge is the virtual to physical address translation on the cards requires the operating system to pin translations.
+This removes paging flexibility from the operating system and allows user processes to (potentially unfairly) grab ahold of physical memory.
+In the long run, the solution here likely lays at the intersection of what progress is occurring in virtual memory translation for I/O with virtual machines and RDMA.
+Our goal will be carefully study the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.
+\comment{weak?
+-M}
+
+The final challenge is that atomic operations via RDMA do not interact atomically with processors in a system.
+Semantically it is not entire clear what happens if a processor issues an atomic operation on a memory address at the same time the network card is attempting to service a remote atomic access.
+\comment{maybe this is not true.
+have to think some more about PCIe an what happens here.
+might not be a problem.
+-M}
 
 \subsection{SSD-only data structures}
+
+\subsection{Compiler and language support}
 

### 2013-02-01 doc/asplos13/paper.tex
initial methodology section
Added initial implementation discussion
add evaluation section, start to UTS subsection/plot
pass on abs,intro
add acknowledgements
paper checklist from asplos13 cfp
more checklist
formatting job
Super redux
removing TODOs
merge conflicts
minor stuff
Remove redundant 'references' section (Preston)

@@
 \input{header}
-
+\usepackage{microtype}
 \begin{document}
 
-\title{All Your Graphs Belong to Us}
-\author{TBD}
+\title{Grappa: Scalable Computing on Commodity Hardware \\for Irregular Applications}
+%\title{All Your Graphs Belong to Us}
+%\authorinfo{\vspace{-2ex}Anonymous for review\vspace{-6ex}}{}{}
+\authorinfo{Jacob Nelson$^{\dagger}$,
+  Brandon Holt$^{\dagger}$,
+  Brandon Myers$^{\dagger}$,
+  Preston Briggs$^{\dagger}$,
+  Luis Ceze$^{\dagger}$,
+  Simon Kahan$^{{\dagger \ddagger}}$,
+  Mark Oskin$^{\dagger}$
+}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
+  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu}
 
@@
 
-\input{whynot}
+\input{relatedwork}
 
-\input{grappa}
+\input{overview}
 
-\subsection{References}
+\input{tasks}
+
+\input{memory}
+
+\input{communication}
+
+\input{method}
+
+\input{evaluation}
+
+\input{conclude}
+
+%\input{acknowledgements}
+
+%\subsection{References}
 \bibliographystyle{abbrv}

### 2013-02-04 doc/xps2013/proposal.tex
for luis

@@
 
+In this proposal we describe a body of research that extends the Grappa~\cite{grappa} runtime system.
+Grappa is a runtime system that compilers target irregular applications to.
+It provides several facilities: a global shared memory abstraction, user-level threading support, and communication and synchronization primitives.
+Applications targeted for Grappa are not expected to have \emph{any} locality (although what locality that exists is favorably exploited), but are expected to provide many thousands of concurrent threads.
+
 Our preliminary work on Grappa shows that it runs several graph-crunching applications (classic examples of irregular behavior) very efficiently on a commodity cluster.
-Our yardstick for comparison is the XMT hardware itself.
-Using the same number of network interfaces, Grappa is in the same ballpark as the XMT: For unbalanced tree search, Grappa is over 3X faster and shows greatly improved scalability; conversely, for breadth first search and betweenness centrality Grappa is 2.5X slower.
-In Section~\ref{sec:evaluation} we explore the factors that underpin this performance.
-Most importantly, however, for significantly less real world cost, users can \emph{add\/} significantly more processors to a commodity cluster than an XMT machine and use Grappa to achieve scalable performance.
+For unbalanced tree search, Grappa is over 3X faster and shows greatly improved scalability compared to the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower.
+Most importantly, however, for significantly less real world cost, users can \emph{add} significantly more processors to a commodity cluster than an XMT machine and use Grappa to achieve scalable performance.
+
+Grappa work is still in its infancy, however.  This research proposal seeks funding to extend Grappa in the following directions:
+
+\begin{itemize}
+\item Presently, applications are hand-coded to the Grappa interface.  A significant research thrust of the proposed work is to ease this programming burden by language and compilation support.
+
+\item Our preliminary work has also revealed limitations of the current Grappa runtime system in the area of network scaling.
+This proposal seeks funding to enhance the network layer and explore the architecture/software interactions of remote direct memory access (RDMA).
+
+\item The size of data being processed by large systems today (hundreds of terabytes) dwarfs what can fit into a single system.
+Obviously multiple system nodes are required to maintain and access all of this data efficiently.
+However, even with hundreds of channels to persistent storage, loading this data into DRAM can prove a bottleneck.
+The Grappa runtime is designed to make applications tolerate extremely high memory latency, thus an idea we wish to explore is using solid state disks as main memory.
+
+\item Presently there does not exist a comprehensive benchmark suite of irregular and graph applications.  We also seeking funding through this proposal to put together just such a
+
+\end{itemize}
+
+%% Roadmap
 
@@
 In the long run, the solution here likely lays at the intersection of what progress is occurring in virtual memory translation for I/O with virtual machines and RDMA.
-Our goal will be carefully study the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.
+Our goal will be to carefully study the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.
 \comment{weak?
@@
 The final challenge is that atomic operations via RDMA do not interact atomically with processors in a system.
-Semantically it is not entire clear what happens if a processor issues an atomic operation on a memory address at the same time the network card is attempting to service a remote atomic access.
+Semantically it is not entirely clear what happens if a processor issues an atomic operation on a memory address at the same time the network card is attempting to service a remote atomic access.
 \comment{maybe this is not true.

### 2013-02-05 doc/xps2013/proposal.tex
some notes on the research section

@@
 
-\subsection{Improved Networking layer}
+\comment{I think we should pitch what have been done so far as essays and a bit part of the research effort is putting everything together. -Luis}
 
-\paragraph{Virtual topology:} Currently Grappa uses a flat network topology where each processing core in the system is individually addressed.
-What this means, for example, is in a 128 node system, where each node is comprised of 16 processing cores, the network has 2048 end-points.
-The aggregator treats each of these end-points individually.
-This design choice is simple, and avoids lock contention within the underlying \gasnet layer, but does not scale.
-As more system nodes are added, a quadratic increase in the number of threads is required in order to sustain aggregator performance.
+\subsection{Networking}
+
+\comment{This research item	is important but right now it sounds really really
+specific. You almost need to know the current code to understand it. -Luis}
+
+\comment{Ins't the argument here something like: If we did aggregation on a
+per-destination basis only, as the number of nodes scales, it takes more
+concurrency (or more time) to aggregate messages. We will address this by
+doing aggrregation hierarchically, choosing representative end-points along
+the way to increase agggregation opporunitites at the cost of more
+communucation hops (making an analogy with changing trains here). -Luis}
+
+\paragraph{Virtual topology:} Currently Grappa uses a flat network topology
+where each processing core in the system is individually addressed. What this
+means, for example, is in a 128 node system, where each node is comprised of
+16 processing cores, the network has 2048 end-points. The aggregator treats
+each of these end-points individually. This design choice is simple, and
+avoids lock contention within the underlying \gasnet layer, but does not
+scale. As more system nodes are added, a quadratic increase in the number of
+threads is required in order to sustain aggregator performance.
 
@@
 
+% this paragraph is great!!
 Hierarchy will reduce scaling requirements from \BigO{n^2} threads to \BigO{n log(n)}.
@@
 Need some more ideas here.
--Mark}
+-Mark.
+Luis: Isn't the problem that the app runtime should be able to determine how the RDMA ops should be done?
+Maybe add examples here like accessing a hash-table, doing collective operations on remote data etc (min, max, avg)?
+How about delegating simple operations to the network card?
+Maybe argue that getting the attention of the remote processor for these operations is too costly?}
 
@@
 \comment{weak?
--M}
+-M If you are 100\% sure this is true, then it is not weak, but if we are wrong, then people can shoot down the proposal saying we know nothing about how RDMA is done today...}
 
@@
 
+\comment{mostly a copy and paste of the pomace?}
+
 \subsection{Compiler and language support}

### 2013-02-05 doc/xps2013/proposal.tex
cbm

--- 2013-02-04 doc/xps2013/proposal.tex
@@
 
-\item Our preliminary work has also revealed limitations of the current Grappa runtime system in the area of network scaling.
-This proposal seeks funding to enhance the network layer and explore the architecture/software interactions of remote direct memory access (RDMA).
+\item Our preliminary work has also revealed limitations of the current Grappa runtime system with regards to network scaling.
+This proposal seeks funding to enhance the network layer and explore the hardware/software interactions of remote direct memory access (RDMA).
 
 \item The size of data being processed by large systems today (hundreds of terabytes) dwarfs what can fit into a single system.
-Obviously multiple system nodes are required to maintain and access all of this data efficiently.
+Multiple system nodes are required to maintain and access all of this data efficiently.
 However, even with hundreds of channels to persistent storage, loading this data into DRAM can prove a bottleneck.
@@
 
-\item Presently there does not exist a comprehensive benchmark suite of irregular and graph applications.  We also seeking funding through this proposal to put together just such a
+\item Presently there does not exist a comprehensive benchmark suite of irregular and graph applications.
+We also seeking funding through this proposal to put together just such a benchmark suite.
+This suite will include benchmarks, reference implementations, and input data-sets.
+In addition a website will be created \texttt{graphbench.org} to distribute the benchmark suite and publish results.
 

### 2013-02-05 doc/xps2013/proposal.tex
for Luis

@@
 
-\paragraph{Virtual topology:} Currently Grappa uses a flat network topology
-where each processing core in the system is individually addressed. What this
-means, for example, is in a 128 node system, where each node is comprised of
-16 processing cores, the network has 2048 end-points. The aggregator treats
-each of these end-points individually. This design choice is simple, and
-avoids lock contention within the underlying \gasnet layer, but does not
-scale. As more system nodes are added, a quadratic increase in the number of
-threads is required in order to sustain aggregator performance.
+\paragraph{Virtual topology:} Currently Grappa uses a flat network topology where each processing core in the system is individually addressed.
+As more nodes are added to the system, a quadratic increase in the number of threads (in aggregate across all nodes) is required in order to compensate for the \emph{same} network delay.
+This is due to the fact that the aggregator treats each of these end-points individually.
+This design choice is simple but does not scale.
 
@@
 The first layer of this hierarchy is to treat all processors at a given system node as a single destination.
-We intend to continue to use the lock-free single-threaded \gasnet communication layer, as we have found the lock-based one to incur too high of an overhead for our purposes.
+We intend to continue to use the lock-free single-threaded \gasnet communication layer that we have been using, as we have found the lock-based one to incur too high of an overhead for our purposes.
 Thus the aggregator will need to very carefully utilize lock-free data-structures.
@@
 
-% this paragraph is great!!
-Hierarchy will reduce scaling requirements from \BigO{n^2} threads to \BigO{n log(n)}.
+Hierarchy will reduce scaling requirements from \BigO{n^2} threads to \BigO{n \times log(n)}.
 Of course, it takes what used to be a single network hop for communication and turns it into \BigO{log(n)} hops.
-More threads will be required to overcome this increased latency, but we believe only \BigO{log(n)} more, hence a real scaling benefit is achievable with the a virtual hierarchical network topology.
+More threads will be required to overcome this increased latency, but we believe only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling benefit is achievable with the a virtual hierarchical network topology.
 
@@
 Read operations take a source address and key (on a remote node), a destination address and key (on the local node), and a length.
-Cards support a address translation for regions of memory that are registered with them.
+Cards support address translation for regions of memory that are registered with them.
 These translations, indexed by the supplied key, are used to perform virtual to physical address translation.
@@
 One possible alternative solution is to borrow ideas from Active Messages~\cite{active-messages} and enable remote nodes to dispatch, essentially, user-level function calls.
-The challenge here is to make those calls interoperate securely with a modern operating system and multicore system and be efficient.
+The challenge here is to make those calls interoperate securely with a modern operating system, and a user-level runtime system.
 \comment{This seems weak.
@@
 The second challenge is the virtual to physical address translation on the cards requires the operating system to pin translations.
-This removes paging flexibility from the operating system and allows user processes to (potentially unfairly) grab ahold of physical memory.
-In the long run, the solution here likely lays at the intersection of what progress is occurring in virtual memory translation for I/O with virtual machines and RDMA.
+This removes paging flexibility from the operating system because runtime systems are (rationally) built to expect data structures to be in memory capable of RDMA or not.
+Those in RDMA memory must be pinned.
+In the long run, the solution here likely lays at the intersection of what progress is occurring in virtual memory translation for I/O within virtual machines.
 Our goal will be to carefully study the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.
-\comment{weak?
--M If you are 100\% sure this is true, then it is not weak, but if we are wrong, then people can shoot down the proposal saying we know nothing about how RDMA is done today...}
 

### 2013-02-07 doc/xps2013/proposal.tex
research items

@@
 
-\comment{mostly a copy and paste of the pomace?}
+Grappa applications are likely to use large amounts of data, 100s of terabytes
+to petabytes. Keeping all these data in DRAM only is unlikely to be cost
+effective. We will explore mechanisms that allow data to reside in solid-state
+storage {\em only}.
+
+Grappa is very effective in hiding latencies of network communication. In
+fact, our measurements and estimates show that we might be able to tolerate
+even more latencies, on the order of milliseconds per request. Therefore, this
+suggests that we might be able to keep data in SSD only and service data
+access requests directly from that, without leaving copies in DRAM.
+
+The natural point of integration for non-volatile memories in Grappa is to
+store the global distributed shared heap in non-volatile memory, since for
+irregular applications the heap often holds data that is infrequently
+modified. All accesses to this shared heap in Grappa programs are done through
+an \textls*{\small API}, so this change could be made transparent to the
+programmer.
+
+If an access is to a chunk of the heap stored in a node's local memory, Grappa
+issues a prefetch for the address and context switches to other work while the
+prefetch is executing~\cite{Nelson:hotpar2011}. If the access is to memory
+stored on another node, Grappa issues a message to that node to request the
+data and context switches to other work until the reply arrives. Non-volatile
+memory would be accessed in the same way: If a request is to memory stored in
+non-volatile memory, Grappa would initiate the request and context switch to
+other work until the request had completed.
+
+This is not straightforward for three reasons: (1) {\bf \textls*{\small
+SSD}\hspace{0.75pt}s assume locality.} In our benchmarks, we find the average
+request size to be less than 64 bytes. Today's \textls*{\small
+SSD}\hspace{0.75pt}s are optimized for reads in the multi-kilobyte range. (2)
+{\bf \textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The
+fastest~\cite{fusionio} can perform on the order of 1 million random reads per
+second, and many are limited to less than 100K/s. Even the individual flash
+chips are slow, on the order of 40K/s~\cite{micronFlash}. In contrast, cores
+may be idle in Grappa if each node cannot perform 100 million random reads per
+second of the global heap. And (3) {\bf \textls*{\small SSD}\hspace{0.75pt}s
+have high overhead.} Mass market \textls*{\small SSD}\hspace{0.75pt}s are
+usually treated as normal disks: accesses must occur through the kernel,
+incurring a high context-switch overhead. Grappa depends on user-mode access
+to the network layer to avoid this overhead; a similar approach has been
+applied to \textls*{\small SSD}\hspace{0.75pt}s in recent
+research~\cite{caulfield:2012}.
+
+We see several ways to address these challenges: \noindent{\bf Scaling up the
+Flash array} Existing Flash-based \textls*{\small SSD}\hspace{0.75pt}s are
+designed for large block requests. An \textls*{\small SSD} for Grappa could be
+designed to support high concurrency, small requests. Where a traditional
+\textls*{\small SSD} might put multiple Flash chips on the same bus, or use
+Flash chips in parallel to build a wider, high-bandwidth bus, a Grappa
+\textls*{\small SSD} would try to maximize parallelism by putting each chip on
+its own bus. Serial Flash might be used to minimize pin count requirements, at
+the cost of some additional latency. This design could scale to request rates,
+but the challenge would be chip count: to support 100 million reads per
+second, an array of 25$\mu$s flash would require 2500 chips. This is possible,
+but expensive.
+
+\noindent{\bf Scaling down the processors.} Alternatively, rather than scaling
+up the flash array, processors could be scaled down to match the flash request
+rate. A low-power processor with few cores might generate only 1 million
+references per second. This could be satisfied with only $25\times 25\mu$s
+Flash chips. There are two problems with this design for our purpose. First,
+while each node requires fewer chips, the overall cluster performance goal,
+and thus chip count requirement, is the same. Second, low single-threaded
+performance is detrimental to performance during periods of diminished
+application concurrency.
+
+\noindent{\bf Buffering requests to increase locality.} Grappa trades latency
+for better throughput by aggregating network communications for better
+utilization and by migrating synchronization operations for reduced
+contention. The same idea could be applied to increase locality and amortize
+the cost of Flash accesses. Requests to nearby flash locations would be
+bucketed by flash page address for a limited time. When a sufficient number of
+requests have arrived for a page or when the requests time out, a single page
+request would be issued to the Flash layer; its result would be used to
+satisfy the individual requests.
 
@@
 
+Language and compiler support are instrumental in making Grappa more usable.
+We envision language and compiler support for three main aspects: (1) specify
+the memory model and semantics of global memory accesses; (2) automatic
+insertion of calls to the runtime, to avoid having the end programmer insert
+explicit calls; and (3) performing code optimizations specific for Grappa as
+well as traditional optimizations in ways that will not break program
+semantics.
+
+Specifying the memory model and access semantics of the Grappa runtime is very
+important for several reasons. The programmer needs to know the behavior of
+accesses to the global heap. The compiler needs to know that too so that code
+optimizations can be performed without breaking program semantics. And the
+model has to be relaxed enough to enable efficient implementations. We opted
+for a sequential consistency for data-race-free programs, similarly to the
+C++0x memory model (something programmers and system software designers are already used to).
+
+We will develop type qualifiers to let programmers declare shared data, e.g.,
+\texttt{shared int}. This will enable the compiler to automatically insert
+runtime calls for accesses to global heap data as well some forms of
+synchronized operations (e.g. atomic fetch-and-add) to global variables. This
+will also enable new analyses and optimizations, for example, the compiler
+could perform spatial and temporal locality analysis to improve request
+aggregation when it can proved that several accesses go to the same
+destination.
+
 \subsection{Graph benchmark suite}
 
+In order to facilitate the development and evaluation of Grappa, we will
+develop a comprehensive benchmark suite for graph analytics applications. This
+will not only be useful for us, but the whole community at large, since there
+is a lack of benchmarks in this application space.
+
+This benchmark suite will consist of application kernels, reference
+implementations, data-sets, data-set generators, documentation, specifications
+for generating standardized performance results, and a hosting service
+(http://graphbench.org) to provide public access to the benchmark and datasets
+as well as aggregate and publish results by all users of the benchmark suite.
+
+This benchmark suite will be a tremendous value to the research community. Not
+only will it enable researchers to evaluate ideas on graph processing system
+techniques, but it will also enable fair comparisons of ideas in a depth not
+currently available from other graph related benchmark suites. The benchmark
+will be released in a completely open fashion, so anyone will be able to use
+it (and contribute to it), there will be no licensing issues even if companies
+(such as Oracle) want to use it. The goal is to get something out relatively
+early in the hope of (i) getting others to cluster around this work rather
+than develop alternatives, and (ii) help establish some of the metrics about
+how to present performance results.
+
+We will build a collection of key computational kernels from graph and
+irregular applications. The benchmark suite will consist of both documentation
+on the actual kernel (ideally a research paper or book reference sourced from
+the community), and a reference implementation. The benchmark suite will also include sample data collected from existing sources, and synthetic
+workload generators. The collected data provides a grounding for performance
+analysis on real-world problems. The workload generators provide cleaner,
+easier to reason about randomly generated data-sets at a programmable scale.
+
 \input{footer}

### 2013-02-07 doc/xps2013/proposal.tex
done for the day

--- 2013-02-05 doc/xps2013/proposal.tex
@@
 
-Irregular applications generate tasks with work, interdependences, or memory accesses that are highly sensitive to input.
+Irregular applications generate tasks with work, interdependences, and memory accesses that are highly sensitive to input.
 Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow.
@@
 
-\noindent{\bf Irregular applications exhibit little spatial locality.} It is not atypical for any given task's data references to be spread randomly across the entire memory of the system.
+\paragraph{Irregular applications exhibit little spatial locality.} It is not atypical for any given task's data references to be spread randomly across the entire memory of the system.
 This makes current memory hierarchy features ineffective.
@@
 
-\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.
+\paragraph{Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.
 Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.
@@
 
-Luckily, many of the important irregular applications (e.g., graph processing, machine learning, ...) naturally offer large amounts of concurrency.
+Luckily, many of the important irregular applications naturally offer large amounts of concurrency.
 This immediately suggests taking advantage of concurrency to tolerate the latency of data movement.
@@
 
-Our preliminary work on Grappa shows that it runs several graph-crunching applications (classic examples of irregular behavior) very efficiently on a commodity cluster.
-For unbalanced tree search, Grappa is over 3X faster and shows greatly improved scalability compared to the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower.
+Our preliminary work on Grappa shows that it runs several graph-crunching applications (classic examples of irregular behavior) efficiently on a commodity cluster.
+For unbalanced tree search, Grappa is over 3X faster and shows greatly improved scalability compared to the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower -- fast compared to implementations without Grappa, but obviously there is room for improvement.
 Most importantly, however, for significantly less real world cost, users can \emph{add} significantly more processors to a commodity cluster than an XMT machine and use Grappa to achieve scalable performance.
@@
 \begin{itemize}
-\item Presently, applications are hand-coded to the Grappa interface.  A significant research thrust of the proposed work is to ease this programming burden by language and compilation support.
 
-\item Our preliminary work has also revealed limitations of the current Grappa runtime system with regards to network scaling.
+\item \textbf{Improved network communication:} Our preliminary work has also revealed limitations of the current Grappa runtime system with regards to network scaling.
 This proposal seeks funding to enhance the network layer and explore the hardware/software interactions of remote direct memory access (RDMA).
 
-\item The size of data being processed by large systems today (hundreds of terabytes) dwarfs what can fit into a single system.
+\item \textbf{Support for \ssd\ only memory systems:} The size of data being processed by large systems today (hundreds of terabytes) dwarfs what can fit into a single system.
 Multiple system nodes are required to maintain and access all of this data efficiently.
@@
 
-\item Presently there does not exist a comprehensive benchmark suite of irregular and graph applications.
+\item \textbf{Language and compiler support:} Presently, applications are hand-coded to the Grappa interface.
+A significant research thrust of the proposed work is to ease this programming burden by language and compilation support.
+
+\item \textbf{Comprehensive benchmark suite:} Presently there does not exist a comprehensive benchmark suite of irregular and graph applications.
 We also seeking funding through this proposal to put together just such a benchmark suite.
@@
 Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
+
+\begin{figure}[t]
+\begin{center}     
+  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
+\begin{minipage}{0.95\columnwidth} 
+  \caption{\label{fig:grappa} Grappa system overview}
+\end{minipage}
+\vspace{-3ex}
+\end{center}
+\end{figure}
 
@@
 Given these restrictions, we can ensure that delegate operations for the same address from multiple requesters are always serialized through a single core in the system, providing atomic semantics without using atomic operations.
-Figure~\ref{fig:delegate-cache} depicts an example of how delegate and cache operations interact.
-
-\begin{figure*}[htb]
-\begin{center}
-  \includegraphics[width=6in]{figs/delegate-cache}
-\begin{minipage}{\columnwidth}
-
-\caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in the global heap.
-It then spawns two tasks on remote cores to increment elements of the array.
-The first task increments two elements of the array using cache operations.
-In step 2, the task is invoked.
-A cache request is issued for two adjacent integers starting at the second element of the array.
-Since these element are stored in the memories of two different cores, this requires the sending of two messages in step 3.
-The task is suspended until both responses arrive in step 4.
-The data carried in these responses is stored in the local buffer.
-The elements are then increment in the buffer.
-Then in step 5, the modified data is sent back to the home node.
-Acknoledgements are returned in step 6 so the task knows when the writes are complete.
-The second task increments an element of the array with a delegate operation.
-In step 7, the task is invoked.
-A delegate request is sent to the home core of the array element with the increment value.
-The task suspends until the response is received.
-In step 8, the increment is executed on the remote core.
-A response is returned in step 9 with the previous value of the array element.
-}
-
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure*}
 
@@
 
+\comment{I really don't know if this is true or if we will be able to to know that is true any time soon -Mark
+
 The final challenge is that atomic operations via RDMA do not interact atomically with processors in a system.
 Semantically it is not entirely clear what happens if a processor issues an atomic operation on a memory address at the same time the network card is attempting to service a remote atomic access.
-\comment{maybe this is not true.
-have to think some more about PCIe an what happens here.
-might not be a problem.
--M}
+}
 
@@
 
-\comment{mostly a copy and paste of the pomace?}
+Grappa applications are likely to use large amounts of data, 100s of terabytes to petabytes.
+Keeping all these data in DRAM only is unlikely to be cost effective.
+We will explore mechanisms that allow data to reside in solid-state storage {\em only}.
+
+Grappa is very effective in hiding latencies of network communication.
+If an access is to a chunk of the heap stored in a node's local memory, Grappa issues a prefetch for the address and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}.
+If the access is to memory stored on another node, Grappa issues a message to that node to request the data and context switches to other work until the reply arrives.
+In fact, our measurements and estimates show that we might be able to tolerate even greater latency -- on the order of milliseconds per request.
+This provides the tantalizing possibility that we will be able to keep data in \ssd\ drives only and service data access requests directly from them, without leaving copies in DRAM.
+The natural point of integration for non-volatile memories in Grappa is to store the global distributed shared heap in non-volatile memory.
+All accesses to this shared heap are done through an \api, so this switch would be transparent to the programmer.
+
+This is not straightforward, however, for three reasons: (1) \ssds\ assume spatial locality.
+In our benchmarks, we find the average request size to be less than 64 bytes.
+Today's \ssds\ are optimized for reads in the multi-kilobyte range.
+(2) \ssds\ have a low request rate.
+The fastest~\cite{fusionio} can perform on the order of 1 million random reads per second, and many are limited to less than 100K/s.
+Even the underlying flash chips are slow, on the order of 40K/s~\cite{micronFlash}.
+In contrast, our measurements suggest cores may be idle in Grappa if each node cannot perform 100 million random reads per second of the global heap.
+And (3) \ssds\ have high overhead.
+Mass market \ssds\ are usually treated as normal disks: accesses must occur through the kernel, incurring a high context-switch overhead.
+Grappa depends on user-mode access to the network layer to avoid this overhead; a similar approach has been applied to \ssds\ in recent research~\cite{caulfield:2012}, however.
+We see several ways to address these challenges:
+
+\paragraph{Scaling up the Flash array:} Existing Flash-based \ssds\ are designed for large block requests.
+An \ssd\ for Grappa could be designed to support high concurrency, small requests.
+Where a traditional \ssd\ might put multiple Flash chips on the same bus, or use Flash chips in parallel to build a wider, high-bandwidth bus, a Grappa \ssd\ would try to maximize message rate by putting each chip on its own bus.
+Serial Flash might be used to minimize pin count requirements, at the cost of some additional latency.
+This design could scale to our necessary request rates, but only at an enormously high chip count: to support 100 million reads per second, an array of 25$\mu$s flash would require 2500 chips.
+
+\paragraph{Scaling down the processors:} Alternatively, rather than scaling up the flash array, processors could be scaled down to match the flash request rate.
+A low-power processor with few cores might generate only 1 million references per second.
+This could be satisfied with only $25\times 25\mu$s Flash chips.
+There are two problems with this design for our purposes, however.
+First, while each node requires fewer chips, the overall cluster performance goal, and thus chip count requirement, is the same.
+More system nodes are thus required to achieve the same level of performance.
+Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
+
+\paragraph{Buffering requests to increase locality:} Grappa trades latency for throughput by aggregating network communication.
+The same idea could be applied to increase locality and amortize the cost of Flash accesses.
+Requests to nearby flash locations would be bucketed by flash page address for a limited time.
+When a sufficient number of requests have arrived for a page or when the requests time out, a single page request would be issued to the Flash layer; its result would be used to satisfy the individual requests.
+The challenge with this approach is as storage size grows, the effectiveness of aggregation correspondingly decreases.
+On the other hand, aggregation occurs for all global memory requests generated by any node in the system, so as aggregate system performance increases, aggregation performance actually \emph{increases} (up to the point that the network is saturated).
+
+\paragraph{Rearchitecting flash:} Clearly the best solution here lays in re-architecting the flash chip itself.
+We will explore new flash microarchitetures that optimize for message rate as well as bandwidth.
+Such chips have value far beyond Grappa.
+Research into persistent key/value stores, and really any persistent high-throughput data-structure will benefit from a redesigned flash chip that can sustain dramatically increased message rates.
 
@@
 
+Language and compiler support are instrumental in making Grappa usable to a wide audience of software developers. We envision language and compiler support can tackle a wide variety of challenges:
+
+\paragraph{Relaxed consistency:} Presently Grappa uses a consistency model that is sequentially consistent for data-race free programs.
+Underlying hooks are available to build even more relaxed consistency models, although we don't expect programmers themselves will want to use them.
+A compiler, however, can see a performance boost by analyzing code and exploiting alias knowledge to program to this underlying Grappa \api, potentially gaining a performance boost by exposing more memory concurrency.
+
+\paragraph{Language support for synchronization operations: } At the moment, runtime library support is provided in Grappa for basic synchronization primitives (lock, barrier, etc).
+Building synchronization directly into the language has the advantage that the compiler can understand and optimize it.
+Several optimizations are possible in the Grappa framework, including exploiting the scheduler and work stealer to elide certain lock operations (scheduling is co-operative on a per-CPU basis), converting synchronization primitives into remote delegate operations, and exploiting the check-in / check-out nature of global memory in Grappa to coalesce multiple synchronization operations together.
+
+\paragraph{Delayed writes: } At the moment when we we write applications for Grappa we use a technique that we call feed-forward delegates.
+Essentially what this technique does is if a loop is reading and writing disjoint sets of data, it is possible and we have found efficient, to dispatch writes asynchronously and ensure those writes are completed only by the end of the loop.
+This increases performance by further overlapping computation with memory access.
+A compiler with alias analysis can do this transformation automatically for the programmer.
+
+\paragraph{Read and write combining: } When multiple remote memory operations are to be dispatched to the same node from the same thread, it is efficient to combine these together into a single request.
+We will develop a compiler pass to combine operations when possible.
+
+\paragraph{Automatic delegate synthesis: } Delegates are a technique in Grappa where one thread can request a small computation to be carried out on a remote node.
+For instance, an increment on a memory location is dispatched to the system node that manages that memory operation and carried out there.
+Currently the decision of what is a delegate is left to the programmer.
+But this is an area rich for compiler optimization.
+We will explore techniques to automatically synthesize delegates from code, potentially leading to more complex delegates and more efficient code.
+
+\paragraph{Compile for a continuations model: } Presently, whenever a task is active it is allocated a complete stack.
+Since several thousand tasks are active per processor at any one time, these stacks occupy a large footprint in memory.
+We wish to explore compiling tasks to a continuation model, whereby when a task dispatches a remote memory operation and is to be contextd out, another task can utilize that stack.
+Combined with the delegate synthesis ideas described above, we will also explore runtime and compiler suppor to convert remote memory operations into thread migrations; instead of moving the data to the system node that has the task, we will move the task to the data in cases where that would be more efficient.
+
+\paragraph{Task-weaving: } Much of the time the same task code is executing for hundreds of thousands of tasks in the Grappa runtime system.
+For example, each task may be processing a loop body for a parallel for loop.
+This is not always the most efficient approach to execution, and we already manually do recursive decomposition of parallel loops with a cut-off to group a collection of loop bodies together into a single task.
+The task then executes each of these loop bodies sequentially.
+This is not the most efficient approach, however.
+An even more efficient technique is for the compiler to unroll and schedule all loop bodies simultaneously within the same task, thereby ``weaving'' these independent tasks together more efficiently to exploit computation / communication overlap.
+
+Presently we are starting work on this language and compiler support.  We are basing our effort on the LLVM~\cite{llvm} infrastructure.
+
+\comment{ three main aspects: (1) easing the interface to the Grappa memory model which is sequentially consistent for data-race free programs; (2) automatic insertion of calls to the runtime to minimize the complexity of accessing globally shared memory; and (3) performing code optimizations specific for Grappa as well as traditional optimizations in ways that will not break program semantics.
+
+We will develop type qualifiers to let programmers declare shared data, e.g., \texttt{shared int}.
+This will enable the compiler to automatically insert runtime calls for accesses to global heap data as well synchronized operations (e.g.
+atomic fetch-and-add) to global variables.
+This will also enable new analyses and optimizations, for example, the compiler could perform spatial and temporal locality analysis to improve request aggregation when it can prove that several accesses go to the same destination.
+
+}
+
+\comment{Luis, don't rant.    -mark
+
+Specifying the memory model and access semantics of the Grappa runtime is very important for several reasons.
+The programmer needs to know the behavior of accesses to the global heap.
+The compiler also must rely on the precise access semantics so that optimizations can be performed without breaking program semantics.
+We opted for a sequential consistency for data-race-free programs, similarly to the C++0x memory model (something programmers and system software designers are already used to).
+}
+
 \subsection{Graph benchmark suite}
 
+In order to facilitate the development and evaluation of Grappa, we will develop a comprehensive benchmark suite for graph analytics applications.
+This will not only be useful for us, but the whole community at large, since there is a lack of benchmarks in this application space.
+
+This benchmark suite will consist of application kernels, reference implementations, data-sets, data-set generators, documentation, specifications for generating standardized performance results, and a hosting service (http://graphbench.org) to provide public access to the benchmark and datasets as well as aggregate and publish results by all users of the benchmark suite.
+
+This benchmark suite will be a tremendous value to the research community.
+Not only will it enable researchers to evaluate ideas on graph processing system techniques, but it will also enable fair comparisons of ideas in a depth not currently available from other graph related benchmark suites.
+The benchmark will be released in a completely open fashion, so anyone will be able to use it (and contribute to it).
+All code will be distributed into the public domain (BSD style licensing) so companies can use it.
+The goal is to get something out relatively early in the hope of (i) getting others to cluster around this work rather than develop alternatives, and (ii) help establish consistent metrics for performance results.
+
+We will build a collection of key computational kernels from graph and irregular applications.
+The benchmark suite will consist of both documentation on the actual kernel (ideally a research paper or book reference sourced from the community), and a reference implementation.
+The benchmark suite will also include sample data collected from existing sources, and synthetic workload generators.
+The collected data provides a grounding for performance analysis on real-world problems.
+The workload generators provide cleaner, easier to reason about randomly generated data-sets at a programmable scale.
+
+\section{Broader impacts}
+
+\paragraph{Increasing involvement of underrepresented groups:} Computer science as a whole has a significant problem attracting women and minorities to the field.
+Computer architecture has an even more extreme problem with diversity (in the PI's view).
+This is a problem the PIs have taken quite seriously since entering the field, and we are quite proud of our PhD students, both female who have gone on to faculty positions in top academic institutions, and male who appreciate the gender and ethnic disparity and who work hard at their own institutions to correct it.
+We continue to recruit a diverse graduate student mix and we continue to include undergraduate students in our research.
+At the moment we are mentoring two women undergraduates involved in the Grappa project, and if this proposal is funded will apply for REU funds to continue their involvement.
+
+\paragraph{Benchmark suite:} As previously described, we are seeking funding support to build and distribute a benchmark suite for irregular and graph applications.
+This benchmark suite has the potential to have a broad impact on academia and industry.
+We will distribute this benchmark through \texttt{graphbench.org} and will also collect and publish results from competing research groups.
+
+\paragraph{Technology transfer:} Bill Dally during his 2010 Eckert-Mauchly award speech said (in effect), technology transfers happen when people transfer.
+Having just come back from a three year stint founding a startup company (where we transferred the Deterministic Multiprocessing technology from UW), we've seen this first hand.
+The students and faculty involved in the Grappa project will spin off a company with their work.
+The exact time frame is not yet known, but everyone involved in the project is excited to see the technology underpin a startup company and make this technology available to the broader commercial world.
+
+\paragraph{Curriculum development activities:} Concurrent with the execution of the research activities described in this grant we will seek to teach an advanced graduate class on advanced concurrency concepts.
+We taught a similar class over six years ago and it is time to revive, update and teach it to the next generation of computer architecture graduate students.
+\comment{Add more here.
+This has no passion.
+-Mark}
+
+\section{Results from Prior NSF Support}
+
+\paragraph{Ceze:} Prof.
+Ceze's research is on computer architecture, programming languages and operating systems.
+His primary focus has been on improving programmability, reliability and energy efficiency of multiprocessor computer systems.
+His research is funded primarily by NSF, PNNL and industry.
+His CAREER award {\em Deterministic Shared Memory Multiprocessing: Vision, Architecture and Impact on Programmability} (CCF-0846004, 2008-13, \$559K) demonstrated that it is possible to provide deterministic execution of arbitrary multithreaded programs without sacrificing performance significantly.
+The results include innovation in architecture, compiler, and operating-systems.
+Two other NSF grants, {\em Code-Centric Approach to Specifying, Checking, and Discovering Shared-Memory Communication} (CCF-1064497, 2011-2014, \$901K) and {\em Precise Concurrency Exceptions: Architecture Support, Semantics and System Implications} (CCF-1016495, 2010-2013, \$500K), are leading to significant results in dealing with concurrency errors and concurrent programming language semantics.
+His work on large-scale graph processing is being funded by the Pacific Northwest Laboratory.
+His work on energy-aware programming models and approximate computing has been primarily funded by a seed NSF grant {\em Disciplined Approximate Programming for Energy-Efficient Computing} (CCF-1216611, 2012-2015, \$300K) and industry, including Microsoft, Google and Qualcomm.
+
+\paragraph{Kahan:} Dr.
+Kahan's research has focused on parallel computation runtimes, for irregular computation and biological applications.
+He is presently an employee of the Department of Energy.
+He has no past or current NSF support.
+
+\paragraph{Oskin:} Prof.
+Oskin's prior research has focused on architectures and systems for scalable computing.
+With support from NSF he has carried out a variety of research activities, leading to over 50 scholarly publications, three PhD students graduated, four masters students graduated, and a company spin-off.
+The past NSF-supported work (WaveScalar) heavily influences the ideas described here.
+In some sense, the work here is a culmination in thought processes about why WaveScalar did not achieve as much ILP as earlier limit studies in ILP suggested were possible.
+NSF support for the PI's work, {\em ITR: WaveScalar, A New Approach to Scalable System Design} (CCF-0325635, 2003-07, \$1.3M), {\em CAREER: Soft-Instruction Set Computing} (CCF-0133188, 2002-05, \$443K), helped to develop WaveScalar.
+He has also spent over a decade exploring architectural implications of quantum computing technologies, with support from NSF {\em NER: Computer Aided Design of Silicon-based Quantum Computers} (CCF-0210373, 2002-04, \$90K), {\em EMT: Microarchitectures for Quantum Computers} (CCF-0523359, 2006-10, \$275K), {\em EMT: Self-Correcting Fault-Tolerant Quantum Computers} (CCF-0523359, 2005-08, \$406K) DARPA {\em QuIST: Architectures and Applications for Scalable Quantum Information Systems}.
+He is currently exploring runtime systems for irregular parallel applications, which is the subject of this proposal and is partially funded by the Department of Energy.
+
+\section{Summary}
+
+We are awesome.  Give us money.
+
 \input{footer}

### 2013-02-10 doc/xps2013/proposal.tex
proposal.tex

@@
 
-Irregular applications generate tasks with work, interdependences, and memory accesses that are highly sensitive to input.
-Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow.
-Important contemporary examples include processing large graphs in the business, national security, machine learning, data-driven science, and social network computing domains.
-For these emerging applications, fast response -- given the sheer amount of data -- requires multinode systems.
-The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband.
-Our goal is to enable scalable performance of irregular applications on these mass market systems.
-Our goal is challenging for two key reasons:
+Irregular applications generate tasks with work, dependences, and memory
+accesses that are highly sensitive to input. Classic examples of irregular
+applications include branch and bound optimization, SPICE circuit simulation,
+contact algorithms in car crash analysis, and network flow. Important
+contemporary examples include processing large graphs in the business,
+national security, machine learning, data-driven science, and social network
+computing domains. For these emerging applications, fast response -- given the
+sheer amount of data -- requires multinode systems. The most broadly available
+multinode systems are those built from x86 commodity computing nodes
+interconnected via ethernet or infiniband. Our goal is to enable scalable
+performance of irregular applications on these mass market systems. Our goal
+is challenging for two key reasons:
 
@@
 
-In this proposal we describe a body of research that extends the Grappa~\cite{grappa} runtime system.
-Grappa is a runtime system that compilers target irregular applications to.
-It provides several facilities: a global shared memory abstraction, user-level threading support, and communication and synchronization primitives.
-Applications targeted for Grappa are not expected to have \emph{any} locality (although what locality that exists is favorably exploited), but are expected to provide many thousands of concurrent threads.
+In this proposal we describe a body of research that extends an early
+prototype of Grappa~\cite{grappa}, a runtime system for irregular
+applications. It provides several facilities: a global shared memory
+abstraction, user-level threading support, and communication and
+synchronization primitives. Applications targeted for Grappa are not expected
+to have \emph{any} locality (although what locality that exists is favorably
+exploited), but are expected to provide many thousands of concurrent threads.
 
@@
 
-Grappa work is still in its infancy, however.  This research proposal seeks funding to extend Grappa in the following directions:
+The work on Grappa is still in its infancy, however. This research proposal
+seeks funding to extend Grappa in the following directions:
 
@@
 
-\item \textbf{Improved network communication:} Our preliminary work has also revealed limitations of the current Grappa runtime system with regards to network scaling.
-This proposal seeks funding to enhance the network layer and explore the hardware/software interactions of remote direct memory access (RDMA).
+\item \textbf{Reinvent network communication:} Our preliminary work has
+revealed limitations of the current Grappa runtime system with regards to
+network scaling. This proposal plans to enhance the network layer with new
+forms of communication scheduling and aggregation, and explore new forms of
+hardware/software interactions of remote direct memory access (RDMA).
 
@@
 However, even with hundreds of channels to persistent storage, loading this data into DRAM can prove a bottleneck.
-The Grappa runtime is designed to make applications tolerate extremely high memory latency, thus an idea we wish to explore is using solid state disks as main memory.
+The Grappa runtime is designed to make applications tolerate extremely high memory latency, thus we will explore using solid state disks as main memory, since and further take advantage of Grappa's latency tolerance properties.
 
@@
 
-\subsection{Communication}\label{sec:communication}
+\subsection{Communication}
+\label{sec:communication}
 
@@
 
-\comment{I think we should pitch what have been done so far as essays and a bit part of the research effort is putting everything together. -Luis}
+As described earlier, our work on Grappa so far has focused on building an initial prototype so we have a baseline system to build upon.
+This early prototype has lead us to think of many research opportunities, from new ways to improve network communication scheduling, to new data storage management ideas to compiler and language support.
+Below we detail our research plans in each of these thrusts.
+
+%I think we should pitch what have been done so far as essays and a %bit part of the research effort is putting everything together. -Luis
 
@@
 As more nodes are added to the system, a quadratic increase in the number of threads (in aggregate across all nodes) is required in order to compensate for the \emph{same} network delay.
-This is due to the fact that the aggregator treats each of these end-points individually.
-This design choice is simple but does not scale.
+This is because the current aggregator treats each of these end-points individually, i.e., it only aggregates messages that go to the same exact end-points.
+This does not scale as the number of nodes go up because the probability of messages going to the same destination goes down (assuming a somewhat uniform communication distribution).
+In summary, this past design choice is simple but does not scale.
 
@@
 
-Hierarchy will reduce scaling requirements from \BigO{n^2} threads to \BigO{n \times log(n)}.
-Of course, it takes what used to be a single network hop for communication and turns it into \BigO{log(n)} hops.
-More threads will be required to overcome this increased latency, but we believe only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling benefit is achievable with the a virtual hierarchical network topology.
+We expect this hierarchical organization will reduce scaling requirements from
+\BigO{n^2} threads to \BigO{n \times log(n)}. Of course, it takes what used to
+be a single network hop for communication and turns it into \BigO{log(n)}
+hops. More threads will be required to overcome this increased latency, but we
+believe only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling
+benefit is achievable with the a virtual hierarchical network topology.
 
-\paragraph{User-level RDMA:} Certain commodity networking hardware today supports remote direct memory access (RDMA).
-Unfortunately, \emph{how} it is implemented is so crude that it is difficult or impossible to efficiently use in practice.
-We start with a short description of the current semantics of RDMA implementations.
-Next we describe the research we wish to undertake to make RDMA more useful.
+% Should we discuss doing HW support for aggregation? Using NetFPGAs?
 
-Networking gear supports remote read and write operations.
+\paragraph{User-level RDMA:} Certain commodity networking hardware today
+supports remote direct memory access (RDMA). Unfortunately, \emph{how} it is
+implemented is so primitive and low-level that it is difficult or impossible
+to efficiently use in practice. We start with a short description of the
+current semantics of RDMA implementations. Next we describe the research we
+wish to undertake to make RDMA more useful.
+
+Networking infrastructure supports remote read and write operations.
 Read operations take a source address and key (on a remote node), a destination address and key (on the local node), and a length.
@@
 The challenge here is to make those calls interoperate securely with a modern operating system, and a user-level runtime system.
-\comment{This seems weak.
+Additionally, the application runtime should be able to control how RDMA operations are carried out and make then semantically richer, e.g., do collective operations (min, max, average) on remote data, or hash-table gets and inserts.
+
+% \comment{This seems weak.
 Need some more ideas here.
@@
 
+%% The second half of this paragraph is really confusing, I have just a minimal clue of what it is talking about. And the little I understand sounds a bit vague and not directly related to grappa.
 The second challenge is the virtual to physical address translation on the cards requires the operating system to pin translations.
@@
 
-\paragraph{Rearchitecting flash:} Clearly the best solution here lays in re-architecting the flash chip itself.
-We will explore new flash microarchitetures that optimize for message rate as well as bandwidth.
+%% Sorry Mark, while the item below is clearly rich and juicy, we can't leave it bhere because it is huuuuge, it does not match the scope of the grant, and we dont have the expertive --- e.g., are you sure nobody is working on it?
+
+% \paragraph{Rearchitecting flash:} Clearly the best solution here lays in re-architecting the flash chip itself.
+We will explore new flash microarchitectures that optimize for message rate as well as bandwidth.
 Such chips have value far beyond Grappa.
@@
 Underlying hooks are available to build even more relaxed consistency models, although we don't expect programmers themselves will want to use them.
-A compiler, however, can see a performance boost by analyzing code and exploiting alias knowledge to program to this underlying Grappa \api, potentially gaining a performance boost by exposing more memory concurrency.
+A compiler, however, can see a performance boost by analyzing code and exploiting alias knowledge to program to this underlying Grappa \api, potentially gaining a performance boost by exposing more memory concurrency and hiding more latency.
 
@@
 
-\paragraph{Compile for a continuations model: } Presently, whenever a task is active it is allocated a complete stack.
+\paragraph{Compile for a continuation model: } Presently, whenever a task is active it is allocated a complete stack.
 Since several thousand tasks are active per processor at any one time, these stacks occupy a large footprint in memory.
-We wish to explore compiling tasks to a continuation model, whereby when a task dispatches a remote memory operation and is to be contextd out, another task can utilize that stack.
-Combined with the delegate synthesis ideas described above, we will also explore runtime and compiler suppor to convert remote memory operations into thread migrations; instead of moving the data to the system node that has the task, we will move the task to the data in cases where that would be more efficient.
+We wish to explore compiling tasks to a continuation model, whereby when a task dispatches a remote memory operation and is to be context-switched out, another task can utilize that stack.
+Combined with the delegate synthesis ideas described above, we will also explore runtime and compiler support to convert remote memory operations into thread migrations; instead of moving the data to the system node that has the task, we will move the task to the data in cases where that would be more efficient.
 
@@
 \comment{Luis, don't rant.    -mark
+Luis: Ehm, gup you!
+It wasnt a rant, we are not even sure what our model is and this will take work...
+and before we know it, we cant cleanly do optimization.
+As you wish, leave it out, the current state is pretty sloppy...
+so if we get a comment on this, you will buy beers.
 

### 2013-02-10 doc/xps2013/proposal.tex
started fleshing out my bio

@@
 \paragraph{Kahan:} Dr.
-Kahan's research has focused on parallel computation runtimes, for irregular computation and biological applications.
+Kahan's research has focused on irregular parallel computation.
+Over the course of twelve years, he developed the Tera MTA's multithreaded runtime system and memory allocator still in use today in the Cray XMT, exploring applicability of the system to production codes and benchmarks in the engineering, science, and government domains.
+His has been driven by application needs.
+The research described herein relies heavily on the experience gained, the innovation being not only to reapply it to mass market technology but also to explore .
 He is presently an employee of the Department of Energy.

### 2013-02-10 doc/xps2013/proposal.tex
changes to intro

@@
 multinode systems are those built from x86 commodity computing nodes
-interconnected via ethernet or infiniband. Our goal is to enable scalable
-performance of irregular applications on these mass market systems. Our goal
-is challenging for two key reasons:
+interconnected via ethernet or infiniband. Our current work enables scalable
+performance of irregular applications on these mass market systems,
+addressing two key challenges:
 
@@
 
-In this proposal we describe a body of research that extends an early
-prototype of Grappa~\cite{grappa}, a runtime system for irregular
+In this proposal we first describe our early
+Grappa~\cite{grappa} prototype, a runtime system for irregular
 applications. It provides several facilities: a global shared memory
@@
 
-Our preliminary work on Grappa shows that it runs several graph-crunching applications (classic examples of irregular behavior) efficiently on a commodity cluster.
+Next, we present some preliminary results.
+Even as an early prototype, Grappa runs several graph-crunching applications (classic examples of irregular behavior) efficiently on a commodity cluster.
 For unbalanced tree search, Grappa is over 3X faster and shows greatly improved scalability compared to the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower -- fast compared to implementations without Grappa, but obviously there is room for improvement.
@@
 
-The work on Grappa is still in its infancy, however. This research proposal
-seeks funding to extend Grappa in the following directions:
+Given these very positive early results,
+we next propose to extend Grappa in the following directions:
 
@@
 
+\end{itemize}
+
+Finally, we propose to explore how the lessons learned performing this research -- specifically knowledge of what ultimately limits Grappa's performance for irregular applications using x86 processors and commodity networks with modified communication protocols -- can be applied to the design of future architectures.
+Doing so will include studying the following research questions:
+
+\being{itemize}
+\item \textbf{TBD}
 \end{itemize}

### 2013-02-12 doc/xps2013/proposal.tex
more work on intro, research and work plan

@@
 
-While some irregular applications can be restructured to better exploit locality, aggregating requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.
+%% I don't think we need the parahraph below. 
+% While some irregular applications can be restructured to better exploit
+% locality, aggregating requests to increase message size, and manage the
+% additional challenges of load balance and synchronization across multinode
+% systems, the work to do so is formidable and requires knowledge and skills
+% pertaining to distributed systems far beyond those of most application
+% programmers.
 
-Luckily, many of the important irregular applications naturally offer large amounts of concurrency.
-This immediately suggests taking advantage of concurrency to tolerate the latency of data movement.
-The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular applications by using concurrency to hide latencies.
-It had a large distributed shared memory with no caches.
-On every clock cycle, each processor would execute a ready instruction chosen from one of its 128 hardware thread contexts, a sufficient number to fully hide memory access latency.
-The network was designed with a single word injection rate that matched the processor clock frequency and sufficient bandwidth to sustain a reference from every processor on every clock cycle.
-Unfortunately, while an excellent match to extremely irregular applications, the MTA was not cost-effective on applications that could exploit locality and had very poor single-thread performance, making it a commercial failure.
-The Cray XMT approximates the Tera MTA-2, substantially reducing its cost but not overcoming its narrow range of applicability.
+Luckily, many of the important irregular applications naturally offer large
+amounts of concurrency. This immediately suggests taking advantage of
+concurrency to tolerate the latency of data movement. The fully custom Tera
+MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular
+applications by using concurrency to hide latencies. It had a large
+distributed shared memory with no caches. On every clock cycle, each processor
+would execute a ready instruction chosen from one of its 128 hardware thread
+contexts, a sufficient number to fully hide memory access latency. The network
+was designed with a single word injection rate that matched the processor
+clock frequency and sufficient bandwidth to sustain a reference from every
+processor on every clock cycle. Unfortunately, while an excellent match to
+extremely irregular applications, the MTA was not cost-effective on
+applications that could exploit locality and had very poor single-thread
+performance, making it a commercial failure. The Cray XMT approximates the
+Tera MTA-2, substantially reducing its cost but not overcoming its narrow
+range of applicability.
 
@@
 
-Next, we present some preliminary results.
-Even as an early prototype, Grappa runs several graph-crunching applications (classic examples of irregular behavior) efficiently on a commodity cluster.
-For unbalanced tree search, Grappa is over 3X faster and shows greatly improved scalability compared to the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower -- fast compared to implementations without Grappa, but obviously there is room for improvement.
-Most importantly, however, for significantly less real world cost, users can \emph{add} significantly more processors to a commodity cluster than an XMT machine and use Grappa to achieve scalable performance.
+Next, we present some preliminary results. Even as an early prototype, Grappa
+runs several graph-crunching applications (classic examples of irregular
+behavior) efficiently on a commodity cluster. For unbalanced tree search,
+Grappa is over 3X faster and shows greatly improved scalability compared to
+the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower --
+fast compared to implementations without Grappa, but obviously there is room
+for improvement. Most importantly, however, for significantly less real world
+cost, users can \emph{add} significantly more processors to a commodity
+cluster than an XMT machine and use Grappa to achieve scalable performance.
 
@@
 
-\item \textbf{Support for \ssd\ only memory systems:} The size of data being processed by large systems today (hundreds of terabytes) dwarfs what can fit into a single system.
-Multiple system nodes are required to maintain and access all of this data efficiently.
-However, even with hundreds of channels to persistent storage, loading this data into DRAM can prove a bottleneck.
-The Grappa runtime is designed to make applications tolerate extremely high memory latency, thus we will explore using solid state disks as main memory, since and further take advantage of Grappa's latency tolerance properties.
+\item \textbf{Language and compiler support:} Presently, applications are
+hand-coded to the Grappa interface. A significant research thrust of the
+proposed work is to ease this programming burden by language and compilation
+support.
 
-\item \textbf{Language and compiler support:} Presently, applications are hand-coded to the Grappa interface.
-A significant research thrust of the proposed work is to ease this programming burden by language and compilation support.
+\item \textbf{Comprehensive benchmark suite:} Presently there does not exist a
+comprehensive benchmark suite of irregular and graph applications. We also
+seeking funding through this proposal to put together just such a benchmark
+suite. This suite will include benchmarks, reference implementations, and
+input data-sets. In addition a website will be created \texttt{graphbench.org}
+to distribute the benchmark suite and publish results.
 
-\item \textbf{Comprehensive benchmark suite:} Presently there does not exist a comprehensive benchmark suite of irregular and graph applications.
-We also seeking funding through this proposal to put together just such a benchmark suite.
-This suite will include benchmarks, reference implementations, and input data-sets.
-In addition a website will be created \texttt{graphbench.org} to distribute the benchmark suite and publish results.
+\item \textbf{Support for \ssd\ only memory systems:} The size of data being
+processed by large systems today (hundreds of terabytes) dwarfs what can fit
+into a single system. Multiple system nodes are required to maintain and
+access all of this data efficiently. However, even with hundreds of channels
+to persistent storage, loading this data into DRAM can prove a bottleneck. The
+Grappa runtime is designed to make applications tolerate extremely high memory
+latency, thus we will explore using solid state disks as main memory, since
+and further take advantage of Grappa's latency tolerance properties.
 
@@
 
-\being{itemize}
-\item \textbf{TBD}
+\begin{itemize}
+\item \textbf What is the right system architecture support for irregular applications, including architecture, communication and storage?
+\item \textbf What are the necessary language and compiler support for large-scale, data-intensive irregular applications?
+\item \textbf How do we comprehensively evaluate performance of graph analytics systems?
 \end{itemize}
@@
 
-Grappa (Figure~\ref{fig:grappa}) has three main software components: \begin{description}
+The current Grappa prototype (Figure~\ref{fig:grappa}) has three main software components: \begin{description}
 
@@
 \comment{Luis, don't rant.    -mark
-Luis: Ehm, gup you!
-It wasnt a rant, we are not even sure what our model is and this will take work...
+Luis: Ehm, it wasnt a rant, we are not even sure what our model is and this will take work...
 and before we know it, we cant cleanly do optimization.
@@
 
+\section{Work Plan}
+\label{sec:plan}
+
+Our research plan can be divided horizontally into three core areas:
+programming languages/compilers; communication and storage; and benchmarking. Below we outline how we
+plan to advance these fronts over the next four years.
+
+\vspace{2ex}
+
+%% TODO: need to expand below.
+
+\noindent\textbf{Year 1:} Benchmarks. Crude language support.
+
+\vspace{1ex}
+\noindent\textbf{Year 2:} Benchmarks. Hierarchical aggregation. Compiler optimizations.
+
+\vspace{1ex}
+\noindent\textbf{Year 3:} Hierarchical aggregation. SSD-only data structures.
+
+\vspace{1ex}
+\noindent\textbf{Year 4:} SSD-only data structures. Detailed system evaluation.
+
 \section{Summary}

### 2013-02-12 doc/xps2013/header.tex
partial commit for Luis the sick boy
done for the day
4l

@@
 
+\newcommand{\BigO}[1]{$O(#1)$}
+
+\newcommand{\api}{\textls*{\small API}}
+\newcommand{\apis}{\textls*{\small API}\hspace{0.75pt}s}
+\newcommand{\ssd}{\textls*{\small SSD}}
+\newcommand{\ssds}{\textls*{\small SSD}\hspace{0.75pt}s}
+
 \setlength{\parskip}{6pt}
@@
 
+\makeatletter
+\renewcommand{\paragraph}{%
+  \@startsection{paragraph}{4}%
+  {\z@}{1.0ex \@plus 1ex \@minus .2ex}{-1em}%
+  {\normalfont\normalsize\bfseries}%
+}
+
+\renewcommand\section{\@startsection{section}{1}{\z@}{-10\p@ \@plus -4\p@ \@minus -4\p@}{5\p@ \@plus 4\p@ \@minus 4\p@}{\normalfont\Large\bfseries\boldmath\rightskip=\z@ \@plus 8em\pretolerance=10000 }}
+
+\renewcommand\subsection{\@startsection{subsection}{2}{\z@}{-10\p@ \@plus -4\p@ \@minus -4\p@}{5\p@ \@plus 4\p@ \@minus 4\p@}{\normalfont\large\bfseries\boldmath\rightskip=\z@ \@plus 8em\pretolerance=10000 }}
+
+\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}{-10\p@ \@plus -4\p@ \@minus -4\p@}{5\p@ \@plus 4\p@ \@minus 4\p@}{\normalfont\normalsize\bfseries\boldmath\rightskip=\z@ \@plus 8em\pretolerance=10000 }}
+
+\makeatother
+
 \begin{document}

### 2013-02-12 doc/xps2013/proposal.tex
4l

@@
 multinode systems are those built from x86 commodity computing nodes
-interconnected via ethernet or infiniband. Our current work enables scalable
+interconnected via ethernet or infiniband. Our focus is to enable scalable
 performance of irregular applications on these mass market systems,
@@
 \begin{itemize}
-\item \textbf What is the right system architecture support for irregular applications, including architecture, communication and storage?
-\item \textbf What are the necessary language and compiler support for large-scale, data-intensive irregular applications?
-\item \textbf How do we comprehensively evaluate performance of graph analytics systems?
+\item \textbf{What is the right system architecture support for irregular applications, including architecture, communication and storage?}
+\item \textbf{What are the necessary language and compiler support for large-scale, data-intensive irregular applications?}
+\item \textbf{How do we comprehensively evaluate performance of graph analytics systems?}
 \end{itemize}
+
+\paragraph{Intellectual Merit:}
+
+Efficiently executing irregular computations on commodity hardware is \emph{hard}.
+Much about commodity hardware is seemingly designed for the opposite application characteristics: caches assume applications have locality; networks are built with low injection rates but high bandwidth, assuming large packet sizes; even the runtimes and operating system is built assuming relatively infrequent communication and synchronization compared to computation.
+This proposal rests on the seemingly outlandish claim that this commodity hardware / software stack can be convinced, through another layer of software, into efficiently executing irregular applications that lack locality, and communicate frequently with small amounts of data.
+A folksy way of saying it is we think we can jam a square peg through a round hole.
+Fortunately we have some preliminary evidence we won't get wedged along the way.
+This is an intellectually challenging endeavor, and one square in the middle of the NSF XPS program: using and exploiting parallelism from irregular computations to efficiently execute them at scale.
+
+\paragraph{Broader Impact:}
+
+Irregular computations are of increasing economic importance.
+Moreover, they are key to several national security applications.
+If commodity hardware can be cajoled into executing such applications efficiently it will open the door for new and rich applications.
+Graph queries, machine-learning and data-driven science will be enabled by the technology we aim to create from the ideas in this grant proposal.
+In addition our work will include the construction of a comprehensive irregular benchmark suite, something to help guide our own and the field's research.
 

### 2013-02-13 doc/xps2013/proposal.tex
Checking in for now.

@@
 \begin{center}
-  \textbf{\Large XPS: Towards an Architecture for Large-Scale, Irregular and Low-Locality Applications}\\ \vspace{5mm}
+  \textbf{\Large XPS-CLCCA: Scalable Parallelism for Irregular and Graph Applications}\\ \vspace{5mm}
   \textbf{Luis Ceze ~~~~~ Simon Kahan ~~~~~ Mark Oskin (PI)}\\
@@
 
-%I think we should pitch what have been done so far as essays and a %bit part of the research effort is putting everything together. -Luis
-
 \subsection{Networking}
 
-\comment{This research item	is important but right now it sounds really really
-specific. You almost need to know the current code to understand it. -Luis}
-
-\comment{Ins't the argument here something like: If we did aggregation on a
-per-destination basis only, as the number of nodes scales, it takes more
-concurrency (or more time) to aggregate messages. We will address this by
-doing aggrregation hierarchically, choosing representative end-points along
-the way to increase agggregation opporunitites at the cost of more
-communucation hops (making an analogy with changing trains here). -Luis}
+Since Grappa is designed for applications that have no or low locality, nearly \emph{every} memory reference is to a remote system node.
+This makes the networking layer key to overall system performance.
+In this section we outline the key directions we wish to take in enhancing the networking layer of the Grappa runtime system.
 
@@
 This is because the current aggregator treats each of these end-points individually, i.e., it only aggregates messages that go to the same exact end-points.
-This does not scale as the number of nodes go up because the probability of messages going to the same destination goes down (assuming a somewhat uniform communication distribution).
+This does not scale as the number of nodes increases because the probability of messages going to the same destination goes down (assuming a somewhat uniform communication distribution).
 In summary, this past design choice is simple but does not scale.
@@
 We intend to continue to use the lock-free single-threaded \gasnet communication layer that we have been using, as we have found the lock-based one to incur too high of an overhead for our purposes.
-Thus the aggregator will need to very carefully utilize lock-free data-structures.
+Thus the aggregator will need to very carefully utilize lock-free data-structures and interact with \gasnet carefully.
 
@@
 hops. More threads will be required to overcome this increased latency, but we
-believe only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling
+hypothesize only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling
 benefit is achievable with the a virtual hierarchical network topology.
@@
 These higher level semantics can, of course, be synthesized from the underlying read and write primitives, but it wastes round trips and synchronization time.
-One possible alternative solution is to borrow ideas from Active Messages~\cite{active-messages} and enable remote nodes to dispatch, essentially, user-level function calls.
-The challenge here is to make those calls interoperate securely with a modern operating system, and a user-level runtime system.
-Additionally, the application runtime should be able to control how RDMA operations are carried out and make then semantically richer, e.g., do collective operations (min, max, average) on remote data, or hash-table gets and inserts.
-
-% \comment{This seems weak.
-Need some more ideas here.
--Mark.
-Luis: Isn't the problem that the app runtime should be able to determine how the RDMA ops should be done?
-Maybe add examples here like accessing a hash-table, doing collective operations on remote data etc (min, max, avg)?
-How about delegating simple operations to the network card?
-Maybe argue that getting the attention of the remote processor for these operations is too costly?}
+One possible alternative solution is to enable remote nodes to dispatch high-level data-structure manipulation requests (e.g.
+insert this item into a queue, remove an item from a set).
+The challenge here is to make those calls programmable and interoperate securely with a modern operating system, and a user-level runtime system.
+Additionally there are questions about \emph{who} should carry out the computation.
+Classic RDMA does not interrupt the host processor, but these richer programmable semantics will require either a processor in the network interface (necessitating issues such as coherence be managed) or complex RDMA tasks need a mechanism to interrupt the host processor \emph{without leaving user space}.
 
@@
 Those in RDMA memory must be pinned.
-In the long run, the solution here likely lays at the intersection of what progress is occurring in virtual memory translation for I/O within virtual machines.
+In the long run, the solution here likely lays at the intersection of what progress is occurring in processor support for virtual machines and virtual memory translation for I/O.
 Our goal will be to carefully study the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.
-
-\comment{I really don't know if this is true or if we will be able to to know that is true any time soon -Mark
-
-The final challenge is that atomic operations via RDMA do not interact atomically with processors in a system.
-Semantically it is not entirely clear what happens if a processor issues an atomic operation on a memory address at the same time the network card is attempting to service a remote atomic access.
-}
 
@@
 
+%% I think this is the only idea that is going to work... maybe we should delete the above three and make this one good.  Seriously, the ideas up there are bunk.  -ark
 % \paragraph{Rearchitecting flash:} Clearly the best solution here lays in re-architecting the flash chip itself.
@@
 
-\comment{ three main aspects: (1) easing the interface to the Grappa memory model which is sequentially consistent for data-race free programs; (2) automatic insertion of calls to the runtime to minimize the complexity of accessing globally shared memory; and (3) performing code optimizations specific for Grappa as well as traditional optimizations in ways that will not break program semantics.
-
-We will develop type qualifiers to let programmers declare shared data, e.g., \texttt{shared int}.
-This will enable the compiler to automatically insert runtime calls for accesses to global heap data as well synchronized operations (e.g.
-atomic fetch-and-add) to global variables.
-This will also enable new analyses and optimizations, for example, the compiler could perform spatial and temporal locality analysis to improve request aggregation when it can prove that several accesses go to the same destination.
-
-}
-
-\comment{Luis, don't rant.    -mark
-Luis: Ehm, it wasnt a rant, we are not even sure what our model is and this will take work...
-and before we know it, we cant cleanly do optimization.
-As you wish, leave it out, the current state is pretty sloppy...
-so if we get a comment on this, you will buy beers.
-
-Specifying the memory model and access semantics of the Grappa runtime is very important for several reasons.
-The programmer needs to know the behavior of accesses to the global heap.
-The compiler also must rely on the precise access semantics so that optimizations can be performed without breaking program semantics.
-We opted for a sequential consistency for data-race-free programs, similarly to the C++0x memory model (something programmers and system software designers are already used to).
-}
-
 \subsection{Graph benchmark suite}
@@
 In order to facilitate the development and evaluation of Grappa, we will develop a comprehensive benchmark suite for graph analytics applications.
-This will not only be useful for us, but the whole community at large, since there is a lack of benchmarks in this application space.
+This will not only be useful for us, but the whole community at large, since there are few benchmarks in this application space.
 
@@
 This benchmark suite will be a tremendous value to the research community.
-Not only will it enable researchers to evaluate ideas on graph processing system techniques, but it will also enable fair comparisons of ideas in a depth not currently available from other graph related benchmark suites.
-The benchmark will be released in a completely open fashion, so anyone will be able to use it (and contribute to it).
+Not only will it enable researchers to evaluate ideas on graph processing system techniques, but it will also enable fair comparisons of ideas in a depth not currently available from other graph related benchmark suites~\cite{graph500}.
+The benchmark suite will be released in a completely open fashion, so anyone will be able to use it (and contribute to it).
 All code will be distributed into the public domain (BSD style licensing) so companies can use it.
-The goal is to get something out relatively early in the hope of (i) getting others to cluster around this work rather than develop alternatives, and (ii) help establish consistent metrics for performance results.
+The goal is to get something out relatively early in the hope of (i) getting others to cluster around this work rather than develop a plethora of overlapping alternatives, and (ii) help establish consistent metrics for performance results.
 
@@
 
-Our research plan can be divided horizontally into three core areas:
-programming languages/compilers; communication and storage; and benchmarking. Below we outline how we
-plan to advance these fronts over the next four years.
+Our research plan can be divided horizontally into three core areas: programming languages/compilers; communication and storage; and benchmarking.
+Below we outline how we plan to advance these fronts over the next four years.
 
@@
 
-\noindent\textbf{Year 1:} Benchmarks. Crude language support.
+\noindent\textbf{Year 1:} The first year of this effort will focus on creating the benchmark suite and initial language and compiler support.
+We aim to pull together 6-8 benchmarks, sample data sets, and scripts to measure performance results on the initial Grappa runtime, UPC~\cite{upc} implementations, and serial reference code.
+For the compiler we will have the front end done enabling easy addition of keywords to annote data-types and structural elements (parallel for, etc).
+The compiler will be generating in-efficient Grappa code by the end of the first year.
 
 \vspace{1ex}
-\noindent\textbf{Year 2:} Benchmarks. Hierarchical aggregation. Compiler optimizations.
+\noindent\textbf{Year 2:} The second year of effort will see the completion of the benchmark suite, a significant effort on compiler optimizations for Grappa, and the beginning of the networking layer research.
+The benchmark suite will include 12-15 applications and/or kernels.
+The website will be live and we will be conducting significant outreach efforts to drive adoption.
+On the compiler front we will have implemented the key optimizations that provide the highest benefit for Grappa code.
+In addition we will have a detailed understanding of what additional work remains to support other languages / language constructs.
+In this year we will start work on rewriting the networking layer of the Grappa runtime, providing a virtual network topology.
+We will also begin work on programmable RDMA by using FPGA-based networking gear.
 
 \vspace{1ex}
-\noindent\textbf{Year 3:} Hierarchical aggregation. SSD-only data structures.
-
-\vspace{1ex}
-\noindent\textbf{Year 4:} SSD-only data structures. Detailed system evaluation.
+\noindent\textbf{Year 3:} In the third and final year of this effort (supported by this grant), we will finish the networking layer rewrite.
+We will also have constructed a test-bed system comprised of mainly SSD-based memory.
+Grappa will be ported to this new style of memory, which we envision will largely be a tuning and not rewriting effort.
+We will explore alternative SSD-memory designs largely through emulation.
+Finally we will complete a detailed system evaluation using our benchmark suite.
 
@@
 
-We are awesome.  Give us money.
+Irregular applications are increasing in importance to the computing industry.
+For a long time they have underpinned computations of interest to the national security arm of the federal government.
+But now, they underpin such questions as ad placement in social networks, and analysis of complex data-sets in medicine and science.
+The defining characteristic of these applications is poor locality and massive available parallelism.
+Our effort is focused on making these applications perform well and scale on commodity hardware.
+The ``secret sauce'' that makes this work is to essentially build a software XMT~\cite{xmt}.
+Instead of relying on locality (which doesn't exist for these applications) we rely on threading to tolerate memory latency.
+Unlike the XMT, however, we must contend with commodity networking hardware, and here we again rely on threading to provide a sufficient number of requests that we can buffer them prior to dispatching on the physical wire.
+
+Our past effort has shown a workable proof of concept of all this runtime smoke and mirrors: a commodity cluster can out due the XMT at its own game, by emulating its key features in software.
+What is needed now is additional effort to improve the underlying performance of this runtime and make it usable to a broad base of developers.
+We will be focusing our efforts in the next three years on the key performance component, the networking layer, and the key programmer efficiency linchpin, high level language support.
+In addition we need a useful mechanism for handling data-sets in the petabyte range, and here we imagine building a cluster of machines that relies primarily on SSD's and not DRAM for data-set storage.
+
+If successful, the Grappa runtime will have a transformative impact on the computing industry.
+Irregular applications are hugely important, and if off the shelf hardware can be efficient with them a new class of questions can be asked of computing systems.
+Additionally our benchmark suite development will further drive innovation among researchers in this space.
 

### 2013-02-13 doc/xps2013/proposal.tex
HW support for aggregation. added sections collaboration plan and data management

@@
 In this proposal we first describe our early
-Grappa~\cite{grappa} prototype, a runtime system for irregular
+Grappa prototype, a runtime system for irregular
 applications. It provides several facilities: a global shared memory
@@
 Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory.
-All global data must be accessed through calls into Grappa's API, shown in Figure~\ref{fig:accessing-memory}.
+All global data must be accessed through calls into Grappa's API.
+%, shown in Figure~\ref{fig:accessing-memory}.
 
@@
 
-% Should we discuss doing HW support for aggregation? Using NetFPGAs?
+\paragraph{Hardware support for aggregation:} If the increased latency caused
+by extra hops in hierarchical aggregation turns out to be a big factor, we
+will explore reducing by move some of the aggregation logic to the network
+card, either by exploiting off-the-shelf programmable network cards or
+experimental systems such ask NetFPGA~\cite{netfpga} (while we are planning to
+focus on Infiniband-based networking, if we explore the NetFPGA route we will
+be using Ethernet).
 
@@
 Having just come back from a three year stint founding a startup company (where we transferred the Deterministic Multiprocessing technology from UW), we've seen this first hand.
-The students and faculty involved in the Grappa project will spin off a company with their work.
+The students and faculty involved in the Grappa project plan to spin off a company with their work.
 The exact time frame is not yet known, but everyone involved in the project is excited to see the technology underpin a startup company and make this technology available to the broader commercial world.
@@
 
-%% TODO: need to expand below.
-
 \noindent\textbf{Year 1:} The first year of this effort will focus on creating the benchmark suite and initial language and compiler support.
 We aim to pull together 6-8 benchmarks, sample data sets, and scripts to measure performance results on the initial Grappa runtime, UPC~\cite{upc} implementations, and serial reference code.
-For the compiler we will have the front end done enabling easy addition of keywords to annote data-types and structural elements (parallel for, etc).
+For the compiler we will have the front end done enabling easy addition of keywords to annotate data-types and structural elements (parallel for, etc).
 The compiler will be generating in-efficient Grappa code by the end of the first year.
@@
 Finally we will complete a detailed system evaluation using our benchmark suite.
+
+\newpage
+
+\section*{Data Management Plan}
+
+The proposed research involves developing experimental systems and
+assessing how well they work. The kinds of data we will produce
+include: characterization data of program behavior; data on system
+performance and reliability; data on how programmers use new language
+constructs and tools. As is our practice, we will regularly submit
+papers for publication detailing our findings and design of our
+systems.
+
+As we have already been doing with all our research, for every
+published paper, we make all the raw data available online. This
+allows other researchers to look at the data in different ways, and it
+enables them to reproduce results. We also often publish
+the experimental software developed, following the rules outlined by
+the University intellectual property office --- which provides ample
+rights for other researchers to build upon our experimental
+systems. We do not anticipate using privileged or confidential
+information.
+
+\newpage
+\section*{Collaboration Plan}
+%\setcounter{page}{1}
+
+Because the PIs already successfully collaborate and co-advise an
+active cadre of spectacular students, the necessary tools and
+coordination mechanisms are already in place for this project to
+proceed successfully.  We first describe the PIs' complementary
+backgrounds, then describe the current organization of our (amorphous)
+research group, then describe the specific technology and facilities
+that enable seamless coordination, and finally describe specific PI roles
+for the research in this proposal.
+
+\subsection*{Complementary PI Experience}
+
+PI Oskin has traditionally participated in the computer-architecture research
+community, co-PI Ceze, in addition to the architecture community also has a
+presence in the systems and programming languages community, and co-PI Kahan
+has participated mainly in the parallel and scientific computing community.
+The move toward multicore architectures has re-invigorated collaboration
+across these divides and recent joint work by Oskin, Ceze, and Kahan
+embodies this important trend. There are currently five graduate students for
+which at last two of the PIs serve as advisors. In addition to the project
+described in this proposal, they are also working together on techniques for
+detecting concurrency errors with machine learning and language
+specifications, using disciplined approximations to save energy in computing,
+and developing new operating-system abstractions for emerging non-volatile
+memory devices. What has made these collaborations so fruitful in such a short
+period of time is the PIs' complementary backgrounds. We now briefly describe
+these backgrounds.
+
+Oskin has... {\bf MARK PUT YOUR STUFF HERE.}
+
+Ceze has been conducting research on architecture support, compiler
+techniques, and programming models to improve the programmability of
+multiprocessor systems for seven years and has co-authored over thirty
+papers in these areas.  His research includes advances in
+architecture~\cite{tlsooo, bulk, bulksc, swbulk, cyclops, dmp,
+  delorean} and programming models~\cite{colorama, ipot, posh} to
+simplify exploiting concurrency in multiprocessors. Some of his recent
+results especially related to this proposal include a programming
+model based on ordered transactions and data annotations~\cite{ipot};
+improving reliability of multithreaded software with architecture
+support for bug detection and avoidance~\cite{atomaid, aatoppicks,
+  bugaboo, oshajava, cs-isca10, ce-isca10}; and deterministic
+execution of multithreaded programs~\cite{dmp, dmptoppicks,
+  asplos10coredet, dmpos}. PI Ceze recently co-organized an NSF-funded
+workshop focused specifically on deterministic multiprocessing. The
+workshop was very successful, bringing leaders in the multicore area
+from industry and academia.
+
+Kahan has tremendous experience in ... from Tera/Cray ... {\bf SIMON PUT YOUR STUFF HERE.}
+
+\subsection*{The SAMPA Research Group: Addressing the Concurrency
+  Challenge Across the Execution Stack}
+
+The PIs organize their joint research activity under the umbrella of
+the ``SAMPA group'' (SAfe MultiProcessing Architectures).  Some brief
+statistics help summarize the group's size, strength, and focus:
+\begin{itemize}
+\item There are roughly ten graduate students, most actively
+  coadvised.
+\item In the last four years, the group has produced twelve papers
+  published in ASPLOS, ISCA, PLDI, OOPSLA, MICRO, and OSDI as well as three
+  papers in the USENIX Workshop on Hot Topics in Parallelism.
+\end{itemize}
+The SAMPA group is an ideal size for collaboration across the system
+execution stack from computer architecture up through software
+engineering.  Compared to large collaborations with several faculty,
+there are no graduate students (or faculty) who specialize entirely in
+particular subtopics.  The students gain expertise across
+computer-science disciplines and use appropriate ideas from hardware,
+software, theory, etc. to solve research problems.
+
+\subsection*{Specific Technology and Facilities}
+
+The research group that will perform the research has the facilities
+needed for successful collaboration.  There is a shared lab facility
+where weekly meetings are held.  Students in the group share computing
+infrastructure, including dedicated machines and regular use of cloud
+services such as Amazon's EC2 for scalability experiments.
+
+The PIs regularly hold joint meetings with each student that they
+co-advise.  They do not split advising duties so much as duplicate
+them.
+
+\subsection*{Specific Roles and Responsibilities}
+
+It should be clear from the preceding discussions that the PIs are
+entirely comfortable jointly managing the work in the proposal and the
+students involved.  Nonetheless, several specific pieces of the work
+more naturally align with the expertise and interests of one of the
+PIs:
+
+\begin{itemize}
+\item PI Oskin will focus on communication issues. 
+\item PI Ceze will focus on compiler and programming languages work.
+\item PI Kahan will focus on organizing benchmarks and performance evaluation efforts.
+\item All PIs expect to solicit guidance and collaboration from
+  researchers such as John Feo (PNNL), Jim Larus (Microsoft Research), Hans
+  Boehm (HP Research), and Preston Briggs,  with whom
+  they have already collaborated on other projects.
+\end{itemize}
 

### 2013-02-13 doc/xps2013/proposal.tex
updated collab section & bib 	modified:   doc/xps2013/proposal.bib 	modified:   doc/xps2013/proposal.tex

@@
 hypothesize only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling
-benefit is achievable with the a virtual hierarchical network topology.
+benefit is achievable with a virtual hierarchical network topology.
 
@@
 
-Kahan has tremendous experience in ... from Tera/Cray ... {\bf SIMON PUT YOUR STUFF HERE.}
+Kahan has been working in the area of parallel computing for irregular
+applications for twenty years, largely in industry.  At Cray (previously Tera
+Computer) he developed the runtime system~\cite{Alverson95schedulingon,Alverson97terahardware-software} and combining memory allocator~\cite{MAMA} still in production use on
+Cray's XMT system in which over 50,000 threads execute concurrently sharing one
+global address space.  He has co-authored patents and publications for both of these
+systems as well as for a scalable connected components graph algorithm that bears
+his name~\cite{Underwood07analyzingthe}.
+Kahan originated and for three years has been co-PI for the Grappa project.
 

### 2013-02-13 doc/xps2013/proposal.tex
fixed bug in bib file & a few bugs in tex file 	modified:   proposal.bib 	modified:   proposal.tex

@@
 embodies this important trend. There are currently five graduate students for
-which at last two of the PIs serve as advisors. In addition to the project
-described in this proposal, they are also working together on techniques for
+which last two of the PIs serve as advisors. In addition to the project
+described in this proposal, they are working together on techniques for
 detecting concurrency errors with machine learning and language
@@
 Our effort is focused on making these applications perform well and scale on commodity hardware.
-The ``secret sauce'' that makes this work is to essentially build a software XMT~\cite{xmt}.
+The ``secret sauce'' that makes this work is to essentially build a software XMT~\cite{feo:xmt}.
 Instead of relying on locality (which doesn't exist for these applications) we rely on threading to tolerate memory latency.

### 2013-02-13 doc/xps2013/proposal.tex
more minor edits

@@
 \item \textbf{Comprehensive benchmark suite:} Presently there does not exist a
-comprehensive benchmark suite of irregular and graph applications. We also
+comprehensive benchmark suite of irregular and graph applications. We are also
 seeking funding through this proposal to put together just such a benchmark
@@
 Grappa runtime is designed to make applications tolerate extremely high memory
-latency, thus we will explore using solid state disks as main memory, since
-and further take advantage of Grappa's latency tolerance properties.
+latency, thus we will explore using solid state disks as main memory, 
+further taking advantage of Grappa's latency tolerance properties.
 
@@
 Efficiently executing irregular computations on commodity hardware is \emph{hard}.
-Much about commodity hardware is seemingly designed for the opposite application characteristics: caches assume applications have locality; networks are built with low injection rates but high bandwidth, assuming large packet sizes; even the runtimes and operating system is built assuming relatively infrequent communication and synchronization compared to computation.
-This proposal rests on the seemingly outlandish claim that this commodity hardware / software stack can be convinced, through another layer of software, into efficiently executing irregular applications that lack locality, and communicate frequently with small amounts of data.
+Much about commodity hardware is seemingly designed for the opposite application characteristics: caches assume applications have locality; networks are built with low injection rates but high bandwidth, assuming large packet sizes; even runtimes and operating systems are built assuming relatively infrequent communication and synchronization compared to computation.
+This proposal rests on the seemingly outlandish claim that this commodity hardware / software stack can be coerced, through another layer of software, into efficiently executing irregular applications that both lack locality and communicate frequently with small amounts of data.
 A folksy way of saying it is we think we can jam a square peg through a round hole.
 Fortunately we have some preliminary evidence we won't get wedged along the way.
-This is an intellectually challenging endeavor, and one square in the middle of the NSF XPS program: using and exploiting parallelism from irregular computations to efficiently execute them at scale.
+This is an intellectually challenging endeavor, landing squarely in the middle of the NSF XPS program: using and exploiting parallelism from irregular computations to efficiently execute them at scale.
 
@@
 Grappa does {\em recursive decomposition} of iterations, similar to Cilk's cilk\_for construct~\cite {cilkforimplementation} and TBB's {\tt parallel\_for}~\cite{intel_tbb}.
-It generates a logarithmically-deep tree of tasks, stopping to execute the loop body when the number of iterations is below the required threshold.
+It generates a logarithmically-deep tree of tasks, stopping to execute the loop body when the number of iterations is below a specified threshold.
 Fourth, a programmer may want to execute an active message; that is, to run a small piece of code on a particular core in the system without waiting for execution resources to be available.
@@
 First, we limit each task to one outstanding delegate operation to avoid the possibility of reordering in the network.
-Second, we limit delegate operations to operate on objects in the 2D address space or objects that fit in a single block of the linear address space so they can be satisfied with a single network request.
+Second, we limit delegate operations to operate on objects in the 2D address space or objects that fit in a single block of the linear address space, so they can be satisfied with a single network request.
 Finally, no context switches are allowed while the data is being modified.
@@
 Grappa's lower networking layer aggregates the upper layer's messages to improve performance.
-Commodity networks including infiniband achieves their peak bisection bandwidth \emph{only} when the packet sizes are relatively large---on the order of multiple kilobytes.
+Commodity networks including infiniband achieve their peak bisection bandwidth \emph{only} when the packet sizes are relatively large---on the order of multiple kilobytes.
 The reason for this discrepancy is the combination of overheads associated with handling each packet (in terms of bytes that form the actual packet, processing time at the card and processing on the CPU within the driver stack).
@@
 The first layer of this hierarchy is to treat all processors at a given system node as a single destination.
-We intend to continue to use the lock-free single-threaded \gasnet communication layer that we have been using, as we have found the lock-based one to incur too high of an overhead for our purposes.
-Thus the aggregator will need to very carefully utilize lock-free data-structures and interact with \gasnet carefully.
+We intend to continue to use the lock-free single-threaded \gasnet~communication layer that we have been using, as we have found the lock-based one to incur too high of an overhead for our purposes.
+Thus the aggregator will need to very carefully utilize lock-free data-structures and interact with \gasnet~carefully.
 
@@
 
-Our past effort has shown a workable proof of concept of all this runtime smoke and mirrors: a commodity cluster can out due the XMT at its own game, by emulating its key features in software.
+Our past effort has shown a workable proof of concept of all this runtime smoke and mirrors: a commodity cluster can out do the XMT at its own game, by emulating its key features in software.
 What is needed now is additional effort to improve the underlying performance of this runtime and make it usable to a broad base of developers.

### 2013-02-13 doc/xps2013/proposal.tex
modified:   proposal.bib 	modified:   proposal.tex

@@
 \paragraph{Hardware support for aggregation:} If the increased latency caused
-by extra hops in hierarchical aggregation turns out to be a big factor, we
-will explore reducing by move some of the aggregation logic to the network
+by extra hops in hierarchical aggregation turns out to be detrimental, we
+will explore mitigation by moving some of the aggregation logic to the network
 card, either by exploiting off-the-shelf programmable network cards or
@@
 The physical address of the destination address is similarly translated using the translation table at the local node.
-Cards also support remote write operations with a similar set of semantics, where addresses are translated on cards using the cards local translation table.
+Cards also support remote write operations with a similar set of semantics, where addresses are translated on a card using the card's local translation table.
 Both read and write operations are asynchronous.
@@
 %% The second half of this paragraph is really confusing, I have just a minimal clue of what it is talking about. And the little I understand sounds a bit vague and not directly related to grappa.
-The second challenge is the virtual to physical address translation on the cards requires the operating system to pin translations.
-This removes paging flexibility from the operating system because runtime systems are (rationally) built to expect data structures to be in memory capable of RDMA or not.
+The second challenge is that the virtual to physical address translation on the cards requires the operating system to pin translations.
+This removes paging flexibility from the operating system, because runtime systems are (rationally) built to expect data structures to be in memory capable of RDMA or not.
 Those in RDMA memory must be pinned.
-In the long run, the solution here likely lays at the intersection of what progress is occurring in processor support for virtual machines and virtual memory translation for I/O.
-Our goal will be to carefully study the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.
+In the long run, the solution here likely lies at the intersection of what progress is occurring in processor support for virtual machines and virtual memory translation for I/O.
+Our goal will be to study carefully the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.
 
@@
 Keeping all these data in DRAM only is unlikely to be cost effective.
-We will explore mechanisms that allow data to reside in solid-state storage {\em only}.
+We will explore mechanisms that use solid-state storage as primary memory, greatly expanding the directly addressible storage at low cost.
 
@@
 \noindent\textbf{Year 1:} The first year of this effort will focus on creating the benchmark suite and initial language and compiler support.
-We aim to pull together 6-8 benchmarks, sample data sets, and scripts to measure performance results on the initial Grappa runtime, UPC~\cite{upc} implementations, and serial reference code.
+We aim to pull together 6-8 benchmarks, sample data sets, and scripts to measure performance results on the initial Grappa runtime, UPC~\cite{upc:2005} implementations, and serial reference code.
 For the compiler we will have the front end done enabling easy addition of keywords to annotate data-types and structural elements (parallel for, etc).

### 2013-02-13 doc/xps2013/proposal.tex
updated nsf support kahan paragraph 	modified:   proposal.tex

@@
 Over the course of twelve years, he developed the Tera MTA's multithreaded runtime system and memory allocator still in use today in the Cray XMT, exploring applicability of the system to production codes and benchmarks in the engineering, science, and government domains.
-His has been driven by application needs.
-The research described herein relies heavily on the experience gained, the innovation being not only to reapply it to mass market technology but also to explore .
-He is presently an employee of the Department of Energy.
+His research has been driven by real-world application needs, resulting in new algorithms and systems to address concurrency bottlenecks.
+He is presently an employee of the Pacific Northwest National Laboratory and an affiliate faculty member of the Computer Science and Engineering Department at the University of Washington.
 He has no past or current NSF support.

### 2013-02-13 doc/xps2013/ceze_bio.tex
my bio

\newcommand \nextthing {\bigskip}

\noindent {\bf Luis Ceze, Ph.D.}\\
Department of Computer Science and Engineering\\
University of Washington\\
Seattle, WA 98195-2350\\
206-543-1896\\
luisceze@cs.washington.edu\\
http://www.cs.washington.edu/homes/luisceze

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\nextthing
\noindent{\bf Education:}

\noindent \begin{tabular}{llll}
University of Illinois at Urbana-Champaign & Computer Science  & Ph.D. & 2007\\
University of S\~{a}o Paulo & Electrical Engineering  & M.Eng. & 2002\\
University of S\~{a}o Paulo & Electrical Engineering  & B.S. & 2000
\end{tabular}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\nextthing
\noindent{\bf Appointments:}

\noindent  \begin{tabular}{lll}
University of Washington &  Associate Professor &  Sep 2012 -- present\\
University of Washington &  Assistant Professor &  Sep 2007 -- Aug 2012\\
University of Illinois at Urbana-Champaign & Research Assistant & 2002 -- 2007 \\
IBM Research, Yorktown & Intern & Summers of 2003, 2004, 2006 \\
IBM Research, Yorktown & Co-op & 2001 -- 2002
\end{tabular}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \nextthing
% \noindent{\bf Relevant Experience:}

% Most of my research has been on improving programmability of
% multiprocessor systems. As part of my thesis, I developed efficient
% and complexity effective support for speculation using signatures
% (ISCA'06). I then built upon this work to develop support for
% high-performance sequential consistency (ISCA'07). Further exploring
% signatures and memory tagging, I worked on data-centric
% synchronization with data coloring (HPCA'07) and on exposing
% signatures and their operations to software to enable better code
% analysis and optimizations (ASPLOS'08). More recently, I have been working on
% architecture support for debugging and concurrency bug avoidance
% (ISCA'08, TopPicks'09, MICRO'10). Finally, I have been working on
% fully deterministic multiprocessor systems (ISCA'08, ASPLOS'09,
% ASPLOS'10, TopPicks'10).

\nextthing
\noindent{\bf Five products most closely related to the proposed project:}
\noindent\begin{enumerate}

\item Nick Hunt, Tom Bergan, Luis Ceze, Steve Gribble, ``DDOS: Taming Nondeterminism in Distributed Systems'', {\em International Conference on
    Architectural Support for Programming Languages and Operating
    Systems ({\bf ASPLOS})}, March 2013.

\item Brandon Lucia, Luis Ceze, Karin Strauss, Shaz Qadeer and
  Hans-J. Boehm ``Conflict Exceptions: Providing Simple Concurrent
  Language Semantics with Precise Hardware Exceptions for Data
  Races'', {\em International Symposium on Computer Architecture ({\bf
      ISCA})}, June 2010.

\item Tom Bergan, Owen Anderson, Joseph Devietti, Luis Ceze, Dan
  Grossman, ``CoreDet: A Compiler and Runtime System for Deterministic
  Multithreaded Execution'', {\em International Conference on
    Architectural Support for Programming Languages and Operating
    Systems ({\bf ASPLOS})}, March 2010.

\item Jose Renau, Karin Strauss, Luis Ceze, Wei Liu, Smruti Sarangi, James Tuck and Josep Torrellas, ``Thread-Level Speculation on a CMP Can Be Energy Efficient'', International Conference on Supercomputing (\textbf{ICS}), June 2005.       Selected to appear
    in {\em IEEE Micro Top Picks.}

\item Calin Cascaval, Jose G. Castanos, Luis Ceze, Monty Denneau, Manish Gupta, Derek Lieber, Jose E. Moreira, Karin Strauss, Henry S. Warren Jr,
``Evaluation of a Multithreaded Architecture for Cellular Computing'',
International Symposium on High-Performance Computer Architecture (\textbf{HPCA}), February 2002.

\end{enumerate}

\noindent{\bf Five other significant products:}

\noindent\begin{enumerate}

\item Hadi Esmaeilzadeh and Adrian Sampson and Luis Ceze and Doug Burger,
``Neural Acceleration for General Purpose Approximate Programs'', {\em
International Symposium on Microarchitecture ({\bf MICRO})}, December 2012.

\item Brandon Lucia, Benjamin Wood, Luis Ceze, ``Isolating and
  Understanding Concurrency Errors Using Reconstructed Execution
  Fragments'', {\em Conference on Programming Language Design and
    Implementation ({\bf PLDI})}, June 2011.

\item Adrian Sampson, Werner Dietl, Emily Fortuna, Dan Gnanapragasam, Luis
Ceze and Dan Grossman, ``EnerJ: Approximate Data Types for Safe and General
Low-Power Computation'', {\em Conference on Programming Language Design and
Implementation ({\bf PLDI})}, June 2011.

\item Tom Bergan, Nick Hunt, Luis Ceze, Steve Gribble, ``Deterministic Process
Groups in dOS'', {\em Symposium on Operating Systems Design and Implementation
({\bf OSDI})}, October 2010.

\item Joseph Devietti, Brandon Lucia, Luis Ceze, Mark Oskin, ``DMP:
Deterministic Shared Memory Multiprocessing'', {\em International Conference
on Architectural Support for Programming Languages and Operating Systems ({\bf
ASPLOS})}, March 2009. Selected to appear in {\em IEEE Micro Top Picks.}

\end{enumerate}

\nextthing
\noindent{\bf Synergistic activities:}

\noindent\begin{itemize}
\item Co-Organizer, Workshop on Deterministic Multiprocessing and Parallel Programming (funded by NSF), 2009 and 2011.

\item PC Co-chair,  USENIX Workshop on Hot Topic in Parallelism ({\bf HotPar}), 2012.

\item PC Member, IEEE Micro Top Picks, 2010, 2013, Annual Symposium on
Principles and Practice of Parallel Programming ({\bf PPoPP}) 2012,
International Conference on Architectural Support for Programming Languages
and Operating Systems ({\bf ASPLOS}) 2011, 2010, 2009, International
Conference on Microarchitecture ({\bf MICRO}) 2011, 2010, International
Symposium on Computer Architecture ({\bf ISCA}), 2008.

\item Released publicly available software: {\bf SESC}, a free cycle-accurate multiprocessor simulator (sesc.sourceforge.net); and {\bf CoreDet}, a compiler and runtime system for deterministic execution (sampa.cs.washington.edu); {\bf Recon}, a system for finding and understanding concurrency bugs.

\item ACM Best Teacher Award, UW CSE, University of Washington, 2010

\end{itemize}

\nextthing
\noindent{\bf Collaborators in the last 48 months:}

\noindent\begin{tabular}{ll}
  Karin Strauss, Shaz Qadeer, Doug Burger & Microsoft Research \\
  Hans-J Boehm  & HP Labs \\
  Dan Grossman, Susan Eggers, Mark Oskin, Steve Gribble, Henry M.  Levy  & University of Washington \\
  C\u{a}lin Ca\c{s}caval & Qualcomm Research \\
  James Tuck & NC State University \\
  Josep Torrellas & UIUC 
\end{tabular}

\nextthing
\noindent{\bf Graduate advisor:} Josep Torrellas, UIUC

\nextthing
\noindent{\bf Graduate advisees:}

\noindent\begin{tabular}{l}
  Current: Tom Bergan, Joseph Devietti, Brandon Lucia, Adrian Sampson \\ 
  \hspace{1cm} Jacob Nelson, Nick Hunt, Hadi Esmaeilzadeh, Ben Wood, Katelin Bailey\\
  Past: Angda Chen (now at Amazon), Cherie Cheung, Nick Murphy (Harvard) \\
  \hspace{1cm} Owen Anderson (Apple), Emily Fortuna (Google) \\
\end{tabular}

### 2013-02-13 doc/xps2013/proposal.tex

@@
 
+\paragraph{Concurrency Control: } Having too few tasks multiplexed per core fails to tolerate fully the latency; having too many increases memory and cache footprints, resulting in excess cache misses or exhausting physical memory unnecessarily.
+Currently, we choose a number at program instantiation:  experiments so far have shown evidence of a broad sweet-spot, though it is of course varies as a function of application behavior.
+At compile-time, we could estimate the number of tasks required based on the density of remote operations and hard-coded measurements of system characteristics; at runtime we can adjust the number of tasks dynamically based on real-time measurement of processor counters.
+Exploring these directions may lead to even more robust performance.
+
 Presently we are starting work on this language and compiler support.  We are basing our effort on the LLVM~\cite{llvm} infrastructure.

### 2013-02-13 doc/xps2013/proposal.tex
Incorporated Holt's suggestions. 	modified:   proposal.tex

@@
 
-\item \textbf{Language and compiler support:} Presently, applications are
-hand-coded to the Grappa interface. A significant research thrust of the
-proposed work is to ease this programming burden by language and compilation
-support.
+\item \textbf{Language and compiler support:} Presently, Grappa is implemented
+strictly as a library.  A significant research thrust of the proposed work is to use
+language and compilation support to provide a cleaner programming model and use
+statically-available knowledge to lessen the runtime overhead automatically.
 
 \item \textbf{Comprehensive benchmark suite:} Presently there does not exist a
-comprehensive benchmark suite of irregular and graph applications. We are also
-seeking funding through this proposal to put together just such a benchmark
-suite. This suite will include benchmarks, reference implementations, and
-input data-sets. In addition a website will be created \texttt{graphbench.org}
-to distribute the benchmark suite and publish results.
+comprehensive benchmark suite of irregular and graph applications. We
+are also seeking funding through this proposal to put together just
+such a benchmark suite. This suite will include benchmarks, reference
+implementations, and input data-sets. In addition a website will be
+created to distribute freely the benchmark suite and publish results.
 
@@
 
-Grappa provides four methods for expressing parallelism.
+Grappa provides two methods for expressing parallelism.
 First, when the programmer identifies work that can be done in parallel, the work may be wrapped up in a function and queued with its arguments for later execution using a \texttt{spawn}.
-Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific core in the system or at the home core of a particular memory location.
-Third, the programmer can invoke a parallel for loop, provided that the trip count is known at loop entry.
+A programmer can further specify using \texttt{spawn\_on} to spawn a task on a specific core in the system or at the home core of a particular memory location.
+Second, the programmer can invoke a parallel for loop, provided that the trip count is known at loop entry.
 The programmer specifies a function pointer along with start and end indices and an optional threshold to control parallel overhead.
@@
 It generates a logarithmically-deep tree of tasks, stopping to execute the loop body when the number of iterations is below a specified threshold.
-Fourth, a programmer may want to execute an active message; that is, to run a small piece of code on a particular core in the system without waiting for execution resources to be available.
-For example, custom synchronization primitives execute this way, as a function executed on the core where the data lives.
 
@@
 
-\paragraph{Read and write combining: } When multiple remote memory operations are to be dispatched to the same node from the same thread, it is efficient to combine these together into a single request.
-We will develop a compiler pass to combine operations when possible.
-
 \paragraph{Automatic delegate synthesis: } Delegates are a technique in Grappa where one thread can request a small computation to be carried out on a remote node.
@@
 We will explore techniques to automatically synthesize delegates from code, potentially leading to more complex delegates and more efficient code.
+
+\paragraph{Read and write combining: } As a special case of automatic delegate synthesis, when multiple remote memory operations are to be dispatched to the same node from the same thread, it may be more efficient to combine these together into a single request.
+We will develop a compiler pass to combine operations when possible.
 
@@
 For example, each task may be processing a loop body for a parallel for loop.
-This is not always the most efficient approach to execution, and we already manually do recursive decomposition of parallel loops with a cut-off to group a collection of loop bodies together into a single task.
-The task then executes each of these loop bodies sequentially.
-This is not the most efficient approach, however.
-An even more efficient technique is for the compiler to unroll and schedule all loop bodies simultaneously within the same task, thereby ``weaving'' these independent tasks together more efficiently to exploit computation / communication overlap.
+However, tasks introduce overhead, so we already manually impose a cutoff on recursive decomposition of parallel loops.
+This way, multiple loop bodies are collected into a single task and executed serially.
+However, this limits the amount of parallelism available for latency tolerance.
+A mmore efficient technique is to allow the compiler to unroll and schedule all loop bodies simultaneously within the same task, including overlapping remote memory operations, thereby ``weaving'' these independent tasks together to overlap computation with communication with minimal task overhead.
+Further, a set of ``woven tasks'' may be amenable to vector execution, allowing even more parallelism to be exploited.
 
@@
 Finally we will complete a detailed system evaluation using our benchmark suite.
+
+\section{Summary}
+
+Irregular applications are increasing in importance to the computing industry.
+For a long time they have underpinned computations of interest to the national security arm of the federal government.
+But now, they underpin such questions as ad placement in social networks, and analysis of complex data-sets in medicine and science.
+The defining characteristic of these applications is poor locality and massive available parallelism.
+Our effort is focused on making these applications perform well and scale on commodity hardware.
+The ``secret sauce'' that makes this work is to essentially build a software XMT~\cite{feo:xmt}.
+Instead of relying on locality (which doesn't exist for these applications) we rely on threading to tolerate memory latency.
+Unlike the XMT, however, we must contend with commodity networking hardware, and here we again rely on threading to provide a sufficient number of requests that we can buffer them prior to dispatching on the physical wire.
+
+Our past effort has shown a workable proof of concept of all this runtime smoke and mirrors: a commodity cluster can out do the XMT at its own game, by emulating its key features in software.
+What is needed now is additional effort to improve the underlying performance of this runtime and make it usable to a broad base of developers.
+We will be focusing our efforts in the next three years on the key performance component, the networking layer, and the key programmer efficiency linchpin, high level language support.
+In addition we need a useful mechanism for handling data-sets in the petabyte range, and here we imagine building a cluster of machines that relies primarily on SSD's and not DRAM for data-set storage.
+
+If successful, the Grappa runtime will have a transformative impact on the computing industry.
+Irregular applications are hugely important, and if off the shelf hardware can be efficient with them a new class of questions can be asked of computing systems.
+Additionally our benchmark suite development will further drive innovation among researchers in this space.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%
+%%%%% Extra stuff: collaboration plan and data-management plan
+%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Summary}
-
-Irregular applications are increasing in importance to the computing industry.
-For a long time they have underpinned computations of interest to the national security arm of the federal government.
-But now, they underpin such questions as ad placement in social networks, and analysis of complex data-sets in medicine and science.
-The defining characteristic of these applications is poor locality and massive available parallelism.
-Our effort is focused on making these applications perform well and scale on commodity hardware.
-The ``secret sauce'' that makes this work is to essentially build a software XMT~\cite{feo:xmt}.
-Instead of relying on locality (which doesn't exist for these applications) we rely on threading to tolerate memory latency.
-Unlike the XMT, however, we must contend with commodity networking hardware, and here we again rely on threading to provide a sufficient number of requests that we can buffer them prior to dispatching on the physical wire.
-
-Our past effort has shown a workable proof of concept of all this runtime smoke and mirrors: a commodity cluster can out do the XMT at its own game, by emulating its key features in software.
-What is needed now is additional effort to improve the underlying performance of this runtime and make it usable to a broad base of developers.
-We will be focusing our efforts in the next three years on the key performance component, the networking layer, and the key programmer efficiency linchpin, high level language support.
-In addition we need a useful mechanism for handling data-sets in the petabyte range, and here we imagine building a cluster of machines that relies primarily on SSD's and not DRAM for data-set storage.
-
-If successful, the Grappa runtime will have a transformative impact on the computing industry.
-Irregular applications are hugely important, and if off the shelf hardware can be efficient with them a new class of questions can be asked of computing systems.
-Additionally our benchmark suite development will further drive innovation among researchers in this space.
-
 \input{footer}

### 2013-02-14 doc/xps2013/proposal.tex
pass on sections 1 and 2. needs some proofreading

@@
 fast compared to implementations without Grappa, but obviously there is room
-for improvement. Most importantly, however, for significantly less real world
-cost, users can \emph{add} significantly more processors to a commodity
-cluster than an XMT machine and use Grappa to achieve scalable performance.
+for improvement.
 
-Given these very positive early results,
-we next propose to extend Grappa in the following directions:
+% Most importantly, however, for a much lower real world cost,
+% users can \emph{add} significantly more processors to a commodity cluster than
+% an XMT machine and use Grappa to achieve scalable performance.
+
+Given these very positive early results, this proposal plans to extend Grappa
+in the following directions:
 
@@
 As we will show later, Grappa can tolerate latencies way beyond that of the network.
-Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
+Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our effective random access memory bandwidth, our synchronization bandwidth, and our ability to improve load imbalance.
 
@@
 The compiler emits an access and the memory is manipulated directly.
-Applications use local accesses for a number of things in Grappa: the stack associated with a task, accesses to localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
+Applications use local accesses for the stack associated with a task, accesses to localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
 Local pointers cannot access memory on other cores, and are valid only on their home core.
 
-Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory.
+Large data that is expected to be shared and accessed with low locality is stored in global memory.
 All global data must be accessed through calls into Grappa's API.
@@
 Any node-local data can be made accessible by other nodes in the system by wrapping the address and node ID into a 2D global address.
-This address can then be accessed with a delegate and can also be cached by other nodes.
+This address can then be accessed with a delegate.
+% and can also be cached by other nodes. 
 At the destination the address is converted into a canonical x86 address by replacing the upper bits with the sign-extended upper bit of the virtual address.
@@
 \paragraph{Global memory access} There are two general approaches Grappa applications use to {\emph access} global memory.
-When the programmer expects a computation on shared data to have spatial locality to exploit, {\em cache} operations may be used.
-When there is no locality to exploit, {\em delegate} operations are used.
+When the programmer expects a computation on shared data to have spatial locality to exploit, explicit {\em cache} operations may be used.
+When there is no locality to exploit, remote {\em delegate} operations are used.
 
@@
 For example, the UPC compiler coalesces struct and array accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's message vectorization hoists small messages out of a loop \cite{FortranD:1992}.
-Caching in Grappa additionally provides a mechanism for exploiting temporal locality by operating on the data locally.
 
-Under the hood, Grappa performs the mechanics of gathering chunks of data from multiple system nodes and presenting a conventional appearing linear block of memory as a local pointer into a cache.
-The strategy employed is to issue all the constituent requests of a cache access request  (as Active Messages) and then yield until all responses have occurred.
+Caching in Grappa additionally provides a mechanism for exploiting temporal locality by operating on the data locally --- even though we aim Grappa at applications with low locality, we want to be able to exploit it when available.
+% Under the hood, Grappa performs the mechanics of gathering chunks of data
+% from multiple system nodes and presenting a conventional appearing linear
+% block of memory as a local pointer into a cache. The strategy employed is to
+% issue all the constituent requests of a cache access request	(as Active
+% Messages) and then yield until all responses have occurred.	
 Currently, Grappa caches are \emph{not} coherent, requiring the programmer to maintain consistent access to data.
@@
 
-Delegate operations are always executed at the home core of their address, and while arbitrary memory operations can be delegated, we restrict the use of delegate operations in three ways to make them more useful for synchronization.
-First, we limit each task to one outstanding delegate operation to avoid the possibility of reordering in the network.
-Second, we limit delegate operations to operate on objects in the 2D address space or objects that fit in a single block of the linear address space, so they can be satisfied with a single network request.
-Finally, no context switches are allowed while the data is being modified.
-Given these restrictions, we can ensure that delegate operations for the same address from multiple requesters are always serialized through a single core in the system, providing atomic semantics without using atomic operations.
+% Delegate operations are always executed at the home core of their address, and
+% while arbitrary memory operations can be delegated, we restrict the use of
+% delegate operations in three ways to make them more useful for
+% synchronization. First, we limit each task to one outstanding delegate
+% operation to avoid the possibility of reordering in the network. Second, we
+% limit delegate operations to operate on objects in the 2D address space or
+% objects that fit in a single block of the linear address space, so they can be
+% satisfied with a single network request. Finally, no context switches are
+% allowed while the data is being modified. Given these restrictions, we can
+% ensure that delegate operations for the same address from multiple requesters
+% are always serialized through a single core in the system, providing atomic
+% semantics without using atomic operations.
 
@@
 
-Grappa's lower networking layer aggregates the upper layer's messages to improve performance.
-Commodity networks including infiniband achieve their peak bisection bandwidth \emph{only} when the packet sizes are relatively large---on the order of multiple kilobytes.
-The reason for this discrepancy is the combination of overheads associated with handling each packet (in terms of bytes that form the actual packet, processing time at the card and processing on the CPU within the driver stack).
-Our measurements confirm manufacturers published data~\cite{infinibandbandwidth}, that with this packet size the bisection bandwidth is only a small fraction, less than 3\% of the peak bisection bandwidth.
+Grappa's lower networking layer aggregates the upper layer's messages to
+improve performance. Commodity networks including
+Infiniband~\cite{infinibandbandwidth} achieve their peak bisection bandwidth
+\emph{only} when the packet sizes are relatively large---on the order of
+multiple kilobytes. The reason for this discrepancy is the combination of
+overheads associated with handling each packet (in terms of bytes that form
+the actual packet, processing time at the card and processing on the CPU
+within the driver stack). Our measurements confirm manufacturers published
+data~\cite{infinibandbandwidth}, that with this packet size the bisection
+bandwidth is only a small fraction, less than 3\% of the peak bisection
+bandwidth.
 
-In our experiments the vast majority of requests were smaller than 44 bytes, far too small to make efficient use of the network.
-To make the best use of the network, we must convert our small messages into large ones.
-When a task sends a message, it is not immediately sent, but rather placed in a queue specific to the destination.
+In our early experiments the vast majority of requests were smaller than 44
+bytes, far too small to make efficient use of the network. To make the best
+use of the network, we must convert our small messages into large ones. When a
+task sends a message, it is not immediately sent, but rather placed in a queue
+specific to the destination. If the size in bytes of a queue is above a
+threshold (e.g., 4096, typically large enough for good bandwidth), the
+contents of the queue are sent immediately. Also, each queue has a wait time
+threshold ($\approx${1ms}). If the oldest message in a queue has been waiting
+longer than this threshold, the contents of the queue are sent immediately,
+even if the queue size is lower than the message size threshold. Finally,
+queues may be explicitly flushed, for situations when latency is more
+important than bandwidth.
 
-There are three situations in which a queue of aggregated messages is sent.
-First, each queue has a message size threshold of 4096 bytes, chosen to give reasonable network performance.
-If the size in bytes of a queue is above the threshold, the contents of the queue are sent immediately.
-Second, each queue has a wait time threshold ($\approx${1ms}).
-If the oldest message in a queue has been waiting longer than this threshold, the contents of the queue are sent immediately, even if the queue size is lower than the message size threshold.
-Third, queues may be explicitly flushed in situations where the programmer wants to minimize the latency of a message at the cost of bandwidth utilization.
-
-The network layer is serviced by polling.
-Periodically when a context switch occurs, the Grappa scheduler switches to the network polling thread.
-This thread has three responsibilities.
-First, it polls the lower-level network layer to ensure it makes progress.
-Second, it deaggregates received messages and executes active message handlers.
-Third, it checks to see if any aggregation queues have messages that have been waiting longer than the threshold; if so, it sends them.
+% The network layer is serviced by polling. Periodically when a context switch
+% occurs, the Grappa scheduler switches to the network polling thread. This
+% thread has three responsibilities. First, it polls the lower-level network
+% layer to ensure it makes progress. Second, it deaggregates received messages
+% and executes active message handlers. Third, it checks to see if any
+% aggregation queues have messages that have been waiting longer than the
+% threshold; if so, it sends them.
 
@@
 
-To evaluate Grappa's performance with respect to the XMT, we ran each of our three benchmarks on up to 16 nodes of each machine.
-Grappa used 6 cores per node, with the best parameters chosen for each point.
-In some cases, the XMT could not run the benchmark with 2 nodes, so the point is omitted.
+To evaluate the performance of our current prototype with respect to the XMT,
+we ran each of our three benchmarks on up to 16 nodes of each machine. Grappa
+used 6 cores per node, with the best parameters chosen for each point. In some
+cases, the XMT could not run the benchmark with 2 nodes, so the point is
+omitted.
 
@@
 
-\hspace{-1in}\begin{minipage}{0.3\textwidth}
+\hspace{-1in}\begin{minipage}{0.2\textwidth}
 \begin{center}
-\includegraphics[width=3in]{figs/uts_performance.pdf}
+\includegraphics[width=2.75in]{figs/uts_performance.pdf}
 \caption{\label{fig:uts_compare} Performance of in-memory unbalanced tree search.}
@@
 \begin{center}
-\includegraphics[width=3in]{figs/bfs_performance}
+\includegraphics[width=2.75in]{figs/bfs_performance}
 \caption{\label{fig:bfs-performance} Performance of breadth-first searcch.}

### 2013-02-14 doc/xps2013/proposal.tex
fixed more stuff...
uupdate

@@
 Injection rate into the network is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.
-
-%% I don't think we need the parahraph below. 
-% While some irregular applications can be restructured to better exploit
-% locality, aggregating requests to increase message size, and manage the
-% additional challenges of load balance and synchronization across multinode
-% systems, the work to do so is formidable and requires knowledge and skills
-% pertaining to distributed systems far beyond those of most application
-% programmers.
 
@@
 Fortunately we have some preliminary evidence we won't get wedged along the way.
-This is an intellectually challenging endeavor, landing squarely in the middle of the NSF XPS program: using and exploiting parallelism from irregular computations to efficiently execute them at scale.
+This is an intellectually challenging endeavor, that we believe is well suited for the NSF XPS program.
+In summary, our goal is to exploit parallelism from irregular computations to efficiently execute them at scale.
 
@@
 
-Underneath the aggregation layer, Grappa uses the \gasnet~communication library~\cite{gasnet} to actually move data.
-All interprocess communication, whether on or off a cluster node, is handled by the \gasnet~library.
-\gasnet~is able to take advantage of many communication mechanisms, including ethernet and infiniband between nodes, as well as shared memory within a node.
+Underneath the aggregation layer, Grappa uses the \gasnet~communication
+library~\cite{gasnet} to actually move data. All interprocess communication,
+whether on or off a cluster node, is handled by the \gasnet~library.
+\gasnet~is able to take advantage of many communication mechanisms, including
+ethernet and infiniband between nodes, as well as shared memory within a node.
 
-Some networks provide access to a remote machine's memory directly.
-This would seem to be a good fit for a programming model focused on global shared memory, but in fact we do not use it.
-In our experiments, we found that RDMA operations are subject to the same message rate limitations as all other messages on these cards, and thus using raw RDMA operations for our small messages would make inefficient use of bandwidth.
-Instead, we implement remote memory operations with active messages.
-A byproduct of this design decision is that Grappa is not limited to RDMA-capable networks.
+Some networks provide access to a remote machine's memory directly. This would
+seem to be a good fit for a programming model focused on global shared memory,
+but in fact we do not use it. In our experiments, we found that RDMA
+operations are subject to the same message rate limitations as all other
+messages on these cards, and thus using raw RDMA operations for our small
+messages would make inefficient use of bandwidth. Instead, we implement remote
+memory operations with active messages. A byproduct of this design decision is
+that Grappa is not limited to RDMA-capable networks.
 
@@
 
-\hspace{-1in}\begin{minipage}{0.2\textwidth}
+\hspace{-1in}\begin{minipage}{0.3\textwidth}
 \begin{center}
-\includegraphics[width=2.75in]{figs/uts_performance.pdf}
+\includegraphics[width=2.5in]{figs/uts_performance.pdf}
 \caption{\label{fig:uts_compare} Performance of in-memory unbalanced tree search.}
@@
 \begin{center}
-\includegraphics[width=2.75in]{figs/bfs_performance}
-\caption{\label{fig:bfs-performance} Performance of breadth-first searcch.}
+\includegraphics[width=2.5in]{figs/bfs_performance}
+\caption{\label{fig:bfs-performance} Performance of breadth-first search.}
 \end{center}
@@
 
-\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex tree (T1L).
-Figure~\ref{fig:uts_compare} shows the performance in terms of number of vertices visited per second versus number of compute nodes.
-Grappa is 3.2 times faster than the XMT at 16 nodes.
-As we will show later, the performance advantage Grappa has over XMT increases as more nodes are added.
-The main reason Grappa performs better is the software-based delegate synchronization obviates the need for the retry-based synchronization that XMT uses.
+\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex
+tree (T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
+number of vertices visited per second versus number of compute nodes. Grappa
+is 3.2 times faster than the XMT at 16 nodes. As we will show later, the
+performance advantage Grappa has over XMT increases as more nodes are added.
+The main reason Grappa performs better is the software-based delegate
+synchronization obviates the need for the retry-based synchronization that XMT
+uses.
 

### 2013-02-14 doc/xps2013/proposal.tex
commit to merge

--- 2013-02-14 doc/xps2013/proposal.tex
@@
 
-Language and compiler support are instrumental in making Grappa usable to a wide audience of software developers. We envision language and compiler support can tackle a wide variety of challenges:
+Language and compiler support are instrumental in making Grappa usable to a
+wide audience of application developers. We plan to use the LLVM~\cite{llvm}
+infrastructure. Below we describe the specific aspects we plan to tackle.
 
@@
 
-\paragraph{Concurrency Control: } Having too few tasks multiplexed per core fails to tolerate fully the latency; having too many increases memory and cache footprints, resulting in excess cache misses or exhausting physical memory unnecessarily.
-Currently, we choose a number at program instantiation:  experiments so far have shown evidence of a broad sweet-spot, though it is of course varies as a function of application behavior.
-At compile-time, we could estimate the number of tasks required based on the density of remote operations and hard-coded measurements of system characteristics; at runtime we can adjust the number of tasks dynamically based on real-time measurement of processor counters.
-Exploring these directions may lead to even more robust performance.
+\paragraph{Task load control:} Having too few tasks per core fails to tolerate
+fully the latency; having too many increases overheads unnecessarily.
+Currently, we choose a number at program instantiation: experiments so far
+have shown evidence of a broad sweet-spot, though it is of course varies as a
+function of application behavior and input data. At compile-time, we might be
+able to estimate the number of tasks required based on the density of remote
+operations and system parameters (e.g., memory size, network latency, etc.);
+at runtime we can adjust the number of tasks dynamically based on periodic
+performance assessments.
 
-Presently we are starting work on this language and compiler support.  We are basing our effort on the LLVM~\cite{llvm} infrastructure.
+% Presently we are starting work on this language and compiler support.  We are basing our effort on the LLVM~\cite{llvm} infrastructure.
 

### 2013-02-14 doc/xps2013/proposal.tex
some edits all over

@@
 
-\paragraph{Scaling up the Flash array:} Existing Flash-based \ssds\ are designed for large block requests.
-An \ssd\ for Grappa could be designed to support high concurrency, small requests.
-Where a traditional \ssd\ might put multiple Flash chips on the same bus, or use Flash chips in parallel to build a wider, high-bandwidth bus, a Grappa \ssd\ would try to maximize message rate by putting each chip on its own bus.
-Serial Flash might be used to minimize pin count requirements, at the cost of some additional latency.
-This design could scale to our necessary request rates, but only at an enormously high chip count: to support 100 million reads per second, an array of 25$\mu$s flash would require 2500 chips.
+\paragraph{Scaling up the Flash array:} Existing Flash-based \ssds\ are
+designed for large block requests. An \ssd\ for Grappa could be designed to
+support high concurrency, small requests. Where a traditional \ssd\ might put
+multiple Flash chips on the same bus, or use Flash chips in parallel to build
+a wider, high-bandwidth bus, a Grappa \ssd\ would try to maximize message rate
+by putting each chip on its own bus. Serial Flash might be used to minimize
+pin count requirements, at the cost of some additional latency. This design
+could scale to our necessary request rates, but only at an enormously high
+chip count: to support 100 million reads per second, an array of 25$\mu$s
+flash would require 2500 chips.
 
-\paragraph{Scaling down the processors:} Alternatively, rather than scaling up the flash array, processors could be scaled down to match the flash request rate.
-A low-power processor with few cores might generate only 1 million references per second.
-This could be satisfied with only $25\times 25\mu$s Flash chips.
-There are two problems with this design for our purposes, however.
-First, while each node requires fewer chips, the overall cluster performance goal, and thus chip count requirement, is the same.
-More system nodes are thus required to achieve the same level of performance.
-Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
+\paragraph{Scaling down the processors:} Alternatively, rather than scaling up
+the flash array, processors could be scaled down to match the flash request
+rate. A low-power processor with few cores might generate only 1 million
+references per second. This could be satisfied with only $25\times 25\mu$s
+Flash chips. There are two problems with this design for our purposes,
+however. First, while each node requires fewer chips, the overall cluster
+performance goal, and thus chip count requirement, is the same. More system
+nodes are thus required to achieve the same level of performance. Second, low
+single-threaded performance is detrimental to performance during periods of
+diminished application concurrency.
 
-\paragraph{Buffering requests to increase locality:} Grappa trades latency for throughput by aggregating network communication.
-The same idea could be applied to increase locality and amortize the cost of Flash accesses.
-Requests to nearby flash locations would be bucketed by flash page address for a limited time.
-When a sufficient number of requests have arrived for a page or when the requests time out, a single page request would be issued to the Flash layer; its result would be used to satisfy the individual requests.
-The challenge with this approach is as storage size grows, the effectiveness of aggregation correspondingly decreases.
-On the other hand, aggregation occurs for all global memory requests generated by any node in the system, so as aggregate system performance increases, aggregation performance actually \emph{increases} (up to the point that the network is saturated).
+% \paragraph{Buffering requests to increase locality:} Grappa trades latency for
+% throughput by aggregating network communication. The same idea could be
+% applied to increase locality and amortize the cost of Flash accesses. Requests
+% to nearby flash locations would be bucketed by flash page address for a
+% limited time. When a sufficient number of requests have arrived for a page or
+% when the requests time out, a single page request would be issued to the Flash
+% layer; its result would be used to satisfy the individual requests. The
+% challenge with this approach is as storage size grows, the effectiveness of
+% aggregation correspondingly decreases. On the other hand, aggregation occurs
+% for all global memory requests generated by any node in the system, so as
+% aggregate system performance increases, aggregation performance actually
+% \emph{increases} (up to the point that the network is saturated).
 

### 2013-02-14 doc/xps2013/proposal.tex
ssd fixes

--- 2013-02-14 doc/xps2013/proposal.tex
@@
 Grappa depends on user-mode access to the network layer to avoid this overhead; a similar approach has been applied to \ssds\ in recent research~\cite{caulfield:2012}, however.
-We see several ways to address these challenges:
 
-\paragraph{Scaling up the Flash array:} Existing Flash-based \ssds\ are designed for large block requests.
-An \ssd\ for Grappa could be designed to support high concurrency, small requests.
-Where a traditional \ssd\ might put multiple Flash chips on the same bus, or use Flash chips in parallel to build a wider, high-bandwidth bus, a Grappa \ssd\ would try to maximize message rate by putting each chip on its own bus.
-Serial Flash might be used to minimize pin count requirements, at the cost of some additional latency.
-This design could scale to our necessary request rates, but only at an enormously high chip count: to support 100 million reads per second, an array of 25$\mu$s flash would require 2500 chips.
-
-\paragraph{Scaling down the processors:} Alternatively, rather than scaling up the flash array, processors could be scaled down to match the flash request rate.
-A low-power processor with few cores might generate only 1 million references per second.
-This could be satisfied with only $25\times 25\mu$s Flash chips.
-There are two problems with this design for our purposes, however.
-First, while each node requires fewer chips, the overall cluster performance goal, and thus chip count requirement, is the same.
-More system nodes are thus required to achieve the same level of performance.
-Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
-
-\paragraph{Buffering requests to increase locality:} Grappa trades latency for throughput by aggregating network communication.
-The same idea could be applied to increase locality and amortize the cost of Flash accesses.
-Requests to nearby flash locations would be bucketed by flash page address for a limited time.
-When a sufficient number of requests have arrived for a page or when the requests time out, a single page request would be issued to the Flash layer; its result would be used to satisfy the individual requests.
-The challenge with this approach is as storage size grows, the effectiveness of aggregation correspondingly decreases.
-On the other hand, aggregation occurs for all global memory requests generated by any node in the system, so as aggregate system performance increases, aggregation performance actually \emph{increases} (up to the point that the network is saturated).
+To address this challenge we will explore building a flash array consisting of several ten's to a hundred disks \emph{per node}.
+This will enable us to increase random accesses to our target 100 million references per second.
+But it will require several active disk controllers per system.
+We do not expect any existing device driver stack will work at this scale.
+Hence we will need to write a custom disk controller device driver and file system that supports such a high degree of concurrent requests across so many parallel channels.
 

### 2013-02-14 doc/xps2013/proposal.tex
edits to summary

--- 2013-02-14 doc/xps2013/proposal.tex
@@
 Keeping all these data in DRAM only is unlikely to be cost effective.
-We will explore mechanisms that use solid-state storage as primary memory, greatly expanding the directly addressible storage at low cost.
+We will explore mechanisms that use solid-state storage as primary memory, greatly expanding the directly addressable storage at low cost.
 
@@
 Irregular applications are increasing in importance to the computing industry.
-For a long time they have underpinned computations of interest to the national security arm of the federal government.
-But now, they underpin such questions as ad placement in social networks, and analysis of complex data-sets in medicine and science.
-The defining characteristic of these applications is poor locality and massive available parallelism.
-Our effort is focused on making these applications perform well and scale on commodity hardware.
-The ``secret sauce'' that makes this work is to essentially build a software XMT~\cite{feo:xmt}.
-Instead of relying on locality (which doesn't exist for these applications) we rely on threading to tolerate memory latency.
-Unlike the XMT, however, we must contend with commodity networking hardware, and here we again rely on threading to provide a sufficient number of requests that we can buffer them prior to dispatching on the physical wire.
+For a long time they have underpinned computations of interest to the national
+security arm of the federal government. But now, they underpin such questions
+as ad placement in social networks, and analysis of complex data-sets in
+medicine and science. The defining characteristic of these applications is
+poor locality and massive available parallelism. Our effort is focused on
+making these applications perform well and scale on commodity hardware. The
+key idea that makes this work is to rely on massive concurrency to tolerate
+memory latency, instead of relying on locality (which doesn't exist for these
+applications). This core idea underpinned the XMT~\cite{feo:xmt} system, a
+fully-custom hardware system. We plan to explore that idea in software only,
+running on commodity processors. And unlike the XMT, however, we must also
+contend with commodity networking hardware, and here we again rely on
+threading to provide a sufficient number of requests that we can buffer them
+prior to dispatching on the network, to overcome bandwidth limitations when
+operating with small messages.
 
-Our past effort has shown a workable proof of concept of all this runtime smoke and mirrors: a commodity cluster can out do the XMT at its own game, by emulating its key features in software.
-What is needed now is additional effort to improve the underlying performance of this runtime and make it usable to a broad base of developers.
-We will be focusing our efforts in the next three years on the key performance component, the networking layer, and the key programmer efficiency linchpin, high level language support.
-In addition we need a useful mechanism for handling data-sets in the petabyte range, and here we imagine building a cluster of machines that relies primarily on SSD's and not DRAM for data-set storage.
+Our preliminary effort has shown a workable proof of concept of all this
+runtime smoke and mirrors: a commodity cluster can out do the XMT at its own
+game, by emulating its key features in software. What is needed now is
+additional effort to improve the underlying performance of this runtime and
+make it usable to a broad base of users. We will be focusing our efforts
+in the next three years on the key performance component, the networking
+layer, and the key programmer efficiency linchpin, high level language
+support. In addition we need a useful mechanism for handling data-sets in the
+petabyte range, and here we imagine building a cluster of machines that relies
+primarily on SSD's and not DRAM as main memory data-set storage.
 
-If successful, the Grappa runtime will have a transformative impact on the computing industry.
-Irregular applications are hugely important, and if off the shelf hardware can be efficient with them a new class of questions can be asked of computing systems.
-Additionally our benchmark suite development will further drive innovation among researchers in this space.
+If successful, the Grappa runtime will have a transformative impact on the
+computing industry. Irregular applications are hugely important, and if off
+the shelf hardware can be efficient with them a new class of questions can be
+asked of computing systems. Additionally our benchmark suite development will
+further drive innovation among researchers in this space.
 

### 2013-02-14 doc/xps2013/proposal.tex
oskin crap

--- 2013-02-14 doc/xps2013/proposal.tex
@@
 
-Oskin has... {\bf MARK PUT YOUR STUFF HERE.}
+Oskin has explored a wide variety of architectural mechanisms for improving application performance, most notably intelligent memory~\cite{activepages} and dataflow computing~\cite{wavescalar-micro, wavescalar-isca}.
+He has developed novel evaluation techniques~\cite{hls} that enable efficient design space exploration of complex superscalar processors.
+His past work has included runtime systems for GPU-based computing~\cite{oskin-gpu}, application specific processor design~\cite{sherpa} and polymorphic on-chip networks~\cite{polynetworks}.
+He has also explored several architectural design options for quantum computers~\cite{quantum-isca03, quantum-isca05, quantum-isca08}.
+Finally he developed mechanisms to make parallel systems more deterministic and easier to program~\cite{dmp}.
+His most recent work is focused on the work described in this proposal~\cite{Nelson:hotpar2011}, which is to make irregular computations efficient on commodity hardware.
 
-Ceze has been conducting research on architecture support, compiler
-techniques, and programming models to improve the programmability of
-multiprocessor systems for seven years and has co-authored over thirty
-papers in these areas.  His research includes advances in
-architecture~\cite{tlsooo, bulk, bulksc, swbulk, cyclops, dmp,
-  delorean} and programming models~\cite{colorama, ipot, posh} to
-simplify exploiting concurrency in multiprocessors. Some of his recent
-results especially related to this proposal include a programming
-model based on ordered transactions and data annotations~\cite{ipot};
-improving reliability of multithreaded software with architecture
-support for bug detection and avoidance~\cite{atomaid, aatoppicks,
-  bugaboo, oshajava, cs-isca10, ce-isca10}; and deterministic
-execution of multithreaded programs~\cite{dmp, dmptoppicks,
-  asplos10coredet, dmpos}. PI Ceze recently co-organized an NSF-funded
-workshop focused specifically on deterministic multiprocessing. The
-workshop was very successful, bringing leaders in the multicore area
-from industry and academia.
+Ceze has been conducting research on architecture support, compiler techniques, and programming models to improve the programmability of multiprocessor systems for seven years and has co-authored over thirty papers in these areas.
+His research includes advances in architecture~\cite{tlsooo, bulk, bulksc, swbulk, cyclops, dmp, delorean} and programming models~\cite{colorama, ipot, posh} to simplify exploiting concurrency in multiprocessors.
+Some of his recent results especially related to this proposal include a programming model based on ordered transactions and data annotations~\cite{ipot}; improving reliability of multithreaded software with architecture support for bug detection and avoidance~\cite{atomaid, aatoppicks, bugaboo, oshajava, cs-isca10, ce-isca10}; and deterministic execution of multithreaded programs~\cite{dmp, dmptoppicks, asplos10coredet, dmpos}.
+PI Ceze recently co-organized an NSF-funded workshop focused specifically on deterministic multiprocessing.
+The workshop was very successful, bringing leaders in the multicore area from industry and academia.
 
-Kahan has been working in the area of parallel computing for irregular
-applications for twenty years, largely in industry.  At Cray (previously Tera
-Computer) he developed the runtime system~\cite{Alverson95schedulingon,Alverson97terahardware-software} and combining memory allocator~\cite{MAMA} still in production use on
-Cray's XMT system in which over 50,000 threads execute concurrently sharing one
-global address space.  He has co-authored patents and publications for both of these
-systems as well as for a scalable connected components graph algorithm that bears
-his name~\cite{Underwood07analyzingthe}.
+Kahan has been working in the area of parallel computing for irregular applications for twenty years, largely in industry.
+At Cray (previously Tera Computer) he developed the runtime system~\cite{Alverson95schedulingon,Alverson97terahardware-software} and combining memory allocator~\cite{MAMA} still in production use on Cray's XMT system in which over 50,000 threads execute concurrently sharing one global address space.
+He has co-authored patents and publications for both of these systems as well as for a scalable connected components graph algorithm that bears his name~\cite{Underwood07analyzingthe}.
 Kahan originated and for three years has been co-PI for the Grappa project.

### 2013-02-14 doc/xps2013/proposal.tex
incorporating Myers suggestions

--- 2013-02-14 doc/xps2013/proposal.tex
@@
 
+%% I don't think we need the parahraph below. 
+% While some irregular applications can be restructured to better exploit
+% locality, aggregating requests to increase message size, and manage the
+% additional challenges of load balance and synchronization across multinode
+% systems, the work to do so is formidable and requires knowledge and skills
+% pertaining to distributed systems far beyond those of most application
+% programmers.
+
 Luckily, many of the important irregular applications naturally offer large
@@
 fast compared to implementations without Grappa, but obviously there is room
-for improvement.
+for improvement. Most importantly, however, for significantly less real world
+cost, users can \emph{add} significantly more processors to a commodity
+cluster than an XMT machine and use Grappa to achieve scalable performance.
 
-% Most importantly, however, for a much lower real world cost,
-% users can \emph{add} significantly more processors to a commodity cluster than
-% an XMT machine and use Grappa to achieve scalable performance.
-
-Given these very positive early results, this proposal plans to extend Grappa
-in the following directions:
+Given these very positive early results,
+we next propose to extend Grappa in the following directions:
 
@@
 \item \textbf{Language and compiler support:} Presently, Grappa is implemented
-strictly as a library.  A significant research thrust of the proposed work is to use
+strictly as a library.  Programmers must express remote operations explicitly as function calls.
+This adds burden high-level language programmers do not usually face and obfuscates opportunities
+for compiler optimizations. A significant research thrust of the proposed work is to use
 language and compilation support to provide a cleaner programming model and use
@@
 Fortunately we have some preliminary evidence we won't get wedged along the way.
-This is an intellectually challenging endeavor, that we believe is well suited for the NSF XPS program.
-In summary, our goal is to exploit parallelism from irregular computations to efficiently execute them at scale.
+This is an intellectually challenging endeavor, landing squarely in the middle of the NSF XPS program: using and exploiting parallelism from irregular computations to efficiently execute them at scale.
 
@@
 If commodity hardware can be cajoled into executing such applications efficiently it will open the door for new and rich applications.
+Furthermore, irregular applications will then reap the benefits of advances in commodity processors and networks, rather than requiring custom hardware that advances only at much higher cost.
 Graph queries, machine-learning and data-driven science will be enabled by the technology we aim to create from the ideas in this grant proposal.
@@
 GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
-To this commodity hardware and software mix Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as few as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
+To this commodity hardware and software mix Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as few as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/}, and synchronizing operations including \emph{fetch-and-add\/}~\cite{fetchandadd} as well as user-defined atomic methods; and (3) a \emph{message aggregation\/} layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
 As we will show later, Grappa can tolerate latencies way beyond that of the network.
-Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our effective random access memory bandwidth, our synchronization bandwidth, and our ability to improve load imbalance.
+Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
 
@@
 
-During execution, a task yields control of its core whenever it performs a long-latency operation, allowing the processor to remain busy while waiting for the operation to complete.
+During execution, a task yields control of its core following initiation of any long-latency operation, allowing the processor to remain busy while waiting for the operation to complete.
 In addition,  a programmer can direct scheduling explicitly via the Grappa API calls.
@@
 Applications written for Grappa utilize two forms of memory: local and global.
-Local memory is local to a single core in the system.
-Accesses occur through conventional pointers.
-The compiler emits an access and the memory is manipulated directly.
-Applications use local accesses for the stack associated with a task, accesses to localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
+Each physical core has a separate local address space.
+The core accesses this local memory through conventional pointers, such as those generated by a standard C compiler.
+We nonetheless refer to these pointers as ``local pointers'' to differentiate them from ``global pointers'' that may be used by other cores to reference the same physical memory as described below.
+Applications use local memory for a number of things in Grappa: the stack associated with a task, localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
 Local pointers cannot access memory on other cores, and are valid only on their home core.
 
-Large data that is expected to be shared and accessed with low locality is stored in global memory.
-All global data must be accessed through calls into Grappa's API.
+Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory.
+All global data must be accessed through Grappa's API.
 %, shown in Figure~\ref{fig:accessing-memory}.
@@
 
-Grappa also allows any local data on a core's stacks or heap to be exported to the global address space to be made accessible to other cores across the system.
+Grappa also allows any local data on a core's stacks or heap to be exported to the global address space, so as to be accessible to other cores across the system.
 Addresses to global memory allocated in this way use \textbf{2D global addresses}.
@@
 Any node-local data can be made accessible by other nodes in the system by wrapping the address and node ID into a 2D global address.
-This address can then be accessed with a delegate.
-% and can also be cached by other nodes. 
 At the destination the address is converted into a canonical x86 address by replacing the upper bits with the sign-extended upper bit of the virtual address.
@@
 \paragraph{Global memory access} There are two general approaches Grappa applications use to {\emph access} global memory.
-When the programmer expects a computation on shared data to have spatial locality to exploit, explicit {\em cache} operations may be used.
-When there is no locality to exploit, remote {\em delegate} operations are used.
+When the programmer expects a computation on shared data to have spatial locality to exploit, {\em cache} operations may be used.
+When there is no locality to exploit, {\em delegate} operations are used.
 
@@
 For example, the UPC compiler coalesces struct and array accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's message vectorization hoists small messages out of a loop \cite{FortranD:1992}.
+Caching in Grappa additionally provides a mechanism for exploiting temporal locality by operating on the data locally.
 
-Caching in Grappa additionally provides a mechanism for exploiting temporal locality by operating on the data locally --- even though we aim Grappa at applications with low locality, we want to be able to exploit it when available.
-% Under the hood, Grappa performs the mechanics of gathering chunks of data
-% from multiple system nodes and presenting a conventional appearing linear
-% block of memory as a local pointer into a cache. The strategy employed is to
-% issue all the constituent requests of a cache access request	(as Active
-% Messages) and then yield until all responses have occurred.	
-Currently, Grappa caches are \emph{not} coherent, requiring the programmer to maintain consistent access to data.
+Under the hood, Grappa performs the mechanics of gathering chunks of data from multiple system nodes and presenting a conventional appearing linear block of memory as a local pointer into a cache.
+The strategy employed is to issue all the constituent requests of a cache access request  (as Active Messages) and then yield until all responses have occurred.
+Currently, the burden is on the programmer to maintain coherency by introducing explicit synchronization.
 Future work will develop a software directory based coherence scheme to simplify consistent access to global data.
@@
 
-% Delegate operations are always executed at the home core of their address, and
-% while arbitrary memory operations can be delegated, we restrict the use of
-% delegate operations in three ways to make them more useful for
-% synchronization. First, we limit each task to one outstanding delegate
-% operation to avoid the possibility of reordering in the network. Second, we
-% limit delegate operations to operate on objects in the 2D address space or
-% objects that fit in a single block of the linear address space, so they can be
-% satisfied with a single network request. Finally, no context switches are
-% allowed while the data is being modified. Given these restrictions, we can
-% ensure that delegate operations for the same address from multiple requesters
-% are always serialized through a single core in the system, providing atomic
-% semantics without using atomic operations.
+Delegate operations are always executed at the home core of their address, and while arbitrary memory operations can be delegated, we restrict the use of delegate operations in three ways to make them more useful for synchronization.
+First, we limit each task to one outstanding delegate operation to avoid the possibility of reordering in the network.
+Second, we limit delegate operations to operate on objects in the 2D address space or objects that fit in a single block of the linear address space, so they can be satisfied with a single network request.
+Finally, no context switches are allowed while the data is being modified.
+Given these restrictions, we can ensure that delegate operations for the same address from multiple requesters are always serialized through a single core in the system, providing atomic semantics without using atomic operations.
 
@@
 
-Grappa's lower networking layer aggregates the upper layer's messages to
-improve performance. Commodity networks including
-Infiniband~\cite{infinibandbandwidth} achieve their peak bisection bandwidth
-\emph{only} when the packet sizes are relatively large---on the order of
-multiple kilobytes. The reason for this discrepancy is the combination of
-overheads associated with handling each packet (in terms of bytes that form
-the actual packet, processing time at the card and processing on the CPU
-within the driver stack). Our measurements confirm manufacturers published
-data~\cite{infinibandbandwidth}, that with this packet size the bisection
-bandwidth is only a small fraction, less than 3\% of the peak bisection
-bandwidth.
+Grappa's lower networking layer aggregates the upper layer's messages to improve performance.
+Commodity networks including infiniband achieve their peak bisection bandwidth \emph{only} when the packet sizes are relatively large---on the order of multiple kilobytes.
+The reason for this discrepancy is the combination of overheads associated with handling each packet (in terms of bytes that form the actual packet, processing time at the card and processing on the CPU within the driver stack).
+Our measurements confirm manufacturers published data~\cite{infinibandbandwidth}, that with this packet size the bisection bandwidth is only a small fraction, less than 3\% of the peak bisection bandwidth.
 
-In our early experiments the vast majority of requests were smaller than 44
-bytes, far too small to make efficient use of the network. To make the best
-use of the network, we must convert our small messages into large ones. When a
-task sends a message, it is not immediately sent, but rather placed in a queue
-specific to the destination. If the size in bytes of a queue is above a
-threshold (e.g., 4096, typically large enough for good bandwidth), the
-contents of the queue are sent immediately. Also, each queue has a wait time
-threshold ($\approx${1ms}). If the oldest message in a queue has been waiting
-longer than this threshold, the contents of the queue are sent immediately,
-even if the queue size is lower than the message size threshold. Finally,
-queues may be explicitly flushed, for situations when latency is more
-important than bandwidth.
+In our experiments the vast majority of requests were smaller than 44 bytes, far too small to make efficient use of the network.
+To make the best use of the network, we must convert our small messages into large ones.
+When a task sends a message, it is not immediately sent, but rather placed in a queue specific to the destination.
 
-% The network layer is serviced by polling. Periodically when a context switch
-% occurs, the Grappa scheduler switches to the network polling thread. This
-% thread has three responsibilities. First, it polls the lower-level network
-% layer to ensure it makes progress. Second, it deaggregates received messages
-% and executes active message handlers. Third, it checks to see if any
-% aggregation queues have messages that have been waiting longer than the
-% threshold; if so, it sends them.
+There are three situations in which a queue of aggregated messages is sent.
+First, each queue has a message size threshold of 4096 bytes, chosen to give reasonable network performance.
+If the size in bytes of a queue is above the threshold, the contents of the queue are sent immediately.
+Second, each queue has a wait time threshold ($\approx${1ms}).
+If the oldest message in a queue has been waiting longer than this threshold, the contents of the queue are sent immediately, even if the queue size is lower than the message size threshold.
+Third, queues may be explicitly flushed in situations where the programmer wants to minimize the latency of a message at the cost of bandwidth utilization.
 
-Underneath the aggregation layer, Grappa uses the \gasnet~communication
-library~\cite{gasnet} to actually move data. All interprocess communication,
-whether on or off a cluster node, is handled by the \gasnet~library.
-\gasnet~is able to take advantage of many communication mechanisms, including
-ethernet and infiniband between nodes, as well as shared memory within a node.
+The network layer is serviced by polling.
+Periodically when a context switch occurs, the Grappa scheduler switches to the network polling thread.
+This thread has three responsibilities.
+First, it polls the lower-level network layer to ensure it makes progress.
+Second, it deaggregates received messages and executes active message handlers.
+Third, it checks to see if any aggregation queues have messages that have been waiting longer than the threshold; if so, it sends them.
 
-Some networks provide access to a remote machine's memory directly. This would
-seem to be a good fit for a programming model focused on global shared memory,
-but in fact we do not use it. In our experiments, we found that RDMA
-operations are subject to the same message rate limitations as all other
-messages on these cards, and thus using raw RDMA operations for our small
-messages would make inefficient use of bandwidth. Instead, we implement remote
-memory operations with active messages. A byproduct of this design decision is
-that Grappa is not limited to RDMA-capable networks.
+Underneath the aggregation layer, Grappa uses the \gasnet~communication library~\cite{gasnet} to actually move data.
+All interprocess communication, whether on or off a cluster node, is handled by the \gasnet~library.
+\gasnet~is able to take advantage of many communication mechanisms, including ethernet and infiniband between nodes, as well as shared memory within a node.
+
+Some networks provide access to a remote machine's memory directly.
+This would seem to be a good fit for a programming model focused on global shared memory, but in fact we do not use it.
+In our experiments, we found that RDMA operations are subject to the same message rate limitations as all other messages on these cards, and thus using raw RDMA operations for our small messages would make inefficient use of bandwidth.
+Namely, we would be limited to the network injection rate providing just a few million accesses per second per node.
+Instead, we implement remote memory operations with active messages in conjuction with aggregation, providing an order of magnitude or more increase in throughput.
+A byproduct of this design decision is that Grappa is not limited to RDMA-capable networks.
 
@@
 
-To evaluate the performance of our current prototype with respect to the XMT,
-we ran each of our three benchmarks on up to 16 nodes of each machine. Grappa
-used 6 cores per node, with the best parameters chosen for each point. In some
-cases, the XMT could not run the benchmark with 2 nodes, so the point is
-omitted.
+To evaluate Grappa's performance with respect to the XMT, we ran each of our three benchmarks on up to 16 nodes of each machine.
+Grappa used 6 cores per node, with the best parameters chosen for each point.
+In some cases, the XMT could not run the benchmark with 2 nodes, so the point is omitted.
 
@@
 \begin{center}
-\includegraphics[width=2.5in]{figs/uts_performance.pdf}
+\includegraphics[width=3in]{figs/uts_performance.pdf}
 \caption{\label{fig:uts_compare} Performance of in-memory unbalanced tree search.}
@@
 \begin{center}
-\includegraphics[width=2.5in]{figs/bfs_performance}
-\caption{\label{fig:bfs-performance} Performance of breadth-first search.}
+\includegraphics[width=3in]{figs/bfs_performance}
+\caption{\label{fig:bfs-performance} Performance of breadth-first searcch.}
 \end{center}
@@
 
-\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex
-tree (T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
-number of vertices visited per second versus number of compute nodes. Grappa
-is 3.2 times faster than the XMT at 16 nodes. As we will show later, the
-performance advantage Grappa has over XMT increases as more nodes are added.
-The main reason Grappa performs better is the software-based delegate
-synchronization obviates the need for the retry-based synchronization that XMT
-uses.
+\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex tree (T1L).
+Figure~\ref{fig:uts_compare} shows the performance in terms of number of vertices visited per second versus number of compute nodes.
+Grappa is 3.2 times faster than the XMT at 16 nodes.
+As we will show later, the performance advantage Grappa has over XMT increases as more nodes are added.
+The main reason Grappa performs better is the software-based delegate synchronization obviates the need for the retry-based synchronization that XMT uses.
 
@@
 The XMT is 2.5 times faster than Grappa at 16 nodes.
-Performance does scale at a constant rate for Grappa, suggesting that adding more nodes will increase performance.
+Performance has not peaked for Grappa, suggesting that adding more nodes will increase performance.
 
@@
 Instead, Grappa, like other partitioned global address space (PGAS) models, implements its DSM at the language, rather than system level.
+%%note:  Myers suggestion is not incorporated here, as this seems ok:
 Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and UPC~\cite{upc:2005} make accesses to shared structures look like normal memory references.
@@
 In our experience, when these networks move small blocks of data (a few bytes), only a fraction of the available bandwidth is achieved.
+%%note:  Myers suggestion is not incorporated here, as this seems ok:
 In addition, the DSM system in Grappa ends up being tightly coupled to the task scheduler in order to overlap long latency memory operations with useful computation.
@@
 As we describe in this paper, Grappa use a lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks on a single processing core.
-The large number of tasks is required because of the extremely high internode latency Grappa is mitigating.
-Grappa's task library employs several optimizations: an extremely fast task switch, a small task size, and judicious use of hardware prefetching to bring task state into the cache long before that task is actually scheduled.
+The large number of tasks is required because of the extremely high internode latency Grappa mitigates.
+Grappa's task library employs several optimizations: an extremely fast task switch, a small task size, and judicious use of hardware prefetching to bring task state into the cache before the task is scheduled to execute.
 
@@
 Thus the aggregator will need to very carefully utilize lock-free data-structures and interact with \gasnet~carefully.
+%%Didn't understand Myers suggestion here.
 
@@
 Grappa depends on user-mode access to the network layer to avoid this overhead; a similar approach has been applied to \ssds\ in recent research~\cite{caulfield:2012}, however.
+We see several ways to address these challenges:
 
-To address this challenge we will explore building a flash array consisting of several ten's to a hundred disks \emph{per node}.
-This will enable us to increase random accesses to our target 100 million references per second.
-But it will require several active disk controllers per system.
-We do not expect any existing device driver stack will work at this scale.
-Hence we will need to write a custom disk controller device driver and file system that supports such a high degree of concurrent requests across so many parallel channels.
+\paragraph{Scaling up the Flash array:} Existing Flash-based \ssds\ are designed for large block requests.
+An \ssd\ for Grappa could be designed to support high concurrency, small requests.
+Where a traditional \ssd\ might put multiple Flash chips on the same bus, or use Flash chips in parallel to build a wider, high-bandwidth bus, a Grappa \ssd\ would try to maximize message rate by putting each chip on its own bus.
+Serial Flash might be used to minimize pin count requirements, at the cost of some additional latency.
+This design could scale to our necessary request rates, but only at an enormously high chip count: to support 100 million reads per second, an array of 25$\mu$s flash would require 2500 chips.
+
+\paragraph{Scaling down the processors:} Alternatively, rather than scaling up the flash array, processors could be scaled down to match the flash request rate.
+A low-power processor with few cores might generate only 1 million references per second.
+This could be satisfied with only $25\times 25\mu$s Flash chips.
+There are two problems with this design for our purposes, however.
+First, while each node requires fewer chips, the overall cluster performance goal, and thus chip count requirement, is the same.
+More system nodes are thus required to achieve the same level of performance.
+Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
+
+\paragraph{Buffering requests to increase locality:} Grappa trades latency for throughput by aggregating network communication.
+The same idea could be applied to increase locality and amortize the cost of Flash accesses.
+Requests to nearby flash locations would be bucketed by flash page address for a limited time.
+When a sufficient number of requests have arrived for a page or when the requests time out, a single page request would be issued to the Flash layer; its result would be used to satisfy the individual requests.
+The challenge with this approach is as storage size grows, the effectiveness of aggregation correspondingly decreases.
+On the other hand, aggregation occurs for all global memory requests generated by any node in the system, so as aggregate system performance increases, aggregation performance actually \emph{increases} (up to the point that the network is saturated).
 
@@
 
-Language and compiler support are instrumental in making Grappa usable to a
-wide audience of application developers. We plan to use the LLVM~\cite{llvm}
-infrastructure. Below we describe the specific aspects we plan to tackle.
+Language and compiler support are instrumental in making Grappa usable to a wide audience of software developers. We envision language and compiler support can tackle a wide variety of challenges:
 
@@
 This increases performance by further overlapping computation with memory access.
-A compiler with alias analysis can do this transformation automatically for the programmer.
+A compiler aided by programmer annotations or capable of alias analysis can do this transformation automatically for the programmer.
 
@@
 We will develop a compiler pass to combine operations when possible.
+%%Myers suggested a reference to UPC or other previous work
 
@@
 
-\paragraph{Task load control:} Having too few tasks per core fails to tolerate
-fully the latency; having too many increases overheads unnecessarily.
-Currently, we choose a number at program instantiation: experiments so far
-have shown evidence of a broad sweet-spot, though it is of course varies as a
-function of application behavior and input data. At compile-time, we might be
-able to estimate the number of tasks required based on the density of remote
-operations and system parameters (e.g., memory size, network latency, etc.);
-at runtime we can adjust the number of tasks dynamically based on periodic
-performance assessments.
+\paragraph{Concurrency Control: } Having too few tasks multiplexed per core fails to tolerate fully the latency; having too many increases memory and cache footprints, resulting in excess cache misses or exhausting physical memory unnecessarily.
+Currently, we choose a number at program instantiation:  experiments so far have shown evidence of a broad sweet-spot, though it is of course varies as a function of application behavior.
+At compile-time, we could estimate the number of tasks required based on the density of remote operations and hard-coded measurements of system characteristics; at runtime we can adjust the number of tasks dynamically based on real-time measurement of processor counters.
+Exploring these directions may lead to even more robust performance.
 
-% Presently we are starting work on this language and compiler support.  We are basing our effort on the LLVM~\cite{llvm} infrastructure.
+Presently we are starting work on this language and compiler support.  We are basing our effort on the LLVM~\cite{llvm} infrastructure.
 
@@
 This benchmark suite will consist of application kernels, reference implementations, data-sets, data-set generators, documentation, specifications for generating standardized performance results, and a hosting service (http://graphbench.org) to provide public access to the benchmark and datasets as well as aggregate and publish results by all users of the benchmark suite.
+Specifically, we aim to provide transparency and an opportunity to build off of others' discoveries by requiring source code be provided with any submission.
 
@@
 \noindent\textbf{Year 1:} The first year of this effort will focus on creating the benchmark suite and initial language and compiler support.
-We aim to pull together 6-8 benchmarks, sample data sets, and scripts to measure performance results on the initial Grappa runtime, UPC~\cite{upc:2005} implementations, and serial reference code.
+We aim to pull together 6-8 benchmarks, sample data sets, and scripts to measure performance results on the initial Grappa runtime, UPC~\cite{upc:2005} and MPI implementations, and serial reference code.
 For the compiler we will have the front end done enabling easy addition of keywords to annotate data-types and structural elements (parallel for, etc).
-The compiler will be generating in-efficient Grappa code by the end of the first year.
+The compiler will be generating non-optimized Grappa code by the end of the first year.
 
@@
 
-Oskin has explored a wide variety of architectural mechanisms for improving application performance, most notably intelligent memory~\cite{activepages} and dataflow computing~\cite{wavescalar-micro, wavescalar-isca}.
-He has developed novel evaluation techniques~\cite{hls} that enable efficient design space exploration of complex superscalar processors.
-His past work has included runtime systems for GPU-based computing~\cite{oskin-gpu}, application specific processor design~\cite{sherpa} and polymorphic on-chip networks~\cite{polynetworks}.
-He has also explored several architectural design options for quantum computers~\cite{quantum-isca03, quantum-isca05, quantum-isca08}.
-Finally he developed mechanisms to make parallel systems more deterministic and easier to program~\cite{dmp}.
-His most recent work is focused on the work described in this proposal~\cite{Nelson:hotpar2011}, which is to make irregular computations efficient on commodity hardware.
+Oskin has... {\bf MARK PUT YOUR STUFF HERE.}
 
-Ceze has been conducting research on architecture support, compiler techniques, and programming models to improve the programmability of multiprocessor systems for seven years and has co-authored over thirty papers in these areas.
-His research includes advances in architecture~\cite{tlsooo, bulk, bulksc, swbulk, cyclops, dmp, delorean} and programming models~\cite{colorama, ipot, posh} to simplify exploiting concurrency in multiprocessors.
-Some of his recent results especially related to this proposal include a programming model based on ordered transactions and data annotations~\cite{ipot}; improving reliability of multithreaded software with architecture support for bug detection and avoidance~\cite{atomaid, aatoppicks, bugaboo, oshajava, cs-isca10, ce-isca10}; and deterministic execution of multithreaded programs~\cite{dmp, dmptoppicks, asplos10coredet, dmpos}.
-PI Ceze recently co-organized an NSF-funded workshop focused specifically on deterministic multiprocessing.
-The workshop was very successful, bringing leaders in the multicore area from industry and academia.
+Ceze has been conducting research on architecture support, compiler
+techniques, and programming models to improve the programmability of
+multiprocessor systems for seven years and has co-authored over thirty
+papers in these areas.  His research includes advances in
+architecture~\cite{tlsooo, bulk, bulksc, swbulk, cyclops, dmp,
+  delorean} and programming models~\cite{colorama, ipot, posh} to
+simplify exploiting concurrency in multiprocessors. Some of his recent
+results especially related to this proposal include a programming
+model based on ordered transactions and data annotations~\cite{ipot};
+improving reliability of multithreaded software with architecture
+support for bug detection and avoidance~\cite{atomaid, aatoppicks,
+  bugaboo, oshajava, cs-isca10, ce-isca10}; and deterministic
+execution of multithreaded programs~\cite{dmp, dmptoppicks,
+  asplos10coredet, dmpos}. PI Ceze recently co-organized an NSF-funded
+workshop focused specifically on deterministic multiprocessing. The
+workshop was very successful, bringing leaders in the multicore area
+from industry and academia.
 
-Kahan has been working in the area of parallel computing for irregular applications for twenty years, largely in industry.
-At Cray (previously Tera Computer) he developed the runtime system~\cite{Alverson95schedulingon,Alverson97terahardware-software} and combining memory allocator~\cite{MAMA} still in production use on Cray's XMT system in which over 50,000 threads execute concurrently sharing one global address space.
-He has co-authored patents and publications for both of these systems as well as for a scalable connected components graph algorithm that bears his name~\cite{Underwood07analyzingthe}.
+Kahan has been working in the area of parallel computing for irregular
+applications for twenty years, largely in industry.  At Cray (previously Tera
+Computer) he developed the runtime system~\cite{Alverson95schedulingon,Alverson97terahardware-software} and combining memory allocator~\cite{MAMA} still in production use on
+Cray's XMT system in which over 50,000 threads execute concurrently sharing one
+global address space.  He has co-authored patents and publications for both of these
+systems as well as for a scalable connected components graph algorithm that bears
+his name~\cite{Underwood07analyzingthe}.
 Kahan originated and for three years has been co-PI for the Grappa project.

### 2013-02-14 doc/xps2013/header.tex

@@
 \makeatletter
+
 \renewcommand{\paragraph}{%
   \@startsection{paragraph}{4}%
-  {\z@}{1.0ex \@plus 1ex \@minus .2ex}{-1em}%
+  {\z@}{1.0ex \@plus 1ex \@minus 1ex}{-1em}%
   {\normalfont\normalsize\bfseries}%
@@
 
+\comment{
 \renewcommand\section{\@startsection{section}{1}{\z@}{-10\p@ \@plus -4\p@ \@minus -4\p@}{5\p@ \@plus 4\p@ \@minus 4\p@}{\normalfont\Large\bfseries\boldmath\rightskip=\z@ \@plus 8em\pretolerance=10000 }}
@@
 \renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}{-10\p@ \@plus -4\p@ \@minus -4\p@}{5\p@ \@plus 4\p@ \@minus 4\p@}{\normalfont\normalsize\bfseries\boldmath\rightskip=\z@ \@plus 8em\pretolerance=10000 }}
+}
 

### 2013-02-14 doc/xps2013/proposal.tex

--- 2013-02-14 doc/xps2013/proposal.tex
@@
 GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
-To this commodity hardware and software mix Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as few as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
+To this commodity hardware and software mix Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as few as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} and synchronizing operations including \emph{fetch-and-add\/}~\cite{fetchandadd} as well as user-defined atomic methods; and (3) a \emph{message aggregation\/} layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
 As we will show later, Grappa can tolerate latencies way beyond that of the network.
@@
 Applications written for Grappa utilize two forms of memory: local and global.
-Local memory is local to a single core in the system.
-Accesses occur through conventional pointers.
-The compiler emits an access and the memory is manipulated directly.
-Applications use local accesses for the stack associated with a task, accesses to localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
+Tasks executing on the same core share an address space local to that core.
+The core accesses this local memory through conventional pointers, such as those generated by a standard C compiler.
+We nonetheless refer to these pointers as ``local pointers'' to differentiate them from ``global pointers'' that may be used by other cores to reference the same physical memory as described below.
+Applications use local memory for a number of things in Grappa: the stack associated with a task, localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
 Local pointers cannot access memory on other cores, and are valid only on their home core.
 
-Large data that is expected to be shared and accessed with low locality is stored in global memory.
-All global data must be accessed through calls into Grappa's API.
+Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory.
+All global data must be accessed through Grappa's API.
 %, shown in Figure~\ref{fig:accessing-memory}.
@@
 Caching in Grappa additionally provides a mechanism for exploiting temporal locality by operating on the data locally --- even though we aim Grappa at applications with low locality, we want to be able to exploit it when available.
-% Under the hood, Grappa performs the mechanics of gathering chunks of data
-% from multiple system nodes and presenting a conventional appearing linear
-% block of memory as a local pointer into a cache. The strategy employed is to
-% issue all the constituent requests of a cache access request	(as Active
-% Messages) and then yield until all responses have occurred.	
-Currently, Grappa caches are \emph{not} coherent, requiring the programmer to maintain consistent access to data.
+Currently, the burden is on the programmer to maintain coherency by introducing explicit synchronization.
 Future work will develop a software directory based coherence scheme to simplify consistent access to global data.
@@
 
-Some networks provide access to a remote machine's memory directly. This would
-seem to be a good fit for a programming model focused on global shared memory,
-but in fact we do not use it. In our experiments, we found that RDMA
-operations are subject to the same message rate limitations as all other
-messages on these cards, and thus using raw RDMA operations for our small
-messages would make inefficient use of bandwidth. Instead, we implement remote
-memory operations with active messages. A byproduct of this design decision is
-that Grappa is not limited to RDMA-capable networks.
+Some networks provide access to a remote machine's memory
+directly. This would seem to be a good fit for a programming model
+focused on global shared memory, but in fact we do not use it. In our
+experiments, we found that RDMA operations are subject to the same
+message rate limitations as all other messages on these cards, and
+thus using raw RDMA operations for our small messages would make
+inefficient use of bandwidth. Namely, we would be limited to the
+network injection rate providing just a few million accesses per
+second per node.  Instead, we implement remote memory operations with
+active messages in conjuction with aggregation, providing an order of
+magnitude or more increase in throughput. A byproduct of this design
+decision is that Grappa is not limited to RDMA-capable networks.
 

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/examples/xed-disas-elf.cpp
Incorporated Preston's suggestions 	modified:   proposal.bib 	modified:   proposal.tex

@@ -1,557 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file disas-elf.cpp
-/// @author Mark Charney   <mark.charney@intel.com>
-
-#include "xed-disas-elf.H"
-
-#if defined(XED_ELF_READER)
-
-////////////////////////////////////////////////////////////////////////////
-#include <elf.h>
-
-extern "C" {
-#include "xed-interface.h"
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/examples/xed-disas-macho.H

@@ -1,50 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-disas-macho.H
-/// @author Mark Charney <mark.charney@intel.com>
-#if !defined(_XED_DISAS_MACHO_H_)
-# define _XED_DISAS_MACHO_H_
-
-#if defined(__APPLE__)
-# define XED_MAC_OSX_FILE_READER
-#endif
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/examples/xed-ex4.c

@@ -1,151 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-ex4.c
-/// @author Mark Charney   <mark.charney@intel.com>
-
-// decoder example. This is the "C" version of xed-ex2.cpp (which is C++).
-
-#include "xed-interface.h"
-#include "xed-examples-util.h"
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h> //strcmp, memset
-#include <assert.h> 
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/examples/xed-symbol-table.H

@@ -1,103 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-
-#if !defined(_XED_SYMBOL_TABLE_H_)
-#define _XED_SYMBOL_TABLE_H_
-extern "C" {
-#include "xed-interface.h"
-#include "xed-examples-util.h"
-}
-#include <vector>
-#include <map>
-#include <iostream>
-#include <iomanip>
-class xed_local_symbol_table_t {
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/examples/xed-tables.c

@@ -1,110 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-
-/// @file xed-tables.c
-/// @author Mark Charney   <mark.charney@intel.com>
-/// @brief a minimal example of accessing the XED internal tables
-
-#include <stdio.h>
-#include "xed-interface.h"
-
-
-void dump_operand(const xed_operand_t* op) {
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/examples/xed-tester.cpp

@@ -1,85 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-tester.cpp
-/// @author Mark Charney   <mark.charney@intel.com>
-
-
-extern "C" {
-#include "xed-interface.h"
-}
-#include <iostream>
-#include <iomanip>
-using namespace std;
-int main(int argc, char** argv);
-
-class xed_test_t
-{
-  public:
-    unsigned int len;
-    unsigned char itext[15];
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/include/xed-error-enum.h

@@ -1,59 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-error-enum.h
-
-// This file was automatically generated.
-// Do not edit this file.
-
-#if !defined(_XED_ERROR_ENUM_H_)
-# define _XED_ERROR_ENUM_H_
-#include "xed-common-hdrs.h"
-typedef enum {
-  XED_ERROR_NONE, ///< There was no error
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/include/xed-inst.h

@@ -1,354 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-inst.h
-/// @author Mark Charney <mark.charney@intel.com>
-
-#if !defined(_XED_INST_H_)
-# define _XED_INST_H_
-
-#include "xed-util.h"
-#include "xed-portability.h"
-#include "xed-category-enum.h" // generated
-#include "xed-extension-enum.h" //generated
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/include/xed-nonterminal-enum.h

@@ -1,194 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/include/xed-operand-ctype-map.h

@@ -1,41 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-operand-ctype-map.h
-
-// This file was automatically generated.
-// Do not edit this file.
-
-#if !defined(_XED_OPERAND_CTYPE_MAP_H_)
-# define _XED_OPERAND_CTYPE_MAP_H_
-#include "xed-internal-header.h"
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/include/xed-operand-enum.h

@@ -1,151 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/include/xed-operand-type-enum.h

@@ -1,54 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-operand-type-enum.h
-
-// This file was automatically generated.
-// Do not edit this file.
-
-#if !defined(_XED_OPERAND_TYPE_ENUM_H_)
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/include/xed-operand-width-enum.h

@@ -1,126 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-ia32/include/xed-reg-class.h

@@ -1,60 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-reg-class.h
-/// @author Mark Charney <mark.charney@intel.com>
-
-#ifndef _XED_REG_CLASS_H_
-# define  _XED_REG_CLASS_H_
-
-#include "xed-types.h"
-#include "xed-reg-enum.h" // a generated file
-#include "xed-reg-class-enum.h" // a generated file
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/examples/udhelp.H

@@ -1,86 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#if !defined(_UDHELP_H_)
-#define _UDHELP_H_
-
-#if defined(_MSC_VER)  && defined(XED_DBGHELP)
-// only try to use dbghelp on MSVS8 (2005) or later versions.
-# if _MSC_VER >= 1400
-#  define XED_USING_DEBUG_HELP
-# endif
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/examples/xed-ex-agen.c

@@ -1,308 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-ex-agen.c  
-/// @author Mark Charney   <mark.charney@intel.com>
-
-// decoder example with agen callbacks.
-
-#include "xed-interface.h"
-#include "xed-examples-util.h"
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h> //strcmp
-#include <assert.h> 
-
-int main(int argc, char** argv);
-
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/examples/xed-ex1.cpp

@@ -1,454 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-ex1.cpp
-/// @author Mark Charney   <mark.charney@intel.com>
-
-extern "C" {
-#include "xed-interface.h"
-}
-#include "xed-examples-ostreams.H"
-#include <iostream>
-#include <iomanip>
-#include <sstream>
-#include <cassert>
-#include <cstring>
-#include <cstdlib>
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/examples/xed-ex6.c

@@ -1,177 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-ex6.c
-/// @author Mark Charney   <mark.charney@intel.com>
-
-// decoder->encoder example. 
-
-#include "xed-interface.h"
-#include "xed-examples-util.h"
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h> //strcmp
-#include <assert.h> 
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/examples/xed-min.c

@@ -1,82 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-min.c
-/// @author Mark Charney   <mark.charney@intel.com>
-/// @brief a minimal toy example of using the decoder. 
-
-#include "xed-interface.h"
-#include <stdio.h>
-
-int main(int argc, char** argv);
-
-int main(int argc, char** argv) {
-    xed_machine_mode_enum_t mmode;
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-common-hdrs.h

@@ -1,74 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-common-hdrs.h
-/// @author  Mark Charney   <mark.charney@intel.com>
-
-
-
-#ifndef _XED_COMMON_HDRS_H_
-# define _XED_COMMON_HDRS_H_
-
-////////////////////////////////////////////////////////////////////////////
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-encoder-gen-defs.h

@@ -1,40 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-encoder-gen-defs.h
-
-// This file was automatically generated.
-// Do not edit this file.
-
-#if !defined(_XED_ENCODER_GEN_DEFS_H_)
-# define _XED_ENCODER_GEN_DEFS_H_
-#define XED_ENCODE_ORDER_MAX_ENTRIES  28
-#define XED_ENCODE_ORDER_MAX_OPERANDS 4
-#endif

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-flag-action-enum.h

@@ -1,56 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-flag-action-enum.h
-
-// This file was automatically generated.
-// Do not edit this file.
-
-#if !defined(_XED_FLAG_ACTION_ENUM_H_)
-# define _XED_FLAG_ACTION_ENUM_H_
-#include "xed-common-hdrs.h"
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-gen-table-defs.h

@@ -1,46 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-gen-table-defs.h
-
-// This file was automatically generated.
-// Do not edit this file.
-
-#if !defined(_XED_GEN_TABLE_DEFS_H_)
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-iclass-enum.h

@@ -1,963 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-iform-enum.h

@@ -1,2438 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-iformfl-enum.h

@@ -1,1861 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-init-pointer-names.h

@@ -1,40 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-init-pointer-names.h
-
-// This file was automatically generated.
-// Do not edit this file.
-
-#if !defined(_XED_INIT_POINTER_NAMES_H_)
-# define _XED_INIT_POINTER_NAMES_H_
-#include "xed-internal-header.h"
-#define XED_MAX_POINTER_NAMES 33
-#endif

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-operand-storage.h

@@ -1,49 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-operand-storage.h 
-/// @author Mark Charney   <mark.charney@intel.com> 
-
-
-
-#ifndef _XED_OPERAND_STORAGE_H_
-# define _XED_OPERAND_STORAGE_H_
-
-#include "xed-common-hdrs.h"
-#include "xed-types.h"
-#include "xed-portability.h"
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-operand-values-interface.h

@@ -1,518 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-operand-values-interface.h
-/// @author Mark Charney <mark.charney@intel.com>
-
-#if !defined(_XED_OPERAND_VALUES_INTERFACE_H_)
-# define _XED_OPERAND_VALUES_INTERFACE_H_
-
-#include "xed-common-hdrs.h"
-#include "xed-common-defs.h"
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-reg-role-enum.h

@@ -1,55 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-reg-role-enum.h
-
-// This file was automatically generated.
-// Do not edit this file.
-
-#if !defined(_XED_REG_ROLE_ENUM_H_)
-# define _XED_REG_ROLE_ENUM_H_
-#include "xed-common-hdrs.h"
-typedef enum {
-  XED_REG_ROLE_INVALID,
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/extras/xed2-intel64/include/xed-util.h

@@ -1,269 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file xed-util.h 
-/// @author Mark Charney   <mark.charney@intel.com> 
-
-
-
-#ifndef _XED_UTIL_H_
-# define _XED_UTIL_H_
-
-#include "xed-common-hdrs.h"
-#include "xed-types.h"
-#include "xed-portability.h"
-
-  
-////////////////////////////////////////////////////////////////////////////
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/include/pin_isa.H

@@ -1,64 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-//
-// @ORIGINAL_AUTHOR: Artur Klauser
-//
-
-/*! @file
- *  This file contains the ISA-dependent PIN iterface specification
- */
-
-#ifndef PIN_ISA_H
-#define PIN_ISA_H
-
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/include/pinapp.h

@@ -1,48 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#ifdef PINAPP_H
-#error duplicate inclusion
-#else
-#define PINAPP_H
-#endif
-
-#if defined(__cplusplus)
-extern "C" 
-{
-#endif
-
-extern void * PIN_NewThread();
-extern void PIN_ExecuteInstrumented(void *);
-extern void PIN_ExecuteUninstrumented();
-
-#if defined(__cplusplus)
-}
-#endif

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AVX/context_ex.h

@@ -1,191 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*  context_ex.h
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AVX/raise_exception.cpp

@@ -1,222 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*! @file
- * This test tool verifies correctness of the PIN_RaiseException 
- * API implementation.
- */
-
-#include "pin.H"
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AVX/set_ymm_scratches.cpp

@@ -1,84 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-
-#include "pin.H"
-#include <stdio.h>
-
-
-KNOB<UINT32> KnobLastInsToInstrument(KNOB_MODE_WRITEONCE, "pintool",
-    "last_ins", "100000000", "LastInsToInstrument");
-KNOB<UINT32> KnobFirstInsToInstrument(KNOB_MODE_WRITEONCE, "pintool",
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AVX/verify_fp_context_in_xrstor.cpp

@@ -1,561 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include <stdio.h>
-#include "pin.H"
-#include "instlib.H"
-#include "portability.H"
-#include <iostream>
-#include <sstream>
-#include <iomanip>
-#include <fstream>
-#include <cstdlib>
-#if defined( __GNUC__)
-
-//typedef unsigned short UINT16;
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AVX/winapp_exception_context.cpp

@@ -1,113 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*! @file
- *  An example of Windows application that raises exception and verifies
- *  the FP/XMM state. 
- */
-#include <windows.h>
-#include <string>
-#include <iostream>
-#include <memory.h>
-
-using namespace std;
-
-typedef unsigned __int8 UINT8 ;
-typedef unsigned __int16 UINT16;
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AVX/ymm_apps_stuff.h

@@ -1,289 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#ifdef YMM_APPS_STUFF_H
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AlignChk/alignchk.cpp

@@ -1,64 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/// @file alignchk.cpp
-/// @author Mark Charney
-
-#include "pin.H"
-#include <iostream>
-#include <iomanip>
-#include <stdlib.h>
-
-#include "alignchk.H"
-void
-usage()
-{
-    cerr << KNOB_BASE::StringKnobSummary();
-    cerr << endl;
-}
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/attach_in_sighandler_app.cpp

@@ -1,366 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-// @ORIGINAL_AUTHOR: Elena Demikhovsky
-
-/*! @file
- *  This application attaches Pin in the middle of signal handler.
- * The test verifies RT signal frame save/restore inside Pin
- */
-#include <assert.h>
-#include <stdio.h>
-#include <dlfcn.h>
-#include <signal.h>
-#include <sys/types.h>
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/detach_probed_tool.cpp

@@ -1,96 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-
-/* ===================================================================== */
-/*
-  @ORIGINAL_AUTHOR: Elena Demikhovsky
-*/
-
-/* ===================================================================== */
-/*! @file
- */
-
-#include "pin.H"
-#include <iostream>
-#include <fstream>
-#include <stdlib.h>
-#include <assert.h>
-
-using namespace std;
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/jit_instr_detach.cpp

@@ -1,102 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include <stdio.h>
-#include "pin.H"
-#include <iostream>
-#include <assert.h>
-
-using namespace std;
-
-UINT64 icount = 0;
-
-
-UINT32 insCountBeforeDetach = 0;
-UINT32 insMax = 100;
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/mt_fork_app.cpp

@@ -1,94 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include <stdio.h>
-#include <stdlib.h>
-#include <pthread.h>
-#include <unistd.h>
-#include <sys/wait.h>
-
-#define NUM_OF_THREADS 5
-
-void RunApp(char *app)
-{
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/my_exe.c

@@ -1,38 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include <stdio.h>
-
-int main()
-{
-	printf(".");
-	return 0;
-}
-

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/reattach_probed_tool.cpp

@@ -1,403 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-
-/* ===================================================================== */
-/*
-  @ORIGINAL_AUTHOR: Elena Demikhovsky
-*/
-
-/* ===================================================================== */
-/*! @file
- * Probe mode detach-reattach test
- */
-
-#include "pin.H"
-#include <iostream>
-#include <fstream>
-#include <stdlib.h>
-#include <assert.h>
-#include <sched.h>
-#include <linux/unistd.h>
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/tls_check_tool.cpp

@@ -1,95 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-
-/* ===================================================================== */
-/*
-  @ORIGINAL_AUTHOR: Elena Demikhovsky
-*/
-
-/* ===================================================================== */
-/*! @file
- */
-
-#include "pin.H"
-#include <iostream>
-#include <fstream>
-#include <stdlib.h>
-#include <assert.h>
-
-using namespace std;
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/verify_fpstate_app.cpp

@@ -1,428 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-// @ORIGINAL_AUTHOR: Elena Demikhovsky
-
-/*! @file
- *  The test verifies that FP state of thread is not changed by Pin injection
- */
-#include <assert.h>
-#include <stdio.h>
-#include <string.h>
-#include <sys/types.h>
-#include <linux/unistd.h>
-#include <pthread.h>
-#include <sys/wait.h>
-#include <stdlib.h>
-#include <string>
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/w_attach_tool1.cpp

@@ -1,379 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include "pin.H"
-#include <iostream>
-using namespace std;
-
-namespace WIND
-{
-#include <windows.h>
-}
-
-/* ===================================================================== */
-/* Commandline Switches */
-/* ===================================================================== */
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/AttachDetach/w_service_app1.cpp

@@ -1,539 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#define UNICODE
-#define _CRT_SECURE_NO_WARNINGS
-#define _CRT_SECURE_NO_DEPRECATE
-
-#include <Windows.h>
-#include <Winsvc.h>
-#include <string>
-#include <iostream>
-#include <fstream>
-using namespace std;
-
-static wchar_t * ServiceNameW         = L"ServiceAppForPinTesting";
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/Buffer/buffer-predicate.cpp

@@ -1,302 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*
- * Sample buffering tool
- * 
- * This tool collects an address trace of instructions that access memory
- * by filling a buffer.  When the buffer overflows,the callback writes all
- * of the collected records to a file.
- *
- */
-#include <iostream>
-#include <fstream>
-#include <stdlib.h>
-
-#include "pin.H"
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/Buffer/buffererror.cpp

@@ -1,195 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*
- * Negative test - Do Not Use!!!
- * 
- */
-#include "pin.H"
-#include <iostream>
-#include <stdio.h>
-
-/* Struct for holding memory references.  Rather than having two separate
- * buffers for loads and stores, we just use one struct that includes a
- * flag for type.
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CacheClient/action_pending_cachefull.cpp

@@ -1,155 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*
- * This test only runs with the application "action-pending-app.cpp".  It helps
- * test the PIN_IsActionPending() API.
- */
- 
-#include <stdio.h> 
-#include <stdlib.h>
-
-#include "pin.H"
-#include "utils.H"
-
-#ifdef TARGET_WINDOWS
-namespace WIND
-{
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CacheClient/action_pending_cachefull_app.c

@@ -1,200 +1,0 @@
-/*NO LEGAL*/
-
-/*
- * This test application is only interesting when run with the tool
- * action_pending_cachefull_tool.cpp. It has two threads, one thread spins in
- * a tool replacement function (non-jitted code), while the other executes
- * a large amount of code in order to cause a high watermark hit in the code
- * cache   
- */
-
-#include <stdio.h>
-#include "../threadlib/threadlib.h"
-#ifndef TARGET_WINDOWS
-#include <unistd.h>
-#endif
-
-static void *Waiter(void *);
-static void *Generator(void *);
-#ifdef TARGET_WINDOWS
-__declspec(dllexport) 
-#else
-extern "C"
-#endif
-void ToolWait();
-
-static int waiterRunning = 0;
-
-
-#define FUNS(name)     \
-    void name()        \
-    {                    \
-        int i;           \
-        for (i = 0; i < n; i++)                 \
-        {                                       \
-            a[i] = i + 2;                       \
-        }                                       \
-    }                                           \
-
-#define FUNS10(number)  \
-    FUNS(number##0)     \
-    FUNS(number##1)     \
-    FUNS(number##2)     \
-    FUNS(number##3)     \
-    FUNS(number##4)     \
-    FUNS(number##5)     \
-    FUNS(number##6)     \
-    FUNS(number##7)     \
-    FUNS(number##8)     \
-    FUNS(number##9)
-
-#define FUNS100(number)  \
-    FUNS10(number##0)     \
-    FUNS10(number##1)     \
-    FUNS10(number##2)     \
-    FUNS10(number##3)     \
-    FUNS10(number##4)     \
-    FUNS10(number##5)     \
-    FUNS10(number##6)     \
-    FUNS10(number##7)     \
-    FUNS10(number##8)     \
-    FUNS10(number##9)
-
-#define FUNS1000(number)  \
-    FUNS100(number##0)     \
-    FUNS100(number##1)     \
-    FUNS100(number##2)     \
-    FUNS100(number##3)     \
-    FUNS100(number##4)     \
-    FUNS100(number##5)     \
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CacheClient/br_test.cpp

@@ -1,154 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-//
-//  ORIGINAL_AUTHOR: Ady Tal
-//
-//  This tool test the big-region binary. We know that the region
-//  is about > 1000 instructions. Since we want to create a region
-//  larger than 64K, we create inlined analysis routine larger 
-//  than 50 bytes for each instructions.
-//  Use fat instcoun.
-
-//  Sample usage:
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CacheClient/deleteTrace.cpp

@@ -1,202 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-//
-//  ORIGINAL_AUTHOR: Kim Hazelwood
-//
-//  This tool deletes a hot program addresses from the code cache
-//      from inside an analysis routine.
-//  Sample usage:
-//    pin -t deleteTrace -- /bin/ls
-
-#include "pin.H"
-#include "utils.H"
-#include "portability.H"
-#include <iostream>
-#include <fstream>
-
-using namespace std;
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CacheClient/flushAndInsert.cpp

@@ -1,158 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-//
-//  ORIGINAL_AUTHOR: Kim Hazelwood
-//
-//  This tool tells you what traces reappear after a code cache
-//      flush. It can be used to identify long-lived traces in a program.
-//      Note: you must limit the code cache size to trigger flushes
-//  Sample usage:
-//    pin -xyzzy -limit_code_cache 131072 -t flushAndInsert -- /bin/ls
-
-#include "pin.H"
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CacheClient/flush_at_if.cpp

@@ -1,110 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*
- * This  tool tests calling a THEN instrumentation function when the if instrumentation function 
-   intiates a code-cache flush
- */
-
-#include "pin.H"
-#include <stdio.h>
-#include <stdlib.h>
-
-int numFlushes = 0;
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CacheClient/utils.cpp

@@ -1,85 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-//
-//  ORIGINAL_AUTHOR: Kim Hazelwood
-//
-//  This is a utility file that contains useful routines for 
-//      all cache client tools
-
-#include "pin.H"
-#include "utils.H"
-#include <iostream>
-
-using namespace std;
-
-/* ================================================================== */
-/*
-  Convert an unsigned integer (representing bytes) into a string 
-  that uses KB or MB as necessary
-*/
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CacheFilter/log-util.H

@@ -1,48 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-// log-util.H
-// Mark Charney   <mark.charney@intel.com>
-// $Id: log-util.H,v 1.2 2003/05/06 20:05:40 rscohn1 Exp $
-
-
-#ifndef _LOG_UTIL_H_
-# define _LOG_UTIL_H_
-
-#include "my-types.H"
-
-UINT ilog(UINT arg);
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CacheFilter/my-types.H

@@ -1,78 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-// my-types.H
-// Mark Charney   <mark.charney@intel.com>
-// $Id: my-types.H,v 1.2 2003/05/06 20:05:40 rscohn1 Exp $
-
-
-#ifndef _MY_TYPES_H_
-# define _MY_TYPES_H_
-
-//#include "types_base.H"
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/arglist.h

@@ -1,59 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#ifndef _ARGLIST_H_
-#define _ARGLIST_H_
-
-#include <string>
-#include <list>
-
-using namespace std;
-
-class ARGUMENTS_LIST
-{
-  public:
-    ARGUMENTS_LIST():m_argv(0) {}
-    ARGUMENTS_LIST(int argc, const char* const*argv);
-    ~ARGUMENTS_LIST();
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/bind_now_tool.cpp

@@ -1,89 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/* ===================================================================== */
-/*
-  @ORIGINAL_AUTHOR: Ady Tal
- */
-
-/* ===================================================================== */
-/*! @file
- * A test for checking if LD_BIND_NOW is set in pintool and not in the application.
- */
-
-#include <stdlib.h>
-#include <stdio.h>
-#include "pin.H"
-
-/* ===================================================================== */
-/* Command line Switches */
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/execvp.c

@@ -1,47 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include <stdio.h>
-#include <sys/types.h>
-#include <unistd.h>
-#include <sys/types.h>
-#include <sys/wait.h>
-#include <errno.h>
-
-extern int errno;
-
-extern **environ;
-
-int main(int argc, char **argv){
-    char * args[] = {"ls", argv[0], NULL};
-    
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/follow_child_3gen_tool_initsym.cpp

@@ -1,63 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*
-This tool is used to output that the child process was created in the correct mode 
-(probe or jit)
-It does PIN_InitSymbols
-*/
-#include "pin.H"
-#include <iostream>
-
-using namespace std;
-
-
-
-int main(INT32 argc, CHAR **argv)
-{
-    PIN_InitSymbols();
-    PIN_Init(argc, argv);
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/fork_app.c

@@ -1,53 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include <stdio.h>
-#include <sys/types.h>
-#include <sys/wait.h>
-#include <unistd.h>
-#include <pthread.h>
-
-
-int main()
-{
-
-    pid_t child_id = fork();
-	if (child_id == 0)
-	{
-		printf("APPLICATION: After fork in child\n");
-	}
-	else
-	{
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/parent_process.cpp

@@ -1,69 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-// Application that creates new process
-
-#include <sys/types.h>
-#include <sys/wait.h>
-#include <unistd.h>
-#include <iostream>
-#include <string>
-using namespace std;
-
-//Wait for a process completion
-//Verify it returned the expected exit code
-
-int main(int argc, char * argv[])
-{
-    char *childArgvArray[5];
-    childArgvArray[0] = argv[1];
-    childArgvArray[1] = argv[2];
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/parent_tool.cpp

@@ -1,157 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include "pin.H"
-#include <iostream>
-using namespace std;
-
-namespace WIND
-{
-#include <windows.h>
-}
-/* ===================================================================== */
-/* Command line Switches */
-/* ===================================================================== */
-
-//General configuration - Pin and Tool full path
-KNOB<string> KnobPinFullPath(KNOB_MODE_WRITEONCE,         "pintool",
-                             "pin_path", "", "pin full path");
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/simple_tool.cpp

@@ -1,58 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include "pin.H"
-#include <iostream>
-using namespace std;
-
-/* ===================================================================== */
-/* Command line Switches */
-/* ===================================================================== */
-
-KNOB<BOOL> KnobToolProbeMode(KNOB_MODE_WRITEONCE, "pintool", "probe", "0", "invoke tool in probe mode");
-
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/unix_parent_tool.cpp

@@ -1,140 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include "pin.H"
-#include <iostream>
-#include "arglist.h"
-
-/* ===================================================================== */
-/* Command line Switches */
-/* ===================================================================== */
-
-//Pin command line
-KNOB<string> KnobPin(KNOB_MODE_WRITEONCE, "pintool", "pin", "", "pin full path");
-
-//Parent configuration - Application name
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/win_launcher_debugged_process.cpp

@@ -1,138 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-// Application that creates new process
-
-#include <Windows.h>
-#include <iostream>
-#include <string>
-using namespace std;
-
-//Wait for a process completion
-//Verify it returned the expected exit code
-bool WaitAndVerify(HANDLE process, HANDLE thread, DWORD pid)
-{
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/win_parent_process_mt.cpp

@@ -1,241 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-// MT Application that creates new processes, from multiple threads simultaneously
-
-#include <Windows.h>
-#include <iostream>
-#include <string>
-using namespace std;
-
-const LONG numthreads = 32;
-
-LONG volatile numThreadsArrived = 0;
-
-HANDLE threadReachedSynchPointSemHandle[numthreads];
-CRITICAL_SECTION critSec;
-
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/ChildProcess/win_parent_process_unicode.cpp

@@ -1,123 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-// Application that creates new process
-#define UNICODE
-#include <Windows.h>
-#include <iostream>
-#include <string>
-using namespace std;
-
-//Wait for a process completion
-//Verify it returned the expected exit code
-bool WaitAndVerify(HANDLE process)
-{
-    if(WaitForSingleObject( process, INFINITE ) == WAIT_FAILED)
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CrossIa32Intel64/follow_child_tool.cpp

@@ -1,63 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include "pin.H"
-#include <iostream>
-using namespace std;
-
-BOOL FollowChild(CHILD_PROCESS childProcess, VOID * userData)
-{
-    cout << "At follow child callback" << endl << flush;
-    return TRUE;
-}
-
-
-
-int main(INT32 argc, CHAR **argv)
-{
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CrossIa32Intel64/follow_child_tool_sym.cpp

@@ -1,68 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include "pin.H"
-#include <iostream>
-using namespace std;
-
-BOOL FollowChild(CHILD_PROCESS childProcess, VOID * userData)
-{
-
-    cout << "At follow child callback" << endl;
-    return TRUE;
-}
-
-
-
-int main(INT32 argc, CHAR **argv)
-{
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/CrossIa32Intel64/w_child_process.cpp

@@ -1,98 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-// Application that creates new process
-
-#include <Windows.h>
-#include <iostream>
-#include <string>
-using namespace std;
-
-//Wait for a process completion
-//Verify it returned the expected exit code
-bool WaitAndVerify(HANDLE process)
-{
-    if(WaitForSingleObject( process, INFINITE ) == WAIT_FAILED)
-    {
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/DeProf/deprof.cpp

@@ -1,369 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include <assert.h>
-#include <stdio.h>
-#include <iostream>
-#include <fstream>
-#include <map>
-#include <set>
-#include "pin.H"
-
-#include <unistd.h>
-#include <sys/times.h>
-
-LOCALVAR ADDRINT GlobalLow;
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/DebugTrace/debugtrace.cpp

@@ -1,860 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-
-/* ===================================================================== */
-/*
-  @ORIGINAL_AUTHOR: Robert Cohn
-*/
-
-/* ===================================================================== */
-/*! @file
- *  This file contains a tool that generates instructions traces with values.
- *  It is designed to help debugging.
- */
-
-
-
-#include "pin.H"
-#include "instlib.H"
-#include "portability.H"
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/Debugger/breaktool.cpp

@@ -1,84 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include <fstream>
-#include <iostream>
-#include "pin.H"
-
-KNOB<BOOL> KnobWaitForDebugger(KNOB_MODE_WRITEONCE, "pintool",
-    "wait_for_debugger", "0", "Wait for debugger to connect");
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/Debugger/fork.cpp

@@ -1,25 +1,0 @@
-/*NO LEGAL*/
-
-#include <sys/types.h>
-#include <sys/wait.h>
-#include <unistd.h>
-
-static void DoChild();
-
-int main()
-{
-    pid_t pid = fork();
-    if (pid != 0)
-        waitpid(pid, 0, 0);
-    else
-        DoChild();
-    return 0;
-}
-
-
-int Glob = 0;
-
-static void DoChild()
-{
-    Glob++;
-}

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/Debugger/intercept-tool.cpp

@@ -1,156 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*
- * This tool only runs with the application "intercept-app.cpp".  It's part
- * of a test for PIN_InterceptDebuggingEvent().
- */
- 
-#include <iostream>
-#include <cstdlib>
-#include "pin.H"
-#include "memlog.hpp"
-
-static void InstrumentRtn(RTN, VOID *);
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/Debugger/interpreterRemove.cpp

@@ -1,65 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*
-  @ORIGINAL_AUTHOR: Daniel Lemel
- */
-
-/*! @file
- *  Check the PIN_RemoveDebugInterpreter interface.
- */
-
-#include <iostream>
-#include "pin.H"
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/Debugger/memlog.hpp

@@ -1,72 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-/*
- * @ORIGINAL_AUTHOR: Greg Lueck
- *
- */
-
-#ifndef MEMLOG_HPP
-#define MEMLOG_HPP
-
-#include <map>
-
-
-class MEMLOG
-{
-public:
-    void Record(ADDRINT addr, ADDRINT size)
-    {
-        UINT8 *ptr = reinterpret_cast<UINT8 *>(addr);
-
[…]

### 2013-02-14 tools/pin-2.8-37300-gcc.3.4.6-ia32_intel64-linux/source/tools/I18N/win_unicode.cpp

@@ -1,61 +1,0 @@
-/*BEGIN_LEGAL 
-Intel Open Source License 
-
-Copyright (c) 2002-2010 Intel Corporation. All rights reserved.
- 
-Redistribution and use in source and binary forms, with or without
-modification, are permitted provided that the following conditions are
-met:
-
-Redistributions of source code must retain the above copyright notice,
-this list of conditions and the following disclaimer.  Redistributions
-in binary form must reproduce the above copyright notice, this list of
-conditions and the following disclaimer in the documentation and/or
-other materials provided with the distribution.  Neither the name of
-the Intel Corporation nor the names of its contributors may be used to
-endorse or promote products derived from this software without
-specific prior written permission.
- 
-THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
-``AS IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
-LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
-A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE INTEL OR
-ITS CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
-SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
-LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
-DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
-THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
-(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
-OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-END_LEGAL */
-#include <iostream>
-#include <fstream>
-using namespace std;
-
-int wmain(int argc, wchar_t * argv[])
-{
-    std::ofstream file;
-    file.open("win_unicode.out");
-    //internationalization in Japanese (encoded in UTF-16)
-    wchar_t internationalization[] = {0x56fd, 0x969B, 0x5316, 0x0000};
-
-    if(argc == 1)
-    {
[…]

### 2013-02-14 doc/xps2013/proposal.tex
Incorporated Preston's suggestions 	modified:   proposal.bib 	modified:   proposal.tex
fl

@@
 
-% Most importantly, however, for a much lower real world cost,
-% users can \emph{add} significantly more processors to a commodity cluster than
-% an XMT machine and use Grappa to achieve scalable performance.
-
 Given these very positive early results, this proposal plans to extend Grappa
@@
 
+%%%%%%%%%%%%%% Tempting to delete from here....
+
 Finally, we propose to explore how the lessons learned performing this research -- specifically knowledge of what ultimately limits Grappa's performance for irregular applications using x86 processors and commodity networks with modified communication protocols -- can be applied to the design of future architectures.
@@
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%  To here.... -Mark
 
@@
 The first is a distributed heap striped across all the machines in the system in a block cyclic fashion.
-The \texttt{global\_malloc} and \texttt{global\_free} calls are used to allocate and deallocate memory in the global heap.
 Addresses to memory in the global heap use \textbf{linear addresses}.
@@
 This address can then be accessed with a delegate.
-% and can also be cached by other nodes. 
 At the destination the address is converted into a canonical x86 address by replacing the upper bits with the sign-extended upper bit of the virtual address.
@@
 Delegate operations, proposed in \cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the primary synchronization method in Grappa.
-
-% Delegate operations are always executed at the home core of their address, and
-% while arbitrary memory operations can be delegated, we restrict the use of
-% delegate operations in three ways to make them more useful for
-% synchronization. First, we limit each task to one outstanding delegate
-% operation to avoid the possibility of reordering in the network. Second, we
-% limit delegate operations to operate on objects in the 2D address space or
-% objects that fit in a single block of the linear address space, so they can be
-% satisfied with a single network request. Finally, no context switches are
-% allowed while the data is being modified. Given these restrictions, we can
-% ensure that delegate operations for the same address from multiple requesters
-% are always serialized through a single core in the system, providing atomic
-% semantics without using atomic operations.
 
@@
 
-% The network layer is serviced by polling. Periodically when a context switch
-% occurs, the Grappa scheduler switches to the network polling thread. This
-% thread has three responsibilities. First, it polls the lower-level network
-% layer to ensure it makes progress. Second, it deaggregates received messages
-% and executes active message handlers. Third, it checks to see if any
-% aggregation queues have messages that have been waiting longer than the
-% threshold; if so, it sends them.
-
 Underneath the aggregation layer, Grappa uses the \gasnet~communication
@@
 
-Some networks provide access to a remote machine's memory
-directly. This would seem to be a good fit for a programming model
-focused on global shared memory, but in fact we do not use it. In our
-experiments, we found that RDMA operations are subject to the same
-message rate limitations as all other messages on these cards, and
-thus using raw RDMA operations for our small messages would make
-inefficient use of bandwidth. Namely, we would be limited to the
-network injection rate providing just a few million accesses per
-second per node.  Instead, we implement remote memory operations with
-active messages in conjuction with aggregation, providing an order of
-magnitude or more increase in throughput. A byproduct of this design
-decision is that Grappa is not limited to RDMA-capable networks.
+Some networks provide access to a remote machine's memory directly.
+This would seem to be a good fit for a programming model focused on global shared memory, however, we do not currently use it.
+In our experiments, we found that RDMA operations are subject to the same message rate limitations as all other messages on these cards, and thus using raw RDMA operations does not measurably increase achieved bandwidth (i.e., we are limited by injection rate not peak bandwidth).
+At the moment, we implement remote memory operations with active messages in conjunction with aggregation, providing an order of magnitude or more increase in throughput compared to straightforward RDMA.
+As we describe later, however, we wish to explore richer semantics for RDMA communication that will make it integrate better with user level runtime systems such as Grappa.
 
@@
 embodies this important trend. There are currently five graduate students for
-which last two of the PIs serve as advisors. In addition to the project
+which the last two of the PIs serve as advisors. In addition to the project
 described in this proposal, they are working together on techniques for

### 2013-02-14 doc/xps2013/proposal.tex
intro fine pass

@@
 
-\paragraph{Irregular applications exhibit little spatial locality.} It is not atypical for any given task's data references to be spread randomly across the entire memory of the system.
-This makes current memory hierarchy features ineffective.
-Caches are of little assistance with such low data re-use and spatial locality.
-Commodity prefetching hardware is effective only when addresses are known many cycles before the data is consumed or the accesses follow a predictable pattern, neither of which occurs in irregular applications.
-As a consequence, commodity microprocessors stall often when executing irregular applications.
+\paragraph{Irregular applications exhibit little spatial locality.} It is not
+atypical for any given task's data references to be spread randomly across the
+entire memory of the system. This makes current memory hierarchy features
+ineffective. Caches are of little assistance with such low data re-use and
+spatial locality. Commodity prefetching hardware is effective only when
+addresses are known many cycles before the data is consumed or the accesses
+follow a predictable pattern, neither of which occurs in irregular
+applications. As a consequence, commodity microprocessors stall often when
+executing irregular applications.
 
-\paragraph{Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.
-Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.
-Injection rate into the network is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.
+\paragraph{Irregular applications frequently request small amounts of off-node
+data.} On multinode systems (e.g., clusters of computers), the challenges
+presented by low locality are analogous, and exacerbated by the increased
+latency of going off-node. Irregular applications also present a challenge to
+commodity network technology, which is designed to transfer large blocks of
+data, not the word-sized references emitted by irregular application tasks.
+Injection rate into the network is insufficient to utilize wire bandwidth when
+blocks are below about two-kilobytes, so any straightforward communication
+strategy severely under-utilizes the network.
 
@@
 revealed limitations of the current Grappa runtime system with regards to
-network scaling. This proposal plans to enhance the network layer with new
-forms of communication scheduling and aggregation, and explore new forms of
-hardware/software interactions of remote direct memory access (RDMA).
+network scaling. This proposal plans to enhance the network layer with whole
+new forms of communication scheduling and aggregation, and explore new forms
+of hardware/software interactions of remote direct memory access (RDMA).
 
 \item \textbf{Language and compiler support:} Presently, Grappa is implemented
-strictly as a library.  Programmers must express remote operations explicitly as function calls.
-This adds burden high-level language programmers do not usually face and obfuscates opportunities
-for compiler optimizations. A significant research thrust of the proposed work is to use
-language and compilation support to provide a cleaner programming model and use
+strictly as a library. Programmers must express remote operations explicitly
+as function calls. This adds burden high-level language programmers do not
+usually face and obfuscates opportunities for compiler optimizations. A
+significant research thrust of the proposed work is to use language and
+compilation support to provide a cleaner programming model and use
 statically-available knowledge to lessen the runtime overhead automatically.
@@
 
-\item \textbf{Support for \ssd\ only memory systems:} The size of data being
+\item \textbf{Support for \ssd\ only data structures:} The size of data being
 processed by large systems today (hundreds of terabytes) dwarfs what can fit
 into a single system. Multiple system nodes are required to maintain and
-access all of this data efficiently. However, even with hundreds of channels
-to persistent storage, loading this data into DRAM can prove a bottleneck. The
+access all of this data efficiently. However, given the lack of locality in the application, loading this data into DRAM is wasteful, in addition to adding overheads. The
 Grappa runtime is designed to make applications tolerate extremely high memory
-latency, thus we will explore using solid state disks as main memory, 
-further taking advantage of Grappa's latency tolerance properties.
+latency, thus we will explore having program data structures reside in solid state disks only.
+This further takes advantage of Grappa's latency tolerance properties to deal with larger data sets and reduce system cost.
 
@@
 
-Efficiently executing irregular computations on commodity hardware is \emph{hard}.
-Much about commodity hardware is seemingly designed for the opposite application characteristics: caches assume applications have locality; networks are built with low injection rates but high bandwidth, assuming large packet sizes; even runtimes and operating systems are built assuming relatively infrequent communication and synchronization compared to computation.
-This proposal rests on the seemingly outlandish claim that this commodity hardware / software stack can be coerced, through another layer of software, into efficiently executing irregular applications that both lack locality and communicate frequently with small amounts of data.
-A folksy way of saying it is we think we can jam a square peg through a round hole.
-Fortunately we have some preliminary evidence we won't get wedged along the way.
-This is an intellectually challenging endeavor, that we believe is well suited for the NSF XPS program.
-In summary, our goal is to exploit parallelism from irregular computations to efficiently execute them at scale.
+Efficiently executing irregular computations on commodity hardware is a
+\emph{difficult} problem. Much about commodity hardware is seemingly designed
+for the opposite application characteristics: caches assume applications have
+locality; networks are built with low injection rates but high bandwidth,
+assuming large packet sizes; even runtimes and operating systems are built
+assuming relatively infrequent communication and synchronization compared to
+computation. This proposal rests on the seemingly outlandish claim that this
+commodity hardware / software stack can be coerced, through another layer of
+software, into efficiently executing irregular applications that both lack
+locality and communicate frequently with small amounts of data. A folksy way
+of saying it is we think we can jam a square peg through a round hole.
+Fortunately we have some preliminary evidence we won't get wedged along the
+way. This is an intellectually challenging endeavor, that we believe is well
+suited for the NSF XPS program. In summary, our goal is to exploit parallelism
+from irregular computations to efficiently execute them at scale.
 
@@
 Furthermore, irregular applications will then reap the benefits of advances in commodity processors and networks, rather than requiring custom hardware that advances only at much higher cost.
-Graph queries, machine-learning and data-driven science will be enabled by the technology we aim to create from the ideas in this grant proposal.
+Graph queries, machine-learning and data-driven science will significantly benefit from the technology we aim to create from the ideas in this grant proposal.
 In addition our work will include the construction of a comprehensive irregular benchmark suite, something to help guide our own and the field's research.

### 2013-02-14 doc/xps2013/summary.tex

--- 2013-02-14 doc/xps2013/proposal.tex
@@
-%% NSF XPS Proposal 2013
+\paragraph{Summary}
 
-\input{header}
+Irregular applications are increasing in importance to the computing industry.
+For a long time they have underpinned computations of interest to the national security arm of the federal government.
+But now, they underpin such questions as ad placement in social networks, and analysis of complex data-sets in medicine and science.
+The defining characteristic of these applications is poor locality and massive available parallelism.
+Our effort is focused on making these applications perform well and scale on commodity hardware.
+The key idea that makes this work is to rely on massive concurrency to tolerate memory latency, instead of relying on locality (which doesn't exist for these applications).
+This core idea underpinned the XMT system, a fully-custom hardware system.
+We plan to explore that idea in software only, running on commodity processors.
+And unlike the XMT, however, we must also contend with commodity networking hardware, and here we again rely on threading to provide a sufficient number of requests that we can buffer them prior to dispatching on the network, to overcome bandwidth limitations when operating with small messages.
 
-\begin{center}
-  \textbf{\Large XPS-CLCCA: Scalable Parallelism for Irregular and Graph Applications}\\ \vspace{5mm}
-  \textbf{Luis Ceze ~~~~~ Simon Kahan ~~~~~ Mark Oskin (PI)}\\
-  \textbf{University of Washington}
-\end{center}
+Our preliminary effort has shown a workable proof of concept of this approach: a commodity cluster can out do the XMT at its own game, by emulating its key features in software.
+What is needed now is additional effort to improve the underlying performance of this runtime and make it usable to a broad base of users.
+We will be focusing our efforts in the next three years on the key performance component, the networking layer, and the key programmer efficiency linchpin, high level language support.
+In addition we need a useful mechanism for handling data-sets in the petabyte range, and here we imagine building a cluster of machines that relies primarily on SSD's and not DRAM as main memory data-set storage.
 
-\section{Introduction}
-
-Irregular applications generate tasks with work, dependences, and memory
-accesses that are highly sensitive to input. Classic examples of irregular
-applications include branch and bound optimization, SPICE circuit simulation,
-contact algorithms in car crash analysis, and network flow. Important
-contemporary examples include processing large graphs in the business,
-national security, machine learning, data-driven science, and social network
-computing domains. For these emerging applications, fast response -- given the
-sheer amount of data -- requires multinode systems. The most broadly available
-multinode systems are those built from x86 commodity computing nodes
-interconnected via ethernet or infiniband. Our focus is to enable scalable
-performance of irregular applications on these mass market systems,
-addressing two key challenges:
-
-\paragraph{Irregular applications exhibit little spatial locality.} It is not atypical for any given task's data references to be spread randomly across the entire memory of the system.
-This makes current memory hierarchy features ineffective.
-Caches are of little assistance with such low data re-use and spatial locality.
-Commodity prefetching hardware is effective only when addresses are known many cycles before the data is consumed or the accesses follow a predictable pattern, neither of which occurs in irregular applications.
-As a consequence, commodity microprocessors stall often when executing irregular applications.
-
-\paragraph{Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.
-Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.
-Injection rate into the network is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.
-
-Luckily, many of the important irregular applications naturally offer large
-amounts of concurrency. This immediately suggests taking advantage of
-concurrency to tolerate the latency of data movement. The fully custom Tera
-MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular
-applications by using concurrency to hide latencies. It had a large
-distributed shared memory with no caches. On every clock cycle, each processor
-would execute a ready instruction chosen from one of its 128 hardware thread
-contexts, a sufficient number to fully hide memory access latency. The network
-was designed with a single word injection rate that matched the processor
-clock frequency and sufficient bandwidth to sustain a reference from every
-processor on every clock cycle. Unfortunately, while an excellent match to
-extremely irregular applications, the MTA was not cost-effective on
-applications that could exploit locality and had very poor single-thread
-performance, making it a commercial failure. The Cray XMT approximates the
-Tera MTA-2, substantially reducing its cost but not overcoming its narrow
-range of applicability.
-
-In this proposal we first describe our early
-Grappa prototype, a runtime system for irregular
-applications. It provides several facilities: a global shared memory
-abstraction, user-level threading support, and communication and
-synchronization primitives. Applications targeted for Grappa are not expected
-to have \emph{any} locality (although what locality that exists is favorably
-exploited), but are expected to provide many thousands of concurrent threads.
-
-Next, we present some preliminary results. Even as an early prototype, Grappa
-runs several graph-crunching applications (classic examples of irregular
-behavior) efficiently on a commodity cluster. For unbalanced tree search,
-Grappa is over 3X faster and shows greatly improved scalability compared to
-the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower --
-fast compared to implementations without Grappa, but obviously there is room
-for improvement.
-
-Given these very positive early results, this proposal plans to extend Grappa
-in the following directions:
-
-\begin{itemize}
-
-\item \textbf{Reinvent network communication:} Our preliminary work has
-revealed limitations of the current Grappa runtime system with regards to
-network scaling. This proposal plans to enhance the network layer with new
-forms of communication scheduling and aggregation, and explore new forms of
-hardware/software interactions of remote direct memory access (RDMA).
-
-\item \textbf{Language and compiler support:} Presently, Grappa is implemented
-strictly as a library.  Programmers must express remote operations explicitly as function calls.
-This adds burden high-level language programmers do not usually face and obfuscates opportunities
-for compiler optimizations. A significant research thrust of the proposed work is to use
-language and compilation support to provide a cleaner programming model and use
-statically-available knowledge to lessen the runtime overhead automatically.
-
-\item \textbf{Comprehensive benchmark suite:} Presently there does not exist a
-comprehensive benchmark suite of irregular and graph applications. We
-are also seeking funding through this proposal to put together just
-such a benchmark suite. This suite will include benchmarks, reference
-implementations, and input data-sets. In addition a website will be
-created to distribute freely the benchmark suite and publish results.
-
-\item \textbf{Support for \ssd\ only memory systems:} The size of data being
-processed by large systems today (hundreds of terabytes) dwarfs what can fit
-into a single system. Multiple system nodes are required to maintain and
-access all of this data efficiently. However, even with hundreds of channels
-to persistent storage, loading this data into DRAM can prove a bottleneck. The
-Grappa runtime is designed to make applications tolerate extremely high memory
-latency, thus we will explore using solid state disks as main memory, 
-further taking advantage of Grappa's latency tolerance properties.
-
-\end{itemize}
-
-%%%%%%%%%%%%%% Tempting to delete from here....
-
-Finally, we propose to explore how the lessons learned performing this research -- specifically knowledge of what ultimately limits Grappa's performance for irregular applications using x86 processors and commodity networks with modified communication protocols -- can be applied to the design of future architectures.
-Doing so will include studying the following research questions:
-
-\begin{itemize}
-\item \textbf{What is the right system architecture support for irregular applications, including architecture, communication and storage?}
-\item \textbf{What are the necessary language and compiler support for large-scale, data-intensive irregular applications?}
-\item \textbf{How do we comprehensively evaluate performance of graph analytics systems?}
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%  To here.... -Mark
+If successful, the Grappa runtime will have a transformative impact on the computing industry.
+Irregular applications are hugely important, and if off the shelf hardware can be efficient with them a new class of questions can be asked of computing systems.
+Additionally our benchmark suite development will further drive innovation among researchers in this space.
 
@@
 
-Efficiently executing irregular computations on commodity hardware is \emph{hard}.
+Efficiently executing irregular computations on commodity hardware is challenging.
 Much about commodity hardware is seemingly designed for the opposite application characteristics: caches assume applications have locality; networks are built with low injection rates but high bandwidth, assuming large packet sizes; even runtimes and operating systems are built assuming relatively infrequent communication and synchronization compared to computation.
@@
 Graph queries, machine-learning and data-driven science will be enabled by the technology we aim to create from the ideas in this grant proposal.
-In addition our work will include the construction of a comprehensive irregular benchmark suite, something to help guide our own and the field's research.
-
-%% Roadmap
-
-\section{Description of the Current Grappa System}
-
-In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.
-This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating in software and inexpensive mass market hardware, the approach taken by Tera.
-We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.
-Ultimately, the opportunity is to cover the spectrum of irregular to regular computation: where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.
-Thus we make the best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.
-
-Grappa is a software runtime system that allows a commodity x86 distributed-memory HPC cluster to be programmed as if it were a single large shared-memory machine and provides scalable performance for irregular applications.
-Grappa is designed to smooth over some of the performance discontinuities in commodity hardware, giving good performance when there is little locality to be exploited while allowing the programmer to exploit it when it is available.
-
-Grappa leverages as much freely available and commodity infrastructure as possible.
-We use unmodified Linux for the operating system and an off-the-shelf user-mode infiniband device driver stack~\cite{OFED}.
-MPI is used for process setup and tear down.
-GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
-To this commodity hardware and software mix Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as few as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} and synchronizing operations including \emph{fetch-and-add\/}~\cite{fetchandadd} as well as user-defined atomic methods; and (3) a \emph{message aggregation\/} layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.
-As we will show later, Grappa can tolerate latencies way beyond that of the network.
-Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our effective random access memory bandwidth, our synchronization bandwidth, and our ability to improve load imbalance.
-
-\begin{figure}[t]
-\begin{center}     
-  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
-\begin{minipage}{0.95\columnwidth} 
-  \caption{\label{fig:grappa} Grappa system overview}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-The current Grappa prototype (Figure~\ref{fig:grappa}) has three main software components: \begin{description}
-
-\item [Tasking system.] Our tasking system supports lightweight multithreading to tolerate communication latency and global distributed workstealing (i.e., tasks can be stolen from any node in the system), which provides automated load balancing.
-
-\item[Distributed shared memory.] Our DSM system provides support for fine-grain access to data anywhere in the system.
-It supports synchronization operations on global data, explicit local caching of any memory in the system, and support for operation on remote data (delegating operations to home node).
-By tight integration with the tasking system and the communication layer, our DSM system offers high aggregate random access bandwidth for accessing remote data.
-
-\item[Communication layer.] As discussed earlier, modern commodity networks support high bandwidth only for large messages.
-Since irregular applications tend to need frequent communication of small requests, the main goal of our communication layer is to aggregate small messages into large ones to better exploit what the network can offer.
-It is largely invisible to the application programmer.
-
-\end{description}
-
-\subsection{Tasks}
-
-The basic unit of execution in Grappa is a {\em task}.
-Each task is represented by a function pointer and its arguments.
-A new task is enqueued at spawn time; when resources are free, it is allocated a stack, bound to a core, and executed.
-
-During execution, a task yields control of its core following initiation of any long-latency operation, allowing the processor to remain busy while waiting for the operation to complete.
-In addition,  a programmer can direct scheduling explicitly via the Grappa API calls.
-To minimize yield overhead, the Grappa scheduler operates entirely in user-space and does little more than store register state of one task and load that of another.
-Context switch times are as low as 38ns even when switching amongst thousands of tasks.
-
-\subsection{Expressing parallelism}
-
-Grappa programmers focus on expressing as much parallelism as possible without concern for where it will execute.
-Grappa then chooses where and when to exploit this parallelism, scheduling as much work as is necessary on each core to keep it busy in the presence of system latencies and task dependences.
-
-Grappa provides two methods for expressing parallelism.
-First, when the programmer identifies work that can be done in parallel, the work may be wrapped up in a function and queued with its arguments for later execution using a \texttt{spawn}.
-A programmer can further specify using \texttt{spawn\_on} to spawn a task on a specific core in the system or at the home core of a particular memory location.
-Second, the programmer can invoke a parallel for loop, provided that the trip count is known at loop entry.
-The programmer specifies a function pointer along with start and end indices and an optional threshold to control parallel overhead.
-Grappa does {\em recursive decomposition} of iterations, similar to Cilk's cilk\_for construct~\cite {cilkforimplementation} and TBB's {\tt parallel\_for}~\cite{intel_tbb}.
-It generates a logarithmically-deep tree of tasks, stopping to execute the loop body when the number of iterations is below a specified threshold.
-
-\subsection{Memory}
-
-Applications written for Grappa utilize two forms of memory: local and global.
-Tasks executing on the same core share an address space local to that core.
-The core accesses this local memory through conventional pointers, such as those generated by a standard C compiler.
-We nonetheless refer to these pointers as ``local pointers'' to differentiate them from ``global pointers'' that may be used by other cores to reference the same physical memory as described below.
-Applications use local memory for a number of things in Grappa: the stack associated with a task, localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.
-Local pointers cannot access memory on other cores, and are valid only on their home core.
-
-Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory.
-All global data must be accessed through Grappa's API.
-%, shown in Figure~\ref{fig:accessing-memory}.
-
-\paragraph{Global memory addressing} Grappa provides two methods for storing data in the global memory.
-The first is a distributed heap striped across all the machines in the system in a block cyclic fashion.
-Addresses to memory in the global heap use \textbf{linear addresses}.
-Choosing the block size involves trading off sequential bandwidth against aggregate random access bandwidth.
-Smaller block sizes help spread data across all the memory controllers in the cluster, but larger block sizes allow the locality-optimized memory controllers to provide increased sequential bandwidth.
-The block size, which is configurable, is typically set to 64 bytes, or the size of a single hardware cache line, in order to exploit spatial locality when available.
-The heap metadata is stored on a single node.
-Currently all heap operations serialize through this node; while this has been sufficient for our benchmarks, in the future Grappa will provide parallel performance through combining~\cite{MAMA,flatcombining}.
-
-Grappa also allows any local data on a core's stacks or heap to be exported to the global address space to be made accessible to other cores across the system.
-Addresses to global memory allocated in this way use \textbf{2D global addresses}.
-This uses a traditional PGAS addressing model, where each address is a tuple of a rank in the job (or global process ID) and an address in that process.
-The lower 48 bits of the address hold a virtual address in the process.
-The top bit is set to indicates that the reference is a 2D address (as opposed to linear address).
-This leaves 15 bits for network endpoint ID, which limits our scalability to $2^{15}$ endpoints.
-Any node-local data can be made accessible by other nodes in the system by wrapping the address and node ID into a 2D global address.
-This address can then be accessed with a delegate.
-At the destination the address is converted into a canonical x86 address by replacing the upper bits with the sign-extended upper bit of the virtual address.
-2D addresses may refer to memory allocated from a single processes' heap or from a task's stack.
-
-\paragraph{Global memory access} There are two general approaches Grappa applications use to {\emph access} global memory.
-When the programmer expects a computation on shared data to have spatial locality to exploit, explicit {\em cache} operations may be used.
-When there is no locality to exploit, remote {\em delegate} operations are used.
-
-\textbf{Explicit caching.} Grappa provides an API to fetch a global pointer of any length and return a local pointer to a cached copy of the global memory.
-Grappa cache operations have the usual read-only and read-write variants, along with a write-only variant used to initialize data structures.
-Languages for distributed shared memory systems have done optimizations to achieve a similar goal.
-For example, the UPC compiler coalesces struct and array accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's message vectorization hoists small messages out of a loop \cite{FortranD:1992}.
-
-Caching in Grappa additionally provides a mechanism for exploiting temporal locality by operating on the data locally --- even though we aim Grappa at applications with low locality, we want to be able to exploit it when available.
-Currently, the burden is on the programmer to maintain coherency by introducing explicit synchronization.
-Future work will develop a software directory based coherence scheme to simplify consistent access to global data.
-
-\textbf{Delegate operations.} When the access pattern has low-locality, it is more efficient to modify the data on its home core rather than bringing a copy to the requesting core and returning it after modification.
-Delegate operations provide this capability.
-Applications can dispatch computation to be performed on individual machine-word sized chunks of global memory to the memory system itself (e.g., \emph{fetch-and-add}).
-Delegate operations, proposed in \cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the primary synchronization method in Grappa.
-
-\subsection{Communication}
-\label{sec:communication}
-
-In order to mitigate the low message injection rate limits of commodity networks, Grappa's communication stack has two layers: one for user-level messages and one for network-level messages.
-
-At the upper layer, Grappa implements asynchronous active messages \cite{vonEicken92}.
-Each message consists of a function pointer, an optional argument payload, and an optional data payload.
-When a task sends a message, the message is copied to a send queue associated with the message's destination and the task continues execution.
-
-Grappa's lower networking layer aggregates the upper layer's messages to
-improve performance. Commodity networks including
-Infiniband~\cite{infinibandbandwidth} achieve their peak bisection bandwidth
-\emph{only} when the packet sizes are relatively large---on the order of
-multiple kilobytes. The reason for this discrepancy is the combination of
-overheads associated with handling each packet (in terms of bytes that form
-the actual packet, processing time at the card and processing on the CPU
-within the driver stack). Our measurements confirm manufacturers published
-data~\cite{infinibandbandwidth}, that with this packet size the bisection
-bandwidth is only a small fraction, less than 3\% of the peak bisection
-bandwidth.
-
-In our early experiments the vast majority of requests were smaller than 44
-bytes, far too small to make efficient use of the network. To make the best
-use of the network, we must convert our small messages into large ones. When a
-task sends a message, it is not immediately sent, but rather placed in a queue
-specific to the destination. If the size in bytes of a queue is above a
-threshold (e.g., 4096, typically large enough for good bandwidth), the
-contents of the queue are sent immediately. Also, each queue has a wait time
-threshold ($\approx${1ms}). If the oldest message in a queue has been waiting
-longer than this threshold, the contents of the queue are sent immediately,
-even if the queue size is lower than the message size threshold. Finally,
-queues may be explicitly flushed, for situations when latency is more
-important than bandwidth.
-
-Underneath the aggregation layer, Grappa uses the \gasnet~communication
-library~\cite{gasnet} to actually move data. All interprocess communication,
-whether on or off a cluster node, is handled by the \gasnet~library.
-\gasnet~is able to take advantage of many communication mechanisms, including
-ethernet and infiniband between nodes, as well as shared memory within a node.
-
-Some networks provide access to a remote machine's memory directly.
-This would seem to be a good fit for a programming model focused on global shared memory, however, we do not currently use it.
-In our experiments, we found that RDMA operations are subject to the same message rate limitations as all other messages on these cards, and thus using raw RDMA operations does not measurably increase achieved bandwidth (i.e., we are limited by injection rate not peak bandwidth).
-At the moment, we implement remote memory operations with active messages in conjunction with aggregation, providing an order of magnitude or more increase in throughput compared to straightforward RDMA.
-As we describe later, however, we wish to explore richer semantics for RDMA communication that will make it integrate better with user level runtime systems such as Grappa.
-
-\subsection{Performance}
-
-To evaluate the performance of our current prototype with respect to the XMT,
-we ran each of our three benchmarks on up to 16 nodes of each machine. Grappa
-used 6 cores per node, with the best parameters chosen for each point. In some
-cases, the XMT could not run the benchmark with 2 nodes, so the point is
-omitted.
-
-\begin{figure}
-\begin{center}
-
-\hspace{-1in}\begin{minipage}{0.3\textwidth}
-\begin{center}
-\includegraphics[width=2.5in]{figs/uts_performance.pdf}
-\caption{\label{fig:uts_compare} Performance of in-memory unbalanced tree search.}
-\end{center}
-\end{minipage}
-\hspace{1.25in}\
-\begin{minipage}{0.3\textwidth}
-\begin{center}
-\includegraphics[width=2.5in]{figs/bfs_performance}
-\caption{\label{fig:bfs-performance} Performance of breadth-first search.}
-\end{center}
-\end{minipage}
-
-\end{center}
-\end{figure}
-
-\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex
-tree (T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
-number of vertices visited per second versus number of compute nodes. Grappa
-is 3.2 times faster than the XMT at 16 nodes. As we will show later, the
-performance advantage Grappa has over XMT increases as more nodes are added.
-The main reason Grappa performs better is the software-based delegate
-synchronization obviates the need for the retry-based synchronization that XMT
-uses.
-
-\paragraph{BFS} We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and $2^{29}$ edges (25 GB of data).
-Figure~\ref{fig:bfs-performance} shows our performance in terms of graph edges traversed per second.
-The XMT is 2.5 times faster than Grappa at 16 nodes.
-Performance has not peaked for Grappa, suggesting that adding more nodes will increase performance.
-
-\section{Related work}
-
-Our efforts are built on many existing ideas in programming languages, systems and architecture. In this section we discuss related frameworks and key enabling technologies that Grappa builds upon.
-
-\paragraph{Comparable frameworks} Distributed graph processing frameworks like Pregel~\cite{pregel:2010} and Distributed GraphLab~\cite{distgraphlab:vldb12} share similar goals as Grappa.
-Pregel adopts a bulk-synchronous parallel (BSP) execution model, which makes it inefficient on workloads that could prioritize vertices.
-GraphLab, on the other hand, schedules vertex computations individually, allowing prioritization, which gives faster convergence in a variety of iterative algorithms.
-GraphLab, however, imposes a rigid computation model where programmers must express computation as transformations on a vertex and its edge list only, with information only from adjacent vertexes.
-Pregel is only slightly less restrictive, as the input data can be any vertex in the graph.
-Grappa also supports dynamic parallelism with asynchronous execution, but parallelism is expressed as tasks or loop iterations, which is a far more general programming model for irregular computation tasks.
-
-\paragraph{Global memory} Grappa includes a custom implementation of a software distributed shared memory (DSM) system.
-Many traditional software DSM systems are page based~\cite{Treadmarks,munin} and aim to hide the fact that they are built in software from applications by exploiting the processor's paging mechanisms, therefore relying heavily on locality.
-Instead, Grappa, like other partitioned global address space (PGAS) models, implements its DSM at the language, rather than system level.
-%%note:  Myers suggestion is not incorporated here, as this seems ok:
-Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and UPC~\cite{upc:2005} make accesses to shared structures look like normal memory references.
-As we describe later, Grappa chooses a middle ground, where global addresses are explicit in the API and local accesses are emitted conventionally by the compiler.
-While Grappa's DSM system is conceptually similar to prior work, its implementation is tuned for irregular computations.
-Past DSM work, being page-based, could exploit the RDMA capabilities of network hardware to move large page-sized blocks of data from node to node.
-In our experience, when these networks move small blocks of data (a few bytes), only a fraction of the available bandwidth is achieved.
-%%note:  Myers suggestion is not incorporated here, as this seems ok:
-In addition, the DSM system in Grappa ends up being tightly coupled to the task scheduler in order to overlap long latency memory operations with useful computation.
-For these two reasons it became necessary to build a new DSM system specifically for Grappa.
-
-\paragraph{Multithreading} Grappa uses multithreading to tolerate memory latency.
-This is a well known technique.
-Hardware implementations include the Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife}, Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}.
-As we describe in this paper, Grappa use a lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks on a single processing core.
-The large number of tasks is required because of the extremely high internode latency Grappa mitigates.
-Grappa's task library employs several optimizations: an extremely fast task switch, a small task size, and judicious use of hardware prefetching to bring task state into the cache before the task is scheduled to execute.
-
-\section{Proposed Research}
-
-As described earlier, our work on Grappa so far has focused on building an initial prototype so we have a baseline system to build upon.
-This early prototype has lead us to think of many research opportunities, from new ways to improve network communication scheduling, to new data storage management ideas to compiler and language support.
-Below we detail our research plans in each of these thrusts.
-
-\subsection{Networking}
-
-Since Grappa is designed for applications that have no or low locality, nearly \emph{every} memory reference is to a remote system node.
-This makes the networking layer key to overall system performance.
-In this section we outline the key directions we wish to take in enhancing the networking layer of the Grappa runtime system.
-
-\paragraph{Virtual topology:} Currently Grappa uses a flat network topology where each processing core in the system is individually addressed.
-As more nodes are added to the system, a quadratic increase in the number of threads (in aggregate across all nodes) is required in order to compensate for the \emph{same} network delay.
-This is because the current aggregator treats each of these end-points individually, i.e., it only aggregates messages that go to the same exact end-points.
-This does not scale as the number of nodes increases because the probability of messages going to the same destination goes down (assuming a somewhat uniform communication distribution).
-In summary, this past design choice is simple but does not scale.
-
-We need to rewrite the aggregator to build in a hierarchical virtual network layer.
-The first layer of this hierarchy is to treat all processors at a given system node as a single destination.
-We intend to continue to use the lock-free single-threaded \gasnet~communication layer that we have been using, as we have found the lock-based one to incur too high of an overhead for our purposes.
-Thus the aggregator will need to very carefully utilize lock-free data-structures and interact with \gasnet~carefully.
-%%Didn't understand Myers suggestion here.
-
-The second, and upper layers of the virtual topology are built from clustering system nodes into groups.
-Instead of aggregating for all end-points, aggregation will occur for system nodes within a group.
-A designated (and dedicated) node will act as a router, processing incoming and outgoing traffic to system nodes within the group and other groups.
-
-We expect this hierarchical organization will reduce scaling requirements from
-\BigO{n^2} threads to \BigO{n \times log(n)}. Of course, it takes what used to
-be a single network hop for communication and turns it into \BigO{log(n)}
-hops. More threads will be required to overcome this increased latency, but we
-hypothesize only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling
-benefit is achievable with a virtual hierarchical network topology.
-
-\paragraph{Hardware support for aggregation:} If the increased latency caused
-by extra hops in hierarchical aggregation turns out to be detrimental, we
-will explore mitigation by moving some of the aggregation logic to the network
-card, either by exploiting off-the-shelf programmable network cards or
-experimental systems such ask NetFPGA~\cite{netfpga} (while we are planning to
-focus on Infiniband-based networking, if we explore the NetFPGA route we will
-be using Ethernet).
-
-\paragraph{User-level RDMA:} Certain commodity networking hardware today
-supports remote direct memory access (RDMA). Unfortunately, \emph{how} it is
-implemented is so primitive and low-level that it is difficult or impossible
-to efficiently use in practice. We start with a short description of the
-current semantics of RDMA implementations. Next we describe the research we
-wish to undertake to make RDMA more useful.
-
-Networking infrastructure supports remote read and write operations.
-Read operations take a source address and key (on a remote node), a destination address and key (on the local node), and a length.
-Cards support address translation for regions of memory that are registered with them.
-These translations, indexed by the supplied key, are used to perform virtual to physical address translation.
-The physical address of the destination address is similarly translated using the translation table at the local node.
-Cards also support remote write operations with a similar set of semantics, where addresses are translated on a card using the card's local translation table.
-Both read and write operations are asynchronous.
-The requesting processor initiates the operation and then can check for completion by querying a completion queue at the local card.
-\emph{Notably}, the processor(s) on the remote node are provided a polling interface whereby they can learn if a remote write operation occurred.
-No such interface exists to learn about remote read or atomic operations.
-
-This interface suffers from multiple shortcomings.
-The biggest of which is that ultimately, moving data to and from a remote node constructively requires higher level semantic operations than simple reads and writes.
-These higher level semantics can, of course, be synthesized from the underlying read and write primitives, but it wastes round trips and synchronization time.
-One possible alternative solution is to enable remote nodes to dispatch high-level data-structure manipulation requests (e.g.
-insert this item into a queue, remove an item from a set).
-The challenge here is to make those calls programmable and interoperate securely with a modern operating system, and a user-level runtime system.
-Additionally there are questions about \emph{who} should carry out the computation.
-Classic RDMA does not interrupt the host processor, but these richer programmable semantics will require either a processor in the network interface (necessitating issues such as coherence be managed) or complex RDMA tasks need a mechanism to interrupt the host processor \emph{without leaving user space}.
-
-%% The second half of this paragraph is really confusing, I have just a minimal clue of what it is talking about. And the little I understand sounds a bit vague and not directly related to grappa.
-The second challenge is that the virtual to physical address translation on the cards requires the operating system to pin translations.
-This removes paging flexibility from the operating system, because runtime systems are (rationally) built to expect data structures to be in memory capable of RDMA or not.
-Those in RDMA memory must be pinned.
-In the long run, the solution here likely lies at the intersection of what progress is occurring in processor support for virtual machines and virtual memory translation for I/O.
-Our goal will be to study carefully the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.
-
-\subsection{SSD-only data structures}
-
-Grappa applications are likely to use large amounts of data, 100s of terabytes to petabytes.
-Keeping all these data in DRAM only is unlikely to be cost effective.
-We will explore mechanisms that use solid-state storage as primary memory, greatly expanding the directly addressable storage at low cost.
-
-Grappa is very effective in hiding latencies of network communication.
-If an access is to a chunk of the heap stored in a node's local memory, Grappa issues a prefetch for the address and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}.
-If the access is to memory stored on another node, Grappa issues a message to that node to request the data and context switches to other work until the reply arrives.
-In fact, our measurements and estimates show that we might be able to tolerate even greater latency -- on the order of milliseconds per request.
-This provides the tantalizing possibility that we will be able to keep data in \ssd\ drives only and service data access requests directly from them, without leaving copies in DRAM.
-The natural point of integration for non-volatile memories in Grappa is to store the global distributed shared heap in non-volatile memory.
-All accesses to this shared heap are done through an \api, so this switch would be transparent to the programmer.
-
-This is not straightforward, however, for three reasons: (1) \ssds\ assume spatial locality.
-In our benchmarks, we find the average request size to be less than 64 bytes.
-Today's \ssds\ are optimized for reads in the multi-kilobyte range.
-(2) \ssds\ have a low request rate.
-The fastest~\cite{fusionio} can perform on the order of 1 million random reads per second, and many are limited to less than 100K/s.
-Even the underlying flash chips are slow, on the order of 40K/s~\cite{micronFlash}.
-In contrast, our measurements suggest cores may be idle in Grappa if each node cannot perform 100 million random reads per second of the global heap.
-And (3) \ssds\ have high overhead.
-Mass market \ssds\ are usually treated as normal disks: accesses must occur through the kernel, incurring a high context-switch overhead.
-Grappa depends on user-mode access to the network layer to avoid this overhead; a similar approach has been applied to \ssds\ in recent research~\cite{caulfield:2012}, however.
-
-To address this challenge we will explore building a flash array consisting of several ten's to a hundred disks \emph{per node}.
-This will enable us to increase random accesses to our target 100 million references per second.
-But it will require several active disk controllers per system.
-We do not expect any existing device driver stack will work at this scale.
-Hence we will need to write a custom disk controller device driver and file system that supports such a high degree of concurrent requests across so many parallel channels.
-
-%% Sorry Mark, while the item below is clearly rich and juicy, we can't leave it bhere because it is huuuuge, it does not match the scope of the grant, and we dont have the expertive --- e.g., are you sure nobody is working on it?
-
-%% I think this is the only idea that is going to work... maybe we should delete the above three and make this one good.  Seriously, the ideas up there are bunk.  -ark
-% \paragraph{Rearchitecting flash:} Clearly the best solution here lays in re-architecting the flash chip itself.
-We will explore new flash microarchitectures that optimize for message rate as well as bandwidth.
-Such chips have value far beyond Grappa.
-Research into persistent key/value stores, and really any persistent high-throughput data-structure will benefit from a redesigned flash chip that can sustain dramatically increased message rates.
-
-\subsection{Compiler and language support}
-
-Language and compiler support are instrumental in making Grappa usable to a
-wide audience of application developers. We plan to use the LLVM~\cite{llvm}
-infrastructure. Below we describe the specific aspects we plan to tackle.
-
-\paragraph{Relaxed consistency:} Presently Grappa uses a consistency model that is sequentially consistent for data-race free programs.
-Underlying hooks are available to build even more relaxed consistency models, although we don't expect programmers themselves will want to use them.
-A compiler, however, can see a performance boost by analyzing code and exploiting alias knowledge to program to this underlying Grappa \api, potentially gaining a performance boost by exposing more memory concurrency and hiding more latency.
-
-\paragraph{Language support for synchronization operations: } At the moment, runtime library support is provided in Grappa for basic synchronization primitives (lock, barrier, etc).
-Building synchronization directly into the language has the advantage that the compiler can understand and optimize it.
-Several optimizations are possible in the Grappa framework, including exploiting the scheduler and work stealer to elide certain lock operations (scheduling is co-operative on a per-CPU basis), converting synchronization primitives into remote delegate operations, and exploiting the check-in / check-out nature of global memory in Grappa to coalesce multiple synchronization operations together.
-
-\paragraph{Delayed writes: } At the moment when we we write applications for Grappa we use a technique that we call feed-forward delegates.
-Essentially what this technique does is if a loop is reading and writing disjoint sets of data, it is possible and we have found efficient, to dispatch writes asynchronously and ensure those writes are completed only by the end of the loop.
-This increases performance by further overlapping computation with memory access.
-A compiler aided by programmer annotations or capable of alias analysis can do this transformation automatically for the programmer.
-
-\paragraph{Automatic delegate synthesis: } Delegates are a technique in Grappa where one thread can request a small computation to be carried out on a remote node.
-For instance, an increment on a memory location is dispatched to the system node that manages that memory operation and carried out there.
-Currently the decision of what is a delegate is left to the programmer.
-But this is an area rich for compiler optimization.
-We will explore techniques to automatically synthesize delegates from code, potentially leading to more complex delegates and more efficient code.
-
-\paragraph{Read and write combining: } As a special case of automatic delegate synthesis, when multiple remote memory operations are to be dispatched to the same node from the same thread, it may be more efficient to combine these together into a single request.
-We will develop a compiler pass to combine operations when possible.
-%%Myers suggested a reference to UPC or other previous work
-
-\paragraph{Compile for a continuation model: } Presently, whenever a task is active it is allocated a complete stack.
-Since several thousand tasks are active per processor at any one time, these stacks occupy a large footprint in memory.
-We wish to explore compiling tasks to a continuation model, whereby when a task dispatches a remote memory operation and is to be context-switched out, another task can utilize that stack.
-Combined with the delegate synthesis ideas described above, we will also explore runtime and compiler support to convert remote memory operations into thread migrations; instead of moving the data to the system node that has the task, we will move the task to the data in cases where that would be more efficient.
-
-\paragraph{Task-weaving: } Much of the time the same task code is executing for hundreds of thousands of tasks in the Grappa runtime system.
-For example, each task may be processing a loop body for a parallel for loop.
-However, tasks introduce overhead, so we already manually impose a cutoff on recursive decomposition of parallel loops.
-This way, multiple loop bodies are collected into a single task and executed serially.
-However, this limits the amount of parallelism available for latency tolerance.
-A mmore efficient technique is to allow the compiler to unroll and schedule all loop bodies simultaneously within the same task, including overlapping remote memory operations, thereby ``weaving'' these independent tasks together to overlap computation with communication with minimal task overhead.
-Further, a set of ``woven tasks'' may be amenable to vector execution, allowing even more parallelism to be exploited.
-
-\paragraph{Task load control:} Having too few tasks per core fails to tolerate
-fully the latency; having too many increases overheads unnecessarily.
-Currently, we choose a number at program instantiation: experiments so far
-have shown evidence of a broad sweet-spot, though it is of course varies as a
-function of application behavior and input data. At compile-time, we might be
-able to estimate the number of tasks required based on the density of remote
-operations and system parameters (e.g., memory size, network latency, etc.);
-at runtime we can adjust the number of tasks dynamically based on periodic
-performance assessments.
-
-% Presently we are starting work on this language and compiler support.  We are basing our effort on the LLVM~\cite{llvm} infrastructure.
-
-\subsection{Graph benchmark suite}
-
-In order to facilitate the development and evaluation of Grappa, we will develop a comprehensive benchmark suite for graph analytics applications.
-This will not only be useful for us, but the whole community at large, since there are few benchmarks in this application space.
-
-This benchmark suite will consist of application kernels, reference implementations, data-sets, data-set generators, documentation, specifications for generating standardized performance results, and a hosting service (http://graphbench.org) to provide public access to the benchmark and datasets as well as aggregate and publish results by all users of the benchmark suite.
-Specifically, we aim to provide transparency and an opportunity to build off of others' discoveries by requiring source code be provided with any submission.
-
-This benchmark suite will be a tremendous value to the research community.
-Not only will it enable researchers to evaluate ideas on graph processing system techniques, but it will also enable fair comparisons of ideas in a depth not currently available from other graph related benchmark suites~\cite{graph500}.
-The benchmark suite will be released in a completely open fashion, so anyone will be able to use it (and contribute to it).
-All code will be distributed into the public domain (BSD style licensing) so companies can use it.
-The goal is to get something out relatively early in the hope of (i) getting others to cluster around this work rather than develop a plethora of overlapping alternatives, and (ii) help establish consistent metrics for performance results.
-
-We will build a collection of key computational kernels from graph and irregular applications.
-The benchmark suite will consist of both documentation on the actual kernel (ideally a research paper or book reference sourced from the community), and a reference implementation.
-The benchmark suite will also include sample data collected from existing sources, and synthetic workload generators.
-The collected data provides a grounding for performance analysis on real-world problems.
-The workload generators provide cleaner, easier to reason about randomly generated data-sets at a programmable scale.
-
-\section{Broader impacts}
-
-\paragraph{Increasing involvement of underrepresented groups:} Computer science as a whole has a significant problem attracting women and minorities to the field.
-Computer architecture has an even more extreme problem with diversity (in the PI's view).
-This is a problem the PIs have taken quite seriously since entering the field, and we are quite proud of our PhD students, both female who have gone on to faculty positions in top academic institutions, and male who appreciate the gender and ethnic disparity and who work hard at their own institutions to correct it.
-We continue to recruit a diverse graduate student mix and we continue to include undergraduate students in our research.
-At the moment we are mentoring two women undergraduates involved in the Grappa project, and if this proposal is funded will apply for REU funds to continue their involvement.
-
-\paragraph{Benchmark suite:} As previously described, we are seeking funding support to build and distribute a benchmark suite for irregular and graph applications.
-This benchmark suite has the potential to have a broad impact on academia and industry.
-We will distribute this benchmark through \texttt{graphbench.org} and will also collect and publish results from competing research groups.
-
-\paragraph{Technology transfer:} Bill Dally during his 2010 Eckert-Mauchly award speech said (in effect), technology transfers happen when people transfer.
-Having just come back from a three year stint founding a startup company (where we transferred the Deterministic Multiprocessing technology from UW), we've seen this first hand.
-The students and faculty involved in the Grappa project plan to spin off a company with their work.
-The exact time frame is not yet known, but everyone involved in the project is excited to see the technology underpin a startup company and make this technology available to the broader commercial world.
-
-\paragraph{Curriculum development activities:} Concurrent with the execution of the research activities described in this grant we will seek to teach an advanced graduate class on advanced concurrency concepts.
-We taught a similar class over six years ago and it is time to revive, update and teach it to the next generation of computer architecture graduate students.
-\comment{Add more here.
-This has no passion.
--Mark}
-
-\section{Results from Prior NSF Support}
-
-\paragraph{Ceze:} Prof.
-Ceze's research is on computer architecture, programming languages and operating systems.
-His primary focus has been on improving programmability, reliability and energy efficiency of multiprocessor computer systems.
-His research is funded primarily by NSF, PNNL and industry.
-His CAREER award {\em Deterministic Shared Memory Multiprocessing: Vision, Architecture and Impact on Programmability} (CCF-0846004, 2008-13, \$559K) demonstrated that it is possible to provide deterministic execution of arbitrary multithreaded programs without sacrificing performance significantly.
-The results include innovation in architecture, compiler, and operating-systems.
-Two other NSF grants, {\em Code-Centric Approach to Specifying, Checking, and Discovering Shared-Memory Communication} (CCF-1064497, 2011-2014, \$901K) and {\em Precise Concurrency Exceptions: Architecture Support, Semantics and System Implications} (CCF-1016495, 2010-2013, \$500K), are leading to significant results in dealing with concurrency errors and concurrent programming language semantics.
-His work on large-scale graph processing is being funded by the Pacific Northwest Laboratory.
-His work on energy-aware programming models and approximate computing has been primarily funded by a seed NSF grant {\em Disciplined Approximate Programming for Energy-Efficient Computing} (CCF-1216611, 2012-2015, \$300K) and industry, including Microsoft, Google and Qualcomm.
-
-\paragraph{Kahan:} Dr.
-Kahan's research has focused on irregular parallel computation.
-Over the course of twelve years, he developed the Tera MTA's multithreaded runtime system and memory allocator still in use today in the Cray XMT, exploring applicability of the system to production codes and benchmarks in the engineering, science, and government domains.
-His research has been driven by real-world application needs, resulting in new algorithms and systems to address concurrency bottlenecks.
-He is presently an employee of the Pacific Northwest National Laboratory and an affiliate faculty member of the Computer Science and Engineering Department at the University of Washington.
-He has no past or current NSF support.
-
-\paragraph{Oskin:} Prof.
-Oskin's prior research has focused on architectures and systems for scalable computing.
-With support from NSF he has carried out a variety of research activities, leading to over 50 scholarly publications, three PhD students graduated, four masters students graduated, and a company spin-off.
-The past NSF-supported work (WaveScalar) heavily influences the ideas described here.
-In some sense, the work here is a culmination in thought processes about why WaveScalar did not achieve as much ILP as earlier limit studies in ILP suggested were possible.
-NSF support for the PI's work, {\em ITR: WaveScalar, A New Approach to Scalable System Design} (CCF-0325635, 2003-07, \$1.3M), {\em CAREER: Soft-Instruction Set Computing} (CCF-0133188, 2002-05, \$443K), helped to develop WaveScalar.
-He has also spent over a decade exploring architectural implications of quantum computing technologies, with support from NSF {\em NER: Computer Aided Design of Silicon-based Quantum Computers} (CCF-0210373, 2002-04, \$90K), {\em EMT: Microarchitectures for Quantum Computers} (CCF-0523359, 2006-10, \$275K), {\em EMT: Self-Correcting Fault-Tolerant Quantum Computers} (CCF-0523359, 2005-08, \$406K) DARPA {\em QuIST: Architectures and Applications for Scalable Quantum Information Systems}.
-He is currently exploring runtime systems for irregular parallel applications, which is the subject of this proposal and is partially funded by the Department of Energy.
-
-\section{Work Plan}
-\label{sec:plan}
-
-Our research plan can be divided horizontally into three core areas: programming languages/compilers; communication and storage; and benchmarking.
-Below we outline how we plan to advance these fronts over the next four years.
-
-\vspace{2ex}
-
-\noindent\textbf{Year 1:} The first year of this effort will focus on creating the benchmark suite and initial language and compiler support.
-We aim to pull together 6-8 benchmarks, sample data sets, and scripts to measure performance results on the initial Grappa runtime, UPC~\cite{upc:2005} and MPI implementations, and serial reference code.
-For the compiler we will have the front end done enabling easy addition of keywords to annotate data-types and structural elements (parallel for, etc).
-The compiler will be generating non-optimized Grappa code by the end of the first year.
-
-\vspace{1ex}
-\noindent\textbf{Year 2:} The second year of effort will see the completion of the benchmark suite, a significant effort on compiler optimizations for Grappa, and the beginning of the networking layer research.
-The benchmark suite will include 12-15 applications and/or kernels.
-The website will be live and we will be conducting significant outreach efforts to drive adoption.
-On the compiler front we will have implemented the key optimizations that provide the highest benefit for Grappa code.
-In addition we will have a detailed understanding of what additional work remains to support other languages / language constructs.
-In this year we will start work on rewriting the networking layer of the Grappa runtime, providing a virtual network topology.
-We will also begin work on programmable RDMA by using FPGA-based networking gear.
-
-\vspace{1ex}
-\noindent\textbf{Year 3:} In the third and final year of this effort (supported by this grant), we will finish the networking layer rewrite.
-We will also have constructed a test-bed system comprised of mainly SSD-based memory.
-Grappa will be ported to this new style of memory, which we envision will largely be a tuning and not rewriting effort.
-We will explore alternative SSD-memory designs largely through emulation.
-Finally we will complete a detailed system evaluation using our benchmark suite.
-
-\section{Summary}
-
-Irregular applications are increasing in importance to the computing industry.
-For a long time they have underpinned computations of interest to the national
-security arm of the federal government. But now, they underpin such questions
-as ad placement in social networks, and analysis of complex data-sets in
-medicine and science. The defining characteristic of these applications is
-poor locality and massive available parallelism. Our effort is focused on
-making these applications perform well and scale on commodity hardware. The
-key idea that makes this work is to rely on massive concurrency to tolerate
-memory latency, instead of relying on locality (which doesn't exist for these
-applications). This core idea underpinned the XMT~\cite{feo:xmt} system, a
-fully-custom hardware system. We plan to explore that idea in software only,
-running on commodity processors. And unlike the XMT, however, we must also
-contend with commodity networking hardware, and here we again rely on
-threading to provide a sufficient number of requests that we can buffer them
-prior to dispatching on the network, to overcome bandwidth limitations when
-operating with small messages.
-
-Our preliminary effort has shown a workable proof of concept of all this
-runtime smoke and mirrors: a commodity cluster can out do the XMT at its own
-game, by emulating its key features in software. What is needed now is
-additional effort to improve the underlying performance of this runtime and
-make it usable to a broad base of users. We will be focusing our efforts
-in the next three years on the key performance component, the networking
-layer, and the key programmer efficiency linchpin, high level language
-support. In addition we need a useful mechanism for handling data-sets in the
-petabyte range, and here we imagine building a cluster of machines that relies
-primarily on SSD's and not DRAM as main memory data-set storage.
-
-If successful, the Grappa runtime will have a transformative impact on the
-computing industry. Irregular applications are hugely important, and if off
-the shelf hardware can be efficient with them a new class of questions can be
-asked of computing systems. Additionally our benchmark suite development will
-further drive innovation among researchers in this space.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%
-%%%%% Extra stuff: collaboration plan and data-management plan
-%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-
-\section*{Data Management Plan}
-
-The proposed research involves developing experimental systems and
-assessing how well they work. The kinds of data we will produce
-include: characterization data of program behavior; data on system
-performance and reliability; data on how programmers use new language
-constructs and tools. As is our practice, we will regularly submit
-papers for publication detailing our findings and design of our
-systems.
-
-As we have already been doing with all our research, for every
-published paper, we make all the raw data available online. This
-allows other researchers to look at the data in different ways, and it
-enables them to reproduce results. We also often publish
-the experimental software developed, following the rules outlined by
-the University intellectual property office --- which provides ample
-rights for other researchers to build upon our experimental
-systems. We do not anticipate using privileged or confidential
-information.
-
-\newpage
-\section*{Collaboration Plan}
-%\setcounter{page}{1}
-
-Because the PIs already successfully collaborate and co-advise an
-active cadre of spectacular students, the necessary tools and
-coordination mechanisms are already in place for this project to
-proceed successfully.  We first describe the PIs' complementary
-backgrounds, then describe the current organization of our (amorphous)
-research group, then describe the specific technology and facilities
-that enable seamless coordination, and finally describe specific PI roles
-for the research in this proposal.
-
-\subsection*{Complementary PI Experience}
-
-PI Oskin has traditionally participated in the computer-architecture research
-community, co-PI Ceze, in addition to the architecture community also has a
-presence in the systems and programming languages community, and co-PI Kahan
-has participated mainly in the parallel and scientific computing community.
-The move toward multicore architectures has re-invigorated collaboration
-across these divides and recent joint work by Oskin, Ceze, and Kahan
-embodies this important trend. There are currently five graduate students for
-which the last two of the PIs serve as advisors. In addition to the project
-described in this proposal, they are working together on techniques for
-detecting concurrency errors with machine learning and language
-specifications, using disciplined approximations to save energy in computing,
-and developing new operating-system abstractions for emerging non-volatile
-memory devices. What has made these collaborations so fruitful in such a short
-period of time is the PIs' complementary backgrounds. We now briefly describe
-these backgrounds.
-
-Oskin has explored a wide variety of architectural mechanisms for improving application performance, most notably intelligent memory~\cite{activepages} and dataflow computing~\cite{wavescalar-micro, wavescalar-isca}.
-He has developed novel evaluation techniques~\cite{hls} that enable efficient design space exploration of complex superscalar processors.
-His past work has included runtime systems for GPU-based computing~\cite{oskin-gpu}, application specific processor design~\cite{sherpa} and polymorphic on-chip networks~\cite{polynetworks}.
-He has also explored several architectural design options for quantum computers~\cite{quantum-isca03, quantum-isca05, quantum-isca08}.
-Finally he developed mechanisms to make parallel systems more deterministic and easier to program~\cite{dmp}.
-His most recent work is focused on the work described in this proposal~\cite{Nelson:hotpar2011}, which is to make irregular computations efficient on commodity hardware.
-
-Ceze has been conducting research on architecture support, compiler techniques, and programming models to improve the programmability of multiprocessor systems for seven years and has co-authored over thirty papers in these areas.
-His research includes advances in architecture~\cite{tlsooo, bulk, bulksc, swbulk, cyclops, dmp, delorean} and programming models~\cite{colorama, ipot, posh} to simplify exploiting concurrency in multiprocessors.
-Some of his recent results especially related to this proposal include a programming model based on ordered transactions and data annotations~\cite{ipot}; improving reliability of multithreaded software with architecture support for bug detection and avoidance~\cite{atomaid, aatoppicks, bugaboo, oshajava, cs-isca10, ce-isca10}; and deterministic execution of multithreaded programs~\cite{dmp, dmptoppicks, asplos10coredet, dmpos}.
-PI Ceze recently co-organized an NSF-funded workshop focused specifically on deterministic multiprocessing.
-The workshop was very successful, bringing leaders in the multicore area from industry and academia.
-
-Kahan has been working in the area of parallel computing for irregular applications for twenty years, largely in industry.
-At Cray (previously Tera Computer) he developed the runtime system~\cite{Alverson95schedulingon,Alverson97terahardware-software} and combining memory allocator~\cite{MAMA} still in production use on Cray's XMT system in which over 50,000 threads execute concurrently sharing one global address space.
-He has co-authored patents and publications for both of these systems as well as for a scalable connected components graph algorithm that bears his name~\cite{Underwood07analyzingthe}.
-Kahan originated and for three years has been co-PI for the Grappa project.
-
-\subsection*{The SAMPA Research Group: Addressing the Concurrency
-  Challenge Across the Execution Stack}
-
-The PIs organize their joint research activity under the umbrella of
-the ``SAMPA group'' (SAfe MultiProcessing Architectures).  Some brief
-statistics help summarize the group's size, strength, and focus:
-\begin{itemize}
-\item There are roughly ten graduate students, most actively
-  coadvised.
-\item In the last four years, the group has produced twelve papers
-  published in ASPLOS, ISCA, PLDI, OOPSLA, MICRO, and OSDI as well as three
-  papers in the USENIX Workshop on Hot Topics in Parallelism.
-\end{itemize}
-The SAMPA group is an ideal size for collaboration across the system
-execution stack from computer architecture up through software
-engineering.  Compared to large collaborations with several faculty,
-there are no graduate students (or faculty) who specialize entirely in
-particular subtopics.  The students gain expertise across
-computer-science disciplines and use appropriate ideas from hardware,
-software, theory, etc. to solve research problems.
-
-\subsection*{Specific Technology and Facilities}
-
-The research group that will perform the research has the facilities
-needed for successful collaboration.  There is a shared lab facility
-where weekly meetings are held.  Students in the group share computing
-infrastructure, including dedicated machines and regular use of cloud
-services such as Amazon's EC2 for scalability experiments.
-
-The PIs regularly hold joint meetings with each student that they
-co-advise.  They do not split advising duties so much as duplicate
-them.
-
-\subsection*{Specific Roles and Responsibilities}
-
-It should be clear from the preceding discussions that the PIs are
-entirely comfortable jointly managing the work in the proposal and the
-students involved.  Nonetheless, several specific pieces of the work
-more naturally align with the expertise and interests of one of the
-PIs:
-
-\begin{itemize}
-\item PI Oskin will focus on communication issues. 
-\item PI Ceze will focus on compiler and programming languages work.
-\item PI Kahan will focus on organizing benchmarks and performance evaluation efforts.
-\item All PIs expect to solicit guidance and collaboration from
-  researchers such as John Feo (PNNL), Jim Larus (Microsoft Research), Hans
-  Boehm (HP Research), and Preston Briggs,  with whom
-  they have already collaborated on other projects.
-\end{itemize}
-
-\input{footer}
+Thus a broad impact of our work will be the transformative effect on the types of scientific and computing questions that commodity hardware can efficiently answer.
+In addition our work will include the construction of a comprehensive irregular benchmark suite to help guide our own and the field's research.

### 2013-02-14 doc/xps2013/proposal.tex
final pass on intro complete

@@
 
-\item \textbf{Reinvent network communication:} Our preliminary work has
+\item \textbf{Accelerated network communication:} Our preliminary work has
 revealed limitations of the current Grappa runtime system with regards to
@@
 strictly as a library. Programmers must express remote operations explicitly
-as function calls. This adds burden high-level language programmers do not
+as function calls. This adds a burden high-level language programmers do not
 usually face and obfuscates opportunities for compiler optimizations. A
@@
 compilation support to provide a cleaner programming model and use
-statically-available knowledge to lessen the runtime overhead automatically.
+statically-available knowledge to lessen the runtime overhead.
 
@@
 Graph queries, machine-learning and data-driven science will significantly benefit from the technology we aim to create from the ideas in this grant proposal.
-In addition our work will include the construction of a comprehensive irregular benchmark suite, something to help guide our own and the field's research.
+In addition our work will include the construction of a comprehensive irregular benchmark suite to help guide our own and the field's research.
 

### 2013-02-14 doc/xps2013/proposal.tex
commit to merge

--- 2013-02-14 doc/xps2013/proposal.tex
@@
 
-Since Grappa is designed for applications that have no or low locality, nearly \emph{every} memory reference is to a remote system node.
+Since Grappa is designed for applications that have no or little locality, nearly \emph{every} memory reference is to a remote system node.
 This makes the networking layer key to overall system performance.
@@
 Keeping all these data in DRAM only is unlikely to be cost effective.
-We will explore mechanisms that use solid-state storage as primary memory, greatly expanding the directly addressable storage at low cost.
+We will explore mechanisms that use solid-state storage as primary memory for program data, greatly expanding the directly addressable storage at low cost.
 

### 2013-02-14 doc/xps2013/proposal.tex
pass on research section.

@@
 
-We need to rewrite the aggregator to build in a hierarchical virtual network layer.
+We need to redesign the aggregator to build in a hierarchical virtual network layer.
 The first layer of this hierarchy is to treat all processors at a given system node as a single destination.
@@
 implemented is so primitive and low-level that it is difficult or impossible
-to efficiently use in practice. We start with a short description of the
+to efficiently use in practice. Below we start with a short description of the
 current semantics of RDMA implementations. Next we describe the research we
@@
 
-Networking infrastructure supports remote read and write operations.
-Read operations take a source address and key (on a remote node), a destination address and key (on the local node), and a length.
-Cards support address translation for regions of memory that are registered with them.
-These translations, indexed by the supplied key, are used to perform virtual to physical address translation.
-The physical address of the destination address is similarly translated using the translation table at the local node.
-Cards also support remote write operations with a similar set of semantics, where addresses are translated on a card using the card's local translation table.
-Both read and write operations are asynchronous.
-The requesting processor initiates the operation and then can check for completion by querying a completion queue at the local card.
-\emph{Notably}, the processor(s) on the remote node are provided a polling interface whereby they can learn if a remote write operation occurred.
-No such interface exists to learn about remote read or atomic operations.
+Networking infrastructure supports remote read and write operations. Read
+operations take a source address and key (on a remote node), a destination
+address and key (on the local node), and a length. Cards support address
+translation for regions of memory that are registered with them. These
+translations, indexed by the supplied key, are used to perform virtual to
+physical address translation. The physical address of the destination address
+is similarly translated using the translation table at the local node. Cards
+also support remote write operations with a similar set of semantics, where
+addresses are translated on a card using the card's local translation table.
+Both read and write operations are asynchronous. The requesting processor
+initiates the operation and then can check for completion by querying a
+completion queue at the local card. \emph{Notably}, the processor(s) on the
+remote node are provided a polling interface whereby they can learn if a
+remote write operation occurred. No such interface exists to learn about
+remote read or atomic operations.
 
-This interface suffers from multiple shortcomings.
-The biggest of which is that ultimately, moving data to and from a remote node constructively requires higher level semantic operations than simple reads and writes.
-These higher level semantics can, of course, be synthesized from the underlying read and write primitives, but it wastes round trips and synchronization time.
-One possible alternative solution is to enable remote nodes to dispatch high-level data-structure manipulation requests (e.g.
-insert this item into a queue, remove an item from a set).
-The challenge here is to make those calls programmable and interoperate securely with a modern operating system, and a user-level runtime system.
-Additionally there are questions about \emph{who} should carry out the computation.
-Classic RDMA does not interrupt the host processor, but these richer programmable semantics will require either a processor in the network interface (necessitating issues such as coherence be managed) or complex RDMA tasks need a mechanism to interrupt the host processor \emph{without leaving user space}.
+This interface suffers from multiple shortcomings. The biggest of which is
+that ultimately, moving data to and from a remote node constructively requires
+higher level semantic operations than simple reads and writes. These higher
+level semantics can, of course, be synthesized from the underlying read and
+write primitives, but it wastes round trips and synchronization time. One
+possible alternative solution is to enable remote nodes to dispatch high-level
+data-structure manipulation requests (e.g. insert this item into a queue,
+remove an item from a set). The challenge here is to make those calls
+programmable and interoperate securely with a modern operating system, and a
+user-level runtime system. Additionally there are questions about \emph{who}
+should carry out the computation. Classic RDMA does not interrupt the host
+processor, but these richer programmable semantics will require either a
+processor in the network interface (necessitating issues such as coherence be
+managed) or complex RDMA tasks need a mechanism to interrupt the host
+processor \emph{without leaving user space}.
 
-%% The second half of this paragraph is really confusing, I have just a minimal clue of what it is talking about. And the little I understand sounds a bit vague and not directly related to grappa.
-The second challenge is that the virtual to physical address translation on the cards requires the operating system to pin translations.
-This removes paging flexibility from the operating system, because runtime systems are (rationally) built to expect data structures to be in memory capable of RDMA or not.
-Those in RDMA memory must be pinned.
-In the long run, the solution here likely lies at the intersection of what progress is occurring in processor support for virtual machines and virtual memory translation for I/O.
-Our goal will be to study carefully the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.
+The second challenge is that the virtual to physical address translation is
+necessary for RDMA operations to be meaningful. Current systems require the OS
+to pin translations so cards can operate directly on physical addresses. This
+removes paging flexibility from the operating system and requires runtime
+systems to be aware that this is happening, which limits overall usability.
+We will explore recent advances in supporting direct memory access with address translation (e.g., AMD IOMMU) to support flexible RDMA operation that go through full physical address translation without having to interrupt a general-purpose core.
+
+% In the long run, the solution here likely
+% lies at the intersection of what progress is occurring in processor support
+% for virtual machines and virtual memory translation for I/O. Our goal will
+% be to study carefully the I/O virtualization proposals put forth by major
+% processor manufactures and devise solutions that enable RDMA to capitalize
+% on these developments.
 

### 2013-02-14 doc/xps2013/proposal.tex
for the win

@@
 
-We expect this hierarchical organization will reduce scaling requirements from
-\BigO{n^2} threads to \BigO{n \times log(n)}. Of course, it takes what used to
-be a single network hop for communication and turns it into \BigO{log(n)}
-hops. More threads will be required to overcome this increased latency, but we
-hypothesize only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling
-benefit is achievable with a virtual hierarchical network topology.
+We expect this hierarchical organization will reduce scaling requirements from \BigO{n^2} threads to \BigO{n \times log(n)}.
+Of course, it takes what used to be a single network hop for communication and turns it into \BigO{log(n)} hops.
+More threads will be required to overcome this increased latency, but we hypothesize only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling benefit is achievable with a virtual hierarchical network topology.
 
-\paragraph{Hardware support for aggregation:} If the increased latency caused
-by extra hops in hierarchical aggregation turns out to be detrimental, we
-will explore mitigation by moving some of the aggregation logic to the network
-card, either by exploiting off-the-shelf programmable network cards or
-experimental systems such ask NetFPGA~\cite{netfpga} (while we are planning to
-focus on Infiniband-based networking, if we explore the NetFPGA route we will
-be using Ethernet).
+\paragraph{Hardware support for aggregation:} If the increased latency caused by extra hops in hierarchical aggregation turns out to be detrimental, we will explore moving some of the aggregation logic to the network card, either by exploiting off-the-shelf programmable network cards or experimental systems such as NetFPGA~\cite{netfpga} (while we are planning to focus on infiniband-based networking, if we use the NetFPGA route we will be using Ethernet as a test-bed).
 
-\paragraph{User-level RDMA:} Certain commodity networking hardware today
-supports remote direct memory access (RDMA). Unfortunately, \emph{how} it is
-implemented is so primitive and low-level that it is difficult or impossible
-to efficiently use in practice. Below we start with a short description of the
-current semantics of RDMA implementations. Next we describe the research we
-wish to undertake to make RDMA more useful.
+\paragraph{User-level RDMA:} Certain commodity networking hardware today supports remote direct memory access (RDMA).
+Unfortunately, \emph{how} it is implemented is so primitive and low-level that it is difficult or impossible to efficiently use in practice.
+Below we start with a short description of the current semantics of RDMA implementations.
+Next we describe the research we wish to undertake to make RDMA more useful.
 
-Networking infrastructure supports remote read and write operations. Read
-operations take a source address and key (on a remote node), a destination
-address and key (on the local node), and a length. Cards support address
-translation for regions of memory that are registered with them. These
-translations, indexed by the supplied key, are used to perform virtual to
-physical address translation. The physical address of the destination address
-is similarly translated using the translation table at the local node. Cards
-also support remote write operations with a similar set of semantics, where
-addresses are translated on a card using the card's local translation table.
-Both read and write operations are asynchronous. The requesting processor
-initiates the operation and then can check for completion by querying a
-completion queue at the local card. \emph{Notably}, the processor(s) on the
-remote node are provided a polling interface whereby they can learn if a
-remote write operation occurred. No such interface exists to learn about
-remote read or atomic operations.
+Networking infrastructure supports remote read and write operations.
+Read operations take a source address and key (on a remote node), a destination address and key (on the local node), and a length.
+Cards support address translation for regions of memory that are registered with them.
+These translations, indexed by the supplied key, are used to perform virtual to physical address translation.
+The physical address of the destination address is similarly translated using the translation table at the local node.
+Cards also support remote write operations with a similar set of semantics, where addresses are translated on a card using the card's local translation table.
+Both read and write operations are asynchronous.
+The requesting processor initiates the operation and then can check for completion by querying a completion queue at the local card.
+\emph{Notably}, the processor(s) on the remote node are provided a polling interface whereby they can learn if a remote write operation occurred.
+No such interface exists to learn about remote read or atomic operations.
 
-This interface suffers from multiple shortcomings. The biggest of which is
-that ultimately, moving data to and from a remote node constructively requires
-higher level semantic operations than simple reads and writes. These higher
-level semantics can, of course, be synthesized from the underlying read and
-write primitives, but it wastes round trips and synchronization time. One
-possible alternative solution is to enable remote nodes to dispatch high-level
-data-structure manipulation requests (e.g. insert this item into a queue,
-remove an item from a set). The challenge here is to make those calls
-programmable and interoperate securely with a modern operating system, and a
-user-level runtime system. Additionally there are questions about \emph{who}
-should carry out the computation. Classic RDMA does not interrupt the host
-processor, but these richer programmable semantics will require either a
-processor in the network interface (necessitating issues such as coherence be
-managed) or complex RDMA tasks need a mechanism to interrupt the host
-processor \emph{without leaving user space}.
+This interface suffers from multiple shortcomings.
+The biggest of which is that ultimately, moving data to and from a remote node constructively requires higher level semantic operations than simple reads and writes.
+These higher level semantics can, of course, be synthesized from the underlying read and write primitives, but it wastes round trips and synchronization time.
+One possible alternative solution is to enable remote nodes to dispatch high-level data-structure manipulation requests (e.g.
+insert this item into a queue, remove an item from a set).
+Past work on Active Messages~\cite{vonEicken92} provides the foundation on which to build these primitives.
+The challenge here is to make those calls programmable and interoperate securely with a modern operating system, and a user-level runtime system.
+Additionally there are questions about \emph{who} should carry out the computation.
+Classic RDMA does not interrupt the host processor, but these richer programmable semantics will require either a processor in the network interface or complex RDMA tasks need a mechanism to interrupt the host processor \emph{without leaving user space}.
 
-The second challenge is that the virtual to physical address translation is
-necessary for RDMA operations to be meaningful. Current systems require the OS
-to pin translations so cards can operate directly on physical addresses. This
-removes paging flexibility from the operating system and requires runtime
-systems to be aware that this is happening, which limits overall usability.
+The second challenge is that the virtual to physical address translation is necessary for RDMA operations to be meaningful.
+Current systems require the OS to pin translations so cards can operate directly on physical addresses.
+This removes paging flexibility from the operating system and requires runtime systems to be aware that this is happening, which limits overall usability.
 We will explore recent advances in supporting direct memory access with address translation (e.g., AMD IOMMU) to support flexible RDMA operation that go through full physical address translation without having to interrupt a general-purpose core.
-
-% In the long run, the solution here likely
-% lies at the intersection of what progress is occurring in processor support
-% for virtual machines and virtual memory translation for I/O. Our goal will
-% be to study carefully the I/O virtualization proposals put forth by major
-% processor manufactures and devise solutions that enable RDMA to capitalize
-% on these developments.
 
@@
 Hence we will need to write a custom disk controller device driver and file system that supports such a high degree of concurrent requests across so many parallel channels.
-
-%% Sorry Mark, while the item below is clearly rich and juicy, we can't leave it bhere because it is huuuuge, it does not match the scope of the grant, and we dont have the expertive --- e.g., are you sure nobody is working on it?
-
-%% I think this is the only idea that is going to work... maybe we should delete the above three and make this one good.  Seriously, the ideas up there are bunk.  -ark
-% \paragraph{Rearchitecting flash:} Clearly the best solution here lays in re-architecting the flash chip itself.
-We will explore new flash microarchitectures that optimize for message rate as well as bandwidth.
-Such chips have value far beyond Grappa.
-Research into persistent key/value stores, and really any persistent high-throughput data-structure will benefit from a redesigned flash chip that can sustain dramatically increased message rates.
 
@@
 However, this limits the amount of parallelism available for latency tolerance.
-A mmore efficient technique is to allow the compiler to unroll and schedule all loop bodies simultaneously within the same task, including overlapping remote memory operations, thereby ``weaving'' these independent tasks together to overlap computation with communication with minimal task overhead.
+A more efficient technique is to allow the compiler to unroll and schedule all loop bodies simultaneously within the same task, including overlapping remote memory operations, thereby ``weaving'' these independent tasks together to overlap computation with communication with minimal task overhead.
 Further, a set of ``woven tasks'' may be amenable to vector execution, allowing even more parallelism to be exploited.
 
-\paragraph{Task load control:} Having too few tasks per core fails to tolerate
-fully the latency; having too many increases overheads unnecessarily.
-Currently, we choose a number at program instantiation: experiments so far
-have shown evidence of a broad sweet-spot, though it is of course varies as a
-function of application behavior and input data. At compile-time, we might be
-able to estimate the number of tasks required based on the density of remote
-operations and system parameters (e.g., memory size, network latency, etc.);
-at runtime we can adjust the number of tasks dynamically based on periodic
-performance assessments.
-
-% Presently we are starting work on this language and compiler support.  We are basing our effort on the LLVM~\cite{llvm} infrastructure.
+\paragraph{Task load control:} Having too few tasks per core fails to fully tolerate remote memory latency; having too many increases overheads unnecessarily.
+Currently, we choose a fixed number at program instantiation.
+This approach is sufficient as a first step, but isn't a workable long term option.
+Our goal will be to develop as part of the runtime, auto-tuning technology that dynamically controls the number of spawned tasks.
 
@@
 This benchmark suite will consist of application kernels, reference implementations, data-sets, data-set generators, documentation, specifications for generating standardized performance results, and a hosting service (http://graphbench.org) to provide public access to the benchmark and datasets as well as aggregate and publish results by all users of the benchmark suite.
-Specifically, we aim to provide transparency and an opportunity to build off of others' discoveries by requiring source code be provided with any submission.
+We will start by adapting and building a collection of key computational kernels from graph and irregular applications.
+We will collect sample data from existing sources, and produce synthetic workload generators.
+The collected data provides a grounding for performance analysis on real-world problems.
+The workload generators provide cleaner, easier to reason about randomly generated data-sets at a programmable scale.
 
@@
 The goal is to get something out relatively early in the hope of (i) getting others to cluster around this work rather than develop a plethora of overlapping alternatives, and (ii) help establish consistent metrics for performance results.
-
-We will build a collection of key computational kernels from graph and irregular applications.
-The benchmark suite will consist of both documentation on the actual kernel (ideally a research paper or book reference sourced from the community), and a reference implementation.
-The benchmark suite will also include sample data collected from existing sources, and synthetic workload generators.
-The collected data provides a grounding for performance analysis on real-world problems.
-The workload generators provide cleaner, easier to reason about randomly generated data-sets at a programmable scale.
 

### 2013-02-14 doc/xps2013/proposal.tex
edits all over

--- 2013-02-14 doc/xps2013/proposal.tex
@@
 
-In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.
-This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating in software and inexpensive mass market hardware, the approach taken by Tera.
-We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.
-Ultimately, the opportunity is to cover the spectrum of irregular to regular computation: where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.
-Thus we make the best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.
+In the twenty years that have elapsed since the Tera MTA, commodity
+microprocessors have become much faster and multicore has driven down the
+absolute price of computation; commodity network price-performance has
+improved as well. This shift has afforded us the opportunity to attack the
+challenges posed by irregular applications by emulating in software and
+inexpensive mass market hardware, the approach taken by Tera. We exploit the
+increased aggregate instruction rate per socket relative to chip bandwidth,
+using what would otherwise be wasted instructions to manage the multiplexing
+of as many as several thousand tasks per core, thus tolerating memory latency,
+reducing stalls, and making better use of available bandwidth. Ultimately, the
+opportunity is to cover the spectrum of irregular to regular computation:
+where tasks exhibit locality, multiplex fewer tasks and expend fewer
+instructions on context switching; where locality is lacking, multiplex more
+tasks at a higher rate to tolerate latency. Thus we make the best use of task
+parallelism -- either to scale to more cores or to tolerate latency -- and of
+caches -- either to exploit application locality or to house more task
+contexts.
 
-Grappa is a software runtime system that allows a commodity x86 distributed-memory HPC cluster to be programmed as if it were a single large shared-memory machine and provides scalable performance for irregular applications.
-Grappa is designed to smooth over some of the performance discontinuities in commodity hardware, giving good performance when there is little locality to be exploited while allowing the programmer to exploit it when it is available.
+Grappa is a software runtime system that allows a commodity x86
+distributed-memory HPC cluster to be programmed as if it were a single large
+shared-memory machine and provides scalable performance for irregular
+applications. Grappa is designed to smooth over some of the performance
+discontinuities in commodity hardware, giving good performance when there is
+little locality to be exploited while allowing the programmer to exploit it
+when it is available.
 
@@
 
-The current Grappa prototype (Figure~\ref{fig:grappa}) has three main software components: \begin{description}
+The current Grappa prototype (Figure~\ref{fig:grappa}) has three main software components: 
+\begin{description}
 
@@
 
-\paragraph{Increasing involvement of underrepresented groups:} Computer science as a whole has a significant problem attracting women and minorities to the field.
-Computer architecture has an even more extreme problem with diversity (in the PI's view).
-This is a problem the PIs have taken quite seriously since entering the field, and we are quite proud of our PhD students, both female who have gone on to faculty positions in top academic institutions, and male who appreciate the gender and ethnic disparity and who work hard at their own institutions to correct it.
-We continue to recruit a diverse graduate student mix and we continue to include undergraduate students in our research.
-At the moment we are mentoring two women undergraduates involved in the Grappa project, and if this proposal is funded will apply for REU funds to continue their involvement.
-
 \paragraph{Benchmark suite:} As previously described, we are seeking funding support to build and distribute a benchmark suite for irregular and graph applications.
@@
 We taught a similar class over six years ago and it is time to revive, update and teach it to the next generation of computer architecture graduate students.
-\comment{Add more here.
-This has no passion.
--Mark}
+We will also reach out to instructors of classes that relate to large-scale irregular computation, e.g., machine learning and computer vision, in order to expose students to parallel programming concepts as well as the Grappa framework.
+
+\paragraph{Increasing involvement of underrepresented groups:} Computer science as a whole has a significant problem attracting women and minorities to the field.
+Computer architecture has an even more extreme problem with diversity (in the PI's view).
+This is a problem the PIs have taken quite seriously since entering the field, and we are quite proud of our PhD students, both female who have gone on to faculty positions in top academic institutions, and male who appreciate the gender and ethnic disparity and who work hard at their own institutions to correct it.
+We continue to recruit a diverse graduate student mix and we continue to include undergraduate students in our research.
+At the moment we are mentoring two women undergraduates involved in the Grappa project, and if this proposal is funded will apply for REU funds to continue their involvement.
 
@@
 \vspace{1ex} 
-\noindent\textbf{Year 3:} In the third and final year of this effort (supported by this grant), we will finish the networking layer rewrite.
-We will also have constructed a test-bed system comprised of mainly SSD-based memory.
-Grappa will be ported to this new style of memory, which we envision will largely be a tuning and not rewriting effort.
-We will explore alternative SSD-memory designs largely through emulation.
-Finally we will complete a detailed system evaluation using our benchmark suite.
+\noindent\textbf{Year 3:} In the third and final year of this
+effort (supported by this grant), we will finish the networking layer rewrite.
+We will also have constructed a test-bed system for the SSD-only data
+structure research. Grappa will be ported to this new style of memory
+hierarchy, which we envision will largely be a tuning and not rewriting
+effort. Finally we will complete a detailed system evaluation using our
+benchmark suite.
 
@@
   coadvised.
-\item In the last four years, the group has produced twelve papers
+\item In the last five years, the group has produced fifteen papers
   published in ASPLOS, ISCA, PLDI, OOPSLA, MICRO, and OSDI as well as three

### 2013-02-14 doc/xps2013/proposal.tex
edits all over

@@
   coadvised.
-\item In the last five years, the group has produced fifteen papers
+\item In the last five years, the group has produced over twenty papers
   published in ASPLOS, ISCA, PLDI, OOPSLA, MICRO, and OSDI as well as three
@@
 \item All PIs expect to solicit guidance and collaboration from
-  researchers such as John Feo (PNNL), Jim Larus (Microsoft Research), Hans
+  researchers such as John Feo (PNNL), David Callahan (Microsoft, recently moved to PNNL), Jim Larus (Microsoft Research), Hans
   Boehm (HP Research), and Preston Briggs,  with whom

### 2013-02-14 doc/xps2013/proposal.tex
cbm

--- 2013-02-14 doc/xps2013/proposal.tex
@@
 
-In the twenty years that have elapsed since the Tera MTA, commodity
-microprocessors have become much faster and multicore has driven down the
-absolute price of computation; commodity network price-performance has
-improved as well. This shift has afforded us the opportunity to attack the
-challenges posed by irregular applications by emulating in software and
-inexpensive mass market hardware, the approach taken by Tera. We exploit the
-increased aggregate instruction rate per socket relative to chip bandwidth,
-using what would otherwise be wasted instructions to manage the multiplexing
-of as many as several thousand tasks per core, thus tolerating memory latency,
-reducing stalls, and making better use of available bandwidth. Ultimately, the
-opportunity is to cover the spectrum of irregular to regular computation:
-where tasks exhibit locality, multiplex fewer tasks and expend fewer
-instructions on context switching; where locality is lacking, multiplex more
-tasks at a higher rate to tolerate latency. Thus we make the best use of task
-parallelism -- either to scale to more cores or to tolerate latency -- and of
-caches -- either to exploit application locality or to house more task
-contexts.
+In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.
+This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating in software and inexpensive mass market hardware, the approach taken by Tera.
+We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.
+Ultimately, the opportunity is to cover the spectrum of irregular to regular computation: where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.
+Thus we make the best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.
 
-Grappa is a software runtime system that allows a commodity x86
-distributed-memory HPC cluster to be programmed as if it were a single large
-shared-memory machine and provides scalable performance for irregular
-applications. Grappa is designed to smooth over some of the performance
-discontinuities in commodity hardware, giving good performance when there is
-little locality to be exploited while allowing the programmer to exploit it
-when it is available.
+Grappa is a software runtime system that allows a commodity x86 distributed-memory HPC cluster to be programmed as if it were a single large shared-memory machine and provides scalable performance for irregular applications.
+Grappa is designed to smooth over some of the performance discontinuities in commodity hardware, giving good performance when there is little locality to be exploited while allowing the programmer to exploit it when it is available.
 
@@
 The current Grappa prototype (Figure~\ref{fig:grappa}) has three main software components:
+
 \begin{description}
@@
 
-Grappa's lower networking layer aggregates the upper layer's messages to
-improve performance. Commodity networks including
-Infiniband~\cite{infinibandbandwidth} achieve their peak bisection bandwidth
-\emph{only} when the packet sizes are relatively large---on the order of
-multiple kilobytes. The reason for this discrepancy is the combination of
-overheads associated with handling each packet (in terms of bytes that form
-the actual packet, processing time at the card and processing on the CPU
-within the driver stack). Our measurements confirm manufacturers published
-data~\cite{infinibandbandwidth}, that with this packet size the bisection
-bandwidth is only a small fraction, less than 3\% of the peak bisection
-bandwidth.
+Grappa's lower networking layer aggregates the upper layer's messages to improve performance.
+Commodity networks including Infiniband~\cite{infinibandbandwidth} achieve their peak bisection bandwidth \emph{only} when the packet sizes are relatively large---on the order of multiple kilobytes.
+The reason for this discrepancy is the combination of overheads associated with handling each packet (in terms of bytes that form the actual packet, processing time at the card and processing on the CPU within the driver stack).
+Our measurements confirm manufacturers published data~\cite{infinibandbandwidth}, that with this packet size the bisection bandwidth is only a small fraction, less than 3\% of the peak bisection bandwidth.
 
-In our early experiments the vast majority of requests were smaller than 44
-bytes, far too small to make efficient use of the network. To make the best
-use of the network, we must convert our small messages into large ones. When a
-task sends a message, it is not immediately sent, but rather placed in a queue
-specific to the destination. If the size in bytes of a queue is above a
-threshold (e.g., 4096, typically large enough for good bandwidth), the
-contents of the queue are sent immediately. Also, each queue has a wait time
-threshold ($\approx${1ms}). If the oldest message in a queue has been waiting
-longer than this threshold, the contents of the queue are sent immediately,
-even if the queue size is lower than the message size threshold. Finally,
-queues may be explicitly flushed, for situations when latency is more
-important than bandwidth.
+In our early experiments the vast majority of requests were smaller than 44 bytes, far too small to make efficient use of the network.
+To make the best use of the network, we must convert our small messages into large ones.
+When a task sends a message, it is not immediately sent, but rather placed in a queue specific to the destination.
+If the size in bytes of a queue is above a threshold (e.g., 4096, typically large enough for good bandwidth), the contents of the queue are sent immediately.
+Also, each queue has a wait time threshold ($\approx${1ms}).
+If the oldest message in a queue has been waiting longer than this threshold, the contents of the queue are sent immediately, even if the queue size is lower than the message size threshold.
+Finally, queues may be explicitly flushed, for situations when latency is more important than bandwidth.
 
-Underneath the aggregation layer, Grappa uses the \gasnet~communication
-library~\cite{gasnet} to actually move data. All interprocess communication,
-whether on or off a cluster node, is handled by the \gasnet~library.
-\gasnet~is able to take advantage of many communication mechanisms, including
-ethernet and infiniband between nodes, as well as shared memory within a node.
+Underneath the aggregation layer, Grappa uses the \gasnet~communication library~\cite{gasnet} to actually move data.
+All interprocess communication, whether on or off a cluster node, is handled by the \gasnet~library.
+\gasnet~is able to take advantage of many communication mechanisms, including ethernet and infiniband between nodes, as well as shared memory within a node.
 
@@
 
-To evaluate the performance of our current prototype with respect to the XMT,
-we ran each of our three benchmarks on up to 16 nodes of each machine. Grappa
-used 6 cores per node, with the best parameters chosen for each point. In some
-cases, the XMT could not run the benchmark with 2 nodes, so the point is
-omitted.
+To evaluate the performance of our current prototype with respect to the XMT, we ran each of our three benchmarks on up to 16 nodes of each machine.
+Grappa used 6 cores per node, with the best parameters chosen for each point.
+In some cases, the XMT could not run the benchmark with 2 nodes, so the point is omitted.
 
@@
 \paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex
-tree (T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
-number of vertices visited per second versus number of compute nodes. Grappa
-is 3.2 times faster than the XMT at 16 nodes. As we will show later, the
-performance advantage Grappa has over XMT increases as more nodes are added.
-The main reason Grappa performs better is the software-based delegate
-synchronization obviates the need for the retry-based synchronization that XMT
-uses.
+tree (T1L).
+Figure~\ref{fig:uts_compare} shows the performance in terms of number of vertices visited per second versus number of compute nodes.
+Grappa is 3.2 times faster than the XMT at 16 nodes.
+The main reason Grappa performs better is the software-based delegate synchronization obviates the need for the retry-based synchronization that XMT uses.
 

### 2013-02-22 doc/xps2013/proposal.tex
minor edits

@@
 \vspace{1ex} 
-\noindent\textbf{Year 3:} In the third and final year of this
-effort (supported by this grant), we will finish the networking layer rewrite.
-We will also have constructed a test-bed system for the SSD-only data
-structure research. Grappa will be ported to this new style of memory
-hierarchy, which we envision will largely be a tuning and not rewriting
-effort. Finally we will complete a detailed system evaluation using our
-benchmark suite.
+\noindent\textbf{Year 3:} In the third and final year of this effort (supported by this grant), we will finish the networking layer rewrite.
+We will also have constructed a test-bed system for the SSD-only data structure research.
+Grappa will be ported to this new style of memory hierarchy.
+Finally we will complete a detailed system evaluation using our benchmark suite.
 
@@
 Irregular applications are increasing in importance to the computing industry.
-For a long time they have underpinned computations of interest to the national
-security arm of the federal government. But now, they underpin such questions
-as ad placement in social networks, and analysis of complex data-sets in
-medicine and science. The defining characteristic of these applications is
-poor locality and massive available parallelism. Our effort is focused on
-making these applications perform well and scale on commodity hardware. The
-key idea that makes this work is to rely on massive concurrency to tolerate
-memory latency, instead of relying on locality (which doesn't exist for these
-applications). This core idea underpinned the XMT~\cite{feo:xmt} system, a
-fully-custom hardware system. We plan to explore that idea in software only,
-running on commodity processors. And unlike the XMT, however, we must also
-contend with commodity networking hardware, and here we again rely on
-threading to provide a sufficient number of requests that we can buffer them
-prior to dispatching on the network, to overcome bandwidth limitations when
-operating with small messages.
+For a long time they have underpinned computations of interest to national security.
+But now, they underpin such questions as ad placement in social networks, and analysis of complex data-sets in medicine and science.
+The defining characteristic of these applications is poor locality and massive available parallelism.
+Our effort is focused on making these applications perform well and scale on commodity hardware.
+The key idea that makes this work is to rely on massive concurrency to tolerate memory latency, instead of relying on locality (which doesn't exist for these applications).
+This core idea underpinned the XMT~\cite{feo:xmt} system, a fully-custom hardware system.
+We plan to explore that idea in software only, running on commodity processors.
+And unlike the XMT, however, we must also contend with commodity networking hardware, and here we again rely on threading to provide a sufficient number of requests that we can buffer them prior to dispatching on the network, to overcome bandwidth limitations when operating with small messages.
 
-Our preliminary effort has shown a workable proof of concept of all this
-runtime smoke and mirrors: a commodity cluster can out do the XMT at its own
-game, by emulating its key features in software. What is needed now is
-additional effort to improve the underlying performance of this runtime and
-make it usable to a broad base of users. We will be focusing our efforts
-in the next three years on the key performance component, the networking
-layer, and the key programmer efficiency linchpin, high level language
-support. In addition we need a useful mechanism for handling data-sets in the
-petabyte range, and here we imagine building a cluster of machines that relies
-primarily on SSD's and not DRAM as main memory data-set storage.
+Our preliminary effort has shown a workable proof of concept of all this runtime smoke and mirrors: a commodity cluster can out do the XMT at its own game, by emulating its key features in software.
+What is needed now is additional effort to improve the underlying performance of this runtime and make it usable to a broad base of users.
+We will be focusing our efforts in the next three years on the key performance component, the networking layer, and the key programmer efficiency linchpin, high level language support.
+In addition we need a useful mechanism for handling data-sets in the petabyte range, and here we imagine building a cluster of machines that relies primarily on SSD's and not DRAM as main memory data-set storage.
 
-If successful, the Grappa runtime will have a transformative impact on the
-computing industry. Irregular applications are hugely important, and if off
-the shelf hardware can be efficient with them a new class of questions can be
-asked of computing systems. Additionally our benchmark suite development will
-further drive innovation among researchers in this space.
+If successful, the Grappa runtime will have a transformative impact on the computing industry.
+Irregular applications are hugely important, and if off the shelf hardware can be efficient with them a new class of questions can be asked of computing systems.
+Additionally our benchmark suite development will further drive innovation among researchers in this space.
 

### 2013-02-22 doc/xps2013/summary.tex

@@
 Irregular applications are increasing in importance to the computing industry.
-For a long time they have underpinned computations of interest to the national security arm of the federal government.
-But now, they underpin such questions as ad placement in social networks, and analysis of complex data-sets in medicine and science.
-The defining characteristic of these applications is poor locality and massive available parallelism.
-Our effort is focused on making these applications perform well and scale on commodity hardware.
-The key idea that makes this work is to rely on massive concurrency to tolerate memory latency, instead of relying on locality (which doesn't exist for these applications).
-This core idea underpinned the XMT system, a fully-custom hardware system.
-We plan to explore that idea in software only, running on commodity processors.
-And unlike the XMT, however, we must also contend with commodity networking hardware, and here we again rely on threading to provide a sufficient number of requests that we can buffer them prior to dispatching on the network, to overcome bandwidth limitations when operating with small messages.
+For a long time they have underpinned computations of interest to the national
+security arm of the federal government. But now, they underpin such questions
+as ad placement in social networks, and analysis of complex data-sets in
+medicine and science. The defining characteristic of these applications is
+poor locality and massive available parallelism. Our effort is focused on
+making these applications perform well and scale on commodity hardware. The
+key idea that makes this work is to rely on massive concurrency to tolerate
+memory latency, instead of relying on locality (which doesn't exist for these
+applications). This core idea underpinned the XMT system, a fully-custom
+hardware system. We plan to explore that idea in software only, running on
+commodity processors. And unlike the XMT, however, we must also contend with
+commodity networking hardware, and here we again rely on threading to provide
+a sufficient number of requests that we can buffer them prior to dispatching
+on the network, to overcome bandwidth limitations when operating with small
+messages.
 
@@
 We will be focusing our efforts in the next three years on the key performance component, the networking layer, and the key programmer efficiency linchpin, high level language support.
-In addition we need a useful mechanism for handling data-sets in the petabyte range, and here we imagine building a cluster of machines that relies primarily on SSD's and not DRAM as main memory data-set storage.
+In addition we need a useful mechanism for handling data-sets in the petabyte range, and here we imagine building a cluster of machines that relies primarily on SSD's and not DRAM as main  data-set storage.
 
 If successful, the Grappa runtime will have a transformative impact on the computing industry.
-Irregular applications are hugely important, and if off the shelf hardware can be efficient with them a new class of questions can be asked of computing systems.
+Irregular applications are hugely important, and if off-the-shelf hardware can be efficient with them a new class of questions can be asked of computing systems.
 Additionally our benchmark suite development will further drive innovation among researchers in this space.

### 2013-02-24 doc/sosp13/relatedwork.tex
sosp repo. put in sosp submission guidelines. new title

--- 2012-07-23 doc/asplos13/relatedwork.tex
@@
 rigid computation model where programmers must express computation as
-transformations on a vertex and its edge list only, with only
-information from adjacent vertexes. Pregel is only slightly less
+transformations on a vertex and its edge list only, with information only
+from adjacent vertexes. Pregel is only slightly less
 restrictive, as the input data can be any vertex in the graph.  Grappa

### 2013-02-24 doc/sosp13/tasks.tex
sosp repo. put in sosp submission guidelines. new title

--- 2012-07-23 doc/asplos13/tasks.tex
@@
 finds one with a non-zero amount of work in its public task queue.  The scheduler steals half of the tasks it finds at the victim.
+Work stealing is particularly interesting in Grappa since performance depends
+on having many active worker threads on each core. Even if there are many
+active threads, if they are all suspended on long-latency operations,
+then the core is underutilized. We choose to initiate one outstanding
+steal whenever the ready queue is empty there are idle worker threads.
 

### 2013-02-26 doc/nvm2013/paper.tex

@@
 We revisit Grappa, our software platform that scales in-memory irregular
-applications on commodity clusters.
+applications on mass market clusters.
 We contemplate options for adapting Grappa to use non-volatile memory technologies.
@@
 
-\item[Communication layer.] Modern commodity networks
+\item[Communication layer.] Modern mass market networks
 support high bandwidth only for large messages, but irregular applications
@@
 expression of fine-grained task parallelism at the application
-programming level, employing that concurrency as needed to sustain throughput in the presence of long latency operations and to better utilize commodity networks.
+programming level, employing that concurrency as needed to sustain throughput in the presence of long latency operations and to better utilize mass market networks.
 Programmers attend only to expressing the parallelism in their program, not to scheduling it.
@@
 temporal locality while maintaining a consistent global view.  With
-these features, we have demonstrated performance on commodity systems
+these features, we have demonstrated performance on mass market systems
 for highly irregular applications approaching that of the Cray \textls*{\small XMT}~\cite{feo:xmt}, a
@@
 this shared heap in Grappa programs are done through an \textls*{\small API}, so this change could be made transparent to the programmer.
+
 If an 
@@
 the access is to memory stored on another node, Grappa issues a
-message to that node requesting the data and context switches to other
+message to that node to request the data and context switches to other
 work until the reply arrives. Non-volatile memory would be accessed in
@@
 
-\subsection{Challenges in using commodity \textls*{\small SSD}\hspace{0.75pt}s}
+\subsection{Challenges in using mass market \textls*{\small SSD}\hspace{0.75pt}s}
 Grappa's goal is to provide good irregular application performance on
-mass market hardware. Unfortunately, commodity \textls*{\small SSD}\hspace{0.75pt}s are
-difficult to use in this way for three reasons:
+mass market hardware. Unfortunately, mass market \textls*{\small SSD}\hspace{0.75pt}s are
+difficult to use directly for three reasons:
 
@@
 less than 64 bytes. Today's \textls*{\small SSD}\hspace{0.75pt}s are optimized for reads in the
-kilobyte range.
+multi-kilobyte range.
 
@@
 are limited to less than 100K/s. Even the individual flash chips are slow, on the order of 40K/s~\cite{micronFlash}. In contrast, cores may be idle in
-Grappa if each node cannot make 100 million random reads per second of
+Grappa if each node cannot perform 100 million random reads per second of
 the global heap.
 
-%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
-\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
+%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Mass market \textls*{\small SSD}\hspace{0.75pt}s are usually treated
+\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Mass market \textls*{\small SSD}\hspace{0.75pt}s are usually treated
 as normal disks: accesses must occur through the kernel, incurring a
 high context-switch overhead. Grappa depends on user-mode access to
-the network layer. This approach has been applied to \textls*{\small SSD}\hspace{0.75pt}s in recent research~\cite{caulfield:2012}.
+the network layer to avoid this overhead; a similar approach has been applied to \textls*{\small SSD}\hspace{0.75pt}s in recent research~\cite{caulfield:2012}.
 
@@
 reasons. First, as discussed earlier, the \textls*{\small SSD} must accept requests at a rate sufficient to
-keep the cores busy; most \textls*{\small SSD}\hspace{0.75pt}s
+keep the cores busy; current \textls*{\small SSD}\hspace{0.75pt}s
 can accept requests at a rate two to four orders of magnitude slower than required.
@@
 processor with few cores might generate only 1 million references per
-second. This could be satisfied with only 25\,x\,25$\mu$s Flash chips.
+%second. This could be satisfied with only 25\,x\,25$\mu$s Flash chips.
+second. This could be satisfied with only $25\times 25\mu$s Flash chips. 
+%Similar have demonstrated energy-efficient operation in recent research~\cite{gordon:2009}.
 There are two problems with this design for our purpose. First, while each node
 requires fewer chips, the overall cluster performance goal, and thus
-chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
+chip count requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
 
@@
 increase locality and amortize the cost of Flash accesses. Requests
-to common flash pages would be bucketed by flash page address for a
+to nearby flash locations would be bucketed by flash page address for a
 limited time. When a sufficient number of requests have arrived for a page
@@
 provide more benefit. In this approach, graph data is constantly
-streamed between \textls*{\small SSD} and \textls*{\small DRAM}. Requests wait to execute until the
+streamed between \textls*{\small SSD} and \textls*{\small DRAM}. Requests would wait to execute until the
 relevant fragment of the graph is in memory. Request density is not a
@@
 
-While Grappa's goal so far has been to leverage commodity hardware, it
+While Grappa's goal so far has been to leverage mass market hardware, it
 is worth exploring the potential benefit of emerging non-volatile
@@
 
-\paragraph{\textls*{\small PCM} as Main Memory.}
-%\noindent{\bf \textls*{\small PCM} as Main Memory.}
+\paragraph{\textls*{\small PCM} as main memory.}
+%\noindent{\bf \textls*{\small PCM} as main memory.}
 Recent research has explored the use of \textls*{\small PCM} as a main memory

### 2013-02-28 doc/sosp13/abstract.tex
edits all over. some eval suggestions

Grappa is a runtime system for commodity clusters of multicore computers that
exposes a massively parallel and single address space abstraction to
applications. Grappa's goal is to provide scalable performance for irregular
parallel applications, such as graph processing. These applications are
challenging because they exhibit little data locality, unbalanced parallel
work and complex communication patterns.

Grappa serves both as a C++ user library and as a foundation upon which higher
level languages can be developed or adapted. Grappa tolerates delays to remote
memory by multiplexing thousands of lightweight tasks to each processor core;
balances load via fine-grained distributed work-stealing; takes full advantage
of network characteristics by aggregating smaller data requests into large
ones; and provides efficient synchronization and remote operations. We present
a detailed description of the Grappa system, programming examples using the
library interface, and compare performance for several irregular benchmarks to
hand-optimized MPI code and the Cray XMT, a custom system used to target the
real time graph analytics market.

### 2013-02-28 doc/sosp13/evaluation.tex
edits all over. some eval suggestions

--- 2012-07-23 doc/asplos13/evaluation.tex
@@
 
+\subsection{Scaling}
+% Got rid of this discussion of UTS by itself, tried to work the highlights in below
+%\paragraph{UTS-Mem}
+%We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
+%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
+%nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
+%Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
+%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
+%The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
+%scaling up through 128 nodes, although scaling
+%declines because of the unscalability of our aggregation mechanism as
+%number of network endpoints increases. 
+%
+%Despite our efforts to tune the UTS implementation specific to the 
+%XMT, performance does not scale well with increasing processor count,
+%flattening out around 60 processors.  When we increase the size of
+%the tree from 100M to 4.2B, we find that performance does not improve,
+%suggesting that performance is not limited by task parallelism.
+%Cray's performance tools show an increasing number of memory
+%retry operations for failed synchronization operations generated by
+%the runtime, which create network contention.
+%
+To determine how Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
+For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
+For Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
+All of the benchmarks continue to improve out to 128 nodes for Grappa.
+UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while Grappa handles this by suspending tasks until messages return.
+For BFS and Centrality, the XMT scales approximately a constant factor better than Grappa.
+We attribute this to a limitation in the current aggregator design and network stack that Grappa uses.
+This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases} performance).
+Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
+Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
+
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/scaling_cropped.pdf}
+    \end{center}
+    \caption{Scaling number of nodes: Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
+    \label{fig:uts_threshold}
+\end{figure}
+
 \subsection{Sensitivity}
@@
 
-\begin{figure}[htb]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/bfs_sweep_threshold}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:bfs-sweep-threshold} Sensitivity to parallel loop threshold. Note the log scale.}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
+%\begin{figure}[htb]
+%\begin{center}
+%  \includegraphics[width=0.95\columnwidth]{figs/bfs_sweep_threshold}
+%\begin{minipage}{0.95\columnwidth}
+%  \caption{\label{fig:bfs-sweep-threshold} Sensitivity to parallel loop threshold. Note the log scale.}
+%\end{minipage}
+%\vspace{-3ex}
+%\end{center}
+%\end{figure}
 
@@
 
-Figure~\ref{fig:bfs-sweep-threshold} shows a sweep of this parameter
-for a 32-node BFS run. We saw the best results serializing 32 or fewer
-iterations.
-
 % uts threshold
@@
 
-\subsection{Scaling}
-% Got rid of this discussion of UTS by itself, tried to work the highlights in below
-%\paragraph{UTS-Mem}
-%We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
-%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
-%nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
-%Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
-%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
-%The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
-%scaling up through 128 nodes, although scaling
-%declines because of the unscalability of our aggregation mechanism as
-%number of network endpoints increases. 
-%
-%Despite our efforts to tune the UTS implementation specific to the 
-%XMT, performance does not scale well with increasing processor count,
-%flattening out around 60 processors.  When we increase the size of
-%the tree from 100M to 4.2B, we find that performance does not improve,
-%suggesting that performance is not limited by task parallelism.
-%Cray's performance tools show an increasing number of memory
-%retry operations for failed synchronization operations generated by
-%the runtime, which create network contention.
-%
-To determine how Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
-For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
-For Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
-All of the benchmarks continue to improve out to 128 nodes for Grappa.
-UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while Grappa handles this by suspending tasks until messages return.
-For BFS and Centrality, the XMT scales approximately a constant factor better than Grappa.
-We attribute this to a limitation in the current aggregator design and network stack that Grappa uses.
-This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases} performance).
-Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
-Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
+\subsection{Characterization}
 
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/scaling_cropped.pdf}
-    \end{center}
-    \caption{Scaling number of nodes: Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
-    \label{fig:uts_threshold}
-\end{figure}
+\paragraph{Context-switch overhead.} Should we also compare with other packages? (Maybe Capriccio, QThreads?, or even real OS threads?)
+
+\paragraph{Latency.} Measure remote date access latency with and without aggregation turned on.
+
+\paragraph{Aggregated message sizes.} Characterization of the resulting message sizes with aggregation. Right now we only have message size vs. bandwidth.
+
+\paragraph{Utilization.} CPU utilization, Memory, Network. It would be great to answer the question of where is our bottleneck right now.
+
+\paragraph{Memory accesses.} Rate of accesses to remote data. Rate of delegate ops.
 

### 2013-02-28 doc/sosp13/intro.tex
edits all over. some eval suggestions

--- 2012-07-24 doc/asplos13/intro.tex
@@
 rate into the network is insufficient to utilize wire bandwidth when blocks
-are below about two-kilobytes, so any straightforward communication strategy
+are below about two-kilobytes, so any naive communication strategy
 severely under-utilizes the network.
@@
 Our evaluation of Grappa shows that it runs several graph-crunching
-applications (classic examples of irregular behavior) very efficiently
+applications very efficiently
 on a commodity cluster. Our yardstick for comparison is the XMT hardware

### 2013-03-10 doc/sosp13/memory.tex
sosp repo. put in sosp submission guidelines. new title
lots of edits. more coming

--- 2012-07-23 doc/asplos13/memory.tex
@@
 
-Applications written for Grappa utilize two forms of memory: local and
-global.  Local memory is local to a single core in the system.  Accesses occur
-through conventional pointers.  The compiler emits an access and the
-memory is manipulated directly.  Applications use local accesses for a
-number of things in Grappa: the stack associated with a task, accesses
-to localized global memory in caches (see below), and accesses to
-debugging infrastructure that is local to each system node.  Local
-pointers cannot access memory on other cores, and are valid only on
-their home core.
+Applications written for Grappa utilize two forms of memory: local and global.
+Local memory is local to a single core \comment{core or processor/node?} in
+the system. Accesses occur through conventional pointers. The compiler emits
+an access and the memory is manipulated directly. Applications use local
+accesses for a number of things in Grappa: the stack associated with a task,
+accesses to localized global memory in caches (see below), and accesses to
+debugging infrastructure that is local to each system node. Local pointers
+cannot access memory on other cores, and are valid only on their home core.
 
@@
   \begin{center}
+    \begin{minipage}{\columnwidth}
+	\small
+	\begin{tabular}{l}
+      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
+      	\texttt{\scriptsize global\_free( global\_address )} \\ \hline 
+      	Allocation in the global heap \\ \\
+      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
+      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
+      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
+      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ \hline
+      	Memory operations at the home core of a global address \\ \\
+      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
+      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ \hline
+		Perform cache operations to acquire/release global data.  \\
+		Acquire copies all data to local node and returns a pointer. \\ 	
+		Release optionally copies data back to global memory. \\
+	\end{tabular}
+      \caption{\label{fig:accessing-memory} Grappa API: accessing memory}     \end{minipage}
+  \end{center}
+\end{figure}
+
+\comment{
     \begin{description}\small
@@
       \item[ \texttt{ delegate\_read( global\_address, local\_var )} ] 
-      \item[ \texttt{ delegate\_write( global\_address, local\_var )} ] %\vspace{-2ex}
-      \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] %\vspace{-2ex}
-      \item[ \texttt{ delegate\_fetch\_inc( global\_address, local\_var )} ] %\vspace{-2ex} 
+      \item[ \texttt{ delegate\_write( global\_address, local\_var )} ] 
+      \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] 
+      \item[ \texttt{ delegate\_fetch\_inc( global\_address, local\_var )} ] 
 \hfill \\
@@
       \item[ \texttt{ cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} ]
-      \item[ \texttt{ cache\_release( global\_address, local\_buf )} ] %\vspace{-2ex} 
+      \item[ \texttt{ cache\_release( global\_address, local\_buf )} ] 
 \hfill \\
-        Perform cache operations \TODO{expand}
+Perform cache operations to acquire/release global data.
+Acquire, returns a local pointer after all data has been copied to the local node.
+Release, optionally writes data back to global memory and frees up management resources.
     \end{description}
-    \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:accessing-memory} Grappa API: accessing memory} %{-4ex}}
-    \end{minipage}
-    %\vspace{-3ex}
-  \end{center}
-\end{figure}
+}
 

### 2013-03-10 doc/sosp13/overview.tex
sosp repo. put in sosp submission guidelines. new title
edits all over. some eval suggestions
lots of edits. more coming

--- 2012-07-23 doc/asplos13/overview.tex
@@
 
-\item [Tasking system.] Our tasking system supports lightweight
-multithreading to tolerate communication latency and global distributed
-workstealing (i.e., tasks can be stolen from any node in the system), which
-provides automated load balancing.
+\item [Tasking system.] Our tasking system supports lightweight multithreading
+to tolerate communication latency and global distributed work-stealing (i.e.,
+tasks can be stolen from any node in the system), which provides automated
+load balancing. As discussed earlier, the reason we built our own custom
+tasking system is because of the need for large number of active tasks and the
+resulting need for careful task context prefetching.
+
+% \TODO{say why we built our own: need to support large number of contexts and just the contexts themselves wont fit in the cache, so being able to prefetch task contexts is very important.}
 
@@
 
+% \paragraph{Exploiting Latency Tolerance} As we will show later, Grappa can tolerate latencies way beyond that of the network.
+Therefore, Grappa can afford to \emph{trade latency for throughput}: by {\em increasing} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
+
+% \TODO{check to see we actually explain these three somewhere.
+Ie, that aggregating increases throughput and latency, that delegating synchronization increases the rate and the latency at which we can eg atomically increment, and that stealing work increases the latency of some tasks (eg when their data is largely at the originating node) but provides greater throughput overall.}
+
 In the next three sections we describe both Grappa's main capabilities and how they are implemented.

### 2013-03-10 doc/sosp13/tasks.tex
lots of edits. more coming

@@
 
-During execution, a task yields control of its core whenever it performs
-a long-latency operation, allowing the processor to remain busy while
-waiting for the operation to complete.  In addition,  a programmer can
-direct scheduling explicitly via the Grappa API calls shown
-in Figure~\ref{fig:scheduling}.
-To minimize yield overhead, the Grappa scheduler operates entirely in
-user-space and does little more than store register state of one task and load
-that of another. Context switch times are as low as 38ns even when switching
-amongst thousands of tasks.
+During execution, a task yields control of the processor core whenever it
+performs a long-latency operation, allowing the processor to remain busy while
+waiting for the operation to complete. In addition, a programmer can direct
+scheduling explicitly via the Grappa API calls shown in
+Figure~\ref{fig:scheduling}. To minimize yield overhead, the Grappa scheduler
+operates entirely in user-space and does little more than store register state
+of one task and load that of another. Context switch times are as low as 38ns
+even when switching amongst thousands of tasks.
 
@@
 
-\subsection{Expressing parallelism}
+\subsection{Expressing Parallelism}
 
@@
 
-Third, the programmer can invoke a parallel for loop, provided that
-the trip count is known at loop entry. The programmer
-specifies a function pointer along with start and end indices and an
-optional threshold to control parallel overhead. Grappa does {\em
-recursive decomposition} of iterations, similar to Cilk's cilk\_for
-construct~\cite {cilkforimplementation}.\comment{could only find slides
-page6 of
-http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-
-Cilk++.pdf    ---- Good enough.  Cite it.  -Mark} and TBB's {\tt parallel\_for}~\cite{intel_tbb}.  It generates a
-logarithmically-deep tree of tasks, stopping to execute the loop body
-when the number of iterations is below the required threshold.
+Third, the programmer can invoke a parallel for loop, provided that the trip
+count is known at loop entry. The programmer specifies a function pointer
+along with start and end indices and an optional threshold to control parallel
+overhead. Grappa does {\em recursive decomposition} of iterations, similar to
+Cilk's cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
+parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
+tasks, stopping to execute the loop body when the number of iterations is
+below the required threshold.
 
@@
 
-\subsection{Implementation tasks}
+\subsection{Task Support Implementation}
 
-\paragraph{Tasks and workers} Grappa tasks are 32-byte entities: a
-64-bit function pointer plus three 64-bit arguments. We use three
-arguments because tasks are often generated via a parallel loop
-decomposition, in which each task needs three kinds of data.
-\begin{description}
-\item[Function pointer] addresses the routine to run. We configure the system nodes to disable address randomization, so that function pointers are valid across process images.
-\item[Private argument] often a loop index.
-\item[Shared argument] typically data shared amongst a group of tasks, or the number of loop iterations to be performed by this task
-\item[Synchronization argument] often used to determine
-  when all tasks that are part of a loop have finished; a global pointer to a synchronization object allocated at the core that spawned a group of tasks.
-\end{description}
-While these are the most common uses of the three task arguments, they
-are treated as arbitrary 64-bit values in the runtime, and can be used
-for any purpose.
+\paragraph{Tasks and workers} Grappa tasks are 32-byte entities: a 64-bit
+function pointer plus three 64-bit arguments. We use three arguments because
+tasks are often generated via a parallel loop decomposition, in which each
+task needs three kinds of data. Thus, tasks include:
+
+\noindent {\it function pointer},
+or the addresses of the routine to run --- we configure the system nodes to
+disable address randomization, so that function pointers are valid across
+process images;
+
+\noindent {\it private argument}, often a loop index;
+
+\noindent {\it shared
+argument}, typically data shared amongst a group of tasks, or the number of
+loop iterations to be performed by this task;
+
+\noindent {\it synchronization argument}, often used to determine when all
+tasks that are part of a loop have finished, in the form of q global pointer
+to a synchronization object allocated at the core that spawned a group of
+tasks.
+
+While these are the most common uses of the three task arguments, they are
+treated as arbitrary 64-bit values in the runtime, and can be used for any
+purpose.
 
@@
 Each scheduler has three queues:
-\begin{description}
-\item{\bf Ready worker queue} This is a FIFO queue of tasks that are
-  matched with workers and are ready to execute.
-\item{\bf Private task queue} This is a FIFO queue of tasks that must run on this core.
-\item{\bf Public task queue} This is a LIFO queue of tasks that are
+
+\noindent {\it ready worker queue}, a FIFO queue of tasks that are
+  matched with workers and are ready to execute;
+
+\noindent {\it private task queue}, a FIFO queue of tasks that must run on this core;
+
+\noindent {\it public task queue},  a LIFO queue of tasks that are
   waiting to be matched with workers. It is a local partition of a shared
   task pool.
-\end{description}
 

### 2013-03-10 doc/sosp13/memory.tex
lots of edits. more coming

@@
 
-\paragraph{Global memory addressing} Grappa provides two methods for
-storing data in the global memory. The first is a distributed heap
-striped across all the machines in the system in a block cyclic fashion.
-The \texttt{global\_malloc} and \texttt{global\_free} calls are used to
-allocate and deallocate memory in the global heap.  Addresses to memory
-in the global heap use \textbf{linear addresses}.  Choosing the block
-size involves trading off sequential bandwidth against aggregate random
-access bandwidth. Smaller block sizes help spread data across all the
-memory controllers in the cluster, but larger block sizes allow the
-locality-optimized memory controllers to provide increased sequential
-bandwidth. The block size, which is configurable, is typically set to 64
-bytes, or the size of a single hardware cache line, in order to exploit
-spatial locality when available. The heap metadata is stored on a single
-node. Currently all heap operations serialize through this node; while
-this has been sufficient for our benchmarks, in the future Grappa will
-provide parallel performance through combining~\cite{MAMA,flatcombining}.
+\paragraph{Global memory addressing} Grappa provides two methods for storing
+data in the global memory. The first is a distributed heap striped across all
+the machines in the system in a block cyclic fashion. The
+\texttt{global\_malloc} and \texttt{global\_free} calls are used to allocate
+and deallocate memory in the global heap. Addresses to memory in the global
+heap use \textbf{linear addresses}. Choosing the block size involves trading
+off sequential bandwidth against aggregate random access bandwidth. Smaller
+block sizes help spread data across all the memory controllers in the cluster,
+but larger block sizes allow the locality-optimized memory controllers to
+provide increased sequential bandwidth. The block size, which is configurable,
+is typically set to 64 bytes, or the size of a single hardware cache line, in
+order to exploit spatial locality when available.
+
+%%% is the below important? -Luis
+\comment{The heap metadata is stored on a single node. Currently all heap
+allocation and free operations are serialized through this node; while this
+has been sufficient for our benchmarks, it is easy to relax in the future
+Grappa will provide parallel performance through
+combining~\cite{MAMA,flatcombining}.}
 
@@
 ID into a 2D global address. This address can then be accessed with a
-delegate and can also be cached by other nodes. At the destination the
-address is converted into a canonical x86 address by replacing the upper
+delegate operation and even be cached by other nodes. The
+address is converted at the destination into a canonical x86 address by replacing the upper
 bits with the sign-extended upper bit of the virtual address. 2D
@@
 \end{figure}
-
-\comment{
-    \begin{description}\small
-      \item[ \texttt{ global\_address global\_malloc( size )} ] \hfill \\
-      \item[ \texttt{ global\_free( global\_address )} ] \hfill \\
-        Allocates and frees memory in the global heap
-      \item[ \texttt{ delegate\_read( global\_address, local\_var )} ] 
-      \item[ \texttt{ delegate\_write( global\_address, local\_var )} ] 
-      \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] 
-      \item[ \texttt{ delegate\_fetch\_inc( global\_address, local\_var )} ] 
-\hfill \\
-        Performs a memory operation at the home core of a global address
-      \item[ \texttt{ cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} ]
-      \item[ \texttt{ cache\_release( global\_address, local\_buf )} ] 
-\hfill \\
-Perform cache operations to acquire/release global data.
-Acquire, returns a local pointer after all data has been copied to the local node.
-Release, optionally writes data back to global memory and frees up management resources.
-    \end{description}
-}
 

### 2013-03-13 doc/sosp13/memory.tex
edits through the memory section. added consistency model discussion

@@
       	\texttt{\scriptsize global\_address global\_malloc( size )} \\
-      	\texttt{\scriptsize global\_free( global\_address )} \\ \hline 
+      	\texttt{\scriptsize global\_free( global\_address )} \\ 
       	Allocation in the global heap \\ \\
@@
       	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
-      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ \hline
+      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ 
       	Memory operations at the home core of a global address \\ \\
       	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
-      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ \hline
+      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ 
 		Perform cache operations to acquire/release global data.  \\
@@
 	\end{tabular}
-      \caption{\label{fig:accessing-memory} Grappa API: accessing memory}     \end{minipage}
+      \caption{\label{fig:accessing-memory} Grappa API for memory accesses}     \end{minipage}
   \end{center}
@@
 \begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:memory-structure} Grappa memory structure}
+  \caption{\label{fig:memory-structure} Global memory referencing in Grappa}
 \end{minipage}
@@
 
-\textbf{Explicit caching.} Grappa provides an API to fetch a global
-pointer of any length and return a local pointer to a cached copy of the
-global memory.  Grappa cache operations have the usual read-only and
-read-write variants, along with a write-only variant used to initialize
-data structures. Languages for distributed shared memory systems have
-done optimizations to achieve a similar goal. For example, the UPC
-compiler coalesces struct and array accesses into remote get/put
-\cite{Chen:2005}, and Fortran D compiler's message vectorization hoists
-small messages out of a loop \cite{FortranD:1992}. Caching in Grappa
-additionally provides a mechanism for exploiting temporal locality by
-operating on the data locally.
+\textit{Explicit caching.} Grappa provides an API to fetch a global pointer of
+any length and return a local pointer to a cached copy of the global memory.
+Grappa cache operations have the usual read-only and read-write variants,
+along with a write-only variant used to initialize data structures. Languages
+for distributed shared memory systems have done optimizations to achieve a
+similar goal. For example, the UPC compiler coalesces struct and array
+accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's
+message vectorization hoists small messages out of a loop
+\cite{FortranD:1992}. Caching in Grappa additionally provides a mechanism for
+exploiting temporal locality when available by operating on the data locally.
 
-Under the hood, Grappa performs the mechanics of gathering chunks of
-data from multiple system nodes and presenting a conventional appearing
-linear block of memory as a local pointer into a cache. The strategy
-employed is to issue all the constituent requests of a cache access
-request  (as Active Messages) and then yield until all responses have
-occurred.  Currently, Grappa caches are \emph{not} coherent, requiring
-the programmer to maintain consistent access to data.  Future work will
-develop a software directory based coherence scheme to simplify
-consistent access to global data.
+Under the hood, Grappa performs the mechanics of gathering chunks of data from
+multiple system nodes and presenting a conventional appearing linear block of
+memory as a local pointer into a cache. The strategy employed is to issue all
+the constituent requests of the desired block (as Active Messages) and then
+yield until all responses have occurred. Currently, Grappa caches are
+\emph{not} kept coherent automatically. If that is desired, the programmer to
+is responsible for pushing the updated block back to its home node. Given that
+synchronization is explicit in Grappa, we don't expect this to be a
+significant burden (programmers can push updates before a relevant
+synchronization operation).
 
-\textbf{Delegate operations.} When the access pattern has low-locality,
+% Future work will develop a software directory based coherence scheme to
+% simplify consistent access to global data.
+
+\textit{Delegate operations.} When the access pattern has low-locality,
 it is more efficient to modify the data on its home core rather than
@@
 serialized through a single core in the system, providing atomic
-semantics without using atomic operations. 
+semantics without using actual atomic operations (and thus avoiding their typical high cost). 
 Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
@@
 
-\begin{figure*}[htb]
-\begin{center}
+\begin{figure*}[htb] \begin{center}
 \includegraphics[width=1.5\columnwidth]{figs/delegate-cache}
-\begin{minipage}{1.9\columnwidth}
-  \caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in
-the global heap. It then spawns two tasks on remote cores to increment
-elements of the array.  The first task increments two elements of the array using cache
-operations. In step 2, the task is invoked. A cache request is issued
-for two adjacent integers starting at the second element of the
-array. Since these element are stored in the memories of two different
-cores, this requires the sending of two messages in step 3. The task
-is suspended until both responses arrive in step 4. The data carried
-in these responses is stored in the local buffer. The elements are
-then increment in the buffer. Then in step 5, the modified data is
-sent back to the home node. Acknoledgements are returned in step 6 so
-the task knows when the writes are complete. The second task increments an element of the array with a delegate
-operation. In step 7, the task is invoked. A delegate request is sent
-to the home core of the array element with the increment value. The
-task suspends until the response is received. In step 8, the increment
-is executed on the remote core. A response is returned in step 9 with
-the previous value of the array element.
-}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure*}
+\begin{minipage}{1.9\columnwidth} \caption{\label{fig:delegate-cache}
+\textbf{Delegation and cache example:} A core allocates an array in the global
+heap (1). It then spawns two tasks on remote cores to increment elements of
+the array. The first task increments two elements of the array using cache
+operations. The first task is then invoked (2). A cache request is issued for two
+adjacent integers starting at the second element of the array. Since these
+element are stored in the memories of two different cores, this requires two messages to be sent (3). The task is suspended until both responses arrive
+(4). The data carried in these responses is stored in the local buffer. The
+elements are then incremented in the buffer. The modified data is then sent back
+to the home node (5). Acknowledgements are returned (6) so the task
+knows when the writes are complete. The second task increments an element of
+the array with a delegate operation. The task is invoked (7). A delegate
+request is sent to the home core of the array element with the increment
+value. The task suspends until the response is received. The increment
+is executed on the remote core (8). A response is returned (9) with the
+previous value of the array element. } \end{minipage} \vspace{-3ex}
+\end{center} \end{figure*}
+
+\paragraph{Memory consistency model discussion} As mentioned earlier, all
+synchronization operations are done via delegate operations. Since they all
+execute in their home node, they are guaranteed to be serialized , with their
+updates visible to all cores across the system in the same order. Also, tasks
+are restricted to a single outstanding delegate operation at a time, and
+therefore are not subject to potential request reordering. Consequently, all
+synchronization operations execute in program order and are made visible in
+the same order to all cores in the system. These properties are sufficient to
+guarantee a memory model that offers sequential consistency for data-race-free
+programs~\cite{drf} (all accesses to shared data are separated by
+synchronization). This is the memory model that underpins C/C++.
+
+Note, however, that if the application code uses explicit caching on shared
+data, all updates must be sent back to the home node before the
+synchronization operation that protects the data is performed.

### 2013-03-13 doc/sosp13/memory.tex
bib for memory consistency model

@@
 guarantee a memory model that offers sequential consistency for data-race-free
-programs~\cite{drf} (all accesses to shared data are separated by
-synchronization). This is the memory model that underpins C/C++.
+programs~\cite{AdveHill1990} (all accesses to shared data are separated by
+synchronization). This is the memory model that underpins C/C++~\cite{N2480,N2800}.
 

### 2013-03-14 doc/sosp13/communication.tex

--- 2012-11-06 doc/asplos13/communication.tex
@@
 In order to mitigate the low message injection rate limits of commodity
-networks, Grappa's communication stack has two layers: one for
-user-level messages and one for network-level messages.
+networks, Grappa's communication stack has two layers: one for user-level
+messages and one for network-level messages.
 
-At the upper layer, Grappa implements asynchronous active messages
-\cite{vonEicken92}. Each message consists of a function pointer, an
-optional argument payload, and an optional data payload. When a task
-sends a message, the message is copied to a send queue associated with
-the message's destination and the task continues execution.
+At the upper (user-level) layer, Grappa implements asynchronous active
+messages~\cite{vonEicken92}. Each message consists of a function pointer, an
+optional argument payload, and an optional data payload. When a task sends a
+message, the message is copied to a send queue associated with the message's
+destination and the task continues execution.
 
-Grappa's lower networking layer aggregates the upper layer's messages
-to improve performance. Commodity networks including infiniband
-achieves their peak bisection bandwidth \emph{only} when the packet
-sizes are relatively large---on the order of multiple kilobytes. The
-reason for this discrepancy is the combination of overheads associated
-with handling each packet (in terms of bytes that form the actual
-packet, processing time at the card and processing on the
-CPU within the driver stack). Our measurements confirm manufacturers
-published data~\cite{infinibandbandwidth}, that with this packet size
-the bisection bandwidth is only a small fraction, less than
-3\% of the peak bisection bandwidth.
+Grappa’s lower networking layer aggregates the upper layer’s messages to
+improve performance. Commodity networks including infiniband achieves their
+peak bisection bandwidth only when the packet sizes are relatively large ---
+on the order of multiple kilobytes. The reason for this discrepancy is the
+combination of overheads associated with handling each packet (in terms of
+bytes that form the actual packet, processing time at the card and processing
+on the CPU within the driver stack).
 
-In our experiments the vast majority of requests were smaller than
-44 bytes, far too small to make
-efficient use of the network. To make the best use of the network, we
-must convert our small messages into large ones.
-When a task sends a message,
-it is not immediately sent, but rather placed in a queue specific to the
+In our experiments the vast majority of requests were smaller than 44 bytes.
+Our measurements confirm manufacturers' published data [15]; with 44-byte
+packets, the available bisection bandwidth is only a small fraction (3\%) of
+the peak bisection bandwidth. To make the best use of the network, we must
+convert our small messages into large ones. When a task sends a message, it is
+not immediately sent, but rather placed in a queue specific to the
 destination.
 
-There are three situations in which a queue of aggregated messages is
-sent. First, each queue has a message size threshold of 4096 bytes, chosen to give reasonable network performance. If the size in
-bytes of a queue is above the threshold, the contents of the queue are
-sent immediately. Second, each queue has a wait time threshold
-($\approx${1ms}). If the oldest message in a queue has been waiting
-longer than this threshold, the contents of the queue are sent
-immediately, even if the queue size is lower than the message size
-threshold.  Third, queues may be explicitly flushed in situations where
-the programmer wants to minimize the latency of a message at the cost of
-bandwidth utilization.
+There are three situations in which a queue of aggregated messages is sent:
+(1) Each queue has a message size threshold of 4096 bytes, chosen to give
+reasonable network performance. If the size in bytes of a queue is above the
+threshold, the contents of the queue are sent immediately. (2) Each queue
+has a wait time threshold ($\approx${1ms}). If the oldest message in a queue
+has been waiting longer than this threshold, the contents of the queue are
+sent immediately, even if the queue size is lower than the message size
+threshold. (3) Queues may be explicitly flushed in situations where the
+programmer wants to minimize the latency of a message at the cost of bandwidth
+utilization.
 
@@
 limited to RDMA-capable networks.
+
+\TODO{We should expand on the zero-copy aggregation stuff?}

### 2013-03-14 doc/sosp13/memory.tex

@@
 	\end{tabular}
-      \caption{\label{fig:accessing-memory} Grappa API for memory accesses}     \end{minipage}
+      \caption{\label{fig:accessing-memory} Grappa API for memory accesses. \TODO{Update table with new calls. And async calls for delegates.}}     \end{minipage}
   \end{center}
@@
 \emph{fetch-and-add}).  Delegate operations, proposed in
-\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the
-primary synchronization method in Grappa.
+\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are used to implement synchronization in Grappa.
 
@@
 cache operations interact.
+
+\TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}
 

### 2013-03-14 doc/sosp13/relatedwork.tex

@@
 
-Grappa is built on many existing ideas in programming languages, systems and
-architecture. In this section we discuss related frameworks and key enabling
-technologies that Grappa builds upon.
+% Grappa is built on many existing ideas in programming languages, systems and
+% architecture. In this section we discuss related frameworks and key enabling
+% technologies that Grappa builds upon.
 
-\paragraph{Comparable frameworks} Distributed graph processing
+\paragraph{Distributed graph processing} While \Grappa is a general runtime
+system for any large-scale concurrent application, one of its killer
+applications is graph analysis. Therefore, distributed graph processing
 frameworks like Pregel~\cite{pregel:2010} and Distributed
-GraphLab~\cite{distgraphlab:vldb12} share similar goals as
-Grappa. Pregel adopts a bulk-synchronous parallel (BSP) execution
-model, which makes it inefficient on workloads that could prioritize
-vertices. GraphLab, on the other hand, schedules vertex computations
-individually, allowing prioritization, which gives faster convergence
-in a variety of iterative algorithms.  GraphLab, however, imposes a
-rigid computation model where programmers must express computation as
-transformations on a vertex and its edge list only, with information only
-from adjacent vertexes. Pregel is only slightly less
-restrictive, as the input data can be any vertex in the graph.  Grappa
-also supports dynamic parallelism with asynchronous execution, but
-parallelism is expressed as tasks or loop iterations, which is a far
-more general programming model for irregular computation tasks.
+GraphLab~\cite{distgraphlab:vldb12} are related. Pregel adopts a
+bulk-synchronous parallel (BSP) execution model, which makes it inefficient on
+workloads that could prioritize vertices. GraphLab, on the other hand,
+schedules vertex computations individually, allowing prioritization, which
+gives faster convergence in a variety of iterative algorithms. GraphLab,
+however, imposes a rigid computation model where programmers must express
+computation as transformations on a vertex and its edge list only, with
+information only from adjacent vertexes. Pregel is only slightly less
+restrictive, as the input data can be any vertex in the graph. Grappa also
+supports dynamic parallelism with asynchronous execution, but parallelism is
+expressed as tasks or loop iterations, which is a far more general programming
+model for irregular computation tasks.
 
-\paragraph{Global memory} Grappa includes a custom implementation of a
-software distributed shared memory (DSM) system. Many traditional
-software DSM systems are page based~\cite{Treadmarks,munin} and aim to
-hide the fact that they are built in software from applications by
-exploiting the processor's paging mechanisms, therefore relying
-heavily on locality. Instead, Grappa, like other partitioned global
-address space (PGAS) models, implements its DSM at the language,
-rather than system level. Languages such as
-Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and
-UPC~\cite{upc:2005} make accesses to shared structures look like
-normal memory references. As we describe later, Grappa chooses a
-middle ground, where global addresses are explicit in the API and
-local accesses are emitted conventionally by the compiler.  While
-Grappa's DSM system is conceptually similar to prior work, its
-implementation is tuned for irregular computations.  Past DSM work,
-being page-based, could exploit the RDMA capabilities of network
-hardware to move large page-sized blocks of data from node to node.
-In our experience, when these networks move small blocks of data (a
-few bytes), only a fraction of the available bandwidth is achieved.
-In addition, the DSM system in Grappa ends up being tightly coupled to
-the task scheduler in order to overlap long latency memory operations
-with useful computation.  For these two reasons it became necessary to
-build a new DSM system specifically for Grappa.
+\paragraph{Global memory} Grappa includes support for global memory namespace,
+effectively a custom implementation of a software distributed shared memory
+(DSM) system.
+
+Many traditional software DSM systems are page based~\cite{Treadmarks,munin}
+and aim to hide the fact that they are built in software from applications by
+exploiting the processor's paging mechanisms, therefore relying heavily on
+locality. Instead, Grappa, like partitioned global address space (PGAS)
+models, implements its DSM at the language, rather than the system level.
+Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and
+UPC~\cite{upc:2005} make accesses to shared structures look like normal
+references. As we describe later, Grappa chooses a middle ground, where global
+addresses are explicit in the API and local accesses are emitted
+conventionally by the compiler. Also, since past DSM work was mostly
+page-based, it could exploit network's RDMA support to move large page-sized
+blocks of data from node to node. However, when commodity networks move small
+blocks of data (a few bytes), only a fraction of the available bandwidth is
+achieved.
+
+In summary, while Grappa's DSM system is conceptually similar to prior work,
+it builds latency tolerance from the ground up (message aggregation, constant
+overlap of long latency memory operations with computation, etc.), so it can
+better support irregular computations with little locality.
+
+% Importantly, global memory support in Grappa
+% ends up being tightly coupled to the task scheduler in order to overlap long
+% latency memory operations with useful computation. For these two reasons it
+% became necessary to build a new DSM system specifically for Grappa.
 
@@
 
-\paragraph{Multithreading} Grappa uses multithreading to tolerate
-memory latency. This is a well known technique. Hardware
-implementations include the Tera MTA~\cite{tera:mta1}, Cray
-XMT~\cite{feo:xmt}, Simultaneous multithreading~\cite{tullsen:smt},
-MIT Alewife~\cite{agarwal:alewife}, Cyclops~\cite{almasi:cyclops}, and
-even GPUs~\cite{gpus}. As we describe in this paper, Grappa use a
-lightweight user-mode task scheduler to multiplex \emph{thousands\/} of
-tasks on a single processing core. The large number of tasks is
-required because of the extremely high internode latency Grappa is
-mitigating.  Grappa's task library employs several optimizations: an
-extremely fast task switch, a small task size, and judicious use of
-hardware prefetching to bring task state into the cache long before
-that task is actually scheduled.
+\paragraph{Multithreading} Grappa uses multithreading to tolerate memory
+latency. This is a well known technique. Hardware implementations include the
+Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous
+multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife},
+Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in
+this paper, Grappa implements its own software-based multithreading with a
+lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
+on a single processing core. The large number of tasks is required because of
+the to tolerate very high inter-node communication latency. Grappa's task
+library employs several optimizations: an extremely fast task switch, a small
+task size, and judicious use of hardware prefetching to bring task state into
+the cache long before that task is actually scheduled.
+The main difference between Grappa's support for lightweight threads and prior work such as QThread~\cite{qt} and Capriccio~\cite{capriccion} is the pervasive use of prefetching (with a large number of active tasks, not even their contexts fit in cache!).

### 2013-03-14 doc/sosp13/abstract.tex
minor mods to abstract, intro, 1st paragraph of relatedwork; added missing citations to bib. 	modified:   abstract.tex 	modified:   intro.tex 	modified:   paper.bib 	modified:   relatedwork.tex

@@
 Grappa is a runtime system for commodity clusters of multicore computers that
-exposes a massively parallel and single address space abstraction to
+presents a massively parallel, single address space abstraction to
 applications. Grappa's goal is to provide scalable performance for irregular
 parallel applications, such as graph processing. These applications are
-challenging because they exhibit little data locality, unbalanced parallel
+challenging because they exhibit little data locality, imbalanced parallel
 work and complex communication patterns.

### 2013-03-15 doc/sosp13/memory.tex
Modifications and notes in memory model/delegates section.

@@
 yield until all responses have occurred. Currently, Grappa caches are
-\emph{not} kept coherent automatically. If that is desired, the programmer to
-is responsible for pushing the updated block back to its home node. Given that
-synchronization is explicit in Grappa, we don't expect this to be a
-significant burden (programmers can push updates before a relevant
-synchronization operation).
+\emph{not} kept coherent automatically. If that is desired, the programmer
+is responsible for ensuring sufficient synchronization (via delegate
+operations) occurs before acquiring or releasing cache objects.
+% Given that synchronization is explicit in Grappa, we don't expect this to be a
+% significant burden (programmers can push updates before a relevant
+% synchronization operation).
 
@@
 restrict the use of delegate operations in three ways to make them more
-useful for synchronization. First, we limit each task to one outstanding
-delegate operation to avoid the possibility of reordering in the
-network. Second, we limit delegate operations to operate on objects in
+useful for synchronization.
+First, the remote operation may not perform any operations that could cause a context switch; this ensures any modifications are atomic due to cooperative multithreading.
+Second, we limit delegate operations to operate on objects in
 the 2D address space or objects that fit in a single block of the linear
 address space so they can be satisfied with a single network request.
-Finally, no context switches are allowed while the data is being
-modified. Given these restrictions, we can ensure that delegate
+Finally, the task issuing the delegate call blocks until the delegate operation
+completes to ensure that synchronization has finished before continuing.
+Using a ``futures''-like mechanism, tasks may issue delegates in parallel and block 
+on the ``promises'' returned by asynchronous delegate invocations.
+Given these restrictions, we can ensure that delegate
 operations for the same address from multiple requesters are always
@@
 \TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}
+\TODO{If we do want a section on GlobalCompletionEvent:
+  \begin{itemize}
+    \item many tasks and operations in the applications we've implemented do not need immediate consistency
+    \item guaranteeing ordering requires blocking a task. this is unavoidable if the task relies on the returned value, or if one task is dependent on others
+    \item however, much of the applications we've implemented (and other big-data things) are data-parallel, so tasks within a data-parallel loop can run in any order
+    \item in addition, often these data-parallel tasks will update values that aren't used until the next data-parallel loop
+    \item this allows us to free up workers/stacks (running out of stacks causes  latency-sensitivity because we can't continue until tasks complete)
+    \item must guarantee all operations that were supposed to happen in the loop complete before continuing, including tasks and delegate operations
+    \item \emph{motivating example:} write to array in one loop, use updated values in next loop (could also mention the ``PushBuffer'' example where we buffer queue appends because the order in the queue doesn't matter)
+    \item GlobalCompletionEvent is complicated by task-stealing and the fact that delegate operations complete on a different core than they started, forcing us to use a ``cancelable barrier''-like approach
+    \item interface to GlobalCompletionEvent:
+      \begin{itemize}
+        \item \texttt{gce.enroll(n)}: enroll n tasks/operations to be done, when goes from 0 to 1, leave cancelable barrier
+        \item \texttt{gce.complete(n)}: complete tasks, when it goes to zero, enter global cancelable barrier
+        \item \texttt{gce.wait()}: block until all cores have no work
+      \end{itemize}
+    \item to prevent early wakeups, stolen tasks must send a completion message back to the core they were started on
+  \end{itemize}
+}
 
@@
 are restricted to a single outstanding delegate operation at a time, and
-therefore are not subject to potential request reordering. Consequently, all
+therefore are not subject to potential request reordering \TODO{I think we can support multiple delegates in parallel from a task as long as we block on them before counting on them being complete.
+see my description above about the \emph{future/promise} delegates we have now.
+Not sure if there's a strong example of when this is useful for synchronization (though it's definitely useful for reads/writes), acquiring multiple locks doesn't work because we have to acquire locks in order to prevent deadlock.} Consequently, all
 synchronization operations execute in program order and are made visible in
@@
 data, all updates must be sent back to the home node before the
-synchronization operation that protects the data is performed.
+synchronization operation that protects the data is performed.\TODO{Not sure this warrants its own paragraph, seems like the explicit caches are just the same as reading and then writing using delegates.}

### 2013-03-16 doc/sosp13/tasks.tex
edits to communication, TODOs all over to reflect developments since the ASPLOS submission.
Add todo: update API examples with what's in "api_examples.cpp"

@@
 even when switching amongst thousands of tasks.
+\TODO{We should expand on the cleverness of our scheduler?}
 
@@
     \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:scheduling} Grappa API: scheduling} %{-4ex}}
+      \caption{\label{fig:scheduling} Grappa API: scheduling } %{-4ex}}
     \end{minipage}
@@
     \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism} % \vspace{-4ex}}
+      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism
+      \TODO{update this with latest C++ code from ``api\_examples.cpp''}
+      } % \vspace{-4ex}}
     \end{minipage}

### 2013-03-16 doc/sosp13/intro.tex

@@
 the network. Therefore, Grappa can afford to \emph{trade latency for
-throughput\/}: by {\em increasing\/} latency in key components of the system we
-are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations
-to remote nodes), and our ability to improve load imbalance (work stealing
-increases latency).
+throughput\/}. By {\em increasing\/} latency in key components of the system we
+are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate 
+%this is kind of subtle -- do we just assert it or instead provide
+%an inadequate explanation?  
+%we delegate synchronization operations on data even when it resides on the
+%local node so as to increase the rate at which we can perform
+%read-modify-write operations in the event of a hot-spot, at the expense
+%of increased synchronization latency as seen by tasks on that node. yuck.
+by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance by tolerating the delays incurred when work-stealing.
 

### 2013-03-16 doc/sosp13/memory.tex

@@
     \item \emph{motivating example:} write to array in one loop, use updated values in next loop (could also mention the ``PushBuffer'' example where we buffer queue appends because the order in the queue doesn't matter)
-    \item GlobalCompletionEvent is complicated by task-stealing and the fact that delegate operations complete on a different core than they started, forcing us to use a ``cancelable barrier''-like approach
+    \item GlobalCompletionEvent is complicated by task-stealing and the fact that delegate operations complete on a different core than they started, forcing us to use a ``cancelable barrier''-like approach (possible citation: Ravichandran, et al.
+``Work stealing for multi-core HPC Clusters'')
     \item interface to GlobalCompletionEvent:

### 2013-03-16 doc/sosp13/relatedwork.tex

@@
 
-\paragraph{Global memory} Grappa includes support for global memory namespace,
-effectively a custom implementation of a software distributed shared memory
-(DSM) system.
+\paragraph{Global memory} Grappa supports a global memory namespace, presenting a programming abstraction similar to that of previous software distributed shared memory (DSM) systems but with differences attuned to irregular application demands.
 
@@
 the cache long before that task is actually scheduled.
-The main difference between Grappa's support for lightweight threads and prior work such as QThread~\cite{qt} and Capriccio~\cite{capriccion} is the pervasive use of prefetching (with a large number of active tasks, not even their contexts fit in cache!).
+The main difference between Grappa's support for lightweight threads and prior work such as QThread~\cite{Wheeler08qthreads:an} and Capriccio~\cite{Behren03capriccio:scalable} is the pervasive use of prefetching (with a large number of active tasks, not even their contexts fit in cache!).

### 2013-03-16 doc/sosp13/tasks.tex

--- 2013-02-24 doc/sosp13/tasks.tex
@@
 
-During execution, a task yields control of its core whenever it performs
-a long-latency operation, allowing the processor to remain busy while
-waiting for the operation to complete.  In addition,  a programmer can
-direct scheduling explicitly via the Grappa API calls shown
-in Figure~\ref{fig:scheduling}.
-To minimize yield overhead, the Grappa scheduler operates entirely in
-user-space and does little more than store register state of one task and load
-that of another. Context switch times are as low as 38ns even when switching
-amongst thousands of tasks.
+During execution, a task yields control of the processor core whenever it
+performs a long-latency operation, allowing the processor to remain busy while
+waiting for the operation to complete. In addition, a programmer can direct
+scheduling explicitly via the Grappa API calls shown in
+Figure~\ref{fig:scheduling}. To minimize yield overhead, the Grappa scheduler
+operates entirely in user-space and does little more than store register state
+of one task and load that of another. Context switch times are as low as 38ns
+even when switching amongst thousands of tasks.
+\TODO{We should expand on the cleverness of our scheduler?}
 

### 2013-03-17 doc/sosp13/memory.tex
lots of edits to memory section.

--- 2013-03-14 doc/sosp13/memory.tex
@@
 Applications written for Grappa utilize two forms of memory: local and global.
-Local memory is local to a single core \comment{core or processor/node?} in
-the system. Accesses occur through conventional pointers. The compiler emits
-an access and the memory is manipulated directly. Applications use local
-accesses for a number of things in Grappa: the stack associated with a task,
-accesses to localized global memory in caches (see below), and accesses to
-debugging infrastructure that is local to each system node. Local pointers
-cannot access memory on other cores, and are valid only on their home core.
+Local memory is local to a single core within a node in the system. Accesses
+occur through conventional pointers. The compiler emits an access and the
+memory is manipulated directly. Applications use local accesses for a number
+of things in Grappa: the stack associated with a task, accesses to localized
+global memory in caches (see below), and accesses to debugging infrastructure
+that is local to each system node. Local pointers cannot access memory on
+other cores, and are valid only on their home core.
 
@@
       	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
+      	\texttt{\scriptsize delegate\_read\_async( global\_address, local\_var )}  \\
       	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
+      	\texttt{\scriptsize delegate\_write\_async( global\_address, local\_var )} \\
       	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
@@
 
-\paragraph{Global memory access} There are two general approaches
-Grappa applications use to {\emph access} global memory. When the
-programmer expects a computation on shared data to have spatial locality
-to exploit, {\em cache} operations may be used. When there is no
-locality to exploit, {\em delegate} operations are used.
+\paragraph{Global memory access} There are two general methods Grappa
+applications use to {\em access} global memory. One method is to use {\em
+delegate} operations, which are short memory accesses performed at the memory
+location's home node. This method is used when there is no locality in the
+access. The other method is to do {\em explicit caching} of memory regions,
+used when the programmer expects that the associated computation has data
+locality. Below we explain each in detail.
 
+\vspace{1ex} \textit{Delegate operations.} When the data access pattern has
+low-locality, it is more efficient to modify the data on its home core rather
+than bringing a copy to the requesting core and returning it after
+modification. Delegate operations~\cite{Nelson:hotpar11, delegated:oopsla11}
+provide this capability. Applications can dispatch computation to be performed
+on individual machine-word sized chunks of global memory to the memory system
+itself. Delegates are general, so we use them to perform simple
+\emph{read}/\emph{write} operations to global memory, as well as more complex \emph{read-modify-write} operations (e.g.,\emph{fetch-and-add}).
+
+Delegate operations are \emph{always} executed at the home core of their
+address. The remote operation may not perform any operations that could cause
+a context switch; this ensures any modifications are atomic. We limit delegate
+operations to operate on objects in the 2D address space or objects that fit
+in a single block of the linear address space so they can be satisfied with a
+single network request. Given these restrictions, we can ensure that delegate
+operations for the same address from multiple requesters are always serialized
+through a single core in the system, providing atomic semantics without using
+actual atomic operations (and thus avoiding their typical high cost).
+
+% We restrict the use of delegate operations in order to make them also
+% useful for synchronization operations.
+
+Delegate operations can be either {\em synchronous} or {\em asynchronous}.
+With synchronous operations, the task issuing the delegate call blocks until
+the delegate operation completes, which is necessary, for example, to ensure
+that synchronization has finished before continuing. Synchronization in Grappa
+is implemented using synchronous delegate operations. Often remote data
+accesses do not need to be performed synchronously. In order to avoid
+unnecessary waiting, we support asynchronous delegate operations. For reads,
+we support a ``futures''-like mechanism. Tasks may issue delegate reads in
+parallel and block on the ``promises'' returned by asynchronous delegate
+invocations. Delegate write operations may also be performed asynchronously,
+and since they do not return data, they need a mechanism to detect completion.
+Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
+outstanding asynchronous operations of the local core. The code and can probe
+the state of this object to determine when all outstanding requests are
+complete. This is necessary to, say, make sure all outstanding operations are
+complete before executing a synchronization operation.
+
+% \TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}
+% \TODO{If we do want a section on GlobalCompletionEvent:
+%   \begin{itemize}
+%     \item many tasks and operations in the applications we've implemented do not need immediate consistency
+%     \item guaranteeing ordering requires blocking a task. this is unavoidable if the task relies on the returned value, or if one task is dependent on others
+%     \item however, much of the applications we've implemented (and other big-data things) are data-parallel, so tasks within a data-parallel loop can run in any order
+%     \item in addition, often these data-parallel tasks will update values that aren't used until the next data-parallel loop
+%     \item this allows us to free up workers/stacks (running out of stacks causes  latency-sensitivity because we can't continue until tasks complete)
+%     \item must guarantee all operations that were supposed to happen in the loop complete before continuing, including tasks and delegate operations
+%     \item \emph{motivating example:} write to array in one loop, use updated values in next loop (could also mention the ``PushBuffer'' example where we buffer queue appends because the order in the queue doesn't matter)
+%     \item GlobalCompletionEvent is complicated by task-stealing and the fact that delegate operations complete on a different core than they started, forcing us to use a ``cancelable barrier''-like approach (possible citation: Ravichandran, et al.
+``Work stealing for multi-core HPC Clusters'')
+%     \item interface to GlobalCompletionEvent:
+%       \begin{itemize}
+%         \item \texttt{gce.enroll(n)}: enroll n tasks/operations to be done, when goes from 0 to 1, leave cancelable barrier
+%         \item \texttt{gce.complete(n)}: complete tasks, when it goes to zero, enter global cancelable barrier
+%         \item \texttt{gce.wait()}: block until all cores have no work
+%       \end{itemize}
+%     \item to prevent early wakeups, stolen tasks must send a completion message back to the core they were started on
+%   \end{itemize}
+% }
+
+\vspace{1ex}
 \textit{Explicit caching.} Grappa provides an API to fetch a global pointer of
@@
 yield until all responses have occurred. Currently, Grappa caches are
-\emph{not} kept coherent automatically. If that is desired, the programmer to
-is responsible for pushing the updated block back to its home node. Given that
-synchronization is explicit in Grappa, we don't expect this to be a
-significant burden (programmers can push updates before a relevant
-synchronization operation).
+\emph{not} kept coherent automatically. If that is desired, the programmer is
+responsible for ensuring sufficient synchronization (via delegate operations)
+occurs before acquiring or releasing cache objects.
 
-% Future work will develop a software directory based coherence scheme to
-% simplify consistent access to global data.
-
-\textit{Delegate operations.} When the access pattern has low-locality,
-it is more efficient to modify the data on its home core rather than
-bringing a copy to the requesting core and returning it after
-modification. Delegate operations provide this capability. Applications
-can dispatch computation to be performed on individual machine-word
-sized chunks of global memory to the memory system itself (e.g.,
-\emph{fetch-and-add}).  Delegate operations, proposed in
-\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are used to implement synchronization in Grappa.
-
-Delegate operations are always executed at the home core of their
-address, and while arbitrary memory operations can be delegated, we
-restrict the use of delegate operations in three ways to make them more
-useful for synchronization. First, we limit each task to one outstanding
-delegate operation to avoid the possibility of reordering in the
-network. Second, we limit delegate operations to operate on objects in
-the 2D address space or objects that fit in a single block of the linear
-address space so they can be satisfied with a single network request.
-Finally, no context switches are allowed while the data is being
-modified. Given these restrictions, we can ensure that delegate
-operations for the same address from multiple requesters are always
-serialized through a single core in the system, providing atomic
-semantics without using actual atomic operations (and thus avoiding their typical high cost). 
-Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
-cache operations interact.
-
-\TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}
+Figure~\ref{fig:delegate-cache} depicts an example how explicit caching and
+delegate operations work and how they interact.
 
@@
 synchronization operations are done via delegate operations. Since they all
-execute in their home node, they are guaranteed to be serialized , with their
+execute in their home node, they are guaranteed to be serialized, with their
 updates visible to all cores across the system in the same order. Also, tasks
-are restricted to a single outstanding delegate operation at a time, and
-therefore are not subject to potential request reordering. Consequently, all
-synchronization operations execute in program order and are made visible in
-the same order to all cores in the system. These properties are sufficient to
-guarantee a memory model that offers sequential consistency for data-race-free
-programs~\cite{AdveHill1990} (all accesses to shared data are separated by
-synchronization). This is the memory model that underpins C/C++~\cite{N2480,N2800}.
+are restricted to a single outstanding synchronous delegate operation at a
+time, which is how synchronization is done. Therefore, synchronization operations are not subject to potential reordering. 
+% \TODO{I think we can support multiple delegates in parallel from a task as
+% long as we block on them before counting on them being complete. see my
+% description above about the \emph{future/promise} delegates we have now. Not
+% sure if there's a strong example of when this is useful for synchronization
+% (though it's definitely useful for reads/writes), acquiring multiple locks
+% doesn't work because we have to acquire locks in order to prevent deadlock.}
+Consequently, all synchronization operations execute in program order and are
+made visible in the same order to all cores in the system. These properties
+are sufficient to guarantee a memory model that offers sequential consistency
+for data-race-free programs~\cite{AdveHill1990} (all accesses to shared data
+are separated by synchronization). This is the memory model that underpins
+C/C++~\cite{N2480,N2800}.
 
-Note, however, that if the application code uses explicit caching on shared
-data, all updates must be sent back to the home node before the
-synchronization operation that protects the data is performed.
+Note, however, that if the application code uses explicit caching or
+asynchronous delegates to access shared data, all updates must be sent back to
+the home node before the synchronization operation that protects the data is
+performed. This is done using release operations on cached regions and using the \texttt{GlobalCompletionEvent} object to determine that asynchronous delegates have been fully executed.
+
+% \TODO{Not sure this warrants its own paragraph, seems like the explicit caches are just the same as reading and then writing using delegates.}

### 2013-03-21 doc/sosp13/intro.tex
small-medium edits over introduction. Made the opportunity section stronger

@@
 
-Irregular applications generate tasks with work, interdependences, or memory
+Irregular applications generate tasks with workloads, interdependences, or memory
 accesses that are highly sensitive to input. Classic examples of irregular
@@
 computing domains. For these emerging applications, fast response -- given the
-sheer amount of data -- requires multinode systems. The most broadly available
+sheer amount of data -- requires multi-node systems. The most broadly available
 multinode systems are those built from x86 commodity computing nodes
@@
 \noindent{\bf Irregular applications exhibit little spatial locality.}
-It is not atypical for any given task's data references to be spread
-randomly across the entire memory of the system. This makes current
+It is common for tasks to perform random access to data that is spread
+across the entire memory of the system. This makes current
 memory hierarchy features ineffective. Caches are of little assistance
@@
 
-\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low
+\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multi-node systems, the challenges presented by low
 locality are analogous, and exacerbated by the increased latency of going
@@
 While some irregular applications can be restructured to better exploit
-locality, aggregating requests to increase message size, and manage the
+locality, aggregate requests to increase message size, and manage the
 additional challenges of load balance and synchronization across multinode
@@
 hardware thread contexts, a sufficient number to fully hide memory
-access latency.  The network was designed with a single word injection
+access latency.  The network was designed with a single-word injection
 rate that matched the processor clock frequency and sufficient bandwidth
@@
 Unfortunately, while an excellent match to extremely irregular
-applications, the MTA was not cost-effective on applications that could
-exploit locality and had very poor single-thread performance, making
-it a commercial failure. The Cray XMT approximates the Tera MTA-2,
-substantially reducing its cost but not overcoming its narrow range of
+applications, it became a commercial failure for two reasons. First, the MTA was not cost-effective on applications that could
+exploit locality. Second, it had very poor single-thread performance: it lacked a data cache and even without memory references when
+there was one thread the processor ran at 21 cycles per instruction.
+The Cray XMT approximates the Tera MTA-2, substantially reducing its cost but not overcoming its narrow range of
 applicability.
@@
 In the twenty years that have elapsed since the Tera MTA, commodity
-microprocessors have become much faster and multicore has driven down the
+microprocessors have become much faster, and multi-core has driven down the
 absolute price of computation; commodity network price-performance has
@@
 challenges posed by irregular applications by emulating in software and
-inexpensive mass market hardware, the approach taken by Tera. We exploit the
-increased aggregate instruction rate per socket relative to chip bandwidth,
-using what would otherwise be wasted instructions to manage the multiplexing
-of as many as several thousand tasks per core, thus tolerating memory latency,
-reducing stalls, and making better use of available bandwidth. Ultimately, the
+inexpensive mass market hardware, the approach taken by Tera. The aggregate instruction rate per socket in commodity processors is unmatched by
+the off-chip bandwidth for workloads with a lot of communication. 
+We can use this otherwise wasted instruction bandwidth on the overhead of multiplexing
+of thousands of tasks on each core. The concurrency provided by
+multithreading enables us to make better use of available network bandwidth. Ultimately, the
 opportunity is to cover the spectrum of irregular to regular computation:
@@
 a commodity x86 distributed-memory HPC cluster to be programmed as if it
-were a single large shared-memory machine and provides scalable
+were a single, large, NUMA shared-memory machine and provides scalable
 performance for irregular applications. Grappa is designed to smooth
@@
 invocations. To this commodity hardware and software mix Grappa adds three
-main software components: (1) a \emph{lightweight tasking\/} layer that supports
-a context switch in as few as 38ns and distributed global load balancing; (2)
-a \emph{distributed shared memory\/} layer that supports normal access operations
-such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer
-that combines short messages to mitigate the aforementioned problem that
-commodity networks are designed to achieve peak bandwidth only on large packet
-sizes, yet irregular applications tend to fetch only a handful of bytes at a
-time. As we will show later, Grappa can tolerate latencies way beyond that of
+main software components: 
+\begin{enumerate}
+\item a \emph{lightweight tasking\/} layer that supports
+a context switch in as few as 38ns and distributed global load balancing
+\item a \emph{distributed shared memory\/} layer that supports normal access operations
+such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}
+\item a \emph{message aggregation\/} layer that combines short messages to achieve peak bandwidth on commodity netorks
+%justification for aggregation doesn't belong here:
+%utilization  the aforementioned problem that
+%commodity networks are designed to achieve peak bandwidth only on large packet
+%sizes, yet irregular applications tend to fetch only a handful of bytes at a
+%time. 
+\end{enumerate}
+As we will show later, Grappa can tolerate latencies way beyond that of
 the network. Therefore, Grappa can afford to \emph{trade latency for
@@
 
-Our evaluation of Grappa shows that it runs several graph-crunching
+Our evaluation of Grappa shows that it runs several graph-crunching \TODO{expand this category if we have sort}
 applications very efficiently
@@
 performance.
+\TODO{revise paragraph after eval, including MPI and possibly programming model}
 

### 2013-03-22 doc/sosp13/relatedwork.tex
Fixed malformed sentence 	modified:   sosp13/relatedwork.tex
related work: mention powergraph and complimentary to grappa, strengthen DSM conclusion, clarify threading

@@
 \paragraph{Distributed graph processing} While \Grappa is a general runtime
-system for any large-scale concurrent application, one of its killer
-applications is graph analysis. Therefore, distributed graph processing
+system for any large-scale concurrent application, one of its \checkme{killer
+applications is graph analysis}. Therefore, distributed graph processing
 frameworks like Pregel~\cite{pregel:2010} and Distributed
@@
 bulk-synchronous parallel (BSP) execution model, which makes it inefficient on
-workloads that could prioritize vertices. GraphLab, on the other hand,
-schedules vertex computations individually, allowing prioritization, which
+workloads that could prioritize vertices. GraphLab overcomes this limitation with an execution mode
+that schedules vertex computations individually, allowing prioritization, which
 gives faster convergence in a variety of iterative algorithms. GraphLab,
@@
 expressed as tasks or loop iterations, which is a far more general programming
-model for irregular computation tasks.
+model for irregular computation tasks. PowerGraph~\cite{powergraph:osdi12} improves
+the performance of GraphLab for real-world graphs with power-law degree distributions by
+using a vertex cut for graph partitioning. If the size of the partial result of 
+applying a subset of edge data to the vertex is constant, then there is a good bound on the communication
+required to keep vertex replicas coherent. We do not explore these algorithmic techniques
+in this work, but they are complimentary and would improve the performance of graph applications on Grappa.
 
@@
 addresses are explicit in the API and local accesses are emitted
-conventionally by the compiler. Also, since past DSM work was mostly
-page-based, it could exploit network's RDMA support to move large page-sized
-blocks of data from node to node. However, when commodity networks move small
+conventionally by the compiler.
+
+Past DSM work was mostly page-based, so it could take advantage of a network's RDMA support to move large page-sized
+blocks of data from node to node \TODO{citation}. However, when commodity networks move small
 blocks of data (a few bytes), only a fraction of the available bandwidth is
-achieved.
+achieved. \TODO{probably want to reframe this discussion in view of Grappa using RDMA differently}
 
 In summary, while Grappa's DSM system is conceptually similar to prior work,
-it builds latency tolerance from the ground up (message aggregation, constant
-overlap of long latency memory operations with computation, etc.), so it can
-better support irregular computations with little locality.
+we accept the random access aspect of irregular applications and optimize for throughput rather than low latency.
+Our DSM must support enough memory concurrency to tolerate the latency of the network.
 
@@
 lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
-on a single processing core. The large number of tasks is required because of
-the to tolerate very high inter-node communication latency. Grappa's task
-library employs several optimizations: an extremely fast task switch, a small
-task size, and judicious use of hardware prefetching to bring task state into
-the cache long before that task is actually scheduled.
-The main difference between Grappa's support for lightweight threads and prior work such as QThread~\cite{Wheeler08qthreads:an} and Capriccio~\cite{Behren03capriccio:scalable} is the pervasive use of prefetching (with a large number of active tasks, not even their contexts fit in cache!).
+on a single processing core. The large number of tasks is required to tolerate
+the very high inter-node communication latency of commodity networks. Grappa's
+task library employs several optimizations: an extremely fast task switch, a
+small task size, and judicious use of software prefetch instructions to bring task state
+into the cache sufficiently long before that task is actually scheduled. The main difference
+between Grappa's support for lightweight threads and prior work such as
+QThreads~\cite{Wheeler08qthreads:an} and
+Capriccio~\cite{Behren03capriccio:scalable} is the use of prefetching by the scheduler, which is needed for good performance
+when multiplexing such a large number of tasks.

### 2013-03-22 doc/sosp13/overview.tex
discuss prefetching in overview; tighten up DSM summary, still may be too vague

@@
 
-\item [Tasking system.] Our tasking system supports lightweight multithreading
+\item [Tasking system.] The tasking system supports lightweight multithreading
 to tolerate communication latency and global distributed work-stealing (i.e.,
 tasks can be stolen from any node in the system), which provides automated
-load balancing. As discussed earlier, the reason we built our own custom
-tasking system is because of the need for large number of active tasks and the
-resulting need for careful task context prefetching.
+load balancing. The scheduler oversubscribes to have more worker
+threads than required for latency tolerance. By having at least
+several threads ready to run at all times, the scheduler can prefetch
+a few stack frames into cache to lower the likelihood of costly main
+memory accesses on every context switch.
 
-% \TODO{say why we built our own: need to support large number of contexts and just the contexts themselves wont fit in the cache, so being able to prefetch task contexts is very important.}
-
-\item[Distributed shared memory.] Our DSM system provides support for
+\item[Distributed shared memory.] The DSM system provides
 fine-grain access to data anywhere in the system. It supports synchronization
 operations on global data, explicit local caching of any memory in the system,
-and support for operation on remote data (delegating operations to home node).
-By tight integration with the tasking system and the
-communication layer, our DSM system offers high aggregate random
+and support for operations on remote data.
+The DSM system design relies on the lightweight tasking system and
+communication layer in order to offer high aggregate random
 access bandwidth for accessing remote data.
 
-\item[Communication layer.] As discussed earlier, modern commodity networks
+\item[Communication layer.] Modern commodity networks
 support high bandwidth only for large messages. Since irregular applications

### 2013-03-22 doc/sosp13/evaluation.tex
little tweaks

@@
 We attribute this to a limitation in the current aggregator design and network stack that Grappa uses.
-This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases} performance).
+This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).
 Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
@@
 
-\paragraph{Latency.} Measure remote date access latency with and without aggregation turned on.
+\paragraph{Latency.} Measure remote data access latency with and without aggregation turned on.
 
@@
 
-\paragraph{Utilization.} CPU utilization, Memory, Network. It would be great to answer the question of where is our bottleneck right now.
+\paragraph{Utilization.} CPU utilization, Memory, Network. It would be great to answer the question of where is our bottleneck right now. Amount of concurrency with and without aggregation.
 
-\paragraph{Memory accesses.} Rate of accesses to remote data. Rate of delegate ops.
+\paragraph{Memory accesses.} Rate of accesses to remote data. Rate of delegate ops. Show limit with GUPs.
 

### 2013-03-22 doc/sosp13/memory.tex
little tweaks

@@
 itself. Delegates are general, so we use them to perform simple
-\emph{read}/\emph{write} operations to global memory, as well as more complex \emph{read-modify-write} operations (e.g.,\emph{fetch-and-add}).
+\emph{read\/}/\emph{write\/} operations to global memory, as well as more complex \emph{read-modify-write\/} operations (e.g., \emph{fetch-and-add\/}).
 
-Delegate operations are \emph{always} executed at the home core of their
+Delegate operations are \emph{always\/} executed at the home core of their
 address. The remote operation may not perform any operations that could cause
@@
 yield until all responses have occurred. Currently, Grappa caches are
-\emph{not} kept coherent automatically. If that is desired, the programmer is
+\emph{not\/} kept coherent automatically. If that is desired, the programmer is
 responsible for ensuring sufficient synchronization (via delegate operations)

### 2013-03-22 doc/sosp13/tasks.tex
little tweaks

@@
 optional threshold to control parallel overhead. Grappa does {\em
-recursive decomposition} of iterations, similar to Cilk's cilk\_for
+recursive decomposition\/} of iterations, similar to Cilk's cilk\_for
 construct~\cite {cilkforimplementation}.\comment{could only find slides
@@
 decides to run them; when this occurs, tasks are matched with {\em
-  worker} threads. Each worker is simply a collection of status bits and a
+  worker\/} threads. Each worker is simply a collection of status bits and a
 stack, allocated at a particular core.

### 2013-03-22 doc/sosp13/tasks.tex
touchups
commit to merge

--- 2013-03-16 doc/sosp13/tasks.tex
@@
 
-The basic unit of execution in Grappa is a {\em task}. Each task is
-represented by a function pointer and its arguments. A new task is enqueued at
-spawn time; when resources are free, it is allocated a stack, bound to a core,
-and executed.
+The basic unit of execution in \Grappa is a {\em task}. Each task is
+represented by a function pointer and its arguments. When a new task is
+spawned, it is enqueued until it is ready for execution. Each hardware core
+has a single operating system thread pinned to it. A {\em worker} is a
+user-level execution thread (stack, \TODO{what else?}) to which tasks are
+mapped onto when they are ready to execute.
 
-During execution, a task yields control of the processor core whenever it
+Once a task is mapped to worker it stays with that worker until it completes. 
+During execution, a worker yields control of the processor core whenever it
 performs a long-latency operation, allowing the processor to remain busy while
-waiting for the operation to complete. In addition, a programmer can direct
+waiting for the operation to complete.
+In addition, a programmer can direct
 scheduling explicitly via the Grappa API calls shown in
@@
 even when switching amongst thousands of tasks.
+
+\Grappa keeps a large number of workers per hardware core, which is at the
+core of \Grappa's latency tolerance properties.
+
 \TODO{We should expand on the cleverness of our scheduler?}
+
+% talk about lifetime of a task. activating. blocking. work stealing. look at
+% Mark's email. Holt's chat.
 
@@
 \paragraph{Work stealing} 
-When the scheduler finds no work to assign to its workers, it
-commences to steal work from other cores using an asynchronous
-\texttt{call\_on} active message.  It chooses a victim at random until it
-finds one with a non-zero amount of work in its public task queue.  The scheduler steals half of the tasks it finds at the victim.
-Work stealing is particularly interesting in Grappa since performance depends
-on having many active worker threads on each core. Even if there are many
-active threads, if they are all suspended on long-latency operations,
-then the core is underutilized. We choose to initiate one outstanding
-steal whenever the ready queue is empty there are idle worker threads.
+When the scheduler finds no work to assign to its workers, it commences to
+steal tasks from other cores using an asynchronous \texttt{call\_on} active
+message. It chooses a victim at random until it finds one with a non-zero
+amount of work in its public task queue. The scheduler steals half of the
+tasks it finds at the victim. Work stealing is particularly interesting in
+Grappa since performance depends on having many active worker threads on each
+core. Even if there are many active threads, if they are all suspended on
+long-latency operations, then the core is underutilized. We choose to initiate
+one outstanding steal whenever the ready queue is empty there are idle worker
+threads.
 

### 2013-03-22 doc/sosp13/method.tex

--- 2012-07-23 doc/asplos13/method.tex
@@
 \section{Methodology} \label{sec:method}
+
+\TODO{Need to rewrite once our evaluations story is solid. We need to talk about MPI and why that comparison makes sense.}
 
@@
 are added to a work queue to be visited. While this benchmark captures
-irregular, dynamic \emph{computation}, we actually want to evaluate
-performance of algorithms with irregular \emph{memory} access patterns.
+irregular, dynamic \emph{computation,} we actually want to evaluate
+performance of algorithms with irregular \emph{memory\/} access patterns.
 Thus we augment UTS by using the existing traversal code to create a
@@
 this benchmark UTS-Mem, and the timed portion is this traversal of the
-in-memory tree. This in-memory traversal has no knowledgeable of the
+in-memory tree. This in-memory traversal has no knowledge of the
 tree structure beforehand.  The Grappa version of the in-memory tree
@@
 a primitive in many other graph algorithms. Performance is measured in
-\emph{traversed edges per second} (TEPS), where the number of edges is
+\emph{traversed edges per second\/} (TEPS), where the number of edges is
 the edges making up the generated BFS tree. One of the reference

### 2013-03-23 doc/sosp13/evaluation.tex
Major reorganization.

@@
 \section{Evaluation} \label{sec:evaluation}
+Our evaluation begins with presentation of microbenchmark results, establishing
+the intrinsic potential of Grappa to provide random access bandwidth and latency tolerance.
+Next, we present application results, both for Grappa and for other paradigms, as well
+as comparing against the Cray XMT.  Finally, we present the impact of increased aggregation delay
+on Grappa results, thus exploring robustness to network scale.
 
-We evaluate Grappa in four ways. First, we evaluate its performance
-with respect to the XMT. Second, we explore the benefit that
-aggregation gives us. Third, we investigate sensitivity of two key
-parameters in the runtime. Fourth, we investigate scaling to more system nodes.
+\subsection{Microbenchmark Results}
+\subsubsection{Random Access}
+\TODO{Random access feed forward results on Grappa.  optional: Results we measured for XMT.  Cite MPI results}
+\subsubsection{Latency Tolerance}
+\TODO{Simple ping test results -- eg, MPI ping, not the full blown aggregation ping test.  Random access blocking results on Grappa.}
+\subsubsection{Scheduling and Robustness}
+\TODO{Yield test results:  latency \& bandwidth.  Discussion of implications of zillions of contexts on robustness of latency tolerance preshadowing ~\ref{sec:scaling}}
 
-\subsection{Performance}
-
+\subsection{Application Results}
 To evaluate Grappa's performance with respect to the XMT, we ran each
@@
 some cases, the XMT could not run the benchmark with 2 nodes, so the
-point is omitted.
-
+point is omitted. \TODO{rewrite this!}
+\subsubsection{Unbalanced Tree Search}\TODO{rewrite with new results}
 %% UTS: performance comparison
@@
 
-\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex tree
+We ran UTS-mem with a geometric 100M-vertex tree
 (T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
@@
 
+\subsubsection{Breadth First Search}\TODO{rewrite with new results}
 \begin{figure}[tH]
@@
 
-\paragraph{BFS}
-
 We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and
@@
 
+\subsubsection{Approximate Betweenness Centrality}\TODO{rewrite with new results}
 \begin{figure}[tH]
@@
 
-\paragraph{Centrality}
-
 We ran Betweenness Centrality on the same scale 25 Kronecker graph as
@@
 
-\subsection{Network aggregation performance}
+% Got rid of this discussion of UTS by itself, tried to work the highlights in below
+%\paragraph{UTS-Mem}
+%We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
+%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
+%nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
+%Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
+%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
+%The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
+%scaling up through 128 nodes, although scaling
+%declines because of the unscalability of our aggregation mechanism as
+%number of network endpoints increases. 
+%
+%Despite our efforts to tune the UTS implementation specific to the 
+%XMT, performance does not scale well with increasing processor count,
+%flattening out around 60 processors.  When we increase the size of
+%the tree from 100M to 4.2B, we find that performance does not improve,
+%suggesting that performance is not limited by task parallelism.
+%Cray's performance tools show an increasing number of memory
+%retry operations for failed synchronization operations generated by
+%the runtime, which create network contention.
+%
+To determine how Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
+For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
+For Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
+All of the benchmarks continue to improve out to 128 nodes for Grappa.
+UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while Grappa handles this by suspending tasks until messages return.
+For BFS and Centrality, the XMT scales approximately a constant factor better than Grappa.
+We attribute this to a limitation in the current aggregator design and network stack that Grappa uses.
+This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases} performance).
+Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
+Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
+
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/scaling_cropped.pdf}
+    \end{center}
+    \caption{Scaling number of nodes: Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
+    \label{fig:uts_threshold}
+\end{figure}
+
+\subsection{Scaling}\label{sec:scaling} \TODO{rewrite with new results}
+
+\subsubsection{Network Aggregation Performance and Robustness}
 
@@
 
-\subsection{Scaling}
-% Got rid of this discussion of UTS by itself, tried to work the highlights in below
-%\paragraph{UTS-Mem}
-%We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
-%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
-%nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
-%Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
-%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
-%The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
-%scaling up through 128 nodes, although scaling
-%declines because of the unscalability of our aggregation mechanism as
-%number of network endpoints increases. 
-%
-%Despite our efforts to tune the UTS implementation specific to the 
-%XMT, performance does not scale well with increasing processor count,
-%flattening out around 60 processors.  When we increase the size of
-%the tree from 100M to 4.2B, we find that performance does not improve,
-%suggesting that performance is not limited by task parallelism.
-%Cray's performance tools show an increasing number of memory
-%retry operations for failed synchronization operations generated by
-%the runtime, which create network contention.
-%
-To determine how Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
-For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
-For Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
-All of the benchmarks continue to improve out to 128 nodes for Grappa.
-UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while Grappa handles this by suspending tasks until messages return.
-For BFS and Centrality, the XMT scales approximately a constant factor better than Grappa.
-We attribute this to a limitation in the current aggregator design and network stack that Grappa uses.
-This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).
-Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
-Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
-
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/scaling_cropped.pdf}
-    \end{center}
-    \caption{Scaling number of nodes: Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
-    \label{fig:uts_threshold}
-\end{figure}
-
-\subsection{Sensitivity}
-
-\paragraph{Aggregator timeout}
+\subsubsection{Aggregator timeout}
 
@@
 
-\paragraph{Number of active tasks}
+\subsubsection{Number of active tasks} \TODO{possibly eliminate}
 
@@
 
-\paragraph{Parallel loop threshold}
+\subsubsection{Parallel loop threshold}
 
@@
 
-\subsection{Characterization}
+\subsection{Summary}\TODO{this is a placeholder: do we need this section?}
 

### 2013-03-23 doc/sosp13/method.tex
Major reorganization.

--- 2012-07-23 doc/asplos13/method.tex
@@
 \section{Methodology} \label{sec:method}
+Towards evaluating how well Grappa enables scalable
+performance of irregular applications on mass market systems, our
+methodology is as follows.  For consistency in measurement, all
+results reported for mass market systems are measured on the same
+platform.  To establish the potential for Grappa performance, we
+identify small microkernel benchmarks that provide insight into the
+limitations of our system's component performance behaviors.  To
+compare meaningfully with the state-of-the-art, we compare Grappa
+performance on well known irregular application benchmarks both
+against the popular high performance programming methodologies MPI and
+UPC for distributed memory clusters, and against a custom shared memory
+architecture used exclusively for irregular computation, the Cray XMT.
+We then examine how sensitive Grappa results are to reduced
+injection rate and increased latency, as often occur when a system is
+scaled up so that nodes are further apart and the effective per
+destination packet injection rate drops.
+The three sections that follow describe this methodology in greater detail.
 
-To explore the performance of the Grappa runtime we have implemented
-three algorithms: breadth first search, betweenness centrality, and
-unbalanced tree search.  These algorithms were implemented for the
-Cray XMT (our baseline) and for Grappa.  The metric we use is
-algorithmic time, which means startup and loading of the data
-structure (from disk) is not included in the measurement.  Data is
-collected on real systems, which means minor variations in runtime
-exist from run to run.  The average of multiple runs are used, and
-where appropriate, confidence in the result is reported.
+\subsection{Systems}
+For commodity system measurements, we run Grappa on a 144-node cluster
+of AMD Interlagos processors. Nodes have 32-core (every pair share a
+floating-point unit) 2.1-GHz processors, 64GB of memory, and 40Gb
+Mellanox ConnectX-2 infiniband network card.  The cluster uses a
+QLogic infiniband switch.  We configure the nodes to have 32 1-GB
+hugepages to minimize TLB misses for the random access patterns we
+expect from irregular applications.
+
+We compare to the MTA using a 128-node Cray XMT (3rd generation MTA).
+Each node consists of a 500-MHz MTA Threadstorm multithreaded processor
+that supports 128~streams. The machine uses Cray's proprietary SeaStar2
+interconnection network.
 
@@
 (which contains up to 32~cores, although we only use 6 in our
-experiments), has a single infiniband interface.  Hence, for our results
+experiments), has a single infiniband interface.
+\TODO{Given we're using number of network interfaces as parity measure,
+say something about the relative bandwidth of these network interfaces.}
+Hence, for our results
 we scale up XMT processors one for one with full system nodes.
 
-\subsection{Systems}
+\subsection{Microbenchmarks}
+With microbenchmarks we aim to evaluate the intrinisc potential of
+Grappa, measured by random access bandwidth, latency tolerance,
+scheduling efficiency, and robustness.  \TODO{do we need to justify
+these are the key determinants of Grappa potential?} Random access
+bandwidth can be measured with GUPS \TODO{describe HPC Random Access}.
+Latency tolerance must be at least sufficient to tolerate the ping
+latency and is limited above by the rate at which we can switch
+context as a function of the number of contexts: were adding contexts
+to slow the context switch rate below the throughput sustainable by
+the network and memory, the processing required to tolerate latency
+would introduce more latency than is tolerated, thereby degrading
+rather than improving performance.  We wish to tolerate latency of
+aggregation and synchronization as well, increasing the number of
+contexts needed.  We measure latency tolerance via a modified GUPS in
+which each update is dependent upon the results of an earlier update,
+not unlike linked-list chasing \TODO{find references?}.  We then
+establish the efficiency of scheduling via direct tests of yield
+performance sans global memory references.  This provides us with
+reassuring insights into how far we could stretch latency tolerance
+with our current system, one aspect of robustness.  Results for these
+microbenchmarks are presented in Section~\ref{sec:evaluation}.
+After presenting application results, we show how this robustness
+in latency tolerance reduces sensitivity of Grappa performance to changes
+in parameters, particularly in the network.  We measure
+this by artificially varying the effective network injection rate by
+sweeping the timeout value in the aggregator.  Results are in Section~\ref{sec:scaling}.
 
-For measurements, we run Grappa on a 144-node cluster of AMD Interlagos
-processors. Nodes have 32-core (every pair share a floating-point unit)
-2.1-GHz processors, 64GB of memory, and 40Gb Mellanox ConnectX-2
-infiniband network card.   The cluster uses a QLogic infiniband switch.
-We configure the nodes to have 32 1-GB hugepages to minimize TLB misses
-for the random access patterns we expect from irregular applications.
+\subsection{Application Kernel Benchmarks}
 
-We compare to the MTA using a 128-node Cray XMT (3rd generation MTA).
-Each node consists of a 500-MHz MTA Threadstorm multithreaded processor
-that supports 128~streams. The machine uses Cray's proprietary SeaStar2
-interconnection network.
-
-\subsection{Applications}
+To explore the performance of the Grappa runtime we have implemented
+three algorithms: breadth first search, betweenness centrality, and
+unbalanced tree search.  These algorithms were implemented for the
+Cray XMT (our baseline) and for Grappa.  The metric we use is
+algorithmic time, which means startup and loading of the data
+structure (from disk) is not included in the measurement.  Data is
+collected on real systems, which means minor variations in runtime
+exist from run to run.  The average of multiple runs are used, and
+where appropriate, confidence in the result is reported.
 

### 2013-03-23 doc/sosp13/tasks.tex

@@
 SSE context/status register. Thus, the minimum amount of state a
-cooperative context switch routine must save according to the ABI is 62
-bytes.
+cooperative context switch routine must save, according to the ABI, is
+62~bytes.
 

### 2013-03-23 doc/sosp13/tasks.tex
new Tasks section

--- 2013-03-22 doc/sosp13/tasks.tex
@@
-\section{Tasks}
+\section{Tasking System}
 
-The basic unit of execution in \Grappa is a {\em task}. Each task is
-represented by a function pointer and its arguments. When a new task is
-spawned, it is enqueued until it is ready for execution. Each hardware core
-has a single operating system thread pinned to it. A {\em worker} is a
-user-level execution thread (stack, \TODO{what else?}) to which tasks are
-mapped onto when they are ready to execute.
+The basic unit of execution in \Grappa is a {\em task}. Each hardware core has
+a single operating system thread pinned to it. When tasks are ready to
+execute, they are mapped to a {\em worker}, which is a user-level thread.
+\Grappa keeps a large number of workers per hardware core, which is at the
+core of \Grappa's latency tolerance properties. Below we discuss the
+implementation of our task management support and then describe how
+applications expose parallelism to the \Grappa runtime.
 
-Once a task is mapped to worker it stays with that worker until it completes. 
-During execution, a worker yields control of the processor core whenever it
-performs a long-latency operation, allowing the processor to remain busy while
-waiting for the operation to complete.
-In addition, a programmer can direct
-scheduling explicitly via the Grappa API calls shown in
-Figure~\ref{fig:scheduling}. To minimize yield overhead, the Grappa scheduler
-operates entirely in user-space and does little more than store register state
-of one task and load that of another. Context switch times are as low as 38ns
-even when switching amongst thousands of tasks.
+\subsection{Task Support Implementation}
 
-\Grappa keeps a large number of workers per hardware core, which is at the
-core of \Grappa's latency tolerance properties.
+\paragraph{Tasks}
+Each task is represented by a function pointer and its arguments, represented
+as 32-byte entities: a 64-bit function pointer plus three 64-bit arguments. We
+use three arguments because tasks are often generated via a parallel loop
+decomposition, in which each task needs three kinds of data. Thus, tasks
+include:
 
-\TODO{We should expand on the cleverness of our scheduler?}
+\noindent {\it function pointer}, or the addresses of the routine to run ---
+we configure the system nodes to disable address randomization, so that
+function pointers are valid across process images;
 
-% talk about lifetime of a task. activating. blocking. work stealing. look at
-% Mark's email. Holt's chat.
+\noindent {\it private argument}, often a loop index;
 
-\comment{
-Delete the spin and move the 2us and ms thing to results if we have data to show it. -Mark
+\noindent {\it shared argument}, typically data shared amongst a group of
+tasks, or the number of loop iterations to be performed by this task;
 
-Fast context switching amongst many tasks is
-crucial to Grappa for two reasons. First, only when the net cost of context
-switching is dominated by the loss incurred by stalling does a latency
-tolerant approach work to increase throughput on computations with intense
-random access. Random access latencies to node local memory are on the order
-of 100ns. Were the scheduler to add more overhead than this, it would actually
-slow down rather than speed up computations that randomly access node local
-data. Second, because we can switch amongst so many tasks without dramatically
-increasing scheduling overhead, Grappa can tolerate system latencies many
-orders of magnitude greater than that of local memory. Network ping latency
-alone is approximately 2us, but our data suggests we can tolerate latencies in the millisecond
-scale. Furthermore, by over-provisioning task parallelism, Grappa promises to
-provide robust throughput that is relatively insensitive to variations in
-latencies within and across our target systems.
-}
+\noindent {\it synchronization argument}, often used to determine when all
+tasks that are part of a loop have finished, in the form of q global pointer
+to a synchronization object allocated at the core that spawned a group of
+tasks.
+
+While these are the most common uses of the three task arguments, they are
+treated as arbitrary 64-bit values in the runtime, and can be used for any
+purpose.
+
+\paragraph{Workers}
+
+When a new task is spawned, it is enqueued until it is ready for execution.
+Tasks are not allocated any execution resources (a worker) until the scheduler
+decides to run them. A worker is simply a collection of status bits and a
+stack, allocated at a particular core. There are two types of workers: (1) user worker; and (2) system worker.
+
+User workers execute application tasks. Once a task is mapped to a worker it
+stays with that worker until it completes. System workers execute runtime
+management activities such as communication and statistics collection. User
+workers and system workers might have different scheduling priorities. System
+workers execute periodically and in idle times, to guarantee timely
+communication services.
+
+\paragraph{Scheduling}
+
+During execution, a worker yields control of its core whenever it performs a
+long-latency operation, allowing the processor to remain busy while waiting
+for the operation to complete. In addition, a programmer can direct scheduling
+explicitly via the Grappa API calls shown in Figure~\ref{fig:scheduling}. To
+minimize yield overhead, the Grappa scheduler operates entirely in user-space
+and does little more than store register state of one task and load that of
+another.
 
@@
     \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:scheduling} Grappa API: scheduling } %{-4ex}}
+      \caption{\label{fig:scheduling} \Grappa scheduling API } 
     \end{minipage}
-    %\vspace{-3ex}
   \end{center}
@@
 
-\subsection{Expressing parallelism}
+Each core in a \Grappa system has its own independent scheduler. Each
+scheduler has three main responsibilities: servicing system workers that
+process communication requests as described in
+Section~\ref{sec:communication}; rescheduling workers that have been waiting
+on long-latency operations; and assigning ready tasks to worker resources that
+have become idle.
 
-Grappa programmers focus on expressing as much
-parallelism as possible without concern for where it will execute.
-Grappa then chooses where and when to exploit this parallelism,
-scheduling as much work as is necessary on each core to keep it busy
-in the presence of system latencies and task dependences.
+Each scheduler has three queues:
 
-Grappa provides four methods for expressing parallelism, shown in
-Figure~\ref{fig:expressing-parallelism}:
+\noindent {\it ready worker queue}, a FIFO queue of tasks that are
+  matched with workers and are ready to execute;
 
-First, when the programmer identifies work
-that can be done in parallel, the work may be wrapped up in a function
-and queued with its arguments for later execution using a
-\texttt{spawn}.
+\noindent {\it private task queue}, a FIFO queue of tasks that must run on this core;
 
-Second, a programmer can use \texttt{spawn\_on} to spawn a task on a
-specific core in the system or at the home core of a particular memory
-location.
+\noindent {\it public task queue},  a LIFO queue of tasks that are
+  waiting to be matched with workers. It is a local partition of a shared
+  task pool.
 
-Third, the programmer can invoke a parallel for loop, provided that
-the trip count is known at loop entry. The programmer
-specifies a function pointer along with start and end indices and an
-optional threshold to control parallel overhead. Grappa does {\em
-recursive decomposition} of iterations, similar to Cilk's cilk\_for
-construct~\cite {cilkforimplementation}.\comment{could only find slides
-page6 of
-http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-
-Cilk++.pdf    ---- Good enough.  Cite it.  -Mark} and TBB's {\tt parallel\_for}~\cite{intel_tbb}.  It generates a
-logarithmically-deep tree of tasks, stopping to execute the loop body
-when the number of iterations is below the required threshold.
+\TODO{don;t we have an actual queue for higher priority system workers?}
 
-Fourth, a programmer may want to execute an active message; that is,
-to run a small piece of code on a particular core in the system
-without waiting for execution resources to be available.  For example,
-custom synchronization primitives execute this way, as a function
-executed on the core where the data lives.  Grappa provides the
-\texttt{call\_on} call for this purpose.
+Whenever a task yields or suspends, the scheduler makes a decision about what
+to do next. Servicing communication requests (system worker) is given priority
+to ensure responsiveness, but to minimize overhead should context switches be
+frequent, servicing is performed only if sufficient time has elapsed. Second,
+the scheduler determines if any workers with running tasks are ready to
+execute; if so, one is scheduled. Finally, if there are no workers ready to
+run, but there are tasks waiting to be matched with workers, an idle worker is
+woken (or a new worker is spawned), matched with a task, and scheduled.
 
-\begin{figure}[htbp]
-  \begin{center}
-    \begin{description}\small
-    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
-      Creates a new stealable task
-    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
-      Creates a new private task that will run on a specific core 
-    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
-      Executes iterations of a loop as stealable tasks 
-    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
-      Runs a limited function on a specific core without consuming
-      Grappa execution resources 
-    \end{description}
-    \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism
-      \TODO{update this with latest C++ code from ``api\_examples.cpp''}
-      } % \vspace{-4ex}}
-    \end{minipage}
-    %\vspace{-3ex}
-  \end{center}
-\end{figure}
-
-\subsection{Implementation tasks}
-
-\paragraph{Tasks and workers} Grappa tasks are 32-byte entities: a
-64-bit function pointer plus three 64-bit arguments. We use three
-arguments because tasks are often generated via a parallel loop
-decomposition, in which each task needs three kinds of data.
-\begin{description}
-\item[Function pointer] addresses the routine to run. We configure the system nodes to disable address randomization, so that function pointers are valid across process images.
-\item[Private argument] often a loop index.
-\item[Shared argument] typically data shared amongst a group of tasks, or the number of loop iterations to be performed by this task
-\item[Synchronization argument] often used to determine
-  when all tasks that are part of a loop have finished; a global pointer to a synchronization object allocated at the core that spawned a group of tasks.
-\end{description}
-While these are the most common uses of the three task arguments, they
-are treated as arbitrary 64-bit values in the runtime, and can be used
-for any purpose.
-
-Tasks are not allocated any execution resources until the scheduler
-decides to run them; when this occurs, tasks are matched with {\em
-  worker} threads. Each worker is simply a collection of status bits and a
-stack, allocated at a particular core.
-
-\paragraph{Context switching} Grappa context switches between tasks
+\paragraph{Context switching} Grappa context switches between workers
 non-preemptively. As with other cooperative multithreading systems, we
@@
 SSE context/status register. Thus, the minimum amount of state a
-cooperative context switch routine must save according to the ABI is 62
-bytes.
+cooperative context switch routine must save, according to the ABI, is
+62~bytes --- importantly, a context is saved in a single 64 cache line.
 
@@
 
-\paragraph{Scheduling} Each core in the Grappa system has its own
-independent scheduler. Each scheduler has three main tasks to perform:
-servicing communication requests as described in
-Section~\ref{sec:communication}; rescheduling tasks that have been
-waiting on long-latency operations; and assigning ready tasks to
-worker resources that have become idle.
+Since \Grappa keeps a very large number of active workers, their context data will not fit in cache.
+Therefore, the scheduler has to perform very careful prefetching in order to reduce context switch cost to a minimum.
+We determined empirically that prefetching the fourth worker in the scheduling order is sufficient.
+This makes context switching effectively free of cache misses.
 
-Each scheduler has three queues:
-\begin{description}
-\item{\bf Ready worker queue} This is a FIFO queue of tasks that are
-  matched with workers and are ready to execute.
-\item{\bf Private task queue} This is a FIFO queue of tasks that must run on this core.
-\item{\bf Public task queue} This is a LIFO queue of tasks that are
-  waiting to be matched with workers. It is a local partition of a shared
-  task pool.
-\end{description}
+% When the scheduler chooses a thread to schedule it first looks to see if it must run a periodic task.
+If so it choose that.
+If not it looks for a ready task.
+If none exists it chooses the idle thread.
+% 
+% If it chooses a task from the ready queue it prefetches the 4'th next thread on the read queue.  Presently 192 bytes of stack are prefetched and 64 bytes of the thread struct itself.
+% 
+% A periodic task is responsible for looking at the size of the ready queue.  If it is too small more not started tasks are started.
+%
 
-Whenever a task yields or suspends, the scheduler makes a decision
-about what to do next. Servicing communication requests is given priority
-to ensure responsiveness, but to minimize overhead should context switches be frequent, servicing is performed only if sufficient time has elapsed.
-Second, the scheduler determines if any workers with running tasks are
-ready to execute; if so, one is scheduled. Finally, if there are no
-workers ready to run, but there are tasks waiting to be matched with
-workers, an idle worker is woken (or a new worker is spawned), matched
-with a task, and scheduled.
+% The context switcher is written entirely in user space.
+A context is saved in a single cache line (64 bytes).
+We prefetch 192 bytes from the stack (which is where the context is saved) in order to prime the stride prefetched on the core and to preload the task state that exists on the stack.
+
+% Fast context switching amongst many tasks is crucial to Grappa for two
+% reasons. First, only when the net cost of context switching is dominated by
+% the loss incurred by stalling does a latency tolerant approach work to
+% increase throughput on computations with intense random access. Random access
+% latencies to node local memory are on the order of 100ns. Were the scheduler
+% to add more overhead than this, it would actually slow down rather than speed
+% up computations that randomly access node local data. Second, because we can
+% switch amongst so many tasks without dramatically increasing scheduling
+% overhead, Grappa can tolerate system latencies many orders of magnitude
+% greater than that of local memory. Network ping latency alone is approximately
+% 2us, but our data suggests we can tolerate latencies in the millisecond scale.
+% Furthermore, by over-provisioning task parallelism, Grappa promises to provide
+% robust throughput that is relatively insensitive to variations in latencies
+% within and across our target systems.
 
@@
 }
+
+\subsection{Expressing Parallelism}
+
+Grappa programmers focus on expressing as much
+parallelism as possible without concern for where it will execute.
+Grappa then chooses where and when to exploit this parallelism,
+scheduling as much work as is necessary on each core to keep it busy
+in the presence of system latencies and task dependences.
+
+Grappa provides four methods for expressing parallelism, shown in
+Figure~\ref{fig:expressing-parallelism}:
+
+First, when the programmer identifies work
+that can be done in parallel, the work may be wrapped up in a function
+and queued with its arguments for later execution using a
+\texttt{spawn}.
+
+Second, a programmer can use \texttt{spawn\_on} to spawn a task on a
+specific core in the system or at the home core of a particular memory
+location.
+
+Third, the programmer can invoke a parallel for loop, provided that the trip
+count is known at loop entry. The programmer specifies a function pointer
+along with start and end indices and an optional threshold to control parallel
+overhead. Grappa does {\em recursive decomposition} of iterations, similar to
+Cilk's cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
+parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
+tasks, stopping to execute the loop body when the number of iterations is
+below the required threshold.
+
+Fourth, a programmer may want to execute an active message; that is,
+to run a small piece of code on a particular core in the system
+without waiting for execution resources to be available.  For example,
+custom synchronization primitives execute this way, as a function
+executed on the core where the data lives.  Grappa provides the
+\texttt{call\_on} call for this purpose.
+
+\begin{figure}[htbp]
+  \begin{center}
+    \begin{description}\small
+    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
+      Creates a new stealable task
+    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
+      Creates a new private task that will run on a specific core 
+    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
+      Executes iterations of a loop as stealable tasks 
+    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
+      Runs a limited function on a specific core without consuming
+      Grappa execution resources 
+    \end{description}
+    \begin{minipage}{0.95\columnwidth}
+      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism
+      \TODO{update this with latest C++ code from ``api\_examples.cpp''}
+      } % \vspace{-4ex}}
+    \end{minipage}
+    %\vspace{-3ex}
+  \end{center}
+\end{figure}

### 2013-03-23 doc/sosp13/abstract.tex
minor edits

@@
 level languages can be developed or adapted. Grappa tolerates delays to remote
-memory by multiplexing thousands of lightweight tasks to each processor core;
-balances load via fine-grained distributed work-stealing; takes full advantage
-of network characteristics by aggregating smaller data requests into large
-ones; and provides efficient synchronization and remote operations. We present
-a detailed description of the Grappa system, programming examples using the
-library interface, and compare performance for several irregular benchmarks to
-hand-optimized MPI code and the Cray XMT, a custom system used to target the
-real time graph analytics market.
+memory by multiplexing thousands of lightweight execution contexts to each
+processor core; balances load via fine-grained distributed work-stealing;
+takes full advantage of network characteristics by aggregating smaller data
+requests into large ones; and provides efficient synchronization and remote
+operations. We present a detailed description of the Grappa system,
+programming examples using the library interface, and compare performance for
+several irregular benchmarks to hand-optimized MPI code and the Cray XMT, a
+custom system used to target the real time graph analytics market.

### 2013-03-23 doc/sosp13/intro.tex
pass on intro

--- 2013-02-28 doc/sosp13/intro.tex
@@
 
-Irregular applications generate tasks with work, interdependences, or memory
-accesses that are highly sensitive to input. Classic examples of irregular
-applications include branch and bound optimization, SPICE circuit simulation,
-contact algorithms in car crash analysis, and network flow. Important
-contemporary examples include processing large graphs in the business,
-national security, machine learning, data-driven science, and social network
-computing domains. For these emerging applications, fast response -- given the
-sheer amount of data -- requires multinode systems. The most broadly available
-multinode systems are those built from x86 commodity computing nodes
-interconnected via ethernet or infiniband. Our goal is to enable scalable
-performance of irregular applications on these mass market systems.
+Irregular applications exhibit workloads, dependences, and memory accesses
+that are highly sensitive to input. Classic examples of such applications
+include branch and bound optimization, SPICE circuit simulation, contact
+algorithms in car crash analysis, and network flow. Important contemporary
+examples include processing large graphs in the business, national security,
+machine learning, and social network computing domains. For these emerging
+applications, fast response -- given the sheer amount of data -- requires
+multinode systems. The most broadly available multinode systems are those
+built from x86 compute nodes interconnected via ethernet or InfiniBand.
+Scalable performance of irregular applications on these systems is challenging
+for two reasons:
 
-Our goal is challenging for two key reasons:
+\noindent{\bf Irregular applications exhibit little data locality.} Data
+accesses tend to be spread across the entire memory of the system. This makes
+current memory hierarchy features ineffective. Caches are of little assistance
+with such low data re-use and spatial locality. Commodity prefetching hardware
+is effective only when addresses are known many cycles before the data is
+consumed or the accesses follow a predictable pattern, neither of which occurs
+in irregular applications. As a consequence, commodity microprocessors stall
+often when executing irregular applications and data-parallel distributed
+frameworks (e.g., MapReduce~\cite{mapreduce:osdi04}) are ineffective.
 
-\noindent{\bf Irregular applications exhibit little spatial locality.}
-It is not atypical for any given task's data references to be spread
-randomly across the entire memory of the system. This makes current
-memory hierarchy features ineffective. Caches are of little assistance
-with such low data re-use and spatial locality. Commodity prefetching
-hardware is effective only when addresses are known many cycles before
-the data is consumed or the accesses follow a predictable pattern,
-neither of which occurs in irregular applications. As a consequence,
-commodity microprocessors stall often when executing irregular
-applications.
-
-\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low
+\noindent{\bf Irregular applications frequently request small amounts of
+off-node data.} On multi-node systems, the challenges presented by low
 locality are analogous, and exacerbated by the increased latency of going
@@
 rate into the network is insufficient to utilize wire bandwidth when blocks
-are below about two-kilobytes, so any naive communication strategy
-severely under-utilizes the network.
+are below about two-kilobytes, so any naive communication strategy severely
+under-utilizes the network.
 
 While some irregular applications can be restructured to better exploit
-locality, aggregating requests to increase message size, and manage the
+locality, aggregate requests to increase message size, and manage the
 additional challenges of load balance and synchronization across multinode
-systems, the work to do so is formidable and requires knowledge and skills
-pertaining to distributed systems far beyond those of most application
-programmers.
+systems, the effort required to do so is formidable and involves knowledge and
+skills pertaining to distributed systems far beyond those of most application
+programmers. Luckily, many of the important irregular applications (e.g.,
+graph processing, our focus in this paper) naturally offer large amounts of
+concurrency. This immediately suggests taking advantage of concurrency to
+tolerate the latency of data movement.
 
-Luckily, many of the important irregular applications (e.g., graph
-processing, our focus in this paper) naturally offer large amounts of
-concurrency. This immediately suggests taking advantage of concurrency
-to tolerate the latency of data movement. The fully custom Tera
-MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular
-applications by using concurrency to hide latencies. It had a large
-distributed shared memory with no caches.  On every clock cycle, each
-processor would execute a ready instruction chosen from one of its 128
-hardware thread contexts, a sufficient number to fully hide memory
-access latency.  The network was designed with a single word injection
-rate that matched the processor clock frequency and sufficient bandwidth
-to sustain a reference from every processor on every clock cycle.
-Unfortunately, while an excellent match to extremely irregular
-applications, the MTA was not cost-effective on applications that could
-exploit locality and had very poor single-thread performance, making
-it a commercial failure. The Cray XMT approximates the Tera MTA-2,
-substantially reducing its cost but not overcoming its narrow range of
+The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of
+supporting irregular applications by using concurrency to hide latencies. It
+had a large distributed shared memory with no caches. On every clock cycle,
+each processor would execute a ready instruction chosen from one of its 128
+hardware thread contexts, a sufficient number to fully hide memory access
+latency. The network was designed with a single-word injection rate that
+matched the processor clock frequency and sufficient bandwidth to sustain a
+reference from every processor on every clock cycle. Unfortunately, the MTA
+was not general enough nor cost-effective. The Cray XMT approximates the Tera
+MTA-2, substantially reducing its cost but not overcoming its narrow range of
 applicability.
 
-% Two opportunities help overcome these challenges without putting more burden on the programmer: (1) Many interesting irregular applications (e.g., graph processing, our focus on this paper) naturally observe large amounts of concurrency; (2)
+We believe we can support irregular applications with good performance and
+cost-effectiveness with commodity hardware for two main reasons. First,
+commodity multicore processors have become extremely fast and cheap. So with
+careful software engineering we can multiplex hundred of thousands of
+execution contexts into a single core. Second, commodity networks offer high
+bandwidth as long as messages are large enough. So with enough data requests
+in flight simultaneously, the system can aggregate small messages into large
+enough ones.
 
-% In addition to large amounts of concurrency in irregular applications, 
-% 
-% Commodity systems offer a large number of cores with good single-thread performance.
-Concurrency can be used to hide network latencies involved in fetching data from remote notes, mitigating the lack of locality.
-% 
-% 
-% ; (3) Commodity networks improved significantly and offer high bandwidth at a relatively low cost
+In this paper we introduce \Grappa, a software runtime system that allows a
+commodity cluster of x86-based nodes connected via an InfiniBand network to be
+programmed as if it were a single, large, shared-memory NUMA (non-uniform
+memory access) machine with scalable performance for irregular applications.
+\Grappa is designed to smooth over some of the performance discontinuities in
+commodity hardware, giving good performance when there is little locality to
+be exploited while allowing the programmer to exploit it when it is available.
 
-In the twenty years that have elapsed since the Tera MTA, commodity
-microprocessors have become much faster and multicore has driven down the
-absolute price of computation; commodity network price-performance has
-improved as well. This shift has afforded us the opportunity to attack the
-challenges posed by irregular applications by emulating in software and
-inexpensive mass market hardware, the approach taken by Tera. We exploit the
-increased aggregate instruction rate per socket relative to chip bandwidth,
-using what would otherwise be wasted instructions to manage the multiplexing
-of as many as several thousand tasks per core, thus tolerating memory latency,
-reducing stalls, and making better use of available bandwidth. Ultimately, the
-opportunity is to cover the spectrum of irregular to regular computation:
-where tasks exhibit locality, multiplex fewer tasks and expend fewer
-instructions on context switching; where locality is lacking, multiplex more
-tasks at a higher rate to tolerate latency. Thus we make the best use of task
-parallelism -- either to scale to more cores or to tolerate latency -- and of
-caches -- either to exploit application locality or to house more task
-contexts.
-
-In this paper we introduce Grappa, a software runtime system that allows
-a commodity x86 distributed-memory HPC cluster to be programmed as if it
-were a single large shared-memory machine and provides scalable
-performance for irregular applications. Grappa is designed to smooth
-over some of the performance discontinuities in commodity hardware,
-giving good performance when there is little locality to be exploited
-while allowing the programmer to exploit it when it is available.
-
-Grappa leverages as much freely available and commodity infrastructure as
+\Grappa leverages as much freely available and commodity infrastructure as
 possible. We use unmodified Linux for the operating system and an
-off-the-shelf user-mode infiniband device driver stack~\cite{OFED}. MPI is
+off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}. MPI is
 used for process setup and tear down. GASnet~\cite{gasnet} is used as the
 underlying mechanism for remote memory reads and writes using active message
-invocations. To this commodity hardware and software mix Grappa adds three
-main software components: (1) a \emph{lightweight tasking\/} layer that supports
-a context switch in as few as 38ns and distributed global load balancing; (2)
-a \emph{distributed shared memory\/} layer that supports normal access operations
-such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer
-that combines short messages to mitigate the aforementioned problem that
-commodity networks are designed to achieve peak bandwidth only on large packet
-sizes, yet irregular applications tend to fetch only a handful of bytes at a
-time. As we will show later, Grappa can tolerate latencies way beyond that of
-the network. Therefore, Grappa can afford to \emph{trade latency for
-throughput\/}: by {\em increasing\/} latency in key components of the system we
-are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations
-to remote nodes), and our ability to improve load imbalance (work stealing
-increases latency).
+invocations. To this commodity hardware and software mix \Grappa adds three
+main software components: (1) a \emph{lightweight tasking\/} layer that
+supports a context switch in as few as 38ns and distributed global load
+balancing; (2) a \emph{distributed shared memory\/} layer that supports normal
+access operations such as \emph{read\/} and \emph{write\/} as well as
+synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd};
+and (3) a \emph{communication\/} layer that combines short messages to
+achieve peak bandwidth on commodity networks. As we will show later, \Grappa
+can tolerate latencies way beyond that of the network. Therefore, \Grappa can
+afford to \emph{trade latency for throughput}. By {\em increasing\/} latency
+in key components of the system we are able to: increase effective random
+access memory bandwidth by delaying and aggregating messages; increase
+synchronization rate by delegating atomic operations to gatekeeper cores, even
+when referencing node-local global data; and improve load balance by
+tolerating the delays incurred when work-stealing.
 
-% 
-% Each of these components will be described in more detail in Section~\ref{sec:grappa}.
-% 
-% {\em Provide evidence we have succeeded -- will rewrite as necessary based on final results:}
-
-Our evaluation of Grappa shows that it runs several graph-crunching
-applications very efficiently
-on a commodity cluster. Our yardstick for comparison is the XMT hardware
-itself. Using the same number of network interfaces, Grappa is in the
-same ballpark as the XMT: For unbalanced tree search, Grappa is over 3X
-faster and shows greatly improved scalability; conversely, for breadth
-first search and betweenness centrality Grappa is 2.5X slower.  In
-Section~\ref{sec:evaluation} we explore the factors that underpin this
-performance. Most importantly, however, for significantly less real
-world cost, users can \emph{add\/} significantly more processors to a
-commodity cluster than an XMT machine and use Grappa to achieve scalable
-performance.
-
-% 
-% \TODO{Add roadmap here. Luis: do we really want a roadmap here? We have enough about what we have done I think, so we might not need to sell things we plan to do.}
+Our evaluation of Grappa shows that it runs several irregular application
+kernels (e.g., graph traversal and sort) very efficiently on a commodity
+cluster. Our yardstick for comparison is the XMT hardware itself. Using the
+same number of network interfaces, \Grappa is in the same ballpark as the XMT:
+For unbalanced tree search, Grappa is over \TODO{3X} faster and shows greatly
+improved scalability; conversely, for breadth first search and betweenness
+centrality Grappa is 2.5X slower. In Section~\ref{sec:evaluation} we explore
+the factors that underpin this performance. Most importantly, however, for
+significantly less real world cost, users can \emph{add\/} significantly more
+processors to a commodity cluster than an XMT machine and use Grappa to
+achieve scalable performance. \TODO{revise paragraph after eval, including MPI
+and possibly programming model}

### 2013-03-24 doc/sosp13/relatedwork.tex
minor edits to related work

@@
 \paragraph{Distributed graph processing} While \Grappa is a general runtime
-system for any large-scale concurrent application, one of its \checkme{killer
-applications is graph analysis}. Therefore, distributed graph processing
+system for any large-scale concurrent application, one of its killer
+applications is graph analysis. Therefore, distributed graph processing
 frameworks like Pregel~\cite{pregel:2010} and Distributed
@@
 bulk-synchronous parallel (BSP) execution model, which makes it inefficient on
-workloads that could prioritize vertices. GraphLab overcomes this limitation with an execution mode
-that schedules vertex computations individually, allowing prioritization, which
-gives faster convergence in a variety of iterative algorithms. GraphLab,
-however, imposes a rigid computation model where programmers must express
-computation as transformations on a vertex and its edge list only, with
-information only from adjacent vertexes. Pregel is only slightly less
-restrictive, as the input data can be any vertex in the graph. Grappa also
-supports dynamic parallelism with asynchronous execution, but parallelism is
-expressed as tasks or loop iterations, which is a far more general programming
-model for irregular computation tasks. PowerGraph~\cite{powergraph:osdi12} improves
-the performance of GraphLab for real-world graphs with power-law degree distributions by
-using a vertex cut for graph partitioning. If the size of the partial result of 
-applying a subset of edge data to the vertex is constant, then there is a good bound on the communication
-required to keep vertex replicas coherent. We do not explore these algorithmic techniques
-in this work, but they are complimentary and would improve the performance of graph applications on Grappa.
+workloads that could prioritize vertices. GraphLab overcomes this limitation
+with an execution mode that schedules vertex computations individually,
+allowing prioritization, which gives faster convergence in a variety of
+iterative algorithms. GraphLab, however, imposes a rigid computation model
+where programmers must express computation as transformations on a vertex and
+its edge list only, with information only from adjacent vertexes. Pregel is
+only slightly less restrictive, as the input data can be any vertex in the
+graph. \Grappa also supports dynamic parallelism with asynchronous execution,
+but parallelism is expressed as tasks or loop iterations, which is a far more
+general programming model for irregular computation tasks.
+PowerGraph~\cite{powergraph:osdi12} improves the performance of GraphLab for
+real-world graphs with power-law degree distributions by using a vertex cut
+for graph partitioning. If the size of the partial result of applying a subset
+of edge data to the vertex is constant, then there is a good bound on the
+communication required to keep vertex replicas coherent. We do not explore
+these algorithmic techniques in this work, but they are complimentary and
+would improve the performance of graph applications on \Grappa.
 
-\paragraph{Global memory} Grappa supports a global memory namespace, presenting a programming abstraction similar to that of previous software distributed shared memory (DSM) systems but with differences attuned to irregular application demands.
-
+\paragraph{Software distributed shared memory}
+% Grappa supports a global memory namespace, presenting a programming
+% abstraction similar to that of previous software DSM systems but with
+% differences attuned to irregular application demands.
 Many traditional software DSM systems are page based~\cite{Treadmarks,munin}
@@
 exploiting the processor's paging mechanisms, therefore relying heavily on
-locality. Instead, Grappa, like partitioned global address space (PGAS)
+locality. Instead, \Grappa, like partitioned global address space (PGAS)
 models, implements its DSM at the language, rather than the system level.
@@
 UPC~\cite{upc:2005} make accesses to shared structures look like normal
-references. As we describe later, Grappa chooses a middle ground, where global
+references. As we describe later, \Grappa chooses a middle ground, where global
 addresses are explicit in the API and local accesses are emitted
@@
 
-Past DSM work was mostly page-based, so it could take advantage of a network's RDMA support to move large page-sized
-blocks of data from node to node \TODO{citation}. However, when commodity networks move small
-blocks of data (a few bytes), only a fraction of the available bandwidth is
-achieved. \TODO{probably want to reframe this discussion in view of Grappa using RDMA differently}
+Past DSM work was mostly page-based, so it could directly take advantage of a
+network's RDMA support to move large page-sized blocks of data from node to
+node \TODO{citation}. However, when commodity networks move small blocks of
+data (a few bytes), only a fraction of the available bandwidth is achieved. \Grappa had to include significant aggregation support to make effective use of RDMA.
 
-In summary, while Grappa's DSM system is conceptually similar to prior work,
-we accept the random access aspect of irregular applications and optimize for throughput rather than low latency.
-Our DSM must support enough memory concurrency to tolerate the latency of the network.
+In summary, while \Grappa's DSM system is conceptually similar to prior work,
+we accept the random access aspect of irregular applications and optimize for
+throughput rather than low latency. Our DSM system must support enough memory
+concurrency to tolerate the latency of the network.
 
@@
 
-\paragraph{Multithreading} Grappa uses multithreading to tolerate memory
+\paragraph{Multithreading} \Grappa uses multithreading to tolerate memory
 latency. This is a well known technique. Hardware implementations include the
@@
 Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in
-this paper, Grappa implements its own software-based multithreading with a
+this paper, \Grappa implements its own software-based multithreading with a
 lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
 on a single processing core. The large number of tasks is required to tolerate
-the very high inter-node communication latency of commodity networks. Grappa's
+the very high inter-node communication latency of commodity networks. \Grappa's
 task library employs several optimizations: an extremely fast task switch, a
@@
 into the cache sufficiently long before that task is actually scheduled. The main difference
-between Grappa's support for lightweight threads and prior work such as
+between \Grappa's support for lightweight threads and prior work such as
 QThreads~\cite{Wheeler08qthreads:an} and

### 2013-03-24 doc/sosp13/abstract.tex
minor edits all over

--- 2013-03-14 doc/sosp13/abstract.tex
@@
-Grappa is a runtime system for commodity clusters of multicore computers that
+\Grappa is a runtime system for commodity clusters of multicore computers that
 presents a massively parallel, single address space abstraction to
-applications. Grappa's goal is to provide scalable performance for irregular
+applications. \Grappa's goal is to provide scalable performance for irregular
 parallel applications, such as graph processing. These applications are
@@
 
-Grappa serves both as a C++ user library and as a foundation upon which higher
-level languages can be developed or adapted. Grappa tolerates delays to remote
-memory by multiplexing thousands of lightweight tasks to each processor core;
-balances load via fine-grained distributed work-stealing; takes full advantage
-of network characteristics by aggregating smaller data requests into large
-ones; and provides efficient synchronization and remote operations. We present
-a detailed description of the Grappa system, programming examples using the
-library interface, and compare performance for several irregular benchmarks to
-hand-optimized MPI code and the Cray XMT, a custom system used to target the
-real time graph analytics market.
+\Grappa serves both as a C++ user library and as a foundation upon which
+higher level languages can be developed or adapted. \Grappa tolerates delays
+to remote memory by multiplexing thousands of lightweight execution contexts
+to each processor core; balances load via fine-grained distributed
+work-stealing; takes full advantage of network characteristics by aggregating
+smaller data requests into large ones; and provides efficient synchronization
+and remote operations. We present a detailed description of the \Grappa
+system, programming examples using the library interface, and compare
+performance for several irregular benchmarks to hand-optimized MPI code and
+the Cray XMT, a custom system used to target the real time graph analytics
+market.

### 2013-03-24 doc/sosp13/overview.tex
little tweaks
minor edits all over

@@
-\section{Grappa Overview}
+\section{\Grappa Overview}
 
@@
 \begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:grappa} Grappa system overview}
+  \caption{\label{fig:grappa} \Grappa system overview}
 \end{minipage}
@@
 
-Grappa (Figure~\ref{fig:grappa}) has three main software components:
+\Grappa (Figure~\ref{fig:grappa}) has three main software components:
 \begin{description}
@@
 tasks can be stolen from any node in the system), which provides automated
-load balancing. The scheduler oversubscribes to have more worker
-threads than required for latency tolerance. By having at least
-several threads ready to run at all times, the scheduler can prefetch
-a few stack frames into cache to lower the likelihood of costly main
-memory accesses on every context switch.
+load balancing. The scheduler oversubscribes to have more worker threads than
+required for latency tolerance. By having at least several threads ready to
+run at all times, the scheduler can prefetch a context data into cache to
+lower the likelihood of costly main memory accesses during a context switch.
 
@@
 % \paragraph{Exploiting Latency Tolerance} As we will show later, Grappa can tolerate latencies way beyond that of the network.
-Therefore, Grappa can afford to \emph{trade latency for throughput}: by {\em increasing} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
+Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
 
@@
 
-In the next three sections we describe both Grappa's main capabilities and how they are implemented.
+In the next three sections we describe both \Grappa's main components and how they are implemented.

### 2013-03-24 doc/sosp13/tasks.tex
minor edits all over

@@
 for the operation to complete. In addition, a programmer can direct scheduling
-explicitly via the Grappa API calls shown in Figure~\ref{fig:scheduling}. To
-minimize yield overhead, the Grappa scheduler operates entirely in user-space
+explicitly via the \Grappa API calls shown in Figure~\ref{fig:scheduling}. To
+minimize yield overhead, the \Grappa scheduler operates entirely in user-space
 and does little more than store register state of one task and load that of
@@
   \begin{center}
-    \begin{description}\small
-    \item[ \texttt{ yield() } ] \hfill \\
-      Yields core to scheduler, enqueuing caller to be scheduled again soon
-    \item[ \texttt{ suspend() } ] \hfill \\
-      Yields core to scheduler, enqueuing caller only once another task calls wake
-    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
-      Enqueues some other task $t$ to be scheduled again soon
-    \end{description}
+	\begin{tabular}{l}
+    \texttt{\scriptsize yield() } \\
+      Yields core to scheduler, enqueuing caller to be \\ scheduled again soon \\
+    \texttt{\scriptsize suspend() }  \\
+      Yields core to scheduler, enqueuing caller only once \\ another task calls wake \\
+    \texttt{\scriptsize wake( task * $t$ ) } \\
+      Enqueues some other task $t$ to be scheduled again soon \\
+	\end{tabular}
     \begin{minipage}{0.95\columnwidth}
@@
 
-\paragraph{Context switching} Grappa context switches between workers
+\paragraph{Context switching} \Grappa context switches between workers
 non-preemptively. As with other cooperative multithreading systems, we
@@
 tasks it finds at the victim. Work stealing is particularly interesting in
-Grappa since performance depends on having many active worker threads on each
+\Grappa since performance depends on having many active worker threads on each
 core. Even if there are many active threads, if they are all suspended on
@@
 tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
-Termination detection is not built into the Grappa task scheduler,
+Termination detection is not built into the \Grappa task scheduler,
 rather, it is considered a programming error for the program to exit
@@
 
-Grappa programmers focus on expressing as much
-parallelism as possible without concern for where it will execute.
-Grappa then chooses where and when to exploit this parallelism,
-scheduling as much work as is necessary on each core to keep it busy
-in the presence of system latencies and task dependences.
+\Grappa programmers focus on expressing as much parallelism as possible
+without concern for where it will execute. \Grappa then chooses where and when
+to exploit this parallelism, scheduling as much work as is necessary on each
+core to keep it busy in the presence of system latencies and task dependences.
 
-Grappa provides four methods for expressing parallelism, shown in
-Figure~\ref{fig:expressing-parallelism}:
-
-First, when the programmer identifies work
-that can be done in parallel, the work may be wrapped up in a function
-and queued with its arguments for later execution using a
-\texttt{spawn}.
-
-Second, a programmer can use \texttt{spawn\_on} to spawn a task on a
-specific core in the system or at the home core of a particular memory
-location.
-
-Third, the programmer can invoke a parallel for loop, provided that the trip
-count is known at loop entry. The programmer specifies a function pointer
-along with start and end indices and an optional threshold to control parallel
-overhead. Grappa does {\em recursive decomposition} of iterations, similar to
-Cilk's cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
+\Grappa provides four methods for expressing parallelism, shown in
+Figure~\ref{fig:expressing-parallelism}. First, when the programmer identifies
+work that can be done in parallel, the work may be wrapped up in a function
+and queued with its arguments for later execution using a \texttt{spawn}.
+Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific
+core in the system or at the home core of a particular memory location. Third,
+the programmer can invoke a parallel for loop, provided that the trip count is
+known at loop entry. The programmer specifies a function pointer along with
+start and end indices and an optional threshold to control parallel overhead.
+\Grappa does {\em recursive decomposition} of iterations, similar to Cilk's
+cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
 parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
 tasks, stopping to execute the loop body when the number of iterations is
-below the required threshold.
-
-Fourth, a programmer may want to execute an active message; that is,
-to run a small piece of code on a particular core in the system
-without waiting for execution resources to be available.  For example,
-custom synchronization primitives execute this way, as a function
-executed on the core where the data lives.  Grappa provides the
-\texttt{call\_on} call for this purpose.
+below the required threshold. Fourth, a programmer may want to execute an active message; that is, to run a
+small piece of code on a particular core in the system without waiting for
+execution resources to be available. For example, custom synchronization
+primitives execute this way, as a function executed on the core where the data
+lives. \Grappa provides the \texttt{call\_on} call for this purpose.
 
@@
       Runs a limited function on a specific core without consuming
-      Grappa execution resources 
+      \Grappa execution resources 
     \end{description}
     \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism
+      \caption{\label{fig:expressing-parallelism} \Grappa API: expressing parallelism
       \TODO{update this with latest C++ code from ``api\_examples.cpp''}

### 2013-03-24 doc/sosp13/tasks.tex
fixed TODO in tasks

@@
 
-Each scheduler has three queues:
+Each scheduler has four queues:
 
@@
 
-\TODO{don;t we have an actual queue for higher priority system workers?}
+\noindent {\it deadline task queue}, a priority queue of tasks that are executed according to task-specific deadline constraints.
 
-Whenever a task yields or suspends, the scheduler makes a decision about what
-to do next. Servicing communication requests (system worker) is given priority
-to ensure responsiveness, but to minimize overhead should context switches be
-frequent, servicing is performed only if sufficient time has elapsed. Second,
-the scheduler determines if any workers with running tasks are ready to
-execute; if so, one is scheduled. Finally, if there are no workers ready to
-run, but there are tasks waiting to be matched with workers, an idle worker is
-woken (or a new worker is spawned), matched with a task, and scheduled.
+Whenever a task yields or suspends, the scheduler makes a decision about what to do next.
+First, any task in the deadline task queue who's deadline constraint is met is chosen for execution.
+This queue manages high priority system tasks, such as periodically servicing communication requests.
+Second, the scheduler determines if any workers with running tasks are ready to execute; if so, one is scheduled.
+Finally, if there are no workers ready to run, but there are tasks waiting to be matched with workers, an idle worker is woken (or a new worker is spawned), matched with a task, and scheduled.
 

### 2013-03-24 doc/sosp13/communication.tex
minor edits
edits/comments on new aggregator

@@
-\section{Communication}\label{sec:communication}
+\section{Communication Support}
+\label{sec:communication}
 
@@
 Grappa’s lower networking layer aggregates the upper layer’s messages to
-improve performance. Commodity networks including infiniband achieves their
-peak bisection bandwidth only when the packet sizes are relatively large ---
-on the order of multiple kilobytes. The reason for this discrepancy is the
-combination of overheads associated with handling each packet (in terms of
-bytes that form the actual packet, processing time at the card and processing
-on the CPU within the driver stack).
+improve performance. As mentioned earlier, commodity networks including
+InfiniBand achieves their peak bisection bandwidth only when the packet sizes
+are relatively large --- on the order of multiple kilobytes. The reason for
+this discrepancy is the combination of overheads associated with handling each
+packet (in terms of bytes that form the actual packet, processing time at the
+card and processing on the CPU within the driver stack).
 
@@
 \TODO{We should expand on the zero-copy aggregation stuff?}
+
+% there is a message pointer list in each core. this list has as many entries
+% as the number of cores in the system. each entry is another list of message
+% with that destination.
+
+% each core in a node is responsible for a set of nodes in the system
+
+% cores in a node share a region of memory
+
+% network buffers also reside in this shared region of memory
+
+% messages ptrs also contain prefetching information to speed up assembling message in the buffer.
+
+% there are as many sending workers in a core as the number of nodes it is responsible for.

### 2013-03-24 doc/sosp13/communication.tex
communication layer pass

@@
 
-In order to mitigate the low message injection rate limits of commodity
-networks, Grappa's communication stack has two layers: one for user-level
-messages and one for network-level messages.
+\Grappa's communication support has two layers: user-level messaging interface
+based on active messages; and network-level transport that supports request
+aggregation for better communication bandwidth.
 
-At the upper (user-level) layer, Grappa implements asynchronous active
-messages~\cite{vonEicken92}. Each message consists of a function pointer, an
-optional argument payload, and an optional data payload. When a task sends a
-message, the message is copied to a send queue associated with the message's
-destination and the task continues execution.
+% In order to mitigate the low message injection rate limits of commodity
+% networks, Grappa's communication stack has two layers:
 
-Grappa’s lower networking layer aggregates the upper layer’s messages to
-improve performance. As mentioned earlier, commodity networks including
-InfiniBand achieves their peak bisection bandwidth only when the packet sizes
-are relatively large --- on the order of multiple kilobytes. The reason for
-this discrepancy is the combination of overheads associated with handling each
+\paragraph{Active messages interface} At the upper (user-level) layer, \Grappa
+implements asynchronous active messages~\cite{vonEicken92}. Each message
+consists of a function pointer, an optional argument payload, and an optional
+data payload.
+
+% When a task sends a message, the message is linked to a send queue
+% associated with the message's destination node and the task continues
+% execution. Note that the message contents stay in the worker's stack until
+% it is sent to the network interface buffer, to avoid multiple copies of the
+% payload.
+
+\paragraph{Message aggregation} In our experiments the vast majority of upper
+layer message requests are smaller than 44 bytes. Our measurements confirm
+manufacturers' published data [15]; with 44-byte packets, the available
+bisection bandwidth is only a small fraction (3\%) of the peak bisection
+bandwidth. As mentioned earlier, commodity networks including InfiniBand
+achieves their peak bisection bandwidth only when the packet sizes are
+relatively large --- on the order of multiple kilobytes. The reason for this
+discrepancy is the combination of overheads associated with handling each
 packet (in terms of bytes that form the actual packet, processing time at the
-card and processing on the CPU within the driver stack).
+card and processing on the CPU within the driver stack). Consequently, to make
+the best use of the network, we must convert small messages into large ones.
 
-In our experiments the vast majority of requests were smaller than 44 bytes.
-Our measurements confirm manufacturers' published data [15]; with 44-byte
-packets, the available bisection bandwidth is only a small fraction (3\%) of
-the peak bisection bandwidth. To make the best use of the network, we must
-convert our small messages into large ones. When a task sends a message, it is
-not immediately sent, but rather placed in a queue specific to the
-destination.
+\paragraph{Message processing mechanics} Since communication is very frequent
+in \Grappa, aggregating and sending messages efficiently is very important. To
+achieve that, \Grappa makes careful use of caches, prefetching and lock-free
+synchronization operations. \TODO{Does the explanation below need a figure?}
 
-There are three situations in which a queue of aggregated messages is sent:
-(1) Each queue has a message size threshold of 4096 bytes, chosen to give
-reasonable network performance. If the size in bytes of a queue is above the
-threshold, the contents of the queue are sent immediately. (2) Each queue
-has a wait time threshold ($\approx${1ms}). If the oldest message in a queue
-has been waiting longer than this threshold, the contents of the queue are
-sent immediately, even if the queue size is lower than the message size
-threshold. (3) Queues may be explicitly flushed in situations where the
-programmer wants to minimize the latency of a message at the cost of bandwidth
-utilization.
+Outgoing messages are organized as an array of linked lists, with one linked
+list per destination node in the system. Each core in a given node is
+responsible for aggregating and sending the resulting message to a set of
+destination nodes. The outgoing message lists are located in a region of
+memory shared across all cores in a \Grappa node (thus enabling cores to peak
+at each other's message lists). When a task sends a message, it allocates a buffer (typically on its stack),  determines the destination system
+node, and links the buffer into the corresponding linked list.
 
-The network layer is serviced by polling. Periodically when a context
-switch occurs, the Grappa scheduler switches to the network polling
-thread. This thread has three responsibilities. First, it polls the
-lower-level network layer to ensure it makes progress. Second, it
-deaggregates received messages and executes active message
-handlers. Third, it checks to see if any aggregation queues have
-messages that have been waiting longer than the threshold; if so, it
-sends them.
+Each core has a system worker per destination node it is responsible for; this
+system worker periodically checks whether messages to the corresponding
+destination node is either large enough or has waited past a time-out period,
+at which point the message is sent out. Aggregating and sending a message
+involves manipulating a shared data-structure (the message list). This is done
+using CAS (compare-and-swap) operations to avoid high synchronization costs.
 
-Underneath the aggregation layer, Grappa uses the \gasnet~communication
-library~\cite{gasnet} to actually move data. All interprocess
-communication, whether on or off a cluster node, is handled by the
-\gasnet~library. \gasnet~is able to take advantage of many communication
-mechanisms, including ethernet and infiniband between nodes, as well as
-shared memory within a node.
+% Threads that dispatch a message allocate a message buffer (typically on their
+% stack) and dispatch a send request. \Grappa examines the destination system
+% node for this field, and adds the message to a linked list of messages
+% destined for that node. Periodically, or when this linked list grows large,
+% the list is transferred using a CAS (compare-and-swap} operation to the
+% processor on the originating system node that is responsible for managing all
+% outbound communication requests to the specified destination node. This
+% two-step process is used because it was determined empirically that we could
+% only infrequently use processor coherent access requests because they remain a
+% bottleneck on modern multicore hardware.
 
-Some networks provide access to a remote machine's memory directly. This
-would seem to be a good fit for a programming model focused on global
-shared memory, but in fact we do not use it. In our experiments, we
-found that RDMA operations are subject to the same message rate
-limitations as all other messages on these cards, and thus using raw
-RDMA operations for our small messages would make inefficient use of
-bandwidth. Instead, we implement remote memory operations with active
-messages. A byproduct of this design decision is that Grappa is not
-limited to RDMA-capable networks.
+Each node has a region of memory with send buffers where the final aggregated
+messages are built. These buffers are visible to the network card, and
+messages can be sent with user-mode operations only. When the worker
+responsible for outbound messages to a given system node has received a
+sufficient number of message send requests, or a timeout is reached, the
+linked list of messages is walked and messages are copied to a send buffer.
+This process requires careful prefetching because most of the outbound
+messages are \emph{not} in the processor cache at this time (recall that a
+core can be aggregating messages originating from other cores in the same
+node). Once the send buffer has been formed it is handed off to GASnet for
+transfer to the remote system node. RDMA is used if the underlying network
+supports it.
 
-\TODO{We should expand on the zero-copy aggregation stuff?}
+There are two useful consequences of forming the send buffer at the time of
+message transmission instead of along the way, as individual upper layer
+message send requests are received. First, as previously mentioned most of the
+messages are not in the cache and prefetching is used to run-ahead in the
+linked list of messages in order to avoid cache misses. But once the send
+buffer is formed it is in the cache (for the most part). Hence, when it is
+handed off to GASnet for transfer across the physical wire, the network card
+can pull the message buffer from the processor cache instead of main memory,
+which we have found speeds performance. The second consequence of this
+decision is that we do not need to pre-allocate buffers for all destination
+nodes in the system, as the buffer can be allocated on the fly. Nevertheless
+we have found it efficient to build a flow-control like protocol of
+outstanding message buffers between pairs of system nodes.
+
+Once the remote system node has received the message buffer it is unpacked.
+Upper level \Grappa messages are delivered to each core at the destination
+system node by examining the message buffer at each processor core and then
+handing the message buffer to the next core for delivery of messages destined
+to that core.
+
+%There are three situations in which a queue of aggregated messages is sent:
+%(1) Each queue has a message size threshold of 4096 bytes, chosen to give
+%reasonable network performance. If the size in bytes of a queue is above the
+%threshold, the contents of the queue are sent immediately. (2) Each queue
+%has a wait time threshold ($\approx${1ms}). If the oldest message in a queue
+%has been waiting longer than this threshold, the contents of the queue are
+%sent immediately, even if the queue size is lower than the message size
+%threshold. (3) Queues may be explicitly flushed in situations where the
+%programmer wants to minimize the latency of a message at the cost of bandwidth
+%utilization.
+
+%The network layer is serviced by polling. Periodically when a context
+%switch occurs, the Grappa scheduler switches to the network polling
+%thread. This thread has three responsibilities. First, it polls the
+%lower-level network layer to ensure it makes progress. Second, it
+%deaggregates received messages and executes active message
+%handlers. Third, it checks to see if any aggregation queues have
+%messages that have been waiting longer than the threshold; if so, it
+%sends them.
+
+%Underneath the aggregation layer, Grappa uses the \gasnet~communication
+%library~\cite{gasnet} to actually move data. All interprocess
+%communication, whether on or off a cluster node, is handled by the
+%\gasnet~library. \gasnet~is able to take advantage of many communication
+%mechanisms, including ethernet and infiniband between nodes, as well as
+%shared memory within a node.
+
+%Some networks provide access to a remote machine's memory directly. This
+%would seem to be a good fit for a programming model focused on global
+%shared memory, but in fact we do not use it. In our experiments, we
+%found that RDMA operations are subject to the same message rate
+%limitations as all other messages on these cards, and thus using raw
+%RDMA operations for our small messages would make inefficient use of
+%bandwidth. Instead, we implement remote memory operations with active
+%messages. A byproduct of this design decision is that Grappa is not
+%limited to RDMA-capable networks.
+
+%\TODO{We should expand on the zero-copy aggregation stuff?}
 

### 2013-03-25 doc/sosp13/abstract.tex

@@
 higher level languages can be developed or adapted. \Grappa tolerates delays
-to remote memory by multiplexing thousands of lightweight execution contexts
+to remote memory by multiplexing thousands of lightweight workers
 to each processor core; balances load via fine-grained distributed
@@
 and remote operations. We present a detailed description of the \Grappa
-system, programming examples using the library interface, and compare
+system,
+%programming examples using the library interface,
+and compare
 performance for several irregular benchmarks to hand-optimized MPI code and

### 2013-03-25 doc/sosp13/communication.tex

@@
 
-\Grappa's communication support has two layers: user-level messaging interface
-based on active messages; and network-level transport that supports request
-aggregation for better communication bandwidth.
+In order to mitigate the low message injection rate limits of commodity networks, Grappa's communication stack has two layers: one for user-level messages and one for network-level messages.
 
-% In order to mitigate the low message injection rate limits of commodity
-% networks, Grappa's communication stack has two layers:
+At the upper (user-level) layer, Grappa implements asynchronous active messages~\cite{vonEicken92}.
+Each message consists of a function pointer, an optional argument payload, and an optional data payload.
+When a task sends a message, the message is copied to a send queue associated with the message's destination and the task continues execution.
+In our experiments the vast majority of upper layer message requests are smaller than 44 bytes.
+Our measurements confirm manufacturers' published data [15]; with 44-byte packets, the available bisection bandwidth is only a small fraction (3\%) of the peak bisection bandwidth.
 
-\paragraph{Active messages interface} At the upper (user-level) layer, \Grappa
-implements asynchronous active messages~\cite{vonEicken92}. Each message
-consists of a function pointer, an optional argument payload, and an optional
-data payload.
+To make the best use of the network, we must convert our small messages into large ones, and this is the responsibility of Grappa’s lower networking layer.
+As mentioned earlier, commodity networks including InfiniBand achieves their peak bisection bandwidth only when the packet sizes are relatively large --- on the order of multiple kilobytes.
+The reason for this discrepancy is the combination of overheads associated with handling each packet (in terms of bytes that form the actual packet, processing time at the card and processing on the CPU within the driver stack).
 
-% When a task sends a message, the message is linked to a send queue
-% associated with the message's destination node and the task continues
-% execution. Note that the message contents stay in the worker's stack until
-% it is sent to the network interface buffer, to avoid multiple copies of the
-% payload.
+Grappa's lower network layer makes careful use of processor cache and coherent access behaviors.
+Threads that dispatch a message allocate a message buffer (typically on their stack) and dispatch a send request.
+Grappa examines the destination system node for this field, and adds the message to a linked list of messages destined for that node.
+Periodically, or when this linked list grows large, the list is transfered using a coherent compare-and-swap operation to the processor on the originating system node that is responsible for managing all outbound communication requests to the specified destination node.
+This two-step process is used because it was determined empirically that we could only infrequently use processor coherent access requests because they remain a bottleneck on modern multicore hardware.
 
-\paragraph{Message aggregation} In our experiments the vast majority of upper
-layer message requests are smaller than 44 bytes. Our measurements confirm
-manufacturers' published data [15]; with 44-byte packets, the available
-bisection bandwidth is only a small fraction (3\%) of the peak bisection
-bandwidth. As mentioned earlier, commodity networks including InfiniBand
-achieves their peak bisection bandwidth only when the packet sizes are
-relatively large --- on the order of multiple kilobytes. The reason for this
-discrepancy is the combination of overheads associated with handling each
-packet (in terms of bytes that form the actual packet, processing time at the
-card and processing on the CPU within the driver stack). Consequently, to make
-the best use of the network, we must convert small messages into large ones.
-
-\paragraph{Message processing mechanics} Since communication is very frequent
-in \Grappa, aggregating and sending messages efficiently is very important. To
-achieve that, \Grappa makes careful use of caches, prefetching and lock-free
-synchronization operations. \TODO{Does the explanation below need a figure?}
-
-Outgoing messages are organized as an array of linked lists, with one linked
-list per destination node in the system. Each core in a given node is
-responsible for aggregating and sending the resulting message to a set of
-destination nodes. The outgoing message lists are located in a region of
-memory shared across all cores in a \Grappa node (thus enabling cores to peak
-at each other's message lists). When a task sends a message, it allocates a buffer (typically on its stack),  determines the destination system
-node, and links the buffer into the corresponding linked list.
-
-Each core has a system worker per destination node it is responsible for; this
-system worker periodically checks whether messages to the corresponding
-destination node is either large enough or has waited past a time-out period,
-at which point the message is sent out. Aggregating and sending a message
-involves manipulating a shared data-structure (the message list). This is done
-using CAS (compare-and-swap) operations to avoid high synchronization costs.
-
-% Threads that dispatch a message allocate a message buffer (typically on their
-% stack) and dispatch a send request. \Grappa examines the destination system
-% node for this field, and adds the message to a linked list of messages
-% destined for that node. Periodically, or when this linked list grows large,
-% the list is transferred using a CAS (compare-and-swap} operation to the
-% processor on the originating system node that is responsible for managing all
-% outbound communication requests to the specified destination node. This
-% two-step process is used because it was determined empirically that we could
-% only infrequently use processor coherent access requests because they remain a
-% bottleneck on modern multicore hardware.
-
-Each node has a region of memory with send buffers where the final aggregated
-messages are built. These buffers are visible to the network card, and
-messages can be sent with user-mode operations only. When the worker
-responsible for outbound messages to a given system node has received a
-sufficient number of message send requests, or a timeout is reached, the
-linked list of messages is walked and messages are copied to a send buffer.
-This process requires careful prefetching because most of the outbound
-messages are \emph{not} in the processor cache at this time (recall that a
-core can be aggregating messages originating from other cores in the same
-node). Once the send buffer has been formed it is handed off to GASnet for
-transfer to the remote system node. RDMA is used if the underlying network
-supports it.
-
-There are two useful consequences of forming the send buffer at the time of
-message transmission instead of along the way, as individual upper layer
-message send requests are received. First, as previously mentioned most of the
-messages are not in the cache and prefetching is used to run-ahead in the
-linked list of messages in order to avoid cache misses. But once the send
-buffer is formed it is in the cache (for the most part). Hence, when it is
-handed off to GASnet for transfer across the physical wire, the network card
-can pull the message buffer from the processor cache instead of main memory,
-which we have found speeds performance. The second consequence of this
-decision is that we do not need to pre-allocate buffers for all destination
-nodes in the system, as the buffer can be allocated on the fly. Nevertheless
-we have found it efficient to build a flow-control like protocol of
-outstanding message buffers between pairs of system nodes.
+Once the processor core responsible for outbound messages to a given system node has received a sufficient number of message send requests, or a timeout is reached, the linked list of messages is walked to form a send buffer.
+This involves making a copy of the messages in memory.
+Careful prefetching is used because most of the outbound messages are \emph{not} in the processor cache at this time.
+Once the send buffer has been formed it is handed off to GASnet for transfer to the remote system node.
+RDMA is used if the underlying network supports it.
+There are two useful consequences of forming the message buffer at the time of message dispatch instead of along the way, as individual upper layer message send requests are received.
+First, as previously mentioned most of the messages are not in the cache and prefetching is used to run-ahead in the linked list of messages in order to avoid cache misses.
+But once the message buffer is formed it is in the cache (for the most part).
+Hence, when it is handed off to GASnet for transfer across the physical wire, the network card can pull the message buffer from the processor cache instead of main memory, which we have found speeds performance.
+The second consequence of this decision we do not need to pre-allocate buffers for all destination nodes in the system, as the buffer can be allocated on the fly.
+Nevertheless we have found it efficient to build a flow-control like protocol of outstanding message buffers between pairs of system nodes.
 
 Once the remote system node has received the message buffer it is unpacked.
-Upper level \Grappa messages are delivered to each core at the destination
-system node by examining the message buffer at each processor core and then
-handing the message buffer to the next core for delivery of messages destined
-to that core.
+Upper level Grappa messages are delivered to each processor at the system node by examining the message buffer at each processor core and then handing the message buffer to the next processor for delivery of messages destined to that processor.
 

### 2013-03-25 doc/sosp13/tasks.tex

@@
 cooperative context switch routine must save, according to the ABI, is
-62~bytes --- importantly, a context is saved in a single 64 cache line.
+62~bytes.
 
-Since the compiler sees all calls to the context switch routine, we
-can save even less state. Our context switch routine appears to the
-compiler as inline assembly; we declare all the registers we need
-to save as ``clobbered'' by the inline assembly routine, and the
-compiler will issue its own save and restore code as needed. This allows the
-compiler to avoid saving any registers that are not used, or are used
-for temporary values that are not needed after the context switch.
+%% No attempt is made to fit them within a single cache line unfortunately :-( Mark
+%%% --- importantly, a context is saved in a single 64 byte cache line.
+
+% Pretty sure this is not true anymore.  Even if it is, it's not important -Mark
+%
+%Since the compiler sees all calls to the context switch routine, we
+%can save even less state. Our context switch routine appears to the
+%compiler as inline assembly; we declare all the registers we need
+%to save as ``clobbered'' by the inline assembly routine, and the
+%compiler will issue its own save and restore code as needed. This allows the
+%compiler to avoid saving any registers that are not used, or are used
+%for temporary values that are not needed after the context switch.
 
@@
 We determined empirically that prefetching the fourth worker in the scheduling order is sufficient.
-This makes context switching effectively free of cache misses.
+Also determined empirically, we prefetch three cache lines from the stack for each worker.
+This makes context switching effectively free of cache misses even to hundreds of thousands of threads.
 

### 2013-03-26 doc/sosp13/communication.tex
edits based on Jacob's feedback

--- 2013-03-24 doc/sosp13/communication.tex
@@
 packet (in terms of bytes that form the actual packet, processing time at the
-card and processing on the CPU within the driver stack). Consequently, to make
-the best use of the network, we must convert small messages into large ones.
+card, multiple roundtrips on the PCI express bus and processing on the CPU
+within the driver stack). Consequently, to make the best use of the network,
+we must convert small messages into large ones.
 
@@
 
-Once the remote system node has received the message buffer it is unpacked.
-Upper level \Grappa messages are delivered to each core at the destination
-system node by examining the message buffer at each processor core and then
-handing the message buffer to the next core for delivery of messages destined
-to that core.
+Once the remote system node has received the message buffer, it is unpacked
+and put int the corresponding incoming message lists. This process is done in
+parallel by all system worker responsible for communication. Upper level
+\Grappa messages are delivered to each core at the destination system node by
+examining the message buffer at each processor core and then handing the
+message buffer to the next core for delivery of messages destined to that
+core.
 

### 2013-03-26 doc/sosp13/abstract.tex
changes to abstract & first 3 paragraphs of intro

@@
 presents a massively parallel, single address space abstraction to
-applications. \Grappa's goal is to provide scalable performance for irregular
-parallel applications, such as graph processing. These applications are
-challenging because they exhibit little data locality, imbalanced parallel
-work and complex communication patterns.
+applications. \Grappa's purpose is scalable performance for irregular
+parallel applications, such as graph processing. Poor data locality, imbalanced parallel
+work and complex communication patterns make these applications difficult to scale.
 
@@
 to each processor core; balances load via fine-grained distributed
-work-stealing; takes full advantage of network characteristics by aggregating
+work-stealing; increases communication throughput by aggregating
 smaller data requests into large ones; and provides efficient synchronization
 and remote operations. We present a detailed description of the \Grappa
-system,
+system and
 %programming examples using the library interface,
-and compare
-performance for several irregular benchmarks to hand-optimized MPI code and
+performance comparisons on several irregular benchmarks to hand-optimized MPI code and to
 the Cray XMT, a custom system used to target the real time graph analytics

### 2013-03-26 doc/sosp13/intro.tex
execution contexts to workers
changes to abstract & first 3 paragraphs of intro

@@
 built from x86 compute nodes interconnected via ethernet or InfiniBand.
-Scalable performance of irregular applications on these systems is challenging
+Scalable performance of irregular applications on these systems is elusive
 for two reasons:
 
-\noindent{\bf Irregular applications exhibit little data locality.} Data
-accesses tend to be spread across the entire memory of the system. This makes
-current memory hierarchy features ineffective. Caches are of little assistance
-with such low data re-use and spatial locality. Commodity prefetching hardware
-is effective only when addresses are known many cycles before the data is
-consumed or the accesses follow a predictable pattern, neither of which occurs
-in irregular applications. As a consequence, commodity microprocessors stall
-often when executing irregular applications and data-parallel distributed
-frameworks (e.g., MapReduce~\cite{mapreduce:osdi04}) are ineffective.
+\noindent{\bf Low Injection Rate} Reference patterns of irregular
+applications are unpredictable, with odds increasing as data is
+spread out more broadly across a larger system that a given reference
+will refer to memory located remotely.  This behavior constrasts
+sharply with that of data-parallel applications for which efficient
+frameworks such as MapReduce~\cite{mapreduce:osdi04} may exploit the
+structure of computation and data affinity to limit internode
+communication to comparably infrequent and larger messages.  Mass
+market network technology is designed to transfer these larger blocks
+of data, not the word-sized references emitted by irregular
+application tasks. Injection rate into the network is insufficient to
+utilize wire bandwidth when blocks are below about two-kilobytes, so
+any naive communication strategy severely under-utilizes the network.
 
-\noindent{\bf Irregular applications frequently request small amounts of
-off-node data.} On multi-node systems, the challenges presented by low
-locality are analogous, and exacerbated by the increased latency of going
-off-node. Irregular applications also present a challenge to mass market
-network technology, which is designed to transfer large blocks of data, not
-the word-sized references emitted by irregular application tasks. Injection
-rate into the network is insufficient to utilize wire bandwidth when blocks
-are below about two-kilobytes, so any naive communication strategy severely
-under-utilizes the network.
+\noindent{\bf High Latency} Unlike data parallel applications, whose
+communications are regular and expected, and can therefore easily be
+overlapped with computation at the application level, irregular
+application tasks encounter remote references throughout their execution
+and must resolve them before making further progress.  The latency of
+commodity networks is anywhere from a one to hundreds of microseconds,
+far more than can be tolerated via the simultaneous multithreading
+available in x86 processors.  Consequently, even with RDMA, processor
+prefetching and caching hardware alone is inadequate to ensure pipeline
+utilization in the presence of high latency networks.
 
@@
 careful software engineering we can multiplex hundred of thousands of
-execution contexts into a single core. Second, commodity networks offer high
+workers into a single core. Second, commodity networks offer high
 bandwidth as long as messages are large enough. So with enough data requests

### 2013-03-26 doc/sosp13/abstract.tex
edited mods from last night

@@
 presents a massively parallel, single address space abstraction to
-applications. \Grappa's purpose is scalable performance for irregular
+applications. \Grappa's purpose is to enable scaling of irregular
 parallel applications, such as graph processing. Poor data locality, imbalanced parallel
-work and complex communication patterns make these applications difficult to scale.
+work and complex communication patterns make scaling these applications difficult.
 

### 2013-03-26 doc/sosp13/intro.tex
edited mods from last night

@@
 structure of computation and data affinity to limit internode
-communication to comparably infrequent and larger messages.  Mass
+communication to less infrequent, larger messages.  Mass
 market network technology is designed to transfer these larger blocks
@@
 
-\noindent{\bf High Latency} Unlike data parallel applications, whose
-communications are regular and expected, and can therefore easily be
-overlapped with computation at the application level, irregular
-application tasks encounter remote references throughout their execution
-and must resolve them before making further progress.  The latency of
-commodity networks is anywhere from a one to hundreds of microseconds,
-far more than can be tolerated via the simultaneous multithreading
-available in x86 processors.  Consequently, even with RDMA, processor
-prefetching and caching hardware alone is inadequate to ensure pipeline
-utilization in the presence of high latency networks.
+\noindent{\bf High Latency} To tolerate latency, data parallel applications make explicit
+their regular, predictable communications so as to minimize interference with computation.
+In contrast, irregular application tasks encounter remote references dynamically during
+execution and must resolve them before making further progress.  The latency of
+commodity networks is anywhere from one to hundreds of microseconds,
+far more than can be tolerated via the hyperthrading available in x86 processors.
+Consequently, even with RDMA, processor prefetching and caching hardware alone is inadequate
+to ensure pipeline utilization in the presence of high latency networks.
 

### 2013-03-26 doc/sosp13/tasks.tex
picture of context switch

@@
 
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/context_switch_time.pdf}
+    \end{center}
+    \caption{Average context switch time with 1 and 6 active cores,
+        with and without prefetching.}
+    \label{fig:context-switch-time}
+\end{figure}
+
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/context_switch_bw.pdf}
+    \end{center}
+    \caption{Context switch rate with prefetching. Once the total
+        size of contexts sufficiently exceeds last-level cache, most
+    prefetches go to main memory and the rate becomes limited to the
+    off-chip bandwidth.}
+    \label{fig:context-switch-time}
+\end{figure}
+
 %% No attempt is made to fit them within a single cache line unfortunately :-( Mark
@@
 This makes context switching effectively free of cache misses even to hundreds of thousands of threads.
+
+\TODO{Summary of what to write please expand: Reference above plots. Bandwidth of single socket is
+    270Mcacheline/s. Each context is 4 cachelines: 1 for worker struct
+    and 3 stack cachelines. We must read and write every context, so 8
+    cacheline transfers per context switch. This asymptotically
+    approaches 33.75Mcontexts/s as you increase the number of contexts
+    (fewer and fewer in L3). Note that only 4+ cores reaches full rate
+    because these westmere chips are balanced to not achieve full off-chip bandwidth until 4
+    cores--bdmyers}
 

### 2013-03-26 doc/sosp13/communication.tex
lots of edits. mostly to comm

@@
 packet (in terms of bytes that form the actual packet, processing time at the
-card, multiple roundtrips on the PCI express bus and processing on the CPU
+card, multiple roundtrips on the PCI Express bus and processing on the CPU
 within the driver stack). Consequently, to make the best use of the network,
@@
 list per destination node in the system. Each core in a given node is
-responsible for aggregating and sending the resulting message to a set of
+responsible for aggregating and sending the resulting messages to a set of
 destination nodes. The outgoing message lists are located in a region of
-memory shared across all cores in a \Grappa node (thus enabling cores to peak
+memory shared across all cores in a \Grappa node (thus enabling cores to peek
 at each other's message lists). When a task sends a message, it allocates a buffer (typically on its stack),  determines the destination system
@@
 core can be aggregating messages originating from other cores in the same
-node). Once the send buffer has been formed it is handed off to GASnet for
+node). Once the send buffer has been formed it is handed off to GASNet for
 transfer to the remote system node. RDMA is used if the underlying network
@@
 buffer is formed it is in the cache (for the most part). Hence, when it is
-handed off to GASnet for transfer across the physical wire, the network card
+handed off to GASNet for transfer across the physical wire, the network card
 can pull the message buffer from the processor cache instead of main memory,
@@
 Once the remote system node has received the message buffer, it is unpacked
-and put int the corresponding incoming message lists. This process is done in
+and put in the corresponding incoming message lists. This process is done in
 parallel by all system worker responsible for communication. Upper level
@@
 
-%There are three situations in which a queue of aggregated messages is sent:
-%(1) Each queue has a message size threshold of 4096 bytes, chosen to give
-%reasonable network performance. If the size in bytes of a queue is above the
-%threshold, the contents of the queue are sent immediately. (2) Each queue
-%has a wait time threshold ($\approx${1ms}). If the oldest message in a queue
-%has been waiting longer than this threshold, the contents of the queue are
-%sent immediately, even if the queue size is lower than the message size
-%threshold. (3) Queues may be explicitly flushed in situations where the
-%programmer wants to minimize the latency of a message at the cost of bandwidth
-%utilization.
+%%%% Comments from Jacob.
 
-%The network layer is serviced by polling. Periodically when a context
-%switch occurs, the Grappa scheduler switches to the network polling
-%thread. This thread has three responsibilities. First, it polls the
-%lower-level network layer to ensure it makes progress. Second, it
-%deaggregates received messages and executes active message
-%handlers. Third, it checks to see if any aggregation queues have
-%messages that have been waiting longer than the threshold; if so, it
-%sends them.
+% Last paragraph:
 
-%Underneath the aggregation layer, Grappa uses the \gasnet~communication
-%library~\cite{gasnet} to actually move data. All interprocess
-%communication, whether on or off a cluster node, is handled by the
-%\gasnet~library. \gasnet~is able to take advantage of many communication
-%mechanisms, including ethernet and infiniband between nodes, as well as
-%shared memory within a node.
-
-%Some networks provide access to a remote machine's memory directly. This
-%would seem to be a good fit for a programming model focused on global
-%shared memory, but in fact we do not use it. In our experiments, we
-%found that RDMA operations are subject to the same message rate
-%limitations as all other messages on these cards, and thus using raw
-%RDMA operations for our small messages would make inefficient use of
-%bandwidth. Instead, we implement remote memory operations with active
-%messages. A byproduct of this design decision is that Grappa is not
-%limited to RDMA-capable networks.
-
-%\TODO{We should expand on the zero-copy aggregation stuff?}
-
-% there is a message pointer list in each core. this list has as many entries
-% as the number of cores in the system. each entry is another list of message
-% with that destination.
-
-% each core in a node is responsible for a set of nodes in the system
-
-% cores in a node share a region of memory
-
-% network buffers also reside in this shared region of memory
-
-% messages ptrs also contain prefetching information to speed up assembling message in the buffer.
-
-% there are as many sending workers in a core as the number of nodes it is responsible for.
+% When a buffer is received, all cores deaggregate their messages in parallel.
+% This is how it works:
+%
+%	- the sending core does an RDMA put into a buffer owned by the receiving
+%	core. an active message is sent to enqueue this buffer to be processed by a
+%	receive worker.
+%
+%	- the receive worker computes the offsets/lengths of the messages in the
+%	buffer for each core on the node. It sends messages through the CAS lists
+%	to each core with their offset and length.
+%
+%	- The cores all deserialize and execute their received active messages.
+%	This is done in parallel. Since messages to the same core are stored next
+%	to each other in the buffer, this deserialization performs well with the
+%	cache and the hardware prefetcher.
+%
+%  - When the cores are done, they send a message back to the receive worker.
+%
+%	- Once the receive worker knows all the cores on the node are done with the
+%	buffer, it's done. (we then return the buffer pointer to the sending node
+%	so it can send again.)
+% 
+% We should make sure we cite the Threaded Abstract Machine paper too.
+% 
+% If we haven't already, we should cite Myrinet, U-Net, EMP, and I think
+% there's one more. Here are some links:
+% 
+% http://dl.acm.org/citation.cfm?id=623898&CFID=128100904&CFTOKEN=36373202
+% http://dl.acm.org/citation.cfm?id=224061&CFID=128100904&CFTOKEN=36373202
+% http://dl.acm.org/citation.cfm?id=582091&CFID=128100904&CFTOKEN=36373202

### 2013-03-26 doc/sosp13/intro.tex
lots of edits. mostly to comm
a couple more changes. 	modified:   intro.tex

@@
 In contrast, irregular application tasks encounter remote references dynamically during
-execution and must resolve them before making further progress.  The latency of
-commodity networks is anywhere from one to hundreds of microseconds,
+execution and must resolve them before making further progress.  Latency of
+commodity networks runs anywhere from one to hundreds of microseconds,
 far more than can be tolerated via the hyperthrading available in x86 processors.
-Consequently, even with RDMA, processor prefetching and caching hardware alone is inadequate
-to ensure pipeline utilization in the presence of high latency networks.
+Consequently, even with RDMA, processors' prefetching and caching hardware alone is inadequate
+to prevent diminishing pipeline utilization as irregular applications are scaled up using
+commodity networks.
 

### 2013-03-26 doc/sosp13/intro.tex
commit to merge

--- 2013-03-26 doc/sosp13/intro.tex
@@
 spread out more broadly across a larger system that a given reference
-will refer to memory located remotely.  This behavior constrasts
+will refer to memory located remotely.  This behavior contrasts
 sharply with that of data-parallel applications for which efficient
@@
 off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}. MPI is
-used for process setup and tear down. GASnet~\cite{gasnet} is used as the
+used for process setup and tear down. GASNet~\cite{gasnet} is used as the
 underlying mechanism for remote memory reads and writes using active message

### 2013-03-27 doc/sosp13/abstract.tex
edits to abstract and intro

@@
 presents a massively parallel, single address space abstraction to
-applications. \Grappa's purpose is to enable scaling of irregular
-parallel applications, such as graph processing. Poor data locality, imbalanced parallel
-work and complex communication patterns make scaling these applications difficult.
+applications. \Grappa's purpose is to enable scalable performance of irregular
+parallel applications, such as graph processing. Poor data locality,
+imbalanced parallel work and complex communication patterns make scaling these
+applications difficult.
 
@@
 higher level languages can be developed or adapted. \Grappa tolerates delays
-to remote memory by multiplexing thousands of lightweight workers
-to each processor core; balances load via fine-grained distributed
-work-stealing; increases communication throughput by aggregating
-smaller data requests into large ones; and provides efficient synchronization
-and remote operations. We present a detailed description of the \Grappa
-system and
-%programming examples using the library interface,
-performance comparisons on several irregular benchmarks to hand-optimized MPI code and to
+to remote memory by multiplexing thousands of lightweight workers to each
+processor core; balances load via fine-grained distributed work-stealing;
+increases communication throughput by aggregating smaller data requests into
+large ones; and provides efficient synchronization and remote operations. We
+present a detailed description of the \Grappa system and performance
+comparisons on several irregular benchmarks to hand-optimized MPI code and to
 the Cray XMT, a custom system used to target the real time graph analytics

### 2013-03-27 doc/sosp13/intro.tex
edits to abstract and intro

--- 2013-03-26 doc/sosp13/intro.tex
@@
 that are highly sensitive to input. Classic examples of such applications
-include branch and bound optimization, SPICE circuit simulation, contact
-algorithms in car crash analysis, and network flow. Important contemporary
-examples include processing large graphs in the business, national security,
-machine learning, and social network computing domains. For these emerging
-applications, fast response -- given the sheer amount of data -- requires
-multinode systems. The most broadly available multinode systems are those
-built from x86 compute nodes interconnected via ethernet or InfiniBand.
-Scalable performance of irregular applications on these systems is elusive
-for two reasons:
+include branch and bound optimization, SPICE circuit simulation, and car crash
+analysis. Important contemporary examples include processing large graphs in
+the business, national security, and social network computing domains. For
+these emerging applications, reasonable response time -- given the sheer
+amount of data -- requires large multinode systems. The most broadly available
+multinode systems are those built from x86 compute nodes interconnected via
+ethernet or InfiniBand. However, scalable performance of irregular
+applications on these systems is elusive for two reasons:
 
-\noindent{\bf Low Injection Rate} Reference patterns of irregular
-applications are unpredictable, with odds increasing as data is
-spread out more broadly across a larger system that a given reference
-will refer to memory located remotely.  This behavior contrasts
-sharply with that of data-parallel applications for which efficient
-frameworks such as MapReduce~\cite{mapreduce:osdi04} may exploit the
-structure of computation and data affinity to limit internode
-communication to less infrequent, larger messages.  Mass
-market network technology is designed to transfer these larger blocks
-of data, not the word-sized references emitted by irregular
-application tasks. Injection rate into the network is insufficient to
-utilize wire bandwidth when blocks are below about two-kilobytes, so
-any naive communication strategy severely under-utilizes the network.
+\noindent{\bf Poor data locality and frequent communication.} Data reference
+patterns of irregular applications are unpredictable and tend to be spread
+across the entire system. This results in frequent requests for remote data.
+Caches are of little assistance because of low data re-use and spatial
+locality. Prefetching is ineffective because request locations are not known
+early enough. Data-parallel frameworks such as
+MapReduce~\cite{mapreduce:osdi04} are also inappropriate because they rely on
+being able to partition data and regular communication patterns. As a
+consequence, irregular applications require frequent communication of
+typically small pieces of data. In contrast, commodity networks are designed
+for large packets and have just a fraction of their peak bandwidth when moving
+small packets.
 
-\noindent{\bf High Latency} To tolerate latency, data parallel applications make explicit
-their regular, predictable communications so as to minimize interference with computation.
-In contrast, irregular application tasks encounter remote references dynamically during
-execution and must resolve them before making further progress.  The latency of
-commodity networks is anywhere from one to hundreds of microseconds,
-far more than can be tolerated via the hyperthrading available in x86 processors.
-Consequently, even with RDMA, processor prefetching and caching hardware alone is inadequate
-to ensure pipeline utilization in the presence of high latency networks.
+\noindent{\bf High network communication latency.} The performance challenges
+of frequent communication is exacerbated by the network latency relative to
+processor performance. Latency of commodity networks runs anywhere from one to
+hundreds of microseconds, which is much more than superscalar execution with
+sophisticated memory hierarchies can tolerate. Since irregular application
+tasks encounter remote references dynamically during execution and must
+resolve them before making further progress, stalls are frequent and lead
+severely underutilized compute resources.
 
 While some irregular applications can be restructured to better exploit
-locality, aggregate requests to increase message size, and manage the
+locality, aggregate requests to increase network message size, and manage the
 additional challenges of load balance and synchronization across multinode
@@
 concurrency. This immediately suggests taking advantage of concurrency to
-tolerate the latency of data movement.
+tolerate the latency of data movement by overlapping computation with
+communication.
 
@@
 was not general enough nor cost-effective. The Cray XMT approximates the Tera
-MTA-2, substantially reducing its cost but not overcoming its narrow range of
-applicability.
+MTA-2, reducing its cost but not overcoming its narrow range of applicability.
 

### 2013-03-28 doc/sosp13/method.tex
todo on methodology
rewritten method.tex

--- 2013-03-22 doc/sosp13/method.tex
@@
 
-\TODO{Need to rewrite once our evaluations story is solid. We need to talk about MPI and why that comparison makes sense.}
+To study the Grappa system we implemented it in C++ for the Linux operating system.
+We ported a number of benchmarks to Grappa as well as collected and optimized a set of comparison benchmarks for alternative approaches.
+We run Grappa on a cluster of AMD Interlagos processors.
+Nodes have 32-core 2.1-GHz processors, 64GB of memory, and 40Gb Mellanox ConnectX-2 infiniband network cards.
+Nodes are connected via a QLogic infiniband switch.
+We use this cluster to evaluate Grappa and native MPI-only performance.
+In addition to a mass-market cluster of machines we compare to the MTA using a 128-node Cray XMT (3rd generation MTA).
+Each node consists of a 500-MHz MTA Threadstorm multithreaded processor that supports 128~streams.
+The machine uses Cray's proprietary SeaStar2 interconnection network.
+To evaluate performance we use a variety of benchmarks:
 
-To explore the performance of the Grappa runtime we have implemented
-three algorithms: breadth first search, betweenness centrality, and
-unbalanced tree search.  These algorithms were implemented for the
-Cray XMT (our baseline) and for Grappa.  The metric we use is
-algorithmic time, which means startup and loading of the data
-structure (from disk) is not included in the measurement.  Data is
-collected on real systems, which means minor variations in runtime
-exist from run to run.  The average of multiple runs are used, and
-where appropriate, confidence in the result is reported.
+\paragraph{Context-switch:} This simple microbenchmark is used to explore context switch overhead.
+A configurable number of workers are used to increment the values in a large array.
+This benchmark stresses the scheduler inside of Grappa.
 
-One question that must be answered when making a comparison between the
-Cray XMT and a typical HPC cluster is what is a fair comparison -- these
-systems are quite different.  Three options immediately come to mind:
-equal number of processing cores, equal number of network interfaces,
-and equal dollars.  We discounted the last option fairly quickly,
-because it isn't a lasting data-point -- the cost of hardware shifts
-over time, skewing the interpretation of results.  The first option,
-cores, has some merit, but Grappa is designed for applications that have
-no locality in their computation.  This means almost all of their memory
-accesses are remote.  The factor that limits their performance is not
-processing, but communicating.  Hence, we have chosen the middle option,
-network interfaces as the way to normalize across the XMT and our
-cluster.  Each processor in the XMT system has its own network interface
-to access shared memory.  In the HPC cluster we use, each processor
-(which contains up to 32~cores, although we only use 6 in our
-experiments), has a single infiniband interface.  Hence, for our results
-we scale up XMT processors one for one with full system nodes.
+\paragraph{GUPS:} This is a faithful implementation of the giga-updates-per-second benchmark.
+Read-modify-write updates are dispatched at random to a large array.
+This benchmark stresses the networking layer of Grappa separately from the scheduler, because only a single worker is used per system node.
 
-\subsection{Systems}
+\paragraph{Unbalanced tree search in-memory (UTS-Mem):} Unbalanced Tree Search (UTS) is a benchmark for evaluating the programmability and performance of systems for parallel applications that require dynamic load balancing~\cite{UTS}.
+It involves traversing an unbalanced implicit tree: at each vertex, its number of children is sampled from some probability distribution, and this number of new nodes are added to a work queue to be visited.
+While this benchmark captures irregular, dynamic \emph{computation,} we actually want to evaluate performance of algorithms with irregular \emph{memory\/} access patterns.
+Thus we augment UTS by using the existing traversal code to create a large tree in memory, and then we traverse the in-memory tree.
+We call this benchmark UTS-Mem, and the timed portion is this traversal of the in-memory tree.
+This in-memory traversal has no knowledge of the tree structure beforehand.
 
-For measurements, we run Grappa on a 144-node cluster of AMD Interlagos
-processors. Nodes have 32-core (every pair share a floating-point unit)
-2.1-GHz processors, 64GB of memory, and 40Gb Mellanox ConnectX-2
-infiniband network card.   The cluster uses a QLogic infiniband switch.
-We configure the nodes to have 32 1-GB hugepages to minimize TLB misses
-for the random access patterns we expect from irregular applications.
+\paragraph{Breadth-first-search (BFS):} This is the primary kernel for the Graph500 benchmark and is what currently determines the ranking of machines on the Graph500 list~\cite{graph500list}.
+As a whole, the Graph500 benchmark suite is designed to bring the focus of system design on data-intensive workloads, particularly large-scale graph analysis problems, that are important among cybersecurity, informatics, and network understanding workloads.
+The BFS benchmark builds a search tree containing parent nodes for each traversed vertex during the search.
+While this is a relatively simple problem to solve, it exercises the random-access and fine-grained synchronization capabilities of a system as well as being a primitive in many other graph algorithms.
+Performance is measured in \emph{traversed edges per second\/} (TEPS), where the number of edges is the edges making up the generated BFS tree.
+One of the reference implementations of Graph500 BFS is for the XMT; this code fails to scale past 16 XMT processors because it does not expose enough parallelism, so we modified the code to use a recursive loop decomposition similar to Grappa's.
+We compare this modified version against a straightforward Grappa implementation.
+We do not employ algorithmic improvements, though there are many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
 
-We compare to the MTA using a 128-node Cray XMT (3rd generation MTA).
-Each node consists of a 500-MHz MTA Threadstorm multithreaded processor
-that supports 128~streams. The machine uses Cray's proprietary SeaStar2
-interconnection network.
+\paragraph{IntSort:}  Write me Brandon Holt!
 
-\subsection{Applications}
+The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.
+Grappa collects statistics about application behavior (packets sent, context switches, etc) and these are discussed where appropriate.
+In addition we use a profiled version of Grappa to report component runtimes, but this profiled version is only used to collect this data and a fully optimized version is used for the bulk of the presented results.
 
-We have used three benchmarks to explore the performance of Grappa:
+%%%% Bit bucket
 
-\paragraph{Unbalanced tree search in-memory (UTS-Mem)} Unbalanced Tree
-Search (UTS) is a benchmark for evaluating the programmability and
-performance of systems for parallel applications that require dynamic
-load balancing~\cite{UTS}. It involves traversing an
-unbalanced implicit tree: at each vertex, its number of children is
-sampled from some probability distribution, and this number of new nodes
-are added to a work queue to be visited. While this benchmark captures
-irregular, dynamic \emph{computation,} we actually want to evaluate
-performance of algorithms with irregular \emph{memory\/} access patterns.
-Thus we augment UTS by using the existing traversal code to create a
-large tree in memory, and then we traverse the in-memory tree. We call
-this benchmark UTS-Mem, and the timed portion is this traversal of the
-in-memory tree. This in-memory traversal has no knowledge of the
-tree structure beforehand.  The Grappa version of the in-memory tree
-search uses the asynchronous parallel for loop over a visited vertex's
-children list.
+%\paragraph{Betweenness Centrality} An important measure of the importance of particular vertices in a network is betweenness centrality (BC)~\cite{freeman1979centrality}.
+By this measure, the ``importance'' of each vertex is computed by finding the fraction of shortest paths that pass through it, which can optionally be approximated by only computing shortest paths using a subset of the vertices as starting points.
+BC can be useful for understanding which vertices may have the greatest impact, so in social networks this could be the primary person linking two communities.
+Because it requires multiple breadth-first traversals across the entire graph and in reverse, on power-law degree graphs, this algorithm exercises random accesses rate, and load balancing.
+It also requires fine-grained synchronization on updates to vertex centrality values because multiple paths will update the same vertex.
+BC is a kernel in the DARPA High Performance Computing Systems (HPCS) Scalable Synthetic Compact Applications graph analysis (SSCA\#2) benchmark~\cite{ssca2}.
+Performance for BC is also measured in TEPS, where the number of traversed edges is the total number of edges in the graph multiplied by the number of random starting vertices used in the approximate computation.
+We use the XMT implementation of BC implemented as part of the GraphCT~\cite{GraphCT} library and a comparable Grappa implementation, both of which use the parallel BC algorithm developed by Bader et al.~\cite{bader:bc}.
+
+%To explore the performance of the Grappa runtime we have implemented three algorithms: breadth first search, betweenness centrality, and unbalanced tree search.
+These algorithms were implemented for the Cray XMT (our baseline) and for Grappa.
+The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.
+Data is collected on real systems, which means minor variations in runtime exist from run to run.
+The average of multiple runs are used, and where appropriate, confidence in the result is reported.
 
@@
 
-\paragraph{Breadth-first-search (BFS)}
-This is the primary kernel for the Graph500 benchmark and is what
-currently determines the ranking of machines on the Graph500
-list~\cite{graph500list}. As a whole, the Graph500 benchmark suite is
-designed to bring the focus of system design on data-intensive
-workloads, particularly large-scale graph analysis problems, that are
-important among cybersecurity, informatics, and network understanding
-workloads. The BFS benchmark builds a search tree containing parent
-nodes for each traversed vertex during the search.  While this is a
-relatively simple problem to solve, it exercises the random-access and
-fine-grained synchronization capabilities of a system as well as being
-a primitive in many other graph algorithms. Performance is measured in
-\emph{traversed edges per second\/} (TEPS), where the number of edges is
-the edges making up the generated BFS tree. One of the reference
-implementations of Graph500 BFS is for the XMT; this code fails to
-scale past 16 XMT processors because it does not expose enough
-parallelism, so we modified the code to use a recursive loop
-decomposition similar to Grappa's. We compare this modified version
-against a straightforward Grappa implementation. We do not employ
-algorithmic improvements, though there are
-many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
+%% Given that we have MPI performance on the cluster this is not
+%% that important to have here anymore.
+%%
+%%
+%One question that must be answered when making a comparison between the Cray XMT and a typical HPC cluster is what is a fair comparison -- these systems are quite different.
+Three options immediately come to mind: equal number of processing cores, equal number of network interfaces, and equal dollars.
+We discounted the last option fairly quickly, because it isn't a lasting data-point -- the cost of hardware shifts over time, skewing the interpretation of results.
+The first option, cores, has some merit, but Grappa is designed for applications that have no locality in their computation.
+This means almost all of their memory accesses are remote.
+The factor that limits their performance is not processing, but communicating.
+Hence, we have chosen the middle option, network interfaces as the way to normalize across the XMT and our cluster.
+Each processor in the XMT system has its own network interface to access shared memory.
+In the HPC cluster we use, each processor (which contains up to 32~cores, although we only use 6 in our experiments), has a single infiniband interface.
+Hence, for our results we scale up XMT processors one for one with full system nodes.
 
-\paragraph{Betweenness Centrality}
-An important measure of the importance of particular vertices in a
-network is betweenness centrality
-(BC)~\cite{freeman1979centrality}. By this measure, the ``importance''
-of each vertex is computed by finding the fraction of shortest paths
-that pass through it, which can optionally be approximated by only
-computing shortest paths using a subset of the vertices as starting
-points. BC can be useful for understanding which vertices may have the
-greatest impact, so in social networks this could be the primary
-person linking two communities. Because it requires multiple
-breadth-first traversals across the entire graph and in reverse, on
-power-law degree graphs, this algorithm exercises random accesses
-rate, and load balancing. It also requires fine-grained
-synchronization on updates to vertex centrality values because
-multiple paths will update the same vertex. BC is a kernel in the
-DARPA High Performance Computing Systems (HPCS) Scalable Synthetic
-Compact Applications graph analysis (SSCA\#2)
-benchmark~\cite{ssca2}. Performance for BC is also measured in TEPS,
-where the number of traversed edges is the total number of edges in
-the graph multiplied by the number of random starting vertices used in
-the approximate computation. We use the XMT implementation of BC
-implemented as part of the GraphCT~\cite{GraphCT} library and a
-comparable Grappa implementation, both of which use the parallel BC
-algorithm developed by Bader et al.~\cite{bader:bc}.
+%Towards evaluating how well Grappa enables scalable performance of irregular applications on mass market systems, our methodology is as follows.
+For consistency in measurement, all results reported for mass market systems are measured on the same platform.
+To establish the potential for Grappa performance, we identify small microkernel benchmarks that provide insight into the limitations of our system's component performance behaviors.
+To compare meaningfully with the state-of-the-art, we compare Grappa performance on well known irregular application benchmarks both against the popular high performance programming methodologies MPI and UPC for distributed memory clusters, and against a custom shared memory architecture used exclusively for irregular computation, the Cray XMT.
+We then examine how sensitive Grappa results are to reduced injection rate and increased latency, as often occur when a system is scaled up so that nodes are further apart and the effective per destination packet injection rate drops.
+The three sections that follow describe this methodology in greater detail.

### 2013-03-28 doc/sosp13/evaluation.tex
moved graphs

@@
 \subsection{Microbenchmark Results}
+
 \subsubsection{Random Access}
@@
 \subsubsection{Scheduling and Robustness}
+
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/context_switch_time.pdf}
+    \end{center}
+    \caption{Average context switch time with 1 and 6 active cores,
+        with and without prefetching.}
+    \label{fig:context-switch-time}
+\end{figure}
+
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/context_switch_bw.pdf}
+    \end{center}
+    \caption{Context switch rate with prefetching. Once the total
+        size of contexts sufficiently exceeds last-level cache, most
+    prefetches go to main memory and the rate becomes limited to the
+    off-chip bandwidth.}
+    \label{fig:context-switch-time}
+\end{figure}
+
+\TODO{Summary of what to write please expand: Reference above plots. Bandwidth of single socket is
+    270Mcacheline/s. Each context is 4 cachelines: 1 for worker struct
+    and 3 stack cachelines. We must read and write every context, so 8
+    cacheline transfers per context switch. This asymptotically
+    approaches 33.75Mcontexts/s as you increase the number of contexts
+    (fewer and fewer in L3). Note that only 4+ cores reaches full rate
+    because these westmere chips are balanced to not achieve full off-chip bandwidth until 4
+    cores--bdmyers}
+
 \TODO{Yield test results:  latency \& bandwidth.  Discussion of implications of zillions of contexts on robustness of latency tolerance preshadowing ~\ref{sec:scaling}}
@@
 
-\subsubsection{Aggregator timeout}
+\subsection{Scaling}
+% Got rid of this discussion of UTS by itself, tried to work the highlights in below
+%\paragraph{UTS-Mem}
+%We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
+%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
+%nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
+%Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
+%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
+%The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
+%scaling up through 128 nodes, although scaling
+%declines because of the unscalability of our aggregation mechanism as
+%number of network endpoints increases. 
+%
+%Despite our efforts to tune the UTS implementation specific to the 
+%XMT, performance does not scale well with increasing processor count,
+%flattening out around 60 processors.  When we increase the size of
+%the tree from 100M to 4.2B, we find that performance does not improve,
+%suggesting that performance is not limited by task parallelism.
+%Cray's performance tools show an increasing number of memory
+%retry operations for failed synchronization operations generated by
+%the runtime, which create network contention.
+%
+To determine how Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
+For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
+For Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
+All of the benchmarks continue to improve out to 128 nodes for Grappa.
+UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while Grappa handles this by suspending tasks until messages return.
+For BFS and Centrality, the XMT scales approximately a constant factor better than Grappa.
+We attribute this to a limitation in the current aggregator design and network stack that Grappa uses.
+This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).
+Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
+Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
+
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{figs/scaling_cropped.pdf}
+    \end{center}
+    \caption{Scaling number of nodes: Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
+    \label{fig:uts_threshold}
+\end{figure}
+
+\subsection{Sensitivity}
+
+\paragraph{Aggregator timeout}
 

### 2013-03-28 doc/sosp13/evaluation.tex
new eval. commit to merge

@@
 \section{Evaluation} \label{sec:evaluation}
-Our evaluation begins with presentation of microbenchmark results, establishing
-the intrinsic potential of Grappa to provide random access bandwidth and latency tolerance.
-Next, we present application results, both for Grappa and for other paradigms, as well
-as comparing against the Cray XMT.  Finally, we present the impact of increased aggregation delay
-on Grappa results, thus exploring robustness to network scale.
 
-\subsection{Microbenchmark Results}
+The goal of our evaluation is to show that the core pieces of the \Grappa
+runtime system, namely our tasking system and the global memory/communication
+layer work as expected and together are able to efficiently run irregular
+applications. We evaluate \Grappa in three basic steps:
+\begin{itemize}
+\item We present results
+that show that the \Grappa runtime is able to sustain very high concurrency
+rates and the communication layer is able to sustain a very high rate of
+global memory operations. We also show the performance of a graph kernel that
+stresses communication and concurrency together.
 
-\subsubsection{Random Access}
-\TODO{Random access feed forward results on Grappa.  optional: Results we measured for XMT.  Cite MPI results}
-\subsubsection{Latency Tolerance}
-\TODO{Simple ping test results -- eg, MPI ping, not the full blown aggregation ping test.  Random access blocking results on Grappa.}
-\subsubsection{Scheduling and Robustness}
+\item We show how some popular irregular applications running on \Grappa compare to the Cray XMT and hand-tuned MPI code.
+
+\item We finish with a characterization of system behavior, including profiling where execution time goes, how aggregation affects message size and rates, how global memory and work stealing behaves.
+
+\end{itemize}
+
+\subsection{Basic Grappa mechanisms}
+
+\subsection{Fast User-Level Context Switching }
 
@@
 
-\TODO{Summary of what to write please expand: Reference above plots. Bandwidth of single socket is
-    270Mcacheline/s. Each context is 4 cachelines: 1 for worker struct
-    and 3 stack cachelines. We must read and write every context, so 8
-    cacheline transfers per context switch. This asymptotically
-    approaches 33.75Mcontexts/s as you increase the number of contexts
-    (fewer and fewer in L3). Note that only 4+ cores reaches full rate
-    because these westmere chips are balanced to not achieve full off-chip bandwidth until 4
-    cores--bdmyers}
+% More specifically, each context is cache lines: 1 for the worker struct and 3
+% for stack data. So the are so 8 cacheline transfers per context switch (write
+% the old context, read the next in). The bandwidth of single socket is 270M cache lines/s.
+% 
+% This asymptotically approaches
+% 33.75Mcontexts/s as you increase the number of contexts (fewer and fewer in
+% L3). Note that only 4+ cores reaches full rate because these westmere chips
+% are balanced to not achieve full off-chip bandwidth until 4 cores.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% OLD %%%%%%%%%%%%%%%%%%
+
+Our evaluation begins with presentation of microbenchmark results, establishing
+the intrinsic potential of Grappa to provide random access bandwidth and latency tolerance. 
+Next, we present application results, both for Grappa and for other paradigms, as well
+as comparing against the Cray XMT.  Finally, we present the impact of increased aggregation delay
+on Grappa results, thus exploring robustness to network scale.
+
+\subsection{Microbenchmark Results}
+\subsubsection{Random Access}
+\TODO{Random access feed forward results on Grappa.  optional: Results we measured for XMT.  Cite MPI results}
+\subsubsection{Latency Tolerance}
+\TODO{Simple ping test results -- eg, MPI ping, not the full blown aggregation ping test.  Random access blocking results on Grappa.}
+\subsubsection{Scheduling and Robustness}
 \TODO{Yield test results:  latency \& bandwidth.  Discussion of implications of zillions of contexts on robustness of latency tolerance preshadowing ~\ref{sec:scaling}}

### 2013-03-28 doc/sosp13/intro.tex
new eval. commit to merge

@@
 
+\vspace{0.5ex}
 \noindent{\bf Poor data locality and frequent communication.} Data reference
@@
 
+\vspace{0.5ex}
 \noindent{\bf High network communication latency.} The performance challenges
@@
 locality, aggregate requests to increase network message size, and manage the
-additional challenges of load balance and synchronization across multinode
-systems, the effort required to do so is formidable and involves knowledge and
-skills pertaining to distributed systems far beyond those of most application
-programmers. Luckily, many of the important irregular applications (e.g.,
-graph processing, our focus in this paper) naturally offer large amounts of
-concurrency. This immediately suggests taking advantage of concurrency to
-tolerate the latency of data movement by overlapping computation with
-communication.
+additional challenges of load balance and synchronization, the effort required
+to do so is formidable and involves knowledge and skills pertaining to
+distributed systems far beyond those of most application programmers. Luckily,
+many of the important irregular applications (e.g., graph processing, our
+focus in this paper) naturally offer large amounts of concurrency. This
+immediately suggests taking advantage of concurrency to tolerate the latency
+of data movement by overlapping computation with communication.
 

### 2013-03-28 doc/sosp13/relatedwork.tex
new eval. commit to merge

@@
 
-% Grappa is built on many existing ideas in programming languages, systems and
-% architecture. In this section we discuss related frameworks and key enabling
-% technologies that Grappa builds upon.
-
-\paragraph{Distributed graph processing} While \Grappa is a general runtime
+\vspace{0.5ex}
+\noindent{\bf Distributed graph processing} While \Grappa is a general runtime
 system for any large-scale concurrent application, one of its killer
@@
 
-\paragraph{Software distributed shared memory}
-% Grappa supports a global memory namespace, presenting a programming
-% abstraction similar to that of previous software DSM systems but with
-% differences attuned to irregular application demands.
-Many traditional software DSM systems are page based~\cite{Treadmarks,munin}
-and aim to hide the fact that they are built in software from applications by
-exploiting the processor's paging mechanisms, therefore relying heavily on
-locality. Instead, \Grappa, like partitioned global address space (PGAS)
-models, implements its DSM at the language, rather than the system level.
-Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and
-UPC~\cite{upc:2005} make accesses to shared structures look like normal
-references. As we describe later, \Grappa chooses a middle ground, where global
-addresses are explicit in the API and local accesses are emitted
-conventionally by the compiler.
-
-Past DSM work was mostly page-based, so it could directly take advantage of a
-network's RDMA support to move large page-sized blocks of data from node to
-node \TODO{citation}. However, when commodity networks move small blocks of
-data (a few bytes), only a fraction of the available bandwidth is achieved. \Grappa had to include significant aggregation support to make effective use of RDMA.
+\vspace{0.5ex}
+\noindent{\bf Software distributed shared memory.} Many traditional software
+DSM systems are page based~\cite{Treadmarks,munin} and aim to hide the fact
+that they are built in software from applications by exploiting the
+processor's paging mechanisms, therefore relying heavily on locality. Instead,
+\Grappa, like partitioned global address space (PGAS) models, implements its
+DSM at the language, rather than the system level. Languages such as
+Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and UPC~\cite{upc:2005}
+make accesses to shared structures look like normal references. As we describe
+later, \Grappa chooses a middle ground, where global addresses are explicit in
+the API and local accesses are emitted conventionally by the compiler. Past
+DSM work was mostly page-based, so it could directly take advantage of moving
+large page-sized blocks of data from node to node. However, when commodity
+networks move small blocks of data (a few bytes), only a fraction of the
+available bandwidth is achieved. \Grappa had to include significant
+aggregation support to make effective use of of the network.
 
@@
 throughput rather than low latency. Our DSM system must support enough memory
-concurrency to tolerate the latency of the network.
+concurrency to tolerate the latency of the network and additional latency
+overhead impose by the runtime system.
 
-% Importantly, global memory support in Grappa
-% ends up being tightly coupled to the task scheduler in order to overlap long
-% latency memory operations with useful computation. For these two reasons it
-% became necessary to build a new DSM system specifically for Grappa.
-
-%% We could be stronger here on why Grappa's task library is different,
-%% but we are running out of time.  -Mark
-
-\paragraph{Multithreading} \Grappa uses multithreading to tolerate memory
+\vspace{0.5ex}
+\noindent{\bf Multithreading} \Grappa uses multithreading to tolerate memory
 latency. This is a well known technique. Hardware implementations include the
@@
 on a single processing core. The large number of tasks is required to tolerate
-the very high inter-node communication latency of commodity networks. \Grappa's
-task library employs several optimizations: an extremely fast task switch, a
-small task size, and judicious use of software prefetch instructions to bring task state
-into the cache sufficiently long before that task is actually scheduled. The main difference
-between \Grappa's support for lightweight threads and prior work such as
+the very high inter-node communication latency of commodity networks.
+\Grappa's task library employs several optimizations: an extremely fast task
+switch, a small task size, and judicious use of software prefetch instructions
+to bring task state into the cache sufficiently long before that task is
+actually scheduled. The main difference between \Grappa's support for
+lightweight threads and prior work such as
 QThreads~\cite{Wheeler08qthreads:an} and
-Capriccio~\cite{Behren03capriccio:scalable} is the use of prefetching by the scheduler, which is needed for good performance
-when multiplexing such a large number of tasks.
+Capriccio~\cite{Behren03capriccio:scalable} is context prefetching, which is needed for good performance when multiplexing such a large
+number of tasks.

### 2013-03-28 doc/sosp13/tasks.tex
moved graphs
new eval. commit to merge

--- 2013-03-25 doc/sosp13/tasks.tex
@@
 a single operating system thread pinned to it. When tasks are ready to
-execute, they are mapped to a {\em worker}, which is a user-level thread.
-\Grappa keeps a large number of workers per hardware core, which is at the
-core of \Grappa's latency tolerance properties. Below we discuss the
-implementation of our task management support and then describe how
+execute, they are mapped to a {\em worker}, which is akin to a user-level
+thread. \Grappa keeps a large number of workers per hardware core, this is the
+core reason why \Grappa is able to tolerate latencies well. Below we discuss
+the implementation of our task management support and then describe how
 applications expose parallelism to the \Grappa runtime.
@@
 \paragraph{Workers}
-
 When a new task is spawned, it is enqueued until it is ready for execution.
@@
 \paragraph{Scheduling} 
-
 During execution, a worker yields control of its core whenever it performs a
@@
 
-Each core in a \Grappa system has its own independent scheduler. Each
-scheduler has three main responsibilities: servicing system workers that
-process communication requests as described in
-Section~\ref{sec:communication}; rescheduling workers that have been waiting
-on long-latency operations; and assigning ready tasks to worker resources that
-have become idle.
+Each core in a \Grappa system has its own independent scheduler. Schedulers
+have three main responsibilities: servicing system workers that process
+communication requests as described in Section~\ref{sec:communication};
+rescheduling workers that have been waiting on long-latency operations; and
+assigning ready tasks to worker resources that have become idle.
 
@@
 
-Whenever a task yields or suspends, the scheduler makes a decision about what to do next.
-First, any task in the deadline task queue who's deadline constraint is met is chosen for execution.
-This queue manages high priority system tasks, such as periodically servicing communication requests.
-Second, the scheduler determines if any workers with running tasks are ready to execute; if so, one is scheduled.
-Finally, if there are no workers ready to run, but there are tasks waiting to be matched with workers, an idle worker is woken (or a new worker is spawned), matched with a task, and scheduled.
+Whenever a task yields or suspends, the scheduler makes a decision about what
+to do next. First, any task in the deadline task queue who's deadline
+constraint is met is chosen for execution. This queue manages high priority
+system tasks, such as periodically servicing communication requests. Second,
+the scheduler determines if any workers with running tasks are ready to
+execute; if so, one is scheduled. Finally, if there are no workers ready to
+run, but there are tasks waiting to be matched with workers, an idle worker is
+woken (or a new worker is spawned), matched with a task, and scheduled.
 
-\paragraph{Context switching} \Grappa context switches between workers
-non-preemptively. As with other cooperative multithreading systems, we
-treat context switches as function calls, saving and restoring only the
-callee-saved state as specified in the x86-64 ABI \cite{amd64:abi:2012}. This
-involves saving six general-purpose 64-bit registers and the stack
-pointer, as well as the 16-bit x87 floating point control word and the
-SSE context/status register. Thus, the minimum amount of state a
-cooperative context switch routine must save, according to the ABI, is
-62~bytes.
-
-%% No attempt is made to fit them within a single cache line unfortunately :-( Mark
-%%% --- importantly, a context is saved in a single 64 byte cache line.
+\paragraph{Context switching} 
+\Grappa context switches between workers non-preemptively. As with other
+cooperative multithreading systems, we treat context switches as function
+calls, saving and restoring only the callee-saved state as specified in the
+x86-64 ABI\~cite{amd64:abi:2012}. This involves saving six general-purpose
+64-bit registers and the stack pointer, as well as the 16-bit x87 floating
+point control word and the SSE context/status register. Thus, the minimum
+amount of state a cooperative context switch routine must save, according to
+the ABI, is 62~bytes.
 
@@
 
-Since \Grappa keeps a very large number of active workers, their context data will not fit in cache.
-Therefore, the scheduler has to perform very careful prefetching in order to reduce context switch cost to a minimum.
-We determined empirically that prefetching the fourth worker in the scheduling order is sufficient.
-Also determined empirically, we prefetch three cache lines from the stack for each worker.
-This makes context switching effectively free of cache misses even to hundreds of thousands of threads.
-
-% When the scheduler chooses a thread to schedule it first looks to see if it must run a periodic task.
-If so it choose that.
-If not it looks for a ready task.
-If none exists it chooses the idle thread.
-% 
-% If it chooses a task from the ready queue it prefetches the 4'th next thread on the read queue.  Presently 192 bytes of stack are prefetched and 64 bytes of the thread struct itself.
-% 
-% A periodic task is responsible for looking at the size of the ready queue.  If it is too small more not started tasks are started.
-%
-
-% The context switcher is written entirely in user space.
-A context is saved in a single cache line (64 bytes).
-We prefetch 192 bytes from the stack (which is where the context is saved) in order to prime the stride prefetched on the core and to preload the task state that exists on the stack.
-
-% Fast context switching amongst many tasks is crucial to Grappa for two
-% reasons. First, only when the net cost of context switching is dominated by
-% the loss incurred by stalling does a latency tolerant approach work to
-% increase throughput on computations with intense random access. Random access
-% latencies to node local memory are on the order of 100ns. Were the scheduler
-% to add more overhead than this, it would actually slow down rather than speed
-% up computations that randomly access node local data. Second, because we can
-% switch amongst so many tasks without dramatically increasing scheduling
-% overhead, Grappa can tolerate system latencies many orders of magnitude
-% greater than that of local memory. Network ping latency alone is approximately
-% 2us, but our data suggests we can tolerate latencies in the millisecond scale.
-% Furthermore, by over-provisioning task parallelism, Grappa promises to provide
-% robust throughput that is relatively insensitive to variations in latencies
-% within and across our target systems.
+Since \Grappa keeps a very large number of active workers, their context data
+will not fit in cache. Therefore, the scheduler has to perform very careful
+prefetching in order to reduce context switch cost to a minimum. We determined
+empirically that prefetching the fourth worker in the scheduling order is
+sufficient. We also prefetch three cache lines from the stack for each worker.
+This makes context switching effectively free of cache misses even to hundreds
+of thousands of threads. We provide a detailed analysis of our context switch
+performance in Section~\ref{eval:ctx}.
 
@@
 threads.
-
-\comment{Commenting out.  Simon things it doesn't add much I worried about its usefulness too.  -M
-
-\TODO{Anything after this I'm not sure adds to the paper -- if it does, please edit -Simon}
-The
-core would be fully-utilized if there is always something on the ready
-queue. Even if there are many active tasks, if they are all suspended
-for long-latency network requests, then the core is underutilized. So,
-there is a choice: should the core use some of this idle time to perform
-work stealing or should it just wait as it is possible that local tasks
-will create more work. We choose to have a core initiate a steal when
-all of the following conditions hold: no workers are ready to run, the
-unstarted task queues are empty, and there are no outstanding steal
-requests. Having only one outstanding steal request throttles steals to
-prevent flooding the network, but other scalable quieting mechanisms
-would be possible, such as a voting
-tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
-Termination detection is not built into the \Grappa task scheduler,
-rather, it is considered a programming error for the program to exit
-without syncing all tasks.
-}
 
@@
   \begin{center}
-    \begin{description}\small
-    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
-      Creates a new stealable task
-    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
-      Creates a new private task that will run on a specific core 
-    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
-      Executes iterations of a loop as stealable tasks 
-    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
-      Runs a limited function on a specific core without consuming
-      \Grappa execution resources 
-    \end{description}
+	\begin{tabular}{l}
+    \texttt{\scriptsize spawn( void (*fp)(args) )} \\
+      Creates a new stealable task \\
+    \texttt{\scriptsize spawn\_on( core, (*fp)(args) )} \\
+      Creates a new task that will run on a specific core \\
+    \texttt{\scriptsize parallel\_for( (*fp)(args), start, end )} \\
+      Executes iterations of a loop as stealable tasks  \\
+    \texttt{\scriptsize call\_on( core, (*fp)(args) )} \\ 
+      Runs a limited function on a specific core without \\
+      consuming \Grappa execution resources 
+	\end{tabular}
     \begin{minipage}{0.95\columnwidth}

### 2013-03-28 doc/sosp13/method.tex

@@
 
-To study the Grappa system we implemented it in C++ for the Linux operating system.
-We ported a number of benchmarks to Grappa as well as collected and optimized a set of comparison benchmarks for alternative approaches.
-We run Grappa on a cluster of AMD Interlagos processors.
-Nodes have 32-core 2.1-GHz processors, 64GB of memory, and 40Gb Mellanox ConnectX-2 infiniband network cards.
-Nodes are connected via a QLogic infiniband switch.
-We use this cluster to evaluate Grappa and native MPI-only performance.
-In addition to a mass-market cluster of machines we compare to the MTA using a 128-node Cray XMT (3rd generation MTA).
-Each node consists of a 500-MHz MTA Threadstorm multithreaded processor that supports 128~streams.
-The machine uses Cray's proprietary SeaStar2 interconnection network.
-To evaluate performance we use a variety of benchmarks:
+We implemented the \Grappa in C++ for the Linux operating system. The core
+runtime system system is about 15K lines of code. We ported a number of
+benchmarks to \Grappa as well as collected and optimized a set of comparison
+benchmarks for XMT, MPI and UPC~\cite{UPC}. We run the \Grappa, MPI and UPC
+experiments on a cluster of AMD Interlagos processors. Nodes have 32 2.1-GHz
+cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand
+network cards. Nodes are connected via a QLogic InfiniBand switch. We also
+compare \Grappa to a 128-node Cray XMT (3rd generation MTA). Each node
+consists of a 500-MHz MTA Threadstorm multithreaded processor that supports
+128~streams. The machine uses Cray's proprietary SeaStar2 interconnection
+network.
 
-\paragraph{Context-switch:} This simple microbenchmark is used to explore context switch overhead.
+We use a variety of benchmarks:
+
+\noindent{\bf Context-switch:} This simple microbenchmark is used to explore context switch overhead.
 A configurable number of workers are used to increment the values in a large array.
@@
 
-\paragraph{GUPS:} This is a faithful implementation of the giga-updates-per-second benchmark.
-Read-modify-write updates are dispatched at random to a large array.
-This benchmark stresses the networking layer of Grappa separately from the scheduler, because only a single worker is used per system node.
+\nonindent{\bf GUPS:} This is a faithful implementation of the
+giga-updates-per-second benchmark. Read-modify-write updates are dispatched at
+random to a large array. This benchmark stresses the networking layer of
+Grappa separately from the scheduler, because only a single worker is used per
+system node.
 
-\paragraph{Unbalanced tree search in-memory (UTS-Mem):} Unbalanced Tree Search (UTS) is a benchmark for evaluating the programmability and performance of systems for parallel applications that require dynamic load balancing~\cite{UTS}.
-It involves traversing an unbalanced implicit tree: at each vertex, its number of children is sampled from some probability distribution, and this number of new nodes are added to a work queue to be visited.
-While this benchmark captures irregular, dynamic \emph{computation,} we actually want to evaluate performance of algorithms with irregular \emph{memory\/} access patterns.
-Thus we augment UTS by using the existing traversal code to create a large tree in memory, and then we traverse the in-memory tree.
-We call this benchmark UTS-Mem, and the timed portion is this traversal of the in-memory tree.
-This in-memory traversal has no knowledge of the tree structure beforehand.
+\nonindent{\bf Unbalanced tree search in-memory (UTS-Mem):} Unbalanced Tree Search
+(UTS) is a benchmark for evaluating the programmability and performance of
+systems for parallel applications that require dynamic load
+balancing~\cite{UTS}. It involves traversing an unbalanced implicit tree: at
+each vertex, its number of children is sampled from some probability
+distribution, and this number of new nodes are added to a work queue to be
+visited. While this benchmark captures irregular, dynamic \emph{computation,}
+we actually want to evaluate performance of algorithms with irregular
+\emph{memory\/} access patterns. Thus we augment UTS by using the existing
+traversal code to create a large tree in memory, and then we traverse the
+in-memory tree. We call this benchmark UTS-Mem, and the timed portion is this
+traversal of the in-memory tree. This in-memory traversal has no knowledge of
+the tree structure beforehand.
 
-\paragraph{Breadth-first-search (BFS):} This is the primary kernel for the Graph500 benchmark and is what currently determines the ranking of machines on the Graph500 list~\cite{graph500list}.
-As a whole, the Graph500 benchmark suite is designed to bring the focus of system design on data-intensive workloads, particularly large-scale graph analysis problems, that are important among cybersecurity, informatics, and network understanding workloads.
-The BFS benchmark builds a search tree containing parent nodes for each traversed vertex during the search.
-While this is a relatively simple problem to solve, it exercises the random-access and fine-grained synchronization capabilities of a system as well as being a primitive in many other graph algorithms.
-Performance is measured in \emph{traversed edges per second\/} (TEPS), where the number of edges is the edges making up the generated BFS tree.
-One of the reference implementations of Graph500 BFS is for the XMT; this code fails to scale past 16 XMT processors because it does not expose enough parallelism, so we modified the code to use a recursive loop decomposition similar to Grappa's.
-We compare this modified version against a straightforward Grappa implementation.
-We do not employ algorithmic improvements, though there are many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
+\nonindent{\bf Breadth-first-search (BFS):} This is the primary kernel for the
+Graph500 benchmark and is what currently determines the ranking of machines on
+the Graph500 list~\cite{graph500list}. As a whole, the Graph500 benchmark
+suite is designed to bring the focus of system design on data-intensive
+workloads, particularly large-scale graph analysis problems, that are
+important among cybersecurity, informatics, and network understanding
+workloads. The BFS benchmark builds a search tree containing parent nodes for
+each traversed vertex during the search. While this is a relatively simple
+problem to solve, it exercises the random-access and fine-grained
+synchronization capabilities of a system as well as being a primitive in many
+other graph algorithms. Performance is measured in \emph{traversed edges per
+second\/} (TEPS), where the number of edges is the edges making up the
+generated BFS tree. One of the reference implementations of Graph500 BFS is
+for the XMT; this code fails to scale past 16 XMT processors because it does
+not expose enough parallelism, so we modified the code to use a recursive loop
+decomposition similar to Grappa's. We compare this modified version against a
+straightforward Grappa implementation. We do not employ algorithmic
+improvements, though there are
+many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
 
-\paragraph{IntSort:}  Write me Brandon Holt!
+\nonindent{\bf IntSort:} \TODO{Write me Brandon Holt!}
 
-The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.
-Grappa collects statistics about application behavior (packets sent, context switches, etc) and these are discussed where appropriate.
-In addition we use a profiled version of Grappa to report component runtimes, but this profiled version is only used to collect this data and a fully optimized version is used for the bulk of the presented results.
+\nonindent{\bf PageRank:} \TODO{Write me Brandon Myers!}
+
+The metric we use is algorithmic time, which means startup and loading of the
+data structure (from disk) is not included in the measurement. \Grappa
+collects statistics about application behavior (packets sent, context
+switches, etc) and these are discussed where appropriate. In addition we use a
+profiled version of \Grappa to report component runtimes, but this profiled
+version is only used to collect this data and a fully optimized version is
+used for the bulk of the presented results.
 

### 2013-03-28 doc/sosp13/method.tex
small edits to method.

@@
 
-\nonindent{\bf GUPS:} This is a faithful implementation of the
+\noindent{\bf GUPS:} This is a faithful implementation of the
 giga-updates-per-second benchmark. Read-modify-write updates are dispatched at
@@
 
-\nonindent{\bf Unbalanced tree search in-memory (UTS-Mem):} Unbalanced Tree Search
+\noindent{\bf Unbalanced tree search in-memory (UTS-Mem):} Unbalanced Tree Search
 (UTS) is a benchmark for evaluating the programmability and performance of
@@
 
-\nonindent{\bf Breadth-first-search (BFS):} This is the primary kernel for the
+\noindent{\bf Breadth-first-search (BFS):} This is the primary kernel for the
 Graph500 benchmark and is what currently determines the ranking of machines on
@@
 
-\nonindent{\bf IntSort:} \TODO{Write me Brandon Holt!}
+\noindent{\bf IntSort:} \TODO{Write me Brandon Holt!}
 
-\nonindent{\bf PageRank:} \TODO{Write me Brandon Myers!}
+\noindent{\bf PageRank:} \TODO{Write me Brandon Myers!}
 

### 2013-03-28 doc/sosp13/evaluation.tex
finished eval structure

@@
 
-\subsection{Basic Grappa mechanisms}
+\subsection{Basic \Grappa Mechanisms}
 
-\subsection{Fast User-Level Context Switching }
+\paragraph{User-level context switching.}
+As discussed earlier, fast context switching is at the heart of \Grappa's
+latency tolerance abilities. We assess context switch overheads using a simple
+microbenchmark that has a configurable number of workers, where each worker 
+just increments values in a large array.
 
@@
       \includegraphics[width=0.5\textwidth]{figs/context_switch_time.pdf}
+%      \includegraphics[width=0.5\textwidth]{figs/context_switch_bw.pdf}
     \end{center}
-    \caption{Average context switch time with 1 and 6 active cores,
-        with and without prefetching.}
-    \label{fig:context-switch-time}
+    \caption{Average context switch time with 1 active, with and without prefetching.}
+    \label{fig:context-switch-exp}
 \end{figure}
+
+Figure~\ref{fig:context-switch-exp} shows the average context switch time as
+the number of workers grow. There are several important takeaway points form
+these data. Context switch time for the number of workers we use currently
+($~$1K) is on the order of 50 ns; and prefetching has significant impact on
+context switch time. And ontext switch time does not grow significantly with
+the number of workers --- even at close to one half million workers context
+switch is around 75ns.
+
+We also measured (not in the plot) the \emph{rate} of context switch for all
+cores in a node, which showed that our rate is limited by off-chip memory
+bandwidth. Each context is 4 cache lines (1 for the worker struct and 3 for
+stack data), leading to 8 cache line transfers per context switch
+(write the previous context, read the next in). The off-chip bandwidth of a
+single socket in our system is 270M cache lines per second, which implies that
+we can sustain at most 34M context switches per second per socket, which is
+almost exactly what we sustain.
+
+In summary, our context switch engine is able to efficiently sustain very high concurrency and we will show later, the amount of concurrency sustained is sufficient the latencies \Grappa needs to hide.
+
+\paragraph{Global memory and communication.} We measure the performance of
+\Grappa's global memory and communication layers using a faithful
+implementation of the giga updates per second (GUPS) benchmark.
+Read-modify-write updates are dispatched at random to a large array. This
+benchmark stresses the communication layer of \Grappa separately from the
+scheduler, because only a single worker is used per system node.
+Figure~\ref{fig:grappa-gups} shows that Grappa is able to sustain on the order of billions of updates per second, which is enough to \TODO{...}.
 
@@
     \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/context_switch_bw.pdf}
+%      \includegraphics[width=0.5\textwidth]{figs/grappa-gups.pdf}
     \end{center}
-    \caption{Context switch rate with prefetching. Once the total
-        size of contexts sufficiently exceeds last-level cache, most
-    prefetches go to main memory and the rate becomes limited to the
-    off-chip bandwidth.}
-    \label{fig:context-switch-time}
+    \caption{GUPS (giga updates per second) for \Grappa as the number of nodes grows.}
+    \label{fig:grappa-gups}
 \end{figure}
 
-% More specifically, each context is cache lines: 1 for the worker struct and 3
-% for stack data. So the are so 8 cacheline transfers per context switch (write
-% the old context, read the next in). The bandwidth of single socket is 270M cache lines/s.
-% 
-% This asymptotically approaches
-% 33.75Mcontexts/s as you increase the number of contexts (fewer and fewer in
-% L3). Note that only 4+ cores reaches full rate because these westmere chips
-% are balanced to not achieve full off-chip bandwidth until 4 cores.
+\paragraph{Putting it all together with Unbalanced Tree Search (UTS).}
+We now show the overall performance of \Grappa running UTS.
+Figure~\ref{fig:grappa-uts}.
+The point of this experiment is to show that \Grappa's context switching and communication layers can in fact be used together to run an irregular application efficiently.
+Looking at the performance as the number of nodes grows shows \TODO{...}.
+
+\begin{figure}[ht]
+    \begin{center}
+%      \includegraphics[width=0.5\textwidth]{figs/grappa-uts.pdf}
+    \end{center}
+    \caption{Vertices per second in UTS on \Grappa as the number of nodes grows.}
+    \label{fig:grappa-uts}
+\end{figure}
+
+\subsection{Comparing \Grappa to Other Systems}
+
+In order to ground \Grappa's performance, we compare it with XMT running BFS,
+PageRank, IntSort, GUPS and UTS. Since XMT is a different hardware platform,
+we also compare \Grappa with hand-tuned MPI versions of BFS and GUPS running
+on the same hardware. Finally, we also compare it with UTS written for UPC. We
+run all experiments with 64 nodes.
+
+\begin{figure}[ht]
+    \begin{center}
+%      \includegraphics[width=0.5\textwidth]{figs/grappa-uts.pdf}
+    \end{center}
+    \caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC.}
+%bar chart, one set of bars per benchmark, one bar per system. runtime normalized to grappa.
+    \label{fig:grappa-comparisons}
+\end{figure}
+
+Figure~\ref{fig:grappa-comparisons} shows the results. Overall, \Grappa is
+within striking distance of XMT and UPC performance. However, it underperforms
+compared to hand-tuned MPI. Nevertheless, this comparison needs to be taken
+with a grain of salt for several reasons. First, we have not spent as much
+time tuning \Grappa's implementations. This is supported by inspecting the BFS
+MPI implementation we used, which clearly shows that it employs a lot of
+algorithmic-specific optimizations. In fact, some of these optimizations
+resemble some of what \Grappa does automatically, like message aggregation.
+Second, the GUPS results presented earlier suggests that \Grappa performance
+can do a lot better. \TODO{update this once we have more results. }
+
+\subsection{Characterization}
+
+\paragraph{Where execution time goes.}
+
+\paragraph{Message size and latency.}
+Distribution if possible. With and without aggregation.
+
+\paragraph{Frequency of remote requests.}
+
+\paragraph{Work stealing behavior.}
+Performance loss of not having stealing. Ratio of \# of steals over total \# of task spawns. UTS needs stealing to work.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% OLD %%%%%%%%%%%%%%%%%%
+
+\comment{
 
@@
 %\end{figure*}
+
+}

### 2013-03-28 doc/sosp13/method.tex
finished eval structure

@@
 
-\noindent{\bf Context-switch:} This simple microbenchmark is used to explore context switch overhead.
+%%% Moved the below to the evaluation sec
+% \vspace{0.5ex}
+% \noindent{\bf Context-switch:} This simple microbenchmark is used to explore context switch overhead.
 A configurable number of workers are used to increment the values in a large array.
 This benchmark stresses the scheduler inside of Grappa.
+% 
+% \vspace{0.5ex}
+% \noindent{\bf GUPS:} This is a faithful implementation of the
+% giga-updates-per-second benchmark. Read-modify-write updates are dispatched at
+% random to a large array. This benchmark stresses the networking layer of
+% Grappa separately from the scheduler, because only a single worker is used per
+% system node.
 
-\noindent{\bf GUPS:} This is a faithful implementation of the
-giga-updates-per-second benchmark. Read-modify-write updates are dispatched at
-random to a large array. This benchmark stresses the networking layer of
-Grappa separately from the scheduler, because only a single worker is used per
-system node.
-
+\vspace{0.5ex}
 \noindent{\bf Unbalanced tree search in-memory (UTS-Mem):} Unbalanced Tree Search
@@
 
+\vspace{0.5ex}
 \noindent{\bf Breadth-first-search (BFS):} This is the primary kernel for the
@@
 
+\vspace{0.5ex}
 \noindent{\bf IntSort:} \TODO{Write me Brandon Holt!}
 
+\vspace{0.5ex}
 \noindent{\bf PageRank:} \TODO{Write me Brandon Myers!}

### 2013-03-28 doc/sosp13/intro.tex
edits

@@
 
-\vspace{0.5ex}
-\noindent{\bf High network communication latency.} The performance challenges
-of frequent communication is exacerbated by the network latency relative to
-processor performance. Latency of commodity networks runs anywhere from one to
-hundreds of microseconds, which is much more than superscalar execution with
-sophisticated memory hierarchies can tolerate. Since irregular application
-tasks encounter remote references dynamically during execution and must
-resolve them before making further progress, stalls are frequent and lead
-severely underutilized compute resources.
+\vspace{0.5ex} \noindent{\bf High network communication latency.} The
+performance challenges of frequent communication is exacerbated by the network
+latency relative to processor performance. Latency of commodity networks runs
+anywhere from a few to hundreds of microseconds, which is much more than
+superscalar execution (even with sophisticated memory hierarchies) can
+tolerate. Since irregular application tasks encounter remote references
+dynamically during execution and must resolve them before making further
+progress, stalls are frequent and lead to severely underutilized compute
+resources.
 
-While some irregular applications can be restructured to better exploit
-locality, aggregate requests to increase network message size, and manage the
-additional challenges of load balance and synchronization, the effort required
-to do so is formidable and involves knowledge and skills pertaining to
-distributed systems far beyond those of most application programmers. Luckily,
-many of the important irregular applications (e.g., graph processing, our
-focus in this paper) naturally offer large amounts of concurrency. This
-immediately suggests taking advantage of concurrency to tolerate the latency
-of data movement by overlapping computation with communication.
+While some irregular applications can be manually restructured to better
+exploit locality, aggregate requests to increase network message size, and
+manage the additional challenges of load balance and synchronization, the
+effort required to do so is formidable and involves knowledge and skills
+pertaining to distributed systems far beyond those of most application
+programmers. Luckily, many of the important irregular applications (e.g.,
+graph processing, our focus in this paper) naturally offer large amounts of
+concurrency. This immediately suggests taking advantage of concurrency to
+tolerate the latency of data movement by overlapping computation with
+communication.
 
@@
 underlying mechanism for remote memory reads and writes using active message
-invocations. To this commodity hardware and software mix \Grappa adds three
-main software components: (1) a \emph{lightweight tasking\/} layer that
-supports a context switch in as few as 38ns and distributed global load
-balancing; (2) a \emph{distributed shared memory\/} layer that supports normal
-access operations such as \emph{read\/} and \emph{write\/} as well as
-synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd};
-and (3) a \emph{communication\/} layer that combines short messages to
-achieve peak bandwidth on commodity networks. As we will show later, \Grappa
-can tolerate latencies way beyond that of the network. Therefore, \Grappa can
-afford to \emph{trade latency for throughput}. By {\em increasing\/} latency
-in key components of the system we are able to: increase effective random
-access memory bandwidth by delaying and aggregating messages; increase
-synchronization rate by delegating atomic operations to gatekeeper cores, even
-when referencing node-local global data; and improve load balance by
-tolerating the delays incurred when work-stealing.
+invocations. \Grappa adds three main software components: (1) a
+\emph{lightweight tasking\/} layer that supports a context switch in as little
+as 38ns and distributed global load balancing; (2) a \emph{distributed shared
+memory\/} layer that supports normal access operations such as \emph{read\/}
+and \emph{write\/} as well as synchronizing operations such as
+\emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{communication\/}
+layer that combines short messages to achieve peak bandwidth on commodity
+networks. As we will show later, \Grappa can tolerate latencies way beyond
+that of the network. Therefore, \Grappa can afford to \emph{trade latency for
+throughput}. By {\em increasing\/} latency in key components of the system we
+are able to: increase effective random access memory bandwidth by delaying and
+aggregating messages; increase synchronization rate by delegating atomic
+operations to gatekeeper cores, even when referencing node-local global data;
+and improve load balance by tolerating the delays incurred when work-stealing.
 

### 2013-03-28 doc/sosp13/conclude.tex
macrotize all of grappa

\section{Conclusion}

Irregular computations are both important and challenging to execute quickly.
Scaling these applications easily to commodity hardware has been a historical
challenge. \Grappa is a runtime framework that simplifies this task for
software developers and compiler writers. This paper describes the \Grappa
framework and its three main components: a task library, a distributed shared
memory system, and a network aggregator to make commodity networks efficient
with small message sizes. \Grappa key aspect is extreme latency tolerance,
which not only hides network latency but also enables the system to spend time
on sophisticated work stealing and network optimizations, trading latency for
even more throughput.

We explored the performance of \Grappa on three algorithms: unbalanced tree
search, breadth first search, and betweenness centrality. Performance
comparisons to the Cray XMT for a small number of nodes (16 nodes / 96 cores)
show great promise: \Grappa performs on-par with the far more expensive and
custom XMT system, sometimes achieving better performance and sometimes worse.
Our initial large scaling experiments to a higher number of nodes shows both promise and room for improvement.  \Grappa trounces XMT on UTS, but while \Grappa performance scales on BFS and betweenness centrality, limitations in the current aggregator design give the performance edge to XMT.  Work is ongoing on further refining this component.

### 2013-03-28 doc/sosp13/evaluation.tex
benchmark plot
macrotize all of grappa

@@
 scheduler, because only a single worker is used per system node.
-Figure~\ref{fig:grappa-gups} shows that Grappa is able to sustain on the order of billions of updates per second, which is enough to \TODO{...}.
+Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain on the order of billions of updates per second, which is enough to \TODO{...}.
 
@@
     \begin{center}
-%      \includegraphics[width=0.5\textwidth]{figs/grappa-uts.pdf}
+      \includegraphics[width=0.5\textwidth]{results/benchmarks.pdf}
     \end{center}
@@
 Our evaluation begins with presentation of microbenchmark results, establishing
-the intrinsic potential of Grappa to provide random access bandwidth and latency tolerance. 
-Next, we present application results, both for Grappa and for other paradigms, as well
+the intrinsic potential of \Grappa to provide random access bandwidth and latency tolerance. 
+Next, we present application results, both for \Grappa and for other paradigms, as well
 as comparing against the Cray XMT.  Finally, we present the impact of increased aggregation delay
-on Grappa results, thus exploring robustness to network scale.
+on \Grappa results, thus exploring robustness to network scale.
 
@@
 \subsubsection{Random Access}
-\TODO{Random access feed forward results on Grappa.  optional: Results we measured for XMT.  Cite MPI results}
+\TODO{Random access feed forward results on \Grappa.  optional: Results we measured for XMT.  Cite MPI results}
 \subsubsection{Latency Tolerance}
-\TODO{Simple ping test results -- eg, MPI ping, not the full blown aggregation ping test.  Random access blocking results on Grappa.}
+\TODO{Simple ping test results -- eg, MPI ping, not the full blown aggregation ping test.  Random access blocking results on \Grappa.}
 \subsubsection{Scheduling and Robustness}
@@
 \subsection{Application Results}
-To evaluate Grappa's performance with respect to the XMT, we ran each
-of our three benchmarks on up to 16 nodes of each machine. Grappa used
+To evaluate \Grappa's performance with respect to the XMT, we ran each
+of our three benchmarks on up to 16 nodes of each machine. \Grappa used
 6 cores per node, with the best parameters chosen for each point. In
@@
 nodes.
-Grappa is 3.2 times faster than the XMT at 16 nodes.
-As we will show later, the performance advantage Grappa has over XMT increases as more nodes are added.
-The main reason Grappa performs better is the software-based delegate synchronization obviates the need for the retry-based synchronization that XMT uses.
+\Grappa is 3.2 times faster than the XMT at 16 nodes.
+As we will show later, the performance advantage \Grappa has over XMT increases as more nodes are added.
+The main reason \Grappa performs better is the software-based delegate synchronization obviates the need for the retry-based synchronization that XMT uses.
 
@@
 our performance in terms of graph edges traversed per second. The XMT
-is 2.5 times faster than Grappa at 16 nodes.  Performance does scale at a constant rate for Grappa, suggesting that adding more nodes will increase performance.
+is 2.5 times faster than \Grappa at 16 nodes.  Performance does scale at a constant rate for \Grappa, suggesting that adding more nodes will increase performance.
 
@@
 performance in terms of graph edges traversed per second. At 16 XMT
-processors/cluster nodes, the XMT is 1.75 times faster than Grappa.
+processors/cluster nodes, the XMT is 1.75 times faster than \Grappa.
 
@@
 %\paragraph{UTS-Mem}
-%We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
+%We ran UTS-Mem on \Grappa and the XMT with a geometric 1.6B-vertex tree
 %(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
-%nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
-%Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
-%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
-%The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
+%nodes---the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
+
+%\Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
+%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, \Grappa adds 1.4 Mvert/s/node.
+%The XMT scales at 850 Kvert/s/node, until it plateaus. \Grappa keeps
 %scaling up through 128 nodes, although scaling
@@
 %
-To determine how Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
+To determine how \Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
 For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
-For Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
-All of the benchmarks continue to improve out to 128 nodes for Grappa.
-UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while Grappa handles this by suspending tasks until messages return.
-For BFS and Centrality, the XMT scales approximately a constant factor better than Grappa.
-We attribute this to a limitation in the current aggregator design and network stack that Grappa uses.
+For \Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
+All of the benchmarks continue to improve out to 128 nodes for \Grappa.
+UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while \Grappa handles this by suspending tasks until messages return.
+For BFS and Centrality, the XMT scales approximately a constant factor better than \Grappa.
+We attribute this to a limitation in the current aggregator design and network stack that \Grappa uses.
 This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases} performance).
-Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
+Ironically, this limitation makes \Grappa applications compute-bound instead of network-bound.
 Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
@@
     \end{center}
-    \caption{Scaling number of nodes: Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
+    \caption{Scaling number of nodes: \Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
     \label{fig:uts_threshold}
@@
   \caption{\label{fig:aggregator-ping} Bandwidth versus message size
-    unidirectional ping test for Grappa with aggregation, Grappa with
+    unidirectional ping test for \Grappa with aggregation, \Grappa with
     raw GASNet messages, and MPI. Aggregation provides an 11x
@@
 
-To implement the ping test, we wrote a simple Grappa application where
+To implement the ping test, we wrote a simple \Grappa application where
 the cores of one node send messages as fast as possible to the cores
@@
 benchmark has the same pattern of communication but doesn't have the
-overhead of Grappa's context switching.
+overhead of \Grappa's context switching.
 
@@
 average message between 32 and 64 bytes. At 32 bytes, the MPI test is
-using less than 7 percent of its peak bandwidth. Grappa sending
+using less than 7 percent of its peak bandwidth. \Grappa sending
 messages directly through GASNet uses less than 3 percent of the peak
@@
 Second, aggregation has the potential to improve this situation by an
-order of magnitude. With aggregation, Grappa is able to send 32-byte
+order of magnitude. With aggregation, \Grappa is able to send 32-byte
 messages over 12 times faster than using GASNet directly. This is a
 more respectable 32 percent of peak bandwidth. Due to expedient design
-decisions, Grappa's aggregator limits its aggregation to 4KB; this
+decisions, \Grappa's aggregator limits its aggregation to 4KB; this
 limits its peak achievable bandwidth to 75 percent of the actual
@@
   \caption{\label{fig:no-aggregation-uts} Performance of UTS on 16
-      nodes with and without Grappa's aggregation.}
+      nodes with and without \Grappa's aggregation.}
 \end{minipage}
@@
 %\paragraph{UTS-Mem}
-%We ran UTS-Mem on Grappa and the XMT with a geometric 1.6B-vertex tree
+%We ran UTS-Mem on \Grappa and the XMT with a geometric 1.6B-vertex tree
 %(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
-%nodes---the maximum we had available for each. Grappa results are for 5 cores per node. Grappa with 20 machines is faster than the entire XMT of 128 processors.
-%Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
-%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, Grappa adds 1.4 Mvert/s/node.
-%The XMT scales at 850 Kvert/s/node, until it plateaus. Grappa keeps
+%nodes---the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
+%\Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
+%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, \Grappa adds 1.4 Mvert/s/node.
+%The XMT scales at 850 Kvert/s/node, until it plateaus. \Grappa keeps
 %scaling up through 128 nodes, although scaling
@@
 %
-To determine how Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
+To determine how \Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
 For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
-For Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
-All of the benchmarks continue to improve out to 128 nodes for Grappa.
-UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while Grappa handles this by suspending tasks until messages return.
-For BFS and Centrality, the XMT scales approximately a constant factor better than Grappa.
-We attribute this to a limitation in the current aggregator design and network stack that Grappa uses.
+For \Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
+All of the benchmarks continue to improve out to 128 nodes for \Grappa.
+UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while \Grappa handles this by suspending tasks until messages return.
+For BFS and Centrality, the XMT scales approximately a constant factor better than \Grappa.
+We attribute this to a limitation in the current aggregator design and network stack that \Grappa uses.
 This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).
-Ironically, this limitation makes Grappa applications compute-bound instead of network-bound.
+Ironically, this limitation makes \Grappa applications compute-bound instead of network-bound.
 Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
@@
 synchronization---can reduce the performance benefit of parallelism.
-Grappa sees a benefit to limiting the amount of parallelism created by
+\Grappa sees a benefit to limiting the amount of parallelism created by
 a recursive loop decomposition. The parallel loop threshold (``parallel
@@
 On a Cray XMT, network bandwidth is limited, so retries create congestion.
-In comparison, Grappa performs synchronization without retries, delaying responses at the receiving end until ready to notify the sender to proceed.
+In comparison, \Grappa performs synchronization without retries, delaying responses at the receiving end until ready to notify the sender to proceed.
 This saves bandwidth and permits scaling of tasks performing synchronization even on low injection rate networks.

### 2013-03-28 doc/sosp13/intro.tex
macrotize all of grappa

@@
 
-Our evaluation of Grappa shows that it runs several irregular application
+Our evaluation of \Grappa shows that it runs several irregular application
 kernels (e.g., graph traversal and sort) very efficiently on a commodity
@@
 same number of network interfaces, \Grappa is in the same ballpark as the XMT:
-For unbalanced tree search, Grappa is over \TODO{3X} faster and shows greatly
+For unbalanced tree search, \Grappa is over \TODO{3X} faster and shows greatly
 improved scalability; conversely, for breadth first search and betweenness
-centrality Grappa is 2.5X slower. In Section~\ref{sec:evaluation} we explore
+centrality \Grappa is 2.5X slower. In Section~\ref{sec:evaluation} we explore
 the factors that underpin this performance. Most importantly, however, for
 significantly less real world cost, users can \emph{add\/} significantly more
-processors to a commodity cluster than an XMT machine and use Grappa to
+processors to a commodity cluster than an XMT machine and use \Grappa to
 achieve scalable performance. \TODO{revise paragraph after eval, including MPI

### 2013-03-28 doc/sosp13/method.tex
macrotize all of grappa

@@
 A configurable number of workers are used to increment the values in a large array.
-This benchmark stresses the scheduler inside of Grappa.
+This benchmark stresses the scheduler inside of \Grappa.
 % 
@@
 % random to a large array. This benchmark stresses the networking layer of
-% Grappa separately from the scheduler, because only a single worker is used per
+% \Grappa separately from the scheduler, because only a single worker is used per
 % system node.
@@
 not expose enough parallelism, so we modified the code to use a recursive loop
-decomposition similar to Grappa's. We compare this modified version against a
-straightforward Grappa implementation. We do not employ algorithmic
+decomposition similar to \Grappa's. We compare this modified version against a
+straightforward \Grappa implementation. We do not employ algorithmic
 improvements, though there are
@@
 Performance for BC is also measured in TEPS, where the number of traversed edges is the total number of edges in the graph multiplied by the number of random starting vertices used in the approximate computation.
-We use the XMT implementation of BC implemented as part of the GraphCT~\cite{GraphCT} library and a comparable Grappa implementation, both of which use the parallel BC algorithm developed by Bader et al.~\cite{bader:bc}.
+We use the XMT implementation of BC implemented as part of the GraphCT~\cite{GraphCT} library and a comparable \Grappa implementation, both of which use the parallel BC algorithm developed by Bader et al.~\cite{bader:bc}.
 
-%To explore the performance of the Grappa runtime we have implemented three algorithms: breadth first search, betweenness centrality, and unbalanced tree search.
-These algorithms were implemented for the Cray XMT (our baseline) and for Grappa.
+%To explore the performance of the \Grappa runtime we have implemented three algorithms: breadth first search, betweenness centrality, and unbalanced tree search.
+These algorithms were implemented for the Cray XMT (our baseline) and for \Grappa.
 The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.
@@
 
-\comment{The Grappa version of tree search UTS-mem
+\comment{The \Grappa version of tree search UTS-mem
 
@@
 We discounted the last option fairly quickly, because it isn't a lasting data-point -- the cost of hardware shifts over time, skewing the interpretation of results.
-The first option, cores, has some merit, but Grappa is designed for applications that have no locality in their computation.
+The first option, cores, has some merit, but \Grappa is designed for applications that have no locality in their computation.
 This means almost all of their memory accesses are remote.
@@
 
-%Towards evaluating how well Grappa enables scalable performance of irregular applications on mass market systems, our methodology is as follows.
+%Towards evaluating how well \Grappa enables scalable performance of irregular applications on mass market systems, our methodology is as follows.
 For consistency in measurement, all results reported for mass market systems are measured on the same platform.
-To establish the potential for Grappa performance, we identify small microkernel benchmarks that provide insight into the limitations of our system's component performance behaviors.
-To compare meaningfully with the state-of-the-art, we compare Grappa performance on well known irregular application benchmarks both against the popular high performance programming methodologies MPI and UPC for distributed memory clusters, and against a custom shared memory architecture used exclusively for irregular computation, the Cray XMT.
-We then examine how sensitive Grappa results are to reduced injection rate and increased latency, as often occur when a system is scaled up so that nodes are further apart and the effective per destination packet injection rate drops.
+To establish the potential for \Grappa performance, we identify small microkernel benchmarks that provide insight into the limitations of our system's component performance behaviors.
+To compare meaningfully with the state-of-the-art, we compare \Grappa performance on well known irregular application benchmarks both against the popular high performance programming methodologies MPI and UPC for distributed memory clusters, and against a custom shared memory architecture used exclusively for irregular computation, the Cray XMT.
+We then examine how sensitive \Grappa results are to reduced injection rate and increased latency, as often occur when a system is scaled up so that nodes are further apart and the effective per destination packet injection rate drops.
 The three sections that follow describe this methodology in greater detail.

### 2013-03-28 doc/sosp13/intro.tex
fixed up intro

@@
 \vspace{0.5ex}
-\noindent{\bf Poor data locality and frequent communication.} Data reference
-patterns of irregular applications are unpredictable and tend to be spread
-across the entire system. This results in frequent requests for remote data.
-Caches are of little assistance because of low data re-use and spatial
-locality. Prefetching is ineffective because request locations are not known
-early enough. Data-parallel frameworks such as
-MapReduce~\cite{mapreduce:osdi04} are also inappropriate because they rely on
-being able to partition data and regular communication patterns. As a
-consequence, irregular applications require frequent communication of
-typically small pieces of data. In contrast, commodity networks are designed
-for large packets and have just a fraction of their peak bandwidth when moving
-small packets.
+\noindent{\bf Poor data locality and frequent communication.} Data reference patterns of irregular applications are unpredictable and tend to be spread across the entire system.
+This results in frequent requests for small pieces of remote data.
+Commodity networks are designed for large packets and achieve just a fraction of their peak bandwidth when moving small packets.
+Caches are of little assistance because of low data re-use and spatial locality.
+Prefetching is ineffective because request locations are not known early enough.
+Data-parallel frameworks such as MapReduce~\cite{mapreduce:osdi04} are also inappropriate because they rely on being able to partition data and regular communication patterns.
 
-\vspace{0.5ex} \noindent{\bf High network communication latency.} The
-performance challenges of frequent communication is exacerbated by the network
-latency relative to processor performance. Latency of commodity networks runs
-anywhere from a few to hundreds of microseconds, which is much more than
-superscalar execution (even with sophisticated memory hierarchies) can
-tolerate. Since irregular application tasks encounter remote references
-dynamically during execution and must resolve them before making further
-progress, stalls are frequent and lead to severely underutilized compute
-resources.
+\vspace{0.5ex} \noindent{\bf High network communication latency.} The performance challenges of frequent communication is exacerbated by the network latency relative to processor performance.
+Latency of commodity networks runs anywhere from a few to hundreds of microseconds -- ten's of thousands of processor clock cycles.
+Since irregular application tasks encounter remote references dynamically during execution and must resolve them before making further progress, stalls are frequent and lead to severely underutilized compute resources.
 
-While some irregular applications can be manually restructured to better
-exploit locality, aggregate requests to increase network message size, and
-manage the additional challenges of load balance and synchronization, the
-effort required to do so is formidable and involves knowledge and skills
-pertaining to distributed systems far beyond those of most application
-programmers. Luckily, many of the important irregular applications (e.g.,
-graph processing, our focus in this paper) naturally offer large amounts of
-concurrency. This immediately suggests taking advantage of concurrency to
-tolerate the latency of data movement by overlapping computation with
-communication.
+While some irregular applications can be manually restructured to better exploit locality, aggregate requests to increase network message size, and manage the additional challenges of load balance and synchronization, the effort required to do so is formidable and involves knowledge and skills pertaining to distributed systems far beyond those of most application programmers.
+Luckily, many of the important irregular applications naturally offer large amounts of concurrency.
+This immediately suggests taking advantage of concurrency to tolerate the latency of data movement by overlapping computation with communication.
 
@@
 
-We believe we can support irregular applications with good performance and
-cost-effectiveness with commodity hardware for two main reasons. First,
-commodity multicore processors have become extremely fast and cheap. So with
-careful software engineering we can multiplex hundred of thousands of
-workers into a single core. Second, commodity networks offer high
-bandwidth as long as messages are large enough. So with enough data requests
-in flight simultaneously, the system can aggregate small messages into large
-enough ones.
+We believe we can support irregular applications with good performance and cost-effectiveness with commodity hardware for two main reasons.
+First, commodity multicore processors have become extremely fast with high clock rates, large caches and robust DRAM bandwidth.
+So with careful software engineering we can multiplex hundred of thousands of workers into a single core.
+Second, commodity networks offer high bandwidth as long as messages are large enough.
+So with enough data requests in flight simultaneously, the system can aggregate small messages into large enough ones.
 
-In this paper we introduce \Grappa, a software runtime system that allows a
-commodity cluster of x86-based nodes connected via an InfiniBand network to be
-programmed as if it were a single, large, shared-memory NUMA (non-uniform
-memory access) machine with scalable performance for irregular applications.
-\Grappa is designed to smooth over some of the performance discontinuities in
-commodity hardware, giving good performance when there is little locality to
-be exploited while allowing the programmer to exploit it when it is available.
-
-\Grappa leverages as much freely available and commodity infrastructure as
-possible. We use unmodified Linux for the operating system and an
-off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}. MPI is
-used for process setup and tear down. GASNet~\cite{gasnet} is used as the
-underlying mechanism for remote memory reads and writes using active message
-invocations. \Grappa adds three main software components: (1) a
-\emph{lightweight tasking\/} layer that supports a context switch in as little
-as 38ns and distributed global load balancing; (2) a \emph{distributed shared
-memory\/} layer that supports normal access operations such as \emph{read\/}
-and \emph{write\/} as well as synchronizing operations such as
-\emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{communication\/}
-layer that combines short messages to achieve peak bandwidth on commodity
-networks. As we will show later, \Grappa can tolerate latencies way beyond
-that of the network. Therefore, \Grappa can afford to \emph{trade latency for
-throughput}. By {\em increasing\/} latency in key components of the system we
-are able to: increase effective random access memory bandwidth by delaying and
-aggregating messages; increase synchronization rate by delegating atomic
-operations to gatekeeper cores, even when referencing node-local global data;
-and improve load balance by tolerating the delays incurred when work-stealing.
+In this paper we introduce \Grappa, a software runtime system that allows a commodity cluster of x86-based nodes connected via an InfiniBand network to be programmed as if it were a single, large, shared-memory NUMA (non-uniform memory access) machine with scalable performance for irregular applications.
+\Grappa leverages as much freely available and commodity infrastructure as possible.
+We use unmodified Linux for the operating system and an off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}.
+MPI is used for process setup and tear down.
+GASNet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
+\Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as little as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{communication\/} layer that combines short messages to achieve peak bandwidth on commodity networks.
+As we will show later, \Grappa can tolerate latencies way beyond that of the network.
+Therefore, \Grappa can afford to \emph{trade latency for throughput}.
+By {\em increasing\/} latency in key components of the system we are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance by tolerating the delays incurred when work-stealing.
 
@@
 and possibly programming model}
+
+%%% bit bucket
+%\Grappa is designed to smooth over some of the performance discontinuities in commodity hardware, giving good performance when there is little locality to be exploited while allowing the programmer to exploit it when it is available.

### 2013-03-28 doc/sosp13/evaluation.tex
uts plot
uts story todo
Added profile table

@@
 \end{figure}
+\TODO{UTS: story here is T1 class is bushy trees similar to properties
+    of real world networks. Larger T1XXL (4.5 B vertices) provides
+    more parallelism for continued scaling up. T3 class is deep skinny
+    trees. Grappa can do no better than the straightforward UPC impl: there very little
+    parallelism and the critical path is long}
 
@@
 \paragraph{Where execution time goes.}
+
+\begin{figure}[ht]
+    \begin{center}
+      \begin{tabular}{c|c c c c}
+        Benchmark     & Communication & User & Idle & Context Switch \\ \hline
+        GUPS          & 6.60  & 42.94   & 47.74 & 2.80 \\
+        BFS           & 54.84 & 30.90   & 10.94 & 3.43 \\ 
+        Intsort       & 34.28 & 42.00   & 21.31 & 2.47 \\ 
+        UTS           & 40.57 & 56.52   &  1.21 & 1.73 \\
+        Pagerank      & 76.79 & 20.71   &  0.06 & 2.50 \\
+      \end{tabular}
+    \end{center}
+    \caption{\Grappa\ execution time profile, in percent.}
+%bar chart, one set of bars per benchmark, one bar per system. runtime normalized to grappa.
+    \label{fig:grappa-profile}
+\end{figure}
 
@@
     \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/uts_performance.pdf}
+      \includegraphics[width=0.5\textwidth]{figs/uts_scale.pdf}
     \end{center}

### 2013-03-28 doc/sosp13/intro.tex
tweaked intro 2

@@
 
+Our evaluation of \Grappa shows that its core components provide scalable performance.
+It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.
+The communication layer provides over 10X the small-message performance compared to the raw underlying message injection rate.
+Combining these two components, \Grappa achieves 1.2 GUPS with 64 system nodes, far above any reported cluster performance~\cite{gups} for comprable systems.
+When we compare \Grappa to native MPI implementations of more complex application kernels, the story is more muddled.
+When the MPI implementation is carefully tuned to aggregate messages inside the application -- essentially duplicating the work \Grappa is doing for the programmer, without the generality and context switch overhead -- then the MPI implementation is between 4-10X faster.
+We see this in our comparisons for breadth first search and integer sorting.
+On the other hand, if the MPI version is implemented in a straightforward manner, as we see with PageRank, GUPS, and unbalanced tree search, \Grappa is 2-5X faster.
+When we compare \Grappa to the custom Cray XMT, we find Grappa to be between 2X faster and 4X slower depending on the benchmark.
+Overall we find this encouraging because \Grappa is scalable and achieves this performance from a mass-market cluster, hence adding more system nodes an inexpensive and attractive option.
+
+\comment{
 Our evaluation of \Grappa shows that it runs several irregular application
@@
 and possibly programming model}
+}
 

### 2013-03-28 doc/sosp13/memory.tex
minor edits

@@
-\section{Memory}
+\section{Distributed Shared Memory}
 
-Applications written for Grappa utilize two forms of memory: local and global.
+Applications written for \Grappa utilize two forms of memory: local and global.
 Local memory is local to a single core within a node in the system. Accesses
@@
 memory is manipulated directly. Applications use local accesses for a number
-of things in Grappa: the stack associated with a task, accesses to localized
+of things in \Grappa: the stack associated with a task, accesses to localized
 global memory in caches (see below), and accesses to debugging infrastructure
@@
 Large data that is expected to be shared and accessed with low locality is
-stored in Grappa's global memory. All global data must be accessed through
-calls into Grappa's API, shown in Figure~\ref{fig:accessing-memory}.
+stored in \Grappa's global memory. All global data must be accessed through
+calls into \Grappa's API, shown in Figure~\ref{fig:accessing-memory}.
 
-\paragraph{Global memory addressing} Grappa provides two methods for storing
+\paragraph{Global memory addressing} \Grappa provides two methods for storing
 data in the global memory. The first is a distributed heap striped across all
@@
 has been sufficient for our benchmarks, it is easy to relax in the future
-Grappa will provide parallel performance through
+\Grappa will provide parallel performance through
 combining~\cite{MAMA,flatcombining}.}
 
-Grappa also allows any local data on a core's stacks or heap to be
+\Grappa also allows any local data on a core's stacks or heap to be
 exported to the global address space to be made accessible to other
@@
 	\end{tabular}
-      \caption{\label{fig:accessing-memory} Grappa API for memory accesses. \TODO{Update table with new calls. And async calls for delegates.}}     \end{minipage}
+      \caption{\label{fig:accessing-memory} \Grappa API for memory accesses. \TODO{Update table with new calls. And async calls for delegates.}}     \end{minipage}
   \end{center}
@@
 \begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:memory-structure} Global memory referencing in Grappa}
+  \caption{\label{fig:memory-structure} Global memory referencing in \Grappa}
 \end{minipage}
@@
 
-\paragraph{Global memory access} There are two general methods Grappa
+\paragraph{Global memory access} There are two general methods \Grappa
 applications use to {\em access} global memory. One method is to use {\em
@@
 the delegate operation completes, which is necessary, for example, to ensure
-that synchronization has finished before continuing. Synchronization in Grappa
+that synchronization has finished before continuing. Synchronization in \Grappa
 is implemented using synchronous delegate operations. Often remote data
@@
 and since they do not return data, they need a mechanism to detect completion.
-Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
+\Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
 outstanding asynchronous operations of the local core. The code and can probe
@@
 \vspace{1ex}
-\textit{Explicit caching.} Grappa provides an API to fetch a global pointer of
+\textit{Explicit caching.} \Grappa provides an API to fetch a global pointer of
 any length and return a local pointer to a cached copy of the global memory.
-Grappa cache operations have the usual read-only and read-write variants,
+\Grappa cache operations have the usual read-only and read-write variants,
 along with a write-only variant used to initialize data structures. Languages
@@
 message vectorization hoists small messages out of a loop
-\cite{FortranD:1992}. Caching in Grappa additionally provides a mechanism for
+\cite{FortranD:1992}. Caching in \Grappa additionally provides a mechanism for
 exploiting temporal locality when available by operating on the data locally.
 
-Under the hood, Grappa performs the mechanics of gathering chunks of data from
+Under the hood, \Grappa performs the mechanics of gathering chunks of data from
 multiple system nodes and presenting a conventional appearing linear block of
@@
 the constituent requests of the desired block (as Active Messages) and then
-yield until all responses have occurred. Currently, Grappa caches are
+yield until all responses have occurred. Currently, \Grappa caches are
 \emph{not\/} kept coherent automatically. If that is desired, the programmer is

### 2013-03-28 doc/sosp13/intro.tex
For my love

@@
 This results in frequent requests for small pieces of remote data.
-Commodity networks are designed for large packets and achieve just a fraction of their peak bandwidth when moving small packets.
 Caches are of little assistance because of low data re-use and spatial locality.
-Prefetching is ineffective because request locations are not known early enough.
-Data-parallel frameworks such as MapReduce~\cite{mapreduce:osdi04} are also inappropriate because they rely on being able to partition data and regular communication patterns.
+Prefetching is of limited value because request locations are not known early enough.
+Data-parallel frameworks such as MapReduce~\cite{mapreduce:osdi04} are ineffective because they rely on being able to partition data and regular communication patterns.
+Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.
 

### 2013-03-28 doc/sosp13/overview.tex
new eval. commit to merge
FL

@@
 \section{\Grappa Overview}
+
+% \TODO{Whats missing before diving into the system
+%     is why we think we can build a system with these
+%     techniques and it will work well--myers}
 
@@
 
-\item [Tasking system.] The tasking system supports lightweight multithreading
-to tolerate communication latency and global distributed work-stealing (i.e.,
-tasks can be stolen from any node in the system), which provides automated
-load balancing. The scheduler oversubscribes to have more worker threads than
-required for latency tolerance. By having at least several threads ready to
-run at all times, the scheduler can prefetch a context data into cache to
-lower the likelihood of costly main memory accesses during a context switch.
+\item [Tasking system.] The tasking system supports lightweight multithreading to tolerate communication latency and global distributed work-stealing (i.e., tasks can be stolen from any node in the system), which provides automated load balancing.
+The scheduler oversubscribes to have more worker threads than required for latency tolerance.
+By having at least four threads per processor core ready to run at all times, the scheduler can prefetch a worker's state into cache to lower the likelihood of costly main memory accesses during a context switch.
 
@@
 
-% \paragraph{Exploiting Latency Tolerance} As we will show later, Grappa can tolerate latencies way beyond that of the network.
-Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
+% \paragraph{Exploiting Latency Tolerance} As we will show later, \Grappa can tolerate latencies way beyond that of the network.
+Therefore, \Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
 

### 2013-03-28 doc/sosp13/intro.tex
move stuff out

@@
 
-In this paper we introduce \Grappa, a software runtime system that allows a commodity cluster of x86-based nodes connected via an InfiniBand network to be programmed as if it were a single, large, shared-memory NUMA (non-uniform memory access) machine with scalable performance for irregular applications.
-\Grappa leverages as much freely available and commodity infrastructure as possible.
-We use unmodified Linux for the operating system and an off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}.
-MPI is used for process setup and tear down.
-GASNet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
-\Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as little as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{communication\/} layer that combines short messages to achieve peak bandwidth on commodity networks.
-As we will show later, \Grappa can tolerate latencies way beyond that of the network.
-Therefore, \Grappa can afford to \emph{trade latency for throughput}.
-By {\em increasing\/} latency in key components of the system we are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance by tolerating the delays incurred when work-stealing.
+The two points above form the cornerstone of \Grappa, a software runtime
+system that allows a commodity cluster of x86-based nodes connected via an
+InfiniBand network to be programmed as if it were a single, large,
+shared-memory NUMA (non-uniform memory access) machine with scalable
+performance for irregular applications. \Grappa tolerates communication
+latencies with lightweight user-level tasking layer that supports a context
+switch in as little as 38ns and can sustain a large number of active contexts.
+It bridges the commodity network bandwidth gap with a communication layer that
+combines short messages into larger ones.
+
+% In this paper we introduce \Grappa, a software runtime system that allows a
+% commodity cluster of x86-based nodes connected via an InfiniBand network to be
+% programmed as if it were a single, large, shared-memory NUMA (non-uniform
+% memory access) machine with scalable performance for irregular applications.
+% \Grappa tolerates communication latencies with lightweight user-level tasking layer that supports a context switch in as little as 38ns and can sustain a large number of active contexts.
+It bridges the commodity network bandwidth gap with a communication layer that combines short messages into larger ones.
+% 
+% , and it addresses the a \emph{communication\/}
+% layer that combines short messages to achieve peak bandwidth on commodity
+% networks
+% 
+% 
+% distributed global load balancing
+% 
+% 
+% \Grappa leverages as much freely available and commodity infrastructure as
+% possible. We use unmodified Linux for the operating system and an
+% off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}. MPI is
+% used for process setup and tear down. GASNet~\cite{gasnet} is used as the
+% underlying mechanism for remote memory reads and writes using active message
+% invocations.
+% 
+% \Grappa adds three main software components: (1) a \emph{lightweight
+% tasking\/} layer that supports a context switch in as little as 38ns and
+% distributed global load balancing; (2) a \emph{distributed shared memory\/}
+% layer that supports normal access operations such as \emph{read\/} and
+% \emph{write\/} as well as synchronizing operations such as
+% \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{communication\/}
+% layer that combines short messages to achieve peak bandwidth on commodity
+% networks. As we will show later, \Grappa can tolerate latencies way beyond
+% that of the network. Therefore, \Grappa can afford to \emph{trade latency for
+% throughput}. By {\em increasing\/} latency in key components of the system we
+% are able to: increase effective random access memory bandwidth by delaying and
+% aggregating messages; increase synchronization rate by delegating atomic
+% operations to gatekeeper cores, even when referencing node-local global data;
+% and improve load balance by tolerating the delays incurred when work-stealing.
 

### 2013-03-28 doc/sosp13/intro.tex
intro/overview reorg

--- 2013-03-28 doc/sosp13/intro.tex
@@
 
-We believe we can support irregular applications with good performance and cost-effectiveness with commodity hardware for two main reasons.
-First, commodity multicore processors have become extremely fast with high clock rates, large caches and robust DRAM bandwidth.
-So with careful software engineering we can multiplex hundred of thousands of workers into a single core.
-Second, commodity networks offer high bandwidth as long as messages are large enough.
-So with enough data requests in flight simultaneously, the system can aggregate small messages into large enough ones.
+We believe we can support irregular applications with good performance and
+cost-effectiveness with commodity hardware for two main reasons. First,
+commodity multicore processors have become extremely fast with high clock
+rates, large caches and robust DRAM bandwidth. So with careful software
+engineering we can multiplex hundred of thousands of workers into a single
+core. Second, commodity networks offer high bandwidth as long as messages are
+large enough. Therefore with enough data requests in flight simultaneously,
+the system can aggregate small messages into large enough ones.
 
-In this paper we introduce \Grappa, a software runtime system that allows a commodity cluster of x86-based nodes connected via an InfiniBand network to be programmed as if it were a single, large, shared-memory NUMA (non-uniform memory access) machine with scalable performance for irregular applications.
-\Grappa leverages as much freely available and commodity infrastructure as possible.
-We use unmodified Linux for the operating system and an off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}.
-MPI is used for process setup and tear down.
-GASNet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.
-\Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as little as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{communication\/} layer that combines short messages to achieve peak bandwidth on commodity networks.
-As we will show later, \Grappa can tolerate latencies way beyond that of the network.
-Therefore, \Grappa can afford to \emph{trade latency for throughput}.
-By {\em increasing\/} latency in key components of the system we are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance by tolerating the delays incurred when work-stealing.
+We build on these two observations and develop \Grappa, a software runtime
+system that allows a commodity cluster of x86-based nodes connected via an
+InfiniBand network to be programmed as if it were a single, large,
+shared-memory NUMA (non-uniform memory access) machine with scalable
+performance for irregular applications. \Grappa tolerates communication
+latencies with lightweight user-level tasking layer that supports a context
+switch in as little as 38ns and can sustain a large number of active workers.
+It bridges the commodity network bandwidth gap with a communication layer that
+combines short messages originating from a large number of active workers
+into larger ones.
 

### 2013-03-28 doc/sosp13/overview.tex
intro/overview reorg

--- 2013-03-24 doc/sosp13/overview.tex
@@
 \section{\Grappa Overview}
+
+\Grappa leverages as much freely available and commodity infrastructure as
+possible. We use unmodified Linux for the operating system and an
+off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}. MPI is
+used for process setup and tear down. GASNet~\cite{gasnet} is used as the
+underlying mechanism for remote memory reads and writes using active message
+invocations. To this infrastructure, \Grappa adds  three main software components, shown in Figure~\ref{fig:grappa}:
 
@@
 
-\Grappa (Figure~\ref{fig:grappa}) has three main software components:
 \begin{description}
@@
 load balancing. The scheduler oversubscribes to have more worker threads than
-required for latency tolerance. By having at least several threads ready to
-run at all times, the scheduler can prefetch a context data into cache to
-lower the likelihood of costly main memory accesses during a context switch.
+required for latency tolerance. By having at least four workers per processor
+core ready to run at all times, the scheduler can prefetch a worker's state
+into cache to lower the likelihood of costly main memory accesses during a
+context switch.
 
-\item[Distributed shared memory.] The DSM system provides
-fine-grain access to data anywhere in the system. It supports synchronization
-operations on global data, explicit local caching of any memory in the system,
-and support for operations on remote data.
-The DSM system design relies on the lightweight tasking system and
-communication layer in order to offer high aggregate random
-access bandwidth for accessing remote data.
+\item[Distributed shared memory.] The DSM system provides fine-grain access to
+data anywhere in the system. It supports normal access operations such as
+\emph{read\/} and \emph{write\/} as well as synchronizing operations such as
+\emph{fetch-and-add\/}~\cite{fetchandadd}. It also offers explicit local
+caching of any memory in the system, and delegates that directly operate on
+remote data. The memory model offered is similar to what underpins
+C/C++~\cite{N2480,N2800}, so it is familiar to programmers. The DSM system
+design relies on the lightweight tasking system and communication layer in
+order to offer high aggregate random access bandwidth for accessing remote
+data.
 
-\item[Communication layer.] Modern commodity networks
-support high bandwidth only for large messages. Since irregular applications
-tend to need frequent communication of small requests, the main goal of our
-communication layer is to aggregate small messages into large ones to better
-exploit what the network can offer. It is largely invisible to the application
-programmer.
-
-% software developer but helps to improve network performance when applications read and write only small pieces of data.
+\item[Communication layer.] The main goal of our communication layer is to
+aggregate small messages into large ones, which invisible to the application
+programmer. Its interface is based on active messages~\cite{vonEicken92}.
+Since aggregation and deaggregation of messages needs to be very efficient, we
+made it highly concurrent with careful use of lock-free synchronization
+operations.
 
@@
 
-% \paragraph{Exploiting Latency Tolerance} As we will show later, Grappa can tolerate latencies way beyond that of the network.
-Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).
-
-% \TODO{check to see we actually explain these three somewhere.
-Ie, that aggregating increases throughput and latency, that delegating synchronization increases the rate and the latency at which we can eg atomically increment, and that stealing work increases the latency of some tasks (eg when their data is largely at the originating node) but provides greater throughput overall.}
-
-In the next three sections we describe both \Grappa's main components and how they are implemented.
+As we will show throughout the paper, \Grappa can tolerate latencies way beyond
+that of the network. Therefore, \Grappa can afford to \emph{trade latency for
+throughput}. By {\em increasing\/} latency in key components of the system we
+are able to: increase effective random access memory bandwidth by delaying and
+aggregating messages; increase synchronization rate by delegating atomic
+operations to gatekeeper cores, even when referencing node-local global data;
+and improve load balance by tolerating the delays incurred when work-stealing.

### 2013-03-28 doc/sosp13/tasks.tex

@@
 
-The basic unit of execution in \Grappa is a {\em task}. Each hardware core has
-a single operating system thread pinned to it. When tasks are ready to
-execute, they are mapped to a {\em worker}, which is akin to a user-level
-thread. \Grappa keeps a large number of workers per hardware core, this is the
-core reason why \Grappa is able to tolerate latencies well. Below we discuss
-the implementation of our task management support and then describe how
-applications expose parallelism to the \Grappa runtime.
+The basic unit of execution in \Grappa is a {\em task}.
+Each hardware core has a single operating system thread pinned to it.
+When tasks are ready to execute, they are mapped to a {\em worker}, which is akin to a user-level thread.
+\Grappa keeps a large pool of workers per hardware core.
+Below we discuss the implementation of our task management support and then describe how applications expose parallelism to the \Grappa runtime.
 
@@
 
-\paragraph{Tasks}
-Each task is represented by a function pointer and its arguments, represented
-as 32-byte entities: a 64-bit function pointer plus three 64-bit arguments. We
-use three arguments because tasks are often generated via a parallel loop
-decomposition, in which each task needs three kinds of data. Thus, tasks
-include:
-
-\noindent {\it function pointer}, or the addresses of the routine to run ---
-we configure the system nodes to disable address randomization, so that
-function pointers are valid across process images;
+\paragraph{Tasks} Each task is represented by a 64-bit function pointer and three 64-bit arguments.
+We use three arguments because tasks are often generated via a parallel loop decomposition, in which each task needs three kinds of data:
 
@@
 \noindent {\it synchronization argument}, often used to determine when all
-tasks that are part of a loop have finished, in the form of q global pointer
+tasks that are part of a loop have finished, in the form of a global pointer
 to a synchronization object allocated at the core that spawned a group of
@@
 
-\paragraph{Workers}
-When a new task is spawned, it is enqueued until it is ready for execution.
-Tasks are not allocated any execution resources (a worker) until the scheduler
-decides to run them. A worker is simply a collection of status bits and a
-stack, allocated at a particular core. There are two types of workers: (1) user worker; and (2) system worker.
+\paragraph{Workers} Workers execute application tasks.
+A worker is simply a collection of status bits and a stack, allocated at a particular core.
+When a task is ready to be executed it is assigned to a worker.
+Once a task is mapped to a worker it stays with that worker until it completes.
 
-User workers execute application tasks. Once a task is mapped to a worker it
-stays with that worker until it completes. System workers execute runtime
-management activities such as communication and statistics collection. User
-workers and system workers might have different scheduling priorities. System
-workers execute periodically and in idle times, to guarantee timely
-communication services.
+%% Note sure this is even true.  And it's confusing and not relevant regardless. -M
+% There are two types of workers: (1) user worker; and (2) system worker. 
+%System workers execute runtime management activities such as communication and statistics collection.
+User workers and system workers might have different scheduling priorities.
+System workers execute periodically and in idle times, to guarantee timely communication services.
 
@@
 minimize yield overhead, the \Grappa scheduler operates entirely in user-space
-and does little more than store register state of one task and load that of
+and does little more than store register state of one worker and load that of
 another.
@@
 
-Each core in a \Grappa system has its own independent scheduler. Schedulers
-have three main responsibilities: servicing system workers that process
-communication requests as described in Section~\ref{sec:communication};
-rescheduling workers that have been waiting on long-latency operations; and
-assigning ready tasks to worker resources that have become idle.
+Each core in a \Grappa system has its own independent scheduler.  The scheduler has a single queue of active workers:
 
-Each scheduler has four queues:
+\noindent {\it ready worker queue}, a FIFO queue of workers that are
+  matched with workers and are ready to execute;
 
-\noindent {\it ready worker queue}, a FIFO queue of tasks that are
-  matched with workers and are ready to execute;
+Each scheduler also has three queues of tasks waiting to be assigned a worker:
+
+\noindent {\it deadline task queue}, a priority queue of tasks that are executed according to task-specific deadline constraints.
 
@@
   task pool.
-
-\noindent {\it deadline task queue}, a priority queue of tasks that are executed according to task-specific deadline constraints.
 
@@
 
-% Pretty sure this is not true anymore.  Even if it is, it's not important -Mark
-%
-%Since the compiler sees all calls to the context switch routine, we
-%can save even less state. Our context switch routine appears to the
-%compiler as inline assembly; we declare all the registers we need
-%to save as ``clobbered'' by the inline assembly routine, and the
-%compiler will issue its own save and restore code as needed. This allows the
-%compiler to avoid saving any registers that are not used, or are used
-%for temporary values that are not needed after the context switch.
-
 Since \Grappa keeps a very large number of active workers, their context data
@@
 core. Even if there are many active threads, if they are all suspended on
-long-latency operations, then the core is underutilized. We choose to initiate
-one outstanding steal whenever the ready queue is empty there are idle worker
-threads.
+long-latency operations, then the core is underutilized.
 
@@
 small piece of code on a particular core in the system without waiting for
-execution resources to be available. For example, custom synchronization
-primitives execute this way, as a function executed on the core where the data
+execution resources to be available. Custom synchronization primitives, for example, execute this way as a function executed on the core where the data
 lives. \Grappa provides the \texttt{call\_on} call for this purpose.

### 2013-03-28 doc/sosp13/intro.tex
ni

@@
 
-We believe we can support irregular applications with good performance and
-cost-effectiveness with commodity hardware for two main reasons. First,
-commodity multicore processors have become extremely fast with high clock
-rates, large caches and robust DRAM bandwidth. So with careful software
-engineering we can multiplex hundred of thousands of workers into a single
-core. Second, commodity networks offer high bandwidth as long as messages are
-large enough. Therefore with enough data requests in flight simultaneously,
-the system can aggregate small messages into large enough ones.
-
-We build on these two observations and develop \Grappa, a software runtime
-system that allows a commodity cluster of x86-based nodes connected via an
-InfiniBand network to be programmed as if it were a single, large,
-shared-memory NUMA (non-uniform memory access) machine with scalable
-performance for irregular applications. \Grappa tolerates communication
-latencies with lightweight user-level tasking layer that supports a context
-switch in as little as 38ns and can sustain a large number of active workers.
-It bridges the commodity network bandwidth gap with a communication layer that
-combines short messages originating from a large number of active workers
-into larger ones.
+We believe we can support irregular applications with good performance and cost-effectiveness with commodity hardware for two main reasons.
+First, commodity multicore processors have become extremely fast with high clock rates, large caches and robust DRAM bandwidth.
+Second, commodity networks offer high bandwidth as long as messages are large enough.
+We build on these two observations and develop \Grappa, a software runtime system that allows a commodity cluster of x86-based nodes connected via an InfiniBand network to be programmed as if it were a single, large, shared-memory NUMA (non-uniform memory access) machine with scalable performance for irregular applications.
+\Grappa exploits fast processors and the memory hierarchy to provide a lightweight user-level tasking layer that supports a context switch in as little as 38ns and can sustain a large number of active workers.
+It bridges the commodity network bandwidth gap with a communication layer that combines short messages originating from a large number of active workers into larger ones.
 

### 2013-03-28 doc/sosp13/tasks.tex
tasks stuff

@@
 
-The basic unit of execution in \Grappa is a {\em task}.
-Each hardware core has a single operating system thread pinned to it.
-When tasks are ready to execute, they are mapped to a {\em worker}, which is akin to a user-level thread.
-\Grappa keeps a large pool of workers per hardware core.
-Below we discuss the implementation of our task management support and then describe how applications expose parallelism to the \Grappa runtime.
+Below we discuss the implementation of our task management support and then
+describe how applications expose parallelism to the \Grappa runtime.
 
@@
 
-\paragraph{Tasks} Each task is represented by a 64-bit function pointer and three 64-bit arguments.
-We use three arguments because tasks are often generated via a parallel loop decomposition, in which each task needs three kinds of data:
+The basic unit of execution in \Grappa is a {\em task}. Each hardware core has
+a single operating system thread pinned to it. When tasks are ready to
+execute, they are mapped to a {\em worker}, which is akin to a user-level
+thread. % \Grappa keeps a large pool of workers per hardware core.
 
-\noindent {\it private argument}, often a loop index;
+\paragraph{Tasks} Each task is represented by a 64-bit function pointer and
+three 64-bit arguments. We use three arguments because tasks are often
+generated via a parallel loop decomposition, in which each task needs three
+kinds of data:
 
-\noindent {\it shared argument}, typically data shared amongst a group of
+\begin{itemize}
+\item {\it private argument}, often a loop index;
+
+\item {\it shared argument}, typically data shared amongst a group of
 tasks, or the number of loop iterations to be performed by this task;
 
-\noindent {\it synchronization argument}, often used to determine when all
+\item {\it synchronization argument}, often used to determine when all
 tasks that are part of a loop have finished, in the form of a global pointer
@@
 tasks.
+\end{itemize}
 
@@
 
-\paragraph{Workers} Workers execute application tasks.
-A worker is simply a collection of status bits and a stack, allocated at a particular core.
-When a task is ready to be executed it is assigned to a worker.
-Once a task is mapped to a worker it stays with that worker until it completes.
+\paragraph{Workers} Workers execute application and system (e.g.,
+communication) tasks. A worker is simply a collection of status bits and a
+stack, allocated at a particular core. When a task is ready to be executed it
+is assigned to a worker. Once a task is mapped to a worker it stays with that
+worker until it completes.
 
-%% Note sure this is even true.  And it's confusing and not relevant regardless. -M
-% There are two types of workers: (1) user worker; and (2) system worker. 
-%System workers execute runtime management activities such as communication and statistics collection.
-User workers and system workers might have different scheduling priorities.
-System workers execute periodically and in idle times, to guarantee timely communication services.
-
-\paragraph{Scheduling} 
-During execution, a worker yields control of its core whenever it performs a
-long-latency operation, allowing the processor to remain busy while waiting
-for the operation to complete. In addition, a programmer can direct scheduling
-explicitly via the \Grappa API calls shown in Figure~\ref{fig:scheduling}. To
-minimize yield overhead, the \Grappa scheduler operates entirely in user-space
-and does little more than store register state of one worker and load that of
-another.
+\paragraph{Scheduling} During execution, a worker yields control of its core
+whenever it performs a long-latency operation, allowing the processor to
+remain busy while waiting for the operation to complete. In addition, a
+programmer can direct scheduling explicitly via the \Grappa API calls shown in
+Figure~\ref{fig:scheduling}. To minimize yield overhead, the \Grappa scheduler
+operates entirely in user-space and does little more than store state of one
+worker and load that of another.
 
@@
 
-Each core in a \Grappa system has its own independent scheduler.  The scheduler has a single queue of active workers:
+Each core in a \Grappa system has its own independent scheduler. The scheduler
+has a single FIFO queue of active \emph{workers} ready to execute, the {\it
+ready worker queue}. Each scheduler also has three queues of \emph{tasks}
+waiting to be assigned a worker:
 
-\noindent {\it ready worker queue}, a FIFO queue of workers that are
-  matched with workers and are ready to execute;
+\begin{itemize}
 
-Each scheduler also has three queues of tasks waiting to be assigned a worker:
+\item {\it deadline task queue}, a priority queue of tasks that are executed according to task-specific deadline constraints;
 
-\noindent {\it deadline task queue}, a priority queue of tasks that are executed according to task-specific deadline constraints.
+\item {\it private task queue}, a FIFO queue of tasks that must run on
+this core and therefore is not subject to stealing;
 
-\noindent {\it private task queue}, a FIFO queue of tasks that must run on this core;
-
-\noindent {\it public task queue},  a LIFO queue of tasks that are
+\item {\it public task queue},  a LIFO queue of tasks that are
   waiting to be matched with workers. It is a local partition of a shared
   task pool.
+
+\end{itemize}
 
@@
 calls, saving and restoring only the callee-saved state as specified in the
-x86-64 ABI\~cite{amd64:abi:2012}. This involves saving six general-purpose
+x86-64 ABI~\cite{amd64:abi:2012}. This involves saving six general-purpose
 64-bit registers and the stack pointer, as well as the 16-bit x87 floating
@@
 This makes context switching effectively free of cache misses even to hundreds
-of thousands of threads. We provide a detailed analysis of our context switch
-performance in Section~\ref{eval:ctx}.
+of thousands of threads. We provide an analysis of our context switch
+performance in Section~\ref{eval:basic}.
 

### 2013-03-28 doc/sosp13/tasks.tex
made up sample code

--- 2013-03-28 doc/sosp13/tasks.tex
@@
 core in the system or at the home core of a particular memory location. Third,
-the programmer can invoke a parallel for loop, provided that the trip count is
+the programmer can invoke a parallel for loop with \texttt{parallel\_for}, provided that the trip count is
 known at loop entry. The programmer specifies a function pointer along with
@@
 \end{figure}
+
+\begin{figure}[htbp]
+\begin{scriptsize}
+\begin{verbatim}
+
+class node_t {
+  key_t   key
+  int64_t numChildren;
+  GlobalAddress<struct node_t *> *children;
+};
+
+void search(GlobalAddress<node_t> node, key_t key) {
+  if (node.key == key) {
+    result = node;
+    return;
+  }
+
+  parallel_for(0, node.numChildren, &index) {
+    search(node.children[index], key);
+  }
+
+}
+\end{verbatim}
+\end{scriptsize}
+\end{figure}

### 2013-03-28 doc/sosp13/memory.tex

@@
 
-Applications written for \Grappa utilize two forms of memory: local and global.
-Local memory is local to a single core within a node in the system. Accesses
-occur through conventional pointers. The compiler emits an access and the
-memory is manipulated directly. Applications use local accesses for a number
-of things in \Grappa: the stack associated with a task, accesses to localized
-global memory in caches (see below), and accesses to debugging infrastructure
-that is local to each system node. Local pointers cannot access memory on
-other cores, and are valid only on their home core.
+Applications written for \Grappa utilize two forms of memory: local and
+global. Local memory is local to a single core within a node in the system.
+Accesses occur through conventional pointers. The compiler emits an access and
+the memory is manipulated directly. Applications use local accesses for a
+number of things in \Grappa: the stack associated with a task, accesses to
+localized global memory in caches (see below), and accesses to debugging
+infrastructure that is local to each system node. Local pointers cannot access
+memory on other cores, and are valid only on their home core.
 
@@
 and deallocate memory in the global heap. Addresses to memory in the global
-heap use \textbf{linear addresses}. Choosing the block size involves trading
-off sequential bandwidth against aggregate random access bandwidth. Smaller
-block sizes help spread data across all the memory controllers in the cluster,
-but larger block sizes allow the locality-optimized memory controllers to
-provide increased sequential bandwidth. The block size, which is configurable,
-is typically set to 64 bytes, or the size of a single hardware cache line, in
+heap use \emph{linear addresses}. Choosing the block size involves trading off
+sequential bandwidth against aggregate random access bandwidth. Smaller block
+sizes help spread data across all the memory controllers in the cluster, but
+larger block sizes allow the locality-optimized memory controllers to provide
+increased sequential bandwidth. The block size, which is configurable, is
+typically set to 64 bytes, or the size of a single hardware cache line, in
 order to exploit spatial locality when available.
 
-%%% is the below important? -Luis
-\comment{The heap metadata is stored on a single node. Currently all heap
-allocation and free operations are serialized through this node; while this
-has been sufficient for our benchmarks, it is easy to relax in the future
-\Grappa will provide parallel performance through
-combining~\cite{MAMA,flatcombining}.}
-
-\Grappa also allows any local data on a core's stacks or heap to be
-exported to the global address space to be made accessible to other
-cores across the system. Addresses to global memory allocated in this
-way use \textbf{2D global addresses}.  This uses a traditional PGAS
-addressing model, where each address is a tuple of a rank in the job (or
-global process ID) and an address in that process. The lower 48 bits of
-the address hold a virtual address in the process. The top bit is set to
-indicates that the reference is a 2D address (as opposed to linear
-address). This leaves 15 bits for network endpoint ID, which limits our
-scalability to $2^{15}$ endpoints. Any node-local data can be made
-accessible by other nodes in the system by wrapping the address and node
-ID into a 2D global address. This address can then be accessed with a
-delegate operation and even be cached by other nodes. The
-address is converted at the destination into a canonical x86 address by replacing the upper
-bits with the sign-extended upper bit of the virtual address. 2D
-addresses may refer to memory allocated from a single processes' heap or
-from a task's stack. Figure~\ref{fig:memory-structure} shows how 2D and
-linear addresses can refer to other cores' memory.
+\Grappa also allows any local data on a core's stacks or heap to be exported
+to the global address space to be made accessible to other cores across the
+system. Addresses to global memory allocated in this way use \emph{2D global
+addresses}. This uses a traditional PGAS (partitioned global address
+space~\cite{upc:2005}) addressing model, where each address is a tuple of a
+rank in the job (or global process ID) and an address in that process. The
+lower 48 bits of the address hold a virtual address in the process. The top
+bit is set to indicates that the reference is a 2D address (as opposed to
+linear address). This leaves 15 bits for network endpoint ID, which limits our
+scalability to $2^{15}$ endpoints. Any node-local data can be made accessible
+by other nodes in the system by wrapping the address and node ID into a 2D
+global address. This address can then be accessed with a delegate operation
+and even be cached by other nodes. The address is converted at the destination
+into a canonical x86 address by replacing the upper bits with the
+sign-extended upper bit of the virtual address. 2D addresses may refer to
+memory allocated from a single processes' heap or from a task's stack.
+Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can refer
+to other cores' memory.
 
@@
       	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\
+      	\texttt{\scriptsize delegate\_fetch\_inc\_async( global\_address, local\_var )} \\ 
       	Memory operations at the home core of a global address \\ \\
@@
 	\end{tabular}
-      \caption{\label{fig:accessing-memory} \Grappa API for memory accesses. \TODO{Update table with new calls. And async calls for delegates.}}     \end{minipage}
+      \caption{\label{fig:accessing-memory} \Grappa API for memory accesses.}     \end{minipage}
   \end{center}

### 2013-03-28 doc/sosp13/tasks.tex
mr

@@
 
-void search(GlobalAddress<node_t> node, key_t key) {
-  if (node.key == key) {
+void search(GlobalAddress<node_t *> node, key_t key) {
+  if (node->key == key) {
     result = node;
@@
 
-  parallel_for(0, node.numChildren, &index) {
-    search(node.children[index], key);
+  parallel_for(0, node->numChildren, &index) {
+    search(node->children[index], key);
   }

### 2013-03-28 doc/sosp13/tasks.tex
Fix up the code and API, remove spawn_on.

@@
 	\begin{tabular}{l}
-    \texttt{\scriptsize spawn( void (*fp)(args) )} \\
+    \texttt{\scriptsize spawn( Functor f )} \\
       Creates a new stealable task \\
-    \texttt{\scriptsize spawn\_on( core, (*fp)(args) )} \\
-      Creates a new task that will run on a specific core \\
-    \texttt{\scriptsize parallel\_for( (*fp)(args), start, end )} \\
-      Executes iterations of a loop as stealable tasks  \\
-    \texttt{\scriptsize call\_on( core, (*fp)(args) )} \\ 
+    \texttt{\scriptsize parallel\_for( start, end, Functor iteration )} \\
+      Executes iterations of a loop as stealable tasks that take the iteration index as an argument  \\
+    \texttt{\scriptsize call\_on( core, Functor f )} \\ 
       Runs a limited function on a specific core without \\
@@
 \begin{verbatim}
-
 class node_t {
@@
   int64_t numChildren;
-  GlobalAddress<struct node_t *> *children;
+  GlobalAddress<node_t *> children;
 };
@@
 
-  parallel_for(0, node->numChildren, &index) {
+  parallel_for(0, node->numChildren, [=](int index) {
+    spawnTask([=]{
       search(node->children[index], key);
-  }
-
+    });
+  });
 }

### 2013-03-28 doc/sosp13/tasks.tex
t

@@
 and queued with its arguments for later execution using a \texttt{spawn}.
-Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific
-core in the system or at the home core of a particular memory location. Third,
+Second,
 the programmer can invoke a parallel for loop with \texttt{parallel\_for}, provided that the trip count is
@@
 tasks, stopping to execute the loop body when the number of iterations is
-below the required threshold. Fourth, a programmer may want to execute an active message; that is, to run a
+below the required threshold. Third, a programmer may want to execute an active message; that is, to run a
 small piece of code on a particular core in the system without waiting for
@@
     \texttt{\scriptsize parallel\_for( start, end, Functor iteration )} \\
-      Executes iterations of a loop as stealable tasks that take the iteration index as an argument  \\
+      Executes iterations of a loop as stealable tasks that \\
+      take the iteration index as an argument  \\
     \texttt{\scriptsize call\_on( core, Functor f )} \\ 
@@
       \caption{\label{fig:expressing-parallelism} \Grappa API: expressing parallelism
-      \TODO{update this with latest C++ code from ``api\_examples.cpp''}
       } % \vspace{-4ex}}
@@
 \begin{figure}[htbp]
+\begin{center}
 \begin{scriptsize}
@@
   }
-
   parallel_for(0, node->numChildren, [=](int index) {
@@
 \end{scriptsize}
+
+    \begin{minipage}{0.95\columnwidth}
+      \caption{\label{fig:sample} Sample \Grappa code illustrating a parallel tree search similar to the unbalanced tree search benchmark we describe later.}
+    \end{minipage}
+
+\end{center}
 \end{figure}

### 2013-03-28 doc/sosp13/method.tex
started int sort paragraph; added a few citations

@@
 \vspace{0.5ex}
-\noindent{\bf IntSort:} \TODO{Write me Brandon Holt!}
+\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS
+Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on
+which the CrayXMT's early predecessor once held the world speed
+record~\cite{TeraRecord}. The class D benchmark ranks a billion
+uniformly distributed random integers using either a bucket or a counting sort
+algorithm: both versions are built into their reference code with the
+choice left to the benchmarker, the difference being the number and iteration
+counts of a series of tight loop computations.  Bucket sort uses less space
+than counting sort but executes a greater number of loops.  We chose
+Bucket sort for Grappa; the Cray XMT benchmark runs faster when using
+counting sort,so our comparison is based on those performance numbers.
 

### 2013-03-28 doc/sosp13/intro.tex

@@
 
+As a general design philosophy, \Grappa trades latency for throughput.
+By \emph{increasing} latency in key components of the system we are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance by tolerating the delays incurred when work-stealing.
+\Grappa then exploits parallelism to overcome latency.
+
 Our evaluation of \Grappa shows that its core components provide scalable performance.

### 2013-03-28 doc/sosp13/overview.tex

@@
 \end{description}
-
-As we will show throughout the paper, \Grappa can tolerate latencies way beyond
-that of the network. Therefore, \Grappa can afford to \emph{trade latency for
-throughput}. By {\em increasing\/} latency in key components of the system we
-are able to: increase effective random access memory bandwidth by delaying and
-aggregating messages; increase synchronization rate by delegating atomic
-operations to gatekeeper cores, even when referencing node-local global data;
-and improve load balance by tolerating the delays incurred when work-stealing.

### 2013-03-28 doc/sosp13/method.tex
pagerank

--- 2013-03-28 doc/sosp13/method.tex
@@
 \vspace{0.5ex}
-\noindent{\bf PageRank:} \TODO{Write me Brandon Myers!}
+\noindent{\bf PageRank:} This is a common centrality metric for
+graphs. Computing PageRank is an iterative algorithm with a common
+pattern of gather, apply, and scatter on the rank of vertex. The algorithm
+is often implemented by sparse linear algebra libraries, with the main
+kernel being the sparse matrix dense vector multiply. For the multiply
+step, \Grappa parallelizes over the rows and parallelizes each dot
+product. PageRank has the fortunate property that the accumulation
+function over the in-edges is associative and commutative, so they can
+be processed in any order in parallel. Rather than the programmer writing the
+parallel dot product as local accumulations with a final all-reduce
+step, we simply send streaming increments to each element of the final
+vector.
 

### 2013-03-28 doc/sosp13/method.tex

--- 2013-03-28 doc/sosp13/method.tex
@@
 \vspace{0.5ex}
-\noindent{\bf IntSort:} \TODO{Write me Brandon Holt!}
+\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS
+Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on
+which the CrayXMT's early predecessor once held the world speed
+record~\cite{TeraRecord}. The largest problem size, class ``D'', ranks two billion
+uniformly distributed random integers using either a bucket or a counting sort
+algorithm, depending on the strengths of the system.
+Bucket sort executes a greater number of loops, but is able to leverage locality and avoid communication completely in the final phase, ranking within buckets.
+For these reasons, the MPI reference version and our \Grappa implementation use bucket sort.
+On the other hand, the Cray XMT cannot take advantage of locality, but has an efficient compiler-supported parallel prefix sum, so it performs best using the counting sort algorithm.
 

### 2013-03-28 doc/sosp13/evaluation.tex
added gups plot
Inserted UTS figure 	modified:   doc/sosp13/evaluation.tex
bunch of little things
uts eval

@@
-\section{Evaluation} \label{sec:evaluation}
+\section{Evaluation}
+\label{sec:evaluation}
 
@@
 applications. We evaluate \Grappa in three basic steps:
+
 \begin{itemize}
-\item We present results
-that show that the \Grappa runtime is able to sustain very high concurrency
-rates and the communication layer is able to sustain a very high rate of
-global memory operations. We also show the performance of a graph kernel that
-stresses communication and concurrency together.
 
-\item We show how some popular irregular applications running on \Grappa compare to the Cray XMT and hand-tuned MPI code.
+\item We present results that show that the \Grappa runtime is able to sustain
+very high concurrency rates and the communication layer is able to sustain a
+very high rate of global memory operations. We also show the performance of a
+graph kernel that stresses communication and concurrency together.
 
-\item We finish with a characterization of system behavior, including profiling where execution time goes, how aggregation affects message size and rates, how global memory and work stealing behaves.
+\item We show how some popular irregular applications running on \Grappa
+compare to the Cray XMT and hand-tuned MPI code.
+
+\item We finish with a characterization of system behavior, including
+profiling where execution time goes, how aggregation affects message size and
+rates, how global memory and work stealing behaves.
 
@@
 \subsection{Basic \Grappa Mechanisms}
+\label{eval:basic}
 
 \paragraph{User-level context switching.}
+
 As discussed earlier, fast context switching is at the heart of \Grappa's
@@
     \begin{center}
-%      \includegraphics[width=0.5\textwidth]{figs/grappa-gups.pdf}
+      \includegraphics[width=0.5\textwidth]{results/gups/gups.pdf}
     \end{center}
@@
 
-\paragraph{Putting it all together with Unbalanced Tree Search (UTS).}
+\paragraph{Putting it all together with Unbalanced Tree Search in
+    memory (UTS).}
 We now show the overall performance of \Grappa running UTS.
-Figure~\ref{fig:grappa-uts}.
-The point of this experiment is to show that \Grappa's context switching and communication layers can in fact be used together to run an irregular application efficiently.
-Looking at the performance as the number of nodes grows shows \TODO{...}.
+Figure~\ref{fig:grappa-uts}. The point of this experiment is to
+demonstrate whether \Grappa's context switching and communication layers can in fact
+be used together, while balancing workload, to run an irregular application efficiently. 
+We look at two classes of trees, T1x and T3x, from the
+original benchmark. T1x trees are very shallow and wide, while T3x
+trees are very deep. When the access to each vertex is a random
+access, the critical path to search T3x trees is very long. On such
+trees, we do not expect there to be sufficient concurrency for any
+system, including \Grappa, to achieve high throughput. \TODO{plot it
+    if time}. To verify this, at the 16-node data point, the average active tasks per
+core over the search is 775 and 13 for T1XL and T3L, respectively.
 
@@
     \begin{center}
-%      \includegraphics[width=0.5\textwidth]{figs/grappa-uts.pdf}
+      \includegraphics[width=0.5\textwidth]{figs/uts_scale.pdf}
     \end{center}
@@
 \end{figure}
-\TODO{UTS: story here is T1 class is bushy trees similar to properties
-    of real world networks. Larger T1XXL (4.5 B vertices) provides
-    more parallelism for continued scaling up. T3 class is deep skinny
-    trees. Grappa can do no better than the straightforward UPC impl: there very little
-    parallelism and the critical path is long}
 
@@
 \end{figure}
+
+\TODO{PAGERANK:
+    Without parallelizing the dot product, \Grappa cannot achieve
+high utilization of workers proportional to the size of the data set.
+Also mention that the cost of this is contention at the target
+elements of vector.  The other parts of pagerank are super fast
+(table?)
+}
 

### 2013-03-28 doc/sosp13/evaluation.tex
commit to merge

--- 2013-03-28 doc/sosp13/evaluation.tex
@@
       \includegraphics[width=0.5\textwidth]{figs/context_switch_time.pdf}
-%      \includegraphics[width=0.5\textwidth]{figs/context_switch_bw.pdf}
     \end{center}
-    \caption{Average context switch time with 1 active, with and without prefetching.}
+    \caption{Average context switch time with and without prefetching.}
     \label{fig:context-switch-exp}
@@
 Figure~\ref{fig:context-switch-exp} shows the average context switch time as
-the number of workers grow. There are several important takeaway points form
-these data. Context switch time for the number of workers we use currently
-($~$1K) is on the order of 50 ns; and prefetching has significant impact on
-context switch time. And ontext switch time does not grow significantly with
-the number of workers --- even at close to one half million workers context
-switch is around 75ns.
+the number of workers grow. There are two important takeaway points from these
+data. Context switch time for the number of workers we use ($~$1K workers) is
+on the order of 50 ns; and prefetching has significant impact on context
+switch time. And context switch time does not grow significantly with the
+number of workers --- even with one-half million workers, context switch
+is around 75ns.
 
@@
 
-In summary, our context switch engine is able to efficiently sustain very high concurrency and we will show later, the amount of concurrency sustained is sufficient the latencies \Grappa needs to hide.
+In summary, our context switch engine is able to efficiently sustain very high
+concurrency and as we will show later, the amount of concurrency sustained is
+sufficient the latencies \Grappa needs to hide.
 
@@
 scheduler, because only a single worker is used per system node.
-Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain on the order of billions of updates per second, which is enough to \TODO{...}.
+Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain well over a
+billion updates per second with 64 nodes. This compares very favorably to
+published results~\footnote{\url{http://icl.cs.utk.edu/hpcc/hpcc_results.cgi}}
+for other high-end HPC systems.
 
@@
 
-\paragraph{Putting it all together with Unbalanced Tree Search (UTS).}
-We now show the overall performance of \Grappa running UTS.
-Figure~\ref{fig:grappa-uts}.
-The point of this experiment is to show that \Grappa's context switching and communication layers can in fact be used together to run an irregular application efficiently.
-Looking at the performance as the number of nodes grows shows \TODO{...}.
+\paragraph{Putting it all together with Unbalanced Tree Search (UTS).} We now
+show the overall performance of \Grappa running UTS.
+Figure~\ref{fig:grappa-uts}. The point of this experiment is to show that
+\Grappa's context switching and communication layers can in fact be used
+together to run an irregular application efficiently. Looking at the
+performance as the number of nodes grows shows \TODO{...}.
 
@@
 \end{figure}
+
 \TODO{UTS: story here is T1 class is bushy trees similar to properties
@@
 
-In order to ground \Grappa's performance, we compare it with XMT running BFS,
-PageRank, IntSort, GUPS and UTS. Since XMT is a different hardware platform,
-we also compare \Grappa with hand-tuned MPI versions of BFS and GUPS running
-on the same hardware. Finally, we also compare it with UTS written for UPC. We
-run all experiments with 64 nodes.
+In order to put \Grappa's performance into a general context, we compare it
+with XMT running BFS, PageRank, IntSort, GUPS and UTS. Since XMT is a
+different hardware platform, we also compare \Grappa with hand-tuned MPI
+versions of BFS and GUPS running on the same hardware. Finally, we also
+compare it with UTS written for UPC. We run all experiments with 64 nodes.
 

### 2013-03-28 doc/sosp13/method.tex
method: Clean up BFS and add IntSort metric.

@@
 second\/} (TEPS), where the number of edges is the edges making up the
-generated BFS tree. One of the reference implementations of Graph500 BFS is
-for the XMT; this code fails to scale past 16 XMT processors because it does
-not expose enough parallelism, so we modified the code to use a recursive loop
-decomposition similar to \Grappa's. We compare this modified version against a
-straightforward \Grappa implementation. We do not employ algorithmic
-improvements, though there are
-many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}.
+generated BFS tree. With some modifications to the XMT reference version of Graph500 BFS,
+the XMT compiler can be made to recognize and apply a Manhattan loop collapse, exposing enough parallelism to allow it to scale out to 64 nodes for the problem scales we show.
+We do not employ algorithmic improvements for any of these versions, though there are
+many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}. \Grappa can be expected to benefit the same as MPI due to decreased communication.
 
@@
 On the other hand, the Cray XMT cannot take advantage of locality, but has an efficient compiler-supported parallel prefix sum, so it performs best using the counting sort algorithm.
+
+The performance metrics for NAS Parallel Benchmarks, including IntSort, are ``millions of operations per second'' (MOPS).
+For IntSort, this ``operation'' is ranking a single key, so it is roughly comparable to ``GUPS'' or ``TEPS''.
 

### 2013-03-28 doc/sosp13/communication.tex
m

@@
 
-% In order to mitigate the low message injection rate limits of commodity
-% networks, Grappa's communication stack has two layers:
-
 \paragraph{Active messages interface} At the upper (user-level) layer, \Grappa
@@
 data payload.
-
-% When a task sends a message, the message is linked to a send queue
-% associated with the message's destination node and the task continues
-% execution. Note that the message contents stay in the worker's stack until
-% it is sent to the network interface buffer, to avoid multiple copies of the
-% payload.
 
@@
 achieve that, \Grappa makes careful use of caches, prefetching and lock-free
-synchronization operations. \TODO{Does the explanation below need a figure?}
+synchronization operations.
 
-Outgoing messages are organized as an array of linked lists, with one linked
-list per destination node in the system. Each core in a given node is
-responsible for aggregating and sending the resulting messages to a set of
-destination nodes. The outgoing message lists are located in a region of
-memory shared across all cores in a \Grappa node (thus enabling cores to peek
-at each other's message lists). When a task sends a message, it allocates a buffer (typically on its stack),  determines the destination system
-node, and links the buffer into the corresponding linked list.
+Each processing core of a system node maintains an array of outgoing message lists.
+The array size is the number of system nodes in the \Grappa system.
+The outgoing message lists and messages are located in a region of memory shared across all cores in a \Grappa node (thus enabling cores to peek at each other's message lists).
+When a task sends a message, it allocates a buffer (typically on its stack), determines the destination system node, and links the buffer into the corresponding linked list.
 
-Each core has a system worker per destination node it is responsible for; this
-system worker periodically checks whether messages to the corresponding
-destination node is either large enough or has waited past a time-out period,
-at which point the message is sent out. Aggregating and sending a message
-involves manipulating a shared data-structure (the message list). This is done
-using CAS (compare-and-swap) operations to avoid high synchronization costs.
-
-% Threads that dispatch a message allocate a message buffer (typically on their
-% stack) and dispatch a send request. \Grappa examines the destination system
-% node for this field, and adds the message to a linked list of messages
-% destined for that node. Periodically, or when this linked list grows large,
-% the list is transferred using a CAS (compare-and-swap} operation to the
-% processor on the originating system node that is responsible for managing all
-% outbound communication requests to the specified destination node. This
-% two-step process is used because it was determined empirically that we could
-% only infrequently use processor coherent access requests because they remain a
-% bottleneck on modern multicore hardware.
+Each processing core in a given system node is responsible for aggregating and sending the resulting messages from all processors on that node to a set of destination nodes.
+Each core periodically executes a task responsible for sending messages.
+This task examines the private (to each core) message lists for each destination node it is responsible for managing and if the list is long enough, or a message has waited past a time-out period, all messages to a given destination system node from that source system node are sent.
+Aggregating and sending a message involves manipulating a set of shared data-structures (the message lists).
+This is done using CAS (compare-and-swap) operations to avoid high synchronization costs.
+Note that the reason we use a per-processor array of message lists that is only periodically modified across processor cores is we empirically determined this approach was faster (sometimes significantly) than a global per-system node array of message lists.
 
@@
 
-Once the remote system node has received the message buffer, it is unpacked
-and put in the corresponding incoming message lists. This process is done in
-parallel by all system worker responsible for communication. Upper level
-\Grappa messages are delivered to each core at the destination system node by
-examining the message buffer at each processor core and then handing the
-message buffer to the next core for delivery of messages destined to that
-core.
+Once the remote system node has received the message buffer, a management task is spawned to manage the unpacking process.
+The management task spawns a task on each processor at the receiving system to simultaneously unpack messages destined for that processor.
+Upon completion, these unpacking tasks synchronize with the management task.
+Once all cores have processed the message buffer the management task sends a reply to the sending system node indicating the successful delivery of the messages.
 

### 2013-03-28 doc/sosp13/evaluation.tex

@@
 
-\paragraph{Putting it all together with Unbalanced Tree Search (UTS).} We now
-show the overall performance of \Grappa running UTS.
-Figure~\ref{fig:grappa-uts}. The point of this experiment is to show that
-\Grappa's context switching and communication layers can in fact be used
-together to run an irregular application efficiently. Looking at the
-performance as the number of nodes grows shows \TODO{...}.
+\paragraph{Putting it all together with Unbalanced Tree Search in
+    memory (UTS).}
+We now show the overall performance of \Grappa running UTS.
+Figure~\ref{fig:grappa-uts}. The point of this experiment is to
+demonstrate whether \Grappa's context switching and communication layers can in fact
+be used together, while balancing workload, to run an irregular application efficiently. 
+We look at two classes of trees, T1x and T3x, from the
+original benchmark. T1x trees are very shallow and wide, while T3x
+trees are very deep. When the access to each vertex is a random
+access, the critical path to search T3x trees is very long. On such
+trees, we do not expect there to be sufficient concurrency for any
+system, including \Grappa, to achieve high throughput. \TODO{plot it
+    if time}. To verify this, at the 16-node data point, the average active tasks per
+core over the search is 775 and 13 for T1XL and T3L, respectively.
 
@@
 \end{figure}
-
-\TODO{UTS: story here is T1 class is bushy trees similar to properties
-    of real world networks. Larger T1XXL (4.5 B vertices) provides
-    more parallelism for continued scaling up. T3 class is deep skinny
-    trees. Grappa can do no better than the straightforward UPC impl: there very little
-    parallelism and the critical path is long}
 

### 2013-03-28 doc/sosp13/intro.tex

@@
 The communication layer provides over 10X the small-message performance compared to the raw underlying message injection rate.
-Combining these two components, \Grappa achieves 1.2 GUPS with 64 system nodes, far above any reported cluster performance~\cite{gups} for comprable systems.
+Combining these two components, \Grappa achieves 1.3 GUPS with 64 system nodes, far above any reported cluster performance~\cite{gups} for comprable systems.
 When we compare \Grappa to native MPI implementations of more complex application kernels, the story is more muddled.
@@
 On the other hand, if the MPI version is implemented in a straightforward manner, as we see with PageRank, GUPS, and unbalanced tree search, \Grappa is 2-5X faster.
-When we compare \Grappa to the custom Cray XMT, we find Grappa to be between 2X faster and 4X slower depending on the benchmark.
+When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2X faster and 4X slower depending on the benchmark.
 Overall we find this encouraging because \Grappa is scalable and achieves this performance from a mass-market cluster, hence adding more system nodes an inexpensive and attractive option.

### 2013-03-28 doc/sosp13/evaluation.tex
g
pe

@@
 billion updates per second with 64 nodes. This compares very favorably to
-published results~\footnote{\url{http://icl.cs.utk.edu/hpcc/hpcc_results.cgi}}
-for other high-end HPC systems.
+published results~\cite{gups} for other high-end HPC systems.
 
@@
       \begin{tabular}{c|c c c c}
-        Benchmark     & Communication & User & Idle & Context Switch \\ \hline
+        Benchmark     & Comm & User & Idle & Sched \\ \hline
         GUPS          & 6.60  & 42.94   & 47.74 & 2.80 \\

### 2013-03-28 doc/sosp13/evaluation.tex
pe2

@@
 the number of workers grow. There are two important takeaway points from these
-data. Context switch time for the number of workers we use ($~$1K workers) is
+data. Context switch time for the number of workers we use ($\approx$1K workers) is
 on the order of 50 ns; and prefetching has significant impact on context

### 2013-03-28 doc/sosp13/method.tex

@@
 \vspace{0.5ex}
-\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS
-Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on
-which the CrayXMT's early predecessor once held the world speed
-record~\cite{TeraRecord}. The largest problem size, class ``D'', ranks two billion
-uniformly distributed random integers using either a bucket or a counting sort
-algorithm, depending on the strengths of the system.
+\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the CrayXMT's early predecessor once held the world speed record~\cite{TeraRecord}.
+The largest problem size, class ``D'', ranks two billion uniformly distributed random integers using either a bucket or a counting sort algorithm, depending on the strengths of the system.
 Bucket sort executes a greater number of loops, but is able to leverage locality and avoid communication completely in the final phase, ranking within buckets.
@@
 On the other hand, the Cray XMT cannot take advantage of locality, but has an efficient compiler-supported parallel prefix sum, so it performs best using the counting sort algorithm.
-
 The performance metrics for NAS Parallel Benchmarks, including IntSort, are ``millions of operations per second'' (MOPS).
@@
 used for the bulk of the presented results.
-
-%%%% Bit bucket
-
-%\paragraph{Betweenness Centrality} An important measure of the importance of particular vertices in a network is betweenness centrality (BC)~\cite{freeman1979centrality}.
-By this measure, the ``importance'' of each vertex is computed by finding the fraction of shortest paths that pass through it, which can optionally be approximated by only computing shortest paths using a subset of the vertices as starting points.
-BC can be useful for understanding which vertices may have the greatest impact, so in social networks this could be the primary person linking two communities.
-Because it requires multiple breadth-first traversals across the entire graph and in reverse, on power-law degree graphs, this algorithm exercises random accesses rate, and load balancing.
-It also requires fine-grained synchronization on updates to vertex centrality values because multiple paths will update the same vertex.
-BC is a kernel in the DARPA High Performance Computing Systems (HPCS) Scalable Synthetic Compact Applications graph analysis (SSCA\#2) benchmark~\cite{ssca2}.
-Performance for BC is also measured in TEPS, where the number of traversed edges is the total number of edges in the graph multiplied by the number of random starting vertices used in the approximate computation.
-We use the XMT implementation of BC implemented as part of the GraphCT~\cite{GraphCT} library and a comparable \Grappa implementation, both of which use the parallel BC algorithm developed by Bader et al.~\cite{bader:bc}.
-
-%To explore the performance of the \Grappa runtime we have implemented three algorithms: breadth first search, betweenness centrality, and unbalanced tree search.
-These algorithms were implemented for the Cray XMT (our baseline) and for \Grappa.
-The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.
-Data is collected on real systems, which means minor variations in runtime exist from run to run.
-The average of multiple runs are used, and where appropriate, confidence in the result is reported.
-
-\comment{The \Grappa version of tree search UTS-mem
-
-\lstset{language=C++,
-       basicstyle=\footnotesize,
-       tabsize=2}
-\begin{lstlisting}
-// base of shared vertex_t[]
-GlobalAddress<vertex_t> Vertices;
-
-// base of shared int64_t[]
-GlobalAddress<int64_t> ChildrenPointers;
-
-void search_vertex( id ) {
-  GlobalAddress<vertex_t> v_addr = Vertices + id;
-  Incoherent<vertex_t>::RO v( v_add, 1 );
-
-  // start index of my children pointers
-  childIndex = v->childIndex;
-
-  // how many children I have
-  numChildren = v->numChildren;
-
-  // parallel loop over the child list for this vertex
-  parallel_for(fn=&search_children, start=childIndex, iters=numChildren);
-}
-
-void search_children( start, iters ) {
-  // take advantage of spatial locality in the array of children
-  GlobalAddress<int64_t> child_base_addr = ChildrenPointers + start;
-  Incoherent<int64_t>::RO childIds( child_base_addr, iters );
-
-  // spawn a task to visit each child
-  for (i = 0..iters) {
-    SoftXMT_publicTask( fn=&search_vertex, id=childIds[i] );
-  }
-}
-\end{lstlisting}
-}
-
-%% Given that we have MPI performance on the cluster this is not
-%% that important to have here anymore.
-%%
-%%
-%One question that must be answered when making a comparison between the Cray XMT and a typical HPC cluster is what is a fair comparison -- these systems are quite different.
-Three options immediately come to mind: equal number of processing cores, equal number of network interfaces, and equal dollars.
-We discounted the last option fairly quickly, because it isn't a lasting data-point -- the cost of hardware shifts over time, skewing the interpretation of results.
-The first option, cores, has some merit, but \Grappa is designed for applications that have no locality in their computation.
-This means almost all of their memory accesses are remote.
-The factor that limits their performance is not processing, but communicating.
-Hence, we have chosen the middle option, network interfaces as the way to normalize across the XMT and our cluster.
-Each processor in the XMT system has its own network interface to access shared memory.
-In the HPC cluster we use, each processor (which contains up to 32~cores, although we only use 6 in our experiments), has a single infiniband interface.
-Hence, for our results we scale up XMT processors one for one with full system nodes.
-
-%Towards evaluating how well \Grappa enables scalable performance of irregular applications on mass market systems, our methodology is as follows.
-For consistency in measurement, all results reported for mass market systems are measured on the same platform.
-To establish the potential for \Grappa performance, we identify small microkernel benchmarks that provide insight into the limitations of our system's component performance behaviors.
-To compare meaningfully with the state-of-the-art, we compare \Grappa performance on well known irregular application benchmarks both against the popular high performance programming methodologies MPI and UPC for distributed memory clusters, and against a custom shared memory architecture used exclusively for irregular computation, the Cray XMT.
-We then examine how sensitive \Grappa results are to reduced injection rate and increased latency, as often occur when a system is scaled up so that nodes are further apart and the effective per destination packet injection rate drops.
-The three sections that follow describe this methodology in greater detail.

### 2013-03-28 doc/sosp13/tasks.tex
Update stuff about how we specify tasks (with lambdas).

@@
 
-\paragraph{Tasks} Each task is represented by a 64-bit function pointer and
-three 64-bit arguments. We use three arguments because tasks are often
-generated via a parallel loop decomposition, in which each task needs three
-kinds of data:
-
-\begin{itemize}
-\item {\it private argument}, often a loop index;
-
-\item {\it shared argument}, typically data shared amongst a group of
-tasks, or the number of loop iterations to be performed by this task;
-
-\item {\it synchronization argument}, often used to determine when all
-tasks that are part of a loop have finished, in the form of a global pointer
-to a synchronization object allocated at the core that spawned a group of
-tasks.
-\end{itemize}
-
-While these are the most common uses of the three task arguments, they are
-treated as arbitrary 64-bit values in the runtime, and can be used for any
-purpose.
+\paragraph{Tasks} 
+Tasks are specified with a ``functor'' object that holds both code to execute and initial state.
+The functor can be specified with a function pointer and explicit arguments, a C++ struct that overloads the parentheses operator, or a C++11 lambda.
+These objects, typically very small (on the order of 64 bytes), hold read-only values such as an iteration index and pointers to common data or synchronization objects.
+These task functors can be serialized and transported around the system, and eventually executed by a Worker, as described next.
 
@@
 stack, allocated at a particular core. When a task is ready to be executed it
-is assigned to a worker. Once a task is mapped to a worker it stays with that
-worker until it completes.
+is assigned to a worker, which executes the task functor on its own stack. 
+Once a task is mapped to a worker it stays with that worker until it finishes.
 

### 2013-03-28 doc/sosp13/evaluation.tex
edits
add histograms
ev

@@
 concurrency and as we will show later, the amount of concurrency sustained is
-sufficient the latencies \Grappa needs to hide.
+sufficient for the latencies \Grappa needs to hide.
 
@@
 Distribution if possible. With and without aggregation.
+
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{results/histograms/latency_cmb.pdf}
+    \end{center}
+    \caption{Round-trip latency of delegate operations.}
+%bar chart, one set of bars per benchmark, one bar per system. runtime normalized to grappa.
+    \label{fig:grappa-latency}
+\end{figure}
+
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{results/histograms/rdma_bytes_sent_histogram_cmb.pdf}
+    \end{center}
+    \caption{Distribution of aggregated message sizes.}
+%bar chart, one set of bars per benchmark, one bar per system. runtime normalized to grappa.
+    \label{fig:grappa-message-size}
+\end{figure}
 

### 2013-03-28 doc/sosp13/tasks.tex
reference to sample code
ct

@@
 These objects, typically very small (on the order of 64 bytes), hold read-only values such as an iteration index and pointers to common data or synchronization objects.
-These task functors can be serialized and transported around the system, and eventually executed by a Worker, as described next.
+These task functors can be serialized and transported around the system, and eventually executed by a worker, as described next.
 
@@
 
+Figure~\ref{fig:sample} shows sample code using \Grappa for a parallel tree
+search. The important aspect to note is that the code looks very similar to
+what would be written for single shared-memory system, without any concern about data locality or communication.
+
 \begin{figure}[htbp]

### 2013-03-28 doc/sosp13/evaluation.tex
Write a bit about why GUPS is interesting.

@@
 
-\subsection{Basic \Grappa Mechanisms}
+\subsection{Basic \Grappa Performance}
 \label{eval:basic}
@@
 Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain well over a
-billion updates per second with 64 nodes. This compares very favorably to
+billion updates per second with 64 nodes.
+
+This compares very favorably to
 published results~\cite{gups} for other high-end HPC systems.
+Though the actual computation done by GUPS is completely non-sensical, irregular, data-intensive applications typically must be able to sustain a high rate of random accesses in order to, for example, visit and mark vertices during a graph traversal.
+High random access rate in a distributed setting has been a long-standing challenge in HPC.
 
@@
     \begin{center}
-      \includegraphics[width=0.5\textwidth]{results/gups/gups.pdf}
+      \includegraphics[width=0.5\textwidth]{figs/gups.pdf}
     \end{center}

### 2013-03-28 doc/sosp13/evaluation.tex
remark on aggregation

@@
 \Grappa's global memory and communication layers using a faithful
-implementation of the giga updates per second (GUPS) benchmark.
+implementation of the giga updates per second (GUPs) benchmark.
 Read-modify-write updates are dispatched at random to a large array. This
@@
 billion updates per second with 64 nodes.
+Note also that when aggregation is turned off, the update rate is flat at the bottom, demonstrating that aggregation is instrumental for good communication performance.
 
-This compares very favorably to
-published results~\cite{gups} for other high-end HPC systems.
-Though the actual computation done by GUPS is completely non-sensical, irregular, data-intensive applications typically must be able to sustain a high rate of random accesses in order to, for example, visit and mark vertices during a graph traversal.
-High random access rate in a distributed setting has been a long-standing challenge in HPC.
+This compares very favorably to published results~\cite{gups} for other
+high-end HPC systems. Though the actual computation done by GUPS is completely
+non-sensical, irregular, data-intensive applications typically must be able to
+sustain a high rate of random accesses in order to, for example, visit and
+mark vertices during a graph traversal. High random access rate in a
+distributed setting has been a long-standing challenge in HPC.
 
@@
     \end{center}
-    \caption{GUPS (giga updates per second) for \Grappa as the number of nodes grows.}
+    \caption{GUPS (giga updates per second) for \Grappa as the number of nodes grows with and without message aggregation.}
     \label{fig:grappa-gups}

### 2013-03-28 doc/sosp13/memory.tex
Update memory api.

@@
 	\begin{tabular}{l}
-      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
-      	\texttt{\scriptsize global\_free( global\_address )} \\ 
+      	\texttt{\scriptsize GlobalAddress<T> global\_malloc<T>( size )} \\
+      	\texttt{\scriptsize global\_free( GlobalAddress<T> )} \\ 
       	Allocation in the global heap \\ \\
-      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
-      	\texttt{\scriptsize delegate\_read\_async( global\_address, local\_var )}  \\
-      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
-      	\texttt{\scriptsize delegate\_write\_async( global\_address, local\_var )} \\
-      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
-      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\
-      	\texttt{\scriptsize delegate\_fetch\_inc\_async( global\_address, local\_var )} \\ 
+      	\texttt{\scriptsize delegate\_read( GlobalAddress<T>, T )}  \\
+      	\texttt{\scriptsize delegate\_read\_async( GlobalAddress<T>, T )}  \\
+      	\texttt{\scriptsize delegate\_write( GlobalAddress<T>, T value)} \\
+      	\texttt{\scriptsize delegate\_write\_async( GlobalAddress<T>, T )} \\
+      	\texttt{\scriptsize delegate\_cas( GlobalAddress<T>, T, T )} \\
+      	\texttt{\scriptsize delegate\_fetch\_inc( GlobalAddress<T>, T )} \\
+      	\texttt{\scriptsize delegate\_fetch\_inc\_async( GlobalAddress<T>, T )} \\ 
       	Memory operations at the home core of a global address \\ \\
-      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
-      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ 
+      	\texttt{\scriptsize cache\_acquire( GlobalAddress<T>, local\_buf, \{RO,RW,WO\})} \\
+      	\texttt{\scriptsize cache\_release( GlobalAddress<T>, local\_buf )} \\ 
 		Perform cache operations to acquire/release global data.  \\
@@
 
-% We restrict the use of delegate operations in order to make them also
-% useful for synchronization operations.
-
 Delegate operations can be either {\em synchronous} or {\em asynchronous}.
@@
 complete before executing a synchronization operation.
-
-% \TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}
-% \TODO{If we do want a section on GlobalCompletionEvent:
-%   \begin{itemize}
-%     \item many tasks and operations in the applications we've implemented do not need immediate consistency
-%     \item guaranteeing ordering requires blocking a task. this is unavoidable if the task relies on the returned value, or if one task is dependent on others
-%     \item however, much of the applications we've implemented (and other big-data things) are data-parallel, so tasks within a data-parallel loop can run in any order
-%     \item in addition, often these data-parallel tasks will update values that aren't used until the next data-parallel loop
-%     \item this allows us to free up workers/stacks (running out of stacks causes  latency-sensitivity because we can't continue until tasks complete)
-%     \item must guarantee all operations that were supposed to happen in the loop complete before continuing, including tasks and delegate operations
-%     \item \emph{motivating example:} write to array in one loop, use updated values in next loop (could also mention the ``PushBuffer'' example where we buffer queue appends because the order in the queue doesn't matter)
-%     \item GlobalCompletionEvent is complicated by task-stealing and the fact that delegate operations complete on a different core than they started, forcing us to use a ``cancelable barrier''-like approach (possible citation: Ravichandran, et al.
-``Work stealing for multi-core HPC Clusters'')
-%     \item interface to GlobalCompletionEvent:
-%       \begin{itemize}
-%         \item \texttt{gce.enroll(n)}: enroll n tasks/operations to be done, when goes from 0 to 1, leave cancelable barrier
-%         \item \texttt{gce.complete(n)}: complete tasks, when it goes to zero, enter global cancelable barrier
-%         \item \texttt{gce.wait()}: block until all cores have no work
-%       \end{itemize}
-%     \item to prevent early wakeups, stolen tasks must send a completion message back to the core they were started on
-%   \end{itemize}
-% }
 
@@
 the home node before the synchronization operation that protects the data is
-performed. This is done using release operations on cached regions and using the \texttt{GlobalCompletionEvent} object to determine that asynchronous delegates have been fully executed.
-
-% \TODO{Not sure this warrants its own paragraph, seems like the explicit caches are just the same as reading and then writing using delegates.}
+performed. This is done using release operations on cached regions and using
+the \texttt{GlobalCompletionEvent} object to determine that asynchronous
+delegates have been fully executed.

### 2013-03-28 doc/sosp13/intro.tex
ci

@@
 It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.
-The communication layer provides over 10X the small-message performance compared to the raw underlying message injection rate.
-Combining these two components, \Grappa achieves 1.3 GUPS with 64 system nodes, far above any reported cluster performance~\cite{gups} for comprable systems.
+The communication layer helps \Grappa achieve over 1.0 GUPS with 64 system nodes.
 When we compare \Grappa to native MPI implementations of more complex application kernels, the story is more muddled.
-When the MPI implementation is carefully tuned to aggregate messages inside the application -- essentially duplicating the work \Grappa is doing for the programmer, without the generality and context switch overhead -- then the MPI implementation is between 4-10X faster.
+When the MPI implementation is carefully tuned to aggregate messages inside the application -- essentially duplicating the work \Grappa is doing for the programmer, without the generality and context switch overhead -- then the MPI implementation is between 6-10X faster.
 We see this in our comparisons for breadth first search and integer sorting.
@@
 When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2X faster and 4X slower depending on the benchmark.
-Overall we find this encouraging because \Grappa is scalable and achieves this performance from a mass-market cluster, hence adding more system nodes an inexpensive and attractive option.
-
-\comment{
-Our evaluation of \Grappa shows that it runs several irregular application
-kernels (e.g., graph traversal and sort) very efficiently on a commodity
-cluster. Our yardstick for comparison is the XMT hardware itself. Using the
-same number of network interfaces, \Grappa is in the same ballpark as the XMT:
-For unbalanced tree search, \Grappa is over \TODO{3X} faster and shows greatly
-improved scalability; conversely, for breadth first search and betweenness
-centrality \Grappa is 2.5X slower. In Section~\ref{sec:evaluation} we explore
-the factors that underpin this performance. Most importantly, however, for
-significantly less real world cost, users can \emph{add\/} significantly more
-processors to a commodity cluster than an XMT machine and use \Grappa to
-achieve scalable performance. \TODO{revise paragraph after eval, including MPI
-and possibly programming model}
-}
-
-%%% bit bucket
-%\Grappa is designed to smooth over some of the performance discontinuities in commodity hardware, giving good performance when there is little locality to be exploited while allowing the programmer to exploit it when it is available.
+Overall we find this encouraging because \Grappa is scalable and achieves this performance from a mass-market cluster, hence adding more system nodes is an inexpensive option for tackling larger problem sizes.

### 2013-03-28 doc/sosp13/conclude.tex
nc

@@
 memory system, and a network aggregator to make commodity networks efficient
-with small message sizes. \Grappa key aspect is extreme latency tolerance,
+with small message sizes. \Grappa's key aspect is extreme latency tolerance,
 which not only hides network latency but also enables the system to spend time
@@
 
-We explored the performance of \Grappa on three algorithms: unbalanced tree
-search, breadth first search, and betweenness centrality. Performance
-comparisons to the Cray XMT for a small number of nodes (16 nodes / 96 cores)
-show great promise: \Grappa performs on-par with the far more expensive and
-custom XMT system, sometimes achieving better performance and sometimes worse.
-Our initial large scaling experiments to a higher number of nodes shows both promise and room for improvement.
-\Grappa trounces XMT on UTS, but while \Grappa performance scales on BFS and betweenness centrality, limitations in the current aggregator design give the performance edge to XMT.
-Work is ongoing on further refining this component.
+Our evaluation of \Grappa reveals that the core components, scheduling and communication, achieve their design goals.
+Thousands of workers can be efficiently context switched on a multicore processor, up to the DRAM bandwidth.
+Aggregating messages enables \Grappa to achieve over 1.0 GUPS on 64 nodes.
+We also explored four other algorithms: unbalanced tree search, breadth first search, PageRank, and integer sort, comparing performance of these algorithms on \Grappa, the Cray XMT, and against hand-optimized MPI implementations.
+As would be expected, when the MPI implementation exploits in the application, many of the same techniques \Grappa provides in the runtime, such as aggregating requests, yet without the overhead of tasking, the MPI implementation excels.
+This comes with enormous implementation complexity burdened on the application developer, however, and significant tuning has yet to be done on the \Grappa runtime system.
+Compared to the Cray XMT \Grappa is between 2X faster and 4X slower.
+Yet when cost-performance is considered, mass-market x86 cluster hardware makes \Grappa a highly attractive option.

### 2013-03-28 doc/sosp13/evaluation.tex
UTS text

@@
 
-\paragraph{Putting it all together with Unbalanced Tree Search in
-    memory (UTS).}
-We now show the overall performance of \Grappa running UTS.
-Figure~\ref{fig:grappa-uts}. The point of this experiment is to
-demonstrate whether \Grappa's context switching and communication layers can in fact
-be used together, while balancing workload, to run an irregular application efficiently. 
-We look at two classes of trees, T1x and T3x, from the
-original benchmark. T1x trees are very shallow and wide, while T3x
-trees are very deep. When the access to each vertex is a random
-access, the critical path to search T3x trees is very long. On such
-trees, we do not expect there to be sufficient concurrency for any
-system, including \Grappa, to achieve high throughput. \TODO{plot it
-    if time}. To verify this, at the 16-node data point, the average active tasks per
-core over the search is 775 and 13 for T1XL and T3L, respectively.
+\paragraph{Putting it all together with Unbalanced Tree Search in memory
+(UTS).} Figure~\ref{fig:grappa-uts} shows the overall performance of \Grappa
+running UTS. The point of this experiment is to demonstrate whether \Grappa's
+context switching and communication layers can in fact be used together, while
+balancing workload, to run an irregular application efficiently. We look at two classes of trees, T1 and T3, from the original benchmark. T1
+trees are very shallow and wide (i.e., significant parallel slack), while T3
+trees are very deep (i.e., little parallel slack). When the access to each
+vertex is a random access, the critical path to search T3 trees is very long,
+hence the low performance and scalability. On such trees, we do not expect there to be
+sufficient concurrency for any system, including \Grappa, to achieve high
+throughput --- at the 16-node data point, the average active
+tasks per core over the search is 775 and 13 for T1 and T3, respectively. On the other hand \Grappa performs and scales very well for T1 trees.
 
@@
     \caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC.}
-%bar chart, one set of bars per benchmark, one bar per system. runtime normalized to grappa.
     \label{fig:grappa-comparisons}

### 2013-03-28 doc/sosp13/intro.tex
Tiny edit to intro.

@@
 Prefetching is of limited value because request locations are not known early enough.
-Data-parallel frameworks such as MapReduce~\cite{mapreduce:osdi04} are ineffective because they rely on being able to partition data and regular communication patterns.
+Data-parallel frameworks such as MapReduce~\cite{mapreduce:osdi04} are ineffective because they rely on effective data partitioning and regular communication patterns.
 Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.

### 2013-03-28 doc/sosp13/evaluation.tex
Edits to description of UTS.

--- 2013-03-28 doc/sosp13/evaluation.tex
@@
 be used together, while balancing workload, to run an irregular application efficiently.
+Visiting vertices in the distributed tree requires mostly remote accesses, and because each vertex in the tree must be visited before it can be expanded, blocking remote reads are required.
+In this case, we are forced to context switch to tolerate the remote access and continue aggregating. 
 We look at two classes of trees, T1x and T3x, from the
-original benchmark. T1x trees are very shallow and wide, while T3x
-trees are very deep. When the access to each vertex is a random
-access, the critical path to search T3x trees is very long. On such
+original benchmark. T1x trees are very shallow and wide, providing plentiful parallelism, while T3x
+trees are very deep, causing the critical path to search T3x trees to be very long. On such
 trees, we do not expect there to be sufficient concurrency for any
@@
 core over the search is 775 and 13 for T1XL and T3L, respectively.
-
+% . When the access to each vertex is a random access
 \begin{figure}[ht]

### 2013-03-28 doc/sosp13/overview.tex
Minor edits to overview.

@@
 \item[Communication layer.] The main goal of our communication layer is to
-aggregate small messages into large ones, which invisible to the application
+aggregate small messages into large ones, invisible to the application
 programmer. Its interface is based on active messages~\cite{vonEicken92}.

### 2013-03-28 doc/sosp13/intro.tex

@@
 
-\vspace{0.5ex} \noindent{\bf High network communication latency.} The performance challenges of frequent communication is exacerbated by the network latency relative to processor performance.
+\vspace{0.5ex} \noindent{\bf High network communication latency.} The performance challenges of frequent communication are exacerbated by high network latency relative to processor performance.
 Latency of commodity networks runs anywhere from a few to hundreds of microseconds -- ten's of thousands of processor clock cycles.
@@
 each processor would execute a ready instruction chosen from one of its 128
-hardware thread contexts, a sufficient number to fully hide memory access
+hardware thread contexts, a sufficient number to fully tolerate memory access
 latency. The network was designed with a single-word injection rate that
@@
 
-We believe we can support irregular applications with good performance and cost-effectiveness with commodity hardware for two main reasons.
+We believe we can support irregular applications with good performance and cost effectively with commodity hardware for two main reasons.
 First, commodity multicore processors have become extremely fast with high clock rates, large caches and robust DRAM bandwidth.
-Second, commodity networks offer high bandwidth as long as messages are large enough.
+Second, commodity networks offer high bandwidth so long as messages are large enough.
 We build on these two observations and develop \Grappa, a software runtime system that allows a commodity cluster of x86-based nodes connected via an InfiniBand network to be programmed as if it were a single, large, shared-memory NUMA (non-uniform memory access) machine with scalable performance for irregular applications.

### 2013-03-28 doc/sosp13/overview.tex
po

@@
 \item[Communication layer.] The main goal of our communication layer is to
-aggregate small messages into large ones, invisible to the application
+aggregate small messages into large ones.  This process is invisible to the application
 programmer. Its interface is based on active messages~\cite{vonEicken92}.
-Since aggregation and deaggregation of messages needs to be very efficient, we
-made it highly concurrent with careful use of lock-free synchronization
+Since aggregation and deaggregation of messages needs to be very efficient, we perform the process in parallel and carefully use lock-free synchronization
 operations.

### 2013-03-28 doc/sosp13/evaluation.tex
minor edits
eval in progres

--- 2013-03-28 doc/sosp13/evaluation.tex
@@
 \paragraph{Putting it all together with Unbalanced Tree Search in memory
-(UTS).} Figure~\ref{fig:grappa-uts} shows the overall performance of \Grappa
-running UTS. The point of this experiment is to demonstrate whether \Grappa's
-context switching and communication layers can in fact be used together, while
-balancing workload, to run an irregular application efficiently. We look at two classes of trees, T1 and T3, from the original benchmark. T1
-trees are very shallow and wide (i.e., significant parallel slack), while T3
-trees are very deep (i.e., little parallel slack). When the access to each
-vertex is a random access, the critical path to search T3 trees is very long,
-hence the low performance and scalability. On such trees, we do not expect there to be
-sufficient concurrency for any system, including \Grappa, to achieve high
-throughput --- at the 16-node data point, the average active
-tasks per core over the search is 775 and 13 for T1 and T3, respectively. On the other hand \Grappa performs and scales very well for T1 trees.
+(UTS).} 
+Figure~\ref{fig:grappa-uts} shows the overall performance of \Grappa running
+UTS. The point of this experiment is to demonstrate whether \Grappa's context
+switching and communication layers can in fact be used together, while
+balancing workload, to run an irregular application efficiently. Visiting
+vertices in the distributed tree requires mostly remote accesses, and because
+each vertex in the tree must be visited before it can be expanded, blocking
+remote reads are required. 
+In this case, we are forced to context switch to tolerate the remote access and continue aggregating. 
+We look at two classes of trees, T1 and T3, from
+the original benchmark. T1 trees are very shallow and wide (i.e., significant
+parallel slack), while T3 trees are very deep (i.e., little parallel slack).
+When the access to each vertex is a random access, the critical path to search
+T3 trees is very long, hence the low performance and scalability. On such
+trees, we do not expect there to be sufficient concurrency for any system,
+including \Grappa, to achieve high throughput --- at the 16-node data point,
+the average active tasks per core over the search is 775 and 13 for T1 and T3,
+respectively. On the other hand \Grappa performs and scales very well for T1
+trees.
 
@@
 \subsection{Comparing \Grappa to Other Systems}
+\label{eval:mainperf}
 
@@
 \paragraph{Message size and latency.}
+
+To characterize the effectiveness of aggregation, we collected
+histograms of aggregated message sizes and application memory request
+latency on BFS and UTS.
+
+Figure \ref{fig:grappa-message-size} shows the distribution of message
+sizes we are able to aggregate before aggregation times out and
+sends. We see different behavior between UTS and BFS.
+
+In UTS, we obtained the best throughput by flushing frequently: on the order 100 microseconds , but
+this mean
+
+In BFS, we see a trimodal distribution. There are two reasons for
+this. First, in BFS, we obtained the best throughput by flushing at
+least 500 microseconds apart, but also flushing whenever the core went
+idle. This leads to a large number of messages that are not
+
+Recall that global memory can be accessed in two ways. For requests of
+locations that are expected to shared, delegate operations are
+used. For requests of non-shared locations where locality can be
+exploited, explicit cache operations are used. We see this reflected
+in the fraction of requests that are handled remotely: less than one
+percent of delegate operations in both BFS and UTS access local
+data. Since explicit cache operations are intended to exploit
+locality; they more often access data on the local node. Only 20
+percent of UTS explicit cache operations are remote, whereas 44
+percent of BFS explict cache operations are remote.
+
+The bimodal nature of the BFS distribution is due to a locality optimization that o
+
+Both
+
+In BFS we exploit a large
+ Since we are traversing power-law graphs,
+some vertices have a very high number of edges. We exploit locality in
+BFS by having tasks visit vertices whose edge lists are stored on the
+local node. When this local traversal hits a high-degree vertex, BFS
+is able to generate a large number of
+
+Figure \ref{fig:grappa-latency} shows
+
+We see different aggregation behavior between BFS and UTS:
+
+In BFS, the bulk of remote accesses are done to take owner
+
+Conversely, UTS attempts to
+
 Distribution if possible. With and without aggregation.
@@
 \end{figure}
-
-\paragraph{Frequency of remote requests.}
 

### 2013-03-28 doc/sosp13/evaluation.tex
overall paragraph cleanup
histogram characterization

@@
 Figure~\ref{fig:grappa-comparisons} shows the results. Overall, \Grappa is
-within striking distance of XMT and UPC performance. However, it underperforms
-compared to hand-tuned MPI. Nevertheless, this comparison needs to be taken
-with a grain of salt for several reasons. First, we have not spent as much
-time tuning \Grappa's implementations. This is supported by inspecting the BFS
-MPI implementation we used, which clearly shows that it employs a lot of
-algorithmic-specific optimizations. In fact, some of these optimizations
-resemble some of what \Grappa does automatically, like message aggregation.
-Second, the GUPS results presented earlier suggests that \Grappa performance
-can do a lot better. \TODO{update this once we have more results. }
+within striking distance of XMT and well beyond UPC performance (UTS only).
+However, it underperforms compared to hand-tuned MPI. Nevertheless, this
+comparison needs to be taken with a grain of salt for several reasons. First,
+we have not spent as much time tuning \Grappa's implementations. This is
+supported by inspecting the BFS MPI implementation we used, which clearly
+shows that it employs a lot of algorithmic-specific optimizations. In fact,
+some of these optimizations resemble some of what \Grappa does automatically,
+like message aggregation. Second, the GUPS results presented earlier suggests
+that \Grappa performance can do a lot better. 
+% \TODO{update this once we have more results. }
 
@@
 
-In UTS, we obtained the best throughput by flushing frequently: on the order 100 microseconds , but
-this mean
-
-In BFS, we see a trimodal distribution. There are two reasons for
-this. First, in BFS, we obtained the best throughput by flushing at
-least 500 microseconds apart, but also flushing whenever the core went
-idle. This leads to a large number of messages that are not
-
 Recall that global memory can be accessed in two ways. For requests of
@@
 
-The bimodal nature of the BFS distribution is due to a locality optimization that o
+In UTS, we obtained the best throughput by flushing frequently: on the
+order 100 microseconds. This, combined with the fewer number of remote
+accesses, means that there are a limited nubmer of requests generated
+per flush interval, leading to packets primarily smaller than 10KB.
 
-Both
+In BFS, we see a trimodal distribution. There are two reasons for
+this. First, in BFS, we obtained the best throughput by flushing at
+least 500 microseconds apart, and also flushing whenever the core went
+idle. Since this process ignores whether another flush has occured
+recently, it leads to a large number of small messages. This leads to
+the first and smallest mode of the distribution.
 
-In BFS we exploit a large
- Since we are traversing power-law graphs,
-some vertices have a very high number of edges. We exploit locality in
-BFS by having tasks visit vertices whose edge lists are stored on the
-local node. When this local traversal hits a high-degree vertex, BFS
-is able to generate a large number of
+The other two modes are due to a locality optimization in the \Grappa
+version of BFS. To minimize unnecessary remote communication, each
+core visits vertices whoes edge lists are stored locally. Since we are
+traversing power-law graphs, a small fraction of vertices have a very
+high number of edges. This means that occasionally a core visits a
+high-degree vertex which allows it to very quickly generate a large
+batch of requests. This interacts well with aggregation; when we find
+high degree vertices, we are able to aggregate packets on the order of
+40KB. In flush intervals without high-degree vertices, we see packets
+on the order of 8KB.
 
-Figure \ref{fig:grappa-latency} shows
-
-We see different aggregation behavior between BFS and UTS:
-
-In BFS, the bulk of remote accesses are done to take owner
-
-Conversely, UTS attempts to
-
-Distribution if possible. With and without aggregation.
+Figure \ref{fig:grappa-latency} shows the latency distribution of
+remote delegate operations. This distribution follows the aggregated
+message size distribution: when we form larger packets, the system
+takes longer to process them.
 
@@
 \end{figure}
-
-\paragraph{Work stealing behavior.}
-Performance loss of not having stealing. Ratio of \# of steals over total \# of task spawns. UTS needs stealing to work.
 

### 2013-03-28 doc/sosp13/evaluation.tex
data in table form

--- 2013-03-28 doc/sosp13/evaluation.tex
@@
 
-\begin{figure}[ht]
+% \begin{figure}[ht]
+%     \begin{center}
+%       \includegraphics[width=0.5\textwidth]{results/benchmarks.pdf}
+%     \end{center}
+%     \caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC.}
+%     \label{fig:grappa-comparisons}
+% \end{figure}
+
+\begin{table}[htb]
 \begin{center}
-      \includegraphics[width=0.5\textwidth]{results/benchmarks.pdf}
+\begin{tabular}{l|c|c|c|c}
+&	Cora &      XMT	&        MPI	&   UPC \\ \hline
+GUPS	& 1	& 2.23	& 0.11& 	-- \\ 
+BFS	& 1 &	4.33	& 14.37 & 	-- \\ 
+Intsort	& 1	& 3.77 &	5.88	& -- \\
+UTS (T1)	& 1	& 0.38	& -- &	0.09 \\ 
+Pagerank	& 1	& 4.35	& 4.87 &	-- \\ 
+\end{tabular}
 \end{center}
-    \caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC.}
-    \label{fig:grappa-comparisons}
-\end{figure}
+\caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC. Numbers are presented as throughput normalized to \Grappa.}
+\label{tab:grappa-comparisons}
+\end{table}
 
-Figure~\ref{fig:grappa-comparisons} shows the results. Overall, \Grappa is
-within striking distance of XMT and UPC performance. However, it underperforms
-compared to hand-tuned MPI. Nevertheless, this comparison needs to be taken
-with a grain of salt for several reasons. First, we have not spent as much
-time tuning \Grappa's implementations. This is supported by inspecting the BFS
-MPI implementation we used, which clearly shows that it employs a lot of
-algorithmic-specific optimizations. In fact, some of these optimizations
-resemble some of what \Grappa does automatically, like message aggregation.
-Second, the GUPS results presented earlier suggests that \Grappa performance
-can do a lot better. \TODO{update this once we have more results. }
+Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
+within striking distance of XMT and well beyond UPC performance (UTS only).
+However, it underperforms compared to hand-tuned MPI. Nevertheless, this
+comparison needs to be taken with a grain of salt for several reasons. First,
+we have not spent as much time tuning \Grappa's implementations. This is
+supported by inspecting the BFS MPI implementation we used, which clearly
+shows that it employs a lot of algorithmic-specific optimizations. In fact,
+some of these optimizations resemble some of what \Grappa does automatically,
+like message aggregation. Second, the GUPS results presented earlier suggests
+that \Grappa performance can do a lot better.
 

### 2013-03-28 doc/sosp13/evaluation.tex

--- 2013-03-28 doc/sosp13/evaluation.tex
@@
 
-\begin{figure}[ht]
+% \begin{figure}[ht]
+%     \begin{center}
+%       \includegraphics[width=0.5\textwidth]{results/benchmarks.pdf}
+%     \end{center}
+%     \caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC.}
+%     \label{fig:grappa-comparisons}
+% \end{figure}
+
+\begin{table}[htb]
 \begin{center}
-      \includegraphics[width=0.5\textwidth]{results/benchmarks.pdf}
+\begin{tabular}{l|c|c|c|c}
+&	Cora &      XMT	&        MPI	&   UPC \\ \hline
+GUPS	& 1	& 2.23	& 0.11& 	-- \\ 
+BFS	& 1 &	4.33	& 14.37 & 	-- \\ 
+Intsort	& 1	& 3.77 &	5.88	& -- \\
+UTS (T1)	& 1	& 0.38	& -- &	0.09 \\ 
+Pagerank	& 1	& 4.35	& 4.87 &	-- \\ 
+\end{tabular}
 \end{center}
-    \caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC.}
-    \label{fig:grappa-comparisons}
-\end{figure}
+\caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC. Numbers are presented as throughput normalized to \Grappa.}
+\label{tab:grappa-comparisons}
+\end{table}
 
-Figure~\ref{fig:grappa-comparisons} shows the results. Overall, \Grappa is
+Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
 within striking distance of XMT and well beyond UPC performance (UTS only).
@@
 that \Grappa performance can do a lot better.
-% \TODO{update this once we have more results. }
 

### 2013-03-28 doc/sosp13/tasks.tex

@@
 
-The basic unit of execution in \Grappa is a {\em task}. Each hardware core has
-a single operating system thread pinned to it. When tasks are ready to
+The basic unit of execution in \Grappa is a {\em task}. When tasks are ready to
 execute, they are mapped to a {\em worker}, which is akin to a user-level
-thread. % \Grappa keeps a large pool of workers per hardware core.
+thread. Each hardware core has a single operating system thread pinned to it.
 
 \paragraph{Tasks} 
-Tasks are specified with a ``functor'' object that holds both code to execute and initial state.
+Tasks are specified by a ``functor'' object that holds both code to execute and initial state.
 The functor can be specified with a function pointer and explicit arguments, a C++ struct that overloads the parentheses operator, or a C++11 lambda.
 These objects, typically very small (on the order of 64 bytes), hold read-only values such as an iteration index and pointers to common data or synchronization objects.
-These task functors can be serialized and transported around the system, and eventually executed by a worker, as described next.
+Task functors can be serialized and transported around the system, and eventually executed by a worker, as described next.
 
@@
 communication) tasks. A worker is simply a collection of status bits and a
-stack, allocated at a particular core. When a task is ready to be executed it
-is assigned to a worker, which executes the task functor on its own stack. 
+stack, allocated at a particular core. When a task is ready to execute it
+is assigned to a worker, that executes the task functor on its own stack. 
 Once a task is mapped to a worker it stays with that worker until it finishes.
@@
 \paragraph{Scheduling} During execution, a worker yields control of its core
-whenever it performs a long-latency operation, allowing the processor to
+whenever performing a long-latency operation, allowing the processor to
 remain busy while waiting for the operation to complete. In addition, a
@@
       Yields core to scheduler, enqueuing caller to be \\ scheduled again soon \\
+    \texttt{\scriptsize wake( task * $t$ ) } \\
+      Enqueues some other task $t$ to be scheduled again soon \\
     \texttt{\scriptsize suspend() }  \\
       Yields core to scheduler, enqueuing caller only once \\ another task calls wake \\
-    \texttt{\scriptsize wake( task * $t$ ) } \\
-      Enqueues some other task $t$ to be scheduled again soon \\
 	\end{tabular}
@@
 Each core in a \Grappa system has its own independent scheduler. The scheduler
-has a single FIFO queue of active \emph{workers} ready to execute, the {\it
-ready worker queue}. Each scheduler also has three queues of \emph{tasks}
+has a single FIFO queue of active workers ready to execute, the {\it
+ready worker queue}. Each scheduler also has three queues of tasks
 waiting to be assigned a worker:

### 2013-03-28 doc/sosp13/evaluation.tex
fix context switching

@@
 number of workers --- even with one-half million workers, context switch
-is around 75ns.
+is around 75ns. In comparison, for the same yield test using kernel-level
+Pthreads on a single core, the switch time is 450ns for a few threads and 800ns for
+1000-32000 threads.
 
-We also measured (not in the plot) the \emph{rate} of context switch for all
-cores in a node, which showed that our rate is limited by off-chip memory
-bandwidth. Each context is 4 cache lines (1 for the worker struct and 3 for
-stack data), leading to 8 cache line transfers per context switch
-(write the previous context, read the next in). The off-chip bandwidth of a
-single socket in our system is 270M cache lines per second, which implies that
-we can sustain at most 34M context switches per second per socket, which is
-almost exactly what we sustain.
+If we calculate aggregate context switch\emph{rate} of all
+cores in a node, we find that with prefetching, \Grappa context
+switching is
+limited \emph{not} by memory latency, as normally assumed, but rather memory bandwidth.
+Specifically, we emprically found that 4 cachelines (1 for worker
+struct and 3 for stack data) was sufficient to avoid cache misses in
+the microbenchmark. Every context switch then requires 8 cacheline transfers. The off-chip
+bandwidth of a single socket in our system is 270M cachelines per
+second \cite{porterfield:bw,Nelson:hotpar11-real}. This implies that
+we can sustain at most 34M context switches per second per socket once
+the space of all contexts is enough to mostly reside outside of L3.
 

### 2013-03-28 doc/sosp13/evaluation.tex
eval

@@
 switch time. And context switch time does not grow significantly with the
-number of workers --- even with one-half million workers, context switch
-is around 75ns. In comparison, for the same yield test using kernel-level
+number of workers --- even with one-half million workers, context
+switch time is around 75ns. In comparison, for the same yield test using kernel-level
 Pthreads on a single core, the switch time is 450ns for a few threads and 800ns for

### 2013-03-28 doc/sosp13/evaluation.tex
data in table form
evaluation.tex

--- 2013-03-28 doc/sosp13/evaluation.tex
@@
 
-% \begin{figure}[ht]
-%     \begin{center}
-%       \includegraphics[width=0.5\textwidth]{results/benchmarks.pdf}
-%     \end{center}
-%     \caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC.}
-%     \label{fig:grappa-comparisons}
-% \end{figure}
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[width=0.5\textwidth]{results/benchmarks.pdf}
+    \end{center}
+    \caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC.}
+    \label{fig:grappa-comparisons}
+\end{figure}
 
-\begin{table}[htb]
-\begin{center}
-\begin{tabular}{l|c|c|c|c}
-&	Cora &      XMT	&        MPI	&   UPC \\ \hline
-GUPS	& 1	& 2.23	& 0.11& 	-- \\ 
-BFS	& 1 &	4.33	& 14.37 & 	-- \\ 
-Intsort	& 1	& 3.77 &	5.88	& -- \\
-UTS (T1)	& 1	& 0.38	& -- &	0.09 \\ 
-Pagerank	& 1	& 4.35	& 4.87 &	-- \\ 
-\end{tabular}
-\end{center}
-\caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC. Numbers are presented as throughput normalized to \Grappa.}
-\label{tab:grappa-comparisons}
-\end{table}
-
-Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
+Figure~\ref{fig:grappa-comparisons} shows the results. Overall, \Grappa is
 within striking distance of XMT and well beyond UPC performance (UTS only).
@@
 that \Grappa performance can do a lot better. 
+% \TODO{update this once we have more results. }
 
@@
 histograms of aggregated message sizes and application memory request
-latency on BFS and UTS.
+latency for BFS and UTS (Figure \ref{fig:grappa-message-size}).
 
-Figure \ref{fig:grappa-message-size} shows the distribution of message
-sizes we are able to aggregate before aggregation times out and
-sends. We see different behavior between UTS and BFS.
-
-Recall that global memory can be accessed in two ways. For requests of
+%%% I couldn't make this paragraph useful in the limited time we have.
+%%% I think it's ok without. -Mark
+\comment{Recall that global memory can be accessed in two ways. For requests of
 locations that are expected to shared, delegate operations are
@@
 percent of UTS explicit cache operations are remote, whereas 44
-percent of BFS explict cache operations are remote.
+percent of BFS explict cache operations are remote.}
 
 In UTS, we obtained the best throughput by flushing frequently: on the
-order 100 microseconds. This, combined with the fewer number of remote
+order of 100 microseconds. This, combined with the fewer number of remote
 accesses, means that there are a limited nubmer of requests generated
@@
 
-In BFS, we see a trimodal distribution. There are two reasons for
-this. First, in BFS, we obtained the best throughput by flushing at
-least 500 microseconds apart, and also flushing whenever the core went
-idle. Since this process ignores whether another flush has occured
-recently, it leads to a large number of small messages. This leads to
-the first and smallest mode of the distribution.
-
-The other two modes are due to a locality optimization in the \Grappa
-version of BFS. To minimize unnecessary remote communication, each
-core visits vertices whoes edge lists are stored locally. Since we are
-traversing power-law graphs, a small fraction of vertices have a very
-high number of edges. This means that occasionally a core visits a
-high-degree vertex which allows it to very quickly generate a large
-batch of requests. This interacts well with aggregation; when we find
-high degree vertices, we are able to aggregate packets on the order of
-40KB. In flush intervals without high-degree vertices, we see packets
-on the order of 8KB.
+In BFS, we see a trimodal distribution.
+There are two reasons for this.
+First, in BFS, we obtained the best throughput by flushing at least 500 microseconds apart, and also flushing whenever a core is idle.
+Since this process ignores whether another flush has occurred recently, it leads to a large number of small messages.
+This leads to the first and smallest mode of the distribution.
+The other two modes are due to a locality optimization in the \Grappa version of BFS.
+To minimize unnecessary remote communication, each core visits vertices whoes edge lists are stored locally.
+Since we are traversing power-law graphs, a small fraction of vertices have a very high number of edges.
+This means that occasionally a core visits a high-degree vertex which allows it to very quickly generate a large batch of requests.
+This interacts well with aggregation; when we find high degree vertices, we are able to aggregate packets on the order of 40KB.
+In flush intervals without high-degree vertices, we see packets on the order of 8KB.
 

### 2013-03-28 doc/sosp13/evaluation.tex

--- 2013-03-28 doc/sosp13/evaluation.tex
@@
 \end{center}
-\caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC. Numbers are presented as throughput normalized to \Grappa.}
+\caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
 \label{tab:grappa-comparisons}

### 2013-03-28 doc/sosp13/evaluation.tex
profile

@@
         Benchmark     & Comm & User & Idle & Sched \\ \hline
-        GUPS          & 6.60  & 42.94   & 47.74 & 2.80 \\
+        GUPS          & 62.9  & 18.9    & 15.4 & 2.80 \\ 
         BFS           & 54.84 & 30.90   & 10.94 & 3.43 \\ 
@@
 
-\TODO{PAGERANK:
-    Without parallelizing the dot product, \Grappa cannot achieve
-high utilization of workers proportional to the size of the data set.
-Also mention that the cost of this is contention at the target
-elements of vector.  The other parts of pagerank are super fast
-(table?)
-}
+Figure \ref{fig:grappa-profile} shows the breakdown of execution
+time. Since \Grappa is focused on frequent communication, it is not
+surprising that many of our benchmarks spend more than half of their
+time doing communication. User time is primarily focused on generating
+more requests; idle time occurs when we run out of parallel work or
+out of workers to execute that work. Scheduler time includes both
+context switch time and deciding what task to run next. Our efficient
+context switching allows us to keep this to a few percent of execution
+time even with thousands of threads per core.
 

### 2013-03-28 doc/sosp13/tasks.tex
pagerank

@@
 Since \Grappa keeps a very large number of active workers, their context data
-will not fit in cache. Therefore, the scheduler has to perform very careful
-prefetching in order to reduce context switch cost to a minimum. We determined
-empirically that prefetching the fourth worker in the scheduling order is
-sufficient. We also prefetch three cache lines from the stack for each worker.
-This makes context switching effectively free of cache misses even to hundreds
+will not fit in cache. By oversubscribing on the number of workers
+beyond what is required for latency tolerance, the scheduler can
+ensure there is always some number of context pointers in the ready
+queue. This allows the scheduler to prefetch contexts into
+cache. The reason this works is because the size of the L1 cache is
+sufficient to hold enough contexts to tolerate the latency to main
+memory. Empirically we find that prefetching the fourth worker in the scheduling order is
+sufficient. This prefetching makes context switching effectively free of cache misses even to hundreds
 of thousands of threads. We provide an analysis of our context switch

### 2013-05-05 doc/pgas13/intro.tex
Setup latex for PGAS paper (hardest part).

\section{Introduction}
The Grappa~\cite{Nelson:hotpar11-real} runtime system is optimized for fine-grained accesses.

### 2013-05-17 doc/pgas13/flat-combining.tex
Write some intro.

Flat combining allows threads to cooperate when accessing a synchronized shared data structure to get greater overall throughput than even each of them operating serially. The trick, however, is designing a way for the threads to communicate their intentions to each other without incurring the same problem they had before.

### 2013-05-17 doc/pgas13/grappa.tex
Write some intro.

One of the persistent challenges with PGAS languages is that existing shared memory applications typically do  fine-grained operations, so naively translating them results in poor performance because many operations must be done via remote communication. Many techniques exist, therefore, to help come up with better ways to partition data, schedule computation, and where necessary, do communication in bulk to best utilize network resources.

Irregular applications, however, are characterized by being highly dependent on the shape of the data, making it difficult to statically partition, causing load imbalance, and requiring a majority of fine-grained random accesses.

### 2013-05-17 doc/pgas13/intro.tex
Write some intro.

@@
 \section{Introduction}
-The Grappa~\cite{Nelson:hotpar11-real} runtime system is optimized for fine-grained accesses.
+% - PGAS languages successfully bring shared memory problems into distributed memory systems
+% - Cost functions are different, but many of the other considerations, such as reducing locking, still apply
+% - With abundant parallelism, we can tolerate some serialization to reduce number of synchronizations and reduce overall communication
+
+% The Grappa~\cite{Nelson:hotpar11-real} runtime system is optimized for fine-grained accesses.
+
+The goal of partitioned global address space (PGAS) languages and runtimes is to provide the illusion of a single shared memory to a program that actually executes on a number of machines each with their own memory.
+This allows programmers to write their algorithms without needing to explicitly manage communication.
+Once the algorithm is correct, many techniques exist to help improve performance improving locality (spatial and temporal) and coarsening communication where possible.
+However, with the shared memory abstraction come all of the difficult concurrency issues that arise in physically-shared memory.
+Luckily, PGAS can leverage the large body of work solving these issues in a shared memory domain, there are just a host of different costs and performance trade-offs to be made in a distributed setting.
+
+When multiple threads of control concurrently access the same object, they have a choice about how to conduct themselves: contend or cooperate.
+Either they all fight to obtain the single global lock, or they attempt a series of potentially-contended atomic operations in the case of a more advanced synchronization strategy such as a lock-free queue.
+In both cases, as the number of concurrent accessors increases, the more contention and therefore failed synchronization operations there are.
+If, however, the threads cooperate, they can delegate one thread to do all the operations and avoid excessive synchronization overhead on highly contended locations.
+In addition, the particular insight of flat-combining is that some operations have semantics such that they can be \emph{combined} and applied more efficiently together than individually.
+With a sufficiently efficient mechanism for delegating and combining, the benefits of reduced synchronization can outweigh the cost of serialization.

### 2013-05-17 doc/pgas13/intro.tex
add to intro, include Grappa and Flat Combining sections

@@
 % - With abundant parallelism, we can tolerate some serialization to reduce number of synchronizations and reduce overall communication
-
-% The Grappa~\cite{Nelson:hotpar11-real} runtime system is optimized for fine-grained accesses.
 
@@
 With a sufficiently efficient mechanism for delegating and combining, the benefits of reduced synchronization can outweigh the cost of serialization.
+
+To honor the shared-memory abstraction, a common goal is for distributed shared data structures to adhere to a sequentially consistent memory model.
+Again, the simplest way to maintain a consistent global view of the data structure is to have a single global lock that is used to implementation mutual exclusion and enforce a serializable order over insertions.
+However, in a system with physically distributed memory, the cost of synchronizing in this way is even greater.
+In addition, with the massive amount of parallelism in a cluster of multiprocessors, the amount of contention on this shared data structure is higher than any shared-memory multiprocessor.
+
+In many cases, algorithms do not require operations to be immediately consistent.
+Many optimizations for message-passing or PGAS-style distributed memory algorithms leverage this observation to minimize the need for fine-grained communication.
+However, they typically must make this visible to programmers, who now must take care to insert their own memory barriers where appropriate, or otherwise express the way in which consistency relaxation can occur.
+
+The goal of this work is to see whether the concept of flat-combining can be used to make the cost of enforcing fine-grained synchronization acceptable.
+To evaluate this possibility, we leverage the Grappa runtime~\cite{Nelson:hotpar11-real}, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads.
+With enough concurrent threads, it is actually possible to allow each thread to maintain sequentiality by blocking on accesses, but still have an opportunity to effectively combine many fine-grained synchronous operations for better performance.
+
+% lay out the upcoming sections of the paper...

### 2013-05-22 doc/prefetch13/model.tex
initial draft of prefetching paper

\section{Analytical Model}

The above equation for processor utilization will hold, but $S_v$ is a
function of $v$ only through the effects of limited bandwidth. We
describe the system, model multithreading as a pipeline, derive
constraints, and discuss the limits on scheduling flexibility.

We make an initial simplification that will be relaxed later. All data
arrivals are one-to-one; that is, a task needs to suspend on one data
item that only that task is waiting on.

\subsection{System}

For the purposes of this paper, we define a \emph{task} as a
schedulable sequence of computation. A task has its own stack, so it
can block and yield the processor when it comes to an unresolved data
dependence.

We consider a system with two levels of memory.
\begin{enumerate}
    \item \emph{Top level store}: closest memory to the processor, where
        executing task state must reside for good performance (e.g., L1 cache)
    \item \emph{Data store}: All heap data and unscheduled tasks will be
        stored here (e.g., main memory)
\end{enumerate}

We must show that the latency to the data store can be overlapped up
to the limit of bandwidth to the data store. This means that the top
level store may only contain a constant amount of state, specifically
a constant number of tasks and scheduler state.

\paragraph{Pipeline}

We will consider a dataflow multithreading system as a pipeline. The
lifecycle of a task is as follows.

Task A resides in the data store, suspended on a data dependence $D_a$.
\begin{enumerate}

    \item Task B resides in top level store and is currently executing. It
        produces $D_a$, so it marks Task A ready and copies the $D_a$ to an
        associated storage location.
    \item At some point in the future, Task A is scheduled. Its stack
        and $D_a$ start moving towards the cpu.
    \item After a time equal to the latency of the data store, the
        stack and $D_a$ arrive in the top level store. Data is
        copied to the stack, the task runs.
    \item Task A suspends on another data dependence
        Repeat 1-4 until Task A exits.

        See figure~\ref{fig:pipeline}.

\end{enumerate}

\begin{figure*}[htbp]
    \begin{center}
      \includegraphics[width=0.7\textwidth,angle=90]{figs/pipeline.pdf}
    \end{center}
    \caption{Concpetual pipeline for dataflow multithreading}
    \label{fig:pipeline}
\end{figure*}

\paragraph{Parameters}
The parameters of the system are: 
\begin{itemize}
    \item $L_m$: latency to the data store
    \item $L_t$: latency to the top level store
    \item $B_m$: bandwidth to the data store
    \item $B_t$: bandwidth to the top level store
    \item $S_m$: size of the data store
    \item $S_t$: size of the top level store
\end{itemize}

The size of each memory level determines its latency and bandwidth
\TODO{where do we discuss bandwidth and latency, cost, physics
    relationships in depth?}

The paramters of the tasking implementation and
user program are:
\begin{itemize}
    \item $S_T$: size of a task, either some top stack frames or at most the whole stack.
    \item $T_S$: syncrhonization time between tasks
    \item $T_E$: execution time between blocking. This is assumed to
        be a single average.
    \item $S_C$: size of a \emph{candidate}, or task reference, for use by the scheduler. It
        includes at least a stack pointer but also may include
        scheduling information (e.g., a priority).
    \item $N_C$: number of candidates in the top level store.
\end{itemize}

\subsection{Constraints}
To operate in saturation mode, the processor needs a large enough
number of tasks $N_T$ to
always have useful work to overlap with memory latency. From Little's
law, $N_T \ge R_m * L_m$ \checkme{make R precise}. The size of the top
level store $S_t$ is proportional to the number of candidates and the
size of the executing task: $S_t \ge S_C * N_C + S_T$. The context
switch rate is upper bounded by the rate at which the system can read
and write tasks to the data store: $R_m \le \frac{B_m}{2*S_T}$, and is
also upper bounded by task run time and synchronization $R_m \le
\frac{1}{T_E + T_S}$.

\subsection{Relaxing one-to-one data matching}

\paragraph{Many data messages to one task}. The task
suspends once for every message that it must wait for.

\paragraph{One data message to many tasks}. When the tasks are on different
processors, there are actually multiple messages and it is the one to
one case. When the tasks are on the same processor, in stage 1 all
stack pointers can be enqueued and copies of the data.

To account for cases where a datum is produced before it is blocked
on, we can simply model a producer/consumer synchronization object as
a task (without a stack). When scheduled, the object is simply marked full.

\TODO{some analog to tag stores in dataflow machines.
swanson's thesis has a good survey}

\section{Scheduling}

The scheduling state used to make a decision must fit in the top level
store to ensure fast context switches. This means the scheduling state
cannot grow with $N_T$, the number of required tasks. A scheduling
decision then can choose among a fixed size subset $S$ of all ready
tasks stored in the data store.

We list the places where scheduling decisions can be made and their 
implications on implementation.

\paragraph{Upon task wake}
The first place to make a scheduling decision for a task is when that
task reference is placed in the ready set. Tasks within $S$ may be
reordered. Note that the task reference must include information used for scheduling
(e.g., a priority) that is passed around by value. Once a task falls
out of $S$, it may be placed in a data structure that
does not require comparisons with other tasks on insertion (e.g., a
FIFO or a priority queue with a fixed number of priorities).
Intuitively, the ready set is a queue of references with some fixed amount of the tail in the top level
store.

\paragraph{When picking a task to prefetch}
The scheduler can pick from a subset $S$ which task to start moving
from the data store. Intuitively, the ready set is a queue of
references with some fixed amount of the head in the top level.

By scheduling at prefetch time, an implementation of pipelined
multithreading that relies on existing software prefetch instructions
cannot support arbitrary scheduling even between a fixed number of
choices. Because a future load instruction reads the task, arbitrary
scheduling would require space proportional to data store latency to
remember all the scheduling decisions in the prefetch window.

On the other hand, a context cache could support arbitrary scheduling
decisions over a fixed window of tasks. \TODO{do we want to expand on
    this}

\paragraph{When picking a task to run}

\subsection{Implementation}
If scheduling at prefetch time, general software prefetch and load instructions in modern processors
cannot provide the required functionality for arbitrary constant size
scheduling. Because the address must be prefetched and later loaded,
the scheduler would need to keep track of the scheduling decisions
made on prefetch to load in the same order.

### 2013-05-22 doc/prefetch13/tflsummary.tex
initial draft of prefetching paper

\section{TFL Argument}

\begin{enumerate}

\item Previous work assumed that in saturation, ie when $v$, the
    number of threads is enough to cover the latency of a remote
    access, the processor utilization is $U=R/(R+S)$, where $R$ is the
    time between remote accesses and $S$ is synchronization time.

\item \cite{culler:93} argued that because there is increasing cost to
    switch between more threads, $S$ must be an increasing function of
    $v$, and the execution of the processor may slow down due to larger
    $v$, leading to $U= \frac{1}{C_v} * \frac{R}{R+S_v}$, where $S_v$ is increasing
    with $v$ and $C_v$, the increased program execution time, is increasing with
    $v$.

\item While this analysis is sound, $S_v$ and $C_v$ are not fundamentally
    increasing due to latency to top level but rather bandwidth to
    thread data. (Not entirely clear to me that they ruled out
    bandwidth or not, hard to tell what they were thinking, see
    ``physics and dollars")

\item This changes how we think about this limit, if we can increase
    bandwidth (cost-based, although I’m not sure I fully believe this)
    then the number of threads can be much higher than if we are
    limited by latency (also cost-based but primarily physics).

\end{enumerate}

### 2013-05-22 doc/prefetch13/intro.tex

\section{Introduction}

\TODO{expand}
Importance of irregular applications today motivates systems that support multithreaded dataflow

 \cite{culler:93} Argues that a fundamental limit of dataflow multiprocessing is that synchronization cost (context switch time) increases with number of threads because access time to the top level store increases with capacity

The observation that only a fixed number of contexts can fit in a fast top level store is correct. However, the assumption that this corresponds to being limited to scheduling with only these contexts to avoid the latency to main memory need not apply. If the scheduling of threads is treated as a pipeline, then synchronization cost is determined by the bandwidth to the memory that holds all ready threads, rather than the latency.

In this paper: prior work, analytical model, an implementation of our ideas in a multithreaded runtime system, Grappa, for commodity processors

### 2013-05-22 doc/pgas13/intro.tex
Rework intro to incorporate Myers' comments.

--- 2013-05-05 doc/pgas13/intro.tex
@@
 \section{Introduction}
-The Grappa~\cite{Nelson:hotpar11-real} runtime system is optimized for fine-grained accesses.
+% - PGAS languages successfully bring shared memory problems into distributed memory systems
+% - Cost functions are different, but many of the other considerations, such as reducing locking, still apply
+% - With abundant parallelism, we can tolerate some serialization to reduce number of synchronizations and reduce overall communication
+
+The goal of partitioned global address space (PGAS) languages and runtimes is to provide the illusion of a single shared memory to a program that actually executes on a number of machines each with their own memory.
+This allows programmers to write their algorithms without needing to explicitly manage communication.
+Once the algorithm is correct, many techniques exist to help improve performance improving locality (spatial and temporal) and coarsening communication where possible.
+However, with this shared memory abstraction come all of the difficult concurrency issues that arise in physically-shared memory.
+Luckily, there exists a large body of work solving these issues in physically-shared memory which the PGAS community can leverage.
+The opportunity here is that, in a distributed setting, the costs of communication and synchronization operations are different, so different performance trade-offs will be made.
+
+Globally-shared data structures are one of the cornerstones of shared-memory and PGAS abstractions. 
+In order for multiple concurrent threads to interact with one another, they must observe shared data consistently.
+It is commonly accepted that the easiest consistency model to reason about enforces that all accesses appear to happen in some serializable order that all threads agree on (known as \emph{sequential consistency or SC}).
+However, in both physically-shared memory and PGAS, maintaining this sequentially consistent view of data among all concurrently accessing threads presents performance challenges.
+The simplest way to maintain SC for a given data structure is to have a single global lock that implements mutual exclusion and enforces a serializable order over read and write operations.
+The cost of literally serializing accesses in this way is typically considered prohibitive, even in physically shared memory.
+With the massive amount of parallelism in a cluster of multiprocessors and with the increased cost of remote synchronization, the problem magnifies.
+
+Prior work on shared memory has explored the scalability challenges of adding more concurrent accessors to shared data structures.
+% One observation was that, when multiple threads of control concurrently access the same object, they can conduct themselves in one of two ways: contend or cooperate.
+In the classic case of contention, either all threads fight to obtain a single global lock, or they attempt a series of potentially-contended atomic operations in the case of a more advanced synchronization strategy such as a lock-free queue.
+In both cases, as the number of concurrent accessors increases, the more they contend and the more failed synchronization operations there are.
+The observation of previous work was that the cost of many of these failed synchronization attempts can be mitigated by having the threads \emph{cooperate} via delegation rather than \emph{contend}.
+If multiple threads delegate a single thread to do all of their operations, they can avoid excessive synchronization overhead on ``hot'' locations.
+
+The trick is coming up with a mechanism for delegating that has lower synchronization overhead than the original contention case.
+Prior work showed that a cleverly-implemented ``publication lists'' can allow multiple accessors to cooperate with minimal synchronization.
+This same work makes the additional observation that, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually, which they dub \emph{flat combining}.
+Together, the reduced synchronization cost and combined operations allow even a data structure with a single global lock to scale better than complicated concurrent data structure implementations with fine-grained locking.
+Further, the same publication list mechanism can be applied to other data structures, and all that must be customized is the particular way in which operations for a given data structure can be combined.
+
+% Make this point later?
+% In many cases, algorithms do not require operations to be immediately consistent.
+Many optimizations for message-passing or PGAS-style distributed memory algorithms leverage this observation to minimize the need for fine-grained communication.
+However, they typically must make this visible to programmers, who now must take care to insert their own memory barriers where appropriate, or otherwise express the way in which consistency relaxation can occur.
+
+The goal of this work is to apply the concept of flat-combining in a PGAS runtime to reduce the cost of maintaining globally consistent data structures in a generic way.
+To evaluate this, we leverage the Grappa runtime~\cite{Nelson:hotpar11-real}, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads.
+With enough concurrent threads, it is actually possible to maintain program order within each thread by blocking on accesses, but still have an opportunity to effectively combine many fine-grained synchronous operations for better performance.
+With the same flat-combining mechanism, multiple global data structures can be implemented efficiently, even with extremely simple locking schemes.
+
+% lay out the upcoming sections of the paper...
+The next section will describe in more detail the Grappa runtime system that is used to implement flat combining for distributed memory machines.
+Then we explain the flat combining mechanism in more depth and describe how it maps to a PGAS model.
+Next we explain how several data structures are implemented in this framework and show how they perform on simple throughput workloads.
+Finally, we evaluate how they affect performance when integrated into a couple application kernels.

### 2013-05-22 doc/pgas13/grappa.tex
add to intro, include Grappa and Flat Combining sections
rework Grappa section a tad.

@@
-One of the persistent challenges with PGAS languages is that existing shared memory applications typically do  fine-grained operations, so naively translating them results in poor performance because many operations must be done via remote communication.
-Many techniques exist, therefore, to help come up with better ways to partition data, schedule computation, and where necessary, do communication in bulk to best utilize network resources.
+\section{Grappa}
+Existing shared memory applications typically do fine-grained operations on data, assuming it is all cheap to access.
+Naively translating these applications to PGAS languages often results in poor performance because many operations must then be done via fine-grained remote communication.
+Much research has been done to improve performance of PGAS implementations by coming up with better ways to partition data and schedule computation, and do communication in bulk to best utilize network resources.
+For regular workloads, most pointedly dense linear algebra, much of this can now be automatically optimized generically.
 
 Irregular applications, however, are characterized by being highly dependent on the shape of the data, making it difficult to statically partition, causing load imbalance, and requiring a majority of fine-grained random accesses.
+The goal of the Grappa runtime is to provide a PGAS implementation that has been designed from the ground up to handle such applications.

### 2013-05-24 doc/pgas13/flat-combining.tex
add to intro, include Grappa and Flat Combining sections
work on flat-combining section, add todos, fix typo in intro

@@
+\section{Flat Combining}
+
 Flat combining allows threads to cooperate when accessing a synchronized shared data structure to get greater overall throughput than even each of them operating serially.
-The trick, however, is designing a way for the threads to communicate their intentions to each other without incurring the same problem they had before.
+Careful engineering must be employed to implement a mechanism for cooperation that does not introduce the same synchronization issues as the original contended data structure.
+However, the same delegation mechanism can be reused in any data structure protected by a single global lock.
+
+\subsection{Physically-shared memory}
+
+The benefits of flat-combining can be broken down into three components: improved locality, reduced synchronization, and data structure-specific optimization.
+Having a single thread bound to a particular core do a string of accesses trivially results in a lower cache miss rate than having multiple threads on multiple cores bring the data structure into cache each in turn.
+
+The story for reducing synchronization is more complicated.
+Locks fail to scale primarily because as contention increases, there will be more failed synchronization attempts (failed \emph{compare and swap}, or \emph{CAS} operations), which have to be retried until they succeed.
+Instead of accessing a data structure directly, threads can delegate their work to another core by adding their requests to the structure's publication list.
+However, in a naive implementation, they still need to perform a synchronization operation every time they add a request to ensure they do not collide with other threads.
+The solution proposed and implemented did not eliminate this problem directly.
+Each thread must pay the synchronization cost to insert a publication record into the list once, but after that, it may reuse the record for subsequent accesses without synchronizing.
+The insight is that threads typically perform many operations on the same data structure, so the synchronization cost is amortized.
+
+Scanning and combining...
+
+\subsection{Grappa Flat Combining Framework}
+
+In a physically-distributed setting, and in Grappa in particular, a number of different choices are made when implementing flat combining.
+
+Because memory is physically distributed, locality must be made explicit in the Grappa version of flat combining.
+
+In Grappa there are orders of magnitude more concurrent threads (``workers'') accessing shared data structures.
+Typically around a thousand are needed per core to tolerate the latency of remote operations, so in a cluster of machines, there are easily millions of workers.
+
+By design, each core in Grappa operates independently of the other cores.
+Workers are scheduled cooperatively, so each worker can automatically assume atomicity between remote calls.
+Therefore workers within a core can cooperate without any explicit synchronization operations.
+This allows them to publish requests to the local data structure proxy cheaply, so the publication list need not amortize the cost of synchronization.
+Instead, it can contain just the workers that are actually blocking on accesses to the data structure, eliminating the cost of detecting and removing unused publication records.

### 2013-05-24 doc/pgas13/grappa.tex
work on flat-combining section, add todos, fix typo in intro

@@
 
-Irregular applications, however, are characterized by being highly dependent on the shape of the data, making it difficult to statically partition, causing load imbalance, and requiring a majority of fine-grained random accesses.
+Irregular applications, on the other hand, are characterized by being highly dependent on the shape of the data, making it difficult to statically partition, causing load imbalance, and requiring a majority of fine-grained random accesses.
 The goal of the Grappa runtime is to provide a PGAS implementation that has been designed from the ground up to handle such applications.
+
+\subsection{Tasks}
+\todo{describe how tasks work}
+
+\subsection{Global Memory}
+\todo{explain how memory is partitioned}
+
+\subsection{Communication}
+\todo{remote delegate operations, aggregation}

### 2013-05-27 doc/pgas13/abstract.tex
Work on flat-combining section (also fix \TODO, add fc papers to bib)

TBD

% Structured concurrency (scheduler, cooperative multithreading + communication) enables hierarchical flat-combining to be implemented efficiently for a software PGAS shared-memory solution.

### 2013-05-27 doc/pgas13/flat-combining.tex
Work on flat-combining section (also fix \TODO, add fc papers to bib)

--- 2013-05-17 doc/pgas13/flat-combining.tex
@@
+\section{Flat Combining}
+
 Flat combining allows threads to cooperate when accessing a synchronized shared data structure to get greater overall throughput than even each of them operating serially.
-The trick, however, is designing a way for the threads to communicate their intentions to each other without incurring the same problem they had before.
+Careful engineering must be employed to implement a mechanism for cooperation that does not introduce the same synchronization issues as the original contended data structure.
+However, the same delegation mechanism can be reused in any data structure protected by a single global lock.
+
+\subsection{Physically-shared memory}
+
+The benefits of flat-combining can be broken down into three components: improved locality, reduced synchronization, and data structure-specific optimization.
+Having a single thread bound to a particular core do a string of accesses trivially results in a lower cache miss rate than having multiple threads on multiple cores bring the data structure into cache each in turn.
+
+The story for reducing synchronization is more complicated.
+The problem with having many threads attempt to acquire a lock is that as contention increases, there will be more failed synchronization attempts (failed \emph{compare and swap}, or \emph{CAS} operations), which must be retried until they succeed.
+Instead of having most of the threads repeated fail on acquiring the lock, when threads observe that the lock has already been taken, they can delegate their work to the core with the lock by adding their requests to the structure's publication list.
+However, in a naive implementation, this operation still needs to perform synchronization every time a request is added to ensure inserts do not collide.
+There is no way to avoid all synchronization, however, the cost may be amortized over many operations on the same data structure.
+Each thread must pay the synchronization cost to insert a publication record into the list once, but after that, it may reuse the record for subsequent accesses without synchronizing.
+
+The final piece of the flat combining story is how operations are semantically combined using knowledge of how data structure accesses compose.
+All of the threads that did not acquire the lock block on their request until it is satisfied.
+Whichever thread did obtain the lock becomes the combiner.
+It iterates over the publication list, merging each non-empty publication record into a combined operation using a combining function defined by the data structure.
+When the pass is complete, the combined operation is applied to the actual data structure.
+The combiner thread then takes the aggregated results of the combined operation and distributes them to the correct requests in the publication list.
+When each request is fulfilled, the thread blocking on it wakes and continues, and finally the combiner thread releases the lock.
+If any publication requests remain, which could have come from threads that published their request after the combining thread had passed, one of them acquires the lock and becomes the new combiner.
+
+\TODO{Add example of doing flat-combining on a global stack, discuss how pushes/pops are flattened.}
+
+\subsection{In Grappa}
+
+In a PGAS setting, and in Grappa in particular, a number of different choices are made when implementing flat combining.
+Because memory is physically distributed, locality must be made explicit.
+In addition, in Grappa there are orders of magnitude more concurrent threads (``workers'') accessing shared data structures.
+Typically around a thousand are needed per core to tolerate the latency of remote operations, so in a cluster of machines, there are easily millions of workers.
+Therefore, a different scheme for managing many threads is necessary.
+
+The cost of synchronizing remotely on a distributed data structure drives the design of the Grappa flat-combining.
+Instead of all of the workers on a core sending messages to do remote synchronization independently, flat-combining can allow all of the workers to share a single remote synchronization operation.
+In a way, the Grappa runtime is already reducing synchronization traffic when it aggregates small messages that are all going to the same destination.
+However, it doesn't know what the operations are doing, so it must still serialize, pack, transmit, perform the operation on the remote side, and send a response for each message.
+Flat combining is about ``teaching'' it how to more efficiently perform many operations remotely in certain special cases.
+
+Flat-combined global data structures have a local publication list on each core which serve as proxies to the global structure and are used to combine requests.
+Workers add their requests to the local publication list by issuing operations on their local proxy and then block waiting for their requests to be satisfied.
+Similar to the shared-memory case, one requesting worker is chosen to combine requests and perform the combined operation globally.
+This worker walks the local publication list, combining requests into a local request object as it goes.
+If it is possible to service a request locally, as it is on some data structures such as matching pushes and pops to a stack, the combiner can satisfy the requests and wake the correct workers even before synchronizing globally.
+The combiner then issues the combined data structure operation globally.
+In most cases, this means communicating with whichever core holds the ``lock-protected'' field that maintains sequential consistency.
+In the case of a global stack, this is the ``master'' \emph{top} pointer which must be incremented on pushes and decremented on pops.
+On the core where global synchronization is occurring, there must be synchronization between the multiple synchronization requests arriving from other cores.
+
+At this point we have traded off one remote synchronization per operation for an additional local synchronization per operation and some number fewer remote synchronizations (depending on the amount of combining that occurs).
+This would likely result in a net benefit due to local synchronization being inherently cheaper.
+However, because of the way the Grappa runtime schedules computation, it is actually possible to get the correct mutual exclusion/atomicity without any additional synchronization overhead.
+
+By design, each core in Grappa operates independently of the other cores.
+Workers are scheduled cooperatively, so each worker can assume atomicity until it performs an operation that yields to the scheduler, such as a remote call.
+Therefore workers within a core can cooperate without any explicit synchronization operations requiring memory fences.
+This allows them to publish requests to the local  proxy very inexpensively, eliminating the need for a complex synchronization-amortizing queue.
+This has the added benefit that there will be no unused publication records, which could have become a problem with thousands of workers.
+Further, remote operations are processed through the same cooperative-multithreading scheduler on the remote side, so combined operations are trivially serialized and atomic.
+
+\paragraph{Correctness}
+\TODO{Make argument about how flat-combining does not break things: sync operations still serialized on head, no order between things happening in different workers/cores unless otherwise synchronized, serializability not broken by satisfying requests locally.}
+
+\section{FC Framework}
+\TODO{Describe how to implement FC data structure}.

### 2013-05-27 doc/pgas13/flat-combining.tex
Add 'correctness' argument, swap two paragraphs in FC section

@@
 
+The cost of performing synchronization remotely on a distributed data structure motivates applying the flat-combining paradigm to Grappa data structures.
+Instead of all of the workers on a core sending messages to do remote synchronization independently, flat-combining can allow all of the workers to share a single remote synchronization operation.
+In a way, the Grappa runtime is already reducing synchronization traffic when it aggregates small messages that are all going to the same destination.
+However, it doesn't know what the operations are doing, so it must still serialize, pack, transmit, perform the operation on the remote side, and send a response for each message.
+Flat combining is about ``teaching'' it how to more efficiently perform many operations remotely in certain special cases.
+
 In a PGAS setting, and in Grappa in particular, a number of different choices are made when implementing flat combining.
@@
 Therefore, a different scheme for managing many threads is necessary.
-
-The cost of synchronizing remotely on a distributed data structure drives the design of the Grappa flat-combining.
-Instead of all of the workers on a core sending messages to do remote synchronization independently, flat-combining can allow all of the workers to share a single remote synchronization operation.
-In a way, the Grappa runtime is already reducing synchronization traffic when it aggregates small messages that are all going to the same destination.
-However, it doesn't know what the operations are doing, so it must still serialize, pack, transmit, perform the operation on the remote side, and send a response for each message.
-Flat combining is about ``teaching'' it how to more efficiently perform many operations remotely in certain special cases.
 
@@
 \paragraph{Correctness}
-\TODO{Make argument about how flat-combining does not break things: sync operations still serialized on head, no order between things happening in different workers/cores unless otherwise synchronized, serializability not broken by satisfying requests locally.}
+
+It may not be immediately obvious why this two-level version of flat combining preserves sequential consistency.
+The combining operations that are performed locally among requests on a single core must preserve the illusion of sequential order; this is unchanged from shared-memory flat-combining.
+When combined operations are applied globally, they are serialized at the ``master'' core for a given synchronization context.
+The global order is essentially the concatenation of each core's serialized updates.
+In the case where operations are satisfied locally (by matching up with other local operations), by definition these must be independent of other operations, so they can conceptually be placed anywhere in the global ordering (or even ignored completely in the global ordering).
+
+In the spirit of the Data-Race-Free-0 Model (guaranteeing sequential consistency for data-race free programs), operations on the global data structure are considered to be made atomic by a conceptual global ``lock'' on the data structure.
+Atomicity and freedom from data races in the case of multiple data structures and other synchronization must be guaranteed externally, just as in any multithreaded shared-memory program.
 

### 2013-05-28 doc/pgas13/flat-combining.tex
Add description of Grappa flat-combining framework

@@
 \paragraph{Correctness}
+\TODO{(are these arguments correct??)}
 
@@
 \section{FC Framework}
-\TODO{Describe how to implement FC data structure}.
+In order to leverage the flat-combining paradigm in Grappa, we implemented a generic flat-combining framework which can be used to improve performance for a number of global data structures.
+The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress.
+When hooking into the framework, each data structure implementation need only define how to combine operations and globally commit them.
+
+The Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine.
+In the original work, an explicit publication list is built up, and when the combiner scans over the list, it has a structure that it uses to accumulate requests in a logical way.
+In the case of the stack, for instance, it is a thread-local stack, which it can use to immediately match up pushes and pops.
+In the Grappa implementation, each worker's actions are atomic anyway, so rather than creating an explicit request list, the combiner object (or \emph{proxy}) is instantiated ahead of time for the data structure, and each worker does its own local combining into that structure before blocking.
+This reduces the space blowout of a large dynamically-sized publication list (with associated caching issues), and does not overly affect the expression of combining.
+
+Using this style, each global data structure must define the following:
+\begin{enumerate}
+  \item A local \emph{proxy} data structure for doing combining.
+  \item \emph{Combining methods} that operate on the local proxy.
+  \item A \emph{sync} method that globally commits the proxy's state.
+\end{enumerate}
+
+Being conservative about SC, workers typically block when issuing these requests, so the framework must also take care of delivering results and waking the correct workers.
+The flat-combining framework is responsible for ensuring that all of the combined operations eventually get committed.
+There are a number of ways progress could be guaranteed, but one of the simplest is to take an approach in the spirit of the original flat-combining, which is to ensure that as long as there are any outstanding requests, there is at least one worker committing a combined operation.
+When that combined operation finishes, if there are still outstanding requests that have built up in the meantime, another blocked worker is chosen to do another combined synchronization.
+
+One could imagine other synchronization policies that would ensure progress but allow more combining to occur before synchronizing.
+For instance, local combining could continue until all workers are blocked.
+However, this would require the flat-combining framework to track the activity of the workers on a core, and additionally would add latency unless the system saved enough active workers to tolerate the latency of the global commit action.
+As it is, the one-in-flight policy implemented performs well, and does not require any coupling with the Grappa scheduler.
+Exploration of other policies is left for future work.
+
+The flat combining framework transparently wraps the \emph{proxy} object.
+Behind this layer of indirection, the flat-combining framework manages multiple copies of the proxy so that, as discussed above, there is always at least one being committed and one to receive additional combining operations.
+
+% explain simple non-combined global data structures in Grappa section...
+
+\subsection{Global Stack}
+To demonstrate the flat-combining framework, we revisit the Global Stack example from earlier.
+Recall that the ``master'' core owns the \texttt{top} index, which indicates where the top of the stack currently is in the stack's allocated global storage.
+In the non-combining case, each call to \texttt{push} must do an atomic increment of \texttt{top} to claim storage, then write to that location, then finally return.

### 2013-06-04 doc/pgas13/evaluation.tex
First figures in place! (queue & stack)

\section{Evaluation}
To evaluate the impact flat combining has on the global data structures, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and further measured their impact on performance of two simple graph benchmarks.

\subsection{Data Structure Throughput}
First we measured the performance of the global data structures on extremely simple workloads to understand their raw performance.

\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
  \centering
  \includegraphics[width=\textwidth]{data/plots/queue_perf.pdf}
  \caption{Queue performance.}
  \label{fig:queue}
  \end{subfigure}%
  \hspace{0.05\textwidth}
  % ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
  %(or a blank line to force the subfigure onto a new line)
  \begin{subfigure}[b]{0.45\textwidth}
  \centering
  \includegraphics[width=\textwidth]{data/plots/stack_perf.pdf}
  \caption{Stack performance.}
  \label{fig:stack}
  \end{subfigure}
  \caption{Raw performance of global data structures.}\label{fig:datastructs}
\end{figure*}

\paragraph{Queue and Stack}
The GlobalQueue and GlobalStack have very similar implementations in terms of how they are synchronized. Both benefit greatly from flat combining. This is because only one round-trip message

However, due to the nature of concurrent stacks, the stack is able to match up pushes and pops locally. This leads to performance on an even mix of pushes and pops far outstripping the case where all pushes are performed due to requiring vastly less communication.

### 2013-06-04 doc/pgas13/flat-combining.tex
First draft of grappa,fc_shared_mem, and non_fc_stack

@@
 \subsection{Physically-shared memory}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.5\textwidth]{figs/fc_shared_mem.pdf}
+  \caption{Flat combining in shared memory. \TODO{walkthrough...}}
+  \label{fig:fc_shared_mem}
+\end{figure}
 
@@
 \subsection{In Grappa}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.5\textwidth]{figs/stack.pdf}
+  \caption{\emph{GlobalStack in Grappa.} Each access must synchronize with the head node (blocking if another worker is performing an operation), put or get its value in the stack's storage, release the lock on the master, and finally continue.}
+  \label{fig:stack}
+\end{figure}
 
@@
 The flat combining framework transparently wraps the \emph{proxy} object.
-Behind this layer of indirection, the flat-combining framework manages multiple copies of the proxy so that, as discussed above, there is always at least one being committed and one to receive additional combining operations.
+Behind this layer of indirection, it manages multiple copies of the proxy so that, as discussed above, there is always at least one being committed and one to receive additional combining operations.
 

### 2013-06-04 doc/pgas13/evaluation.tex
Add Hash{Set,Map} plots.

@@
   \centering
-  \includegraphics[width=\textwidth]{data/plots/queue_perf.pdf}
-  \caption{Queue performance.}
-  \label{fig:queue}
+  \includegraphics[width=\textwidth]{data/plots/stack_perf.pdf}
+  \caption{Stack performance.}
+  \label{fig:stack}
   \end{subfigure}%
@@
   \centering
-  \includegraphics[width=\textwidth]{data/plots/stack_perf.pdf}
-  \caption{Stack performance.}
-  \label{fig:stack}
+  \includegraphics[width=\textwidth]{data/plots/queue_perf.pdf}
+  \caption{Queue performance.}
+  \label{fig:queue}
   \end{subfigure}
-  \caption{Raw performance of global data structures.}\label{fig:datastructs}
+
+  \begin{subfigure}[b]{0.45\textwidth}
+  \centering
+  \includegraphics[width=\textwidth]{data/plots/hashset_perf.pdf}
+  \caption{GlobalHashSet performance.}
+  \label{fig:hashset}
+  \end{subfigure}
+  %
+  \hspace{0.05\textwidth}
+  %
+  \begin{subfigure}[b]{0.45\textwidth}
+  \centering
+  \includegraphics[width=\textwidth]{data/plots/hashmap_perf.pdf}
+  \caption{GlobalHashMap performance.}
+  \label{fig:hashmap}
+  \end{subfigure}%
+  %
+  \caption{Raw performance of global data structures.
+    \TODO{describe results.}
+    \TODO{either all with error bars or none.}
+  }\label{fig:datastructs}
 \end{figure*}

### 2013-06-04 doc/pgas13/data/plot.cc.R

@@ -40,7 +40,7 @@ g <- ggplot(subset(d, cc_hash_size <= 16384 & scale == 26 & ppn == 16), aes(
   # ylab("")+
   expand_limits(y=0)+
   my_theme
-ggsave(plot=g, filename="plots/cc_mess.pdf", scale=1.4)
+# ggsave(plot=g, filename="plots/cc_mess.pdf", scale=1.4)
 
 g <- ggplot(subset(d, 
     cc_hash_size == 16384 & scale == 26 & ppn == 16
@@ -57,11 +57,12 @@ g <- ggplot(subset(d,
   # geom_smooth(aes(linetype=fc_version),stat=stat_smooth(se=FALSE))+
   # geom_smooth(aes(linetype=fc_version),fill=NA)+
   geom_line()+
+  xlab("Nodes")+
   # geom_text(size=2,hjust=-0.2,vjust=1)+
   # facet_grid(scales="free", labeller=label_pretty)+
   ylab("MTEPS")+expand_limits(y=0)+
   my_theme
-ggsave(plot=g, filename="plots/cc_perf.pdf", scale=1.4)
+ggsave(plot=g, filename="plots/cc_perf.pdf", width=8, height=6)
 
 # g <- ggplot(d.m, aes(
 #     x=cc_concurrent_roots,
@@ -94,7 +95,7 @@ g.t <- ggplot(subset(d.t, cc_hash_size == 16384 & ppn == 16),
   facet_grid(scale+cc_hash_size+fc_version~nnode+ppn, labeller=label_pretty)+
   geom_bar(stat="identity")+
   my_theme
-ggsave(plot=g.t, filename="plots/cc_times.pdf", width=16, height=10)
+# ggsave(plot=g.t, filename="plots/cc_times.pdf", width=16, height=10)
 
 d.stat <- melt(d, measure=c('mteps', 'ce_remote_completions', 'DelegateStats_ops', 'gce_flats'))
 g.stat <- ggplot(subset(d.stat, scale == 26 & fc_version != 'async'), aes(
@@ -110,7 +111,7 @@ g.stat <- ggplot(subset(d.stat, scale == 26 & fc_version != 'async'), aes(
   # geom_smooth(fill=NA)+
   facet_grid(~scale~variable~nnode, labeller=label_pretty, scales="free")+
   ylab("")+expand_limits(y=0)+my_theme
[…]

### 2013-06-04 doc/pgas13/data/plot.hashset.R

@@ -34,7 +34,7 @@ d$ops_per_msg <- with(d, (hashset_insert_ops+hashset_lookup_ops)/(hashset_insert
 d$throughput <- with(d, (hashset_insert_ops+hashset_lookup_ops)/trial_time_mean)
 
 d$fc_version <- sapply(paste('v',d$flat_combining,d$insert_async,sep=''),switch,
-  v0NA='none', v1NA='fc', v11='async', v10='fc', v00='none', v01='?'
+  v0NA='none', v1NA='local', v11='async', v10='local', v00='none', v01='?'
 )
 
 d.melt <- melt(d, measure=c("ops_per_msg", "throughput"))
@@ -50,7 +50,7 @@ g <- ggplot(subset(d.melt,
     label=trial_time_mean,
     linetype=fc_version,
   ))+
-  geom_point()+
+  # geom_point()+
   # geom_text()+
   geom_line()+
   # geom_smooth(aes(linetype=fc_version), fill=NA)+
@@ -68,22 +68,23 @@ gg <- ggplot(subset(d, log_nelems==28 & ppn==16 & num_starting_workers==8192
     x=nnode,
     y=throughput,
     color=fc_version,
-    group=x(fc_version,version,fraction_lookups,trial),
+    group=x(fc_version,version,fraction_lookups),
     shape=fraction_lookups,
     linetype=fraction_lookups,
   ))+
-  geom_point()+
-  geom_line()+
+  # geom_point()+
+  # geom_line()+
+  geom_smooth(size=1)+
   # facet_grid(log_nelems~., scales="free", labeller=label_pretty)+
   xlab("Nodes")+
-  scale_color_discrete(name="Combining mode:")+
+  scale_color_discrete(name="Flat Combining")+
   scale_linetype_discrete(name="Fraction lookups:")+
   scale_shape_discrete(name="Fraction lookups:")+
-  ylab("Throughput")+expand_limits(y=0)+my_theme+
-  theme(strip.text=element_text(size=rel(0.4)),
-        axis.text.x=element_text(size=rel(0.65)))
+  ylab("Throughput (ops/sec)")+expand_limits(y=0)+my_theme+
+  theme(strip.text=element_text(size=rel(0.7)),
+        axis.text.x=element_text(size=rel(0.85)))
 
-ggsave(plot=gg, filename="plots/hashset_perf.pdf", width=8, height=6)
[…]

### 2013-06-04 doc/pgas13/data/plot.queue.R

@@ -44,16 +44,16 @@ gg <- ggplot(subset(d, log_nelems==28 & ppn==16 & num_starting_workers==2048
     y=throughput,
     color=fc_version,
     group=x(fc_version,version,fraction_push,trial),
-    shape=fraction_push,
+    shape=fc_version,
     linetype=fraction_push,
   ))+
-  geom_point()+
-  geom_line()+
+  geom_point(size=3)+
+  geom_line(size=1)+
   # facet_grid(log_nelems~., scales="free", labeller=label_pretty)+
   xlab("Nodes")+scale_x_continuous(breaks=c(0,8,16,32,48,64))+
   scale_color_discrete(name="Flat Combining")+
+  scale_shape_discrete(name="Flat Combining")+
   scale_linetype_discrete(name="Fraction pushes")+
-  scale_shape_discrete(name="Fraction pushes")+
   ylab("Throughput (ops/sec)")+expand_limits(y=0)+my_theme+
   theme(strip.text=element_text(size=rel(0.4)),
         axis.text.x=element_text(size=rel(0.75)))

### 2013-06-04 doc/pgas13/data/plot.stack.R

@@ -50,15 +50,15 @@ g <- ggplot(subset(d, log_nelems==28 & ppn==16 & num_starting_workers==2048
     y=throughput,
     color=fc_version,
     group=x(fc_version,version,fraction_push),
-    shape=fraction_push,
+    shape=fc_version,
     linetype=fraction_push,
   ))+
-  geom_point()+
-  geom_line()+
+  geom_point(size=3)+
+  geom_line(size=1)+
   xlab("Nodes")+scale_x_continuous(breaks=c(0,8,16,32,48,64))+
   scale_color_discrete(name="Flat Combining")+
+  scale_shape_discrete(name="Flat Combining")+
   scale_linetype_discrete(name="Fraction pushes")+
-  scale_shape_discrete(name="Fraction pushes")+
   # facet_grid(.~fraction_push, scales="free", labeller=label_pretty)+
   ylab("Throughput (ops/sec)")+expand_limits(y=0)+my_theme+
   theme(strip.text=element_text(size=rel(0.4)),

### 2013-06-05 doc/pgas13/evaluation.tex
Add plots for BFS and CC performance.

@@
 This leads to performance on an even mix of pushes and pops far outstripping the case where all pushes are performed due to requiring vastly less communication.
+
+\paragraph{HashSet and HashMap}
+The GlobalHashSet and GlobalHashMap have the same synchronization strategy (serialization happens at each hashed location).
+
+\subsection{Application Kernel Performance}
+The impetus for this work was that naive implementations of standard data structures are insufficient for use in high-performance kernels.
+
+\paragraph{Breadth-First Search}
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=0.5\textwidth]{data/plots/bfs_perf.pdf}
+  \caption{BFS on a Graph500-spec graph of scale 26 (64 million vertices, 1 billion edges), with the direction-optimizing BFS algorithm.
+Performance is shown in millions of Traversed Edges Per Second (MTEPS).}
+  \label{fig:bfs_perf}
+\end{figure}
+
+\paragraph{Connected Components}
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=0.5\textwidth]{data/plots/cc_perf.pdf}
+  \caption{Connected Components on the same scale 26 Graph500 graph. Performance is measured in TEPS, parameterized only by the total number of edges in the graph, regardless of the algorithm used.}
+  \label{fig:cc_perf}
+\end{figure}

### 2013-06-05 doc/pgas13/data/plot.bfs.R

@@ -4,21 +4,21 @@ source("common.R")
 
 q2 <- "select * from bfs where bfs_version like 'beamer%'"
 
-d <- db(q2, c("nnode", "num_starting_workers", "aggregator_autoflush_ticks", "periodic_poll_ticks"))
+d <- db(q2, c("nnode", "periodic_poll_ticks", "flat_combining"))
 d$global_vector_push_ops_per_msg <- with(d, global_vector_push_ops/global_vector_push_msgs)
 
 d$version_combining <- paste(d$bfs_version,":flat<",d$flat_combining,">",sep="")
 d$mteps <- d$max_teps/1e6
 
-d <- melt(d, measure=c("global_vector_push_ops_per_msg", "mteps")) 
+d.melt <- melt(d, measure=c("global_vector_push_ops_per_msg", "mteps")) 
 
-g <- ggplot(subset(d,
-    ppn == 16
+g <- ggplot(subset(d.melt,
+    ppn == 16 & scale == 26
   ), aes(
     x=num_starting_workers,
     y=value,
     color=x(version_combining,ppn),
-    shape=aggregator_autoflush_ticks,
+    # shape=aggregator_autoflush_ticks,
     group=x(version_combining,ppn)
     # label=nnode~ppn,
   ))+
@@ -29,26 +29,39 @@ g <- ggplot(subset(d,
   ylab("")+
   expand_limits(y=0)+
   my_theme
+ggsave(plot=g, filename="plots/bfs_mess.pdf", width=10, height=8)
 
-ggsave(plot=g, filename="plots/bfs_mess.pdf", scale=1.2)
+d$nw <- d$num_starting_workers
+
+d$fc_version <- sapply(paste(d$bfs_version,d$flat_combining,sep='_'),switch,
+  beamer_0='async', beamer_1='async', beamer_queue_0='none', beamer_queue_1='full'
+)
+# d$fc_version <- paste(d$bfs_version,d$flat_combining,sep='_')
+# d$fc_version
 
 g <- ggplot(subset(d,
-    ppn == 16 & scale == 26 & num_starting_workers == 1024
+    ppn == 16 & scale == 26
+    & (num_starting_workers == 1024 & grepl('full|none', fc_version) 
+      | num_starting_workers >= 256 & grepl('async', fc_version))
+    & aggregator_autoflush_ticks > 100000
   ), aes(
     x=nnode,
-    y=value,
-    color=x(version_combining,ppn),
[…]

### 2013-06-05 doc/pgas13/data/plot.cc.R

@@ -7,7 +7,6 @@ d <- db(
    ",
   c("nnode", "ppn",
     "num_starting_workers",
-    "aggregator_autoflush_ticks",
     "periodic_poll_ticks",
     "flat_combining",
     "cc_concurrent_roots"
@@ -20,7 +19,7 @@ d$nroots <- d$cc_concurrent_roots
 d$gce_flats <- with(d, gce_total_remote_completions/gce_completions_sent)
 
 d$fc_version <- sapply(paste('v',d$flat_combining,d$cc_insert_async,sep=''),switch,
-  v0NA='none', v1NA='fc', v11='async', v10='fc', v00='none'
+  v0NA='none', v1NA='fc', v11='async', v10='fc', v00='none', '??'
 )
 
 g <- ggplot(subset(d, cc_hash_size <= 16384 & scale == 26 & ppn == 16), aes(
@@ -28,41 +27,49 @@ g <- ggplot(subset(d, cc_hash_size <= 16384 & scale == 26 & ppn == 16), aes(
     y=mteps,
     color=nroots,
     shape=fc_version,
-    # label=trial,
-    group=x(cc_concurrent_roots,fc_version)
+    label=x(aggregator_autoflush_ticks/1e6,num_starting_workers),
+    group=x(cc_concurrent_roots,fc_version,aggregator_autoflush_ticks),
+    linetype=fc_version,
   ))+
   geom_point()+
   # geom_smooth(aes(linetype=fc_version),stat=stat_smooth(se=FALSE))+
   # geom_smooth(aes(linetype=fc_version),fill=NA)+
-  geom_line(aes(linetype=fc_version))+
-  # geom_text(size=2,hjust=-0.2,vjust=1)+
+  geom_line()+
+  geom_text(size=2,hjust=-0.2,vjust=1)+
   facet_grid(~cc_hash_size~ppn~scale~nnode, scales="free", labeller=label_pretty)+
   # ylab("")+
   expand_limits(y=0)+
+  theme(strip.text=element_text(size=rel(0.4)),
+        axis.text.x=element_text(size=rel(0.75)))+
   my_theme
 # ggsave(plot=g, filename="plots/cc_mess.pdf", scale=1.4)
 
 g <- ggplot(subset(d, 
     cc_hash_size == 16384 & scale == 26 & ppn == 16
-    & num_starting_workers == 2048 & cc_concurrent_roots == 2048
[…]

### 2013-06-05 doc/pgas13/evaluation.tex
Add descriptions of diagrams, plots.

@@
   \includegraphics[width=\textwidth]{data/plots/stack_perf.pdf}
-  \caption{Stack performance.}
+  \caption{\emph{GlobalStack.} Even with Grappa's aggregation, the stack without combining completely fails to scale.
+With combining, both workloads scale out to 64 nodes.
+The workload with roughly even pushes and pops performs best because it can do matching locally.}
   \label{fig:stack}
@@
   \includegraphics[width=\textwidth]{data/plots/queue_perf.pdf}
-  \caption{Queue performance.}
+  \caption{\emph{GlobalQueue.} Same parameters as stack performance results.
+The queue is unable to do matching locally, but benefits from reducing the amount of synchronization that must globally serialize.
+The mixed workload performs worse because the current implementation serializes combined enqueue and dequeue operations.}
   \label{fig:queue}
@@
   \includegraphics[width=\textwidth]{data/plots/hashset_perf.pdf}
-  \caption{GlobalHashSet performance.}
+  \caption{\emph{GlobalHashSet.} Throughput workload inserting and looking up random keys in the range 0-1024 into a global hash set with 1024 cells.
+   Performance without combining scales better than the global queue or stack because synchronization happens at each hash cell, but drops off as number of destinations increases.
+Combining version is able to match inserts and lookups of the same key.}
+    \TODO{if time: add 'delete' operation, too (show that it doesn't affect correctness or performance)}
   \label{fig:hashset}
@@
   \includegraphics[width=\textwidth]{data/plots/hashmap_perf.pdf}
-  \caption{GlobalHashMap performance.}
+  \caption{\emph{GlobalHashMap.} Same workload as for the Set, but random integers as the values in the map.
+Performance matches that of the Set.} \TODO{try increasing size of value "payload", currently is just a tiny int64.}
   \label{fig:hashmap}
@@
   %
-  \caption{Raw performance of global data structures.
-    \TODO{describe results.}
-    \TODO{either all with error bars or none.}
+  \caption{
+    \emph{Raw performance of global data structures on simple throughput workloads.}
+    Results shown for workloads with all writes and a mix of writing and reading operations. All results are for 2048 workers per core and 16 cores per node.
+    \TODO{either all with error bars or none?}
+    \TODO{combine stack/queue \& set/map plots, no point in saying the same things twice.}
   }\label{fig:datastructs}
@@
 \paragraph{Queue and Stack}
-The GlobalQueue and GlobalStack have very similar implementations in terms of how they are synchronized. Both benefit greatly from flat combining. This is because only one round-trip message
+The GlobalQueue and GlobalStack have very similar implementations in terms of how they are synchronized.
+Both benefit greatly from flat combining.
+This is because only one round-trip message \TODO{...}
 
-However, due to the nature of concurrent stacks, the stack is able to match up pushes and pops locally.
+% However, the stack is able to match up pushes and pops locally.
 This leads to performance on an even mix of pushes and pops far outstripping the case where all pushes are performed due to requiring vastly less communication.
@@
 The GlobalHashSet and GlobalHashMap have the same synchronization strategy (serialization happens at each hashed location).
+\TODO{...}
 
@@
   \includegraphics[width=0.5\textwidth]{data/plots/bfs_perf.pdf}
-  \caption{BFS on a Graph500-spec graph of scale 26 (64 million vertices, 1 billion edges), with the direction-optimizing BFS algorithm.
-Performance is shown in millions of Traversed Edges Per Second (MTEPS).}
+  \caption{\emph{BFS} on a Graph500-spec graph of scale 26 (64 million vertices, 1 billion edges), with the direction-optimizing BFS algorithm.
+Performance is measured in millions of Traversed Edges Per Second (MTEPS).}
   \label{fig:bfs_perf}
@@
   \includegraphics[width=0.5\textwidth]{data/plots/cc_perf.pdf}
-  \caption{Connected Components on the same scale 26 Graph500 graph. Performance is measured in TEPS, parameterized only by the total number of edges in the graph, regardless of the algorithm used.}
+  \caption{\emph{Connected Components} on the same scale 26 Graph500 graph.
+Performance is measured in MTEPS (parameterized by the total number of edges in the graph, independent of the algorithm used).}
   \label{fig:cc_perf}

### 2013-06-05 doc/pgas13/flat-combining.tex
Add diagram of stack w/ FC, update one w/o
Add descriptions of diagrams, plots.

@@
   \includegraphics[width=0.5\textwidth]{figs/fc_shared_mem.pdf}
-  \caption{Flat combining in shared memory. \TODO{walkthrough...}}
+  \caption{\emph{Flat combining in shared memory.}
+    To access the shared stack, each thread adds its request to the publication list.
+Thread 4 acquires the stack's lock and becomes the combiner, while the remaining threads spin waiting for their request to be satisfied.
+The combiner walks the publication list from the head, matching up Thread 3's push with Thread 2's pop on its own private stack.
+The 2 remaining pushes are added to the shared stack, the top is updated, and finally Thread 4 releases the lock and continues its normal execution.
+  }
   \label{fig:fc_shared_mem}
@@
 
-\begin{figure}[ht]
+\begin{figure*}[t]
   \centering
-  \includegraphics[width=0.5\textwidth]{figs/stack.pdf}
-  \caption{\emph{GlobalStack in Grappa.} Each access must synchronize with the head node (blocking if another worker is performing an operation), put or get its value in the stack's storage, release the lock on the master, and finally continue.}
+  \begin{subfigure}[b]{0.43\textwidth}
+    \includegraphics[width=\textwidth]{figs/stack_nofc.pdf}
+    \caption{\emph{Without flat combining.}
+      To do its push, Worker 1 sends a message to synchronize with the master on Core 0, sends another message to write its value to the top of the stack, bumps the synchronized \texttt{top} pointer, and finally continues.
+Worker 2, and all other workers on all cores accessing the stack, must block at the master and wait for Worker 1 to complete its push before doing their operations.
+    }
+    \label{fig:stacknofc}
+  \end{subfigure}%
+  \hspace{0.05\textwidth}
+  \begin{subfigure}[b]{0.43\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{figs/stack_fc.pdf}
+    \caption{\emph{Local flat combining.}
+      Workers within the core add their requests to a local proxy object.
+Worker 3's pop matches with Worker 2's push, requiring no global communication.
+After combining locally, worker 1 and 4's pushes remain, so Worker 4 becomes the core's combiner, synchronizes with the master, adds both new values to the global stack, bumps the top pointer and releases the lock on master, and finally wakes Worker 1 and continues.
+    }
+    \label{fig:stackfc}
+  \end{subfigure}
+  \caption{\emph{Global stack in Grappa} with and without \emph{flat combining}.}
   \label{fig:stack}
-\end{figure}
+\end{figure*}
 

### 2013-06-05 doc/pgas13/grappa.tex
Work on flat-combining section (also fix \TODO, add fc papers to bib)
First draft of grappa,fc_shared_mem, and non_fc_stack
Add descriptions of diagrams, plots.

@@
 Much research has been done to improve performance of PGAS implementations by coming up with better ways to partition data and schedule computation, and do communication in bulk to best utilize network resources.
-For regular workloads, most pointedly dense linear algebra, much of this can now be automatically optimized generically.
+For workloads with regular access patterns, for instance dense linear algebra or structured grids, there are proven techniques that can automatically optimize communication effectively.
+\TODO{cite or remove prev.
+statement}
 
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.5\textwidth]{figs/grappa_system.pdf}
+  \caption{\emph{Grappa System Overview:}
+    Thousands of workers are multiplexed onto each core, context switching to tolerate the additional latency of aggregating remote operations.
+Each core has exclusive access to a slice of the global heap which is partitioned across all nodes, as well as core-private data and worker stacks.
+Tasks can be spawned and executed anywhere, but are bound to a worker in order to be executed.
+  }
+  \label{fig:system}
+\end{figure}
+
 \subsection{Tasks}
-\todo{describe how tasks work}
+\TODO{describe how tasks work}
 
 \subsection{Global Memory}
-\todo{explain how memory is partitioned}
+\TODO{explain how memory is partitioned}
 
 \subsection{Communication}
-\todo{remote delegate operations, aggregation}
+\TODO{remote delegate operations, aggregation}

### 2013-06-05 doc/pgas13/intro.tex
work on flat-combining section, add todos, fix typo in intro
Add descriptions of diagrams, plots.

@@
 The trick is coming up with a mechanism for delegating that has lower synchronization overhead than the original contention case.
-Prior work showed that a cleverly-implemented ``publication lists'' can allow multiple accessors to cooperate with minimal synchronization.
+Prior work showed that a cleverly-implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization.
 This same work makes the additional observation that, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually, which they dub \emph{flat combining}.
@@
 With enough concurrent threads, it is actually possible to maintain program order within each thread by blocking on accesses, but still have an opportunity to effectively combine many fine-grained synchronous operations for better performance.
-With the same flat-combining mechanism, multiple global data structures can be implemented efficiently, even with extremely simple locking schemes.
+With the same flat-combining mechanism, multiple global data structures can be implemented efficiently, even with simple locking schemes.
 
@@
 Next we explain how several data structures are implemented in this framework and show how they perform on simple throughput workloads.
-Finally, we evaluate how they affect performance when integrated into a couple application kernels.
+Finally, we evaluate how they affect performance when used in a simple application kernels.

### 2013-06-06 doc/pgas13/grappa.tex
First attempt at Grappa section. Still needs code examples?

@@
 \section{Grappa}
-Existing shared memory applications typically do fine-grained operations on data, assuming it is all cheap to access.
-Naively translating these applications to PGAS languages often results in poor performance because many operations must then be done via fine-grained remote communication.
-Much research has been done to improve performance of PGAS implementations by coming up with better ways to partition data and schedule computation, and do communication in bulk to best utilize network resources.
-For workloads with regular access patterns, for instance dense linear algebra or structured grids, there are proven techniques that can automatically optimize communication effectively.
-\TODO{cite or remove prev.
-statement}
 
-Irregular applications, on the other hand, are characterized by being highly dependent on the shape of the data, making it difficult to statically partition, causing load imbalance, and requiring a majority of fine-grained random accesses.
-The goal of the Grappa runtime is to provide a PGAS implementation that has been designed from the ground up to handle such applications.
+Irregular applications are characterized by having unpredictable data-dependent access patterns and poor spatial and temporal locality.
+Applications in this class, including data mining, graph analytics, and various learning algorithms, are becoming increasingly prevalent in high-performance computing.
+These programs typically perform many fine-grained accesses to disparate sources of data, which is a problem even at multicore scales, but is further exacerbated on distributed memory machines.
+It is often the case that naively porting an application to a PGAS system results in excessive communication and poor access patterns, but this class of applications defies typical optimization techniques such as data partitioning, shadow objects, and bulk-synchronous communication transformations.
+Luckily, applications in this class have another thing in common: abundant amounts of data parallelism.
+This parallelism can be exploited in a number of different ways to improve overall throughput of these applications.
 
-\begin{figure}[ht]
+% The goal of the Grappa runtime is to provide a PGAS implementation that has been designed from the ground up to leverage massive amounts of parallelism to achieve high performance on irregular applications.
+
+Grappa is a PGAS-style runtime for commodity clusters which has been designed from the ground up to achieve high performance on irregular applications.
+The key is latency tolerance---long-latency operations such as remote communication can be tolerated by switching to another concurrent thread of execution.
+Given abundant  concurrency, there are opportunities to increase throughput by sacrificing latency.
+In particular, throughput of random accesses to remote memory can be improved by delaying communication requests and aggregating them into larger packets.
+Highly-tuned implementations of irregular applications in shared-memory, PGAS and message-passing paradigms, typically end up implementing constructs to allow for aggregation and leverage data parallelism.
+The goal of Grappa is to provide a simple-to-use PGAS abstraction with mechanisms for expressing the concurrency needed to achieve high performance.
+
+The programming interface, implemented as a C++11 library, provides high-level operations to access and synchronize through global shared memory, task spawning capabilities and parallel loop constructs for expressing concurrency.
+In addition, the Grappa ``standard library'' includes functions to manipulate a global heap as well as provides several synchronized global data structures.
+The following sections will explain the execution model of Grappa and the current C++ programming interface.
+
+\begin{figure}[t]
   \centering
@@
 
-\subsection{Tasks}
-\TODO{describe how tasks work}
+\subsection{Tasks and Workers}
+Core to Grappa is the expression of concurrency.
+The Grappa runtime has a user-level threading system that uses an efficient context-switching mechanism with prefetching to allow it to scale up to thousands of threads on a single core with minimal increase in context-switch time.
+This is exposed at the programming level as a simple task-parallel model.
+A \emph{task} is simply a function object with some state and a function to execute.
+In the runtime, ``Worker'' threads pull these programmer-specified tasks off of a queue and execute them to completion.
+Tasks may block on remote accesses or synchronization operations.
+When tasks block, the worker thread executing them is simply suspended and consumes no computational resources until woken again by the completion of communication or synchronization.
+
+\subsection{Aggregated Communication}
+Communication in Grappa is achieved via \emph{active messages} in conjunction with bulk remote direct memory access (RDMA) operations.
+Various components of Grappa's runtime contribute different kinds of messages to move and modify remote data, perform synchronization, or spawn tasks.
+To make efficient use of the networks in high-performance systems, which perform better with larger messages (typically 64 KB), all communication in Grappa is sent through an aggregation layer that automatically accumulates messages to the same destination.
+When enough messages are ready to go to a particular destination node, they are serialized into a buffer and sent using RDMA.
 
 \subsection{Global Memory}
-\TODO{explain how memory is partitioned}
+In the PGAS style, Grappa has a notion of a global address space which is partitioned across the physical memories of the nodes in a cluster.
+Each core owns a portion of memory, which is further divided for three different uses:
+\begin{itemize}
+  \item Execution stacks for the core's Workers.
+  \item A core-local heap.
+  \item A slice of the global shared heap.
+\end{itemize}
+All of these can be addressed by any core in the system using a Grappa \texttt{GlobalAddress}, which encodes both the owning core and the address on that core.
+Additionally, addresses into the global heap are partitioned in a block-cyclic fashion, so that a large allocation is automatically distributed among many nodes.
+For irregular applications, this helps avoid hot spots and is typically sufficient for random access.
 
-\subsection{Communication}
-\TODO{remote delegate operations, aggregation}
+All accesses to memory owned by another core must be done by that core via a message.
+In this way, the runtime maintains strict isolation even between processes running on the same physical memory.
+At the programming level, however, this is hidden behind higher-level remote operations, which in Grappa are called ``delegates''.
+Figure~\ref{fig:system} shows an example of a delegate \texttt{read()} which blocks the calling task and sends a message to the owner, who sends a reply with the data and wakes the caller.

### 2013-06-06 doc/pgas13/flat-combining.tex
Overhaul of FC section.

@@
 
-Flat combining allows threads to cooperate when accessing a synchronized shared data structure to get greater overall throughput than even each of them operating serially.
-Careful engineering must be employed to implement a mechanism for cooperation that does not introduce the same synchronization issues as the original contended data structure.
+At the most basic level, the concept of flat combining is about enabling cooperation among threads rather than contention.
+The benefits can be broken down into three components: improved locality, reduced synchronization, and data structure-specific optimization.
+We will explore how this works in a traditional shared memory system, and then describe how the same concepts can be applied to distributed memory.
+
+% Flat combining allows threads to cooperate when accessing a synchronized shared data structure to get greater overall throughput than even each of them operating serially.
 However, the same delegation mechanism can be reused in any data structure protected by a single global lock.
@@
 
-The benefits of flat-combining can be broken down into three components: improved locality, reduced synchronization, and data structure-specific optimization.
-Having a single thread bound to a particular core do a string of accesses trivially results in a lower cache miss rate than having multiple threads on multiple cores bring the data structure into cache each in turn.
+One of the issues with shared data structures in a shared-memory multicore is that threads executing on different cores will force the hottest parts of the data structure to thrash between their caches.
+Having a single thread do all of the operations on the data structure allows it to keep everything in its cache.
 
-The story for reducing synchronization is more complicated.
-The problem with having many threads attempt to acquire a lock is that as contention increases, there will be more failed synchronization attempts (failed \emph{compare and swap}, or \emph{CAS} operations), which must be retried until they succeed.
-Instead of having most of the threads repeated fail on acquiring the lock, when threads observe that the lock has already been taken, they can delegate their work to the core with the lock by adding their requests to the structure's publication list.
-However, in a naive implementation, this operation still needs to perform synchronization every time a request is added to ensure inserts do not collide.
-There is no way to avoid all synchronization, however, the cost may be amortized over many operations on the same data structure.
-Each thread must pay the synchronization cost to insert a publication record into the list once, but after that, it may reuse the record for subsequent accesses without synchronizing.
+% Improving locality in a shared-memory multicore is achieved by 
+% Having a single thread bound to a particular core do a string of accesses trivially results in a lower cache miss rate than having multiple threads on multiple cores bring the data structure into cache each in turn.
 
-The final piece of the flat combining story is how operations are semantically combined using knowledge of how data structure accesses compose.
-All of the threads that did not acquire the lock block on their request until it is satisfied.
-Whichever thread did obtain the lock becomes the combiner.
-It iterates over the publication list, merging each non-empty publication record into a combined operation using a combining function defined by the data structure.
-When the pass is complete, the combined operation is applied to the actual data structure.
-The combiner thread then takes the aggregated results of the combined operation and distributes them to the correct requests in the publication list.
-When each request is fulfilled, the thread blocking on it wakes and continues, and finally the combiner thread releases the lock.
-If any publication requests remain, which could have come from threads that published their request after the combining thread had passed, one of them acquires the lock and becomes the new combiner.
+% Careful engineering must be employed to implement a mechanism for cooperation that does not introduce the same synchronization issues as the original contended data structure.
 
-\TODO{Add example of doing flat-combining on a global stack, discuss how pushes/pops are flattened.}
+Reduced synchronization comes from delegating work to another core.
+To illustrate the issue, consider the shared synchronous queue, shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock.
+Without flat combining, whenever a thread attempts to push something on the stack, it must acquire the lock, put its value into the storage array, bump the top, and then release the lock.
+When many threads contend for the lock, all but one will fail and will have to retry.
+Each try forces a memory fence for consistency and consumes bandwidth, and as the number of threads increases, the fraction of successes plummets.
+Under flat combining, instead, threads add requests to the publication list.
+They each try to acquire the lock, and the one that succeeds becomes the combiner.
+Instead of retrying, the rest spin waiting for their request to be filled.
+The combiner walks the publication list, performs all of the requests, and when done, releases the lock.
+This has greatly reduced the synchronization on the stack's lock, but has introduced a new point of synchronization---threads must synchronize to add to the publication list.
+However, if a thread performs multiple operations on the stack, it can leave its publication record in the list and amortize the synchronization cost.
+
+The above example of delegation is compelling in itself.
+However, the crux of the prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently than separately.
+As the combiner walks the publication list, it merges each non-empty publication record into a combined operation.
+In the case of the stack example shown in Figure~\ref{fig:fc_shared_mem}, as it walks the list, Thread 4 keeps track of the operations on its own temporary stack.
+When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills each of their records and allows each of them to proceed.
+After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the remaining two elements are pushed onto the top of the stack, and the requesting threads' records are marked as complete.
+In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be combined for greater performance.
 
@@
 
-The cost of performing synchronization remotely on a distributed data structure motivates applying the flat-combining paradigm to Grappa data structures.
-Instead of all of the workers on a core sending messages to do remote synchronization independently, flat-combining can allow all of the workers to share a single remote synchronization operation.
-In a way, the Grappa runtime is already reducing synchronization traffic when it aggregates small messages that are all going to the same destination.
-However, it doesn't know what the operations are doing, so it must still serialize, pack, transmit, perform the operation on the remote side, and send a response for each message.
-Flat combining is about ``teaching'' it how to more efficiently perform many operations remotely in certain special cases.
+In a PGAS setting, and in Grappa in particular, the cost of global synchronization and the amount of concurrency is even greater than in shared memory.
+With thousands of workers per core, in a reasonably sized cluster there are easily millions of workers.
+This presents an opportunity for flat combining to pay off greatly, but also poses new challenges.
 
-In a PGAS setting, and in Grappa in particular, a number of different choices are made when implementing flat combining.
+% a number of different choices are made when implementing flat combining.
 Because memory is physically distributed, locality must be made explicit.
@@
 
-Flat-combined global data structures have a local publication list on each core which serve as proxies to the global structure and are used to combine requests.
-Workers add their requests to the local publication list by issuing operations on their local proxy and then block waiting for their requests to be satisfied.
-Similar to the shared-memory case, one requesting worker is chosen to combine requests and perform the combined operation globally.
-This worker walks the local publication list, combining requests into a local request object as it goes.
-If it is possible to service a request locally, as it is on some data structures such as matching pushes and pops to a stack, the combiner can satisfy the requests and wake the correct workers even before synchronizing globally.
-The combiner then issues the combined data structure operation globally.
-In most cases, this means communicating with whichever core holds the ``lock-protected'' field that maintains sequential consistency.
+To illustrate how flat combining can be applied to Grappa, we must first describe what a global data structure might look like.
+Figure~\ref{fig:stacknofc} shows a simple PGAS translation of the shared-memory stack from earlier.
+A storage array is allocated out of the global heap, so its elements are striped across all the cores in the system.
+One core is designated the ``master'' to enforce global synchronization, and holds the elements of the data structure that all concurrent accessors must agree on, in this case, the \texttt{top} pointer.
+
+All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object.
+To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read()} delegate described earlier, blocks the task until complete.
+To do a \texttt{push}, the task must send a message to the master to acquire the lock.
+If it succeeds, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker.
+All other workers performing operations block at the first message until the lock is released.
+Because of Grappa's user-level blocking mechanisms, blocked requests do not have to retry or busy-wait.
+However, all thousand workers on each core all perform this synchronization and serialize on the master core.
+
+With flat combining, instead of all workers on a core sending independent messages to do remote synchronization, they can do combining locally and share a single remote synchronization operation. 
+As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and as bonus, allows for data structure-specific combining to magnify the benefits even further.
+
+% allows synchronization to happen in parallel on all the cores, 
+% Flat combining in Grappa is about allowing synchronization and  distributing synchronization 
+% In a way, Grappa's automatic message aggregation is already trying to reduce synchronization traffic; however, the generic mechanism for serializing and packing has a cost, and in the end, the operations must still serialize on the master.
+% Flat combining is about ``teaching'' it how to more efficiently perform many operations remotely in certain special cases.
+
+% In order for workers to cooperate locally, a local ``proxy'' object is allocated in the core-private heap. This proxy serves as the
+
+When a flat-combined global data structure is created, each core allocates a local proxy object in it's core-local heap.
+This proxy serves the role of the publication list, allowing workers within a core to coordinate their efforts.
+Conceptually, workers add their requests to the local publication list, and one is chosen to do the combining.
+However, because workers are scheduled cooperatively, each worker has atomic access to the local proxy ``for free'', so an explicit publication list is unnecessary.
+Instead, each worker performs its own combining, and blocks if it was unable to satisfy its request immediately.
+Figure~\ref{fig:stackfc} shows how in the global stack example, pushes and pops can be matched locally, avoiding global synchronization for those operations.
+
+After all local combining has been done, one requesting worker is chosen to perform the combined operation globally.
+In the same figure, Worker 4 becomes the combiner and performs much the same synchronization as in the un-combined case, but is able to push multiple elements with one synchronization.
+As in the uncombined case, the global lock must still be acquired, and concurrent combined requests must still block and serialize on the master, though a technique later will describe how flat combining can be applied here, too.
+Combining in this way allows synchronization to be distributed and performed in parallel, coarsening the granularity of the actual global synchronization and reducing the amount of serialization that occurs.
+
+% In most cases, this means communicating with whichever core holds the ``lock-protected'' field that maintains sequential consistency.
 In the case of a global stack, this is the ``master'' \emph{top} pointer which must be incremented on pushes and decremented on pops.
@@
 
-At this point we have traded off one remote synchronization per operation for an additional local synchronization per operation and some number fewer remote synchronizations (depending on the amount of combining that occurs).
+% At this point we have traded off one remote synchronization per operation for an additional local synchronization per operation and some number fewer remote synchronizations (depending on the amount of combining that occurs).
 This would likely result in a net benefit due to local synchronization being inherently cheaper.
@@
 
-By design, each core in Grappa operates independently of the other cores.
+% By design, each core in Grappa operates independently of the other cores.
 Workers are scheduled cooperatively, so each worker can assume atomicity until it performs an operation that yields to the scheduler, such as a remote call.
@@
 \section{FC Framework}
+Many different data structures can integrate flat combining into their operation using the same delegation mechanism, making it easier to implement concurrent data structures.
+
 In order to leverage the flat-combining paradigm in Grappa, we implemented a generic flat-combining framework which can be used to improve performance for a number of global data structures.

### 2013-06-06 doc/pgas13/evaluation.tex
Put stack/queue and hash{set/map} plots together.

@@
   \centering
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.9\textwidth}
   \centering
-  \includegraphics[width=\textwidth]{data/plots/stack_perf.pdf}
-  \caption{\emph{GlobalStack.} Even with Grappa's aggregation, the stack without combining completely fails to scale.
-With combining, both workloads scale out to 64 nodes.
-The workload with roughly even pushes and pops performs best because it can do matching locally.}
-  \label{fig:stack}
+  \includegraphics[width=\textwidth]{data/plots/vector_perf.pdf}
+  \caption{\emph{GlobalStack and GlobalQueue} Even with Grappa's aggregation, without combining, both the stack and queue completely fail to scale.
+With combining, both scale out to 64 nodes.
+On the mixed workload, the stack is able to do matching locally, drastically improving global performance.
+  On the other hand, though still benefiting from reduced synchronization, the queue is unable to do matching locally, and in fact the mixed workload performs worse because the current implementation serializes combined enqueue and dequeue operations.
+  }
+  \label{fig:vector}
   \end{subfigure}%
-  \hspace{0.05\textwidth}
-  % ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-  %(or a blank line to force the subfigure onto a new line)
-  \begin{subfigure}[b]{0.45\textwidth}
-  \centering
-  \includegraphics[width=\textwidth]{data/plots/queue_perf.pdf}
-  \caption{\emph{GlobalQueue.} Same parameters as stack performance results.
+  % \hspace{0.05\textwidth}
+  % % ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+  % %(or a blank line to force the subfigure onto a new line)
+  % \begin{subfigure}[b]{0.45\textwidth}
+  % \centering
+  % \includegraphics[width=\textwidth]{data/plots/queue_perf.pdf}
+  % \caption{\emph{GlobalQueue.} Same parameters as stack performance results.
 The queue is unable to do matching locally, but benefits from reducing the amount of synchronization that must globally serialize.
 The mixed workload performs worse because the current implementation serializes combined enqueue and dequeue operations.}
-  \label{fig:queue}
+  % \label{fig:queue}
+  % \end{subfigure}
+  % \hspace{0.05\textwidth}
+
+  %  
+  \begin{subfigure}[b]{0.9\textwidth}
+  \centering
+  \includegraphics[width=\textwidth]{data/plots/hash_perf.pdf}
+  \caption{\emph{GlobalHashSet and GlobalHashMap.} Results are shown for a throughput workload inserting and looking up random keys in the range 0-1024 into a global hash with 1024 cells.
+Performance without combining scales better than the global queue or stack because synchronization happens at each hash cell, but drops off as number of destinations increases.
+Combining version is able to match inserts and lookups of the same key, enabling it to have a steeper scaling curve.}
+    \TODO{if time: add 'delete' operation, too (show that it doesn't affect correctness or performance)}
+  \label{fig:hash_perf}
   \end{subfigure}
-
-  \begin{subfigure}[b]{0.45\textwidth}
-  \centering
-  \includegraphics[width=\textwidth]{data/plots/hashset_perf.pdf}
-  \caption{\emph{GlobalHashSet.} Throughput workload inserting and looking up random keys in the range 0-1024 into a global hash set with 1024 cells.
-   Performance without combining scales better than the global queue or stack because synchronization happens at each hash cell, but drops off as number of destinations increases.
-Combining version is able to match inserts and lookups of the same key.}
-    \TODO{if time: add 'delete' operation, too (show that it doesn't affect correctness or performance)}
-  \label{fig:hashset}
-  \end{subfigure}
-  %
-  \hspace{0.05\textwidth}
-  %
-  \begin{subfigure}[b]{0.45\textwidth}
-  \centering
-  \includegraphics[width=\textwidth]{data/plots/hashmap_perf.pdf}
-  \caption{\emph{GlobalHashMap.} Same workload as for the Set, but random integers as the values in the map.
+  % %
+  % \hspace{0.05\textwidth}
+  % %
+  % \begin{subfigure}[b]{0.45\textwidth}
+  % \centering
+  % \includegraphics[width=\textwidth]{data/plots/hashmap_perf.pdf}
+  % \caption{\emph{GlobalHashMap.} Same workload as for the Set, but random integers as the values in the map.
 Performance matches that of the Set.} \TODO{try increasing size of value "payload", currently is just a tiny int64.}
-  \label{fig:hashmap}
-  \end{subfigure}%
+  % \label{fig:hashmap}
+  % \end{subfigure}%
   %
@@
     \emph{Raw performance of global data structures on simple throughput workloads.}
-    Results shown for workloads with all writes and a mix of writing and reading operations. All results are for 2048 workers per core and 16 cores per node.
+    Results shown for workloads with varying mixtures of writing and reading operations.
+All experiments were performed with 16 cores per node, and a suitable set of Grappa runtime parameters fixed for a given plot.
     \TODO{either all with error bars or none?}

### 2013-06-06 doc/pgas13/data/plot.stack.R

@@ -10,41 +10,52 @@ db <- function(query, factors, db="pgas.sqlite") {
 
 d <- db("select * from stack where ppn == 16",
       c("nnode", "num_starting_workers", "aggregator_autoflush_ticks", "periodic_poll_ticks", "flat_combining", "ppn", "fraction_push"))
+dq <- db("select * from queue where ppn == 16",
+    c("nnode", "num_starting_workers", "aggregator_autoflush_ticks", "periodic_poll_ticks", "flat_combining", "ppn", "fraction_push"))
 
 d$ops_per_msg <- with(d, (global_vector_push_ops+global_vector_pop_ops)/(global_vector_push_msgs+global_vector_pop_msgs))
 d$throughput <- with(d, nelems/trial_time_mean)
+
+dq$throughput <- with(dq, nelems/trial_time_mean)
 
 d$fc <- with(d, x(flat_combining,flat_combining_local_only))
 
 d$fc_version <- sapply(paste('v',d$flat_combining,d$flat_combining_local_only,sep=''),switch,
   v0NA='none', v1NA='full', v10='full', v11='local', v00='none'
 )
+dq <- within(dq,
+fc_version <- sapply(paste('v',flat_combining,flat_combining_local_only,sep=''),switch,
+  v0NA='none', v1NA='full', v10='full', v11='local', v00='none'
+))
 
-d.melt <- melt(d, measure=c("global_vector_push_ops", "global_vector_pop_ops", "ops_per_msg", "throughput"))
+d.melt <- melt(d, measure=c("global_vector_push_latency", "global_vector_pop_latency", "ops_per_msg", "throughput"))
 
 g <- ggplot(subset(d.melt,
-    nnode == 16 & log_nelems == 27
+    # & nnode == 16 
+    log_nelems == 28
   ), aes(
-    x=num_starting_workers,
+    x=nnode,
     y=value,
-    color=x(fc_version,fraction_push),
+    color=fc_version,
     shape=x(ppn,version),
-    group=x(flat_combining,fraction_push,ppn,version),
+    group=x(fc_version,fraction_push,version),
+    linetype=fraction_push,
     # label=global_vector_buffer,
   ))+
   geom_point()+
   geom_line()+
   # geom_text()+
-  facet_grid(variable~log_nelems+nnode+version, scales="free", labeller=label_pretty)+
[…]

### 2013-06-06 doc/pgas13/evaluation.tex
Improve descriptions of results.

@@
   \includegraphics[width=\textwidth]{data/plots/vector_perf.pdf}
-  \caption{\emph{GlobalStack and GlobalQueue} Even with Grappa's aggregation, without combining, both the stack and queue completely fail to scale.
-With combining, both scale out to 64 nodes.
-On the mixed workload, the stack is able to do matching locally, drastically improving global performance.
-  On the other hand, though still benefiting from reduced synchronization, the queue is unable to do matching locally, and in fact the mixed workload performs worse because the current implementation serializes combined enqueue and dequeue operations.
+  \caption{\emph{GlobalStack and GlobalQueue.}
+    Results are shown on a log scale for a throughput workload performing 256 million operations with 2048 workers per core and 16 cores per node.
+Flat combining improves throughput by at least an order of magnitude and allows performance to scale.
+Matching pushes and pops enables the stack to perform even better on a mixed workload.
   }
@@
   \includegraphics[width=\textwidth]{data/plots/hash_perf.pdf}
-  \caption{\emph{GlobalHashSet and GlobalHashMap.} Results are shown for a throughput workload inserting and looking up random keys in the range 0-1024 into a global hash with 1024 cells.
-Performance without combining scales better than the global queue or stack because synchronization happens at each hash cell, but drops off as number of destinations increases.
-Combining version is able to match inserts and lookups of the same key, enabling it to have a steeper scaling curve.}
+  \caption{\emph{GlobalHashSet and GlobalHashMap.}
+    Results are shown for a throughput workload inserting and looking up 256 million random keys in the range 0-1024 into a global hash with 1024 cells. 
+    Performance without combining scales out to 32 nodes because synchronization happens at each hash cell, but drops off as the number of destinations increases.
+Due to eliminating duplicate inserts and lookups, the combining version is able to continue to scale.}
     \TODO{if time: add 'delete' operation, too (show that it doesn't affect correctness or performance)}
@@
 The GlobalQueue and GlobalStack have very similar implementations in terms of how they are synchronized.
-Both benefit greatly from flat combining.
-This is because only one round-trip message \TODO{...}
+Even with Grappa's aggregation, without combining, both the stack and queue completely fail to scale because of serialization of all workers' updates on the master core.
+With combining, both scale well out to 64 nodes.
+On the mixed workload, the stack is able to do matching locally, allowing it to reduce the amount of communication drastically, greatly improving its performance.
 
-% However, the stack is able to match up pushes and pops locally.
-This leads to performance on an even mix of pushes and pops far outstripping the case where all pushes are performed due to requiring vastly less communication.
+The queue also benefits from reduced synchronization and batching operations, and its all-push workload performs identically to the stack's.
+However, the queue is unable to do matching locally, and in fact, the mixed workload performs worse because the current implementation serializes combined enqueue and dequeue operations.
+This restriction could be lifted with more careful synchronization at the master core allowing enqueues and dequeues to proceed in parallel as long as they do not conflict.
 
@@
 The GlobalHashSet and GlobalHashMap have the same synchronization strategy (serialization happens at each hashed location).
-\TODO{...}
+Performance without combining scales better than the global queue or stack because synchronization happens at each hash cell, but drops off as number of destinations increases.
+Combining version is able to match inserts and lookups of the same key, enabling it to have a steeper scaling curve.
 

### 2013-06-07 doc/pgas13/evaluation.tex
bunch of writing for evaluation section.

@@
 
+Experiments were run on a cluster of AMD Interlagos processors.
+Nodes have 32 2.1-GHz cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand network cards, connected via a QLogic InfiniBand switch.
+
 \subsection{Data Structure Throughput}
-First we measured the performance of the global data structures on extremely simple workloads to understand their raw performance.
+First we measured the raw performance of the global data structures on extremely simple throughput workloads.
 
-\begin{figure*}[t]
+% \begin{figure*}[t]
+%   \centering
+\begin{figure}[t] %{0.5\textwidth}
   \centering
-  \begin{subfigure}[b]{0.9\textwidth}
-  \centering
-  \includegraphics[width=\textwidth]{data/plots/vector_perf.pdf}
+  \includegraphics[width=0.5\textwidth]{data/plots/vector_perf.pdf}
   \caption{\emph{GlobalStack and GlobalQueue.}
@@
   \label{fig:vector}
-  \end{subfigure}%
+\end{figure}
   % \hspace{0.05\textwidth}
@@
   %  
-  \begin{subfigure}[b]{0.9\textwidth}
+\begin{figure}[t]
   \centering
-  \includegraphics[width=\textwidth]{data/plots/hash_perf.pdf}
+  \includegraphics[width=0.5\textwidth]{data/plots/hash_perf.pdf}
   \caption{\emph{GlobalHashSet and GlobalHashMap.}
-    Results are shown for a throughput workload inserting and looking up 256 million random keys in the range 0-1024 into a global hash with 1024 cells. 
+    Results are shown for a throughput workload inserting and looking up 256 million random keys in the range 0-1024 into a global hash with 1024 cells, with 2048 workers per core and 16 cores per node.
     Performance without combining scales out to 32 nodes because synchronization happens at each hash cell, but drops off as the number of destinations increases.
@@
   \label{fig:hash_perf}
-  \end{subfigure}
+\end{figure}
   % %
@@
   %
-  \caption{
-    \emph{Raw performance of global data structures on simple throughput workloads.}
-    Results shown for workloads with varying mixtures of writing and reading operations.
+%   \caption{
+%     \emph{Raw performance of global data structures on simple throughput workloads.}
+%     Results shown for workloads with varying mixtures of writing and reading operations.
 All experiments were performed with 16 cores per node, and a suitable set of Grappa runtime parameters fixed for a given plot.
-    \TODO{either all with error bars or none?}
-    \TODO{combine stack/queue \& set/map plots, no point in saying the same things twice.}
-  }\label{fig:datastructs}
-\end{figure*}
+%     \TODO{either all with error bars or none?}
+%     \TODO{combine stack/queue \& set/map plots, no point in saying the same things twice.}
+%   }\label{fig:datastructs}
+% \end{figure*}
 
@@
 \paragraph{HashSet and HashMap}
-The GlobalHashSet and GlobalHashMap have the same synchronization strategy (serialization happens at each hashed location).
-Performance without combining scales better than the global queue or stack because synchronization happens at each hash cell, but drops off as number of destinations increases.
-Combining version is able to match inserts and lookups of the same key, enabling it to have a steeper scaling curve.
+The GlobalHashSet and GlobalHashMap have the same synchronization strategy (serialization happens at each hashed location) which allows them to scale fairly well even without combining because serialization only happens on conflicts.
+However, after 32 nodes, scaling drops off significantly due to increased number of concurrent accessors and more destinations.
+The combining version is able to perform repeated inserts and repeated lookups with a single remote operation, enabling it to continue scaling further.
 
 \subsection{Application Kernel Performance}
-The impetus for this work was that naive implementations of standard data structures are insufficient for use in high-performance kernels.
+The original goal was to have scalable global data structures in the Grappa library that could be used in applications.
+Naive implementations of the data structures are insufficient for scaling, so HPC applications often implement their own highly customized versions.
+However, this forces each application to carefully reason about the ways in which the structures will be accessed.
+Additionally, many optimizations rely on relaxing global consistency in a way that does not affect the program.
+The Grappa data structures are synchronized to provide the most general use and match the expectations of programmers and algorithms.
 
@@
 \end{figure}
+The first application kernel is the Graph500 Breadth-First-Search (BFS) benchmark.
+This benchmark does a search starting from a random vertex in a synthetic graph and builds a search tree of parent vertices for each vertex traversed during the search.
+While this is a relatively simple problem, it exercises the random-access throughput of a system as well as being a primitive in many other graph algorithms.
+The BFS algorithm contains a global queue which represents the frontier of vertices to be visited in each level.
+Our implementation employs the direction-optimizing algorithm by Beamer et al.\cite{Beamer:Graph500} which performs particularly well for the scale-free RMAT graphs generated by the benchmark.
+The frontier queue in BFS is amenable to further optimization to take advantage of the fact that the algorithm does pushes and pops in separate phases, allowing consistency to be relaxed.
+We compare our implementation of BFS using the flat-combined global queue described above with a highly tuned Grappa implementation that uses a custom asynchronous queue.
+
+Figure~\ref{fig:bfs_perf} shows the results of scaling the BFS kernel up to 64 nodes.
+The simple queue implementation without flat combining is completely unscalable.
+However, the flat combining queue tracks the highly tuned asynchronous version.
+This illustrates that providing a safe, synchronized data structure for initially developing algorithms for PGAS is possible, while further optimizations can be applied incrementally.
 
 \paragraph{Connected Components}
-
 \begin{figure}[t]
@@
   \includegraphics[width=0.5\textwidth]{data/plots/cc_perf.pdf}
-  \caption{\emph{Connected Components} on the same scale 26 Graph500 graph.
-Performance is measured in MTEPS (parameterized by the total number of edges in the graph, independent of the algorithm used).}
+  \caption{\emph{Connected Components} on the same scale 26 Graph500 graph. Performance is measured in MTEPS.}
   \label{fig:cc_perf}
 \end{figure}
+Connected Components (CC) is another simple graph analysis kernel that illustrates another use of global data structures in irregular applications.
+We implement the three-phase CC algorithm~\cite{mtgl} which was designed for the massively-parallel MTA-2 machine.
+In first phase, the algorithm begins many traversals in parallel from random starting vertices, labeling vertices with the root vertex.
+Whenever two traversals encounter each other, their searches are pruned and an edge between the two roots is inserted in a set.
+After all edges have been traversed in this way, the set of edges forms a new, typically much smaller, graph.
+The second phase performs the classical Shiloach-Vishkin parallel algorithm\cite{shiloach1982n} on this reduced graph, and the final phase propagates the component labels out to the full graph.
+The creation of the reduced-graph edge set dominates the runtime of this algorithm, so improving the implementation of the set operations has a significant impact on performance.
+As in the case of BFS, further optimizations involving relaxation of consistency can be applied, in this case, to the global set.
+Therefore, we compare our straightforward implementation using the generic GlobalHashSet with and without flat combining against a tuned asynchronous implementation.
+
+The results in Figure~\ref{fig::cc_perf} show that none of these three implementations scale very well out to 64 nodes.
+However, performing combining does improves the performance of the algorithm overall and improves scaling.
+The tuned version outperforms the synchronous version because it is able to build up most of the set locally on each core before merging them at the end of the first phase.
+An implementation of the GlobalHashSet that did not provide synchronized semantics could potentially relax consistency in a more general way, but this is left for future work.

### 2013-06-07 doc/pgas13/evaluation.tex
Improve evaluation discussion a bunch.

@@
 First we measured the raw performance of the global data structures on extremely simple throughput workloads.
+In each experiment, a Grappa parallel loop spawns an equal number of tasks on all cores.
+Each task randomly chooses an operation based on the predetermined ``operation mix'', selecting either a push or pop for the Stack and Queue, or an insert or lookup for the Set and Map.
 
-% \begin{figure*}[t]
-%   \centering
-\begin{figure}[t] %{0.5\textwidth}
+\paragraph{Queue and Stack}
+\begin{figure}[t]
   \centering
   \includegraphics[width=0.5\textwidth]{data/plots/vector_perf.pdf}
-  \caption{\emph{GlobalStack and GlobalQueue.}
+  \caption{\emph{Queue and Stack performance.}
     Results are shown on a log scale for a throughput workload performing 256 million operations with 2048 workers per core and 16 cores per node.
@@
 \end{figure}
-  % \hspace{0.05\textwidth}
-  % % ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-  % %(or a blank line to force the subfigure onto a new line)
-  % \begin{subfigure}[b]{0.45\textwidth}
-  % \centering
-  % \includegraphics[width=\textwidth]{data/plots/queue_perf.pdf}
-  % \caption{\emph{GlobalQueue.} Same parameters as stack performance results.
-The queue is unable to do matching locally, but benefits from reducing the amount of synchronization that must globally serialize.
-The mixed workload performs worse because the current implementation serializes combined enqueue and dequeue operations.}
-  % \label{fig:queue}
-  % \end{subfigure}
-  % \hspace{0.05\textwidth}
 
-  %  
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=0.5\textwidth]{data/plots/stack_stats.pdf}
+  \caption{\emph{GlobalStack Statistics.}
+    Measured number of combined messages sent by the Stack with a fixed combining buffer of 1024 elements. Being above the buffer size indicates the amount of matched pushes and pops that occurred.
+  }
+  \label{fig:stack_stats}
+\end{figure}
+
+Figure~\ref{fig:vector} shows the results of the throughput experiments for the global Stack and Queue.
+Results are shown with flat combining completely disabled, only combining at the master core, only combining locally, and full two-level combining.
+
+Despite Grappa's automatic aggregation, without combining, both the stack and queue completely fail to scale because all workers' updates must serialize on the master core.
+Though combining on the master core alleviates some of the serialization, it does not provide significant benefit because all of the synchronization must still be performed on one core.
+However, with local flat combining, synchronization is able to be done mostly in parallel, with much less-frequent bulk synchronization at the master.
+Doing the additional level of combining at the master seems to not affect performance significantly compared to the impact of local combining.\TODO{Add data showing how many are combined in one pass on the master.}
+
+On the mixed workload, the stack is able to do matching locally, allowing it to reduce the amount of communication drastically, greatly improving its performance.
+Figure~\ref{fig:stack_stats} corroborates this, showing that the amount of combining that occurs directly correlates with the observed throughput.
+
+The queue also benefits from reduced synchronization and batching operations, and its all-push workload performs identically to the stack's.
+However, the queue is unable to do matching locally, and in fact, the mixed workload performs worse because the current implementation serializes combined enqueue and dequeue operations.
+This restriction could be lifted with more careful synchronization at the master core allowing enqueues and dequeues to proceed in parallel as long as enough elements are in the queue to allow them to proceed without conflicting.
+
+\paragraph{HashSet and HashMap}
 \begin{figure}[t]
@@
 \end{figure}
-  % %
-  % \hspace{0.05\textwidth}
-  % %
-  % \begin{subfigure}[b]{0.45\textwidth}
-  % \centering
-  % \includegraphics[width=\textwidth]{data/plots/hashmap_perf.pdf}
-  % \caption{\emph{GlobalHashMap.} Same workload as for the Set, but random integers as the values in the map.
-Performance matches that of the Set.} \TODO{try increasing size of value "payload", currently is just a tiny int64.}
-  % \label{fig:hashmap}
-  % \end{subfigure}%
-  %
-%   \caption{
-%     \emph{Raw performance of global data structures on simple throughput workloads.}
-%     Results shown for workloads with varying mixtures of writing and reading operations.
-All experiments were performed with 16 cores per node, and a suitable set of Grappa runtime parameters fixed for a given plot.
-%     \TODO{either all with error bars or none?}
-%     \TODO{combine stack/queue \& set/map plots, no point in saying the same things twice.}
-%   }\label{fig:datastructs}
-% \end{figure*}
 
-\paragraph{Queue and Stack}
-The GlobalQueue and GlobalStack have very similar implementations in terms of how they are synchronized.
-Even with Grappa's aggregation, without combining, both the stack and queue completely fail to scale because of serialization of all workers' updates on the master core.
-With combining, both scale well out to 64 nodes.
-On the mixed workload, the stack is able to do matching locally, allowing it to reduce the amount of communication drastically, greatly improving its performance.
-
-The queue also benefits from reduced synchronization and batching operations, and its all-push workload performs identically to the stack's.
-However, the queue is unable to do matching locally, and in fact, the mixed workload performs worse because the current implementation serializes combined enqueue and dequeue operations.
-This restriction could be lifted with more careful synchronization at the master core allowing enqueues and dequeues to proceed in parallel as long as they do not conflict.
-
-\paragraph{HashSet and HashMap}
-The GlobalHashSet and GlobalHashMap have the same synchronization strategy (serialization happens at each hashed location) which allows them to scale fairly well even without combining because serialization only happens on conflicts.
+Figure~\ref{fig:hash_perf} shows the throughput results for the GlobalQueue and GlobalStack.
+Both data structures have the same synchronization strategy (serialization happens only at each hashed location) which allows them to scale fairly well even without combining because serialization only happens on conflicts.
 However, after 32 nodes, scaling drops off significantly due to increased number of concurrent accessors and more destinations.

### 2013-06-07 doc/pgas13/evaluation.tex
Get rid of std error shadows, add keys facetting in Set/Map plot

@@
 However, with local flat combining, synchronization is able to be done mostly in parallel, with much less-frequent bulk synchronization at the master.
-Doing the additional level of combining at the master seems to not affect performance significantly compared to the impact of local combining.\TODO{Add data showing how many are combined in one pass on the master.}
+Doing the additional level of combining at the master seems to not affect performance significantly.
+This is somewhat surprising as all combined updates must completely serialize on the master still.
+However, with full combining enabled, we see that relatively few requests are combined together---for an all-push workload, on the order of a thousand are combined each pass, and for the mixed push/pop workload, on average less than 20 are combined.
 
@@
   \caption{\emph{GlobalHashSet and GlobalHashMap.}
-    Results are shown for a throughput workload inserting and looking up 256 million random keys in the range 0-1024 into a global hash with 1024 cells, with 2048 workers per core and 16 cores per node.
-    Performance without combining scales out to 32 nodes because synchronization happens at each hash cell, but drops off as the number of destinations increases.
-Due to eliminating duplicate inserts and lookups, the combining version is able to continue to scale.}
-    \TODO{if time: add 'delete' operation, too (show that it doesn't affect correctness or performance)}
+    Results are shown for a throughput workload inserting and looking up 256 million random keys in a particular range into a global hash with the same number of cells, with 2048 workers per core and 16 cores per node.
+  }
   \label{fig:hash_perf}
@@
 
-Figure~\ref{fig:hash_perf} shows the throughput results for the GlobalQueue and GlobalStack.
-Both data structures have the same synchronization strategy (serialization happens only at each hashed location) which allows them to scale fairly well even without combining because serialization only happens on conflicts.
-However, after 32 nodes, scaling drops off significantly due to increased number of concurrent accessors and more destinations.
-The combining version is able to perform repeated inserts and repeated lookups with a single remote operation, enabling it to continue scaling further.
+Without combining, performance scales out to 32 nodes because synchronization is distributed over hash cells, but drops off as the number of destinations increases.
+
+Figure~\ref{fig:hash_perf} shows the throughput results for the GlobalHashSet and GlobalHashMap.
+Both data structures synchronize at each hash cell, which allows them to scale fairly well even without combining.
+However, after 32 nodes, scaling drops off significantly due to increased number of concurrent accessors and destinations.
+Combining allows duplicate inserts and lookups to be eliminated, so performs better the smaller the key range. This reduction in message traffic allows scaling to continue out to 64 nodes.
 

### 2013-06-07 doc/pgas13/data/plot.cc.R

@@ -66,7 +66,7 @@ g <- ggplot(subset(d,
     label=x(aggregator_autoflush_ticks/1e6,num_starting_workers),
   ))+
   # geom_point()+
-  geom_smooth(size=1)+
+  geom_smooth(size=1, fill=NA)+
   # geom_text(size=2,hjust=-0.2,vjust=1)+
   # geom_line()+
   xlab("Nodes")+

### 2013-06-07 doc/pgas13/data/plot.hashset.R

@@ -36,7 +36,7 @@ dset$ops_per_msg <- with(dset, (hashset_insert_ops+hashset_lookup_ops)/(hashset_
 dset$throughput <- with(dset, (hashset_insert_ops+hashset_lookup_ops)/trial_time_mean)
 
 dset$fc_version <- sapply(paste('v',dset$flat_combining,dset$insert_async,sep=''),switch,
-  v0NA='none', v1NA='local', v11='async', v10='local', v00='none', v01='?'
+  v0NA='none', v1NA='local only', v11='async', v10='local only', v00='none', v01='?'
 )
 
 d.melt <- melt(dset, measure=c("ops_per_msg", "throughput"))
@@ -80,8 +80,8 @@ gg <- ggplot(subset(dset, log_nelems==28 & ppn==16 & num_starting_workers==8192
   # facet_grid(log_nelems~., scales="free", labeller=label_pretty)+
   xlab("Nodes")+
   scale_color_discrete(name="Flat Combining")+
-  scale_linetype_discrete(name="Fraction lookups:")+
-  scale_shape_discrete(name="Fraction lookups:")+
+  scale_linetype_discrete(name="Operation Mix")+
+  scale_shape_discrete(name="Operation Mix")+
   ylab("Throughput (ops/sec)")+expand_limits(y=0)+my_theme+
   theme(strip.text=element_text(size=rel(0.7)),
         axis.text.x=element_text(size=rel(0.85)))
@@ -94,7 +94,7 @@ gg <- ggplot(subset(dset, log_nelems==28 & ppn==16 & num_starting_workers==8192
 dmap$ops_per_msg <- with(dmap, (hashmap_insert_ops+hashmap_lookup_ops)/(hashmap_insert_msgs+hashmap_lookup_msgs))
 dmap$throughput <- with(dmap, (hashmap_insert_ops+hashmap_lookup_ops)/trial_time_mean)
 dmap$fc_version <- sapply(paste('v',dmap$flat_combining,dmap$insert_async,sep=''),switch,
-  v0NA='none', v1NA='local', v11='async', v10='local', v00='none', v01='?'
+  v0NA='none', v1NA='local only', v11='async', v10='local only', v00='none', v01='?'
 )
 
 gg <- ggplot(subset(dmap, log_nelems==28 & ppn==16 & num_starting_workers==8192
@@ -114,8 +114,8 @@ gg <- ggplot(subset(dmap, log_nelems==28 & ppn==16 & num_starting_workers==8192
   xlab("Nodes")+
[…]

### 2013-06-07 doc/pgas13/data/plot.stack.R

@@ -22,11 +22,11 @@ dq$throughput <- with(dq, nelems/trial_time_mean)
 d$fc <- with(d, x(flat_combining,flat_combining_local_only))
 
 d$fc_version <- sapply(paste('v',d$flat_combining,d$flat_combining_local_only,sep=''),switch,
-  v0NA='none', v1NA='full', v10='full', v11='local', v00='master', v01='none'
+  v0NA='none', v1NA='full', v10='full', v11='local only', v00='master only', v01='none'
 )
 dq <- within(dq,
 fc_version <- sapply(paste('v',flat_combining,flat_combining_local_only,sep=''),switch,
-  v0NA='none', v1NA='full', v10='full', v11='local', v00='master', v01='none'
+  v0NA='none', v1NA='full', v10='full', v11='local only', v00='master only', v01='none'
 ))
 
 
@@ -94,8 +94,8 @@ ggsave(plot=g, filename="plots/stack_perf.pdf", width=7, height=5)
 # dq <- within(dq,
 #   mix <- sapply(fraction_push, switch, '1'='100% pop', '0.5'='50% push, 50% pop', '?')
 # )
-dq$mix <- sapply(dq$fraction_push, function(f){ return(ifelse(f == '1', '100% push', '50% push, 50% pop')) })
-d$mix <- sapply(d$fraction_push, function(f){ return(ifelse(f == '1', '100% push', '50% push, 50% pop')) })
+dq$mix <- sapply(dq$fraction_push, function(f){ return(ifelse(f == '1', '100% push', '50% push,\n50% pop')) })
+d$mix <- sapply(d$fraction_push, function(f){ return(ifelse(f == '1', '100% push', '50% push,\n50% pop')) })
 
 d$struct <- sapply( d$nnode, function(i){ return('GlobalStack') })
 dq$struct <- sapply(  dq$nnode, function(i){ return('GlobalQueue') })
@@ -128,7 +128,7 @@ gg <- ggplot(subset(d.c,
   # scale_x_continuous(breaks=c(8,16,32,48,64))+
   scale_color_discrete(name="Flat Combining")+
   scale_shape_discrete(name="Flat Combining")+
-  scale_linetype_discrete(name="Operation mix")+
+  scale_linetype_discrete(name="Operation Mix")+
   expand_limits(y=0)+my_theme+
   theme(strip.text=element_text(size=rel(1)),
         axis.text.x=element_text(size=rel(0.85)))
[…]

### 2013-06-07 doc/pgas13/flat-combining.tex
Rework SC section.

@@
 
-\paragraph{Correctness}
-\TODO{(are these arguments correct??)}
+\paragraph{Sequential Consistency}
+The sequential consistency guarantee is that within a task, operations will be in task order, so to enforce ordering between operations in different tasks, one task must perform an operation that another one can observe.
+An example is shown in Figure~\ref{fig:sync}.\TODO{make sync example}
+The Grappa memory model follows the Data-Race-Free-0 Model (guaranteeing sequential consistency for data-race free programs).
+Being conservative, delegate operations block the calling worker until they have completed globally, ensuring they can be used for synchronization---delegate operations within a task are guaranteed to be globally visible in program order.
+Because it is not immediately obvious that this distributed version of flat combining preserves sequential consistency, here we will make arguments for why it does.
 
-It may not be immediately obvious why this two-level version of flat combining preserves sequential consistency.
-The combining operations that are performed locally among requests on a single core must preserve the illusion of sequential order; this is unchanged from shared-memory flat-combining.
+% \begin{itemize}
+%   \item Block the calling worker until the operation is guaranteed to have a valid global order.
+%   \item Local combining must not allow workers to observe a different order than is committed globally.
+% \end{itemize}
+To preserve sequentially-consistent semantics for flat-combined operations, they must obey a consistent global order.
+To start with, local combined operations must have a serializable order among each other; this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
+When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map).
+The global order, then, is essentially the concatenation of the cores' sequential orders.
+For this ``concatenation'' to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.
+
+In the case of the Stack and Queue, this guarantee comes from applying a batch of push or pop operations atomically in a contiguous block on the global data structure.
+The second level of combining at the master preserves this as well by only allowing updates to proceed when they will not conflict and blocking all of them until they have completed.
+Matching pushes and pops locally is one exception to the rule that operations must block until globally committed.
+Because a pop ``destroys'' the corresponding push, these operations together are independent of all others and can conceptually be placed anywhere in a total global ordering, so they can be safely matched locally.
+
+The correctness of combining set and map operations is more nuanced.
+Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally.
+Therefore, a batch of these operations can be committed globally in parallel.
+If an insert finds that its key is already in the combiner, it does not need to send its own message.
+However, it must still block until that insert is done, otherwise it may assume that the key has been inserted and perform some other operation that requires the insert to have happened.
+In the same way, lookups can piggy-back on other lookups of the same key.
+It is tempting to allow a lookup to be satisfied locally by inspecting the keys to be inserted.
+However, this allows the local order in which keys were inserted to be observed.
+If this is allowed, then that same order must be respected globally, forcing each batch to ensure atomicity from all other cores' batches.
+This is prohibitively expensive, so a cheaper solution is chosen: lookups must get their values from the global data structure, but can be performed in parallel in the same way as inserts.
+
+% (wrong!) A lookup can be satisfied locally provided that the insert it relies on is guaranteed to eventually be committed globally.
+This allows the task to continue.
+If it does another lookup to a different key that is satisfied locally, then that task has observed one particular ordering of the two inserts.
+When those inserts commit globally, they will proceed in parallel to their respective hashed locations.
+Another task could
+
+% The combining operations that are performed locally among requests on a single core must preserve the illusion of sequential order; this is unchanged from shared-memory flat-combining.
 When combined operations are applied globally, they are serialized at the ``master'' core for a given synchronization context.
-The global order is essentially the concatenation of each core's serialized updates.
-In the case where operations are satisfied locally (by matching up with other local operations), by definition these must be independent of other operations, so they can conceptually be placed anywhere in the global ordering (or even ignored completely in the global ordering).
+The global order is essentially the concatenation of each core's batch of serialized accesses.
+% In the case where operations are satisfied locally (by matching up with other local operations), by definition these must be independent of other operations, so they can conceptually be placed anywhere in the global ordering (or even ignored completely in the global ordering).
 
-In the spirit of the Data-Race-Free-0 Model (guaranteeing sequential consistency for data-race free programs), operations on the global data structure are considered to be made atomic by a conceptual global ``lock'' on the data structure.
-Atomicity and freedom from data races in the case of multiple data structures and other synchronization must be guaranteed externally, just as in any multithreaded shared-memory program.
+% In the spirit of the Data-Race-Free-0 Model (guaranteeing sequential consistency for data-race free programs), operations on the global data structure are considered to be made atomic by a conceptual global ``lock'' on the data structure.
+
+These requirements only guarantee sequential consistency for a data structure in isolation.
+Atomicity and data race freedom in the case of multiple data structures must be guaranteed externally, just as in any multithreaded shared-memory program.
 
 \section{FC Framework}
+\TODO{add code examples, and add discussion of "master" combining}
 Many different data structures can integrate flat combining into their operation using the same delegation mechanism, making it easier to implement concurrent data structures.

### 2013-06-07 doc/pgas13/flat-combining.tex
Update diagrams with Luis's suggestions (add numbers fix locks)

@@
   \caption{\emph{Flat combining in shared memory.}
-    To access the shared stack, each thread adds its request to the publication list.
-Thread 4 acquires the stack's lock and becomes the combiner, while the remaining threads spin waiting for their request to be satisfied.
-The combiner walks the publication list from the head, matching up Thread 3's push with Thread 2's pop on its own private stack.
-The 2 remaining pushes are added to the shared stack, the top is updated, and finally Thread 4 releases the lock and continues its normal execution.
+    To access the shared stack, each thread adds its request to the publication list (1).
+Thread 4 acquires the stack's lock and becomes the combiner (2), while the remaining threads spin waiting for their request to be satisfied.
+The combiner walks the publication list from the head, matching up Thread 3's push with Thread 2's pop on its own private stack (3).
+The 2 remaining pushes are added to the top of the shared stack (4).
+Finally, the top is updated, and Thread 4 releases the lock and continues its normal execution.
   }
@@
     \caption{\emph{Without flat combining.}
-      To do its push, Worker 1 sends a message to synchronize with the master on Core 0, sends another message to write its value to the top of the stack, bumps the synchronized \texttt{top} pointer, and finally continues.
-Worker 2, and all other workers on all cores accessing the stack, must block at the master and wait for Worker 1 to complete its push before doing their operations.
+      To do its push, Worker 1 sends a message to synchronize with the master on Core 0 (1), which sends another message to write the value to the top of the stack (2), bumps the synchronized \texttt{top} pointer (3), and finally continues.
+Worker 2, and all other workers on all cores accessing the stack, must block at the master and wait for Worker 1 to complete its push before doing their operations (4).
     }
@@
       Workers within the core add their requests to a local proxy object.
-Worker 3's pop matches with Worker 2's push, requiring no global communication.
-After combining locally, worker 1 and 4's pushes remain, so Worker 4 becomes the core's combiner, synchronizes with the master, adds both new values to the global stack, bumps the top pointer and releases the lock on master, and finally wakes Worker 1 and continues.
+Worker 3's pop matches with Worker 2's push, requiring no global communication (1).
+After combining locally, Worker 1 and 4's pushes remain, so Worker 4 becomes the core's combiner (2), sends a message to synchronize with the master (3), adds both new values to the global stack (4), bumps the top pointer and releases the lock on master (5), and finally wakes Worker 1 and continues (6).
     }

### 2013-06-08 doc/pgas13/flat-combining.tex
Version sent to grappa@cs for feedback.
Working on code snippets.

@@
 After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the remaining two elements are pushed onto the top of the stack, and the requesting threads' records are marked as complete.
-In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be combined for greater performance.
+In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched locally.
 
@@
 
+\begin{figure}[t]
+\centering
+\begin{lstlisting}[style=grappa]
+class GlobalStack {
+  GlobalAddress<T> top;
+  Grappa::Mutex lock;
+};
+void push(GlobalAddress<GlobalStack> stack, T val) {
+  // perform operation on the stack's master core
+  delegate::call(stack.core(), [stack,val]{
+    // force other operations to wait
+    stack->lock.acquire();
+    // do remote write to the top of the stack
+    // (so other workers may interleave)
+    delegate::write(stack->top, val);
+    // bump the top pointer
+    stack->top++;
+    stack->lock.release();
+  }); // blocks until response arrives
+}
+\end{lstlisting}
+
+\caption{Snippet of code for pushing to a global stack (without flat combining).}
+\label{fig:push}
+\end{figure}
+
 All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object.
 To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read()} delegate described earlier, blocks the task until complete.
-To do a \texttt{push}, the task must send a message to the master to acquire the lock.
+Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}, the task must send a message to the master to acquire the lock.
 If it succeeds, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker.
@@
 
-\section{FC Framework}
+\section{Grappa FC Framework}
 \TODO{add code examples, and add discussion of "master" combining}
-Many different data structures can integrate flat combining into their operation using the same delegation mechanism, making it easier to implement concurrent data structures.
-
+% Many different data structures can integrate flat combining into their operation using the same delegation mechanism, making it easier to implement concurrent data structures.
 In order to leverage the flat-combining paradigm in Grappa, we implemented a generic flat-combining framework which can be used to improve performance for a number of global data structures.
@@
 
+% As discussed before, Grappa's distributed combining takes a different approach to expressing how operation locally  flat combining design takes a different approach to performing 
 The Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine.
-In the original work, an explicit publication list is built up, and when the combiner scans over the list, it has a structure that it uses to accumulate requests in a logical way.
+As mentioned before, the local ``publication list'' is trivial to access atomically in Grappa due to workers being cooperatively scheduled.
+Instead, in the Grappa FC Framework, a \emph{proxy} object is instantiated ahead of time on each core and each worker merges its update into that structure before blocking or becoming the combiner.
+% In the original work, an explicit publication list is built up, and when the combiner scans over the list, it has a structure that it uses to accumulate requests in a logical way.
 In the case of the stack, for instance, it is a thread-local stack, which it can use to immediately match up pushes and pops.
@@
 This reduces the space blowout of a large dynamically-sized publication list (with associated caching issues), and does not overly affect the expression of combining.
-
-Using this style, each global data structure must define the following:
+In this style, each global data structure must define the following:
 \begin{enumerate}
-  \item A local \emph{proxy} data structure for doing combining.
+  \item A local \emph{proxy} data structure for tracking updates to the global structure.
   \item \emph{Combining methods} that operate on the local proxy.
@@
 \end{enumerate}
+
+\begin{figure}[t]
+\centering
+\begin{lstlisting}[style=grappa]
+struct StackProxy {
+  T values[CAPACITY];
+  T* results[CAPACITY];
+  ConditionVariable waitlist;
+  int npush, npop;
+  void push(T e) {
+    if (npop > 0) { // satisfy locally
+      *results[--npop] = e;
+      waitlist.signal();
+    } else {
+      values[npush++] = e;
+      block(waitlist);
+    }
+  }
+  T pop() {
+    T val;
+    if (npush > 0) { // satisfy locally
+      val = values[--npush];
+      waitlist.signal();
+    } else {
+      pops[npop++] = &val;
+      block(waitlist);
+    }
+  }
+  void sync() {
+    if (npush > 0) {
+
+    } else if (npop > 0) {
+
+      for (int i = 0; i < npop; i++) *results[i] = values[i];
+    }
+  }
+};
+\end{lstlisting}
+\end{figure}
 

### 2013-06-08 doc/pgas13/flat-combining.tex
Worked on FC framework section (w/ code examples)

@@
 One core is designated the ``master'' to enforce global synchronization, and holds the elements of the data structure that all concurrent accessors must agree on, in this case, the \texttt{top} pointer.
-
-\begin{figure}[t]
-\centering
-\begin{lstlisting}[style=grappa]
-class GlobalStack {
-  GlobalAddress<T> top;
-  Grappa::Mutex lock;
-};
-void push(GlobalAddress<GlobalStack> stack, T val) {
-  // perform operation on the stack's master core
-  delegate::call(stack.core(), [stack,val]{
-    // force other operations to wait
-    stack->lock.acquire();
-    // do remote write to the top of the stack
-    // (so other workers may interleave)
-    delegate::write(stack->top, val);
-    // bump the top pointer
-    stack->top++;
-    stack->lock.release();
-  }); // blocks until response arrives
-}
-\end{lstlisting}
-
-\caption{Snippet of code for pushing to a global stack (without flat combining).}
-\label{fig:push}
-\end{figure}
 
@@
 \end{enumerate}
+An example proxy declaration for the GlobalStack is shown in Figure~\ref{fig:stack_proxy}.
 
@@
 \begin{lstlisting}[style=grappa]
-struct StackProxy {
-  T values[CAPACITY];
-  T* results[CAPACITY];
-  ConditionVariable waitlist;
+class GlobalStack {
+  GlobalAddress<T> top;
+  Grappa::Mutex lock;
+};
+
+void push(GlobalAddress<GlobalStack> stack, T val) {
+  // perform operation on the stack's master core
+  delegate::call(stack.core(), [stack,val]{
+    // force other operations to wait
+    stack->lock.acquire();
+    // do remote write to the top of the stack
+    // (so other workers may interleave)
+    delegate::write(stack->top, val);
+    // bump the top pointer
+    stack->top++;
+    stack->lock.release();
+  }); // blocks until response arrives
+}
+\end{lstlisting}
+
+\caption{Snippet of code for pushing to a global stack (without flat combining).}
+\label{fig:push}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\begin{lstlisting}[style=grappa]
+class GlobalStackProxy : Grappa::FCProxy {
+  // pointer to data structure master
+  GlobalAddress<GlobalStack> master;
+
+  // local state for tracking pushes and pops
+  T  pushed_values[1024];
+  T* popped_results[1024];
   int npush, npop;
-  void push(T e) {
-    if (npop > 0) { // satisfy locally
-      *results[--npop] = e;
-      waitlist.signal();
-    } else {
-      values[npush++] = e;
-      block(waitlist);
-    }
-  }
-  T pop() {
-    T val;
-    if (npush > 0) { // satisfy locally
-      val = values[--npush];
-      waitlist.signal();
-    } else {
-      pops[npop++] = &val;
-      block(waitlist);
-    }
-  }
-  void sync() {
+
+  // Combining Methods 
+  // (either match locally or block for results)
+  void push(T val);
+  T pop();
+
+  // Global sync (called by FC framework)
+  void sync() override {
     if (npush > 0) {
-
-    } else if (npop > 0) {
-
-      for (int i = 0; i < npop; i++) *results[i] = values[i];
-    }
+      // on master, acquire lock and return GlobalAddress to the top of the stack
+      auto top = delegate::call(master.core(),[=]{
+        master->lock.acquire();
+        return master->top;
+      });
+      // copy values to top of stack
+      Grappa::memcpy(top, pushed_values, npush);
+      // release lock at the master core
+      delegate::call(master.core(),[master,npush]{
+        master->top += npush;
+        master->lock.release();
+      });
+    } else if (npop > 0) {} //...elided for space
   }
@@
 \end{lstlisting}
+\caption{Snippet of the Proxy object for the GlobalStack. The code to synchronize pushes should resemble that from Figure~\ref{fig:push} but for multiple elements.}
+\label{fig:proxy}
 \end{figure}
 
-Being conservative about SC, workers typically block when issuing these requests, so the framework must also take care of delivering results and waking the correct workers.
-The flat-combining framework is responsible for ensuring that all of the combined operations eventually get committed.
-There are a number of ways progress could be guaranteed, but one of the simplest is to take an approach in the spirit of the original flat-combining, which is to ensure that as long as there are any outstanding requests, there is at least one worker committing a combined operation.
+With this data structure-specific proxy object, the FC framework does the bookkeeping to block and wake Workers, deliver results, manage when to synchronize, and choose the combiner.
+The FC framework is responsible for ensuring that all of the combined operations eventually get committed.
+There are a number of ways progress could be guaranteed, but one of the simplest is to ensure that as long as there are any outstanding requests, at least one worker is committing a combined operation.
 When that combined operation finishes, if there are still outstanding requests that have built up in the meantime, another blocked worker is chosen to do another combined synchronization.
@@
 
-The flat combining framework transparently wraps the \emph{proxy} object.
+While a combining operation is in flight, new requests are likely to continue to accumulate.
+The framework transparently wraps the proxy object so that when it starts a sync on one, it can direct requests to a fresh combiner.
+It is important to note that combiners must start out fresh each time---it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them.
+
+% The flat combining framework transparently wraps the \emph{proxy} object.
 Behind this layer of indirection, it manages multiple copies of the proxy so that, as discussed above, there is always at least one being committed and one to receive additional combining operations.
@@
 
-\subsection{Global Stack}
-To demonstrate the flat-combining framework, we revisit the Global Stack example from earlier.
-Recall that the ``master'' core owns the \texttt{top} index, which indicates where the top of the stack currently is in the stack's allocated global storage.
-In the non-combining case, each call to \texttt{push} must do an atomic increment of \texttt{top} to claim storage, then write to that location, then finally return.
+\subsection{Global Stack and Queue}
+Figure~\ref{fig:proxy} shows an excerpt from the definition of the proxy object for the Grappa GlobalStack.
+The proxy tracks pushes locally in the \texttt{pushed\_values} array.
+When \texttt{pop()} is called, if there are pushes in the local array, it immediately takes the value and wakes the last worker to push.
+Otherwise, it adds a pointer to a location on its stack where the \texttt{sync} operation, will eventually write the result.
+Because of local matching, when a Stack proxy is synchronized, it will have either all pops or all pushes, which makes the implementation of sync straightforward.
+Note that the operation to synchronize a batch of pushes looks almost identical to the code to do a single push from Figure~\ref{fig:push}.
+
+The GlobalQueue has nearly the same implementation as the stack, but is unable to match locally, so \texttt{sync} first does the pushes, then does the pops.
+
+\subsection{Global Set, and Map}
+The Grappa GlobalSet uses a simple chaining design, implemented with a global array of cells (allocated from the global heap), which are partitioned evenly among all the cores, and indexed with a globally-agreed hash function.
+Our flat-combining version supports both \texttt{insert} and \texttt{lookup}.
+To track all of the keys waiting to be inserted, we use the hash set implementation from the C++11 standard library (\texttt{std::unordered\_set}).
+Similar to the case with pops, for lookups, we must keep track of pointers for where to put the results of lookup operations, which the proxy does with a hash map (again from the C++ standard library) from keys to a list of pointers.
+As discussed in Section~\ref{sec:sc}, matching lookups with inserts locally would force a particular sequential order, so instead we only allow matching inserts with inserts and lookups with lookups.
+This allows \texttt{sync} to simply issue all of the inserts and lookups in parallel and block until all have completed.
+
+Our implementation of the GlobalMap matches that of the Set, but of course stores values of a given type.
+This does not change the requirements of the proxy except that the return values of lookups must be a copy of the value rather than just a boolean.

### 2013-06-08 doc/pgas13/evaluation.tex
Remove 'full' from plots, fix discusion, rename to 'dist' and 'centralized'

@@
     Results are shown on a log scale for a throughput workload performing 256 million operations with 2048 workers per core and 16 cores per node.
-Flat combining improves throughput by at least an order of magnitude and allows performance to scale.
+Local flat combining improves throughput by at least an order of magnitude and allows performance to scale.
 Matching pushes and pops enables the stack to perform even better on a mixed workload.
@@
 However, with local flat combining, synchronization is able to be done mostly in parallel, with much less-frequent bulk synchronization at the master.
-Doing the additional level of combining at the master seems to not affect performance significantly.
-This is somewhat surprising as all combined updates must completely serialize on the master still.
-However, with full combining enabled, we see that relatively few requests are combined together---for an all-push workload, on the order of a thousand are combined each pass, and for the mixed push/pop workload, on average less than 20 are combined.
 
@@
 
-The queue also benefits from reduced synchronization and batching operations, and its all-push workload performs identically to the stack's.
-However, the queue is unable to do matching locally, and in fact, the mixed workload performs worse because the current implementation serializes combined enqueue and dequeue operations.
-This restriction could be lifted with more careful synchronization at the master core allowing enqueues and dequeues to proceed in parallel as long as enough elements are in the queue to allow them to proceed without conflicting.
+We also experimented with performing both forms of combining together; however, combining at the master did not affect performance significantly.
+This was somewhat surprising as combined updates will still force others to wait.
+However, with both levels of combining enabled, relatively few requests are combined at the master---for the all-push workload, on the order of a thousand, and for the mixed workload on the stack, the average is less than 20.
+
+The queue benefits in the same way from reducing synchronization and batching, and its all-push workload performs identically to the stack's.
+However, the queue is unable to do matching locally, and in fact, the mixed workload performs worse because the current implementation serializes combined pushes and combined pops.
+This restriction could be lifted with more careful synchronization at the master core allowing pushes and pops to proceed in parallel as long as there are enough elements in the queue.
 

### 2013-06-08 doc/pgas13/flat-combining.tex
Remove 'full' from plots, fix discusion, rename to 'dist' and 'centralized'

@@
 
-With flat combining, instead of all workers on a core sending independent messages to do remote synchronization, they can do combining locally and share a single remote synchronization operation. 
-As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and as bonus, allows for data structure-specific combining to magnify the benefits even further.
+\paragraph{Centralized Combining.}
+A first thought might be to directly apply the idea of flat combining to the serialization at the master node.
+The worker that acquires the lock on the master can walk through the requests of all the other workers waiting to acquire the global lock and perform combining on them.
+In the case of the GlobalStack, this would mean taking all of the push requests, matching them with the pops, and sending messages back to the remote workers with the results.
+Any leftover pushes or pops would be applied to the data structure, and another round of combining would begin.
+This approach, combining on the master, reduces traffic to the data structure storage, but the single ``master'' core must still process every single request.
+This approach will not scale when every other core can generate requests just as fast as the master can combine them.
 
+\paragraph{Distributed Combining}
+Instead of all workers on a core sending independent synchronization messages and putting the burden all on the master, what if each core did its own combining first, and synchronized with the master in bulk?
+Distributing the combining to each core allows the majority of the work to be performed in parallel and without communication.
+As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and allows operations to be composed in data structure-specific ways for even greater performance.
 % allows synchronization to happen in parallel on all the cores, 
@@
 
-When a flat-combined global data structure is created, each core allocates a local proxy object in it's core-local heap.
-This proxy serves the role of the publication list, allowing workers within a core to coordinate their efforts.
-Conceptually, workers add their requests to the local publication list, and one is chosen to do the combining.
+In distributed flat-combining, each core needs its own publication list to track all of its outstanding operations to be performed on the global data structure.
+In Grappa, this takes the form of a local \emph{proxy} object allocated from the core-local heap.
+Conceptually, workers add their requests to the local publication list, one is chosen to do the combining, and the rest block until their request is satisfied.
 However, because workers are scheduled cooperatively, each worker has atomic access to the local proxy ``for free'', so an explicit publication list is unnecessary.
-Instead, each worker performs its own combining, and blocks if it was unable to satisfy its request immediately.
+Instead, each worker performs its own combining, pushing the 
+, and blocks if it was unable to satisfy its request immediately.
 Figure~\ref{fig:stackfc} shows how in the global stack example, pushes and pops can be matched locally, avoiding global synchronization for those operations.
@@
 \section{Grappa FC Framework}
-\TODO{add code examples, and add discussion of "master" combining}
+% One of the strengths of flat combining is that the same combining mechanism can be applied in multiple data structures. Each structure must only specify how its operations compose.
 % Many different data structures can integrate flat combining into their operation using the same delegation mechanism, making it easier to implement concurrent data structures.
-In order to leverage the flat-combining paradigm in Grappa, we implemented a generic flat-combining framework which can be used to improve performance for a number of global data structures.
+In order to leverage the flat-combining paradigm in Grappa, we implemented a generic framework which can be used to improve performance for a number of global data structures.
 The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress.
@@
 \end{enumerate}
-An example proxy declaration for the GlobalStack is shown in Figure~\ref{fig:stack_proxy}.
+An example proxy declaration for the GlobalStack is shown in Figure~\ref{fig:proxy}.
 
@@
 
-void push(GlobalAddress<GlobalStack> stack, T val) {
+void push(GlobalAddress<GlobalStack> stack, T e) {
   // perform operation on the stack's master core
-  delegate::call(stack.core(), [stack,val]{
+  delegate::call(stack.core(), [stack,e]{
     // force other operations to wait
@@
     // (so other workers may interleave)
-    delegate::write(stack->top, val);
+    delegate::write(stack->top, e);
     // bump the top pointer
@@
   }); // blocks until response arrives
-}
-\end{lstlisting}
-
+}\end{lstlisting}
 \caption{Snippet of code for pushing to a global stack (without flat combining).}

### 2013-06-08 doc/pgas13/data/plot.bfs.R

@@ -34,15 +34,15 @@ ggsave(plot=g, filename="plots/bfs_mess.pdf", width=10, height=8)
 d$nw <- d$num_starting_workers
 
 d$fc_version <- sapply(paste(d$bfs_version,d$flat_combining,sep='_'),switch,
-  beamer_0='async', beamer_1='async', beamer_queue_0='none', beamer_queue_1='full'
+  beamer_0='custom', beamer_1='custom', beamer_queue_0='none', beamer_queue_1='distributed'
 )
 # d$fc_version <- paste(d$bfs_version,d$flat_combining,sep='_')
 # d$fc_version
 
 g <- ggplot(subset(d,
     ppn == 16 & scale == 26
-    & (num_starting_workers == 1024 & grepl('full|none', fc_version) 
-      | num_starting_workers >= 256 & grepl('async', fc_version))
+    & (num_starting_workers == 1024 & grepl('distributed|none', fc_version) 
+      | num_starting_workers >= 256 & grepl('custom', fc_version))
     & aggregator_autoflush_ticks > 100000
   ), aes(
     x=nnode,
@@ -53,7 +53,7 @@ g <- ggplot(subset(d,
     group=x(fc_version,num_starting_workers),
     label=x(num_starting_workers,aggregator_autoflush_ticks),
   ))+
-  geom_point()+
+  # geom_point()+
   # geom_text(size=2,position="jitter")+
   geom_smooth(size=1, fill=NA)+
   xlab("Nodes")+

### 2013-06-08 doc/pgas13/data/plot.cc.R

@@ -19,7 +19,7 @@ d$nroots <- d$cc_concurrent_roots
 d$gce_flats <- with(d, gce_total_remote_completions/gce_completions_sent)
 
 d$fc_version <- sapply(paste('v',d$flat_combining,d$cc_insert_async,sep=''),switch,
-  v0NA='none', v1NA='fc', v11='async', v10='fc', v00='none', '??'
+  v0NA='none', v1NA='distributed', v11='custom', v10='distributed', v00='none', '??'
 )
 
 g <- ggplot(subset(d, cc_hash_size <= 16384 & scale == 26 & ppn == 16
@@ -49,8 +49,8 @@ ggsave(plot=g, filename="plots/cc_mess.pdf", scale=1.4)
 
 g <- ggplot(subset(d, 
     cc_hash_size == 16384 & scale == 26 & ppn == 16
-    & ( fc_version == 'async' & num_starting_workers == 2048 & aggregator_autoflush_ticks == 5e5
-      | fc_version == 'fc' & num_starting_workers == 2048 & aggregator_autoflush_ticks == 5e5
+    & ( fc_version == 'custom' & num_starting_workers == 2048 & aggregator_autoflush_ticks == 5e5
+      | fc_version == 'distributed' & num_starting_workers == 2048 & aggregator_autoflush_ticks == 5e5
       | fc_version == 'none' & num_starting_workers == 2048 & aggregator_autoflush_ticks == 1e5
     )
     & cc_concurrent_roots == 2048
@@ -62,7 +62,7 @@ g <- ggplot(subset(d,
     color=fc_version,
     shape=fc_version,
     linetype=fc_version,
-    group=x(cc_concurrent_roots,fc_version,aggregator_autoflush_ticks,num_starting_workers),
+    group=x(cc_concurrent_roots,fc_version),
     label=x(aggregator_autoflush_ticks/1e6,num_starting_workers),
   ))+
   # geom_point()+
@@ -70,6 +70,8 @@ g <- ggplot(subset(d,
   # geom_text(size=2,hjust=-0.2,vjust=1)+
   # geom_line()+
   xlab("Nodes")+
+  scale_color_discrete(name="Flat Combining")+
+  scale_linetype_discrete(name="Flat Combining")+
   # facet_grid(scales="free", labeller=label_pretty)+
   ylab("MTEPS")+expand_limits(y=0)+
   my_theme
[…]

### 2013-06-08 doc/pgas13/data/plot.hashset.R

@@ -36,7 +36,7 @@ dset$ops_per_msg <- with(dset, (hashset_insert_ops+hashset_lookup_ops)/(hashset_
 dset$throughput <- with(dset, (hashset_insert_ops+hashset_lookup_ops)/trial_time_mean)
 
 dset$fc_version <- sapply(paste('v',dset$flat_combining,dset$insert_async,sep=''),switch,
-  v0NA='none', v1NA='local only', v11='async', v10='local only', v00='none', v01='?'
+  v0NA='none', v1NA='distributed', v11='async', v10='distributed', v00='none', v01='?'
 )
 
 d.melt <- melt(dset, measure=c("ops_per_msg", "throughput"))
@@ -94,7 +94,7 @@ gg <- ggplot(subset(dset, log_nelems==28 & ppn==16 & num_starting_workers==8192
 dmap$ops_per_msg <- with(dmap, (hashmap_insert_ops+hashmap_lookup_ops)/(hashmap_insert_msgs+hashmap_lookup_msgs))
 dmap$throughput <- with(dmap, (hashmap_insert_ops+hashmap_lookup_ops)/trial_time_mean)
 dmap$fc_version <- sapply(paste('v',dmap$flat_combining,dmap$insert_async,sep=''),switch,
-  v0NA='none', v1NA='local only', v11='async', v10='local only', v00='none', v01='?'
+  v0NA='none', v1NA='distributed', v11='async', v10='distributed', v00='none', v01='?'
 )
 
 gg <- ggplot(subset(dmap, log_nelems==28 & ppn==16 & num_starting_workers==8192

### 2013-06-08 doc/pgas13/data/plot.stack.R

@@ -22,11 +22,11 @@ dq$throughput <- with(dq, nelems/trial_time_mean)
 d$fc <- with(d, x(flat_combining,flat_combining_local_only))
 
 d$fc_version <- sapply(paste('v',d$flat_combining,d$flat_combining_local_only,sep=''),switch,
-  v0NA='none', v1NA='full', v10='full', v11='local only', v00='master only', v01='none'
+  v0NA='none', v1NA='full', v10='full', v11='distributed', v00='centralized', v01='none'
 )
 dq <- within(dq,
 fc_version <- sapply(paste('v',flat_combining,flat_combining_local_only,sep=''),switch,
-  v0NA='none', v1NA='full', v10='full', v11='local only', v00='master only', v01='none'
+  v0NA='none', v1NA='full', v10='full', v11='distributed', v00='centralized', v01='none'
 ))
 
 
@@ -110,6 +110,7 @@ gg <- ggplot(subset(d.c,
   & ( ((struct == 'GlobalQueue' | fraction_push != 0.5) & version == 'fixed_random')
     | version == 'matching_better'
     )
+  & fc_version != 'full'
   ),aes(
     x=nnode,
     y=throughput/1e6,
@@ -152,7 +153,7 @@ gg <- ggplot(subset(d.melt,
   log_nelems==28 & ppn==16 & num_starting_workers==2048
   & (( ( fraction_push != 0.5) & version == 'fixed_random')
     | version == 'matching_better')
-  # & fc_version == 'local only'
+  & grepl('none|distributed', fc_version)
   ),aes(
     x=nnode,
     y=value,

### 2013-06-08 doc/pgas13/intro.tex
rework intro a tiny bit

@@
 
-Prior work on shared memory has explored the scalability challenges of adding more concurrent accessors to shared data structures.
+Much prior work has explored the scalability challenges of adding more concurrent accessors to shared data structures in shared memory systems.
 % One observation was that, when multiple threads of control concurrently access the same object, they can conduct themselves in one of two ways: contend or cooperate.
-In the classic case of contention, either all threads fight to obtain a single global lock, or they attempt a series of potentially-contended atomic operations in the case of a more advanced synchronization strategy such as a lock-free queue.
+In the classic multithreaded case, either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of potentially-contended atomic operations, and retry if they fail.
 In both cases, as the number of concurrent accessors increases, the more they contend and the more failed synchronization operations there are.
@@
 The trick is coming up with a mechanism for delegating that has lower synchronization overhead than the original contention case.
-Prior work showed that a cleverly-implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization.
-This same work makes the additional observation that, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually, which they dub \emph{flat combining}.
+Hendler et al.\cite{flatCombining} showed that a cleverly-implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization.
+The same work makes the additional observation that, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually, which they dub \emph{flat combining}.
 Together, the reduced synchronization cost and combined operations allow even a data structure with a single global lock to scale better than complicated concurrent data structure implementations with fine-grained locking.
@@
 
-The goal of this work is to apply the concept of flat-combining in a PGAS runtime to reduce the cost of maintaining globally consistent data structures in a generic way.
-To evaluate this, we leverage the Grappa runtime~\cite{Nelson:hotpar11-real}, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads.
-With enough concurrent threads, it is actually possible to maintain program order within each thread by blocking on accesses, but still have an opportunity to effectively combine many fine-grained synchronous operations for better performance.
-With the same flat-combining mechanism, multiple global data structures can be implemented efficiently, even with simple locking schemes.
+The goal of this work is to apply flat-combining paradigm to a PGAS runtime to reduce the cost of maintaining sequentially consistent data structures.
+We leverage the Grappa runtime~\cite{Nelson:hotpar11-real}, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads.
+By leveraging lots of concurrency, we find that it is actually possible to scalably enforce strict consistency among threads.
+We leverage Grappa's latency tolerance mechanisms to allow many fine-grained synchronized operations to be combined to reduce communication and achieve higher overall throughput.
+In addition, we show how a generic flat-combining framework can be used to implement multiple scalable sequentially-consistent global data structures.
 
@@
 The next section will describe in more detail the Grappa runtime system that is used to implement flat combining for distributed memory machines.
-Then we explain the flat combining mechanism in more depth and describe how it maps to a PGAS model.
-Next we explain how several data structures are implemented in this framework and show how they perform on simple throughput workloads.
-Finally, we evaluate how they affect performance when used in a simple application kernels.
+Then we explain the flat combining paradigm in more depth and describe how it maps to a PGAS model.
+Next we explain how several data structures are implemented in our framework and show how they perform on simple throughput workloads.
+Finally, we evaluate how this affects application performance in two simple graph analysis kernels.

### 2013-06-08 doc/pgas13/flat-combining.tex
Related work and citations galore.

@@
 An example is shown in Figure~\ref{fig:sync}.\TODO{make sync example}
-The Grappa memory model follows the Data-Race-Free-0 Model (guaranteeing sequential consistency for data-race free programs).
-Being conservative, delegate operations block the calling worker until they have completed globally, ensuring they can be used for synchronization---delegate operations within a task are guaranteed to be globally visible in program order.
+The Grappa memory model is in the style of the C++ memory model~\cite{boehm:drf0,N2480,N2800} memory model (guaranteeing sequential consistency for data-race free programs).
+Being conservative, delegate operations block the calling worker until they have completed globally, ensuring they can be used for synchronization.
+As such, delegate operations within a task are guaranteed to be globally visible in program order.
 Because it is not immediately obvious that this distributed version of flat combining preserves sequential consistency, here we will make arguments for why it does.

### 2013-06-08 doc/pgas13/grappa.tex
Overhaul of FC section.
Related work and citations galore.

@@
 \subsection{Aggregated Communication}
-Communication in Grappa is achieved via \emph{active messages} in conjunction with bulk remote direct memory access (RDMA) operations.
+Communication in Grappa is achieved via \emph{active messages}~\cite{vonEicken92} in conjunction with bulk remote direct memory access (RDMA)~\TODO{cite?} operations.
 Various components of Grappa's runtime contribute different kinds of messages to move and modify remote data, perform synchronization, or spawn tasks.

### 2013-06-08 doc/pgas13/intro.tex
Related work and citations galore.

@@
 
-The goal of partitioned global address space (PGAS) languages and runtimes is to provide the illusion of a single shared memory to a program that actually executes on a number of machines each with their own memory.
+The goal of partitioned global address space (PGAS)~\cite{upc:2005} languages and runtimes is to provide the illusion of a single shared memory to a program that actually executes on a number of machines each with their own memory.
 This allows programmers to write their algorithms without needing to explicitly manage communication.

### 2013-06-08 doc/pgas13/relatedwork.tex
Related work and citations galore.

\section{Related Work}
Attempts to improve the scalability of synchronized data structures are as old as the data structures themselves. Though much attention has gone to lock-free data structures such as the Trieber Stack~\cite{trieber} and Michael-Scott Queue~\cite{msqueue}, a good body of work has explored ways of scaling globally-locked data structures.

Flat combining~\cite{flatCombining} is one of long trail of techniques for allowing a single thread to perform the updates of others. Previous techniques performed combining hierarchically using fixed trees~\cite{yew:combining-trees}, dynamic trees (or ``funnels'')~\cite{funnels}, and randomized trees~\cite{edtrees}. However, the observation of the flat combining work was that in practice, a cleverly-implemented \emph{flat} publication list can outperform more complicated schemes.
Follow-on work introduced parallel flat-combining~\cite{scalableFCQueues}, in which multiple publication lists are combined in parallel and their remainders serialized.
Flat combining was also applied in the NUMA (non-uniform memory access) domain with scalable hierarchical locks~\cite{fcNUMALocks}, which improves on prior work on hierarchical locking by leveraging flat-combining's publication mechanism which amortizes the cost of synchronization.

Our work extends the flat-combining paradigm even further to the PGAS domain, where only software provides the illusion of a physically-distributed memory. The relatively higher cost of communication in PGAS makes flat combining even more compelling. In addition, the physical isolation per node  combined with Grappa's engineered per-core isolation guarantees introduce a novel opportunity to apply combining, which led to our ``combining proxy'' design.

In the distributed memory and PGAS domains, implementing scalable and distributed synchronization is the name of the game.
Besides Grappa, other work has proposed leveraging latency tolerance in PGAS runtimes, including MADNESS~\cite{shet:async-upc} which proposes ways of using asynchronous operations in UPC, and Zhang et al.~\cite{zhang:barnes-hut} who use asynchronous messages in a UPC Barnes Hut implementation to overlap computation with communication and allow for buffering.
Jose et al. introduced UPC Queues~\cite{jose:upc-queues}, a library which provides explicit queues as a replacement for locks to synchronize more efficiently and allow for buffering communication. This work demonstrates a similar use case for synchronized queues as in our work, and in fact uses them in the Graph500 benchmark as well. In contrast, Grappa performs much of the same message aggregation invisibly to the programmer and even data structure designer, and the flat combining framework described in this work can be used to implement many different data structures for use in synchronizing.

### 2013-06-08 doc/pgas13/relatedwork.tex
Ah, that's how Treiber is spelled...

@@
 Attempts to improve the scalability of synchronized data structures are as old as the data structures themselves.
-Though much attention has gone to lock-free data structures such as the Trieber Stack~\cite{trieber} and Michael-Scott Queue~\cite{msqueue}, a good body of work has explored ways of scaling globally-locked data structures.
+Though much attention has gone to lock-free data structures such as the Trieber Stack~\cite{treiber} and Michael-Scott Queue~\cite{msqueue}, a good body of work has explored ways of scaling globally-locked data structures.
 

### 2013-06-08 doc/pgas13/abstract.tex
Write abstract, add placeholder for conclusion.

@@
-TBD
-
 % Structured concurrency (scheduler, cooperative multithreading + communication) enables hierarchical flat-combining to be implemented efficiently for a software PGAS shared-memory solution.
+The implementation of scalable synchronized data structures is notoriously difficult.
+Recent work in shared-memory multicores introduced a new synchronization paradigm called \emph{flat combining} that allows many concurrent accessors to efficiently cooperate to reduce contention on shared locks.
+In this work we introduce this paradigm to a domain where reducing communication is paramount: distributed memory systems.
+We implement a flat combining framework for Grappa, a latency-tolerant PGAS runtime, and show how it can be used to implement synchronized global data structures.
+Even using simple locking schemes, we find that these flat-combining data structures scale out to 64 nodes with 2x-100x improvement in throughput.
+We also demonstrate that this translates to application performance with two simple graph analysis kernels.
+The higher communication cost and structured concurrency of Grappa lead to a new form of distributed flat combining that drastically reduces the amount of communication necessary for maintaining global sequential consistency.

### 2013-06-08 doc/pgas13/conclusion.tex
First crack at conclusion.

\section{Conclusion}
In PGAS implementations there is typically a large discrepancy between the first, simple, description of an algorithm and the final, optimized one. At the core of this work is the desire to have a library of data structures which can be used without worrying about when they will be consistent or whether they will limit scalability.
Bringing findings from shared memory into the PGAS domain brings with it new challenges and opportunities.
We have shown that the additional concurrency that comes with a latency-tolerant runtime, rather than compounding the problem, provides a new opportunity for reducing communication by combining locally.
% perform some of the locality optimizations that finely-tuned implementations would.
This allows flat-combined sequentially-consistent global data structures to scale well out to a thousand cores with millions of concurrent threads.

### 2013-06-08 doc/pgas13/flat-combining.tex
Write abstract, add placeholder for conclusion.
edits to the sequential consistency

@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
   \centering
@@
 
-\paragraph{Sequential Consistency}
-The sequential consistency guarantee is that within a task, operations will be in task order, so to enforce ordering between operations in different tasks, one task must perform an operation that another one can observe.
-An example is shown in Figure~\ref{fig:sync}.\TODO{make sync example}
-The Grappa memory model is in the style of the C++ memory model~\cite{boehm:drf0,N2480,N2800} memory model (guaranteeing sequential consistency for data-race free programs).
-Being conservative, delegate operations block the calling worker until they have completed globally, ensuring they can be used for synchronization.
-As such, delegate operations within a task are guaranteed to be globally visible in program order.
-Because it is not immediately obvious that this distributed version of flat combining preserves sequential consistency, here we will make arguments for why it does.
+\paragraph{Memory Consistency Model}
+
+In the context of this work, sequential consistency guarantees that within a
+task, operations will be in task order and that all tasks will observe the
+same global order of operations. An example is shown in
+Figure~\ref{fig:sync}.\TODO{make sync example}
+
+The Grappa memory model is essentially the same as the C++ memory
+model~\cite{boehm:drf0,N2480,N2800} memory model (guaranteeing sequential
+consistency for data-race free programs). Being conservative, delegate
+operations block the calling worker until they have completed globally,
+ensuring they can be used for synchronization. As such, delegate operations
+within a task are guaranteed to be globally visible in program order and all
+tasks observe the same global order. Because it is not immediately obvious
+that this distributed version of flat combining preserves sequential
+consistency, we will now argue why it does so.
 
@@
 
-The correctness of combining set and map operations is more nuanced.
-Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally.
-Therefore, a batch of these operations can be committed globally in parallel.
-If an insert finds that its key is already in the combiner, it does not need to send its own message.
-However, it must still block until that insert is done, otherwise it may assume that the key has been inserted and perform some other operation that requires the insert to have happened.
+The guarantees of combining set and map operations is more nuanced.
+Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally (i.e., we assume data-race freedom).
+Therefore, a batch of these operations can be committed globally in parallel, since they are guaranteed not to conflict with each other.
+Note that, if an insert finds that its key is already in the combiner, it does not need to send its own message.
+However, it must still block until that insert is done, otherwise it may assume that the key has been inserted and perform some other operation that requires the insert to have happened, effectively reordering operations.
 In the same way, lookups can piggy-back on other lookups of the same key.
@@
 
-These requirements only guarantee sequential consistency for a data structure in isolation.
-Atomicity and data race freedom in the case of multiple data structures must be guaranteed externally, just as in any multithreaded shared-memory program.
+These requirements only guarantee serializability for operations on a data structure in isolation.
+To guarantee sequential consistency with respect to accesses to all data structures, data race freedom  must be guaranteed externally, just as in any multithreaded shared-memory program.
 

### 2013-06-08 doc/pgas13/flat-combining.tex
Could more edits to solidify things in memory model section

@@
 After all local combining has been done, one requesting worker is chosen to perform the combined operation globally.
-In the same figure, Worker 4 becomes the combiner and performs much the same synchronization as in the un-combined case, but is able to push multiple elements with one synchronization.
+In Figure~\ref{fig:stackfc} again, Worker 4 becomes the combiner and performs much the same synchronization as in the un-combined case, but is able to push multiple elements with one synchronization.
 As in the uncombined case, the global lock must still be acquired, and concurrent combined requests must still block and serialize on the master, though a technique later will describe how flat combining can be applied here, too.
@@
 The Grappa memory model is essentially the same as the C++ memory
-model~\cite{boehm:drf0,N2480,N2800} memory model (guaranteeing sequential
-consistency for data-race free programs). Being conservative, delegate
+model~\cite{boehm:drf0,N2480,N2800} (guaranteeing sequential
+consistency for data-race free programs). To be conservative, delegate
 operations block the calling worker until they have completed globally,
@@
 within a task are guaranteed to be globally visible in program order and all
-tasks observe the same global order. Because it is not immediately obvious
+tasks observe the same global order.
+Operations on synchronized data structures must provide the same guarantees.
+Because it is not immediately obvious
 that this distributed version of flat combining preserves sequential
@@
 However, it must still block until that insert is done, otherwise it may assume that the key has been inserted and perform some other operation that requires the insert to have happened, effectively reordering operations.
-In the same way, lookups can piggy-back on other lookups of the same key.
-It is tempting to allow a lookup to be satisfied locally by inspecting the keys to be inserted.
+This ensures that if one wanted to use the insert to signal synchronization, it is guaranteed to have been committed globally before the task continues.
+
+In the same way as inserts, lookups can piggy-back on other lookups of the same key.
+It is tempting to allow a lookup to be satisfied locally by inspecting the keys being inserted.
 However, this allows the local order in which keys were inserted to be observed.

### 2013-06-08 doc/pgas13/evaluation.tex
Related work and citations galore.
Pretty minor wording and citation fixes

@@
 \section{Evaluation}
-To evaluate the impact flat combining has on the global data structures, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and further measured their impact on performance of two simple graph benchmarks.
+To evaluate the impact flat combining has on the global data structures, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and then measure their impact on performance of two simple graph benchmarks.
 
@@
 Figure~\ref{fig:vector} shows the results of the throughput experiments for the global Stack and Queue.
-Results are shown with flat combining completely disabled, only combining at the master core, only combining locally, and full two-level combining.
+Results are shown with flat combining completely disabled, only combining at the master core (``centralize''), and combining locally (``distributed'').
 
-Despite Grappa's automatic aggregation, without combining, both the stack and queue completely fail to scale because all workers' updates must serialize on the master core.
-Though combining on the master core alleviates some of the serialization, it does not provide significant benefit because all of the synchronization must still be performed on one core.
-However, with local flat combining, synchronization is able to be done mostly in parallel, with much less-frequent bulk synchronization at the master.
+Despite Grappa's automatic aggregation, without combining, both the stack and queue completely fail to scale because all workers' updates must serialize.
+Though centralized combining alleviates some of the serialization, its benefit is limited because all synchronization must still be performed on one core.
+However, with local flat combining, synchronization is able to be done mostly in parallel, with less-frequent bulk synchronization at the master.
 
-On the mixed workload, the stack is able to do matching locally, allowing it to reduce the amount of communication drastically, greatly improving its performance.
+On the mixed workload, the stack is able to do matching locally, allowing it to reduce the amount of communication drastically, greatly improving its performance on the mixed workload.
 Figure~\ref{fig:stack_stats} corroborates this, showing that the amount of combining that occurs directly correlates with the observed throughput.
@@
 We also experimented with performing both forms of combining together; however, combining at the master did not affect performance significantly.
-This was somewhat surprising as combined updates will still force others to wait.
+This was somewhat surprising as combined updates still force others to wait.
 However, with both levels of combining enabled, relatively few requests are combined at the master---for the all-push workload, on the order of a thousand, and for the mixed workload on the stack, the average is less than 20.
@@
 However, the queue is unable to do matching locally, and in fact, the mixed workload performs worse because the current implementation serializes combined pushes and combined pops.
-This restriction could be lifted with more careful synchronization at the master core allowing pushes and pops to proceed in parallel as long as there are enough elements in the queue.
+This restriction could be lifted with more careful synchronization at the master core.
 
@@
 
-Without combining, performance scales out to 32 nodes because synchronization is distributed over hash cells, but drops off as the number of destinations increases.
-
-Figure~\ref{fig:hash_perf} shows the throughput results for the GlobalHashSet and GlobalHashMap.
+% Without combining, performance scales out to 32 nodes because synchronization is distributed over hash cells, but drops off as the number of destinations increases.
+Figure~\ref{fig:hash_perf} shows the throughput results for the Set and Map.
 Both data structures synchronize at each hash cell, which allows them to scale fairly well even without combining.
-However, after 32 nodes, scaling drops off significantly due to increased number of concurrent accessors and destinations.
+However, after 32 nodes, scaling drops off significantly due to the increased number of destinations.
 Combining allows duplicate inserts and lookups to be eliminated, so performs better the smaller the key range. This reduction in message traffic allows scaling to continue out to 64 nodes.
@@
 The original goal was to have scalable global data structures in the Grappa library that could be used in applications.
-Naive implementations of the data structures are insufficient for scaling, so HPC applications often implement their own highly customized versions.
-However, this forces each application to carefully reason about the ways in which the structures will be accessed.
-Additionally, many optimizations rely on relaxing global consistency in a way that does not affect the program.
+Generic implementations of data structures are often insufficient for scaling, so HPC applications often implement their own customized versions.
+However, this forces each application to carefully reason about the ways in which the structures will be accessed, and often relies on relaxing consistency.
 The Grappa data structures are synchronized to provide the most general use and match the expectations of programmers and algorithms.
+In these evaluations we compare the flat-combining structures against tuned, relaxed-consistency versions.
 
@@
 \end{figure}
-The first application kernel is the Graph500 Breadth-First-Search (BFS) benchmark.
+The first application kernel is the Graph500 Breadth-First-Search (BFS) benchmark~\cite{graph500list}.
 This benchmark does a search starting from a random vertex in a synthetic graph and builds a search tree of parent vertices for each vertex traversed during the search.
@@
 The BFS algorithm contains a global queue which represents the frontier of vertices to be visited in each level.
-Our implementation employs the direction-optimizing algorithm by Beamer et al.\cite{Beamer:Graph500} which performs particularly well for the scale-free RMAT graphs generated by the benchmark.
+Our implementation employs the direction-optimizing algorithm by Beamer et al.~\cite{Beamer:Graph500} which performs particularly well for the scale-free RMAT graphs generated by the benchmark.
 The frontier queue in BFS is amenable to further optimization to take advantage of the fact that the algorithm does pushes and pops in separate phases, allowing consistency to be relaxed.
@@
 The simple queue implementation without flat combining is completely unscalable.
-However, the flat combining queue tracks the highly tuned asynchronous version.
-This illustrates that providing a safe, synchronized data structure for initially developing algorithms for PGAS is possible, while further optimizations can be applied incrementally.
+However, the flat combining queue tracks the tuned version.
+This illustrates that providing a safe, synchronized data structure for initially developing algorithms for PGAS is possible without sacrificing scaling.
+% , while further optimizations can be applied incrementally.
 
@@
 \end{figure}
-Connected Components (CC) is another simple graph analysis kernel that illustrates another use of global data structures in irregular applications.
+Connected Components (CC) is another simple graph analysis kernel that illustrates a different use of global data structures in irregular applications.
 We implement the three-phase CC algorithm~\cite{mtgl} which was designed for the massively-parallel MTA-2 machine.
-In first phase, the algorithm begins many traversals in parallel from random starting vertices, labeling vertices with the root vertex.
-Whenever two traversals encounter each other, their searches are pruned and an edge between the two roots is inserted in a set.
+In the first phase, parallel traversals attempt to claim vertices and label them.
+Whenever two traversals encounter each other, an edge between the two roots is inserted in a set.
 After all edges have been traversed in this way, the set of edges forms a new, typically much smaller, graph.
-The second phase performs the classical Shiloach-Vishkin parallel algorithm\cite{shiloach1982n} on this reduced graph, and the final phase propagates the component labels out to the full graph.
-The creation of the reduced-graph edge set dominates the runtime of this algorithm, so improving the implementation of the set operations has a significant impact on performance.
-As in the case of BFS, further optimizations involving relaxation of consistency can be applied, in this case, to the global set.
-Therefore, we compare our straightforward implementation using the generic GlobalHashSet with and without flat combining against a tuned asynchronous implementation.
+The second phase performs the classical Shiloach-Vishkin parallel algorithm~\cite{shiloach1982n} on this reduced graph, and the final phase propagates the component labels out to the full graph.
+The creation of the reduced-graph dominates the runtime of this algorithm, in includes many repeated inserts at the boundary between traversals.
+As in BFS, further optimizations involving relaxation of consistency can be applied, in this case to the Set.
+Therefore, we compare our straightforward implementation using the generic Set with and without flat combining against a tuned asynchronous implementation.
 
-The results in Figure~\ref{fig::cc_perf} show that none of these three implementations scale very well out to 64 nodes.
-However, performing combining does improves the performance of the algorithm overall and improves scaling.
+The results in Figure~\ref{fig:cc_perf} show that none of these three scale well out to 64 nodes.
+However, performing combining does improve the performance over the uncombined case.
 The tuned version outperforms the synchronous version because it is able to build up most of the set locally on each core before merging them at the end of the first phase.
-An implementation of the GlobalHashSet that did not provide synchronized semantics could potentially relax consistency in a more general way, but this is left for future work.
+An implementation that did not provide synchronized semantics could potentially relax consistency in a more general way, but this is left for future work.

### 2013-06-08 doc/pgas13/flat-combining.tex
Pretty minor wording and citation fixes

@@
 On the core where global synchronization is occurring, there must be synchronization between the multiple synchronization requests arriving from other cores.
-
 % At this point we have traded off one remote synchronization per operation for an additional local synchronization per operation and some number fewer remote synchronizations (depending on the amount of combining that occurs).
@@
 However, because of the way the Grappa runtime schedules computation, it is actually possible to get the correct mutual exclusion/atomicity without any additional synchronization overhead.
-
 % By design, each core in Grappa operates independently of the other cores.
@@
 \paragraph{Memory Consistency Model}
-
 In the context of this work, sequential consistency guarantees that within a
 task, operations will be in task order and that all tasks will observe the
-same global order of operations. An example is shown in
-Figure~\ref{fig:sync}.\TODO{make sync example}
+same global order of operations.
+% An example is shown in Figure~\ref{fig:sync}.\TODO{make sync example}
 

### 2013-06-08 doc/pgas13/grappa.tex
Pretty minor wording and citation fixes

--- 2013-06-06 doc/pgas13/grappa.tex
@@
 These programs typically perform many fine-grained accesses to disparate sources of data, which is a problem even at multicore scales, but is further exacerbated on distributed memory machines.
-It is often the case that naively porting an application to a PGAS system results in excessive communication and poor access patterns, but this class of applications defies typical optimization techniques such as data partitioning, shadow objects, and bulk-synchronous communication transformations.
+It is often the case that naively porting an application to a PGAS system results in excessive communication and poor access patterns~\cite{pgas-analysis}, but this class of applications defies typical optimization techniques such as data partitioning, shadow objects, and bulk-synchronous communication transformations.
 Luckily, applications in this class have another thing in common: abundant amounts of data parallelism.
@@
 
-Grappa is a PGAS-style runtime for commodity clusters which has been designed from the ground up to achieve high performance on irregular applications.
+Grappa is a global-view PGAS-style runtime for commodity clusters which has been designed from the ground up to achieve high performance on irregular applications.
 The key is latency tolerance---long-latency operations such as remote communication can be tolerated by switching to another concurrent thread of execution.
@@
 \subsection{Aggregated Communication}
-Communication in Grappa is achieved via \emph{active messages} in conjunction with bulk remote direct memory access (RDMA) operations.
+Communication in Grappa is achieved via \emph{active messages}~\cite{vonEicken92} in conjunction with bulk remote direct memory access (RDMA) operations.
 Various components of Grappa's runtime contribute different kinds of messages to move and modify remote data, perform synchronization, or spawn tasks.
@@
 In the PGAS style, Grappa has a notion of a global address space which is partitioned across the physical memories of the nodes in a cluster.
-Each core owns a portion of memory, which is further divided for three different uses:
-\begin{itemize}
-  \item Execution stacks for the core's Workers.
-  \item A core-local heap.
-  \item A slice of the global shared heap.
-\end{itemize}
+Each core owns a portion of memory, which is divided among execution stacks for the core's Workers, a core-local heap, and a slice of the global heap.
+
 All of these can be addressed by any core in the system using a Grappa \texttt{GlobalAddress}, which encodes both the owning core and the address on that core.
@@
 In this way, the runtime maintains strict isolation even between processes running on the same physical memory.
-At the programming level, however, this is hidden behind higher-level remote operations, which in Grappa are called ``delegates''.
+At the programming level, however, this is hidden behind higher-level remote operations, which in Grappa are called ``delegate'' operations.
 Figure~\ref{fig:system} shows an example of a delegate \texttt{read()} which blocks the calling task and sends a message to the owner, who sends a reply with the data and wakes the caller.

### 2013-06-08 doc/pgas13/intro.tex
Pretty minor wording and citation fixes

@@
 Luckily, there exists a large body of work solving these issues in physically-shared memory which the PGAS community can leverage.
-The opportunity here is that, in a distributed setting, the costs of communication and synchronization operations are different, so different performance trade-offs will be made.
+In a distributed setting, the costs of communication and synchronization operations are different, providing an opportunity for findings in shared-memory to be re-evaluated.
 
@@
 The trick is coming up with a mechanism for delegating that has lower synchronization overhead than the original contention case.
-Hendler et al.\cite{flatCombining} showed that a cleverly-implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization.
-The same work makes the additional observation that, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually, which they dub \emph{flat combining}.
-Together, the reduced synchronization cost and combined operations allow even a data structure with a single global lock to scale better than complicated concurrent data structure implementations with fine-grained locking.
-Further, the same publication list mechanism can be applied to other data structures, and all that must be customized is the particular way in which operations for a given data structure can be combined.
+Hendler et al.\cite{flatCombining} showed that a cleverly-implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization cost.
+In addition, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually.
+Together, the reduced synchronization cost and combined operations allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.
+The same publication list mechanism can be applied to multiple data structures; all that must be customized is the particular way in which the data structure's operations can be combined.
 
@@
 
-The goal of this work is to apply flat-combining paradigm to a PGAS runtime to reduce the cost of maintaining sequentially consistent data structures.
+The goal of this work is to apply the flat-combining paradigm to a PGAS runtime to reduce the cost of maintaining sequentially consistent data structures.
 We leverage the Grappa runtime~\cite{Nelson:hotpar11-real}, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads.
-By leveraging lots of concurrency, we find that it is actually possible to scalably enforce strict consistency among threads.
+By leveraging lots of concurrency, we find that it is actually possible to scalably enforce strict sequential consistency.
 We leverage Grappa's latency tolerance mechanisms to allow many fine-grained synchronized operations to be combined to reduce communication and achieve higher overall throughput.
-In addition, we show how a generic flat-combining framework can be used to implement multiple scalable sequentially-consistent global data structures.
+In addition, we show how a generic flat-combining framework can be used to implement multiple global data structures.
 
@@
 The next section will describe in more detail the Grappa runtime system that is used to implement flat combining for distributed memory machines.
-Then we explain the flat combining paradigm in more depth and describe how it maps to a PGAS model.
+We then explain the flat combining paradigm in more depth and describe how it maps to a PGAS model.
 Next we explain how several data structures are implemented in our framework and show how they perform on simple throughput workloads.

### 2013-06-08 doc/pgas13/relatedwork.tex
Pretty minor wording and citation fixes

@@
 Attempts to improve the scalability of synchronized data structures are as old as the data structures themselves.
-Though much attention has gone to lock-free data structures such as the Trieber Stack~\cite{treiber} and Michael-Scott Queue~\cite{msqueue}, a good body of work has explored ways of scaling globally-locked data structures.
+Though much attention has gone to lock-free data structures such as the Treiber Stack~\cite{treiber} and Michael-Scott Queue~\cite{msqueue}, a good body of work has explored ways of scaling globally-locked data structures.
 

### 2013-06-09 doc/pgas13/evaluation.tex
incorporating Mark's feedback and stuff

@@
 Figure~\ref{fig:vector} shows the results of the throughput experiments for the global Stack and Queue.
-Results are shown with flat combining completely disabled, only combining at the master core (``centralize''), and combining locally (``distributed'').
+Results are shown with flat combining completely disabled, only combining at the master core (``centralized''), and combining locally (``distributed'').
 

### 2013-06-09 doc/pgas13/flat-combining.tex
incorporating Mark's feedback and stuff

--- 2013-06-08 doc/pgas13/flat-combining.tex
@@
 Reduced synchronization comes from delegating work to another core.
-To illustrate the issue, consider the shared synchronous queue, shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock.
+To illustrate the issue, consider the shared synchronous stack shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock.
 Without flat combining, whenever a thread attempts to push something on the stack, it must acquire the lock, put its value into the storage array, bump the top, and then release the lock.
 When many threads contend for the lock, all but one will fail and will have to retry.
-Each try forces a memory fence for consistency and consumes bandwidth, and as the number of threads increases, the fraction of successes plummets.
+Each attempt forces an expensive memory fence and consumes bandwidth, and as the number of threads increases, the fraction of successes plummets.
 Under flat combining, instead, threads add requests to the publication list.
 They each try to acquire the lock, and the one that succeeds becomes the combiner.
-Instead of retrying, the rest spin waiting for their request to be filled.
+Instead of retrying, the rest spin on their request waiting for it to be filled.
 The combiner walks the publication list, performs all of the requests, and when done, releases the lock.
@@
 However, if a thread performs multiple operations on the stack, it can leave its publication record in the list and amortize the synchronization cost.
+This publication list mechanism can be re-used in different data structures, saving each one from needing to implement its own clever synchronization mechanism.
 
@@
 In the case of the stack example shown in Figure~\ref{fig:fc_shared_mem}, as it walks the list, Thread 4 keeps track of the operations on its own temporary stack.
-When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills each of their records and allows each of them to proceed.
-After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the remaining two elements are pushed onto the top of the stack, and the requesting threads' records are marked as complete.
-In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched locally.
+When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills both of their records and allows them to proceed.
+After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the two unmatched pushes are added to the top of the stack, and the requesting threads' records are marked as complete.
+In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched.
 
@@
 
-To illustrate how flat combining can be applied to Grappa, we must first describe what a global data structure might look like.
+To illustrate how flat combining can be applied to Grappa, we must first describe what a global data structure looks like.
 Figure~\ref{fig:stacknofc} shows a simple PGAS translation of the shared-memory stack from earlier.
@@
 All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object.
-To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read()} delegate described earlier, blocks the task until complete.
+To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read} delegate described earlier, blocks the task until complete.
 Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}, the task must send a message to the master to acquire the lock.
@@
 Because of Grappa's user-level blocking mechanisms, blocked requests do not have to retry or busy-wait.
-However, all thousand workers on each core all perform this synchronization and serialize on the master core.
+However, all workers on each core perform this synchronization and serialize on a single core, causing that core to become the bottleneck.
 
@@
 
-\paragraph{Distributed Combining}
-Instead of all workers on a core sending independent synchronization messages and putting the burden all on the master, what if each core did its own combining first, and synchronized with the master in bulk?
+\paragraph{Distributed Combining.}
+Instead of all workers on a core sending independent synchronization messages and putting the burden all on one core, what if each core did its own combining first, and synchronized with the master in bulk?
 Distributing the combining to each core allows the majority of the work to be performed in parallel and without communication.
 As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and allows operations to be composed in data structure-specific ways for even greater performance.
-% allows synchronization to happen in parallel on all the cores, 
-% Flat combining in Grappa is about allowing synchronization and  distributing synchronization 
+
 % In a way, Grappa's automatic message aggregation is already trying to reduce synchronization traffic; however, the generic mechanism for serializing and packing has a cost, and in the end, the operations must still serialize on the master.
@@
 
-% In order for workers to cooperate locally, a local ``proxy'' object is allocated in the core-private heap. This proxy serves as the
+In distributed flat-combining, each core builds its own publication list to track all the operations waiting to be committed to the global data structure.
+In Grappa, for each global data structure, a local \emph{proxy} object is allocated from the core-local heap to intercept requests.
+Conceptually, workers add their requests to a local publication list in the proxy, then one is chosen to do the combining, and the rest block until their request is satisfied.
+However, because Grappa's scheduler is non-preemptive, each worker has atomicity ``for free'' until it performs a blocking operation (such as communication).
+This means that an explicit publication list with clever synchronization is unnecessary.
+Instead, workers perform their own combining, merging their operations directly into the local proxy, and block if unable to satisfy their requests immediately.
+The proxy structure is chosen to be able to compactly aggregate operations and efficiently perform matching in cases where it is allowed.
+Figure~\ref{fig:stackfc} shows how in the global stack example, pushes and pops are matched on the proxy's local stack.
 
-In distributed flat-combining, each core needs its own publication list to track all of its outstanding operations to be performed on the global data structure.
-In Grappa, this takes the form of a local \emph{proxy} object allocated from the core-local heap.
-Conceptually, workers add their requests to the local publication list, one is chosen to do the combining, and the rest block until their request is satisfied.
-However, because workers are scheduled cooperatively, each worker has atomic access to the local proxy ``for free'', so an explicit publication list is unnecessary.
-Instead, each worker performs its own combining, pushing the 
-, and blocks if it was unable to satisfy its request immediately.
-Figure~\ref{fig:stackfc} shows how in the global stack example, pushes and pops can be matched locally, avoiding global synchronization for those operations.
-
-After all local combining has been done, one requesting worker is chosen to perform the combined operation globally.
-In Figure~\ref{fig:stackfc} again, Worker 4 becomes the combiner and performs much the same synchronization as in the un-combined case, but is able to push multiple elements with one synchronization.
-As in the uncombined case, the global lock must still be acquired, and concurrent combined requests must still block and serialize on the master, though a technique later will describe how flat combining can be applied here, too.
+After all local combining has been done, one requesting worker is chosen to commit the combined operation globally.
+Back to the example in Figure~\ref{fig:stackfc}, Worker 4 becomes the combiner and performs much the same synchronization as in the uncombined case, but is able to push multiple elements with one synchronization.
+The global lock must still be acquired, so concurrent combined requests (from different cores) must block and serialize on the master.
 Combining in this way allows synchronization to be distributed and performed in parallel, coarsening the granularity of the actual global synchronization and reducing the amount of serialization that occurs.
 
-% In most cases, this means communicating with whichever core holds the ``lock-protected'' field that maintains sequential consistency.
-In the case of a global stack, this is the ``master'' \emph{top} pointer which must be incremented on pushes and decremented on pops.
-On the core where global synchronization is occurring, there must be synchronization between the multiple synchronization requests arriving from other cores.
-
-% At this point we have traded off one remote synchronization per operation for an additional local synchronization per operation and some number fewer remote synchronizations (depending on the amount of combining that occurs).
-This would likely result in a net benefit due to local synchronization being inherently cheaper.
-However, because of the way the Grappa runtime schedules computation, it is actually possible to get the correct mutual exclusion/atomicity without any additional synchronization overhead.
-
-% By design, each core in Grappa operates independently of the other cores.
-Workers are scheduled cooperatively, so each worker can assume atomicity until it performs an operation that yields to the scheduler, such as a remote call.
-Therefore workers within a core can cooperate without any explicit synchronization operations requiring memory fences.
-This allows them to publish requests to the local proxy very inexpensively, eliminating the need for a complex synchronization-amortizing queue.
-This has the added benefit that there will be no unused publication records, which could have become a problem with thousands of workers.
-Further, remote operations are processed through the same cooperative-multithreading scheduler on the remote side, so combined operations are trivially serialized and atomic.
-
 \paragraph{Memory Consistency Model}
-
+\label{sec:memory-model}
 In the context of this work, sequential consistency guarantees that within a
 task, operations will be in task order and that all tasks will observe the
-same global order of operations. An example is shown in
-Figure~\ref{fig:sync}.\TODO{make sync example}
+same global order of operations.
 
@@
 that this distributed version of flat combining preserves sequential
-consistency, we will now argue why it does so.
+consistency, we will now argue why it does.
 
-% \begin{itemize}
-%   \item Block the calling worker until the operation is guaranteed to have a valid global order.
-%   \item Local combining must not allow workers to observe a different order than is committed globally.
-% \end{itemize}
 To preserve sequentially-consistent semantics for flat-combined operations, they must obey a consistent global order.
 To start with, local combined operations must have a serializable order among each other; this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
-When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map).
-The global order, then, is essentially the concatenation of the cores' sequential orders.
+When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the ``master'' core for the Stack).
+% haven't discussed queue, set, or map yet...
+%(the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map). The global order, then, is essentially the concatenation of the cores' sequential orders.
 For this ``concatenation'' to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.
 
-In the case of the Stack and Queue, this guarantee comes from applying a batch of push or pop operations atomically in a contiguous block on the global data structure.
+In the case of a stack or queue, this guarantee comes from applying a batch of push or pop operations atomically in a contiguous block on the global data structure.
 The second level of combining at the master preserves this as well by only allowing updates to proceed when they will not conflict and blocking all of them until they have completed.
-Matching pushes and pops locally is one exception to the rule that operations must block until globally committed.
+For the \emph{stack}, matching pushes and pops locally is one exception to the rule that operations must block until globally committed.
 Because a pop ``destroys'' the corresponding push, these operations together are independent of all others and can conceptually be placed anywhere in a total global ordering, so they can be safely matched locally.
 
-The guarantees of combining set and map operations is more nuanced.
-Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally (i.e., we assume data-race freedom).
+Combining set and map operations expose more nuances in the consistency guarantees.
+Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally.
 Therefore, a batch of these operations can be committed globally in parallel, since they are guaranteed not to conflict with each other.
@@
 
-% (wrong!) A lookup can be satisfied locally provided that the insert it relies on is guaranteed to eventually be committed globally.
-This allows the task to continue.
-If it does another lookup to a different key that is satisfied locally, then that task has observed one particular ordering of the two inserts.
-When those inserts commit globally, they will proceed in parallel to their respective hashed locations.
-Another task could
-
-% The combining operations that are performed locally among requests on a single core must preserve the illusion of sequential order; this is unchanged from shared-memory flat-combining.
-When combined operations are applied globally, they are serialized at the ``master'' core for a given synchronization context.
-The global order is essentially the concatenation of each core's batch of serialized accesses.
-% In the case where operations are satisfied locally (by matching up with other local operations), by definition these must be independent of other operations, so they can conceptually be placed anywhere in the global ordering (or even ignored completely in the global ordering).
-
-% In the spirit of the Data-Race-Free-0 Model (guaranteeing sequential consistency for data-race free programs), operations on the global data structure are considered to be made atomic by a conceptual global ``lock'' on the data structure.
+Whenever a global commit starts, a fresh combiner must be created for subsequent operations to use.
+If they were to use the old combiner's state to satisfy requests locally, it would violate global ordering because the local object would not reflect other cores' updates.
+This guarantees, for example, that destructive operations only happen once (i.e., a worker cannot locally pop the same value that another core's combined operation does).
 
@@
 \section{Grappa FC Framework}
-% One of the strengths of flat combining is that the same combining mechanism can be applied in multiple data structures. Each structure must only specify how its operations compose.
-% Many different data structures can integrate flat combining into their operation using the same delegation mechanism, making it easier to implement concurrent data structures.
 In order to leverage the flat-combining paradigm in Grappa, we implemented a generic framework which can be used to improve performance for a number of global data structures.
@@
 
-% As discussed before, Grappa's distributed combining takes a different approach to expressing how operation locally  flat combining design takes a different approach to performing 
 The Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine.
-As mentioned before, the local ``publication list'' is trivial to access atomically in Grappa due to workers being cooperatively scheduled.
+As mentioned before, the local ``publication list'' is trivial to access atomically in Grappa's cooperative scheduling.
 Instead, in the Grappa FC Framework, a \emph{proxy} object is instantiated ahead of time on each core and each worker merges its update into that structure before blocking or becoming the combiner.
-% In the original work, an explicit publication list is built up, and when the combiner scans over the list, it has a structure that it uses to accumulate requests in a logical way.
-In the case of the stack, for instance, it is a thread-local stack, which it can use to immediately match up pushes and pops.
-In the Grappa implementation, each worker's actions are atomic anyway, so rather than creating an explicit request list, the combiner object (or \emph{proxy}) is instantiated ahead of time for the data structure, and each worker does its own local combining into that structure before blocking.
-This reduces the space blowout of a large dynamically-sized publication list (with associated caching issues), and does not overly affect the expression of combining.
 In this style, each global data structure must define the following:
@@
   }); // blocks until response arrives
-}\end{lstlisting}
+}
+\end{lstlisting}
 \caption{Snippet of code for pushing to a global stack (without flat combining).}
@@
 While a combining operation is in flight, new requests are likely to continue to accumulate.
-The framework transparently wraps the proxy object so that when it starts a sync on one, it can direct requests to a fresh combiner.
+The framework transparently wraps the proxy object so that when it starts a \texttt{sync} on one, it can direct requests to a fresh combiner.
+This is done by hiding instances of proxy objects behind a C++ smart-pointer-like object which provides the pointer to the current combiner, and whenever it detects that it should send, allocates a new combiner and points subsequent references to that.
 It is important to note that combiners must start out fresh each time---it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them.
-
-% The flat combining framework transparently wraps the \emph{proxy} object.
-Behind this layer of indirection, it manages multiple copies of the proxy so that, as discussed above, there is always at least one being committed and one to receive additional combining operations.
-
-% explain simple non-combined global data structures in Grappa section...
+We refer readers to Section~\ref{sec:memory-model} for why this must be so.
 
@@
 The proxy tracks pushes locally in the \texttt{pushed\_values} array.
-When \texttt{pop()} is called, if there are pushes in the local array, it immediately takes the value and wakes the last worker to push.
+When \texttt{pop} is called, if there are pushes in the local array, it immediately takes the value and wakes the last worker to push.
 Otherwise, it adds a pointer to a location on its stack where the \texttt{sync} operation, will eventually write the result.
@@
 \subsection{Global Set, and Map}
-The Grappa GlobalSet uses a simple chaining design, implemented with a global array of cells (allocated from the global heap), which are partitioned evenly among all the cores, and indexed with a globally-agreed hash function.
+The Grappa GlobalSet uses a simple chaining design, implemented with a global array of cells (allocated from the global heap), which are partitioned evenly among all the cores, and indexed with a hash function.
 Our flat-combining version supports both \texttt{insert} and \texttt{lookup}.
@@
 Similar to the case with pops, for lookups, we must keep track of pointers for where to put the results of lookup operations, which the proxy does with a hash map (again from the C++ standard library) from keys to a list of pointers.
-As discussed in Section~\ref{sec:sc}, matching lookups with inserts locally would force a particular sequential order, so instead we only allow matching inserts with inserts and lookups with lookups.
+As discussed in Section~\ref{sec:memory-model}, matching lookups with inserts locally would force a particular sequential order, so instead we only allow matching inserts with inserts and lookups with lookups.
 This allows \texttt{sync} to simply issue all of the inserts and lookups in parallel and block until all have completed.
 
-Our implementation of the GlobalMap matches that of the Set, but of course stores values of a given type.
+Our implementation of the GlobalMap matches that of the Set but of course stores values of a given type.
 This does not change the requirements of the proxy except that the return values of lookups must be a copy of the value rather than just a boolean.

### 2013-06-09 doc/pgas13/grappa.tex
incorporating Mark's feedback and stuff

@@
 In particular, throughput of random accesses to remote memory can be improved by delaying communication requests and aggregating them into larger packets.
-Highly-tuned implementations of irregular applications in shared-memory, PGAS and message-passing paradigms, typically end up implementing constructs to allow for aggregation and leverage data parallelism.
-The goal of Grappa is to provide a simple-to-use PGAS abstraction with mechanisms for expressing the concurrency needed to achieve high performance.
+Highly-tuned implementations of irregular applications in shared-memory, PGAS and message-passing paradigms, typically end up implementing similar constructs.
+Grappa includes these as part of its core infrastructure and simply asks the programmer to express concurrency which it can leverage to provide performance.
 
-The programming interface, implemented as a C++11 library, provides high-level operations to access and synchronize through global shared memory, task spawning capabilities and parallel loop constructs for expressing concurrency.
+Grappa's programming interface, implemented as a C++11 library, provides high-level operations to access and synchronize through global shared memory, task spawning capabilities and parallel loop constructs for expressing concurrency.
 In addition, the Grappa ``standard library'' includes functions to manipulate a global heap as well as provides several synchronized global data structures.
@@
 \subsection{Aggregated Communication}
-Communication in Grappa is achieved via \emph{active messages}~\cite{vonEicken92} in conjunction with bulk remote direct memory access (RDMA) operations.
+The most basic unit of communication in Grappa is an \emph{active message}~\cite{vonEicken92}.
+To make efficient use of the networks in high-performance systems, which typically achieve maximum bandwidth only for messages on the order of 64 KB, all communication in Grappa is sent via an aggregation layer that automatically buffers messages to the same destination.
+% Communication in Grappa is achieved via \emph{active messages}~\cite{vonEicken92} in conjunction with bulk remote direct memory access (RDMA) operations.
 Various components of Grappa's runtime contribute different kinds of messages to move and modify remote data, perform synchronization, or spawn tasks.
@@
 
-All of these can be addressed by any core in the system using a Grappa \texttt{GlobalAddress}, which encodes both the owning core and the address on that core.
+All of these can be addressed by any core in the system using a \texttt{GlobalAddress}, which encodes both the owning core and the address on that core.
 Additionally, addresses into the global heap are partitioned in a block-cyclic fashion, so that a large allocation is automatically distributed among many nodes.
@@
 
-All accesses to memory owned by another core must be done by that core via a message.
-In this way, the runtime maintains strict isolation even between processes running on the same physical memory.
+All accesses to memory owned by another core must be done by that core via a message, even enforcing strict isolation between processes on the same physical memory.
 At the programming level, however, this is hidden behind higher-level remote operations, which in Grappa are called ``delegate'' operations.
-Figure~\ref{fig:system} shows an example of a delegate \texttt{read()} which blocks the calling task and sends a message to the owner, who sends a reply with the data and wakes the caller.
+Figure~\ref{fig:system} shows an example of a delegate \texttt{read} which blocks the calling task and sends a message to the owner, who sends a reply with the data and wakes the caller.

### 2013-06-09 doc/pgas13/relatedwork.tex
incorporating Mark's feedback and stuff

@@
 
-Flat combining~\cite{flatCombining} is one of long trail of techniques for allowing a single thread to perform the updates of others.
+Flat combining~\cite{flatCombining} is the latest of many techniques for allowing a single thread to perform the updates of others.
 Previous techniques performed combining hierarchically using fixed trees~\cite{yew:combining-trees}, dynamic trees (or ``funnels'')~\cite{funnels}, and randomized trees~\cite{edtrees}.
@@
 The relatively higher cost of communication in PGAS makes flat combining even more compelling.
-In addition, the physical isolation per node combined with Grappa's engineered per-core isolation guarantees introduce a novel opportunity to apply combining, which led to our ``combining proxy'' design.
+In addition, the physical isolation per node combined with Grappa's engineered per-core isolation guarantees, introduce a novel opportunity to apply combining, which led to our ``combining proxy'' design.
 
@@
 introduced UPC Queues~\cite{jose:upc-queues}, a library which provides explicit queues as a replacement for locks to synchronize more efficiently and allow for buffering communication.
-This work demonstrates a similar use case for synchronized queues as in our work, and in fact uses them in the Graph500 benchmark as well.
+This work demonstrates a similar use case for synchronized queues as in our work, and uses them with the Graph500 benchmark as well.
 In contrast, Grappa performs much of the same message aggregation invisibly to the programmer and even data structure designer, and the flat combining framework described in this work can be used to implement many different data structures for use in synchronizing.

### 2013-06-11 doc/pgas13/grappa.tex
Clean up intro and Grappa sections a bit.

@@
 It is often the case that naively porting an application to a PGAS system results in excessive communication and poor access patterns~\cite{pgas-analysis}, but this class of applications defies typical optimization techniques such as data partitioning, shadow objects, and bulk-synchronous communication transformations.
-Luckily, applications in this class have another thing in common: abundant amounts of data parallelism.
+Luckily, applications in this class have another thing in common: abundant amounts of data access parallelism.
 This parallelism can be exploited in a number of different ways to improve overall throughput of these applications.
@@
 
-Grappa is a global-view PGAS-style runtime for commodity clusters which has been designed from the ground up to achieve high performance on irregular applications.
-The key is latency tolerance---long-latency operations such as remote communication can be tolerated by switching to another concurrent thread of execution.
+Grappa is a global-view PGAS runtime for commodity clusters which has been designed from the ground up to achieve high performance on irregular applications.
+The key is latency tolerance---long-latency operations such as reads of remote memory can be tolerated by switching to another concurrent thread of execution.
 Given abundant concurrency, there are opportunities to increase throughput by sacrificing latency.
 In particular, throughput of random accesses to remote memory can be improved by delaying communication requests and aggregating them into larger packets.
+
 Highly-tuned implementations of irregular applications in shared-memory, PGAS and message-passing paradigms, typically end up implementing similar constructs.
@@
 Grappa's programming interface, implemented as a C++11 library, provides high-level operations to access and synchronize through global shared memory, task spawning capabilities and parallel loop constructs for expressing concurrency.
-In addition, the Grappa ``standard library'' includes functions to manipulate a global heap as well as provides several synchronized global data structures.
+In addition, the Grappa ``standard library'' includes functions to manipulate a global heap, stock remote synchronization operations such as \emph{compare-and-swap}, and several synchronized global data structures.
 The following sections will explain the execution model of Grappa and the current C++ programming interface.
@@
 \subsection{Tasks and Workers}
-Core to Grappa is the expression of concurrency.
-The Grappa runtime has a user-level threading system that uses an efficient context-switching mechanism with prefetching to allow it to scale up to thousands of threads on a single core with minimal increase in context-switch time.
-This is exposed at the programming level as a simple task-parallel model.
+Grappa uses a task-parallel programming model to make it easy to express concurrency.
 A \emph{task} is simply a function object with some state and a function to execute.
-In the runtime, ``Worker'' threads pull these programmer-specified tasks off of a queue and execute them to completion.
 Tasks may block on remote accesses or synchronization operations.
-When tasks block, the worker thread executing them is simply suspended and consumes no computational resources until woken again by the completion of communication or synchronization.
+The Grappa runtime has a lightweight threading system that uses prefetching to scale up to thousands of threads on a single core with minimal increase in context-switch time.
+In the runtime, \emph{Worker} threads pull these programmer-specified tasks off of a queue and execute them to completion.
+When tasks block, the worker thread executing them is simply suspended and consumes no computational resources until woken again by some event.
 

### 2013-06-11 doc/pgas13/intro.tex
Clean up intro and Grappa sections a bit.

@@
 \section{Introduction}
+\TODO{Work in discussion of \emph{linearizability} a la~\cite{herlihy1990linearizability}}.
 % - PGAS languages successfully bring shared memory problems into distributed memory systems
@@
 
-The goal of partitioned global address space (PGAS)~\cite{upc:2005} languages and runtimes is to provide the illusion of a single shared memory to a program that actually executes on a number of machines each with their own memory.
+The goal of partitioned global address space (PGAS)~\cite{upc:2005} languages and runtimes is to provide the illusion of a single shared memory to a program actually executing on a distributed memory machine such as a cluster.
 This allows programmers to write their algorithms without needing to explicitly manage communication.
-Once the algorithm is correct, many techniques exist to help improve performance improving locality (spatial and temporal) and coarsening communication where possible.
+% Once the algorithm is correct, many techniques exist to help improve performance improving locality (spatial and temporal) and coarsening communication where possible.
 However, with this shared memory abstraction come all of the difficult concurrency issues that arise in physically-shared memory.
-Luckily, there exists a large body of work solving these issues in physically-shared memory which the PGAS community can leverage.
-In a distributed setting, the costs of communication and synchronization operations are different, providing an opportunity for findings in shared-memory to be re-evaluated.
+Luckily, the PGAS community can leverage a large body of work solving these issues in shared memory.
+The same techniques can be applied in PGAS, but the costs are different, providing an opportunity for different trade-offs to be made.
 
 Globally-shared data structures are one of the cornerstones of shared-memory and PGAS abstractions. 
-In order for multiple concurrent threads to interact with one another, they must observe shared data consistently.
+In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently.
 It is commonly accepted that the easiest consistency model to reason about enforces that all accesses appear to happen in some serializable order that all threads agree on (known as \emph{sequential consistency or SC}).
@@
 % One observation was that, when multiple threads of control concurrently access the same object, they can conduct themselves in one of two ways: contend or cooperate.
-In the classic multithreaded case, either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of potentially-contended atomic operations, and retry if they fail.
+In the classic multithreading case, increasing concurrency increases contention and damages performance.
+Either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of potentially-contended atomic operations, and retry if they fail.
 In both cases, as the number of concurrent accessors increases, the more they contend and the more failed synchronization operations there are.
-The observation of previous work was that the cost of many of these failed synchronization attempts can be mitigated by having the threads \emph{cooperate} via delegation rather than \emph{contend}.
-If multiple threads delegate a single thread to do all of their operations, they can avoid excessive synchronization overhead on ``hot'' locations.
 
-The trick is coming up with a mechanism for delegating that has lower synchronization overhead than the original contention case.
-Hendler et al.\cite{flatCombining} showed that a cleverly-implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization cost.
-In addition, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually.
-Together, the reduced synchronization cost and combined operations allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.
-The same publication list mechanism can be applied to multiple data structures; all that must be customized is the particular way in which the data structure's operations can be combined.
+At the core of a new synchronization technique called \emph{flat combining}~\cite{flatCombining} is the idea of having threads \emph{cooperate} rather than \emph{contend}.
+Threads delegate their work to a single thread, which gives it the opportunity to combine multiple requests in data-structure specific ways and perform them free from contention.
+Together, the reduced synchronization cost and combining allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.
+
+% The observation of previous work was that the cost of many of these failed synchronization attempts can be mitigated by having the threads \emph{cooperate} via delegation rather than \emph{contend}.
+% If multiple threads delegate a single thread to do all of their operations, they can avoid excessive synchronization overhead on ``hot'' locations.
+% The trick is coming up with a mechanism for delegating that has lower synchronization overhead than the original contention case.
+% Hendler et al.\cite{flatCombining} showed that a cleverly-implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization cost.
+% In addition, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually.
+% Together, the reduced synchronization cost and combined operations allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.
+% The same publication list mechanism can be applied to multiple data structures; all that must be customized is the particular way in which the data structure's operations can be combined.
 

### 2013-06-14 doc/pgas13/relatedwork.tex
Rework related work based on Simon's feedback.

@@
 
-Flat combining~\cite{flatCombining} is the latest of many techniques for allowing a single thread to perform the updates of others.
-Previous techniques performed combining hierarchically using fixed trees~\cite{yew:combining-trees}, dynamic trees (or ``funnels'')~\cite{funnels}, and randomized trees~\cite{edtrees}.
-However, the observation of the flat combining work was that in practice, a cleverly-implemented \emph{flat} publication list can outperform more complicated schemes.
+Many techniques performed combining hierarchically to reduce synchronization and hot-spotting, differing mainly in the structures used to do combining: fixed trees~\cite{yew:combining-trees}, dynamic trees (or ``funnels'')~\cite{funnels,MAMA}, and randomized trees~\cite{edtrees}.
+In particular, MAMA~\cite{MAMA}, built for the MTA-2, leveraged funnels to better utilize hardware optimized for massive concurrency.
+On the other hand, \emph{flat} combining~\cite{flatCombining} observed that in multicore systems, hierarchical schemes have too much overhead, and a cleverly-implemented \emph{flat} publication list can perform better.
 Follow-on work introduced parallel flat-combining~\cite{scalableFCQueues}, in which multiple publication lists are combined in parallel and their remainders serialized.
@@
 The relatively higher cost of communication in PGAS makes flat combining even more compelling.
-In addition, the physical isolation per node combined with Grappa's engineered per-core isolation guarantees, introduce a novel opportunity to apply combining, which led to our ``combining proxy'' design.
+Within-core combining is naturally done in a \emph{flat} way.
+In addition, the physical isolation per node combined with Grappa's engineered per-core isolation guarantees, create a novel situation for combining, which led to our \emph{combining proxy} design.
 

### 2013-06-14 doc/pgas13/evaluation.tex
Fix typos and improve wording and typography (based on Preston's feedback)

@@
 In each experiment, a Grappa parallel loop spawns an equal number of tasks on all cores.
-Each task randomly chooses an operation based on the predetermined ``operation mix'', selecting either a push or pop for the Stack and Queue, or an insert or lookup for the Set and Map.
+Each task randomly chooses an operation based on the predetermined ``operation mix,'' selecting either a push or pop for the Stack and Queue, or an insert or lookup for the Set and Map.
 
@@
   \caption{\emph{GlobalStack Statistics.}
-    Measured number of combined messages sent by the Stack with a fixed combining buffer of 1024 elements. Being above the buffer size indicates the amount of matched pushes and pops that occurred.
+    Measured number of combined messages sent by the Stack with a fixed combining buffer of 1024 elements.
+    Matched pushes and pops result in ops/message being greater than the buffer size.
   }
@@
 This was somewhat surprising as combined updates still force others to wait.
-However, with both levels of combining enabled, relatively few requests are combined at the master---for the all-push workload, on the order of a thousand, and for the mixed workload on the stack, the average is less than 20.
+However, with both levels of combining enabled, relatively few requests are combined at the master--for the all-push workload, on the order of a thousand, and for the mixed workload on the stack, the average is less than 20.
 
@@
 The Grappa data structures are synchronized to provide the most general use and match the expectations of programmers and algorithms.
-In these evaluations we compare the flat-combining structures against tuned, relaxed-consistency versions.
+In these evaluations, we compare the flat-combining structures against tuned, relaxed-consistency versions.
 
@@
 Connected Components (CC) is another simple graph analysis kernel that illustrates a different use of global data structures in irregular applications.
-We implement the three-phase CC algorithm~\cite{mtgl} which was designed for the massively-parallel MTA-2 machine.
+We implement the three-phase CC algorithm~\cite{mtgl} which was designed for the massively parallel MTA-2 machine.
 In the first phase, parallel traversals attempt to claim vertices and label them.
@@
 The second phase performs the classical Shiloach-Vishkin parallel algorithm~\cite{shiloach1982n} on this reduced graph, and the final phase propagates the component labels out to the full graph.
-The creation of the reduced-graph dominates the runtime of this algorithm, in includes many repeated inserts at the boundary between traversals.
+Creation of the reduced graph dominates the runtime of this algorithm, but includes many repeated inserts at the boundary between traversals, so is a prime target for the flat-combining Set.
 As in BFS, further optimizations involving relaxation of consistency can be applied, in this case to the Set.

### 2013-06-14 doc/pgas13/flat-combining.tex
Clean up intro and Grappa sections a bit.
Fix typos and improve wording and typography (based on Preston's feedback)

@@
 At the most basic level, the concept of flat combining is about enabling cooperation among threads rather than contention.
-The benefits can be broken down into three components: improved locality, reduced synchronization, and data structure-specific optimization.
-We will explore how this works in a traditional shared memory system, and then describe how the same concepts can be applied to distributed memory.
+The benefits can be broken down into three components: improved locality, reduced synchronization, and data-structure-specific optimization.
+We will explore how this works in a traditional shared-memory system, and then describe how the same concepts can be applied to distributed memory.
 
@@
 
-\subsection{Physically-shared memory}
+\subsection{Physically shared memory}
 
@@
 The combiner walks the publication list from the head, matching up Thread 3's push with Thread 2's pop on its own private stack (3).
-The 2 remaining pushes are added to the top of the shared stack (4).
+The two remaining pushes are added to the top of the shared stack (4).
 Finally, the top is updated, and Thread 4 releases the lock and continues its normal execution.
@@
 The combiner walks the publication list, performs all of the requests, and when done, releases the lock.
-This has greatly reduced the synchronization on the stack's lock, but has introduced a new point of synchronization---threads must synchronize to add to the publication list.
+This has greatly reduced the synchronization on the stack's lock, but has introduced a new point of synchronization--threads must synchronize to add to the publication list.
 However, if a thread performs multiple operations on the stack, it can leave its publication record in the list and amortize the synchronization cost.
@@
 The above example of delegation is compelling in itself.
-However, the crux of the prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently than separately.
+However, the crux of the prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently.
 As the combiner walks the publication list, it merges each non-empty publication record into a combined operation.
@@
 Figure~\ref{fig:stacknofc} shows a simple PGAS translation of the shared-memory stack from earlier.
-A storage array is allocated out of the global heap, so its elements are striped across all the cores in the system.
+A storage array is allocated in the global heap, so its elements are striped across all the cores in the system.
 One core is designated the ``master'' to enforce global synchronization, and holds the elements of the data structure that all concurrent accessors must agree on, in this case, the \texttt{top} pointer.
@@
 Any leftover pushes or pops would be applied to the data structure, and another round of combining would begin.
-This approach, combining on the master, reduces traffic to the data structure storage, but the single ``master'' core must still process every single request.
+This approach, combining on the master, reduces traffic to the data structure storage, but the single ``master'' core must still process every request.
 This approach will not scale when every other core can generate requests just as fast as the master can combine them.
@@
 
-\paragraph{Memory Consistency Model}
+\paragraph{Memory Consistency Model.}
 \label{sec:memory-model}
@@
 
-To preserve sequentially-consistent semantics for flat-combined operations, they must obey a consistent global order.
+To preserve sequentially consistent semantics for flat-combined operations, they must obey a consistent global order.
 To start with, local combined operations must have a serializable order among each other; this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
@@
 These requirements only guarantee serializability for operations on a data structure in isolation.
-To guarantee sequential consistency with respect to accesses to all data structures, data race freedom  must be guaranteed externally, just as in any multithreaded shared-memory program.
+To guarantee sequential consistency with respect to accesses to all data structures, data-race freedom  must be guaranteed externally, just as in any multithreaded shared-memory program.
 
@@
 
-With this data structure-specific proxy object, the FC framework does the bookkeeping to block and wake Workers, deliver results, manage when to synchronize, and choose the combiner.
+With this data-structure-specific proxy object, the FC framework does the bookkeeping to block and wake Workers, deliver results, manage when to synchronize, and choose the combiner.
 The FC framework is responsible for ensuring that all of the combined operations eventually get committed.
@@
 This is done by hiding instances of proxy objects behind a C++ smart-pointer-like object which provides the pointer to the current combiner, and whenever it detects that it should send, allocates a new combiner and points subsequent references to that.
-It is important to note that combiners must start out fresh each time---it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them.
+It is important to note that combiners must start out fresh each time; it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them.
 We refer readers to Section~\ref{sec:memory-model} for why this must be so.
@@
 
-\subsection{Global Set, and Map}
+\subsection{Global Set and Map}
 The Grappa GlobalSet uses a simple chaining design, implemented with a global array of cells (allocated from the global heap), which are partitioned evenly among all the cores, and indexed with a hash function.

### 2013-06-14 doc/pgas13/grappa.tex
Fix typos and improve wording and typography (based on Preston's feedback)

@@
 Grappa is a global-view PGAS runtime for commodity clusters which has been designed from the ground up to achieve high performance on irregular applications.
-The key is latency tolerance---long-latency operations such as reads of remote memory can be tolerated by switching to another concurrent thread of execution.
+The key is latency tolerance--long-latency operations such as reads of remote memory can be tolerated by switching to another concurrent thread of execution.
 Given abundant concurrency, there are opportunities to increase throughput by sacrificing latency.
@@
 
-Highly-tuned implementations of irregular applications in shared-memory, PGAS and message-passing paradigms, typically end up implementing similar constructs.
+Highly tuned implementations of irregular applications in shared-memory, PGAS and message-passing paradigms, typically end up implementing similar constructs.
 Grappa includes these as part of its core infrastructure and simply asks the programmer to express concurrency which it can leverage to provide performance.
@@
 The Grappa runtime has a lightweight threading system that uses prefetching to scale up to thousands of threads on a single core with minimal increase in context-switch time.
-In the runtime, \emph{Worker} threads pull these programmer-specified tasks off of a queue and execute them to completion.
-When tasks block, the worker thread executing them is simply suspended and consumes no computational resources until woken again by some event.
+In the runtime, \emph{Worker} threads pull these programmer-specified tasks from a queue and execute them to completion.
+When a task blocks, the worker thread executing it is simply suspended and consumes no computational resources until woken again by some event.
 
@@
 All accesses to memory owned by another core must be done by that core via a message, even enforcing strict isolation between processes on the same physical memory.
-At the programming level, however, this is hidden behind higher-level remote operations, which in Grappa are called ``delegate'' operations.
+At the programming level, however, this is hidden behind higher-level remote operations, which in Grappa are called \emph{delegates}.
 Figure~\ref{fig:system} shows an example of a delegate \texttt{read} which blocks the calling task and sends a message to the owner, who sends a reply with the data and wakes the caller.

### 2013-06-14 doc/pgas13/intro.tex
Fix typos and improve wording and typography (based on Preston's feedback)

@@
 % Once the algorithm is correct, many techniques exist to help improve performance improving locality (spatial and temporal) and coarsening communication where possible.
-However, with this shared memory abstraction come all of the difficult concurrency issues that arise in physically-shared memory.
+However, with this shared memory abstraction come all of the difficult concurrency issues that arise in physically shared memory.
 Luckily, the PGAS community can leverage a large body of work solving these issues in shared memory.
@@
 
-Globally-shared data structures are one of the cornerstones of shared-memory and PGAS abstractions. 
+Globally shared data structures are one of the cornerstones of shared-memory and PGAS abstractions. 
 In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently.
 It is commonly accepted that the easiest consistency model to reason about enforces that all accesses appear to happen in some serializable order that all threads agree on (known as \emph{sequential consistency or SC}).
-However, in both physically-shared memory and PGAS, maintaining this sequentially consistent view of data among all concurrently accessing threads presents performance challenges.
+However, in both physically shared memory and PGAS, maintaining this sequentially consistent view of data among all concurrently accessing threads presents performance challenges.
 The simplest way to maintain SC for a given data structure is to have a single global lock that implements mutual exclusion and enforces a serializable order over read and write operations.
@@
 In the classic multithreading case, increasing concurrency increases contention and damages performance.
-Either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of potentially-contended atomic operations, and retry if they fail.
+Either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of potentially contended atomic operations, and retry if they fail.
 In both cases, as the number of concurrent accessors increases, the more they contend and the more failed synchronization operations there are.
@@
 % The trick is coming up with a mechanism for delegating that has lower synchronization overhead than the original contention case.
-% Hendler et al.\cite{flatCombining} showed that a cleverly-implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization cost.
+% Hendler et al.~\cite{flatCombining} showed that a cleverly implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization cost.
 % In addition, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually.
@@
 The next section will describe in more detail the Grappa runtime system that is used to implement flat combining for distributed memory machines.
-We then explain the flat combining paradigm in more depth and describe how it maps to a PGAS model.
-Next we explain how several data structures are implemented in our framework and show how they perform on simple throughput workloads.
+We then explain the flat-combining paradigm in more depth and describe how it maps to a PGAS model.
+Next, we explain how several data structures are implemented in our framework and show how they perform on simple throughput workloads.
 Finally, we evaluate how this affects application performance in two simple graph analysis kernels.

### 2013-06-14 doc/pgas13/relatedwork.tex
Fix typos and improve wording and typography (based on Preston's feedback)

@@
 Attempts to improve the scalability of synchronized data structures are as old as the data structures themselves.
-Though much attention has gone to lock-free data structures such as the Treiber Stack~\cite{treiber} and Michael-Scott Queue~\cite{msqueue}, a good body of work has explored ways of scaling globally-locked data structures.
+Though much attention has gone to lock-free data structures such as the Treiber Stack~\cite{treiber} and Michael-Scott Queue~\cite{msqueue}, a significant body of work has explored ways of scaling globally locked data structures.
 
@@
 In particular, MAMA~\cite{MAMA}, built for the MTA-2, leveraged funnels to better utilize hardware optimized for massive concurrency.
-On the other hand, \emph{flat} combining~\cite{flatCombining} observed that in multicore systems, hierarchical schemes have too much overhead, and a cleverly-implemented \emph{flat} publication list can perform better.
+On the other hand, \emph{flat} combining~\cite{flatCombining} observed that in multicore systems, hierarchical schemes have too much overhead, and a cleverly implemented \emph{flat} publication list can perform better.
 Follow-on work introduced parallel flat-combining~\cite{scalableFCQueues}, in which multiple publication lists are combined in parallel and their remainders serialized.
@@
 
-Our work extends the flat-combining paradigm even further to the PGAS domain, where only software provides the illusion of a physically-distributed memory.
+Our work extends the flat-combining paradigm further, to the PGAS domain, where only software provides the illusion of a physically distributed memory.
 The relatively higher cost of communication in PGAS makes flat combining even more compelling.
@@
 In the distributed memory and PGAS domains, implementing scalable and distributed synchronization is the name of the game.
-Besides Grappa, other work has proposed leveraging latency tolerance in PGAS runtimes, including MADNESS~\cite{shet:async-upc} which proposes ways of using asynchronous operations in UPC, and Zhang et al.~\cite{zhang:barnes-hut} who use asynchronous messages in a UPC Barnes Hut implementation to overlap computation with communication and allow for buffering.
+Besides Grappa, other work has proposed leveraging latency tolerance in PGAS runtimes, including MADNESS~\cite{shet:async-upc} which proposes ways of using asynchronous operations in UPC, and Zhang et al.~\cite{zhang:barnes-hut} who use asynchronous messages in a UPC Barnes-Hut implementation to overlap computation with communication and allow for buffering.
 Jose et al.

### 2013-06-14 doc/pgas13/abstract.tex
Incorporate Simon's feedback.

@@
 The implementation of scalable synchronized data structures is notoriously difficult.
-Recent work in shared-memory multicores introduced a new synchronization paradigm called \emph{flat combining} that allows many concurrent accessors to efficiently cooperate to reduce contention on shared locks.
+Recent work in shared-memory multicores introduced a new synchronization paradigm called \emph{flat combining} that allows many concurrent accessors to cooperate efficiently to reduce contention on shared locks.
 In this work we introduce this paradigm to a domain where reducing communication is paramount: distributed memory systems.
@@
 Even using simple locking schemes, we find that these flat-combining data structures scale out to 64 nodes with 2x-100x improvement in throughput.
-We also demonstrate that this translates to application performance with two simple graph analysis kernels.
+We also demonstrate that this translates to application performance via two simple graph analysis kernels.
 The higher communication cost and structured concurrency of Grappa lead to a new form of distributed flat combining that drastically reduces the amount of communication necessary for maintaining global sequential consistency.

### 2013-06-14 doc/pgas13/flat-combining.tex
Incorporate Simon's feedback.

@@
 
-The above example of delegation is compelling in itself.
-However, the crux of the prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently.
+The above example of delegation is compelling in itself. However, the crux of prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently.
 As the combiner walks the publication list, it merges each non-empty publication record into a combined operation.
@@
 To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read} delegate described earlier, blocks the task until complete.
-Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}, the task must send a message to the master to acquire the lock.
-If it succeeds, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker.
-All other workers performing operations block at the first message until the lock is released.
+Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}.
+The task must send a message to the master to acquire the lock.
+If successful, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker.
+All others block at the first message until the lock is released.
 Because of Grappa's user-level blocking mechanisms, blocked requests do not have to retry or busy-wait.
@@
 To start with, local combined operations must have a serializable order among each other; this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
-When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the ``master'' core for the Stack).
+When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the \emph{master} core for the Stack).
 % haven't discussed queue, set, or map yet...
@@
 However, this allows the local order in which keys were inserted to be observed.
-If this is allowed, then that same order must be respected globally, forcing each batch to ensure atomicity from all other cores' batches.
-This is prohibitively expensive, so a cheaper solution is chosen: lookups must get their values from the global data structure, but can be performed in parallel in the same way as inserts.
+If this is allowed, then that same order must be respected globally, forcing each batch to commit atomically with respect to other cores' batches.
+Enforcing this is prohibitively expensive, so a cheaper solution is chosen: if lookups must get their values from the global data structure, then batches can be performed in parallel.
 

### 2013-06-14 doc/pgas13/grappa.tex
Incorporate Simon's feedback.

@@
 \subsection{Global Memory}
-In the PGAS style, Grappa has a notion of a global address space which is partitioned across the physical memories of the nodes in a cluster.
+In the PGAS style, Grappa provides a global address space partitioned across the physical memories of the nodes in a cluster.
 Each core owns a portion of memory, which is divided among execution stacks for the core's Workers, a core-local heap, and a slice of the global heap.

### 2013-06-14 doc/pgas13/intro.tex
Incorporate Simon's feedback.

@@
 In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently.
-It is commonly accepted that the easiest consistency model to reason about enforces that all accesses appear to happen in some serializable order that all threads agree on (known as \emph{sequential consistency or SC}).
-However, in both physically shared memory and PGAS, maintaining this sequentially consistent view of data among all concurrently accessing threads presents performance challenges.
-The simplest way to maintain SC for a given data structure is to have a single global lock that implements mutual exclusion and enforces a serializable order over read and write operations.
+It is commonly accepted that the easiest consistency model to reason about enforces that all accesses appear to happen in some global serializable order (known as \emph{sequential consistency or SC}).
+To preserve SC, operations on shared data structures should be \emph{linearizable}~\cite{herlihy1990linearizability}; that is, appear to happen atomically in some global total order.
+However, in both physically shared memory and PGAS, maintaining linearizability presents performance challenges.
+The simplest way is to have a single global lock that implements mutual exclusion and enforces a serializable order over read and write operations.
 The cost of literally serializing accesses in this way is typically considered prohibitive, even in physically shared memory.
-With the massive amount of parallelism in a cluster of multiprocessors and with the increased cost of remote synchronization, the problem magnifies.
+With the massive amount of parallelism in a cluster of multiprocessors and with the increased cost of remote synchronization, the problem is magnified.
 
@@
 In the classic multithreading case, increasing concurrency increases contention and damages performance.
-Either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of potentially contended atomic operations, and retry if they fail.
-In both cases, as the number of concurrent accessors increases, the more they contend and the more failed synchronization operations there are.
+Either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of contentious atomic operations, and retry if they fail.
+In both cases, as the number of concurrent accessors increases, the more they contend and the more synchronization operations fail.
 
 At the core of a new synchronization technique called \emph{flat combining}~\cite{flatCombining} is the idea of having threads \emph{cooperate} rather than \emph{contend}.
-Threads delegate their work to a single thread, which gives it the opportunity to combine multiple requests in data-structure specific ways and perform them free from contention.
+Threads delegate their work to a single thread, giving it the opportunity to combine multiple requests in data-structure specific ways and performing them free from contention.
 Together, the reduced synchronization cost and combining allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.

### 2013-06-15 doc/pgas13/evaluation.tex
Incorporate Simon's feedback.
Incorporate more of Myers' feedback.

@@
 \section{Evaluation}
-To evaluate the impact flat combining has on the global data structures, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and then measure their impact on performance of two simple graph benchmarks.
+To evaluate the impact flat combining has on the global data structures, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and then measure their impact on performance of two graph benchmarks.
 
@@
 \subsection{Data Structure Throughput}
-First we measured the raw performance of the global data structures on extremely simple throughput workloads.
+First we measured the raw performance of the global data structures on synthetic throughput workloads.
 In each experiment, a Grappa parallel loop spawns an equal number of tasks on all cores.
@@
 Despite Grappa's automatic aggregation, without combining, both the stack and queue completely fail to scale because all workers' updates must serialize.
-Though centralized combining alleviates some of the serialization, its benefit is limited because all synchronization must still be performed on one core.
-However, with local flat combining, synchronization is able to be done mostly in parallel, with less-frequent bulk synchronization at the master.
+Though centralized combining alleviates some of the serialization, its benefit is limited 
+because all operations involve synchronization through a single core.
+However, with local flat combining, synchronization is done mostly in parallel, with less-frequent bulk synchronization at the master.
 

### 2013-06-15 doc/pgas13/flat-combining.tex
Incorporate more of Myers' feedback.

@@
 A storage array is allocated in the global heap, so its elements are striped across all the cores in the system.
-One core is designated the ``master'' to enforce global synchronization, and holds the elements of the data structure that all concurrent accessors must agree on, in this case, the \texttt{top} pointer.
+One core is designated the \emph{master} to enforce global synchronization, and holds the elements of the data structure that all concurrent accessors must agree on, in this case, the \texttt{top} pointer.
 
@@
 All others block at the first message until the lock is released.
-Because of Grappa's user-level blocking mechanisms, blocked requests do not have to retry or busy-wait.
+Grappa's user-level threading allows requests to block without consuming compute resources.
 However, all workers on each core perform this synchronization and serialize on a single core, causing that core to become the bottleneck.
@@
 \paragraph{Centralized Combining.}
-A first thought might be to directly apply the idea of flat combining to the serialization at the master node.
+A first thought might be to directly apply the idea of flat combining to the serialization at the \emph{master}.
 The worker that acquires the lock on the master can walk through the requests of all the other workers waiting to acquire the global lock and perform combining on them.
@@
 Any leftover pushes or pops would be applied to the data structure, and another round of combining would begin.
-This approach, combining on the master, reduces traffic to the data structure storage, but the single ``master'' core must still process every request.
+This approach reduces traffic to the data structure storage, but a single core must still process every request.
 This approach will not scale when every other core can generate requests just as fast as the master can combine them.
@@
 \paragraph{Distributed Combining.}
-Instead of all workers on a core sending independent synchronization messages and putting the burden all on one core, what if each core did its own combining first, and synchronized with the master in bulk?
+Instead of all workers on a core sending independent synchronization messages and putting the burden all on one core, each core can instead do its own combining first then synchronize with the master in bulk.
 Distributing the combining to each core allows the majority of the work to be performed in parallel and without communication.
@@
 In Grappa, for each global data structure, a local \emph{proxy} object is allocated from the core-local heap to intercept requests.
-Conceptually, workers add their requests to a local publication list in the proxy, then one is chosen to do the combining, and the rest block until their request is satisfied.
+Conceptually, workers add their requests to a local publication list in the proxy, one is chosen to do the combining, and the rest block until their request is satisfied.
 However, because Grappa's scheduler is non-preemptive, each worker has atomicity ``for free'' until it performs a blocking operation (such as communication).
@@
 
-After all local combining has been done, one requesting worker is chosen to commit the combined operation globally.
-Back to the example in Figure~\ref{fig:stackfc}, Worker 4 becomes the combiner and performs much the same synchronization as in the uncombined case, but is able to push multiple elements with one synchronization.
+After all local combining has been done, one requesting worker on the core is chosen to commit the combined operation globally.
+In Figure~\ref{fig:stackfc}, Worker 4 becomes the combiner and performs much the same synchronization as in the uncombined case, but is able to push multiple elements with one synchronization.
 The global lock must still be acquired, so concurrent combined requests (from different cores) must block and serialize on the master.
-Combining in this way allows synchronization to be distributed and performed in parallel, coarsening the granularity of the actual global synchronization and reducing the amount of serialization that occurs.
+Combining in this way allows synchronization to be performed in parallel, coarsening the granularity of the actual global synchronization and reducing the amount of serialization that occurs.
 
@@
 \label{sec:memory-model}
-In the context of this work, sequential consistency guarantees that within a
+In the context of Grappa, sequential consistency guarantees that within a
 task, operations will be in task order and that all tasks will observe the
-same global order of operations.
+same global order.
 
@@
 consistency for data-race free programs). To be conservative, delegate
-operations block the calling worker until they have completed globally,
+operations block the calling worker until they have become globally visible,
 ensuring they can be used for synchronization. As such, delegate operations
@@
 
-To preserve sequentially consistent semantics for flat-combined operations, they must obey a consistent global order.
-To start with, local combined operations must have a serializable order among each other; this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
+To preserve linearizability for flat-combined operations, they must obey a consistent global order.
+First of all, the execution of local combined operations must be serializable;
+this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
 When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the \emph{master} core for the Stack).
@@
 In the case of a stack or queue, this guarantee comes from applying a batch of push or pop operations atomically in a contiguous block on the global data structure.
-The second level of combining at the master preserves this as well by only allowing updates to proceed when they will not conflict and blocking all of them until they have completed.
+The second level of combining at the master preserves this as well by only allowing updates to proceed when they will not conflict and blocking a batch of operations until all have completed.
 For the \emph{stack}, matching pushes and pops locally is one exception to the rule that operations must block until globally committed.
-Because a pop ``destroys'' the corresponding push, these operations together are independent of all others and can conceptually be placed anywhere in a total global ordering, so they can be safely matched locally.
+Because a pop ``destroys'' the corresponding push, these operations executed atomically together are independent of all others and can conceptually be placed anywhere in a total global ordering that preserves program order, so they can be safely matched locally.
 
@@
 Note that, if an insert finds that its key is already in the combiner, it does not need to send its own message.
-However, it must still block until that insert is done, otherwise it may assume that the key has been inserted and perform some other operation that requires the insert to have happened, effectively reordering operations.
-This ensures that if one wanted to use the insert to signal synchronization, it is guaranteed to have been committed globally before the task continues.
+However, it must still block until that insert is done to ensure that a subsequent operation it performs cannot be reordered with it, providing the same guarantee as all other delegate operations.
+% This ensures that if one wanted to use the insert to signal synchronization, it is guaranteed to have been committed globally before the task continues.
 
 In the same way as inserts, lookups can piggy-back on other lookups of the same key.
-It is tempting to allow a lookup to be satisfied locally by inspecting the keys being inserted.
+It is tempting to allow a lookup to be satisfied locally by inspecting the keys being inserted, using the same intuition as for store/write buffers in modern processors.
 However, this allows the local order in which keys were inserted to be observed.
-If this is allowed, then that same order must be respected globally, forcing each batch to commit atomically with respect to other cores' batches.
+If this is allowed, then to preserve SC, that same order must be respected globally, forcing each batch to commit atomically with respect to other cores' batches.
 Enforcing this is prohibitively expensive, so a cheaper solution is chosen: if lookups must get their values from the global data structure, then batches can be performed in parallel.
@@
 These requirements only guarantee serializability for operations on a data structure in isolation.
-To guarantee sequential consistency with respect to accesses to all data structures, data-race freedom  must be guaranteed externally, just as in any multithreaded shared-memory program.
+To guarantee sequential consistency with respect to accesses to all data structures, data-race freedom  must be guaranteed externally, as in the C++ memory model for shared memory.
 
@@
 The Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine.
-As mentioned before, the local ``publication list'' is trivial to access atomically in Grappa's cooperative scheduling.
+As mentioned before, the local \emph{publication list} is trivial to access atomically in Grappa's cooperative scheduling.
 Instead, in the Grappa FC Framework, a \emph{proxy} object is instantiated ahead of time on each core and each worker merges its update into that structure before blocking or becoming the combiner.
@@
 This is done by hiding instances of proxy objects behind a C++ smart-pointer-like object which provides the pointer to the current combiner, and whenever it detects that it should send, allocates a new combiner and points subsequent references to that.
-It is important to note that combiners must start out fresh each time; it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them.
-We refer readers to Section~\ref{sec:memory-model} for why this must be so.
+Requests cannot be combined with others that have been committed or are currently being committed; we refer
+% Updates cannot be added to combiners that are in the process of being committed and must start out fresh each time; it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them.
+ readers to Section~\ref{sec:memory-model} for why this must be so.
 
@@
 
-Our implementation of the GlobalMap matches that of the Set but of course stores values of a given type.
+Our implementation of the GlobalMap matches that of the Set but of course stores values.
 This does not change the requirements of the proxy except that the return values of lookups must be a copy of the value rather than just a boolean.

### 2013-06-15 doc/pgas13/relatedwork.tex
Incorporate more of Myers' feedback.

@@
 This work demonstrates a similar use case for synchronized queues as in our work, and uses them with the Graph500 benchmark as well.
-In contrast, Grappa performs much of the same message aggregation invisibly to the programmer and even data structure designer, and the flat combining framework described in this work can be used to implement many different data structures for use in synchronizing.
+In contrast, Grappa performs much of the same message aggregation invisibly to the programmer, and the FC framework can be used to implement many different linearizable data structures.

### 2013-06-15 doc/pgas13/flat-combining.tex
Simple space hacking w/o sacrificing content.

@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figs/fc_shared_mem.pdf}
+  \includegraphics[width=0.38\textwidth]{figs/fc_shared_mem.pdf}
   \caption{\emph{Flat combining in shared memory.}
@@
 
-\begin{figure*}[t]
+% \begin{figure*}[t]
+%   \begin{subfigure}[b]{0.38\textwidth}
+\begin{figure}[t]
     \centering
-  \begin{subfigure}[b]{0.43\textwidth}
-    \includegraphics[width=\textwidth]{figs/stack_nofc.pdf}
-    \caption{\emph{Without flat combining.}
+    \includegraphics[width=0.35\textwidth]{figs/stack_nofc.pdf}
+    \caption{\emph{Stack without combining.}
       To do its push, Worker 1 sends a message to synchronize with the master on Core 0 (1), which sends another message to write the value to the top of the stack (2), bumps the synchronized \texttt{top} pointer (3), and finally continues.
-Worker 2, and all other workers on all cores accessing the stack, must block at the master and wait for Worker 1 to complete its push before doing their operations (4).
+Worker 2, and workers on other cores, must block at the master and wait for Worker 1 to complete its push before doing their operations (4).
     }
     \label{fig:stacknofc}
-  \end{subfigure}%
-  \hspace{0.05\textwidth}
-  \begin{subfigure}[b]{0.43\textwidth}
+  % \end{subfigure}%
+  % \hspace{0.12\textwidth}
+\end{figure}
+\begin{figure}
+  % \begin{subfigure}[b]{0.38\textwidth}
     \centering
-    \includegraphics[width=\textwidth]{figs/stack_fc.pdf}
-    \caption{\emph{Local flat combining.}
-      Workers within the core add their requests to a local proxy object.
+    \includegraphics[width=0.38\textwidth]{figs/stack_fc.pdf}
+    \caption{\emph{Stack with distributed combining.}
       Worker 3's pop matches with Worker 2's push, requiring no global communication (1).
-After combining locally, Worker 1 and 4's pushes remain, so Worker 4 becomes the core's combiner (2), sends a message to synchronize with the master (3), adds both new values to the global stack (4), bumps the top pointer and releases the lock on master (5), and finally wakes Worker 1 and continues (6).
+After combining locally, Worker 1 and 4's pushes remain, so Worker 4 becomes the core's combiner (2), sends a message to synchronize with the master (3), adds both new values to global stack (4), bumps the top pointer and releases the lock on master (5), and finally wakes Worker 1 (6).
     }
     \label{fig:stackfc}
-  \end{subfigure}
-  \caption{\emph{Global stack in Grappa} with and without \emph{flat combining}.}
-  \label{fig:stack}
-\end{figure*}
+  % \end{subfigure}
+\end{figure}
+  % \caption{\emph{Global stack in Grappa} with and without \emph{flat combining}.}
+  % \label{fig:stack}
+% \end{figure*}
 
@@
     // do remote write to the top of the stack
-    // (so other workers may interleave)
     delegate::write(stack->top, e);
@@
   // Combining Methods 
-  // (either match locally or block for results)
   void push(T val);
@@
     if (npush > 0) {
-      // on master, acquire lock and return GlobalAddress to the top of the stack
+      // on master: acquire lock, return top ptr
       auto top = delegate::call(master.core(),[=]{
@@
       });
-    } else if (npop > 0) {} //...elided for space
+    } else if (npop > 0) {} // elided for space...
   }

### 2013-06-15 doc/pgas13/intro.tex
Simple space hacking w/o sacrificing content.

@@
 \section{Introduction}
-\TODO{Work in discussion of \emph{linearizability} a la~\cite{herlihy1990linearizability}}.
-% - PGAS languages successfully bring shared memory problems into distributed memory systems
-% - Cost functions are different, but many of the other considerations, such as reducing locking, still apply
-% - With abundant parallelism, we can tolerate some serialization to reduce number of synchronizations and reduce overall communication
-
 The goal of partitioned global address space (PGAS)~\cite{upc:2005} languages and runtimes is to provide the illusion of a single shared memory to a program actually executing on a distributed memory machine such as a cluster.
 This allows programmers to write their algorithms without needing to explicitly manage communication.
-% Once the algorithm is correct, many techniques exist to help improve performance improving locality (spatial and temporal) and coarsening communication where possible.
 However, with this shared memory abstraction come all of the difficult concurrency issues that arise in physically shared memory.
@@
 Much prior work has explored the scalability challenges of adding more concurrent accessors to shared data structures in shared memory systems.
-% One observation was that, when multiple threads of control concurrently access the same object, they can conduct themselves in one of two ways: contend or cooperate.
 In the classic multithreading case, increasing concurrency increases contention and damages performance.
@@
 
-% The observation of previous work was that the cost of many of these failed synchronization attempts can be mitigated by having the threads \emph{cooperate} via delegation rather than \emph{contend}.
-% If multiple threads delegate a single thread to do all of their operations, they can avoid excessive synchronization overhead on ``hot'' locations.
-% The trick is coming up with a mechanism for delegating that has lower synchronization overhead than the original contention case.
-% Hendler et al.~\cite{flatCombining} showed that a cleverly implemented ``publication list'' can allow multiple accessors to cooperate with minimal synchronization cost.
-% In addition, given the semantics of some particular data structures, the set of accesses being serialized can be composed and performed more efficiently \emph{combined} than individually.
-% Together, the reduced synchronization cost and combined operations allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.
-% The same publication list mechanism can be applied to multiple data structures; all that must be customized is the particular way in which the data structure's operations can be combined.
-
-% Make this point later?
-% In many cases, algorithms do not require operations to be immediately consistent.
-Many optimizations for message-passing or PGAS-style distributed memory algorithms leverage this observation to minimize the need for fine-grained communication.
-However, they typically must make this visible to programmers, who now must take care to insert their own memory barriers where appropriate, or otherwise express the way in which consistency relaxation can occur.
-
 The goal of this work is to apply the flat-combining paradigm to a PGAS runtime to reduce the cost of maintaining sequentially consistent data structures.
@@
 
-% lay out the upcoming sections of the paper...
 The next section will describe in more detail the Grappa runtime system that is used to implement flat combining for distributed memory machines.

### 2013-06-15 doc/pgas13/flat-combining.tex
cut down code samples a tiny bit

@@
 
-void push(GlobalAddress<GlobalStack> stack, T e) {
+void push(GlobalAddress<GlobalStack> master, T e){
   // perform operation on the stack's master core
-  delegate::call(stack.core(), [stack,e]{
+  delegate::call(master.core(), [master,e]{
     // force other operations to wait
-    stack->lock.acquire();
+    master->lock.acquire();
     // do remote write to the top of the stack
-    delegate::write(stack->top, e);
-    // bump the top pointer
-    stack->top++;
-    stack->lock.release();
+    delegate::write(master->top, e);
+    // bump top and release lock
+    master->top++;
+    master->lock.release();
   }); // blocks until response arrives
@@
   GlobalAddress<GlobalStack> master;
-
   // local state for tracking pushes and pops
@@
       Grappa::memcpy(top, pushed_values, npush);
-      // release lock at the master core
+      // bump top and release lock
       delegate::call(master.core(),[master,npush]{

### 2013-06-15 doc/pgas13/conclusion.tex
Fix typos and improve wording and typography (based on Preston's feedback)
rewriting to conserve space, tighten language

@@
 In PGAS implementations there is typically a large discrepancy between the first, simple, description of an algorithm and the final, optimized one.
-At the core of this work is the desire to have a library of data structures which can be used without worrying about when they will be consistent or whether they will limit scalability.
+% At the core of this work is the desire to have a library of data structures which can be used without worrying about when they will be consistent or whether they will limit scalability.
 Bringing findings from shared memory into the PGAS domain brings with it new challenges and opportunities.
 We have shown that the additional concurrency that comes with a latency-tolerant runtime, rather than compounding the problem, provides a new opportunity for reducing communication by combining locally.
-% perform some of the locality optimizations that finely-tuned implementations would.
-This allows flat-combined sequentially-consistent global data structures to scale well out to a thousand cores with millions of concurrent threads.
+% perform some of the locality optimizations that finely tuned implementations would.
+This allows easy implementation of a library of flat-combined linearizable global data structures that scale well out to a thousand cores and millions of concurrent threads.

### 2013-06-15 doc/pgas13/evaluation.tex
Simple space hacking w/o sacrificing content.
rewriting to conserve space, tighten language

@@
 To evaluate the impact flat combining has on the global data structures, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and then measure their impact on performance of two graph benchmarks.
-
 Experiments were run on a cluster of AMD Interlagos processors.
@@
 
-\paragraph{Queue and Stack}
+\paragraph{Queue and Stack.}
 \begin{figure}[t]
@@
 This was somewhat surprising as combined updates still force others to wait.
-However, with both levels of combining enabled, relatively few requests are combined at the master--for the all-push workload, on the order of a thousand, and for the mixed workload on the stack, the average is less than 20.
+However, with both levels of combining enabled, relatively few requests are combined at the master--for the stack, a maximum of 6\% per pass for the all-push workload and 0.1\% for the mixed workload.
 
@@
 
-\paragraph{HashSet and HashMap}
+\paragraph{HashSet and HashMap.}
 \begin{figure}[t]
@@
 \subsection{Application Kernel Performance}
-The original goal was to have scalable global data structures in the Grappa library that could be used in applications.
+% The original goal was to have scalable global data structures in the Grappa library that could be used in applications.
 Generic implementations of data structures are often insufficient for scaling, so HPC applications often implement their own customized versions.
@@
 The Grappa data structures are synchronized to provide the most general use and match the expectations of programmers and algorithms.
-In these evaluations, we compare the flat-combining structures against tuned, relaxed-consistency versions.
+In these evaluations, we compare the flat-combining structures against custom, tuned versions that leverage relaxed consistency needs of the applications.
 
-\paragraph{Breadth-First Search}
+\paragraph{Breadth-First Search.}
 \begin{figure}[t]
   \centering
-  \includegraphics[width=0.5\textwidth]{data/plots/bfs_perf.pdf}
+  \includegraphics[width=0.38\textwidth]{data/plots/bfs_perf.pdf}
   \caption{\emph{BFS} on a Graph500-spec graph of scale 26 (64 million vertices, 1 billion edges), with the direction-optimizing BFS algorithm.
@@
 This benchmark does a search starting from a random vertex in a synthetic graph and builds a search tree of parent vertices for each vertex traversed during the search.
-While this is a relatively simple problem, it exercises the random-access throughput of a system as well as being a primitive in many other graph algorithms.
+% While this is a relatively simple problem, it exercises the random-access throughput of a system as well as being a primitive in many other graph algorithms.
 The BFS algorithm contains a global queue which represents the frontier of vertices to be visited in each level.
-Our implementation employs the direction-optimizing algorithm by Beamer et al.~\cite{Beamer:Graph500} which performs particularly well for the scale-free RMAT graphs generated by the benchmark.
-The frontier queue in BFS is amenable to further optimization to take advantage of the fact that the algorithm does pushes and pops in separate phases, allowing consistency to be relaxed.
-We compare our implementation of BFS using the flat-combined global queue described above with a highly tuned Grappa implementation that uses a custom asynchronous queue.
+Our implementation employs the direction-optimizing algorithm by Beamer et al.~\cite{Beamer:Graph500}.
+ % which performs particularly well for the scale-free RMAT graphs generated by the benchmark.
+The frontier queue is write-only in one phase and read-only in the next, making it amenable to relaxed consistency.
+We compare performance of BFS using the Grappa FC queue described above with a highly tuned Grappa implementation that uses a custom asynchronous queue.
 
-Figure~\ref{fig:bfs_perf} shows the results of scaling the BFS kernel up to 64 nodes.
-The simple queue implementation without flat combining is completely unscalable.
-However, the flat combining queue tracks the tuned version.
+Figure~\ref{fig:bfs_perf} shows the results of scaling up to 64 nodes.
+Using the simple queue without flat combining is completely unscalable, but with FC, it tracks the tuned version.
 This illustrates that providing a safe, synchronized data structure for initially developing algorithms for PGAS is possible without sacrificing scaling.
@@
 
-\paragraph{Connected Components}
+\paragraph{Connected Components.}
 \begin{figure}[t]
   \centering
-  \includegraphics[width=0.5\textwidth]{data/plots/cc_perf.pdf}
+  \includegraphics[width=0.38\textwidth]{data/plots/cc_perf.pdf}
   \caption{\emph{Connected Components} on the same scale 26 Graph500 graph. Performance is measured in MTEPS.}
@@
 Whenever two traversals encounter each other, an edge between the two roots is inserted in a set. 
-After all edges have been traversed in this way, the set of edges forms a new, typically much smaller, graph.
-The second phase performs the classical Shiloach-Vishkin parallel algorithm~\cite{shiloach1982n} on this reduced graph, and the final phase propagates the component labels out to the full graph.
-Creation of the reduced graph dominates the runtime of this algorithm, but includes many repeated inserts at the boundary between traversals, so is a prime target for the flat-combining Set.
+% After all edges have been traversed in this way, the set of edges forms a new, typically much smaller, graph.
+The second phase performs the classical Shiloach-Vishkin parallel algorithm~\cite{shiloach1982n} on the reduced graph formed by phase one's edge set, and the final phase propagates the component labels out to the full graph.
+Creation of the reduced edge set dominates the runtime of this algorithm, but includes many repeated inserts at the boundary between traversals, so is a prime target for the flat-combining Set.
 As in BFS, further optimizations involving relaxation of consistency can be applied, in this case to the Set.

### 2013-06-15 doc/pgas13/flat-combining.tex
rewriting to conserve space, tighten language

@@
 
-One of the issues with shared data structures in a shared-memory multicore is that threads executing on different cores will force the hottest parts of the data structure to thrash between their caches.
+% One of the issues with shared data structures in a shared-memory multicore is that threads executing on different cores will force the hottest parts of the data structure to thrash between their caches.
 Having a single thread do all of the operations on the data structure allows it to keep everything in its cache.
@@
 
-Reduced synchronization comes from delegating work to another core.
+Simply by delegating work to another core, locality is improved and synchronization is reduced.
+% Reduced synchronization comes from delegating work to another core.
 To illustrate the issue, consider the shared synchronous stack shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock.
 Without flat combining, whenever a thread attempts to push something on the stack, it must acquire the lock, put its value into the storage array, bump the top, and then release the lock.
-When many threads contend for the lock, all but one will fail and will have to retry.
+When many threads contend for the lock, all but one will fail and have to retry.
 Each attempt forces an expensive memory fence and consumes bandwidth, and as the number of threads increases, the fraction of successes plummets.
-Under flat combining, instead, threads add requests to the publication list.
+Under flat combining, instead, threads add requests to a \emph{publication list}.
 They each try to acquire the lock, and the one that succeeds becomes the combiner.
@@
 The combiner walks the publication list, performs all of the requests, and when done, releases the lock.
-This has greatly reduced the synchronization on the stack's lock, but has introduced a new point of synchronization--threads must synchronize to add to the publication list.
-However, if a thread performs multiple operations on the stack, it can leave its publication record in the list and amortize the synchronization cost.
-This publication list mechanism can be re-used in different data structures, saving each one from needing to implement its own clever synchronization mechanism.
+This allows the one thread to keep the data structure in cache, reducing thrashing between threads.
+It also greatly reduces contention on the lock, but introduces a new point of synchronization--adding to the publication list.
+However, if a thread performs multiple operations, it can leave its publication record in the list and amortize the synchronization cost.
+This publication list mechanism can be re-used in other data structures, saving each from needing to implement its own clever synchronization.
 
@@
 When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills both of their records and allows them to proceed.
-After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the two unmatched pushes are added to the top of the stack, and the requesting threads' records are marked as complete.
+After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the two unmatched pushes are added to the top of the stack.
 In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched.
@@
 
-All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object.
-To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read} delegate described earlier, blocks the task until complete.
+All tasks accessing the stack hold a \texttt{GlobalAddress} to the master object, and invoke custom \emph{delegate methods} that, like the \texttt{read} delegate described earlier, block the task until complete.
 Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}.
@@
 A first thought might be to directly apply the idea of flat combining to the serialization at the \emph{master}.
-The worker that acquires the lock on the master can walk through the requests of all the other workers waiting to acquire the global lock and perform combining on them.
-In the case of the GlobalStack, this would mean taking all of the push requests, matching them with the pops, and sending messages back to the remote workers with the results.
-Any leftover pushes or pops would be applied to the data structure, and another round of combining would begin.
-This approach reduces traffic to the data structure storage, but a single core must still process every request.
-This approach will not scale when every other core can generate requests just as fast as the master can combine them.
+The worker that acquires the lock can walk through the requests of all the other workers waiting to acquire the lock and combine them.
+In the case of the Stack, this would mean matching pushes and pops, applying the remainder, and sending messages back to all remote workers with results, before starting another round of combining.
+This approach reduces traffic to the data structure storage, but a single core must still process every request, so it cannot scale if every other core is generating requests at the same rate.
 
 \paragraph{Distributed Combining.}
-Instead of all workers on a core sending independent synchronization messages and putting the burden all on one core, each core can instead do its own combining first then synchronize with the master in bulk.
+Instead of all workers sending independent synchronization messages and putting the burden all on one core, each core can instead do its own combining first then synchronize with the master in bulk.
 Distributing the combining to each core allows the majority of the work to be performed in parallel and without communication.
-As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and allows operations to be composed in data structure-specific ways for even greater performance.
+% As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and allows operations to be composed in data structure-specific ways for even greater performance.
 
@@
 This means that an explicit publication list with clever synchronization is unnecessary.
-Instead, workers perform their own combining, merging their operations directly into the local proxy, and block if unable to satisfy their requests immediately.
+Instead, workers merging their operations directly into the local proxy, and block if unable to satisfy their requests immediately.
 The proxy structure is chosen to be able to compactly aggregate operations and efficiently perform matching in cases where it is allowed.
-Figure~\ref{fig:stackfc} shows how in the global stack example, pushes and pops are matched on the proxy's local stack.
+Figure~\ref{fig:stackfc} shows how pushes and pops are matched on the proxy's local stack.
 
@@
 In Figure~\ref{fig:stackfc}, Worker 4 becomes the combiner and performs much the same synchronization as in the uncombined case, but is able to push multiple elements with one synchronization.
-The global lock must still be acquired, so concurrent combined requests (from different cores) must block and serialize on the master.
-Combining in this way allows synchronization to be performed in parallel, coarsening the granularity of the actual global synchronization and reducing the amount of serialization that occurs.
+The global lock must still be acquired, so concurrent combined requests (from different cores) must block and serialize on the master, but the granularity of global synchronization is coarser, reducing actual serialization.
 
@@
 same global order.
-
 The Grappa memory model is essentially the same as the C++ memory
@@
 that this distributed version of flat combining preserves sequential
-consistency, we will now argue why it does.
+consistency, we now argue why it does.
 
@@
 When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the \emph{master} core for the Stack).
+Whenever a global commit starts, a fresh combiner must be created for subsequent operations to use.
+If they were to use the old combiner's state to satisfy requests locally, it would violate global ordering because the local object would not reflect other cores' updates.
+% This guarantees, for example, that destructive operations only happen once (i.e., a worker cannot locally pop the same value that another core's combined operation does).
 % haven't discussed queue, set, or map yet...
 %(the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map). The global order, then, is essentially the concatenation of the cores' sequential orders.
-For this ``concatenation'' to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.
+For this serial ``concatenation'' of serialized batches to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.
 
 In the case of a stack or queue, this guarantee comes from applying a batch of push or pop operations atomically in a contiguous block on the global data structure.
-The second level of combining at the master preserves this as well by only allowing updates to proceed when they will not conflict and blocking a batch of operations until all have completed.
-For the \emph{stack}, matching pushes and pops locally is one exception to the rule that operations must block until globally committed.
-Because a pop ``destroys'' the corresponding push, these operations executed atomically together are independent of all others and can conceptually be placed anywhere in a total global ordering that preserves program order, so they can be safely matched locally.
+% Centralized combining preserves this by only allowing updates to proceed when they will not conflict and blocking a batch of operations until all have completed.
+Matching pushes and pops locally for the Stack is one exception to the rule that operations must block until globally committed.
+Because a pop ``destroys'' the corresponding push, these operations together are independent of all others and can conceptually be placed anywhere in a global order.
+It is trivial to respect program order with this placement, so they can be safely matched locally.
 
@@
 Therefore, a batch of these operations can be committed globally in parallel, since they are guaranteed not to conflict with each other.
-Note that, if an insert finds that its key is already in the combiner, it does not need to send its own message.
-However, it must still block until that insert is done to ensure that a subsequent operation it performs cannot be reordered with it, providing the same guarantee as all other delegate operations.
+Note that if an insert finds that its key is already in the combiner, it does not need to send its own message.
+However, it must still block until that insert is done to ensure that a subsequent operation it performs cannot be reordered with it, preserving program order.
+% providing the same guarantee as all other delegate operations. 
 % This ensures that if one wanted to use the insert to signal synchronization, it is guaranteed to have been committed globally before the task continues.
 
-In the same way as inserts, lookups can piggy-back on other lookups of the same key.
-It is tempting to allow a lookup to be satisfied locally by inspecting the keys being inserted, using the same intuition as for store/write buffers in modern processors.
+Similarly, lookups can piggy-back on other lookups of the same key.
+Using intuition from store/write buffers in modern processors, it is tempting to allow a lookup to be satisfied locally by inspecting outstanding inserts.
 However, this allows the local order in which keys were inserted to be observed.
@@
 
-Whenever a global commit starts, a fresh combiner must be created for subsequent operations to use.
-If they were to use the old combiner's state to satisfy requests locally, it would violate global ordering because the local object would not reflect other cores' updates.
-This guarantees, for example, that destructive operations only happen once (i.e., a worker cannot locally pop the same value that another core's combined operation does).
-
-These requirements only guarantee serializability for operations on a data structure in isolation.
-To guarantee sequential consistency with respect to accesses to all data structures, data-race freedom  must be guaranteed externally, as in the C++ memory model for shared memory.
+These requirements only guarantee linearizability of operations on a single data structure.
+To guarantee sequential consistency with respect to all accesses, data-race freedom must be guaranteed by the program, as in the C++ memory model for shared memory.
 
 \section{Grappa FC Framework}
-In order to leverage the flat-combining paradigm in Grappa, we implemented a generic framework which can be used to improve performance for a number of global data structures.
+To leverage the flat-combining paradigm in Grappa, we implemented a generic framework to improve performance for a number of global data structures.
 The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress.
@@
 
-The Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine.
-As mentioned before, the local \emph{publication list} is trivial to access atomically in Grappa's cooperative scheduling.
-Instead, in the Grappa FC Framework, a \emph{proxy} object is instantiated ahead of time on each core and each worker merges its update into that structure before blocking or becoming the combiner.
-In this style, each global data structure must define the following:
+As mentioned before, the Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine.
+% As mentioned before, the local \emph{publication list} is trivial to access atomically in Grappa's cooperative scheduling.
+% Instead, in the Grappa FC Framework, 
+A \emph{proxy} object is instantiated ahead of time on each core and each worker merges its request into that structure before blocking or becoming the combiner.
+Each global data structure must define a \emph{proxy} with the following:
 \begin{enumerate}
-  \item A local \emph{proxy} data structure for tracking updates to the global structure.
+  \item A local \emph{proxy} data structure for tracking combined requests.
   \item \emph{Combining methods} that operate on the local proxy.
@@
 class GlobalStackProxy : Grappa::FCProxy {
-  // pointer to data structure master
   GlobalAddress<GlobalStack> master;
-  // local state for tracking pushes and pops
+  // Local state for tracking requests
   T  pushed_values[1024];
@@
 
-With this data-structure-specific proxy object, the FC framework does the bookkeeping to block and wake Workers, deliver results, manage when to synchronize, and choose the combiner.
+With this data-structure-specific proxy object, the FC framework does the bookkeeping to block and wake workers, deliver results, manage when to synchronize, and choose the combiner.
 The FC framework is responsible for ensuring that all of the combined operations eventually get committed.
@@
 
-One could imagine other synchronization policies that would ensure progress but allow more combining to occur before synchronizing.
+% % left out for space
+% One could imagine other synchronization policies that would ensure progress but allow more combining to occur before synchronizing.
 For instance, local combining could continue until all workers are blocked.
@@
 
-While a combining operation is in flight, new requests are likely to continue to accumulate.
-The framework transparently wraps the proxy object so that when it starts a \texttt{sync} on one, it can direct requests to a fresh combiner.
+While a combining operation is in flight, new requests continue to accumulate.
+The framework transparently wraps the proxy object so when one \texttt{sync} starts, it can direct requests to a fresh combiner.
 This is done by hiding instances of proxy objects behind a C++ smart-pointer-like object which provides the pointer to the current combiner, and whenever it detects that it should send, allocates a new combiner and points subsequent references to that.
-Requests cannot be combined with others that have been committed or are currently being committed; we refer
+% Requests cannot be combined with others that have been committed or are currently being committed; we refer readers to Section~\ref{sec:memory-model} for why this must be so.
 % Updates cannot be added to combiners that are in the process of being committed and must start out fresh each time; it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them.
- readers to Section~\ref{sec:memory-model} for why this must be so.
 
@@
 
-The GlobalQueue has nearly the same implementation as the stack, but is unable to match locally, so \texttt{sync} first does the pushes, then does the pops.
+The GlobalQueue has nearly the same implementation as the stack, but is unable to match locally, so \texttt{sync} first commits the pushes, then the pops.
 
@@
 To track all of the keys waiting to be inserted, we use the hash set implementation from the C++11 standard library (\texttt{std::unordered\_set}).
-Similar to the case with pops, for lookups, we must keep track of pointers for where to put the results of lookup operations, which the proxy does with a hash map (again from the C++ standard library) from keys to a list of pointers.
-As discussed in Section~\ref{sec:memory-model}, matching lookups with inserts locally would force a particular sequential order, so instead we only allow matching inserts with inserts and lookups with lookups.
-This allows \texttt{sync} to simply issue all of the inserts and lookups in parallel and block until all have completed.
+As with pops, lookups must provide pointers for results to be put, which is done with a hash map (again from the C++ standard library) from keys to lists of pointers.
+As discussed in Section~\ref{sec:memory-model}, matching lookups with inserts locally would force a particular sequential order.
+Instead, we only allow matching inserts with inserts and lookups with lookups, allowing \texttt{sync} to simply issue all inserts and lookups in parallel and block until all have completed.
 

### 2013-06-15 doc/pgas13/intro.tex
rewriting to conserve space, tighten language

@@
 However, with this shared memory abstraction come all of the difficult concurrency issues that arise in physically shared memory.
-Luckily, the PGAS community can leverage a large body of work solving these issues in shared memory.
-The same techniques can be applied in PGAS, but the costs are different, providing an opportunity for different trade-offs to be made.
+Luckily, the PGAS community can leverage a large body of work solving these issues in shared memory, where the costs are different, providing an opportunity for different trade-offs to be made.
 
@@
 In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently.
-It is commonly accepted that the easiest consistency model to reason about enforces that all accesses appear to happen in some global serializable order (known as \emph{sequential consistency or SC}).
+It is commonly accepted that the easiest consistency model to reason about is \emph{sequential consistency} (SC), which enforces that all accesses appear to happen in some global serializable order.
 To preserve SC, operations on shared data structures should be \emph{linearizable}~\cite{herlihy1990linearizability}; that is, appear to happen atomically in some global total order.
-However, in both physically shared memory and PGAS, maintaining linearizability presents performance challenges.
-The simplest way is to have a single global lock that implements mutual exclusion and enforces a serializable order over read and write operations.
+In both physically shared memory and PGAS, maintaining linearizability presents performance challenges.
+The simplest way is to have a single global lock for implementing mutual exclusion for all operations, enforcing atomicity and linearizability.
 The cost of literally serializing accesses in this way is typically considered prohibitive, even in physically shared memory.
@@
 
-At the core of a new synchronization technique called \emph{flat combining}~\cite{flatCombining} is the idea of having threads \emph{cooperate} rather than \emph{contend}.
-Threads delegate their work to a single thread, giving it the opportunity to combine multiple requests in data-structure specific ways and performing them free from contention.
+A new synchronization technique called \emph{flat combining}~\cite{flatCombining} coerces threads to \emph{cooperate} rather than \emph{contend}.
+Threads delegate their work to a single thread, giving it the opportunity to combine multiple requests in data-structure specific ways and perform them free from contention.
 Together, the reduced synchronization cost and combining allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.
@@
 We leverage the Grappa runtime~\cite{Nelson:hotpar11-real}, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads.
-By leveraging lots of concurrency, we find that it is actually possible to scalably enforce strict sequential consistency.
-We leverage Grappa's latency tolerance mechanisms to allow many fine-grained synchronized operations to be combined to reduce communication and achieve higher overall throughput.
+By leveraging lots of concurrency, we find that it is actually possible to scalably provide sequential consistency.
+We leverage Grappa's latency tolerance mechanisms to allow many fine-grained synchronized operations to be combined to achieve higher overall throughput.
 In addition, we show how a generic flat-combining framework can be used to implement multiple global data structures.

### 2013-06-15 doc/pgas13/evaluation.tex
More aggressive cutting to get down to 8 pages.

@@
 Experiments were run on a cluster of AMD Interlagos processors.
-Nodes have 32 2.1-GHz cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand network cards, connected via a QLogic InfiniBand switch.
+Nodes have 32 2.1-GHz cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand network cards connected via a QLogic switch.
 
@@
 
-On the mixed workload, the stack is able to do matching locally, allowing it to reduce the amount of communication drastically, greatly improving its performance on the mixed workload.
+On the mixed workload, the stack is able to do matching locally, allowing it to reduce the amount of communication drastically, greatly improving its performance.
 Figure~\ref{fig:stack_stats} corroborates this, showing that the amount of combining that occurs directly correlates with the observed throughput.
 
-We also experimented with performing both forms of combining together; however, combining at the master did not affect performance significantly.
-This was somewhat surprising as combined updates still force others to wait.
-However, with both levels of combining enabled, relatively few requests are combined at the master--for the stack, a maximum of 6\% per pass for the all-push workload and 0.1\% for the mixed workload.
+% % left out for space
+% We also experimented with performing both forms of combining together.
+Combining at the master did not affect performance significantly, which is somewhat surprising as all combined requests must still serialize.
+% However, with both levels of combining enabled, relatively few requests are combined at the master--for the stack, a maximum of 6\% per pass for the all-push workload and 0.1\% for the mixed workload.
 
@@
 However, after 32 nodes, scaling drops off significantly due to the increased number of destinations.
-Combining allows duplicate inserts and lookups to be eliminated, so performs better the smaller the key range. This reduction in message traffic allows scaling to continue out to 64 nodes.
+Combining allows duplicate inserts and lookups to be eliminated, so performs better the smaller the key range. This reduction in message traffic allows scaling out to 64 nodes.
 
@@
 
-Figure~\ref{fig:bfs_perf} shows the results of scaling up to 64 nodes.
+Figure~\ref{fig:bfs_perf} shows the scaling results.
 Using the simple queue without flat combining is completely unscalable, but with FC, it tracks the tuned version.
@@
 The tuned version outperforms the synchronous version because it is able to build up most of the set locally on each core before merging them at the end of the first phase.
-An implementation that did not provide synchronized semantics could potentially relax consistency in a more general way, but this is left for future work.
+An implementation that did not provide synchronized semantics could potentially relax consistency in a more general way.

### 2013-06-15 doc/pgas13/flat-combining.tex
More aggressive cutting to get down to 8 pages.

@@
 Simply by delegating work to another core, locality is improved and synchronization is reduced.
-% Reduced synchronization comes from delegating work to another core.
-To illustrate the issue, consider the shared synchronous stack shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock.
+Consider the shared synchronous stack shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock.
 Without flat combining, whenever a thread attempts to push something on the stack, it must acquire the lock, put its value into the storage array, bump the top, and then release the lock.
@@
 % Instead, in the Grappa FC Framework, 
-A \emph{proxy} object is instantiated ahead of time on each core and each worker merges its request into that structure before blocking or becoming the combiner.
-Each global data structure must define a \emph{proxy} with the following:
-\begin{enumerate}
-  \item A local \emph{proxy} data structure for tracking combined requests.
-  \item \emph{Combining methods} that operate on the local proxy.
-  \item A \emph{sync} method that globally commits the proxy's state.
-\end{enumerate}
-An example proxy declaration for the GlobalStack is shown in Figure~\ref{fig:proxy}.
+For each global structure instance, a \emph{proxy} object is created on each core and each worker merges its request into that structure before blocking or becoming the combiner.
+Each global data structure must define a \emph{proxy} that has \emph{state} to track combined requests, \emph{methods} to add new requests, and a way to \emph{sync} the state globally.
+% \begin{enumerate}
+%   \item A local \emph{proxy} data structure for tracking combined requests.
+%   \item \emph{Combining methods} that operate on the local proxy.
+%   \item A \emph{sync} method that globally commits the proxy's state.
+% \end{enumerate}
+An example proxy declaration for the GlobalStack is shown in Figure~\ref{fig:push}.
 
@@
 \begin{lstlisting}[style=grappa]
+// Global master state
 class GlobalStack {
@@
 };
-
+// Push *without* combining
 void push(GlobalAddress<GlobalStack> master, T e){
-  // perform operation on the stack's master core
+  // from stack's master core, perform write to top and increment atomically
   delegate::call(master.core(), [master,e]{
-    // force other operations to wait
     master->lock.acquire();
-    // do remote write to the top of the stack
     delegate::write(master->top, e);
-    // bump top and release lock
     master->top++;
@@
 }
-\end{lstlisting}
-\caption{Snippet of code for pushing to a global stack (without flat combining).}
-\label{fig:push}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\begin{lstlisting}[style=grappa]
+// Definition of proxy
 class GlobalStackProxy : Grappa::FCProxy {
@@
 \end{lstlisting}
-\caption{Snippet of the Proxy object for the GlobalStack. The code to synchronize pushes should resemble that from Figure~\ref{fig:push} but for multiple elements.}
-\label{fig:proxy}
+\caption{Snippet of code from GlobalStack.
+% Note how the combined \texttt{push} resembles the uncombined version but for multiple elements.
+}
+\label{fig:push}
 \end{figure}
 
-With this data-structure-specific proxy object, the FC framework does the bookkeeping to block and wake workers, deliver results, manage when to synchronize, and choose the combiner.
+% The FC framework does the bookkeeping to block and wake workers, deliver results, manage when to synchronize, and choose the combiner.
 The FC framework is responsible for ensuring that all of the combined operations eventually get committed.
@@
 \subsection{Global Stack and Queue}
-Figure~\ref{fig:proxy} shows an excerpt from the definition of the proxy object for the Grappa GlobalStack.
-The proxy tracks pushes locally in the \texttt{pushed\_values} array.
-When \texttt{pop} is called, if there are pushes in the local array, it immediately takes the value and wakes the last worker to push.
-Otherwise, it adds a pointer to a location on its stack where the \texttt{sync} operation, will eventually write the result.
+Figure~\ref{fig:push} shows an excerpt from the definition of the proxy object for the Grappa GlobalStack.
+The proxy tracks pushes in the \texttt{pushed\_values} array.
+When \texttt{pop} is called, if there are pushes, it immediately takes the top value and wakes the last pusher.
+Otherwise, it adds a pointer to a location on its stack where the \texttt{sync} operation will write the result.
 Because of local matching, when a Stack proxy is synchronized, it will have either all pops or all pushes, which makes the implementation of sync straightforward.
@@
 
-The GlobalQueue has nearly the same implementation as the stack, but is unable to match locally, so \texttt{sync} first commits the pushes, then the pops.
+The GlobalQueue has nearly the same implementation as the stack, but is unable to match locally.
+%, so \texttt{sync} first commits the pushes, then the pops.
 
@@
 Our implementation of the GlobalMap matches that of the Set but of course stores values.
-This does not change the requirements of the proxy except that the return values of lookups must be a copy of the value rather than just a boolean.
+ % This does not change the requirements of the proxy except that the return values of lookups must be a copy of the value rather than just a boolean.

### 2013-06-15 doc/pgas13/intro.tex
More aggressive cutting to get down to 8 pages.

@@
 This allows programmers to write their algorithms without needing to explicitly manage communication.
-However, with this shared memory abstraction come all of the difficult concurrency issues that arise in physically shared memory.
-Luckily, the PGAS community can leverage a large body of work solving these issues in shared memory, where the costs are different, providing an opportunity for different trade-offs to be made.
+However, it does not alleviate the need for reasoning about consistency among concurrent threads.
+Luckily, the PGAS community can leverage a large body of work solving these issues in shared memory and explore how differing costs lead to different design choices.
 
-Globally shared data structures are one of the cornerstones of shared-memory and PGAS abstractions. 
-In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently.
+% Globally shared data structures are one of the cornerstones of shared-memory and PGAS abstractions. 
+% In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently.
 It is commonly accepted that the easiest consistency model to reason about is \emph{sequential consistency} (SC), which enforces that all accesses appear to happen in some global serializable order.
@@
 The cost of literally serializing accesses in this way is typically considered prohibitive, even in physically shared memory.
+% Much prior work has explored the scalability challenges of adding more concurrent accessors to shared data structures in shared memory systems.
+In the classic multithreading case, as the number of concurrent accessors increases, the more they contend and the more synchronization operations fail.
 With the massive amount of parallelism in a cluster of multiprocessors and with the increased cost of remote synchronization, the problem is magnified.
-
-Much prior work has explored the scalability challenges of adding more concurrent accessors to shared data structures in shared memory systems.
-In the classic multithreading case, increasing concurrency increases contention and damages performance.
-Either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of contentious atomic operations, and retry if they fail.
-In both cases, as the number of concurrent accessors increases, the more they contend and the more synchronization operations fail.
 
@@
 Threads delegate their work to a single thread, giving it the opportunity to combine multiple requests in data-structure specific ways and perform them free from contention.
-Together, the reduced synchronization cost and combining allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.
+This allows even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.
 
@@
 We leverage the Grappa runtime~\cite{Nelson:hotpar11-real}, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads.
-By leveraging lots of concurrency, we find that it is actually possible to scalably provide sequential consistency.
-We leverage Grappa's latency tolerance mechanisms to allow many fine-grained synchronized operations to be combined to achieve higher overall throughput.
+We leverage Grappa's latency tolerance mechanisms to allow many fine-grained synchronized operations to be combined to achieve higher, scalable throughput while maintaining sequential consistency.
 In addition, we show how a generic flat-combining framework can be used to implement multiple global data structures.
@@
 We then explain the flat-combining paradigm in more depth and describe how it maps to a PGAS model.
-Next, we explain how several data structures are implemented in our framework and show how they perform on simple throughput workloads.
-Finally, we evaluate how this affects application performance in two simple graph analysis kernels.
+Next, we explain how several data structures are implemented in our framework and show how they perform on simple throughput workloads as well as in two simple graph analysis kernels.

### 2013-06-15 doc/pgas13/relatedwork.tex
More aggressive cutting to get down to 8 pages.

@@
 \section{Related Work}
-Attempts to improve the scalability of synchronized data structures are as old as the data structures themselves.
-Though much attention has gone to lock-free data structures such as the Treiber Stack~\cite{treiber} and Michael-Scott Queue~\cite{msqueue}, a significant body of work has explored ways of scaling globally locked data structures.
-
-Many techniques performed combining hierarchically to reduce synchronization and hot-spotting, differing mainly in the structures used to do combining: fixed trees~\cite{yew:combining-trees}, dynamic trees (or ``funnels'')~\cite{funnels,MAMA}, and randomized trees~\cite{edtrees}.
+% Attempts to improve the scalability of synchronized data structures are as old as the data structures themselves.
+Though much attention has gone to lock-free data structures such as the Treiber Stack~\cite{treiber} and Michael-Scott Queue~\cite{msqueue}, a significant body of work has explored ways of scaling globally locked data structures using combining to reduce synchronization and hot-spotting.
+Techniques differ mainly in the structures used to do combining: fixed trees~\cite{yew:combining-trees}, dynamic trees (or ``funnels'')~\cite{funnels,MAMA}, and randomized trees~\cite{edtrees}.
 In particular, MAMA~\cite{MAMA}, built for the MTA-2, leveraged funnels to better utilize hardware optimized for massive concurrency.
@@
 The relatively higher cost of communication in PGAS makes flat combining even more compelling.
-Within-core combining is naturally done in a \emph{flat} way.
 In addition, the physical isolation per node combined with Grappa's engineered per-core isolation guarantees, create a novel situation for combining, which led to our \emph{combining proxy} design.
@@
 This work demonstrates a similar use case for synchronized queues as in our work, and uses them with the Graph500 benchmark as well.
-In contrast, Grappa performs much of the same message aggregation invisibly to the programmer, and the FC framework can be used to implement many different linearizable data structures.
+In contrast, Grappa performs much of the same message aggregation invisibly to the programmer, and the FC framework can be used to implement many data structures.

### 2013-06-15 doc/pgas13/evaluation.tex
edits from proofreading, use new-found space for larger pics

@@
 \section{Evaluation}
-To evaluate the impact flat combining has on the global data structures, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and then measure their impact on performance of two graph benchmarks.
+To evaluate the impact flat combining has, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and then measure the impact on performance of two graph benchmarks.
 Experiments were run on a cluster of AMD Interlagos processors.
@@
 Combining at the master did not affect performance significantly, which is somewhat surprising as all combined requests must still serialize.
-% However, with both levels of combining enabled, relatively few requests are combined at the master--for the stack, a maximum of 6\% per pass for the all-push workload and 0.1\% for the mixed workload.
+% However, with both levels of combining enabled, relatively few requests are combined at the master---for the stack, a maximum of 6\% per pass for the all-push workload and 0.1\% for the mixed workload.
 
@@
 % After all edges have been traversed in this way, the set of edges forms a new, typically much smaller, graph.
-The second phase performs the classical Shiloach-Vishkin parallel algorithm~\cite{shiloach1982n} on the reduced graph formed by phase one's edge set, and the final phase propagates the component labels out to the full graph.
+The second phase performs the classical Shiloach-Vishkin parallel algorithm~\cite{shiloach1982n} on the reduced graph formed by the edge set from the first phase, and the final phase propagates the component labels out to the full graph.
 Creation of the reduced edge set dominates the runtime of this algorithm, but includes many repeated inserts at the boundary between traversals, so is a prime target for the flat-combining Set.
-As in BFS, further optimizations involving relaxation of consistency can be applied, in this case to the Set.
+Similar to BFS, the Set is write-only first, then read-only later, so further optimizations involving relaxation of consistency can be applied.
 Therefore, we compare our straightforward implementation using the generic Set with and without flat combining against a tuned asynchronous implementation.

### 2013-06-15 doc/pgas13/flat-combining.tex
edits from proofreading, use new-found space for larger pics

@@
   \centering
-  \includegraphics[width=0.38\textwidth]{figs/fc_shared_mem.pdf}
+  \includegraphics[width=0.42\textwidth]{figs/fc_shared_mem.pdf}
   \caption{\emph{Flat combining in shared memory.}
@@
 The combiner walks the publication list, performs all of the requests, and when done, releases the lock.
-This allows the one thread to keep the data structure in cache, reducing thrashing between threads.
-It also greatly reduces contention on the lock, but introduces a new point of synchronization--adding to the publication list.
+This allows the one thread to keep the data structure in cache, reducing thrashing between threads on different cores.
+It also greatly reduces contention on the lock, but introduces a new point of synchronization---adding to the publication list.
 However, if a thread performs multiple operations, it can leave its publication record in the list and amortize the synchronization cost.
@@
     \centering
-    \includegraphics[width=0.35\textwidth]{figs/stack_nofc.pdf}
+    \includegraphics[width=0.40\textwidth]{figs/stack_nofc.pdf}
     \caption{\emph{Stack without combining.}
@@
     \centering
-    \includegraphics[width=0.38\textwidth]{figs/stack_fc.pdf}
+    \includegraphics[width=0.42\textwidth]{figs/stack_fc.pdf}
     \caption{\emph{Stack with distributed combining.}
@@
 This means that an explicit publication list with clever synchronization is unnecessary.
-Instead, workers merging their operations directly into the local proxy, and block if unable to satisfy their requests immediately.
+Instead, workers merge their operations directly into the local proxy, and block if unable to satisfy their requests immediately.
 The proxy structure is chosen to be able to compactly aggregate operations and efficiently perform matching in cases where it is allowed.
@@
 The Grappa memory model is essentially the same as the C++ memory
-model~\cite{boehm:drf0,N2480,N2800} (guaranteeing sequential
-consistency for data-race free programs). To be conservative, delegate
+model~\cite{boehm:drf0,N2480,N2800}, guaranteeing sequential
+consistency for data-race free programs. To be conservative, delegate
 operations block the calling worker until they have become globally visible,
@@
 To track all of the keys waiting to be inserted, we use the hash set implementation from the C++11 standard library (\texttt{std::unordered\_set}).
-As with pops, lookups must provide pointers for results to be put, which is done with a hash map (again from the C++ standard library) from keys to lists of pointers.
+As with pops, lookups must provide pointers to stack space where results should be put, which is done with a hash map (again from the C++ standard library) from keys to lists of pointers.
 As discussed in Section~\ref{sec:memory-model}, matching lookups with inserts locally would force a particular sequential order.

### 2013-06-15 doc/pgas13/grappa.tex
Simple space hacking w/o sacrificing content.
edits from proofreading, use new-found space for larger pics

@@
 Luckily, applications in this class have another thing in common: abundant amounts of data access parallelism.
-This parallelism can be exploited in a number of different ways to improve overall throughput of these applications.
+This parallelism can be exploited in a number of different ways to improve overall throughput.
 
@@
 Grappa is a global-view PGAS runtime for commodity clusters which has been designed from the ground up to achieve high performance on irregular applications.
-The key is latency tolerance--long-latency operations such as reads of remote memory can be tolerated by switching to another concurrent thread of execution.
+The key is latency tolerance---long-latency operations such as reads of remote memory can be tolerated by switching to another concurrent thread of execution.
 Given abundant concurrency, there are opportunities to increase throughput by sacrificing latency.
@@
 
-Grappa's programming interface, implemented as a C++11 library, provides high-level operations to access and synchronize through global shared memory, task spawning capabilities and parallel loop constructs for expressing concurrency.
+Grappa's programming interface, implemented as a C++11 library, provides high-level operations to access and synchronize through global shared memory, and task and parallel loop constructs for expressing concurrency.
 In addition, the Grappa ``standard library'' includes functions to manipulate a global heap, stock remote synchronization operations such as \emph{compare-and-swap}, and several synchronized global data structures.
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figs/grappa_system.pdf}
+  \includegraphics[width=0.35\textwidth]{figs/grappa_system.pdf}
   \caption{\emph{Grappa System Overview:}
@@
 The Grappa runtime has a lightweight threading system that uses prefetching to scale up to thousands of threads on a single core with minimal increase in context-switch time.
-In the runtime, \emph{Worker} threads pull these programmer-specified tasks from a queue and execute them to completion.
-When a task blocks, the worker thread executing it is simply suspended and consumes no computational resources until woken again by some event.
+In the runtime, \emph{worker} threads pull these programmer-specified tasks from a queue and execute them to completion.
+When a task blocks, the worker thread executing it is suspended and consumes no computational resources until woken again by some event.
 
@@
 In the PGAS style, Grappa provides a global address space partitioned across the physical memories of the nodes in a cluster.
-Each core owns a portion of memory, which is divided among execution stacks for the core's Workers, a core-local heap, and a slice of the global heap.
+Each core owns a portion of memory, which is divided among execution stacks for the core's workers, a core-local heap, and a slice of the global heap.
 
@@
 
-All accesses to memory owned by another core must be done by that core via a message, even enforcing strict isolation between processes on the same physical memory.
+Grappa enforces strict isolation---all accesses must be done by the core that owns it via a message, even between processes on the same physical memory.
 At the programming level, however, this is hidden behind higher-level remote operations, which in Grappa are called \emph{delegates}.

### 2013-06-15 doc/pgas13/intro.tex
edits from proofreading, use new-found space for larger pics

@@
 In both physically shared memory and PGAS, maintaining linearizability presents performance challenges.
-The simplest way is to have a single global lock for implementing mutual exclusion for all operations, enforcing atomicity and linearizability.
-The cost of literally serializing accesses in this way is typically considered prohibitive, even in physically shared memory.
+The simplest way is to have a single global lock to enforce atomicity and linearizability through simple mutual exclusion.
+Literally serializing accesses in this way is typically considered prohibitively expensive, even in physically shared memory.
 % Much prior work has explored the scalability challenges of adding more concurrent accessors to shared data structures in shared memory systems.
-In the classic multithreading case, as the number of concurrent accessors increases, the more they contend and the more synchronization operations fail.
+However, even for fine-grained lock-free synchronization schemes, as the number of concurrent accessors increases, the more they contend and the more synchronization operations fail.
 With the massive amount of parallelism in a cluster of multiprocessors and with the increased cost of remote synchronization, the problem is magnified.
@@
 The goal of this work is to apply the flat-combining paradigm to a PGAS runtime to reduce the cost of maintaining sequentially consistent data structures.
-We leverage the Grappa runtime~\cite{Nelson:hotpar11-real}, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads.
+We leverage Grappa, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads as sketched out in prior work~\cite{Nelson:hotpar11-real}.
 We leverage Grappa's latency tolerance mechanisms to allow many fine-grained synchronized operations to be combined to achieve higher, scalable throughput while maintaining sequential consistency.

### 2013-06-19 doc/pgas13/intro.tex
Some edits based on feedback from Brad.

@@
 % Much prior work has explored the scalability challenges of adding more concurrent accessors to shared data structures in shared memory systems.
-However, even for fine-grained lock-free synchronization schemes, as the number of concurrent accessors increases, the more they contend and the more synchronization operations fail.
+However, as the number of concurrent accessors increases, the more they contend and the more synchronization operations fail, even for fine-grained lock-free synchronization schemes.
 With the massive amount of parallelism in a cluster of multiprocessors and with the increased cost of remote synchronization, the problem is magnified.
@@
 The goal of this work is to apply the flat-combining paradigm to a PGAS runtime to reduce the cost of maintaining sequentially consistent data structures.
-We leverage Grappa, a PGAS-style runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads as sketched out in prior work~\cite{Nelson:hotpar11-real}.
+We leverage Grappa, a PGAS runtime library optimized for fine-grained random access, which provides the ability to tolerate long latencies by efficiently switching between many lightweight threads as sketched out in prior work~\cite{Nelson:hotpar11-real}.
 We leverage Grappa's latency tolerance mechanisms to allow many fine-grained synchronized operations to be combined to achieve higher, scalable throughput while maintaining sequential consistency.
@@
 We then explain the flat-combining paradigm in more depth and describe how it maps to a PGAS model.
-Next, we explain how several data structures are implemented in our framework and show how they perform on simple throughput workloads as well as in two simple graph analysis kernels.
+Next, we explain how several data structures are implemented in our framework and show how they perform on simple throughput workloads as well as in two graph analysis kernels.

### 2013-06-20 doc/pgas13/flat-combining.tex
Some edits based on feedback from Brad.
A couple minor things based on Jacob's feedback.

@@
 
-\subsection{In Grappa}
+\subsection{Grappa}
 
@@
 This means that an explicit publication list with clever synchronization is unnecessary.
-Instead, workers merge their operations directly into the local proxy, and block if unable to satisfy their requests immediately.
+Instead, workers merge their operations directly into the local proxy, and 
+% block if unable to satisfy their requests immediately.
+block, except in restricted cases where they are able to satisfy their requirements immediately without violating ordering rules for consistency (discussed in the next section).
 The proxy structure is chosen to be able to compactly aggregate operations and efficiently perform matching in cases where it is allowed.
@@
 Therefore, a batch of these operations can be committed globally in parallel, since they are guaranteed not to conflict with each other.
-Note that if an insert finds that its key is already in the combiner, it does not need to send its own message.
+Note that if an insert finds that its key is already in the combiner object, it does not need to send its own message.
 However, it must still block until that insert is done to ensure that a subsequent operation it performs cannot be reordered with it, preserving program order.
@@
 If this is allowed, then to preserve SC, that same order must be respected globally, forcing each batch to commit atomically with respect to other cores' batches.
-Enforcing this is prohibitively expensive, so a cheaper solution is chosen: if lookups must get their values from the global data structure, then batches can be performed in parallel.
+Enforcing this is prohibitively expensive, so a cheaper solution is chosen: lookups must get their values from the global data structure, so batches can be performed in parallel.
 
@@
 The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress.
-When hooking into the framework, each data structure implementation need only define how to combine operations and globally commit them.
+When hooking into the framework, each data structure need only define how to combine operations and globally commit them.
 

### 2013-06-20 doc/pgas13/intro.tex
A couple minor things based on Jacob's feedback.

@@
 % In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently.
-It is commonly accepted that the easiest consistency model to reason about is \emph{sequential consistency} (SC), which enforces that all accesses appear to happen in some global serializable order.
+It is commonly accepted that the easiest consistency model to reason about is \emph{sequential consistency} (SC), which enforces that all accesses are visible in program order and appear to happen in some global serializable order.
 To preserve SC, operations on shared data structures should be \emph{linearizable}~\cite{herlihy1990linearizability}; that is, appear to happen atomically in some global total order.

### 2013-06-30 doc/pgas13/evaluation.tex

@@
 \end{figure}
-Connected Components (CC) is another simple graph analysis kernel that illustrates a different use of global data structures in irregular applications.
+Connected Components (CC) is another core graph analysis kernel that illustrates a different use of global data structures in irregular applications.
 We implement the three-phase CC algorithm~\cite{mtgl} which was designed for the massively parallel MTA-2 machine.

### 2013-06-30 doc/pgas13/grappa.tex

@@
 
-Highly tuned implementations of irregular applications in shared-memory, PGAS and message-passing paradigms, typically end up implementing similar constructs.
+Highly tuned implementations of irregular applications in shared-memory, PGAS, and message-passing paradigms, typically end up implementing similar constructs.
 Grappa includes these as part of its core infrastructure and simply asks the programmer to express concurrency which it can leverage to provide performance.
@@
 In addition, the Grappa ``standard library'' includes functions to manipulate a global heap, stock remote synchronization operations such as \emph{compare-and-swap}, and several synchronized global data structures.
+These features make it suitable for implementing some next-generation PGAS languages like Chapel~\cite{chapel} and X10~\cite{x10}.
 The following sections will explain the execution model of Grappa and the current C++ programming interface.

### 2013-06-30 doc/pgas13/relatedwork.tex

@@
 
-Our work extends the flat-combining paradigm further, to the PGAS domain, where only software provides the illusion of a physically distributed memory.
-The relatively higher cost of communication in PGAS makes flat combining even more compelling.
+Our work extends the flat-combining paradigm further, to the PGAS domain, where only software provides the illusion of global memory to physically distributed memories.
+The relatively higher cost of fine-grained data-driven access patterns in PGAS makes flat combining even more compelling.
 In addition, the physical isolation per node combined with Grappa's engineered per-core isolation guarantees, create a novel situation for combining, which led to our \emph{combining proxy} design.

### 2013-07-13 doc/sosp13/evaluation.tex
incorporate todos reviews

@@
 
+Our questions to answer with experimentation are:
 The goal of our evaluation is to show that the core pieces of the \Grappa
@@
 
-\item We present results that show that the \Grappa runtime is able to sustain
-very high concurrency rates and the communication layer is able to sustain a
-very high rate of global memory operations. We also show the performance of a
+\item We present results that show that \Grappa can
+    support large amounts of concurrency: sufficient for remote memory
+    access and aggregation. The communication layer is able to
+    achievesustain a very high rate of global memory operations. We also show the performance of a
 graph kernel that stresses communication and concurrency together.
+\TODO{compare to key-value stores?}
 
-\item We show how some popular irregular applications running on \Grappa
+\item We characterize system behavior, including
+profiling where execution time goes, how aggregation affects message size and
+rates, and \checkme{the relationship between aggregation and
+    concurrency.}
+
+\item Finally, we show how some more realistic irregular workloads on \Grappa
 compare to the Cray XMT and hand-tuned MPI code.
 
-\item We finish with a characterization of system behavior, including
-profiling where execution time goes, how aggregation affects message size and
-rates, how global memory and work stealing behaves.
+\TODO{Grappa is well-suited to applications where there is
+    abundant data parallelism but unpredictable or imbalanced data
+    access. BSP is efficient where there is sufficient parallel slack.
+    Although Grappa loses some asynchrony by relying heavily on
+    aggregation, it still provides more flexibility for sections of the
+    computation to proceed more quickly.
+
+\TODO{how much concurrency is needed on apps. Our basic plot of
+    performance vs workers (with and without aggro?)}
+\TODO{relationship between user-thread-visible latency and aggregate throughput.}
+\TODO{characterization of aggregation message size vs nodes}
+\TODO{show our scalability plots; if they don't scale, show
+    specifically why. It is okay if its because aggregation stops
+    working after a point if we can discuss future scalability.
+    Perhaps reference ``MPI on millions of cores'' type work}
 
@@
 
+\TODO{go into more detail about what Grappa is doing when executing
+    GUPs. Specifically note what the rate per core is, what the
+    aggregation and deaggregation rates are. Show on the plot the
+    message size distribution (some box plot)}
+
 \paragraph{Putting it all together with Unbalanced Tree Search in memory
@@
 the original benchmark. T1 trees are very shallow and wide (i.e., significant
-parallel slack), while T3 trees are very deep (i.e., little parallel slack).
+parallel slack \TODO{cite Valiant and say more somewhere about BSP}), while T3 trees are very deep (i.e., little parallel slack).
 When the access to each vertex is a random access, the critical path to search
@@
 
+\TODO{addressing reviwer2: sweep of irregularness in one aggregate or multiple dimensions
+    and how a specific aspect of performance changes in different
+    systems}
+
 Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
@@
 \end{figure}
+\TODO{ shorten the execution profile section in favor of deeper
+    discussion about where the message rate comes from}
 
@@
 
-In BFS, we see a trimodal distribution.
-There are two reasons for this.
-First, in BFS, we obtained the best throughput by flushing at least 500 microseconds apart, and also flushing whenever a core is idle.
-Since this process ignores whether another flush has occurred recently, it leads to a large number of small messages.
-This leads to the first and smallest mode of the distribution.
-The other two modes are due to a locality optimization in the \Grappa version of BFS.
-To minimize unnecessary remote communication, each core visits vertices whoes edge lists are stored locally.
+In BFS, we see a trimodal distribution. There are two reasons for
+this. First, in BFS, we obtained the best throughput by flushing at
+least 500 microseconds apart, and also flushing whenever a core is
+idle. Since this process ignores whether another flush has occurred
+recently, it leads to a large number of small messages. This leads to
+the first and smallest mode of the distribution.  The other two modes
+are due to a locality optimization in the \Grappa version of BFS. To
+minimize unnecessary remote communication, each core visits vertices
+whose edge lists are stored locally.
 Since we are traversing power-law graphs, a small fraction of vertices have a very high number of edges.

### 2013-07-13 doc/sosp13/memory.tex
incorporate todos reviews

@@
 
+\TODO{eliminate the word cache from the paper since the main point is
+    contiguity and not unpredictable temporal reuse. We
+    must make it clear that this is not user-level caching.}
+
 \paragraph{Memory consistency model discussion} As mentioned earlier, all
@@
 delegates have been fully executed.
+
+\TODO{admit the lack of a synchronization feature for lightweight
+    transactions across two or more domains. Talk about using the same
+    philosophy of moving data along with the sync (same as a real
+    cache) and utilizing latency tolerance. Then say further is out of
+    scope for this paper.}

### 2013-07-19 doc/asplos14/conclude.tex
make asplos14 directory from sosp13

--- 2013-03-28 doc/sosp13/conclude.tex
@@
 Irregular computations are both important and challenging to execute quickly.
-Scaling these applications easily to commodity hardware has been a historical
+Scaling these applications easily on commodity hardware has been a historical
 challenge. \Grappa is a runtime framework that simplifies this task for
@@
 framework and its three main components: a task library, a distributed shared
-memory system, and a network aggregator to make commodity networks efficient
-with small message sizes. \Grappa's key aspect is extreme latency tolerance,
+memory system, and a network aggregator. \Grappa's key aspect is extreme latency tolerance,
 which not only hides network latency but also enables the system to spend time
@@
 Aggregating messages enables \Grappa to achieve over 1.0 GUPS on 64 nodes.
-We also explored four other algorithms: unbalanced tree search, breadth first search, PageRank, and integer sort, comparing performance of these algorithms on \Grappa, the Cray XMT, and against hand-optimized MPI implementations.
+We also explored four other algorithms: unbalanced tree search, breadth first search, PageRank, and integer sort, comparing performance of these algorithms on \Grappa, the Cray XMT, and hand-optimized MPI implementations.
 As would be expected, when the MPI implementation exploits in the application, many of the same techniques \Grappa provides in the runtime, such as aggregating requests, yet without the overhead of tasking, the MPI implementation excels.
-This comes with enormous implementation complexity burdened on the application developer, however, and significant tuning has yet to be done on the \Grappa runtime system.
-Compared to the Cray XMT \Grappa is between 2X faster and 4X slower.
+This comes with enormous implementation complexity, burdening the application developer.
+Moreover, significant tuning has yet to be done on the \Grappa runtime system.
+Compared to the Cray XMT \Grappa is between 2.6X faster and 4.3X slower.
 Yet when cost-performance is considered, mass-market x86 cluster hardware makes \Grappa a highly attractive option.

### 2013-07-19 doc/asplos14/evaluation.tex
make asplos14 directory from sosp13

--- 2013-07-13 doc/sosp13/evaluation.tex
@@
     aggregation, it still provides more flexibility for sections of the
-    computation to proceed more quickly.
+    computation to proceed more quickly. }
 
@@
 
-As discussed earlier, fast context switching is at the heart of \Grappa's
+Fast context switching is at the heart of \Grappa's
 latency tolerance abilities. We assess context switch overheads using a simple
-microbenchmark that has a configurable number of workers, where each worker 
-just increments values in a large array.
+microbenchmark that runs a configurable number of workers on a single core, where each worker 
+increments values in a large array.
 
@@
 
-Figure~\ref{fig:context-switch-exp} shows the average context switch time as
-the number of workers grow. There are two important takeaway points from these
-data. Context switch time for the number of workers we use ($\approx$1K workers) is
-on the order of 50 ns; and prefetching has significant impact on context
-switch time. And context switch time does not grow significantly with the
-number of workers --- even with one-half million workers, context
-switch time is around 75ns. In comparison, for the same yield test using kernel-level
-Pthreads on a single core, the switch time is 450ns for a few threads and 800ns for
-1000-32000 threads.
+Figure~\ref{fig:context-switch-exp} shows the average context switch
+time as the number of workers grow. At our standard operating point
+($\approx$1K workers), context switch time is on the order of 50
+ns. As we add workers, the time increases slowly, but levels off: we
+also ran with 500,000 workers (10 times what is shown in the figure)
+and found that context switch time was around 75 ns. In comparison,
+for the same yield test using kernel-level Pthreads on a single core,
+the switch time is 450ns for a few threads and 800ns for 1000-32000
+threads.
 
-If we calculate aggregate context switch\emph{rate} of all
+If we calculate aggregate context switch \emph{rate} of all
 cores in a node, we find that with prefetching, \Grappa context
@@
 limited \emph{not} by memory latency, as normally assumed, but rather memory bandwidth.
-Specifically, we emprically found that 4 cachelines (1 for worker
+Specifically, we empirically found that 4 cache lines (1 for worker
 struct and 3 for stack data) was sufficient to avoid cache misses in
-the microbenchmark. Every context switch then requires 8 cacheline transfers. The off-chip
-bandwidth of a single socket in our system is 270M cachelines per
-second \cite{porterfield:bw,Nelson:hotpar11-real}. This implies that
+the microbenchmark. Every context switch then requires 8 cache line transfers. The off-chip
+bandwidth of a single socket in our system is 270M cache lines per
+second \cite{porterfield:bw,Nelson:hotpar11}. This implies that
 we can sustain at most 34M context switches per second per socket once
@@
 \Grappa's global memory and communication layers using a faithful
-implementation of the giga updates per second (GUPs) benchmark.
+implementation of the giga updates per second (GUPs) benchmark, which measures cluster-wide random access bandwidth.
 Read-modify-write updates are dispatched at random to a large array. This
@@
 Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain well over a
-billion updates per second with 64 nodes.
-Note also that when aggregation is turned off, the update rate is flat at the bottom, demonstrating that aggregation is instrumental for good communication performance.
+billion updates per second with 64 nodes. Note also that when aggregation is turned off, the update rate is nearly flat. Aggregation is instrumental for good communication performance.
 
 This compares very favorably to published results~\cite{gups} for other
-high-end HPC systems. Though the actual computation done by GUPS is completely
-non-sensical, irregular, data-intensive applications typically must be able to
+high-end HPC systems. Though the actual computation done by GUPS is not 
+useful, irregular, data-intensive applications typically must be able to
 sustain a high rate of random accesses in order to, for example, visit and
@@
 Figure~\ref{fig:grappa-uts} shows the overall performance of \Grappa running
-UTS. The point of this experiment is to demonstrate whether \Grappa's context
-switching and communication layers can in fact be used together, while
-balancing workload, to run an irregular application efficiently. Visiting
-vertices in the distributed tree requires mostly remote accesses, and because
-each vertex in the tree must be visited before it can be expanded, blocking
-remote reads are required. 
-In this case, we are forced to context switch to tolerate the remote access and continue aggregating. 
+UTS. This experiment demonstrates that \Grappa's context
+switching and communication layers can be used together, while
+balancing workload, to run an irregular application efficiently.
+
+Visiting vertices in the distributed tree requires mostly remote
+accesses, and because each vertex in the tree must be visited before
+it can be expanded, blocking remote reads are required. In this case,
+we are forced to context switch to tolerate the remote access and
+continue aggregating.
+
 We look at two classes of trees, T1 and T3, from
@@
 the average active tasks per core over the search is 775 and 13 for T1 and T3,
-respectively. On the other hand \Grappa performs and scales very well for T1
+respectively. On the other hand, \Grappa performs and scales very well for T1
 trees.
@@
 within striking distance of XMT and well beyond UPC performance (UTS only).
-However, it underperforms compared to hand-tuned MPI. Nevertheless, this
-comparison needs to be taken with a grain of salt for several reasons. First,
-we have not spent as much time tuning \Grappa's implementations. This is
-supported by inspecting the BFS MPI implementation we used, which clearly
-shows that it employs a lot of algorithmic-specific optimizations. In fact,
-some of these optimizations resemble some of what \Grappa does automatically,
-like message aggregation. Second, the GUPS results presented earlier suggests
-that \Grappa performance can do a lot better.
+However, it underperforms compared to hand-tuned MPI. We believe there are two reasons for this:
+First, Grappa is young; much optimization and tuning of the runtime
+and the application is yet to be done. Second, the MPI implementations
+we used employ algorithmic-specific optimizations---some of which
+resemble of what \Grappa does automatically, like message aggregation.
+Grappa's GUPS results presented earlier suggest there is significant
+room for improvement.
 

### 2013-07-19 doc/asplos14/intro.tex
make asplos14 directory from sosp13

--- 2013-03-28 doc/sosp13/intro.tex
@@
 matched the processor clock frequency and sufficient bandwidth to sustain a
-reference from every processor on every clock cycle. Unfortunately, the MTA
+reference from every processor on every clock cycle. Unfortunately, the MTA's relatively low single-threaded performance meant that it
 was not general enough nor cost-effective. The Cray XMT approximates the Tera
@@
 
-Our evaluation of \Grappa shows that its core components provide scalable performance.
+Our evaluation of \Grappa shows that its core components provide scalable performance for irregular applications.
 It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.
@@
 On the other hand, if the MPI version is implemented in a straightforward manner, as we see with PageRank, GUPS, and unbalanced tree search, \Grappa is 2-5X faster.
-When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2X faster and 4X slower depending on the benchmark.
-Overall we find this encouraging because \Grappa is scalable and achieves this performance from a mass-market cluster, hence adding more system nodes is an inexpensive option for tackling larger problem sizes.
+When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.

### 2013-07-19 doc/asplos14/relatedwork.tex
make asplos14 directory from sosp13

\section{Related Work}

\vspace{0.5ex}
\noindent{\bf Multithreading} 
%\section{Multithreading}
\Grappa uses multithreading to tolerate memory
latency. This is a well known technique. Hardware implementations include the
Denelcor HEP~\cite{hep}, Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous
multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife},
Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}.

Grappa's closest ancestor is the Threaded Abstract
Machine~\cite{CullerGSvE93}. This was a software runtime system
designed as a prototyping platform for dataflow execution models on
distributed memory supercomputers, and contained support for
inter-node communication and management of the memory hierarchy as
well as context switching and scheduling computation. The Active
Messages~\cite{vonEicken:1992:AMM:139669.140382} work that grew out of
this project inspired our communication layer. One of the conclusions
of this work~\cite{Culler:1993:TFL:647025.714362} was that context
switch costs can be low only when contexts are in cache, and that
latency tolerance was not sufficient to guarantee performance on
commodity processors. \Grappa demonstrates that times have changed:
modern commodity processors have sufficient bandwidth and prefetch
capacity to stream contexts from DRAM.

\Grappa implements its own software-based multithreading with a
lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
on a single processing core. The large number of tasks is required to tolerate
the very high inter-node communication latency of commodity networks.
\Grappa's task library employs several optimizations: an extremely fast task
switch, a small task size, and judicious use of software prefetch instructions
to bring task state into the cache sufficiently long before that task is
actually scheduled. The main difference between \Grappa's support for
lightweight threads and prior work such as
QThreads~\cite{Wheeler08qthreads:an} and
Capriccio~\cite{Behren03capriccio:scalable} is context prefetching, which is needed for good performance when multiplexing such a large
number of tasks.

\vspace{0.5ex}
\noindent{\bf Software distributed shared memory.} 
%\section{Software distributed shared memory} 
The goal of providing a shared memory abstraction for a distributed
memory system goes back nearly 30 years. Here I present a few of the
ideas explored in this area.

Much of the innovation in SDSM has occurred around reducing the
synchronization cost of doing updates. The first DSM systems,
including IVY~\cite{Li:1989:MCS:75104.75105}, used invalidations to
provide sequential consistency, which imposed significant
communication cost for write-heavy workloads.  Later systems supported
more relaxed consistency models, including release consistency, which
reduced the cost by allowing updates to be buffered until
synchronization occurred. Furthermore, multiple writer protocols that
sent only modified data were developed to help combat false
sharing. The
Munin~\cite{Bennett:1990:MDS:99163.99182,Carter:1991:IPM:121132.121159}
and TreadMarks~\cite{Keleher:1994:TDS:1267074.1267084} systems
exploited both of these ideas, but some coherence overhead was still
required. In contrast, Grappa's delegate-based approach to updates
avoids synchronization overhead entirely, providing sequential
consistency for data-race-free programs without the cost of a
coherence protocol. This cost of this communication is mitigated through latency
tolerance.

Another way in which SDSM systems differ is in the granularity of
access control. Many SDSM systems, including IVY and TreadMarks,
tracked ownership at a granularity of a kilobyte or more. There are
two main justifications for this design choice: first, networks are
more efficient with large packets, and second, multi-kilobyte
granularity allowed systems to reuse the processor's paging mechanisms
to accelerate access control checks and to provide shared memory
transparently. Unfortunately, this meant that these systems depended
on lots of locality to amortize the cost of moving these large
blocks, and the systems were very susceptible to false sharing. Other
systems, including Munin and
Blizzard~\cite{Schoinas:1994:FAC:195473.195575}, allowed tracking
ownership with variable granularity to address these problems.
Grappa's delegate-based approach is similar; since updates are always
performed at data's home node, only modified data needs to be moved.

SDSM systems often required extensive modifications to the system
software stack, including the OS (IVY, Blizzard) and compilation
infrastructure (Munin). This made these systems difficult to port to
new platforms. Grappa follows the lead of TreadMarks and provides
SDSM entirely at user-level through a library and runtime.

In summary, while \Grappa's DSM system is conceptually similar to prior work,
we accept the random access aspect of irregular applications and optimize for
throughput rather than low latency. Our DSM system must support enough memory
concurrency to tolerate the latency of the network and additional latency
overhead impose by the runtime system.

\vspace{0.5ex}
\noindent{\bf Partitioned Global Address Space languages.} 
%\section{Partitioned Global Address Space languages} 
The high-performance computing community has largely discarded the
coherent distributed shared memory approach in favor of the
Partitioned Global Address Space (PGAS)
model. Split-C~\cite{Krishnamurthy:1993:PPS:169627.169724} is the
earliest example I know of this style of language; contemporary
examples include Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005},
Co-array Fortran~\cite{Numrich:1998:CFP:289918.289920} and
UPC~\cite{upc:2005}. Grappa shares many parts of its design philosophy
with these languages.

There are two key ideas Grappa draws from PGAS languages. First, PGAS
languages implement DSM at the language, rather than the system
level. This allows for a number of optimizations, including efficient
split-phase access, variable data granularity, and support for
user-customizable synchronization operations. Grappa follows this
approach for this reason.

Second, in PGAS languages, each piece of data has a single canonical
location on a particular node. The language designers expect
programmers to modify algorithms to take advantage of locality by
processing node-local data as much as possible; access to data stored
on other nodes is possible but is seen as something to avoid if
possible. Grappa is designed to support the opposite view: Grappa is
optimized for random access to data anywhere in the cluster, and
locality can be exploited when it is available.

Most PGAS languages adopt a SPMD programming model: the programmer
must reason about what is happening on every node in the
system. Chapel is the exception: it provides a global view of control,
while allowing programmers to direct individual cores when
necessary. Grappa follows the same approach, providing a
single-machine abstraction to the programmer along with support for
controlling the locality of computation when requested.

\vspace{0.5ex}
\noindent{\bf Distributed graph processing systems.} 
%\section{Distributed graph processing systems} 
While \Grappa is a general runtime system for any large-scale
concurrent application, it has been designed to perform especially
well on graph analysis. Other distributed graph processing frameworks
include Pregel~\cite{pregel:2010} and Distributed
GraphLab~\cite{distgraphlab:vldb12}. Pregel adopts a bulk-synchronous
parallel (BSP) execution model, which makes it inefficient on
workloads that could prioritize vertices. GraphLab overcomes this
limitation with an execution mode that schedules vertex computations
individually, allowing prioritization, which gives faster convergence
in a variety of iterative algorithms. GraphLab, however, imposes a
rigid computation model where programmers must express computation as
transformations on a vertex and its edge list, with information only
from adjacent vertexes. Pregel is only slightly less restrictive, as
the input data can be any vertex in the graph. \Grappa also supports
dynamic parallelism with asynchronous execution, but parallelism is
expressed as tasks or loop iterations, which is a far more general
programming model for irregular computation tasks.
PowerGraph~\cite{powergraph:osdi12} improves the performance of
GraphLab for real-world graphs with power-law degree distributions by
using a vertex cut for graph partitioning. Algorithmic transformations
like this  and would also improve the
performance of graph applications on \Grappa.

While the bulk-synchronous MapReduce~\cite{Dean:2008:MSD:1327452.1327492} model and
related systems such as Hadoop~\cite{hadoop} are not a good fit for
irregular or graph problems, the ideas have been extended to support a
subset of related problems that require iteration, including some
graph-based machine learning
applications. HaLoop~\cite{Bu:2010:HEI:1920841.1920881} and
Spark~\cite{Zaharia:2010:SCC:1863103.1863113} are two examples of this
sort of system. This class of systems is generally characterized by
having a restricted programming model and restricted interface to
access data, focused on streaming data off of disk. In particular,
direct communication between nodes in the cluster is not supported;
communications must happen through the provided parallel data
structures. Grappa is general enough to run these jobs, but the
MapReduce-derived systems have much more complete IO support and the
ability to tolerate from node failures.

Grappa's latency-tolerant approach to irregular and graph applications is
catching on. Colleagues at PNNL have developed GMT~\cite{GMT}, a system that
follows the same basic approach as Grappa but with more of a focus on
node-shared resources. After discussions with the Grappa team, the next version of GraphLab will also include
lightweight context switching for latency tolerance \cite{graphlab-next}.

### 2013-07-19 doc/asplos14/tasks.tex
make asplos14 directory from sosp13

--- 2013-03-28 doc/sosp13/tasks.tex
@@
 queue. This allows the scheduler to prefetch contexts into
-cache. The reason this works is because the size of the L1 cache is
+cache using software prefetch instructions. The reason this works is because the size of the L1 cache is
 sufficient to hold enough contexts to tolerate the latency to main
@@
 
+\TODO{addressing reviewer2: make it clear that yes Grappa is a low
+    level programming model, but it provides the mechanisms for
+    implementing a variety of execution models. Further, in some cases
+    like the parallel loops, we have allowed for may parallelism to
+    give flexibility to the implementation.}
 \Grappa programmers focus on expressing as much parallelism as possible

### 2013-07-19 doc/asplos14/results/histograms/histogram.R

@@ -1,0 +1,137 @@
+#!/usr/bin/env Rscript
+library(sqldf)
+library(ggplot2)
+library(scales)
+library(grid)
+#library(reshape)
+
+#source("../common.R")
+
+#d.intsort <- sqldf("select * from histograms where core == 1 and stat == 'app_bytes_sent_histogram' and value > 0", dbname="hist.intsort.db")
+# d.intsort <- sqldf("select * from histograms where core == 1 and stat == 'rdma_bytes_sent_histogram' and value > 0", dbname="hist.intsort.db")
+# d.intsort <- sqldf("select * from histograms where core == 1 and stat == 'delegate_op_latency_histogram' and value > 0", dbname="hist.intsort.db")
+d.intsort <- sqldf("select * from histograms", dbname="hist.intsort.db")
+
+#d.intsort <- sample("hist.intsort.db",1000)
+#d.intsort$benchmark <- "intsort"
+
+#d.bfs <- sqldf("select * from histograms where core % 8 == 0", dbname="hist.bfs.db")
+#d.bfs <- sample("hist.bfs.db",1000)
+# d.bfs$benchmark <- "bfs"
+
+
+hist <- function(data,title,maxx,bin) {
+  p <- ggplot(data, aes(x=value))+
+    ggtitle(title)+
+    sosp_theme+
+    geom_histogram(binwidth=bin)+
+    coord_cartesian(xlim=c(0,maxx))+
+    facet_grid(~stat,labeller=label_both)
+  return(p)
+}
+
+# plot_rdma_bytes <- function(data,title) {
+#   p <- ggplot(subset(data,stat == 'app_bytes_sent_histogram'), aes(x=value))+
+#     ggtitle(title)+
+#     sosp_theme+
+#     geom_histogram(binwidth=2000)+
+#     coord_cartesian(xlim=c(0,64000))+
+#     facet_grid(~stat,labeller=label_both)
+#   return(p)
+# }
+# 
+# plot_app_bytes <- function(data,title) {
+#   p <- ggplot(subset(data,stat == 'app_bytes_sent_histogram'), aes(x=value))+
+#     ggtitle(title)+
+#     sosp_theme+
+#     geom_histogram(binwidth=20)+
+#     coord_cartesian(xlim=c(0,5000))+
+#     facet_grid(~stat,labeller=label_both)
+#   return(p)
+# }
+# 
[…]

### 2013-07-19 doc/asplos14/asplos14-template/template.tex
Use ASPLOS14 style

\documentclass[pageno]{jpaper}

%replace XXX with the submission number you are given from the ASPLOS submission site.
\newcommand{\asplossubmissionnumber}{XXX}

\usepackage[normalem]{ulem}

\begin{document}

\title{
Instructions for Submissions to ASPLOS 2014}

\date{}
\maketitle

\thispagestyle{empty}

\begin{abstract}

This document provides instructions for submitting papers to the 19th
International Conference on Architectural Support for Programming Languages
and Operating Systems (ASPLOS), 2014.  It also provides a sample for how
papers must be formatted.
\end{abstract}

\section{Introduction}

This document provides instructions for submitting papers to the 19th
International Conference on Architectural Support for Programming Languages
and Operating Systems (ASPLOS), 2014.  Submissions that violate these
instructions may not be reviewed, at the discretion of the program chair.

An example submission (formatted using the ASPLOS'14 submission format)
that contains the submission and formatting guidelines can be downloaded
from here:
\href{http://www.cs.utah.edu/asplos14/files/asplos14-template.pdf}{Sample
PDF}. The contents of this document are the same as the contents of the
submission instructions that appear on
\href{http://www.cs.utah.edu/asplos14/submission.html}{this website}.  The
paper submission site is \href{http://asplos14.cs.illinois.edu}{here.}

All questions regarding paper formatting and submission should be directed
to the program chair.

\section{What's New?}

We specifically call attention to the following instructions/features new
to ASPLOS this year.

\begin{itemize} 
\item Text must be in 10pt ({\bf not} 9pt) font.
\item Papers must be at most 12 pages, not including references and an
optional appendix. 
\item Each reference must specify {\em all} authors (no {\em et al.}).
\item Authors may optionally suggest reviewers. 
\item Authors may optionally provide information on previous rejections.
\item Authors of {\em all} accepted papers will be required to give a
lightning presentation (about 90s) and a poster in addition to the regular
conference talk.
\item Proceedings will appear in the ACM digital library up to two weeks
before the conference.
\end{itemize}

\section{Acceptance Philosophy and Authors' Commitments}

Every effort will be made to judge a paper on its own merits. There will be
no target acceptance rate. We expect to accept a wide range of papers with
appropriate expectations for evaluation -- while papers that build on
significant past work with strong evaluations are valuable, papers that
open new areas with less rigorous evaluation are even more so.  The mantra
for this ASPLOS is: {\bf A}lways {\bf S}upport {\bf P}otentially {\bf
L}ong-term and {\bf O}ut-of-the-box {\bf S}ystems.

Accepting all worthy papers sometimes results in multitrack sessions,
reduced talk lengths, and reduced interaction. To mitigate these downsides
and in keeping with the spirit of ASPLOS as a melting pot for multiple
disciplines, we will have the following (borrowed from Micro'12).  We will
kick off the conference with a {\bf lightning presentation} session with a
(roughly) 90s presentation of each accepted paper. There will also be a {\bf
poster/demo session} where {\em all} accepted papers will be required to
participate with a poster and/or demo to facilitate closer interactions. Of
course, there will also be the regular conference talks for each paper.
Authors should be prepared to participate in all three sessions (lightning
presentation, poster/demo, and regular talk).

\section{The Paper}

\subsection{Formatting}

Submitted manuscripts should contain a maximum of 12 pages of single-spaced
two-column text, {\bf not including references and an optional
appendix}. You may include any number of pages for references and the
appendix, but see below for more instructions.  If you are using
\LaTeX~\cite{lamport94} to typeset your paper, then we suggest that you use
the template here:
\href{http://www.cs.utah.edu/asplos14/files/asplos14-latex-template.tar.gz}{\LaTeX~Template}.
(\href{http://www.cs.utah.edu/asplos14/files/asplos14-template.pdf}{This
document} was prepared with that template.)  If you are using a different
software package to typeset your paper, then please adhere to the
guidelines given in Table~\ref{table:formatting}.\footnote{One exception is
that authors may use the SIGPLAN style/class file
\href{http://classic.sigplan.org/sigplanconf.cls}{here}, but {\bf only with
the 10pt body font option (9pt will be rejected)} and modified as needed
for the requirements of the references section below.  This is marginally
different from the specified template, but will be accepted due to its
widespread use.}

\begin{table}[h!]
  \centering
  \begin{tabular}{|l|l|}
    \hline
    \textbf{Field} & \textbf{Value}\\
    \hline
    \hline
    Page limit & 12 pages, {\bf not including}\\
               & {\bf references \& appendix}\\
    \hline
    Paper size & US Letter 8.5in $\times$ 11in\\
    \hline
    Top margin & 1in\\
    \hline
    Bottom margin & 1in\\
    \hline
    Left margin & 0.75in\\
    \hline
    Right margin & 0.75in\\
    \hline
    Separation between columns & 0.25in\\
    \hline
    Body font & 10pt\\
    \hline
    Abstract font & 10pt, italicized\\
    \hline
    Section heading font & 12pt, bold\\
    \hline
    Subsection heading font & 10pt, bold\\
    \hline
    Caption font & 9pt, bold\\
    \hline
    References & 8pt\\
    \hline
  \end{tabular}
  \caption{Formatting guidelines for submission. }
  \label{table:formatting}
\end{table}

\textbf{Please ensure that you include page numbers with your
submission}. This makes it easier for the reviewers to refer to different
parts of your paper when they provide comments.

Please ensure that your submission has a banner at the top of the title
page, similar to
\href{http://www.cs.utah.edu/asplos14/files/asplos14-template.pdf}{this
one}, which contains the submission number and the notice of
confidentiality.  If using the template, just replace XXX with your
submission number.

\subsection{Content}

\noindent\textbf{\sout{Author List.}} Reviewing will be double blind;
therefore, please do not include any author names on any submitted
documents except in the space provided on the submission form.  You must
also ensure that the metadata included in the PDF does not give away the
authors. If you are improving upon your prior work, refer to your prior
work in the third person and include a full citation for the work in the
bibliography.  For example, if you are building on {\em your own} prior
work in the papers \cite{nicepaper1,nicepaper2,nicepaper3}, you would say
something like: "While the authors of
\cite{nicepaper1,nicepaper2,nicepaper3} did X, Y, and Z, this paper
additionally does W, and is therefore much better."  Do NOT omit or
anonymize references for blind review.  There is one exception to this for
your own prior work that appeared in IEEE CAL, workshops without archived
proceedings, etc.\, as discussed later in this document.

\noindent\textbf{Figures and Tables.} Ensure that the figures and tables
are legible.  Please also ensure that you refer to your figures in the main
text.  Many reviewers print the papers in gray-scale. Therefore, if you use
colors for your figures, ensure that the different colors are highly
distinguishable in gray-scale.

\noindent\textbf{References.}  There is no length limit for references.
{\bf Each reference must explicitly list all authors of the paper.  Papers
not meeting this requirement will be rejected.} Authors of NSF proposals
should be familiar with this requirement. Knowing all authors of related
work will help find the best reviewers.

\noindent\textbf{Optional Appendix.}  Authors may optionally include an
appendix (no length limit) as the last section of the manuscript; however,
{\em reviewers are not obliged to read the appendix.}  An appendix may
include proofs of theorems, more details on methodology, more results, and
anything else that can potentially answer reviewer questions.  The rest of
the manuscript may cite the appendix, but the paper should stand on its own
without the appendix.  Authors need not feel compelled to include an
appendix -- we understand the authors' time is best spent on the main
manuscript.

\section{Other Information for the Submission Site}

\subsection{Declaring Authors}

Declare all the authors of the paper upfront. Addition/removal of authors
once the paper is accepted will have to be approved by the program chair,
since it potentially undermines the goal of eliminating conflicts for
reviewer assignment.

\subsection{Areas and Topics}

ASPLOS emphasizes multidisciplinary research. Submissions should ideally
emphasize synergy of two or more ASPLOS areas: architecture, programming
languages, operating systems, and related areas (broadly
interpreted). Authors should indicate these areas on the submission form as
well as specific topics covered by the paper for optimal reviewer match. If
you are unsure whether your paper falls within the scope of ASPLOS, please
check with the program chair -- ASPLOS is a broad, multidisciplinary
conference and encourages new topics.

\subsection{Declaring Conflicts of Interest}

Authors must register all their conflicts on the paper submission site.
Conflicts are needed to ensure appropriate assignment of reviewers.  
If a paper is found to have an undeclared conflict that causes
a problem OR if a paper is found to declare false conflicts in order to
abuse or ``game'' the review system, the paper may be rejected.

Please declare a conflict of interest (COI) with the following 
for any author of your paper:

\begin{enumerate}
\item Your Ph.D. advisor, post-doctoral advisor, and Ph.D. students.
\item Family relations by blood or marriage (if they might be potential reviewers).
\item People with whom you co-authored accepted/rejected/pending papers in the last five years.
\item People with whom you co-authored accepted/pending grant proposals
  in the last five years.
\item People who shared your primary institution in the last five years. 
\end{enumerate}

``Service'' collaborations like co-authoring a CSTB report or co-presenting
tutorials do not constitute conflicts.  Co-authoring a paper that is a
compendium of various projects with no true collaboration among the
projects does not constitute a conflict among the authors of the different
projects.  On the other hand, there may be others not covered by the above
with whom you know a COI exists.  Please report such COIs; however, you
will need to justify them. Please be reasonable.  For example, just because
a reviewer works on similar topics as your paper, you cannot declare a COI
with that reviewer. We will carefully check the justification of conflicts.

We hope to draw most reviewers from the PC and the ERC, but others from the
community may also write reviews.  Please declare all your conflicts (not
just restricted to the PC and ERC).  When in doubt, contact the program
chair.

\subsection{Optional Information on Previous Submissions}

In addition to your paper, you may optionally upload your past similar
(rejected) submissions, their reviews, the author responses, and/or a
statement of how your ASPLOS'14 submission improved over the previous
submissions or addressed the previous reviews. Please use only one PDF file
for all this information and include the venue(s) where the previous work
was submitted.  Please do not include any material that will compromise
author anonymity.

Each paper will undergo a full new review process independent of whether
such a PDF file is submitted. The ASPLOS'14 reviewers will see this
document (and learn about its existence) only after they have finalized
their review.  It is, however, possible that reading this document will
influence the ASPLOS reviewers to request a change of their review or
request the PC chair to get another opinion from one of the previous
reviewers. Please consider all such implications carefully before
submitting the PDF file.

Authors are under no obligation to submit such a document. Reviewers
will be given explicit instructions not to hold the lack of such a document
against the paper in any way. Similarly, reviewers are under no obligation
to read such a document (unless a review brings up a past
submission explicitly).

\subsection{Optional Reviewer Suggestions}

Authors may optionally mark (non-conflicted) PC and ERC members that they
believe could provide expert reviews for their submission.  If authors
believe there is insufficient expertise on the PC and ERC for the topic of
their paper, they may suggest alternate reviewers.  The program chair will
use the authors' input at her discretion.  We provide this opportunity
for input mostly for papers on non-traditional and emerging topics.

\subsection{Concurrent Submissions and Workshops}

By submitting a manuscript to ASPLOS'14, the authors guarantee that the
manuscript has not been previously published or accepted for publication in
a substantially similar form in any conference, journal, or the archived
proceedings of a workshop (e.g., in the ACM digital library) -- see
exceptions below. The authors also guarantee that no paper that contains
significant overlap with the contributions of the submitted paper will be
under review for any other conference or journal or an archived proceedings
of a workshop during the ASPLOS'14 review period. Violation of any of these
conditions will lead to rejection.

The only exceptions to the above rules are for the authors' own papers in
(1) workshops without archived proceedings such as in the ACM digital
library (or where the authors chose not to have their paper appear in the
archived proceedings), or (2) venues such as IEEE CAL where there is an
explicit policy that such publication does not preclude longer conference
submissions.  In all such cases, the submitted manuscript may ignore the
above work to preserve author anonymity. This information must, however, be
provided on the submission form -- the PC chair will make this information
available to reviewers if it becomes necessary to ensure a fair review.
(This policy will be explicitly conveyed to the reviewers.)

As always, if you are in doubt, it is best to contact the program chair.

Finally, we also note that the ACM Plagiarism Policy ({\em
http://www.acm.org/publications/policies/plagiarism\_policy}) covers a
range of ethical issues concerning the misrepresentation of other works or
one's own work.

\section{Early Access in the Digital Library}

The ASPLOS'14 proceedings will be freely available via the ACM Digital
Library for up to two weeks before and up to a month after the
conference. {\bf Authors must consider any implications of this early
disclosure of their work {\em before} submitting their papers.}

\section{Acknowledgements}
Several ideas in this document and parts of the document text have been
taken from previous conferences and we thank their program chairs;
especially, Christos Kozyrakis (Micro'13), Margaret Martonosi (ISCA'13),
and Onur Mutlu (Micro'12).

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2013-07-19 doc/asplos14/overview.tex
take a stab at making the overview intro more compelling and tell our story

--- 2013-03-28 doc/sosp13/overview.tex
@@
 
-\Grappa leverages as much freely available and commodity infrastructure as
-possible. We use unmodified Linux for the operating system and an
-off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}. MPI is
-used for process setup and tear down. GASNet~\cite{gasnet} is used as the
-underlying mechanism for remote memory reads and writes using active message
-invocations. To this infrastructure, \Grappa adds  three main software components, shown in Figure~\ref{fig:grappa}:
+Commodity hardware systems provide fast processors and high-bandwidth
+interconnects but suffer when they encounter unpredictable data access
+patterns. Maximizing throughput in such a situation depends upon proper
+management of concurrency and communication. The \Grappa runtime
+components handle thousands of concurrent accesses and batch them to
+efficiently use the network. The imbalance between network bandwidth
+and core-cycles for communication-intensive codes makes supporting
+these functions in software an attractive design point. 
+\Grappa specifically employs user-mode threading and networking for
+low overhead because context switches and messages are very frequent.
+
+As \Grappa depends on concurrency for performance, its programming
+model focuses on enabling programmers to express large amounts of
+concurrency. The overall goal is for the system to provide the
+critical components for a well-performing implementation of an
+irregular application, allowing the programmer to foremost focus on
+expressing concurrency. While the programmer can initially be
+productive in a shared memory environment and rely on the
+underlying runtime components for reasonable performance, \Grappa
+provides constructs for optimizing for locality and overlapping
+communication.
+
+\Grappa comprises three main software components, shown in Figure~\ref{fig:grappa}:
 
@@
 operations.
+\TODO{Re-include the mention of GASNET,MPI spawning,and inifiniband
+    somewhere but not in the overview intro}
 

### 2013-07-19 doc/asplos14/abstract.tex

--- 2013-03-27 doc/sosp13/abstract.tex
@@
 applications. \Grappa's purpose is to enable scalable performance of irregular
-parallel applications, such as graph processing. Poor data locality,
+parallel applications, such as  branch and bound optimization and SPICE circuit simulation, and graph processing. Poor data locality,
 imbalanced parallel work and complex communication patterns make scaling these
@@
 
-\Grappa serves both as a C++ user library and as a foundation upon which
-higher level languages can be developed or adapted. \Grappa tolerates delays
-to remote memory by multiplexing thousands of lightweight workers to each
-processor core; balances load via fine-grained distributed work-stealing;
-increases communication throughput by aggregating smaller data requests into
-large ones; and provides efficient synchronization and remote operations. We
-present a detailed description of the \Grappa system and performance
-comparisons on several irregular benchmarks to hand-optimized MPI code and to
-the Cray XMT, a custom system used to target the real time graph analytics
-market.
+\Grappa serves both as a C++ user library and as a foundation for higher level
+languages. \Grappa tolerates delays to remote memory by multiplexing thousands
+of lightweight workers to each processor core; balances load via fine-grained
+distributed work-stealing; increases communication throughput by aggregating
+smaller data requests into large ones; and provides efficient synchronization
+and remote operations. We present a detailed description of the \Grappa system
+and performance comparisons on several irregular benchmarks to hand-optimized
+MPI code and to the Cray XMT, a custom system used to target the real time
+graph analytics market.

### 2013-07-19 doc/asplos14/communication.tex

--- 2013-03-28 doc/sosp13/communication.tex
@@
 packet (in terms of bytes that form the actual packet, processing time at the
-card, multiple roundtrips on the PCI Express bus and processing on the CPU
+card, multiple round-trips on the PCI Express bus and processing on the CPU
 within the driver stack). Consequently, to make the best use of the network,

### 2013-07-19 doc/asplos14/intro.tex

@@
 It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.
-The communication layer helps \Grappa achieve over 1.0 GUPS with 64 system nodes.
-When we compare \Grappa to native MPI implementations of more complex application kernels, the story is more muddled.
-When the MPI implementation is carefully tuned to aggregate messages inside the application -- essentially duplicating the work \Grappa is doing for the programmer, without the generality and context switch overhead -- then the MPI implementation is between 6-10X faster.
+The communication layer helps \Grappa achieve over 1.0 Giga-Updates per second (GUPS) with 64 system nodes.
+Comparing \Grappa and native MPI implementations for more complex kernels is more complex.
+When the MPI implementation is carefully tuned to aggregate messages in application code -- essentially duplicating the work \Grappa is doing for the programmer, without the generality and context switch overhead -- then the MPI implementation is between \checkme{6-10X} faster.
 We see this in our comparisons for breadth first search and integer sorting.
@@
 When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.
+
+ \TODO{appA and tuned compared to appB and untuned is not ideal}

### 2013-07-19 doc/asplos14/memory.tex

--- 2013-07-13 doc/sosp13/memory.tex
@@
 calls into \Grappa's API, shown in Figure~\ref{fig:accessing-memory}.
+
+\TODO{worth mentioning localized access for data-parallel stuff, along
+    with reasoning about local vs remote accesses as in the PGAS world}
 
@@
       	Memory operations at the home core of a global address \\ \\
-      	\texttt{\scriptsize cache\_acquire( GlobalAddress<T>, local\_buf, \{RO,RW,WO\})} \\
-      	\texttt{\scriptsize cache\_release( GlobalAddress<T>, local\_buf )} \\ 
-		Perform cache operations to acquire/release global data.  \\
+      	\texttt{\scriptsize buffer\_acquire( GlobalAddress<T>, local\_buf, \{RO,RW,WO\})} \\
+      	\texttt{\scriptsize buffer\_release( GlobalAddress<T>, local\_buf )} \\ 
+		Perform buffer operations to acquire/release global data.  \\
 		Acquire copies all data to local node and returns a pointer. \\ 	
-		Release optionally copies data back to global memory. \\
+		For write acquire, release copies data back to global memory. \\
 	\end{tabular}
@@
 location's home node. This method is used when there is no locality in the
-access. The other method is to do {\em explicit caching} of memory regions,
+access. The other method is to do {\em explicit buffering} of memory regions,
 used when the programmer expects that the associated computation has data
@@
 
-Delegate operations can be either {\em synchronous} or {\em asynchronous}.
-With synchronous operations, the task issuing the delegate call blocks until
+Delegate operations can be either {\em blocking} or {\em non-blocking}.
+With blocking operations, the task issuing the delegate call blocks until
 the delegate operation completes, which is necessary, for example, to ensure
 that synchronization has finished before continuing. Synchronization in \Grappa
-is implemented using synchronous delegate operations. Often remote data
-accesses do not need to be performed synchronously. In order to avoid
-unnecessary waiting, we support asynchronous delegate operations. For reads,
+is implemented using blocking delegate operations. On the other hand, remote data
+accesses often can overlap. In order to avoid
+unnecessary waiting, we support non-blocking delegate operations. For reads,
 we support a ``futures''-like mechanism. Tasks may issue delegate reads in
-parallel and block on the ``promises'' returned by asynchronous delegate
-invocations. Delegate write operations may also be performed asynchronously,
-and since they do not return data, they need a mechanism to detect completion.
+parallel and block on the ``promises'' returned by non-blocking delegate
+invocations. Delegate write operations may also be performed as
+non-blocking,
+and since they do not return data, a mechanism to detect completion is needed.
 \Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
-outstanding asynchronous operations of the local core. The code and can probe
-the state of this object to determine when all outstanding requests are
-complete. This is necessary to, say, make sure all outstanding operations are
-complete before executing a synchronization operation.
+outstanding non-blocking operations of the local core. Threads that depend on completion may block on this object
+the state of this object, and they will be woken when all outstanding requests are
+complete. 
+\TODO{augment this with our bulk futures ie global completions}
 
 \vspace{1ex}
-\textit{Explicit caching.} \Grappa provides an API to fetch a global pointer of
+\textit{Explicit buffering.} \Grappa provides an API to fetch a global pointer of
 any length and return a local pointer to a cached copy of the global memory.
-\Grappa cache operations have the usual read-only and read-write variants,
+\Grappa buffer operations have the usual read-only and read-write variants,
 along with a write-only variant used to initialize data structures. Languages
@@
 message vectorization hoists small messages out of a loop
-\cite{FortranD:1992}. Caching in \Grappa additionally provides a mechanism for
+\cite{FortranD:1992}. Buffering in \Grappa additionally provides a mechanism for
 exploiting temporal locality when available by operating on the data locally.
@@
 multiple system nodes and presenting a conventional appearing linear block of
-memory as a local pointer into a cache. The strategy employed is to issue all
+memory as a local pointer into a buffer. The strategy employed is to issue all
 the constituent requests of the desired block (as Active Messages) and then
-yield until all responses have occurred. Currently, \Grappa caches are
+yield until all responses have occurred. In \Grappa, cached regions are
 \emph{not\/} kept coherent automatically. If that is desired, the programmer is
 responsible for ensuring sufficient synchronization (via delegate operations)
-occurs before acquiring or releasing cache objects.
+occurs before acquiring or releasing buffer objects.
 
-Figure~\ref{fig:delegate-cache} depicts an example how explicit caching and
+Figure~\ref{fig:delegate-cache} depicts an example how explicit
+buffering and
 delegate operations work and how they interact.

### 2013-07-19 doc/asplos14/tasks.tex

--- 2013-03-28 doc/sosp13/tasks.tex
@@
 
+\TODO{either explain the usage of yield vs suspend in talking about
+    short and longer latency tolerance or eliminate the
+    distinction}
 \begin{figure}[htbp]
@@
 Each core in a \Grappa system has its own independent scheduler. The scheduler
-has a single FIFO queue of active workers ready to execute, the {\it
+has a collection of active workers ready to execute, the {\it
 ready worker queue}. Each scheduler also has three queues of tasks
@@
 
-\item {\it private task queue}, a FIFO queue of tasks that must run on
+\item {\it private task queue}, a queue of tasks that must run on
 this core and therefore is not subject to stealing;
 
-\item {\it public task queue},  a LIFO queue of tasks that are
+\item {\it public task queue}, a queue of tasks that are
   waiting to be matched with workers. It is a local partition of a shared
@@
 
-Whenever a task yields or suspends, the scheduler makes a decision about what
+Whenever a task yields, the scheduler makes a decision about what
 to do next. First, any task in the deadline task queue who's deadline
-constraint is met is chosen for execution. This queue manages high priority
+is imminent is chosen for execution. This queue manages high priority
 system tasks, such as periodically servicing communication requests. Second,
@@
 queue. This allows the scheduler to prefetch contexts into
-cache. The reason this works is because the size of the L1 cache is
+cache using software prefetch instructions. The reason this works is because the size of the L1 cache is
 sufficient to hold enough contexts to tolerate the latency to main
 memory. Empirically we find that prefetching the fourth worker in the scheduling order is
-sufficient. This prefetching makes context switching effectively free of cache misses even to hundreds
-of thousands of threads. We provide an analysis of our context switch
+sufficient. This prefetching constrains the types of task scheduling
+decisions that can be made but makes context switching effectively
+free of cache misses, even to hundreds
+of thousands of workers. We provide an analysis of our context switch
 performance in Section~\ref{eval:basic}.
@@
 core. Even if there are many active threads, if they are all suspended on
-long-latency operations, then the core is underutilized.
+long-latency operations, then the core is underutilized. Similarly to
+\cite{the_hierarchical_workstealing} the
+stealing policy must predict whether local tasks will likely
+generate enough new work soon.
 
@@
 
+\TODO{addressing reviewer2: Grappa provides mechanisms for many
+    programming idioms. Also, may parallelism to
+    give flexibility to the implementation.}
 \Grappa programmers focus on expressing as much parallelism as possible
@@
       Runs a limited function on a specific core without \\
-      consuming \Grappa execution resources 
+      consuming a stack or context switching
 	\end{tabular}
@@
 void search(GlobalAddress<node_t *> node, key_t key) {
+  // dereference is expected cache miss
   if (node->key == key) { 
@@
   parallel_for(0, node->numChildren, [=](int index) {
-    spawnTask([=]{
+    // first dereference of children is expected cache miss
     search(node->children[index], key); 
-    });
   });
@@
     \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:sample} Sample \Grappa code illustrating a parallel tree search similar to the unbalanced tree search benchmark we describe later.}
+      \caption{\label{fig:sample} Sample \Grappa code illustrating a
+          parallel tree search similar to the unbalanced tree search
+          benchmark we describe later. The labeled cache misses occur
+          if we assume the tree is not layed out in a regular fashion, which
+      models more general graphs.}
     \end{minipage}

### 2013-07-22 doc/asplos14/intro.tex
Rework end of intro to be more clear about how some MPI beats us (and reorder these)

@@
 First, commodity multicore processors have become extremely fast with high clock rates, large caches and robust DRAM bandwidth.
-Second, commodity networks offer high bandwidth so long as messages are large enough.
+Second, commodity networks offer high bandwidth as long as messages are large enough.
 We build on these two observations and develop \Grappa, a software runtime system that allows a commodity cluster of x86-based nodes connected via an InfiniBand network to be programmed as if it were a single, large, shared-memory NUMA (non-uniform memory access) machine with scalable performance for irregular applications.
 \Grappa exploits fast processors and the memory hierarchy to provide a lightweight user-level tasking layer that supports a context switch in as little as 38ns and can sustain a large number of active workers.
-It bridges the commodity network bandwidth gap with a communication layer that combines short messages originating from a large number of active workers into larger ones.
+It bridges the commodity network bandwidth gap with a communication layer that combines short messages originating from many concurrent workers into larger packets.
 
@@
 It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.
-The communication layer helps \Grappa achieve over 1.0 Giga-Updates per second (GUPS) with 64 system nodes.
-Comparing \Grappa and native MPI implementations for more complex kernels is more complex.
-When the MPI implementation is carefully tuned to aggregate messages in application code -- essentially duplicating the work \Grappa is doing for the programmer, without the generality and context switch overhead -- then the MPI implementation is between \checkme{6-10X} faster.
-We see this in our comparisons for breadth first search and integer sorting.
-On the other hand, if the MPI version is implemented in a straightforward manner, as we see with PageRank, GUPS, and unbalanced tree search, \Grappa is 2-5X faster.
+The communication layer helps \Grappa achieve over 1.0 Giga-Updates Per Second (GUPS) with 64 system nodes.
+Comparing \Grappa and native MPI implementations for more complex kernels is nuanced.
+Some applications use MPI in a straightforward manner, such as GUPS, PageRank, and Unbalanced Tree Search, for which \Grappa is 2-5X faster.
+On the other hand, some MPI implementations of irregular applications recognize its limitations and are written specially to aggregate messages and do some communication asynchronously, performing much the same work \Grappa does behind the scenes, but without paying the cost of generality imposed by context switches or message serialization, to name a few.
+We see this in our comparisons for Breadth First Search and Integer Sort, where the MPI implementation is \checkme{6-10X} faster.
 When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.
-
  \TODO{appA and tuned compared to appB and untuned is not ideal}

### 2013-07-22 doc/asplos14/overview.tex
Overview: Reorder DSM and Communication, and put more emphasis on delegation in DSM
(sorry about the bad diff because I moved text)
note: also removed mention of caches here

@@
 load balancing. The scheduler oversubscribes to have more worker threads than
-required for latency tolerance. By having at least four workers per processor
+required for latency tolerance. By keeping at least four workers per 
 core ready to run at all times, the scheduler can prefetch a worker's state
-into cache to lower the likelihood of costly main memory accesses during a
+into cache to reduce the chance of stalling on memory accesses during a
 context switch.
-
-\item[Distributed shared memory.] The DSM system provides fine-grain access to
-data anywhere in the system. It supports normal access operations such as
-\emph{read\/} and \emph{write\/} as well as synchronizing operations such as
-\emph{fetch-and-add\/}~\cite{fetchandadd}. It also offers explicit local
-caching of any memory in the system, and delegates that directly operate on
-remote data. The memory model offered is similar to what underpins
-C/C++~\cite{N2480,N2800}, so it is familiar to programmers. The DSM system
-design relies on the lightweight tasking system and communication layer in
-order to offer high aggregate random access bandwidth for accessing remote
-data.
 
@@
 
+\item[Distributed shared memory.] The DSM system provides fine-grain access to
+data anywhere in the system, with delegated operation at the core of its design.
+Every piece of global memory is owned by a particular core in the system, and all others may only access that memory by delegating their requests to the owning core.
+It supports normal access operations such as
+\emph{read\/} and \emph{write\/} as well as synchronizing operations such as
+\emph{fetch-and-add\/}~\cite{fetchandadd}. Due to delegation, the memory model offered is similar to what underpins
+C/C++~\cite{N2480,N2800}, so it is familiar to programmers. The DSM system
+design relies on the lightweight tasking system and communication layer in
+order to offer high aggregate random access bandwidth for accessing remote
+data.
+
 \end{description}

### 2013-07-22 doc/asplos14/tasks.tex
Use Inconsolata and fix up look and feel of listings.

@@
 \begin{scriptsize}
-\begin{verbatim}
+\begin{lstlisting}[style=grappa]
 class node_t {
@@
   int64_t numChildren;
-  GlobalAddress<node_t *> children;
+  GlobalAddress<node_t> children;
 };
 
-void search(GlobalAddress<node_t *> node, key_t key) {
-  // dereference is expected cache miss
+void search(GlobalAddress<node_t> node, key_t key, GlobalAddress<node_t> result) {
+  // deref of key is expected cache miss
   if (node->key == key) {
-    result = node;
-    return;
-  }
-  parallel_for(0, node->numChildren, [=](int index) {
-    // first dereference of children is expected cache miss
+    delegate::write(result, node);
+  } else {
+    parallel_for(0, node->numChildren,
+    [=](int index) {
+      // first deref of children is likely miss
       search(node->children[index], key); 
@@
   }
-\end{verbatim}
+}
+\end{lstlisting}
 \end{scriptsize}

### 2013-07-22 doc/asplos14/memory.tex
Fix access API, use `description` env for API's throughout

@@
   \begin{center}
-    \begin{minipage}{\columnwidth}
+    \begin{minipage}{0.95\columnwidth}
 	\small
-	\begin{tabular}{l}
-      	\texttt{\scriptsize GlobalAddress<T> global\_malloc<T>( size )} \\
-      	\texttt{\scriptsize global\_free( GlobalAddress<T> )} \\ 
-      	Allocation in the global heap \\ \\
-      	\texttt{\scriptsize delegate\_read( GlobalAddress<T>, T )}  \\
-      	\texttt{\scriptsize delegate\_read\_async( GlobalAddress<T>, T )}  \\
-      	\texttt{\scriptsize delegate\_write( GlobalAddress<T>, T value)} \\
-      	\texttt{\scriptsize delegate\_write\_async( GlobalAddress<T>, T )} \\
-      	\texttt{\scriptsize delegate\_cas( GlobalAddress<T>, T, T )} \\
-      	\texttt{\scriptsize delegate\_fetch\_inc( GlobalAddress<T>, T )} \\
-      	\texttt{\scriptsize delegate\_fetch\_inc\_async( GlobalAddress<T>, T )} \\ 
-      	Memory operations at the home core of a global address \\ \\
-      	\texttt{\scriptsize buffer\_acquire( GlobalAddress<T>, local\_buf, \{RO,RW,WO\})} \\
-      	\texttt{\scriptsize buffer\_release( GlobalAddress<T>, local\_buf )} \\ 
-		Perform buffer operations to acquire/release global data.  \\
-		Acquire copies all data to local node and returns a pointer. \\ 	
-		For write acquire, release copies data back to global memory. \\
-	\end{tabular}
+  \begin{description}
+    \item[Allocation in the global heap:] \hfill \\
+      	\lstinline[style=grappa]{GlobalAddress<T> global_malloc<T>( size )} \hfill \\
+      	\lstinline[style=grappa]{global_free( GlobalAddress<T> )}  \hfill
+    \item[Delegate operations:] \emph{(at the owning core of a global address)} \hfill \\
+      	\lstinline|T    delegate_read( GlobalAddress<T>)|  \hfill \\
+      	\lstinline|void delegate_write( GlobalAddress<T>, T value)| \hfill \\
+      	\lstinline|void delegate_write_async( GlobalAddress<T>, T value)| \hfill \\
+      	\lstinline|bool delegate_cas( GlobalAddress<T>, T cmp, T set)| \hfill \\
+      	\lstinline|T    delegate_fetch_inc( GlobalAddress<T>, T inc)| \hfill \\
+      	\lstinline|void delegate_inc_async( GlobalAddress<T>, T inc)|  \hfill\\ 
+        \lstinline|void buffer_acquire( GlobalAddress<T>, T * local_buf, {RO,RW,WO})| \hfill \\
+        \lstinline|void buffer_release( GlobalAddress<T>, T * local_buf )| \hfill
+    % Perform buffer operations to acquire/release global data.
+    % Acquire copies all data to local node and returns a pointer.
+    % For write acquire, release copies data back to global memory.
+	\end{description}
       \caption{\label{fig:accessing-memory} \Grappa API for memory accesses.}     \end{minipage}

### 2013-07-22 doc/asplos14/tasks.tex
Fix access API, use `description` env for API's throughout

@@
 stack, allocated at a particular core. When a task is ready to execute it
-is assigned to a worker, that executes the task functor on its own stack. 
+is assigned to a worker, which executes the task functor on its own stack. 
 Once a task is mapped to a worker it stays with that worker until it finishes.
@@
 worker and load that of another.
+\TODO{mention how we block on response (``scheduler activations'' or whatever) not busy wait}
 
@@
   \begin{center}
-	\begin{tabular}{l}
-    \texttt{\scriptsize yield() } \\
-      Yields core to scheduler, enqueuing caller to be \\ scheduled again soon \\
-    \texttt{\scriptsize wake( task * $t$ ) } \\
-      Enqueues some other task $t$ to be scheduled again soon \\
-    \texttt{\scriptsize suspend() }  \\
-      Yields core to scheduler, enqueuing caller only once \\ another task calls wake \\
-	\end{tabular}
   \begin{minipage}{0.95\columnwidth}  
-      \caption{\label{fig:scheduling} \Grappa scheduling API } 
+  \begin{description}
+    \item[\texttt{yield() }] \hfill \\
+      Yields core to scheduler, enqueuing caller to be scheduled again soon
+    \item[\texttt{wake( Worker * $w$ ) }] \hfill \\
+      Enqueues some other worker $w$ to be scheduled again soon
+    \item[\texttt{suspend() } ] \hfill \\
+      Yields core to scheduler, enqueuing caller only once another task calls wake
+	\end{description}
+  \end{minipage}
+    \begin{minipage}{0.95\columnwidth}
+      \caption{\label{fig:scheduling} \Grappa scheduling API
+      \TODO{Drop this? Doesn't seem non-obvious or important.}}
     \end{minipage}
@@
   \begin{center}
-	\begin{tabular}{l}
-    \texttt{\scriptsize spawn( Functor f )} \\
-      Creates a new stealable task \\
-    \texttt{\scriptsize parallel\_for( start, end, Functor iteration )} \\
-      Executes iterations of a loop as stealable tasks that \\
-      take the iteration index as an argument  \\
-    \texttt{\scriptsize call\_on( core, Functor f )} \\ 
-      Runs a limited function on a specific core without \\
+	\begin{description}
+    \item[\texttt{spawn( Functor f )}] \hfill \\
+      Creates a new stealable task
+    \item[\texttt{parallel\_for( start, end, Functor iteration )}] \hfill \\
+      Executes iterations of a loop as stealable tasks that
+      take the iteration index as an argument
+    \item[\texttt{call\_on( core, Functor f )}] \hfill \\
+      Runs a limited function on a specific core without
       consuming a stack or context switching
-	\end{tabular}
+	\end{description}
     \begin{minipage}{0.95\columnwidth}

### 2013-07-22 doc/asplos14/memory.tex
add delegate_read_async back (with Promise<T>)

@@
       	\lstinline|T    delegate_read( GlobalAddress<T>)|  \hfill \\
+      	\lstinline|Promise<T> delegate_read_async( GlobalAddress<T> )|  \hfill \\
       	\lstinline|void delegate_write( GlobalAddress<T>, T value)| \hfill \\
@@
       	\lstinline|void delegate_inc_async( GlobalAddress<T>, T inc)|  \hfill\\ 
-        \lstinline|void buffer_acquire( GlobalAddress<T>, T * local_buf, {RO,RW,WO})| \hfill \\
-        \lstinline|void buffer_release( GlobalAddress<T>, T * local_buf )| \hfill
+        \lstinline|void buffer_acquire( GlobalAddress<T>, local_buf, {RO,RW,WO})| \hfill \\
+        \lstinline|void buffer_release( GlobalAddress<T>, local_buf )| \hfill
     % Perform buffer operations to acquire/release global data.

### 2013-07-22 doc/asplos14/memory.tex
some minor fixes to Memory, plus added citation for 'linearizability'

@@
 global. Local memory is local to a single core within a node in the system.
-Accesses occur through conventional pointers. The compiler emits an access and
-the memory is manipulated directly. Applications use local accesses for a
+Accesses occur through conventional pointers. Applications use local accesses for a
 number of things in \Grappa: the stack associated with a task, accesses to
@@
 lower 48 bits of the address hold a virtual address in the process. The top
-bit is set to indicates that the reference is a 2D address (as opposed to
+bit is set to indicate that the reference is a 2D address (as opposed to
 linear address). This leaves 15 bits for network endpoint ID, which limits our
 scalability to $2^{15}$ endpoints. Any node-local data can be made accessible
-by other nodes in the system by wrapping the address and node ID into a 2D
+to other cores in the system by wrapping the address and node ID into a 2D
 global address. This address can then be accessed with a delegate operation
-and even be cached by other nodes. The address is converted at the destination
+and even be buffered by other cores. The address is converted at the destination
 into a canonical x86 address by replacing the upper bits with the
@@
 and since they do not return data, a mechanism to detect completion is needed.
-\Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
-outstanding non-blocking operations of the local core. Threads that depend on completion may block on this object
-the state of this object, and they will be woken when all outstanding requests are
-complete. 
+\Grappa provides a \texttt{GlobalCompletionEvent} synchronization object, which non-blocking operations (including tasks) can be enrolled with and which other tasks can block on to be notified and woken when all enrolled operations in the collection are complete.
 \TODO{augment this with our bulk futures ie global completions}
@@
 \vspace{1ex}
-\textit{Explicit buffering.} \Grappa provides an API to fetch a global pointer of
+\textit{Explicit buffering.}
+\TODO{de-emphasize buffering? requires rewording above sections to talk about Delegate ops as the primary form of use, and then we can mention this functionality offhand}
+\Grappa provides an API to fetch a global pointer of
 any length and return a local pointer to a cached copy of the global memory.
@@
 the constituent requests of the desired block (as Active Messages) and then
-yield until all responses have occurred. In \Grappa, cached regions are
+block until all responses have occurred. In \Grappa, buffered regions are
 \emph{not\/} kept coherent automatically. If that is desired, the programmer is
 responsible for ensuring sufficient synchronization (via delegate operations)
-occurs before acquiring or releasing buffer objects.
+occurs before acquiring or releasing buffers.
 
-Figure~\ref{fig:delegate-cache} depicts an example how explicit
+Figure~\ref{fig:delegate-cache} depicts an example of how explicit
 buffering and
@@
 \begin{minipage}{1.9\columnwidth} \caption{\label{fig:delegate-cache}
+\TODO{Redo this figure to simplify it and focus on delegates and latency tolerance.}
 \textbf{Delegation and cache example:} A core allocates an array in the global
@@
 synchronization operations are done via delegate operations. Since they all
-execute in their home node, they are guaranteed to be serialized, with their
-updates visible to all cores across the system in the same order. Also, tasks
-are restricted to a single outstanding synchronous delegate operation at a
-time, which is how synchronization is done. Therefore, synchronization operations are not subject to potential reordering. 
+execute on their home core in some serial order, they are guaranteed to be globally linearizable~\cite{herlihy1990linearizability}, with their
+updates visible to all cores across the system in the same order.
+In addition, only one synchronous delegate will be in flight at a time from a particular task.
+Therefore, synchronization operations are not subject to potential reordering.
 % \TODO{I think we can support multiple delegates in parallel from a task as
@@
 
-Note, however, that if the application code uses explicit caching or
-asynchronous delegates to access shared data, all updates must be sent back to
-the home node before the synchronization operation that protects the data is
+Note, however, that if the application code uses explicit buffers or
+asynchronous delegates to access shared data, all updates must be published back to
+the home core before the synchronization operation that protects the data is
 performed. This is done using release operations on cached regions and using
 the \texttt{GlobalCompletionEvent} object to determine that asynchronous
-delegates have been fully executed.
+delegates have completed.
 

### 2013-07-22 doc/asplos14/relatedwork.tex
change vocie of related work

@@
 The goal of providing a shared memory abstraction for a distributed
-memory system goes back nearly 30 years. Here I present a few of the
-ideas explored in this area.
-
-Much of the innovation in SDSM has occurred around reducing the
+memory system goes back nearly 30 years. Much of the innovation in SDSM has occurred around reducing the
 synchronization cost of doing updates. The first DSM systems,
@@
 Partitioned Global Address Space (PGAS)
-model. Split-C~\cite{Krishnamurthy:1993:PPS:169627.169724} is the
-earliest example I know of this style of language; contemporary
-examples include Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005},
+model. Example include Split-C~\cite{Krishnamurthy:1993:PPS:169627.169724},
+Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005},
 Co-array Fortran~\cite{Numrich:1998:CFP:289918.289920} and
@@
 ability to tolerate from node failures.
-
-Grappa's latency-tolerant approach to irregular and graph applications is
-catching on. Colleagues at PNNL have developed GMT~\cite{GMT}, a system that
-follows the same basic approach as Grappa but with more of a focus on
-node-shared resources. After discussions with the Grappa team, the next version of GraphLab will also include
-lightweight context switching for latency tolerance \cite{graphlab-next}.

### 2013-07-22 doc/asplos14/abstract.tex
remove caching and clean up some wording

@@
 applications. \Grappa's purpose is to enable scalable performance of irregular
-parallel applications, such as  branch and bound optimization and SPICE circuit simulation, and graph processing. Poor data locality,
+parallel applications, such as  branch and bound optimization, SPICE circuit simulation, and graph processing. Poor data locality,
 imbalanced parallel work and complex communication patterns make scaling these

### 2013-07-22 doc/asplos14/intro.tex
remove caching and clean up some wording

@@
 the business, national security, and social network computing domains. For
-these emerging applications, reasonable response time -- given the sheer
-amount of data -- requires large multinode systems. The most broadly available
+these emerging applications, reasonable response time---given the sheer
+amount of data---requires large multinode systems. The most broadly available
 multinode systems are those built from x86 compute nodes interconnected via
-ethernet or InfiniBand. However, scalable performance of irregular
+Ethernet or InfiniBand. However, scalable performance of irregular
 applications on these systems is elusive for two reasons:
@@
 \vspace{0.5ex} \noindent{\bf High network communication latency.} The performance challenges of frequent communication are exacerbated by high network latency relative to processor performance.
-Latency of commodity networks runs anywhere from a few to hundreds of microseconds -- ten's of thousands of processor clock cycles.
+Latency of commodity networks runs anywhere from a few to hundreds of microseconds---tens of thousands of processor clock cycles.
 Since irregular application tasks encounter remote references dynamically during execution and must resolve them before making further progress, stalls are frequent and lead to severely underutilized compute resources.
@@
 
-The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of
+The fully custom Tera MTA-2~\cite{tera:mta1,Alverson:1992:EHP:143369.143408} system is a classic example of
 supporting irregular applications by using concurrency to hide latencies. It
@@
 matched the processor clock frequency and sufficient bandwidth to sustain a
-reference from every processor on every clock cycle. Unfortunately, the MTA's relatively low single-threaded performance meant that it
+reference from every processor on every clock cycle. Unfortunately, the MTA-2's relatively low single-threaded performance meant that it
 was not general enough nor cost-effective. The Cray XMT approximates the Tera
@@
 It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.
-The communication layer helps \Grappa achieve over 1.0 Giga-Updates Per Second (GUPS) with 64 system nodes.
-Comparing \Grappa and native MPI implementations for more complex kernels is nuanced.
+The communication layer helps \Grappa achieve over 1 billion random updates per second across 64 system nodes.
+\TODO{update with new performance numbers}.
+% Comparing \Grappa and native MPI implementations for more complex kernels is nuanced.
 Some applications use MPI in a straightforward manner, such as GUPS, PageRank, and Unbalanced Tree Search, for which \Grappa is 2-5X faster.
@@
 We see this in our comparisons for Breadth First Search and Integer Sort, where the MPI implementation is \checkme{6-10X} faster.
-When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.
- \TODO{appA and tuned compared to appB and untuned is not ideal}
+
+%When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.
+% \TODO{appA and tuned compared to appB and untuned is not ideal}

### 2013-07-22 doc/asplos14/memory.tex
remove caching and clean up some wording

@@
       	\lstinline[style=grappa]{global_free( GlobalAddress<T> )}  \hfill
-    \item[Delegate operations:] \emph{(at the owning core of a global address)} \hfill \\
+    \item[Delegate operations:] \hfill \\
       	\lstinline|T    delegate_read( GlobalAddress<T>)|  \hfill \\
@@
       	\lstinline|void delegate_inc_async( GlobalAddress<T>, T inc)|  \hfill\\ 
-        \lstinline|void buffer_acquire( GlobalAddress<T>, local_buf, {RO,RW,WO})| \hfill \\
-        \lstinline|void buffer_release( GlobalAddress<T>, local_buf )| \hfill
+%        \lstinline|void buffer_acquire( GlobalAddress<T>, local_buf, {RO,RW,WO})| \hfill \\
+%        \lstinline|void buffer_release( GlobalAddress<T>, local_buf )| \hfill
     % Perform buffer operations to acquire/release global data.
@@
 
-\paragraph{Global memory access} There are two general methods \Grappa
-applications use to {\em access} global memory. One method is to use {\em
+\paragraph{Global memory access} 
+Access to \Grappa's distributed shared memory is provided through  {\em
 delegate} operations, which are short memory accesses performed at the memory
-location's home node. This method is used when there is no locality in the
-access. The other method is to do {\em explicit buffering} of memory regions,
-used when the programmer expects that the associated computation has data
-locality. Below we explain each in detail.
-
-\vspace{1ex} \textit{Delegate operations.} When the data access pattern has
+location's home node. When the data access pattern has
 low-locality, it is more efficient to modify the data on its home core rather
@@
 on individual machine-word sized chunks of global memory to the memory system
-itself. Delegates are general, so we use them to perform simple
+itself. Delegates can execute arbitrary non-blocking code, so we use them to perform simple
 \emph{read\/}/\emph{write\/} operations to global memory, as well as more complex \emph{read-modify-write\/} operations (e.g., \emph{fetch-and-add\/}).
@@
 
-\vspace{1ex}
-\textit{Explicit buffering.}
-\TODO{de-emphasize buffering? requires rewording above sections to talk about Delegate ops as the primary form of use, and then we can mention this functionality offhand}
-\Grappa provides an API to fetch a global pointer of
-any length and return a local pointer to a cached copy of the global memory.
-\Grappa buffer operations have the usual read-only and read-write variants,
-along with a write-only variant used to initialize data structures. Languages
-for distributed shared memory systems have done optimizations to achieve a
-similar goal. For example, the UPC compiler coalesces struct and array
-accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's
-message vectorization hoists small messages out of a loop
-\cite{FortranD:1992}. Buffering in \Grappa additionally provides a mechanism for
-exploiting temporal locality when available by operating on the data locally.
-
-Under the hood, \Grappa performs the mechanics of gathering chunks of data from
-multiple system nodes and presenting a conventional appearing linear block of
-memory as a local pointer into a buffer. The strategy employed is to issue all
-the constituent requests of the desired block (as Active Messages) and then
-block until all responses have occurred. In \Grappa, buffered regions are
-\emph{not\/} kept coherent automatically. If that is desired, the programmer is
-responsible for ensuring sufficient synchronization (via delegate operations)
-occurs before acquiring or releasing buffers.
-
-Figure~\ref{fig:delegate-cache} depicts an example of how explicit
-buffering and
-delegate operations work and how they interact.
-
-\begin{figure*}[htb] \begin{center}
-\includegraphics[width=1.5\columnwidth]{figs/delegate-cache}
-\begin{minipage}{1.9\columnwidth} \caption{\label{fig:delegate-cache}
-\TODO{Redo this figure to simplify it and focus on delegates and latency tolerance.}
-\textbf{Delegation and cache example:} A core allocates an array in the global
-heap (1). It then spawns two tasks on remote cores to increment elements of
-the array. The first task increments two elements of the array using cache
-operations. The first task is then invoked (2). A cache request is issued for two
-adjacent integers starting at the second element of the array. Since these
-element are stored in the memories of two different cores, this requires two messages to be sent (3). The task is suspended until both responses arrive
-(4). The data carried in these responses is stored in the local buffer. The
-elements are then incremented in the buffer. The modified data is then sent back
-to the home node (5). Acknowledgements are returned (6) so the task
-knows when the writes are complete. The second task increments an element of
-the array with a delegate operation. The task is invoked (7). A delegate
-request is sent to the home core of the array element with the increment
-value. The task suspends until the response is received. The increment
-is executed on the remote core (8). A response is returned (9) with the
-previous value of the array element. } \end{minipage} \vspace{-3ex}
-\end{center} \end{figure*}
-
-\TODO{eliminate the word cache from the paper since the main point is
-    contiguity and not unpredictable temporal reuse. We
-    must make it clear that this is not user-level caching.}
+When programmers want to operate on data structures spread across
+multiple nodes, accesses must be expressed as multiple delegate
+operations along with with appropriate synchronization
+operations. \Grappa's API also includes calls for gathering and scattering
+contiguous blocks in the global heap, but the user is responsible for
+ensuring correct synchronization.
 
@@
 In addition, only one synchronous delegate will be in flight at a time from a particular task.
-Therefore, synchronization operations are not subject to potential reordering.
+Therefore, synchronization operations from a particular task are not subject to reordering.
 % \TODO{I think we can support multiple delegates in parallel from a task as

### 2013-07-22 doc/asplos14/overview.tex
remove caching and clean up some wording

@@
 
-Commodity hardware systems provide fast processors and high-bandwidth
+Commodity clusters provide fast processors and high-bandwidth
 interconnects but suffer when they encounter unpredictable data access
-patterns. Maximizing throughput in such a situation depends upon proper
-management of concurrency and communication. The \Grappa runtime
-components handle thousands of concurrent accesses and batch them to
-efficiently use the network. The imbalance between network bandwidth
-and core-cycles for communication-intensive codes makes supporting
-these functions in software an attractive design point. 
-\Grappa specifically employs user-mode threading and networking for
-low overhead because context switches and messages are very frequent.
+patterns. Maximizing throughput in such a situation depends upon
+proper management of concurrency and communication. \Grappa's design is
+motivated by four relationsips we observe within commodity clusters.
+
+First, cores in our cluster can execte hundreds of thousands of
+instructions in the few microseconds that it takes for one RDMA read
+to complete. Given the high degree of fine-grained random access in
+irregular applications, this pushes us to overlap communication with
+other computation.
+
+Second, cores can perfom random access to local DRAM over two orders
+of magnitude faster than they can issue RDMA requests through the
+network. In addition, cores can execute instructions three orders of
+magnitude faster than they can issue RDMA requests. On the other hand,
+per-node network bandwidth (measured in bytes rather than messages) is
+within an order of magnitude of local DRAM bandwidth. The combination
+of these facts motivates us to delay and aggregate remote operations
+to run on remote cores.
 
@@
 model focuses on enabling programmers to express large amounts of
-concurrency. The overall goal is for the system to provide the
-critical components for a well-performing implementation of an
-irregular application, allowing the programmer to foremost focus on
-expressing concurrency. While the programmer can initially be
-productive in a shared memory environment and rely on the
-underlying runtime components for reasonable performance, \Grappa
-provides constructs for optimizing for locality and overlapping
-communication.
+concurrency. \Grappa is intended to support what we think is a
+desriable workflow: programmers start by expressing a computation
+using a multithreaded, shared-memory model on a single node. \Grappa
+then provides the primitves to do a straightforward translation of
+this code to run on a distributed-memory cluster, through constructs
+for optimizing locality and overlapping communication.
 

### 2013-07-22 doc/asplos14/tasks.tex
remove caching and clean up some wording

@@
 \paragraph{Tasks} 
-Tasks are specified by a ``functor'' object that holds both code to execute and initial state.
-The functor can be specified with a function pointer and explicit arguments, a C++ struct that overloads the parentheses operator, or a C++11 lambda.
+Tasks are specified by a closure (or ``function object'' in C++ parlance) that holds both code to execute and initial state.
+The functor can be specified with a function pointer and explicit arguments, a C++ struct that overloads the parentheses operator, or a C++11 lambda construct.
 These objects, typically very small (on the order of 64 bytes), hold read-only values such as an iteration index and pointers to common data or synchronization objects.
@@
 remain busy while waiting for the operation to complete. In addition, a
-programmer can direct scheduling explicitly via the \Grappa API calls shown in
-Figure~\ref{fig:scheduling}. To minimize yield overhead, the \Grappa scheduler
+programmer can direct scheduling explicitly.
+% via the \Grappa API calls shown in
+%Figure~\ref{fig:scheduling}. 
+To minimize context-switch overhead, the \Grappa scheduler
 operates entirely in user-space and does little more than store state of one
@@
     distinction}
-\begin{figure}[htbp]
-  \begin{center}
-  \begin{minipage}{0.95\columnwidth}  
-  \begin{description}
-    \item[\texttt{yield() }] \hfill \\
-      Yields core to scheduler, enqueuing caller to be scheduled again soon
-    \item[\texttt{wake( Worker * $w$ ) }] \hfill \\
-      Enqueues some other worker $w$ to be scheduled again soon
-    \item[\texttt{suspend() } ] \hfill \\
-      Yields core to scheduler, enqueuing caller only once another task calls wake
-	\end{description}
-  \end{minipage}
-    \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:scheduling} \Grappa scheduling API
-      \TODO{Drop this? Doesn't seem non-obvious or important.}}
-    \end{minipage}
-  \end{center}
-\end{figure}
+%% \begin{figure}[htbp]
+%%   \begin{center}
+%%   \begin{minipage}{0.95\columnwidth}  
+%%   \begin{description}
+%%     \item[\texttt{yield() }] \hfill \\
+%%       Yields core to scheduler, enqueuing caller to be scheduled again soon
+%%     \item[\texttt{wake( Worker * $w$ ) }] \hfill \\
+%%       Enqueues some other worker $w$ to be scheduled again soon
+%%     \item[\texttt{suspend() } ] \hfill \\
+%%       Yields core to scheduler, enqueuing caller only once another task calls wake
+%% 	\end{description}
+%%   \end{minipage}
+%%     \begin{minipage}{0.95\columnwidth}
+%%       \caption{\label{fig:scheduling} \Grappa scheduling API
+%%       \TODO{Drop this? Doesn't seem non-obvious or important.}}
+%%     \end{minipage}
+%%   \end{center}
+%% \end{figure}
 
 Each core in a \Grappa system has its own independent scheduler. The scheduler
-has a collection of active workers ready to execute, the {\it
+has a collection of active workers ready to execute called the {\it
 ready worker queue}. Each scheduler also has three queues of tasks
@@
 Whenever a task yields, the scheduler makes a decision about what
-to do next. First, any task in the deadline task queue who's deadline
+to do next. First, any task in the deadline task queue whose deadline
 is imminent is chosen for execution. This queue manages high priority
@@
 the scheduler determines if any workers with running tasks are ready to
-execute; if so, one is scheduled. Finally, if there are no workers ready to
-run, but there are tasks waiting to be matched with workers, an idle worker is
+execute; if so, one is scheduled. Finally, if no workers are ready to
+run, but tasks are waiting to be matched with workers, an idle worker is
 woken (or a new worker is spawned), matched with a task, and scheduled.
@@
 will not fit in cache. By oversubscribing on the number of workers
-beyond what is required for latency tolerance, the scheduler can
+beyond what is required for local DRAM latency tolerance, the scheduler can
 ensure there is always some number of context pointers in the ready
 queue. This allows the scheduler to prefetch contexts into
-cache using software prefetch instructions. The reason this works is because the size of the L1 cache is
+cache using software prefetch instructions; the size of the L1 cache is
 sufficient to hold enough contexts to tolerate the latency to main
@@
 tasks, stopping to execute the loop body when the number of iterations is
-below the required threshold. Third, a programmer may want to execute an active message; that is, to run a
-small piece of code on a particular core in the system without waiting for
-execution resources to be available. Custom synchronization primitives, for example, execute this way as a function executed on the core where the data
+below the required threshold. Third, a programmer may want to execute an active message~\cite{vonEicken:1992:AMM:139669.140382}; that is, to run a
+small non-blocking piece of code on a particular core in the system without waiting for
+ a free worker to become available. For example, custom synchronization primitives are implemented as an active message executed on the core where the data
 lives. \Grappa provides the \texttt{call\_on} call for this purpose.

### 2013-07-23 doc/asplos14/overview.tex
edits mostly to overview

--- 2013-07-22 doc/asplos14/overview.tex
@@
 
-Commodity hardware systems provide fast processors and high-bandwidth
-interconnects but suffer when they encounter unpredictable data access
-patterns. Maximizing throughput in such a situation depends upon proper
-management of concurrency and communication. The \Grappa runtime
-components handle thousands of concurrent accesses and batch them to
-efficiently use the network. The imbalance between network bandwidth
-and core-cycles for communication-intensive codes makes supporting
-these functions in software an attractive design point. 
-\Grappa specifically employs user-mode threading and networking for
-low overhead because context switches and messages are very frequent.
+% Commodity clusters provide fast processors and high-bandwidth
+% interconnects but suffer when they encounter unpredictable data access
+% patterns. Maximizing throughput in such a situation depends upon
+% proper management of concurrency and communication.
 
-As \Grappa depends on concurrency for performance, its programming
-model focuses on enabling programmers to express large amounts of
-concurrency. The overall goal is for the system to provide the
-critical components for a well-performing implementation of an
-irregular application, allowing the programmer to foremost focus on
-expressing concurrency. While the programmer can initially be
-productive in a shared memory environment and rely on the
-underlying runtime components for reasonable performance, \Grappa
-provides constructs for optimizing for locality and overlapping
-communication.
+\Grappa's design is motivated by two key observations of today's systems.
+First, modern compute cores can execute hundreds of thousands of instructions
+in the few microseconds that it takes for one RDMA read to complete. Second,
+the injection rate of modern commodity networks is low and the bandwidth is
+optimized for large messages. Given the frequent communication caused by high
+degree of fine-grained random access in irregular applications, it is
+necessary to both overlap communication with computation as much as possible
+and to aggregate network messages efficiently. Providing these two properties
+depend on exploiting large amounts of concurrency. Hence, \Grappa's
+programming model focuses on enabling programmers to easily express
+concurrency.
+
+% \Grappa is intended to support what we
+% think is a desirable workflow: programmers start by expressing a computation
+% using a multithreaded, shared-memory model on a single node. \Grappa then
+% provides the primitives to do a straightforward translation of this code to
+% run on a distributed-memory cluster, through constructs for optimizing
+% locality and overlapping communication.
 
@@
 \begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:grappa} \Grappa system overview}
+  \caption{\label{fig:grappa} \Grappa system overview\TODO{Cora->Manta}}
 \end{minipage}

### 2013-07-23 doc/asplos14/overview.tex
added text to communication item on details

@@
 \item[Communication layer.] The main goal of our communication layer is to
-aggregate small messages into large ones.  This process is invisible to the application
-programmer. Its interface is based on active messages~\cite{vonEicken92}.
-Since aggregation and deaggregation of messages needs to be very efficient, we perform the process in parallel and carefully use lock-free synchronization
-operations.
-\TODO{Re-include the mention of GASNET,MPI spawning,and inifiniband
-    somewhere but not in the overview intro}
+aggregate small messages into large ones. This process is invisible to the
+application programmer. Its interface is based on active
+messages~\cite{vonEicken92}. Since aggregation and deaggregation of messages
+needs to be very efficient, we perform the process in parallel and carefully
+use lock-free synchronization operations.
+
+GASNet~\cite{gasnet} is used as the
+underlying mechanism for remote memory reads and writes using active message
+invocations, with an off-the-shelf user-mode InfiniBand device driver
+stack~\cite{OFED}. MPI is used for process setup and tear down.
 
@@
 \end{description}
+
+% \Grappa leverages as much freely available and commodity infrastructure as
+% possible. We use unmodified Linux for the operating system and an
+% off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}. MPI is
+% used for process setup and tear down. GASNet~\cite{gasnet} is used as the
+% underlying mechanism for remote memory reads and writes using active message
+% invocations.

### 2013-07-23 doc/asplos14/relatedwork.tex
pass on related work

@@
 
-Grappa's closest ancestor is the Threaded Abstract
+\Grappa's closest ancestor is the Threaded Abstract
 Machine~\cite{CullerGSvE93}. This was a software runtime system
@@
 modern commodity processors have sufficient bandwidth and prefetch
-capacity to stream contexts from DRAM.
+capacity to stream contexts from DRAM and sustain a very large number of active contexts.
 
@@
 synchronization cost of doing updates. The first DSM systems,
-including IVY~\cite{Li:1989:MCS:75104.75105}, used invalidations to
+including IVY~\cite{Li:1989:MCS:75104.75105}, used frequent invalidations to
 provide sequential consistency, which imposed significant
@@
 exploited both of these ideas, but some coherence overhead was still
-required. In contrast, Grappa's delegate-based approach to updates
+required. In contrast, \Grappa's delegate-based approach to updates
 avoids synchronization overhead entirely, providing sequential
 consistency for data-race-free programs without the cost of a
-coherence protocol. This cost of this communication is mitigated through latency
-tolerance.
+full coherence protocol. 
+% This cost of this communication is mitigated through latency
+% tolerance.
 
@@
 transparently. Unfortunately, this meant that these systems depended
-on lots of locality to amortize the cost of moving these large
+on a fair ammount of locality to amortize the cost of moving these large
 blocks, and the systems were very susceptible to false sharing. Other
@@
 ownership with variable granularity to address these problems.
-Grappa's delegate-based approach is similar; since updates are always
+\Grappa's delegate-based approach is similar; since updates are always
 performed at data's home node, only modified data needs to be moved.
@@
 infrastructure (Munin). This made these systems difficult to port to
-new platforms. Grappa follows the lead of TreadMarks and provides
+new platforms. \Grappa follows the lead of TreadMarks and provides
 SDSM entirely at user-level through a library and runtime.
@@
 we accept the random access aspect of irregular applications and optimize for
-throughput rather than low latency. Our DSM system must support enough memory
+throughput rather than low latency. Our DSM system must support enough 
 concurrency to tolerate the latency of the network and additional latency
@@
 Co-array Fortran~\cite{Numrich:1998:CFP:289918.289920} and
-UPC~\cite{upc:2005}. Grappa shares many parts of its design philosophy
+UPC~\cite{upc:2005}. \Grappa shares many parts of its design philosophy
 with these languages.
 
-There are two key ideas Grappa draws from PGAS languages. First, PGAS
+There are two key ideas \Grappa draws from PGAS languages. First, PGAS
 languages implement DSM at the language, rather than the system
@@
 split-phase access, variable data granularity, and support for
-user-customizable synchronization operations. Grappa follows this
-approach for this reason.
+user-customizable synchronization operations. \Grappa follows this
+approach for the same reason.
 
-Second, in PGAS languages, each piece of data has a single canonical
-location on a particular node. The language designers expect
-programmers to modify algorithms to take advantage of locality by
-processing node-local data as much as possible; access to data stored
-on other nodes is possible but is seen as something to avoid if
-possible. Grappa is designed to support the opposite view: Grappa is
-optimized for random access to data anywhere in the cluster, and
-locality can be exploited when it is available.
+Second, in PGAS languages, each piece of data has a single canonical location
+on a particular node. \Grappa takes the same approach. However, PGAS languages
+often expect programmers to modify algorithms to take advantage of locality by
+processing node-local data as much as possible; access to data stored on other
+nodes is possible but is seen as something to avoid if possible. \Grappa takes
+the opposite view, as it optimizes for random access to data anywhere in the
+cluster, and exploiting any available locality in the application is a
+secondary concern.
 
@@
 while allowing programmers to direct individual cores when
-necessary. Grappa follows the same approach, providing a
+necessary. \Grappa follows the same approach as Chapel, providing a
 single-machine abstraction to the programmer along with support for
@@
 workloads that could prioritize vertices. GraphLab overcomes this
-limitation with an execution mode that schedules vertex computations
+limitation with an execution model that schedules vertex computations
 individually, allowing prioritization, which gives faster convergence
 in a variety of iterative algorithms. GraphLab, however, imposes a
-rigid computation model where programmers must express computation as
+rigid model that requires programmers to express computation as
 transformations on a vertex and its edge list, with information only
@@
 having a restricted programming model and restricted interface to
-access data, focused on streaming data off of disk. In particular,
+access data, focused on streaming data from disk. In particular,
 direct communication between nodes in the cluster is not supported;
-communications must happen through the provided parallel data
-structures. Grappa is general enough to run these jobs, but the
+communications must happen through the  parallel data
+structures provided. \Grappa is general enough to run these jobs, but the
 MapReduce-derived systems have much more complete IO support and the

### 2013-07-23 doc/asplos14/abstract.tex
many small fixes

@@
 languages. \Grappa tolerates delays to remote memory by multiplexing thousands
-of lightweight workers to each processor core; balances load via fine-grained
-distributed work-stealing; increases communication throughput by aggregating
-smaller data requests into large ones; and provides efficient synchronization
+of lightweight workers to each processor core, balances load via fine-grained
+distributed work-stealing, increases communication throughput by aggregating
+smaller data requests into large ones, and provides efficient synchronization
 and remote operations. We present a detailed description of the \Grappa system
 and performance comparisons on several irregular benchmarks to hand-optimized
-MPI code and to the Cray XMT, a custom system used to target the real time
-graph analytics market.
+MPI code and to the Cray XMT, a custom system used to target the real-time
+graph-analytics market.

### 2013-07-23 doc/asplos14/communication.tex
remove caching and clean up some wording
many small fixes

@@
 \paragraph{Message aggregation} In our experiments the vast majority of upper
-layer message requests are smaller than 44 bytes. Our measurements confirm
+layer message requests are smaller than 44~bytes. Our measurements confirm
 manufacturers' published data [15]; with 44-byte packets, the available
@@
 achieves their peak bisection bandwidth only when the packet sizes are
-relatively large --- on the order of multiple kilobytes. The reason for this
+relatively large -- on the order of multiple kilobytes. The reason for this
 discrepancy is the combination of overheads associated with handling each
@@
 in \Grappa, aggregating and sending messages efficiently is very important. To
-achieve that, \Grappa makes careful use of caches, prefetching and lock-free
+achieve that, \Grappa makes careful use of caches, prefetching, and lock-free
 synchronization operations.
 
-Each processing core of a system node maintains an array of outgoing message lists.
-The array size is the number of system nodes in the \Grappa system.
-The outgoing message lists and messages are located in a region of memory shared across all cores in a \Grappa node (thus enabling cores to peek at each other's message lists).
-When a task sends a message, it allocates a buffer (typically on its stack), determines the destination system node, and links the buffer into the corresponding linked list.
+Each processing core of a system node maintains an array of outgoing
+message lists.  The array size is the number of system cores in the
+\Grappa system.  The outgoing message lists and messages are located
+in a region of memory shared across all cores in a \Grappa node (thus
+enabling cores to peek at each other's message lists). When a task
+sends a message, it allocates a buffer (typically on its stack),
+determines the destination system node, and links the buffer into the
+corresponding linked list.
 
-Each processing core in a given system node is responsible for aggregating and sending the resulting messages from all processors on that node to a set of destination nodes.
-Each core periodically executes a task responsible for sending messages.
-This task examines the private (to each core) message lists for each destination node it is responsible for managing and if the list is long enough, or a message has waited past a time-out period, all messages to a given destination system node from that source system node are sent.
-Aggregating and sending a message involves manipulating a set of shared data-structures (the message lists).
-This is done using CAS (compare-and-swap) operations to avoid high synchronization costs.
-Note that the reason we use a per-processor array of message lists that is only periodically modified across processor cores is we empirically determined this approach was faster (sometimes significantly) than a global per-system node array of message lists.
+Each processing core in a given system node is responsible for
+aggregating and sending the resulting messages from all processors on
+that node to a set of destination nodes.  Each core periodically
+executes a task responsible for sending messages.  This task examines
+the private (to each core) message lists for each destination node it
+is responsible for managing and, if the list is long enough or a
+message has waited past a time-out period, all messages to a given
+destination system node from that source system node are sent.
+Aggregating and sending a message involves manipulating a set of
+shared data-structures (the message lists). This is done using CAS
+(compare-and-swap) operations to avoid high synchronization costs.
+Note that we use a per-processor array of message lists
+that is only periodically modified across processor cores because we
+empirically determined this approach was faster (sometimes
+significantly) than a global per-system node array of message lists.
 
@@
 responsible for outbound messages to a given system node has received a
-sufficient number of message send requests, or a timeout is reached, the
+sufficient number of message send requests or a timeout is reached, the
 linked list of messages is walked and messages are copied to a send buffer.
 This process requires careful prefetching because most of the outbound
-messages are \emph{not} in the processor cache at this time (recall that a
+messages are \emph{not\/} in the processor cache at this time (recall that a
 core can be aggregating messages originating from other cores in the same
-node). Once the send buffer has been formed it is handed off to GASNet for
+node). Once the send buffer has been formed, it is handed off to GASNet for
 transfer to the remote system node. RDMA is used if the underlying network
@@
 There are two useful consequences of forming the send buffer at the time of
-message transmission instead of along the way, as individual upper layer
-message send requests are received. First, as previously mentioned most of the
-messages are not in the cache and prefetching is used to run-ahead in the
+message transmission instead of along the way as individual upper layer
+message send requests are received. First, as previously mentioned, most of the
+messages are not in the cache and prefetching is used to run ahead in the
 linked list of messages in order to avoid cache misses. But once the send
-buffer is formed it is in the cache (for the most part). Hence, when it is
+buffer is formed, it is in the cache (for the most part). Hence, when it is
 handed off to GASNet for transfer across the physical wire, the network card
@@
 nodes in the system, as the buffer can be allocated on the fly. Nevertheless
-we have found it efficient to build a flow-control like protocol of
+we have found it efficient to build a flow-control-like protocol of
 outstanding message buffers between pairs of system nodes.
 
-Once the remote system node has received the message buffer, a management task is spawned to manage the unpacking process.
-The management task spawns a task on each processor at the receiving system to simultaneously unpack messages destined for that processor.
-Upon completion, these unpacking tasks synchronize with the management task.
-Once all cores have processed the message buffer the management task sends a reply to the sending system node indicating the successful delivery of the messages.
+Once the remote system node has received the message buffer, a
+management task is spawned to manage the unpacking process.  The
+management task spawns a task on each processor at the receiving
+system to simultaneously unpack messages destined for that processor.
+Upon completion, these unpacking tasks synchronize with the management
+task.  Once all cores have processed the message buffer, the management
+task sends a reply to the sending system node indicating the
+successful delivery of the messages.
 

### 2013-07-23 doc/asplos14/evaluation.tex
many small fixes

@@
 
-Our questions to answer with experimentation are:
 The goal of our evaluation is to show that the core pieces of the \Grappa
@@
 \item We present results that show that \Grappa can
-    support large amounts of concurrency: sufficient for remote memory
+    support large amounts of concurrency, sufficient for remote memory
     access and aggregation. The communication layer is able to
-    achievesustain a very high rate of global memory operations. We also show the performance of a
+    sustain a very high rate of global memory operations. We also show the performance of a
 graph kernel that stresses communication and concurrency together.
@@
 ns. As we add workers, the time increases slowly, but levels off: we
-also ran with 500,000 workers (10 times what is shown in the figure)
-and found that context switch time was around 75 ns. In comparison,
+also ran with 500,000 workers (10~times what is shown in the figure)
+and found that context switch time was around 75ns. In comparison,
 for the same yield test using kernel-level Pthreads on a single core,
@@
 
-If we calculate aggregate context switch \emph{rate} of all
+If we calculate aggregate context switch \emph{rate\/} of all
 cores in a node, we find that with prefetching, \Grappa context
 switching is
-limited \emph{not} by memory latency, as normally assumed, but rather memory bandwidth.
-Specifically, we empirically found that 4 cache lines (1 for worker
+limited \emph{not\/} by memory latency, as normally assumed, but rather memory bandwidth.
+Specifically, we empirically found that 4~cache lines (1~for worker
 struct and 3 for stack data) was sufficient to avoid cache misses in
-the microbenchmark. Every context switch then requires 8 cache line transfers. The off-chip
+the microbenchmark. Every context switch then requires 8~cache-line transfers. The off-chip
 bandwidth of a single socket in our system is 270M cache lines per
-second \cite{porterfield:bw,Nelson:hotpar11}. This implies that
-we can sustain at most 34M context switches per second per socket once
-the space of all contexts is enough to mostly reside outside of L3.
+second~\cite{porterfield:bw,Nelson:hotpar11}. This implies that, in the limit,
+we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).
 
 In summary, our context switch engine is able to efficiently sustain very high
-concurrency and as we will show later, the amount of concurrency sustained is
+concurrency and, as we will show later, the amount of concurrency sustained is
 sufficient for the latencies \Grappa needs to hide.
@@
 \Grappa's global memory and communication layers using a faithful
-implementation of the giga updates per second (GUPs) benchmark, which measures cluster-wide random access bandwidth.
-Read-modify-write updates are dispatched at random to a large array. This
-benchmark stresses the communication layer of \Grappa separately from the
+implementation of the giga updates per second (GUPs) benchmark, which
+measures cluster-wide random access bandwidth.  Read-modify-write
+updates are dispatched at random to a large array. This benchmark
+stresses the communication layer of \Grappa separately from the
 scheduler, because only a single worker is used per system node.
-Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain well over a
-billion updates per second with 64 nodes. Note also that when aggregation is turned off, the update rate is nearly flat. Aggregation is instrumental for good communication performance.
+Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain
+well over a billion updates per second with 64~nodes. Note also that
+when aggregation is turned off, the update rate is nearly
+flat. Clearly aggregation is instrumental for good communication
+performance.
 
@@
 trees, we do not expect there to be sufficient concurrency for any system,
-including \Grappa, to achieve high throughput --- at the 16-node data point,
-the average active tasks per core over the search is 775 and 13 for T1 and T3,
-respectively. On the other hand, \Grappa performs and scales very well for T1
+including \Grappa, to achieve high throughput -- at the 16-node data point,
+the average active tasks per core over the search is 775 for T1 and 13 for T3.
+On the other hand, \Grappa performs and scales very well for T1
 trees.
@@
 \begin{tabular}{l|c|c|c|c}
-&	Cora &      XMT	&        MPI	&   UPC \\ \hline
+&	\Grappa & XMT	&        MPI	&   UPC \\ \hline
 GUPS	& 1	& 2.23	& 0.11& 	-- \\ 
@@
 and the application is yet to be done. Second, the MPI implementations
-we used employ algorithmic-specific optimizations---some of which
-resemble of what \Grappa does automatically, like message aggregation.
+we used employ algorithmic-specific optimizations -- some of which
+resemble what \Grappa does automatically, like message aggregation.
 Grappa's GUPS results presented earlier suggest there is significant
@@
 %%% I think it's ok without. -Mark
-\comment{Recall that global memory can be accessed in two ways. For requests of
+\grappacomment{Recall that global memory can be accessed in two ways. For requests of
 locations that are expected to shared, delegate operations are
@@
 In UTS, we obtained the best throughput by flushing frequently: on the
-order of 100 microseconds. This, combined with the fewer number of remote
-accesses, means that there are a limited nubmer of requests generated
+order of 100~microseconds. This, combined with the fewer number of remote
+accesses, means that there are a limited number of requests generated
 per flush interval, leading to packets primarily smaller than 10KB.
@@
 this. First, in BFS, we obtained the best throughput by flushing at
-least 500 microseconds apart, and also flushing whenever a core is
+least 500~microseconds apart, and also flushing whenever a core is
 idle. Since this process ignores whether another flush has occurred
@@
 
-\comment{
+\grappacomment{
 
@@
 %(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
-%nodes---the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
+%nodes -- the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
 
@@
 We attribute this to a limitation in the current aggregator design and network stack that \Grappa uses.
-This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases} performance).
+This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).
 Ironically, this limitation makes \Grappa applications compute-bound instead of network-bound.
@@
 %(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
-%nodes---the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
+%nodes -- the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
 %\Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
@@
 
-Parallel overhead---in the form of context switches, task spawns, and
-synchronization---can reduce the performance benefit of parallelism.
+Parallel overhead -- in the form of context switches, task spawns, and
+synchronization -- can reduce the performance benefit of parallelism.
 \Grappa sees a benefit to limiting the amount of parallelism created by

### 2013-07-23 doc/asplos14/intro.tex
many small fixes

@@
 the business, national security, and social network computing domains. For
-these emerging applications, reasonable response time---given the sheer
-amount of data---requires large multinode systems. The most broadly available
+these emerging applications, reasonable response time -- given the sheer
+amount of data -- requires large multinode systems. The most broadly available
 multinode systems are those built from x86 compute nodes interconnected via
@@
 This results in frequent requests for small pieces of remote data.
-Caches are of little assistance because of low data re-use and spatial locality.
+Caches are of little assistance because of low temporal and spatial locality.
 Prefetching is of limited value because request locations are not known early enough.
@@
 \vspace{0.5ex} \noindent{\bf High network communication latency.} The performance challenges of frequent communication are exacerbated by high network latency relative to processor performance.
-Latency of commodity networks runs anywhere from a few to hundreds of microseconds---tens of thousands of processor clock cycles.
+Latency of commodity networks runs anywhere from a few to hundreds of microseconds -- tens of thousands of processor clock cycles.
 Since irregular application tasks encounter remote references dynamically during execution and must resolve them before making further progress, stalls are frequent and lead to severely underutilized compute resources.
@@
 As a general design philosophy, \Grappa trades latency for throughput.
-By \emph{increasing} latency in key components of the system we are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance by tolerating the delays incurred when work-stealing.
-\Grappa then exploits parallelism to overcome latency.
+By \emph{increasing\/} latency in key components of the system we are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance via work-stealing.
+\Grappa then exploits parallelism to tolerate the increased latency.
 

### 2013-07-23 doc/asplos14/memory.tex
many small fixes

@@
 localized global memory in caches (see below), and accesses to debugging
-infrastructure that is local to each system node. Local pointers cannot access
+infrastructure local to each system node. Local pointers cannot access
 memory on other cores, and are valid only on their home core.
@@
 data in the global memory. The first is a distributed heap striped across all
-the machines in the system in a block cyclic fashion. The
+the machines in the system in a block-cyclic fashion. The
 \texttt{global\_malloc} and \texttt{global\_free} calls are used to allocate
@@
 bit is set to indicate that the reference is a 2D address (as opposed to
-linear address). This leaves 15 bits for network endpoint ID, which limits our
-scalability to $2^{15}$ endpoints. Any node-local data can be made accessible
+linear address). This leaves 15~bits for network endpoint ID, which limits our
+scalability to $2^{15}$ cores. Any node-local data can be made accessible
 to other cores in the system by wrapping the address and node ID into a 2D
@@
 Access to \Grappa's distributed shared memory is provided through  {\em
-delegate} operations, which are short memory accesses performed at the memory
+delegate\/} operations, which are short memory accesses performed at the memory
 location's home node. When the data access pattern has
@@
 
-Delegate operations can be either {\em blocking} or {\em non-blocking}.
+Delegate operations can be either {\em blocking\/} or {\em non-blocking}.
 With blocking operations, the task issuing the delegate call blocks until

### 2013-07-23 doc/asplos14/method.tex
many small fixes

--- 2013-03-28 doc/sosp13/method.tex
@@
 benchmarks to \Grappa as well as collected and optimized a set of comparison
-benchmarks for XMT, MPI and UPC~\cite{UPC}. We run the \Grappa, MPI and UPC
+benchmarks for XMT, MPI, and UPC~\cite{UPC}. We run the \Grappa, MPI and UPC
 experiments on a cluster of AMD Interlagos processors. Nodes have 32 2.1-GHz
@@
 workloads, particularly large-scale graph analysis problems, that are
-important among cybersecurity, informatics, and network understanding
+important among cybersecurity, informatics, and network-understanding
 workloads. The BFS benchmark builds a search tree containing parent nodes for
@@
 The performance metrics for NAS Parallel Benchmarks, including IntSort, are ``millions of operations per second'' (MOPS).
-For IntSort, this ``operation'' is ranking a single key, so it is roughly comparable to ``GUPS'' or ``TEPS''.
+For IntSort, this ``operation'' is ranking a single key, so it is roughly comparable to ``GUPS'' or ``TEPS.''
 
@@
 function over the in-edges is associative and commutative, so they can
-be processed in any order in parallel. Rather than the programmer writing the
+be processed in any order or in parallel. Rather than the programmer writing the
 parallel dot product as local accumulations with a final all-reduce
 step, we simply send streaming increments to each element of the final
-vector.
+vector. We compare PageRank to published results for the Trilinos linear algebra library implemented in MPI, and multithreaded PageRank for the XMT \cite{MTGL:pagerank}.
+For \Grappa, we run on a scale 29 graph using the Graph500 generator.
 

### 2013-07-23 doc/asplos14/tasks.tex
many small fixes

@@
 The basic unit of execution in \Grappa is a {\em task}. When tasks are ready to
-execute, they are mapped to a {\em worker}, which is akin to a user-level
+execute, they are mapped to a {\em worker,} which is akin to a user-level
 thread. Each hardware core has a single operating system thread pinned to it.
@@
 \item {\it private task queue}, a queue of tasks that must run on
-this core and therefore is not subject to stealing;
+this core and is therefore not subject to stealing;
 
@@
 \end{itemize}
-
 Whenever a task yields, the scheduler makes a decision about what
@@
 
-\Grappa provides four methods for expressing parallelism, shown in
+\Grappa provides three methods for expressing parallelism, shown in
 Figure~\ref{fig:expressing-parallelism}. First, when the programmer identifies
@@
 tasks, stopping to execute the loop body when the number of iterations is
-below the required threshold. Third, a programmer may want to execute an active message~\cite{vonEicken:1992:AMM:139669.140382}; that is, to run a
+below the required threshold. Third, a programmer may want to execute an active message~\cite{vonEicken92}; that is, to run a
 small non-blocking piece of code on a particular core in the system without waiting for
  a free worker to become available. For example, custom synchronization primitives are implemented as an active message executed on the core where the data
-lives. \Grappa provides the \texttt{call\_on} call for this purpose.
+lives. \Grappa provides the \texttt{call\_on} routine for this purpose.
 

### 2013-07-23 doc/asplos14/intro.tex
fixed bib references

@@
 Prefetching is of limited value because request locations are not known early enough.
-Data-parallel frameworks such as MapReduce~\cite{mapreduce:osdi04} are ineffective because they rely on effective data partitioning and regular communication patterns.
+Data-parallel frameworks such as MapReduce~\cite{Dean:2008:MSD:1327452.1327492} are ineffective because they rely on effective data partitioning and regular communication patterns.
 Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.

### 2013-07-23 doc/asplos14/tasks.tex
pass on tasks

@@
 % via the \Grappa API calls shown in
-%Figure~\ref{fig:scheduling}. 
+% Figure~\ref{fig:scheduling}. 
 To minimize context-switch overhead, the \Grappa scheduler
 operates entirely in user-space and does little more than store state of one
-worker and load that of another.
-\TODO{mention how we block on response (``scheduler activations'' or whatever) not busy wait}
+worker and load that of another. When a tasks encounters a long-latency operation, its worker is suspended and subsequently woken when the operation completes.
 
-\TODO{either explain the usage of yield vs suspend in talking about
-    short and longer latency tolerance or eliminate the
-    distinction}
+%% Maybe the sentence above is enough for the TODOs below.
+%
+% \TODO{mention how we block on response (``scheduler activations'' or whatever) not busy wait}
+% 
+% \TODO{either explain the usage of yield vs suspend in talking about
+%     short and longer latency tolerance or eliminate the
+%     distinction}
+
 %% \begin{figure}[htbp]
@@
 Each core in a \Grappa system has its own independent scheduler. The scheduler
-has a collection of active workers ready to execute called the {\it
-ready worker queue}. Each scheduler also has three queues of tasks
-waiting to be assigned a worker:
+has a collection of active workers ready to execute called the {\it ready
+worker queue}. Each scheduler also has three queues of tasks waiting to be
+assigned a worker:
 
@@
 
-\TODO{addressing reviewer2: Grappa provides mechanisms for many
-    programming idioms. Also, may parallelism to
-    give flexibility to the implementation.}
+% \TODO{addressing reviewer2: Grappa provides mechanisms for many
+%     programming idioms. Also, may parallelism to
+%     give flexibility to the implementation.}
+
 \Grappa programmers focus on expressing as much parallelism as possible

### 2013-07-23 doc/asplos14/evaluation.tex
small fixes

@@
 time as the number of workers grow. At our standard operating point
-($\approx$1K workers), context switch time is on the order of 50
-ns. As we add workers, the time increases slowly, but levels off: we
-also ran with 500,000 workers (10~times what is shown in the figure)
-and found that context switch time was around 75ns. In comparison,
-for the same yield test using kernel-level Pthreads on a single core,
-the switch time is 450ns for a few threads and 800ns for 1000-32000
-threads.
+($\approx$1K workers), context switch time is on the order of
+50~ns. As we add workers, the time increases slowly, but levels off:
+we also ran with 500,000 workers (10~times what is shown in the
+figure) and found that context switch time was around 75~ns. In
+comparison, for the same yield test using kernel-level Pthreads on a
+single core, the switch time is 450~ns for a few threads and 800~ns for
+1000--32000 threads.
 
@@
 second~\cite{porterfield:bw,Nelson:hotpar11}. This implies that, in the limit,
-we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).
+we can sustain at most 34M context switches per second per socket (a context-switch time of 29~ns).
 
@@
 within striking distance of XMT and well beyond UPC performance (UTS only).
-However, it underperforms compared to hand-tuned MPI. We believe there are two reasons for this:
+However, it underperforms compared to hand-tuned MPI\@. We believe there are two reasons for this:
 First, Grappa is young; much optimization and tuning of the runtime
@@
 data. Since explicit cache operations are intended to exploit
-locality; they more often access data on the local node. Only 20
-percent of UTS explicit cache operations are remote, whereas 44
-percent of BFS explict cache operations are remote.}
+locality; they more often access data on the local node. Only 20\%
+of UTS explicit cache operations are remote, whereas 44\%
+of BFS explict cache operations are remote.}
 
 In UTS, we obtained the best throughput by flushing frequently: on the
-order of 100~microseconds. This, combined with the fewer number of remote
+order of 100~$\mu$s. This, combined with the fewer number of remote
 accesses, means that there are a limited number of requests generated
-per flush interval, leading to packets primarily smaller than 10KB.
+per flush interval, leading to packets primarily smaller than~10KB.
 
@@
 this. First, in BFS, we obtained the best throughput by flushing at
-least 500~microseconds apart, and also flushing whenever a core is
+least 500~$\mu$s apart, and also flushing whenever a core is
 idle. Since this process ignores whether another flush has occurred
@@
 applications at scale, we ran a small (100M node tree) UTS-Mem
-with the aggregator disabled, on 16 nodes.
+with the aggregator disabled, on 16~nodes.
 Figure~\ref{fig:no-aggregation-uts}. At this configuration, the aggregator

### 2013-07-23 doc/asplos14/method.tex
small fixes

@@
 \vspace{0.5ex}
-\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the CrayXMT's early predecessor once held the world speed record~\cite{TeraRecord}.
+\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the Cray XMT's early predecessor once held the world speed record~\cite{TeraRecord}.
 The largest problem size, class ``D'', ranks two billion uniformly distributed random integers using either a bucket or a counting sort algorithm, depending on the strengths of the system.
@@
 step, we simply send streaming increments to each element of the final
-vector. We compare PageRank to published results for the Trilinos linear algebra library implemented in MPI, and multithreaded PageRank for the XMT \cite{MTGL:pagerank}.
+vector. We compare PageRank to published results for the Trilinos linear algebra library implemented in MPI, and multithreaded PageRank for the XMT~\cite{MTGL:pagerank}.
 For \Grappa, we run on a scale 29 graph using the Graph500 generator.

### 2013-07-23 doc/asplos14/conclude.tex
reworded confusing sentance

@@
 
-Irregular computations are both important and challenging to execute quickly.
-Scaling these applications easily on commodity hardware has been a historical
-challenge. \Grappa is a runtime framework that simplifies this task for
-software developers and compiler writers. This paper describes the \Grappa
-framework and its three main components: a task library, a distributed shared
-memory system, and a network aggregator. \Grappa's key aspect is extreme latency tolerance,
-which not only hides network latency but also enables the system to spend time
-on sophisticated work stealing and network optimizations, trading latency for
-even more throughput.
+Irregular computations are both important and challenging to execute
+quickly.  Scaling these applications easily on commodity hardware has
+been a historical challenge. \Grappa is a runtime framework that
+simplifies this task for software developers and compiler
+writers. This paper describes the \Grappa framework and its three main
+components: a task library, a distributed shared memory system, and a
+network aggregator. \Grappa's key aspect is extreme latency tolerance,
+which not only hides network latency but also enables the system to
+spend time on sophisticated work stealing and network optimizations,
+trading latency for even more throughput.
 
-Our evaluation of \Grappa reveals that the core components, scheduling and communication, achieve their design goals.
-Thousands of workers can be efficiently context switched on a multicore processor, up to the DRAM bandwidth.
-Aggregating messages enables \Grappa to achieve over 1.0 GUPS on 64 nodes.
-We also explored four other algorithms: unbalanced tree search, breadth first search, PageRank, and integer sort, comparing performance of these algorithms on \Grappa, the Cray XMT, and hand-optimized MPI implementations.
-As would be expected, when the MPI implementation exploits in the application, many of the same techniques \Grappa provides in the runtime, such as aggregating requests, yet without the overhead of tasking, the MPI implementation excels.
-This comes with enormous implementation complexity, burdening the application developer.
-Moreover, significant tuning has yet to be done on the \Grappa runtime system.
-Compared to the Cray XMT \Grappa is between 2.6X faster and 4.3X slower.
-Yet when cost-performance is considered, mass-market x86 cluster hardware makes \Grappa a highly attractive option.
+Our evaluation of \Grappa reveals that the core components, scheduling
+and communication, achieve their design goals.  Thousands of workers
+can be efficiently context switched on a multicore processor, up to
+the DRAM bandwidth.  Aggregating messages enables \Grappa to achieve
+over 1.0~GUPS on 64~nodes.  We also explored four other algorithms:
+unbalanced tree search, breadth first search, PageRank, and integer
+sort, comparing performance of these algorithms on \Grappa, the Cray
+XMT, and hand-optimized MPI implementations.  As might be expected,
+when an MPI implementation exploits, in the application code, many of the
+same techniques \Grappa provides in the runtime, such as aggregating
+requests, yet without the overhead of tasking, the MPI implementation
+excels.  This comes with enormous implementation complexity, burdening
+the application developer. Moreover, significant tuning has yet to be
+done on the \Grappa runtime system.  Compared to the Cray XMT, \Grappa
+is between 2.6X faster and 4.3X slower.  Yet when cost-performance is
+considered, mass-market x86-based cluster hardware makes \Grappa a
+highly attractive option.

### 2013-07-24 doc/asplos14/communication.tex
more small fixes

@@
 Note that we use a per-processor array of message lists
-that is only periodically modified across processor cores because we
-empirically determined this approach was faster (sometimes
+that is only periodically modified across processor cores after
+experimentally determining that this approach was faster (sometimes
 significantly) than a global per-system node array of message lists.

### 2013-07-24 doc/asplos14/evaluation.tex
more small fixes

@@
 ($\approx$1K workers), context switch time is on the order of
-50~ns. As we add workers, the time increases slowly, but levels off:
+50ns. As we add workers, the time increases slowly, but levels off:
 we also ran with 500,000 workers (10~times what is shown in the
-figure) and found that context switch time was around 75~ns. In
+figure) and found that context switch time was around 75ns. In
 comparison, for the same yield test using kernel-level Pthreads on a
-single core, the switch time is 450~ns for a few threads and 800~ns for
+single core, the switch time is 450ns for a few threads and 800ns for
 1000--32000 threads.
@@
 second~\cite{porterfield:bw,Nelson:hotpar11}. This implies that, in the limit,
-we can sustain at most 34M context switches per second per socket (a context-switch time of 29~ns).
+we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).
 
@@
 including \Grappa, to achieve high throughput -- at the 16-node data point,
-the average active tasks per core over the search is 775 for T1 and 13 for T3.
+the average active tasks per core over the search is 775 for T1 and
+only 13 for T3.
 On the other hand, \Grappa performs and scales very well for T1
@@
 In UTS, we obtained the best throughput by flushing frequently: on the
-order of 100~$\mu$s. This, combined with the fewer number of remote
+order of 100$\mu$s. This, combined with the fewer number of remote
 accesses, means that there are a limited number of requests generated
@@
 this. First, in BFS, we obtained the best throughput by flushing at
-least 500~$\mu$s apart, and also flushing whenever a core is
+least 500$\mu$s apart, and also flushing whenever a core is
 idle. Since this process ignores whether another flush has occurred

### 2013-07-24 doc/asplos14/intro.tex
more small fixes

@@
 \vspace{0.5ex}
-\noindent{\bf Poor data locality and frequent communication.} Data reference patterns of irregular applications are unpredictable and tend to be spread across the entire system.
+\noindent{\bf Poor data locality and frequent communication} Data reference patterns of irregular applications are unpredictable and tend to be spread across the entire system.
 This results in frequent requests for small pieces of remote data.
@@
 
-\vspace{0.5ex} \noindent{\bf High network communication latency.} The performance challenges of frequent communication are exacerbated by high network latency relative to processor performance.
+\vspace{0.5ex} \noindent{\bf High network communication latency} The performance challenges of frequent communication are exacerbated by high network latency relative to processor performance.
 Latency of commodity networks runs anywhere from a few to hundreds of microseconds -- tens of thousands of processor clock cycles.

### 2013-07-24 doc/asplos14/method.tex
more small fixes

@@
 benchmarks to \Grappa as well as collected and optimized a set of comparison
-benchmarks for XMT, MPI, and UPC~\cite{UPC}. We run the \Grappa, MPI and UPC
+benchmarks for XMT, MPI, and UPC~\cite{UPC}. We ran the \Grappa, MPI, and UPC
 experiments on a cluster of AMD Interlagos processors. Nodes have 32 2.1-GHz
@@
 \noindent{\bf IntSort:} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the Cray XMT's early predecessor once held the world speed record~\cite{TeraRecord}.
-The largest problem size, class ``D'', ranks two billion uniformly distributed random integers using either a bucket or a counting sort algorithm, depending on the strengths of the system.
+The largest problem size, class D, ranks two billion uniformly distributed random integers using either a bucket or a counting sort algorithm, depending on the strengths of the system.
 Bucket sort executes a greater number of loops, but is able to leverage locality and avoid communication completely in the final phase, ranking within buckets.

### 2013-07-24 doc/asplos14/overview.tex
fixed todo
Cora->Manta
more small fixes

@@
 \begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:grappa} \Grappa system overview\TODO{Cora->Manta}}
+  \caption{\label{fig:grappa} \Grappa system overview}
 \end{minipage}
@@
 
-\item [Tasking system.] The tasking system supports lightweight multithreading
+\item [Tasking system] The tasking system supports lightweight multithreading
 to tolerate communication latency and global distributed work-stealing (i.e.,
@@
 
-\item[Communication layer.] The main goal of our communication layer is to
+\item[Communication layer] The main goal of our communication layer is to
 aggregate small messages into large ones. This process is invisible to the
@@
 
-\item[Distributed shared memory.] The DSM system provides fine-grain access to
+\item[Distributed shared memory] The DSM system provides fine-grain access to
 data anywhere in the system, with delegated operation at the core of its design.

### 2013-07-24 doc/asplos14/tasks.tex
more small fixes

@@
 
-\begin{itemize}
+\begin{description}
 
-\item {\it deadline task queue}, a priority queue of tasks that are executed according to task-specific deadline constraints;
+\item[deadline task queue] a priority queue of tasks that are executed according to task-specific deadline constraints;
 
-\item {\it private task queue}, a queue of tasks that must run on
+\item[private task queue] a queue of tasks that must run on
 this core and is therefore not subject to stealing;
 
-\item {\it public task queue}, a queue of tasks that are
+\item[public task queue] a queue of tasks that are
   waiting to be matched with workers. It is a local partition of a shared
@@
 
-\end{itemize}
+\end{description}
 Whenever a task yields, the scheduler makes a decision about what

### 2013-07-24 doc/asplos14/evaluation.tex
insert new BFS numbers; expand on discussion

@@
 with XMT running BFS, PageRank, IntSort, GUPS and UTS. Since XMT is a
-different hardware platform, we also compare \Grappa with hand-tuned MPI
+different hardware platform, we also compare \Grappa with optimized MPI
 versions of BFS and GUPS running on the same hardware. Finally, we also
-compare it with UTS written for UPC. We run all experiments with 64 nodes.
+compare it with UTS written for UPC. We run all experiments with 64 nodes and 16 cores per node.
 
@@
          & \Grappa & XMT   & MPI  & UPC \\ \hline
-GUPS	& 1	& 2.23	& 0.11& 	-- \\ 
-BFS	& 1 &	4.33	& 14.37 & 	-- \\ 
+GUPS     & 1       & 2.23  & 0.11 & -- \\ 
+BFS      & 1       & 1.63  & 3.52 & -- \\ 
 Intsort  & 1       & 3.77  & 5.88 & -- \\
@@
 \end{center}
-\caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
+\caption{Comparing \Grappa with XMT, optimized MPI and UPC. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
 \label{tab:grappa-comparisons}
@@
 
-\TODO{addressing reviwer2: sweep of irregularness in one aggregate or multiple dimensions
-    and how a specific aspect of performance changes in different
-    systems}
+Table~\ref{tab:grappa-comparisons} shows the results. \Grappa is able
+to provide performance approaching and sometimes exceeding that of the
+other systems. This data shows that, while \Grappa is good at the
+operations for which it was designed, there is a cost to providing
+\Grappa's generality. In general, the benchmarks whose MPI
+implementations are faster than \Grappa include a implementation of a
+subset of \Grappa's functionality specialized for their particular
+problem. We now discuss each benchmark in more detail.
 
-Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
-within striking distance of XMT and well beyond UPC performance (UTS only).
-However, it underperforms compared to hand-tuned MPI\@. We believe there are two reasons for this:
-First, Grappa is young; much optimization and tuning of the runtime
-and the application is yet to be done. Second, the MPI implementations
-we used employ algorithmic-specific optimizations -- some of which
-resemble what \Grappa does automatically, like message aggregation.
-Grappa's GUPS results presented earlier suggest there is significant
-room for improvement.
+\paragraph{UTS.}
+It is perhaps not surprising that the XMT is faster than
+\Grappa on many of the benchmarks; after all, the XMT is custom
+hardware optimized for irregular applications. The XMT can
+context-switch every cycle, while \Grappa takes $~$50 ns. The XMT can
+complete 100 million random-access remote reads per second per node in
+hardware, while \Grappa must spend many instructions managing and
+aggregating each of these reads in software.
+
+What is more surprising is that Grappa is able to exceed the XMT's
+performance on UTS. This is a demonstration of the inflexibility of
+custom hardware. Recall that the UTS benchmark searches an unbalanced
+tree and spawns a data-dependent number of child tasks at each
+vertex. The XMT's spawn/join semantics, whose implementation is spread
+across the hardware, OS, and runtime, require a task visiting a vertex
+to block until all its child tasks have completed, even though the
+benchmark does not require those tasks to have completed until the end
+of the tree search. Furthermore, each new task is immediately made
+available to all other processors in the system, even though it is
+likely that processor that spawned a task will end up executing it.
+
+Since \Grappa's tasks are implemented entirely in software, it is easy
+to support arbitrary synchronization semantics. \Grappa allows both
+XMT-like per-vertex task joins as well as whole-problem joins that are
+a better fit for UTS. Furthermore, \Grappa's work-stealing distributed
+task queue and recursive loop decomposition encourage locality in task
+execution, so that the system is optimized for the common case of a
+processor executing a task it spawned.
+
+The UPC UTS implementation includes an implementation of work-stealing
+but has limited ability to overlap communication with
+computation. Furthermore, it must execute heavyweight synchronization
+operations when stealing. \Grappa benefits from its lightweight
+context switching and support for fine-grained synchronization.
+
+\paragraph{BFS.}
+The Grappa and MPI BFS implementations are quite similar. They both
+use the same graph representation, and they both depend on aggregation
+for performance. The difference is that the BFS' aggregation is
+integrated with the traversal code and specialized for the problem:
+since the only messages required are edge traversals, the aggregation
+buffers store only pairs of 8-byte vertex IDs.  In contrast, \Grappa
+pays execution and space overhead in order to support aggregation of
+arbitrary closures; the messages \Grappa aggregates include the two
+8-byte vertex IDs along with 8 bytes of deserialization information as
+well as 8 bytes of synchronization information. Furthermore, the MPI
+implementation writes the edge information directly into the
+aggregation buffer while \Grappa executes additional code to serialize
+and deserialize the messages. \TODO{Include message counts, size, and
+  rates; code size too?}.
+
+\paragraph{Other benchmarks.}
+
+The MPI GUPS implementation includes an implementation of aggregation,
+but it is not optimized to work when the aggregated data exceeds
+cache, and it has limited support for concurrent communication with
+multiple destinations. \Grappa benefits from being optimized for both
+of these cases.
+
+The MPI Pagerank implementation it built on top of the
+highly-optimized sparse matrix support in the Trilinos~\cite{trilinos}
+library. In contrast, the \Grappa implementation is fairly
+straightforward.
+
+\TODO{Describe Intsort}
+
+\paragraph{Summary.}
+
+Overall, \Grappa provides a general programming model at a moderate
+performance cost. While \Grappa's core functionality performs well,
+applications can specialize similar functionality for their problems
+and obtain better performance than \Grappa on the same hardware. This
+is, however, not the end of the story for \Grappa's performance;
+\Grappa is a young library and, as discussed in the next section, is
+limited by its implentation rather than the hardware on which it
+runs. We believe there are opportunities for optimization that will
+improve its performance.
 

### 2013-07-24 doc/asplos14/evaluation.tex
rearrange headings

@@
 
-\paragraph{Other benchmarks.}
-
+\paragraph{GUPS.}
 The MPI GUPS implementation includes an implementation of aggregation,
@@
 
+\paragraph{Pagerank.}
 The MPI Pagerank implementation it built on top of the
@@
 
+\paragraph{Intsort.}
 \TODO{Describe Intsort}
@@
 \Grappa is a young library and, as discussed in the next section, is
-limited by its implentation rather than the hardware on which it
+limited by its implementation rather than the hardware on which it
 runs. We believe there are opportunities for optimization that will

### 2013-07-24 doc/asplos14/method.tex

@@
 the XMT compiler can be made to recognize and apply a Manhattan loop collapse, exposing enough parallelism to allow it to scale out to 64 nodes for the problem scales we show.
-We do not employ algorithmic improvements for any of these versions, though there are
+In order to make comparison easier, we do not employ algorithmic improvements for any of these versions, though there are
 many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}. \Grappa can be expected to benefit the same as MPI due to decreased communication.
@@
 step, we simply send streaming increments to each element of the final
-vector. We compare PageRank to published results for the Trilinos linear algebra library implemented in MPI, and multithreaded PageRank for the XMT~\cite{MTGL:pagerank}.
+vector. We compare PageRank to published results for the Trilinos linear algebra library implemented in MPI~\cite{Plimpton:2011:MML:2286659.2286704},
+ and multithreaded PageRank for the XMT~\cite{MTGL:pagerank}.
 For \Grappa, we run on a scale 29 graph using the Graph500 generator.

### 2013-07-24 doc/asplos14/conclude.tex
todo in conclusion
new data, partial characterization

@@
 sort, comparing performance of these algorithms on \Grappa, the Cray
-XMT, and hand-optimized MPI implementations.  As might be expected,
-when an MPI implementation exploits, in the application code, many of the
-same techniques \Grappa provides in the runtime, such as aggregating
-requests, yet without the overhead of tasking, the MPI implementation
-excels.  This comes with enormous implementation complexity, burdening
-the application developer. Moreover, significant tuning has yet to be
-done on the \Grappa runtime system.  Compared to the Cray XMT, \Grappa
-is between 2.6X faster and 4.3X slower.  Yet when cost-performance is
-considered, mass-market x86-based cluster hardware makes \Grappa a
-highly attractive option.
+XMT, and optimized MPI implementations.
+
+Depending on the benchmark, \Grappa is between 11X faster and 5.4X
+slower than MPI and UPC, and between 2.6X faster and 4.3X slower than
+the XMT. When \Grappa is slower than MPI, we find that the MPI
+implementation includes a specialized implementation of a feature
+\Grappa provides generally to all applications. This generality comes
+at a performance cost but eases the application developer's
+task. Moreover, significant tuning has yet to be done on the \Grappa
+runtime system.  Yet when cost-performance is considered, mass-market
+x86-based cluster hardware makes \Grappa a highly attractive option.

### 2013-07-24 doc/asplos14/evaluation.tex
notes about evaluation context switching
new data, partial characterization

@@
 we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).
+\TODO{based on this and GUPS, our message rate will be limited by
+    context switching for small+single messages. One experiment is
+    GUPs with context switching, where we vary number of reads/suspend}
 
@@
 BFS      & 1       & 1.63  & 3.52 & -- \\ 
-Intsort  & 1       & 3.77  & 5.88 & -- \\
+IntSort  & 1       & 3.59  & 5.36 & -- \\
 UTS (T1) & 1       & 0.38  & --   & 0.09 \\ 
@@
 \paragraph{Pagerank.}
-The MPI Pagerank implementation it built on top of the
+The MPI Pagerank implementation is built on top of the
 highly-optimized sparse matrix support in the Trilinos~\cite{trilinos}
-library. In contrast, the \Grappa implementation is fairly
-straightforward.
+library. In contrast, the \Grappa implementation is a straightforward nested
+loop.
 
-\paragraph{Intsort.}
-\TODO{Describe Intsort}
+\paragraph{IntSort.}
+Both MPI and \Grappa IntSort implement a bucket sort. For \Grappa,
+distributing keys into buckets accounts for the bulk of the execution
+time; \Grappa does this using asynchronous delegate operations to
+write the keys directly to their final destination. The MPI
+implementation, on the other hand, first does a local sort of the keys
+on each core, and then uses an \texttt{MPI\_Alltoallv()} collective
+operation to move all the keys to their destination nodes in
+bulk. This is essentially specialized aggregation combined with
+collective communication. When we measure the time spent in this
+region of both benchmarks, we find that the the sort and
+\texttt{MPI\_Alltoallv()} are $~5\times$ faster than the \Grappa
+version using generic aggregation and point-to-point communication. An
+interesting future direction would be to extend \Grappa to support
+aggregation via collective operations.
 
@@
 
-\paragraph{Where execution time goes.}
+\begin{table*}[htb]
+\begin{center}
+\begin{tabular}{l|c|c|c|c|c}
+% remember: bfs time is mean_time * nbfs
+                               &   GUPS            &   BFS   & IntSort & UTS & Pagerank \\ \hline
+Application message rate/core  & 984K/s            &  983K   &           \\
+%Application message bytes     & 31.8 $\pm$ 1.2    & \\
+Avg. Application message bytes & 31.8              & 33.9    &       \\
+Network bandwidth per node     & 478MB/s           & 511MB/s &          \\
+%Network message bytes         & 23.2K $\pm$ 12.4K & \\ \hline
+Avg. Network message bytes     & 23.2K             & 4.3K    &     \\ \hline
+Avg. active tasks/core         & 0.9               & 58.2    &    \\
+Max. active tasks/core         & 1                 & 128     &     \\
+Avg. ready queue length/core   & 2.3               & 6.1     &      \\
+Avg. Context switch rate/core & 34.2K/s           & 539K/s  &       \\ 
+Steal count                    & 0                 & 0       &       \\
+\end{tabular}
+\end{center}
+\caption{Internal runtime metrics, for 64-node, 16-core-per-node benchmark runs \TODO{Finish} }
+\label{tab:grappa-metrics}
+\end{table*}
+
+\paragraph{Runtime metrics}
+Table~\ref{tab:grappa-metrics} show a number of internal runtime
+metrics collected while executing the benchmarks from the previous
+section. These are per-core averages computed over all 1024 cores of
+the 64-node, 16-core-per-node jobs.
+
+The first group of four metrics relate to the communication
+layer. Application messages refer to those issued by the user code;
+network message refer to the aggregated packes sent over the wire. The
+data show that most application messages are only a handful of bytes,
+but our aggregator is able to turn them into packets of many kilobytes.
+
+The next group of five metrics relate to the scheduler. We show the
+average and maximum number of concurrently-executing tasks, along with
+the average length of the ready queue and the average context switch
+rate.  The last line shows the number of time a core stole work from
+another core. Only UTS depends on work-stealing for performance; the
+other applications exploit locality by binding tasks to specific cores
+Nevertheless, even in UTS, which depends on work-stealing for
+performance, steals are an infrequent occurance and account for a
+small fraction of the execution time.
 
@@
     \begin{center}
-      \begin{tabular}{c|c c c c}
-        Benchmark     & Comm & User & Idle & Sched \\ \hline
-        GUPS          & 62.9  & 18.9    & 15.4 & 2.80 \\ 
-        BFS           & 54.84 & 30.90   & 10.94 & 3.43 \\ 
-        Intsort       & 34.28 & 42.00   & 21.31 & 2.47 \\ 
-        UTS           & 40.57 & 56.52   &  1.21 & 1.73 \\
-        Pagerank      & 76.79 & 20.71   &  0.06 & 2.50 \\
-      \end{tabular}
+      \includegraphics[width=0.5\textwidth]{results/plot_concurrency.pdf}
     \end{center}
-    \caption{\Grappa\ execution time profile, in percent.}
-%bar chart, one set of bars per benchmark, one bar per system. runtime normalized to grappa.
-    \label{fig:grappa-profile}
-\end{figure}
-\TODO{ shorten the execution profile section in favor of deeper
-    discussion about where the message rate comes from}
-
-Figure \ref{fig:grappa-profile} shows the breakdown of execution
-time. Since \Grappa is focused on frequent communication, it is not
-surprising that many of our benchmarks spend more than half of their
-time doing communication. User time is primarily focused on generating
-more requests; idle time occurs when we run out of parallel work or
-out of workers to execute that work. Scheduler time includes both
-context switch time and deciding what task to run next. Our efficient
-context switching allows us to keep this to a few percent of execution
-time even with thousands of threads per core.
-
-\paragraph{Message size and latency.}
-
-To characterize the effectiveness of aggregation, we collected
-histograms of aggregated message sizes and application memory request
-latency for BFS and UTS (Figure \ref{fig:grappa-message-size}).
-
-%%% I couldn't make this paragraph useful in the limited time we have.
-%%% I think it's ok without. -Mark
-\grappacomment{Recall that global memory can be accessed in two ways. For requests of
-locations that are expected to shared, delegate operations are
-used. For requests of non-shared locations where locality can be
-exploited, explicit cache operations are used. We see this reflected
-in the fraction of requests that are handled remotely: less than one
-percent of delegate operations in both BFS and UTS access local
-data. Since explicit cache operations are intended to exploit
-locality; they more often access data on the local node. Only 20\%
-of UTS explicit cache operations are remote, whereas 44\%
-of BFS explict cache operations are remote.}
-
-In UTS, we obtained the best throughput by flushing frequently: on the
-order of 100$\mu$s. This, combined with the fewer number of remote
-accesses, means that there are a limited number of requests generated
-per flush interval, leading to packets primarily smaller than~10KB.
-
-In BFS, we see a trimodal distribution. There are two reasons for
-this. First, in BFS, we obtained the best throughput by flushing at
-least 500$\mu$s apart, and also flushing whenever a core is
-idle. Since this process ignores whether another flush has occurred
-recently, it leads to a large number of small messages. This leads to
-the first and smallest mode of the distribution.  The other two modes
-are due to a locality optimization in the \Grappa version of BFS. To
-minimize unnecessary remote communication, each core visits vertices
-whose edge lists are stored locally.
-Since we are traversing power-law graphs, a small fraction of vertices have a very high number of edges.
-This means that occasionally a core visits a high-degree vertex which allows it to very quickly generate a large batch of requests.
-This interacts well with aggregation; when we find high degree vertices, we are able to aggregate packets on the order of 40KB.
-In flush intervals without high-degree vertices, we see packets on the order of 8KB.
-
-Figure \ref{fig:grappa-latency} shows the latency distribution of
-remote delegate operations. This distribution follows the aggregated
-message size distribution: when we form larger packets, the system
-takes longer to process them.
-
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{results/histograms/latency_cmb.pdf}
-    \end{center}
-    \caption{Round-trip latency of delegate operations.}
-%bar chart, one set of bars per benchmark, one bar per system. runtime normalized to grappa.
-    \label{fig:grappa-latency}
+    \caption{Throughput, request latency, and idleness as number of concurrent workers is varied}
+    \label{fig:grappa-concurrency}
 \end{figure}
 
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{results/histograms/rdma_bytes_sent_histogram_cmb.pdf}
-    \end{center}
-    \caption{Distribution of aggregated message sizes.}
-%bar chart, one set of bars per benchmark, one bar per system. runtime normalized to grappa.
-    \label{fig:grappa-message-size}
-\end{figure}
+\paragraph{How much concurrency does \Grappa require? How much latency does it add?}
+Figure~\ref{fig:grappa-concurrency} shows 
+UTS blocks at ea
+\todo{FINISH}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% OLD %%%%%%%%%%%%%%%%%%
+\paragraph{Does \Grappa scale?}
 
-\grappacomment{
+\todo{FINISH}
+Our current
 
-Our evaluation begins with presentation of microbenchmark results, establishing
-the intrinsic potential of \Grappa to provide random access bandwidth and latency tolerance. 
-Next, we present application results, both for \Grappa and for other paradigms, as well
-as comparing against the Cray XMT.  Finally, we present the impact of increased aggregation delay
-on \Grappa results, thus exploring robustness to network scale.
+\paragraph{What limits \Grappa's performance?}
+The most common operation in \Grappa is sending a message. Three key
+operations occur in a message's lifetime: creating and enqueuing a
+message, serializing a message in the aggregator, and deserializing a
+message at the destination. We benchmarked each of these steps
+individually to find the bottleneck.
 
-\subsection{Microbenchmark Results}
-
-\subsubsection{Random Access}
-\TODO{Random access feed forward results on \Grappa.  optional: Results we measured for XMT.  Cite MPI results}
-\subsubsection{Latency Tolerance}
-\TODO{Simple ping test results -- eg, MPI ping, not the full blown aggregation ping test.  Random access blocking results on \Grappa.}
-\subsubsection{Scheduling and Robustness}
-
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/context_switch_time.pdf}
-    \end{center}
-    \caption{Average context switch time with 1 and 6 active cores,
-        with and without prefetching.}
-    \label{fig:context-switch-time}
-\end{figure}
-
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/context_switch_bw.pdf}
-    \end{center}
-    \caption{Context switch rate with prefetching. Once the total
-        size of contexts sufficiently exceeds last-level cache, most
-    prefetches go to main memory and the rate becomes limited to the
-    off-chip bandwidth.}
-    \label{fig:context-switch-time}
-\end{figure}
-
-\TODO{Summary of what to write please expand: Reference above plots. Bandwidth of single socket is
-    270Mcacheline/s. Each context is 4 cachelines: 1 for worker struct
-    and 3 stack cachelines. We must read and write every context, so 8
-    cacheline transfers per context switch. This asymptotically
-    approaches 33.75Mcontexts/s as you increase the number of contexts
-    (fewer and fewer in L3). Note that only 4+ cores reaches full rate
-    because these westmere chips are balanced to not achieve full off-chip bandwidth until 4
-    cores--bdmyers}
-
-\TODO{Yield test results:  latency \& bandwidth.  Discussion of implications of zillions of contexts on robustness of latency tolerance preshadowing ~\ref{sec:scaling}}
-
-\subsection{Application Results}
-To evaluate \Grappa's performance with respect to the XMT, we ran each
-of our three benchmarks on up to 16 nodes of each machine. \Grappa used
-6 cores per node, with the best parameters chosen for each point. In
-some cases, the XMT could not run the benchmark with 2 nodes, so the
-point is omitted. \TODO{rewrite this!}
-\subsubsection{Unbalanced Tree Search}\TODO{rewrite with new results}
-%% UTS: performance comparison
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/uts_scale.pdf}
-    \end{center}
-    \caption{Performance of in-memory unbalanced tree search.}
-    \label{fig:uts_compare}
-\end{figure}
-
-We ran UTS-mem with a geometric 100M-vertex tree
-(T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
-number of vertices visited per second versus number of compute
-nodes.
-\Grappa is 3.2 times faster than the XMT at 16 nodes.
-As we will show later, the performance advantage \Grappa has over XMT increases as more nodes are added.
-The main reason \Grappa performs better is the software-based delegate synchronization obviates the need for the retry-based synchronization that XMT uses.
-
-\subsubsection{Breadth First Search}\TODO{rewrite with new results}
-\begin{figure}[tH]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/bfs_performance}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:bfs-performance} BFS performance}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and
-$2^{29}$ edges (25 GB of data). Figure~\ref{fig:bfs-performance} shows
-our performance in terms of graph edges traversed per second. The XMT
-is 2.5 times faster than \Grappa at 16 nodes.  Performance does scale at a constant rate for \Grappa, suggesting that adding more nodes will increase performance.
-
-\subsubsection{Approximate Betweenness Centrality}\TODO{rewrite with new results}
-\begin{figure}[tH]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/centrality_performance}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:centrality-performance} Centrality performance}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-We ran Betweenness Centrality on the same scale 25 Kronecker graph as
-we did for BFS. Figure~\ref{fig:centrality-performance} shows our
-performance in terms of graph edges traversed per second. At 16 XMT
-processors/cluster nodes, the XMT is 1.75 times faster than \Grappa.
-
-% Got rid of this discussion of UTS by itself, tried to work the highlights in below
-%\paragraph{UTS-Mem}
-%We ran UTS-Mem on \Grappa and the XMT with a geometric 1.6B-vertex tree
-%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
-%nodes -- the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
-
-%\Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
-%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, \Grappa adds 1.4 Mvert/s/node.
-%The XMT scales at 850 Kvert/s/node, until it plateaus. \Grappa keeps
-%scaling up through 128 nodes, although scaling
-%declines because of the unscalability of our aggregation mechanism as
-%number of network endpoints increases. 
-%
-%Despite our efforts to tune the UTS implementation specific to the 
-%XMT, performance does not scale well with increasing processor count,
-%flattening out around 60 processors.  When we increase the size of
-%the tree from 100M to 4.2B, we find that performance does not improve,
-%suggesting that performance is not limited by task parallelism.
-%Cray's performance tools show an increasing number of memory
-%retry operations for failed synchronization operations generated by
-%the runtime, which create network contention.
-%
-To determine how \Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
-For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
-For \Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
-All of the benchmarks continue to improve out to 128 nodes for \Grappa.
-UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while \Grappa handles this by suspending tasks until messages return.
-For BFS and Centrality, the XMT scales approximately a constant factor better than \Grappa.
-We attribute this to a limitation in the current aggregator design and network stack that \Grappa uses.
-This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).
-Ironically, this limitation makes \Grappa applications compute-bound instead of network-bound.
-Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
-
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/scaling_cropped.pdf}
-    \end{center}
-    \caption{Scaling number of nodes: \Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
-    \label{fig:uts_threshold}
-\end{figure}
-
-\subsection{Scaling}\label{sec:scaling} \TODO{rewrite with new results}
-
-\subsubsection{Network Aggregation Performance and Robustness}
-
-\begin{figure}[htb]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/aggregator_ping}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:aggregator-ping} Bandwidth versus message size
-    unidirectional ping test for \Grappa with aggregation, \Grappa with
-    raw GASNet messages, and MPI. Aggregation provides an 11x
-    bandwidth benefit at our common operating point.}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-To evaluate the benefits of network aggregation, we ran two experiments.
-First, we ran a simple unidirectional ping test to see the maximum
-benefit the aggregator can provide in terms of improved network
-efficiency. Second, we ran BFS with the aggregator disabled in order to
-measure its benefit on an application.
-
-To implement the ping test, we wrote a simple \Grappa application where
-the cores of one node send messages as fast as possible to the cores
-of another node. We vary the size of the payload up the maximum
-payload size supported by the aggregator (nearly 4KB). Each core has a
-single task sending to a single destination, so this is a best case
-scenario for the aggregator. To see the benefit of the aggregator, we
-added a bypass that lets us send messages directly through GASNet. We
-also compare against the OSU \texttt{osu\_mbw\_mr} benchmark
-\cite{osu:mpi}  compiled against OpenMPI 1.5.3; this
-benchmark has the same pattern of communication but doesn't have the
-overhead of \Grappa's context switching.
-
-The results are shown in Figure~\ref{fig:aggregator-ping}. There are
-two key observations.
-
-First, small message performance against the existing libraries is, as expected, poor. The MPI application test shows us that peak per node
-bandwidth supported by our infiniband card is 3.4GB/s. This is
-achievable only with large messages; we must send 16KB packets to get
-within 5 percent of peak bandwidth. But in our benchmarks, we saw
-average message between 32 and 64 bytes. At 32 bytes, the MPI test is
-using less than 7 percent of its peak bandwidth. \Grappa sending
-messages directly through GASNet uses less than 3 percent of the peak
-bandwidth.
-
-Second, aggregation has the potential to improve this situation by an
-order of magnitude. With aggregation, \Grappa is able to send 32-byte
-messages over 12 times faster than using GASNet directly. This is a
-more respectable 32 percent of peak bandwidth. Due to expedient design
-decisions, \Grappa's aggregator limits its aggregation to 4KB; this
-limits its peak achievable bandwidth to 75 percent of the actual
-peak.
-
-This comparison is the best possible case for the aggregator. In
-order to verify that the aggregator still has value on actual
-applications at scale, we ran a small (100M node tree) UTS-Mem
-with the aggregator disabled, on 16~nodes.
-Figure~\ref{fig:no-aggregation-uts}. At this configuration, the aggregator
-improves our application performance by 10x.
-
-\begin{figure}[htb]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/no_aggregation_uts.pdf}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:no-aggregation-uts} Performance of UTS on 16
-      nodes with and without \Grappa's aggregation.}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-\subsection{Scaling}
-% Got rid of this discussion of UTS by itself, tried to work the highlights in below
-%\paragraph{UTS-Mem}
-%We ran UTS-Mem on \Grappa and the XMT with a geometric 1.6B-vertex tree
-%(T1XL) and a geometric 4.2B-vertex tree (T1XXL), using up to 128
-%nodes -- the maximum we had available for each. \Grappa results are for 5 cores per node. \Grappa with 20 machines is faster than the entire XMT of 128 processors.
-%\Grappa achieves \checkme{188Mvert/s} with 128 nodes and the XMT
-%achieves only 50Mvert/s, plateauing at 60 nodes. Beyond 90 nodes, \Grappa adds 1.4 Mvert/s/node.
-%The XMT scales at 850 Kvert/s/node, until it plateaus. \Grappa keeps
-%scaling up through 128 nodes, although scaling
-%declines because of the unscalability of our aggregation mechanism as
-%number of network endpoints increases. 
-%
-%Despite our efforts to tune the UTS implementation specific to the 
-%XMT, performance does not scale well with increasing processor count,
-%flattening out around 60 processors.  When we increase the size of
-%the tree from 100M to 4.2B, we find that performance does not improve,
-%suggesting that performance is not limited by task parallelism.
-%Cray's performance tools show an increasing number of memory
-%retry operations for failed synchronization operations generated by
-%the runtime, which create network contention.
-%
-To determine how \Grappa's performance scales compared to the performance of the entire XMT, we ran a set of experiments up to all 128 XMT processors and 128 cluster nodes.
-For the XMT, the number of allowed processors was varied up to the entire machine, with some minor tuning of stream parameters needed to get optimal performance.
-For \Grappa, parameters such as cores per node, aggregator timeouts, and parallel threshold were tuned to get the best performance for each node count.
-All of the benchmarks continue to improve out to 128 nodes for \Grappa.
-UTS continues to fare better than the XMT with large node counts, with the XMT appearing to plateau at 60 processors due to contention from synchronization retries, while \Grappa handles this by suspending tasks until messages return.
-For BFS and Centrality, the XMT scales approximately a constant factor better than \Grappa.
-We attribute this to a limitation in the current aggregator design and network stack that \Grappa uses.
-This limits the practical number of cores we can use to 6 per node (adding more cores per node \emph{decreases\/} performance).
-Ironically, this limitation makes \Grappa applications compute-bound instead of network-bound.
-Work is ongoing to rework the Infiniband driver stack and aggregation interface to remove this limitation and improve aggregation addressing using local routing.
-
-\begin{figure}[ht]
-    \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/scaling_cropped.pdf}
-    \end{center}
-    \caption{Scaling number of nodes: \Grappa continues to perform significantly better than XMT for UTS but scales a constant factor slower than XMT for BFS (4x slower) and Centrality (2x slower). }
-    \label{fig:uts_threshold}
-\end{figure}
-
-\subsection{Sensitivity}
-
-\paragraph{Aggregator timeout}
-
-\begin{figure}[htb]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/flushticks_sweep}
-\begin{minipage}{0.95\columnwidth}
-  \caption{\label{fig:bfs-sweep-flushticks} Sensitivity to aggregation delay}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-One of the key parameters of the aggregator is the message
-timeout. All messages that are queued must eventually be sent in order
-to ensure progress. In the best case, we are able to aggregate enough
-messages to fill an aggregation buffer and cause it to be sent, but as
-we scale up, the average rate of messages heading to a common
-destination decreases, and this gets harder. To bound the problem, the
-aggregator includes a timeout. Any packet waiting this long is sent
-the next time the communications layer is serviced.
-
-Figure~\ref{fig:bfs-sweep-flushticks} shows a sweep of this parameter
-for UTS, BFS, and Betweenness Centrality on 16 nodes, using the
-datasets described previously. The maximum number of workers is fixed
-at 2048. All the benchmarks show a performance peak with a 2
-millisecond timeout; at this point we are delaying long enough to
-aggregate the largest packets we can; setting the parameter higher
-causes tasks to wait longer for responses, but few new requests are
-being generated.
-
-\begin{figure}[htb]
-\begin{center}
-  \includegraphics[width=0.95\columnwidth]{figs/worker_sweep}
-\begin{minipage}{0.95\columnwidth} 
-  \caption{\label{fig:bfs-sweep-workers} Sensitivity to maximum active tasks}
-\end{minipage}
-\vspace{-3ex}
-\end{center}
-\end{figure}
-
-\subsubsection{Number of active tasks} \TODO{possibly eliminate}
-
-When a task issues a request that requires a response, it blocks to
-allow other tasks to utilize its core. These tasks may also block. To
-support the many milliseconds of latency aggregation adds, we need to
-support many thousands of blocked tasks. One of the key parameters of
-the runtime is the number of blocked tasks allowed; we need enough to
-cover the network and aggregation latency, but too many running tasks
-can add extra latency as they all must be multiplexed onto the same core.
-
-Figure~\ref{fig:bfs-sweep-workers} shows a sweep of the maximum number
-of active tasks (workers) per core for each of our three benchmarks on
-16 nodes. The aggregator timeout is set at 1 ms for UTS and 2 ms for
-BFS and Betweenness Centrality. The performance peak shifts in this
-case, with UTS peaking at 1536 workers, BFS peaking at 2048 workers,
-and Betweenness Centrality peaking at 3072 workers. This is the point
-where we have enough workers to cover the latency of aggreation. The
-different values reflect the different amounts of work done by a task
-in each benchmark; UTS does the least, while Betweenness Centrality does the most.
-
-%\subsubsection{Work stealing parameters}
-%
-%\paragraph{Chunk size}
-%
-%It is important to steal multiple tasks at a time to both amortize the
-%cost of stealing over the network and to spread out work quickly in a
-%large system. Figure~\ref{fig:ut_chunksize} shows performance and
-%stealing statistics for UTS on \checkme{30} nodes as we increase the stealing chunk size. Recall
-%that a thief will take a number of tasks equal to the minimum of half
-%the available work or the chunk size; steals fail only when the victim
-%has fewer than 2 available tasks. As the scheduler is allowed to
-%steal more work beyond 1 task, we see that performance increases up to 6x. This
-%shows that the heuristic of stealing the oldest task from victims is
-%insufficient alone when a tree-structured computation is imbalanced,
-%as observed in \cite{UTS}. By observing sampled state in the execution
-%trace, we find that a chunk size as low as 1 allows stealing to
-%spread the load evenly across the cluster but cores spend much time
-%underutilized as multiple workers wait for steal replies that
-%utlimately return little new work.
-%
-%Performance plateaus before maximum steal amount is limited by the
-%size of the victim's task queue. This indicates that artificially limiting steals
-%to \checkme{128} tasks does not limit performance. Although a lower
-%chunk size limits how quickly work spreads, for sufficient chunk size,
-%the heuristic of stealing the oldest tasks from victims in tree-based computations allows for
-%stolen work to expand quickly.
-
-%% UTS: chunk size
-%\begin{figure}[ht]
-%    \begin{center}
-%      \includegraphics[width=0.5\textwidth]{figs/uts_chunksize.pdf}
-%    \end{center}
-%    \caption{Performance of UTS-Mem with varying maximum chunk size of
-%    steals, run with 30 nodes, 6 cores per node, 4000 workers,
-%    \checkme{6M flush ticks}}
-%    \label{fig:uts_chunksize}
-%\end{figure}
-
-%\TODO{(difference with BFS)}
-
-\subsubsection{Parallel loop threshold}
-
-%\begin{figure}[htb]
-%\begin{center}
-%  \includegraphics[width=0.95\columnwidth]{figs/bfs_sweep_threshold}
-%\begin{minipage}{0.95\columnwidth}
-%  \caption{\label{fig:bfs-sweep-threshold} Sensitivity to parallel loop threshold. Note the log scale.}
-%\end{minipage}
-%\vspace{-3ex}
-%\end{center}
-%\end{figure}
-
-Parallel overhead -- in the form of context switches, task spawns, and
-synchronization -- can reduce the performance benefit of parallelism.
-\Grappa sees a benefit to limiting the amount of parallelism created by
-a recursive loop decomposition. The parallel loop threshold (``parallel
-granularity'') parameter tells the runtime when to stop creating new tasks and just execute iterations
-sequentially. This allows us to amortize the overhead of task
-creation. In addition, assigning sequential iterations to a single
-task provides the potential to exploit locality when data for adjacent iterations is
-also adjacent in memory. The ability to exploit this locality that
-exists in the application is an important advantage. We found that in UTS and BFS, increasing the
-threshold from 1 up to 8 or 16, respectively, increases performance by
-more than 60\%.
-
-% uts threshold
-%\begin{figure}[ht]
-%    \begin{center}
-%      \includegraphics[width=0.5\textwidth]{figs/uts_threshold.pdf}
-%    \end{center}
-%    \caption{Performance of UTS-Mem with varying parallel loop
-%        threshold, run with 30 nodes, 6 cores per node, 4000 workers,
-%    \checkme{6M flush ticks}}
-%    \label{fig:uts_threshold}
-%\end{figure}
-
-\subsection{Summary}\TODO{this is a placeholder: do we need this section?}
-
-\paragraph{Context-switch overhead.} Should we also compare with other packages? (Maybe Capriccio, QThreads?, or even real OS threads?)
-
-\paragraph{Latency.} Measure remote data access latency with and without aggregation turned on.
-
-\paragraph{Aggregated message sizes.} Characterization of the resulting message sizes with aggregation. Right now we only have message size vs. bandwidth.
-
-\paragraph{Utilization.} CPU utilization, Memory, Network. It would be great to answer the question of where is our bottleneck right now. Amount of concurrency with and without aggregation.
-
-\paragraph{Memory accesses.} Rate of accesses to remote data. Rate of delegate ops. Show limit with GUPs.
-
-%Retries are performed by the memory controller when remote synchronization operations fail to find the full-bit associated with each memory location in the unavailable state.
-Retries are issued at low priority relative to new memory operations issued by the processor by other contexts, so they consume what would otherwise be unused injection bandwidth.
-On a full-bandwidth system such as the MTA-2, retries have no impact on the progress of tasks other than their own.
-On a Cray XMT, network bandwidth is limited, so retries create congestion.
-In comparison, \Grappa performs synchronization without retries, delaying responses at the receiving end until ready to notify the sender to proceed.
-This saves bandwidth and permits scaling of tasks performing synchronization even on low injection rate networks.
-
-%\begin{figure*}[ht]
-%    \begin{minipage}{0.3\linewidth}
-%        \centering
-%        \includegraphics[width=\textwidth]{figs/chunksize-uts.pdf}
-%        \caption{chunksize caption}
-%        \label{fig:chunksize-uts}
-%    \end{minipage}
-%    \begin{minipage}{0.3\linewidth}
-%        \centering
-%        \includegraphics[width=\textwidth]{figs/workers-uts.pdf}
-%        \caption{workers caption}
-%        \label{fig:workers-uts}
-%    \end{minipage}
-%    \begin{minipage}{0.3\linewidth}
-%        \centering
-%        \includegraphics[width=\textwidth]{figs/thresh-uts.pdf}
-%        \caption{threshold caption}
-%        \label{fig:thresh-uts}
-%    \end{minipage}
-%\end{figure*}
-
-}
+Minimum-sized messages can be created and enqueued at a rate of 16M/s,
+serialized into an aggregation buffer at 32M/s, and deserialized from
+an aggregation buffer at 210M/s.  Together, these rates limit us to
+$~$10M/s per core in the best case. In a blocking application, the
+context switch costs (50ns, or 20M/s) bring this maximum down to
+7M/s. In practice, \Grappa does not achieve this maximum; the current
+aggregator design communicates inefficiently between cores within a
+node, limiting our message rate to to $~$1M/s per core with messages
+of common sizes. Future work will reduce this cost.

### 2013-07-24 doc/asplos14/intro.tex
new data, partial characterization

@@
 The communication layer helps \Grappa achieve over 1 billion random updates per second across 64 system nodes.
-\TODO{update with new performance numbers}.
-% Comparing \Grappa and native MPI implementations for more complex kernels is nuanced.
-Some applications use MPI in a straightforward manner, such as GUPS, PageRank, and Unbalanced Tree Search, for which \Grappa is 2-5X faster.
-On the other hand, some MPI implementations of irregular applications recognize its limitations and are written specially to aggregate messages and do some communication asynchronously, performing much the same work \Grappa does behind the scenes, but without paying the cost of generality imposed by context switches or message serialization, to name a few.
-We see this in our comparisons for Breadth First Search and Integer Sort, where the MPI implementation is \checkme{6-10X} faster.
-
-%When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.
-% \TODO{appA and tuned compared to appB and untuned is not ideal}
+\Grappa provides performance approaching and in some cases exceeding that of the custom-hardware XMT, as well as MPI and UPC implementations of the same benchmarks.
+When we compare \Grappa to the XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.
+When compared with MPI and UPC, \Grappa is between 11X faster and 5.4X slower depending on the benchmark.
+We find that the benchmarks that are faster contain specialized implementations of features \Grappa provides generally.

### 2013-07-24 doc/asplos14/evaluation.tex
add latency exploration

@@
     \begin{center}
-      \includegraphics[width=0.5\textwidth]{results/plot_concurrency.pdf}
+      \includegraphics[height=0.5\textwidth]{results/plot_concurrency.pdf}
     \end{center}
@@
 \paragraph{How much concurrency does \Grappa require? How much latency does it add?}
-Figure~\ref{fig:grappa-concurrency} shows 
-UTS blocks at ea
-\todo{FINISH}
+\Grappa depends on concurrency to cover the latency of aggregation and
+remote communication. How much is required for good performance?
+
+Figure~\ref{fig:grappa-concurrency} shows a 48-node, 16-core-per-node
+run of UTS, varying the number of concurrently executing tasks on each
+core. The top pane shows the overall throughput of the tree search.
+The middle pane shows average blocking delegate operation latency in
+microseconds. The bottom pane shows idle time; that is, the fraction
+of the time the scheduler could not find a ready task to run.
+
+We can observe three things from this figure. First, above 512
+concurrently executing tasks per core, idle time is practically zero:
+these tasks generate requests fast enough to cover the latency of
+aggregation and communication. This matches the results seen in the
+throughput plot; throughput peaks at 512 workers and gradually
+decreases after that due to the overhead of unnecessary context
+switches. Finally, we see that with 512 workers, the average
+per-request latency is $~$1.8ms.
 
@@
 
-\todo{FINISH}
+\TODO{FINISH}
 Our current

### 2013-07-24 doc/asplos14/evaluation.tex
scaling

@@
 
-\TODO{FINISH}
-Our current
+\begin{figure}[ht]
+    \begin{center}
+      \includegraphics[height=0.5\textwidth]{results/plot_scaling.pdf}
+    \end{center}
+    \caption{Scaling for BFS, IntSort, and Pagerank. Refer to Figure~\cite{fig:grappa-gups} for GUPS scaling and Figure~\cite{fig:grappa-uts} for UTS scaling}
+    \label{fig:grappa-scaling}
+\end{figure}
+
+Figures~\cite{fig:grappa-scaling}, \cite{fig:grappa-gups}, and
+\cite{fig:grappa-uts} show scaling out to 64 nodes. Grappa scales best
+on BFS and worst on IntSort, but is in general able to make use of
+more nodes. Unfortunately memory limitations in our current network
+library keep us from exploring scaling beyond 64 nodes; this will be
+addressed in future work.
+
+In the limit, aggregation does not scale: the time it takes to
+aggregate enough random requests to build a buffer of reasonable size
+scales with the size of the cluster. However, we believe our current
+approach will work for clusters with hundreds of nodes. In the future,
+we will explore hierarchical, collective techniques to aggregate
+requests from multiple nodes as well; we believe this can apply to
+clusters with thousands of nodes.
 

### 2013-07-24 doc/asplos14/evaluation.tex
scaling

@@
 layer. Application messages refer to those issued by the user code;
-network message refer to the aggregated packes sent over the wire. The
+network message refer to the aggregated packets sent over the wire. The
 data show that most application messages are only a handful of bytes,
@@
 Nevertheless, even in UTS, which depends on work-stealing for
-performance, steals are an infrequent occurance and account for a
+performance, steals are an infrequent occurrence and account for a
 small fraction of the execution time.
@@
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
     \begin{center}
-      \includegraphics[height=0.5\textwidth]{results/plot_scaling.pdf}
+      \subfloat[BFS]{ \includegraphics[height=0.25\textwidth]{results/plot_scaling_bfs.pdf} }\qquad
+      \subfloat[IntSort]{ \includegraphics[height=0.25\textwidth]{results/plot_scaling_intsort.pdf} }\qquad
+      \subfloat[Pagerank]{ \includegraphics[height=0.25\textwidth]{results/plot_scaling_pagerank.pdf} }\qquad
     \end{center}
-    \caption{Scaling for BFS, IntSort, and Pagerank. Refer to Figure~\cite{fig:grappa-gups} for GUPS scaling and Figure~\cite{fig:grappa-uts} for UTS scaling}
+    \caption{Scaling for BFS, IntSort, and Pagerank. Refer to Figure~\ref{fig:grappa-gups} for GUPS scaling and Figure~\ref{fig:grappa-uts} for UTS scaling}
     \label{fig:grappa-scaling}
-\end{figure}
+\end{figure*}
 
-Figures~\cite{fig:grappa-scaling}, \cite{fig:grappa-gups}, and
-\cite{fig:grappa-uts} show scaling out to 64 nodes. Grappa scales best
+Figures~\ref{fig:grappa-scaling}, \ref{fig:grappa-gups}, and
+\ref{fig:grappa-uts} show scaling out to 64 nodes. Grappa scales best
 on BFS and worst on IntSort, but is in general able to make use of

### 2013-07-24 doc/asplos14/results/plot_scaling_uts.R
Incorporate new UTS plot, add todo to rewrite text to match

@@ -38,7 +38,7 @@ g <- ggplot(d.sub,
   scale_color_discrete(name="Tree:")+
   # geom_point()+
   # geom_smooth(fill=NA)+
-  geom_line(size=1.6)+
+  geom_line(size=1.4)+
   # geom_text(size=3, hjust=-0.4, vjust=0.3)+
   # facet_grid(variable~scale, scales="free", labeller=label_pretty)+
   expand_limits(y=0)+
@@ -47,4 +47,5 @@ g <- ggplot(d.sub,
   sosp_theme
 g
 
-ggsave(plot=g, filename="plot_scaling_uts.pdf", width=3.5, height=3.5)
+ggsave(plot=g, filename="plot_scaling_uts.pdf", width=4.25, height=4.25)
+I

### 2013-07-24 doc/asplos14/evaluation.tex
Incorporate new UTS plot, add todo to rewrite text to match
eval update

@@
 
-The goal of our evaluation is to show that the core pieces of the \Grappa
+The goal of our evaluation is to understand whether the core pieces of the \Grappa
 runtime system, namely our tasking system and the global memory/communication
-layer work as expected and together are able to efficiently run irregular
+layer work as expected and whether together they are able to efficiently run irregular
 applications. We evaluate \Grappa in three basic steps:
@@
 graph kernel that stresses communication and concurrency together.
-\TODO{compare to key-value stores?}
+%\TODO{compare to key-value stores?}
 
 \item We characterize system behavior, including
-profiling where execution time goes, how aggregation affects message size and
-rates, and \checkme{the relationship between aggregation and
-    concurrency.}
+profiling where execution time goes, and how aggregation affects message size and
+rates.
+
+%, and \checkme{the relationship between aggregation and
+%    concurrency.}
 
@@
 
+\comment{
 \TODO{Grappa is well-suited to applications where there is
@@
     Perhaps reference ``MPI on millions of cores'' type work}
-
+}
 \end{itemize}
@@
 we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).
-\TODO{based on this and GUPS, our message rate will be limited by
-    context switching for small+single messages. One experiment is
-    GUPs with context switching, where we vary number of reads/suspend}
+%\TODO{based on this and GUPS, our message rate will be limited by
+%    context switching for small+single messages. One experiment is
+%    GUPs with context switching, where we vary number of reads/suspend}
 
-In summary, our context switch engine is able to efficiently sustain very high
+In summary, our tasking layer is able to efficiently sustain very high
 concurrency and, as we will show later, the amount of concurrency sustained is
@@
 measures cluster-wide random access bandwidth.  Read-modify-write
-updates are dispatched at random to a large array. This benchmark
+updates are dispatched at random to a global large array. This benchmark
 stresses the communication layer of \Grappa separately from the
@@
 
-\TODO{go into more detail about what Grappa is doing when executing
-    GUPs. Specifically note what the rate per core is, what the
-    aggregation and deaggregation rates are. Show on the plot the
-    message size distribution (some box plot)}
+%\TODO{go into more detail about what Grappa is doing when executing
+%    GUPs. Specifically note what the rate per core is, what the
+%    aggregation and deaggregation rates are. Show on the plot the
+%    message size distribution (some box plot)}
 
@@
 the original benchmark. T1 trees are very shallow and wide (i.e., significant
-parallel slack \TODO{cite Valiant and say more somewhere about BSP}), while T3 trees are very deep (i.e., little parallel slack).
+parallel slack \comment{\TODO{cite Valiant and say more somewhere about BSP}}), while T3 trees are very deep (i.e., little parallel slack).
 When the access to each vertex is a random access, the critical path to search
@@
 pays execution and space overhead in order to support aggregation of
-arbitrary closures; the messages \Grappa aggregates include the two
+arbitrary messages; the messages \Grappa aggregates include the two
 8-byte vertex IDs along with 8 bytes of deserialization information as
@@
 aggregation buffer while \Grappa executes additional code to serialize
-and deserialize the messages. \TODO{Include message counts, size, and
-  rates; code size too?}.
+and deserialize the messages.
 
@@
 message at the destination. We benchmarked each of these steps
-individually to find the bottleneck.
+individually to shed light on the mechanism.
 
@@
 node, limiting our message rate to to $~$1M/s per core with messages
-of common sizes. Future work will reduce this cost.
+of common sizes. Future work will reduce this cost.  \comment{Ack! -Mark}

### 2013-07-24 doc/asplos14/evaluation.tex
dont ever put an explanation point in a comment block

@@
 
-\comment{
-\TODO{Grappa is well-suited to applications where there is
-    abundant data parallelism but unpredictable or imbalanced data
-    access. BSP is efficient where there is sufficient parallel slack.
-    Although Grappa loses some asynchrony by relying heavily on
-    aggregation, it still provides more flexibility for sections of the
-    computation to proceed more quickly. }
-
-\TODO{how much concurrency is needed on apps. Our basic plot of
-    performance vs workers (with and without aggro?)}
-\TODO{relationship between user-thread-visible latency and aggregate throughput.}
-\TODO{characterization of aggregation message size vs nodes}
-\TODO{show our scalability plots; if they don't scale, show
-    specifically why. It is okay if its because aggregation stops
-    working after a point if we can discuss future scalability.
-    Perhaps reference ``MPI on millions of cores'' type work}
-}
+%\comment{
+%\TODO{Grappa is well-suited to applications where there is
+%    abundant data parallelism but unpredictable or imbalanced data
+%    access. BSP is efficient where there is sufficient parallel slack.
+%    Although Grappa loses some asynchrony by relying heavily on
+%    aggregation, it still provides more flexibility for sections of the
+%    computation to proceed more quickly. }
+%
+%\TODO{how much concurrency is needed on apps. Our basic plot of
+%    performance vs workers (with and without aggro?)}
+%\TODO{relationship between user-thread-visible latency and aggregate throughput.}
+%\TODO{characterization of aggregation message size vs nodes}
+%\TODO{show our scalability plots; if they don't scale, show
+%    specifically why. It is okay if its because aggregation stops
+%    working after a point if we can discuss future scalability.
+%    Perhaps reference ``MPI on millions of cores'' type work}
+%}
 \end{itemize}
@@
 we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).
-%\TODO{based on this and GUPS, our message rate will be limited by
-%    context switching for small+single messages. One experiment is
-%    GUPs with context switching, where we vary number of reads/suspend}
 
@@
 the original benchmark. T1 trees are very shallow and wide (i.e., significant
-parallel slack \comment{\TODO{cite Valiant and say more somewhere about BSP}}), while T3 trees are very deep (i.e., little parallel slack).
+parallel slack), while T3 trees are very deep (i.e., little parallel slack).
 When the access to each vertex is a random access, the critical path to search
@@
     \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/uts_scale.pdf}
+      \includegraphics[width=0.36\textwidth]{results/plot_scaling_uts.pdf}
     \end{center}
     \caption{Vertices per second in UTS on \Grappa as the number of nodes grows.}
+    %\TODO{modify discussion inline to match new plot}}
     \label{fig:grappa-uts}
@@
 \end{center}
-\caption{Internal runtime metrics, for 64-node, 16-core-per-node benchmark runs \TODO{Finish} }
+\caption{Internal runtime metrics, for 64-node, 16-core-per-node benchmark runs}
 \label{tab:grappa-metrics}
@@
 node, limiting our message rate to to $~$1M/s per core with messages
-of common sizes. Future work will reduce this cost.  \comment{Ack! -Mark}
+of common sizes. Future work will reduce this cost.

### 2013-07-24 doc/asplos14/conclude.tex
back to space limit

@@
 at a performance cost but eases the application developer's
-task. Moreover, significant tuning has yet to be done on the \Grappa
-runtime system.  Yet when cost-performance is considered, mass-market
-x86-based cluster hardware makes \Grappa a highly attractive option.
+task. Moreover, there are significant tuning opportunities the \Grappa
+runtime system.  When cost-performance is considered, \Grappa on mass-market
+x86-based clusters is a highly attractive option.

### 2013-07-24 doc/asplos14/evaluation.tex
fixed figure rotation
back to space limit

@@
 
-\begin{table*}[htb]
+\begin{table}[htb]
+\small
 \begin{center}
+\setlength{\tabcolsep}{1pt}
 \begin{tabular}{l|c|c|c|c|c}
@@
                                &   GUPS            &   BFS   & IntSort & UTS & Pagerank \\ \hline
-Application message rate/core  & 984K/s            &  983K   &           \\
+App. message rate/core (K/s)  & 984            &  983   &      &     \\
 %Application message bytes     & 31.8 $\pm$ 1.2    & \\
-Avg. Application message bytes & 31.8              & 33.9    &       \\
-Network bandwidth per node     & 478MB/s           & 511MB/s &          \\
+Avg. App. message bytes & 31.8              & 33.9    &      & \\
+Network BW per node (MB/s)    & 478           & 511 &    &      \\
 %Network message bytes         & 23.2K $\pm$ 12.4K & \\ \hline
-Avg. Network message bytes     & 23.2K             & 4.3K    &     \\ \hline
-Avg. active tasks/core         & 0.9               & 58.2    &    \\
-Max. active tasks/core         & 1                 & 128     &     \\
-Avg. ready queue length/core   & 2.3               & 6.1     &      \\
-Avg. Context switch rate/core & 34.2K/s           & 539K/s  &       \\ 
-Steal count                    & 0                 & 0       &       \\
+Avg. Network message bytes     & 23.2K             & 4.3K    & &    \\ \hline
+Avg. active tasks/core         & 0.9               & 58.2    &  &  \\
+Max. active tasks/core         & 1                 & 128     &  &   \\
+Avg. ready queue length/core   & 2.3               & 6.1     &  &    \\
+Avg. Ctx switch rate/core (K/s) & 34.2           & 539  &     &  \\ 
+Steal count                    & 0                 & 0       &  &     \\
 \end{tabular}
@@
 \label{tab:grappa-metrics}
-\end{table*}
+\end{table}
 
@@
     \begin{center}
-      \includegraphics[height=0.5\textwidth]{results/plot_concurrency.pdf}
+      \includegraphics[height=0.5\textwidth, angle=90]{results/plot_concurrency.pdf}
     \end{center}
+    \vspace*{-0.25in}
     \caption{Throughput, request latency, and idleness as number of concurrent workers is varied}

### 2013-07-24 doc/asplos14/evaluation.tex
moved UPC data to text only, more pruning

@@
 \begin{center}
-\begin{tabular}{l|c|c|c|c}
-         & \Grappa & XMT   & MPI  & UPC \\ \hline
-GUPS     & 1       & 2.23  & 0.11 & -- \\ 
-BFS      & 1       & 1.63  & 3.52 & -- \\ 
-IntSort  & 1       & 3.59  & 5.36 & -- \\
-UTS (T1) & 1       & 0.38  & --   & 0.09 \\ 
-Pagerank & 1       & 4.35  & 4.87 & -- \\ 
+\begin{tabular}{l|c|c|c}
+         & \Grappa & XMT   & MPI   \\ \hline
+GUPS     & 1       & 2.23  & 0.11 \\ 
+BFS      & 1       & 1.63  & 3.52 \\ 
+IntSort  & 1       & 3.59  & 5.36 \\
+UTS (T1) & 1       & 0.38  & --   \\  % 0.09 for UPC
+Pagerank & 1       & 4.35  & 4.87 \\ 
 \end{tabular}
 \end{center}
-\caption{Comparing \Grappa with XMT, optimized MPI and UPC. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
+\caption{Comparing \Grappa with XMT and optimized MPI. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
 \label{tab:grappa-comparisons}
@@
 
-The UPC UTS implementation includes an implementation of work-stealing
-but has limited ability to overlap communication with
-computation. Furthermore, it must execute heavyweight synchronization
-operations when stealing. \Grappa benefits from its lightweight
-context switching and support for fine-grained synchronization.
+We also compare with an UPC UTS implementation, which is nearly 10$\times$
+\emph{slower} than \Grappa. Our UPC version supports work-stealing but has
+limited ability to overlap communication with computation. Also, it
+employs heavyweight synchronization operations when stealing. \Grappa
+benefits from its lightweight context switching and support for fine-grained
+synchronization.
 
 \paragraph{BFS.}
-The Grappa and MPI BFS implementations are quite similar. They both
+The \Grappa and MPI BFS implementations are quite similar. They both
 use the same graph representation, and they both depend on aggregation

### 2013-07-24 doc/asplos14/evaluation.tex
more edits. more fixes to macro usage

@@
 
-What is more surprising is that Grappa is able to exceed the XMT's
+What is more surprising is that \Grappa is able to exceed the XMT's
 performance on UTS. This is a demonstration of the inflexibility of
@@
 Figures~\ref{fig:grappa-scaling}, \ref{fig:grappa-gups}, and
-\ref{fig:grappa-uts} show scaling out to 64 nodes. Grappa scales best
+\ref{fig:grappa-uts} show scaling out to 64 nodes. \Grappa scales best
 on BFS and worst on IntSort, but is in general able to make use of

### 2013-07-24 doc/asplos14/evaluation.tex
eval uts

--- 2013-07-24 doc/asplos14/evaluation.tex
@@
 the original benchmark. T1 trees are very shallow and wide (i.e., significant
-parallel slack), while T3 trees are very deep (i.e., little parallel slack).
-When the access to each vertex is a random access, the critical path to search
+parallel slack \cite{Valiant:1990}), while T3 trees are very deep (i.e., little parallel slack).
+Given that access to each vertex is a random access, the critical path to search
 T3 trees is very long, hence the low performance and scalability. On such
@@
 the average active tasks per core over the search is 775 for T1 and
-only 13 for T3.
+only 13 for T3. Given the lack of parallelism, scaling up only serves
+to reduce throughput by distributing the tree to more machines.
 On the other hand, \Grappa performs and scales very well for T1
 trees.
+
+We compared this to a in-memory modification of the UPC implementation
+of UTS \cite{UTS}, where each core hosts a UPC ``THREAD''. Berkeley
+UPC (BUPC) is built upon GASNet, like Grappa, but it uses RDMA transfers directly
+and the compiler overlaps remote communication when possible. We expect BUPC to
+perform similarly on the T3 tree, where there is little parallelism to
+overlap communication and computation, and no better than T3 on T1,
+given the inability to overlap vertices on each core without
+multithreading (a multithreaded variant of UPC has been evaluated in
+\cite{HotSLAW}). Our experiments confirm that BUPC achieves the lower throughput of
+Grappa on T3 for \emph{both} the T1 and T3 trees.
 
@@
 
-\begin{table}[htb]
-\small
+\begin{table*}[htb]
 \begin{center}
-\setlength{\tabcolsep}{1pt}
 \begin{tabular}{l|c|c|c|c|c}
@@
                                &   GUPS            &   BFS   & IntSort & UTS & Pagerank \\ \hline
-App. message rate/core (K/s)  & 984            &  983   &      &     \\
+Application message rate/core  & 984K/s            &  983K   &           \\
 %Application message bytes     & 31.8 $\pm$ 1.2    & \\
-Avg. App. message bytes & 31.8              & 33.9    &      & \\
-Network BW per node (MB/s)    & 478           & 511 &    &      \\
+Avg. Application message bytes & 31.8              & 33.9    &       \\
+Network bandwidth per node     & 478MB/s           & 511MB/s &          \\
 %Network message bytes         & 23.2K $\pm$ 12.4K & \\ \hline
-Avg. Network message bytes     & 23.2K             & 4.3K    & &    \\ \hline
-Avg. active tasks/core         & 0.9               & 58.2    &  &  \\
-Max. active tasks/core         & 1                 & 128     &  &   \\
-Avg. ready queue length/core   & 2.3               & 6.1     &  &    \\
-Avg. Ctx switch rate/core (K/s) & 34.2           & 539  &     &  \\ 
-Steal count                    & 0                 & 0       &  &     \\
+Avg. Network message bytes     & 23.2K             & 4.3K    &     \\ \hline
+Avg. active tasks/core         & 0.9               & 58.2    &    \\
+Max. active tasks/core         & 1                 & 128     &     \\
+Avg. ready queue length/core   & 2.3               & 6.1     &      \\
+Avg. Context switch rate/core & 34.2K/s           & 539K/s  &       \\ 
+Steal count                    & 0                 & 0       &       \\
 \end{tabular}
@@
 \label{tab:grappa-metrics}
-\end{table}
+\end{table*}
 

### 2013-07-24 doc/asplos14/conclude.tex

@@
 quickly.  Scaling these applications easily on commodity hardware has
-been a historical challenge. \Grappa is a runtime framework that
+been a historical challenge. \Grappa 
 simplifies this task for software developers and compiler
-writers. This paper describes the \Grappa framework and its three main
-components: a task library, a distributed shared memory system, and a
-network aggregator. \Grappa's key aspect is extreme latency tolerance,
+writers. 
+% This paper describes \Grappa's framework and its main
+% components: a task library, a distributed shared memory system, and a
+% network aggregator. 
+\Grappa's key aspect is extreme latency tolerance,
 which not only hides network latency but also enables the system to
@@
 
-Our evaluation of \Grappa reveals that the core components, scheduling
+Our evaluation of \Grappa shows that the core components, scheduling
 and communication, achieve their design goals.  Thousands of workers
-can be efficiently context switched on a multicore processor, up to
-the DRAM bandwidth.  Aggregating messages enables \Grappa to achieve
+can be efficiently context switched on a multicore processor, limited by DRAM bandwidth.  Aggregating messages enables \Grappa to achieve
 over 1.0~GUPS on 64~nodes.  We also explored four other algorithms:
@@
 sort, comparing performance of these algorithms on \Grappa, the Cray
-XMT, and optimized MPI implementations.
-
-Depending on the benchmark, \Grappa is between 11X faster and 5.4X
+XMT, and optimized MPI implementations. \Grappa is between 11X faster and 5.4X
 slower than MPI and UPC, and between 2.6X faster and 4.3X slower than

### 2013-07-24 doc/asplos14/relatedwork.tex

@@
 well as context switching and scheduling computation. The Active
-Messages~\cite{vonEicken:1992:AMM:139669.140382} work that grew out of
+Messages~\cite{vonEicken92} work that grew out of
 this project inspired our communication layer. One of the conclusions
@@
 While \Grappa is a general runtime system for any large-scale
-concurrent application, it has been designed to perform especially
+concurrent application, it performs especially
 well on graph analysis. Other distributed graph processing frameworks

### 2013-07-24 doc/asplos14/evaluation.tex
remove UPC from places it doesnt belong

@@
 different hardware platform, we also compare \Grappa with optimized MPI
-versions of BFS and GUPS running on the same hardware. Finally, we also
-compare it with UTS written for UPC. We run all experiments with 64 nodes and 16 cores per node.
+versions of BFS and GUPS running on the same hardware. We run all experiments with 64 nodes and 16 cores per node.
 
@@
 
-We also compare with an UPC UTS implementation, which is nearly 10$\times$
-\emph{slower} than \Grappa. Our UPC version supports work-stealing but has
-limited ability to overlap communication with computation. Also, it
-employs heavyweight synchronization operations when stealing. \Grappa
-benefits from its lightweight context switching and support for fine-grained
-synchronization.
+%bdmyers: the below is already covered specially in the UTS section
+%We also compare with an UPC UTS implementation, which is nearly 10$\times$
+%\emph{slower} than \Grappa. Our UPC version supports work-stealing but has
+%limited ability to overlap communication with computation. Also, it
+%employs heavyweight synchronization operations when stealing. \Grappa
+%benefits from its lightweight context switching and support for fine-grained
+%synchronization.
 

### 2013-07-24 doc/asplos14/evaluation.tex
fixed some grappas

@@
 %\comment{
-%\TODO{Grappa is well-suited to applications where there is
+%\TODO{\Grappa is well-suited to applications where there is
 %    abundant data parallelism but unpredictable or imbalanced data
 %    access. BSP is efficient where there is sufficient parallel slack.
-%    Although Grappa loses some asynchrony by relying heavily on
+%    Although \Grappa loses some asynchrony by relying heavily on
 %    aggregation, it still provides more flexibility for sections of the
@@
 
-%\TODO{go into more detail about what Grappa is doing when executing
+%\TODO{go into more detail about what \Grappa is doing when executing
 %    GUPs. Specifically note what the rate per core is, what the
@@
 We compared this to a in-memory modification of the UPC implementation
-of UTS \cite{UTS}, where each core hosts a UPC ``THREAD''. Berkeley
-UPC (BUPC) is built upon GASNet, like Grappa, but it uses RDMA transfers directly
+of UTS \cite{UTS}, where each core hosts a UPC ``THREAD.'' Berkeley
+UPC (BUPC) is built upon GASNet, like \Grappa, but it uses RDMA transfers directly
 and the compiler overlaps remote communication when possible. We expect BUPC to
@@
 \cite{HotSLAW}). Our experiments confirm that BUPC achieves the lower throughput of
-Grappa on T3 for \emph{both} the T1 and T3 trees.
+\Grappa on T3 for \emph{both} the T1 and T3 trees.
 
@@
 In order to put \Grappa's performance into a general context, we compare it
-with XMT running BFS, PageRank, IntSort, GUPS and UTS. Since XMT is a
+with XMT running BFS, PageRank, IntSort, GUPS, and UTS\@. Since XMT is a
 different hardware platform, we also compare \Grappa with optimized MPI
@@
 \end{center}
-\caption{Comparing \Grappa with XMT and optimized MPI. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
+\caption{Comparing \Grappa with XMT and optimized MPI\@. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
 \label{tab:grappa-comparisons}
@@
 hardware optimized for irregular applications. The XMT can
-context-switch every cycle, while \Grappa takes $~$50 ns. The XMT can
+context-switch every cycle, while \Grappa takes $~$50ns. The XMT can
 complete 100 million random-access remote reads per second per node in
@@
 XMT-like per-vertex task joins as well as whole-problem joins that are
-a better fit for UTS. Furthermore, \Grappa's work-stealing distributed
+a better fit for UTS\@. Furthermore, \Grappa's work-stealing distributed
 task queue and recursive loop decomposition encourage locality in task
@@
 
-\paragraph{BFS.}
+\paragraph{BFS}
 The \Grappa and MPI BFS implementations are quite similar. They both
 use the same graph representation, and they both depend on aggregation
-for performance. The difference is that the BFS' aggregation is
+for performance. The difference is that the BFS aggregation is
 integrated with the traversal code and specialized for the problem:
@@
 
-\paragraph{GUPS.}
+\paragraph{GUPS}
 The MPI GUPS implementation includes an implementation of aggregation,
@@
 
-\paragraph{Pagerank.}
+\paragraph{Pagerank}
 The MPI Pagerank implementation is built on top of the
@@
 
-\paragraph{IntSort.}
+\paragraph{IntSort}
 Both MPI and \Grappa IntSort implement a bucket sort. For \Grappa,
@@
 
-\paragraph{Summary.}
+\paragraph{Summary}
 
@@
 Figures~\ref{fig:grappa-scaling}, \ref{fig:grappa-gups}, and
-\ref{fig:grappa-uts} show scaling out to 64 nodes. \Grappa scales best
+\ref{fig:grappa-uts} show scaling out to 64~nodes. \Grappa scales best
 on BFS and worst on IntSort, but is in general able to make use of
 more nodes. Unfortunately memory limitations in our current network
-library keep us from exploring scaling beyond 64 nodes; this will be
+library keep us from exploring scaling beyond 64~nodes; this will be
 addressed in future work.

### 2013-07-24 doc/asplos14/evaluation.tex
better description of Grappa advantages in UTS over XMT

@@
 available to all other processors in the system, even though it is
-likely that processor that spawned a task will end up executing it.
+likely the processor that spawned a task will end up executing it.
 
@@
 XMT-like per-vertex task joins as well as whole-problem joins that are
-a better fit for UTS\@. Furthermore, \Grappa's work-stealing distributed
-task queue and recursive loop decomposition encourage locality in task
-execution, so that the system is optimized for the common case of a
-processor executing a task it spawned.
+a better fit for UTS\@. \Grappa's work-stealing distributed
+task queue encourages locality in task
+execution so that the system is optimized for the common case of a
+processor executing a task it spawned. Furthermore, Grappa's parallel
+loop granularity and buffering features provide a disciplined way for the programmer to easily take better
+advantage of spatial locality when it oviously exists,
+specifically when reading a vertex's edgelist.
+>>>>>>> better description of Grappa advantages in UTS over XMT
 

### 2013-07-24 doc/asplos14/evaluation.tex
oops fixed some Grappas

@@
 execution so that the system is optimized for the common case of a
-processor executing a task it spawned. Furthermore, Grappa's parallel
+processor executing a task it spawned. Furthermore, \Grappa's parallel
 loop granularity and buffering features provide a disciplined way for the programmer to easily take better
@@
 specifically when reading a vertex's edgelist.
->>>>>>> better description of Grappa advantages in UTS over XMT
 

### 2013-07-24 doc/asplos14/tasks.tex
fixed broken cite.
workstealing citation
use delegate_write to be consistent

@@
 core. Even if there are many active threads, if they are all suspended on
-long-latency operations, then the core is underutilized. Similarly to
-\cite{the_hierarchical_workstealing} the
-stealing policy must predict whether local tasks will likely
-generate enough new work soon.
+long-latency operations, then the core is underutilized. 
+The
+stealing policy 
+must predict whether local tasks will likely
+generate enough new work soon; a similar problem is addressed in
+\cite{vanNieuwpoort:2001}.
 
@@
   if (node->key == key) {
-    delegate::write(result, node);
+    delegate_write(result, node);
   } else {

### 2013-07-24 doc/asplos14/evaluation.tex
don't rotate concurrent plot
rest of characterization data

@@
                                 &  GUPS  &  BFS & IntSort &   UTS & Pagerank \\ \hline
-App. message rate/core (K/s)  & 984            &  983   &      &     \\
-%Application message bytes     & 31.8 $\pm$ 1.2    & \\
-Avg. App. message bytes & 31.8              & 33.9    &      & \\
-Network BW per node (MB/s)    & 478           & 511 &    &      \\
-%Network message bytes         & 23.2K $\pm$ 12.4K & \\ \hline
-Avg. Network message bytes     & 23.2K             & 4.3K    & &    \\ \hline
-Avg. active tasks/core         & 0.9               & 58.2    &  &  \\
-Max. active tasks/core         & 1                 & 128     &  &   \\
-Avg. ready queue length/core   & 2.3               & 6.1     &  &    \\
-Avg. Ctx switch rate/core (K/s) & 34.2           & 539  &     &  \\ 
-Steal count                    & 0                 & 0       &  &     \\
+App. message rate/core (K/s)    & 984    &  983 &     732 &   411 &      474 \\
+Avg. App. message bytes         & 31.8   & 33.9 &    30.6 &  47.3 &     45.5 \\
+Network BW per node (MB/s)      & 478    & 511  &     343 &   298 &      332 \\
+Avg. Network message bytes      & 23.2K  & 4.3K &   12.8K &  4.2K &     3.0K \\ \hline
+Avg. active tasks/core          & 0.9    & 58.2 &     0.9 & 326.2 &    429.3 \\
+Max. active tasks/core          & 1      & 128  &       1 &   507 &     1024 \\
+Avg. ready queue length/core    & 2.3    & 6.1  &     1.3 & 186.1 &    300.4 \\
+Avg. Ctx switch rate/core (K/s) & 34.2   & 539  &     127 &   543 &      336 \\ 
+Steal attempts/core/s           & 0      & 0    &       0 &  54.4 &        0 \\
+Steal successes/core/s          & 0      & 0    &       0 &  17.0 &        0 \\
 \end{tabular}
@@
     \begin{center}
-      \includegraphics[height=0.5\textwidth, angle=90]{results/plot_concurrency.pdf}
+      % \includegraphics[height=0.5\textwidth, angle=90]{results/plot_concurrency.pdf}
+      \includegraphics[height=0.4\textwidth]{results/plot_concurrency.pdf}
     \end{center}

### 2013-07-24 doc/asplos14/evaluation.tex
Fix up search code example and refer to it from UTS Eval section

@@
 
-\paragraph{Putting it all together with Unbalanced Tree Search in memory
-(UTS).} 
+\paragraph{Putting it all together with Unbalanced Tree Search (UTS).} 
 Figure~\ref{fig:grappa-uts} shows the overall performance of \Grappa running
@@
 we are forced to context switch to tolerate the remote access and
-continue aggregating.
+continue aggregating. Figure~\ref{fig:sample} shows a closely analogous tree search in Grappa.
 
@@
 multithreading (a multithreaded variant of UPC has been evaluated in
-\cite{HotSLAW}). Our experiments confirm that BUPC achieves the lower throughput of
-\Grappa on T3 for \emph{both} the T1 and T3 trees.
+\cite{HotSLAW}). Our experiments confirm that BUPC achieves lower throughput than
+\Grappa for both T1 and T3 trees.
 

### 2013-07-24 doc/asplos14/tasks.tex
Fix up search code example and refer to it from UTS Eval section

@@
 Figure~\ref{fig:sample} shows sample code using \Grappa for a parallel tree
-search. The important aspect to note is that the code looks very similar to
-what would be written for single shared-memory system, without any concern about data locality or communication.
+search.
+Note how the code looks very similar to a recursive search procedure for a shared-memory system, without regard for communication, and \Grappa's parallel loop construct allows easy parallelization of the search.
 
@@
 \begin{lstlisting}[style=grappa]
-class node_t {
-  key_t   key
+class Vertex {
+  Key key
   int64_t numChildren;
-  GlobalAddress<node_t> children;
+  GlobalAddress<Vertex> children;
 };
 
-void search(GlobalAddress<node_t> node, key_t key, GlobalAddress<node_t> result) {
-  // deref of key is expected cache miss
-  if (node->key == key) {
-    delegate_write(result, node);
+void search(GlobalAddress<Vertex> vtx_addr,
+      Key key, GlobalAddress<Vertex> result) {
+  // blocking remote read to get vertex info
+  Vertex vtx = delegate_read(vtx_addr);
+  if (vtx.key == key) {
+    // key found
+    delegate_write(result, vtx);
   } else {
-    parallel_for(0, node->numChildren,
-    [=](int index) {
-      // first deref of children is likely miss
-      search(node->children[index], key); 
+    // spawn stealable tasks for iterations
+    parallel_for(0,vtx.numChildren,[=](int i) {
+      // recursive search
+      search(vtx.children+i, key, result);       
     });
@@
 
-    \begin{minipage}{0.95\columnwidth}
+    \begin{minipage}{\columnwidth}
       \caption{\label{fig:sample} Sample \Grappa code illustrating a
         parallel tree search similar to the unbalanced tree search
-          benchmark we describe later. The labeled cache misses occur
-          if we assume the tree is not layed out in a regular fashion, which
-      models more general graphs.}
+        benchmark we describe later. Children are spread over the system, so each parallel recursive search performs a delegate read to get vertex data.
+      }
     \end{minipage}

### 2013-07-24 doc/asplos14/memory.tex
edits and todos
Rework memory section to use `asyncronous` instead of non-blocking.

@@
 on individual machine-word sized chunks of global memory to the memory system
-itself. Delegates can execute arbitrary non-blocking code, so we use them to perform simple
+itself.
+Delegates can execute arbitrary code, provided they do not block, to ensure communicator workers make progress.
+Provided they touch only memory owned by a single core, we can use them to perform simple
 \emph{read\/}/\emph{write\/} operations to global memory, as well as more complex \emph{read-modify-write\/} operations (e.g., \emph{fetch-and-add\/}).
+We use these primitive operations to implement higher-level synchronization mechanisms such as mutexes, condition variables, and full-empty bits.
 
@@
 
-Delegate operations can be either {\em blocking\/} or {\em non-blocking}.
+Delegate operations can be either {\em blocking\/} or {\em asynchronous}.
 With blocking operations, the task issuing the delegate call blocks until
 the delegate operation completes, which is necessary, for example, to ensure
-that synchronization has finished before continuing. Synchronization in \Grappa
-is implemented using blocking delegate operations. On the other hand, remote data
-accesses often can overlap. In order to avoid
-unnecessary waiting, we support non-blocking delegate operations. For reads,
-we support a ``futures''-like mechanism. Tasks may issue delegate reads in
-parallel and block on the ``promises'' returned by non-blocking delegate
-invocations. Delegate write operations may also be performed as
-non-blocking,
-and since they do not return data, a mechanism to detect completion is needed.
-\Grappa provides a \texttt{GlobalCompletionEvent} synchronization object, which non-blocking operations (including tasks) can be enrolled with and which other tasks can block on to be notified and woken when all enrolled operations in the collection are complete.
-\TODO{augment this with our bulk futures ie global completions}
+that synchronization has finished before continuing. On the other hand, remote data
+accesses often can overlap, and delegates with no return value may not need to block the caller. 
+To avoid unnecessary waiting, we support asynchronous delegate operations. For reads,
+we support a ``futures''-like mechanism which allows tasks to issue reads in parallel and block on the ``promises'' returned.
+Delegate write operations may also be performed asynchronously, but synchronization is still needed to ensure that asynchronous operations have completed.
+\Grappa provides a \texttt{GlobalCompletionEvent} synchronization object, which asyncronous operations (including tasks) can be enrolled.
+Tasks can block on these objects to be woken when all enrolled operations are complete.
+% \TODO{augment this with our bulk futures ie global completions}
 

### 2013-07-24 doc/asplos14/overview.tex
small changes

@@
 First, modern compute cores can execute hundreds of thousands of instructions
-in the few microseconds that it takes for one RDMA read to complete. Second,
+in the few microseconds that it takes for one remote memory read to complete. Second,
 the injection rate of modern commodity networks is low and the bandwidth is
@@
 GASNet~\cite{gasnet} is used as the
-underlying mechanism for remote memory reads and writes using active message
+underlying mechanism for bulk remote memory reads and writes using active message
 invocations, with an off-the-shelf user-mode InfiniBand device driver

### 2013-07-24 doc/asplos14/evaluation.tex
shrink two plots

@@
 runtime system, namely our tasking system and the global memory/communication
-layer work as expected and whether together they are able to efficiently run irregular
+layer, work as expected and whether together they are able to efficiently run irregular
 applications. We evaluate \Grappa in three basic steps:
@@
     \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/context_switch_time.pdf}
+      \includegraphics[width=0.4\textwidth]{figs/context_switch_time.pdf}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=0.5\textwidth]{figs/gups.pdf}
+      \includegraphics[width=0.4\textwidth]{figs/gups.pdf}
     \end{center}

### 2013-07-24 doc/asplos14/method.tex
method changes

@@
 traversal code to create a large tree in memory, and then we traverse the
-in-memory tree. We call this benchmark UTS-Mem, and the timed portion is this
+in-memory tree. In our modified UTS, the timed portion is this
 traversal of the in-memory tree. This in-memory traversal has no knowledge of
@@
 In order to make comparison easier, we do not employ algorithmic improvements for any of these versions, though there are
-many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}. \Grappa can be expected to benefit the same as MPI due to decreased communication.
+many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}; this makes our results difficult to compare with published Graph500 results.
+\Grappa can be expected to benefit the same as MPI due to decreased communication.
 
@@
 \noindent{\bf PageRank:} This is a common centrality metric for
-graphs. Computing PageRank is an iterative algorithm with a common
+graphs. PageRank is an iterative algorithm with a common
 pattern of gather, apply, and scatter on the rank of vertex. The algorithm
@@
 collects statistics about application behavior (packets sent, context
-switches, etc) and these are discussed where appropriate. In addition we use a
-profiled version of \Grappa to report component runtimes, but this profiled
-version is only used to collect this data and a fully optimized version is
-used for the bulk of the presented results.
+switches, etc) and these are discussed where appropriate.

### 2013-07-24 doc/asplos14/evaluation.tex
oops fix Grappa
fix punctuation

@@
 
-\paragraph{User-level context switching.}
+\paragraph{User-level context switching}
 
@@
 
-\paragraph{Global memory and communication.} We measure the performance of
+\paragraph{Global memory and communication} We measure the performance of
 \Grappa's global memory and communication layers using a faithful
@@
 
-\paragraph{Putting it all together with Unbalanced Tree Search (UTS).} 
+\paragraph{Putting it all together with Unbalanced Tree Search (UTS)} 
 Figure~\ref{fig:grappa-uts} shows the overall performance of \Grappa running
@@
 we are forced to context switch to tolerate the remote access and
-continue aggregating. Figure~\ref{fig:sample} shows a closely analogous tree search in Grappa.
+continue aggregating. Figure~\ref{fig:sample} shows a closely analogous tree search in \Grappa.
 
@@
 
-\paragraph{UTS.}
+\paragraph{UTS}
 It is perhaps not surprising that the XMT is faster than

### 2013-07-24 doc/asplos14/method.tex
fix punctuation

@@
 \vspace{0.5ex}
-\noindent{\bf Unbalanced tree search in-memory (UTS-Mem):} Unbalanced Tree Search
+\noindent{\bf Unbalanced tree search in-memory (UTS-Mem)} Unbalanced Tree Search
 (UTS) is a benchmark for evaluating the programmability and performance of
@@
 \vspace{0.5ex}
-\noindent{\bf Breadth-first-search (BFS):} This is the primary kernel for the
+\noindent{\bf Breadth-first-search (BFS)} This is the primary kernel for the
 Graph500 benchmark and is what currently determines the ranking of machines on
@@
 \vspace{0.5ex}
-\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the Cray XMT's early predecessor once held the world speed record~\cite{TeraRecord}.
+\noindent{\bf IntSort} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the Cray XMT's early predecessor once held the world speed record~\cite{TeraRecord}.
 The largest problem size, class D, ranks two billion uniformly distributed random integers using either a bucket or a counting sort algorithm, depending on the strengths of the system.
@@
 \vspace{0.5ex}
-\noindent{\bf PageRank:} This is a common centrality metric for
+\noindent{\bf PageRank} This is a common centrality metric for
 graphs. PageRank is an iterative algorithm with a common

### 2013-07-24 doc/asplos14/evaluation.tex
teal

@@
 the average length of the ready queue and the average context switch
-rate.  The last line shows the number of time a core stole work from
-another core. Only UTS depends on work-stealing for performance; the
+rate.  The last two lines shows the rate of work-stealing from
+other cores. Only UTS depends on work-stealing for performance; the
 other applications exploit locality by binding tasks to specific cores
-Nevertheless, even in UTS, which depends on work-stealing for
-performance, steals are an infrequent occurrence and account for a
+Nevertheless, even in UTS, steals are an infrequent occurrence and account for a
 small fraction of the execution time.
@@
 an aggregation buffer at 210M/s.  Together, these rates limit us to
-$~$10M/s per core in the best case. In a blocking application, the
-context switch costs (50ns, or 20M/s) bring this maximum down to
-7M/s. In practice, \Grappa does not achieve this maximum; the current
-aggregator design communicates inefficiently between cores within a
-node, limiting our message rate to to $~$1M/s per core with messages
-of common sizes. Future work will reduce this cost.
+$~$10M/s per core in the best case. In practice, \Grappa does not
+achieve this maximum; the overhead of cross-core communication in the
+current aggregator design limits our message rate to to $~$1M/s per
+core with messages of common sizes. Future work will reduce this cost.

### 2013-07-29 doc/asplos14/evaluation.tex

@@
 including \Grappa, to achieve high throughput -- at the 16-node data point,
-the average active tasks per core over the search is 775 for T1 and
-only 13 for T3. Given the lack of parallelism, scaling up only serves
+the average active tasks per core over the search is an order of magnitude larger for T1 than for T3.
+ Given the lack of parallelism, scaling up only serves
 to reduce throughput by distributing the tree to more machines.

### 2013-07-29 doc/asplos14/communication.tex

--- 2013-07-23 doc/asplos14/communication.tex
@@
 Each processing core in a given system node is responsible for
-aggregating and sending the resulting messages from all processors on
+aggregating and sending the resulting messages from all cores on
 that node to a set of destination nodes.  Each core periodically
@@
 (compare-and-swap) operations to avoid high synchronization costs.
-Note that we use a per-processor array of message lists
-that is only periodically modified across processor cores because we
-empirically determined this approach was faster (sometimes
+Note that we use a per-core array of message lists
+that is only periodically modified across processor cores after
+experimentally determining that this approach was faster (sometimes
 significantly) than a global per-system node array of message lists.
@@
 messages are built. These buffers are visible to the network card, and
-messages can be sent with user-mode operations only. When the worker
+messages are sent with user-mode operations only. When the worker
 responsible for outbound messages to a given system node has received a
@@
 management task is spawned to manage the unpacking process.  The
-management task spawns a task on each processor at the receiving
-system to simultaneously unpack messages destined for that processor.
+management task spawns a task on each core at the receiving
+system to simultaneously unpack messages destined for that core.
 Upon completion, these unpacking tasks synchronize with the management

### 2013-07-29 doc/asplos14/intro.tex
abstract intro numbers

@@
 Prefetching is of limited value because request locations are not known early enough.
-Data-parallel frameworks such as MapReduce~\cite{Dean:2008:MSD:1327452.1327492} are ineffective because they rely on effective data partitioning and regular communication patterns.
+Data-parallel frameworks such as MapReduce~\cite{Dean:2008:MSD:1327452.1327492} are ineffective because they rely on data partitioning and regular communication patterns.
 Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.
@@
 
-We believe we can support irregular applications with good performance and cost effectively with commodity hardware for two main reasons.
+We believe we can support irregular applications with good performance and cost-effectiveness with commodity hardware for two main reasons.
 First, commodity multicore processors have become extremely fast with high clock rates, large caches and robust DRAM bandwidth.
@@
 When we compare \Grappa to the XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.
-When compared with MPI and UPC, \Grappa is between 11X faster and 5.4X slower depending on the benchmark.
+When compared with MPI, \Grappa is 9X faster on a random access microbenchmark and 3.5X to 5.4X slower on applications.
 We find that the benchmarks that are faster contain specialized implementations of features \Grappa provides generally.

### 2013-07-29 doc/asplos14/tasks.tex

@@
 
-\Grappa provides three methods for expressing parallelism, shown in
-Figure~\ref{fig:expressing-parallelism}. First, when the programmer identifies
-work that can be done in parallel, the work may be wrapped up in a function
-and queued with its arguments for later execution using a \texttt{spawn}.
+\Grappa provides three methods for expressing parallelism.
+First, a single task can be created to execute in parallel with the current task by calling \texttt{spawn} with a functor.
+This adds it to the queue of tasks which will be executed the next time a worker is available.
 Second,
 the programmer can invoke a parallel for loop with \texttt{parallel\_for}, provided that the trip count is
-known at loop entry. The programmer specifies a function pointer along with
-start and end indices and an optional threshold to control parallel overhead.
+known at loop entry. The programmer specifies a functor which takes the loop index as a parameter, and an optional threshold to control parallel overhead.
 \Grappa does {\em recursive decomposition} of iterations, similar to Cilk's
@@
 tasks, stopping to execute the loop body when the number of iterations is
-below the required threshold. Third, a programmer may want to execute an active message~\cite{vonEicken92}; that is, to run a
-small non-blocking piece of code on a particular core in the system without waiting for
- a free worker to become available. For example, custom synchronization primitives are implemented as an active message executed on the core where the data
-lives. \Grappa provides the \texttt{call\_on} routine for this purpose.
+below the required threshold.
+Third, parallelism can be expressed via asynchronous delegate operations, which are explained next, in Section~\ref{sec:memory}.
+% Third, a programmer may want to execute an active message~\cite{vonEicken92}; that is, to run a
+% small non-blocking piece of code on a particular core in the system without waiting for
+ % a free worker to become available. For example, custom synchronization primitives are implemented as an active message executed on the core where the data
+% lives. \Grappa provides the \texttt{call\_on} routine for this purpose.
 
-\begin{figure}[htbp]
-  \begin{center}
-	\begin{description}
-    \item[\texttt{spawn( Functor f )}] \hfill \\
-      Creates a new stealable task
-    \item[\texttt{parallel\_for( start, end, Functor iteration )}] \hfill \\
-      Executes iterations of a loop as stealable tasks that
-      take the iteration index as an argument
-    \item[\texttt{call\_on( core, Functor f )}] \hfill \\
-      Runs a limited function on a specific core without
-      consuming a stack or context switching
-	\end{description}
-    \begin{minipage}{0.95\columnwidth}
-      \caption{\label{fig:expressing-parallelism} \Grappa API: expressing parallelism
-      } % \vspace{-4ex}}
-    \end{minipage}
-    %\vspace{-3ex}
-  \end{center}
-\end{figure}
+% \begin{figure}[htbp]
+%   \begin{center}
+%   \begin{description}
+%     \item[\texttt{spawn( Functor f )}] \hfill \\
+%       Creates a new stealable task
+%     \item[\texttt{parallel\_for( start, end, Functor iteration )}] \hfill \\
+%       Executes iterations of a loop as stealable tasks that
+%       take the iteration index as an argument
+%     \item[\texttt{call\_on( core, Functor f )}] \hfill \\
+%       Runs a limited function on a specific core without
+%       consuming a stack or context switching
+%   \end{description}
+%     \begin{minipage}{0.95\columnwidth}
+%       \caption{\label{fig:expressing-parallelism} \Grappa API: expressing parallelism
+%       } % \vspace{-4ex}}
+%     \end{minipage}
+%     %\vspace{-3ex}
+%   \end{center}
+% \end{figure}
 

### 2013-07-29 doc/sosp13/memory.tex

@@
 that synchronization has finished before continuing. Synchronization in \Grappa
-is implemented using synchronous delegate operations. Often remote data
-accesses do not need to be performed synchronously. In order to avoid
+is implemented using synchronous delegate operations. On the other hand, remote data
+accesses often do not need to be performed synchronously. In order to avoid
 unnecessary waiting, we support asynchronous delegate operations. For reads,
@@
 invocations. Delegate write operations may also be performed asynchronously,
-and since they do not return data, they need a mechanism to detect completion.
+and since they do not return data, a mechanism to detect completion is needed.
 \Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
-outstanding asynchronous operations of the local core. The code and can probe
-the state of this object to determine when all outstanding requests are
-complete. This is necessary to, say, make sure all outstanding operations are
-complete before executing a synchronization operation.
+outstanding asynchronous operations of the local core. Threads that depend on completion may block on this object
+the state of this object, and they will be woken when all outstanding requests are
+complete.
 

### 2013-08-28 doc/pgas13/flat-combining.tex
improve explanation of linearizability

@@
 
-To preserve linearizability for flat-combined operations, they must obey a consistent global order.
-First of all, the execution of local combined operations must be serializable;
+To behave in accordance with sequential consistency, operations on a particular data structure must obey a consistent global order.
+One way to provide this is to guarantee \emph{linearizability}~\cite{herlihy1990linearizability} of operations, which requires that the operation appear to take effect globally at some instantaneous point during invocation of the API call.
+This ensures that a consistent total order can be imposed on operations on a single data structure.
+% To preserve linearizability for flat-combined operations, they must obey a consistent global order.
+For operations to be globally linearizable, first of all the execution of local combined operations must be serializable;
 this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
-When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the \emph{master} core for the Stack).
+Second, combined operations must be committed atomically in some globally serializable order.
+When committing, combined operations are serialized at the particular core that owns the synchronization (the \emph{master} core for the Stack).
 Whenever a global commit starts, a fresh combiner must be created for subsequent operations to use.
@@
 
-Combining set and map operations expose more nuances in the consistency guarantees.
+Combining set and map operations exposes more nuances in the requirements for consistency.
 Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally.
@@
 % This ensures that if one wanted to use the insert to signal synchronization, it is guaranteed to have been committed globally before the task continues.
+Similarly, lookups can piggy-back on other lookups of the same key.
 
-Similarly, lookups can piggy-back on other lookups of the same key.
 Using intuition from store/write buffers in modern processors, it is tempting to allow a lookup to be satisfied locally by inspecting outstanding inserts.
-However, this allows the local order in which keys were inserted to be observed.
-If this is allowed, then to preserve SC, that same order must be respected globally, forcing each batch to commit atomically with respect to other cores' batches.
-Enforcing this is prohibitively expensive, so a cheaper solution is chosen: lookups must get their values from the global data structure, so batches can be performed in parallel.
+However, this would allow the local order in which keys were inserted to be observed.
+Then to preserve SC, that same order would need to be respected globally, forcing each batch to commit atomically with respect to other cores' batches.
+Enforcing this would be prohibitively expensive, so a cheaper solution is chosen: lookups only get their values from the global data structure, therefore batches can be performed in parallel.
 
@@
 % As mentioned before, the local \emph{publication list} is trivial to access atomically in Grappa's cooperative scheduling.
-% Instead, in the Grappa FC Framework, 
+% Instead, in the Grappa FC Framework, definition
 For each global structure instance, a \emph{proxy} object is created on each core and each worker merges its request into that structure before blocking or becoming the combiner.

### 2013-08-28 doc/pgas13/intro.tex
improve explanation of linearizability

@@
 % Much prior work has explored the scalability challenges of adding more concurrent accessors to shared data structures in shared memory systems.
-However, as the number of concurrent accessors increases, the more they contend and the more synchronization operations fail, even for fine-grained lock-free synchronization schemes.
+However, even in fine-grained lock-free synchronization schemes, as the number of concurrent accessors increases, there is more contention, resulting in more failed synchronization operations.
 With the massive amount of parallelism in a cluster of multiprocessors and with the increased cost of remote synchronization, the problem is magnified.

### 2013-08-28 doc/pgas13/evaluation.tex
resize and subfigure-ize figures

@@
   \centering
-  \includegraphics[width=0.5\textwidth]{data/plots/vector_perf.pdf}
+  \includegraphics[width=0.8\textwidth]{data/plots/vector_perf.pdf}
   \caption{\emph{Queue and Stack performance.}
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{data/plots/stack_stats.pdf}
+  \includegraphics[width=0.8\textwidth]{data/plots/stack_stats.pdf}
   \caption{\emph{GlobalStack Statistics.}
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{data/plots/hash_perf.pdf}
+  \includegraphics[width=0.8\textwidth]{data/plots/hash_perf.pdf}
   \caption{\emph{GlobalHashSet and GlobalHashMap.}
@@
 
-\paragraph{Breadth-First Search.}
 \begin{figure}[t]
   \centering
-  \includegraphics[width=0.38\textwidth]{data/plots/bfs_perf.pdf}
-  \caption{\emph{BFS} on a Graph500-spec graph of scale 26 (64 million vertices, 1 billion edges), with the direction-optimizing BFS algorithm.
-Performance is measured in millions of Traversed Edges Per Second (MTEPS).}
+  \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{data/plots/bfs_perf.pdf}
+    \caption{\emph{BFS} with the direction-optimizing algorithm.}
     \label{fig:bfs_perf}
+  \end{subfigure}
+  ~
+  \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{data/plots/cc_perf.pdf}
+    \caption{\emph{Connected Components} using Kahan's algorithm.}
+    \label{fig:cc_perf}
+  \end{subfigure}
+  \caption{Performance of graph kernels on a Graph500-spec graph of scale 26 (64 million vertices, 1 billion edges).
+Performance for both is measured in Millions of Traversed Edges Per Second (MTEPS).}
 \end{figure}
+
+\paragraph{Breadth-First Search.}
 The first application kernel is the Graph500 Breadth-First-Search (BFS) benchmark~\cite{graph500list}.
@@
 \paragraph{Connected Components.}
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.38\textwidth]{data/plots/cc_perf.pdf}
-  \caption{\emph{Connected Components} on the same scale 26 Graph500 graph. Performance is measured in MTEPS.}
-  \label{fig:cc_perf}
-\end{figure}
 Connected Components (CC) is another core graph analysis kernel that illustrates a different use of global data structures in irregular applications.

### 2013-08-28 doc/pgas13/flat-combining.tex
resize and subfigure-ize figures
add note about doing both centralized and dist combining being insignificant

@@
   \centering
-  \includegraphics[width=0.42\textwidth]{figs/fc_shared_mem.pdf}
+  \includegraphics[width=0.6\textwidth]{figs/fc_shared_mem.pdf}
   \caption{\emph{Flat combining in shared memory.}
@@
     \centering
-    \includegraphics[width=0.40\textwidth]{figs/stack_nofc.pdf}
+    \includegraphics[width=0.6\textwidth]{figs/stack_nofc.pdf}
     \caption{\emph{Stack without combining.}
@@
     \centering
-    \includegraphics[width=0.42\textwidth]{figs/stack_fc.pdf}
+    \includegraphics[width=0.6\textwidth]{figs/stack_fc.pdf}
     \caption{\emph{Stack with distributed combining.}
@@
 The global lock must still be acquired, so concurrent combined requests (from different cores) must block and serialize on the master, but the granularity of global synchronization is coarser, reducing actual serialization.
+
+Centralized combining could be applied on top of distributed combining, combining the combined operations at the master node.
+However, in our experiments, performing the additional combining on the master did not significantly affect performance, so it is left out of our evaluation for space.
 

### 2013-09-10 doc/pgas13/conclusion.tex

@@
 \section{Conclusion}
-In PGAS implementations there is typically a large discrepancy between the first, simple, description of an algorithm and the final, optimized one.
 % At the core of this work is the desire to have a library of data structures which can be used without worrying about when they will be consistent or whether they will limit scalability.
-Bringing findings from shared memory into the PGAS domain brings with it new challenges and opportunities.
+% Bringing findings from shared memory into the PGAS domain brings with it new challenges and opportunities.
+Coming from the multi-core domain, the flat-combining paradigm provides a new perspective to global synchronization, bringing with it both challenges and new opportunities.
 We have shown that the additional concurrency that comes with a latency-tolerant runtime, rather than compounding the problem, provides a new opportunity for reducing communication by combining locally.
 % perform some of the locality optimizations that finely tuned implementations would.
-This allows easy implementation of a library of flat-combined linearizable global data structures that scale well out to a thousand cores and millions of concurrent threads.
+In PGAS implementations there is typically a large discrepancy between the first simple description of an algorithm and the final optimized one.
+Our distributed flat-combining framework allows easy implementation of a library of flat-combined linearizable global data structures, allowing even simple applications that use them to scale out to a thousand cores and millions of concurrent threads.

### 2013-09-10 doc/pgas13/intro.tex

@@
 % In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently.
-It is commonly accepted that the easiest consistency model to reason about is \emph{sequential consistency} (SC), which enforces that all accesses are visible in program order and appear to happen in some global serializable order.
+It is commonly accepted that the easiest consistency model to reason about is \emph{sequential consistency} (SC), which enforces that all accesses are committed in program order and appear to happen in some global serializable order.
 To preserve SC, operations on shared data structures should be \emph{linearizable}~\cite{herlihy1990linearizability}; that is, appear to happen atomically in some global total order.
